### 2022-08-04 background.tex
paper structure

\section{Background \& Related Work}\label{sec:background}

### 2022-08-04 model.tex
paper structure

\section{A model of human behavior for \ac{WCA}}\label{sec:model}

In order to accurately emulate the behavior of a human, a model for \ac{WCA} needs to implement two main behaviors.
One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
\todo[inline]{Improve}

In this section, we present and discuss our approaches to both these challenges.

\subsection{Generating realistic timings}

\input{tables/empirical_exec_times.tex}

In our previous work~\cite{olguinmunoz:impact2021}, we identified four main results concerning the step execution times of humans in \ac{WCA}:

\begin{enumerate}
    \item System slow-down induces additional behavioral slow-down which scales with the decrease in system responsiveness.
    Moreover, users become progressively slower the longer they spend in a degraded system state.

    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.

    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.

    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
\end{enumerate}

Using these insights, we design a generator of realistic execution times for step-based \ac{WCA}.
The generator further builds upon the original data collected for~\cite{olguinmunoz:impact2021}; this data is first preprocessed and then sampled according to a series of rules in order to produce novel,realistic sequences of execution times for a task.
Alternatively, instead of sampling directly from the preprocessed data, a theoretical distribution can be fitted, thus allowing for generation of realistic synthetic execution times.

\subsubsection{Data pre-processing}

A sample of the step timing data obtained for~\cite{olguinmunoz:impact2021} is presented in \cref{tab:exec_times:empirical}.
This data corresponds to real step execution times for \num{40} repetitions, by different human subjects, of the same \num{169}-step task.
During each repetition of the task, system responsiveness was altered in real time by setting the round-trip-times of all frames to a fixed value (corresponding to the \emph{delay} column in \cref{tab:exec_times:empirical}).
For more details on the experimental design and data collection procedure, please see the original article~\cite{olguinmunoz:impact2021}.

We process this data in the following manner:

\input{tables/processed_exec_times.tex}

\begin{enumerate}
    \item Normalized neuroticism values are binned into predefined, contiguous bins.
    We use three bins with edges
    \( \left(-\infty, \frac{1}{3}\right) \),
    \( \left[\frac{1}{3}, \frac{2}{3}\right) \),
    and \( \left[\frac{2}{3}, \infty\right) \),
    as these give the most even distribution of number of samples across bins.

    \item System impairment (\emph{delay}) is binned into predefined, contiguous bins.
    Once again, we use three bins giving a somewhat even distribution of samples:
    \( \left(-\infty, 1.0\right) \),
    \( \left[1.0, 2.0\right) \),
    and \( \left[2.0, \infty\right) \) seconds.

    \item A predefined number of steps (by default, \num{8}) immediately after each change in binned impairment are marked according to the relative change in impairment: \emph{low-to-high}, or \emph{high-to-low} impairment.
    We label this value the \emph{transition tag}.
    Steps at the beginning of the task (where there is no previous impairment), or too far away from a transition in impairment, are tagged as \emph{no transition}.

    \item Each step in a sequence tagged with the same transition tag is assigned a sequential \emph{duration} number.
    This value tracks how long the system has been in a specific state, and is in turn binned into three bins:
    \( \left[0, 5\right) \),
    \( \left[5, 10\right) \),
    and \( \left[10, \infty\right) \) steps.

    \item Finally, execution times are shifted by one step.
    This is because the execution time of a step depends on the state of the previous step, and not the current one.
    As explained previously, humans only becomes aware of the responsiveness of the system when they finish a step and have to wait for feedback from the system.
\end{enumerate}

\subsection{Generating a synthetic trace of frames}

\subsection{Model implementation}

The model is provided as a Python~\num{3.10} library which includes \acp{API} for the generation of realistic execution times and synthetic traces of video frames.

\subsection{Verifying the model}

### 2022-08-04 tables/processed_exec_times.tex
paper structure

\begin{table*}
    \centering
    \caption{}\label{tab:exectime:proc}
\begin{tabular}{rrrllll}
\toprule
subject ID &  step &  next exec. time & neuroticism &  impairment & transition tag & duration \\
\midrule
\num{136516} &   \num{67} & \SI{4.663}{\second} & medium & high & no transition & short \\
%
\num{137299} &  \num{114} & \SI{2.230}{\second} & low & low & higher-to-lower &  short \\
\num{137344} &   \num{45} & \SI{1.775}{\second} & low &  medium & no transition & medium \\
\num{135955} &   \num{77} & \SI{4.326}{\second} & low &  high & no transition & long \\
\num{135385} &  \num{129} & \SI{2.499}{\second} & high &  medium & no transition & long \\
\bottomrule
\end{tabular}
\todo[inline]{Use same samples as first table for clarity!}
\end{table*}

### 2022-08-05 outline.tex
working on outline

\section{Outline of paper}

Key research question:

\emph{How do different sampling schemes affect system load and performance of WCA applications?}

Secondary question answered:

\emph{How can an artificial model of human behavior simplify reasearch in WCA?}

\begin{enumerate}
    \item Introduction
    \item Background and related work
        \begin{enumerate}
            \item What is WCA?
            \item Why is it interesting?
            \item Example applications and systems.
            \item Studying the impact of WCA on edge computing systems (why do we need a model for WCA?).
            \item Sampling strategy as an important factor driving system load.
            \item Impact of sampling schemes on system load.
            \item Include examples from other systems (control theory).
            \item Previous work on sampling in WCA.
            \item Research on sampling is hampered by the lack of a model for WCA.
            \item Previous approaches to understanding WCA behavior and modeling it.
        \end{enumerate}
    \item Model for WCA
        \begin{enumerate}
            \item Reiterate reasons for model
            \item Discuss base data from PLOS paper
                \begin{enumerate}
                    \item Timing
                    \item Frames
                \end{enumerate}
            \item Timing model
            \item Frame statistics
            \item Discuss effects of neuroticism (Bobby)
            \item Implementation in Python
        \end{enumerate}
    \item Validation of WCA Model
    \begin{enumerate}
        \item Reference model implementation (naive model)
        \item Plain comparison between naive, empirical, theoretical model using zero-wait sampling.
        \item Scaling behavior? 
    \end{enumerate}

    \item Sampling strategies
    \begin{enumerate}
        \item Reference strategies: zero-wait, ideal.
        \item Basic strategies: periodic, hold.
        \item Advanced strategy: adaptive aperiodic. Need to reference Vishnu's paper, indicate we're using a modified approximate solution.
    \end{enumerate}

    \item Experimental setup
    \begin{enumerate}
        \item Brief introduction to testbed
        \item Architectural design of experiment
    \end{enumerate}

    \item Results
    \item Discussion
    \item Conclusion
\end{enumerate}
\newpage

### 2022-08-09 outline.tex
outline

@@
 
-Key research question:
+Main research question:
 
-\emph{How do different sampling schemes affect system load and performance of WCA applications?}
-
-Secondary question answered:
-
-\emph{How can an artificial model of human behavior simplify reasearch in WCA?}
+\emph{Implications of the interdependence between human behavior and system responsiveness in WCA applications}
 
@@
             \item Studying the impact of WCA on edge computing systems (why do we need a model for WCA?).
-            \item Sampling strategy as an important factor driving system load.
-            \item Impact of sampling schemes on system load.
-            \item Include examples from other systems (control theory).
-            \item Previous work on sampling in WCA.
-            \item Research on sampling is hampered by the lack of a model for WCA.
-            \item Previous approaches to understanding WCA behavior and modeling it.
+            \item Trade-offs: responsiveness (i.e. sampling speed) vs network load.
+            \item Discuss how responsiveness relates to execution times and subsequently to system utilization.
+            \item How to study these effects? Experimentally, with a model.
+            \item Introduce model of WCA, refer back to PLOS paper. 
+            % \item Impact of sampling schemes on system load.
+            % \item Include examples from other systems (control theory).
+            % \item Previous work on sampling in WCA.
+            % \item Research on sampling is hampered by the lack of a model for WCA.
+            % \item Previous approaches to understanding WCA behavior and modeling it.
         \end{enumerate}
@@
         \begin{enumerate}
-            \item Reiterate reasons for model
+            \item Reiterate reasons for model: correlations between step execution time and system responsiveness
             \item Discuss base data from PLOS paper
@@
             \item Timing model
-            \item Frame statistics
-            \item Discuss effects of neuroticism (Bobby)
+            \item Frame generation model
+            \item Discuss effects of neuroticism (? for this we would need Bobby I think)
             \item Implementation in Python
@@
         \item Reference model implementation (naive model)
-        \item Plain comparison between naive, empirical, theoretical model using zero-wait sampling.
-        \item Scaling behavior? 
+        \item Comparison between naive, empirical, theoretical model using zero-wait sampling.
+        \begin{enumerate}
+            \item Scaling behavior (1, 5, and 10 clients?)
+        \end{enumerate}
+        \item Highlight behaviors captured by realistic models that are not present in naive model.
+        \item Primarily, the spread in durations. Realistic models get faster in optimal conditions, but also react more strongly to lowered responsiveness.
     \end{enumerate}
 
-    \item Sampling strategies
+    \item Studying implications of responsiveness-network load trade-offs using the model
     \begin{enumerate}
-        \item Reference strategies: zero-wait, ideal.
-        \item Basic strategies: periodic, hold.
-        \item Advanced strategy: adaptive aperiodic. Need to reference Vishnu's paper, indicate we're using a modified approximate solution.
+        \item Sampling strategies as a way to alter this trade-off.
+        \item Present basic strategies: zero-wait, ideal, hold, regular.
+        \begin{enumerate}
+            \item Only realistic models, not naive.
+            \item Present comparisons between basic strategies, without any additional load: total task duration, time-to-feedback breakdowns, total number of samples.
         \end{enumerate}
-
-    \item Experimental setup
+        \item Can we do better? Introduce and discuss advanced strategy: adaptive aperiodic approach.
         \begin{enumerate}
-        \item Brief introduction to testbed
-        \item Architectural design of experiment
+            \item Maybe fold in Vishnu's globecom paper here, with modifications to target responsiveness instead of power consumption.
+            \item Discuss parameters, and how values were settled on.
+            \item Compare adaptive aperiodic with 2 best basic strategies without additional load.
+            \item Introduce additional load to showcase the adaptability of the adaptive aperiodic approach.
         \end{enumerate}
-
-    \item Results
-    \item Discussion
+    \end{enumerate}
     \item Conclusion
+    \begin{enumerate}
+        \item Future work and open questions
+        \item Effects of individual differences (neuroticism, age, etc)
+        \item Extend to other WCAs
+        \item Extend to other applications
+    \end{enumerate}
 \end{enumerate}

### 2022-08-10 examplethesis.tex
base

%%
%% forked from https://gits-15.sys.kth.se/giampi/kthlatex kthlatex-0.2rc4 on 2020-02-13
%% expanded upon by Gerald Q. Maguire Jr.
%% This template has been adapted by Anders Sjögren to the University
%% Engineering Program in Computer Science at KTH ICT. Adaptation is the
%% translation of English headings into Swedish as the addition of Swedish
%% text. Original body text is deliberately left in English.

%% Conventions for todo notes:
% \todo[inline]{Comments/directions/... in English}
% \todo[inline, backgroundcolor=kth-lightblue]{Text på svenska}
% \todo[inline, backgroundcolor=kth-lightgreen]{English descriptions about formatting}

%% The template is designed to handle a thesis in English or Swedish
% set the default language to english or swedish by passing an option to the documentclass - this handles the inside tile page
% To optimize for digital output (this changes the color palette add the option: digitaloutput
% To use bibtex or biblatex - include one of these as an option
% For a licentiate thesis or a doctoral dissertation - select g5paper for the booklet format

\documentclass[english, g5paper, bibtex]{kththesis}
%\documentclass[swedish, g5paper, biblatex]{kththesis}

\ifbiblatex
    \usepackage[bibstyle=authoryear,citestyle=authoryear, maxbibnames=99,language=english]{biblatex}
    \addbibresource{references.bib}
    %\DeclareLanguageMapping{norsk}{norwegian}
\else
    % The line(s) below are for BibTeX
    \bibliographystyle{bibstyle/myIEEEtran}
    %\bibliographystyle{apalike}
\fi

\include{lib/includes}
\include{lib/kthcolors}

\include{lib/defines}  % load some additional definitions to make writing more consistent

% The following is needed in conjunction with generating the DiVA data with abstracts and keywords using the scontents package and a modified listings environment
%\usepackage{listings}   %  already included
\ExplSyntaxOn
\newcommand\typestoredx[2]{\expandafter\__scontents_typestored_internal:nn\expandafter{#1} {#2}}
\ExplSyntaxOff
\makeatletter
\let\verbatimsc\@undefined
\let\endverbatimsc\@undefined
\lst@AddToHook{Init}{\hyphenpenalty=50\relax}
\makeatother

\lstnewenvironment{verbatimsc}
    {
    \lstset{%
        basicstyle=\ttfamily\tiny,
        %basicstyle=\tiny,
        %columns=fullflexible,
        columns=[l]fixed,
        language=[LaTeX]TeX,
        %numbers=left,
        %numberstyle=\tiny\color{gray},
        keywordstyle=\color{red},
        breaklines=true,                 % sets automatic line breaking
        breakatwhitespace=true,          % sets if automatic breaks should only happen at whitespace
        %keepspaces=false,
        breakindent=0em,
        %fancyvrb=true
    }
}{}

%% Reduce hyphenation as much as possible
\hyphenpenalty=15000 
\tolerance=1000

%% definition of new command for bytefield package
\newcommand{\colorbitbox}[3]{%
	\rlap{\bitbox{#2}{\color{#1}\rule{\width}{\height}}}%
	\bitbox{#2}{#3}}

% to find unicode characters that LaTeX complains about
%\DeclareUnicodeCharacter{F0DE}{XXXX Here I am?????XXXX}

\include{lib/includes-after-hyperref}

%% Acronyms
%\usepackage[acronym]{polyglossia}
%\usepackage[acronym]{glossaries}
\usepackage[acronym, section=section, nonumberlist, nomain, nopostdot]{glossaries}

%% Acronyms
% note that nonumberlist - removes the cross references to the pages where the acronym appears
% note that nomain - does not produce a main gloassay, this only acronyms will be in the glossary
% note that nopostdot - will present there being a period at the end of each entry
\ifinswedish
    %\usepackage{glossaries-swedish}
\fi
%\glsdisablehyper
\makeglossaries
\include{lib/acronyms}                %load the acronyms file

\begin{document}
%% if using XeLaTex uncomment the following
%\selectlanguage[variant=american]{english}
%% otherwise
%\selectlanguage{UKenglish}
%\selectlanguage{english}

%% Information for inside title page
\ifinswedish
  \selectlanguage{swedish}
  \title{Titeln på avhandlingens språk}
  \subtitle{En undertiteln på avhandlingens språk}
%You: Note that the commands you are using are for biblatex and not bibtex. So you have to change the include files to include biblatex also you have to change the way that the reference bib file is included.

  \alttitle{English translation of the title}
  \altsubtitle{English translation of the subtitle}
\else
   \selectlanguage{english}
   \title{The title in the language of the thesis}
   \subtitle{A subtitle in the language of the thesis}
    % give the alternative title - i.e., if the thesis is in English, then give a Swedish title
   \alttitle{Detta är den svenska översättningen av titeln}
   \altsubtitle{Detta är den svenska översättningen av undertiteln}
% alternative, if the thesis is in Swedish, then give an English title

\fi

%%% Uncomment the program that you are in
%\programcode{KTHARK}% Architecture
%\programcode{KTHBIO}% Biotechnology
%\programcode{KTHBYV}% Civil and Architectural Engineering
\programcode{KTHDAT}% Computer Science
%\programcode{KTHEST}% Electrical Engineerin
%\programcode{KTHEGI}% Energy Technology and Systems
%\programcode{KTHFTK}% Vehicle and Maritime Engineering
%\programcode{KTHFYS}% Physics
%\programcode{KTHGEO}% Geodesy and Geoinformatics
%\programcode{KTHHFL}% Solid Mechanics
%\programcode{KTHIEO}% Industrial Economics and Management
%\programcode{KTHIIP}% Production Engineering
%\programcode{KTHIKT}% Information and Communication Technology
%\programcode{KTHKEV}% Chemical Science and Engineering
%\programcode{KTHKON}% Art, Technology and Design
%\programcode{KTHMAT}% Mathematics
%\programcode{KTHKOM}% Mediated Communication
%\programcode{KTHPBA}% Planning and Decision Analysis
%\programcode{KTHSHB}% The Built Environment and Society: Management, Economics and Law
%\programcode{KTHTMV}% Engineering Materials Science
%\programcode{KTHMEK}% Engineering Mechanics
%\programcode{KTHTKB}% Theoretical Chemistry and Biology

%% Alternatively, you can say \programme{xx} to directly set the programme string

%\schoolAcronym{ABE}
%\schoolAcronym{CBH}
\schoolAcronym{EECS}
%\schoolAcronym{ITM}
%\schoolAcronym{SCI}
%% Alternatively, you can say \school{School of xx} to directly set the school string

\authorsLastname{Student}
\authorsFirstname{Oscar}
\email{oscars@kth.se}
\kthid{u1xxxxxx}
\orcid{0000-0002-0007-746X}
\authorsDepartment{Computer Science}

\ifinswedish
    \examen{doktorsexamen}
    %\examen{Ekonomie doktorsexamen}
\else
    \examen{Doctor of Technology}
    %\examen{Doctor of Philosophy}
\fi

\ifinswedish
    \thesistype{Doktorsavhandling}
\else
    \thesistype{Doctoral Thesis}
\fi

\ifinswedish
    \disputationsdatum{Måndag den 15 Mars 2021 klockan 13:00 CET}
\else
    \disputationsdatum{Monday the 15\textsuperscript{th} March 2021 at 13:00 CET}
\fi

\ifinswedish
    \disputationslokal{ett Zoom-Webinar}
\else
    \disputationslokal{via Zoom}
\fi

%	"Room": "Seminar room Grimeton at COM",
%	"Address": "Kistagången 16, East, Floor 4, Elevator B",
%	"City": "Kista"

\presentationDateAndTimeISO{2021-03-15 13:00}
\presentationLanguage{eng}
\presentationRoom{via Zoom}
%\presentationAddress{}
\presentationCity{Stockholm}

\isbn{ISBN 978-91-7873-xxx-y}
\trita{TRITA-ITM-AVL 2021:x}
\publisher{Universitetsservice US AB}

\publicationYear{\the\year{}}
\publicationMonth{\the\month{}}

%% Supervisors' information
% first supervisor (A)
\supervisorAsLastname{Maguire Jr.}
\supervisorAsFirstname{Gerald Q.}
\supervisorAsEmail{maguire@kth.se}
% if not with KTH then enter the name of the external organization
%\supervisorAsOrganization{University of Nowhere, School of Hard, Department of Knocks}
\supervisorAsKTHID{u1d13i2c}
\supervisorAsSchool{EECS}
\supervisorAsDepartment{Computer Science}
% second supervisor (B)
\supervisorBsLastname{Adviser}
\supervisorBsFirstname{Secondary}
\supervisorBsEmail{secondary.adviser@foo.edu}
% if not with KTH then enter the name of the external organization
\supervisorBsOrganization{University of Nowhere, School of Hard, Department of Knocks}
% otherwise, enter the supervisor's KTHID, KTH School and Department information
%\supervisorBsKTHID{u10000c}
%\supervisorBsSchool{School of xxx}
%\supervisorBsDepartment{dept}

% Opponent's information
\opponentsLastname{Normal}
\opponentsFirstname{A. B.}
\opponentsEmail{ABNormal@example.org}
% for an opponent outside KTH define:
\opponentsOrganization{Famous Anvils}
% for an opponent inside KTH define:
%\opponentsSchool{Circus school}
%\opponentsDepartment{Clown department}

%%%%% for DiVA's National Subject Category information
%%% Enter one or more 3 or 5 digit codes
%%% See https://www.scb.se/contentassets/3a12f556522d4bdc887c4838a37c7ec7/standard-for-svensk-indelning--av-forskningsamnen-2011-uppdaterad-aug-2016.pdf
%%% See https://www.scb.se/contentassets/10054f2ef27c437884e8cde0d38b9cc4/oversattningsnyckel-forskningsamnen.pdf
%%%%
%%%% Some examples of these codes are shown below:
% 102 Data- och informationsvetenskap (Datateknik)    Computer and Information Sciences
% 10201 Datavetenskap (datalogi) Computer Sciences 
% 10202 Systemvetenskap, informationssystem och informatik (samhällsvetenskaplig inriktning under 50804)
% Information Systems (Social aspects to be 50804)
% 10203 Bioinformatik (beräkningsbiologi) (tillämpningar under 10610)
% Bioinformatics (Computational Biology) (applications to be 10610)
% 10204 Människa-datorinteraktion (interaktionsdesign) (Samhällsvetenskapliga aspekter under 50803) Human Computer Interaction (Social aspects to be 50803)
% 10205 Programvaruteknik Software Engineering
% 10206 Datorteknik Computer Engineering
% 10207 Datorseende och robotik (autonoma system) Computer Vision and Robotics (Autonomous Systems)
% 10208 Språkteknologi (språkvetenskaplig databehandling) Language Technology (Computational Linguistics)
% 10209 Medieteknik Media and Communication Technology
% 10299 Annan data- och informationsvetenskap Other Computer and Information Science
%%%
% 202 Elektroteknik och elektronik Electrical Engineering, Electronic Engineering, Information Engineering
% 20201 Robotteknik och automation Robotics
% 20202 Reglerteknik Control Engineering
% 20203 Kommunikationssystem Communication Systems
% 20204 Telekommunikation Telecommunications
% 20205 Signalbehandling Signal Processing
% 20206 Datorsystem Computer Systems
% 20207 Inbäddad systemteknik Embedded Systems
% 20299 Annan elektroteknik och elektronik Other Electrical Engineering, Electronic Engineering, Information Engineering
%% Example for a thesis in Computer Science and Computer Systems
\nationalsubjectcategories{10201, 10206}

% Enter the English and Swedish keywords here for use in the PDF meta data _and_ for later use
% following the respective abstract.
% Try to put the words in the same order in both to facilitate matching.
\EnglishKeywords{Canvas Learning Management System, Docker containers, Performance tuning}
\SwedishKeywords{Canvas Lärplattform, Dockerbehållare, Prestandajustering}

% Put the title, author, and keyword information into the PDF meta information
\include{lib/pdf_related_includes}

\titlepage
% document/book information page
\bookinfopage

% Frontmatter includes the abstracts and table-of-contents
\frontmatter
\setcounter{page}{1}

\begin{abstract}
% The first abstract should be in the language of the thesis.
% Abstract fungerar på svenska också.
  \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
eng
\end{scontents}
%%% The contents of the abstract will be saved in LaTeX format and output on the page(s) at the end of the thesis with information for DiVA facilitating the correct entry of the meta data for your thesis. These page(s) will be removed before the thesis is to be inserted into DiVA.
\begin{scontents}[store-env=abstracts,print-env=true]
\todo[inline, backgroundcolor=kth-lightgreen]{All theses at KTH are \textbf{required} to have an abstract in both \textit{English} and \textit{Swedish}.}

\todo[inline, backgroundcolor=kth-lightgreen]{Keep in mind that most of your potential readers are only going to read your \texttt{title} and \texttt{abstract}. This is why it is important that the abstract give them enough information that they can decide is this document relevant to them or not. Otherwise the likely default choice is to ignore the rest of your document.\\

A abstract should stand on its own, i.e., no citations, cross references to the body of the document, acronyms must be spelled out, \ldots .\\

Write this early and revise as necessary. This will help keep you focused on what you are trying to do.}

Write an abstract that is about 250 and 350 words (1/2 A4-page)  with the following components:
\begin{itemize}
  \item What is the topic area? (optional) Introduces the subject area for the project.
  \item Short problem statement
  \item Why was this problem worth a third cycle thesis/dissertation? (\ie, why is the problem both significant and of a suitable degree of difficulty for the degree? Why has no one else solved it yet?)
  \item How did you solve the problem? What was your method/insight?
  \item Results/Conclusions/Consequences/Impact: What are your key results and conclusions? What will others do based upon your results? What can be done now that you have finished - that could not be done before your thesis project was completed?
  \textbf{The presentation of the results should be the main part of the abstract.}
\end{itemize}
\end{scontents}

\subsection*{Keywords}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 keywords\\
% If you set the EnglishKeywords earlier, you can retrieve them with:
% Alternative 1:
%\makeatletter
%\@EnglishKeywords
%\makeatother
%
% Alternative 2:
\InsertKeywords{english}
% If you did not set the EnglishKeywords earlier then simply enter the keywords here:
% Alternative 3:
% comma separate keywords, such as: Canvas Learning Management System, Docker containers, performance tuning

\todo[inline, backgroundcolor=kth-lightgreen]{\textbf{Choosing good keywords can help others to locate your paper, thesis, dissertation, \ldots and related work.}}
Choose the most specific keyword from those used in your domain, see for example: the ACM Computing Classification System ({\small \url{https://www.acm.org/publications/computing-classification-system/how-to-use})},
the IEEE Taxonomy ({\small \url{https://www.ieee.org/publications/services/thesaurus-thank-you.html}}), PhySH (Physics Subject Headings)\linebreak[4] ({\small \url{https://physh.aps.org/}}), \ldots or keyword selection tools such as the  National Library of Medicine's Medical Subject Headings (MeSH)  ({\small \url{https://www.nlm.nih.gov/mesh/authors.html}}) or Google's Keyword Tool ({\small \url{https://keywordtool.io/}})\\

\textbf{Mechanics}:
\begin{itemize}
  \item The first letter of a keyword should be set with a capital letter and proper names should be capitalized as usual.
  \item Spell out acronyms and abbreviations.
  \item Avoid "stop words" - as they generally carry little or no information.
  \item List your keywords separated by commas (",").
\end{itemize}    
Since you should have both English and Swedish keywords - you might think of ordering them in corresponding order (\ie, so that the n\textsuperscript{th} word in each list correspond) - this makes it easier to mechanically find matching keywords.

\end{scontents}
\end{abstract}

\cleardoublepage
\babelpolyLangStart{swedish}
  \begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
swe
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
\todo[inline, backgroundcolor=kth-lightblue]{Alla avhandlingar vid KTH \textbf{måste ha} ett abstrakt på både \textit{engelska} och \textit{svenska}.\\

Om du skriver din avhandling på svenska ska detta göras först (och placera det som det första abstraktet) - och du bör revidera det vid behov.}

\todo[inline]{If you are writing your thesis in English, you can leave this until the final version. If you are writing your thesis in Swedish then this should be done first – and you should revise as necessary along the way.\\
If you are writing your thesis in English, then this section can be a summary targeted at a more general reader. However, if you are writing your thesis in Swedish, then the reverse is true – your abstract should be for your target audience, while an English summary can be written targeted at a more general audience.\\
This means that the English abstract and Swedish sammnfattning  
or Swedish abstract and English summary need not be literal translations of each other.}

\todo[inline, backgroundcolor=kth-lightgreen]{The abstract in the language used for the thesis should be the first abstract, while the Summary/Sammanfattning in the other language can follow.}

\todo[inline, backgroundcolor=kth-lightgreen]{Exchange students many want to include one or more abstracts in the language(s) used in their home institutions to avoid the neeed to write another thesis when returning to their home institution.}

\end{scontents}
\subsection*{Nyckelord}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 nyckelord\\
% SwedishKeywords were set earlier, hence we can use alternative 2
\InsertKeywords{swedish}
\todo[inline, backgroundcolor=kth-lightblue]{Nyckelord som beskriver innehållet i uppsatsrapporten}
\end{scontents}
\end{abstract}
\babelpolyLangStop{swedish} % Ends language block
\cleardoublepage
\todo[inline, backgroundcolor=kth-lightgreen]{Note that you may need to augment the set of language used in polyglossia or
babel (see the file kththesis.cls). The following languages include those languages that were used in theses at KTH in 2018-2019, except for one in Chinese.\\
Remove those versions that you do not need.\\
If adding a new language, when specifying the language for the abstract use the three letter ISO 639-2 Code – specifically the "B" (bibliographic) variant of these codes (note that this is the same language code used in DiVA).}
\todo[inline]{Use the relevant language for abstracts for your home university.}
\babelpolyLangStart{french} 
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
fre
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Résumé en français
\end{scontents}
\subsection*{Mots clés}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 mots-clés
\end{scontents}
\end{abstract}
\babelpolyLangStop{french}
\cleardoublepage
\babelpolyLangStart{spanish}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
spa
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Résumé en espagnol.
\end{scontents}
Résumé en espagnol

\subsection*{Palabras claves}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 Palabras claves
\end{scontents}
\end{abstract}
\babelpolyLangStop{spanish}
\cleardoublepage
\babelpolyLangStart{italian}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
ita
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Sommario in italiano.
\end{scontents}
\subsection*{parole chiave}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 parole chiave
\end{scontents}
\end{abstract}
\babelpolyLangStop{italian}
\cleardoublepage
% note that a macro is used to avoid Overleaf parsing problems
\babelpolyLangStart{norsk}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
nor
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Sammendrag på norsk.
\end{scontents}
\subsection*{Nøkkelord}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 nøkkelord
\end{scontents}
\end{abstract}
\babelpolyLangStop{norsk}
\cleardoublepage
% note that a macro is used to avoid Overleaf parsing problems
\babelpolyLangStart{ngerman}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
ger
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Zusammenfassung in Deutsch.
\end{scontents}
\subsection*{Schlüsselwörter}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 Schlüsselwörter
\end{scontents}
\end{abstract}
\babelpolyLangStop{ngerman}
\cleardoublepage
\babelpolyLangStart{danish}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
dan
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Abstrakt på dansk.
\end{scontents}
\subsection*{Søgeord}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 Søgeord
\end{scontents}
\end{abstract}
\babelpolyLangStop{danish}
\cleardoublepage
\babelpolyLangStart{dutch}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
dut
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Samenvatting in het Nederlands.
\end{scontents}
\subsection*{Trefwoorden}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 trefwoorden
\end{scontents}
\end{abstract}
\babelpolyLangStop{dutch}
\cleardoublepage
\babelpolyLangStart{estonian}
\begin{abstract}
    \markboth{\abstractname}{}
\begin{scontents}[store-env=lang]
est
\end{scontents}
\begin{scontents}[store-env=abstracts,print-env=true]
Eesti keeles kokkuvõte.
\end{scontents}
\subsection*{Märksõnad}
\begin{scontents}[store-env=keywords,print-env=true]
5-6 Märksõnad
\end{scontents}
\end{abstract}
\babelpolyLangStop{estonian}
\cleardoublepage
% set to the language of the body of the thesis

%\selectlanguage{swedish}
\selectlanguage{english}

\section*{Acknowledgments }
\markboth{Acknowledgments}{}
\todo[inline]{It is nice to acknowledge the people that have helped you. It is
  also necessary to acknowledge any special permissions that you have gotten –
  for example getting permission from the copyright owner to reproduce a
  figure. In this case you should acknowledge them and this permission here
  and in the figure’s caption. \\
  Note: If you do not have the copyright owner’s permission, then you cannot use any copyrighted figures/tables/… .
}
\todo[inline, backgroundcolor=kth-lightblue]{I detta kapitel kan du e v nämna något om
  din bakgrund om det påverkar rapporten på något sätt. Har du t ex inte
  möjlighet att skriva perfekt svenska för att du är nyanländ till landet kan
  det vara på sin plats att nämna detta här. OBS, detta får dock inte vara en
  ursäkt för att lämna in en rapport med undermåligt språk, grammatik och
  stavning (t ex får fel som en automatisk stavningskontroll och
  grammatikkontroll kan upptäcka inte förekomma)\\

En dualism som måste hanteras i hela rapporten och projektet
}

I would like to thank xxxx for having yyyy.\\

\acknowlegmentssignature

\fancypagestyle{plain}{}
\renewcommand{\chaptermark}[1]{ \markboth{#1}{}} 
\tableofcontents
  \markboth{\contentsname}{}

\cleardoublepage
\listoffigures

\cleardoublepage

\listoftables
\cleardoublepage
\lstlistoflistings\todo[inline, backgroundcolor=kth-lightgreen]{If you have listings in your thesis, otherwise delete this preface page.}
\cleardoublepage
\printglossary[type=\acronymtype, title={List of acronyms and abbreviations}]
\todo[inline, backgroundcolor=kth-lightgreen]{The list of acronyms and abbreviations should be in alphabetical order based on the spelling of the acronym or abbreviation.
}
\label{pg:lastPageofPreface}
% Mainmatter is where the actual contents of the thesis goes
\mainmatter

\renewcommand{\chaptermark}[1]{\markboth{#1}{}}
\chapter{Introduction}
\label{ch:introduction}
\todo[inline, backgroundcolor=kth-lightblue]{Introduktion}
\todo[inline]{This chapter describes the
specific problem that this thesis addresses, the context of the problem, the
goals of this thesis project, and outlines the structure of the thesis.}
\todo[inline, backgroundcolor=kth-lightgreen]{The first paragraph after a heading is not indented, all of the
  subsequent paragraphs have their first line indented.}

\todo[inline]{Give a general introduction to the area. (Remember to use appropriate references in this and all other sections.)}

% One can use either biblatex or bibtex - set as the option for the document at the top of this file
\ifbiblatex
\todo[inline, backgroundcolor=kth-lightgreen]{We use the \emph{biblatex} package to handle our references.  We
use the command \texttt{parencite} to get a reference in parenthesis, like
this \textbackslash parencite\{heisenberg2015\} resulting in \parencite{heisenberg2015}.  It is also possible to include the author as part of the sentence using \texttt{textcite}, like talking about the work of \textbackslash textcite\{einstein2016\} resulting in \textcite{einstein2016}.\\
This also means that you have to change the include files to include biblatex and change the way that the reference.bib file is included.}
\else
\todo[inline, backgroundcolor=kth-lightgreen]{We use the \emph{bibtex} package to handle our references.  We therefore
use the command \textbackslash cite\{farshin\_make\_2019\}. For example, Farshin, \etal described how to improve LLC
cache performance in \cite{farshin_make_2019} in the context of links running
at \SI{200}{Gbps}.}
\fi

\todo[inline, backgroundcolor=kth-lightgreen]{Use the glossaries package to help yourself and your readers.
Add the acronyms and abbreviations to lib/acronyms.tex. Some examples are shown below:}
In this thesis we will examine the use of \glspl{LAN}. In this thesis we will
assume that \glspl{LAN} include \glspl{WLAN}, such as \gls{WiFi}.

\section{Background}
\label{sec:background}
\todo[inline, backgroundcolor=kth-lightblue]{Bakgrund}
\todo[inline]{Present the background for the area. Set the context for your project – so that your reader can understand both your project and this thesis. (Give detailed background information in Chapter 2 - together with related work.)\\
Sometimes it is useful to insert a system diagram here so that the reader knows what are the different elements and their relationship to each other. This also introduces the names/terms/… that you are going to usethroughout your thesis (be consistent). This figure will also help you later delimit what you are going to do and what others have done or will do.}

As one can find in RFC 1235\,\cite{ioannidis_coherent_1991} multicast is useful for xxxx .

\section{Problem}
\label{sec:problem}
\todo[inline, backgroundcolor=kth-lightblue]{Problemdefinition / Frågeställning}

\todo[inline]{Longer problem statement\\
If possible, end this section with a question as a problem statement.}

% Research Question
%\label{sec:researchQuestion}
\section{Purpose}
\label{sec:purpose}
\todo[inline, backgroundcolor=kth-lightblue]{Syfte}
\todo[inline]{State the purpose  of your thesis and research.

Describe who benefits and how they benefit if you achieve your goals. Include anticipated ethical, sustainability, social issues, \etc related to your project. (Return to these in your reflections in Section~\ref{sec:reflections}.)}

\section{Goals}
\label{sec:goals}
\todo[inline, backgroundcolor=kth-lightblue]{Mål}

\todo[inline]{State the goal/goals of this degree project.

In addition to presenting the goal(s), you might also state what the deliverables and results of the project are.}

\section{Research Methodology}
\label{sec:researchMethodology}
\todo[inline, backgroundcolor=kth-lightblue]{Forskningsmetodik}

\todo[inline]{Introduce your choice of methodology/methodologies and method/methods – and the reason why you chose them. Contrast them with and explain why you did not choose other methodologies or methods. (The details of the actual methodology and method you have chosen will be given in Chapter~\ref{ch:methods}. Note that in Chapter~\ref{ch:methods}, the focus could be research strategies, data collection, data analysis, and quality assurance.)\\
In this section you should present your philosophical assumption(s), research method(s), and research approach(es).}

\section{Delimitations}
\label{sec:deliminations}
\todo[inline, backgroundcolor=kth-lightblue]{Avgränsningar}
\todo[inline]{Describe the boundary/limits of your thesis project and what you are explicitly not going to do. This will help you bound your efforts – as you have clearly defined what is out of the scope of this thesis project. Explain the delimitations. These are all the things that could affect the study if they were examined and included in the degree project.}

\section{Structure of the thesis}
\todo[inline, backgroundcolor=kth-lightblue]{ Avhandlingens disposition}
Chapter~\ref{ch:background} presents relevant background information about xxx.  Chapter~\ref{ch:methods} presents the methodology and method used to solve the problem. …

\cleardoublepage
\chapter{Background}
\label{ch:background}
\todo[inline, backgroundcolor=kth-lightblue]{Bakgrund}

\todo[inline]{When you do your literature study, you should have a nearly complete Chapters 1 and 2.\\
You may also find it convenient to introduce the future work section into your report early – so that you can put things that you think about but decide not to do now into this section.\\
Note that later you can move things between this future work section and what you have done as you may change your mind about what to do now versus what to put off to future work.
}

This chapter provides basic background information about xxx. Additionally, this chapter describes xxx. The chapter also describes related work xxxx.

\todo[inline]{What does a reader (another x student -- where x is your program) need to know to understand your report?
What have others already done? (This is the “related work”.) Explain what and
how prior work / prior research will be applied on or used in the degree
project /work (described in this thesis). Explain why and what is not used in
the degree project and give valid reasons for rejecting the work/research.}

\section{Major background area 1}
\todo[inline, backgroundcolor=kth-lightblue]{Viktigt bakgrundsområde 1}
There are xxx characteristics that distinguish yyy from other information and communication technology (ICT) system, as shown in Figure~\ref{fig:lotsofstars}. Table \ref{tab:tablecaracteristics} summarizes these characteristics.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/lots_of_stars.png}
  \end{center}
  \caption{Lots of stars  (Inspired by Figure x.y on page z of [xxx])}
  \label{fig:lotsofstars}
\end{figure}

\begin{table}[!ht]
  \begin{center}
    \caption{xxx characteristics}
    \label{tab:tablecaracteristics}
    \begin{tabular}{l|S[table-format=4.6]} % <-- Alignments: 1st column left, 2nd middle, with vertical lines in between
      \textbf{Characteristics} & \textbf{Description}\\
      $\alpha$ & $\beta$ \\
      \hline
      1 & 1110.1\\
      2 & 10.1\\
      3 & 23.113231\\
    \end{tabular}
  \end{center}
\end{table}

\subsection{Subarea 1.1}
Entangled states are an important part of quantum cryptography, but also relevant in other domains. This concept might be relevant for neutrinos, see for example \cite{kim_small-mass_2016}.

\subsection{Subarea 1.1.2}
Computational methods are increasingly used as a third method of carrying out
scientific investigations. For example, computational experiments were used to
find the amount of wear in a polyethylene liner of a hip prosthesis in \cite{maguire_jr_new_2014}.
…

\subsection{Subarea 1.1.2}
Using the nearest data center may improve performance, see \cite{bogdanov_nearest_2015}

\subsection{Link layer Encapsulation}
\label{sec:llencap}

See Figure~\ref{fig:ieee8023-data-packet} which uses the \textsf{bytefield}  \LaTeX\ package.

% Note that one could use kth-coolgray40 instead of lightgray, but this KTH color is not defined for digital output
\begin{figure}[!ht]
	\centering
\begin{bytefield}{21}
\bitbox[]{7}{} & \bitbox[]{3}{\tiny octets:} & \bitbox[]{4}{\tiny 6} & \bitbox[]{4}{\tiny 6} & \bitbox[]{3}{\tiny 2} & \bitbox[]{5}{\tiny 46 to 1500} & \bitbox[]{3}{\tiny 0 to 46} & \bitbox[]{2}{\tiny 4}\\

\bitbox[]{8}{\textbf{ETHERNET \\[-1ex] \tiny{data link-layer}}} & \bitbox[]{2}{} &

\bitbox{4}{\tiny Destination Address} & \bitbox{4}{\tiny Source Address} & \bitbox{3}{\tiny Length/ Type} & 
\bitbox{5}{\tiny Data Payload} & \bitbox{3}{\tiny Padding} &
\bitbox{2}{\tiny CRC} \\

\bitbox[]{1}{} &\bitbox[]{3}{\tiny octets:} & \bitbox[]{4}{\tiny 7} & \bitbox[]{2}{\tiny 1} & \bitbox[]{0}{$\vdots$ \\[1ex]} & \bitbox[]{16}{} & \bitbox[]{0}{$\vdots$ \\[1ex]} & \bitbox[]{5}{} & \bitbox[]{4}{\tiny Variable}\\

\bitbox[]{4}{\textbf{MAC \\[-1ex] \tiny{packet}}} & \colorbitbox{lightgray}{4}{\tiny Preamble} & \colorbitbox{lightgray}{2}{\tiny SFD} & \colorbitbox{lightgray}{16}{\tiny MAC Client Data} & \colorbitbox{lightgray}{3}{\tiny Padding} &
\colorbitbox{lightgray}{2}{\tiny CRC} & \colorbitbox{lightgray}{4}{\tiny Extension}
\end{bytefield}
     \label{fig:ieee8023-data-packet}
     \caption{Ethernet data link layer protocol encapsulated into a IEEE~802.3 MAC packet}
\end{figure}

\subsection{IP packet headers}
\label{sec:ipheaders}
The data link layer will receive a packet from the IP layer. The layout of
an IPv4 packet is shown in Figure~\ref{fig:ipv4-header}. This should be
contrasted with the IPv6 header shown in Figure~\ref{fig:ipv6-header}.

%
% IPv4 packet header
%
\begin{figure}[!ht]
	\centering
\begin{bytefield}{32}
\bitheader{0-31} \\
\bitbox{4}{\footnotesize{Version}} & \bitbox{4}{IHL} & \bitbox{6}{\tiny{Type of Service}} & \bitbox{2}{{\scriptsize ECN}} \bitbox{16}{Total Length}\\
\bitbox{16}{Identification} & \bitbox{3}{Flags} & \bitbox{13}{Fragment Offset}\\
\bitbox{8}{Time to Live} & \bitbox{8}{Protocol} & \bitbox{16}{Header Checksum}\\
\wordbox{1}{Source Address}\\
\wordbox{1}{Destination Address}\\
\colorbitbox{lightgray}{24}{Options} & \colorbitbox{lightgray}{8}{Padding}
\end{bytefield}
     \label{fig:ipv4-header} 
     \caption[IPv4 datagram header]{IPv4 datagram header. Light grey coloured fields are optional.}
\end{figure}

%
% IPv6 packet header
%
\begin{figure}[!ht]
	\centering
\begin{bytefield}{32}
\bitheader{0-31} \\
\bitbox{4}{\footnotesize{Version}} & \bitbox{8}{Traffic Class} & \bitbox{20}{Flow Label}\\
\bitbox{16}{Payload Length} & \bitbox{8}{Next Header} & \bitbox{8}{Hop Limit}\\
\wordbox{4}{Source Address}\\
\wordbox{4}{Destination Address}\\
\end{bytefield}
     \label{fig:ipv6-header}
     \caption{IPv6 datagram header}
\end{figure}

...
\section{Major background area 2}\todo[inline, backgroundcolor=kth-lightblue]{Viktigt bakgrundsområde 2}
...

\section{Related work area}\todo[inline, backgroundcolor=kth-lightblue]{Relaterande arbeten}

\subsection{Major related work 1}
\todo[inline, backgroundcolor=kth-lightblue]{Relaterande arbeten 1}
Carrier clouds have been suggested as a way to reduce the delay between the users and the cloud server that is providing them with content. However, there is a question of how to find the available resources in such a carrier cloud. One approach has been to disseminate resource information using an extension to OSPF-TE, see Roozbeh, Sefidcon, and Maguire \cite{roozbeh_resource_2013}.

\subsection{Major related work}
\todo[inline, backgroundcolor=kth-lightblue]{Relaterande arbeten}

\subsection{Minor related work 1}
\todo[inline, backgroundcolor=kth-lightblue]{Mindre relaterat arbete 1}

…
\subsection{Minor related work n}
\todo[inline, backgroundcolor=kth-lightblue]{Mindre relaterat arbete n}

\section{Summary}\todo[inline, backgroundcolor=kth-lightblue]{Sammanfattning}

\todo[inline]{It is nice to have this chapter conclude with a summary. For
  example, you can include a table that summarizes other people's ideas and
  benefits and drawbacks with each - so as later you can compare your solution
  to each of them. This will also help you define the variables that you will
  use for your evaluation.}

\cleardoublepage
\chapter{Method / Methods}
\label{ch:methods}
\todo[inline, backgroundcolor=kth-lightblue]{Metod / Metodval}
\todo[inline]{This chapter is about Engineering-related
  content, Methodologies and Methods.  Use a self-explaining title.\\The
  contents and structure of this chapter will change with your choice of
  methodology and methods.}

\todo[inline]{Describe the engineering-related contents (preferably with models) and the research methodology and methods that are used in the degree project.

Give a theoretical description of the scientific or engineering methodology are you going to use and why have you chosen this method. What other methods did you consider and why did you reject them.

In this chapter, you describe what engineering-related and scientific skills you are going to apply, such as modeling, analyzing, developing, and evaluating engineering-related and scientific content. The choice of these methods should be appropriate for the problem . Additionally, you should be consciousness of aspects relating to society and ethics (if applicable). The choices should also reflect your goals and what you (or someone else) should be able to do as a result of your solution - which could not be done well before you started.}

The purpose of this chapter is to provide an overview of the research method
used in this thesis. Section~\ref{sec:researchProcess} describes the research
process. Section~\ref{sec:researchParadigm} details the research
paradigm. Section~\ref{sec:dataCollection} focuses on the data collection
techniques used for this research. Section~\ref{sec:experimentalDesign}
describes the experimental design. Section~\ref{sec:assessingReliability}
explains the techniques used to evaluate the reliability and validity of the
data collected. Section~\ref{sec:plannedDataAnalysis} describes the method
used for the data analysis. Finally, Section~\ref{sec:evaluationFramework}
describes the framework selected to evaluate xxx.

\section{Research Process}
\label{sec:researchProcess}
\todo[inline, backgroundcolor=kth-lightblue]{Forskningsprocess}

Figure~\ref{fig:researchprocess} shows the steps conducted in order to carry out this research.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.5\textwidth]{figures/researchprocess.png}
  \end{center}
  \caption{Research Process}
  \label{fig:researchprocess}
\end{figure}

\section{Research Paradigm}
\label{sec:researchParadigm}
\todo[inline, backgroundcolor=kth-lightblue]{Undersökningsparadigm}

\section{Data Collection}
\todo[inline]{This should also show that you are aware of the social and ethical concerns that might be relevant to your data collection method.}
\label{sec:dataCollection}
\todo[inline, backgroundcolor=kth-lightblue]{Datainsamling}

\section{Experimental design/Measurements}
\label{sec:experimentalDesign}
\todo[inline, backgroundcolor=kth-lightblue]{Experimentdesign/mätningar}

\section{Assessing reliability and validity of the data collected}
\todo[inline, backgroundcolor=kth-lightblue]{Bedömning av tillförlitlighet och validitet hos använda metoder och insamlade data }
\label{sec:assessingReliability}

\section{Planned Data Analysis}
\todo[inline, backgroundcolor=kth-lightblue]{Planerad analys av data}
\label{sec:plannedDataAnalysis}

\section{Evaluation framework}
\todo[inline, backgroundcolor=kth-lightblue]{Utvärdering}
\label{sec:evaluationFramework}

\section{System documentation}\todo[inline]{If this is going to be a complete document consider putting it in as an appendix, then just put the highlights here.}
\todo[inline, backgroundcolor=kth-lightblue]{Systemdokumentation}

\cleardoublepage
\chapter{What you did}\todo[inline]{Choose your own chapter title to describe this}
\todo[inline, backgroundcolor=kth-lightblue]{Välj lämplig rubrik, till exempel Genomförande, Konstruktion, Utveckling}
\label{ch:whatYouDid}

\todo[inline]{What have you done? How did you do it? What design decisions did you make? How did what you did help you to meet your goals?}

\section{Hardware/Software design …/Model/Simulation model \& parameters/…}

Figure~\ref{fig:homepageicon} shows a simple icon for a home page. The time
to access this page when served will be quantified in a series of
experiments. The configurations that have been tested in the test bed are
listed in Table~\ref{tab:configstested}.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=0.25\textwidth]{figures/Homepage-icon.png}
  \end{center}
  \caption{Homepage icon}
  \label{fig:homepageicon}
\end{figure}

\begin{table}[!ht]
  \begin{center}
    \caption{Configurations tested}
    \label{tab:configstested}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      \textbf{Configuration} & \textbf{Description} \\
      \hline
      1 & Simple test with one server\\
      2 & Simple test with one server\\
    \end{tabular}
  \end{center}
\end{table}

\section{Implementation …/Modeling/Simulation/…}

\subsection{Some examples of coding}

Listing~\ref{lst:helloWorldInC} shows an example of a simple program written
in C code.

\begin{lstlisting}[language={C}, caption={Hello world in C code}, label=lst:helloWorldInC]
int main() {
printf("hello, world");
return 0;
}
\end{lstlisting}

In contrast, Listing~\ref{lst:programmes} is an example of code in Python to
get a list of all of the programs at KTH.

\lstset{extendedchars=true}
\begin{lstlisting}[language={Python}, caption={Using a Python program to
    access the KTH API to get all of the programs at KTH}, label=lst:programmes]
KOPPSbaseUrl = 'https://www.kth.se'

def v1_get_programmes():
    global Verbose_Flag
    #
    # Use the KOPPS API to get the data
    # note that this returns XML
    url = "{0}/api/kopps/v1/programme".format(KOPPSbaseUrl)
    if Verbose_Flag:
        print("url: " + url)
    #
    r = requests.get(url)
    if Verbose_Flag:
        print("result of getting v1 programme: {}".format(r.text))
    #
    if r.status_code == requests.codes.ok:
        return r.text           # simply return the XML
    #
    return None
\end{lstlisting}

\cleardoublepage
\chapter{Results and Analysis}
\label{ch:resultsAndAnalysis}
\todo[inline, backgroundcolor=kth-lightblue]{Resultat och analys}

\todo[inline]{
Sometimes this is split into two chapters.\\

Keep in mind: How you are going to evaluate what you have done? What are your metrics?\\
Analysis of your data and proposed solution\\
Does this meet the goals which you had when you started?
}

In this chapter, we present the results and discuss them.

\section{Major results}
\todo[inline, backgroundcolor=kth-lightblue]{Huvudsakliga resultat}

Some statistics of the delay measurements are shown in Table~\ref{tab:delayMeasurements}.
The delay has been computed from the time the GET request is received until the response is sent.

\begin{table}[!ht]
  \begin{center}
    \caption{Delay measurement statistics}
    \label{tab:delayMeasurements}
    \begin{tabular}{l|S[table-format=4.2]|S[table-format=3.2]} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      \textbf{Configuration} & \textbf{Average delay (ns)} & \textbf{Median delay (ns)}\\
      \hline
      1 & 467.35 & 450.10\\
      2 & 1687.5 & 901.23\\
    \end{tabular}
  \end{center}
\end{table}

Figure \ref{fig:processing_vs_payload_length} shows and example of the
performance as measured in the experiments.

\begin{figure}[!ht]
% GNUPLOT: LaTeX picture
\setlength{\unitlength}{0.240900pt}
\ifx\plotpoint\undefined\newsavebox{\plotpoint}\fi
\begin{picture}(1500,900)(0,0)
\sbox{\plotpoint}{\rule[-0.200pt]{0.400pt}{0.400pt}}%
\put(171.0,131.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,131){\makebox(0,0)[r]{ 1.5}}
\put(1419.0,131.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,212.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,212){\makebox(0,0)[r]{ 2}}
\put(1419.0,212.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,292.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,292){\makebox(0,0)[r]{ 2.5}}
\put(1419.0,292.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,373.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,373){\makebox(0,0)[r]{ 3}}
\put(1419.0,373.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,454.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,454){\makebox(0,0)[r]{ 3.5}}
\put(1419.0,454.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,534.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,534){\makebox(0,0)[r]{ 4}}
\put(1419.0,534.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,615.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,615){\makebox(0,0)[r]{ 4.5}}
\put(1419.0,615.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,695.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,695){\makebox(0,0)[r]{ 5}}
\put(1419.0,695.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,776.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(151,776){\makebox(0,0)[r]{ 5.5}}
\put(1419.0,776.0){\rule[-0.200pt]{4.818pt}{0.400pt}}
\put(171.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(171,90){\makebox(0,0){ 0}}
\put(171.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(298.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(298,90){\makebox(0,0){ 10}}
\put(298.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(425.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(425,90){\makebox(0,0){ 20}}
\put(425.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(551.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(551,90){\makebox(0,0){ 30}}
\put(551.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(678.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(678,90){\makebox(0,0){ 40}}
\put(678.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(805.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(805,90){\makebox(0,0){ 50}}
\put(805.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(932.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(932,90){\makebox(0,0){ 60}}
\put(932.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1059.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1059,90){\makebox(0,0){ 70}}
\put(1059.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1185.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1185,90){\makebox(0,0){ 80}}
\put(1185.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1312.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1312,90){\makebox(0,0){ 90}}
\put(1312.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1439.0,131.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(1439,90){\makebox(0,0){ 100}}
\put(1439.0,756.0){\rule[-0.200pt]{0.400pt}{4.818pt}}
\put(171.0,131.0){\rule[-0.200pt]{0.400pt}{155.380pt}}
\put(171.0,131.0){\rule[-0.200pt]{305.461pt}{0.400pt}}
\put(1439.0,131.0){\rule[-0.200pt]{0.400pt}{155.380pt}}
\put(171.0,776.0){\rule[-0.200pt]{305.461pt}{0.400pt}}
\put(30,453){\rotatebox{-270}{\makebox(0,0){Processing time (ms)}}}
\put(805,29){\makebox(0,0){Payload size (bytes)}}
\put(868.0,131.0){\rule[-0.200pt]{0.400pt}{84.074pt}}
\put(995.0,131.0){\rule[-0.200pt]{0.400pt}{98.287pt}}
\put(1173.0,131.0){\rule[-0.200pt]{0.400pt}{118.041pt}}
\put(1325.0,131.0){\rule[-0.200pt]{0.400pt}{134.904pt}}
\put(1350.0,131.0){\rule[-0.200pt]{0.400pt}{137.795pt}}
\put(1439.0,131.0){\rule[-0.200pt]{0.400pt}{155.380pt}}
\end{picture}
\caption[A GNUplot figure]{Processing time vs. payload length}\vspace{0.5cm}
\label{fig:processing_vs_payload_length}
\end{figure}

Given these measurements, we can calculate our processing bit rate as the inverse of the time it takes to process an additional byte divided by 8 bits per byte:

\[
	bitrate = \frac{1}{\frac{time_{byte}}{8}} = 20.03 \quad kb/s
\]

\section{Reliability Analysis}
\todo[inline, backgroundcolor=kth-lightblue]{Analys av reliabilitet}

\section{Validity Analysis}
\todo[inline, backgroundcolor=kth-lightblue]{Analys av validitet}

\cleardoublepage
\chapter{Discussion}
\label{ch:discussion}
\todo[inline, backgroundcolor=kth-lightblue]{Diskussion}

\cleardoublepage
\chapter{Conclusions and Future work}
\label{ch:conclusionsAndFutureWork}
\todo[inline, backgroundcolor=kth-lightblue]{Slutsatser och framtida arbete}

\todo[inline]{Add text to introduce the subsections of this chapter.}

\section{Conclusions}
\label{sec:conclusions}
\todo[inline, backgroundcolor=kth-lightblue]{Slutsatser}
\todo[inline]{Describe the conclusions (reflect on the whole introduction given in Chapter 1).}

\todo[inline]{Discuss the positive effects and the drawbacks.\\
Describe the evaluation of the results of the degree project.\\
Did you meet your goals?\\
What insights have you gained?\\
What suggestions can you give to others working in this area?\\
If you had it to do again, what would you have done differently?}

\section{Limitations}
\label{sec:limitations}
\todo[inline, backgroundcolor=kth-lightblue]{Begränsande faktorer}
\todo[inline]{What did you find that limited your
  efforts? What are the limitations of your results?}

\section{Future work}
\label{sec:futureWork}
\todo[inline, backgroundcolor=kth-lightblue]{Framtida arbete}
\todo[inline]{Describe valid future work that you or someone else could or should do.\\
Consider: What you have left undone? What are the next obvious things to be done? What hints can you give to the next person who is going to follow up on your work?
}

Due to the breadth of the problem, only some of the initial goals have been
met. In these section we will focus on some of the remaining issues that
should be addressed in future work. ...

\subsection{What has been left undone?}
\label{what-has-been-left-undone}

The prototype does not address the third requirement, \ie, a yearly
unavailability of less than 3 minutes, this remains an open problem. ...

\subsubsection{Cost analysis}

The current prototype works, but the performance from a cost perspective makes
this an impractical solution. Future work must reduce the cost of this
solution, to do so a cost analysis needs to first be done. ...

\subsubsection{Security}

A future research effort is needed to address the security holes that results
from using a self-signed certificate. Page filling text mass. Page filling
text mass. ...

\subsection{Next obvious things to be done}

In particular, the author of this thesis wishes to point out xxxxxx remains as
a problem to be solved. Solving this problem is the next thing that should be
done. ...

\section{Reflections}
\label{sec:reflections}
\todo[inline, backgroundcolor=kth-lightblue]{Reflektioner}
\todo[inline]{What are the relevant economic, social,
  environmental, and ethical aspects of your work?
}

One of the most important results is the reduction in the amount of
energy required to process each packet while at the same time reducing the
time required to process each packet.

The thesis contributes to the \gls{UN}\enspace\glspl{SDG} numbers 1 and 9 by
xxxx.

\noindent\rule{\textwidth}{0.4mm}
\todo[inline]{In the references, let Zotero or other tool fill this
  in for you. I suggest an extended version of the IEEE  style, to include
  URLs, DOIs, ISBNs, \etc, to make it easier for your reader to find
  them. This will make life easier for your opponents and examiners. \\

  IEEE Editorial Style Manual: \url{https://www.ieee.org/content/dam/ieee-org/ieee/web/org/conferences/style_references_manual.pdf}
}

\cleardoublepage
% Print the bibliography (and make it appear in the table of contents)

\renewcommand{\bibname}{References}
\addcontentsline{toc}{chapter}{References}

\ifbiblatex
    \printbibliography[heading=bibintoc]
\else
    \bibliography{references}
\fi

\cleardoublepage
\appendix
\renewcommand{\chaptermark}[1]{\markboth{Appendix \thechapter\relax:\thinspace\relax#1}{}}
\chapter{Something Extra}
\todo[inline, backgroundcolor=kth-lightblue]{Bilaga}

\section{Just for testing KTH colors}
\ifdigitaloutput
    \textbf{You have selected to optimize for digital output}
\else
    \textbf{You have selected to optimize for print output}
\fi
\begin{itemize}[noitemsep]
    \item Primary color
    \begin{itemize}
    \item \textcolor{kth-blue}{kth-blue \ifdigitaloutput
    actually Deep sea
    \fi} {\color{kth-blue} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-blue80}{kth-blue80} {\color{kth-blue80} \rule{0.3\linewidth}{1mm} }\\
\end{itemize}

\item  Secondary colors
\begin{itemize}[noitemsep]
    \item \textcolor{kth-lightblue}{kth-lightblue \ifdigitaloutput
    actually Stratosphere
    \fi} {\color{kth-lightblue} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-lightred}{kth-lightred \ifdigitaloutput
    actually Fluorescence\fi} {\color{kth-lightred} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-lightred80}{kth-lightred80} {\color{kth-lightred80} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-lightgreen}{kth-lightgreen \ifdigitaloutput
    actually Front-lawn\fi} {\color{kth-lightgreen} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-coolgray}{kth-coolgray \ifdigitaloutput
    actually Office\fi} {\color{kth-coolgray} \rule{0.3\linewidth}{1mm} }\\

    \item \textcolor{kth-coolgray80}{kth-coolgray80} {\color{kth-coolgray80} \rule{0.3\linewidth}{1mm} }
\end{itemize}
\end{itemize}

\textcolor{black}{black} {\color{black} \rule{\linewidth}{1mm} }
\cleardoublepage
\fancyhead{}  % Do not use header on the following pages
% as the separator pages do not have page numbers or headings and
%the individual included papers will have their own page headings
\chapter{Included papers} % just for some experiments with kthpaper package

\paper{Title of paper}{citation for published or submitted paper}{permission or copyright information}

\paper{A New Automated Way to Measure Polyethylene Wear in THA Using a High Resolution CT Scanner: Method and Analysis}{G. Q. Maguire Jr., M. E. Noz, H. Olivecrona, M. P. Zeleznik, and L. Weidenhielm, “A New Automated Way to Measure Polyethylene Wear in THA Using a High Resolution CT Scanner: Method and Analysis,” \textit{The Scientific World Journal}, vol. 2014, pp. 1–9, 2014. doi: 10.1155/2014/528407.}{Copyright © 2014 Gerald Q. Maguire Jr. \etal. This is an open access article distributed under the \href{http://creativecommons.org/licenses/by/3.0/}{Creative Commons Attribution License}, which permits unrestricted use, distribution, and reproduction in any medium, provided the original work is properly cited.}
\cleardoublepage
\includepdf[pages=-,pagecommand=\thispagestyle{plain}]{./papers/automated_2014-528407.pdf}

\paper{Yet another title}{Oscar Student and A. B. Normal, “Yet another title,” \textit{A Fictitious Journal}, vol. 2021, pp. 1024-2048, 2021. doi: 10.123456789/foo.2021.xxx.}{Copyright © 2021 XXXX. Reprinted with permission.}

\label{pg:lastPageofMainmatter}

\clearpage
\fancyhead{}  % Do not use header on this extra page or pages
\section*{For DIVA}
\divainfo{pg:lastPageofPreface}{pg:lastPageofMainmatter}
\end{document}

### 2022-08-19 publications/2018DemoScalingOnTheEdge/paper_body.tex
hotmbile

\section{Introduction}

Previous works on cloudlets~\cite{satya2009case,Ha:ImpactMobile}, one of the earliest incarnation of edge computing, enable small data-centers at the edge of the Internet. 
Many futuristic applications become viable with these clusters that are only one wireless hop away. 
One of the most promising genres of these emerging applications is human-in-the-loop applications such as wearable cognitive assistance~\cite{Ha:TowardsWearableCogAssist}.

In these applications, sensor data, for example video and audio, are continuously streamed to a cloudlet, where they are analyzed in realtime in order to assist users to complete a particular task. 
Researchers have built prototypes of these applications to help users assemble LEGO models and IKEA furniture, and even learn how to play ping-pong~\cite{satya2009case,Chen:EarlyImplementation}.
% In general, all human-in-the-loop applications capture in some form user and environment information and typically leverage compute-intensive algorithms to analyze the data in order to provide feedback to the user.
% On the side of the sensory input, typically video, but also audio as well as further user-related data (like orientation and movement) is captured.
% This  data is forwarded to the processing unit where the first step is usually to run detection algorithms on the sensory input.
% In case of a positive result from the detection, the next step is then the generation of feedback to be sent back to the user.

Cognitive assistance applications are highly interactive. 
Feedback is sent back to the user once the application detects interesting events, for example, when the user places the wrong LEGO block on the model.
% For this the detected features are put into the context of the specific application and its task model (for instance for assistance applications a comparison to the current task goal and potentially the required step towards the next task goal) which then leads potentially to feedback that has to be provided to the user.
% Thus, the next step is to generate this feedback and forward this back to the user, where it is exposed to the user.
%, as shown in Figure~\ref{fig:app-pipeline}.
%, from which on a user takes further actions, triggering the next loop of the application. 
The feedback loop is then repeated until the user finishes the task.
It is important to note that not all sensory input triggers feedback.
Take for instance, an application which relies on image recognition.
Inevitably, some frames are going to receive confidence values below a set threshold from the image recognition algorithms, and thus do not generate feedback.
We will refer to these inputs as \emph{feedback-poor}, and conversely, refer to inputs which do generate feedback as \emph{feedback-rich}.

% In the following, we distinguish between input (i.e. frames) triggering feedback whereas frames for which no new state is detected are referred to as ones that do not provide feedback (i.e. frames without feedback).

These principles result in the following characteristics of human-in-the-loop applications powered by edge computing:

\begin{description}[labelindent=\parindent, listparindent=\parindent, style=unboxed, leftmargin=0cm]
	\item[Latency Sensitive:] Given their tight interaction with the physical world, the quality of human-in-the-loop applications is determined by the latencies experienced by users. 
    These applications are different from conventional mobile applications by the low latency requirements inherent to the applications themselves~\cite{Suzuki,Chen:AnEmpiricalStudyOfLatency}. 
    For example, consider a Ping-Pong Assistance application that instructs a user where to hit a ball -- any instruction delivered after the user has made a hit is useless.
  Hence, the average and the distribution of end-to-end latency, in particular for \emph{feedback-rich} inputs, can serve as good metrics for a benchmark tool. 
% Previous work has shown that offload to cloudlet reduces 100ms to 200ms in end-to-end latency from offload to the cloud \cite{Suzuki,Chen:AnEmpiricalStudyOfLatency}.
% Thus, the low latency benefits provided by cloudlets distinguish human-in-the-loop applications from conventional mobile or web applications. 
% Therefore, low latency is a salient characteristic of cloudlet-enabled applications.

    \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly Deep Neural Networks (DNNs). 
    Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{flinn2012cyber}. 
    While state-of-the-art DNN object detectors can run at more than 20 FPS on a server GPU, their performances are much worse on mobile GPUs - some models cannot even be loaded due to memory constraints. 
    Cloudlet-based applications overcome these challenges by offloading the computation. 
%    Offloading makes sense when the computation is intensive or special hardwares are needed.

\end{description}

% Experimentally, it has been shown \cite{Chen:AnEmpiricalStudyOfLatency} that human-in-the-loop applications are perceived within one of three coarse quality-of-experience ranges: Flawless, impaired and useless. If such an application experiences a round-trip delay below a certain threshold $\tau_1$, further improvememt of the round-trip delay does not lead to a better perceived quality of experience (at least with respect to subjective user experience). The delay range below threshold $\tau_1$ is hence referred to as flawless. Furthermore, there is a second round-trip delay threshold $\tau_2$ which represents a cut-off beyond which a human-in-the-loop application is perceived as useless, i.e. an assistance application can not provide any help due to the delays in this case. The range in between $\tau_1$ and $\tau_2$ is generally perceived as a range where the application can still be operated but a degradation in the quality of experience is perceived, which degrades with the absolute value of the round-trip time within the given range. 
% In terms of absolute values, $\tau_1$ and $\tau_2$ are generally application-dependent while spanning from a few tens to a few hundreds of milliseconds.
% Finally, note that for the round-trip times in particular the sensory input triggering feedback is of key relevance for the perceived quality-of-experience.

Benchmarking infrastructures for these human-in-the-loop applications is challenging -- the main issue arises from the involvement of humans. % in the application.
Applications' execution path and resource utilization vary among users. 
For example, in a task guidance application, the reaction speed of the human to a new instruction governs the inter-arrival time of the next \emph{feedback-rich} input.
Furthermore, large scale evaluation of these applications require the involvement of many human users. % and provide feedback.
Both these aspects significantly limit experimental studies that could be done to improve architectures, algorithms, and protocols due to costs, efforts as well as reproducibility.

\section{The Measurement Framework}\label{sec:framework}

% \begin{figure*}[t]
% 	\centering
%     \begin{minipage}{.7\linewidth}
%     	\centering
%         \includegraphics[width=.9\textwidth]{img/app_pipeline.pdf}
%         \captionof{figure}{Generic Human-in-the-loop Application Pipeline}
%         \label{fig:app-pipeline}
%     \end{minipage}%
% 	\begin{minipage}{.3\linewidth}
%     	\centering
% 		\includegraphics[width=\linewidth]{img/TraceReplay_GenArch}
%         \captionof{figure}{General architecture of the benchmarking suite.}
% 		\label{fig:TraceReplayArch}
% 	\end{minipage}
% \end{figure*}

% We furthermore argue that human-in-the-loop applications like AR potentially require costly gadgets for providing feedback to the human user. 
% Hence, this is another factor that contributes to the costs of such trials. 
% We hence strive for an approach to scalability tests of human-in-the-loop applications that can overcome these challenges while still sticking to an experimental approach in general.

%We address these challenges through an emulation approach.
%We address the discussed challenges on different levels.

To establish a reproducible and comparable workload, the first step in our methodology is to record a trace of the sensory input data while having a human user operate the target application. 
% The trace simply consists of the payload packets passed to the socket of the sensory device, which can be collected by TCP dump.
The collected data consists of the sensory inputs provided to the system at runtime, for instance, in the case of a visual application, the video feed from the camera.

To use the trace for reproducible experiments, we developed a benchmarking suite which can replay the trace to the original application, which results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
During the replay, many system level metrics can be collected, such as round-trip and processing times.
By enabling the client-side to play out the trace from a file, it becomes independent from human operation.
% For this, in particular a model of the human behavior needs to be defined and considered in the way the recorded trace is provided to the backend compute process.

%\subsection{User Model}

In order to imitate a human behavior as close as possible, we propose the following user model as shown in Figure~\ref{fig:usermodel}.
We assume a user that is patient and does not make mistakes; any error message received from the application backend is ignored. We first divide a trace into steps that corresponds to individual events that should trigger feedback. If a positive feedback is received from the application, we jump ahead in the trace to replay the next step as if a human user reacts to the feedback. 
On the other hand, whenever the end of the current step is reached without having received any \emph{positive} feedback, the step is rewound a number $\tau$ of seconds.
To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.

\begin{figure}
    % Define block styles
    %         \tikzstyle{block} = [circle, draw, fill=red!20, 
    %         text width=5em, text centered, rounded corners, minimum size=4em]
    %         \tikzstyle{decision} = [diamond, aspect=2, draw, fill=yellow!20,
    %         text width=5em, text centered, rounded corners, minimum size=4em]
    %        \tikzstyle{line} = [draw, -{Latex[length=3mm]}]
    \centering
    \adjustbox{scale=0.75}{
        \begin{tikzpicture}[align=center,
        node distance=.5cm and 1.5cm,
        every initial by arrow/.style={-{Latex[length=2mm]}}]
        % Place nodes              
        \node [initial, state, minimum size=6em, initial text=] (play) {Play};               
        \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
        \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
        \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};

        % Draw edges
        \path[draw, -{Latex[length=2mm]}]
        (play) edge [bend right=20] node[left] {Step done} (rewind)
        edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
        edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)

        (change) edge [bend left=20] node[right] {Step changed} (play)
        edge [bend left=20] node[right] {All steps done} (shutdown)

        (rewind) edge [bend right=20] node[right] {Rewound} (play)
        edge [bend right=20] node[right] {Too many rewinds} (shutdown);

        \end{tikzpicture}
    }
    \caption{State diagram of the user model}
    \label{fig:usermodel}
\end{figure}

%A state diagram for the behavior of this user can be seen in Figure \ref{fig:usermodel}.

% Finally, by enabling this benchmark suite to run on commodity-of-the-shelf hardware, in our example we implemented the suite as an Android app, costs can furthermore be reduced by avoiding expensive gadgets to be available. 
% The suite is also based on a well-established cognitive assistance application framework, \emph{Gabriel} \cite{Chen:EarlyImplementation,Ha:TowardsWearableCogAssist}, which we believe will further aid in its adoption.

% In the following, we discuss the architecture and implementation of our benchmarking suite in more detail in particular with respect to wearable cognitive assistance applications.
% However, we stress that our methodology and in fact also our benchmark suite is applicable to most human-in-the-loop application as long as the backend code is available and the sensory input data can be captured.
% \footnote{We plan to make the benchmarking suite available to the community as Free and Open Source Software. Additionally, we plan to release the recorded traces under a Creative Commons License.}

\subsection{Architecture}

The suite has three elements, as shown in Figure~\ref{fig:TraceReplayArch}:
% the \emph{application backend}, the \emph{client emulator}, and the \emph{control backend} or \emph{server}, which interact in the following ways:

\begin{description}[labelindent=\parindent, listparindent=\parindent, style=unboxed, leftmargin=0cm]
  \item [The application backend] consists of instances of the target application running on \textcite{docker} containers. 
  These correspond to real, unaltered instances of said cognitive assistance applications -- we do not model or emulate them in any way -- and they are containerized in order to be able to execute an arbitrary number of them on the same cloudlet.
  \item [The client emulator] consists of an Android application which emulates the behavior of a user operating the target cognitive assistance application while following the previously discussed user model.
  This Android app replays the previously recorded sensory data over the network to a specific \emph{application backend}, while collecting statistics and measurements of the system status. 
  %{\color{blue} Junjue: Should there be an application client in the picture? Logically, the benchmark android app feeds the application client the sensory information it needs, almost like hijacking the camera stream.}
  \item [The control backend] also runs on the cloudlet, although it could be executed in a separate cloud or cloudlet. It controls the execution of the experiments, by controlling the \emph{client emulators} over the network, initializing the \emph{application backends} and finally aggregating collected data when the experiments are completed. 
\end{description}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{img/TraceReplay_GenArch}
  \captionof{figure}{General architecture of the benchmarking suite.}
  \label{fig:TraceReplayArch}
\end{figure}

\section{Demo Overview}

We will demonstrate the benchmarking framework using the \emph{LEGO Task Guidance} application previously developed by \textcite{Chen:EarlyImplementation}.
This application guides a user step by step in the assembly of a LEGO model; input consists exclusively of video feed of the current state of the assembly, whereas feedback includes visual and auditory components in the form of animations and speech, respectively.

The benchmarking tool will be employed to extract key real-time metrics from this application, \emph{e.g.} average computation and network times, as well as the estimated \emph{user experience} level of the system as a whole (\emph{flawless}, \emph{impaired} or \emph{unusable}, based on the categorization in \cite{Chen:AnEmpiricalStudyOfLatency}).

\section{Discussion and Future Work}

Some open questions and challenges remain to be tackled in the design and development of the presented tool.
To start with, the benchmarking suite is relatively narrow in the types of applications it can be applied to, currently only targeting event based cognitive assistance applications.
The tool needs to be extended to work on a much a broader spectrum of applications, in particular those that do not have a clear task model.
Furthermore, the current implementation does not support hardware accelerators (e.g. GPUs) that are commonly used for DNN inference.

In addition, our current user model is simplistic. 
This model could be expanded to emulate a human user more accurately and realistically, e.g. making mistakes and responding to feedback to correct them.

We plan to extend our work in two directions. First, in addition to emulating user behaviors, we plan to simulate all the components in the system in order to generate reproducible experiments, test individual components of a real system, and identify performance bottlenecks when many users are using an application concurrently. Second, we are going to develop a statistical characterization of the application footprint, based on the data obtained from the tool.

### 2022-08-19 publications/2019ScalingOnTheEdge/img/legoflow.tex
hotmbile

\adjustbox{scale=0.65}{
    \ttfamily\centering%\fbox{%
    \begin{tikzpicture}[align=center,
    node distance=0.7cm and 0.7cm,
    every initial by arrow/.style={draw=none, minimum size=0em}]
    % Place nodes       
    \node [state, draw=none, minimum size=0em] (input) {};       
    \node [state, right=1cmof input, minimum size=7em] (detect) {Detection};               
    \node [state, right=2cm of detect, minimum size=7em] (symb) {Symbolic\\Repr.};
    \node [state, right=.5cm of symb, minimum size=7em] (feedback) {Feedback\\Gen.};
    \node [state, accepting, draw=none, right=1cm of feedback, minimum size=0em] (output1) {};
    \node [state, accepting, draw=none, below=1.4cm of output1, minimum size=0em] (output2) {};

    % Draw edges
    \path[draw, -{Latex[length=3mm]}]
    (input) edge node[above left] {Input} (detect)
    (detect) edge node[above] {LEGO bricks\\in input} (symb)
    (symb) edge node {} (feedback)
    (feedback) edge node[above right] {Feedback} (output1)
    (detect) edge [bend right=10] node[below left] {No LEGO bricks, thus no feedback} (output2) ;
    \end{tikzpicture}
}%}

### 2022-08-19 publications/2019ScalingOnTheEdge/img/pipeline.tex
hotmbile

\adjustbox{scale=0.65}{
        \ttfamily\centering%\fbox{%
        \begin{tikzpicture}[align=center,
                        node distance=0.7cm and 0.7cm,
                        every initial by arrow/.style={draw=none, minimum size=0em},
                        %node styles:
                        block_center/.style ={rectangle, draw=black, thick, fill=white,
                                        text width=8em, text centered, minimum height=4em},
                        block_left/.style ={rectangle, draw=black, thick, fill=white,
                                        text width=16em, text ragged, minimum height=4em, inner sep=6pt},
                        block_noborder/.style ={rectangle, draw=none, thick, fill=none,
                                        text width=18em, text centered, minimum height=1em},
                        block_assign/.style ={rectangle, draw=black, thick, fill=white,
                                        text width=18em, text ragged, minimum height=3em, inner sep=6pt},
                        block_lost/.style ={rectangle, draw=black, thick, fill=white,
                                        text width=16em, text ragged, minimum height=3em, inner sep=6pt},
                        block_rounded/.style ={rectangle, draw=black, thick, fill=white,
                                        text width=8em, text centered, rounded corners=.55cm, minimum height=4em},
                        line/.style ={draw, thick, -{Latex[length=3mm]}, shorten >=0pt}
                ]

                \matrix [column sep=5mm,row sep=3mm] (top) {
                        \node [block_center] (detect) {Detection};
                         & \node [block_center] (repr) {Symbolic\\Repr.};
                         & \node [block_center] (model) {Task Model\\$M\{S, E\}$};
                         & \node [block_center] (feedback) {Feedback\\Generation}; \\[3ex]
                };

                \matrix [column sep=13mm,row sep=3mm, below=of top] (bottom) {
                        \node [block_rounded] (sensors) {On-body\\Sensors};
                         %& \node [block_rounded] (user) {Human User};
                         %&\node[cloud, draw, black, text centered, cloud ignores aspect=false] (user) {Human User};
                         & \node[maninblack, minimum height=5em] (user) {\Large Human User};
                         & \node [block_rounded] (hud) {HUD, Speakers,\\etc.}; \\
                };

                %\Smiley[1, below of=bottom] (smiley);

                \begin{scope}[every path/.style=line]
                        \path (detect) -- (repr);
                        \path (repr) -- (model);
                        \path (model) -- (feedback);
                        \path (feedback) -- (hud);
                        \path (hud) -- (user);
                        \path (user) -- (sensors);
                        \path (sensors) -- (detect);
                \end{scope}

                % % Place nodes       
                % \node [initial, state, draw=none, initial text=, minimum size=0em] (input) {};
                % \node [state, right=1cmof input, minimum size=7em] (detect) {Detection};
                % \node [state, right=2cm of detect, minimum size=7em] (symb) {Symbolic\\Repr.};
                % \node [state, right=.5cm of symb, minimum size=7em] (feedback) {Feedback\\Gen.};
                % \node [state, accepting, draw=none, right=1cm of feedback, minimum size=0em] (output1) {};
                % \node [state, accepting, draw=none, below=1.4cm of output1, minimum size=0em] (output2) {};

                % % Draw edges
                % \path[draw, -{Latex[length=3mm]}]
                % (input) edge node[above left] {Input} (detect)
                % (detect) edge node[above] {LEGO bricks\\in input} (symb)
                % (symb) edge node {} (feedback)
                % (feedback) edge node[above right] {Feedback} (output1)
                % (detect) edge [bend right=10] node[below left] {No LEGO bricks, thus no feedback} (output2) ;
        \end{tikzpicture}
}%}

### 2022-08-19 publications/2019ScalingOnTheEdge/paper_body.tex
hotmbile

\section{Introduction}\label{sec:intro}

There is increasing interest from academia and industry in novel applications such as immersive augmented reality (AR) or wearable cognitive assistance (WCA)~\cite{Chatzopoulos:Hyperion,Ha:TowardsWearableCogAssist}, also known as human-in-the-loop applications.
These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as Deep Neural Networks (DNN)~\cite{Ha:TowardsWearableCogAssist}.
\Cref{fig:pipeline2} depicts such a system.
These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{Chen:AnEmpiricalStudyOfLatency}.

In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{Bittmann_Edge,flinn2012cyber,Chen:AnEmpiricalStudyOfLatency,Ha:JITProvisioning}.
However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
Finally, in production use these applications will most likely run concurrently with others.
How does this, together with other best-effort applications, impact the latencies perceived by the human user?
Existing studies~\cite{Ha:TowardsWearableCogAssist, Chen:EarlyImplementation, satya2009case,Chatzopoulos:Hyperion} of this class of applications have only lightly touched upon these issues~\cite{Chen:AnEmpiricalStudyOfLatency}.
On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{Zubaidy15,Schiessl17} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.

The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.

Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
This sensory information is then processed by the original compute process at the backend, generating feedback.
However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications

The rest of the paper is structured as follows:
\Cref{sec:background} presents some background on human-in-the-loop applications.
\Cref{sec:approach} discusses the general approach taken with EdgeDroid, while \Cref{sec:implementation} exposes the implementation details for EdgeDroid 1.0.
We show the value of the tool through use case analysis in \Cref{sec:usecases,sec:results}, before discussing future work and concluding in \Cref{sec:conclusions}.

%\section{Design \& Implementation}\label{sec:design_impl}
\section{Background}\label{sec:background}

Human-in-the-loop applications are novel applications aiming to seamlessly integrate into the lives of users and provide real-time, context-aware information.

Given the wide range of applications which fall under this concept, we have chosen to focus our initial efforts on one particular category: task-guidance wearable cognitive assistance (WCA)~\cite{Ha:TowardsWearableCogAssist}.
We have chosen this type of application for two reasons: one, their relative simplicity in terms of execution flow, and two, the already well-established existing body of work~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation,Chen:AnEmpiricalStudyOfLatency}.
Task-guidance WCA applications aim to guide users in the execution of a task by monitoring user actions and providing real-time instructions, usually through wearable sensors and gadgets such as the Google Glass platform.
They have many potential use cases including training and assistance for professionals performing complex tasks.
Imagine for instance a specialized technician performing field repairs on a complex piece of machinery.
A task-guidance WCA application could offer real-time guidance in this task, by analyzing a real-time video feed from the technician's head-mounted camera and providing step-by-step repair instructions.

Broadly speaking, task-guidance WCA applications (and human-in-the-loop in general) process a multitude of inputs in parallel, which are also usually continuous, multidimensional and time-sensitive, e.g.\ video and audio feeds.
These inputs are passed to the computation backend, where they are processed in the pipeline depicted in \cref{fig:pipeline2}.
The first step of the backend processing is detection, which acts as a filter for irrelevant inputs.
For example, this step could consist of a computer vision detector which discards all frames for which the relevant features were not detected or which were below a set threshold.
%Given the often complex nature of the inputs, this stage usually employs Machine Learning to process them.
The remaining inputs pass on to the symbolic representation stage, where features are extracted and parameterized for subsequent computation.
For video frames, this parametrization would usually convert the visual data into a matrix representation of the  relevant features.
This representation of the inputs then continues on to the task model, where the actual application logic resides, before finally passing through the feedback generation stage at which point human-parseable feedback is generated; for instance, animations and voice commands.

\begin{figure}[b]
    \centering
    \input{img/pipeline.tex}
    \caption{Pipeline of human-in-the-loop applications.}\label{fig:pipeline2}
\end{figure}

The task model of a task-guidance WCA is a parameterized representation of the task in question and the steps required to complete it.
It can be represented as a Finite State Machine (FSM) \( M\{S, E\} \), where \(S\) is a set of states and \(E\) is a set of edges connecting these states.
Each state \(s_i \in S\) represents a configuration the application could potentially reach in an arbitrary execution, and each edge \(e_{i,j} \in E\) corresponds to the ability of the application to switch from state \(i\) to state \(j\) based on some user input.
We make a distinction between the set of \emph{steps} required to complete the task, \(S_s\), and \(S\), since the latter may also contain states which represent user mistakes in the execution of the task.
Thus, if we call the set of errors \(S_e\), we can further define \(S := S_s \cup S_e\).

\begin{figure}[tb]
    \centering
    \input{img/taskmodel.tex}
    \captionof{figure}{Segment of the internal task model of a linear task-guidance WCA.}%
    \label{fig:taskmodel}
\end{figure}

These formalisms are exemplified in \cref{fig:taskmodel}, which represents an arbitrary segment of a linear task-guidance WCA application.
\(s_{i}, s_{j} \in S\) represent sequential states in the execution of the task, with the edge \(e_{i,j}\) symbolizing the unique correct way of transitioning between them.
While in \(s_i\) and in the absence of inputs, the task model continuously provides instructions to the user on how to move to \(s_j\).
We will refer to this type of output as \emph{positive} feedback, since it guides the user forward in the execution of the task.
On the other hand, in the case of an erroneous input by the user, the task model moves to \(s^{e}_{k}\), where it will constantly provide instructions until the error is corrected and normal execution can resume.
This type of output directs the user to move backwards in the task model, and thus we will refer to it as \emph{negative} feedback.

\section{Approach of EdgeDroid}\label{sec:approach}

A system for benchmarking human-in-the-loop applications needs thus not only to be able to generate realistic, real-time inputs that follow the behavior of a real user but also be able to correctly react to feedback from the target application.
The design of EdgeDroid 1.0 tackles these challenges from two angles.
One, the generation of realistic inputs is delegated to a human user and provided to the suite in the form of a trace.
This ensures that the raw sensory input data is realistic.
Two, we propose the use of a \emph{user model} to adapt the replay of the trace to the task model and current system conditions.

Concretely, our proposed approach works in the following way:
\begin{enumerate}
    \item A trace of the sensory inputs for an execution of the task is recorded.
          For example, for a video-based application, this trace could consist of a recording from the point of view of a user performing the task.
          % Thanks to the user model, a single trace can be used for wildly different scenarios.
    \item This trace is \emph{manually} segmented into the logical steps which lead to the completion of the task.
          In other words for the generation of the trace the task model must be known to the human operator.
    \item The benchmarking suite is then configured to use this trace and a certain number of virtual users.
          This is done through a TOML~\cite{toml} configuration file on the backend.
    \item The benchmark is executed.
          Mobile devices are used to emulate users with wearable devices connecting to the \emph{real cloudlet} over the \emph{real network}.
          These devices replay the aforementioned trace, employing a \emph{user model} to adapt the trace to changing system conditions and navigate the task model to reach the desired system state.
          %\item The benchmarking suite finally outputs statistics and measurements once the benchmark finalizes.
\end{enumerate}

The extraction of metrics pertaining to the distribution of latencies across the application pipeline is the initial focus of EdgeDroid 1.0.
We calculate latencies by synchronizing clocks across the system components and storing timestamps at key points in the feedback loop.
These points include when input is sent and received, as well as when feedback is sent and received.
We collect raw data about each input-feedback cycle between user and cloudlet, which includes metrics on all the major steps in the feedback loop: uplink, downlink and processing time, presence of feedback, payload size, and so on.
This allows us to aggregate and obtain relevant statistics in postprocessing, such as average delays per step or distribution of average delays across the feedback loop.
We also store metadata regarding the task model with each measurement, for instance to link the state of the system with the current step being performed.

Note that this work does not directly target the \emph{motion-to-photon} latency metric.
This metric includes components such as sensing time and display time, which we do not consider.
We also do not evaluate the accuracy of the applications themselves, as we consider them to be \emph{black boxes}.
The system \emph{can} however be used to evaluate the trade-off between accuracy and performance, by comparing benchmarks performed using traces corresponding to different levels of accuracy.

\section{Implementation of EdgeDroid}\label{sec:implementation}

\begin{figure}[tb]
    \centering
    \includegraphics[width=.7\columnwidth]{img/TraceReplay_GenArch}
    \caption{Architecture of the benchmarking suite.}\label{fig:TraceReplayArch}
\end{figure}

%In the following, we will detail the implementation of EdgeDroid 1.0.

The system architecture is composed of a \emph{cloudlet}, where one or more instances of the target application run, and one or more client devices, as pictured in \cref{fig:TraceReplayArch}.
These correspond to completely unaltered instances of the target application, running inside containers~\cite{docker}.
This allows the suite to extract metrics from them while remaining transparent to the application.
Here we also deploy the central component of EdgeDroid 1.0, the \emph{control backend}.

The \emph{control backend} is implemented in Python 3.6.
Its main purpose is to act as a central point of control and configuration of the experiments, as well as to aggregate results.
It controls the execution of the \emph{application instances} and configures the experiment and client devices, while also collecting system metrics during the execution of the benchmarks.
The experiments are configured in a centralized manner to simplify scaling.
Backend and clients communicate over TCP, using a simple application-layer protocol which allows for remote configuration, transmission of the trace data and result data, and synchronization of clocks across the system.
Note that although in the current implementation of the system the backend runs co-located with the \emph{application instances} on the cloudlet, we plan on implementing functionality for distributed monitoring from a separate computer and/or across multiple cloudlets.

The \emph{client emulators} correspond to Android applications with the main purpose of emulating a real user utilizing the target cognitive assistance application.
The emulators achieve this by mimicking the operation of real clients, but instead of obtaining the input data from the on-board sensors, they extract it from the previously recorded trace.
They replay this data over the network to the \emph{application instances} employing, as mentioned in \Cref{sec:background}, a \emph{user model}, while simultaneously collecting client-side statistics and measurements.

The \emph{client emulators} are remote-controlled --- no interaction with a user is necessary once the application is initialized, simplifying large scale scenarios.

For our initial iteration of the EdgeDroid 1.0 benchmarking suite, we implement a preliminary user model, depicted in \Cref{fig:usermodel}, to run on the \emph{client emulators}, designed with a linear task model in mind.
In EdgeDroid 1.0, our model is that of a user who is totally impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
This leads to a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
In the future, we envision building upon this approach with many more human-like user models that more accurately emulate attributes such as fatigue and annoyance.

Each \emph{step} (i.e.\ segment) of the trace is played out to the backend until \emph{feedback} is received. If no feedback is received before the end of the step, it is replayed a pre-configured numbered of times before completely aborting the task (i.e.\ giving up).
Note that this model makes no distinction between positive and negative feedback, as both simply require the user to perform a specific action which is included in the trace.

\begin{figure}[tb]
    \centering
    \input{img/usermodel.tex}
    \captionof{figure}{Preliminary user model.}\label{fig:usermodel}
\end{figure}

\section{Use Cases}\label{sec:usecases}

\begin{table}[tb]
    \centering%
    \small%\ttfamily%
    \caption{Hardware used in the experiments.}%
    \label{tbl:cloudletclienthardware}%
    \begin{tabular}{@{}l@{\qquad}lclcl@{}}
        \toprule
         & \textbf{CPU}
         & \makecell{\textbf{Freq.}\\\textbf{[\si{\giga\hertz}]}}
         & \textbf{Cores}
         & \makecell[c]{\textbf{RAM}\\\textbf{[\si{\giga\byte}]}}
         & \makecell{\textbf{Operating}\\\textbf{System}}                                               \\
        \midrule
        \textbf{Cloudlet}
         & \makecell[bl]{Intel{\textregistered}\ Core{\texttrademark}\\i7--6700}
         & 3.4
         & 4
         & 32
         & \makecell[bl]{Ubuntu 17.10,\\kernel v4.13.0}                             \\
        \textbf{Clients}
         & \makecell[bl]{ARM{\textregistered}\\Cortex{\texttrademark}--A53}
         & 1.3
         & 4
         & 2
         & Android 7.0                                                             \\
        \bottomrule
    \end{tabular}%
\end{table}

In this section, we demonstrate the practical utility of EdgeDroid 1.0 through scalability measurements of real cognitive assistance application running on a cloudlet.
The questions we aim to answer relate to the ability of EdgeDroid 1.0 to provide relevant and accurate metrics on the performance of human-in-the-loop applications running on edge computing infrastructure.

Of particular interest is the ability to provide information about scaling limits in terms of latencies at both micro and macro levels.
We envision for instance a system designer performing the set of experiments detailed in this section.
The presented results would allow them to identify a bottleneck in performance.
From that they could extrapolate to how they need to scale their system hardware to manage their expected load, or they might conclude that their wireless link is not good enough for the average case.
They can also obtain real-time measurements to determine runtime measures to optimize performance, such as load balancing.
On the other hand, imagine an application developer who wishes to optimize their application.
EdgeDroid 1.0 would allow them to extract valuable information on where to focus their efforts.

%\todo[inline]{Also keep in mind that a perhaps more important question relates to on-line application optimization. Think about measures that can be applied at run-time on the side of the infrastructure to improve application performance - this is what one would like to explore here in this section.}

We chose the open source \emph{Gabriel} platform~\cite{Ha:TowardsWearableCogAssist} running the LEGO assistance application~\cite{Chen:AnEmpiricalStudyOfLatency} for our experimentation.
This application guides a user through the assembly of a 2-dimensional LEGO model with visual and auditory instructions.
This functionality can be observed online.\footnote{\url{https://www.youtube.com/watch?v=7L9U-n29abg}}
We chose it due to its maturity and stability compared to other applications which run on the \emph{Gabriel} platform, as well as the relative simplicity of the task it performs.
The application employs a straightforward, linear task model as the one described in \Cref{sec:background}.
Each state only allows for one specific correct action by the user, and any other action triggers negative feedback until the application is reset to the previous state.
The application has several different LEGO models to choose from, all of them roughly equal in complexity.
The specific task chosen for the experiments detailed in this section consists of the assembly of a 7-piece LEGO model.
The task has 7 distinct steps, and takes an average of approximately 2 minutes for a normal, untrained user to perform.
The input to the task model consist of a video stream captured either by an Android phone or a Google Glass wearable.

It should be noted that the \emph{Gabriel} platform by design always sends an acknowledgment for each input it receives, even when it is discarded, allowing us to measure latency for all inputs.

\Cref{tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
The components communicated through a single consumer-grade WiFi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
We considered exclusively off-the-shelves hardware and software.
To minimize interference during the measurements, we locked the wireless link to the least congested available channel.
%Intel\textregistered \@ TurboBoost\texttrademark \@ technology was activated on the cloudlet CPU and the frequency of all cores locked to \SI{3.7}{\giga\hertz} to ensure consistency across experiments.

To ensure statistical independence between each emulated user, the initialization of each client emulator was subject to a random delay within a predefined window of time at the beginning of each run.
This way, the time between the start of any two clients in any repetition of the experiment is stochastic, leading to independent samples from each.
We also only take into account metrics collected while \emph{all} clients were concurrently running.

Our measurements were obtained from series of scenarios, repeated 100 times each.
We performed three \emph{optimal} scenarios with 1, 5 and 10 clients each, the signal strength of the WiFi link being an excellent \SI{-40}{\dBm}.
Next, a scenario where the 10 clients were moved to another room roughly \SI{10}{\meter} away, thus degrading the network link to an average measured strength of \SI{-73}{\dBm}.
Finally, we studied an additional optimal single-client scenario focused on latency distributions and jitters within application execution path.
This scenario is used to showcase the utility of EdgeDroid 1.0 for application developers.

\section{Results}\label{sec:results}

\begin{figure}
    \centering%
    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/rtt_fb_vs_nofb}%
    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
    \label{fig:comparison:rtt}%
\end{figure}%
\begin{figure}
    \centering%
    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/box_feedback}%
    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
    \label{fig:comparison:feedback}%
\end{figure}%
\begin{figure}%
    \centering%
    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/box_taskstep}%
    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
    \label{fig:comparison:tasksteps}
\end{figure}%

%In \cref{fig:comparison:feedback,fig:comparison:rtt,fig:comparison:tasksteps} we show some of the results obtained from the use case scenarios.
The results presented in this section provide valuable insights on both the system limits as well as on the application itself.

\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.

In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.

For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.

We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.

We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
One, scaling behavior for the application seems to be linear with respect to the number of clients.
Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.

It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.

EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
Here, the downlink time sometimes almost reaches the ideal bound by itself.
A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.

Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.

\section{Conclusions \& Future Work}\label{sec:conclusions}

Benchmarking human-in-the-loop applications is hard, given their tight interaction with human users who complicate the scaling and repeatability of experiments.
In this paper, we have presented a benchmarking suite for this type of applications, called EdgeDroid 1.0, capable of cutting out the need for users in performance evaluations.
We achieve this by employing pre-recorded sensory input traces which we play over the network to the real application backend, employing a parameterized user model to react to feedback.
We demonstrate its utility through a series of use case scenarios, from which we are able to extract metrics regarding latency both in regards to the application itself and the hardware stack.
We believe the EdgeDroid 1.0 suite thus represents an important first step towards enabling inexpensive and low-complexity large-scale research on the scaling limits of this type of applications, a requirement for wide adoption of the technology.

Nonetheless, there is still future work to be done.

The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with WCA applications in order to present a more complete model in the future.
As mentioned in \Cref{sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
Rather, EdgeDroid 1.0 models a perfectly stoic user who is like an automaton and responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
Of course, no real human user is an automaton.
In the future, we envision creating  many versions of EdgeDroid (i.e., EdgeDroid 2.0, EdgeDroid 3.0, etc.) that embody more human-like user models that more accurately emulate attributes such as those mentioned above.
Experimental validation of these human-like user models via user studies will be an important part of our future work.

We are also working on expanding the benchmarking suite to also work first with other types of Wearable Cognitive Assistance, and later with other categories of human-in-the-loop applications.
Other types of WCA we will consider in future iterations of the tool include real-time task-assistance WCA applications (such as the Ping-Pong application described in~\cite{Chen:EarlyImplementation}), which don't have a linear task model like task-guidance WCA and have tighter latency bounds and context- and information-providing WCA applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
The latter also do not have a linear task model, but present more lax latency bounds.

\begin{acks}
    We thank Bobby Klatzky and Dan Siewiorek for many valuable technical discussions relating to this research.
    We also thank our shepherd, Wenjun Hu, and the anonymous reviewers for helping us improve the paper.
    This research was supported in part by the National Science Foundation (NSF), grant number CNS-1518865, the VINNOVA grant MERIT (2017--05232).
    Additional support was provided by Intel,\ Vodafone,\ Deutsche~Telekom,\ Verizon,\ Crown~Castle,\ NTT,\ and the Conklin~Kistler family fund.
    Opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or the mentioned funding sources.
\end{acks}

### 2022-08-19 publications/2019EdgeDroid/abstract.tex
rename

%Benchmarking human-in-the-loop applications is complex.
%This limits reproducibility as well as feasibility of performance evaluations.
%In this paper we present EdgeDroid, a benchmarking suite that addresses these challenges.
%Our core idea rests on recording traces of these applications \jjw{-> application traces} which are then replayed out \jjw{remove out} in a controlled fashion based on an underlying model of human behavior \jjw{-> emulating human behaviors}.
%The traces are then exposed to the original backend compute process of the respective human-in-the-loop application, generating realistic feedback.
%This allows for an automated system that greatly simplifies benchmarking large scale scenarios.
%Our results confirm the utility of EdgeDroid as a tool for system designers, application developers and researchers.

%\jjw{Alternative version:
Many emerging mobile applications, including augmented reality (AR) and wearable cognitive assistance (WCA), aim to provide seamless user interaction. 
However, the complexity of benchmarking these human-in-the-loop applications limits reproducibility and makes performance evaluation difficult. In this paper, we present EdgeDroid, a benchmarking suite designed to reproducibly evaluate these applications.

Our core idea rests on recording traces of user interaction, which are then replayed at benchmarking time in a controlled fashion based on an underlying model of human behavior. 
This allows for an automated system that greatly simplifies benchmarking large scale scenarios and stress testing the application.
Our results show the benefits of EdgeDroid as a tool for both system designers and application developers.
%}

### 2022-08-19 publications/2019EdgeDroid/img/taskmodel.tex
rename

\adjustbox{scale=0.75}{
    \ttfamily\centering%\fbox{%
    \begin{tikzpicture}[align=center,
        node distance=.5cm and 1cm,
        every initial by arrow/.style={-{Latex[length=2mm]}},
        line/.style ={draw, thick, -latex', shorten >=0pt}]
        % Place nodes       
        \node [state, minimum size=3em] (si) {\Large $s_{i}$};
        \node [state, above right=of si, minimum size=3em] (sj) {\Large $s_{j}$};
        \node [state, below right=of si, minimum size=3em] (se) {\Large $s^{e}_{k}$};

        \path[draw, -{Latex[length=2mm]}]
        (si)
        edge [out=140,in=220,looseness=6] node[left] {\emph{In:} none\\\emph{Out:} positive feedback} (si)
        edge [bend left=20] node[above left] {\emph{In:} correct action\\\emph{Out:} none} (sj)
        edge [bend left=20] node[right] {\emph{In:} incorrect action\\\emph{Out:} none} (se)

        (se)
        edge [bend left=20] node[below left] {\emph{In:} undo action\\\emph{Out:} none} (si)
        edge [out=230,in=310,looseness=6] node[below] {\emph{In:} none\\\emph{Out:} negative feedback} (se)

        (sj) edge [out=50,in=130,looseness=6] node[above] {\emph{In:} none\\\emph{Out:} positive feedback} (sj)
        ;

    \end{tikzpicture}
}%}

### 2022-08-19 publications/2019EdgeDroid/img/usermodel.tex
rename

\adjustbox{scale=0.7}{
    \ttfamily\centering%\fbox{%
    \begin{tikzpicture}[align=center,
        node distance=.5cm and 1.5cm,
        every initial by arrow/.style={-{Latex[length=2mm]}}]
        % Place nodes              
        \node [initial, state, minimum size=5em, initial text=] (play) {\textbf{Play}};
        \node [state, above right=of play, minimum size=5em] (change) {\textbf{Change}\\\textbf{step}};
        \node [state, below right=of play, minimum size=5em] (rewind) {\textbf{Rewind}};
        \node [state, accepting, above right=of rewind, minimum size=5em] (shutdown) {\textbf{Shut}\\\textbf{down}};

        % Draw edges
        \path[draw, -{Latex[length=2mm]}]
        (play) edge [out=270, in=180] node[below left] {Finished step but\\no feedback received} (rewind)
        edge [out=90, in=180] node[above left] {Received feedback} (change)
        edge [out=140, in=220,looseness=6] node[left] {Step\\not finished} (play)

        (change) edge [bend left=5] node[below right] {Step changed} (play)
        edge [out=0, in=90] node[above right] {All steps\\finished} (shutdown)

        (rewind) edge [bend right=5] node[above right] {Rewound} (play)
        edge [out=0, in=270] node[below right] {Too many\\rewinds} (shutdown);

    \end{tikzpicture}
}%}

### 2022-08-19 publications/2021ImpactDelayedResponse/Response_to_Reviewers.tex
plos

\documentclass[10pt, letterpaper]{letter}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[]{todonotes}
\usepackage[]{geometry}
\usepackage[inline]{enumitem}
\usepackage[hang]{footmisc}
\usepackage{todonotes}
\usepackage{siunitx}
\usepackage[capitalize]{cleveref}
\usepackage{multicol}
\usepackage{csquotes}

\newcommand{\todoplaceholder}{\todo[inline]{TODO}}

\makeatletter
% Default:
% \def\@makefnmark{\hbox{\@textsuperscript{\normalfont\@thefnmark}}}
\renewcommand{\@makefnmark}{\makebox{\normalfont[\@thefnmark]}}
\makeatother

% \usepackage[citestyle=numeric,
%             backend=bibtex,
%             bibstyle=authoryear,
%             sorting=none,
%             sortcites,
%             firstinits,
%             terseinits,
%             maxbibnames=6,
%             autocite=footnote,
%             dashed=false]{biblatex}
\usepackage[
    backend=bibtex,
    style=ieee,
    maxcitenames=3,
    mincitenames=1,
    % maxbibnames=3,
    % minbibnames=1,
    citestyle=numeric-comp % for [1, 2] instead of [1], [2]
]{biblatex}
\bibliography{bibliography.bib}
\defbibheading{letterbib}[\refname]{#1}

% question/answer environment
\newenvironment{QandA}
{\begin{enumerate}[
    label={\bfseries Comment \arabic*:}, 
    wide,
    ref={Comment \arabic*}]}
{\end{enumerate}}

\newenvironment{revised}[2]
{%
\begin{displayquote}
    \medskip
    \begin{flushright}
        \itshape{(Lines \numrange{#1}{#2} in the revised manuscript.)}\normalfont%
    \end{flushright}
    \medskip

}
{\end{displayquote}}

\newenvironment{answered}
{\medskip\par\bfseries Reply: \normalfont}
{\par\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}\bigskip}

\signature{
    Manuel {Olguín Muñoz}\\
    Division of Information~Science and Engineering, EECS\\
    KTH Royal Institute of Technology\\
    Malvinas väg 10, 7th floor\\
    100 44 Stockholm, Sweden\\
    \href{mailto:molguin@kth.se}{\url{molguin@kth.se}}
}
% \address{}
\begin{document}

\begin{letter}{
    M. Usman Ashraf, Ph.D\\
    Academic Editor\\
    PLOS ONE\\
    \medskip
    Manuscript PONE-D-20-32471\\
    Response to reviewers\\
}

\opening{Dear Dr.\ Usman Ashraf:}

We would like to begin by thanking you for giving us the opportunity to submit a revised version of our manuscript \emph{``Impact of delayed response on Wearable Cognitive Assistance''}.
We also wish to thank you and the reviewers for the effort you have dedicated to providing us with relevant and insightful comments to improve the manuscript.

After careful consideration, most of these recommendations have been included and highlighted in the revised paper.
In particular, we have reworked extensive segments of the Introduction --- following Reviewer 1's comments --- as well as significantly summarized subsections 3.1, 3.2 and 3.3.3 --- as per the combined suggestions of both reviewers.
Please see below for a detailed, point-by-point response to each of the comments and suggestions made by the reviewers.

With regards to the dataset associated with this research, we have chosen to make it publicly available in full, as a repository on the Zenodo platform:
\begin{itemize}
    \item DOI: \texttt{10.5281/zenodo.4494912}
    \item URL: \href{https://zenodo.org/record/4494912}{\url{https://zenodo.org/record/4494912}}
\end{itemize}

% \todo[inline]{Please ensure that your manuscript meets PLOS ONE's style requirements, including those for file naming.}

% \todo[inline]{If there are no restrictions, please upload the minimal anonymized data set necessary to replicate your study findings as either Supporting Information files or to a stable, public repository and provide us with the relevant URLs, DOIs, or accession numbers.}

Once again, we thank the reviewers for their comments and hope we have managed to sufficiently address their concerns in this revision.

\closing{Sincerely,}

\pagebreak

\textbf{Responses to reviewers' comments:}

\textit{Reviewer 1}
\begin{QandA}
    \item Please bring references for (page 2, lines 13--17):
    \begin{itemize}
        \item \emph{One is providing quality of life improvements to the millions of people around the world affected by some form of cognitive decline [\ldots]}
        \item \emph{WCA can, for instance, provide assistance to people recovering from traumatic brain injuries, smoothly guiding them through day-to-day interactions with the world which would otherwise be extremely challenging [\ldots]}
    \end{itemize}
    \begin{answered}
        We agree with the reviewer that these claims need further evidence.
        As such we have decided to rework this segment to highlight the fact that these use cases inspire the work, but they are not everyday reality yet.
        The paragraphs in question now read:

        \begin{revised}{13}{19}
            WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\autocite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
            More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\autocite{Chen:AnEmpiricalStudyOfLatency}.

            These systems have also found success as companion tools for specialists and as a means for guiding their training. 
            Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\autocite{funk2015caworkplace,gorecky2011cognito}.
            Detethering this assistance from its current fixed location will surely make it available to many more fields.
        \end{revised}
    \end{answered}

    \item \emph{Characterizing the relationships between system responsiveness and user behavior and experience is of paramount importance for the design and evaluation of these applications} (page 2, lines 54--56).

    This sentence needs some more clarifications regarding user behaviour and experience:
    bringing some examples for user behaviour and specifying the experience may help to better understand it.
    For instance, if the experience is an experience about the use of such applications then it needs to add ``use experience'' to make it clear.

    \begin{answered}
        We agree with the reviewer, and have thus modified this section of the text to make it more clear by adding some examples:

        \begin{revised}{53}{59}
            Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
            It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\autocite{dabrowsky:2011:40years}. 
            Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
            A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
        \end{revised}
    \end{answered}

    \item\label{it:questions} The research questions cannot be found easily.
    They should be explained.
    The aim of the study is mentioned (p3, lines 64--65) i.e.\ how human behavior changes with system responsiveness. 
    \emph{We aim to tackle this question…}, but the research questions need to be clarified too.

    \begin{answered}
        The reviewer is correct in noting that research questions were not explicitly mentioned in the manuscript.
        We have chosen to address this comment together with \labelcref{it:hypothesis}.
        Please refer to said comment for details.
    \end{answered}

    \item \emph{We present in this paper an experimental WCA test-bed of our design and making} (p3, line 70).
    It seems the sentence is not complete: and making (??)

    \begin{answered}
        We accept that the sentence is unclear and strange.
        In the revised manuscript it has been changed to:

        \begin{revised}{71}{72}
            We present in this paper the design and elaboration of an experimental WCA test-bed.
        \end{revised}
    \end{answered}

    \item \emph{This test-bed was subsequently employed in a study in which undergraduate students were asked to interact with and follow the instructions given to them by a cognitive assistant} (p3, lines 71--73).
    Please bring more information about undergraduate students for example in which field, gender\ldots

    \begin{answered}
        Yes, for purposes of clarity we have amended the sentence:

        \begin{revised}{72}{74}
            This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
        \end{revised}

        We also note that more detailed information about the participants is provided in Section 3.
    \end{answered}

    \item\label{it:hypothesis} Research hypotheses should be mentioned and explained in a clear way.

    \begin{answered}
        We agree with the reviewer, and have addressed this comment, together with \labelcref{it:questions}, by reformulating a significant segment of the introduction and explicitly stating the research questions along with the associated hypotheses.

        \begin{revised}{8}{121}
            Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.\\

            \begin{itemize}
            \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}\\

            In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
            The extent or form of these changes were however unknown.
            We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.\\

            Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
            While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
            To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
            Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.\\

            \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}\\

            We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.\\

            The results we present, however, seem to refute this hypothesis. 
            We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.\\

            \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}\\

            In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
            In particular, we expected emotional effects to be correlated with the strength of the added delay.\\

            The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
            This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.\\

            \item \emph{Are these effects mediated by personality indicators in any way?}\\

            Finally, we hypothesized that the individual trait of \emph{neuroticism}~\autocite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
            We also expected \emph{focus} and \emph{involvement}~\autocite{witmer1998:itq} to play roles in this.\\

            The results obtained agree with our hypothesis. 
            We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.
            \end{itemize}
        \end{revised}
    \end{answered}

    \item The introduction part needs to be more elaborated with theoretical frame.

    \begin{answered}
        Agreed.
        We believe this is partly addressed by the changes made to the introduction according to \labelcref{it:questions,it:hypothesis}.
        Additionally, we have included the following paragraph at the beginning of Section 2:

        \begin{revised}{146}{154}
            Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
            In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
            Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.\\

            Accordingly, in the following we review background work in the areas of 
            \begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
            \item[(Section 2.1)] time perception
            \item[(Section 2.2)] mechanisms relating human performance to delay.
            \end{enumerate*}
        \end{revised}
    \end{answered}

    \item As the authors wrote a \emph{third potential explanation of delay effects} (p5, line 167).
    Similar types of writing for all three explanation may help readers grasp the messages easier: thus, I suggest that the authors bring at the beginning of previous relevant phrases ``first explanation'' and ``second explanation'' too.

    \begin{answered}
        Yes.
        The explanations in question have been modified in the revised manuscript:

        \begin{revised}{191}{192}
            A first possible explanation comes from research on cognitive and motor planning. 
            Delay may move users from relatively automatic to more attention-demanding processing. 
            [\ldots]
        \end{revised}

        \begin{revised}{205}{206}
            A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
            [\ldots]
        \end{revised}

        \begin{revised}{208}{210}
            Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort.
            [\ldots]
        \end{revised}
    \end{answered}

    \item Key characteristics of the sample (Section 3) should be clarified.

    \begin{answered}
        Agreed.
        We assume here that by \emph{sample} the reviewer refers to the \emph{participant population}.
        We have amended the text to specify that our participants correspond to undergraduate volunteers enrolled in an introductory-level psychology course at CMU:\@

        \begin{revised}{240}{250}

            This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board under record number \texttt{STUDY2019\_00000247}.
            Written consent was obtained.\\

            Subjects were recruited from a pool of undergraduate students at Carnegie Mellon University.
            Students enrolled in an introductory-level psychology course fulfill a research requirement as part of the plan of study.
            This requirement may be fulfilled either through the elaboration of a written essay or by volunteering as a participant in a small number of research experiments.\\

            No particular exclusion criteria were applied, and as specified by our approved data-collection protocol, no gender- or sex-related statistics were collected.
            In total, 40 participants were recruited, all of them of college student age (\numrange{18}{25} years old).

        \end{revised}
    \end{answered}

    \item\label{it:shorten31} Sections 3.1, 3.2 need to be summarized.
    This part regarding definitions are too long and need to be shortened.

    \begin{answered}
        Yes, we agree with the reviewer, and have edited sections 3.1 and 3.2 to be more concise.
        As the changes are too extensive to fully detail here, please refer to the revised manuscript, specifically lines \numrange{250}{324}.
    \end{answered}

    \item\label{it:shorten333} Subsection 3.3.3 also need to be shortened and summarized in a clearer way.
    This part is very long too.
    For instance, EEG is well known, it doesn't need to be defined very much.

    \begin{answered}
        We agree.
        Subsection 3.3.3 has been shortened by summarizing descriptions and trimming away unnecessary details.
        It now reads as follows:

        \begin{revised}{357}{385}
            The participants wore devices to acquire four physiological measures:
            \begin{enumerate*}[itemjoin={{, }},
                            itemjoin*={{, and }},
                            label={{(\arabic*)}}]
            \item galvanic skin response (\emph{GSR})
            \item accelerometer data from the dominant wrist
            \item brain activity in the form of electroencephalography (\emph{EEG})
            \item heart rate.
            \end{enumerate*}\\

            These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\autocite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
            GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\autocite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
            EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\autocite{Antonenko2010,Grimes2008,kumar2016measurement}.\\

            Wrist acceleration, GSR and heart rate data were obtained using the Empatica E4~\autocite{empatica:e4} biosensing wristband.
            Accelerometer data was sampled at \SI{32}{\hertz},
            GSR was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} (BVP) signal sampled at \SI{64}{\hertz}.
            Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach a stable equilibrium and establish a baseline for the signals.\\

            The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\autocite{ragot2017emotion, mccarthy2016validation}.
            The E4 also includes a skin temperature thermometer, however the measure was not used for the present study.\\

            For the EEG data we employed the OpenBCI EEG Headband Kit\autocite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
            It provides a quick and non-invasive way of obtaining EEG signals from participants.
            Electrodes were placed according to the \emph{10--20 Electrode System}~\autocite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
            The signals were sampled at \SI{200}{\hertz} and postprocessed to 
            \begin{enumerate*}[itemjoin={{, }},
                            itemjoin*={{, and }},
                            label={{(\arabic*)}}]
            \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
            \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
            \end{enumerate*}
        \end{revised}
    \end{answered}

    \item\ [The Discussion] part is very short in comparison to other parts of this study.
    The reflection of the authors of the current paper on their results is missing (the reflection of authors regarding why some results were significant while other not are important).
    Thus the authors should elaborate the Discussion part as well.

    \begin{answered}
        We partially agree with the reviewer.
        We believe that  this concern of the reviewer was addressed in the original discussion section (paragraph beginning with line 597 in the old manuscript), but apparently not with sufficient clarity.
        We pointed out that the failure of the physiological measures to show effects of delay indicates that resource depletion and emotionally induced arousal  are not impacted.
        We concluded, ``Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.''

        In response to the review, we have now expanded this paragraph to specifically point to the statistical outcomes.
        That is, we now say:

        \begin{revised}{596}{602}
            However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
            Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
            The acceleration data further did not indicate that extended delay significantly increases erratic movement.
        \end{revised}
    \end{answered}

\end{QandA}

\textit{Reviewer 2}
\begin{QandA}
    \item The explanations have been used too much in somewhere.
    For example, it is not necessary to explain the ANOVA test.
    This test is well known.

    \begin{answered}
        We agree with the reviewer.
        The explanation of the ANOVA test has been removed, and the paragraph has been reworked with a new reference:

        \begin{revised}{411}{413}
            We confirmed this effect through an \emph{analysis of variance} (ANOVA~\autocite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in Table 2.
        \end{revised}

        We also understand this remark as relating to the same issues pointed out by \emph{Reviewer 1} in \labelcref{it:shorten31} and \labelcref{it:shorten333}.
        As such, we refer to our responses to these comments.
    \end{answered}

    \item Use more resolution for figures.

    \begin{answered}
        Yes, this has been addressed in the revised manuscript.
    \end{answered}

    \item Some figures can be inserted in a figure as subplots and or subpanels.

    \begin{answered}
        Agreed.
        This has been addressed in the revised manuscript.
        In particular, figure 4 has been reorganized and merged with figure 5 (in the old manuscript).
    \end{answered}

    \item In line 286 at the end of line the ``A Latin'' is written, what is that?

    \begin{answered}
        This is an artifact of the line-wrapping of the document.
        The segment of text actually reads ``A \emph{Latin~square}-type design [\ldots]'', but \LaTeX decided to automatically wrap the line right after \emph{Latin}.
        This has been fixed in the revised manuscript.
    \end{answered}

\end{QandA}

\pagebreak

\begingroup\setlength{\multicolsep}{0pt}
\begin{multicols}{2}[]
    \printbibliography[heading=letterbib]
\end{multicols}\endgroup
\end{letter}
\end{document}

### 2022-08-19 publications/2021ImpactDelayedResponse/cover_letter.tex
plos

\documentclass[10pt, letterpaper]{letter}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[]{todonotes}
\usepackage[]{geometry}
\usepackage[inline]{enumitem}
\signature{
    Manuel {Olguín Muñoz}\\
    Division of Information~Science and Engineering, EECS\\
    KTH Royal Institute of Technology\\
    Malvinas väg 10, 7th floor\\
    100 44 Stockholm, Sweden\\
    \href{mailto:molguin@kth.se}{\url{molguin@kth.se}}
}
% \address{}
\begin{document}

\begin{letter}{
    Joerg Heber, PhD\\
    Editor-in-Chief\\
    PLOS ONE
}
\opening{Dear Dr.\ Heber:}

Please find enclosed a research article submission to PLOS ONE entitled \emph{``Impact of delayed response on Wearable Cognitive Assistance''}.
This manuscript presents a study on the physiological and behavioral reactions of users of Wearable Cognitive Assistance (WCA) to delays in the application pipeline.
This question was approached with real-time measures in a controlled experiment, using an instrumented WCA testbed.
We found that delay appears to affect the cognitive plan of users, preventing them from automating the task, and that the strength of this effect seems to be modulated by individual differences in users.
We believe these results provide key insights for developers and system designers working on cost-efficient strategies for resource allocation in these systems, as well as both new and existing synthetic models of human behavior for WCA.\@

The paper connects the current research to previously published work in two broad areas:  human time perception and response to system delays in human-computer interaction.
Our review establishes that time perception is subject to contextual influences and that self-paced behavior is not regulated by an internal clock.
Our review of delays in human-computer interaction describes consequences known in the literature, like attention lapses and frustration, and further uses prior research to propose cognitive and emotional mechanisms that might mediate these effects.

We do not have any opposed reviewers, and we suggest the following Academic Editors to handle this submission (in order of preference)
\begin{enumerate*}[label={(\arabic*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item Borsci, Simone
    \item Lv, Zhihan
    \item Ostadabbas, Sarah 
\end{enumerate*}

I will be serving as the corresponding author for the manuscript, assuming all the responsibilities this entails. 
All of the authors listed have agreed to the byline order and to submission of the manuscript, and understand that, if accepted for publication, a certification of authorship form will be required that will be signed by all of us.

\closing{Sincerely,}

\end{letter}
\end{document}

### 2022-08-19 publications/2021ImpactDelayedResponse/paper.tex
plos

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
\usepackage{tabularx}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% % Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% % Remove brackets from numbering in List of References
% \makeatletter
% \renewcommand{\@biblabel}[1]{\quad#1.}
% \makeatother

% BibLatex and PLOS One hacks
\usepackage[citestyle=numeric,
            backend=bibtex,
            bibstyle=authoryear,
            sorting=none,
            sortcites,
            firstinits,
            terseinits,
            maxbibnames=6,
            dashed=false]{biblatex}
\bibliography{bibliography}
\usepackage[colorlinks]{hyperref}

% Name list format
\renewcommand*{\labelnamepunct}{\addspace}
\renewcommand*{\finalnamedelim}{%
  \ifbibliography{\addcomma\space}{\addspace\&\space}}
\renewcommand*{\revsdnamepunct}{}
\DeclareNameAlias{sortname}{last-first}

% No quotes or italics in titles
\DeclareFieldFormat{title}{#1}
\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
                    thesis,unpublished]{title}{#1}
\DeclareFieldFormat{journaltitle}{#1}
\DeclareFieldFormat{issuetitle}{#1}
\DeclareFieldFormat{maintitle}{#1}

\renewbibmacro*{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

% Remove page prefixes
\DeclareFieldFormat{pages}{#1}

% Only print journal volume
\renewbibmacro*{volume+number+eid}{%
  \setunit*{\addspace}%
  \printfield{volume}}

% Use colon for volume-pages delimiter
\renewcommand*{\bibpagespunct}{%
  \ifentrytype{article}{\addcolon\space}{\addcomma\space}}

% Add labelnumbers to bibliography
\DeclareFieldFormat{labelnumberwidth}{#1\adddot}
\defbibenvironment{bibliography}
  {\list
     {\printtext[labelnumberwidth]{%
      \printfield{prefixnumber}%
      \printfield{labelnumber}}}
     {\setlength{\labelwidth}{\labelnumberwidth}%
      \setlength{\leftmargin}{\labelwidth}%
      \setlength{\labelsep}{\biblabelsep}%
      \addtolength{\leftmargin}{\labelsep}%
      \setlength{\itemsep}{\bibitemsep}%
      \setlength{\parsep}{\bibparsep}}%
      \renewcommand*{\makelabel}[1]{\hss##1}}
  {\endlist}
  {\item}

% Omit authoryear disambiguation
\AtEveryBibitem{\clearfield{extrayear}}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

% other packages
\usepackage[capitalize, nameinlink]{cleveref}
\usepackage[inline]{enumitem}
\usepackage[binary-units, range-units=single, range-phrase=~to~, list-units=single]{siunitx}
\sisetup{
    detect-all,
    detect-inline-family=math,
    detect-inline-weight=math,
    detect-display-math=true,
    list-final-separator = {, and }
}
\usepackage{subcaption}
\usepackage{amsthm}
\usepackage{thmtools}

\declaretheoremstyle[
notefont=\bfseries, notebraces={}{},
bodyfont=\normalfont,
headformat=\NAME~\NUMBER~---~\NOTE%
]{nopar}
\declaretheorem[style=nopar]{definition}

\usepackage[export]{adjustbox}
\usepackage{multirow}
\usepackage{booktabs}

\usepackage{pgf-umlsd}
\usepackage{tikz}
\usepackage{tikzpeople}
\usetikzlibrary{shapes,arrows.meta,positioning,automata,matrix,shapes.symbols,trees}
\usepackage{tikz-qtree}
\usepackage[]{todonotes}
\reversemarginpar  % todonotes on left

\newenvironment{acks}{
  \section*{Acknowledgements}
}{}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Impact of delayed response on Wearable Cognitive Assistance} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Manuel~{Olguín~Muñoz}\textsuperscript{1, *},
Roberta~Klatzky\textsuperscript{2, 3},
Junjue~Wang\textsuperscript{4},
Padmanabhan~Pillai\textsuperscript{5},
Mahadev~Satyanarayanan\textsuperscript{4},
James~Gross\textsuperscript{1}
\\
\bigskip
\begin{tabularx}{\textwidth}{lX}
  \textbf{1} & Division of Information~Science and Engineering, School of Electrical~Engineering and Computer~Science, KTH Royal Institute~of~Technology, Stockholm, Sweden\\
  \textbf{2} & Department of Psychology, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
  \textbf{3} & Human-Computer Interaction Institute, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
  \textbf{4} & School of Computer Science, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
  \textbf{5} & Intel Labs, Pittsburgh, PA, USA\\
\end{tabularx}
\bigskip

\begin{tabularx}{\textwidth}{lX}
  % \Yinyang & These authors contributed equally to this work.\\
  % \ddag & Senior author.\\
  * & Corresponding author: \href{mailto:molguin@kth.se}{\texttt{molguin@kth.se}}\\
\end{tabularx}

\end{flushleft}
% Please keep the abstract below 300 words

\section*{Abstract}

Wearable Cognitive Assistants (WCA) are anticipated to become a widely-used application class, in conjunction with emerging network infrastructures like 5G that incorporate edge computing capabilities.
While prototypical studies of such applications exist today, the relationship between infrastructure service provisioning and its implication for WCA usability is largely unexplored despite the relevance that these applications have for future networks.
This paper presents an experimental study assessing how WCA users react to varying end-to-end delays induced by the application pipeline or infrastructure.
Participants interacted directly with an instrumented task-guidance WCA as delays were introduced into the system in a controllable fashion.
System and task state were tracked in real time, and biometric data from wearable sensors on the participants were recorded.
Our results show that periods of extended system delay cause users to correspondingly (and substantially) slow down in their guided task execution, an effect that persists for a time after the system returns to a more responsive state.
Furthermore, the slow-down in task execution is correlated with a personality trait, neuroticism, associated with intolerance for time delays.
We show that our results implicate impaired cognitive planning, as contrasted with resource depletion or emotional arousal, as the reason for slowed user task executions under system delay.
The findings have several implications for the design and operation of WCA applications as well as computational and communication infrastructure, and additionally for the development of performance analysis tools for WCA.\@

\linenumbers%
\section{Introduction}

Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\autocite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
An example is an \textcite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.

WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\autocite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\autocite{Chen:AnEmpiricalStudyOfLatency}.
Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\autocite{funk2015caworkplace,gorecky2011cognito}.
Detethering this assistance from its current fixed location will surely make it available to many more fields.

Based on these use cases, we identify three main requirements for WCA:\@
\begin{enumerate}
    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.

    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.

    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
    This means that users will have expectations of constant, immediate feedback as they progress through the task.
    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\autocite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
\end{enumerate}

\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.

% However, additional challenges remain in the path toward widespread adoption.
\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\autocite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.

This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
We expand upon the findings of \textcite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of \textcite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
This is an important step toward a more systematic understanding of human behavior in this domain.

We present in this paper the design and elaboration of an experimental WCA test-bed.
This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\autocite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@

Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.

\begin{itemize}
  \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}

  In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
  The extent or form of these changes were however unknown.
  We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.

  Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
  While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
  To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
  Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.

  \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}

  We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.

  The results we present, however, seem to refute this hypothesis. 
  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.

  \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}

  In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
  In particular, we expected emotional effects to be correlated with the strength of the added delay.

  The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
  This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.

  \item \emph{Are these effects mediated by personality indicators in any way?}

  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\autocite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
  We also expected \emph{focus} and \emph{involvement}~\autocite{witmer1998:itq} to play roles in this.

  The results obtained agree with our hypothesis. 
  We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.

\end{itemize}

% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\autocite{john1999:bfi}.

We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and direct implications for resource and power consumption.
Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished --- this has potential consequences for resource allocation strategies.
Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource allocation across applications potentially looks different from what could be considered ``fair''.

Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \textcite{olguin:2018, olguin:2019}.
Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@

The structure of this paper is as follows.
We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
Then, in \cref{sec:results} we detail the results of our experiment.
Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.

\section{Background and Related Work}\label{sec:background}

% Overarching theo construct: people perform temp seq task in the framework of cognition and emotion and individual differences mediate response to delay. In particular we view cognition as described by models such as ACT-R as a seq of procedures.

% Section 2.
% Accordingly we review time perception...

% Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
% Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
% In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.

Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.

Accordingly, in the following we review background work in the areas of 
\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item[(\cref{ssec:timeperception})] time perception
  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
\end{enumerate*}

\subsection{Time perception in computing systems}\label{ssec:timeperception}

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\autocite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\autocite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\autocite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\autocite{matthews2011stimulus}
    \item arousal can expand durations~\autocite{droit2011emotion}.
\end{enumerate*}

Expectations play a critical role in time perception as well~\autocite{zakay1995attentional,zakay1996role}.
It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\autocite{Loewenstein1992anomaliesintertemporalchoice}.
Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
Violations of expectancy have a strong impact on the acceptability of computer systems.
Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
In immersive systems like WCA, which aim to provide seamless interaction, delays are particularly noticeable.

It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\autocite{dabrowsky:2011:40years}.
However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\autocite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.

For WCA the acceptable range for latencies was explored by \textcite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
They then proposed a latency tolerance zone according to the task demands.
For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
Beyond that, users could begin to show the negative outcomes previously catalogued~\autocite{dabrowsky:2011:40years}.

\subsection{Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}

While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood. 
These mechanisms could be cognitive or emotional in origin.

A first possible explanation comes from research on cognitive and motor planning. 
Delay may move users from relatively automatic to more attention-demanding processing.
Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
As competency in a task increases, execution of the hierarchy becomes increasingly automated.
Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\autocite{neves1981knowledge}.
Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\autocite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\autocite{lewis2003distinct,koch2009neural,lee2019limiting}.

Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
Delays, in contrast, would disrupt such a plan through the loss of automated control~\autocite{lee2019limiting}.

A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\autocite{lee2019limiting}.

Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\autocite{baumeister74tice,lin2020strong}.

The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
Arousal can also reduce movement smoothness or add erratic gestures~\autocite{pijpers2003anxiety}.
Ego depletion has been found to produce premature responses culminating in error~\autocite{lin2020strong}, or to lead to abandoning a task entirely~\autocite{baumeister74tice}.

Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality. 
Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
\textcite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting. Extraversion could also  be  a moderating factor among the broader target audience of WCA, which are intended for relatively inexperienced users of an application.
These and other measures of individual variation were considered here.

\section{Experimental Design}\label{sec:experimentaldesign}

The core elements of our experiment are shown in \cref{fig:experimentaltestbed}:
\begin{itemize}
    \item Subjects interact with a WCA while wearing an array of biometric sensors.
    \item The \emph{responsiveness} of the application, i.e.\ the interval of time between an input being provided to the system and the associated output returned to the user, is manipulated in real time. 
    The effects of these manipulations on the subjects are recorded and subsequently analyzed.    
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{images/experiment_arch.pdf}
  \caption{Experimental test-bed.
  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
  \label{fig:experimentaltestbed}
\end{figure}

This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board.

Subjects were recruited from a pool of undergraduate students at Carnegie Mellon University.
Students enrolled in an introductory-level psychology course fulfill a research requirement as part of the plan of study.
This requirement may be fulfilled either through the elaboration of a written essay or by volunteering as a participant in a small number of research experiments.

No particular exclusion criteria were applied, and as specified by our approved data-collection protocol, no gender- or sex-related statistics were collected.
In total, 40 participants were recruited, all of them of college student age (\numrange{18}{25} years old).

\subsection{The Cognitive Assistance Application}

We used a modified version of the LEGO Assistant application introduced by \textcite{Chen:EarlyImplementation}.
This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
% The results of the user performing these instructions are provided as inputs to the system.
% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.

% In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
In the following, we will provide some formal definitions relating to the WCA application.
% \theoremstyle{definition}
% \begin{definition}[Task and subtask]
A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{fig:task:steps}).
% We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
At that point, the assistant either provides a new instruction to the user, or the task ends.
  % See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
% \end{definition}

\begin{figure*}[h]
  \centering
  \includegraphics[width=.8\textwidth]{images/step_task_description.pdf}
  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
\end{figure*}

% \begin{definition}[Step]

% While the subtask remains unfinished, the assistant discards results of the processing of the sampled inputs.
% At the beginning of a step, an instruction is given to the user.
% The user then proceeds to perform the subtask specified by the instruction, while the cognitive assistant is continuously sampling the subtask state at specified intervals.

% Once the user finishes the required action, the next sample taken will contain a valid input, and thus the cognitive assistant will provide a new instruction. 
% This finishes the current step and potentially begins a new one if there remain instructions to be performed in the task.
% \end{definition}

In the base LEGO Assistant, the task consists of the assembly of a LEGO model; each step requires the user to append a LEGO brick to the model at a specific location and orientation.
The system monitors progress through a video feed and provides timely feedback in the form of visual and textual instructions to guide the user.

The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
In particular
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
    \item it is easily understood by users and requires essentially no training
    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
\end{enumerate*}

\begin{figure*}[h]
  \centering
  % \input{plots/lego_hierarchy_diagram.tex}
  \includegraphics[width=\textwidth]{plots/lego_hierarchy_diagram_standalone.eps}
  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
  \label{fig:lego:hierarchical}
\end{figure*}

% The original design of the LEGO assistance application was based on a client-server model communicating over a wireless network, with the client software running on a wearable device and the server software deployed on a cloudlet. 
For the purposes of this study, the original client-server design of the LEGO Assistant was altered to be locally executable.
This was done in order to eliminate the stochastic effects of jitter and latency from the network link, and greatly simplify instrumentation.
% By this means, we achieve much more fine-grained control over the latencies to which the system is subject.
% Additionally, this greatly simplified the instrumentation of the application.
Instructions were output in image and text form to a computer display situated on a table directly in front of the participants.
Participants performed the instructions on the table, these actions being captured by a high-definition camera located on top of the display.

% Finally, a data collection and experimental layer was implemented between the user interface and video capture and the core processing component of the LEGO Assistant.
% This layer controlled the manipulated experimental variables and recorded the measured biometric and task-related effects.

\subsection{The Experimental LEGO Task}

For the experimental LEGO task, a key modification was made to the structure of the steps.
After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
The length of this delay is one of two independent variables we manipulated for the experiment.
Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by \textcite{Chen:AnEmpiricalStudyOfLatency}.
% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.

In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
% \begin{definition}[Block]
%   A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
We used values of \numlist{4;8;12} steps, chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block.
% ; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
% \end{definition}

To assign tasks to the participants, a pseudo-random permutation of the combinations of block length and delay was first generated, and a unique sequence of steps assigned to each of these combinations to create 21 unique blocks.
A \emph{Latin~square}-type design was then used to reorder this initial permutation in order to generate a task for each participant.
This ensures a counterbalance of the order of the blocks across participants and avoids systematic learning effects. 
% The design rotates the block types (as defined by length and delay) across participants so that each type is tested in each ordinal position, but it only coarsely samples from the \( 21 \times 20 \) possible sequences from one block type to another.
Note that unlike the base LEGO assistant task, in which instructions led participants through the assembly of a specific model, the experimental LEGO task consisted of a sequence of instructions to either add or remove pieces with no evident goal.
% Users were directed to either add a piece to the ongoing model or to remove a piece, and blocks were designed in such as way so as to ensure that the transitions between them were invisible to the user.

% In this paper we will thus consider blocks to be our basic element of study, and most aggregations will be done at this level (with a few exceptions).
% For this, we will need additional definitions:

% \begin{definition}[Delay associated with a block]
%   We will refer to the delay of a block as the delay applied to every frame of every step in that block.
% \end{definition}

% \begin{definition}[Execution time]

Given the variability in the system latency to detect step completion, correcting step-completion time for system response time by a fixed amount would not enable sufficient precision.
We therefore define the \emph{execution time} empirically for an individual step as the total time between the user receiving the instruction for the step and the time instant of the sample capturing the completed step.
% their presenting the completed subtask to the system.

% We define the \emph{execution time} of a step as the time between the user receiving an instruction for a subtask and them presenting the completed subtask to the system.
% This, however, could not be measured directly and has to be inferred from the records of inputs processed by the system.

For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{fig:exectime:diagram}).
\( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
If we define \( t_{c} \) as the instant marking when the user presented the finished step to the system, it must follow that \( t_{n-1} < t_{c} < t_{n} \).
Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
However, this introduces no bias in the results, as we can assume that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
% This is also aggregated into an average at the block level.

% \begin{figure}
%   \centering
%   \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
%   \caption{Visualization of the execution time of a step.}%
%   \label{fig:exectime:diagram}
% \end{figure}

% \end{definition}

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/step_structure.pdf}
    \caption{Structure of a step in a generic cognitive assistance application. 
    The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
    % by the assistant, as they do not contain relevant information.
    Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
    % which will subsequently be provided to the user.}%
    }
    \label{fig:cogassist:step}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/step_structure_delay.pdf}
    \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
    \label{fig:cogassist:step:delay}
  \end{subfigure}
  \medskip%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/block.pdf}
    \caption{Structure of a block in the experimental task.}%
    \label{fig:cogassist:block}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{images/exec_time_diagram.pdf}
    \caption{Visualization of the execution time of a step.}%
    \label{fig:exectime:diagram}
  \end{subfigure}%
  % \medskip%
  \caption{Components of the cognitive assistance task.}
\end{figure}

\subsection{Collected Data}

The collected data from the experiments fall into four categories: behavioral and personality indicators, \emph{frame-to-frame} metrics, biometric data, and video recordings.

\subsubsection{Behavioral and Personality Indicators}

Before beginning the experimental procedure, participants were asked to fill out two questionnaires.

The first of these, the Big Five Inventory of Personality (\emph{BFI}~\autocite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
Of these, extroversion and neuroticism have been related to tolerance for delay~\autocite{hirsh2008delay}.

The second survey, the Immersive Tendencies Questionnaire (\emph{ITQ}~\autocite{witmer1998:itq}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
These questions use a 7-point horizontal scale  with opposing descriptors anchoring  the ends.
Participants were asked to mark the appropriate point in the scale, and these responses were converted to a numerical value between 1 and 7 for processing.

In post-processing, the obtained scores for both questionnaires were normalized to fall in the \( [0, 1] \) range for ease of interpretation. 
See \cref{tab:qscores} for their means and standard deviations.

\begin{table}[h]
  \centering
  \caption{Means and standard deviations of normalized questionnaire scores.}\label{tab:qscores}
  \setlength{\tabcolsep}{0pt} 
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}clrr@{}}
    \toprule
    \textbf{Questionnaire} & \textbf{Metric} & \textbf{Mean} & \textbf{SD}\\
    \midrule
    \multirow{5}{*}{BFI} 
        & Agreeableness &               0.705 &                  0.136 \\
        & Conscientiousness &               0.562 &                  0.149 \\
        & Extroversion &               0.491 &                  0.180 \\
        & Neuroticism &               0.524 &                  0.230 \\
        & Openness &               0.677 &                  0.152 \\
    \cline{1-4}
    \multirow{4}{*}{ITQ} 
        & Focus &               0.626 &                  0.126 \\
        & Games &               0.463 &                  0.261 \\
        & Involvement &               0.568 &                  0.175 \\
        & Total &               0.540 &                  0.092 \\
    \bottomrule
  \end{tabular*}
\end{table}

\subsubsection{Frame-to-frame metrics}
During the execution of the task, we logged every event occurring in the application pipeline.
Each incoming frame (including frames discarded by the assistant), as well as its associated outputs, was logged at multiple points in the process along with associated metadata such as currently implemented delay as specified by the experimental design.
This allowed us to extract metrics relating to the performance of the task, such as the time spent by participants on particular steps, any mistakes made, etc.
In particular, it made possible the easy segmentation of the other time-series data we collected into our main unit of analysis, the aforementioned \emph{block}.

\subsubsection{Biometric Data}
The participant wore devices to acquire four physiological measures:
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item galvanic skin response (\emph{GSR})
  \item accelerometer data from the dominant wrist
  \item brain activity in the form of electroencephalography (\emph{EEG})
  \item heart rate.
\end{enumerate*}

These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\autocite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\autocite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\autocite{Antonenko2010,Grimes2008,kumar2016measurement}.

Wrist acceleration, GSR and heart rate data were obtained using the \textcite{empatica:e4} biosensing wristband.
Accelerometer data was sampled at \SI{32}{\hertz},
GSR was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} (BVP) signal sampled at \SI{64}{\hertz}.
Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach a stable equilibrium and establish a baseline for the signals.

The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\autocite{ragot2017emotion, mccarthy2016validation}.
The E4 also includes a skin temperature thermometer, however the measure was not used for the present study.

For the EEG data we employed the \textcite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
It provides a quick and non-invasive way of obtaining EEG signals from participants.
Electrodes were placed according to the \emph{10--20 Electrode System}~\autocite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
The signals were sampled at \SI{200}{\hertz} and postprocessed to 
\begin{enumerate*}[itemjoin={{, }},
                  itemjoin*={{, and }},
                  label={{(\arabic*)}}]
  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
  \item smooth out noise (using the technique proposed by~\textcite{agarwal2017eeg}).
\end{enumerate*}

\subsubsection{Video recordings}

During the task, participants were recorded by two separate cameras.
One camera was angled downwards, towards the table, the LEGO board, and the participant's hands.
This camera was used to capture the necessary inputs for the LEGO assembly task as well as to record the actions performed by the user.
The second camera was angled horizontally, towards the participant, in order to record their facial expressions during the execution of the task.

Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
Examples of the captured frames can be seen in \cref{fig:sampleframes}.
The video feeds were not used for the present study, but may be utilized in future analysis.

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.35\textwidth}
    \centering
    \includegraphics[height=140pt]{images/participant.eps}
    \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
  \end{subfigure}%
  % \par\bigskip
  % \hspace{.04\textwidth}%
  \hfill%
  \begin{subfigure}[t]{.60\textwidth}
    \centering
    \includegraphics[height=140pt]{images/board_and_hand.eps}
    \caption{Frame from the board recording of the same participant.}
  \end{subfigure}
  \caption{Sample frames captured during the experiment.}\label{fig:sampleframes}
\end{figure}

\section{Results}\label{sec:results}

The results consist of execution time per step. and the outcome of physiological variables: heart rate, GSR, and EEG.\@
Each will be discussed in turn.
%The temperature data could not be analyzed due to a progressive increase in the sensor data over the course of the session, which obscured effects of the experimental variables.

\subsection{Execution Time}

% Before describing the analysis and results relating to execution time, it should be noted that accuracy for this metric was essentially perfect.
% No mistakes were made by participants during the execution of the task and no major outliers could be identified in the samples.

Before describing the analysis and results relating to execution time, it should be noted that participants' performance during the execution of the task was error free; all steps were completed as instructed.

\cref{fig:exectime} shows the mean per-step execution time per block, averaged over block length (number of steps) and artificial delay.
We can clearly see a trend for the execution time to increase with the delay, increasingly so for longer blocks. 
Since the per-step execution time  compensates for the added delay in the measure \emph{per se}, this trend must result from the participants' behavioral adjustment to the delay.
This leads to one of the key outcomes from this study: 
\emph{participants tend to act more slowly on steps affected by longer delays} --- i.e.\ there is evidence of a pacing effect in users' behavior with respect to the responsiveness of the system.

We confirmed this effect through an \emph{analysis of variance} (ANOVA~\autocite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
% An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
% The degrees of freedom represent the number of observations going into each of these variability measures.
% As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
% The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
% Conventionally \( p < .05 \) is the criterion for significance.
% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=.8\textwidth]{plots/exec_time_delay_length.eps}
%     \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
% \end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{plots/exec_time_delay_length.eps}
    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
\end{figure}

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{tab:anova:exectime}
  \setlength{\tabcolsep}{0pt}%
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    length         &    \(F(2, 78) = 9.59\) &   \(< 0.001\) &           \(0.20\) \\
    delay          &  \(F(6, 234) = 15.52\) &  \(< 0.0001\) &           \(0.28\) \\
    length \(\times\) delay &  \(F(12, 468) = 3.84\) &  \(< 0.0001\) &           \(0.09\) \\
    \bottomrule
  \end{tabular*}
\end{table}

Further analysis focused on the progressive effect on per-step execution time as additional steps occurred at a constant delay.
For this purpose, the steps within a block were aggregated over sequences of 4, constituting a \emph{slice}.
Note that the first slice within a block is procedurally identical for all block lengths, in the sense that a participant currently performing a step in the first slice of a block has no way of predicting if the block ends after step 4 or not.
The same logic can be applied to slice 2 for blocks of length 8 and 12.
Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{tab:anova:exectime:slice}.
As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
% We refer to the measure at slice 3 as a percentage of the measure at slice 1 as the ``change factor''.
With the longest delays, where the execution time per step was longest, the slow-down persisted; that is, the system response time hindered the participant's execution throughout the course of the block.

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{tab:anova:exectime:slice}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    slice         &   \( F(2, 78) = 88.79 \) &  \( < 0.0001 \) &         \( 0.69 \) \\
    delay         &  \( F(6, 234) = 14.13 \) &  \( < 0.0001 \) &         \( 0.27 \) \\
    slice \( \times \) delay &  \( F(12, 468) = 2.49 \) &    \( < 0.01 \) &         \( 0.06 \) \\
    \bottomrule
  \end{tabular*}%
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{plots/exec_time_delay_slice.eps}
    \caption{Mean per-step execution time vs.\ delay, by step slice.
    Error bars indicate S.E.M.}
    \label{fig:exectime:delay:slice}
\end{figure}

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{enumerate*}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{plots/exec_time_delay_change.eps}
  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
\end{figure}

In summary, we observe two direct effects on execution time due to added delays.
Firstly, we see a clear hampering of the improvement of execution time across steps that is otherwise evident across blocks.
Secondly, we notice that this effect lingers on even after the delay is removed, affecting subsequent blocks in the task.

\subsection{Acceleration}

Acceleration data from the E4 wristband were taken in 3 axes defined relative to the device.
As our interest was in the amount of movement rather than the direction in space, we calculated a ``movement score'' for each block which we defined as:
\begin{equation}
    M_B = \frac{ \sum_{A_B} |\: \overrightarrow{\alpha_j} \:| }{\Delta T_{B}}
\end{equation}

where \( B \) is an arbitrary block, and \(A_{B} = \{ \overrightarrow{\alpha_0}, \overrightarrow{\alpha_1}, \ldots, \overrightarrow{\alpha_k} \} \) represents the set of acceleration vector samples collected for said block.
This score would include the time imposed by the delay and the time to decode the instructions and respond by moving the pieces so that the next state was recognized.
It is only during the last part of the step that explicit movement is required, so any additional accelerations would derive from arbitrary movements while processing and waiting.
We normalized the sum of the accelerations by the duration of the step to correct for differences in delay and the execution time, which tends to increase with delay as described above.
An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{tab:anova:acc} and \cref{fig:acc:delaylength}.
There was also a significant delay by length interaction, reflecting that delay effects occurred particularly for the longer lengths.

\begin{table}[h]
  \centering
  \caption{Significant effects on accelerometer data from ANOVA on factors delay and block length.}\label{tab:anova:acc}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\(\eta^{2}_{p}\)} \\ 
    \midrule
    delay & \(F(6, 234) = 18.56\) & \(<0.001\) & \(0.32\) \\
    length \( \times \) delay & \(F(12, 468) = 3.04\) & \(<0.001\) & \(0.07\) \\ 
    \bottomrule
  \end{tabular*}%
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{plots/acc_vs_delay_per_length.eps}
    \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{fig:acc:delaylength}
\end{figure}

We next conducted the same analysis as for execution time, dividing blocks into three ``slices'' of four steps each, such that slice 1 comprised the first four steps of all block lengths, slice 2 the second four steps of blocks of length 8 and 12, and slice 3 the last four steps of blocks 12 steps long.
As shown in \cref{fig:acc:delayslice} the effect of delay was attributable only to the later slices, yielding main effect of slice number and delay and an interaction (\cref{tab:anova:acc:slice}).

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{plots/acc_vs_delay_per_slice.eps}
  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
\end{figure}

\begin{table}[h]
  \centering
  \caption{Significant effects on accelerometer data from ANOVA on factors delay and slice number.}\label{tab:anova:acc:slice}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\(\eta^{2}_{p}\)} \\
    \midrule 
    slice & \(F(2, 78) = 8.39\) & \(<0.01\) & \(0.18\) \\
    delay & \(F(6, 234) = 26.92\) & \(<0.001\) & \(0.41\) \\
    delay \(\times\) slice & \(F(12, 468) = 4.08\) & \(<0.001\) & \(0.1\) \\
    \bottomrule
  \end{tabular*}%
\end{table}

These results present further evidence of the aforementioned pacing effect.
As a sequence of steps with a long delay unfolds, more of the step duration is spent without moving.
This can be interpreted in the context of the increased execution time at long delays demonstrated in the previous analysis.
Assuming that adding or deleting a LEGO block takes essentially the same amount of time and accelerates the wrist similarly at any one step, it appears that the participant simply remains stationary during the extra time that is induced by a series of long delays.
Accordingly, the acceleration per unit time, our movement score, is reduced.

\subsection{EEG}
Analyses were conducted on the log EEG power in the alpha band, beta band, and total of all bands measured.
Readings from the two frontally placed poles were highly correlated and were pooled into an average.
Logs were taken for analysis because the EEG distribution tended to have a rightward tail.
Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 s.d.s.\ from the mean of all participants).

The analyses then comprised 28 participants.\@
Omnibus ANOVAs were conducted with delay and block length (number of steps) as factors, on the EEG data from the alpha and beta band.
The analysis of alpha EEG found no significant effects.
Beta EEG showed only an effect of block length, \( F(2,54) = 3.56 \), \( p = .035 \), \( \eta^{2}_{p} = .12 \), reflecting a tendency for the 4-step length to produce lower log power (mean 3.9 vs. 4.1 for lengths of 8 and 12).
However, this effect was small and not consistent across delays.

Again the analysis dividing blocks into 4-step slices was conducted, with delay and slice number as factors.  
For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{tab:anova:egg:alphabeta}.
Both bands showed the same tendency:  EEG declined as a sequence of steps with the same delay progressed.
These effects are shown in \cref{fig:logeegpow:slices}.
Thus, EEG taken from frontal locations mimics the execution time data in showing a decline over the course of a block, but unlike the execution time, there was no tendency for the decline in EEG to be reduced at longer delays.

\begin{table}[h]
  \centering
  \caption{Significant effects on log EEG power from ANOVA on factors delay and block slice.}\label{tab:anova:egg:alphabeta}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}llrrr}
    \toprule
    \textbf{Band} & \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\( \eta^{2}_{p} \)} \\ \midrule
    alpha   & slice & \( F(2, 78) = 28.26 \) & \( <0.001 \) & \( 0.51 \) \\ %\midrule
    beta    & slice & \( F(2, 78) = 29.18 \) & \( <0.001 \) & \( 0.52 \) \\ 
    \bottomrule
  \end{tabular*}
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{plots/eeg_alpha_beta_vs_block_slice.eps}
    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
\end{figure}

\subsection{Galvanic Skin Response (GSR) and Heart Rate}
The measure of GSR was the log of total amplitude in the signal.
To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, GSR data were normalized by the temporal duration of the block.
Due to sensor failure and the elimination of one subject with extreme values of GSR amplitude using the same rule as for EEG, the analysis comprised 34 participants.
No systematic trend due to delay or block length was observed.
In addition, the same analysis of slice position within block length as was performed for the EEG revealed no effects, indicating that GSR was stable across step positions within a block.

Heart rate, measured in beats per minute, averaged \num{90.2}~BPM and showed no trend related to the experimental variables of delay and block length.
The same was true of the variability in beats per minute (average standard deviation across conditions was \num{20.3}).

The absence of systematic trends in both these results is interesting in the context of our initial suggestions of potential mechanisms relating delay in the application to human behavior.
In section \cref{ssec:potentialmechs} we proposed that emotional arousal during the execution of the task was a potential explanation for the effects observed in the human.
However, these results seem to refute this hypothesis, and will be further discussed in \cref{sec:discussion}.

\subsection{Individual Difference Analysis}

The Big 5 personality inventory and Immersive Tendencies Questionnaires (ITQ) were combined with outcome variables in an analysis of individual differences.
Given the previous results, we initially considered the following outcome variables:
\begin{itemize}
    \item execution time in the most demanding block (length 12, delay \SI{3.0}{\second}); 
    \item total of alpha and beta bands EEG in the first four steps at all lengths (``Slice 1'');
    \item average heart rate;
    \item and average log GSR.\@
\end{itemize}

Although the heart rate and GSR data had produced no significant effects in the analysis of experimental variables, they could in principle correlate with experimental outcomes across individuals. 
A principal-components analysis (\emph{PCA}) on a subset of these variables, shown in \cref{tab:pca}, was ultimately conducted on the 28 participants for which there was EEG data. 
Three of the personality measures were excluded after initial analyses indicated significant correlations among neuroticism, openness, agreeableness, and extroversion.
Neuroticism (with poles of sensitivity and security) was selected as most relevant to the issue of response to system delay and was included along with conscientiousness.
GSR was used in a subsidiary analysis, because only 25 participants had both EEG and GSR measures that were reliable.

\begin{table}[h]
  \centering
  \caption{Principal Component Analysis}\label{tab:pca}
  \begin{subtable}[h]{\textwidth}
    \centering
    \caption{Main components identified.}
    \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
    \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
      \toprule
      \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} \\
      \midrule
      BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
      BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
      ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
      ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
      Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
      Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
      \bottomrule
    \end{tabular*}
  \end{subtable}
  \newline
  \medskip
  \newline
  \begin{subtable}[h]{\textwidth}
    \centering
    \caption{Percentages of variance explained by the components.}
    \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
      \toprule
      {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total} \\
      \midrule
      Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent} & \SI{73.13}{\percent} \\
      \bottomrule
    \end{tabular*}
  \end{subtable}
\end{table}

The PCA produced three components that accounted for \SI{73.13}{\percent} of the variance in the six factors considered.
Component 1 included neuroticism, both ITQ scores, and execution time, indicating that more sensitive and immersed individuals tended to slow their responses under extended delay.
Component 2 included conscientiousness and the absence of neuroticism along with immersive involvement, indicating that efficient and secure participants tended to be more involved.
The third component had positive loadings on EEG in Slice 1 (first four steps of a block) and the focus component of the ITQ.\@
An additional analysis including heart rate added a component but improved the PCA fit by only \SI{4.5}{\percent}, indicating that any variability in this measure across individuals is unrelated to personality, immersiveness, or outcome.
A further analysis including GSR produced a solution in which GSR loaded with the first component, along with execution time.
% A further analysis including GSR again produced a three-component solution, in which GSR loaded with the first component (execution time, sensitivity, immersion).

\cref{fig:neuro:exectime:reg} shows the correlation between neuroticism and execution time for the block with extreme values of length (12 steps) and delay (\SI{3.0}{\second}), for the full set of 40 participants.
It confirms the relationship indicated by the first component in the PCA using the smaller sample of 28 participants; that is, higher neuroticism is associated with more responsiveness to delay.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{plots/reg_exectime_vs_neuroticism.eps}
    \caption{Correlation between neuroticism score of participants and their execution time in the longest block at the longest delay.
    Pearson correlation coefficient \( r = .40 \); 2-tailed \( p = .01 \).}
    \label{fig:neuro:exectime:reg}
\end{figure}

On the whole, these results suggest that individual differences in widely accepted personality variables and immersive tendencies moderate the response to delay.
This fact could have practical implications in the future.
It could, for instance, provide a tunable parameter for eventual models aiming to emulate human interaction with a WCA.\@
In addition, physiological measures of heart rate and EEG appear not to be direct indicators of  behavioral response to delay, although GSR may be more promising in this regard.

\section{Discussion}\label{sec:discussion}

% First summarize the main findings
We start our discussion with the main results of our experimentation presented in the previous section:

\begin{itemize}
  \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional} behavioral slow-down.
  That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their execution of the task.
  This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
  Moreover, there is a temporal component to this effect; users become progressively slower the more time passes with reduced system responsiveness.

  \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
  The question of whether any lingering effect can be measured after these four steps remains open.

  \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at performing steps as the task progresses.
  However, the strength of this effect decreases as delay increases.
  Whereas for blocks without delay users performed the last four steps of a 12-step block on average \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.

  \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay. 
  The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
  Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables related to attentiveness, respectively.
  In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\autocite{quesnel2018you}.
  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\autocite{hirsh2008delay}.
\end{itemize}

A central question therefore arises: to which physio- and psychological mechanisms can these findings, most importantly the substantial slow-down in task execution, be attributed?

In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional reactions.
These could in turn elicit generalized arousal.
We also postulated that adapting to delay might progressively deplete cognitive resources in users.
However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
The acceleration data further did not indicate that extended delay significantly increases erratic movement.
% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
% The acceleration data further do not indicate that extended delay increases erratic movement.
To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
That is, users fail to capitalize on the new information as quickly as they could.
Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.
We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
This also provides a possible explanation for the lingering effects of delay after an acceptable system responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.

As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.

The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.

\begin{itemize}

  \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to significantly extended application lifetimes.
  This has direct implications for resource and power consumption.

  \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a high-responsive state could have unconventional consequences for resource allocation.
  This is of particular importance, for instance, for cases where the user may be able to finish the task before these effects subside.
  In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired application.

  \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of system responsiveness across applications may not ultimately be beneficial to the system as a whole.
  Take for instance two applications on the same system which negatively interfere with each other. 
  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop. 
  In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.

  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
  This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.
  However, this remains an open challenge.

\end{itemize}

To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of WCA deployments.
Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable tool to tackle inefficiencies in these systems.
Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the relationship between application responsiveness and human behavior.
More research in this area will surely uncover more complex and interesting behaviors.
Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\autocite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.

\section{Conclusion}\label{sec:conclusion}

In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.

We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
This was evidenced by an increase in task execution times, even after accounting for the artificially introduced delays, as well as accelerometer data from participants wrists.
Additionally, we found that the strength of this effect is modulated by individual differences between subjects.

These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
In this context, we believe there are two direct and important next steps to be performed.
The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
We believe these need to be implemented, tested and validated in real setups.
In particular, we identify two of these implications to be prime candidates for their own experimental studies.
One, our postulation that in cases where an impaired application is close to finishing, diverting resources to it might not be the most optimal course of action; and two, the possibility that ``fair'' degradation of system responsiveness across applications may be, in some cases, undesirable.
Both of these questions could be answered with straightforward setups.

The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
Incorporating the results presented here would allow for much more realistic workloads.
For instance, our findings relating to the effects of delay on execution times could be directly adapted to modulate timings in the input stream.
Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.

\begin{acks}
  This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.

  Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or the above-mentioned funding sources.
\end{acks}

\nolinenumbers%

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
%

\printbibliography%
\end{document}

### 2022-08-19 publications/2021ImpactDelayedResponse/paper_old.tex
plos

--- 2022-08-19 publications/2021ImpactDelayedResponse/paper.tex
@@
 
-Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\autocite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
-They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
+Wearable Cognitive Assistants, \emph{WCA} for short, have recently started to garner attention from the research community~\autocite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
+They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to augment human cognition in both day-to-day tasks and professional settings.
 Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
@@
 
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\autocite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\autocite{Chen:AnEmpiricalStudyOfLatency}.
+WCA systems show great potential for two main use-cases.
+One is providing quality of life improvements to the millions of people around the world affected by some form of cognitive decline. 
+WCA can, for instance, provide assistance to people recovering from traumatic brain injuries, smoothly guiding them through day-to-day interactions with the world which would otherwise be extremely challenging.
+
+The second use-case is as a companion tool for specialists and as a means for guiding their training. 
 Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\autocite{funk2015caworkplace,gorecky2011cognito}.
@@
 % Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
-Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\autocite{dabrowsky:2011:40years}.
-Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
-A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
+Characterizing the relationships between system responsiveness and user behavior and experience is of paramount importance for the design and evaluation of these applications.
+A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployments of WCA.\@
 % Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.
@@
 We expand upon the findings of \textcite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of \textcite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
-While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
-We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
+While these bounds present a general understanding of when it is likely that a user will drop an application, they do not provide any insights as to what happens \emph{before} that --- i.e.
+how human behavior changes with system responsiveness.
+We aim to tackle this question through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
 This is an important step toward a more systematic understanding of human behavior in this domain.
 
-We present in this paper the design and elaboration of an experimental WCA test-bed.
-This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
+We present in this paper an experimental WCA test-bed of our design and making.
+This test-bed was subsequently employed in a study in which undergraduate students were asked to interact with and follow the instructions given to them by a cognitive assistant.
 Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
@@
 
-Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
-
-\begin{itemize}
-  \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}
-
-  In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
-  The extent or form of these changes were however unknown.
-  We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.
-
-  Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
+Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
+This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
 While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
 To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-  Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.
-
-  \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}
-
-  We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.
-
-  The results we present, however, seem to refute this hypothesis. 
-  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.
-
-  \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}
-
-  In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
-  In particular, we expected emotional effects to be correlated with the strength of the added delay.
-
-  The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
-  This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
-
-  \item \emph{Are these effects mediated by personality indicators in any way?}
-
-  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\autocite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
-  We also expected \emph{focus} and \emph{involvement}~\autocite{witmer1998:itq} to play roles in this.
-
-  The results obtained agree with our hypothesis. 
-  We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.
-
-\end{itemize}
-
-% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
-% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
-% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
-% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\autocite{john1999:bfi}.
+Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\autocite{john1999:bfi}, which has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
 
@@
 \section{Background and Related Work}\label{sec:background}
-
-% Overarching theo construct: people perform temp seq task in the framework of cognition and emotion and individual differences mediate response to delay.
-In particular we view cognition as described by models such as ACT-R as a seq of procedures.
-
-% Section 2.
-% Accordingly we review time perception...
-
-% Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
-% Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
-% In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
-
-Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
-In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
-Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
-
-Accordingly, in the following we review background work in the areas of 
-\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-  \item[(\cref{ssec:timeperception})] time perception
-  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
-\end{enumerate*}
-
-\subsection{Time perception in computing systems}\label{ssec:timeperception}
+\subsection{Time perception in computing systems}
 
@@
 
-\subsection{Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}
+\subsection{Potential Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}
 
@@
 
-A first possible explanation comes from research on cognitive and motor planning. 
-Delay may move users from relatively automatic to more attention-demanding processing.
+Research on cognitive and motor planning suggests that delay may move users from relatively automatic to more attention-demanding processing.
 Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
+
 As competency in a task increases, execution of the hierarchy becomes increasingly automated.
@@
 
-A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
+An alternative view of delay effects appeals to emotional systems rather than cognitive processes.
 As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\autocite{lee2019limiting}.
 
-Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\autocite{baumeister74tice,lin2020strong}.
+A third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\autocite{baumeister74tice,lin2020strong}.
 
@@
 This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board.
-
-Subjects were recruited from a pool of undergraduate students at Carnegie Mellon University.
-Students enrolled in an introductory-level psychology course fulfill a research requirement as part of the plan of study.
-This requirement may be fulfilled either through the elaboration of a written essay or by volunteering as a participant in a small number of research experiments.
-
-No particular exclusion criteria were applied, and as specified by our approved data-collection protocol, no gender- or sex-related statistics were collected.
+Subjects were recruited from a pool of undergraduate students fulfilling a course requirement at Carnegie Mellon University. 
+No particular exclusion criteria were applied. 
 In total, 40 participants were recruited, all of them of college student age (\numrange{18}{25} years old).
@@
 We used a modified version of the LEGO Assistant application introduced by \textcite{Chen:EarlyImplementation}.
-This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
-% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
-The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
-Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
-In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
-% The results of the user performing these instructions are provided as inputs to the system.
-% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.
-
-% In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
-In the following, we will provide some formal definitions relating to the WCA application.
-% \theoremstyle{definition}
-% \begin{definition}[Task and subtask]
+This application belongs to a category of \emph{WCA}s that are designed to guide users through the execution of a sequential task.
+Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
+A set of instructions are to be performed by the user in a semi-predetermined order.
+The results of the user performing these instructions are provided as inputs to the system.
+These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.
+
+In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
+\theoremstyle{definition}
+\begin{definition}[Task and subtask]
   A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
-\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{fig:task:steps}).
-% We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
-A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
-At that point, the assistant either provides a new instruction to the user, or the task ends.
-  % See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
-% \end{definition}
+  \emph{Subtask} will refer to a specific action to be performed by the user, described by a single instruction.  
+  See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
+\end{definition}
 
@@
   \includegraphics[width=.8\textwidth]{images/step_task_description.pdf}
-  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
+  \caption{Example of a Cognitive Assistance task and its component subtasks and steps.}\label{fig:task:steps}
 \end{figure*}
 
-% \begin{definition}[Step]
-
-% While the subtask remains unfinished, the assistant discards results of the processing of the sampled inputs.
-% At the beginning of a step, an instruction is given to the user.
-% The user then proceeds to perform the subtask specified by the instruction, while the cognitive assistant is continuously sampling the subtask state at specified intervals.
-
-% Once the user finishes the required action, the next sample taken will contain a valid input, and thus the cognitive assistant will provide a new instruction. 
-% This finishes the current step and potentially begins a new one if there remain instructions to be performed in the task.
-% \end{definition}
-
-In the base LEGO Assistant, the task consists of the assembly of a LEGO model; each step requires the user to append a LEGO brick to the model at a specific location and orientation.
-The system monitors progress through a video feed and provides timely feedback in the form of visual and textual instructions to guide the user.
-
-The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
-In particular
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-    \item it is easily understood by users and requires essentially no training
-    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
-    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
-    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
-\end{enumerate*}
+\begin{definition}[Step]
+  Interval of time delimited by two consecutive instructions, see \cref{fig:task:steps,fig:cogassist:step}.
+  Steps are characterized by the actions of the user and the assistant.
+  At the beginning of a step, an instruction is given to the user.
+  The user then proceeds to perform the subtask specified by the instruction, while the cognitive assistant is continuously sampling the subtask state at specified intervals. 
+  While the subtask remains unfinished, the results of the processing of the sampled inputs are discarded.
+  Once the user finishes the required action, the next sample taken will contain a valid input, and thus the cognitive assistant will provide a new instruction. 
+  This finishes the current step and potentially begins a new one if there remain instructions to be performed in the task.
+\end{definition}
+
+In the base LEGO Assistant, the task consists of steps leading the user through the assembly of a LEGO model; each subtask requires the user to append a LEGO brick to the model at a specific location and orientation.
+The system monitors progress through a video feed and provides timely feedback in the form of visual and textual instructions to guide the user towards the desired end result.
+
+The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant:
+\begin{enumerate}
+    \item \ It is easily understood by users and requires essentially no training. 
+    \item\ The step-by-step nature of these applications simplifies the isolation of the relevant experimental variables and the effects of the delay.
+    \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
+    \item Each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
+\end{enumerate}
 
@@
   \centering
-  % \input{plots/lego_hierarchy_diagram.tex}
-  \includegraphics[width=\textwidth]{plots/lego_hierarchy_diagram_standalone.eps}
+  \input{plots/lego_hierarchy_diagram.tex}
   \caption{Hierarchical cognitive structure of a step in the LEGO task.}
@@
 
-% The original design of the LEGO assistance application was based on a client-server model communicating over a wireless network, with the client software running on a wearable device and the server software deployed on a cloudlet.
-For the purposes of this study, the original client-server design of the LEGO Assistant was altered to be locally executable.
-This was done in order to eliminate the stochastic effects of jitter and latency from the network link, and greatly simplify instrumentation.
-% By this means, we achieve much more fine-grained control over the latencies to which the system is subject.
-% Additionally, this greatly simplified the instrumentation of the application.
+The original design of the LEGO assistance application was based on a client-server model communicating over a wireless network, with the client software running on a wearable device and the server software deployed on a cloudlet.
+For the purposes of this study, this design was altered to be executable on a single, non-networked computer, in order to eliminate the stochastic effects of jitter and latency on the network link.
+By this means, we achieve much more fine-grained control over the latencies to which the system is subject.
+Additionally, this greatly simplified the instrumentation of the application.
 Instructions were output in image and text form to a computer display situated on a table directly in front of the participants.
@@
 
-% Finally, a data collection and experimental layer was implemented between the user interface and video capture and the core processing component of the LEGO Assistant.
-% This layer controlled the manipulated experimental variables and recorded the measured biometric and task-related effects.
+Finally, a data collection and experimental layer was implemented between the user interface and video capture and the core processing component of the LEGO Assistant.
+This layer controlled the manipulated experimental variables and recorded the measured biometric and task-related effects.
 
@@
 For the experimental LEGO task, a key modification was made to the structure of the steps.
-After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
+After each the processing of each input frame is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in figure \cref{fig:cogassist:step:delay}.
 The length of this delay is one of two independent variables we manipulated for the experiment.
 Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by \textcite{Chen:AnEmpiricalStudyOfLatency}.
-% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
-% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
-Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
-% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.
-
-In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
-% \begin{definition}[Block]
-%   A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
-The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
-We used values of \numlist{4;8;12} steps, chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
-Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block.
-% ; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
-% \end{definition}
-
-To assign tasks to the participants, a pseudo-random permutation of the combinations of block length and delay was first generated, and a unique sequence of steps assigned to each of these combinations to create 21 unique blocks.
-A \emph{Latin~square}-type design was then used to reorder this initial permutation in order to generate a task for each participant.
-This ensures a counterbalance of the order of the blocks across participants and avoids systematic learning effects. 
-% The design rotates the block types (as defined by length and delay) across participants so that each type is tested in each ordinal position, but it only coarsely samples from the \( 21 \times 20 \) possible sequences from one block type to another.
-Note that unlike the base LEGO assistant task, in which instructions led participants through the assembly of a specific model, the experimental LEGO task consisted of a sequence of instructions to either add or remove pieces with no evident goal.
-% Users were directed to either add a piece to the ongoing model or to remove a piece, and blocks were designed in such as way so as to ensure that the transitions between them were invisible to the user.
-
-% In this paper we will thus consider blocks to be our basic element of study, and most aggregations will be done at this level (with a few exceptions).
-% For this, we will need additional definitions:
-
-% \begin{definition}[Delay associated with a block]
-%   We will refer to the delay of a block as the delay applied to every frame of every step in that block.
-% \end{definition}
-
-% \begin{definition}[Execution time]
-
-Given the variability in the system latency to detect step completion, correcting step-completion time for system response time by a fixed amount would not enable sufficient precision.
-We therefore define the \emph{execution time} empirically for an individual step as the total time between the user receiving the instruction for the step and the time instant of the sample capturing the completed step.
-% their presenting the completed subtask to the system.
-
-% We define the \emph{execution time} of a step as the time between the user receiving an instruction for a subtask and them presenting the completed subtask to the system.
-% This, however, could not be measured directly and has to be inferred from the records of inputs processed by the system.
-
-For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{fig:exectime:diagram}).
-\( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
-If we define \( t_{c} \) as the instant marking when the user presented the finished step to the system, it must follow that \( t_{n-1} < t_{c} < t_{n} \).
-Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
-However, this introduces no bias in the results, as we can assume that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
-We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
-% This is also aggregated into an average at the block level.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
-%   \caption{Visualization of the execution time of a step.}%
-%   \label{fig:exectime:diagram}
-% \end{figure}
-
-% \end{definition}
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}[t]{.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{images/step_structure.pdf}
+A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
+Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
+Thus, our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state, while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.
+
+Additionally, in order to study the effects of a delay applied across multiple steps, we implemented an experimental design component called a \emph{block}:
+
+\begin{definition}[Block]
+  A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
+  The \emph{length} of a block corresponds to the number of steps it encompasses, and is the second of the two independent variables manipulated in our experiment.
+  We used values of \numlist{4;8;12} steps for the lengths, values chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
+  Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
+\end{definition}
+
+\begin{figure}[hp]
+  \centering
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=.8\textwidth]{images/step_structure.pdf}
     \caption{Structure of a step in a generic cognitive assistance application. 
-    The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
-    % by the assistant, as they do not contain relevant information.
-    Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
-    % which will subsequently be provided to the user.}%
-    }
+    The assistant provides an instruction to the user and continuously samples the subtask state; inputs captured while the subtask is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction) by the assistant, as they do not contain relevant information.
+    However, once the user finishes performing the given instruction, the next input \emph{will} cause the generation of a new instruction, which will subsequently be provided to the user.}%
     \label{fig:cogassist:step}
-  \end{subfigure}%
-  % \medskip%
-  \hfill%
-  \begin{subfigure}[t]{.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{images/step_structure_delay.pdf}
-    \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
+  \end{subfigure}
+  \medskip%
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=.8\textwidth]{images/step_structure_delay.pdf}
+    \caption{Modified structure of a step in the experimental task.
+In contrast to \cref{fig:cogassist:step}, an additional variable segment of time is introduced immediately following the processing of the input frame in the cognitive assistant, in order to extend the perceived processing time of the input to a specific target delay.}%
     \label{fig:cogassist:step:delay}
@@
   \medskip%
-  \begin{subfigure}[t]{.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{images/block.pdf}
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=.8\textwidth]{images/block.pdf}
     \caption{Structure of a block in the experimental task.}%
@@
   \end{subfigure}%
-  % \medskip%
-  \hfill%
-  \begin{subfigure}[t]{.49\textwidth}
-    \centering
-    \includegraphics[width=\textwidth]{images/exec_time_diagram.pdf}
+  \caption{Components of the cognitive assistance task.}
+\end{figure}
+
+To assign tasks to the participants, a pseudo-random permutation of the combinations of block length and delay was first generated, and a unique sequence of steps assigned to each of these combinations to create 21 unique blocks.
+A \emph{Latin square}-type design was then used to reorder this initial permutation in order to generate a task for each participant.
+This ensures a counterbalance of the order of the blocks across participants and avoids systematic learning effects. 
+The design rotates the block types (as defined by length and delay) across participants so that each type is tested in each ordinal position, but it only coarsely samples from the \( 21 \times 20 \) possible sequences from one block type to another.
+Note that unlike the base LEGO assistant task, in which instructions led participants through the assembly of a specific model, the experimental LEGO task consisted of a sequence of instructions with no evident goal.
+Users were directed to either add a piece to the ongoing model or to remove a piece, and blocks were designed in such as way so as to ensure that the transitions between them were invisible to the user.
+
+In this paper we will thus consider blocks to be our basic element of study, and most aggregations will be done at this level (with a few exceptions).
+For this, we will need additional definitions:
+
+\begin{definition}[Delay associated with a block]
+  We will refer to the delay of a block as the delay applied to every frame of every step in that block.
+\end{definition}
+
+\begin{definition}[Execution time]
+
+  Given the variability in the system latency to detect step completion, correcting step-completion time for system response time by a fixed amount would not enable sufficient precision.
+  We therefore estimated the \emph{execution time} empirically for an individual step as the total time between the user receiving the instruction for a subtask and their presenting the completed subtask to the system.
+
+  % We define the \emph{execution time} of a step as the time between the user receiving an instruction for a subtask and them presenting the completed subtask to the system.
+  % This, however, could not be measured directly and has to be inferred from the records of inputs processed by the system.
+
+  Formally, for an arbitrary step, we define the sequence \( S = \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step.
+  \( t_{0} \) corresponds then to the instant where the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given (or conversely, the sample which captured the finished subtask as presented by the user).
+  If we define \( t_{c} \) as the instant marking when the user finished and presented the task to the system, we infer the following:
+
+  \begin{itemize}
+    \item \( t_{c} < t_{n} \), since by definition the user must have finished the task before the system took the sample that triggered a new instruction.
+    \item \( t_{c} > t_{n - 1} \), since otherwise the system would have triggered a new instruction after some \( t_{k}, k \in [0, n-1] \) instead of after \( t_{n} \).
+  \end{itemize}
+
+  Therefore, \( t_{n-1} < t_{c} < t_{n} \) (see also \cref{fig:exectime:diagram}).
+  Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
+  However, this introduces no bias in the results, as we can infer that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
+  We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
+  This is also aggregated into an average at the block level.
+
+  \begin{figure}
+    \centering
+    \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
     \caption{Visualization of the execution time of a step.}%
     \label{fig:exectime:diagram}
-  \end{subfigure}%
-  % \medskip%
-  \caption{Components of the cognitive assistance task.}
   \end{figure}
+
+\end{definition}
 
@@
 The participant wore devices to acquire four physiological measures:
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-  \item galvanic skin response (\emph{GSR})
-  \item accelerometer data from the dominant wrist
-  \item brain activity in the form of electroencephalography (\emph{EEG})
-  \item heart rate.
-\end{enumerate*}
+
+\begin{itemize}
+  \item galvanic skin response (\emph{GSR});
+  \item accelerometer data from the dominant wrist;
+  \item brain activity in the form of electroencephalography (\emph{EEG});
+  \item and heart rate.
+\end{itemize}
 
 These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\autocite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
-GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\autocite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
-EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\autocite{Antonenko2010,Grimes2008,kumar2016measurement}.
+More specifically, galvanic skin response (\emph{GSR}, also known as electrodermal activity) is the measure of the variation of the conductive properties of human skin due to changes in the state of the sweat glands.
+It is interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\autocite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
+Electroencephalography (\emph{EEG}) refers to the monitoring of brain activity through the measurement of the fluctuations of the electric field surrounding the brain.
+\emph{EEG} measures the voltage fluctuations due to electrical activity within neurons in the brain, which result in distinct waves of specific frequencies associated with different contexts, emotions and actions.
+\emph{EEG} has  been used to measure cognitive load in the context of human-computer interactions~\autocite{Antonenko2010,Grimes2008,kumar2016measurement}.
 
@@
 GSR was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} (BVP) signal sampled at \SI{64}{\hertz}.
-Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach a stable equilibrium and establish a baseline for the signals.
-
-The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\autocite{ragot2017emotion, mccarthy2016validation}.
-The E4 also includes a skin temperature thermometer, however the measure was not used for the present study.
-
-For the EEG data we employed the \textcite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
-It provides a quick and non-invasive way of obtaining EEG signals from participants.
-Electrodes were placed according to the \emph{10--20 Electrode System}~\autocite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
-The signals were sampled at \SI{200}{\hertz} and postprocessed to 
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
-  \item smooth out noise (using the technique proposed by~\textcite{agarwal2017eeg}).
-\end{enumerate*}
+Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach an stable equilibrium and establish a baseline for the signals.
+
+The E4 wristband was chosen due to its small, non-invasive and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous research~\autocite{ragot2017emotion, mccarthy2016validation}.
+The E4 also includes a skin temperature thermometer; however, due to failure to reach equilibrium, the measure was not used for the present study.
+
+For the EEG data we employed the \textcite{openbci:headbandkit}, a kit consisting of a number of dry electrodes fastened to a Velcro headband.
+It thus provides a quick, easy and non-invasive way of obtaining EEG signals from participants.
+Electrodes were placed according to the \emph{10--20 Electrode System}~\autocite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe.
+Ground and reference electrodes were positioned on the right and left earlobes respectively.
+
+This kit was paired with the \textcite{openbci:ganglion} for the actual acquisition of the signals, which were sampled at \SI{200}{\hertz} and streamed over Bluetooth to the experiment computer.
+
+Following capture, the  EEG signal was post-processed in the following fashion:
+% 
+\begin{enumerate}
+  \item Since our main interest was in the \( \alpha \) (\SIrange{8}{12}{\hertz}) and \( \beta \) (\SIrange{12}{30}{\hertz}) bands, a low-pass Butterworth filter of order 8 and cutoff frequency \SI{40}{\hertz} was applied to filter out noise in the higher end of the spectrum (in particular, noise from the board power supply at \SI{60}{\hertz}).
+  \item A high-pass Butterworth filter of order 8 and cutoff frequency \SI{0.1}{\hertz} was then applied to filter out noise at the low end of the spectrum.
+  \item Subsequently, a pair of cascaded Savitsky-Golay filters~\autocite{savitzky1964smoothing}, both of order 8 and window size 21, were applied to the signal in order to smooth out noise, as proposed by \textcite{agarwal2017eeg}.
+  \item A spectrogram was then calculated for the signal.
+  \item Finally, the power over time for each EEG frequency band is obtained by integrating over the relevant frequencies for each time-step in the spectrogram.
+\end{enumerate}
+
+An example of the effects of the filtering on the raw EEG signal can be observed in \cref{fig:eeg:raw_vs_filt}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.8\textwidth]{plots/raw_vs_filt_eeg.eps}
+  \caption{Comparison of an EEG signal before and after filtering.
+  Note in particular the artifact from the DC power supply of the capture board around \SI{60}{\hertz} in the raw signal, and how it essentially disappears after filtering.}%
+  \label{fig:eeg:raw_vs_filt}
+\end{figure}
 
@@
 
-Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
+Both video feeds were captured at a rate of \num{24}~FPS (i.e.\ with a sampling interval of \( \SI[parse-numbers=false,per-mode=symbol]{41.\overline{6}}{\milli\second} \)) in parallel processes to ensure a constant rate of capture.
 Examples of the captured frames can be seen in \cref{fig:sampleframes}.
@@
 
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\autocite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
-% An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
-% The degrees of freedom represent the number of observations going into each of these variability measures.
-% As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
-% The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
-% Conventionally \( p < .05 \) is the criterion for significance.
-% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
-
-% \begin{figure}[h]
-%     \centering
-%     \includegraphics[width=.8\textwidth]{plots/exec_time_delay_length.eps}
-%     \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
-% \end{figure}
+We confirmed this effect through an \emph{analysis of variance} (ANOVA) with factors of block length and delay.
+An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
+The degrees of freedom represent the number of observations going into each of these variability measures.
+As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
+The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
+Conventionally \( p < .05 \) is the criterion for significance.
+The current length X delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
 
@@
     \includegraphics[width=.8\textwidth]{plots/eeg_alpha_beta_vs_block_slice.eps}
-    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
+    \caption{Mean Log EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
 \end{figure}
@@
 We also postulated that adapting to delay might progressively deplete cognitive resources in users.
-However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
-Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-The acceleration data further did not indicate that extended delay significantly increases erratic movement.
-% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-% The acceleration data further do not indicate that extended delay increases erratic movement.
+However, the present data provide relatively little support for these alternatives, in that physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+The acceleration data further do not indicate that extended delay increases erratic movement.
 To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
@@
   Take for instance two applications on the same system which negatively interfere with each other. 
-  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop. 
+  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a negative feedback loop. 
   In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.
 
-  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
+  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristic from measured execution times.
   This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.

### 2022-08-19 publications/2021ImpactDelayedResponse/plots/lego_hierarchy_diagram.tex
plos

\begin{tikzpicture}[grow=down, scale=.6]
    \tikzset{edge from parent/.style= 
            {draw, thick, edge from parent fork down},
            every tree node/.style=
            {draw,very thick,align=center,scale=1,fill=black!10!white},
            every leaf node/.style=
            {fill=green!60!black!25!white,text width=2cm,minimum width=2cm},
            level distance=1in,sibling distance=.05in,
            execute at begin node=\strut
            }

    \node[draw,dashed,align=center](command){
        \footnotesize{Example instruction:}\\
        \small\textbf{``Find a $1 \times 1$ black piece and add it to}\\
        \textbf{the top right of the current model.''}
    % \\
    % \large{\textbf{``Remove the $1 \times 3$ red piece from the top of the current model.''}}
    };

    \begin{scope}[shift={($(command.south)+(0,-.25in)$)}]
        \Tree [.\node[fill=red!50!white]{\textbf{\Large{Change model}}};
                \node[diamond, aspect=2, fill=yellow!25!white](choice){Add or remove a piece?};
                ]
    \end{scope}

    \begin{scope}[shift={($(choice.west)-(1.5in,0)$)}]
        \Tree [.\node[draw=none, fill=none](add){Add a piece};
                [.{Select piece $X$} 
                    {Set $X$ size and color}
                    {Search box for matching pieces} ]
                [.{Select location $Y$} 
                    {Set $Y$ vertical and horizontal position}
                    {Find location on model} ]
                [.{Place $X$ at $Y$} 
                    {Pick up $X$ and move to model}
                    {Attach $X$ to model} ]
            ]
    \end{scope}

    \begin{scope}[shift={($(choice.east)+(1.5in,0)$)}]
        \Tree [.\node[draw=none, fill=none](remove){Remove a piece};
                \node[draw=none, fill=none]{\Huge{\ldots}}; 
                % [.{Select location $Y$} 
                %     {Set $Y$\\vertical and\\horizontal\\position}
                %     {Find\\location\\on model} ]
                % [.{Select piece $X$ at $Y$}
                %     {Set $X$\\size and\\color}
                %     {Search $Y$ for\\matching\\pieces} ]
                % [.{Remove piece $X$} 
                %     {Detach\\$X$}
                %     {Return $X$\\to box} ]
            ]
    \end{scope}

    \draw [thick] (choice.west) -- (add.east);
    \draw [thick] (choice.east) -- (remove.west);

\end{tikzpicture}%

### 2022-08-19 publications/2021ImpactDelayedResponse/prepare.py

@@ -1,0 +1,153 @@
+import shutil
+import subprocess
+from dataclasses import dataclass
+from io import StringIO
+from pathlib import Path
+from typing import Iterator, Tuple
+
+import click
+import parse
+
+
+class Figure:
+    @dataclass
+    class SubFig:
+        sub_index: str
+        line: int
+        original_path: str
+
+    def __init__(self,
+                 index: int):
+        self._index = f'Fig{index}'
+        self._subfigs = []
+
+    def iterfigs(self) -> Iterator[Tuple[int, str, str]]:
+        n_subfigs = len(self._subfigs)
+        if n_subfigs == 1:
+            sfig = self._subfigs[0]
+            yield sfig.line, self._index, sfig.original_path
+        elif n_subfigs > 1:
+            for sfig in self._subfigs:
+                yield sfig.line, \
+                      f'{self._index}{sfig.sub_index}', \
+                      sfig.original_path
+        else:
+            raise RuntimeError()
+
+    def add_subfig(self, line: int, graphics: str) -> None:
+        sub_index = chr(ord('a') + len(self._subfigs))
+        self._subfigs.append(Figure.SubFig(sub_index, line, graphics))
+
+    def __str__(self) -> str:
+        n_subfigs = len(self._subfigs)
+        if n_subfigs == 1:
+            return f'{self._index} {self._subfigs[0].original_path}'
+        elif n_subfigs > 1:
+            return '\n'.join([f'{self._index}{s.sub_index} '
+                              f'{s.original_path}' for s in self._subfigs])
+        else:
+            raise RuntimeError()
+
+
+class DirectoryPath(click.ParamType):
+    name = 'directory'
+
+    def __init__(self, clean: bool = False,
+                 exist_ok: bool = True,
+                 parents: bool = False):
+        super(DirectoryPath, self).__init__()
+        self._clean = clean
+        self._exist_ok = exist_ok
+        self._parents = parents
+
+    def convert(self, value, param, ctx) -> Path:
+        p = Path(value).resolve()
+        if self._clean and p.exists():
+            subprocess.run(['rm', '-rf', str(p)])
+
+        p.mkdir(exist_ok=self._exist_ok, parents=self._parents)
+        return p
+
+
+@click.command()
+@click.option('--cls-file',
+              required=False,
+              default='./plos2015.bst',
+              show_default=True,
+              type=click.Path(exists=True, file_okay=True, dir_okay=False))
+@click.option('--bib-file',
+              required=False,
+              default='./bibliography.bib',
[…]

### 2022-08-19 publications/2021ImpactDelayedResponse/tables/generate.py

@@ -1,0 +1,76 @@
+#!/usr/bin/env python3
+import shutil
+import subprocess
+from pathlib import Path
+from tempfile import TemporaryDirectory
+from typing import Collection, IO
+
+import click
+import jinja2
+
+
+def generate_table_texs(template_str: str, tables: Collection[str]) \
+        -> Collection[str]:
+    latex_jinja_env = jinja2.Environment(
+        block_start_string='\BLOCK{',
+        block_end_string='}',
+        variable_start_string='\VAR{',
+        variable_end_string='}',
+        comment_start_string='\#{',
+        comment_end_string='}',
+        line_statement_prefix='%%',
+        line_comment_prefix='%#',
+        trim_blocks=True,
+        autoescape=False)
+    temp = latex_jinja_env.from_string(template_str)
+    return list(map(lambda t: temp.render(table=t), tables))
+
+
+@click.command()
+@click.argument('template', nargs=1, type=click.File())
+@click.argument('output_dir', nargs=1,
+                type=click.Path(exists=True, file_okay=False,
+                                dir_okay=True, resolve_path=True))
+@click.argument('tables', nargs=-1,
+                type=click.Path(exists=True, file_okay=True,
+                                dir_okay=False, resolve_path=True))
+def generate(template: IO, output_dir: str, tables: Collection[str]):
+    template_str = template.read()
+    tab_str = []
+    tab_names = []
+    for table in tables:
+        tab_path = Path(table)
+        with tab_path.open('r') as fp:
+            tab_str.append(fp.read())
+        tab_names.append(tab_path.name)
+
+    output_dir = Path(output_dir).resolve()
+    texs = generate_table_texs(template_str, tab_str)
+    # make temp dir
+    with TemporaryDirectory() as tmp_dir_str:
+        tmp_dir = Path(tmp_dir_str).resolve()
+        for tex, table_name in zip(texs, tab_names):
+            tex_file_path = (tmp_dir / table_name).resolve()
+            with tex_file_path.open('w') as fp:
+                fp.write(tex)
+
+            for i in range(4):
+                subprocess.Popen([
+                    '/usr/bin/latexmk',
+                    '-interaction=nonstopmode',
+                    '-shell-escape',
+                    '-synctex=1',
+                    '-f',
+                    f'-jobname={table_name}',
+                    f'-output-directory={tmp_dir}',
+                    f'{tex_file_path}'
+                ], cwd=str(tmp_dir)).wait()
+
[…]

### 2022-08-22 model.tex
working

@@
 \begin{enumerate}
-    \item System slow-down induces additional behavioral slow-down which scales with the decrease in system responsiveness.
+    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
     Moreover, users become progressively slower the longer they spend in a degraded system state.
@@
 
-Using these insights, we design a generator of realistic execution times for step-based \ac{WCA}.
-The generator further builds upon the original data collected for~\cite{olguinmunoz:impact2021}; this data is first preprocessed and then sampled according to a series of rules in order to produce novel,realistic sequences of execution times for a task.
+Based on these insights and the data collected for~\cite{olguinmunoz:impact2021}, we design a generator of realistic execution times for step-based \ac{WCA}.
+
+\todo[inline]{Diagram of model?}
+
+\begin{algorithm}
+    \caption{Timing model algorithm}
+    \KwIn{%
+        \begin{itemize}
+            \item $T_\text{start}$: start timestamp of the previous step.
+            \item $T_\text{success}$: capture timestamp of the succesful sample of the previous step.
+            \item $T_\text{end}$: end timestamp of the previous step.
+            \item $t_\text{exec}$: execution time (in seconds) of the previous step.
+        \end{itemize}%
+    }
+    \KwOut{$t_\text{exec}'$: execution time for the current step.}
+
+    \Comment{%
+        Initial values:\\
+        $d \leftarrow 0$\\
+        $I_\text{prev} \leftarrow \text{NULL}$
+    }
+    $t_\text{wait} \leftarrow T_\text{success} - (T_\text{start} + t_\text{exec})$\;
+    $t_\text{round-trip} \leftarrow T_\text{end} - T_\text{success}$\;
+    $i \leftarrow t_\text{wait} + t_\text{round-trip}$ \Comment*[r]{raw impairment}
+    $I \leftarrow \text{BIN}(i)$ \Comment*[r]{bin impairment into\\discrete levels}
+
+    \If{$I_\text{prev} == \text{NULL}$}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{NO TRANSITION}$\;
+    } \ElseIf{$I_\text{prev} < I$}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{LOWER TO HIGHER} $
+    } \ElseIf{$I_\text{prev} > I$}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{HIGHER TO LOWER} $  
+    }
+
+\end{algorithm}
+
+This data is first preprocessed and then sampled according to a series of rules in order to produce novel,realistic sequences of execution times for a task.
 Alternatively, instead of sampling directly from the preprocessed data, a theoretical distribution can be fitted, thus allowing for generation of realistic synthetic execution times.

### 2022-08-22 publications/2019EdgeDroid/paper_body.tex
working

--- 2022-08-19 publications/2019ScalingOnTheEdge/paper_body.tex
@@
     \centering
-    \input{img/pipeline.tex}
+    \input{publications/2019EdgeDroid/img/pipeline.tex}
     \caption{Pipeline of human-in-the-loop applications.}\label{fig:pipeline2}
@@
     \centering
-    \input{img/taskmodel.tex}
+    \input{publications/2019EdgeDroid/img/taskmodel.tex}
     \captionof{figure}{Segment of the internal task model of a linear task-guidance WCA.}%
@@
     \centering
-    \includegraphics[width=.7\columnwidth]{img/TraceReplay_GenArch}
+    \includegraphics[width=.7\columnwidth]{publications/2019EdgeDroid/img/TraceReplay_GenArch}
     \caption{Architecture of the benchmarking suite.}\label{fig:TraceReplayArch}
@@
     \centering
-    \input{img/usermodel.tex}
+    \input{publications/2019EdgeDroid/img/usermodel.tex}
     \captionof{figure}{Preliminary user model.}\label{fig:usermodel}
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/rtt_fb_vs_nofb}%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
     \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/box_feedback}%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
     \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{plots/comparison/nofonts/box_taskstep}%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
     \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%

### 2022-08-23 model.tex
explaining

@@
-\section{A model of human behavior for \ac{WCA}}\label{sec:model}
+\section{A model of human behavior for \acl*{WCA}}\label{sec:model}
 
-In order to accurately emulate the behavior of a human, a model for \ac{WCA} needs to implement two main behaviors.
+In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
+We employed a modified and instrumented version of the LEGO Assistant application devised by~\textcite{Chen2015LEGO}, which belongs to a category of \acp{WCA} with the goal of guiding a user through a sequential task.
+The application monitors the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
+Whenever the applications detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+Subjects interacted with this \ac{WCA} while the responsiveness of the system was altered in realtime and we captured key application and task performance metrics.
+Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
+
+From this experimentation, we identified four main results concerning the step execution times of humans in \ac{WCA}:
+
+\begin{enumerate}
+    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
+    Moreover, users become progressively slower the longer they spend in a degraded system state.
+
+    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
+
+    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
+
+    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
+\end{enumerate}
+
+In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+In order to accurately emulate the behavior of a human, such a model for needs to implement two main behaviors.
 One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
 And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
-\todo[inline]{Improve}
-
-In this section, we present and discuss our approaches to both these challenges.
 
 \subsection{Generating realistic timings}
+
+We begin by providing some formal definitions relating to step-based \acl{WCA}.
+\begin{description}
+    \item[Task:] understood as a finite sequence of instructions to be executed in order.
+    \item[Step:] specific action to be performed by the user, described by a single instruction; e.g. put a green $1 \times 3$ brick on the board.
+    \item[Step duration:] the time interval between 
+    \item[Step execution time:]
+    \item[Wait time:]
+    \item[\ac{TTF}:]   
+\end{description}
+
+The step starts at \(t_0\), when the instruction is presented to the user.
+We call the interval \(t_c - t_0\) the ``execution time'', corresponding to the time the user takes to complete the instruction.
+The final sample of the step is taken at \(t_n\) and it takes \( t_{n+1} - t_n \) time units for the user to receive a reply from the backend (``last sample \ac{RTT}'').
+Finally, we call the time interval between when the user finishes the instruction and the capturing of the final sample the ``wait time''; together with the last sample \ac{RTT} it composes the ``Time-to-Feedback'' of the step.%
+
+\begin{figure}
+    \centering
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{figs/task.png}
+        \caption{%
+            Overview of a task in a \ac{WCA}, composed of a series of steps.
+            Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
+            The \ac{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
+        }\label{}
+    \end{subfigure}\\
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{figs/step_time.png}
+        \caption{%
+            Breakdown of a step into its timing components.
+            The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
+            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \ac{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
+        }\label{}
+    \end{subfigure}
+    \caption{}
+\end{figure}
+
+A task will be understood as a finite sequence of instructions to be performed in order.
+Step will refer to a specific action to be performed by the user, described by a single instruction; e.g.
+put a green 1 × 3 brick on the board (see Fig 2).
+A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
+At that point, the assistant either provides a new instruction to the user, or the task ends.
 

### 2022-08-23 model.tex
writing

@@
 
-We begin by providing some formal definitions relating to step-based \acl{WCA}.
-\begin{description}
-    \item[Task:] understood as a finite sequence of instructions to be executed in order.
-    \item[Step:] specific action to be performed by the user, described by a single instruction; e.g. put a green $1 \times 3$ brick on the board.
-    \item[Step duration:] the time interval between 
-    \item[Step execution time:]
-    \item[Wait time:]
-    \item[\ac{TTF}:]   
-\end{description}
+We begin by providing some definitions relating to step-based \acl{WCA}.
+First of all, a \emph{step} is understood as a specific action to be performed by the user, described by a single instruction; a \emph{task} consists of a series of steps to be executed in sequence (see \cref{fig:task}).
+More formally, a step starts when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided.
+We call the time interval between these two events the \emph{step duration}.
 
-The step starts at \(t_0\), when the instruction is presented to the user.
-We call the interval \(t_c - t_0\) the ``execution time'', corresponding to the time the user takes to complete the instruction.
-The final sample of the step is taken at \(t_n\) and it takes \( t_{n+1} - t_n \) time units for the user to receive a reply from the backend (``last sample \ac{RTT}'').
-Finally, we call the time interval between when the user finishes the instruction and the capturing of the final sample the ``wait time''; together with the last sample \ac{RTT} it composes the ``Time-to-Feedback'' of the step.%
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals at which the \ac{WCA} captures the state of the physical system --- see \cref{fig:step}.
+\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample for said step is taken, and \( t_{n+1} \) to the instant at which the instruction for step \( M + 1 \) is provided.
+Furthermore, \( t_c \) represents the instant at which the user finishes performing the instruction for step \( M \); the interval \( t_c - t_0 \) we call the \emph{execution time} of \( M \).
+From the discrete sampling nature of the \ac{WCA} application, it follows that \( t_{n - 1} \leq t_c \leq t_{n} \); in plain English, the final state of the step is --- by definition --- always reached \emph{before} the final sample of the step is taken but after the next-to-last sampling instant.
+We define the intervals \( t_n - t_c \) and \( t_{n + 1} - t_n \) as the \emph{wait time} and \emph{last sample \ac{RTT}}, respectively, for step \( M \).
+The sum of these two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{time-to-feedback}.
+
+\emph{Execution time} and \emph{time-to-feedback} correspond to the two key metrics necessary to characterize the interactions between system responsiveness and human behavior.
 
@@
             The \ac{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
-        }\label{}
+        }\label{fig:task}
     \end{subfigure}\\
@@
             \( t_k | k \in \{1, \ldots, n \} \) correspond to the \ac{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
-        }\label{}
+        }\label{fig:step}
     \end{subfigure}
@@
 \end{figure}
-
-A task will be understood as a finite sequence of instructions to be performed in order.
-Step will refer to a specific action to be performed by the user, described by a single instruction; e.g.
-put a green 1 × 3 brick on the board (see Fig 2).
-A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
-At that point, the assistant either provides a new instruction to the user, or the task ends.
-
-\input{tables/empirical_exec_times.tex}
-
-In our previous work~\cite{olguinmunoz:impact2021}, we identified four main results concerning the step execution times of humans in \ac{WCA}:
-
-\begin{enumerate}
-    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
-    Moreover, users become progressively slower the longer they spend in a degraded system state.
-
-    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
-
-    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
-
-    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
-\end{enumerate}
-
-Based on these insights and the data collected for~\cite{olguinmunoz:impact2021}, we design a generator of realistic execution times for step-based \ac{WCA}.
 

### 2022-08-24 model.tex
writing

@@
 
-\emph{Execution time} and \emph{time-to-feedback} correspond to the two key metrics necessary to characterize the interactions between system responsiveness and human behavior.
+\emph{Execution time} and \emph{time-to-feedback} correspond to the two key metrics necessary to characterize the interactions between human behavior and system responsiveness.
 

### 2022-08-25 model.tex
explaining ttf

@@
 Furthermore, \( t_c \) represents the instant at which the user finishes performing the instruction for step \( M \); the interval \( t_c - t_0 \) we call the \emph{execution time} of \( M \).
-From the discrete sampling nature of the \ac{WCA} application, it follows that \( t_{n - 1} \leq t_c \leq t_{n} \); in plain English, the final state of the step is --- by definition --- always reached \emph{before} the final sample of the step is taken but after the next-to-last sampling instant.
+From the discrete sampling nature of the \ac{WCA} application, it follows that
+
+\begin{equation}\label{eq:tc}
+    t_{n - 1} \leq t_c \leq t_{n}
+\end{equation}
+
+In plain English, the final state of the step is --- by definition --- always reached \emph{before} the final sample of the step is taken but after the next-to-last sampling instant.
 We define the intervals \( t_n - t_c \) and \( t_{n + 1} - t_n \) as the \emph{wait time} and \emph{last sample \ac{RTT}}, respectively, for step \( M \).
@@
 
-\emph{Execution time} and \emph{time-to-feedback} correspond to the two key metrics necessary to characterize the interactions between human behavior and system responsiveness.
+In~\cite{olguinmunoz:impact2021} we used \emph{delay} and \emph{execution time} as our key variables of interest.
+\emph{Delay} corresponded to a set time duration the processing of each frame was padded to --- i.e.
+if during a series of steps delay was set to \( X \) seconds, the feedback for each frame was provided to the user at exactly \( X \) seconds after frame capture.
+The correlation between delay and execution time was then studied.
+Delay, however, was merely an experimental tool, and in the present paper we argue that the underlying metric driving the human behavior is rather \emph{time-to-feedback}.
+If we consider that
+\begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
+    \item \( t_c \) is uniformly distributed in the interval \( [t_{n - 1}, t_{n}] \) (see \cref{eq:tc} above and~\cite{olguinmunoz:impact2021})
+    \item for a step subject to \emph{delay} \( X \), \( t_k - t_{k - 1} = X \) for all \( k \in \{1, 2, \ldots, n\} \)
+\end{enumerate*},
+then it must follow that, \emph{on average}, for all steps subject to a \emph{delay} \( X \):
 
+\begin{equation}\label{eq:ttf}
+    \text{\emph{time-to-feedback}} = \frac{3}{2}X
+\end{equation}
+
+Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use time-to-feedback instead of delay by simply multiplying the delay value of each step by \( 1.5 \).
+
+\newpage
 \begin{figure}

### 2022-08-25 background.tex
comments

@@
-\section{Background \& Related Work}\label{sec:background}
+\section{A background on \ac{WCA}}\label{sec:background}
+
+In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
+We employed a modified and instrumented version of the LEGO Assistant application devised by~\textcite{Chen2015LEGO}, which belongs to a category of \acp{WCA} with the goal of guiding a user through a sequential task.
+\todo[inline]{Talk more generally about step-based-WCA, time-to-feedback in more general sense.}
+\todo[inline]{WCA tracks progress autonoumosly, provides feedback.}
+\todo[inline]{Token based flow, zero-wait sampling, discarded samples.}
+The application monitors the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
+Whenever the applications detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+Subjects interacted with this \ac{WCA} while the responsiveness of the system was altered in realtime and we captured key application and task performance metrics.
+Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
+
+From this experimentation, we identified four main results concerning the step execution times of humans in \ac{WCA}:
+
+\begin{enumerate}
+    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
+    Moreover, users become progressively slower the longer they spend in a degraded system state.
+
+    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
+
+    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
+
+    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
+\end{enumerate}
+\todo[inline]{Needs more precision.}
+
+In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+In order to accurately emulate the behavior of a human, such a model for needs to implement two main behaviors.
+One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
+And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
+
+\subsection{Formal definitions}
+\todo[inline]{Reorganize to beginning of section}
+
+We begin by providing some definitions relating to step-based \acl{WCA}.
+First of all, a \emph{step} is understood as a specific action to be performed by the user, described by a single instruction; a \emph{task} consists of a series of steps to be executed in sequence (see \cref{fig:task}).
+More formally, a step starts when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided.
+We call the time interval between these two events the \emph{step duration}.
+
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals at which the \ac{WCA} captures the state of the physical system --- see \cref{fig:step}.
+\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample for said step is taken, and \( t_{n+1} \) to the instant at which the instruction for step \( M + 1 \) is provided.
+Furthermore, \( t_c \) represents the instant at which the user finishes performing the instruction for step \( M \); the interval \( t_c - t_0 \) we call the \emph{execution time} of \( M \).
+From the discrete sampling nature of the \ac{WCA} application, it follows that (\( U(a, b) \) represents the continuous uniform distribution in the open interval \( (a, b) \))
+
+\begin{align}\label{eq:tc}
+    t_c &\in U(t_{n - 1}, t_n)
+\end{align}
+
+In plain English, the user will always finish the instruction at some instant \( t_c \) which can be assumed to be randomly distributed between the next-to-last sampling instant \( t_{n - 1} \) and the last sample of a step, \( t_n \).
+
+We further define the intervals \( t_n - t_c \) and \( t_{n + 1} - t_n \) as the \emph{wait time} and \emph{last sample \ac{RTT}}, respectively, for step \( M \).
+The sum of these two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{time-to-feedback}.
+
+In~\cite{olguinmunoz:impact2021} we used \emph{delay} and \emph{execution time} as our key variables of interest.
+\emph{Delay} corresponded to a set time duration the processing of each frame was padded to --- i.e.
+if during a series of steps delay was set to \( X \) seconds, the feedback for each frame was provided to the user at exactly \( X \) seconds after frame capture.
+The correlation between delay and execution time was then studied.
+Delay, however, was merely an experimental tool, and in the present paper we argue that the underlying metric driving  human behavior is actually \emph{time-to-feedback}.
+If we consider that
+\begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
+    \item \( t_c \) is uniformly distributed in the interval \( [t_{n - 1}, t_{n}] \) (see~~\cite{olguinmunoz:impact2021} and \cref{eq:tc} above)
+    \item for a step subject to \emph{delay} \( X \), \( t_k - t_{k - 1} = X \) for all \( k \in \{1, 2, \ldots, n\} \)
+\end{enumerate*},
+then it must follow that, \emph{on average}, for all steps subject to a \emph{delay} \( X \):
+
+\begin{equation}\label{eq:ttf}
+    \text{\emph{time-to-feedback}} = \frac{3}{2}X
+\end{equation}
+
+Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use time-to-feedback instead of delay by simply multiplying the delay value of each step by \( 1.5 \) without loss of accuracy.
+
+\newpage

### 2022-08-25 intro.tex
comments

\section{Introduction \& Related Work}\label{sec:intro}

### 2022-08-25 model.tex
comments

--- 2022-08-24 model.tex
@@
-\section{A model of human behavior for \acl*{WCA}}\label{sec:model}
-
-In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
-We employed a modified and instrumented version of the LEGO Assistant application devised by~\textcite{Chen2015LEGO}, which belongs to a category of \acp{WCA} with the goal of guiding a user through a sequential task.
-The application monitors the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
-Whenever the applications detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
-Subjects interacted with this \ac{WCA} while the responsiveness of the system was altered in realtime and we captured key application and task performance metrics.
-Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
-
-From this experimentation, we identified four main results concerning the step execution times of humans in \ac{WCA}:
-
-\begin{enumerate}
-    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
-    Moreover, users become progressively slower the longer they spend in a degraded system state.
-
-    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
-
-    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
-
-    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
-\end{enumerate}
-
-In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
-In order to accurately emulate the behavior of a human, such a model for needs to implement two main behaviors.
-One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
-And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
-
-\subsection{Generating realistic timings}
-
-We begin by providing some definitions relating to step-based \acl{WCA}.
-First of all, a \emph{step} is understood as a specific action to be performed by the user, described by a single instruction; a \emph{task} consists of a series of steps to be executed in sequence (see \cref{fig:task}).
-More formally, a step starts when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided.
-We call the time interval between these two events the \emph{step duration}.
-
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals at which the \ac{WCA} captures the state of the physical system --- see \cref{fig:step}.
-\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample for said step is taken, and \( t_{n+1} \) to the instant at which the instruction for step \( M + 1 \) is provided.
-Furthermore, \( t_c \) represents the instant at which the user finishes performing the instruction for step \( M \); the interval \( t_c - t_0 \) we call the \emph{execution time} of \( M \).
-From the discrete sampling nature of the \ac{WCA} application, it follows that \( t_{n - 1} \leq t_c \leq t_{n} \); in plain English, the final state of the step is --- by definition --- always reached \emph{before} the final sample of the step is taken but after the next-to-last sampling instant.
-We define the intervals \( t_n - t_c \) and \( t_{n + 1} - t_n \) as the \emph{wait time} and \emph{last sample \ac{RTT}}, respectively, for step \( M \).
-The sum of these two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{time-to-feedback}.
-
-\emph{Execution time} and \emph{time-to-feedback} correspond to the two key metrics necessary to characterize the interactions between human behavior and system responsiveness.
+\section{A model of human behavior for \ac{WCA}}\label{sec:model}
 
@@
 \todo[inline]{Diagram of model?}
+\todo[inline]{Verification: do we need the model? (Compare to naive)}
 

### 2022-08-26 background.tex
reformulating section 3

--- 2022-08-04 background.tex
@@
-\section{Background \& Related Work}\label{sec:background}
+\section{A background on \acs*{WCA}}\label{sec:background}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{figs/task.png}
+        \caption{%
+            Overview of a task in a \ac{WCA}, composed of a series of steps.
+            Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
+            The \ac{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
+        }\label{fig:task}
+    \end{subfigure}\\
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{figs/step_time.png}
+        \caption{%
+            Breakdown of a step into its timing components.
+            The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
+            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \ac{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
+        }\label{fig:step}
+    \end{subfigure}
+    \caption{}
+\end{figure}
+
+\acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \ac{AR} applications.
+In this work we focus on a particular category of these applications --- denoted ``step-based'' \ac{WCA} --- which have as their ultimate goal the guiding of a user through a sequential task.
+Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
+
+Step-based \acp{WCA} operate analogously to how \ac{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
+The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
+Whenever the system detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.
+because they captured an intermediate or unfinished state, or because they contain noise due to user movement) are silently discarded.
+Herein lies one of key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
+
+In order to discuss these applications with precision and fully understand them, we provide some definitions and explanations relating to their operation.
+First of all, a \emph{step} is formally understood as a specific action to be performed by the user, described by a single instruction, and a \emph{task} consists of a series of steps to be executed in sequence (see \cref{fig:task}).
+A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided; we call the time interval between these two events the \emph{step duration}.
+
+Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{zero-wait} sampling, where a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
+This means that
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item the interval between these sampling instants is not necessarily constant, as it is subject to fluctuations due to resource contention on both the network and compute side
+    \item an acknowledgement must be provided even for those samples which do not cause the generation of a new instruction --- this acknowledgement is simply used for flow control and does not generate user-noticeable feedback
+\end{enumerate*}.
+
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals at which the \ac{WCA} captures the state of the physical system --- see \cref{fig:step}.
+\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.
+which captures the final state of step \( M \)) is taken.
+\( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
+If we define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), it must follow that (see also~\cite{olguinmunoz:impact2021})\footnote{\( U(a, b) \) represents the continuous uniform distribution in the open interval \( (a, b) \).}
+\begin{align}\label{eq:tc}
+    t_c &\thicksim U(t_{n - 1}, t_n)
+\end{align}
+
+We define the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
+The sum of the latter two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{time-to-feedback}.
+Once again, refer to \cref{fig:step} for a visual breakdown of these values.
+
+\subsection{The effects of changes in responsiveness on human behavior}
+
+In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
+We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA} devised by~\textcite{Chen2015LEGO},
+Subjects interacted with the assistant while the responsiveness of the system was altered in realtime and we captured key application and task performance metrics.
+Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
+
+We used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
+\emph{Delay} corresponded to a temporarily fixed time duration the processing of each frame was padded to --- i.e.
+if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
+The correlation between this variable and execution time was then studied.
+Delay, however, was merely an experimental tool, and if we consider
+\begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
+    \item \cref{eq:tc} above
+    \item that for a step subject to \emph{delay} \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
+\end{enumerate*},
+then it must follow that, \emph{on average}, for \emph{all} steps subject to a \emph{delay} \( D \):
+
+\begin{equation}\label{eq:ttf}
+    \text{\emph{time-to-feedback}} = \frac{3}{2}D
+\end{equation}
+
+Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use time-to-feedback instead of delay by simply multiplying the delay value of each step by \( 1.5 \) without loss of accuracy.
+We can thus summarize the results from~\cite{olguinmunoz:impact2021} as follows:
+
+\begin{enumerate}
+    \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
+    Moreover, users become progressively slower the longer they spend in a degraded system state.
+
+    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
+
+    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
+
+    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
+\end{enumerate}
+\todo[inline]{Needs more precision.}
+
+In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+In order to accurately emulate the behavior of a human, such a model for needs to implement two main behaviors.
+One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
+And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
+
+\newpage

### 2022-08-26 model.tex
reformulating section 3

@@
 \section{A model of human behavior for \ac{WCA}}\label{sec:model}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}{\columnwidth}
-        \centering
-        \includegraphics[width=\columnwidth]{figs/task.png}
-        \caption{%
-            Overview of a task in a \ac{WCA}, composed of a series of steps.
-            Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
-            The \ac{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
-        }\label{fig:task}
-    \end{subfigure}\\
-    \begin{subfigure}{\columnwidth}
-        \centering
-        \includegraphics[width=\columnwidth]{figs/step_time.png}
-        \caption{%
-            Breakdown of a step into its timing components.
-            The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
-            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \ac{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
-        }\label{fig:step}
-    \end{subfigure}
-    \caption{}
-\end{figure}
 

### 2022-08-26 background.tex
model

@@
 The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
-Whenever the system detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+Whenever the assistant application detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
 The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.
-because they captured an intermediate or unfinished state, or because they contain noise due to user movement) are silently discarded.
+because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
 Herein lies one of key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
@@
 
-Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{zero-wait} sampling, where a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
-This means that
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the interval between these sampling instants is not necessarily constant, as it is subject to fluctuations due to resource contention on both the network and compute side
-    \item an acknowledgement must be provided even for those samples which do not cause the generation of a new instruction --- this acknowledgement is simply used for flow control and does not generate user-noticeable feedback
-\end{enumerate*}.
+Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{zero-wait} sampling.
+In this scheme, a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
+This means that the interval between these sampling instants is not necessarily constant, as it is subject to fluctuations due to resource contention on both the network and compute side.
+It also has as a consequence that acknowledgements must be provided even for those samples which do not cause the generation of a new instruction --- these acknowledgements are simply used for flow control and do not generate user-noticeable feedback.
 
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals at which the \ac{WCA} captures the state of the physical system --- see \cref{fig:step}.
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals (following a zero-wait scheme) at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
 \( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.
@@
 We define the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
-The sum of the latter two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{time-to-feedback}.
+The sum of the latter two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{\ac{TTF}}.
 Once again, refer to \cref{fig:step} for a visual breakdown of these values.
@@
 In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
-We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA} devised by~\textcite{Chen2015LEGO},
-Subjects interacted with the assistant while the responsiveness of the system was altered in realtime and we captured key application and task performance metrics.
+We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA} devised by \textcite{Chen2015LEGO}.
+Subjects interacted with the assistant while we altered the responsiveness in realtime and captured key application and task performance metrics.
 Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
 
-We used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
-\emph{Delay} corresponded to a temporarily fixed time duration the processing of each frame was padded to --- i.e.
+For this work, we used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
+\emph{Delay} corresponded to a temporarily fixed time duration of the processing of each frame was padded to --- i.e.
 if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
@@
     \item \cref{eq:tc} above
-    \item that for a step subject to \emph{delay} \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
+    \item that for a step subject to delay \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
 \end{enumerate*},
-then it must follow that, \emph{on average}, for \emph{all} steps subject to a \emph{delay} \( D \):
+then it follows that, \emph{on average}, for \emph{all} steps subject to a delay \( D \):
 
-\begin{equation}\label{eq:ttf}
-    \text{\emph{time-to-feedback}} = \frac{3}{2}D
-\end{equation}
+\begin{align}\label{eq:ttf}
+    \text{\emph{\ac{TTF}}} = \frac{3}{2}D
+\end{align}
 
-Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use time-to-feedback instead of delay by simply multiplying the delay value of each step by \( 1.5 \) without loss of accuracy.
-We can thus summarize the results from~\cite{olguinmunoz:impact2021} as follows:
+Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5} without loss of accuracy.
 
+With this in mind, we can reformulate the results this work as follows:
 \begin{enumerate}
     \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
-    Moreover, users become progressively slower the longer they spend in a degraded system state.
+    Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \ac{TTF} of \SI{2.475}{\second}, and a further \num{14} percentage points slower at a mean \ac{TTF} of \SI{4.5}{\second}.
+    Users also become progressively slower the longer they spend in a degraded system state.
 
     \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
+    These effects are noticeable for at least \num{4} steps after the return to a high-responsiveness state.
 
-    \item Humans get faster at performing steps as the task progresses. However, this effect is dampened by reduced system responsiveness, even disappearing at higher levels of system impairment.
+    \item Humans get faster at performing steps as the task progresses.
+    For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.
+    However, this effect is dampened by reduced system responsiveness, and disappears at the highest levels of system impairment.
 
-    \item The above effects seem to be modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
+    \item The above effects are modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}, as evidenced by a \ac{PCA}.
 \end{enumerate}
-\todo[inline]{Needs more precision.}
 
 In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
-In order to accurately emulate the behavior of a human, such a model for needs to implement two main behaviors.
-One, it needs to generate realistic \emph{execution times} for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
-And two, it needs to produce sequences of video frames for each step which mimic what a human generate.
-
-\newpage

### 2022-08-26 model.tex
model

@@
 \section{A model of human behavior for \ac{WCA}}\label{sec:model}
+
+In order to accurately emulate the behavior of a human, a model for \ac{WCA} needs to implement two main behaviors.
+One, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
+And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate.
+
+\subsection{Generating realistic execution times}
+
+For the sub-model for the generation of realistic execution times, we employ a cleaned and re-parameterized copy of the timing data collected for~\cite{olguinmunoz:impact2021}.
+This data consists of a table containing \num{6760} rows, where each row contains data for a single step and consists of four columns:
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item a unique identifier for the participant from which the data was recorded
+    \item the recorded \acl{TTF} for the current step
+    \item the recorded execution time for the current step
+    \item a normalized value for the neuroticism of the participant the data is from
+\end{enumerate*}.
+
+\subsection{Generating realistic frame samples}
 

### 2022-08-28 model.tex
sample data

@@
 For the sub-model for the generation of realistic execution times, we employ a cleaned and re-parameterized copy of the timing data collected for~\cite{olguinmunoz:impact2021}.
-This data consists of a table containing \num{6760} rows, where each row contains data for a single step and consists of four columns:
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item a unique identifier for the participant from which the data was recorded
-    \item the recorded \acl{TTF} for the current step
-    \item the recorded execution time for the current step
-    \item a normalized value for the neuroticism of the participant the data is from
-\end{enumerate*}.
+The data consists of timing information for each of the \num{169} steps performed by the \num{40} participants in the human-subject trials.
+The resulting \num{6760} data points are arranged in a table, a sample of which can be seen in \cref{tab:data:exectime}, together with identifiers for the subjects and a normalized value of their neuroticism.
+
+\begin{table}
+\centering
+\caption[A]{%
+    Sample of the execution time data  used for the elaboration of the timing model.
+    Rows contain the execution times and \acp{TTF} for each of the \num{169} steps performed by the \num{40} participants in \textcite{olguinmunoz:impact2021}, expressed in seconds, together with a unique identifier for the subject, their associated level of normalized neuroticism, and a sequence number identifying the position of the step in the task.
+}\label{tab:data:exectime}
+\begin{tabular}{rrrrrr}
+    \toprule
+    {seq} & {run\_id} & {exec\_time} & {ttf} & {neuroticism}\\
+    \midrule
+    1 & 134146 & 3.655 & 0.597 & 0.375\\
+    2 & 134146 & 4.439 & 0.554 & 0.375\\
+    3 & 134146 & 2.943 & 0.562 & 0.375\\
+    \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) \\
+    169 & 134146 & 3.532 & 4.571 & 0.375\\
+    1 & 134470 & 3.461 & 0.561 & 0.594\\
+    2 & 134470 & 6.678 & 0.572 & 0.594\\
+    \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) \\
+    \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) \\
+    169 & 137353 & 4.615 & 0.537 & 0.625\\
+    \bottomrule
+    \end{tabular}
+\end{table}
 

### 2022-08-29 background.tex
plotting model

@@
 
-\subsection{The effects of changes in responsiveness on human behavior}
+\subsection{The effects of changes in responsiveness on human behavior}\label{ssec:plos}
 
@@
 
-    \item The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
+    \item\label{item:remain} The effects of system slow-down on human behavior remain for a while even after system responsiveness improves.
     These effects are noticeable for at least \num{4} steps after the return to a high-responsiveness state.

### 2022-08-29 model.tex
plotting model

@@
 
-For the sub-model for the generation of realistic execution times, we employ a cleaned and re-parameterized copy of the timing data collected for~\cite{olguinmunoz:impact2021}.
-The data consists of timing information for each of the \num{169} steps performed by the \num{40} participants in the human-subject trials.
-The resulting \num{6760} data points are arranged in a table, a sample of which can be seen in \cref{tab:data:exectime}, together with identifiers for the subjects and a normalized value of their neuroticism.
+As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
+For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
+The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
+The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
+These distributions can then be sampled to produce new, realistic execution times.
+In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.
 
@@
 
-\subsection{Generating realistic frame samples}
+We employ a cleaned and re-parameterized copy of the aforementioned timing data.
+The \num{6760} data points are arranged in a table, a sample of which can be seen in \cref{tab:data:exectime}, together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
+
+We begin by discretizing neuroticism and \ac{TTF} each into terciles; we will refer to the first, second, and third terciles of neuroticism as \emph{low}, \emph{medium}, and \emph{high} neuroticism respectively, and to the corresponding terciles of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+
+The next steps of processing are slightly more complex.
+For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \ac{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
+This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
+Users notice the impairment of the current step only when they have finished it, and thus system impairment has no effect on the execution time of the current step, but directly affects the next one.
+
+We then assign a label to each sample indicating the class of the most recent transition in impairment level in the run.
+In other words, for each recorded step, we look back until we find a step with a different level of impairment and assign a label to the current step according to whether that change was from a higher to a lower level of impairment or vice-versa.
+However, if the current step is too far from the latest transition in impairment (by default, more than \num{8} steps), it is instead marked as \emph{no transition}; this label is also used for steps at the beginning of each run which have no previous transition in impairment.
+This labeling allows us to capture the behavior described in \cref{item:remain} of \cref{ssec:plos}.
+
+Finally, we assign a number to each sample corresponding to \emph{how long} the current combination of level of impairment and transition has been in effect, in number of steps.
+This count resets whenever this combination changes.
+This value is subsequently discretized into bins with edges \( [0, 5) \), \( [5, 10) \), and \( [10, \infty) \); we will refer to these discretized levels as \emph{short}, \emph{medium}, and \emph{long duration}, respectively.
+
+The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}
+
+\begin{table*}
+\caption{Sample of the processed execution time data table.}\label{tab:data:exectime:processed}
+\begin{tabular}{rrrrcrccrc}
+\toprule
+{seq} & {run\_id} & {next\_exec\_time} & {ttf} & {impairment} & {neuroticism\_raw} & {neuroticism} & {transition} & {duration\_raw} & {duration} \\
+\midrule
+1   & 134146 & 3.655 & 0.000 & low & 0.375  & low& NoTransition & 1 & short \\
+2   & 134146 & 4.439 & 0.597 & low & 0.375 & low & NoTransition & 2 & short \\
+3   & 134146 & 2.943 & 0.554 & low & 0.375 & low & NoTransition & 3 & short \\
+\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
+169 & 134146 & 3.532 & 3.689 & high & 0.375 & low & Lower2Higher & 2 & short \\
+1   & 134470 & 3.461 & 0.000 & low & 0.594 & med & NoTransition & 1 & short \\
+2   & 134470 & 6.678 & 0.561 & low & 0.594 & med & NoTransition & 2 & short \\
+\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
+\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
+169 & 137353 & 4.615 & 0.579 & low & 0.625 & med & NoTransition & 2 & short \\
+\bottomrule
+\end{tabular}
+\end{table*}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{./model_data/duration_vs_exectime.png}
+    \caption{}\label{fig:timing:durvsetime}
+\end{figure}
 
@@
 
-This data is first preprocessed and then sampled according to a series of rules in order to produce novel,realistic sequences of execution times for a task.
-Alternatively, instead of sampling directly from the preprocessed data, a theoretical distribution can be fitted, thus allowing for generation of realistic synthetic execution times.
-
-\subsubsection{Data pre-processing}
-
-A sample of the step timing data obtained for~\cite{olguinmunoz:impact2021} is presented in \cref{tab:exec_times:empirical}.
-This data corresponds to real step execution times for \num{40} repetitions, by different human subjects, of the same \num{169}-step task.
-During each repetition of the task, system responsiveness was altered in real time by setting the round-trip-times of all frames to a fixed value (corresponding to the \emph{delay} column in \cref{tab:exec_times:empirical}).
-For more details on the experimental design and data collection procedure, please see the original article~\cite{olguinmunoz:impact2021}.
-
-We process this data in the following manner:
-
-\input{tables/processed_exec_times.tex}
-
-\begin{enumerate}
-    \item Normalized neuroticism values are binned into predefined, contiguous bins.
-    We use three bins with edges
-    \( \left(-\infty, \frac{1}{3}\right) \),
-    \( \left[\frac{1}{3}, \frac{2}{3}\right) \),
-    and \( \left[\frac{2}{3}, \infty\right) \),
-    as these give the most even distribution of number of samples across bins.
-
-    \item System impairment (\emph{delay}) is binned into predefined, contiguous bins.
-    Once again, we use three bins giving a somewhat even distribution of samples:
-    \( \left(-\infty, 1.0\right) \),
-    \( \left[1.0, 2.0\right) \),
-    and \( \left[2.0, \infty\right) \) seconds.
-
-    \item A predefined number of steps (by default, \num{8}) immediately after each change in binned impairment are marked according to the relative change in impairment: \emph{low-to-high}, or \emph{high-to-low} impairment.
-    We label this value the \emph{transition tag}.
-    Steps at the beginning of the task (where there is no previous impairment), or too far away from a transition in impairment, are tagged as \emph{no transition}.
-
-    \item Each step in a sequence tagged with the same transition tag is assigned a sequential \emph{duration} number.
-    This value tracks how long the system has been in a specific state, and is in turn binned into three bins:
-    \( \left[0, 5\right) \),
-    \( \left[5, 10\right) \),
-    and \( \left[10, \infty\right) \) steps.
-
-    \item Finally, execution times are shifted by one step.
-    This is because the execution time of a step depends on the state of the previous step, and not the current one.
-    As explained previously, humans only becomes aware of the responsiveness of the system when they finish a step and have to wait for feedback from the system.
-\end{enumerate}
-
-\subsection{Generating a synthetic trace of frames}
-
 \subsection{Model implementation}

### 2022-08-29 model.tex
workng on model ;

@@
 
-\begin{figure}
+\begin{figure*}
     \centering
-    \includegraphics[width=\columnwidth]{./model_data/duration_vs_exectime.png}
+    \begin{subfigure}[t]{\columnwidth}
+        \centering
+        \includegraphics[width=.8\columnwidth]{./model_data/duration_vs_exectime.png}
         \caption{}\label{fig:timing:durvsetime}
-\end{figure}
+    \end{subfigure}%
+    \begin{subfigure}[t]{\columnwidth}
+        \centering
+        \includegraphics[width=.8\columnwidth]{./model_data/impairment_transition_vs_exectime.png}
+        \caption{}\label{fig:timing:imptransvsetime}
+    \end{subfigure}
+    \caption{\cref{fig:timing:durvsetime} shows how the model captures the speed-up and slow-down behaviors due to impairment and duration described in \textcite{olguinmunoz:impact2021}}
+\end{figure*}
 

### 2022-08-30 model.tex
plotting

@@
 
-We begin by discretizing neuroticism and \ac{TTF} each into terciles; we will refer to the first, second, and third terciles of neuroticism as \emph{low}, \emph{medium}, and \emph{high} neuroticism respectively, and to the corresponding terciles of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, \( [0.0, 0.5) \) and \( [0.5, 1.0) \) respectively. 
+We also bin \ac{TTF} into terciles; we will refer to the first, second, and third terciles of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
 
@@
 
-The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}
+The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}.
+Three examples of the effect of this discretization and processing on the data are also presented in \cref{fig:timing}.
 
@@
 \midrule
-1   & 134146 & 3.655 & 0.000 & low & 0.375  & low& NoTransition & 1 & short \\
+1   & 134146 & 3.655 & 0.000 & low & 0.375  & low & NoTransition & 1 & short \\
 2   & 134146 & 4.439 & 0.597 & low & 0.375 & low & NoTransition & 2 & short \\
@@
 169 & 134146 & 3.532 & 3.689 & high & 0.375 & low & Lower2Higher & 2 & short \\
-1   & 134470 & 3.461 & 0.000 & low & 0.594 & med & NoTransition & 1 & short \\
-2   & 134470 & 6.678 & 0.561 & low & 0.594 & med & NoTransition & 2 & short \\
+1   & 134470 & 3.461 & 0.000 & low & 0.594 & high & NoTransition & 1 & short \\
+2   & 134470 & 6.678 & 0.561 & low & 0.594 & high & NoTransition & 2 & short \\
 \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
 \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
-169 & 137353 & 4.615 & 0.579 & low & 0.625 & med & NoTransition & 2 & short \\
+169 & 137353 & 4.615 & 0.579 & low & 0.625 & high & NoTransition & 2 & short \\
 \bottomrule
@@
     \centering
+    \begin{subfigure}[t]{\columnwidth}
+        \centering
+        \includegraphics[width=.8\columnwidth]{./model_data/imp_neur_vs_exectime.png}
+        \caption{}\label{fig:timing:impneurvsetime}
+    \end{subfigure}\\
     \begin{subfigure}[t]{\columnwidth}
@@
     \end{subfigure}
-    \caption{\cref{fig:timing:durvsetime} shows how the model captures the speed-up and slow-down behaviors due to impairment and duration described in \textcite{olguinmunoz:impact2021}}
+    \caption{%
+        Visualizations of the model data after discretization and post-processing.
+        \subref{fig:timing:impneurvsetime} illustrates the effect of impairment on execution times, and how neuroticism modulates this behavior;
+        \subref{fig:timing:durvsetime} exemplifies how the model captures the speed-up and slow-down behaviors due to impairment and duration described in \textcite{olguinmunoz:impact2021};
+        finally, \subref{fig:timing:imptransvsetime} shows the lingering effects of impairment after a transition.
+    }\label{fig:timing}
 \end{figure*}

### 2022-08-30 model.tex
tweaking

@@
 
-We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, \( [0.0, 0.5) \) and \( [0.5, 1.0) \) respectively. 
-We also bin \ac{TTF} into terciles; we will refer to the first, second, and third terciles of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, \( [0.0, 0.5) \) and \( [0.5, 1.0] \) respectively. 
+We also bin \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
 
@@
 
-\begin{algorithm}
-    \caption{Timing model algorithm}
-    \KwIn{%
-        \begin{itemize}
-            \item $T_\text{start}$: start timestamp of the previous step.
-            \item $T_\text{success}$: capture timestamp of the succesful sample of the previous step.
-            \item $T_\text{end}$: end timestamp of the previous step.
-            \item $t_\text{exec}$: execution time (in seconds) of the previous step.
-        \end{itemize}%
-    }
-    \KwOut{$t_\text{exec}'$: execution time for the current step.}
+% \begin{algorithm}
+%     \caption{Timing model algorithm}
+%     \KwIn{%
+%         \begin{itemize}
+%             \item $T_\text{start}$: start timestamp of the previous step.
+%             \item $T_\text{success}$: capture timestamp of the succesful sample of the previous step.
+%             \item $T_\text{end}$: end timestamp of the previous step.
+%             \item $t_\text{exec}$: execution time (in seconds) of the previous step.
+%         \end{itemize}%
+%     }
+%     \KwOut{$t_\text{exec}'$: execution time for the current step.}
 
-    \Comment{%
-        Initial values:\\
-        $d \leftarrow 0$\\
-        $I_\text{prev} \leftarrow \text{NULL}$
-    }
-    $t_\text{wait} \leftarrow T_\text{success} - (T_\text{start} + t_\text{exec})$\;
-    $t_\text{round-trip} \leftarrow T_\text{end} - T_\text{success}$\;
-    $i \leftarrow t_\text{wait} + t_\text{round-trip}$ \Comment*[r]{raw impairment}
-    $I \leftarrow \text{BIN}(i)$ \Comment*[r]{bin impairment into\\discrete levels}
+%     \Comment{%
+%         Initial values:\\
+%         $d \leftarrow 0$\\
+%         $I_\text{prev} \leftarrow \text{NULL}$
+%     }
+%     $t_\text{wait} \leftarrow T_\text{success} - (T_\text{start} + t_\text{exec})$\;
+%     $t_\text{round-trip} \leftarrow T_\text{end} - T_\text{success}$\;
+%     $i \leftarrow t_\text{wait} + t_\text{round-trip}$ \Comment*[r]{raw impairment}
+%     $I \leftarrow \text{BIN}(i)$ \Comment*[r]{bin impairment into\\discrete levels}
 
-    \If{$I_\text{prev} == \text{NULL}$}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{NO TRANSITION}$\;
-    } \ElseIf{$I_\text{prev} < I$}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{LOWER TO HIGHER} $
-    } \ElseIf{$I_\text{prev} > I$}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{HIGHER TO LOWER} $  
-    }
+%     \If{$I_\text{prev} == \text{NULL}$}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{NO TRANSITION}$\;
+%     } \ElseIf{$I_\text{prev} < I$}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{LOWER TO HIGHER} $
+%     } \ElseIf{$I_\text{prev} > I$}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{HIGHER TO LOWER} $  
+%     }
 
-\end{algorithm}
+% \end{algorithm}
 

### 2022-08-31 model.tex
algorithm

@@
 Three examples of the effect of this discretization and processing on the data are also presented in \cref{fig:timing}.
+Finally, in \cref{alg:timings} we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
+In plain English, a model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
+At the beginning of each step, the model
+\begin{enumerate}
+    \item Checks whether any change in impairment has occurred.
+    If it has, it resets duration to \num{1} and updates the transition variable accordingly.
+    If not, it increases the duration variable by one unless that increase causes it to exceed the threshold of steps required to forget a transition.
+    If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
+    \item Discretizes the updated duration value.
+    \item Finds all samples with binned neuroticism, impairment, duration, and transition values matching the current state of the system.
+    \item Randomly samples the filtered data to obtain a realistic execution time value.
+    \begin{itemize}
+        \item Alternatively, a distribution can be first be fitted to the filtered samples.
+        The model then samples the theoretical distribution instead of the empirical samples.
+    \end{itemize}
+    \item Finally, updates the stored impairment value with the new impairment and proceeds to the next step.
+\end{enumerate}
+
+\begin{algorithm}
+    \caption{Generating realistic execution times}\label{alg:timings}
+
+    \SetKwInput{KwState}{Initial State}
+
+    \KwIn{$\tau$: \ac{TTF} of the previous step.}
+    \KwOut{$t_\text{exec}$: next execution time.}
+    \KwState{%
+        \begin{itemize}
+            \item Binned neuroticism $N$.
+            \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
+            \item Duration counter: $d \leftarrow 0$.
+            \item Previous impairment: $I' \leftarrow \varnothing$.
+            \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
+            \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
+        \end{itemize}
+    }
+
+    $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}
+
+    \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{``NO TRANSITION''}$\;
+    } \uElseIf{$I' < I$}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{``LOWER TO HIGHER''} $
+    } \uElseIf{$I' > I$}{%
+        $d \leftarrow 1$\;
+        $T \leftarrow \text{``HIGHER TO LOWER''} $  
+    } \Else {%
+        $d \leftarrow d + 1$\;
+    }
+    $I' \leftarrow I$ \tcc*{store impairment for later}
+    $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}
+
+    \tcc{next, filter the data to select only the matching samples}
+    $E \leftarrow F_\text{data}(N, I, D, T)$ \;
+    \tcc{finally, sample the data to return a realistic execution time}
+    $t_\text{exec} \leftarrow {SAMPLE}(E)$\;
+
+\end{algorithm}
 
@@
 
+\subsection{Generating realistic samples}
+
 \todo[inline]{Diagram of model?}
 \todo[inline]{Verification: do we need the model? (Compare to naive)}
-
-% \begin{algorithm}
-%     \caption{Timing model algorithm}
-%     \KwIn{%
-%         \begin{itemize}
-%             \item $T_\text{start}$: start timestamp of the previous step.
-%             \item $T_\text{success}$: capture timestamp of the succesful sample of the previous step.
-%             \item $T_\text{end}$: end timestamp of the previous step.
-%             \item $t_\text{exec}$: execution time (in seconds) of the previous step.
-%         \end{itemize}%
-%     }
-%     \KwOut{$t_\text{exec}'$: execution time for the current step.}
-
-%     \Comment{%
-%         Initial values:\\
-%         $d \leftarrow 0$\\
-%         $I_\text{prev} \leftarrow \text{NULL}$
-%     }
-%     $t_\text{wait} \leftarrow T_\text{success} - (T_\text{start} + t_\text{exec})$\;
-%     $t_\text{round-trip} \leftarrow T_\text{end} - T_\text{success}$\;
-%     $i \leftarrow t_\text{wait} + t_\text{round-trip}$ \Comment*[r]{raw impairment}
-%     $I \leftarrow \text{BIN}(i)$ \Comment*[r]{bin impairment into\\discrete levels}
-
-%     \If{$I_\text{prev} == \text{NULL}$}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{NO TRANSITION}$\;
-%     } \ElseIf{$I_\text{prev} < I$}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{LOWER TO HIGHER} $
-%     } \ElseIf{$I_\text{prev} > I$}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{HIGHER TO LOWER} $  
-%     }
-
-% \end{algorithm}
 

### 2022-08-31 publications/2021ImpactDelayedResponse/Manuscript.tex
plos

--- 2022-08-19 publications/2021ImpactDelayedResponse/paper.tex
@@
 % BibLatex and PLOS One hacks
-\usepackage[citestyle=numeric,
-            backend=bibtex,
-            bibstyle=authoryear,
-            sorting=none,
-            sortcites,
-            firstinits,
-            terseinits,
-            maxbibnames=6,
-            dashed=false]{biblatex}
-\bibliography{bibliography}
+% \usepackage[citestyle=numeric,
+%             backend=bibtex,
+%             bibstyle=authoryear,
+%             sorting=none,
+%             sortcites,
+%             firstinits,
+%             terseinits,
+%             maxbibnames=6,
+%             dashed=false]{biblatex}
+% \bibliography{bibliography}
+\usepackage{cite}
+
 \usepackage[colorlinks]{hyperref}
@@
 % Name list format
-\renewcommand*{\labelnamepunct}{\addspace}
-\renewcommand*{\finalnamedelim}{%
-  \ifbibliography{\addcomma\space}{\addspace\&\space}}
-\renewcommand*{\revsdnamepunct}{}
-\DeclareNameAlias{sortname}{last-first}
+% \renewcommand*{\labelnamepunct}{\addspace}
+% \renewcommand*{\finalnamedelim}{%
+%   \ifbibliography{\addcomma\space}{\addspace\&\space}}
+% \renewcommand*{\revsdnamepunct}{}
+% \DeclareNameAlias{sortname}{last-first}
 
 % No quotes or italics in titles
-\DeclareFieldFormat{title}{#1}
-\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
-                    thesis,unpublished]{title}{#1}
-\DeclareFieldFormat{journaltitle}{#1}
-\DeclareFieldFormat{issuetitle}{#1}
-\DeclareFieldFormat{maintitle}{#1}
-
-\renewbibmacro*{in:}{%
-  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
+% \DeclareFieldFormat{title}{#1}
+% \DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
+%                     thesis,unpublished]{title}{#1}
+% \DeclareFieldFormat{journaltitle}{#1}
+% \DeclareFieldFormat{issuetitle}{#1}
+% \DeclareFieldFormat{maintitle}{#1}
+
+% \renewbibmacro*{in:}{%
+%   \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
 
 % Remove page prefixes
-\DeclareFieldFormat{pages}{#1}
+% \DeclareFieldFormat{pages}{#1}
 
 % Only print journal volume
-\renewbibmacro*{volume+number+eid}{%
-  \setunit*{\addspace}%
-  \printfield{volume}}
+% \renewbibmacro*{volume+number+eid}{%
+%   \setunit*{\addspace}%
+%   \printfield{volume}}
 
 % Use colon for volume-pages delimiter
-\renewcommand*{\bibpagespunct}{%
-  \ifentrytype{article}{\addcolon\space}{\addcomma\space}}
+% \renewcommand*{\bibpagespunct}{%
+%   \ifentrytype{article}{\addcolon\space}{\addcomma\space}}
 
 % Add labelnumbers to bibliography
-\DeclareFieldFormat{labelnumberwidth}{#1\adddot}
-\defbibenvironment{bibliography}
-  {\list
-     {\printtext[labelnumberwidth]{%
-      \printfield{prefixnumber}%
-      \printfield{labelnumber}}}
-     {\setlength{\labelwidth}{\labelnumberwidth}%
-      \setlength{\leftmargin}{\labelwidth}%
-      \setlength{\labelsep}{\biblabelsep}%
-      \addtolength{\leftmargin}{\labelsep}%
-      \setlength{\itemsep}{\bibitemsep}%
-      \setlength{\parsep}{\bibparsep}}%
-      \renewcommand*{\makelabel}[1]{\hss##1}}
-  {\endlist}
-  {\item}
+% \DeclareFieldFormat{labelnumberwidth}{#1\adddot}
+% \defbibenvironment{bibliography}
+%   {\list
+%      {\printtext[labelnumberwidth]{%
+%       \printfield{prefixnumber}%
+%       \printfield{labelnumber}}}
+%      {\setlength{\labelwidth}{\labelnumberwidth}%
+%       \setlength{\leftmargin}{\labelwidth}%
+%       \setlength{\labelsep}{\biblabelsep}%
+%       \addtolength{\leftmargin}{\labelsep}%
+%       \setlength{\itemsep}{\bibitemsep}%
+%       \setlength{\parsep}{\bibparsep}}%
+%       \renewcommand*{\makelabel}[1]{\hss##1}}
+%   {\endlist}
+%   {\item}
 
 % Omit authoryear disambiguation
-\AtEveryBibitem{\clearfield{extrayear}}
+% \AtEveryBibitem{\clearfield{extrayear}}
 
@@
 
-Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\autocite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
+Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
 They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
@@
 Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
-An example is an \textcite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
-
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\autocite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\autocite{Chen:AnEmpiricalStudyOfLatency}.
-Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\autocite{funk2015caworkplace,gorecky2011cognito}.
+An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
+
+WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
+More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
+Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
 Detethering this assistance from its current fixed location will surely make it available to many more fields.
@@
     In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
-    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\autocite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
+    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
 \end{enumerate}
@@
 Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\autocite{dabrowsky:2011:40years}.
+It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}.
 Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
@@
 This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
-We expand upon the findings of \textcite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of \textcite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
+We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
 While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
@@
 Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
-The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\autocite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
+The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
 
@@
 
-  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\autocite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
-  We also expected \emph{focus} and \emph{involvement}~\autocite{witmer1998:itq} to play roles in this.
+  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
+  We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
 
@@
 % To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\autocite{john1999:bfi}.
+% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
 
@@
 These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
-Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \textcite{olguin:2018, olguin:2019}.
+Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
 Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
@@
 % Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
-% In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
+% In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
 
 Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
-In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
+In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
 Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
@@
 The question of how people respond to delay in a computer system is grounded in how people perceive time.
-Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\autocite{zakay1995attentional,zakay1996role}.
-More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\autocite{bruno2016multiple,wiener2011multiple}.
+Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
+More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
 The estimation of an event's duration varies with context of various types
 \begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-    \item events subsequent to a long or short interval are contracted or extended, respectively~\autocite{heron2012duration}
-    \item repeated events tend to be perceived as shorter than novel ones~\autocite{matthews2011stimulus}
-    \item arousal can expand durations~\autocite{droit2011emotion}.
+    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
+    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
+    \item arousal can expand durations~\cite{droit2011emotion}.
 \end{enumerate*}
 
-Expectations play a critical role in time perception as well~\autocite{zakay1995attentional,zakay1996role}.
-It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\autocite{Loewenstein1992anomaliesintertemporalchoice}.
+Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
+It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{Loewenstein1992anomaliesintertemporalchoice}.
 Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
@@
 
-It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\autocite{dabrowsky:2011:40years}.
-However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\autocite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
+It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowsky:2011:40years}.
+However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
 Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.
 
-For WCA the acceptable range for latencies was explored by \textcite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+For WCA the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
 They then proposed a latency tolerance zone according to the task demands.
 For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
-Beyond that, users could begin to show the negative outcomes previously catalogued~\autocite{dabrowsky:2011:40years}.
+Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowsky:2011:40years}.
 
@@
 As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\autocite{neves1981knowledge}.
-Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\autocite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\autocite{lewis2003distinct,koch2009neural,lee2019limiting}.
+Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
+Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
 
 Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-Delays, in contrast, would disrupt such a plan through the loss of automated control~\autocite{lee2019limiting}.
+Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
 
 A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\autocite{lee2019limiting}.
-
-Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\autocite{baumeister74tice,lin2020strong}.
+As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
+
+Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
 
@@
 Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-Arousal can also reduce movement smoothness or add erratic gestures~\autocite{pijpers2003anxiety}.
-Ego depletion has been found to produce premature responses culminating in error~\autocite{lin2020strong}, or to lead to abandoning a task entirely~\autocite{baumeister74tice}.
+Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
+Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
 
@@
 High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-\textcite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
+Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
 Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
@@
   \centering
-  \includegraphics[width=.8\textwidth]{images/experiment_arch.pdf}
+%   \includegraphics[width=.8\textwidth]{Fig1.eps}
   \caption{Experimental test-bed.
@@
 
-This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board.
+This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board under record number \texttt{STUDY2019\_00000247}. 
+Written consent was obtained.
 
@@
 
-We used a modified version of the LEGO Assistant application introduced by \textcite{Chen:EarlyImplementation}.
+We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{Chen:EarlyImplementation}.
 This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
@@
   \centering
-  \includegraphics[width=.8\textwidth]{images/step_task_description.pdf}
+%   \includegraphics[width=.8\textwidth]{Fig2.eps}
   \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
@@
   % \input{plots/lego_hierarchy_diagram.tex}
-  \includegraphics[width=\textwidth]{plots/lego_hierarchy_diagram_standalone.eps}
+%   \includegraphics[width=\textwidth]{Fig3.eps}
   \caption{Hierarchical cognitive structure of a step in the LEGO task.}
@@
 The length of this delay is one of two independent variables we manipulated for the experiment.
-Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by \textcite{Chen:AnEmpiricalStudyOfLatency}.
+Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
 % A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
@@
     \centering
-    \includegraphics[width=\textwidth]{images/step_structure.pdf}
+%     \includegraphics[width=\textwidth]{Fig4a.eps}
     \caption{Structure of a step in a generic cognitive assistance application. 
@@
     \centering
-    \includegraphics[width=\textwidth]{images/step_structure_delay.pdf}
+%     \includegraphics[width=\textwidth]{Fig4b.eps}
     \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
@@
     \centering
-    \includegraphics[width=\textwidth]{images/block.pdf}
+%     \includegraphics[width=\textwidth]{Fig4c.eps}
     \caption{Structure of a block in the experimental task.}%
@@
     \centering
-    \includegraphics[width=\textwidth]{images/exec_time_diagram.pdf}
+%     \includegraphics[width=\textwidth]{Fig4d.eps}
     \caption{Visualization of the execution time of a step.}%
@@
 
-The first of these, the Big Five Inventory of Personality (\emph{BFI}~\autocite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
-Of these, extroversion and neuroticism have been related to tolerance for delay~\autocite{hirsh2008delay}.
-
-The second survey, the Immersive Tendencies Questionnaire (\emph{ITQ}~\autocite{witmer1998:itq}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
+The first of these, the Big Five Inventory of Personality (\emph{BFI}~\cite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
+Of these, extroversion and neuroticism have been related to tolerance for delay~\cite{hirsh2008delay}.
+
+The second survey, the Immersive Tendencies Questionnaire (\emph{ITQ}~\cite{witmer1998:itq}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
 These questions use a 7-point horizontal scale  with opposing descriptors anchoring  the ends.
@@
 \subsubsection{Biometric Data}
-The participant wore devices to acquire four physiological measures:
+The participants wore devices to acquire four physiological measures:
 \begin{enumerate*}[itemjoin={{, }},
@@
 
-These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\autocite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
-GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\autocite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
-EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\autocite{Antonenko2010,Grimes2008,kumar2016measurement}.
-
-Wrist acceleration, GSR and heart rate data were obtained using the \textcite{empatica:e4} biosensing wristband.
+These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\cite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
+GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\cite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
+EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\cite{Antonenko2010,Grimes2008,kumar2016measurement}.
+
+Wrist acceleration, GSR and heart rate data were obtained using the Empatica E4~\cite{empatica:e4} biosensing wristband.
 Accelerometer data was sampled at \SI{32}{\hertz},
@@
 
-The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\autocite{ragot2017emotion, mccarthy2016validation}.
+The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\cite{ragot2017emotion, mccarthy2016validation}.
 The E4 also includes a skin temperature thermometer, however the measure was not used for the present study.
 
-For the EEG data we employed the \textcite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
+For the EEG data we employed the OpenBCI EEG Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
 It provides a quick and non-invasive way of obtaining EEG signals from participants.
-Electrodes were placed according to the \emph{10--20 Electrode System}~\autocite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
+Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
 The signals were sampled at \SI{200}{\hertz} and postprocessed to 
@@
   \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
-  \item smooth out noise (using the technique proposed by~\textcite{agarwal2017eeg}).
+  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
 \end{enumerate*}
@@
     \centering
-    \includegraphics[height=140pt]{images/participant.eps}
+%     \includegraphics[height=140pt]{Fig5a.eps}
     \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
@@
     \centering
-    \includegraphics[height=140pt]{images/board_and_hand.eps}
+%     \includegraphics[height=140pt]{Fig5b.eps}
     \caption{Frame from the board recording of the same participant.}
@@
 
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\autocite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
+We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
 % An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
@@
     \centering
-    \includegraphics[width=.8\textwidth]{plots/exec_time_delay_length.eps}
+%     \includegraphics[width=.8\textwidth]{Fig6.eps}
     \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
@@
     \centering
-    \includegraphics[width=.8\textwidth]{plots/exec_time_delay_slice.eps}
+%     \includegraphics[width=.8\textwidth]{Fig7.eps}
     \caption{Mean per-step execution time vs.\ delay, by step slice.
@@
   \centering
-  \includegraphics[width=.8\textwidth]{plots/exec_time_delay_change.eps}
+%   \includegraphics[width=.8\textwidth]{Fig8.eps}
   \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
@@
     \centering
-    \includegraphics[width=.8\textwidth]{plots/acc_vs_delay_per_length.eps}
+%     \includegraphics[width=.8\textwidth]{Fig9.eps}
     \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{fig:acc:delaylength}
@@
   \centering
-  \includegraphics[width=.8\textwidth]{plots/acc_vs_delay_per_slice.eps}
+%   \includegraphics[width=.8\textwidth]{Fig10.eps}
   \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
@@
     \centering
-    \includegraphics[width=.8\textwidth]{plots/eeg_alpha_beta_vs_block_slice.eps}
+%     \includegraphics[width=.8\textwidth]{Fig11.eps}
     \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
@@
     \centering
-    \includegraphics[width=.8\textwidth]{plots/reg_exectime_vs_neuroticism.eps}
+%     \includegraphics[width=.8\textwidth]{Fig12.eps}
     \caption{Correlation between neuroticism score of participants and their execution time in the longest block at the longest delay.
@@
   In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
-  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\autocite{quesnel2018you}.
-  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\autocite{hirsh2008delay}.
+  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
+  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
 \end{itemize}
@@
 Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
-These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\autocite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
+These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
 
@@
 \begin{acks}
-  This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
-
-  Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or the above-mentioned funding sources.
+  % This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
+  Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or funding sources.
 \end{acks}
@@
 
-\printbibliography%
+% \bibliographystyle{plos2015}
+% \bibliography{bibliography}
+\begin{thebibliography}{10}
+
+  \bibitem{Ha:TowardsWearableCogAssist}
+  Ha K, Chen Z, Hu W, Richter W, Pillai P, Satyanarayanan M.
+  \newblock Towards Wearable Cognitive Assistance.
+  \newblock In: Proceedings of the 12th Annual International Conference on Mobile
+    Systems, Applications, and Services. MobiSys '14. New York, NY, USA: ACM;
+    2014. p. 68--81.
+  \newblock Available from: \url{http://doi.acm.org/10.1145/2594368.2594383}.
+
+  \bibitem{Chen:EarlyImplementation}
+  Chen Z, Jiang L, Hu W, Ha K, Amos B, Pillai P, et~al.
+  \newblock Early Implementation Experience with Wearable Cognitive Assistance
+    Applications.
+  \newblock In: Proceedings of the 2015 Workshop on Wearable Systems and
+    Applications. WearSys '15. New York, NY, USA: ACM; 2015. p. 33--38.
+  \newblock Available from: \url{http://doi.acm.org/10.1145/2753509.2753517}.
+
+  \bibitem{IKEAAssistant}
+  {IKEA assistant}; 2020.
+  \newblock Available from: \url{https://www.youtube.com/watch?v=vMzTGgQHjpo}.
+
+  \bibitem{satya2019augmenting}
+  {Satyanarayanan} M, {Davies} N.
+  \newblock {Augmenting Cognition Through Edge Computing}.
+  \newblock Computer. 2019;52(7):37--46.
+  \newblock doi:{10.1109/MC.2019.2911878}.
+
+  \bibitem{Chen:AnEmpiricalStudyOfLatency}
+  Chen Z, Hu W, Wang J, Zhao S, Amos B, Wu G, et~al.
+  \newblock An Empirical Study of Latency in an Emerging Class of Edge Computing
+    Applications for Wearable Cognitive Assistance.
+  \newblock In: Proceedings of the Second ACM/IEEE Symposium on Edge Computing.
+    SEC '17. New York, NY, USA: ACM; 2017. p. 14:1--14:14.
+  \newblock Available from: \url{http://doi.acm.org/10.1145/3132211.3134458}.
+
+  \bibitem{funk2015caworkplace}
+  {Funk} M, {Schmidt} A.
+  \newblock Cognitive Assistance in the Workplace.
+  \newblock IEEE Pervasive Computing. 2015;14(3):53--55.
+
+  \bibitem{gorecky2011cognito}
+  Gorecky D, Worgan SF, Meixner G.
+  \newblock {COGNITO: A Cognitive Assistance and Training System for Manual Tasks
+    in Industry}.
+  \newblock In: Proceedings of the 29th Annual European Conference on Cognitive
+    Ergonomics. ECCE '11. New York, NY, USA: Association for Computing Machinery;
+    2011. p. 53--6.
+  \newblock Available from: \url{https://doi.org/10.1145/2074712.2074723}.
+
+  \bibitem{PingPongAssistant}
+  {Ping Pong assistant}; 2020.
+  \newblock Available from: \url{https://www.youtube.com/watch?v=_lp32sowyUA}.
+
+  \bibitem{dabrowsky:2011:40years}
+  Dabrowski J, Munson EV.
+  \newblock {40 years of Searching for the Best Computer System Response Time}.
+  \newblock Interact Comput. 2011;23(5):555–564.
+  \newblock doi:{10.1016/j.intcom.2011.05.008}.
+
+  \bibitem{haapalainen2010psycho}
+  Haapalainen E, Kim S, Forlizzi JF, Dey AK.
+  \newblock Psycho-physiological measures for assessing cognitive load.
+  \newblock In: Proceedings of the 12th ACM international conference on
+    Ubiquitous computing; 2010. p. 301--310.
+
+  \bibitem{kumar2016measurement}
+  Kumar N, Kumar J.
+  \newblock Measurement of cognitive load in {HCI} systems using {EEG} power
+    spectrum: an experimental study.
+  \newblock Procedia Computer Science. 2016;84:70--78.
+
+  \bibitem{john1999:bfi}
+  John OP, Srivastava S, et~al.
+  \newblock The Big Five trait taxonomy: History, measurement, and theoretical
+    perspectives.
+  \newblock Handbook of personality: Theory and research. 1999;2(1999):102--138.
+
+  \bibitem{hirsh2008delay}
+  Hirsh JB, Morisano D, Peterson JB.
+  \newblock Delay discounting: Interactions between personality and cognitive
+    ability.
+  \newblock Journal of research in personality. 2008;42(6):1646--1650.
+
+  \bibitem{witmer1998:itq}
+  Witmer BG, Singer MJ.
+  \newblock Measuring presence in virtual environments: A presence questionnaire.
+  \newblock Presence. 1998;7(3):225--240.
+
+  \bibitem{olguin:2018}
+  {Olguín Muñoz} M, {Wang} J, {Satyanarayanan} M, {Gross} J.
+  \newblock Demo: Scaling on the Edge --- A Benchmarking Suite for
+    Human-in-the-Loop Applicationss.
+  \newblock In: 2018 IEEE/ACM Symposium on Edge Computing (SEC); 2018. p.
+    323--325.
+
+  \bibitem{olguin:2019}
+  {Olguín Muñoz} M, Wang J, Satyanarayanan M, Gross J.
+  \newblock EdgeDroid: An Experimental Approach to Benchmarking Human-in-the-Loop
+    Applications.
+  \newblock In: Proceedings of the 20th International Workshop on Mobile
+    Computing Systems and Applications. HotMobile '19. New York, NY, USA:
+    Association for Computing Machinery; 2019. p. 93–98.
+  \newblock Available from: \url{https://doi.org/10.1145/3301293.3302353}.
+
+  \bibitem{neves1981knowledge}
+  Neves DM, Anderson JR.
+  \newblock Knowledge Compilation: Mechanisms for the.
+  \newblock Cognitive skills and their acquisition. 1981;6:57.
+
+  \bibitem{zakay1995attentional}
+  Zakay D, Block RA.
+  \newblock An attentional-gate model of prospective time estimation.
+  \newblock Time and the dynamic control of behavior. 1995; p. 167--178.
+
+  \bibitem{zakay1996role}
+  Zakay D, Block RA.
+  \newblock The role of attention in time estimation processes.
+  \newblock In: Advances in psychology. vol. 115. Elsevier; 1996. p. 143--164.
+
+  \bibitem{bruno2016multiple}
+  Bruno A, Cicchini GM.
+  \newblock Multiple channels of visual time perception.
+  \newblock Current opinion in behavioral sciences. 2016;8:131--139.
+
+  \bibitem{wiener2011multiple}
+  Wiener M, Matell MS, Coslett H.
+  \newblock Multiple mechanisms for temporal processing.
+  \newblock Frontiers in integrative neuroscience. 2011;5:31.
+
+  \bibitem{heron2012duration}
+  Heron J, Aaen-Stockdale C, Hotchkiss J, Roach NW, McGraw PV, Whitaker D.
+  \newblock Duration channels mediate human time perception.
+  \newblock Proceedings of the Royal Society B: Biological Sciences.
+    2012;279(1729):690--698.
+
+  \bibitem{matthews2011stimulus}
+  Matthews WJ.
+  \newblock Stimulus repetition and the perception of time: The effects of prior
+    exposure on temporal discrimination, judgment, and production.
+  \newblock PLoS one. 2011;6(5).
+
+  \bibitem{droit2011emotion}
+  Droit-Volet S, Fayolle SL, Gil S.
+  \newblock Emotion and time perception: effects of film-induced mood.
+  \newblock Frontiers in integrative neuroscience. 2011;5:33.
+
+  \bibitem{Loewenstein1992anomaliesintertemporalchoice}
+  Loewenstein G, Prelec D.
+  \newblock {Anomalies in Intertemporal Choice: Evidence and an Interpretation*}.
+  \newblock The Quarterly Journal of Economics. 1992;107(2):573--597.
+  \newblock doi:{10.2307/2118482}.
+
+  \bibitem{nielsen1994usability}
+  Nielsen J.
+  \newblock Usability engineering.
+  \newblock Morgan Kaufmann; 1994.
+
+  \bibitem{shneiderman2016designing}
+  Shneiderman B, Plaisant C, Cohen M, Jacobs S, Elmqvist N, Diakopoulos N.
+  \newblock Designing the user interface: strategies for effective human-computer
+    interaction.
+  \newblock Pearson; 2016.
+
+  \bibitem{seow2008designing}
+  Seow SC.
+  \newblock Designing and engineering time: The psychology of time perception in
+    software.
+  \newblock Addison-Wesley Professional; 2008.
+
+  \bibitem{jeon2015degree}
+  Jeon H, Friederici AD.
+  \newblock Degree of automaticity and the prefrontal cortex.
+  \newblock Trends in cognitive sciences. 2015;19(5):244--250.
+
+  \bibitem{puttemans2005changes}
+  Puttemans V, Wenderoth N, Swinnen SP.
+  \newblock Changes in brain activation during the acquisition of a
+    multifrequency bimanual coordination task: from the cognitive stage to
+    advanced levels of automaticity.
+  \newblock Journal of Neuroscience. 2005;25(17):4270--4278.
+
+  \bibitem{lewis2003distinct}
+  Lewis PA, Miall RC.
+  \newblock Distinct systems for automatic and cognitively controlled time
+    measurement: evidence from neuroimaging.
+  \newblock Current opinion in neurobiology. 2003;13(2):250--255.
+
+  \bibitem{koch2009neural}
+  Koch G, Oliveri M, Caltagirone C.
+  \newblock Neural networks engaged in milliseconds and seconds time processing:
+    evidence from transcranial magnetic stimulation and patients with cortical or
+    subcortical dysfunction.
+  \newblock Philosophical Transactions of the Royal Society B: Biological
+    Sciences. 2009;364(1525):1907--1918.
+
+  \bibitem{lee2019limiting}
+  Lee TG, Acuña DE, Kording KP, Grafton ST.
+  \newblock Limiting motor skill knowledge via incidental training protects
+    against choking under pressure.
+  \newblock Psychonomic bulletin \& review. 2019;26(1):279--290.
+
+  \bibitem{baumeister74tice}
+  Baumeister RE, Bratslavsky E, Muraven M.
+  \newblock Tice (1998). Ego Depletion: Is the Active Self a Limited Resource.
+  \newblock Journal of personality and social psychology;74(5):1252.
+
+  \bibitem{lin2020strong}
+  Lin H, Saunders B, Friese M, Evans NJ, Inzlicht M.
+  \newblock Strong effort manipulations reduce response caution: A preregistered
+    reinvention of the ego-depletion paradigm.
+  \newblock Psychological science. 2020; p. 0956797620904990.
+
+  \bibitem{pijpers2003anxiety}
+  Pijpers JRR, Oudejans RRD, Holsheimer F, Bakker FC.
+  \newblock Anxiety--performance relationships in climbing: a process-oriented
+    approach.
+  \newblock Psychology of sport and exercise. 2003;4(3):283--304.
+
+  \bibitem{khawadi2015:usinggsrtrust}
+  Khawaji A, Zhou J, Chen F, Marcus N.
+  \newblock {Using Galvanic Skin Response (GSR) to Measure Trust and Cognitive
+    Load in the Text-Chat Environment}.
+  \newblock In: Proceedings of the 33rd Annual ACM Conference Extended Abstracts
+    on Human Factors in Computing Systems. CHI EA ’15. New York, NY, USA:
+    Association for Computing Machinery; 2015. p. 1989–1994.
+  \newblock Available from:
+    \url{https://doi-org.focus.lib.kth.se/10.1145/2702613.2732766}.
+
+  \bibitem{kuikkaniemi2010:biofeedback}
+  Kuikkaniemi K, Laitinen T, Turpeinen M, Saari T, Kosunen I, Ravaja N.
+  \newblock The Influence of Implicit and Explicit Biofeedback in First-Person
+    Shooter Games.
+  \newblock In: Proceedings of the SIGCHI Conference on Human Factors in
+    Computing Systems. CHI ’10. New York, NY, USA: Association for Computing
+    Machinery; 2010. p. 859–868.
+  \newblock Available from:
+    \url{https://doi-org.focus.lib.kth.se/10.1145/1753326.1753453}.
+
+  \bibitem{solovey2014:classifyingdriverworkload}
+  Solovey ET, Zec M, {Garcia Perez} EA, Reimer B, Mehler B.
+  \newblock Classifying Driver Workload Using Physiological and Driving
+    Performance Data: Two Field Studies.
+  \newblock In: Proceedings of the SIGCHI Conference on Human Factors in
+    Computing Systems. CHI ’14. New York, NY, USA: Association for Computing
+    Machinery; 2014. p. 4057–4066.
+  \newblock Available from:
+    \url{https://doi-org.focus.lib.kth.se/10.1145/2556288.2557068}.
+
+  \bibitem{peterson1907psycho}
+  Peterson F, Jung CG.
+  \newblock Psycho-physical investigations with the galvanometer and pneumograph
+    in normal and insane individuals.
+  \newblock Brain. 1907;30:153--218.
+
+  \bibitem{Healey2005}
+  Healey JA, Picard RW.
+  \newblock Detecting Stress During Real-World Driving Tasks Using Physiological
+    Sensors.
+  \newblock {IEEE} Transactions on Intelligent Transportation Systems.
+    2005;6(2):156--166.
+  \newblock doi:{10.1109/tits.2005.848368}.
+
+  \bibitem{Son2010}
+  Son J, Reimer B, Mehler B, Pohlmeyer AE, Godfrey KM, Orszulak J, et~al.
+  \newblock Age and cross-cultural comparison of drivers' cognitive workload and
+    performance in simulated urban driving.
+  \newblock International Journal of Automotive Technology. 2010;11(4):533--539.
+  \newblock doi:{10.1007/s12239-010-0065-6}.
+
+  \bibitem{Antonenko2010}
+  Antonenko P, Paas F, Grabner R, van Gog T.
+  \newblock Using Electroencephalography to Measure Cognitive Load.
+  \newblock Educational Psychology Review. 2010;22(4):425--438.
+  \newblock doi:{10.1007/s10648-010-9130-y}.
+
+  \bibitem{Grimes2008}
+  Grimes D, Tan DS, Hudson SE, Shenoy P, Rao RPN.
+  \newblock Feasibility and pragmatics of classifying working memory load with an
+    electroencephalograph.
+  \newblock In: Proceeding of the twenty-sixth annual {CHI} conference on Human
+    factors in computing systems - {CHI} {\textquotesingle}08. {ACM} Press;
+    2008.Available from: \url{https://doi.org/10.1145/1357054.1357187}.
+
+  \bibitem{empatica:e4}
+  {Empatica E4}; 2020.
+  \newblock Available from: \url{https://www.empatica.com/research/e4/}.
+
+  \bibitem{ragot2017emotion}
+  Ragot M, Martin N, Em S, Pallamin N, Diverrez JM.
+  \newblock Emotion recognition using physiological signals: laboratory vs.
+    wearable sensors.
+  \newblock In: International Conference on Applied Human Factors and Ergonomics.
+    Springer; 2017. p. 15--22.
+
+  \bibitem{mccarthy2016validation}
+  McCarthy C, Pradhan N, Redpath C, Adler A.
+  \newblock Validation of the {Empatica E4} wristband.
+  \newblock In: 2016 IEEE EMBS International Student Conference (ISC). IEEE;
+    2016. p. 1--4.
+
+  \bibitem{openbci:headbandkit}
+  {OpenBCI EEG Headband Kit}; 2020.
+  \newblock Available from:
+    \url{https://shop.openbci.com/products/openbci-eeg-headband-kit?variant=8120393760782}.
+
+  \bibitem{eeg1020system:1961}
+  The Ten Twenty Electrode System: International Federation of Societies for
+    Electroencephalography and Clinical Neurophysiology.
+  \newblock American Journal of EEG Technology. 1961;1(1):13--19.
+  \newblock doi:{10.1080/00029238.1961.11080571}.
+
+  \bibitem{agarwal2017eeg}
+  Agarwal S, Rani A, Singh V, Mittal AP.
+  \newblock {EEG signal enhancement using cascaded S-Golay filter}.
+  \newblock Biomedical Signal Processing and Control. 2017;36:194--204.
+
+  \bibitem{FUJIKOSHI1993ANOVA}
+  Fujikoshi Y.
+  \newblock {Two-way ANOVA models with unbalanced data}.
+  \newblock Discrete Mathematics. 1993;116(1):315 -- 334.
+  \newblock doi:{https://doi.org/10.1016/0012-365X(93)90410-U}.
+
+  \bibitem{quesnel2018you}
+  Quesnel D, Riecke BE.
+  \newblock Are you awed yet? How virtual reality gives us awe and goose bumps.
+  \newblock Frontiers in psychology. 2018;9:2158.
+
+  \end{thebibliography}
 \end{document}

### 2022-08-31 publications/2021ImpactDelayedResponse/body.tex
plos

--- 2022-08-31 publications/2021ImpactDelayedResponse/Manuscript.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-% \usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-\usepackage{tabularx}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% % Use the PLoS provided BiBTeX style
-% \bibliographystyle{plos2015}
-
-% % Remove brackets from numbering in List of References
-% \makeatletter
-% \renewcommand{\@biblabel}[1]{\quad#1.}
-% \makeatother
-
-% BibLatex and PLOS One hacks
-% \usepackage[citestyle=numeric,
-%             backend=bibtex,
-%             bibstyle=authoryear,
-%             sorting=none,
-%             sortcites,
-%             firstinits,
-%             terseinits,
-%             maxbibnames=6,
-%             dashed=false]{biblatex}
-% \bibliography{bibliography}
-\usepackage{cite}
-
-\usepackage[colorlinks]{hyperref}
-
-% Name list format
-% \renewcommand*{\labelnamepunct}{\addspace}
-% \renewcommand*{\finalnamedelim}{%
-%   \ifbibliography{\addcomma\space}{\addspace\&\space}}
-% \renewcommand*{\revsdnamepunct}{}
-% \DeclareNameAlias{sortname}{last-first}
-
-% No quotes or italics in titles
-% \DeclareFieldFormat{title}{#1}
-% \DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
-%                     thesis,unpublished]{title}{#1}
-% \DeclareFieldFormat{journaltitle}{#1}
-% \DeclareFieldFormat{issuetitle}{#1}
-% \DeclareFieldFormat{maintitle}{#1}
-
-% \renewbibmacro*{in:}{%
-%   \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
-
-% Remove page prefixes
-% \DeclareFieldFormat{pages}{#1}
-
-% Only print journal volume
-% \renewbibmacro*{volume+number+eid}{%
-%   \setunit*{\addspace}%
-%   \printfield{volume}}
-
-% Use colon for volume-pages delimiter
-% \renewcommand*{\bibpagespunct}{%
-%   \ifentrytype{article}{\addcolon\space}{\addcomma\space}}
-
-% Add labelnumbers to bibliography
-% \DeclareFieldFormat{labelnumberwidth}{#1\adddot}
-% \defbibenvironment{bibliography}
-%   {\list
-%      {\printtext[labelnumberwidth]{%
-%       \printfield{prefixnumber}%
-%       \printfield{labelnumber}}}
-%      {\setlength{\labelwidth}{\labelnumberwidth}%
-%       \setlength{\leftmargin}{\labelwidth}%
-%       \setlength{\labelsep}{\biblabelsep}%
-%       \addtolength{\leftmargin}{\labelsep}%
-%       \setlength{\itemsep}{\bibitemsep}%
-%       \setlength{\parsep}{\bibparsep}}%
-%       \renewcommand*{\makelabel}[1]{\hss##1}}
-%   {\endlist}
-%   {\item}
-
-% Omit authoryear disambiguation
-% \AtEveryBibitem{\clearfield{extrayear}}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
-
-% other packages
-\usepackage[capitalize, nameinlink]{cleveref}
-\usepackage[inline]{enumitem}
-\usepackage[binary-units, range-units=single, range-phrase=~to~, list-units=single]{siunitx}
-\sisetup{
-    detect-all,
-    detect-inline-family=math,
-    detect-inline-weight=math,
-    detect-display-math=true,
-    list-final-separator = {, and }
-}
-\usepackage{subcaption}
-\usepackage{amsthm}
-\usepackage{thmtools}
-
-\declaretheoremstyle[
-notefont=\bfseries, notebraces={}{},
-bodyfont=\normalfont,
-headformat=\NAME~\NUMBER~---~\NOTE%
-]{nopar}
-\declaretheorem[style=nopar]{definition}
-
-\usepackage[export]{adjustbox}
-\usepackage{multirow}
-\usepackage{booktabs}
-
-\usepackage{pgf-umlsd}
-\usepackage{tikz}
-\usepackage{tikzpeople}
-\usetikzlibrary{shapes,arrows.meta,positioning,automata,matrix,shapes.symbols,trees}
-\usepackage{tikz-qtree}
-\usepackage[]{todonotes}
-\reversemarginpar  % todonotes on left
-
-\newenvironment{acks}{
-  \section*{Acknowledgements}
-}{}
-
-\begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Impact of delayed response on Wearable Cognitive Assistance} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Manuel~{Olguín~Muñoz}\textsuperscript{1, *},
-Roberta~Klatzky\textsuperscript{2, 3},
-Junjue~Wang\textsuperscript{4},
-Padmanabhan~Pillai\textsuperscript{5},
-Mahadev~Satyanarayanan\textsuperscript{4},
-James~Gross\textsuperscript{1}
-\\
-\bigskip
-\begin{tabularx}{\textwidth}{lX}
-  \textbf{1} & Division of Information~Science and Engineering, School of Electrical~Engineering and Computer~Science, KTH Royal Institute~of~Technology, Stockholm, Sweden\\
-  \textbf{2} & Department of Psychology, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
-  \textbf{3} & Human-Computer Interaction Institute, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
-  \textbf{4} & School of Computer Science, Carnegie~Mellon~University, Pittsburgh, PA, USA\\
-  \textbf{5} & Intel Labs, Pittsburgh, PA, USA\\
-\end{tabularx}
-\bigskip
-
-\begin{tabularx}{\textwidth}{lX}
-  % \Yinyang & These authors contributed equally to this work.\\
-  % \ddag & Senior author.\\
-  * & Corresponding author: \href{mailto:molguin@kth.se}{\texttt{molguin@kth.se}}\\
-\end{tabularx}
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-
 \section*{Abstract}
@@
 
-\linenumbers%
 \section{Introduction}
@@
 \end{acks}
-
-\nolinenumbers%
-
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for 
-% step-by-step instructions.
-%
-
-% \bibliographystyle{plos2015}
-% \bibliography{bibliography}
-\begin{thebibliography}{10}
-
-  \bibitem{Ha:TowardsWearableCogAssist}
-  Ha K, Chen Z, Hu W, Richter W, Pillai P, Satyanarayanan M.
-  \newblock Towards Wearable Cognitive Assistance.
-  \newblock In: Proceedings of the 12th Annual International Conference on Mobile
-    Systems, Applications, and Services. MobiSys '14. New York, NY, USA: ACM;
-    2014. p. 68--81.
-  \newblock Available from: \url{http://doi.acm.org/10.1145/2594368.2594383}.
-
-  \bibitem{Chen:EarlyImplementation}
-  Chen Z, Jiang L, Hu W, Ha K, Amos B, Pillai P, et~al.
-  \newblock Early Implementation Experience with Wearable Cognitive Assistance
-    Applications.
-  \newblock In: Proceedings of the 2015 Workshop on Wearable Systems and
-    Applications. WearSys '15. New York, NY, USA: ACM; 2015. p. 33--38.
-  \newblock Available from: \url{http://doi.acm.org/10.1145/2753509.2753517}.
-
-  \bibitem{IKEAAssistant}
-  {IKEA assistant}; 2020.
-  \newblock Available from: \url{https://www.youtube.com/watch?v=vMzTGgQHjpo}.
-
-  \bibitem{satya2019augmenting}
-  {Satyanarayanan} M, {Davies} N.
-  \newblock {Augmenting Cognition Through Edge Computing}.
-  \newblock Computer. 2019;52(7):37--46.
-  \newblock doi:{10.1109/MC.2019.2911878}.
-
-  \bibitem{Chen:AnEmpiricalStudyOfLatency}
-  Chen Z, Hu W, Wang J, Zhao S, Amos B, Wu G, et~al.
-  \newblock An Empirical Study of Latency in an Emerging Class of Edge Computing
-    Applications for Wearable Cognitive Assistance.
-  \newblock In: Proceedings of the Second ACM/IEEE Symposium on Edge Computing.
-    SEC '17. New York, NY, USA: ACM; 2017. p. 14:1--14:14.
-  \newblock Available from: \url{http://doi.acm.org/10.1145/3132211.3134458}.
-
-  \bibitem{funk2015caworkplace}
-  {Funk} M, {Schmidt} A.
-  \newblock Cognitive Assistance in the Workplace.
-  \newblock IEEE Pervasive Computing. 2015;14(3):53--55.
-
-  \bibitem{gorecky2011cognito}
-  Gorecky D, Worgan SF, Meixner G.
-  \newblock {COGNITO: A Cognitive Assistance and Training System for Manual Tasks
-    in Industry}.
-  \newblock In: Proceedings of the 29th Annual European Conference on Cognitive
-    Ergonomics. ECCE '11. New York, NY, USA: Association for Computing Machinery;
-    2011. p. 53--6.
-  \newblock Available from: \url{https://doi.org/10.1145/2074712.2074723}.
-
-  \bibitem{PingPongAssistant}
-  {Ping Pong assistant}; 2020.
-  \newblock Available from: \url{https://www.youtube.com/watch?v=_lp32sowyUA}.
-
-  \bibitem{dabrowsky:2011:40years}
-  Dabrowski J, Munson EV.
-  \newblock {40 years of Searching for the Best Computer System Response Time}.
-  \newblock Interact Comput. 2011;23(5):555–564.
-  \newblock doi:{10.1016/j.intcom.2011.05.008}.
-
-  \bibitem{haapalainen2010psycho}
-  Haapalainen E, Kim S, Forlizzi JF, Dey AK.
-  \newblock Psycho-physiological measures for assessing cognitive load.
-  \newblock In: Proceedings of the 12th ACM international conference on
-    Ubiquitous computing; 2010. p. 301--310.
-
-  \bibitem{kumar2016measurement}
-  Kumar N, Kumar J.
-  \newblock Measurement of cognitive load in {HCI} systems using {EEG} power
-    spectrum: an experimental study.
-  \newblock Procedia Computer Science. 2016;84:70--78.
-
-  \bibitem{john1999:bfi}
-  John OP, Srivastava S, et~al.
-  \newblock The Big Five trait taxonomy: History, measurement, and theoretical
-    perspectives.
-  \newblock Handbook of personality: Theory and research. 1999;2(1999):102--138.
-
-  \bibitem{hirsh2008delay}
-  Hirsh JB, Morisano D, Peterson JB.
-  \newblock Delay discounting: Interactions between personality and cognitive
-    ability.
-  \newblock Journal of research in personality. 2008;42(6):1646--1650.
-
-  \bibitem{witmer1998:itq}
-  Witmer BG, Singer MJ.
-  \newblock Measuring presence in virtual environments: A presence questionnaire.
-  \newblock Presence. 1998;7(3):225--240.
-
-  \bibitem{olguin:2018}
-  {Olguín Muñoz} M, {Wang} J, {Satyanarayanan} M, {Gross} J.
-  \newblock Demo: Scaling on the Edge --- A Benchmarking Suite for
-    Human-in-the-Loop Applicationss.
-  \newblock In: 2018 IEEE/ACM Symposium on Edge Computing (SEC); 2018. p.
-    323--325.
-
-  \bibitem{olguin:2019}
-  {Olguín Muñoz} M, Wang J, Satyanarayanan M, Gross J.
-  \newblock EdgeDroid: An Experimental Approach to Benchmarking Human-in-the-Loop
-    Applications.
-  \newblock In: Proceedings of the 20th International Workshop on Mobile
-    Computing Systems and Applications. HotMobile '19. New York, NY, USA:
-    Association for Computing Machinery; 2019. p. 93–98.
-  \newblock Available from: \url{https://doi.org/10.1145/3301293.3302353}.
-
-  \bibitem{neves1981knowledge}
-  Neves DM, Anderson JR.
-  \newblock Knowledge Compilation: Mechanisms for the.
-  \newblock Cognitive skills and their acquisition. 1981;6:57.
-
-  \bibitem{zakay1995attentional}
-  Zakay D, Block RA.
-  \newblock An attentional-gate model of prospective time estimation.
-  \newblock Time and the dynamic control of behavior. 1995; p. 167--178.
-
-  \bibitem{zakay1996role}
-  Zakay D, Block RA.
-  \newblock The role of attention in time estimation processes.
-  \newblock In: Advances in psychology. vol. 115. Elsevier; 1996. p. 143--164.
-
-  \bibitem{bruno2016multiple}
-  Bruno A, Cicchini GM.
-  \newblock Multiple channels of visual time perception.
-  \newblock Current opinion in behavioral sciences. 2016;8:131--139.
-
-  \bibitem{wiener2011multiple}
-  Wiener M, Matell MS, Coslett H.
-  \newblock Multiple mechanisms for temporal processing.
-  \newblock Frontiers in integrative neuroscience. 2011;5:31.
-
-  \bibitem{heron2012duration}
-  Heron J, Aaen-Stockdale C, Hotchkiss J, Roach NW, McGraw PV, Whitaker D.
-  \newblock Duration channels mediate human time perception.
-  \newblock Proceedings of the Royal Society B: Biological Sciences.
-    2012;279(1729):690--698.
-
-  \bibitem{matthews2011stimulus}
-  Matthews WJ.
-  \newblock Stimulus repetition and the perception of time: The effects of prior
-    exposure on temporal discrimination, judgment, and production.
-  \newblock PLoS one. 2011;6(5).
-
-  \bibitem{droit2011emotion}
-  Droit-Volet S, Fayolle SL, Gil S.
-  \newblock Emotion and time perception: effects of film-induced mood.
-  \newblock Frontiers in integrative neuroscience. 2011;5:33.
-
-  \bibitem{Loewenstein1992anomaliesintertemporalchoice}
-  Loewenstein G, Prelec D.
-  \newblock {Anomalies in Intertemporal Choice: Evidence and an Interpretation*}.
-  \newblock The Quarterly Journal of Economics. 1992;107(2):573--597.
-  \newblock doi:{10.2307/2118482}.
-
-  \bibitem{nielsen1994usability}
-  Nielsen J.
-  \newblock Usability engineering.
-  \newblock Morgan Kaufmann; 1994.
-
-  \bibitem{shneiderman2016designing}
-  Shneiderman B, Plaisant C, Cohen M, Jacobs S, Elmqvist N, Diakopoulos N.
-  \newblock Designing the user interface: strategies for effective human-computer
-    interaction.
-  \newblock Pearson; 2016.
-
-  \bibitem{seow2008designing}
-  Seow SC.
-  \newblock Designing and engineering time: The psychology of time perception in
-    software.
-  \newblock Addison-Wesley Professional; 2008.
-
-  \bibitem{jeon2015degree}
-  Jeon H, Friederici AD.
-  \newblock Degree of automaticity and the prefrontal cortex.
-  \newblock Trends in cognitive sciences. 2015;19(5):244--250.
-
-  \bibitem{puttemans2005changes}
-  Puttemans V, Wenderoth N, Swinnen SP.
-  \newblock Changes in brain activation during the acquisition of a
-    multifrequency bimanual coordination task: from the cognitive stage to
-    advanced levels of automaticity.
-  \newblock Journal of Neuroscience. 2005;25(17):4270--4278.
-
-  \bibitem{lewis2003distinct}
-  Lewis PA, Miall RC.
-  \newblock Distinct systems for automatic and cognitively controlled time
-    measurement: evidence from neuroimaging.
-  \newblock Current opinion in neurobiology. 2003;13(2):250--255.
-
-  \bibitem{koch2009neural}
-  Koch G, Oliveri M, Caltagirone C.
-  \newblock Neural networks engaged in milliseconds and seconds time processing:
-    evidence from transcranial magnetic stimulation and patients with cortical or
-    subcortical dysfunction.
-  \newblock Philosophical Transactions of the Royal Society B: Biological
-    Sciences. 2009;364(1525):1907--1918.
-
-  \bibitem{lee2019limiting}
-  Lee TG, Acuña DE, Kording KP, Grafton ST.
-  \newblock Limiting motor skill knowledge via incidental training protects
-    against choking under pressure.
-  \newblock Psychonomic bulletin \& review. 2019;26(1):279--290.
-
-  \bibitem{baumeister74tice}
-  Baumeister RE, Bratslavsky E, Muraven M.
-  \newblock Tice (1998). Ego Depletion: Is the Active Self a Limited Resource.
-  \newblock Journal of personality and social psychology;74(5):1252.
-
-  \bibitem{lin2020strong}
-  Lin H, Saunders B, Friese M, Evans NJ, Inzlicht M.
-  \newblock Strong effort manipulations reduce response caution: A preregistered
-    reinvention of the ego-depletion paradigm.
-  \newblock Psychological science. 2020; p. 0956797620904990.
-
-  \bibitem{pijpers2003anxiety}
-  Pijpers JRR, Oudejans RRD, Holsheimer F, Bakker FC.
-  \newblock Anxiety--performance relationships in climbing: a process-oriented
-    approach.
-  \newblock Psychology of sport and exercise. 2003;4(3):283--304.
-
-  \bibitem{khawadi2015:usinggsrtrust}
-  Khawaji A, Zhou J, Chen F, Marcus N.
-  \newblock {Using Galvanic Skin Response (GSR) to Measure Trust and Cognitive
-    Load in the Text-Chat Environment}.
-  \newblock In: Proceedings of the 33rd Annual ACM Conference Extended Abstracts
-    on Human Factors in Computing Systems. CHI EA ’15. New York, NY, USA:
-    Association for Computing Machinery; 2015. p. 1989–1994.
-  \newblock Available from:
-    \url{https://doi-org.focus.lib.kth.se/10.1145/2702613.2732766}.
-
-  \bibitem{kuikkaniemi2010:biofeedback}
-  Kuikkaniemi K, Laitinen T, Turpeinen M, Saari T, Kosunen I, Ravaja N.
-  \newblock The Influence of Implicit and Explicit Biofeedback in First-Person
-    Shooter Games.
-  \newblock In: Proceedings of the SIGCHI Conference on Human Factors in
-    Computing Systems. CHI ’10. New York, NY, USA: Association for Computing
-    Machinery; 2010. p. 859–868.
-  \newblock Available from:
-    \url{https://doi-org.focus.lib.kth.se/10.1145/1753326.1753453}.
-
-  \bibitem{solovey2014:classifyingdriverworkload}
-  Solovey ET, Zec M, {Garcia Perez} EA, Reimer B, Mehler B.
-  \newblock Classifying Driver Workload Using Physiological and Driving
-    Performance Data: Two Field Studies.
-  \newblock In: Proceedings of the SIGCHI Conference on Human Factors in
-    Computing Systems. CHI ’14. New York, NY, USA: Association for Computing
-    Machinery; 2014. p. 4057–4066.
-  \newblock Available from:
-    \url{https://doi-org.focus.lib.kth.se/10.1145/2556288.2557068}.
-
-  \bibitem{peterson1907psycho}
-  Peterson F, Jung CG.
-  \newblock Psycho-physical investigations with the galvanometer and pneumograph
-    in normal and insane individuals.
-  \newblock Brain. 1907;30:153--218.
-
-  \bibitem{Healey2005}
-  Healey JA, Picard RW.
-  \newblock Detecting Stress During Real-World Driving Tasks Using Physiological
-    Sensors.
-  \newblock {IEEE} Transactions on Intelligent Transportation Systems.
-    2005;6(2):156--166.
-  \newblock doi:{10.1109/tits.2005.848368}.
-
-  \bibitem{Son2010}
-  Son J, Reimer B, Mehler B, Pohlmeyer AE, Godfrey KM, Orszulak J, et~al.
-  \newblock Age and cross-cultural comparison of drivers' cognitive workload and
-    performance in simulated urban driving.
-  \newblock International Journal of Automotive Technology. 2010;11(4):533--539.
-  \newblock doi:{10.1007/s12239-010-0065-6}.
-
-  \bibitem{Antonenko2010}
-  Antonenko P, Paas F, Grabner R, van Gog T.
-  \newblock Using Electroencephalography to Measure Cognitive Load.
-  \newblock Educational Psychology Review. 2010;22(4):425--438.
-  \newblock doi:{10.1007/s10648-010-9130-y}.
-
-  \bibitem{Grimes2008}
-  Grimes D, Tan DS, Hudson SE, Shenoy P, Rao RPN.
-  \newblock Feasibility and pragmatics of classifying working memory load with an
-    electroencephalograph.
-  \newblock In: Proceeding of the twenty-sixth annual {CHI} conference on Human
-    factors in computing systems - {CHI} {\textquotesingle}08. {ACM} Press;
-    2008.Available from: \url{https://doi.org/10.1145/1357054.1357187}.
-
-  \bibitem{empatica:e4}
-  {Empatica E4}; 2020.
-  \newblock Available from: \url{https://www.empatica.com/research/e4/}.
-
-  \bibitem{ragot2017emotion}
-  Ragot M, Martin N, Em S, Pallamin N, Diverrez JM.
-  \newblock Emotion recognition using physiological signals: laboratory vs.
-    wearable sensors.
-  \newblock In: International Conference on Applied Human Factors and Ergonomics.
-    Springer; 2017. p. 15--22.
-
-  \bibitem{mccarthy2016validation}
-  McCarthy C, Pradhan N, Redpath C, Adler A.
-  \newblock Validation of the {Empatica E4} wristband.
-  \newblock In: 2016 IEEE EMBS International Student Conference (ISC). IEEE;
-    2016. p. 1--4.
-
-  \bibitem{openbci:headbandkit}
-  {OpenBCI EEG Headband Kit}; 2020.
-  \newblock Available from:
-    \url{https://shop.openbci.com/products/openbci-eeg-headband-kit?variant=8120393760782}.
-
-  \bibitem{eeg1020system:1961}
-  The Ten Twenty Electrode System: International Federation of Societies for
-    Electroencephalography and Clinical Neurophysiology.
-  \newblock American Journal of EEG Technology. 1961;1(1):13--19.
-  \newblock doi:{10.1080/00029238.1961.11080571}.
-
-  \bibitem{agarwal2017eeg}
-  Agarwal S, Rani A, Singh V, Mittal AP.
-  \newblock {EEG signal enhancement using cascaded S-Golay filter}.
-  \newblock Biomedical Signal Processing and Control. 2017;36:194--204.
-
-  \bibitem{FUJIKOSHI1993ANOVA}
-  Fujikoshi Y.
-  \newblock {Two-way ANOVA models with unbalanced data}.
-  \newblock Discrete Mathematics. 1993;116(1):315 -- 334.
-  \newblock doi:{https://doi.org/10.1016/0012-365X(93)90410-U}.
-
-  \bibitem{quesnel2018you}
-  Quesnel D, Riecke BE.
-  \newblock Are you awed yet? How virtual reality gives us awe and goose bumps.
-  \newblock Frontiers in psychology. 2018;9:2158.
-
-  \end{thebibliography}
-\end{document}

### 2022-08-31 publications/2021ImpactDelayedResponse/prepare.py

@@ -1,153 +1,0 @@
-import shutil
-import subprocess
-from dataclasses import dataclass
-from io import StringIO
-from pathlib import Path
-from typing import Iterator, Tuple
-
-import click
-import parse
-
-
-class Figure:
-    @dataclass
-    class SubFig:
-        sub_index: str
-        line: int
-        original_path: str
-
-    def __init__(self,
-                 index: int):
-        self._index = f'Fig{index}'
-        self._subfigs = []
-
-    def iterfigs(self) -> Iterator[Tuple[int, str, str]]:
-        n_subfigs = len(self._subfigs)
-        if n_subfigs == 1:
-            sfig = self._subfigs[0]
-            yield sfig.line, self._index, sfig.original_path
-        elif n_subfigs > 1:
-            for sfig in self._subfigs:
-                yield sfig.line, \
-                      f'{self._index}{sfig.sub_index}', \
-                      sfig.original_path
-        else:
-            raise RuntimeError()
-
-    def add_subfig(self, line: int, graphics: str) -> None:
-        sub_index = chr(ord('a') + len(self._subfigs))
-        self._subfigs.append(Figure.SubFig(sub_index, line, graphics))
-
-    def __str__(self) -> str:
-        n_subfigs = len(self._subfigs)
-        if n_subfigs == 1:
-            return f'{self._index} {self._subfigs[0].original_path}'
-        elif n_subfigs > 1:
-            return '\n'.join([f'{self._index}{s.sub_index} '
-                              f'{s.original_path}' for s in self._subfigs])
-        else:
-            raise RuntimeError()
-
-
-class DirectoryPath(click.ParamType):
-    name = 'directory'
-
-    def __init__(self, clean: bool = False,
-                 exist_ok: bool = True,
-                 parents: bool = False):
-        super(DirectoryPath, self).__init__()
-        self._clean = clean
-        self._exist_ok = exist_ok
-        self._parents = parents
-
-    def convert(self, value, param, ctx) -> Path:
-        p = Path(value).resolve()
-        if self._clean and p.exists():
-            subprocess.run(['rm', '-rf', str(p)])
-
-        p.mkdir(exist_ok=self._exist_ok, parents=self._parents)
-        return p
-
-
-@click.command()
-@click.option('--cls-file',
-              required=False,
-              default='./plos2015.bst',
-              show_default=True,
-              type=click.Path(exists=True, file_okay=True, dir_okay=False))
-@click.option('--bib-file',
-              required=False,
-              default='./bibliography.bib',
[…]

### 2022-08-31 publications/2021ImpactDelayedResponse/tables/generate.py

@@ -1,76 +1,0 @@
-#!/usr/bin/env python3
-import shutil
-import subprocess
-from pathlib import Path
-from tempfile import TemporaryDirectory
-from typing import Collection, IO
-
-import click
-import jinja2
-
-
-def generate_table_texs(template_str: str, tables: Collection[str]) \
-        -> Collection[str]:
-    latex_jinja_env = jinja2.Environment(
-        block_start_string='\BLOCK{',
-        block_end_string='}',
-        variable_start_string='\VAR{',
-        variable_end_string='}',
-        comment_start_string='\#{',
-        comment_end_string='}',
-        line_statement_prefix='%%',
-        line_comment_prefix='%#',
-        trim_blocks=True,
-        autoescape=False)
-    temp = latex_jinja_env.from_string(template_str)
-    return list(map(lambda t: temp.render(table=t), tables))
-
-
-@click.command()
-@click.argument('template', nargs=1, type=click.File())
-@click.argument('output_dir', nargs=1,
-                type=click.Path(exists=True, file_okay=False,
-                                dir_okay=True, resolve_path=True))
-@click.argument('tables', nargs=-1,
-                type=click.Path(exists=True, file_okay=True,
-                                dir_okay=False, resolve_path=True))
-def generate(template: IO, output_dir: str, tables: Collection[str]):
-    template_str = template.read()
-    tab_str = []
-    tab_names = []
-    for table in tables:
-        tab_path = Path(table)
-        with tab_path.open('r') as fp:
-            tab_str.append(fp.read())
-        tab_names.append(tab_path.name)
-
-    output_dir = Path(output_dir).resolve()
-    texs = generate_table_texs(template_str, tab_str)
-    # make temp dir
-    with TemporaryDirectory() as tmp_dir_str:
-        tmp_dir = Path(tmp_dir_str).resolve()
-        for tex, table_name in zip(texs, tab_names):
-            tex_file_path = (tmp_dir / table_name).resolve()
-            with tex_file_path.open('w') as fp:
-                fp.write(tex)
-
-            for i in range(4):
-                subprocess.Popen([
-                    '/usr/bin/latexmk',
-                    '-interaction=nonstopmode',
-                    '-shell-escape',
-                    '-synctex=1',
-                    '-f',
-                    f'-jobname={table_name}',
-                    f'-output-directory={tmp_dir}',
-                    f'{tex_file_path}'
-                ], cwd=str(tmp_dir)).wait()
-
[…]

### 2022-08-31 publications/2022Ainur/abstract.tex
cleave and ainur

\begin{abstract}
  Experimental research on wireless networking in combination with edge and cloud computing has been the subject of explosive interest in the last decade.
  This development has been driven by the increasing complexity of modern wireless technologies and the  extensive softwarization of these through projects such as a \acf*{O-RAN}.
  In this context, a number of small- to mid-scale testbeds have emerged, employing a variety of technologies to target a wide array of use-cases and scenarios in the context of novel mobile communication technologies such as 5G and beyond-5G.
  Little work, however, has yet been devoted to developing a standard framework for wireless testbed automation which is hardware-agnostic and compatible with edge- and cloud-native technologies.
  Such a solution would simplify the development of new testbeds by completely or partially removing the requirement for custom management and orchestration software.

  In this paper, we present the first such mostly hardware-agnostic wireless testbed automation framework, \emph{Ainur}.
  It is designed to configure, manage, orchestrate, and deploy workloads from an end-to-end perspective.
  Ainur is built on top of cloud-native technologies such as Docker, and is provided as \acs*{FOSS} to the community through the {KTH-EXPECA/Ainur} repository on GitHub.
  We demonstrate the utility of the platform with a series of scenarios, showcasing in particular its flexibility with respect to physical link definition, computation placement, and automation of arbitrarily complex experimental scenarios.
\end{abstract}

### 2022-08-31 publications/2022Ainur/ainur.tex
cleave and ainur

\section{The Ainur Framework}\label{sec:ainur}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
    \caption{Network structure assumed by Ainur.}\label{fig:network}
\end{figure*}

Ainur is designed as an end-to-end wireless network testbed management framework, fully flexible in terms of the communication network stack, computation hosts, and the distributed application deployed on top.
The core goal of the framework is to facilitate the creation of the desired communication and computation elements to discover their implications to the performance of end-to-end distributed applications.
Ainur achieves this by offering a Python \ac{API} which is used to describe an end-to-end experiment in a procedural manner, and we plan to eventually provide toolkits for declarative configuration of experiments.

Conceptually, in Ainur, an experiment is decomposed into a layered structure, consisting of
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item the distributed application (workload)
    \item computation hosts
    \item communication networks connecting the hosts
\end{enumerate*}.
In \cref{fig:overview}, the workload is represented by a client-server process pair.
These could correspond, for instance, to the emulation of an inverted pendulum and a matching controller.
However, Ainur makes no assumptions about the nature of the workloads deployed on the framework, and virtually any process or combination of processes can be used.

Hosts correspond to either bare-metal machines or cloud instances on \ac{AWS} \ac{EC2}.
Users are free to combine and interconnect these in any configuration.
Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including WiFi, 4G \ac{LTE}, and 5G.
To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \ac{USRP} \ac{SDR}s and some computation hosts dedicated to signal processing and/or radio management.

To collect data from workloads, Ainur configures both a shared, distributed, storage location which all processes can reach, as well as a logging service which automatically captures structured and unstructured data from the standard output of workload processes.
The logging service additionally collects data from the other two layers, and the resulting dataset could, for instance, be analyzed to relate the performance of the control loop to the quality of the wireless link.

\subsection{Ainur Software Stack}

To realize our vision for an automated, flexible, cloud-native, and workload-agnostic framework for cloud and edge computing experimentation, we built Ainur on top of a combination of well-established tools and frameworks.
Below we briefly touch on the most important of these:

\begin{description}
    \item[Python:]
    Ainur is built on-top of Python 3.8.
    We chose this language for its flexibility, ease of prototyping, and for the extensive ecosystem of third-party cloud-native frameworks and libraries.

    \item[Ansible:]
    One of the ``cloud-native frameworks'' mentioned above, Red Hat Ansible is a powerful Python framework for bare-metal configuration, provisioning, and automation.
    We use it extensively in Ainur for configuration of network interfaces and services on managed hosts.

    \item[Containers:]
    We leverage Docker containers to great extent for the virtualization and orchestration of workloads, as well as the encapsulation and management of complex network configurations.
    See \cref{ssec:containers} for more details.

    \item[\ac{AWS} and \texttt{boto3}:]
    Ainur employs the Amazon \texttt{boto3} library for Python to directly interface with \ac{AWS} \ac{EC2} and deploy, configure, and manage remote cloud instances.

    \item[\ac{OAI}:]
    \ac{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \acp{SDR} like the \ac{USRP}~\cite{KALTENBERGER2020107284}. 
    Ainur can deploy, configure, and manage \ac{OAI} software components that implement 4G \ac{LTE} and 5G \ac{NR}.

    \item[Mango Communications 802.11:]
    Finally, the Mango Communications project implements real-time 802.11 (WiFi) MAC and PHY in Xilinx \acp{FPGA}.
    It can be used on a variety of hardware platforms including \ac{USRP} \acp{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
\end{description}

\subsection{Main Software Components}

Ainur follows a layered architecture which closely mimics the conceptual layers of the \acs{TCP}/\acs{IP} stack.
Components are deployed in bottom-up order, starting with the establishment of physical links, through the establishment of \ac{IP} connectivity and deployment of links to the cloud, and ending with the distribution and initialization of workloads on top of a container orchestration layer.
The architectural modules of the framework can broadly be classified according to the below categories:

\begin{description}[]
    \item[Configuration Layer:]
    The lowest layer of Ainur. It handles the parsing of configuration files describing experimental scenarios.
    Optional, as it is only needed when Ainur is running experiments described in a declarative manner.

    \item[Logging layer:]
    This layer handles logging across all layers of the system to a central repository.
    Concretely, it manages the configuration and lifecycle of a Fluentd server to which any component in the network can send logs.
    Currently, the two main components relying on this scaffolding are the physical layer and the workloads.

    \item[Physical Layer:] 
    Creates and deploys the underlying physical connections of the workload data network.
    This layer interacts with hardware such as managed switches and \acp{SDR} (and their associated computation hosts) to create links between the desired workload hosts.
    These links correspond to ethernet, WiFi, and even an entire software-defined 4G \ac{LTE} and 5G networks.

    \item[Cloud Layer:]
    Handles integration with cloud services (currently, \ac{AWS} \ac{EC2}). 
    Only deployed in the case of an experiment requiring cloud instances, it manages the instantiation and configuration of remote cloud resources.

    \item[\ac{IP} Layer:]
    Configures and establishes \ac{IP} layer connectivity of packets between hosts in the experimental setup, both local and cloud.
    This includes assigning valid \ac{IP} addresses to hosts, establishing \ac{VPN} routes to cloud instances through a pair of pre-configured \ac{VPN} gateways, and configuring routing tables to ensure any two hosts in the workload network can communicate with each other.

    \item[Workload Layer:]
    Finally, this layer deploys, scales, and orchestrates containerized workloads, as well as configures a shared, distributed storage for workloads to store data.
    This components leverages Docker Swarm to spin up and manage workload containers on desired hosts.
    It also establishes overlay networks abstracting away the physical topology and allowing containers to interconnect through dynamically assigned hostnames.
\end{description}

\subsection{Containerization}\label{ssec:containers}

Containers are a key cloud-native technology for the virtualization and sandboxing of arbitrary processes~\cite{merkel2014docker}.
They allow for easy packaging of software in predefined, consistent, and conflict-free execution environments, including all necessary dependencies.
Containers are lightweight and portable compared to \ac{VM}-based virtualization, making them an ideal solution for the distribution, deployment, and orchestration of software in distributed computing environments.
They deploy quickly and are very configurable, while abstracting away the complexities and delays that come with in creating, managing, and moving (potentially huge) \ac{VM} disk images.

Ainur leverages containerization extensively across the framework, in order to
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item deploy and orchestrate workloads
    \item support a wide spectrum of different physical layer configurations out-of-the-box, as well as allow for easy extension to new ones
    \item support automated collection of logs
\end{enumerate*}.
Workloads in the framework are deployed packaged in Docker containers, orchestrated through \texttt{docker-compose} and Docker Swarm.
We have selected Docker Swarm as our container orchestration layer instead of more advanced toolkits such as Kubernetes due to
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item it being more lightweight
    \item its comparative simplicity in terms of configuration and setup
    \item the fact that it is included in the default Docker runtime installation
\end{enumerate*}.
This allows the framework to remain mostly agnostic to the nature of the workloads, and thus support a wide ranged of different applications and system architectures.
It also allows for easy deployment to different compute nodes in the network, without having to take into consideration details such as required libraries and versions.

As expressed above, Ainur also leverages containers for the communication stack.
With the advent of \ac{O-RAN} and \acp{SDR}, all components of the wireless network stack can run on general-purpose processors.
They can thus be containerized and distributed across multiple hosts.

As a final point, the automation of logging is another advantage of using containers and an orchestration framework.
Ainur employs the Fluentd unified layer for log collection.
It natively integrates with Docker containers and allows for the automatic collection of text data from the standard output and error streams of processes executing inside containers.
It automatically decouples data sources from containers running on different systems and allows Ainur to collect and classify the logs from any components of the network, whether on the wireless stack or workload.

\subsection{Obtaining Ainur and deploying it on a new testbed}

\begin{figure}
    \centering
    \includegraphics[width=.8\columnwidth]{publications/2022Ainur/figures/network-minimal.png}
    \caption{Minimal testbed setup on which Ainur can be deployed.}\label{fig:network:minimal}
\end{figure}

The framework is released as \acl{FOSS} under a permissive Apache license.
It can be downloaded from the Ainur repository~\cite{ainur:github} of the {KTH-EXPECA} organization on GitHub.
Given the software's complexity, the repository also includes detailed documentation on its requirements, configuration, deployment, and execution, as well as links to external resources containing information and guides about related concepts and services.

Ainur makes a number of assumptions about its execution environment.
Apart from generic ones regarding the presence of necessary authentication information for remote access to local and remote hosts and services, the framework assumes a split network architecture.
Ainur runs in a dedicated control host and the management plane resides in a physically distinct network from the workload data.
This is a key requirement to be able to reconfigure the physical links in the workload data network without disrupting management traffic.

A full-featured, and rather complex deployment is depicted in \cref{fig:network}, including a number of optional testbed features such as offload to the cloud through \ac{VPN} gateways.
However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{fig:network:minimal}.
The only requirements for the aforementioned hosts are that a general purpose Linux distribution such as Ubuntu can be deployed on them, and --- in the case of the workload hosts --- the presence of two network interfaces.
Thus, a minimal testbed can be achieved cheaply and quickly by employing small single-board computers such as Raspberry Pi, Jetson Nano, or BeagleBone units.
Finally, the configuration and deployment of an Ainur-compatible testbed can be easily automated with tools such as Ansible, and playbooks for this effect can be made available to the community upon request.

% Also assumed is the existence of \ac{VPN} gateways to the cloud, time synchronization, and configured hostname lookups (possibly using a \ac{DNS} service).

### 2022-08-31 publications/2022Ainur/assumptions.tex
cleave and ainur

\subsection{Underlying Assumptions}\label{sec:assumptions}

In order to be able to provide as general a platform as possible for containerized workloads, and target a broad spectrum of applications and network configuration, the design of Ainur needs to make a number of assumptions about the underlying testbed platform.
In this section we will detail these assumptions, and try to provide some general context and reasoning behind each of them.

\subsubsection{Time synchronization}
Hosts on the testbed synchronize their clocks to a central \ac{NTP} server.
This is however only required for consistent time measurements \emph{across} devices; measurements with respect to the clock of a single device (e.g.\ \ac{RTT} measurements) do not require synchronization.

\subsubsection{Network architecture}
Ainur makes strong assumptions about the layout of the testbed network.
A general overview of these can be seen in \cref{fig:network}.

In particular, the network follows a split architecture composed of two sub-networks: a control network for control plane data, and a workload data network for workload payload traffic.
It is designed like this to
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item be able to dynamically reconfigure, segment, and route the workload data network for different experiments
    \item ensure consistent and accurate network traffic measurements (\acp{RTT}, bandwidth, etc.) on the workload data network, uncontaminated by control traffic.
\end{enumerate*}

\begin{description}[style=unboxed]
    \item[Control network]

    The control network, whose links are represented in orange in \cref{fig:network}, carries control plane data between the control host and the rest of the hosts on the testbed.
    It also provides internet access through a gateway router, and control plane access to the cloud instances through the control \ac{VPN} gateway (more details on this below).

    \item[Workload data network]

    This network carries data generated by the workloads deployed on the testbed, and is represented by blue links in \cref{fig:network}
    It also provides access to the workload data \ac{VPN} link to cloud instances through the workload data gateway.
    Although the physical links between devices are always present, interfaces on this network by default have no \ac{IP} addresses --- these are managed dynamically by Ainur.

    \item[Managed switch]

    Devices on the workload data network are interconnected through a managed switch.
    This switch is used by Ainur to change the virtual topology of the workload data network at runtime using \acp{VLAN}, for instance separating the network into front-end and back-end networks.

    In order to manage it through software, the managed switch is also connected to the control network through a single link.
    The associated interface is on a separate, exclusive \ac{VLAN}, and no data from the workload data network is allowed to enter the control network.

    \item[\ac{VPN} gateways]

    Two \ac{VPN} gateways are present on the testbed; one for control traffic, and one for workload data traffic to \ac{AWS} cloud instances.
    More details on these gateways are given in \cref{sec:layer3}.

\end{description}

\subsubsection{Separate control host}

Ainur is executed from a host separate from hosts where the workload is actually run.
This is to avoid issues that could arise from having the control plane residing on the same physical device as the workload data plane, such as traffic cross-contamination, routing issues, etc.

\subsubsection{Workload host network capabilities}

Hosts on which workloads are to be deployed must at the very least have \emph{two} network interfaces. 
One interface is connected to the control network for remote management, and the other to the workload data network, used for workload traffic.

Workload nodes may additionally have WiFi interfaces, but that is not a requirement.

\subsubsection{Software requirements}

Hosts controlled by Ainur are assumed to have a certain set of software packages installed on them.
Most of the below assumptions can be met simply by having hosts run a modern version of Ubuntu (18.04 or later, ideally 20.04).

\begin{description}[style=unboxed]
    % \item[\ac{APT}]

    % Hosts use \ac{APT} to manage installed packages.

    \item[Netplan]

    Hosts use Netplan\footnote{\url{https://netplan.io/}} to manage network connections and interfaces.

    \item[systemd]

    Hosts use systemd\footnote{\url{https://www.freedesktop.org/wiki/Software/systemd/}} to manage system configurations and services.

    \item[Docker]

    Docker\footnote{\url{https://docs.docker.com/}} is installed on the hosts.
    It is configured to allow regular users to run the Docker client and the server is set up to listen on unencrypted \ac{TCP} port \num{2375} on all interfaces, in addition to the default local UNIX socket.
\end{description}

\subsubsection{Authentication}
As Ainur interacts with hosts remotely, some assumptions are made about the access to these hosts.

\begin{description}[style=unboxed]
    \item[SSH access]

    The \ac{SSH} client on the control host and the \ac{SSH} servers on the controlled hosts (i.e.\ workload hosts, \ac{VPN} gateways) have been configured to allow password-less, public-key based authentication.

    \item[\texttt{sudo} access]

    Some of the actions executed on the remote hosts by Ainur require elevated privileges.
    The \texttt{sudo} command on the remote hosts is thus configured to allow password-less privilege escalation.

\end{description}

\subsubsection{Fluentbit logging}

A Fluentd\footnote{\url{https://www.fluentd.org/}} or Fluentbit\footnote{\url{https://fluentbit.io/}`'} server is running on a host in the control network.
At runtime, this server receives data streams of logs from all services deployed on the workload data network and collects them into log files.

\subsubsection{Cloud integration}

Cloud integration in Ainur is done in \ac{AWS}\footnote{\url{https://aws.amazon.com/}} and the associated \texttt{boto3}\footnote{\url{https://aws.amazon.com/sdk-for-python/}} Python library.

\begin{description}[style=unboxed]
    \item[Authentication to \ac{AWS}]

    Credentials to \ac{AWS} exist in the execution environment of Ainur on the control hosts.
    Specifically, these are credentials that allow Ainur to fully manage:
    \begin{itemize}
        \item \ac{EC2} instances, to create and terminate required compute nodes on-the-fly;
        \item security groups, to allow remote access over \ac{SSH} and \ac{VPN} to the compute nodes;
        \item and private access keys, to securely log in to the nodes over \ac{SSH}.
    \end{itemize}

    \item[\acp{AMI}]

    \Ac{EC2} instances deployed by Ainur are expected to be based on a Linux \ac{AMI} equipped with the similar software as the local hosts.
    Specifically:

    \begin{itemize}
        \item Docker is installed and configured for regular users and over-the-network access on \ac{TCP} port \num{2375}.
        \item Netplan and systemd are present and in use.
        \item VPNCloud\footnote{\url{https://vpncloud.ddswd.de/}} is installed (see below).
    \end{itemize}

    \item[\ac{VPN} gateways]

    In order to transparently integrate remote compute instances with the local testbed resources, Ainur uses VPNCLoud to configure tunnels to the cloud for control and workload data traffic.
    This allows the control host and hosts on the workload data network to treat cloud instances as if they resided locally on the control and workload data networks.

    \begin{description}[style=unboxed]
        \item[One gateway per network]

        Two gateways are present on the testbed networks, as illustrated in \cref{fig:network}; one for the control traffic \ac{VPN} link, and another for the workload data traffic \ac{VPN} link.

        \item[Public \ac{IP} address]

        Both gateways are reachable on the same public \ac{IP} address of the gateway router.
        The gateway router listens on two different \ac{UDP} ports and forwards data to the corresponding \ac{VPN} gateway.

        \item[Pre-shared keys]

        Both links share the same VPNCloud authentication parameters, in particular the pre-shared key used to initiate connections.

        \item[Routes]

        Static routes to the control \ac{VPN} link through the corresponding gateway are configured on all devices attached to the control network.

    \end{description} 
\end{description}

### 2022-08-31 publications/2022Ainur/conclusion.tex
cleave and ainur

\section{Conclusions and Future Directions}\label{sec:conclusion}

In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
    \item provisioning of and connection to remote cloud instances
    \item initialization of \ac{IP} layer connectivity between hosts
    \item collection of logs and data
    \item deployment, scaling, and lifecycle management of containerized processes
\end{enumerate*}.
We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\

In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \ac{CHI}~\cite{keahey2020lessons}.
While we believe Ainur to be unique and valuable in its focus on fully-automated end-to-end wireless edge computing experimentation, integrating with the above software stacks will provide a number of features and functionalities which will greatly expand Ainur's potential.
Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \ac{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\ac{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.

### 2022-08-31 publications/2022Ainur/demo.tex
cleave and ainur

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/demo_hardware.png}
    \caption[caption]{%
        The possible workload offloading setups and physical layer configurations used in the demo.
        Not pictured, in Phy Option 2:
        \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
            \item \ac{SDR} processing hosts 
            \item additional base-station host for the Core Network and \ac{eNodeB}
        \end{enumerate*}.
    }\label{fig:democonfigs}
\end{figure*}

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_1.png}
        \caption{Workload 1: \acs*{MNIST} image classifier}\label{fig:wkld:mnist}
    \end{subfigure}%
    \vspace{1em}
    \begin{subfigure}{\columnwidth}
        \centering
        \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_2.png}
        \caption{Workload 2: inverted pendulum \acs*{NCS}}\label{fig:wkld:ncs}
    \end{subfigure}
    \caption{Demonstration workloads}\label{fig:wkld}
\end{figure}

\section{Demo}\label{sec:demo}

In this demo, we will show the flexibility of Ainur for running end-to-end experimental workloads on both the Edge and the Cloud.
This will be done interactively, and members of the audience will be invited to propose testbed configurations.

\Cref{fig:democonfigs} illustrates the possible configurations for the testbed.
Workloads are deployed on client hosts, and computation is offloaded either to an edge server (cloudlet), or to \ac{AWS} \ac{EC2} instances on the cloud.
Communication between the client- and server-sides of the workload will occur over one of three possible physical link layer setups:
\begin{description}
    \item[WiFi:] \iac{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board WiFi.
    \item[4G \ac{LTE}:] \iac{SDR} is configured as an 4G \ac{LTE} base station, and another is configured as an 4G \ac{LTE} \ac{UE}.
    Together these radios act as an \ac{LTE} bridge between the client-side and the server-side of the network, and all client-server traffic is routed through them.
    \item[Ethernet:] connects everything through plain ethernet.
\end{description}

The number of workload instances (and therefore client hosts) deployed in each execution of this demonstration will range from \numrange[]{1}{10}.
Workloads deployed to the cloud will be able to target any of \ac{AWS}'s datacenters.

We will employ two different workloads for this demonstration, illustrated in \cref{fig:wkld}.
The first of these (\cref{fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \ac{MNIST} dataset~\cite{mnist}.
The application consists of a client with a web interface, and \iac{HTTP} server which responds to requests from the client and performs the actual image recognition.
This workload is intended to showcase in an interactive manner the effects of placing computation at different points of the network, and how Ainur simplifies these deployments.

The second workload (\cref{fig:wkld:ncs}) corresponds to \iac{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \acp{NCS} using cloud-native technologies~\cite{cleave}.
It consists of an emulation of the physical inverted pendulum system plant and a software-implemented proportional-differential controller.
These components communicate with each other over the \ac{UDP}.
This workload will be used to showcase the utility of Ainur for automating the execution of batches of experiments potentially including multiple different clients, servers, and physical layers.

\subsection{Testbed Setup}

This demonstration will be performed on a testbed consisting of
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and finally }}]
    \item \num{10} Raspberry Pi 4 Model B boards, acting as client-side workload hosts
    \item a Raspberry Pi 4 Model B acting as \ac{VPN} gateway for the workload network
    \item a Raspberry Pi 4 Model B acting as \ac{VPN} gateway for the management network, as well as hosting the necessary \ac{NTP} and \ac{DNS} server software
    \item a \texttt{i386} workstation, acting as an edge-side workload host
    \item a configurable number of  \ac{AWS} \ac{EC2} cloud instances acting as cloud servers
    \item a separate \texttt{i386} workstation hosting the Fluent server and on which Ainur is deployed as well.
\end{enumerate*}
These nodes are interconnected using a combination of managed switches, \acp{SDR}, and \ac{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.

\subsection{Demo Procedure}

\subsubsection{Workload 1}

\begin{figure}
    \centering
    \includegraphics[width=.65\columnwidth]{publications/2022Ainur/figures/demo_mockup.png}
    \caption{Screenshot of the user interface of demo workload 1. After deciding on the placement of the compute backend and the type of physical layer connecting client and backend, participants will use this interface to interact with the system.}\label{fig:mockup}
\end{figure}

Participants will be asked to choose
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item a location to offload computation to (local (i.e.\ no offloading), edge, or any \ac{AWS} datacenter)
    \item a physical layer for the first hop of the network (WiFi, 4G \ac{LTE}, or ethernet). 
\end{enumerate*}
Through the Ainur command line, we will deploy a single client-server pair according to the specifications.
Next, participants will be able to access the client web interface and interact with the server by requesting classification of images from the \ac{MNIST} database.
The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{fig:mockup}.

\subsubsection{Workload 2}

Participants will be asked to specify the full, arbitrarily complex, experimental scenario, including the number of clients, physical layers (single or multiple, and which clients are on each physical link), and offloading configuration (number of instances on the edge vs.\ on the cloud, which \ac{AWS} datacenters to deploy to).
Configuration will be specified through a \acs{YAML} file, which will then be parsed by Ainur for the automatic execution of the scenario.

### 2022-08-31 publications/2022Ainur/futurework.tex
cleave and ainur

\section{Future Work (ExPECA 2.0)}

\begin{description}[wide,style=nextline]
    \item[Client-Server Architecture and Microservices]
    Our current design follows a monolithic approach, in which ``layers'' of software build upon each other to reach a desired state.
    This approach is easy to understand and somewhat straightforward to implement, and works well for end-to-end experimentation.
    However, it causes a couple of problems, for example:

    \begin{enumerate}
        \item Components in the implementation are quite tightly coupled, so adding new functionality or layers in many cases implies rewritting large chunks of code.
        This also requires quite a lot of synchronization between people working on different parts of the code.
        \item It makes experimentation synchronous, and when running things over \ac{SSH}, it requires the use of tools such as \texttt{screen} or \texttt{tmux} to be able to run long experiments without having to keep the connection open at all times.
        \item We currently have no way of ensuring mutual exclusion of multiple instances of Ainur on the testbed, and implementing such functionality would probably require some sort of centralized instance tracking mechanism.
    \end{enumerate}

    I (Manuel) suggest then that for ExPECA 2.0 we switch to a distributed client-server architecture, using a microservice for each core functionality.
    In other words, we would have several smaller server applications running on potentially different hosts on the network, each handling a specific part of the functionality we need.
    For instance, we could have a ``Layer 3'' service which handles setting up \ac{IP}-layer connectivity, and another ``SDR'' service which manages the \acp{SDR}.
    Services could communicate with each other using simple HTTP \acp{API} for easy debugging, and we would have an ``Ainur'' client which orchestrates the services into an experimental run.

    This has several advantages:

    \begin{enumerate}
        \item It decouples the implementation details for each service.
        As long as the \ac{API} is well-defined, it doesn't matter how the service is implemented - we could even use different languages for different functionalities.
        \item Adding new functionality becomes simply a matter of writing a new microservice and plugging it into the existing setup.
        \item Experimentation can be done asynchronously through an ``experiment scheduler'' service which queues, runs, and then collects results from experiments.
        This service would also then ensure mutual exclusion of testbed resources.
        \item Things could be parallelized.
        Our current design is sequential, even when it doesn't need to be.
        Things that do not depend on each other could be executed in parallel (e.g. cloud and local configuration), cutting down on experiment time overhead.
        \item It would fit squarely into the design of both Docker and Kubernetes, both of which are distributed client-service architectures.
    \end{enumerate}

    \item[OpenStack/Canonical MaaS]

    We should really consider switching to a mature framework for bare-metal provisioning instead of doing it manually with Ansible.
    Potential options are OpenStack Ironic and Canonical Metal-as-a-Service.

    \item[Raspberry Pi SSDs]

    Currently our Raspberry Pis boot Ubuntu 20.04 LTS from SD cards; however this is very taxing on these storage devices, and their estimated lifetime doing this is less than a year.
    We should consider switching to the slightly more expensive option of booting the Raspberry Pis from solid-state drives.
    An alternative which I am studying would be to boot all the Raspberry Pis from a SINGLE shared storage device using iSCI and network booting (PXE).

\end{description}

% - microservices
% - openstack
% - raspberry pi rack, shared storage?

### 2022-08-31 publications/2022Ainur/introduction.tex
cleave and ainur

\section{Introduction}
Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications. 
On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.

Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the 
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item \acs*{COSMOS}
    \item \acs*{POWDER}
    \item \acs*{ARA}
    \item Drexel Grid \ac{SDR}
\end{enumerate*} testbeds.
\acs{COSMOS}~(\emph{\acl{COSMOS}})\acused{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \acp{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
\ac{COSMOS} relies on the \ac{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \ac{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.

\acs{POWDER}~(\emph{\acl{POWDER}})\acused{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes. 
\ac{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \ac{VM} images containing the necessary software and configurations.
These profiles are defined through using \ac{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.

The \ac{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
It includes a broad range of wireless platforms ranging from low-\ac{UHF} massive \ac{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \ac{COTS} radios, as well as automated ground vehicles, cameras and sensors.
\ac{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \ac{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
This affords the \ac{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.

Finally, the \emph{Drexel Grid \ac{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
Primarily intended for \ac{SDR}-centric research, it does not integrate any core, cloud or edge components.
However, the testbed extensively employs the \ac{LXC} runtime for the deployment of both experimental code and \ac{SDR} software, which affords users great freedom when it comes to the development of experiments.

Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
This is, for instance, the case with \ac{COSMOS} and \ac{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.

To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \acp{VM} containing the necessary software for experiments.
Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \iac{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.

In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
Furthermore, we make Ainur available to the community as \ac{FOSS}.
It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.

The rest of this paper is structured as follows.
\cref{sec:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
Next, \cref{sec:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
Finally, in \cref{sec:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.

% \todo[inline]{
%     COSMOS: Experimenters work on these resources using the \ac*{OMF}~\cite{omf} tool, originally leveraged for ORBIT~\cite{orbit}, and \ac{SSH} sessions, thus allowing for free-form scripting of experimental procedures.

%     CONS: OMF/OEDL is domain specific. Requires learning a new language. High barrier of entry, does not integrate with existing technologies.
%     Imperative specification.

%     POWDER: To abstract and redistribute resources and \acp{API}, \ac{POWDER} leverages the concept of \emph{profiles}, corresponding shareable packages containing experimental configurations and descriptions\cite{emulab}. 
%     Automation and web interface control are limited to some test experiments, and thus regular users must interact with the platform through \ac{SSH}.

%     Profiles: complex description of an experiment. Can be produced from a Python script. Experiments built on top of VMs or bare metal. Uses OpenStack.

%     Drexel: Orchestration in this testbed is facilitated through a container-based approach, leveraging LXC containers, while deployment and configuration is facilitated through Ansible.
%     Container deployment is requested by users through a web-based reservation system or a \ac{REST} web \ac{API}.
%     \ac{OS} deployment on host computers is facilitated through Canonical \ac{MaaS}.

%     REST web API, containers for access to hardware. Metal-as-a-Service and Ansible for managing hardware (but not for API I think)
% }

### 2022-08-31 publications/2022Ainur/paper.tex
cleave and ainur

\documentclass[conference,10pt,letterpaper,twocolumn]{IEEEtran}

% begin packages
% ==============
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[htt]{hyphenat}  % for linebreaks in texttt
\usepackage[pdftex]{graphicx}
% \graphicspath{{./images/}{./plots/}{./figures}}
% \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\usepackage{amsmath}
\usepackage{url}
\usepackage{orcidlink}
\usepackage[inline]{enumitem}
\setlist[description]{font=\itshape\underline}
\usepackage{todonotes}
\usepackage{caption}
\usepackage{subcaption}
% \usepackage[newfloat=true]{minted}
% \setminted{breaklines=true, frame=lines}
% \usemintedstyle{xcode}
% \SetupFloatingEnvironment{listing}{name=List.,within=none}
% \captionsetup[table]{justification=centerlast,
%                      labelsep=newline,
%                      font=sf,
%                      textfont=footnotesize}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage[detect-weight=true,detect-family=true,binary-units=true,list-units=single,range-units=single]{siunitx}
\usepackage[nolist]{acronym}
\input{acronyms.tex}

% references
\usepackage[
  style=numeric-comp,
  sorting=none,
  sortcites,
  hyperref,
  mincitenames=1,
  maxcitenames=2,
  maxbibnames=2,
  minbibnames=1,
  citestyle=numeric-comp, % for [1, 2] instead of [1], [2]
  backend=bibtex
]{biblatex}
\bibliography{references.bib}
\AtBeginBibliography{\small}
\AtEveryBibitem{\clearfield{day}}
\AtEveryBibitem{\clearfield{isbn}}
% \AtEveryBibitem{\clearfield{url}}
\AtEveryBibitem{\clearfield{series}}
\AtEveryBibitem{\clearlist{location}}
\AtEveryBibitem{\clearfield{doi}}

% for dealing with ACM's stupid bibtex requirements
% \usepackage{biblatex2bibitem}
% \usepackage[nosort,compress]{cite}

\usepackage{easyReview}
\usepackage{outline}
\usepackage{orcidlink}
\usepackage[all]{hypcap}
\usepackage{makecell}
\usepackage[capitalize,nameinlink,noabbrev]{cleveref}
% \crefname{subfigure}{Subfigure}{Subfigures}
% \Crefname{subfigure}{Subfigure}{Subfigures}
% \crefname{listing}{Listing}{Listings}
% \Crefname{listing}{Listing}{Listings}

\hypersetup{
  hidelinks,
  colorlinks=true,
  allcolors=black,
  pdfstartview=Fit,
  breaklinks=true
}

\newcommand{\kthemail}[1]{\href{mailto:#1@kth.se}{#1}}

\begin{document}
\title{Ainur:~A Framework for Repeatable End-to-End Wireless Edge Computing Testbed Research}

\author{%
  \IEEEauthorblockN{%
  Manuel {Olguín Muñoz}\IEEEauthorrefmark{1}~\orcidlink{0000-0002-3383-2335}, Seyed Samie Mostafavi\IEEEauthorrefmark{2}~\orcidlink{0000-0001-9316-0414}, Vishnu N. Moothedath\IEEEauthorrefmark{3}~\orcidlink{0000-0002-2739-5060}, James Gross\IEEEauthorrefmark{4}~\orcidlink{0000-0001-6682-6559}%
  }%
  \IEEEauthorblockA{%
    School of Electrical Engineering \& Computer Science\\%
    KTH Royal Institute of Technology, Sweden\\%
    Email:~\{\IEEEauthorrefmark{1}\kthemail{molguin}, \IEEEauthorrefmark{2}\kthemail{ssmos}, \IEEEauthorrefmark{3}\kthemail{vnmo}, \IEEEauthorrefmark{4}\kthemail{jamesgr}\}\href{mailto:molguin@kth.se}{@kth.se}
    % \IEEEauthorrefmark{1}Corresponding author.%
  }
}
\maketitle

\input{abstract.tex}
\input{introduction}
\input{ainur}
\input{demo}
\input{conclusion}
\input{acks.tex}
% \input{futurework}

\printbibliography{}
\end{document}

### 2022-08-31 publications/2022Ainur/software_layers/cloud.tex
cleave and ainur

\subsubsection{Cloud Instances}\label{sec:cloud}

Integration with \ac{AWS} is handled by the \mintinline{python}{CloudInstances} class.
The implementation of this class uses \verb|boto3| for interaction with the \ac{AWS}, and thus needs access to appropriate credentials\footnote{%
    \url{https://boto3.amazonaws.com/v1/documentation/api/latest/guide/credentials.html}%
}.

Each instance of this class is associated with a specific \ac{AWS} region, and manages the end-to-end lifecycle of an arbitrary number of \ac{AWS} \ac{EC2} instances on said region.
Specifically, \mintinline{python}{CloudInstances} handles
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item configuring the necessary prerequisites for deploying the instances (security groups and private access keys)
    \item deploying the instances and waiting for them to successfully boot
    \item shutting down and terminating the instances
    \item cleaning up access keys and security groups
\end{enumerate*}.
Instances of this class can be used as context managers, automating this way the termination of instances and cleanup of resources when execution exits the \mintinline{python}{with} block.

By default, \mintinline{python}{CloudInstances} spawns instances of type \verb|t3.micro|, with at least one security group allowing inbound \ac{SSH} traffic and unrestricted traffic between instances.
The type of instances spawned can be modified at runtime, and additional security groups can be added at will.

The spawned \ac{EC2} instances are created from a preconfigured \ac{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
This \ac{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \ac{AMI} --- or a custom \ac{AMI} which has been preconfigured by the user.
Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \ac{AMI} used for deploying (with the exception that it needs to be a Linux-based \ac{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.

### 2022-08-31 publications/2022Ainur/software_layers/flow.tex
cleave and ainur

\begin{figure}
    \centering
    \includegraphics[width=.7\columnwidth]{figures/flow.png}
    \caption{Execution flow of Ainur.}\label{fig:flow}
\end{figure}

\Cref{fig:flow} illustrates the execution flow of Ainur.
The software follows a mostly linear approach given its direct interaction with distinct layers of hardware and networking.
Immediately after start, Ainur reads and parses the deployment configuration and parses it into a set of various dataclasses and Python dictionaries.
Next, if deploying anything to the cloud, Ainur spawns the required number of \ac{AWS} \ac{EC2} instance.
Execution then goes on to configure the physical links, connecting hosts with each other through clever management of \acp{VLAN} and \acp{SDR}.
Once physical and/or \ac{VLAN} connections are configure, \ac{IP} addresses are assigned to devices and static routes between them are configured.
If cloud instances are used, \ac{VPN} connections are established to them.
At this stage, the testbed network is fully configured, and bidirectional traffic between any two hosts is possible over the workload data network.

The next stages deal with workload deployment, scaling, and orchestration.
A \emph{Docker Swarm}\footnote{\url{https://docs.docker.com/engine/swarm/}} is deployed on the relevant hosts on the workload data network.
On top of this Swarm, required container images are fetched from a remote repository, and distributed storage is configured across all workload hosts.
Finally, the actual containerized workload is deployed as a Docker Swarm Service Stack\footnote{\url{https://docs.docker.com/engine/swarm/stack-deploy/}}, allowing us to leverage the flexibility and utilities of this orchestration platform for automatic scaling, placement, health checking, and coordination of workloads.

In terms of actual code, in practice, most steps in \cref{fig:flow} are implemented in software as abstractions of ``layers'': each step builds upon previous steps in the flow, and provides the prerequisites steps down the line expect.
Our Python implementations of these components also frequently use the \emph{context manager}\footnote{%
    For reference, see the Python documentation on \url{https://docs.python.org/3/library/contextlib.html} and associated pages.
} pattern to ensure clean teardown of the state of the testbed and the end of the execution of Ainur.

### 2022-08-31 publications/2022Ainur/software_layers/ip_layer.tex
cleave and ainur

\subsubsection{\ac{IP} Layer Connectivity}\label{sec:layer3}

Layer 3 connectivity, that is, \ac{IP} addressing and routing, is handled by two main components in Ainur: \mintinline{python}{LANLayer} and \mintinline{python}{VPNCloudMesh}.

\mintinline{python}{LANLayer} handles local network connectivity.
It can be used as a context manager for automatic network teardown, as well as an iterator over the hosts it manages.
This class exposes two main \ac{API} methods, \mintinline{python}{self.add_hosts()} and \mintinline{python}{self.tear_down()}:
\begin{description}[]
    \item[\mintinline{python}{LANLayer.add_hosts(self, layer2)}]
    This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \ac{IP} layer connectivity on the hosts within.
    This includes assigning \ac{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
    The underlying implementation uses Ansible to write Netplan YAML configuration files on the target hosts and subsequently apply the network configuration.

    \item[\mintinline{python}{LANLayer.tear_down(self)}]
    Performs a clean teardown of \ac{IP} layer connectivity of the hosts managed by the instance, removing the created Netplan configuration files and reverting host network configuration to its initial state.
    When \mintinline{python}{LANLayer} is used as a context manager, this method is automatically called the moment execution leaves the \mintinline{python}{with} block.
\end{description}

\mintinline{python}{VPNCloudMesh}\label{sec:vpn} handles \ac{VPN} connectivity to running cloud instances.
It makes some strong assumptions about the presence of two preconfigured \ac{VPN} gateways on the testbed (some of these have been mentioned above, but we will repeat them here for clarity):

\begin{itemize}
    \item Two VPNCloud gateways, one for control connections and one for workload data streams, are configured and accessible on the control network.
    \item One of these gateways is additionally connected to the workload data network through a separate interface.
    \item These gateways are both listening on separate \ac{UDP} ports on a single public IP address.
    In other words, both gateways access the public internet through a single \ac{NAT} gateway, on which appropriate \ac{UDP} port forwarding rules are configured to allow \ac{VPN} traffic back into the control network.
    \item The VPNCloud connections configured on the the gateways share the same pre-shared key.
    \item The VPNCloud links belong to mutually \emph{disjoint} \acs{IP}v4 address ranges.
    % \item The gateways are assigned the first host address in the available address space of their respective \ac{VPN} links.
    % For example, if the gateway \ac{VPN} link address space is \texttt{10.0.0.0/24}, the gateway has address \texttt{10.0.0.1} on the \ac{VPN} link.
    \item Devices on the control network have been preconfigured with appropriate \ac{IP} routes to reach the control gateway and its associated \ac{VPN} link.
    \item Finally, cloud instances intended to be used with this class have VPNCloud installed on them.
\end{itemize}

After spawning cloud instances with an instance of the \mintinline{python}{CloudInstances} class described in \cref{sec:cloud}, the \mintinline{python}{VPNCloudMesh} class can be used to connect said cloud instances to the control and workload data networks of the testbed.
Instances of this class can further be used as context managers for automatic cleanup of stale links, and implement an iterator interface to access connected hosts.

The \ac{API} of this class exposes the below relevant methods:

\begin{description}[]
    \item[The \mintinline{python}{VPNCloudMesh} constructor]

    The constructor takes the following arguments:

    \begin{description}[font=\normalfont\ttfamily\underline]
        \item[gateway\_ip] An \texttt{IPv4Address} object corresponding to the shared public \ac{IP}v4 address of the \ac{VPN} gateways (i.e.\ the public \ac{IP} address of the \ac{NAT} gateway.)

        \item[vpn\_psk]
        The pre-shared key used to authenticate and encrypt traffic on the \ac{VPN} links.

        \item[ansible\_ctx~\normalfont{and} ansible\_quiet]
        Correspond to an Ansible context to use for the execution of this instances' methods (see above), and a boolean flag to suppress verbose debug output, respectively.

        \item[gw\_mgmt\_ip~\normalfont{and} gw\_wkld\_ip]
        Full \ac{IP} addresses of the \ac{VPN} links for the control and workload data networks, respectively, including subnet masks.
        These values are passed as \mintinline[breakafter=4]{python}{IPv4Interface} objects, and are used to set up correct routing tables to the cloud instances.

        \item[mgmt\_local\_net~\normalfont{and} wkld\_local\_net]
        The base addresses of the local control and workload data networks, respectively, as \mintinline[breakafter=4]{python}{IPv4Network}.
        Used to configure routing back from the cloud.

        \item[mgmt\_port~\normalfont{and} wkld\_port]
        \ac{UDP} ports on the \ac{NAT} gateway forwarded to the control and workload gateways, respectively.
    \end{description}

    \item[\mintinline{python}{VPNCloudMesh.connect_cloud(self, ...)}]

    The full signature of this method is \mintinline[breakafter=.,breakafter=_]{python}{VPNCloudMesh.connect_cloud(self, cloud_layer, host_configs)}.
    The \texttt{cloud\_layer} argument corresponds to a \mintinline{python}{CloudInstances} object instance, and the \texttt{host\_configs} to a collection of host configuration dataclasses.
    This method matches the host configurations to cloud instances in the \mintinline{python}{CloudInstances} object and connects them to to the local control and workload data networks through the two aforementioned separate \ac{VPN} links.
    The underlying implementation uses a combination of the \texttt{boto3} library and Ansible to configure appropriate security groups on the instances and subsequently initialize and connect the \ac{VPN} links.

    \item[\mintinline{python}{VPNCloudMesh.tear_down(self)}]
    This method tears down the \ac{VPN} links, disconnecting the remote instances from the gateways and removing any created routes to the cloud from local devices.
    When an instance of this class is used as a context manager, this method is automatically called at the end of the \mintinline{python}{with} block, ensuring automatic cleanup of resources.

\end{description}

### 2022-08-31 publications/2022Ainur/software_layers/physical.tex
cleave and ainur

\subsubsection{Physical Layer}

This layer creates and deploy the underlying physical connections of the data network and is implemented in the \mintinline{python}{PhysicalLayer} class. 
Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \ac{LTE} network.

In order to ensure flexibility, we make no assumptions on the wired connections between the data hosts or radios.
On the other hand, an arbitrary data plane could require wired connections between two hosts, or a host and a radio.
Therefore, we use a managed switch at the heart of the physical layer and connect all the hosts and radios to it.
According to the data network description, the physical layer then creates the required \acp{VLAN} in the switch.

At instantiation time, this class first configures ethernet connections by creating \acp{VLAN} in the workload data network switch.
It then triggers the deployment of potential wireless connections using \mintinline{python}{WirelessConnection} objects.

A set of \mintinline{python}{WirelessConnection} instances must be provided to the constructor for deployment.
This class can be used as a context manager in a \mintinline{python}{with} statement; at the end of the block, it tears down all connections and cleans the added \acp{VLAN} from the switch.

\subsubsection{Wireless Connections}
To integrate various wireless connection schemes, we have defined the Python interface \mintinline{python}{WirelessConnection}.
It unifies the interaction of Ainur with the wireless implementation in the following aspects: 
\begin{enumerate}
    \item \textbf{\mintinline{python}{start()} and \mintinline{python}{stop()} methods:} In addition to the constructor and destructor, the implementations need to override these methods which could be called by the \mintinline{python}{PhysicalLayer}.
    The constructor is used to set the configurations by the user and these methods handle the deployment.

    \item \textbf{Data gates:} Every wireless connection implementation must identify certain network interfaces whether on radios or hosts, to act as the gate for the data plane.
    Hence, any host can exchange packets through those gates if the appropriate VLANs are created on the data network managed switch.
\end{enumerate}

Currently, WiFi and \ac{LTE} implementations are available. 
User-defined implementation for other wireless connections are possible and encouraged.

We introduce how the current implementations work in the following.

\begin{description}[style=nextline]
    \item[SDR-based WiFi Network]
    The \mintinline{python}{SDRWifi} class handles the configuration and deployment of a software-defined IEEE 802.11 network using the Mangocomm software package and its supported software-defined radios \footnote{See https://mangocomm.com/802.11-mac-phy/}.
    It lunches a Docker container on the same machine to communicate with the \acp{SDR} over the control network.
    This communication is for the configuration and monitoring the WiFi nodes.
    In a WiFi network, a radio is either \emph{station} or \emph{access point}.
    To initialize a \mintinline{python}{SDRWifi} class, the user is required to provide two lists containing the access points and station nodes instances as described below.

    \mintinline{python}{APSoftwareDefinedRadio} class represents a WiFi access point.
    It advertises a network with a specific SSID on a radio channel and waits for the incoming connections.

    \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station WiFi nodes.
    It searches for an WiFi \ac{SSID} among the networks that are being advertised and attempts for establishing the connection.

    When a WiFi network is up and an station is connected, the data network interfaces of the radios behave as if an ethernet cable is connecting them.
    Thus, the client hosts and the server hosts can be on the same ip layer subnets.
    \begin{figure}[t]
        \centering
        \includegraphics[width=0.8\linewidth]{figures/wifi.png}
        \caption{An example of WiFi network topology in data network and its corresponding \acp{VLAN}}
        \label{fig:wifi}
    \end{figure}

    \item[\ac{SDR}-based \ac{LTE} Network]
    \mintinline{python}{SDRLTE} class can configure and deploy a complete software-defined \ac{LTE} network.
    The constructor must be provided with an \mintinline{python}{EPC} class, a set of \mintinline{python}{ENodeB} classes, and a set of \mintinline{python}{LTEUE} classes from the \emph{autoran} Python package\footnote{See https://github.com/samiemostafavi/autoran}.
    This package exploits the \emph{docker-py} framework to run containerized OpenAirInterface \ac{LTE} functions on remote hosts.
    Network configurations are needed in the form of Python dictionaries for the construction of \emph{autoran} classes.

    \ac{LTE} \ac{EPC} consists of HSS, MME, SPGW-U, and SPGW-C services which can be run on the same host.
    For each \ac{eNodeB} or \ac{UE}, a software-defined radio and a host for signal processing must be specified additionally.
    The resulting topology of an \ac{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
    Unlike WiFi wireless connection, due to the fact that \ac{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.

    \begin{figure}[t]
        \centering
        \includegraphics[width=0.8\linewidth]{figures/lte.png}
        \caption{An example of \ac{LTE} data network topology in data network and its corresponding \acp{VLAN}}
        \label{fig:lte}
    \end{figure}

\end{description}

### 2022-08-31 publications/2022Ainur/software_layers/storage_logging.tex
cleave and ainur

\subsubsection{Storage}\label{sec:storage}

Dynamic, distributed storage shared across testbed hosts is managed by the \mintinline{python}{ExperimentStorage} class.
This class exposes a very simple \ac{API}:
\begin{itemize}
    \item At instantiation time, the instance deploys a samba\footnote{\url{https://www.samba.org/}} server container on a specified host, using a host folder mounted inside the container for persistent storage.
    Then, on each host configured on the workload data network, it configures a Docker Volume\footnote{\url{https://docs.docker.com/storage/volumes/}} pointing to the samba server.
    These Docker Volumes are only locally available, but since they all share the same identifier, they can be included in deployment configurations and mounted on applications orchestrated with Docker Swarm (see \cref{sec:swarm} below).

    The above operations are achieved using a combination of Ansible and the \texttt{docker-py} library.

    \item The server and local volumes can all be cleanly torn down using the \mintinline{python}{tear_down()} method of the class.
    This method is also automatically called at the end of the block if this class is used in a \mintinline{python}{with} statement.
\end{itemize}

\subsubsection{Logging}\label{sec:logging}

Logging of the experiment is carried out using a containerized approach.
This is achieved through the combination of \texttt{docker-py} and \textit{FluentBit} -- an open-source and multi-platform log processor and forwarding tool.
At deployment time, Ainur configures a FluentBit client container in each workload host which captures logs produced by the workload.
These FluentBit clients are configured to listen on a \ac{TCP} port to which workload logs are forwarded by Dockers built-in FluentD logging driver\footnote{\url{https://docs.docker.com/config/containers/logging/fluentd/}}.
This means that logs are automatically captured by Docker from the \emph{standard output} of the containerized workload process.
The FluentBit client containers append metadata such as timestamp, workload name, workload host ID, to the collected logs and forward them to a central FluentBit or FluentD server.

% On the other hand, the Fluent Bit server container is configured to run persistently in the background at the logging server, unless and until a restart is necessary for some development, debug or error handling purposes. They listen to a pre-defined fluent forward port for log messages from arbitrary number of fluent clients. These logs are then segregated based on various parameters like the workload name, client name and/or log category, and are accordingly written to files in a directory structure determined by the metadata.

### 2022-08-31 publications/2022Ainur/software_layers/swarm.tex
cleave and ainur

\subsubsection{Swarm Layer}\label{sec:swarm}

\begin{listing}[tb]
\caption{Example usage of the \mintinline{python}{DockerSwarm} class.}\label{lst:swarm:example}
\begin{minted}{python}
with DockerSwarm() as swarm:
    swarm.deploy_managers(
        hosts={
            cloudlet: {'location': 'edge'},
            aws01: {'location': 'cloud'}
        }
    )
    swarm.deploy_workers(
        hosts={
            client1: {},
            client2: {}
        },
        location='user'
    )
    swarm.pull_image('ubuntu', tag='20.04')
    swarm.deploy_workload(
        specification=workload_spec
    )
\end{minted}
\end{listing}

This layer handles the configuration and deployment of a Docker Swarm over the controlled hosts (including cloud instances) for the orchestration of containerized workloads.
It is implemented as a class \mintinline{python}{DockerSwarm} with appropriate methods for
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item connecting and configuring manager and worker nodes to the Swarm
    \item fetching required Docker container images to all configured nodes
    \item finally, deploying distributed workloads as stacks of Docker containers
\end{enumerate*}.
Additionally, the class defines appropriate \mintinline{python}{__enter__} and \mintinline{python}{__exit__} magic methods allowing it to be used as a context manager.

A simple example of the usage of this class and its basic \ac{API} can be observed in \cref{lst:swarm:example}, and in the following we will detail the major steps in operating this class:
\begin{description}[]
    \item[Instantiating a \mintinline{python}{DockerSwarm} object.]
    Obtaining an instance of a Swarm object is straightforward, as the constructor takes no parameters.
    \item[Adding Swarm nodes.]
    Swarm manager and worker nodes are added through the \mintinline{python}{deploy_managers()} and \mintinline{python}{deploy_workers()} methods respectively.
    These methods take a single mandatory parameter \mintinline{python}{hosts}, corresponding a mapping from \mintinline{python}{AinurHost} instances to dictionaries of node variables in the form \mintinline[breaklines, breakafter={_}]{python}{{'label_name': 'label_value'}}.
    Additional keyword arguments are parsed as default values for labels which are assigned to all nodes.

    Note that the actual underlying Docker Swarm cluster is created on-demand at the time of the first call to \mintinline{python}{deploy_managers()}.
    This means that at least one call to \mintinline{python}{deploy_managers()} \emph{must} be made before any call to any other method of the Swarm.
    \item[Fetching required container images.]
    \mintinline{python}{DockerSwarm} provides a \mintinline{python}{pull_images()} method which takes a mandatory repository parameter, as well as an optional tag identifier, and triggers a download of the specified container image on all nodes in the Swarm.
    This method supports all repository and tag formats supported by the \verb|docker-py| library, and defaults to fetching from Docker Hub\footnote{\url{https://hub.docker.com/}} when the given repository name string does not include remote registry information.
    \item[Workload Deployment.]
    Finally the Swarm object also provides a method \mintinline[breaklines, breakafter={_}]{python}{deploy_workload()}, which takes a \emph{workload definition} and deploys the workload on the Swarm.
    Details on the workload definition and running workloads are given in \cref{sec:workload} 
\end{description}

### 2022-08-31 publications/2022Ainur/software_layers/workload.tex
cleave and ainur

\subsubsection{Workloads}\label{sec:workload}

\begin{listing}[tb]
\caption{Workload specification example.}\label{lst:compose}
\begin{minted}{yaml}
name: FooBar
duration: '3h 15m'
version: '1.2a'
# ...
compose:
  version: '3.9'
  services:
  fooServer:
    image: 'foo:server'
    hostname: 'server.{{.Task.Slot}}'
    command: '/usr/bin/server'
    environment:
      PORT: 5000
    deploy:
      replicas: 5
      placement:
        constraints:
        - "node.labels.type==cloudlet"
    volumes:
    - type: volume
      source: FooBar
      target: /opt/foobar_data
      volume:
        nocopy: true
    # ...
\end{minted}
\end{listing}

% As previously explained, in the context of Ainur and ExPECA, we use the term \emph{workload} to refer to a collection of containerized, distributed applications deployed on the testbed for a specific amount of time.
% Ainur employ Docker Swarm for this purpose, thus leveraging

Workloads in Ainur are defined using a dictionary of metadata containing a nested \verb|docker-compose| definition, usually provided as a YAML file.
This dictionary is parsed into a \mintinline{python}{WorkloadSpecification} dataclass, which validates the structure of the definition, and which in turn is passed to a \mintinline{python}{DockerSwarm} class instance to actually deploy the workload.

In terms of structure, the workload definition dictionary is expected to contain the following keys:

\begin{description}
    \item[name]
    A string representing an identifier for the workload, used to tag logs and stored data, as well as to identify the generated dynamic distributed storage volumes.
    All leading and trailing whitespace will be trimmed from this value at parse time.
    \item[author]
    An identifier for the author of the workload specification.
    Optional, used in logs.
    \item[email]
    Contact email for the author.
    Optional, used in logs.
    \item[url]
    URL for information related to the workload; optional.
    \item[version]
    String representing a version tag for this workload.
    Optional, but defaults to \texttt{1.0} if not specified, and is parsed into a Python \mintinline{python}{LooseVersion}object for easy manipulation in code.
    \item[max\_duration]
    Represents the maximum duration for the workload, and can be specified as any string value supported by the \texttt{pytimeparse} library\footnote{\url{https://pypi.org/project/pytimeparse/}}.
    For instance, a maximum duration of \SI{1}{\hour} and \SI{30}{\minute} could be specified as any of the following strings: \mintinline{python}{'1h 30m'}, \mintinline{python}{'1.5 h'}, \mintinline{python}{'90m'}, \mintinline{python}{'5400s'}, and so on --- see the \verb|pytimeparse| library documentation for more supported formats.

    This value is used as a timeout for the workload.
    The \mintinline{python}{DockerSwarm} in charge of the execution and orchestration keeps track of the time the workload has been running, forcefully tearing it down if it does not exit cleanly before it the timeout.
    This prevents workloads stuck in non-ending loops from running forever and hogging up testbed resources.

    Optional, defaults to a very liberal \mintinline{python}{'1d'} (i.e.\ \SI{24}{\hour}) if not explicitly specified.
    Note that users are strongly encouraged to explicitly define this value, as the default is subject to arbitrary change as the usage patterns of the testbed evolve.

    \item[compose]

    Corresponds to a nested \texttt{docker-compose v3}\footnote{\url{https://docs.docker.com/compose/compose-file/compose-file-v3/}} definition, and is the single most important component of the workload specification.
    This definition must be compliant with at least version 3 of the \texttt{docker-compose} file specification standard.
    It must additionally be fully compatible with Docker Swarm service stacks\footnote{%
        \url{https://docs.docker.com/engine/reference/commandline/stack_deploy/}%
    }, as the underlying implementation of the \mintinline{python}{DockerSwarm} class uses the \texttt{docker stack deploy} command to deploy the workload on the cluster.
    Each service in this compose specification must contain a \texttt{deploy} configuration segment detailing the scaling and placement requirements for it.
    Finally, if distributed storage has been configured using the components detailed in \cref{sec:storage}, it should be mounted on the appropriate services.
    Examples of these configurations can be seen in \cref{lst:compose}; for more details refer to the \texttt{docker-compose} online documentation.
\end{description}

### 2022-08-31 publications/2022CLEAVE/cleave.tex
cleave and ainur

\section{The \ac{CLEAVE} framework}\label{sec:approach}

In this section, we detail our \ac{CLEAVE} framework for the performance evaluation of \aclp{NCS}, with an emphasis on edge deployments.
Our design follows a virtualized emulation approach; providing a framework for the real-time \emph{emulation} of physical control systems and the easy implementation of softwarized controllers interacting with \emph{real} networks.
\ac{CLEAVE} is thus a platform with an easy-to-use \ac{API} on which a multitude of \ac{NCS} types can be emulated.
This allows for great flexibility, as the core components of the \ac{NCS} can be switched out while maintaining the realism of the network, the most complex and limiting component of edge \acp{NCS}.

The framework is distributed as \emph{free and open-source} software through the \emph{GitHub} organization of the \emph{KTH ExPECA} research group\footnote{\url{https://github.com/KTH-EXPECA/CLEAVE}}.
We also plan to provide a library of \ac{NCS} implementations, making the tool even more accessible.

\subsection{Design and Implementation}\label{sec:design}

The design of \ac{CLEAVE} attempts to target a broad a set of different control systems as possible, and tries to minimize the number of assumptions made about the systems implemented on top of the framework.
These are:
\begin{enumerate}
    \item Controllers can be stateful or stateless, but they exclusively control a single plant.
    \item Plants and controllers interact in a ``request-response'' manner.
    We currently don't support setups in which the controller generates an actuation command without there first being a request from the plant.
    \item Controller and plant communicate over a TCP/IP network.
    \item The physical state behavior of the plant can be implemented in a discrete-time manner.
    \item Values associated to sensors and actuators can be represented as integers, floating point numbers, booleans, or sequences of bytes.
    \item A plant has zero or more sensors attached to it, with potentially different sampling rates.
    \item A plant has zero or more actuators attached to it.
\end{enumerate}

\cref{fig:cleave:ncs:struct} shows an overview of the conceptual structure and operation of \ac{CLEAVE}.
The \emph{State} implements the discrete-time behavior of the Plant.
At the beginning of each time-step, actuated variables in the \emph{State} are updated with values obtained from the \emph{Actuator} objects.
Conversely, at the end of each time-step, sensed variables are sampled and used to update values associated with \emph{Sensor} objects.
These values are processed at the \emph{Sensors} (e.g.\ to add noise) before being sent over the network to the \emph{Controller}.
The \emph{Controller} calculates actuation values and updates the \emph{Actuators} over the network.
Finally, the \emph{Actuators} process the actuation values and hold the results until they are read at the next time-step.

Implementation-wise, \ac{CLEAVE} is built on top of the well-established and stable \emph{Twisted}\footnote{\url{https://twistedmatrix.com}} asynchronous programming and networking framework.
The aforementioned core components (\emph{State}, \emph{Controller}, \emph{Sensors}, and \emph{Actuators}) are provided as \acp{ABC} with core functionality that users must extend for their specific use cases and \acp{NCS}.
The framework makes no assumptions about the inner workings of the user implementations of these classes other than the interfaces they expose.
Users are therefore free to implement arbitrarily complex functionality in these components.
On the other hand, \ac{CLEAVE} handles network communication and metric collection autonomously; users only need to select the desired transport protocol (we do however provide an \ac{API} for extending the available transports).

Below we present a brief overview of the design and implementation of the core components.

%\begin{description}[style=nextline]
%\begin{itemize}
\textbf{State Base Class}:
\ac{ABC} with an abstract method
\mintinline{python}{advance()} which must be extended with the implementation of discrete-time emulation of the plant.
The \mintinline{python}{advance()} method is called by the framework periodically according to the configured emulation update rate, and should perform a single time-step of the discrete-time emulation.
Two optionally extensible methods, \mintinline{python}{initialize()} and \mintinline{python}{shutdown()} are provided to implement single-fire set-up and tear-down procedures.

    % This class also provides functionality for the automatic tracking and updating of sensed and actuated values.
    % This is handled through special auxiliary classes, \emph{SensorVariable} and \emph{ActuatorVariable}, which when instantiated and assigned to an attribute of the \emph{State} allow the framework to track and modify them.

\textbf{Controller Base Class}:
Defines a single abstract method \mintinline{python}{process()}, taking a mapping of names to values corresponding to the sensed variables of the \emph{State}, and returning a similar mapping corresponding to the actuated variables.
This method is called by \ac{CLEAVE} in an event-based manner whenever samples are received from the plant side; however, we intend to extend this to time-interval-based approaches.

\textbf{Sensor Base Class}:
Implements processing of monitored stated variables before sending them over the network to the \emph{Controller}.
This base class must be initialized with a property name (corresponding to the monitored \emph{State} attribute) and a sampling rate in \si{\hertz}, and provides an abstract method \mintinline{python}{process_sample()} which must be extended with an implementation of the processing performed on each sample.

\textbf{Actuator Base Class}:
Performs the processing of actuation values obtained over the network from the \emph{Controller}.
It is initialized with a property name corresponding to the name of the actuated \emph{State} attribute, and provides two abstract methods which must be implemented by extending classes:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
\item \mintinline{python}{set_value()}, called by the framework whenever a new value for the corresponding actuated attribute is received
\item \mintinline{python}{get_actuation()}, called on every time-step and must return a value to apply to the actuated attribute.
\end{enumerate*}
%\end{itemize}
%\end{description}

\subsection{Configuring and Executing \iac{NCS}}

Once the core components for the emulation have been defined and implemented, they need to be registered with the framework.
This is done through Python scripts which define a number of required (and some optional) top-level variables, corresponding to emulation parameters such as the \emph{State}, \emph{Controller}, \emph{Sensor}, and \emph{Actuator} subclasses to use; the emulation update rate; the controller-side host and address; data output directories; \emph{et cetera}.
These Python configuration files are then provided as arguments to the \texttt{cleave} command line provided by the framework.
\ac{CLEAVE} imports and parses them, and then sets up and executes the emulation.

Although the core design of \ac{CLEAVE} follows a single-loop approach, multi-loop scenarios can be easily implemented through the combination of multiple \ac{CLEAVE} instances.
The framework does not make any assumptions about the individual components other than the signatures of methods on the \acp{ABC} that must be extended and implemented.
Combined with the Python emulation configuration scripts, this makes development of more complex scenarios merely a matter of implementing the desired functionality within the framework itself.

\subsection{Current Implementations}

\begin{figure}
    \centering
    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
    \caption{
        The 2D inverted pendulum system.
    }\label{fig:invpend}
\end{figure}

For the purpose of this work, we have used \ac{CLEAVE} to implement an emulated \emph{inverted pendulum} control loop (see \cref{fig:invpend}), the goal of which is to balance the vertically free-swinging pendulum by applying horizontal forces on the cart.
We have chosen this system as an initial benchmark for its relative simplicity as well as prevalence in the field of automatic control as one of the fundamental examples of linear control.

The inverted pendulum plant \emph{State} is implemented as a real-time discrete-time physical emulation using CLEAVE's API and a 2D physics library\footnote{Pymunk: \url{http://www.pymunk.org/en/latest/}}, updated at a constant \SI{120}{\hertz} (this value is configurable; in this case it corresponds to the maximum achievable stable rate on our available hardware).
\emph{Sensors} for the angle, position, angular velocity, and velocity of the \emph{State}, and an \emph{Actuator} for the horizontal force applied to the cart are implemented as ``perfect'', i.e.\ values are returned as-is, without any added noise.
For the controller side, a proportional-differential strategy is implemented using the framework \emph{Controller} API and the \emph{NumPy} numeric computation library\footnote{NumPy: \url{https://numpy.org/}}.
Plant and controller are then packaged into containers, for ease of orchestration and reparametrization, and to mimic real-world deployment.

A previously mentioned, we eventually plan to build an open library of emulated control loops.\footnote{%
    At the moment of writing, we are in the process of implementing \emph{model-predictive} as well as \emph{deep neural network} controllers for the inverted pendula plants.%
}
The inverted pendulum plant is merely a proof-of-concept, and \ac{CLEAVE} could be used to implement much more complex physical systems with even more stringent latency requirements.
For instance, the flexibility afforded by the choice of programming language and the abstractions detailed in \cref{sec:design}, allow for relatively straightforward emulation of autonomous drone dynamics.
Propeller dynamics could be implemented by \emph{Actuator} objects, gyroscopes by \emph{Sensors}, and the physical behavior of the plant could be modeled using a general purpose 3D physics engine (\emph{e.g.} the Panda3D\footnote{\url{https://docs.panda3d.org/1.10/python/programming/physics/builtin/index}{https://docs.panda3d.org/1.10/python/programming/physics/builtin/index}} engine).

### 2022-08-31 publications/2022CLEAVE/experiments.tex
cleave and ainur

\section{Use case validation}\label{sec:experiments}

In this section, we demonstrate the utility of \ac{CLEAVE} by walking readers through an example use case.
With this, we aim to showcase the ability of our framework to provide accurate and repeatable measurements of the performance of \acp{NCS} deployed on edge computing infrastructure.

We present the following use case scenario.
A simple \ac{NCS} consisting of an \emph{inverted pendulum} plant controlled by a \emph{proportional-differential} controller is to be deployed on an Edge server.
The connection between plant and server is over a WiFi (IEEE 802.11n) link.
Moreover, there are a number of video-streaming applications running concurrently on the Edge setup.

We believe this to be an interesting and representative use-case scenario for a number of reasons.
Firstly, the inverted pendulum is broadly used as a benchmark in control-system literature, see for instance\ \cite{Baumann2018LowPower} and\ \cite{Natale2004InvPendEthernet}.
A key reason for this is that such a relatively simplistic system allows for straightforward reparametrization to obtain varied system dynamics, which in turn makes a broader range of experiments possible.
Secondly, similar, albeit non-Edge-enabled, setups have been a reality in industry for almost two decades, as evidenced by\ \cite{Gupta2010NCSOverview}.
Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \acp{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
4G, for instance, has average \acp{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \ac{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \acp{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
Finally, video analytics is one of the main proposed use cases for edge computing~\cite{Ananthanarayanan2017Analytics,Yi2017Analytics,Wang2018Analytics}, and thus we foresee edge \ac{NCS} deployments being deployed in parallel with such applications in the future.
However, before deployment can become a reality, key questions such as: 
``what are the baseline requirements of the \ac{NCS}?''; ``what are the achievable best-case end-to-end latencies in the system?''; and
``how does the concurrent video-streaming traffic affect \ac{NCS} stability?'' need to be answered.

We attempt to study these questions using the experimental setup in \cref{fig:cleave:expsetup}.
\ac{CLEAVE} is deployed on a testbed consisting of \num{10} Raspberry Pi 4B  clients connected wirelessly to an IEEE 802.11n \ac{AP}.
Connected to the Ethernet backbone of this \ac{AP} is a general-purpose \verb|x86_64| desktop computer acting as a Cloudlet/Edge server.

% \input{tables/hardware.tex}

\begin{figure}
    \centering
    \includegraphics[width=.95\columnwidth]{publications/2022CLEAVE/images/CLEAVE_experiment_setup}
    \caption{
        The setup used for our experimentation. 
        % Containerized versions of the core CLEAVE emulation components are deployed inside a Docker Swarm Overlay Network spanning \num{10} Raspberry Pi 4B clients connected to a single Cloudlet over an IEEE 802.11n \ac{AP}.
    }\label{fig:cleave:expsetup}
\end{figure}

The first step in our case study is to evaluate the baseline performance of the control system, both locally and over the network.
We configure \ac{CLEAVE} to run a single loop with both plant and controller co-located on the Edge server, and we also configure a separate setup with an identical loop executed over the wireless link.
For both of these setups, we execute a series of scenarios varying:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item the sampling rate of the Plant state, setting it to \SIlist[list-final-separator={, or }]{5;10;20;40;60}{\hertz}
    \item the responsiveness of the Controller, adding fixed delays of \SIlist[list-final-separator={, or }]{0;25;50}{\milli\second} after the processing of each sample.
\end{enumerate*}

We run repeated executions of each combination of these parameters at least \num{10} times, for both the networked and ``local-only'' setups.
Each execution lasts for \SI{5}{\minute}, during which we collect detailed data on both the state of the controlled system as well as on the data sent over the network.
After the initial \num{10} runs, we identify that setups with sampling rates of \SIlist{5;10}{\hertz} were consistently too unstable to consider, and disregard them in further analyses.
We further note that setups with \SI{0}{\milli\second} additional delay are always stable and thus also disregard them.
Remaining setups, i.e.\ sampling rates \( \geq \) \SI{20}{\hertz} and artificial delays \( \geq \) \SI{25}{\milli\second}, are repeated an additional \num{20} times for better statistical significance.

This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \ac{CLEAVE}'s design.
Scenarios are executed automatically in batches using a simple Python script which interacts with Docker through the widely adopted \emph{docker-py}\footnote{Docker SDK for Python: \url{https://docker-py.readthedocs.io/en/stable/}} library.
This is \ac{CLEAVE}'s first advantage over existing frameworks; it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_toppled}
        \caption{Fraction of toppled plants.}\label{fig:single:topple}
    \end{subfigure}\\
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_rms}
        \caption{Mean angle \acs*{RMS}; red line indicates \( y = 180 \).}\label{fig:single:rms}
    \end{subfigure}\\
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_rtts}
        \caption{
            Mean \acsp*{RTT}.
        }\label{fig:single:rtt}
    \end{subfigure}%
    \caption[caption]{
        Baseline results.
        Error bars indicate \SI{95}{\percent} \acp{CI}.
        }%
    \label{fig:single}
\end{figure}

\cref{fig:single} shows a summary of the results relating to the single-loop scenarios:
\cref{fig:single:topple} shows the fraction of plants that toppled in each scenario, and \cref{fig:single:rms} shows the average \ac{RMS} for the absolute pendula angles.
\cref{fig:single:rtt} shows average latency due to network and processing (excluding synthetic delays) for both single-loop scenarios.
Packet losses were below \SI{0.2}{\percent} for all parametrizations of the single-loop scenario over the network, and \SI{0}{\percent} for all parametrizations of the local-only setup.

As expected, higher sampling rates tend to correlate with better quality of control; at higher sampling rates the system was able to reach stability at higher \acp{RTT}.
These initial results already hint at interesting consequences for such an edge-bound \ac{NCS} deployment.
For instance, it is clear from \cref{fig:single:topple,fig:single:rms} that network delays can, to a certain extent, be compensated for by increasing the sampling rate of the system.
A corollary of this is, conversely, that at lower network latencies \acp{NCS} are able to stabilize at lower sampling rates.
Adaptive sampling might thus be a viable method for optimizing resource utilization.

Once baselines for single loops in the use case scenario have been established, we can proceed to studying the interaction between the \acp{NCS} and the video-streaming applications.
We deploy \num{6} control loops on the experimental setup depicted in \cref{fig:cleave:expsetup}.
On the remaining \num{4} clients we run the \emph{iperf3}\footnote{iperf3: \url{https://iperf.fr/}} traffic load generator, each generating \SI[per-mode=symbol]{6.5}{\mega\bit\per\second} of uplink \ac{UDP} traffic.
This emulates the load generated on the network by \SI{1080}{p} Full-HD video streaming, originating from the clients and terminating in the cloudlet.
Based on the baseline results, we execute setups with \ac{NCS} plant sampling rates of \SIlist{20;40;60}{\hertz}.
Each sampling rate configuration is run for \SI{5}{\minute}, and then repeated \num{30} times to obtain statistical significance.
Once again, repetitions of this scenario are executed automatically in batches using a simple Python script and \emph{docker-py}.

\begin{figure}[t]
    \centering
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
        \caption{Toppled plants.}\label{fig:video:toppled}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
    \end{subfigure}\\
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
        \caption{Packet losses.}\label{fig:video:drop}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
        \caption{\acsp{RTT}.}\label{fig:video:rtt}
    \end{subfigure}%
    \caption{
        Multi-loop, resource-constrained setup results.
        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
    }\label{fig:video:results}
\end{figure}

\cref{fig:video:results} shows a summary of the results obtained.
\cref{fig:video:toppled} shows the fraction of repetitions of each setup in which \emph{at least} one plant failed to maintain stability and toppled.
\cref{fig:video:rms} shows the \ac{RMS} for the pendulum angles for each setup, only considering data from plants that did not topple.
\cref{fig:video:drop} shows the fraction of \ac{UDP} datagrams dropped, averaged over all plants and repetitions per setup.
\cref{fig:video:rtt} shows the measured end-to-end plant-side \ac{RTT}, averaged over all plants and repetitions per setup.

These results are interesting in their counter-intuitiveness compared to the single-loop baseline results.
While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, \cref{fig:video:toppled,fig:video:rms} show this not to be the case for \ac{NCS} in resource-constrained scenarios.
\SI{60}{\hertz} was the least stable configuration, with at least one pendulum toppling in around \SI{16}{\percent} of the repetitions, and mean pendulum angle \ac{RMS} circa \num{3} times that of the \SI{40}{\hertz} scenario.
\SI{40}{\hertz} was in turn the second worst configuration --- although it presented no toppled pendula, average angle \ac{RMS} doubled that of the \SI{20}{\hertz} setup.

\cref{fig:video:drop,fig:video:rtt} explains these behaviors.
Whereas both the \SIlist{20;40}{\hertz} setups show losses well below \SI{5}{\percent}, the \SI{60}{\hertz} scenario shows an average of around \SI{13}{\percent} of datagrams lost.
The differences in \acp{RTT} results are equally telling; \acp{RTT} for the \SI{60}{\hertz} setup were on average approx.\ \num{3} times those for the \SIlist{20;40}{\hertz} setups.
These results stem from the contention for network resources, and hint at important trade-offs system designers will have to take into consideration when designing and developing \acp{NCS} for deployment on the Edge.
The Edge will be \emph{multi-tenant} and \emph{multi-instance}. 
\ac{NCS} deployments will have to be designed with shared resources in mind, and given the complexity of these systems, experimental tools like \ac{CLEAVE} will be key for their succesful adoption and massification.

### 2022-08-31 publications/2022CLEAVE/intro.tex
cleave and ainur

\section{Introduction}\label{sec:intro}

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
    \caption{
        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
    }\label{fig:cleave:ncs:struct}
\end{figure*}

The number and applications of \acp{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
However, this rapid increase in adoption has mostly been limited to industrial contexts.
Although \acp{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \ac{CPS} in the future.

\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \ac{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.

%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \acp{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.

One the one hand, related literature in \acp{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level. 
%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
On the other, there exist several approaches when considering experimental methodologies. 
%A number of works concerning \acp{NCS} deal with experimental studies.
%\acp{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
One such approach uses setups in which the complete system is built on top of real hardware.
This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
Conversely, other studies choose to instead use completely \emph{simulated} \ac{NCS} setups.
The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \acp{NCS}.
Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
\end{enumerate*}

As evidenced above, experimental research in \acp{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \ac{NCS}.
This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.

Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\ 
Their methodology utilizes joint knowledge of control, computation, and communication. 
In their work various architectural elements and the corresponding delays associated with the NCS are modelled. 
Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation. 
This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.

We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.

Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.

\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.

The rest of this paper is structured as follows.
\cref{sec:approach} presents the design principles and implementation of the framework.
In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \ac{CLEAVE}.
Finally, in \cref{sec:conclusion} we conclude this work.

### 2022-08-31 publications/2021ImpactDelayedResponse/body.tex
figs

@@
   \centering
-%   \includegraphics[width=.8\textwidth]{Fig1.eps}
+  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig1.eps}
   \caption{Experimental test-bed.
@@
   \centering
-%   \includegraphics[width=.8\textwidth]{Fig2.eps}
+  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig2.eps}
   \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
@@
   % \input{plots/lego_hierarchy_diagram.tex}
-%   \includegraphics[width=\textwidth]{Fig3.eps}
+  \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig3.eps}
   \caption{Hierarchical cognitive structure of a step in the LEGO task.}
@@
     \centering
-%     \includegraphics[width=\textwidth]{Fig4a.eps}
+    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
     \caption{Structure of a step in a generic cognitive assistance application. 
@@
     \centering
-%     \includegraphics[width=\textwidth]{Fig4b.eps}
+    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
     \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
@@
     \centering
-%     \includegraphics[width=\textwidth]{Fig4c.eps}
+    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
     \caption{Structure of a block in the experimental task.}%
@@
     \centering
-%     \includegraphics[width=\textwidth]{Fig4d.eps}
+    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
     \caption{Visualization of the execution time of a step.}%
@@
     \centering
-%     \includegraphics[height=140pt]{Fig5a.eps}
+    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5a.eps}
     \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
@@
     \centering
-%     \includegraphics[height=140pt]{Fig5b.eps}
+    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5b.eps}
     \caption{Frame from the board recording of the same participant.}
@@
     \centering
-%     \includegraphics[width=.8\textwidth]{Fig6.eps}
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
     \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
@@
     \centering
-%     \includegraphics[width=.8\textwidth]{Fig7.eps}
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
     \caption{Mean per-step execution time vs.\ delay, by step slice.
@@
   \centering
-%   \includegraphics[width=.8\textwidth]{Fig8.eps}
+  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
   \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
@@
     \centering
-%     \includegraphics[width=.8\textwidth]{Fig9.eps}
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig9.eps}
     \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{fig:acc:delaylength}
@@
   \centering
-%   \includegraphics[width=.8\textwidth]{Fig10.eps}
+  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
   \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
@@
     \centering
-%     \includegraphics[width=.8\textwidth]{Fig11.eps}
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig11.eps}
     \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
@@
     \centering
-%     \includegraphics[width=.8\textwidth]{Fig12.eps}
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig12.eps}
     \caption{Correlation between neuroticism score of participants and their execution time in the longest block at the longest delay.
@@
 \begin{acks}
-  % This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
+  This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
   Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or funding sources.

### 2022-09-01 model.tex
values

@@
 
-We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, \( [0.0, 0.5) \) and \( [0.5, 1.0] \) respectively. 
-We also bin \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+\begin{table}[]
+\centering
+\caption{Default bins used for model parameter levels.}
+\label{tab:defaultbins}
+% \resizebox{\columnwidth}{!}{%
+\begin{tabular}{@{}lccc@{}}
+\toprule
+\textbf{Parameter} & \textbf{Low}     & \textbf{Medium}      & \textbf{High}         \\ \midrule
+Neuroticism        & \( [0, 0.5) \)   &                      & \( [0.5, 1.0] \)      \\
+Impairment         & \( [0, 1.482) \) & \( [1.482, 3.198) \) & \( [3.198, \infty) \) \\
+Duration           & \( [0, 5) \)     & \( [5, 10) \)        & \( [10, \infty) \)    \\ \bottomrule
+\end{tabular}%
+% }
+\end{table}
+
+We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+\cref{tab:defaultbins} shows the exact values used for these ranges.
 
@@
 This count resets whenever this combination changes.
-This value is subsequently discretized into bins with edges \( [0, 5) \), \( [5, 10) \), and \( [10, \infty) \); we will refer to these discretized levels as \emph{short}, \emph{medium}, and \emph{long duration}, respectively.
+This value is subsequently discretized into three levels; \emph{short}, \emph{medium}, and \emph{long duration} (again, refer to \cref{tab:defaultbins} for the exact values used).
 

### 2022-09-01 model.tex
figure explanation

@@
 \centering
-\caption{Default bins used for model parameter levels.}
+\caption{%
+    Default bins used for model parameter levels.
+    For impairment, values were rounded to three decimal places.
+}
 \label{tab:defaultbins}
@@
 Three examples of the effect of this discretization and processing on the data are also presented in \cref{fig:timing}.
+Specifically, we showcase how
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and}}]
+    \item higher neuroticism leads to higher mean execution times as impairment goes up (\cref{fig:timing:impneurvsetime})
+    \item duration has both a speed-up and slow-down effect on execution times, depending on the level of impairment (\cref{fig:timing:durvsetime})
+    \item transitions from higher to lower levels of impairment lead to higher mean execution times when compared to steps far away from a transition, or from a transition from lower to higher levels of impairment (\cref{fig:timing:imptransvsetime})
+\end{enumerate*}.
+
 Finally, in \cref{alg:timings} we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
@@
     \centering
-    \begin{subfigure}[t]{\columnwidth}
+    \begin{subfigure}[t]{.28\textwidth}
         \centering
-        \includegraphics[width=.8\columnwidth]{./model_data/imp_neur_vs_exectime.png}
+        \includegraphics[width=\textwidth]{./model_data/imp_neur_vs_exectime.png}
         \caption{}\label{fig:timing:impneurvsetime}
-    \end{subfigure}\\
-    \begin{subfigure}[t]{\columnwidth}
+    \end{subfigure}%
+    \hspace{.05\textwidth}%
+    \begin{subfigure}[t]{.28\textwidth}
         \centering
-        \includegraphics[width=.8\columnwidth]{./model_data/duration_vs_exectime.png}
+        \includegraphics[width=\textwidth]{./model_data/duration_vs_exectime.png}
         \caption{}\label{fig:timing:durvsetime}
     \end{subfigure}%
-    \begin{subfigure}[t]{\columnwidth}
+    \hspace{.05\textwidth}%
+    \begin{subfigure}[t]{.28\textwidth}
         \centering
-        \includegraphics[width=.8\columnwidth]{./model_data/impairment_transition_vs_exectime.png}
+        \includegraphics[width=\textwidth]{./model_data/impairment_transition_vs_exectime.png}
         \caption{}\label{fig:timing:imptransvsetime}

### 2022-09-01 model.tex
frame probs

@@
 One, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
-And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate.
+We detail this in \cref{ssec:model:exectimes}.
+And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}.
 
-\subsection{Generating realistic execution times}
+\subsection{Generating realistic execution times}\label{ssec:model:exectimes}
 
@@
 
-\subsection{Generating realistic samples}
+\subsection{Generating realistic samples}\label{ssec:model:frames}
+
+Apart from the aforementioned timing and performance data, for~\cite{olguinmunoz:impact2021} we also recorded all collected video frame samples together with matching metadata.
+For each video frame submitted to the \ac{WCA} during the tasks, we recorded
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item the raw video frame captured
+    \item sample submission timestamp
+    \item \ac{WCA} processing completed timestamp
+    \item result or acknowledgement returned timestamp
+    \item a tag representing the result of the \ac{WCA} processing
+\end{enumerate*}.
+The tags assigned corresponded to:
+\begin{description}[font={\bfseries\ttfamily}]
+    \item[SUCCESS:] frames which triggered a transition to a new step in the logical task model of the \ac{WCA}, and thus cause the generation of feedback to the user.
+    This tag is also used for frames which corrected a previous mistake.
+    \item[REPEAT:] frames which contained the same board state as the previous successful frame, and thus produced no feedback.
+    \item[LOW\_CONFIDENCE] frames for which the image recognition algorithm in the \ac{WCA} did not reach the necessary confidence threshold to interpret it as a valid board state.
+    These frames also produce no feedback.
+    \item[BLANK] frames in which not enough of the board is visible due to noise, movement, occlusion, etc.
+    These frames produce no feedback either.
+    \item[TASK\_ERROR] frames which contained an incorrect board state and thus triggered a transition to a procedural corrective step and the generation of feedback to the user.
+    However, it must be noted that none of the \num{40} participants made any mistakes during the task, and thus no such frames were encountered in the data.
+\end{description}
+
+We correlate this frame data with the step timing data described in \cref{ssec:model:exectimes} to match frames with their corresponding step execution times.
+We assign to each frame a normalized instant value corresponding to its capture instant (expressed in seconds since the start of the step) divided by the total execution time of the step.
+To exemplify, for a step with execution time \( t_\text{exec} = \SI{10}{\second} \), a frame captured at time \( \tau = \SI{3}{\second} \) from the start of the step, will have a normalized instant value \( t_\text{norm} \):
+\begin{align}
+    t_\text{norm} = \frac{\tau}{t_\text{exec}} = \frac{\SI{3}{\second}}{\SI{10}{\second}} = 0.3
+\end{align}
+
+This allows us to then analyze the distribution of frame tag probabilities as a step progresses, independently of execution times.
+This is illustrated in \cref{fig:frameprobs}. 
+Intuitively, \texttt{REPEAT} frames dominate the early instants after a step transition, as the user has not had time to start performing the new instruction and thus the \ac{WCA} keeps capturing frames representing the previous state of the board.
+As the user starts moving and performing actions, \texttt{BLANK} frames start to dominate, as this activity prevents the \ac{WCA} from capturing ``clean'' frames.
+Finally, it must be noted that \texttt{SUCCESS} frames are not included in this probability density plot, as, by definition (\( t_\text{norm} \) represents the normalized instant value):
+\begin{equation}
+    P(\text{\texttt{SUCCESS}} | t_\text{norm}) =
+    \left\{ \begin{array}{ll}
+        0 & t_\text{norm} < 1.0 \\
+        1 & t_\text{norm} \geq 1.0    
+    \end{array} \right.
+\end{equation}
+That is, any frame captured immediately at or after the execution time has been reached will contain the finished board state and thus correspond to a \texttt{SUCCESS} frame.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{model_data/frame_probabilities.png}
+    \caption{%
+        Probability density of frame result tags as a step progresses.
+        Note that \texttt{SUCCESS} frames are not included as --- by definition --- the probability for success frames is \num{1} for all normalized instant values greater than or equal to \num{1.0}.
+    }\label{fig:frameprobs}
+\end{figure}
 

### 2022-09-05 model.tex
explaoining frames

@@
 
+Using the above insights, together with the corresponding recorded video frames, we devise an scheme for the procedural generation of a synthetic trace for any step in \ac{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
+We first prepare a discretized representation of the probability density map in \cref{fig:frameprobs}.
+We segment the normalized instant value into a number of discrete bins (\num{25} in this work), and calculate the relative fraction of frames for each category in each bin.
+Then, given
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item a step execution time \( t_\text{exec} \)
+    \item a collection of random non-\texttt{SUCCESS}, non-\texttt{REPEAT} frames (at least one frame for each of the \texttt{BLANK} and \texttt{LOW\_CONFIDENCE} categories)
+    \item an appropriate \texttt{SUCCESS} video frame containing the correct state for the step
+    \item an appropriate \texttt{REPEAT} video frame containing the correct state for the \emph{previous} step
+\end{enumerate*},
+we can procedurally generate a trace.
+For each sampling instant in the step:
+\begin{enumerate}
+    \item We calculate the normalized instant value \( t_\text{norm} = \frac{\tau}{t_\text{exec}} \).
+    \item If \( t_\text{norm} \ge 1.0 \), we select the \texttt{SUCCESS} frame and finish the trace.
+    \item If instead \( t_\text{norm} < 1.0 \), we find the appropriate bin for \( t_\text{norm} \) and then select a frame by performing a weighted random sampling of the frame categories in the normalized instant bin using the relative proportions for each category.
+\end{enumerate}
+
 \todo[inline]{Diagram of model?}

### 2022-09-05 model.tex
writing

@@
 We segment the normalized instant value into a number of discrete bins (\num{25} in this work), and calculate the relative fraction of frames for each category in each bin.
-Then, given
+For each step, given
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item a step execution time \( t_\text{exec} \)
     \item a collection of random non-\texttt{SUCCESS}, non-\texttt{REPEAT} frames (at least one frame for each of the \texttt{BLANK} and \texttt{LOW\_CONFIDENCE} categories)
@@
 \end{enumerate*},
-we can procedurally generate a trace.
-For each sampling instant in the step:
+we can then procedurally generate a trace by randomly selecting appropriate frames according to the distributions presented in \cref{fig:frameprobs}.
+In other words, for each sampling instant in a step with a given execution time \( t_\text{exec} \):
 \begin{enumerate}
@@
 
-\todo[inline]{Diagram of model?}
-\todo[inline]{Verification: do we need the model? (Compare to naive)}
-
 \subsection{Model implementation}
 
-The model is provided as a Python~\num{3.10} library which includes \acp{API} for the generation of realistic execution times and synthetic traces of video frames.
+We provide this model as a Python~\num{3.10} library which includes \acp{API} for the generation of realistic execution times and synthetic traces of video frames.
+We also include a reference implementation of a server-client loop for realistic load generation using the model.
+
+We have made this code available online as \ac{FOSS} --- under a permissive Apache version 2 license --- at the \href{https://github.com/KTH-EXPECA/EdgeDroid2}{\texttt{KTH-EXPECA/EdgeDroid2}} repository on GitHub.
+Furthermore, we provide the reference load generator component implementations as pre-packaged Docker containers available on Docker Hub, at the \href{https://hub.docker.com/r/expeca/edgedroid2}{\texttt{expeca/edgedroid2}} repository.
 
 \subsection{Verifying the model}
+
+\todo[inline]{Verification: do we need the model? (Compare to naive)}

### 2022-09-06 model.tex
model
impl

@@
 And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}.
+
+In the following subsections, \cref{ssec:model:exectimes,ssec:model:frames}, we explain how we design a model which fulfills these requirements.
+We briefly discuss where implementations of this model can be obtained in \cref{ssec:model:impl}
 
@@
 
-\subsection{Model implementation}
+\subsection{Model verification}
+\todo[inline]{Needs better name.}
+
+We use Python~\num{3.10} to implement two variants of the above described model.
+The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
+The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
+We also implement a reference model which uses a constant execution time of \SI{5.23}{\second}, which we in turn call the \emph{naive model}.
+This value of \SI{5.23}{\second} corresponds to the mean execution time of steps with the lowest level of impairment and a transition tag of \emph{no transition}.
+\todo[inline]{Transition fade distance was 4 instead of 8, need to mention?}
+
+\subsection{Model implementation}\label{ssec:model:impl}
 
@@
 
-\subsection{Verifying the model}
-
-\todo[inline]{Verification: do we need the model? (Compare to naive)}
+\todo[inline]{Variants?}

### 2022-09-06 model.tex
started verification

@@
 We also implement a reference model which uses a constant execution time of \SI{5.23}{\second}, which we in turn call the \emph{naive model}.
-This value of \SI{5.23}{\second} corresponds to the mean execution time of steps with the lowest level of impairment and a transition tag of \emph{no transition}.
+This value corresponds to the mean execution time of steps with the lowest level of impairment and a transition tag of \emph{no transition}.
 \todo[inline]{Transition fade distance was 4 instead of 8, need to mention?}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
+    \caption{%
+        Experimental setup we used to study the implications of the realistic models of human behavior for \ac{WCA}.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
+    }\label{fig:expsetup}
+\end{figure}
+
+These models are integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
+We use these client-server loops to illustrate the implications of the realistic timing models, which react to changes in system responsiveness in real-time, when compared to the static, naive model.
+Up to \num{10} of these loops are deployed on the testbed shown in \cref{fig:expsetup}, consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n link.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.8\columnwidth]{figs/numclients_vs_duration.png}
+    \caption{}\label{fig:scaling_duration}
+\end{figure}
 

### 2022-09-07 model.tex
scenarios

@@
     \caption{%
-        Experimental setup we used to study the implications of the realistic models of human behavior for \ac{WCA}.
+        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
         We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
@@
 These models are integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
-We use these client-server loops to illustrate the implications of the realistic timing models, which react to changes in system responsiveness in real-time, when compared to the static, naive model.
-Up to \num{10} of these loops are deployed on the testbed shown in \cref{fig:expsetup}, consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n link.
+In the following, we use these client-server loops to illustrate the implications of the realistic timing models, which react to changes in system responsiveness in real-time, when compared to the static, naive model.
+Two scenarios are studied, using the testbed shown in \cref{fig:expsetup}:
+\begin{enumerate}
+    \item\label{verification:scaling} A scenario to study the effects of decreasing system responsiveness due to resource and network congestion.
+    An increasing number of client-server loops (\numlist{1;4;7;10} loops) are deployed in parallel for each model.
+    Neuroticism is set to a constant value of \num{0.5}, and each combination of model and number of clients is repeated \num{5} times for better statistical representativeness.
+    \item\label{verification:neuro} A scenario in which the effects of neuroticism are studied.
+    Two different configurations of the realistic models, with neuroticism values of \num{0.0} (\emph{low}) and \num{1.0} (\emph{high}), are compared to a reference setup using the naive model.
+    In all these setups, \num{9} loops are deployed in parallel, and an \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} client-server pair is deployed between the tenth Raspberry Pi client and the cloudlet.
+    The \emph{iperf3} deployment is configured to generate additional network load by constantly sending \SI{1460}{\byte} \ac{UDP} datagrams on the uplink; the rate at which these datagrams are sent is being limited by the \acs{CPU} of the Raspberry Pi client.
+
+    We repeat each (model, neuroticism value) configuration --- (naive, \acs{N/A}), (empirical, low), (empirical, high), (theoretical, low), and (theoretical, high) --- \num{5} times for better statistical accuracy.
+\end{enumerate}
+
+Both scenarios described above use a variant of one of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps.
 
@@
     \includegraphics[width=.8\columnwidth]{figs/numclients_vs_duration.png}
-    \caption{}\label{fig:scaling_duration}
+    \caption{
+        Consequences of decreased system responsiveness due to resource and network congestion.
+        We can evidence both how the realistic models speed up at with respect to the naive model when completely unimpaired, and how these same models strongly slow down as system impairment increases.
+    }\label{fig:scaling_duration}
 \end{figure}
+
+\cref{fig:scaling_duration} shows the results for scenario \labelcref{verification:scaling}.
 

### 2022-09-07 background.tex
tweaks, zero-wait to greedy

@@
 
-Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{zero-wait} sampling.
+Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{greedy} sampling.
 In this scheme, a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
@@
 
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals (following a zero-wait scheme) at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals (following a greedy scheme) at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
 \( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.

### 2022-09-07 model.tex
tweaks, zero-wait to greedy

@@
         finally, \subref{fig:timing:imptransvsetime} shows the lingering effects of impairment after a transition.
+        Error bars indicate the \SI{95}{\percent} \ac{CI}.
     }\label{fig:timing}
@@
 
-\subsection{Model verification}
+\subsection{Model verification}\label{ssec:model:verification}
 \todo[inline]{Needs better name.}
@@
     In all these setups, \num{9} loops are deployed in parallel, and an \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} client-server pair is deployed between the tenth Raspberry Pi client and the cloudlet.
-    The \emph{iperf3} deployment is configured to generate additional network load by constantly sending \SI{1460}{\byte} \ac{UDP} datagrams on the uplink; the rate at which these datagrams are sent is being limited by the \acs{CPU} of the Raspberry Pi client.
+    This \emph{iperf3} deployment starts \SI{120}{\second} after the start of the \num{9} model loops, and is configured to generate additional network load by constantly sending a stream of \SI{1460}{\byte} \ac{UDP} datagrams on the uplink; the rate at which these datagrams are sent is only limited by the \acs{CPU} of the Raspberry Pi client.
 
@@
 Both scenarios described above use a variant of one of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps.
+\medskip
 
-\begin{figure}
+\begin{figure*}
     \centering
-    \includegraphics[width=.8\columnwidth]{figs/numclients_vs_duration.png}
+    \begin{subfigure}[t]{\columnwidth}
+        \centering
+        \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
         \caption{
             Consequences of decreased system responsiveness due to resource and network congestion.
-        We can evidence both how the realistic models speed up at with respect to the naive model when completely unimpaired, and how these same models strongly slow down as system impairment increases.
+            We can evidence both how the realistic models speed up with respect to the naive model when completely unimpaired, and how these same models strongly slow down as system impairment increases.
         }\label{fig:scaling_duration}
-\end{figure}
+    \end{subfigure}%
+    \hspace{\fill}%
+    \begin{subfigure}[t]{\columnwidth}
+        \centering
+        \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
+        \caption{}\label{fig:neuro_duration}
+    \end{subfigure}%
+    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \ac{CI}.}
+\end{figure*}
 
-\cref{fig:scaling_duration} shows the results for scenario \labelcref{verification:scaling}.
+\cref{fig:scaling_duration} illustrates the results for scenario \labelcref{verification:scaling}. 
+It showcases the important behaviors of the realistic models which are not captured by naive, non-adaptive model.
+On one hand, the realistic models (and by extension, humans) are actually faster than the naive model by up to \SI{4.37}{\percent} at the lowest level of impairment.
+This is due to the speed-up effect described in \textcite{olguinmunoz:impact2021} --- humans get faster as a task progresses with a high system responsiveness.
+
+On the other, the realistic models experience a slowdown of about \SI{5.13}{\percent} at the highest levels of impairment when compared to the naive model.
+This is explained by the reduced system responsiveness either negating the speed-up effect, or directly leading to the \emph{additional} behavioral slowdown previously described~\cite{olguinmunoz:impact2021}.
+It should be noted that for the \numlist{7;10} client setups only about \SI{6}{\percent} and \SI{22}{\percent} of \acp{TTF} fell into the medium impairment level, and practically no \acp{TTF}
 

### 2022-09-07 model.tex
discussing

@@
 Both scenarios described above use a variant of one of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps.
+
 \medskip
@@
 \cref{fig:scaling_duration} illustrates the results for scenario \labelcref{verification:scaling}. 
-It showcases the important behaviors of the realistic models which are not captured by naive, non-adaptive model.
-On one hand, the realistic models (and by extension, humans) are actually faster than the naive model by up to \SI{4.37}{\percent} at the lowest level of impairment.
+It showcases the important behaviors of the realistic models which are not captured by the naive, non-adaptive model.
+On one hand, the realistic models (and by extension, humans) are actually faster than the naive model by up to \SI{4.37}{\percent} in a deployment with no resource contention.
 This is due to the speed-up effect described in \textcite{olguinmunoz:impact2021} --- humans get faster as a task progresses with a high system responsiveness.
 
-On the other, the realistic models experience a slowdown of about \SI{5.13}{\percent} at the highest levels of impairment when compared to the naive model.
-This is explained by the reduced system responsiveness either negating the speed-up effect, or directly leading to the \emph{additional} behavioral slowdown previously described~\cite{olguinmunoz:impact2021}.
-It should be noted that for the \numlist{7;10} client setups only about \SI{6}{\percent} and \SI{22}{\percent} of \acp{TTF} fell into the medium impairment level, and practically no \acp{TTF}
+On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \acp{TTF} in this scenario reach the medium impairment level and practically no \acp{TTF} pass threshold for the highest level of impairment.
+This showcases that the additional user slow-down due to reduced system responsiveness described in \textcite{olguinmunoz:impact2021} causes a noticeable and significant  even at lower and medium levels of impairment.
+\todo[inline]{Add another setup with additional load and discuss}
+
+\medskip
+
+We illustrate the results of scenario \labelcref{verification:neuro} in \cref{fig:neuro_duration}.
 

### 2022-09-07 model.tex
writing

@@
 In the following subsections, \cref{ssec:model:exectimes,ssec:model:frames}, we explain how we design a model which fulfills these requirements.
-We briefly discuss where implementations of this model can be obtained in \cref{ssec:model:impl}
+We briefly discuss where implementations of this model can be obtained in \cref{ssec:model:obtaining}
+\todo[inline]{Finish}
 
@@
         \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
-        \caption{
+        \caption{%
             Consequences of decreased system responsiveness due to resource and network congestion.
-            We can evidence both how the realistic models speed up with respect to the naive model when completely unimpaired, and how these same models strongly slow down as system impairment increases.
+            At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
+            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
         }\label{fig:scaling_duration}
@@
         \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
-        \caption{}\label{fig:neuro_duration}
+        \caption{%
+            Modulating effects of neuroticism on task duration.
+            Realistic models parameterized with high neuroticism are on average \SI{7.3}{\percent} slower than the reference model, and \SI{3.37}{\percent} slower than the same models instead parameterized with low neuroticism.
+        }\label{fig:neuro_duration}
     \end{subfigure}%
@@
 
-We illustrate the results of scenario \labelcref{verification:neuro} in \cref{fig:neuro_duration}.
+We illustrate the results of scenario \labelcref{verification:neuro} in \cref{fig:neuro_duration}, showcasing the modulating effects of neuroticism in a medium-impairment setup.
+The realistic models were on average \SI{3.8}{\percent} slower than the reference naive model when configured with a low level of neuroticism, and \SI{7.3}{\percent} slower when configured with a high neuroticism.
 
-\subsection{Model implementation}\label{ssec:model:impl}
+\medskip
 
-We provide this model as a Python~\num{3.10} library which includes \acp{API} for the generation of realistic execution times and synthetic traces of video frames.
-We also include a reference implementation of a server-client loop for realistic load generation using the model.
+The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \ac{WCA} applications.
 
-We have made this code available online as \ac{FOSS} --- under a permissive Apache version 2 license --- at the \href{https://github.com/KTH-EXPECA/EdgeDroid2}{\texttt{KTH-EXPECA/EdgeDroid2}} repository on GitHub.
-Furthermore, we provide the reference load generator component implementations as pre-packaged Docker containers available on Docker Hub, at the \href{https://hub.docker.com/r/expeca/edgedroid2}{\texttt{expeca/edgedroid2}} repository.
+\todo[inline]{Finish this subsection.}
 
-\todo[inline]{Variants?}
+\subsection{Obtaining the model}\label{ssec:model:obtaining}
+
+We provide the model implementations in Python~\num{3.10}, the base data, as well as the client and server implementations used for \cref{ssec:model:verification} to the community as \ac{FOSS}.
+All of these are published on the \href{https://github.com/KTH-EXPECA/EdgeDroid2}{\texttt{KTH-EXPECA/EdgeDroid2}} repository on GitHub under a permissive Apache version 2 license.
+Furthermore, the client and server implementations are also made available as pre-packaged Docker containers on Docker Hub, under the \href{https://hub.docker.com/r/expeca/edgedroid2}{\texttt{expeca/edgedroid2}} repository.

### 2022-09-08 model.tex
some formatting

@@
         \toprule
-    {seq} & {run\_id} & {exec\_time} & {ttf} & {neuroticism}\\
+        {seq}        & {run\_id}    & {exec\_time} & {ttf}        & {neuroticism} \\
         \midrule
-    1 & 134146 & 3.655 & 0.597 & 0.375\\
-    2 & 134146 & 4.439 & 0.554 & 0.375\\
-    3 & 134146 & 2.943 & 0.562 & 0.375\\
+        1            & 134146       & 3.655        & 0.597        & 0.375         \\
+        2            & 134146       & 4.439        & 0.554        & 0.375         \\
+        3            & 134146       & 2.943        & 0.562        & 0.375         \\
         \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
-    169 & 134146 & 3.532 & 4.571 & 0.375\\
-    1 & 134470 & 3.461 & 0.561 & 0.594\\
-    2 & 134470 & 6.678 & 0.572 & 0.594\\
+        169          & 134146       & 3.532        & 4.571        & 0.375         \\
+        1            & 134470       & 3.461        & 0.561        & 0.594         \\
+        2            & 134470       & 6.678        & 0.572        & 0.594         \\
         \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
         \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
-    169 & 137353 & 4.615 & 0.537 & 0.625\\
+        169          & 137353       & 4.615        & 0.537        & 0.625         \\
         \bottomrule
@@
         3            & 134146       & 2.943              & 0.554        & low          & 0.375              & low           & NoTransition & 3               & short        \\
-\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
+        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
         169          & 134146       & 3.532              & 3.689        & high         & 0.375              & low           & Lower2Higher & 2               & short        \\
@@
         2            & 134470       & 6.678              & 0.561        & low          & 0.594              & high          & NoTransition & 2               & short        \\
-\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
-\( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)\\
+        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
+        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
         169          & 137353       & 4.615              & 0.579        & low          & 0.625              & high          & NoTransition & 2               & short        \\
@@
 \begin{figure*}
-    \centering
+    \centering%
+    % \hspace{\fill}%
     \begin{subfigure}[t]{.28\textwidth}
@@
     \end{subfigure}%
-    \hspace{.05\textwidth}%
+    \hspace{\fill}%
     \begin{subfigure}[t]{.28\textwidth}
@@
         \caption{}\label{fig:timing:imptransvsetime}
-    \end{subfigure}
+    \end{subfigure}%
+    % \hspace{\fill}%
     \caption{%

### 2022-09-11 model.tex
working

@@
 On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \acp{TTF} in this scenario reach the medium impairment level and practically no \acp{TTF} pass threshold for the highest level of impairment.
-This showcases that the additional user slow-down due to reduced system responsiveness described in \textcite{olguinmunoz:impact2021} causes a noticeable and significant  even at lower and medium levels of impairment.
+This showcases that the additional user slow-down due to reduced system responsiveness, originally described in \textcite{olguinmunoz:impact2021}, causes a noticeable and significant extension of system utilization time, even at lower and medium levels of impairment.
 \todo[inline]{Add another setup with additional load and discuss}
@@
 The realistic models were on average \SI{3.8}{\percent} slower than the reference naive model when configured with a low level of neuroticism, and \SI{7.3}{\percent} slower when configured with a high neuroticism.
+In this scenario, an average of \SI{50}{\percent} of all recorded \acp{TTF} fell within the medium impairment range, and about \SI{26.6}{\percent} within the high impairment range, showing that the effects of behavioral slow-down are considerable even in moderately impaired systems.
 

### 2022-09-24 model.tex
notes
updating a bit

@@
 \subsection{Generating realistic execution times}\label{ssec:model:exectimes}
+\todo[inline]{Rework with new tweaked model.}
 
@@
 In other words, for each recorded step, we look back until we find a step with a different level of impairment and assign a label to the current step according to whether that change was from a higher to a lower level of impairment or vice-versa.
-However, if the current step is too far from the latest transition in impairment (by default, more than \num{8} steps), it is instead marked as \emph{no transition}; this label is also used for steps at the beginning of each run which have no previous transition in impairment.
-This labeling allows us to capture the behavior described in \cref{item:remain} of \cref{ssec:plos}.
+% However, if the current step is too far from the latest transition in impairment (by default, more than \num{8} steps), it is instead marked as \emph{no transition}; this label is also used for steps at the beginning of each run which have no previous transition in impairment.
+% This labeling allows us to capture the behavior described in \cref{item:remain} of \cref{ssec:plos}.
 
@@
           If it has, it resets duration to \num{1} and updates the transition variable accordingly.
-          If not, it increases the duration variable by one unless that increase causes it to exceed the threshold of steps required to forget a transition.
-          If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
+          If not, it increases the duration variable by one.
+        %   unless that increase causes it to exceed the threshold of steps required to forget a transition.
+        %   If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
     \item Discretizes the updated duration value.
-    \item Finds all samples with binned neuroticism, impairment, duration, and transition values matching the current state of the system.
+    \item Finds all samples with binned neuroticism, impairment and duration matching the state of the system.
+    If, and only if, the current duration falls within the \emph{short} duration bin, this data is further bisected according to the current value of the transition variable.
     \item Randomly samples the filtered data to obtain a realistic execution time value.
@@
 \begin{algorithm}
+    \todo[inline]{Update algorithm.}
     \caption{Generating realistic execution times}\label{alg:timings}
@@
 \begin{table*}
+    \todo[inline]{Remove this table, or re-do it.}
     \caption{Sample of the processed execution time data table.}\label{tab:data:exectime:processed}
@@
         \subref{fig:timing:durvsetime} exemplifies how the model captures the speed-up and slow-down behaviors due to impairment and duration described in \textcite{olguinmunoz:impact2021};
-        finally, \subref{fig:timing:imptransvsetime} shows the lingering effects of impairment after a transition.
+        finally, \subref{fig:timing:imptransvsetime} shows the lingering effects of impairment after a transition, by comparing the mean execution times of the steps in the first bin of duration after a transition.
         Error bars indicate the \SI{95}{\percent} \ac{CI}.
@@
 \todo[inline]{Needs better name.}
+\todo[inline]{Rework with new tweaked models.}
 

### 2022-09-24 publications/2018DemoScalingOnTheEdge/paper_body.tex
starting to include papers
fixing references

@@
 
-Previous works on cloudlets~\cite{satya2009case,Ha:ImpactMobile}, one of the earliest incarnation of edge computing, enable small data-centers at the edge of the Internet. 
+Previous works on cloudlets~\cite{Satya2009Case,Ha2013Impact}, one of the earliest incarnation of edge computing, enable small data-centers at the edge of the Internet. 
 Many futuristic applications become viable with these clusters that are only one wireless hop away. 
-One of the most promising genres of these emerging applications is human-in-the-loop applications such as wearable cognitive assistance~\cite{Ha:TowardsWearableCogAssist}.
+One of the most promising genres of these emerging applications is human-in-the-loop applications such as wearable cognitive assistance~\cite{Ha2014Towards}.
 
 In these applications, sensor data, for example video and audio, are continuously streamed to a cloudlet, where they are analyzed in realtime in order to assist users to complete a particular task. 
-Researchers have built prototypes of these applications to help users assemble LEGO models and IKEA furniture, and even learn how to play ping-pong~\cite{satya2009case,Chen:EarlyImplementation}.
+Researchers have built prototypes of these applications to help users assemble LEGO models and IKEA furniture, and even learn how to play ping-pong~\cite{Satya2009Case,Chen2015Early}.
 % In general, all human-in-the-loop applications capture in some form user and environment information and typically leverage compute-intensive algorithms to analyze the data in order to provide feedback to the user.
@@
 	\item[Latency Sensitive:] Given their tight interaction with the physical world, the quality of human-in-the-loop applications is determined by the latencies experienced by users. 
-    These applications are different from conventional mobile applications by the low latency requirements inherent to the applications themselves~\cite{Suzuki,Chen:AnEmpiricalStudyOfLatency}. 
+    These applications are different from conventional mobile applications by the low latency requirements inherent to the applications themselves~\cite{Suzuki2016Vehicle,Chen2017Empirical}. 
     For example, consider a Ping-Pong Assistance application that instructs a user where to hit a ball -- any instruction delivered after the user has made a hit is useless.
@@
     \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly Deep Neural Networks (DNNs).
-    Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{flinn2012cyber}. 
+    Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{Flinn2012Cyber}. 
     While state-of-the-art DNN object detectors can run at more than 20 FPS on a server GPU, their performances are much worse on mobile GPUs - some models cannot even be loaded due to memory constraints.
@@
 \begin{description}[labelindent=\parindent, listparindent=\parindent, style=unboxed, leftmargin=0cm]
-  \item [The application backend] consists of instances of the target application running on \textcite{docker} containers. 
+  \item [The application backend] consists of instances of the target application running on Docker containers. 
   These correspond to real, unaltered instances of said cognitive assistance applications -- we do not model or emulate them in any way -- and they are containerized in order to be able to execute an arbitrary number of them on the same cloudlet.
@@
   \centering
-  \includegraphics[width=.7\columnwidth]{img/TraceReplay_GenArch}
+  \includegraphics[width=.7\columnwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
   \captionof{figure}{General architecture of the benchmarking suite.}
@@
 
-We will demonstrate the benchmarking framework using the \emph{LEGO Task Guidance} application previously developed by \textcite{Chen:EarlyImplementation}.
+We will demonstrate the benchmarking framework using the \emph{LEGO Task Guidance} application previously developed by \textcite{Chen2015Early}.
 This application guides a user step by step in the assembly of a LEGO model; input consists exclusively of video feed of the current state of the assembly, whereas feedback includes visual and auditory components in the form of animations and speech, respectively.
 
-The benchmarking tool will be employed to extract key real-time metrics from this application, \emph{e.g.} average computation and network times, as well as the estimated \emph{user experience} level of the system as a whole (\emph{flawless}, \emph{impaired} or \emph{unusable}, based on the categorization in \cite{Chen:AnEmpiricalStudyOfLatency}).
+The benchmarking tool will be employed to extract key real-time metrics from this application, \emph{e.g.} average computation and network times, as well as the estimated \emph{user experience} level of the system as a whole (\emph{flawless}, \emph{impaired} or \emph{unusable}, based on the categorization in \cite{Chen2017Empirical}).
 

### 2022-09-24 publications/2018DemoScalingOnTheEdge/paper_body.tex
acronyms

@@
 
-    \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly Deep Neural Networks (DNNs).
+    \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly \acp{DNN}.
     Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{Flinn2012Cyber}. 
-    While state-of-the-art DNN object detectors can run at more than 20 FPS on a server GPU, their performances are much worse on mobile GPUs - some models cannot even be loaded due to memory constraints.
+    While state-of-the-art DNN object detectors can run at more than 20 \acs{FPS} on a server \acs{GPU}, their performances are much worse on mobile \acsp{GPU} - some models cannot even be loaded due to memory constraints.
     Cloudlet-based applications overcome these challenges by offloading the computation. 
@@
 The tool needs to be extended to work on a much a broader spectrum of applications, in particular those that do not have a clear task model.
-Furthermore, the current implementation does not support hardware accelerators (e.g. GPUs) that are commonly used for DNN inference.
+Furthermore, the current implementation does not support hardware accelerators (e.g. \acsp*{GPU}) that are commonly used for \ac{DNN} inference.
 

### 2022-09-24 publications/2021ImpactDelayedResponse/body.tex
abstracts

@@
-\section*{Abstract}
-
+\begin{abstract}
   Wearable Cognitive Assistants (WCA) are anticipated to become a widely-used application class, in conjunction with emerging network infrastructures like 5G that incorporate edge computing capabilities.
@@
   The findings have several implications for the design and operation of WCA applications as well as computational and communication infrastructure, and additionally for the development of performance analysis tools for WCA.\@
+\end{abstract}
 

### 2022-09-29 model.tex
updating to new models

@@
 
-\begin{table}
-    \centering
-    \caption[A]{%
-        Sample of the execution time data  used for the elaboration of the timing model.
-        Rows contain the execution times and \acp{TTF} for each of the \num{169} steps performed by the \num{40} participants in \textcite{olguinmunoz:impact2021}, expressed in seconds, together with a unique identifier for the subject, their associated level of normalized neuroticism, and a sequence number identifying the position of the step in the task.
-    }\label{tab:data:exectime}
-    \begin{tabular}{rrrrrr}
-        \toprule
-        {seq}        & {run\_id}    & {exec\_time} & {ttf}        & {neuroticism} \\
-        \midrule
-        1            & 134146       & 3.655        & 0.597        & 0.375         \\
-        2            & 134146       & 4.439        & 0.554        & 0.375         \\
-        3            & 134146       & 2.943        & 0.562        & 0.375         \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
-        169          & 134146       & 3.532        & 4.571        & 0.375         \\
-        1            & 134470       & 3.461        & 0.561        & 0.594         \\
-        2            & 134470       & 6.678        & 0.572        & 0.594         \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \) & \( \cdots \)  \\
-        169          & 137353       & 4.615        & 0.537        & 0.625         \\
-        \bottomrule
-    \end{tabular}
-\end{table}
-
 We employ a cleaned and re-parameterized copy of the aforementioned timing data.
-The \num{6760} data points are arranged in a table, a sample of which can be seen in \cref{tab:data:exectime}, together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
+The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
 
@@
 
-Finally, we assign a number to each sample corresponding to \emph{how long} the current combination of level of impairment and transition has been in effect, in number of steps.
-This count resets whenever this combination changes.
+Finally, we assign a number to each sample corresponding to \emph{how long} the system has been under the current level of impairment, in number of steps.
+% This count resets whenever this combination changes.
 This value is subsequently discretized into three levels; \emph{short}, \emph{medium}, and \emph{long duration} (again, refer to \cref{tab:defaultbins} for the exact values used).
 
-The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}.
-Three examples of the effect of this discretization and processing on the data are also presented in \cref{fig:timing}.
+% The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}.
+Three examples of the effect of this discretization and processing on the data are presented in \cref{fig:timing}.
 Specifically, we showcase how
@@
 \end{algorithm}
-
-\begin{table*}
-    \todo[inline]{Remove this table, or re-do it.}
-    \caption{Sample of the processed execution time data table.}\label{tab:data:exectime:processed}
-    \begin{tabular}{rrrrcrccrc}
-        \toprule
-        {seq}        & {run\_id}    & {next\_exec\_time} & {ttf}        & {impairment} & {neuroticism\_raw} & {neuroticism} & {transition} & {duration\_raw} & {duration}   \\
-        \midrule
-        1            & 134146       & 3.655              & 0.000        & low          & 0.375              & low           & NoTransition & 1               & short        \\
-        2            & 134146       & 4.439              & 0.597        & low          & 0.375              & low           & NoTransition & 2               & short        \\
-        3            & 134146       & 2.943              & 0.554        & low          & 0.375              & low           & NoTransition & 3               & short        \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
-        169          & 134146       & 3.532              & 3.689        & high         & 0.375              & low           & Lower2Higher & 2               & short        \\
-        1            & 134470       & 3.461              & 0.000        & low          & 0.594              & high          & NoTransition & 1               & short        \\
-        2            & 134470       & 6.678              & 0.561        & low          & 0.594              & high          & NoTransition & 2               & short        \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
-        \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \) & \( \cdots \) & \( \cdots \)       & \( \cdots \)  & \( \cdots \) & \( \cdots \)    & \( \cdots \) \\
-        169          & 137353       & 4.615              & 0.579        & low          & 0.625              & high          & NoTransition & 2               & short        \\
-        \bottomrule
-    \end{tabular}
-\end{table*}
 
@@
 \todo[inline]{Rework with new tweaked models.}
+\todo[inline]{Needs to include first the synthetic setup, then "real" setup.}
 

### 2022-10-04 model.tex
nice table

@@
 
-Finally, in \cref{alg:timings} we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
-In plain English, a model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
-At the beginning of each step, the model
+Finally, we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
+A model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
+At the beginning of each step, it:
 \begin{enumerate}
@@
 
-\begin{algorithm}
-    \todo[inline]{Update algorithm.}
-    \caption{Generating realistic execution times}\label{alg:timings}
+% \begin{algorithm}
+%     \todo[inline]{Update algorithm.}
+%     \caption{Generating realistic execution times}\label{alg:timings}
 
-    \SetKwInput{KwState}{Initial State}
+%     \SetKwInput{KwState}{Initial State}
 
-    \KwIn{$\tau$: \ac{TTF} of the previous step.}
-    \KwOut{$t_\text{exec}$: next execution time.}
-    \KwState{%
-        \begin{itemize}
-            \item Binned neuroticism $N$.
-            \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
-            \item Duration counter: $d \leftarrow 0$.
-            \item Previous impairment: $I' \leftarrow \varnothing$.
-            \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
-            \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
-        \end{itemize}
-    }
+%     \KwIn{$\tau$: \ac{TTF} of the previous step.}
+%     \KwOut{$t_\text{exec}$: next execution time.}
+%     \KwState{%
+%         \begin{itemize}
+%             \item Binned neuroticism $N$.
+%             % \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
+%             \item Duration counter: $d \leftarrow 0$.
+%             \item Previous impairment: $I' \leftarrow \varnothing$.
+%             \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
+%             \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
+%         \end{itemize}
+%     }
 
-    $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}
+%     $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}
 
-    \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{``NO TRANSITION''}$\;
-    } \uElseIf{$I' < I$}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{``LOWER TO HIGHER''} $
-    } \uElseIf{$I' > I$}{%
-        $d \leftarrow 1$\;
-        $T \leftarrow \text{``HIGHER TO LOWER''} $
-    } \Else {%
-        $d \leftarrow d + 1$\;
-    }
-    $I' \leftarrow I$ \tcc*{store impairment for later}
-    $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}
+%     \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{``NO TRANSITION''}$\;
+%     } \uElseIf{$I' < I$}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{``LOWER TO HIGHER''} $
+%     } \uElseIf{$I' > I$}{%
+%         $d \leftarrow 1$\;
+%         $T \leftarrow \text{``HIGHER TO LOWER''} $
+%     } \Else {%
+%         $d \leftarrow d + 1$\;
+%     }
+%     $I' \leftarrow I$ \tcc*{store impairment for later}
+%     $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}
 
-    \tcc{next, filter the data to select only the matching samples}
-    $E \leftarrow F_\text{data}(N, I, D, T)$ \;
-    \tcc{finally, sample the data to return a realistic execution time}
-    $t_\text{exec} \leftarrow {SAMPLE}(E)$\;
+%     \tcc{next, filter the data to select only the matching samples}
+%     $E \leftarrow F_\text{data}(N, I, D, T)$ \;
+%     \tcc{finally, sample the data to return a realistic execution time}
+%     $t_\text{exec} \leftarrow {SAMPLE}(E)$\;
 
-\end{algorithm}
+% \end{algorithm}
 
@@
 
-We use Python~\num{3.10} to implement two variants of the above described model.
+We use Python~\num{3.10} to implement two variants of the above described timing model.
 The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
 The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
-We also implement a reference model which uses a constant execution time of \SI{5.23}{\second}, which we in turn call the \emph{naive model}.
-This value corresponds to the mean execution time of steps with the lowest level of impairment and a transition tag of \emph{no transition}.
-\todo[inline]{Transition fade distance was 4 instead of 8, need to mention?}
+We also implement three simpler reference timing models:
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
+    % \item a model which randomly samples from the raw data without any grouping
+    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
+\end{enumerate*}.
+We will refer to these models as the \emph{constant} and \emph{fitted-naive} model, respectively.
+
+\todo[inline]{Insert synthetic, controlled setup here.}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
+    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
+    \toprule
+    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
+    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
+    \midrule
+    fitted-naive & n/a & \SI{576.74}{\second} & & \SI{1084.57}{\second} & \\
+    \multirow[c]{2}{*}{empirical} & low & \SI{511.18}{\second} & \SI{-11.37}{\percent} & \SI{1083.40}{\second} & \SI{-0.11}{\percent} \\
+    & high & \SI{533.29}{\second} & \SI{-7.53}{\percent} & \SI{1245.46}{\second} & +\SI{14.84}{\percent} \\
+    \multirow[c]{2}{*}{theoretical} & low & \SI{504.73}{\second} & \SI{-12.49}{\percent} & \SI{1080.16}{\second} & \SI{-0.41}{\percent} \\
+    & high & \SI{539.00}{\second} & \SI{-6.54}{\percent} & \SI{1225.68}{\second} & +\SI{13.01}{\percent} \\
+    \bottomrule
+    \end{tabular}
+    \caption{
+        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
+        The results are separated for clarity, with the top plot showing the results for the realistic models, and the bottom one for the reference models.
+    }\label{fig:ctrlttf}
+\end{figure*}
+
+We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
+For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.
+
+We show the result of these runs in \cref{fig:ctrlttf}.
+At the highest level of neuroticism, the realistic models result in \textasciitilde{}\SI{14}{\percent} longer total durations when compared to the \emph{fitted-naive} reference model.
 

### 2022-10-04 model.tex
results of synthetic setup

@@
         Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
-        The results are separated for clarity, with the top plot showing the results for the realistic models, and the bottom one for the reference models.
+        The results are separated for clarity, with the left plot showing the results for the realistic models, and the right the reference models.
+        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
     }\label{fig:ctrlttf}

### 2022-10-10 model.tex
badsbdfhbhfbg

@@
     \midrule
-    fitted-naive & n/a & \SI{576.74}{\second} & & \SI{1084.57}{\second} & \\
-    \multirow[c]{2}{*}{empirical} & low & \SI{511.18}{\second} & \SI{-11.37}{\percent} & \SI{1083.40}{\second} & \SI{-0.11}{\percent} \\
-    & high & \SI{533.29}{\second} & \SI{-7.53}{\percent} & \SI{1245.46}{\second} & +\SI{14.84}{\percent} \\
-    \multirow[c]{2}{*}{theoretical} & low & \SI{504.73}{\second} & \SI{-12.49}{\percent} & \SI{1080.16}{\second} & \SI{-0.41}{\percent} \\
-    & high & \SI{539.00}{\second} & \SI{-6.54}{\percent} & \SI{1225.68}{\second} & +\SI{13.01}{\percent} \\
+    fitted-naive & & \SI{575.54}{\second} & & \SI{1082.28}{\second} & \\
+    \multirow[c]{2}{*}{empirical} & low & \SI{509.39}{\second} & \SI{-11.49}{\percent} & \SI{1071.99}{\second} & \SI{-0.95}{\percent} \\
+    & high & \SI{536.48}{\second} & \SI{-6.79}{\percent} & \SI{1235.65}{\second} & +\SI{14.17}{\percent} \\
+    \multirow[c]{2}{*}{theoretical} & low & \SI{505.22}{\second} & \SI{-12.22}{\percent} & \SI{1073.49}{\second} & \SI{-0.81}{\percent} \\
+    & high & \SI{536.31}{\second} & \SI{-6.82}{\percent} & \SI{1235.70}{\second} & +\SI{14.18}{\percent} \\
     \bottomrule
@@
         Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
-        The results are separated for clarity, with the left plot showing the results for the realistic models, and the right the reference models.
         The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
@@
 
-We show the result of these runs in \cref{fig:ctrlttf}.
-At the highest level of neuroticism, the realistic models result in \textasciitilde{}\SI{14}{\percent} longer total durations when compared to the \emph{fitted-naive} reference model.
+We show the results of these runs in \cref{fig:ctrlttf}, and we can notice two interesting behaviors of the realistic models.
+One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
+When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.
+
+Two, at the highest \ac{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
+For \ac{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
+
+\todo[inline]{explain whyyyyy}
 

### 2022-10-21 background.tex
writing

@@
 
-With this in mind, we can reformulate the results this work as follows:
+With this in mind, we can reformulate the conclusions of this work as follows:
 \begin{enumerate}
@@
 
-    \item Humans get faster at performing steps as the task progresses.
+    \item\label{item:speedup} Humans get faster at performing steps as the task progresses.
     For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.

### 2022-10-21 implications_optimization.tex
writing

\section{Implications for the optimization of \acs{WCA}}\label{sec:implications}

In this section we will discuss potential consequences of the

### 2022-10-21 model.tex
writing

@@
 \subsection{Model verification}\label{ssec:model:verification}
-\todo[inline]{Needs better name.}
-\todo[inline]{Rework with new tweaked models.}
-\todo[inline]{Needs to include first the synthetic setup, then "real" setup.}
+\todo[inline]{Needs better name?}
 
@@
 The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
-We also implement three simpler reference timing models:
+The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
+\todo[inline]{Explain why using ExGaussian?}
+
+We also implement two simpler reference timing models:
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
-    % \item a model which randomly samples from the raw data without any grouping
+    % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
+    \item a model which randomly samples from the raw data without any grouping
     \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
 \end{enumerate*}.
-We will refer to these models as the \emph{constant} and \emph{fitted-naive} model, respectively.
-
-\todo[inline]{Insert synthetic, controlled setup here.}
+We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.
 
@@
 
-We show the results of these runs in \cref{fig:ctrlttf}, and we can notice two interesting behaviors of the realistic models.
+We show the results of these runs in \cref{fig:ctrlttf}.
+Two interesting behaviors of the realistic models can be noticed in this plot.
 One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
@@
 
-\todo[inline]{explain whyyyyy}
+These behaviors clearly stem from the effects identified in \textcite{olguinmunoz:impact2021}, described in \cref{ssec:plos}.
+At low levels of impairments, i.e.\ small \acp{TTF}, the speed-up effect dominates, bringing execution times to a minimum and consequently resulting in shorter total durations.
+On the other hand, at high levels of impairment, the slow-down (pacing) effect starts to take over, pushing step execution times and total durations up.
+However, only in the high neuroticism case do execution times eventually surpass the reference model, which is in line with our findings with respect to the modulating effects of neuroticism.
+
+\medskip
 
@@
 
-These models are integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
-In the following, we use these client-server loops to illustrate the implications of the realistic timing models, which react to changes in system responsiveness in real-time, when compared to the static, naive model.
-Two scenarios are studied, using the testbed shown in \cref{fig:expsetup}:
-\begin{enumerate}
-    \item\label{verification:scaling} A scenario to study the effects of decreasing system responsiveness due to resource and network congestion.
-    An increasing number of client-server loops (\numlist{1;4;7;10} loops) are deployed in parallel for each model.
-    Neuroticism is set to a constant value of \num{0.5}, and each combination of model and number of clients is repeated \num{5} times for better statistical representativeness.
-    \item\label{verification:neuro} A scenario in which the effects of neuroticism are studied.
-    Two different configurations of the realistic models, with neuroticism values of \num{0.0} (\emph{low}) and \num{1.0} (\emph{high}), are compared to a reference setup using the naive model.
-    In all these setups, \num{9} loops are deployed in parallel, and an \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} client-server pair is deployed between the tenth Raspberry Pi client and the cloudlet.
-    This \emph{iperf3} deployment starts \SI{120}{\second} after the start of the \num{9} model loops, and is configured to generate additional network load by constantly sending a stream of \SI{1460}{\byte} \ac{UDP} datagrams on the uplink; the rate at which these datagrams are sent is only limited by the \acs{CPU} of the Raspberry Pi client.
+These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
+In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
+Our goal is to understand the consequences of taking second order effects into consideration when modeling execution times, versus simply relying on first-order statistics.
 
-    We repeat each (model, neuroticism value) configuration --- (naive, \acs{N/A}), (empirical, low), (empirical, high), (theoretical, low), and (theoretical, high) --- \num{5} times for better statistical accuracy.
-\end{enumerate}
+\todo[inline]{Update with new deployment, 802.11b/g wifi.}
 
-Both scenarios described above use a variant of one of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps.
+We deploy an increasing number of client-server loops (\numlist{1;4;7;10} loops) in parallel on the testbed and record the resulting timing metrics.
+These setups are repeated \num{5} times per model for better statistical representativeness; realistic models parameterized with \emph{low} (\num{0.0}) and \emph{high} (\num{1.0}) neuroticism are considered separate models for the purpose of this experiment.
+The task used for all repetitions corresponds to a variant of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps plus a special initial trigger step.
 
@@
 This showcases that the additional user slow-down due to reduced system responsiveness, originally described in \textcite{olguinmunoz:impact2021}, causes a noticeable and significant extension of system utilization time, even at lower and medium levels of impairment.
-\todo[inline]{Add another setup with additional load and discuss}
 

### 2022-10-25 model.tex
verification

@@
 \subsection{Generating realistic execution times}\label{ssec:model:exectimes}
-\todo[inline]{Rework with new tweaked model.}
+\todo[inline]{Rewrite! New model!}
 
-As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
-For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
-The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
-These distributions can then be sampled to produce new, realistic execution times.
-In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.
+\medskip{}
+Two versions of the aforementioned model for the generation of execution times are implemented in Python 3.10.
+Apart from a version which follows the above description exactly, we also implement a version which after grouping the data fits an \ac{exGaussian} distribution to each group.
+Execution times are then sampled directly from these distributions instead of from the empirical data.
+The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
 
-We employ a cleaned and re-parameterized copy of the aforementioned timing data.
-The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
+\subsubsection{Verifying the behavior of the timing model}\label{ssec:model:verification}
+% \todo[inline]{Based on the 4 main conclusions of the PLOS paper.}
 
-\begin{table}[]
-    \centering
-    \caption{%
-        Default bins used for model parameter levels.
-        For impairment, values were rounded to three decimal places.
-    }
-    \label{tab:defaultbins}
-    % \resizebox{\columnwidth}{!}{%
-    \begin{tabular}{@{}lccc@{}}
-        \toprule
-        \textbf{Parameter} & \textbf{Low}     & \textbf{Medium}      & \textbf{High}         \\ \midrule
-        Neuroticism        & \( [0, 0.5) \)   &                      & \( [0.5, 1.0] \)      \\
-        Impairment         & \( [0, 1.482) \) & \( [1.482, 3.198) \) & \( [3.198, \infty) \) \\
-        Duration           & \( [0, 5) \)     & \( [5, 10) \)        & \( [10, \infty) \)    \\ \bottomrule
-    \end{tabular}%
-    % }
-\end{table}
+We verify the behavior of the above described model with respect to the four main conclusions of our work in~\cite{olguinmunoz:impact2021}, which were previously discussed in \cref{ssec:plos}.
+These are reformulated as objectives below:
 
-We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
-\cref{tab:defaultbins} shows the exact values used for these ranges.
-
-The next steps of processing are slightly more complex.
-For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \ac{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
-This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
-Users notice the impairment of the current step only when they have finished it, and thus system impairment has no effect on the execution time of the current step, but directly affects the next one.
-
-We then assign a label to each sample indicating the class of the most recent transition in impairment level in the run.
-In other words, for each recorded step, we look back until we find a step with a different level of impairment and assign a label to the current step according to whether that change was from a higher to a lower level of impairment or vice-versa.
-% However, if the current step is too far from the latest transition in impairment (by default, more than \num{8} steps), it is instead marked as \emph{no transition}; this label is also used for steps at the beginning of each run which have no previous transition in impairment.
-% This labeling allows us to capture the behavior described in \cref{item:remain} of \cref{ssec:plos}.
-
-Finally, we assign a number to each sample corresponding to \emph{how long} the system has been under the current level of impairment, in number of steps.
-% This count resets whenever this combination changes.
-This value is subsequently discretized into three levels; \emph{short}, \emph{medium}, and \emph{long duration} (again, refer to \cref{tab:defaultbins} for the exact values used).
-
-% The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}.
-Three examples of the effect of this discretization and processing on the data are presented in \cref{fig:timing}.
-Specifically, we showcase how
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and}}]
-    \item higher neuroticism leads to higher mean execution times as impairment goes up (\cref{fig:timing:impneurvsetime})
-    \item duration has both a speed-up and slow-down effect on execution times, depending on the level of impairment (\cref{fig:timing:durvsetime})
-    \item transitions from higher to lower levels of impairment lead to higher mean execution times when compared to steps far away from a transition, or from a transition from lower to higher levels of impairment (\cref{fig:timing:imptransvsetime})
-\end{enumerate*}.
-
-Finally, we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
-A model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
-At the beginning of each step, it:
 \begin{enumerate}
-    \item Checks whether any change in impairment has occurred.
-          If it has, it resets duration to \num{1} and updates the transition variable accordingly.
-          If not, it increases the duration variable by one.
-        %   unless that increase causes it to exceed the threshold of steps required to forget a transition.
-        %   If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
-    \item Discretizes the updated duration value.
-    \item Finds all samples with binned neuroticism, impairment and duration matching the state of the system.
-    If, and only if, the current duration falls within the \emph{short} duration bin, this data is further bisected according to the current value of the transition variable.
-    \item Randomly samples the filtered data to obtain a realistic execution time value.
-          \begin{itemize}
-              \item Alternatively, a distribution can be first be fitted to the filtered samples.
-                    The model then samples the theoretical distribution instead of the empirical samples.
-          \end{itemize}
-    \item Finally, updates the stored impairment value with the new impairment and proceeds to the next step.
+    \item\label{it:ttftoexectime} Higher \acp{TTF} should result in higher execution times.
+    \item\label{it:duration} When subject to a series of steps at the same level of impairment, desired model behavior depends on the impairment:
+    \begin{enumerate}
+        \item At low \acp{TTF}, the model should speed up; i.e.\ execution times should decrease.
+        \item At medium \acp{TTF}, execution times should remain more or less the same.
+        \item At high \acp{TTF}, execution times should progressively increase.
+    \end{enumerate}
+    \item The effects on execution times due to past reductions or improvements in system responsiveness should linger for at least a few steps whenever system responsiveness changes/
+    \item\label{it:neuro} Finally, neuroticism should act as a modulating factor for the above effects.
 \end{enumerate}
 
-% \begin{algorithm}
-%     \todo[inline]{Update algorithm.}
-%     \caption{Generating realistic execution times}\label{alg:timings}
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\columnwidth]{figs/new_model/ttf_to_exectime.png}
+    \caption{%
+        Effects of feeding three different \acp{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}) into the model on the generated execution times.
+        Higher \acp{TTF} directly lead to higher execution times.
+        Error bars indicate the \SI{95}{\percent} \ac{CI}.
+    }\label{fig:ttf_to_exectime}
+\end{figure}
 
-%     \SetKwInput{KwState}{Initial State}
+\cref{fig:ttf_to_exectime} shows the mean execution time outputted by the model when fed three different levels of \ac{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}).
+These results were generated by first warming up the model by feeding it \num{25} \acp{TTF} selected at random from the data before feeding it the desired input \ac{TTF}, and recording the generated execution time.
+This procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
+The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \acp{TTF} reaching \SI{14}{\percent} (\textasciitilde\SI{5.6}{\second} to \textasciitilde\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
+Additionally, we can observe the effects of neuroticism on generated execution times, as specified in \cref{it:neuro}.
+At low neuroticism, the average difference between execution times at low versus high \acp{TTF} was of roughly \SI{6.2}{\percent}, compared to \SI{12.5}{\percent} at high neuroticism, clearly putting into evidence the modulating effect of this variable.
 
-%     \KwIn{$\tau$: \ac{TTF} of the previous step.}
-%     \KwOut{$t_\text{exec}$: next execution time.}
-%     \KwState{%
-%         \begin{itemize}
-%             \item Binned neuroticism $N$.
-%             % \item Number of steps to remember transitions for: $\delta \leftarrow 8$.
-%             \item Duration counter: $d \leftarrow 0$.
-%             \item Previous impairment: $I' \leftarrow \varnothing$.
-%             \item $B_\text{TTF}$ and $B_\text{duration}$ are functions for binning \ac{TTF} and duration into discrete levels, respectively.
-%             \item $F_\text{data}$ is a function which filters the data and returns only execution time samples which match the given levels of binned neuroticism, binned impairment, binned duration, and transition label.
-%         \end{itemize}
-%     }
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\columnwidth]{figs/new_model/exectime_over_steps.png}
+    \caption{%
+    Effects of prolonged exposure to constant levels of system impairment on the model.
+    At low (\SI{0}{\second}) \ac{TTF}, the models speed-up over time; conversely at high (\SI{5.0}{\second}) \ac{TTF}, the models either present no change or drastically increase their generated execution times, depending on the level of neuroticism.
+    Error bars indicate \SI{95}{\percent} \ac{CI}.
+    }\label{fig:exectimeduration}
+\end{figure}
 
-%     $I \leftarrow B_\text{TTF}(\tau)$ \tcc*{discretize \ac{TTF}}
-
-%     \uIf{($I' == \varnothing$) {\normalfont\textbf{or}}\\\hspace{1em}($T \neq \text{``NO TRANSITION''}$ {\normalfont\textbf{and}} $d + 1 > \delta$)}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{``NO TRANSITION''}$\;
-%     } \uElseIf{$I' < I$}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{``LOWER TO HIGHER''} $
-%     } \uElseIf{$I' > I$}{%
-%         $d \leftarrow 1$\;
-%         $T \leftarrow \text{``HIGHER TO LOWER''} $
-%     } \Else {%
-%         $d \leftarrow d + 1$\;
-%     }
-%     $I' \leftarrow I$ \tcc*{store impairment for later}
-%     $D \leftarrow B_\text{duration}(d)$ \tcc*{discretize duration}
-
-%     \tcc{next, filter the data to select only the matching samples}
-%     $E \leftarrow F_\text{data}(N, I, D, T)$ \;
-%     \tcc{finally, sample the data to return a realistic execution time}
-%     $t_\text{exec} \leftarrow {SAMPLE}(E)$\;
-
-% \end{algorithm}
+Next, \cref{fig:exectimeduration} shows the evolution of generated execution times while the model is subject to a fixed \ac{TTF}, either \emph{low} (\SI{0}{\second}) or \emph{high} (\SI{5}{\second}).
+These results were generated by first warming up the model with \num{25} random \acp{TTF}, and recording the generated execution times over a sequence of \num{12} steps at a fixed \ac{TTF}; this procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
+Once again, we see here behavior matching what is expected of the model, in particular with respect to \cref{it:duration}.
+At low \acp{TTF}, the model is on average, across all configurations, \SI{8.2}{\percent} faster at step \num{12} when compared to step \num{1}.
+At high \acp{TTF}, the behavior changes depending on the level of neuroticism of the model.
+Low neuroticism models basically do not change their execution times, whereas high neuroticism configurations are on average \SI{10}{\percent} slower after \num{12} steps.
+This is once again in line with our previous findings, as we had previously concluded that humans tend to speed up during a task, but that this speed-up is hindered and eventually reversed as system responsiveness decreases, and that the strength of this effect is correlated with neuroticism.
 
 \begin{figure*}
-    \centering%
-    % \hspace{\fill}%
-    \begin{subfigure}[t]{.28\textwidth}
     \centering
-        \includegraphics[width=\textwidth]{./model_data/imp_neur_vs_exectime.png}
-        \caption{}\label{fig:timing:impneurvsetime}
-    \end{subfigure}%
-    \hspace{\fill}%
-    \begin{subfigure}[t]{.28\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{./model_data/duration_vs_exectime.png}
-        \caption{}\label{fig:timing:durvsetime}
-    \end{subfigure}%
-    \hspace{\fill}%
-    \begin{subfigure}[t]{.28\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{./model_data/impairment_transition_vs_exectime.png}
-        \caption{}\label{fig:timing:imptransvsetime}
-    \end{subfigure}%
-    % \hspace{\fill}%
+    \includegraphics[width=\textwidth]{figs/new_model/transitions.png}
     \caption{%
-        Visualizations of the model data after discretization and post-processing.
-        \subref{fig:timing:impneurvsetime} illustrates the effect of impairment on execution times, and how neuroticism modulates this behavior;
-        \subref{fig:timing:durvsetime} exemplifies how the model captures the speed-up and slow-down behaviors due to impairment and duration described in \textcite{olguinmunoz:impact2021};
-        finally, \subref{fig:timing:imptransvsetime} shows the lingering effects of impairment after a transition, by comparing the mean execution times of the steps in the first bin of duration after a transition.
-        Error bars indicate the \SI{95}{\percent} \ac{CI}.
-    }\label{fig:timing}
+        Effects of changes in system impairment on subsequently generated execution times.
+        These effects linger on after the change, and thus execution times immediately after a transition are either consistently lower or higher than otherwise at the new \ac{TTF}, depending on the old \ac{TTF}.
+        Error bars indicate \SI{95}{\percent} \ac{CI}.
+    }\label{fig:transitions}
 \end{figure*}
+
+Finally, in \cref{fig:transitions} we showcase the behavior of the model when comparing execution times generated immediately after a change in system responsiveness.
+We generate these results by first warming up the model by feeding it a fixed \ac{TTF} (which we will refer to as the \emph{origin} \ac{TTF}) \num{25} times.
+Next, another \ac{TTF} value (the \emph{destination} \ac{TTF}) is fed to the model, and we record the output execution time.
+Each sample is tagged according to the relation between origin and destination \ac{TTF}, either lower to higher, higher to lower, or equal.
+As before, we run \num{600} repetitions of this procedure for each combination of model configuration, origin \ac{TTF}, and destination \ac{TTF}.
+Execution times generated immediately after a transition from a higher \ac{TTF} into a lower one are consistently higher than execution times generated without an preceding change in \acp{TTF}.
+Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \ac{TTF} into a higher one.
+These results are once again in line with our findings in~\cite{olguinmunoz:impact2021}, in which we found lingering effects of transitions between levels of system impairment on human execution times.
 
@@
 
-\subsection{Model verification}\label{ssec:model:verification}
-\todo[inline]{Needs better name?}
+\subsection{Implementation}
+\todo[inline]{Maybe not needed.}
 
-We use Python~\num{3.10} to implement two variants of the above described timing model.
-The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
-The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
-The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
-Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
-\todo[inline]{Explain why using ExGaussian?}
-
-We also implement two simpler reference timing models:
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
-    \item a model which randomly samples from the raw data without any grouping
-    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
-\end{enumerate*}.
-We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
-    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
-    \toprule
-    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
-    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
-    \midrule
-    fitted-naive & & \SI{575.54}{\second} & & \SI{1082.28}{\second} & \\
-    \multirow[c]{2}{*}{empirical} & low & \SI{509.39}{\second} & \SI{-11.49}{\percent} & \SI{1071.99}{\second} & \SI{-0.95}{\percent} \\
-    & high & \SI{536.48}{\second} & \SI{-6.79}{\percent} & \SI{1235.65}{\second} & +\SI{14.17}{\percent} \\
-    \multirow[c]{2}{*}{theoretical} & low & \SI{505.22}{\second} & \SI{-12.22}{\percent} & \SI{1073.49}{\second} & \SI{-0.81}{\percent} \\
-    & high & \SI{536.31}{\second} & \SI{-6.82}{\percent} & \SI{1235.70}{\second} & +\SI{14.18}{\percent} \\
-    \bottomrule
-    \end{tabular}
-    \caption{
-        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
-        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
-    }\label{fig:ctrlttf}
-\end{figure*}
-
-We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
-For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.
-
-We show the results of these runs in \cref{fig:ctrlttf}.
-Two interesting behaviors of the realistic models can be noticed in this plot.
-One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
-When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.
-
-Two, at the highest \ac{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
-For \ac{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
-
-These behaviors clearly stem from the effects identified in \textcite{olguinmunoz:impact2021}, described in \cref{ssec:plos}.
-At low levels of impairments, i.e.\ small \acp{TTF}, the speed-up effect dominates, bringing execution times to a minimum and consequently resulting in shorter total durations.
-On the other hand, at high levels of impairment, the slow-down (pacing) effect starts to take over, pushing step execution times and total durations up.
-However, only in the high neuroticism case do execution times eventually surpass the reference model, which is in line with our findings with respect to the modulating effects of neuroticism.
-
-\medskip
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
-    \caption{%
-        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
-    }\label{fig:expsetup}
-\end{figure}
-
-These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
-In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
-Our goal is to understand the consequences of taking second order effects into consideration when modeling execution times, versus simply relying on first-order statistics.
-
-\todo[inline]{Update with new deployment, 802.11b/g wifi.}
-
-We deploy an increasing number of client-server loops (\numlist{1;4;7;10} loops) in parallel on the testbed and record the resulting timing metrics.
-These setups are repeated \num{5} times per model for better statistical representativeness; realistic models parameterized with \emph{low} (\num{0.0}) and \emph{high} (\num{1.0}) neuroticism are considered separate models for the purpose of this experiment.
-The task used for all repetitions corresponds to a variant of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps plus a special initial trigger step.
-
-\medskip
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[t]{\columnwidth}
-        \centering
-        \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
-        \caption{%
-            Consequences of decreased system responsiveness due to resource and network congestion.
-            At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
-            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
-        }\label{fig:scaling_duration}
-    \end{subfigure}%
-    \hspace{\fill}%
-    \begin{subfigure}[t]{\columnwidth}
-        \centering
-        \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
-        \caption{%
-            Modulating effects of neuroticism on task duration.
-            Realistic models parameterized with high neuroticism are on average \SI{7.3}{\percent} slower than the reference model, and \SI{3.37}{\percent} slower than the same models instead parameterized with low neuroticism.
-        }\label{fig:neuro_duration}
-    \end{subfigure}%
-    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \ac{CI}.}
-\end{figure*}
-
-\cref{fig:scaling_duration} illustrates the results for scenario \labelcref{verification:scaling}.
-It showcases the important behaviors of the realistic models which are not captured by the naive, non-adaptive model.
-On one hand, the realistic models (and by extension, humans) are actually faster than the naive model by up to \SI{4.37}{\percent} in a deployment with no resource contention.
-This is due to the speed-up effect described in \textcite{olguinmunoz:impact2021} --- humans get faster as a task progresses with a high system responsiveness.
-
-On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \acp{TTF} in this scenario reach the medium impairment level and practically no \acp{TTF} pass threshold for the highest level of impairment.
-This showcases that the additional user slow-down due to reduced system responsiveness, originally described in \textcite{olguinmunoz:impact2021}, causes a noticeable and significant extension of system utilization time, even at lower and medium levels of impairment.
-
-\medskip
-
-We illustrate the results of scenario \labelcref{verification:neuro} in \cref{fig:neuro_duration}, showcasing the modulating effects of neuroticism in a medium-impairment setup.
-The realistic models were on average \SI{3.8}{\percent} slower than the reference naive model when configured with a low level of neuroticism, and \SI{7.3}{\percent} slower when configured with a high neuroticism.
-In this scenario, an average of \SI{50}{\percent} of all recorded \acp{TTF} fell within the medium impairment range, and about \SI{26.6}{\percent} within the high impairment range, showing that the effects of behavioral slow-down are considerable even in moderately impaired systems.
-
-\medskip
-
-The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \ac{WCA} applications.
-
-\todo[inline]{Finish this subsection.}
+\subsection{Model validation}
+\todo[inline]{Implications for task durations, wrt to naive scheme.}
 

### 2022-10-25 old_model.tex
verification

\todo[inline]{Rewrite! New model!}

As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
These distributions can then be sampled to produce new, realistic execution times.
In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.

We employ a cleaned and re-parameterized copy of the aforementioned timing data.
The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.

\begin{table}[]
    \centering
    \caption{%
        Default bins used for model parameter levels.
        For impairment, values were rounded to three decimal places.
    }
    \label{tab:defaultbins}
    % \resizebox{\columnwidth}{!}{%
    \begin{tabular}{@{}lccc@{}}
        \toprule
        \textbf{Parameter} & \textbf{Low}     & \textbf{Medium}      & \textbf{High}         \\ \midrule
        Neuroticism        & \( [0, 0.5) \)   &                      & \( [0.5, 1.0] \)      \\
        Impairment         & \( [0, 1.482) \) & \( [1.482, 3.198) \) & \( [3.198, \infty) \) \\
        Duration           & \( [0, 5) \)     & \( [5, 10) \)        & \( [10, \infty) \)    \\ \bottomrule
    \end{tabular}%
    % }
\end{table}

We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
\cref{tab:defaultbins} shows the exact values used for these ranges.

The next steps of processing are slightly more complex.
For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \ac{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
Users notice the impairment of the current step only when they have finished it, and thus system impairment has no effect on the execution time of the current step, but directly affects the next one.

We then assign a label to each sample indicating the class of the most recent transition in impairment level in the run.
In other words, for each recorded step, we look back until we find a step with a different level of impairment and assign a label to the current step according to whether that change was from a higher to a lower level of impairment or vice-versa.
% However, if the current step is too far from the latest transition in impairment (by default, more than \num{8} steps), it is instead marked as \emph{no transition}; this label is also used for steps at the beginning of each run which have no previous transition in impairment.
% This labeling allows us to capture the behavior described in \cref{item:remain} of \cref{ssec:plos}.

Finally, we assign a number to each sample corresponding to \emph{how long} the system has been under the current level of impairment, in number of steps.
% This count resets whenever this combination changes.
This value is subsequently discretized into three levels; \emph{short}, \emph{medium}, and \emph{long duration} (again, refer to \cref{tab:defaultbins} for the exact values used).

% The results of this processing on the samples from \cref{tab:data:exectime} can be seen in \cref{tab:data:exectime:processed}.
Three examples of the effect of this discretization and processing on the data are presented in \cref{fig:timing}.
Specifically, we showcase how
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and}}]
    \item higher neuroticism leads to higher mean execution times as impairment goes up (\cref{fig:timing:impneurvsetime})
    \item duration has both a speed-up and slow-down effect on execution times, depending on the level of impairment (\cref{fig:timing:durvsetime})
    \item transitions from higher to lower levels of impairment lead to higher mean execution times when compared to steps far away from a transition, or from a transition from lower to higher levels of impairment (\cref{fig:timing:imptransvsetime})
\end{enumerate*}.

Finally, we present a step-by-step breakdown of how this data is subsequently used to generate realistic execution times.
A model which uses this processed data to generate execution times needs to maintain variables for duration, previous impairment, and most recent transition.
At the beginning of each step, it:
\begin{enumerate}
    \item Checks whether any change in impairment has occurred.
          If it has, it resets duration to \num{1} and updates the transition variable accordingly.
          If not, it increases the duration variable by one.
        %   unless that increase causes it to exceed the threshold of steps required to forget a transition.
        %   If that's the case, instead the model resets duration to \num{1} and sets the transition variable to ``no transition''.
    \item Discretizes the updated duration value.
    \item Finds all samples with binned neuroticism, impairment and duration matching the state of the system.
    If, and only if, the current duration falls within the \emph{short} duration bin, this data is further bisected according to the current value of the transition variable.
    \item Randomly samples the filtered data to obtain a realistic execution time value.
          \begin{itemize}
              \item Alternatively, a distribution can be first be fitted to the filtered samples.
                    The model then samples the theoretical distribution instead of the empirical samples.
          \end{itemize}
    \item Finally, updates the stored impairment value with the new impairment and proceeds to the next step.
\end{enumerate}

We use Python~\num{3.10} to implement two variants of the above described timing model.
The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
\todo[inline]{Explain why using ExGaussian?}

We also implement two simpler reference timing models:
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
    \item a model which randomly samples from the raw data without any grouping
    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
\end{enumerate*}.
We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.

\begin{figure*}
    \centering
    \includegraphics[width=.9\textwidth]{figs/model_comp_controlled_ttf.png}
    \begin{tabular}{ll >{\columncolor[HTML]{EAEAEA}}r >{\columncolor[HTML]{EAEAEA}}r rr}
    \toprule
    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
    {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
    \midrule
    fitted-naive & & \SI{575.54}{\second} & & \SI{1082.28}{\second} & \\
    \multirow[c]{2}{*}{empirical} & low & \SI{509.39}{\second} & \SI{-11.49}{\percent} & \SI{1071.99}{\second} & \SI{-0.95}{\percent} \\
    & high & \SI{536.48}{\second} & \SI{-6.79}{\percent} & \SI{1235.65}{\second} & +\SI{14.17}{\percent} \\
    \multirow[c]{2}{*}{theoretical} & low & \SI{505.22}{\second} & \SI{-12.22}{\percent} & \SI{1073.49}{\second} & \SI{-0.81}{\percent} \\
    & high & \SI{536.31}{\second} & \SI{-6.82}{\percent} & \SI{1235.70}{\second} & +\SI{14.18}{\percent} \\
    \bottomrule
    \end{tabular}
    \caption{
        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
    }\label{fig:ctrlttf}
\end{figure*}

We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.

We show the results of these runs in \cref{fig:ctrlttf}.
Two interesting behaviors of the realistic models can be noticed in this plot.
One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.

Two, at the highest \ac{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
For \ac{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.

These behaviors clearly stem from the effects identified in \textcite{olguinmunoz:impact2021}, described in \cref{ssec:plos}.
At low levels of impairments, i.e.\ small \acp{TTF}, the speed-up effect dominates, bringing execution times to a minimum and consequently resulting in shorter total durations.
On the other hand, at high levels of impairment, the slow-down (pacing) effect starts to take over, pushing step execution times and total durations up.
However, only in the high neuroticism case do execution times eventually surpass the reference model, which is in line with our findings with respect to the modulating effects of neuroticism.

\medskip

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
    \caption{%
        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
    }\label{fig:expsetup}
\end{figure}

These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
Our goal is to understand the consequences of taking second order effects into consideration when modeling execution times, versus simply relying on first-order statistics.

\todo[inline]{Update with new deployment, 802.11b/g wifi.}

We deploy an increasing number of client-server loops (\numlist{1;4;7;10} loops) in parallel on the testbed and record the resulting timing metrics.
These setups are repeated \num{5} times per model for better statistical representativeness; realistic models parameterized with \emph{low} (\num{0.0}) and \emph{high} (\num{1.0}) neuroticism are considered separate models for the purpose of this experiment.
The task used for all repetitions corresponds to a variant of the \num{169}-step tasks used in~\cite{olguinmunoz:impact2021}, truncated to \num{45} steps plus a special initial trigger step.

\medskip

\begin{figure*}
    \centering
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/numclients_vs_duration.png}
        \caption{%
            Consequences of decreased system responsiveness due to resource and network congestion.
            At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
        }\label{fig:scaling_duration}
    \end{subfigure}%
    \hspace{\fill}%
    \begin{subfigure}[t]{\columnwidth}
        \centering
        \includegraphics[width=.8\textwidth]{figs/neuroticism_vs_duration.png}
        \caption{%
            Modulating effects of neuroticism on task duration.
            Realistic models parameterized with high neuroticism are on average \SI{7.3}{\percent} slower than the reference model, and \SI{3.37}{\percent} slower than the same models instead parameterized with low neuroticism.
        }\label{fig:neuro_duration}
    \end{subfigure}%
    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \ac{CI}.}
\end{figure*}

\cref{fig:scaling_duration} illustrates the results for scenario \labelcref{verification:scaling}.
It showcases the important behaviors of the realistic models which are not captured by the naive, non-adaptive model.
On one hand, the realistic models (and by extension, humans) are actually faster than the naive model by up to \SI{4.37}{\percent} in a deployment with no resource contention.
This is due to the speed-up effect described in \textcite{olguinmunoz:impact2021} --- humans get faster as a task progresses with a high system responsiveness.

On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \acp{TTF} in this scenario reach the medium impairment level and practically no \acp{TTF} pass threshold for the highest level of impairment.
This showcases that the additional user slow-down due to reduced system responsiveness, originally described in \textcite{olguinmunoz:impact2021}, causes a noticeable and significant extension of system utilization time, even at lower and medium levels of impairment.

\medskip

We illustrate the results of scenario \labelcref{verification:neuro} in \cref{fig:neuro_duration}, showcasing the modulating effects of neuroticism in a medium-impairment setup.
The realistic models were on average \SI{3.8}{\percent} slower than the reference naive model when configured with a low level of neuroticism, and \SI{7.3}{\percent} slower when configured with a high neuroticism.
In this scenario, an average of \SI{50}{\percent} of all recorded \acp{TTF} fell within the medium impairment range, and about \SI{26.6}{\percent} within the high impairment range, showing that the effects of behavioral slow-down are considerable even in moderately impaired systems.

\medskip

The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \ac{WCA} applications.

\todo[inline]{Finish this subsection.}

### 2022-10-26 background.tex
verification
writing

--- 2022-09-07 background.tex
@@
 
-With this in mind, we can reformulate the results this work as follows:
+With this in mind, we can reformulate the conclusions of this previous work as follows:
 \begin{enumerate}
@@
     Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \ac{TTF} of \SI{2.475}{\second}, and a further \num{14} percentage points slower at a mean \ac{TTF} of \SI{4.5}{\second}.
-    Users also become progressively slower the longer they spend in a degraded system state.
+
+    \item\label{item:speedup} At lower \acp{TTF}, humans get faster at performing steps as the task progresses, 
+    For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.
+    However, this effect is dampened by reduced system responsiveness, and actually inverts at the highest levels of system impairment; humans actually become progressively slower the longer they spend in a degraded system state.
 
@@
     These effects are noticeable for at least \num{4} steps after the return to a high-responsiveness state.
-
-    \item Humans get faster at performing steps as the task progresses.
-    For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.
-    However, this effect is dampened by reduced system responsiveness, and disappears at the highest levels of system impairment.
 

### 2022-10-26 model.tex
writing

@@
 \subsection{Generating realistic execution times}\label{ssec:model:exectimes}
-\todo[inline]{Rewrite! New model!}
 
-\medskip{}
-Two versions of the aforementioned model for the generation of execution times are implemented in Python 3.10.
-Apart from a version which follows the above description exactly, we also implement a version which after grouping the data fits an \ac{exGaussian} distribution to each group.
-Execution times are then sampled directly from these distributions instead of from the empirical data.
-The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
+For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
+We group the data previously collected according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
+The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
+These distributions can then be sampled to produce new, realistic execution times.
+In the following, we detail the step-by-step processing of this data and construction of the probabilistic model.
+
+We employ a cleaned and re-parameterized copy of the aforementioned timing data.
+The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
+
+\begin{table*}[]
+    \centering
+    \caption{%
+        Default bins used for model parameter levels.
+        Values have been rounded to two decimal places.
+    }
+    \label{tab:defaultbins}
+    \begin{tabular}{@{}lccccccc@{}}
+        \toprule
+        \textbf{Parameter} & \textbf{Low} & & & \textbf{Medium} & & & \textbf{High}         \\ \midrule
+        Neuroticism        & \( [0, 0.5) \) & & & & & & \( [0.5, 1.0] \)      \\
+        Weighted TTF & \((-\infty, 0.82]\) & \((0.82, 1.53]\) & \((1.53, 2.08]\) & \((2.08, 2.67]\) & \((2.67, 3.45]\) & \((3.45, 4.13]\) & \((4.13, \infty]\)
+    \end{tabular}
+\end{table*}
+
+We begin by calculating rolling weighted averages of the \acp{TTF} of each of the \num{40} individual repetitions of the data.
+We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \ac{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \textasciitilde\SI{25}{\percent}, the third-most \textasciitilde\SI{12}{\percent}, and so on.
+We additionally pad the data for each run with \num{12} copies of the first \ac{TTF} in order to ensure sensible values for the first twelve steps; this padding is removed after the weighted values have been calculated.
+This weighing ensures that drastic changes in \ac{TTF} are significantly remembered by the model for at least \num{4} steps --- in line with our previous findings on human behavior --- and are subsequently quickly forgotten.
+
+\begin{equation}\label{eq:weights}
+    w_{n - i} = 
+    \left\{ \begin{array}{ll}
+        \frac{e^{0.7 i}}{\sum\limits^{12}_{j=1} e^{0.7 j}} & 1 \leq i \leq 12 \\
+        & \\
+        0 & i > 12
+    \end{array} \right.
+\end{equation}
+
+The resulting weighted \acp{TTF} are then binned into continuous ranges by splitting the data on the \num{7}-quantiles.
+The exact resulting bins of this operation on the data can be seen in \cref{tab:defaultbins}.
+
+Next, the data is further tagged according to its associated level of normalized neuroticism.
+For this work, we use two levels, \emph{low} and \emph{high}, the exact values for which can be seen in \cref{tab:defaultbins}.
+
+After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \acp{TTF}.
+We devise two variants of a model which uses the above-described pre-processed data to generate realistic timings for \ac{WCA}.
+These models operate as follows:
+
+\begin{enumerate}
+    \item At the beginning of each step, the model is fed the measured \ac{TTF} for the previous step.
+    \item The model calculates a weighted average over the latest \num{12} steps recorded using the weights defined in \cref{eq:weights}.
+    \item The resulting weighted \ac{TTF} is binned into the corresponding range (\cref{tab:defaultbins}).
+    \item The model then filters the pre-processed data to find execution time samples associated with this same discretized weighted \ac{TTF}.
+    \item The selected samples are further filtered to match the desired level of neuroticism.
+    \item The remaining samples are then used to output a realistic execution time, either by
+    \begin{itemize}
+        \item directly sampling the empirical execution time values;
+        \item or, first fitting a distribution to the execution time samples and then sampling the distribution instead.
+    \end{itemize}
+\end{enumerate}
+
+The distribution chosen for the second variant described above corresponds to the \ac{exGaussian} distribution.
+We've chosen this distribution due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+
+We implement these models in Python 3.10, and verify their correct behavior in the following.
 

### 2022-10-26 implications_optimization.tex
verification
writing

@@
-\section{Implications for the optimization of \acs{WCA}}\label{sec:implications}
+\section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
+\todo[inline]{Implications for task durations, wrt to naive scheme.}
+\todo[inline]{Implications for optimization in terms of sampling, energy.}
+\todo[inline]{Introduction to this section}
 
-In this section we will discuss potential consequences of the
+We begin by studying the implications of such a model on the estimation of application lifetimes.
+In the context of \ac{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
+This is an important metric for \ac{WCA} optimization, as it directly relates to system resource utilization and contention, and to energy consumption.
+
+In order to illustrate the consequences of using a less realistic, less accurate model for this kind of research, we introduce here a reference model to which we will compare our results.
+This reference model represents one of the most crude possible approximations the empirical execution time modeling.
+It consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021} which is randomly sampled at runtime without any sort of adjustment to the current state of the system.
+
+We start by studying application lifetimes in a controlled setup.

### 2022-10-26 model.tex
writing

@@
 
-\subsection{Implementation}
-\todo[inline]{Maybe not needed.}
-
-\subsection{Model validation}
-\todo[inline]{Implications for task durations, wrt to naive scheme.}
-
 \subsection{Obtaining the model}\label{ssec:model:obtaining}
 
-We provide the model implementations in Python~\num{3.10}, the base data, as well as the client and server implementations used for \cref{ssec:model:verification} to the community as \ac{FOSS}.
+We provide the model implementations in Python~\num{3.10} as well as the base data to the community as \ac{FOSS}.
 All of these are published on the \href{https://github.com/KTH-EXPECA/EdgeDroid2}{\texttt{KTH-EXPECA/EdgeDroid2}} repository on GitHub under a permissive Apache version 2 license.
-Furthermore, the client and server implementations are also made available as pre-packaged Docker containers on Docker Hub, under the \href{https://hub.docker.com/r/expeca/edgedroid2}{\texttt{expeca/edgedroid2}} repository.

### 2022-10-27 implications_optimization.tex
writing

@@
 
+\subsection{Application lifetimes}
+
 We begin by studying the implications of such a model on the estimation of application lifetimes.
@@
 
-In order to illustrate the consequences of using a less realistic, less accurate model for this kind of research, we introduce here a reference model to which we will compare our results.
-This reference model represents one of the most crude possible approximations the empirical execution time modeling.
-It consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021} which is randomly sampled at runtime without any sort of adjustment to the current state of the system.
+In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
+This new model represents a first-order approximations the empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
+This distribution is then randomly sampled at runtime to obtain execution times for each step, without any sort of adjustment to the current state of the system.
 
-We start by studying application lifetimes in a controlled setup.
+We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
+These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \acp{TTF} and record the generated execution time.
+We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
+This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \ac{TTF}.
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_all_ttfs.png}
+        \caption{%
+            Evolution of application lifetimes as \acp{TTF} increase.
+            Error bars indicate \SI{95}{\percent} \acp{CI}.
+        }
+    \end{subfigure}
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_minmax.png}
+        \caption{Detailed view at application lifetimes at the lowest and highest \acp{TTF} studied.}
+    \end{subfigure}
+    \caption{\acl{TTF} versus application lifetime.}\label{fig:lifetimes}
+\end{figure}
+
+The results of this investigation are presented in \cref{fig:lifetimes}.
+Compared to the reference model, our realistic models are, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
+Conversely, at higher \acp{TTF} the relationship inverts, with the realistic models resulting in durations about \SI{2}{\minute} longer than the reference model, an increase of roughly \SI{11}{\percent}.
+
+\todo[inline]{Discuss}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
+    \caption{%
+        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
+    }\label{fig:expsetup}
+\end{figure}
+
+Next we study the effects of first- versus second-order models in a realistic setting.
+
+\subsection{Energy consumption}
+
+\todo[inline]{Add description of aperiodic sampling scheme}

### 2022-10-28 implications_optimization.tex
clean plot

@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_minmax.png}
-        \caption{Detailed view at application lifetimes at the lowest and highest \acp{TTF} studied.}
+        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_diff.png}
+        \caption{%
+            Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+        }
     \end{subfigure}

### 2022-10-28 implications_optimization.tex
junjues

@@
         Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}b/g access point.
     }\label{fig:expsetup}
@@
 
-\subsection{Energy consumption}
+\subsection{Optimizing for mean number of samples per step}
+
+In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \ac{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
+At every sampling instant \( t \), the scheme adapts the sampling rate \( R(t) \) of the system according to the estimated likelihood of the user having finished the step,
+following the formula
+
+\begin{equation}
+    R(t) = R_\text{min} + \alpha\left( R_\text{max} - R_\text{min} \right) * CDF(t)
+\end{equation}
+
+\( R_\text{max} \) and \( R_\text{min} \) correspond to the maximum and minimum sampling rates of the system, respectively.
+\( R_\text{max} \) can directly be assumed to correspond to one over the mean \ac{RTT} of the system, whereas \( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually.
+\( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
+Finally, \( CDF \) corresponds to the \acl*{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
+
+In the following, we will show the effects of more accurate execution time estimation using our timing models on the performance of this adaptive scheme.
+For this, we implement two variants of the adaptive scheme.
+The first one corresponds to \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
+The second variant instead uses an embedded internal execution time model to keep track of the \acp{TTF} of the system and adjust the distribution of execution times following the state of the timing model.
+
+We proceed to set up an experiment where these sampling scheme variants are deployed on identical, simulated, \num{100}-step tasks.
+The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
+We set \( R_\text{min} \) to \SI{0.5}{\hertz}, \( \alpha \) to \num{3.0}, and the \ac{RTT} of the system to a constant \SI{150}{\milli\second}, and then run \num{90} repetitions of the task for the original design of the sampling scheme as well as for each possible configuration of it with a different parameterization of the timing model.
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_junjue.png}
+        \caption{%
+            Mean number of samples per step over a \num{100}-step task, \num{90} independent repetitions per configuration.
+            Error bars indicate \SI{95}{\percent} \acp{CI}.
+        }
+    \end{subfigure}
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/diff_sampling_junjue.png}
+        \caption{%
+            Percentage difference in mean number of samples per step.
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+        }
+    \end{subfigure}
+    \caption{Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.}\label{fig:samples}
+\end{figure}
+
+The results of this investigation are presented in \cref{fig:samples}, and they clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
+When using the realistic model for execution time estimation, the sampling scheme achieve between \SI{21}{\percent} and \SI{25}{\percent} better average performance.
+This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
+
+\subsection{Optimizing for energy consumption}
 

### 2022-10-28 model.tex
junjues

@@
 We've chosen this distribution due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+Fitting is done using \ac{MLE}.
 
@@
     \end{enumerate}
-    \item The effects on execution times due to past reductions or improvements in system responsiveness should linger for at least a few steps whenever system responsiveness changes/
+    \item The effects on execution times due to past reductions or improvements in system responsiveness should linger for at least a few steps whenever system responsiveness changes.
     \item\label{it:neuro} Finally, neuroticism should act as a modulating factor for the above effects.

### 2022-10-29 sampling.tex
Update on Overleaf.

\section{Aperiodic Sampling}

For sampling the system, we use the optimal aperiodic sampling policy proposed in \cite{sec2022}. The approach followed by the authors is to first express the energy in terms of the expected number of samples and expected wait time experienced by the users and then find the optimum sampling interval that minimises this energy analytically and algorithmically.
The authors express the energy $\mathrm{E}$ required at the terminal to detect an event as \cite{optSampling_tmc}
\begin{alignat}{2}
    \mathrm{E}&=\mathcal{S}\tau_{\text{c}} (P_{\text{c}} -P_0)+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{c}} +\tau_{\text{s}} ) P_0+\tau_{\text{c}} P_{\text{c}}\,,\nonumber
\end{alignat}
and the relevant part of the energy, or energy penalty $\mathcal{E}$ as 
\begin{alignat}{1}
\Rightarrow\mathcal{E}=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
% \\
% % \intertext{where we use the constants $\alpha$ and $\beta$ for mathematical tractability in the upcoming sections. Here,}
% \text{where}\quad\alpha&=\tau_{\text{c}} (P_{\text{c}} -P_0) \;\text{ and }\;\beta=P_0\,.\label{eq:alphaBeta}
\end{alignat}
Here the parameters XYZ corresponds to XYZ \todo[inline]{Need to read the full paper and fill the parameters that are not explained already.}

The authors first prove the two properties that need to be satisfied by the set of sampling instants, $\{t_n,\, n=1,2,\dots,n\}$, which are 
\begin{subequations}\label{cond}
\begin{alignat*}{3}
\!\!&(a)\text{ positive sampling intervals}\,&:&\, t_{n+1}\!-\!t_n\!>\!0,\text{ and}\\
\!\!&(b)\text{ decreasing sampling intervals}\,&:&\, t_{n+1}\!-\!t_n\!<\!t_{n}\!-\!t_{n-1}.
\end{alignat*}
The authors then derive a recursive solution to find the optimum sampling intervals $t_2,t_3\dots$ for a given $t_1$, given by
\begin{alignat}{1}
t_{n+1}=t_n+\dfrac{\sigma^2}{t_n}\Big(\mathrm{exp}\big({\tfrac{t_n^2-t_{n-1}^2}{2\sigma^2}}\big)-1\Big)-\dfrac{\alpha}{\beta},\quad\forall n\geq1.
\end{alignat}
Finally, to find the optimum $t_1$, an algorithm that resembles a simple bisection algorithm is used. In each iteration of the algorithm, the validity of the sequence is repeatedly checked using the above properties and the $t_1$ is updated based on the property that is being violated. The algorithm converges as the power of two and is omitted due to space constraints.
\end{subequations}

### 2022-10-30 sampling.tex
Update on Overleaf.

@@
 \section{Aperiodic Sampling}
+% !TEX root = 00_main.tex
+In this section we will explain the method that we used to find the optimum set of sampling instants (or equivalently sampling intervals) for the given distribution of TTE, denoted by $\mathcal{T}$.
+\todo[inline]{aka execution times.
+Need to recheck the terminologies used in this section and make it consistent with the paper.} We take the modelling and partial solution approach from the recent works on energy efficient sampling of edge-based feedback systems.
+In \cite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user, where the wait time $\mathcal{W}$ -- in the terminology of the authors -- is the time between the event completion and the immediate next sample taken.
+The authors then find the optimum periodic sampling interval that minimises this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
+In this optimisation, the constant parts are excluded as it does not affect the optimisation.
+With a value much smaller than the execution time of the event or TTE, the RTT of the final sample is included in this constant part, which is the reason why $\mathcal{W}$ is computed without including this RTT.
+Next, in \cite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimises the same energy penalty.
+In this work, we use the modelling from \cite{secAperiodic} to find the optimum aperiodic sampling interval.
+However, instead of using their two-step approach to find the solution which include a recursive solution followed by an algorithm, we develop a novel approximate but easier solution to finding the set of optimum aperiodic sampling interval.
+In what follows, we first recap the system modelling used by the authors \cite{secAperiodic} in and then give our independent approximate solution.
 
-For sampling the system, we use the optimal aperiodic sampling policy proposed in \cite{sec2022}.
-The approach followed by the authors is to first express the energy in terms of the expected number of samples and expected wait time experienced by the users and then find the optimum sampling interval that minimises this energy analytically and algorithmically.
-The authors express the energy $\mathrm{E}$ required at the terminal to detect an event as \cite{optSampling_tmc}
-\begin{alignat}{2}
-    \mathrm{E}&=\mathcal{S}\tau_{\text{c}} (P_{\text{c}} -P_0)+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{c}} +\tau_{\text{s}} ) P_0+\tau_{\text{c}} P_{\text{c}}\,,\nonumber
+The energy penalty can be expressed as \cite{secAperiodic}
+\begin{alignat}{1}
+\Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
 \end{alignat}
-and the relevant part of the energy, or energy penalty $\mathcal{E}$ as 
+where $\alpha\!=\!\tau_{\text{c}} (P_{\text{c}}\!-\!P_0)$ and $\beta\!=\!P_0$ are constants. \todo[inline]{Define and explain the parameters if this is not done earlier in other sections.}
+Here, note that $\alpha\mathbb{E}[\mathcal{S}]$ and $\beta\,\mathbb{E}[\mathcal{W}]$ corresponds to the energy wasted per discarded sample and the additional energy expended for waiting, respectively.
+As noted down by the authors, it is important to note that this optimisation approach can also be applied to optimise metrics other than energy, simply by formulating this new metrics in the same format as follows, that is by expressing the metric as a linear combination of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and expected wait time $\mathbb{E}[\mathcal{W}]$.
+Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimises the energy penalty given in \eqref{eq:epsilon_terminal}.
+\subsection{Optimum sampling instants}\label{sec:aprxSol}
+First, we borrow the idea of checkpointing density from \cite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is
+related to $\{t_n\}$ such that,
 \begin{alignat}{1}
-\Rightarrow\mathcal{E}=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
-% \\
-% % \intertext{where we use the constants $\alpha$ and $\beta$ for mathematical tractability in the upcoming sections. Here,}
-% \text{where}\quad\alpha&=\tau_{\text{c}} (P_{\text{c}} -P_0) \;\text{ and }\;\beta=P_0\,.\label{eq:alphaBeta}
+{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
 \end{alignat}
-Here the parameters XYZ corresponds to XYZ \todo[inline]{Need to read the full paper and fill the parameters that are not explained already.}
-
-The authors first prove the two properties that need to be satisfied by the set of sampling instants, $\{t_n,\, n=1,2,\dots,n\}$, which are 
-\begin{subequations}\label{cond}
-\begin{alignat*}{3}
-\!\!&(a)\text{ positive sampling intervals}\,&:&\, t_{n+1}\!-\!t_n\!>\!0,\text{ and}\\
-\!\!&(b)\text{ decreasing sampling intervals}\,&:&\, t_{n+1}\!-\!t_n\!<\!t_{n}\!-\!t_{n-1}.
+We find $r^*(t)$, the $r(t)$ that minimises $\mathcal{E}$.
+Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
+Thus, we get the expected number of samples $\mathbb{E}[\mathcal{S}]$ as
+\begin{alignat}{1}
+\mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{Es}
+\end{alignat}
+To find 
+% the expected wait 
+$\mathbb{E}[\mathcal{W}]$, we use the TTE conditional CDF. We get,
+\begin{multline*}
+\mathbb{P}(\mathcal{W}=t_n-\mathcal{T}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)\\=\dfrac{\mathbb{P}(\mathcal{T}\geq t_n-t\,,\,t_{n-1}<\mathcal{T}\leq t_n)}{\mathbb{P}(t_{n-1}<\mathcal{T}\leq t_n)}.
+\end{multline*}
+The numerator is degenerate when $t\!<\!0$ or $t\!>\!(t_n\!-\!t_{n-1})$.
+Thus, we are only interested in $0\!\leq\!t\!\leq\!
+(t_n-t_{n-1})$.
+Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ corresponds to the CDF, CCDF and pdf of the TTE distribution.
+\begin{alignat*}{1}
+\!\!\!\Rightarrow\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&=\dfrac{\mathbb{P}(t_n-t\leq\mathcal{T}\leq t_n)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\\
+&\approx\dfrac{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n}-t)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\tag{$A_1$}\label{Apx1}
 \end{alignat*}
-The authors then derive a recursive solution to find the optimum sampling intervals $t_2,t_3\dots$ for a given $t_1$, given by
+Here, \ref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator. Expanding the CDF using Taylor series, 
+\begin{multline*}
+    =\Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)\Big)\\
+    \div \Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)\\+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)\Big).
+\end{multline*}
+Simplifying and using the first order approximation, we get
+\begin{alignat*}{1}
+% &=\dfrac{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)}{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)}\\&
+\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&\approx\dfrac{tf_\mathcal{T}(t_n)}{(t_{n}-t_{n-1})f_\mathcal{T}(t_n)}\tag{$A_2$}\label{Apx2}\\
+\Rightarrow\mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&= 1-\dfrac{t}{(t_{n}-t_{n-1})}.
+\end{alignat*}
+Using this CCDF, we find the conditional expectation of $\mathcal{W}$.
+\begin{alignat*}{1}
+% \mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&\approx 1-\dfrac{t}{(t_{n}-t_{n-1})}\\
+\Rightarrow \mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]&=\smashoperator[r]{\int_{0}^{t_n-t_{n-1}}}\Big(1-\dfrac{t}{(t_{n}-t_{n-1})}\Big)\,\mathrm{d}t\\
+&=\dfrac{(t_n-t_{n-1})}{2}.
+\end{alignat*}
+% However, we can also approximate $(t_n-t_{n-1})$, the sampling interval to the inverse of the instantaneous sampling frequency $r(t)$.
+This approximation (\ref{Apx3}) is valid as long as $r(t)$ is varying slowly between two consecutive sampling instants.
+That is,
+If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $t_n\!-\!t_{n-1}$ as,
 \begin{alignat}{1}
-t_{n+1}=t_n+\dfrac{\sigma^2}{t_n}\Big(\mathrm{exp}\big({\tfrac{t_n^2-t_{n-1}^2}{2\sigma^2}}\big)-1\Big)-\dfrac{\alpha}{\beta},\quad\forall n\geq1.
+ (t_n-t_{n-1})&\approx\dfrac{1}{r(t)},\;\forall t,n:t_{n-1}\!<\!t\!\leq\!t_n.\tag{$A_3$}\label{Apx3}\\
+% \end{alignat}
+% As a result, we can compute the expected wait as,
+% \begin{alignat}{1}
+\Rightarrow \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]f_\mathcal{T}(t)\,\mathrm{d}t\nonumber\\
+&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\label{Ew}
 \end{alignat}
-Finally, to find the optimum $t_1$, an algorithm that resembles a simple bisection algorithm is used.
-In each iteration of the algorithm, the validity of the sequence is repeatedly checked using the above properties and the $t_1$ is updated based on the property that is being violated.
-The algorithm converges as the power of two and is omitted due to space constraints.
-\end{subequations}
+We can thus find the energy penalty using \eqref{Es} and \eqref{Ew} as
+\begin{alignat*}{1}
+\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha\int_{x=0}^{t}r(x)\,\mathrm{d}x+\beta\dfrac{1}{2r(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{epsilon_eulerForm}\\
+\intertext{Let $g(t)=\int_{0}^{t}r(x)\dif x$. Then  $g'(t)=\tfrac{\mathrm{d}}{\mathrm{d}t}g(t)=r(t)$. That is,}
+\vspace{-4cm}\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha g(t)+\dfrac{\beta}{2g'(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.
+\end{alignat*}
+As per the Euler-Lagrange equation from the calculus of variations\cite{calcVariation2,calcVariation1}, the extreme value of $\mathcal{E}$ is obtained at 
+\begin{alignat}{1}
+r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\\
+\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
+r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\\
+\Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
+\Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
+\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.
+\end{alignat}

### 2022-10-31 implications_optimization.tex
durations

@@
 \section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
-\todo[inline]{Implications for task durations, wrt to naive scheme.}
-\todo[inline]{Implications for optimization in terms of sampling, energy.}
-\todo[inline]{Introduction to this section}
+% \todo[inline]{Introduction to this section?}
 
@@
 In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This new model represents a first-order approximations the empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
+This new model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
 This distribution is then randomly sampled at runtime to obtain execution times for each step, without any sort of adjustment to the current state of the system.
@@
 We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
-These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \acp{TTF} and record the generated execution time.
+These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \ac{TTF} and record the generated execution time.
 We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
@@
         \caption{%
-            Evolution of application lifetimes as \acp{TTF} increase.
+            Evolution of mean application lifetimes as \acp{TTF} increase.
             Error bars indicate \SI{95}{\percent} \acp{CI}.
@@
 The results of this investigation are presented in \cref{fig:lifetimes}.
-Compared to the reference model, our realistic models are, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
-Conversely, at higher \acp{TTF} the relationship inverts, with the realistic models resulting in durations about \SI{2}{\minute} longer than the reference model, an increase of roughly \SI{11}{\percent}.
-
-\todo[inline]{Discuss}
+Compared to the reference model, our realistic model is, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
+At higher \acp{TTF}, the behavior of the model depends on its level of normalized neuroticism.
+At low neuroticism, the behavior of the realistic model results in total task durations that are basically indistinguishable from the reference model.
+However, at high neuroticism, the model once again results in a considerable difference in total task duration with respect to the reference --- this time extending durations by \textasciitilde\SI{11}{\percent} on average.
 
@@
         Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}b/g access point.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \acs*{COTS} \acs{IEEE} \num{802.11}b/g access point.
     }\label{fig:expsetup}
@@
 
-Next we study the effects of first- versus second-order models in a realistic setting.
+Next we study the effects of first- versus second-order models in a more realistic setting.
+The models, reference and realistic, are deployed on the Raspberry Pi clients of the testbed depicted in \cref{fig:expsetup}.
+For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \ac{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
+Clients and cloudlet communicate over an \acs{IEEE} \num{802.11}b/g wireless network.
+Our choice of wireless standard is simply motivated by a desire to amplify the potential effects of network congestion.
+\todo[inline]{More justification?}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_testbed.png}
+        \caption{%
+            Mean application lifetimes per testbed configuration.
+            Note that due to the low number of samples, means have been calculated using the geometric instead of arithmetic average. 
+            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using bootstrapping.
+        }
+    \end{subfigure}
+    \par\bigskip
+    \begin{subtable}{\columnwidth}
+        \centering
+        \begin{tabular}{lrrrr}
+            \toprule
+            \# clients & 1 & 4 & 7 & 10 \\
+            \ac{RTT} & \SI{0.42}{\second} & \SI{1.12}{\second} & \SI{1.92}{\second} & \SI{2.68}{\second} \\
+            \bottomrule
+        \end{tabular}
+        \caption{Mean measured \aclp{RTT} for each testbed configuration.}
+    \end{subtable}
+    \par\bigskip
+    \begin{subfigure}[]{\columnwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_testbed_diff.png}
+        \caption{%
+            Percentage difference in mean application lifetimes with respect to the reference model.
+            Confidence intervals have been omitted due to the low number of samples and the use of the geometric mean.
+        }
+    \end{subfigure}
+    \caption{Application lifetimes in the realistic scenarios.}\label{fig:testbed_lifetimes}
+\end{figure}
+
+We deploy configurations running \num{45}-step versions of the LEGO task described in \textcite{olguinmunoz:impact2021}.
+The testbed configurations include setups with \numlist{1;4;7;10} clients.
+Due to having only limited time, each combination of testbed and timing model configuration is only repeated \num{10} times.
+The results are presented in \cref{fig:testbed_lifetimes}.
+Owing to the low number of samples, we opt here for the geometric rather than arithmetic mean to represent our results.
+The results are nonetheless clear, and follow the same pattern as the previously discussed results under ideal, controlled circumstances.
+With just a single client and mean \ac{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
+At \num{10} clients, the results mimic those at higher \acp{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
+
+These results highlight the importance of accurate execution time modeling when studying \ac{WCA} application lifetimes.
+Not only do we see considerable differences in lifetimes at relatively moderate levels of system congestion, but the sign of these differences depends on the load placed on the system.
+Imagine thus a system designer studying resource consumption optimization in a \ac{WCA}.
+If they were to employ the reference model for their study on an unimpaired system, it could lead them to significantly underestimate the potential for optimization of resource consumption, leaving performance (and, potentially, monetary) gains on the table.
+On the other hand, under heavy load, they would instead underestimate system resource occupation, again leading to performance losses.
+
+% .as well as an additional setup with \num{10} clients during which an instance of the \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} network load generator is also running on each client.
+% The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \ac{TCP} traffic, further congesting the wireless network.
 
@@
     \centering
+    \begin{subtable}[]{\columnwidth}
+        \centering
+        \begin{tabular}{rrr}
+            \toprule
+            \( R_\text{min} \) & \( \alpha \) & \ac{RTT} \\
+            \midrule
+            \SI{0.5}{\hertz} & \num{3.0} & \SI{150}{\milli\second} \\
+            \bottomrule
+        \end{tabular}
+        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
+    \end{subtable}
     \begin{subfigure}[]{\columnwidth}
@@
 The results of this investigation are presented in \cref{fig:samples}, and they clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling scheme achieve between \SI{21}{\percent} and \SI{25}{\percent} better average performance.
+When using the realistic model for execution time estimation, the sampling scheme achieve between \SI{22}{\percent} and \SI{27}{\percent} better average performance.
 This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.

### 2022-10-31 implications_optimization.tex
title
massaging sampling section

@@
 
-\todo[inline]{Add description of aperiodic sampling scheme}
+Energy consumption is another yet unexplored dimension of \ac{WCA} with great opportunities for optimization.
+To explore the potential implications of our realistic model of human timings for energy consumption, we will introduce here a novel scheme for sampling in step-based \ac{WCA} systems.
+
+We begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
+In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
+The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
+In this optimization, the constant parts are excluded as it does not affect the process.
+With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is the reason why $\mathcal{W}$ is computed without including it.
+Next, in \textcite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+
+In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
+However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of optimum aperiodic sampling interval.
+In what follows, we first recap the system modeling used by the authors \textcite{secAperiodic} in and then give our independent approximate solution.
+
+% \begin{table}
+%     \centering
+%     \caption{}
+%     \begin{tabular}{ll}
+%         \toprule
+%         Parameter & Explanation \\
+%         \midrule
+%         \( P_0 \) & Idle power consumption \\
+%         \( P_\text{c} \) & Power consumption during communication with backend \\
+%         \bottomrule 
+%     \end{tabular}
+% \end{table}
+
+The energy penalty can be expressed as
+\begin{alignat}{1}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
+\end{alignat}
+where \( P_0 \) corresponds to the power consumed by the \ac{WCA} client device during idle periods and \( P_\text{c} \) to the power consumption during communication with the backend (i.e.\ transmitting a sample or receiving a reply).
+$\alpha = \tau_{\text{c}} (P_{\text{c}} - P_0)$ and $\beta = P_0$ are constants.
+Here, note that $\alpha\mathbb{E}[\mathcal{S}]$ and $\beta\,\mathbb{E}[\mathcal{W}]$ correspond to the energy wasted per discarded sample and the additional energy expended while waiting, respectively.
+Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimizes the energy penalty given in \cref{eq:epsilon_terminal}.
+
+\subsubsection{Optimum sampling instants}\label{sec:aprxSol}
+
+First, we borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is
+related to $\{t_n\}$ such that,
+\begin{alignat}{1}
+{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
+\end{alignat}
+We find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
+Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
+Thus, we get the expected number of samples $\mathbb{E}[\mathcal{S}]$ as
+\begin{alignat}{1}
+\mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{Es}
+\end{alignat}
+To find 
+% the expected wait 
+$\mathbb{E}[\mathcal{W}]$, we use the conditional \ac{CDF} of the execution time.
+We get,
+\begin{multline*}
+\mathbb{P}(\mathcal{W}=t_n-\mathcal{T}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)\\=\dfrac{\mathbb{P}(\mathcal{T}\geq t_n-t\,,\,t_{n-1}<\mathcal{T}\leq t_n)}{\mathbb{P}(t_{n-1}<\mathcal{T}\leq t_n)}.
+\end{multline*}
+The numerator is degenerate when $t\!<\!0$ or $t\!>\!(t_n\!-\!t_{n-1})$.
+Thus, we are only interested in $0\!\leq\!t\!\leq\! (t_n-t_{n-1})$.
+Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ correspond to the \ac{CDF}, \ac{CCDF} and \ac{PDF} of the execution time distribution.
+\begin{alignat*}{1}
+\!\!\!\Rightarrow\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&=\dfrac{\mathbb{P}(t_n-t\leq\mathcal{T}\leq t_n)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\\
+&\approx\dfrac{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n}-t)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\tag{$A_1$}\label{Apx1}
+\end{alignat*}
+Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
+Expanding the \ac{CDF} using Taylor series, 
+\begin{multline*}
+    =\Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)\Big)\\
+    \div \Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)\\+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)\Big).
+\end{multline*}
+Simplifying and using the first order approximation, we get
+\begin{alignat*}{1}
+% &=\dfrac{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)}{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)}\\&
+\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&\approx\dfrac{tf_\mathcal{T}(t_n)}{(t_{n}-t_{n-1})f_\mathcal{T}(t_n)}\tag{$A_2$}\label{Apx2}\\
+\Rightarrow\mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&= 1-\dfrac{t}{(t_{n}-t_{n-1})}.
+\end{alignat*}
+Using this \ac{CCDF}, we find the conditional expectation of $\mathcal{W}$.
+\begin{alignat*}{1}
+% \mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&\approx 1-\dfrac{t}{(t_{n}-t_{n-1})}\\
+\Rightarrow \mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]&=\smashoperator[r]{\int_{0}^{t_n-t_{n-1}}}\Big(1-\dfrac{t}{(t_{n}-t_{n-1})}\Big)\,\mathrm{d}t\\
+&=\dfrac{(t_n-t_{n-1})}{2}.
+\end{alignat*}
+% However, we can also approximate $(t_n-t_{n-1})$, the sampling interval to the inverse of the instantaneous sampling frequency $r(t)$.
+This approximation (\ref{Apx3}) is valid as long as $r(t)$ is varying slowly between two consecutive sampling instants.
+That is,
+If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $t_n\!-\!t_{n-1}$ as,
+\begin{alignat}{1}
+ (t_n-t_{n-1})&\approx\dfrac{1}{r(t)},\;\forall t,n:t_{n-1}\!<\!t\!\leq\!t_n.\tag{$A_3$}\label{Apx3}\\
+% \end{alignat}
+% As a result, we can compute the expected wait as,
+% \begin{alignat}{1}
+\Rightarrow \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]f_\mathcal{T}(t)\,\mathrm{d}t\nonumber\\
+&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\label{Ew}
+\end{alignat}
+We can thus find the energy penalty using \cref{Es} and \cref{Ew} as
+\begin{alignat*}{1}
+\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha\int_{x=0}^{t}r(x)\,\mathrm{d}x+\beta\dfrac{1}{2r(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{epsilon_eulerForm}\\
+\intertext{Let $g(t)=\int_{0}^{t}r(x)\dif x$. Then  $g'(t)=\tfrac{\mathrm{d}}{\mathrm{d}t}g(t)=r(t)$. That is,}
+\vspace{-4cm}\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha g(t)+\dfrac{\beta}{2g'(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.
+\end{alignat*}
+As per the Euler-Lagrange equation from the calculus of variations~\cite{calcVariation2,calcVariation1}, the extreme value of $\mathcal{E}$ is obtained at 
+\begin{alignat}{1}
+r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\\
+\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
+r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\\
+\Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
+\Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
+\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.
+\end{alignat}

### 2022-10-31 model.tex
massaging sampling section

@@
 
-In order to accurately emulate the behavior of a human, a model for \ac{WCA} needs to implement two main behaviors.
+In the following subsections, we detail the construction of a model which uses the data collected for \textcite{olguinmunoz:impact2021} to generate, at runtime, realistic execution times. 
+In order to accurately emulate the behavior of a human, such a model needs to implement two main behaviors.
 One, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
@@
 And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}.
-
-In the following subsections, \cref{ssec:model:exectimes,ssec:model:frames}, we explain how we design a model which fulfills these requirements.
-We briefly discuss where implementations of this model can be obtained in \cref{ssec:model:obtaining}
-\todo[inline]{Finish}
+Finally, in \cref{ssec:model:obtaining} we briefly discuss where implementations of this model can be obtained.
 

### 2022-10-31 sampling.tex
massaging sampling section

@@
 \subsection{Optimum sampling instants}\label{sec:aprxSol}
+
 First, we borrow the idea of checkpointing density from \cite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is

### 2022-11-01 intro.tex
reformulating section 3
massaging sampling section
writing intro

@@
-\section{Introduction \& Related Work}\label{sec:intro}
+\section{Introduction}\label{sec:intro}
+
+\acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
+These systems aim to seamlessly integrate into the day-to-day of users to provide real-time, context-aware information and feedback.
+They do so by capturing user and environment information, and leveraging compute-intensive algorithms to analyze the data and produce timely context-dependent feedback.
+A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses --- they are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers by tracking their progress and providing feedback and instructions at appropriate times.
+
+\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{Ha2014towards,Satya2019augmenting}, and have since then expanded to a greater range of use cases.
+In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical}.
+
+As their name suggests, a defining characteristic of these applications is their wearability, which implies the use of lightweight, battery-powered, low-energy consumption devices.
+On the other hand, their context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs such as real-time video feeds which require intensive compute capabilities to process.
+Furthermore, as with any \ac{AR} application, \ac{WCA} systems are sensitive to latency.
+Delays and jitter can be jarring to the user, causing them discomfort, leading them to make mistakes and potentially even give up on the application altogether.
+
+The combination of these characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+
+\todo[inline]{WIP}
+
+This paper builds on previous work on the characterization of human perception of delay in \ac{WCA} systems.
+We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.
+Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}
+In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
+We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
+In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we use a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
+In \textcite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
+We found that humans generally pace themselves according to the perceived system responsiveness.
+When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.
+
+In this paper, we introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
+Furthermore, we introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+We name this model \emph{EdgeDroid \num{2.0}}, as it corresponds to a direct, more realistic evolution of our initial EdgeDroid approach.
+
+\todo[inline]{What is our goal with this?}
+
+This paper is structured as follows.
+In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
+In \cref{sec:model} we detail our model for the generation of realistic human timings in \ac{WCA}.
+We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
+Next in \cref{sec:implications} we discuss the potential implications of such a model by studying a small series of representative scenarios.
+Finally, in \cref{sec:conclusion} we summarize and conclude this paper.

### 2022-11-01 intro.tex
intro draft

@@
 They do so by capturing user and environment information, and leveraging compute-intensive algorithms to analyze the data and produce timely context-dependent feedback.
-A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses --- they are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers by tracking their progress and providing feedback and instructions at appropriate times.
-
 \ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{Ha2014towards,Satya2019augmenting}, and have since then expanded to a greater range of use cases.
@@
 
-As their name suggests, a defining characteristic of these applications is their wearability, which implies the use of lightweight, battery-powered, low-energy consumption devices.
+A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses --- they are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers by tracking their progress and providing feedback and instructions at appropriate times.
+Moreover, as their name suggests, another key attribute of these applications is their wearability, which implies the use of lightweight, battery-powered, low-energy consumption devices.
 On the other hand, their context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs such as real-time video feeds which require intensive compute capabilities to process.
 Furthermore, as with any \ac{AR} application, \ac{WCA} systems are sensitive to latency.
-Delays and jitter can be jarring to the user, causing them discomfort, leading them to make mistakes and potentially even give up on the application altogether.
+Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
 
 The combination of these characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
+One key gap in knowledge pertains to the current lack of tools and methodologies for scalable and repeatable study of \ac{WCA} application performance and resource utilization in realistic deployments.
+Due to their human-in-the-loop nature, these applications present a challenge to benchmark and characterize, in particular in real-scale deployments where dozens or even hundreds of users might concurrently use the system.
+Recruiting a large enough cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers.
+Furthermore, humans are unpredictable, and human-subject research is thus often hard to repeat and replicate.
+There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do away with the human-the-loop.
 
-\todo[inline]{WIP}
+In this paper, we introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
+Furthermore, we introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
 
-This paper builds on previous work on the characterization of human perception of delay in \ac{WCA} systems.
+Our contributions build on previous work on the characterization of human perception of delay in \ac{WCA} systems.
 We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.
@@
 We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
-In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we use a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
+In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
 In \textcite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
@@
 
-In this paper, we introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
-Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
-Furthermore, we introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-We name this model \emph{EdgeDroid \num{2.0}}, as it corresponds to a direct, more realistic evolution of our initial EdgeDroid approach.
-
+We name the new model presented in this paper \emph{EdgeDroid \num{2.0}}; a direct, more realistic evolution of our initial EdgeDroid approach.
 \todo[inline]{What is our goal with this?}

### 2022-11-01 conclusion.tex
working

\section{Conclusion}\label{sec:conclusion}

The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic human-in-the-loop nature.
Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming and prohibitively expensive for many research groups.
In this paper, we have presented the \edgedroid{} model of human timing behavior for \ac{WCA}

### 2022-11-02 background.tex
massaging sampling section
writing intro
working
discussing neuroticism

@@
     \end{subfigure}
-    \caption{}
+    \caption{Key concepts in \acl{WCA}}
 \end{figure}
@@
 
-    \item The above effects are modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}, as evidenced by a \ac{PCA}.
+    \item The above effects are modulated by the \acs*{BFI}~\cite{oliver:bfi1999} trait of \emph{neuroticism}.
+    This is discussed in more detail below in \cref{ssec:moderationeffects}
+
+    % , as evidenced by a \ac{PCA}.
 \end{enumerate}
 
+\subsubsection{Moderation of effects by individual characteristics}\label{ssec:moderationeffects}
+
+As mentioned above, we collected variables related to well-known individual differences.
+These encompassed both the \acf{BFI}, introduced by \textcite{oliver:bfi1999}, and the \acf{ITQ}, described by \textcite{witmer1998measuring}.
+Out of the collected individual-difference variables, the most salient one corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
+Delay discounting is the tendency to devalue rewards for which one must wait; high rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
+    \caption{%
+        Correlation between neuroticism and the mean execution time of the last four steps in blocks of \num{12} steps subject to the same \ac{TTF} in ~\textcite{olguinmunoz:impact2021}.
+    }\label{fig:neuroexectimecorrel}
+\end{figure}
+
+Neuroticism was identified as a modulating factor for the previously discussed pacing effects through a \ac{PCA}.
+Out of the three identified components, which cumulatively accounted for \SI{73.13}{\percent} of the variance in the results, neuroticism was included in the first two.
+The effect of neuroticism was transversal to all \acp{TTF} and impairment durations in the tasks, and we show an example of this correlation between execution time and neuroticism in \cref{fig:neuroexectimecorrel}
+
+Furthermore, we found that execution times were well-fit by an \ac{exGaussian} distribution, a distribution with ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+Fits were performed using the \ac{MLE} method.
+
 In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+
+\todo[inline]{Include section on neuroticism? Ask Bobby?}

### 2022-11-02 model.tex
discussing neuroticism

@@
 
-The distribution chosen for the second variant described above corresponds to the \ac{exGaussian} distribution.
-We've chosen this distribution due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+The distribution chosen for the second variant described above corresponds to the \ac{exGaussian} distribution, previously discussed in \cref{ssec:moderationeffects}.
 Fitting is done using \ac{MLE}.

### 2022-11-02 background.tex
comments
Update on Overleaf.
neuroticism

--- 2022-10-26 background.tex
@@
 
-    \item The above effects are modulated by the Big Five~\cite{oliver:bfi1999} trait of \emph{neuroticism}, as evidenced by a \ac{PCA}.
+    \item The above effects are modulated by an individual personality trait.
+    We will discuss this in detail below in \cref{ssec:moderationeffects}
 \end{enumerate}
 
-In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+\subsubsection{Moderation of effects by individual characteristics}\label{ssec:moderationeffects}
+
+We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}, introduced by \textcite{oliver:bfi1999}, and the \acf{ITQ}, described by \textcite{witmer1998measuring}.
+Out of the collected individual-difference variables, the most salient one corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
+Delay discounting is the tendency to devalue rewards for which one must wait; high rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
+    \caption{%
+        Correlation between neuroticism and the mean execution time of the last four steps in segments of \num{12} steps subject to the same \ac{TTF} in~\textcite{olguinmunoz:impact2021}.
+        Pearson correlation coefficient \( \rho = 0.418 \), 2-tailed \( p < 0.05 \).
+    }\label{fig:neuroexectimecorrel}
+\end{figure}
+
+Neuroticism was identified as a modulating factor for the pacing effects through a \ac{PCA}.
+Out of the three identified components, which cumulatively accounted for \SI{73.13}{\percent} of the variance in the results, neuroticism was included in the first two.
+The effect of neuroticism was transversal to all \acp{TTF} and impairment durations in the tasks --- we show an example of the correlation between execution time and neuroticism in \cref{fig:neuroexectimecorrel}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\columnwidth]{figs/new_model/mu_fits_exgaussian_slice0.png}
+    \caption{%
+        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \ac{TTF} in \textcite{olguinmunoz:impact2021}.
+        Fits were performed using the \ac{MLE} method.
+    }\label{fig:muexgaussian}
+\end{figure}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figs/new_model/dist_fits_neuro.png}
+    \caption{%
+        Example \ac{exGaussian} fits on execution times from steps \numrange{4}{8} in a segment of steps at the maximum experimental \ac{TTF}.
+        The effects of neuroticism are clearly visible in the tail and the mean of the distributions.
+    }\label{fig:fitsneuro}
+\end{figure*}
+
+Furthermore, we found that execution times, when grouping by experimental variables such as neuroticism, \ac{TTF}, and continuous segments of steps subject to the same \ac{TTF}, were well-fit by an \ac{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
+This distribution has an ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+We found that the effects of neuroticism on execution times were clearly identifiable in the fitted distributions, in particular in their means and tails.
+\cref{fig:muexgaussian} shows an example of this modulating effect, illustrating the behavior of the mean (\( \mu \) parameter) of \ac{exGaussian} distributions fitted to the execution times of the first four steps of segments of steps subject to the same \ac{TTF}.
+Finally, \cref{fig:fitsneuro} shows an example of the effects of neuroticism on the fitted \ac{exGaussian} distributions for a specific group of execution times.
+Higher neuroticism directly translates into a higher mean and longer tail.

### 2022-11-02 model.tex
neuroticism

@@
 
-In the following subsections, we detail the construction of a model which uses the data collected for \textcite{olguinmunoz:impact2021} to generate, at runtime, realistic execution times. 
+In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+We detail the construction of a model which uses the data collected to generate, at runtime, realistic execution times. 
 In order to accurately emulate the behavior of a human, such a model needs to implement two main behaviors.

### 2022-11-03 background.tex
greedy

@@
 If we define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), it must follow that (see also~\cite{olguinmunoz:impact2021})\footnote{\( U(a, b) \) represents the continuous uniform distribution in the open interval \( (a, b) \).}
+% \todo[inline]{I think it's an assumption rather than a must. Specifically, it is not uniform if the general execution time distribution is Rayleigh/ExpGaussian -vishnu}
 \begin{align}\label{eq:tc}
@@
     \centering
-    \includegraphics[width=.9\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
+    \includegraphics[width=\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=.9\columnwidth]{figs/new_model/mu_fits_exgaussian_slice0.png}
+    \includegraphics[width=\columnwidth]{figs/new_model/mu_fits_exgaussian_slice0.png}
     \caption{%

### 2022-11-03 implications_optimization.tex
writing
writing intro
greedy

@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_all_ttfs.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_all_ttfs.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_diff.png}
         \caption{%
             Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
         }
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_testbed.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/lifetime_testbed_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed_diff.png}
         \caption{%
@@
 
+\todo[inline]{%
+    Ergo we need a second order model. Everything that comes afterwards is only with second order model. Play with RTT.
+}
+
 % .as well as an additional setup with \num{10} clients during which an instance of the \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} network load generator is also running on each client.
@@
 \( R_\text{max} \) and \( R_\text{min} \) correspond to the maximum and minimum sampling rates of the system, respectively.
-\( R_\text{max} \) can directly be assumed to correspond to one over the mean \ac{RTT} of the system, whereas \( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually.
-\( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
-Finally, \( CDF \) corresponds to the \acl*{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
+\( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
+\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
+Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
 
@@
 For this, we implement two variants of the adaptive scheme.
-The first one corresponds to \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
+The first one corresponds a reference implementation using \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
+This variant the same \ac{CDF} for every step.
 The second variant instead uses an embedded internal execution time model to keep track of the \acp{TTF} of the system and adjust the distribution of execution times following the state of the timing model.
+In other words, in this variant the \ac{CDF} used for instantaneous calculation of the sampling rate \( R \) changes over time according to the state of the execution time model.
+For timing model configurations which do not fit a distribution to execution time data, the \ac{CDF} corresponds to the \ac{ECDF} of the selected execution times.
 
 We proceed to set up an experiment where these sampling scheme variants are deployed on identical, simulated, \num{100}-step tasks.
+Also included for reference in this experiment is greedy (also known as \emph{zero-wait}) sampling, where samples are taken as fast as possible; a new sample is taken as soon as the result for the previous one has been received.
+This is the default sampling strategy used in most existing \ac{WCA} prototypes.
 The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_junjue.png}
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_junjue.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/diff_sampling_junjue.png}
+        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue.png}
         \caption{%
             Percentage difference in mean number of samples per step.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
+            \todo[inline]{%
+                Add greedy.
+                Develop optimal sampling scheme?
+                Explain grouping better?
+            }
         }
     \end{subfigure}
-    \caption{Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.}\label{fig:samples}
+    \caption{%
+        Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.
+        \emph{ED2} indicates a sampling scheme using the \edgedroid{} model for \ac{CDF} estimation.
+        \emph{HN} and \emph{LN} indicate high or low neuroticism, respectively, and \emph{ExG fit} indicates a timing model which fits an \ac{exGaussian} distribution to the execution time data for each step.
+    }\label{fig:samples}
 \end{figure}
 
-The results of this investigation are presented in \cref{fig:samples}, and they clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling scheme achieve between \SI{22}{\percent} and \SI{27}{\percent} better average performance.
+The results of this investigation are presented in \cref{fig:samples}
+They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
+When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{22}{\percent} and \textasciitilde\SI{25}{\percent} better average performance.
 This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
@@
 \begin{multline*}
-    =\Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)\Big)\\
+    =\Big(F_\mathcal{T}(t_n)-\\\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)\Big)\\
     \div \Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)\\+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)\Big).
@@
 \end{alignat}
+
+\subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}
+
+\todo[inline]{
+    TODO.
+
+    Plot energy consumption over RTT, mean energy per step over a task.
+    Ground truth empirical high.
+    For greedy, ideal, junjues sampling, vishnus model with gt as estimator.
+}

### 2022-11-03 model.tex
greedy

@@
     \centering
-    \includegraphics[width=.9\columnwidth]{figs/new_model/ttf_to_exectime.png}
+    \includegraphics[width=\columnwidth]{figs/new_model/ttf_to_exectime.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=.9\columnwidth]{figs/new_model/exectime_over_steps.png}
+    \includegraphics[width=\columnwidth]{figs/new_model/exectime_over_steps.png}
     \caption{%

### 2022-11-03 implications_optimization.tex
sampling

@@
             Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
         }
@@
 The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
-We set \( R_\text{min} \) to \SI{0.5}{\hertz}, \( \alpha \) to \num{3.0}, and the \ac{RTT} of the system to a constant \SI{150}{\milli\second}, and then run \num{90} repetitions of the task for the original design of the sampling scheme as well as for each possible configuration of it with a different parameterization of the timing model.
+We set \( R_\text{min} \) to \SI{0.5}{\hertz} and \( \alpha \) to \num{3.0}.
+The \ac{RTT} of the system is set to one of \SIlist[list-final-separator={, or}]{150;300;600}{\milli\second}, constant for all steps of the task.
+For each combination of sampling scheme configuration (original adaptive by \textcite{Wang2019Towards}, adaptive with \edgedroid{} model, greedy) and \ac{RTT}, we run \num{100} repetitions of the task for good statistical significance.
 
-\begin{figure}
+\begin{figure*}
     \centering
-    \begin{subtable}[]{\columnwidth}
+    \begin{subtable}[t]{.3\textwidth}
         \centering
-        \begin{tabular}{rrr}
+        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
+        \medskip
+        \begin{tabular}{lr}
             \toprule
-            \( R_\text{min} \) & \( \alpha \) & \ac{RTT} \\
-            \midrule
-            \SI{0.5}{\hertz} & \num{3.0} & \SI{150}{\milli\second} \\
+            \( R_\text{min} \)  & \SI{0.5}{\hertz} \\
+            \( \alpha \)        & \num{3.0} \\
+            \ac{RTT}            & \SIlist{150;300;600}{\milli\second} \\
             \bottomrule
         \end{tabular}
-        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
-    \end{subtable}
-    \begin{subfigure}[]{\columnwidth}
+    \end{subtable}%
+    \hspace{.05\textwidth}%
+    % \begin{subfigure}[]{\columnwidth}
+    %     \centering
+    %     \includegraphics[width=\textwidth]{figs/new_model/sampling_junjue.png}
+    %     \caption{%
+    %         Mean number of samples per step over a \num{100}-step task, \num{90} independent repetitions per configuration.
+    %         Error bars indicate \SI{95}{\percent} \acp{CI}.
+    %     }
+    % \end{subfigure}
+    \begin{subtable}[t]{.6\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_junjue.png}
+        \caption{Mean number of captured samples per sampling scheme at the three different \acp{RTT}.}
+        \medskip
+        \begin{tabular}{lrrr}
+            \toprule
+            Sampling scheme & \SI{150}{\milli\second} & \SI{300}{\milli\second} & \SI{600}{\milli\second} \\
+            \midrule
+            Adaptive (ED2 LN) & \num{14.69} & \num{8.07} & \num{5.10} \\
+            Adaptive (ED2 HN) & \num{14.60} & \num{8.03} & \num{5.05} \\
+            Adaptive (ED2 LN, ExG fit) & \num{15.30} & \num{8.23} & \num{5.09} \\
+            Adaptive (ED2 HN, ExG fit) & \num{15.05} & \num{8.47} & \num{5.20} \\
+            Adaptive (reference, gaussian fit) & \num{20.17} & \num{10.65} & \num{6.04} \\
+            Greedy & \num{36.68} & \num{19.11} & \num{10.36} \\
+            \bottomrule
+        \end{tabular}
+    \end{subtable}\\
+    \bigskip
+    \begin{subfigure}[]{\textwidth}
+        \centering
+        \includegraphics[width=.8\textwidth]{figs/new_model/diff_sampling_junjue.png}
         \caption{%
-            Mean number of samples per step over a \num{100}-step task, \num{90} independent repetitions per configuration.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
-        }
-    \end{subfigure}
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue.png}
-        \caption{%
-            Percentage difference in mean number of samples per step.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
-            \todo[inline]{%
-                Add greedy.
-                Develop optimal sampling scheme?
-                Explain grouping better?
-            }
+            Percentage reduction in mean number of captured samples per step (higher is better).
+            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+            % \todo[inline]{%
+            %     Add greedy.
+            %     Develop optimal sampling scheme?
+            %     Explain grouping better?
+            % }
         }
@@
     }\label{fig:samples}
-\end{figure}
+\end{figure*}
 
@@
 They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{22}{\percent} and \textasciitilde\SI{25}{\percent} better average performance.
+When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{14}{\percent} and \textasciitilde\SI{27}{\percent} better average performance, depending on the \edgedroid{} model parameterization and the system \ac{RTT}.
+\todo[inline]{Discuss that lower RTTs lead to better results?}
 This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.

### 2022-11-04 implications_optimization.tex
update plots

@@
     \centering
-    \begin{subtable}[t]{.3\textwidth}
+    \begin{subtable}[t]{.25\textwidth}
         \centering
@@
     \hspace{.05\textwidth}%
-    % \begin{subfigure}[]{\columnwidth}
-    %     \centering
-    %     \includegraphics[width=\textwidth]{figs/new_model/sampling_junjue.png}
-    %     \caption{%
-    %         Mean number of samples per step over a \num{100}-step task, \num{90} independent repetitions per configuration.
-    %         Error bars indicate \SI{95}{\percent} \acp{CI}.
-    %     }
-    % \end{subfigure}
-    \begin{subtable}[t]{.6\textwidth}
+    \begin{subtable}[t]{.65\textwidth}
         \centering
@@
         \medskip
-        \begin{tabular}{lrrr}
+        \begin{tabular}{llrrr}
             \toprule 
-            Sampling scheme & \SI{150}{\milli\second} & \SI{300}{\milli\second} & \SI{600}{\milli\second} \\
+            Timing model neuro. & Sampling scheme & \SI{150}{\milli\second} & \SI{300}{\milli\second} & \SI{600}{\milli\second} \\
            \midrule
-            Adaptive (ED2 LN) & \num{14.69} & \num{8.07} & \num{5.10} \\
-            Adaptive (ED2 HN) & \num{14.60} & \num{8.03} & \num{5.05} \\
-            Adaptive (ED2 LN, ExG fit) & \num{15.30} & \num{8.23} & \num{5.09} \\
-            Adaptive (ED2 HN, ExG fit) & \num{15.05} & \num{8.47} & \num{5.20} \\
-            Adaptive (reference, gaussian fit) & \num{20.17} & \num{10.65} & \num{6.04} \\
-            Greedy & \num{36.68} & \num{19.11} & \num{10.36} \\
+           \multirow[c]{2}{*}{Low} & Greedy & \num{36.53} & \num{18.88} & \num{10.52} \\
+            & Adaptive Gaussian fit & \num{19.97} & \num{10.43} & \num{6.39} \\
+           \multirow[c]{2}{*}{High} & Greedy & \num{37.05} & \num{19.04} & \num{10.89} \\
+            & Adaptive Gaussian fit & \num{20.29} & \num{10.79} & \num{6.67} \\
             \bottomrule
@@
         \centering
-        \includegraphics[width=.8\textwidth]{figs/new_model/diff_sampling_junjue.png}
+        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue_both.png}
         \caption{%
@@
             Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
-            % \todo[inline]{%
-            %     Add greedy.
-            %     Develop optimal sampling scheme?
-            %     Explain grouping better?
-            % }
+            \todo[inline]{%
+                % Vishnu: minimize number of samples.
+                I don't like the two plots, I think the message gets diluted and confusing.
+                I suggest only keeping one ``ground truth'', high neuroticism.
+                Otherwise, every plot and table needs to be duplicated.
+            }
         }
@@
 When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{14}{\percent} and \textasciitilde\SI{27}{\percent} better average performance, depending on the \edgedroid{} model parameterization and the system \ac{RTT}.
-\todo[inline]{Discuss that lower RTTs lead to better results?}
+\todo[inline]{Update with new ground truths.}
 This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
@@
 
+\begin{figure*}
+    \centering
+    \begin{subfigure}[t]{\textwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy.png}
+        \caption{}
+    \end{subfigure}\\
+    \medskip
+    \begin{subfigure}[t]{\textwidth}
+        \centering
+        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy_diff.png}
+        \caption{}
+    \end{subfigure}
+\end{figure*}
+
 \todo[inline]{
@@
 
-    Plot energy consumption over RTT, mean energy per step over a task.
-    Ground truth empirical high.
-    For greedy, ideal, junjues sampling, vishnus model with gt as estimator.
+    Explain plots.
 }

### 2022-11-04 sampling.tex
Update on Overleaf.

@@
 \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.
+\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.\label{tnFinal}
 \end{alignat}
+
+Now, if we have a constrained optimisation with an upper bound $w_0$ for the expected wait time, we can find the optimum set of sampling instants by equating \eqref{Ew} to $w_0$ and find the ratio of $\frac{\alpha}{\beta}$ that provides the optimum set of sampling instants when plugged into \eqref{tnFinal}.
+For this, we make use of the facts that (1) the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals and (2)one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimisation discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+\begin{alignat*}{1}
+\mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber\\
+&=\int_{0}^{\infty}\frac{1}{2}\sqrt{\frac{2\alpha\sigma^2}{\beta t}}\cdot \frac{t}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\int_{0}^{\infty}\frac{\sqrt{t}}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\int_{0}^{\infty}y^{-1/4}e^{-y}\mathrm{d}y\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4})\\
+% &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+\intertext{Thus, when the upper bound $w_0$ is tight,}
+w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+w_0&=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
+\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\&=1.88355\frac{w_0^2}{\sigma}
+\end{alignat*}

### 2022-11-05 background.tex
add James comments as todonotes

@@
 
+\todo[inline]{%
+Talk about sampling, but don’t introduce greedy right away.
+First explain the sampling process generally (why is sampling necessary?) and then mention that different sampling strategies exist, for instance greedy ….
+}
+
 Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals (following a greedy scheme) at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
@@
 
+\todo[inline]{%
+Again, turn it around. The uniform distribution is less relevant than the terms execution time, wait time and last sample etc. Introduce the distribution at the end of the paragraph.
+}
+
 We define the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
@@
 \subsection{The effects of changes in responsiveness on human behavior}\label{ssec:plos}
+\todo[inline]{Shorten}
 

### 2022-11-05 intro.tex
working
add James comments as todonotes

@@
 
-We name the new model presented in this paper \emph{EdgeDroid \num{2.0}}; a direct, more realistic evolution of our initial EdgeDroid approach.
+We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
 \todo[inline]{What is our goal with this?}
@@
 Finally, in \cref{sec:conclusion} we summarize and conclude this paper.
+
+\todo[inline]{%
+Nice intro.
+Perhaps too long, and a concise presentation - preferably as bullet list - of the contributions is missing.
+Try to shorten in particular for the first three paragraphs.
+Another question would be if there is no other related work.
+For instance, Zorzi recently published a model for AR and VR - trace-based - that would be a good reference to have.
+}

### 2022-11-05 appendices.tex

--- 2022-11-03 implications_optimization.tex
@@
-\section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
-% \todo[inline]{Introduction to this section?}
-
-\subsection{Application lifetimes}
-
-We begin by studying the implications of such a model on the estimation of application lifetimes.
-In the context of \ac{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
-This is an important metric for \ac{WCA} optimization, as it directly relates to system resource utilization and contention, and to energy consumption.
-
-In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This new model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
-This distribution is then randomly sampled at runtime to obtain execution times for each step, without any sort of adjustment to the current state of the system.
-
-We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
-These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \ac{TTF} and record the generated execution time.
-We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
-This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \ac{TTF}.
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_all_ttfs.png}
-        \caption{%
-            Evolution of mean application lifetimes as \acp{TTF} increase.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
-        }
-    \end{subfigure}
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_diff.png}
-        \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
-        }
-    \end{subfigure}
-    \caption{\acl{TTF} versus application lifetime.}\label{fig:lifetimes}
-\end{figure}
-
-The results of this investigation are presented in \cref{fig:lifetimes}.
-Compared to the reference model, our realistic model is, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
-At higher \acp{TTF}, the behavior of the model depends on its level of normalized neuroticism.
-At low neuroticism, the behavior of the realistic model results in total task durations that are basically indistinguishable from the reference model.
-However, at high neuroticism, the model once again results in a considerable difference in total task duration with respect to the reference --- this time extending durations by \textasciitilde\SI{11}{\percent} on average.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
-    \caption{%
-        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \acs*{COTS} \acs{IEEE} \num{802.11}b/g access point.
-    }\label{fig:expsetup}
-\end{figure}
-
-Next we study the effects of first- versus second-order models in a more realistic setting.
-The models, reference and realistic, are deployed on the Raspberry Pi clients of the testbed depicted in \cref{fig:expsetup}.
-For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \ac{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
-Clients and cloudlet communicate over an \acs{IEEE} \num{802.11}b/g wireless network.
-Our choice of wireless standard is simply motivated by a desire to amplify the potential effects of network congestion.
-\todo[inline]{More justification?}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed.png}
-        \caption{%
-            Mean application lifetimes per testbed configuration.
-            Note that due to the low number of samples, means have been calculated using the geometric instead of arithmetic average. 
-            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using bootstrapping.
-        }
-    \end{subfigure}
-    \par\bigskip
-    \begin{subtable}{\columnwidth}
-        \centering
-        \begin{tabular}{lrrrr}
-            \toprule
-            \# clients & 1 & 4 & 7 & 10 \\
-            \ac{RTT} & \SI{0.42}{\second} & \SI{1.12}{\second} & \SI{1.92}{\second} & \SI{2.68}{\second} \\
-            \bottomrule
-        \end{tabular}
-        \caption{Mean measured \aclp{RTT} for each testbed configuration.}
-    \end{subtable}
-    \par\bigskip
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed_diff.png}
-        \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model.
-            Confidence intervals have been omitted due to the low number of samples and the use of the geometric mean.
-        }
-    \end{subfigure}
-    \caption{Application lifetimes in the realistic scenarios.}\label{fig:testbed_lifetimes}
-\end{figure}
-
-We deploy configurations running \num{45}-step versions of the LEGO task described in \textcite{olguinmunoz:impact2021}.
-The testbed configurations include setups with \numlist{1;4;7;10} clients.
-Due to having only limited time, each combination of testbed and timing model configuration is only repeated \num{10} times.
-The results are presented in \cref{fig:testbed_lifetimes}.
-Owing to the low number of samples, we opt here for the geometric rather than arithmetic mean to represent our results.
-The results are nonetheless clear, and follow the same pattern as the previously discussed results under ideal, controlled circumstances.
-With just a single client and mean \ac{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
-At \num{10} clients, the results mimic those at higher \acp{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
-
-These results highlight the importance of accurate execution time modeling when studying \ac{WCA} application lifetimes.
-Not only do we see considerable differences in lifetimes at relatively moderate levels of system congestion, but the sign of these differences depends on the load placed on the system.
-Imagine thus a system designer studying resource consumption optimization in a \ac{WCA}.
-If they were to employ the reference model for their study on an unimpaired system, it could lead them to significantly underestimate the potential for optimization of resource consumption, leaving performance (and, potentially, monetary) gains on the table.
-On the other hand, under heavy load, they would instead underestimate system resource occupation, again leading to performance losses.
-
-\todo[inline]{%
-    Ergo we need a second order model. Everything that comes afterwards is only with second order model. Play with RTT.
-}
-
-% .as well as an additional setup with \num{10} clients during which an instance of the \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} network load generator is also running on each client.
-% The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \ac{TCP} traffic, further congesting the wireless network.
-
-\subsection{Optimizing for mean number of samples per step}
-
-In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \ac{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
-At every sampling instant \( t \), the scheme adapts the sampling rate \( R(t) \) of the system according to the estimated likelihood of the user having finished the step,
-following the formula
-
-\begin{equation}
-    R(t) = R_\text{min} + \alpha\left( R_\text{max} - R_\text{min} \right) * CDF(t)
-\end{equation}
-
-\( R_\text{max} \) and \( R_\text{min} \) correspond to the maximum and minimum sampling rates of the system, respectively.
-\( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
-\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
-Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
-
-In the following, we will show the effects of more accurate execution time estimation using our timing models on the performance of this adaptive scheme.
-For this, we implement two variants of the adaptive scheme.
-The first one corresponds a reference implementation using \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
-This variant the same \ac{CDF} for every step.
-The second variant instead uses an embedded internal execution time model to keep track of the \acp{TTF} of the system and adjust the distribution of execution times following the state of the timing model.
-In other words, in this variant the \ac{CDF} used for instantaneous calculation of the sampling rate \( R \) changes over time according to the state of the execution time model.
-For timing model configurations which do not fit a distribution to execution time data, the \ac{CDF} corresponds to the \ac{ECDF} of the selected execution times.
-
-We proceed to set up an experiment where these sampling scheme variants are deployed on identical, simulated, \num{100}-step tasks.
-Also included for reference in this experiment is greedy (also known as \emph{zero-wait}) sampling, where samples are taken as fast as possible; a new sample is taken as soon as the result for the previous one has been received.
-This is the default sampling strategy used in most existing \ac{WCA} prototypes.
-The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
-We set \( R_\text{min} \) to \SI{0.5}{\hertz}, \( \alpha \) to \num{3.0}, and the \ac{RTT} of the system to a constant \SI{150}{\milli\second}, and then run \num{90} repetitions of the task for the original design of the sampling scheme as well as for each possible configuration of it with a different parameterization of the timing model.
-
-\begin{figure}
-    \centering
-    \begin{subtable}[]{\columnwidth}
-        \centering
-        \begin{tabular}{rrr}
-            \toprule
-            \( R_\text{min} \) & \( \alpha \) & \ac{RTT} \\
-            \midrule
-            \SI{0.5}{\hertz} & \num{3.0} & \SI{150}{\milli\second} \\
-            \bottomrule
-        \end{tabular}
-        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
-    \end{subtable}
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_junjue.png}
-        \caption{%
-            Mean number of samples per step over a \num{100}-step task, \num{90} independent repetitions per configuration.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
-        }
-    \end{subfigure}
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue.png}
-        \caption{%
-            Percentage difference in mean number of samples per step.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sample T-test.
-            \todo[inline]{%
-                Add greedy.
-                Develop optimal sampling scheme?
-                Explain grouping better?
-            }
-        }
-    \end{subfigure}
-    \caption{%
-        Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.
-        \emph{ED2} indicates a sampling scheme using the \edgedroid{} model for \ac{CDF} estimation.
-        \emph{HN} and \emph{LN} indicate high or low neuroticism, respectively, and \emph{ExG fit} indicates a timing model which fits an \ac{exGaussian} distribution to the execution time data for each step.
-    }\label{fig:samples}
-\end{figure}
-
-The results of this investigation are presented in \cref{fig:samples}
-They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{22}{\percent} and \textasciitilde\SI{25}{\percent} better average performance.
-This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
-
-\subsection{Optimizing for energy consumption}
-
-Energy consumption is another yet unexplored dimension of \ac{WCA} with great opportunities for optimization.
-To explore the potential implications of our realistic model of human timings for energy consumption, we will introduce here a novel scheme for sampling in step-based \ac{WCA} systems.
-
-We begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
-In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
-The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
-In this optimization, the constant parts are excluded as it does not affect the process.
-With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is the reason why $\mathcal{W}$ is computed without including it.
-Next, in \textcite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
-
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
-However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of optimum aperiodic sampling interval.
-In what follows, we first recap the system modeling used by the authors \textcite{secAperiodic} in and then give our independent approximate solution.
-
-% \begin{table}
-%     \centering
-%     \caption{}
-%     \begin{tabular}{ll}
-%         \toprule
-%         Parameter & Explanation \\
-%         \midrule
-%         \( P_0 \) & Idle power consumption \\
-%         \( P_\text{c} \) & Power consumption during communication with backend \\
-%         \bottomrule 
-%     \end{tabular}
-% \end{table}
-
+\subsection*{Optimisation}
 The energy penalty can be expressed as
@@
 
-First, we borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is
+First, we borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is
 related to $\{t_n\}$ such that,
@@
 \end{alignat}
-
-\subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}
-
-\todo[inline]{
-    TODO.
-
-    Plot energy consumption over RTT, mean energy per step over a task.
-    Ground truth empirical high.
-    For greedy, ideal, junjues sampling, vishnus model with gt as estimator.
-}

### 2022-11-05 implications_optimization.tex

@@
 In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
-The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
-In this optimization, the constant parts are excluded as it does not affect the process.
-With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is the reason why $\mathcal{W}$ is computed without including it.
-Next, in \textcite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
+% In this optimization, the constant parts are excluded as it does not affect the process.
+With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
+Next, in \textcite{secAperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
 
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
-However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of optimum aperiodic sampling interval.
-In what follows, we first recap the system modeling used by the authors \textcite{secAperiodic} in and then give our independent approximate solution.
+In this work, we use the modelling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
+However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
+Furthermore, instead of directly optimising for energy, we start by noting that an objective metric of a variety of optimisation problems that are related to sampling, RTT, responsiveness or wait time present themselves
+% in a format similar to that of the energy, 
+as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
+Let $\mathcal{E}$ correspond to such an objective metric.
+Thus,
+\begin{alignat}{1}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
+\end{alignat}
+Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
+For instance, in the modelling used by Moothedath et.al~\cite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterisation results in a metric which is equal to the energy penalty.
+The outline of our solution that minimises \eqref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in APPENDIX \ref{appx1}.
 
-% \begin{table}
-%     \centering
-%     \caption{}
-%     \begin{tabular}{ll}
-%         \toprule
-%         Parameter & Explanation \\
-%         \midrule
-%         \( P_0 \) & Idle power consumption \\
-%         \( P_\text{c} \) & Power consumption during communication with backend \\
-%         \bottomrule 
-%     \end{tabular}
-% \end{table}
-
-The energy penalty can be expressed as
-\begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
-\end{alignat}
-where \( P_0 \) corresponds to the power consumed by the \ac{WCA} client device during idle periods and \( P_\text{c} \) to the power consumption during communication with the backend (i.e.\ transmitting a sample or receiving a reply).
-$\alpha = \tau_{\text{c}} (P_{\text{c}} - P_0)$ and $\beta = P_0$ are constants.
-Here, note that $\alpha\mathbb{E}[\mathcal{S}]$ and $\beta\,\mathbb{E}[\mathcal{W}]$ correspond to the energy wasted per discarded sample and the additional energy expended while waiting, respectively.
-Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimizes the energy penalty given in \cref{eq:epsilon_terminal}.
-
-\subsubsection{Optimum sampling instants}\label{sec:aprxSol}
-
-First, we borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is
-related to $\{t_n\}$ such that,
+\subsubsection{General solution}\label{sec:aprxSol}
+Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
@@
 \end{alignat}
-We find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
-Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
-Thus, we get the expected number of samples $\mathbb{E}[\mathcal{S}]$ as
+For instance, for periodic sampling, the rate function is a constant equal to the sampling frequency. 
+% Using this rate function, we rewrite the expected number of samples as
+% \begin{alignat}{1}
+% \mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t,\nonumber
+% \end{alignat}
+% and the approximate the expected wait time as
+% \begin{alignat}{1}
+% \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber
+% \end{alignat}
+Now, rewriting \eqref{eq:epsilon_terminal} in terms of $r(t)$ and minimising it gives as an optimum rate function $r^*(t)$ as
 \begin{alignat}{1}
-\mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{Es}
+r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\nonumber\\
+\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
+r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
+% \Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
+% \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
+\Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.\label{tn_approx_rayleigh}
 \end{alignat}
-To find 
-% the expected wait 
-$\mathbb{E}[\mathcal{W}]$, we use the conditional \ac{CDF} of the execution time.
-We get,
-\begin{multline*}
-\mathbb{P}(\mathcal{W}=t_n-\mathcal{T}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)\\=\dfrac{\mathbb{P}(\mathcal{T}\geq t_n-t\,,\,t_{n-1}<\mathcal{T}\leq t_n)}{\mathbb{P}(t_{n-1}<\mathcal{T}\leq t_n)}.
-\end{multline*}
-The numerator is degenerate when $t\!<\!0$ or $t\!>\!(t_n\!-\!t_{n-1})$.
-Thus, we are only interested in $0\!\leq\!t\!\leq\! (t_n-t_{n-1})$.
-Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ correspond to the \ac{CDF}, \ac{CCDF} and \ac{PDF} of the execution time distribution.
-\begin{alignat*}{1}
-\!\!\!\Rightarrow\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&=\dfrac{\mathbb{P}(t_n-t\leq\mathcal{T}\leq t_n)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\\
-&\approx\dfrac{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n}-t)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\tag{$A_1$}\label{Apx1}
-\end{alignat*}
-Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
-Expanding the \ac{CDF} using Taylor series, 
-\begin{multline*}
-    =\Big(F_\mathcal{T}(t_n)-\\\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)\Big)\\
-    \div \Big(F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)\\+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)\Big).
-\end{multline*}
-Simplifying and using the first order approximation, we get
-\begin{alignat*}{1}
-% &=\dfrac{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)}{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)}\\&
-\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&\approx\dfrac{tf_\mathcal{T}(t_n)}{(t_{n}-t_{n-1})f_\mathcal{T}(t_n)}\tag{$A_2$}\label{Apx2}\\
-\Rightarrow\mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&= 1-\dfrac{t}{(t_{n}-t_{n-1})}.
-\end{alignat*}
-Using this \ac{CCDF}, we find the conditional expectation of $\mathcal{W}$.
-\begin{alignat*}{1}
-% \mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&\approx 1-\dfrac{t}{(t_{n}-t_{n-1})}\\
-\Rightarrow \mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]&=\smashoperator[r]{\int_{0}^{t_n-t_{n-1}}}\Big(1-\dfrac{t}{(t_{n}-t_{n-1})}\Big)\,\mathrm{d}t\\
-&=\dfrac{(t_n-t_{n-1})}{2}.
-\end{alignat*}
-% However, we can also approximate $(t_n-t_{n-1})$, the sampling interval to the inverse of the instantaneous sampling frequency $r(t)$.
-This approximation (\ref{Apx3}) is valid as long as $r(t)$ is varying slowly between two consecutive sampling instants.
-That is,
-If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $t_n\!-\!t_{n-1}$ as,
+\subsubsection{Energy minimisation}\label{sec:aprxSol_Energy}
+With the general solution available, minimising energy $E$ boils down to finding the appropriate $\alpha$ and $\beta$ that $\mathcal{E}=E$ in \eqref{eq:epsilon_terminal}.
+For this, we directly take the modelling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
+With feedback given even to the discarded samples in our model, we have,
+\begin{alignat}{2}
+    \mathrm{E}=&\;2\mathcal{S}\tau_cP_c+(\mathcal{T}+\mathcal{W}+\tau_\mathrm{p}+2\tau_\mathrm{c}-2\mathcal{S}\tau_c)P_0\nonumber\\
+    % &=s\tau_c(P_c-P_0^{(t)})+wP_0^{(t)}+(\tau+\tau_c+\tau_p) P_0^{(t)}+\tau_cP_c\\
+    =&\;2\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +2\tau_{\text{c}}) P_0\nonumber\\
+&\Rightarrow \alpha=2\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0.\nonumber
+\end{alignat}
+\todo[inline]{is communication delay and processing delay defined already?}
+\subsubsection{Minimising number of samples }\label{sec:aprxSol_Energy_samples}
+Apart from energy, another common optimisation metric is simply the number of samples.
+However, an unconstrained optimisation of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
+We look at the constrained optimisation of the expected number of samples with an upper bound $w_0$ for the expected wait.
+That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$.
+We show in APPENDIX \ref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
 \begin{alignat}{1}
- (t_n-t_{n-1})&\approx\dfrac{1}{r(t)},\;\forall t,n:t_{n-1}\!<\!t\!\leq\!t_n.\tag{$A_3$}\label{Apx3}\\
-% \end{alignat}
-% As a result, we can compute the expected wait as,
-% \begin{alignat}{1}
-\Rightarrow \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\mathbb{E}[\mathcal{W}\,|\,t_{n-1}<\mathcal{T}\leq t_n]f_\mathcal{T}(t)\,\mathrm{d}t\nonumber\\
-&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\label{Ew}
+\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma},\nonumber
 \end{alignat}
-We can thus find the energy penalty using \cref{Es} and \cref{Ew} as
-\begin{alignat*}{1}
-\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha\int_{x=0}^{t}r(x)\,\mathrm{d}x+\beta\dfrac{1}{2r(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{epsilon_eulerForm}\\
-\intertext{Let $g(t)=\int_{0}^{t}r(x)\dif x$. Then  $g'(t)=\tfrac{\mathrm{d}}{\mathrm{d}t}g(t)=r(t)$. That is,}
-\vspace{-4cm}\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha g(t)+\dfrac{\beta}{2g'(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.
-\end{alignat*}
-As per the Euler-Lagrange equation from the calculus of variations~\cite{calcVariation2,calcVariation1}, the extreme value of $\mathcal{E}$ is obtained at 
-\begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\\
-\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
-r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\\
-\Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
-\Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.
-\end{alignat}
-
+where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
 \subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}

### 2022-11-05 implications_optimization.tex
fixing Vishnus stuff

@@
 The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
+
 % In this optimization, the constant parts are excluded as it does not affect the process.
@@
 
-In this work, we use the modelling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
+In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
 However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
-Furthermore, instead of directly optimising for energy, we start by noting that an objective metric of a variety of optimisation problems that are related to sampling, RTT, responsiveness or wait time present themselves
+Furthermore, instead of directly optimizing for energy, we start by noting that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
 % in a format similar to that of the energy, 
@@
 Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
-For instance, in the modelling used by Moothedath et.al~\cite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterisation results in a metric which is equal to the energy penalty.
-The outline of our solution that minimises \eqref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in APPENDIX \ref{appx1}.
+For instance, in the modeling used by \textcite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
+The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
 
 \subsubsection{General solution}\label{sec:aprxSol}
+
 Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
@@
 % \end{alignat}
-Now, rewriting \eqref{eq:epsilon_terminal} in terms of $r(t)$ and minimising it gives as an optimum rate function $r^*(t)$ as
+Now, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
 \begin{alignat}{1}
@@
 \end{alignat}
-\subsubsection{Energy minimisation}\label{sec:aprxSol_Energy}
-With the general solution available, minimising energy $E$ boils down to finding the appropriate $\alpha$ and $\beta$ that $\mathcal{E}=E$ in \eqref{eq:epsilon_terminal}.
-For this, we directly take the modelling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
+
+\subsubsection{Energy minimization}\label{sec:aprxSol_Energy}
+
+With the general solution available, minimizing energy $E$ boils down to finding the appropriate $\alpha$ and $\beta$ that $\mathcal{E}=E$ in \cref{eq:epsilon_terminal}.
+For this, we directly take the modeling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
 With feedback given even to the discarded samples in our model, we have, 
@@
 \end{alignat}
+
 \todo[inline]{is communication delay and processing delay defined already?}
-\subsubsection{Minimising number of samples }\label{sec:aprxSol_Energy_samples}
-Apart from energy, another common optimisation metric is simply the number of samples.
-However, an unconstrained optimisation of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
-We look at the constrained optimisation of the expected number of samples with an upper bound $w_0$ for the expected wait.
-That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$.
-We show in APPENDIX \ref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
+
+\subsubsection{Minimizing number of samples }\label{sec:aprxSol_Energy_samples}
+
+Apart from energy, another common optimization metric is simply the number of samples.
+However, an unconstrained optimization of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
+We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
+That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
 \begin{alignat}{1}
@@
 where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
+
 \subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}

### 2022-11-05 sampling.tex
fixing Vishnus stuff

@@
 \section{Aperiodic Sampling}
-% !TEX root = 00_main.tex
+
 In this section we will explain the method that we used to find the optimum set of sampling instants (or equivalently sampling intervals) for the given distribution of TTE, denoted by $\mathcal{T}$.
-\todo[inline]{aka execution times.
-Need to recheck the terminologies used in this section and make it consistent with the paper.} We take the modelling and partial solution approach from the recent works on energy efficient sampling of edge-based feedback systems.
-In \cite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user, where the wait time $\mathcal{W}$ -- in the terminology of the authors -- is the time between the event completion and the immediate next sample taken.
-The authors then find the optimum periodic sampling interval that minimises this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
-In this optimisation, the constant parts are excluded as it does not affect the optimisation.
+\todo[inline]{aka execution times. Need to recheck the terminologies used in this section and make it consistent with the paper.}
+We take the modeling and partial solution approach from the recent works on energy efficient sampling of edge-based feedback systems.
+In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user, where the wait time $\mathcal{W}$ --- in the terminology of the authors --- is the time between the event completion and the immediate next sample taken.
+The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty} $\mathcal{E}$.
+In this optimization, the constant parts are excluded as they do not affect the optimization.
 With a value much smaller than the execution time of the event or TTE, the RTT of the final sample is included in this constant part, which is the reason why $\mathcal{W}$ is computed without including this RTT.
-Next, in \cite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimises the same energy penalty.
-In this work, we use the modelling from \cite{secAperiodic} to find the optimum aperiodic sampling interval.
-However, instead of using their two-step approach to find the solution which include a recursive solution followed by an algorithm, we develop a novel approximate but easier solution to finding the set of optimum aperiodic sampling interval.
-In what follows, we first recap the system modelling used by the authors \cite{secAperiodic} in and then give our independent approximate solution.
+Next, in \textcite{secAperiodic}, the authors use this exact model and find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimizes the same energy penalty.
+In this work, we use the modeling from \textcite{secAperiodic} to find the optimum aperiodic sampling interval.
+However, instead of using their two-step approach to find the solution which include a recursive solution followed by an algorithm, we develop a novel approximate but easier solution to finding the set of optimum aperiodic sampling intervals.
+In what follows, we first recap the system modeling used by the authors in \textcite{secAperiodic} and then give our independent approximate solution.
 
-The energy penalty can be expressed as \cite{secAperiodic}
+The energy penalty can be expressed as~\cite{secAperiodic}
 \begin{alignat}{1}
@@
 \end{alignat}
-where $\alpha\!=\!\tau_{\text{c}} (P_{\text{c}}\!-\!P_0)$ and $\beta\!=\!P_0$ are constants. \todo[inline]{Define and explain the parameters if this is not done earlier in other sections.}
+where $\alpha\!=\!\tau_{\text{c}} (P_{\text{c}}\!-\!P_0)$ and $\beta\!=\!P_0$ are constants.
+\todo[inline]{Define and explain the parameters if this is not done earlier in other sections.}
 Here, note that $\alpha\mathbb{E}[\mathcal{S}]$ and $\beta\,\mathbb{E}[\mathcal{W}]$ corresponds to the energy wasted per discarded sample and the additional energy expended for waiting, respectively.
-As noted down by the authors, it is important to note that this optimisation approach can also be applied to optimise metrics other than energy, simply by formulating this new metrics in the same format as follows, that is by expressing the metric as a linear combination of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and expected wait time $\mathbb{E}[\mathcal{W}]$.
-Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimises the energy penalty given in \eqref{eq:epsilon_terminal}.
+As noted down by the authors, it is important to note that this optimization approach can also be applied to optimize metrics other than energy, simply by formulating this new metrics in the same format as follows, that is by expressing the metric as a linear combination of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and expected wait time $\mathbb{E}[\mathcal{W}]$.
+Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimizes the energy penalty given in \cref{eq:epsilon_terminal}.
+
 \subsection{Optimum sampling instants}\label{sec:aprxSol}
 
-First, we borrow the idea of checkpointing density from \cite{CP_C,CP_C} and call it instantaneous sampling rate function $r(t)$ which is
+First, we borrow the idea of checkpointing density from \textcite{CP_C} and call it instantaneous sampling rate function $r(t)$ which is
 related to $\{t_n\}$ such that,
@@
 \end{alignat}
-We find $r^*(t)$, the $r(t)$ that minimises $\mathcal{E}$.
+We find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
 Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
@@
 % the expected wait 
-$\mathbb{E}[\mathcal{W}]$, we use the TTE conditional CDF. We get,
+$\mathbb{E}[\mathcal{W}]$, we use the TTE conditional CDF.
+We get,
 \begin{multline*}
@@
 The numerator is degenerate when $t\!<\!0$ or $t\!>\!(t_n\!-\!t_{n-1})$.
-Thus, we are only interested in $0\!\leq\!t\!\leq\!
-(t_n-t_{n-1})$.
+Thus, we are only interested in $0\!\leq\!t\!\leq\! (t_n-t_{n-1})$.
 Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ corresponds to the CDF, CCDF and pdf of the TTE distribution.
@@
 \end{alignat*}
-Here, \ref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator. Expanding the CDF using Taylor series, 
+Here, \ref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
+Expanding the CDF using Taylor series, 
 \begin{multline*}
@@
 
-Now, if we have a constrained optimisation with an upper bound $w_0$ for the expected wait time, we can find the optimum set of sampling instants by equating \eqref{Ew} to $w_0$ and find the ratio of $\frac{\alpha}{\beta}$ that provides the optimum set of sampling instants when plugged into \eqref{tnFinal}.
-For this, we make use of the facts that (1) the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals and (2)one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimisation discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+Now, if we have a constrained optimization with an upper bound $w_0$ for the expected wait time, we can find the optimum set of sampling instants by equating \cref{Ew} to $w_0$ and find the ratio of $\frac{\alpha}{\beta}$ that provides the optimum set of sampling instants when plugged into \cref{tnFinal}.
+For this, we make use of the facts that
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
+    \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+\end{enumerate*}
 \begin{alignat*}{1}

### 2022-11-05 intro.tex
tweaking intro

@@
 
-In this paper, we introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+\medskip
+
+With that context, we can express the contributions of this paper as follows:
+\begin{enumerate}
+    \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
     Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
-Furthermore, we introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    \item\label{item:contrib:frames} We introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    \item Finally, using \cref{item:contrib:model,item:contrib:frames}, we explore the potential for optimization in these systems, and propose a novel strategy for the optimum sampling of \ac{WCA} applications with respect to the number of video frames captured per step.
+\end{enumerate}
 
@@
 We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.
-Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}
+
+Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}, as well as on optimum sampling of edge-bound interactive systems.
 In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
@@
 When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.
-
 We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
-\todo[inline]{What is our goal with this?}
+Finally, in \textcite{ICCperiodic1,TMCperiodic} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
+This is further extended into an aperiodic strategy in \textcite{secAperiodic}.
 

### 2022-11-06 implications_optimization.tex
adding new plots

@@
 \section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
+\todo[inline]{Implications for \ac{WCA} footprint.}
+\todo[inline]{Two subsections (simulation and real system.)}
+\todo[inline]{Leave it for now.}
 % \todo[inline]{Introduction to this section?}
@@
 \end{alignat}
+\todo[inline]{
+Optimization is own section.    
+First thing in optimization.
+Represents a lot of tradeoffs in these systems.
+For two particular metrics, we discuss solutions of this equations for different optimization purposes.
+We compare to other approaches.
+State of the art and offline optimum.
+Optimization business should go into own section.}
 Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy.png}
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
+        \caption{\todo[inline]{Note logarithmic yscale.}}
+    \end{subfigure}\\
+    \medskip
+    \begin{subfigure}[t]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
+        \caption{}
+    \end{subfigure}
+    \caption{Sampling optimization}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \begin{subfigure}[t]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization.png}
         \caption{}
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization_diff.png}
         \caption{}
     \end{subfigure}
+    \caption{Energy optimization}
 \end{figure*}
-
-\todo[inline]{
-    TODO.
-
-    Explain plots.
-}

### 2022-11-06 implications_footprint.tex
fix

--- 2022-11-05 implications_optimization.tex
@@
-\section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
+\section{Implications for the study of \ac{WCA} footprints}\label{sec:implications:footprint}
+\todo[inline]{Two subsections (simulation and real system.)}
+\todo[inline]{Leave it for now.}
 % \todo[inline]{Introduction to this section?}
@@
 % The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \ac{TCP} traffic, further congesting the wireless network.
-
-\subsection{Optimizing for mean number of samples per step}
-
-In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \ac{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
-At every sampling instant \( t \), the scheme adapts the sampling rate \( R(t) \) of the system according to the estimated likelihood of the user having finished the step,
-following the formula
-
-\begin{equation}
-    R(t) = R_\text{min} + \alpha\left( R_\text{max} - R_\text{min} \right) * CDF(t)
-\end{equation}
-
-\( R_\text{max} \) and \( R_\text{min} \) correspond to the maximum and minimum sampling rates of the system, respectively.
-\( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
-\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
-Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
-
-In the following, we will show the effects of more accurate execution time estimation using our timing models on the performance of this adaptive scheme.
-For this, we implement two variants of the adaptive scheme.
-The first one corresponds a reference implementation using \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
-This variant the same \ac{CDF} for every step.
-The second variant instead uses an embedded internal execution time model to keep track of the \acp{TTF} of the system and adjust the distribution of execution times following the state of the timing model.
-In other words, in this variant the \ac{CDF} used for instantaneous calculation of the sampling rate \( R \) changes over time according to the state of the execution time model.
-For timing model configurations which do not fit a distribution to execution time data, the \ac{CDF} corresponds to the \ac{ECDF} of the selected execution times.
-
-We proceed to set up an experiment where these sampling scheme variants are deployed on identical, simulated, \num{100}-step tasks.
-Also included for reference in this experiment is greedy (also known as \emph{zero-wait}) sampling, where samples are taken as fast as possible; a new sample is taken as soon as the result for the previous one has been received.
-This is the default sampling strategy used in most existing \ac{WCA} prototypes.
-The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
-We set \( R_\text{min} \) to \SI{0.5}{\hertz} and \( \alpha \) to \num{3.0}.
-The \ac{RTT} of the system is set to one of \SIlist[list-final-separator={, or}]{150;300;600}{\milli\second}, constant for all steps of the task.
-For each combination of sampling scheme configuration (original adaptive by \textcite{Wang2019Towards}, adaptive with \edgedroid{} model, greedy) and \ac{RTT}, we run \num{100} repetitions of the task for good statistical significance.
-
-\begin{figure*}
-    \centering
-    \begin{subtable}[t]{.25\textwidth}
-        \centering
-        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
-        \medskip
-        \begin{tabular}{lr}
-            \toprule
-            \( R_\text{min} \)  & \SI{0.5}{\hertz} \\
-            \( \alpha \)        & \num{3.0} \\
-            \ac{RTT}            & \SIlist{150;300;600}{\milli\second} \\
-            \bottomrule
-        \end{tabular}
-    \end{subtable}%
-    \hspace{.05\textwidth}%
-    \begin{subtable}[t]{.65\textwidth}
-        \centering
-        \caption{Mean number of captured samples per sampling scheme at the three different \acp{RTT}.}
-        \medskip
-        \begin{tabular}{llrrr}
-            \toprule 
-            Timing model neuro. & Sampling scheme & \SI{150}{\milli\second} & \SI{300}{\milli\second} & \SI{600}{\milli\second} \\
-           \midrule
-           \multirow[c]{2}{*}{Low} & Greedy & \num{36.53} & \num{18.88} & \num{10.52} \\
-            & Adaptive Gaussian fit & \num{19.97} & \num{10.43} & \num{6.39} \\
-           \multirow[c]{2}{*}{High} & Greedy & \num{37.05} & \num{19.04} & \num{10.89} \\
-            & Adaptive Gaussian fit & \num{20.29} & \num{10.79} & \num{6.67} \\
-            \bottomrule
-        \end{tabular}
-    \end{subtable}\\
-    \bigskip
-    \begin{subfigure}[]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue_both.png}
-        \caption{%
-            Percentage reduction in mean number of captured samples per step (higher is better).
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
-            \todo[inline]{%
-                % Vishnu: minimize number of samples.
-                I don't like the two plots, I think the message gets diluted and confusing.
-                I suggest only keeping one ``ground truth'', high neuroticism.
-                Otherwise, every plot and table needs to be duplicated.
-            }
-        }
-    \end{subfigure}
-    \caption{%
-        Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.
-        \emph{ED2} indicates a sampling scheme using the \edgedroid{} model for \ac{CDF} estimation.
-        \emph{HN} and \emph{LN} indicate high or low neuroticism, respectively, and \emph{ExG fit} indicates a timing model which fits an \ac{exGaussian} distribution to the execution time data for each step.
-    }\label{fig:samples}
-\end{figure*}
-
-The results of this investigation are presented in \cref{fig:samples}
-They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{14}{\percent} and \textasciitilde\SI{27}{\percent} better average performance, depending on the \edgedroid{} model parameterization and the system \ac{RTT}.
-\todo[inline]{Update with new ground truths.}
-This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
-
-\subsection{Optimizing for energy consumption}
-
-Energy consumption is another yet unexplored dimension of \ac{WCA} with great opportunities for optimization.
-To explore the potential implications of our realistic model of human timings for energy consumption, we will introduce here a novel scheme for sampling in step-based \ac{WCA} systems.
-
-We begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
-In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
-The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
-
-% In this optimization, the constant parts are excluded as it does not affect the process.
-With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in \textcite{secAperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
-
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
-However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
-Furthermore, instead of directly optimizing for energy, we start by noting that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
-% in a format similar to that of the energy, 
-as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
-Let $\mathcal{E}$ correspond to such an objective metric.
-Thus,
-\begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
-\end{alignat}
-Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
-For instance, in the modeling used by \textcite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
-The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
-
-\subsubsection{General solution}\label{sec:aprxSol}
-
-Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
-\begin{alignat}{1}
-{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
-\end{alignat}
-For instance, for periodic sampling, the rate function is a constant equal to the sampling frequency. 
-% Using this rate function, we rewrite the expected number of samples as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t,\nonumber
-% \end{alignat}
-% and the approximate the expected wait time as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber
-% \end{alignat}
-Now, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
-\begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\nonumber\\
-\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
-r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
-% \Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
-% \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.\label{tn_approx_rayleigh}
-\end{alignat}
-
-\subsubsection{Energy minimization}\label{sec:aprxSol_Energy}
-
-With the general solution available, minimizing energy $E$ boils down to finding the appropriate $\alpha$ and $\beta$ that $\mathcal{E}=E$ in \cref{eq:epsilon_terminal}.
-For this, we directly take the modeling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
-With feedback given even to the discarded samples in our model, we have, 
-\begin{alignat}{2}
-    \mathrm{E}=&\;2\mathcal{S}\tau_cP_c+(\mathcal{T}+\mathcal{W}+\tau_\mathrm{p}+2\tau_\mathrm{c}-2\mathcal{S}\tau_c)P_0\nonumber\\
-    % &=s\tau_c(P_c-P_0^{(t)})+wP_0^{(t)}+(\tau+\tau_c+\tau_p) P_0^{(t)}+\tau_cP_c\\
-    =&\;2\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +2\tau_{\text{c}}) P_0\nonumber\\
-&\Rightarrow \alpha=2\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0.\nonumber
-\end{alignat}
-
-\todo[inline]{is communication delay and processing delay defined already?}
-
-\subsubsection{Minimizing number of samples }\label{sec:aprxSol_Energy_samples}
-
-Apart from energy, another common optimization metric is simply the number of samples.
-However, an unconstrained optimization of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
-We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
-That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
-\begin{alignat}{1}
-\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma},\nonumber
-\end{alignat}
-where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
-
-\subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy.png}
-        \caption{}
-    \end{subfigure}\\
-    \medskip
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \includegraphics[width=.9\textwidth]{figs/new_model/sampling_energy_diff.png}
-        \caption{}
-    \end{subfigure}
-\end{figure*}
-
-\todo[inline]{
-    TODO.
-
-    Explain plots.
-}

### 2022-11-06 implications_optimization.tex
fix

@@
-\section{Implications for the study and optimization of \acs{WCA}}\label{sec:implications}
-\todo[inline]{Implications for \ac{WCA} footprint.}
-\todo[inline]{Two subsections (simulation and real system.)}
-\todo[inline]{Leave it for now.}
-% \todo[inline]{Introduction to this section?}
-
-\subsection{Application lifetimes}
-
-We begin by studying the implications of such a model on the estimation of application lifetimes.
-In the context of \ac{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
-This is an important metric for \ac{WCA} optimization, as it directly relates to system resource utilization and contention, and to energy consumption.
-
-In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This new model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
-This distribution is then randomly sampled at runtime to obtain execution times for each step, without any sort of adjustment to the current state of the system.
-
-We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
-These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \ac{TTF} and record the generated execution time.
-We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
-This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \ac{TTF}.
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_all_ttfs.png}
-        \caption{%
-            Evolution of mean application lifetimes as \acp{TTF} increase.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
-        }
-    \end{subfigure}
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_diff.png}
-        \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
-        }
-    \end{subfigure}
-    \caption{\acl{TTF} versus application lifetime.}\label{fig:lifetimes}
-\end{figure}
-
-The results of this investigation are presented in \cref{fig:lifetimes}.
-Compared to the reference model, our realistic model is, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
-At higher \acp{TTF}, the behavior of the model depends on its level of normalized neuroticism.
-At low neuroticism, the behavior of the realistic model results in total task durations that are basically indistinguishable from the reference model.
-However, at high neuroticism, the model once again results in a considerable difference in total task duration with respect to the reference --- this time extending durations by \textasciitilde\SI{11}{\percent} on average.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
-    \caption{%
-        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \acs*{COTS} \acs{IEEE} \num{802.11}b/g access point.
-    }\label{fig:expsetup}
-\end{figure}
-
-Next we study the effects of first- versus second-order models in a more realistic setting.
-The models, reference and realistic, are deployed on the Raspberry Pi clients of the testbed depicted in \cref{fig:expsetup}.
-For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \ac{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
-Clients and cloudlet communicate over an \acs{IEEE} \num{802.11}b/g wireless network.
-Our choice of wireless standard is simply motivated by a desire to amplify the potential effects of network congestion.
-\todo[inline]{More justification?}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed.png}
-        \caption{%
-            Mean application lifetimes per testbed configuration.
-            Note that due to the low number of samples, means have been calculated using the geometric instead of arithmetic average. 
-            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using bootstrapping.
-        }
-    \end{subfigure}
-    \par\bigskip
-    \begin{subtable}{\columnwidth}
-        \centering
-        \begin{tabular}{lrrrr}
-            \toprule
-            \# clients & 1 & 4 & 7 & 10 \\
-            \ac{RTT} & \SI{0.42}{\second} & \SI{1.12}{\second} & \SI{1.92}{\second} & \SI{2.68}{\second} \\
-            \bottomrule
-        \end{tabular}
-        \caption{Mean measured \aclp{RTT} for each testbed configuration.}
-    \end{subtable}
-    \par\bigskip
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed_diff.png}
-        \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model.
-            Confidence intervals have been omitted due to the low number of samples and the use of the geometric mean.
-        }
-    \end{subfigure}
-    \caption{Application lifetimes in the realistic scenarios.}\label{fig:testbed_lifetimes}
-\end{figure}
-
-We deploy configurations running \num{45}-step versions of the LEGO task described in \textcite{olguinmunoz:impact2021}.
-The testbed configurations include setups with \numlist{1;4;7;10} clients.
-Due to having only limited time, each combination of testbed and timing model configuration is only repeated \num{10} times.
-The results are presented in \cref{fig:testbed_lifetimes}.
-Owing to the low number of samples, we opt here for the geometric rather than arithmetic mean to represent our results.
-The results are nonetheless clear, and follow the same pattern as the previously discussed results under ideal, controlled circumstances.
-With just a single client and mean \ac{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
-At \num{10} clients, the results mimic those at higher \acp{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
-
-These results highlight the importance of accurate execution time modeling when studying \ac{WCA} application lifetimes.
-Not only do we see considerable differences in lifetimes at relatively moderate levels of system congestion, but the sign of these differences depends on the load placed on the system.
-Imagine thus a system designer studying resource consumption optimization in a \ac{WCA}.
-If they were to employ the reference model for their study on an unimpaired system, it could lead them to significantly underestimate the potential for optimization of resource consumption, leaving performance (and, potentially, monetary) gains on the table.
-On the other hand, under heavy load, they would instead underestimate system resource occupation, again leading to performance losses.
-
-\todo[inline]{%
-    Ergo we need a second order model. Everything that comes afterwards is only with second order model. Play with RTT.
-}
-
-% .as well as an additional setup with \num{10} clients during which an instance of the \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} network load generator is also running on each client.
-% The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \ac{TCP} traffic, further congesting the wireless network.
+\section{Implications for the optimization of resource consumption and responsiveness trade-offs in \ac{WCA}}\label{sec:implications:optimization}
 

### 2022-11-06 intro.tex
fix

@@
 We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
-Next in \cref{sec:implications} we discuss the potential implications of such a model by studying a small series of representative scenarios.
+Next in \cref{sec:implications:footprint} we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
+This is followed by a more in-depth investigation in\cref{sec:implications:optimization} on the potential consequences of such a model on the general optimization of \ac{WCA} systems.
 Finally, in \cref{sec:conclusion} we summarize and conclude this paper.

### 2022-11-06 appendices.tex
fixing Vishnus stuff
Update on Overleaf.

@@
-\subsection*{Optimisation}
-The energy penalty can be expressed as
+\appendix{}\label[appendixwithoutnumber]{appx1}
+\subsection*{Sampling optimisation}
+In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in section \ref{sec:aprxSol}.
+We start with the general solution where the objective function to be minimised is given by \eqref{eq:epsilon_terminal}.
 \begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}],\;\label{eq:epsilon_terminal}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
 \end{alignat}
-where \( P_0 \) corresponds to the power consumed by the \ac{WCA} client device during idle periods and \( P_\text{c} \) to the power consumption during communication with the backend (i.e.\ transmitting a sample or receiving a reply).
-$\alpha = \tau_{\text{c}} (P_{\text{c}} - P_0)$ and $\beta = P_0$ are constants.
-Here, note that $\alpha\mathbb{E}[\mathcal{S}]$ and $\beta\,\mathbb{E}[\mathcal{W}]$ correspond to the energy wasted per discarded sample and the additional energy expended while waiting, respectively.
-Using this model, we find the approximate, near-optimal set of sampling instants $\{t_n^*\}$ that minimizes the energy penalty given in \cref{eq:epsilon_terminal}.
-
-\subsubsection{Optimum sampling instants}\label{sec:aprxSol}
-
-First, we borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is
-related to $\{t_n\}$ such that,
+To proceed with the optimisation, we first borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}

### 2022-11-06 sampling.tex
Update on Overleaf.

@@
 \end{alignat*}
+
+Now, we modify the problem (see section \ref{sec:aprxSol_Energy_samples}) and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+First note that, the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
+The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
+As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to thewith all possible real values in between
+
+We do this by equating \cref{Ew} to $w_0$ and find the ratio of $\frac{\alpha}{\beta}$ that provides the optimum set of sampling instants when plugged into \cref{tnFinal}.
+This is possible because, wit
+For this, we make use of the facts that
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
+    \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+\end{enumerate*}
+\begin{alignat*}{1}
+\mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber\\
+&=\int_{0}^{\infty}\frac{1}{2}\sqrt{\frac{2\alpha\sigma^2}{\beta t}}\cdot \frac{t}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\int_{0}^{\infty}\frac{\sqrt{t}}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\int_{0}^{\infty}y^{-1/4}e^{-y}\mathrm{d}y\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4})\\
+% &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+\intertext{Thus, when the upper bound $w_0$ is tight,}
+w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+w_0&=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
+\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\&=1.88355\frac{w_0^2}{\sigma}
+\end{alignat*}

### 2022-11-06 sampling.tex
Update on Overleaf.

@@
 The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
-As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to thewith all possible real values in between
+As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modwith all possible real values in between
 

### 2022-11-06 sampling.tex
Update on Overleaf.

--- 2022-11-05 sampling.tex
@@
 \end{alignat*}
+
+Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+First note that, the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
+The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
+As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
+Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and use the general sooptimum set of sampling instants when plugged into \cref{tnFinal}.
+This is possible because, wit
+For this, we make use of the facts that
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
+    \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+\end{enumerate*}
+\begin{alignat*}{1}
+\mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber\\
+&=\int_{0}^{\infty}\frac{1}{2}\sqrt{\frac{2\alpha\sigma^2}{\beta t}}\cdot \frac{t}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\int_{0}^{\infty}\frac{\sqrt{t}}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\int_{0}^{\infty}y^{-1/4}e^{-y}\mathrm{d}y\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4})\\
+% &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+\intertext{Thus, when the upper bound $w_0$ is tight,}
+w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+w_0&=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
+\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\&=1.88355\frac{w_0^2}{\sigma}
+\end{alignat*}

### 2022-11-06 sampling.tex
Update on Overleaf.

@@
 As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
-Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and use the general sooptimum set of sampling instants when plugged into \cref{tnFinal}.
+Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and use the general solution to sooptimum set of sampling instants when plugged into \cref{tnFinal}.
 This is possible because, wit

### 2022-11-06 sampling.tex
Update on Overleaf.

--- 2022-11-06 sampling.tex
@@
 As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
-Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and use the general sooptimum set of sampling instants when plugged into \cref{tnFinal}.
+Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
 This is possible because, wit

### 2022-11-06 sampling.tex
Update on Overleaf.

--- 2022-11-06 sampling.tex
@@
 The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
-As any positive real value is valid for this ratio, we can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
-Thus, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and use the general sooptimum set of sampling instants when plugged into \cref{tnFinal}.
-This is possible because, wit
-For this, we make use of the facts that
+Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
 \begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
@@
 \end{enumerate*}
+Thus, can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
+Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
+
 \begin{alignat*}{1}

### 2022-11-06 appendices.tex
Update on Overleaf.

@@
 \end{alignat}
-We find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
+Note that, for periodic sampling, this function is a constant, which is equal to the sampling frequency. In the aperiodic case, we find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
 Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
@@
 \end{alignat}
+
+Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+First note that the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
+The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
+Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
+    \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
+\end{enumerate*}
+Thus, can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
+Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
+
+\begin{alignat*}{1}
+\mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber\\
+&=\int_{0}^{\infty}\frac{1}{2}\sqrt{\frac{2\alpha\sigma^2}{\beta t}}\cdot \frac{t}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\int_{0}^{\infty}\frac{\sqrt{t}}{\sigma^2}e^{-{t^2}/{2\sigma^2}}\mathrm{d}t\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\int_{0}^{\infty}y^{-1/4}e^{-y}\mathrm{d}y\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4})\\
+% &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+\intertext{Thus, when the upper bound $w_0$ is tight,}
+w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
+w_0&=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
+\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\&=1.88355\frac{w_0^2}{\sigma}
+\end{alignat*}

### 2022-11-06 sampling.tex
Update on Overleaf.

@@
 
-Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+By making use of this general solution, we can also prove the results given in section \ref{sec:aprxSol_Energy_samples}, Now, for proving the results we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
 First note that, the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.

### 2022-11-06 sampling.tex
Update on Overleaf.

--- 2022-11-06 sampling.tex
@@
 
-Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+By making use of this general solution, we can also prove the results given in section \ref{sec:aprxSol_Energy_samples}, where Now, for proving the results we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
 First note that, the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.

### 2022-11-06 appendices.tex
Update on Overleaf.

@@
 
-Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+By making use of this general solution, we can also prove the results given in section \ref{sec:aprxSol_Energy_samples}, where we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound $w_0$ for the expected wait time.
 First note that the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
@@
 Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
-    \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
+\begin{enumerate}
+    \item the bound will be tight at the optimum
     \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
-\end{enumerate*}
+\end{enumerate}
 Thus, can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.

### 2022-11-06 sampling.tex
Update on Overleaf.

--- 2022-11-06 sampling.tex
@@
 
-Now, we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound for the expected wait time.
+By making use of this general solution, we can also prove the results given in section \ref{sec:aprxSol_Energy_samples}, where we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound $w_0$ for the expected wait time.
 First note that, the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
@@
 Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{enumerate*}
     \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals

### 2022-11-06 appendices.tex
Update on Overleaf.

@@
 The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
-Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
-\begin{enumerate}
-    \item the bound will be tight at the optimum
-    \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
-\end{enumerate}
-Thus, can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.
-Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$ and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
+Since any positive real value is valid for this ratio, one can achieve any valid point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via simply by varying the ratio $\frac{\alpha}{\beta}$.
+Furthermore, since the sampling instants are aperiodic and can take any positive real values, the bound will be tight at the optimum.
+Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
 

### 2022-11-06 appendices.tex
Update on Overleaf.

@@
 \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.
+\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tnRayleigh}.
 \end{alignat}
@@
 Furthermore, since the sampling instants are aperiodic and can take any positive real values, the bound will be tight at the optimum.
-Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnFinal}.
-
+Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnRayleigh}.
 \begin{alignat*}{1}
@@
 &=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\int_{0}^{\infty}y^{-1/4}e^{-y}\mathrm{d}y\\
-&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4})\\
+&=\sqrt{\frac{\alpha\sigma^2}{2\beta}}\left({\frac{1}{{2}\sigma^2}}\right)^{1/4}\cdot\mathlarger{\Gamma}(\tfrac{3}{4}),\\
 % &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
-\intertext{Thus, when the upper bound $w_0$ is tight,}
+\intertext{where $\mathlarger{\Gamma(x)}$ is the gamma function. Thus, when the upper bound $w_0$ is tight,}
 w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\

### 2022-11-06 implications_optimization.tex
reworking optimizarion

@@
 \section{Implications for the optimization of resource consumption and responsiveness trade-offs in \ac{WCA}}\label{sec:implications:optimization}
+
+As with any application intended for mass deployment on multi-tenant systems such as the edge, \ac{WCA} applications will have to be configured and designed to minimize resource consumption at runtime.
+At the same time, imposing too stringent requirements on resource consumption can adversely affect application responsiveness, which in turn has the potential to drastically reduce quality of experience for users.
+Optimizing the deployment of these applications thus intrinsically involves achieving a balance between the reduction of resource consumption while maintaining an acceptable level of system responsiveness.
+
+In the following section we will study the optimization of resource consumption-responsiveness trade-offs in two dimensions of \ac{WCA}, number of samples captured per step and energy consumption per step.
+
+\todo[inline]{Finish}
+
+\subsection{Optimization framework}
+
+Again, our optimization approaches focus on individual steps.
+Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
+In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
+The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
+
+% In this optimization, the constant parts are excluded as it does not affect the process.
+With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
+Next, in \textcite{secAperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+
+In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
+However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
+Furthermore, instead of directly optimizing for energy, we start by noting that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
+% in a format similar to that of the energy, 
+as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
+Let $\mathcal{E}$ correspond to such an objective metric.
+Thus,
+\begin{alignat}{1}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
+\end{alignat}
+\todo[inline]{
+Optimization is own section.    
+First thing in optimization.
+Represents a lot of tradeoffs in these systems.
+For two particular metrics, we discuss solutions of this equations for different optimization purposes.
+We compare to other approaches.
+State of the art and offline optimum.
+Optimization business should go into own section.}
+Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
+For instance, in the modeling used by \textcite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
+The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
+
+\bigskip
+
+Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+\begin{alignat}{1}
+{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
+\end{alignat}
+For instance, for periodic sampling, the rate function is a constant equal to the sampling frequency. 
+% Using this rate function, we rewrite the expected number of samples as
+% \begin{alignat}{1}
+% \mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t,\nonumber
+% \end{alignat}
+% and the approximate the expected wait time as
+% \begin{alignat}{1}
+% \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber
+% \end{alignat}
+Now, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
+\begin{alignat}{1}
+r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\nonumber\\
+\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
+r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
+% \Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
+% \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
+\Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tn_approx_rayleigh}
+\end{alignat}
 
@@
 
+We first look at the application of \cref{eq:epsilon_terminal} for the optimization of number of captured samples per step.
+We start with this metric as its implications for resource consumption and responsiveness are straightforward to understand.
+On the one hand, higher sampling rates directly lead to perceived increased system responsiveness, as smaller sampling intervals translate into smaller maximum wait times.
+On the other, the relationship between number of samples captured and sent and network congestion is exponential, and too high sampling rates quickly lead to bottlenecks on the network, particularly in multi-tenant environments.
+Additionally, the energy cost of capturing a sample on a \ac{WCA} client device is often much higher than remaining in an idle state, and thus excessive sampling leads to drastically increased energy consumption.
+Optimizing the number of samples captured per step can thus be a straightforward way of reducing resource consumption and contention in \ac{WCA} applications.
+
+However, an unconstrained optimization of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
+We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
+That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
+\begin{alignat}{1}
+\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma},\nonumber
+\end{alignat}
+where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
+
+\medskip
+
+We introduce here a reference scheme to which we will compare our approach.
 In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \ac{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
@@
 
-We begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
-In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
-The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
-
-% In this optimization, the constant parts are excluded as it does not affect the process.
-With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in \textcite{secAperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
-
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
-However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
-Furthermore, instead of directly optimizing for energy, we start by noting that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
-% in a format similar to that of the energy, 
-as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
-Let $\mathcal{E}$ correspond to such an objective metric.
-Thus,
-\begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
-\end{alignat}
-\todo[inline]{
-Optimization is own section.    
-First thing in optimization.
-Represents a lot of tradeoffs in these systems.
-For two particular metrics, we discuss solutions of this equations for different optimization purposes.
-We compare to other approaches.
-State of the art and offline optimum.
-Optimization business should go into own section.}
-Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
-For instance, in the modeling used by \textcite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
-The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
-
-\subsubsection{General solution}\label{sec:aprxSol}
-
-Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
-\begin{alignat}{1}
-{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
-\end{alignat}
-For instance, for periodic sampling, the rate function is a constant equal to the sampling frequency. 
-% Using this rate function, we rewrite the expected number of samples as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t,\nonumber
-% \end{alignat}
-% and the approximate the expected wait time as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber
-% \end{alignat}
-Now, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
-\begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\nonumber\\
-\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
-r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
-% \Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
-% \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}.\label{tn_approx_rayleigh}
-\end{alignat}
-
 \subsubsection{Energy minimization}\label{sec:aprxSol_Energy}
@@
 \subsubsection{Minimizing number of samples }\label{sec:aprxSol_Energy_samples}
-
-Apart from energy, another common optimization metric is simply the number of samples.
-However, an unconstrained optimization of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
-We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
-That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
-\begin{alignat}{1}
-\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma},\nonumber
-\end{alignat}
-where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
 

### 2022-11-06 appendices.tex
working

@@
 \begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C
 \end{alignat}
-To proceed with the optimisation, we first borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+To proceed with the optimization, we first borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
@@
 \intertext{where $\mathlarger{\Gamma(x)}$ is the gamma function. Thus, when the upper bound $w_0$ is tight,}
-w_0&=0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
-w_0&=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
-\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\&=1.88355\frac{w_0^2}{\sigma}
+\mathbb{E}[\mathcal{W}]&=w_0=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\
+\Rightarrow\frac{\alpha}{\beta}&=\frac{w_0^2}{\sigma}\frac{2\sqrt{2}}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2}\\
+% &=1.88355\frac{w_0^2}{\sigma}
+&\approx1.9\frac{w_0^2}{\sigma}
 \end{alignat*}

### 2022-11-06 implications_optimization.tex
working

@@
 That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
-\begin{alignat}{1}
-\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma},\nonumber
+\begin{alignat}{1}\label{eq:optimization:sampling}
+\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma}
 \end{alignat}
@@
 \begin{equation}
-    R(t) = R_\text{min} + \alpha\left( R_\text{max} - R_\text{min} \right) * CDF(t)
+    R(t) = R_\text{min} + \varphi\left( R_\text{max} - R_\text{min} \right) * CDF(t)
 \end{equation}
@@
 \( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
-\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \alpha \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
+\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \varphi \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
 Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
 
-In the following, we will show the effects of more accurate execution time estimation using our timing models on the performance of this adaptive scheme.
-For this, we implement two variants of the adaptive scheme.
-The first one corresponds a reference implementation using \citeauthor{Wang2019Towards}'s original design, using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
-This variant the same \ac{CDF} for every step.
-The second variant instead uses an embedded internal execution time model to keep track of the \acp{TTF} of the system and adjust the distribution of execution times following the state of the timing model.
-In other words, in this variant the \ac{CDF} used for instantaneous calculation of the sampling rate \( R \) changes over time according to the state of the execution time model.
-For timing model configurations which do not fit a distribution to execution time data, the \ac{CDF} corresponds to the \ac{ECDF} of the selected execution times.
+\medskip
 
-We proceed to set up an experiment where these sampling scheme variants are deployed on identical, simulated, \num{100}-step tasks.
-Also included for reference in this experiment is greedy (also known as \emph{zero-wait}) sampling, where samples are taken as fast as possible; a new sample is taken as soon as the result for the previous one has been received.
-This is the default sampling strategy used in most existing \ac{WCA} prototypes.
-The execution times for each step are generated by a realistic model parameterized with high neuroticism and without any distribution fitting on the data (this timing model is separate from the internal timing models inside the sampling schemes).
+In the following, we will show the effects of our optimization approach combined with our timing models compared to \textcite{Wang2019Towards}'s state-of-the-art approach.
+\todo[inline]{Sounds clunky}
+For this we implement these sampling schemes in Python and run a number of simulations with them.
+The first scheme uses our approach and \cref{eq:epsilon_terminal,eq:optimization:sampling} to determine the optimum sampling instants and the values for \( \alpha \) and \( \beta \) at each step.
+It includes an embedded timing model to provide updated estimates of \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
+
+We implement \citeauthor{Wang2019Towards}'s original design using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
+This scheme does not include an embedded timing model, and uses the same \ac{CDF} for every step.
+
+Finally, we also implement two reference sampling schemes representing best- and worst-case extremes.
+The first of these corresponds to the offline optimum which uses an embedded \emph{oracle} to perfectly predict the execution time of each step.
+Such an ideal scheme is thus able to always sample exactly once per step, with a constant wait time of zero.
+The second reference scheme corresponds to one which \emph{greedily} samples as much as possible.
+This represents a completely unoptimized design with no considerations for resource-consumption trade-offs; it simply attempts to maximize the number of captured samples per step.
+This is an interesting approach to include given that it corresponds to the default sampling strategy used in most existing \ac{WCA} prototypes.
+
+\begin{table*}
+    \centering
+    \caption{Experimental parameters}\label{tab:params}
+    \begin{tabular}{lrl}
+        \toprule
+        Parameter & Value & Clarification \\
+        \midrule
+        Number of task steps & \num{100} & \\
+        Repetitions & \num{100} & \\
+        \acp{RTT} & \SIlist{0.3375;0.675;1.25;2.5;5.0}{\second} & \\
+        \( t_\text{p} \) & \SI{300}{\milli\second} & Processing delay \\
+        \( t_\text{c} \) & \( \text{\ac{RTT}} - t_\text{p} \) & Communication delay \\
+        \( w_0 \) & \SI{1.0}{\second} & \\
+        \( R_\text{min} \) & \SI{0.5}{\hertz} & Derived as \( R_\text{min} = {(2 w_0)}^{-1} \) \\
+        \( \varphi \) & 3.0 & Scaling factor, \textcite{Wang2019Towards}\\
+        \( P_0 \) & \SI{15}{\milli\watt} & Idle power \\
+        \( P_\text{c} \) & \SI{45}{\milli\watt} & Communication power \\
+        \bottomrule
+    \end{tabular}
+\end{table*}
+
+We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks.
+Experimental parameters are detailed in \cref{tab:params}.
+The execution times for each step are generated by a realistic model without any distribution fitting on the data.
+
+ (this timing model is separate from the internal timing models inside the sampling schemes).
 We set \( R_\text{min} \) to \SI{0.5}{\hertz} and \( \alpha \) to \num{3.0}.

### 2022-11-06 implications_optimization.tex
working on sampling

@@
 The first scheme uses our approach and \cref{eq:epsilon_terminal,eq:optimization:sampling} to determine the optimum sampling instants and the values for \( \alpha \) and \( \beta \) at each step.
-It includes an embedded timing model to provide updated estimates of \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
+It includes an embedded timing model (without any distribution fitting) to provide updated estimates of \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
+We will refer to this scheme as the \emph{sample-count-optimized aperiodic} sampling scheme.
 
@@
 
-\begin{table*}
+\begin{table}
     \centering
@@
         \midrule
-        Number of task steps & \num{100} & \\
+        \# of steps & \num{100} & \\
         Repetitions & \num{100} & \\
-        \acp{RTT} & \SIlist{0.3375;0.675;1.25;2.5;5.0}{\second} & \\
+        \acp{RTT} & \makecell[cr]{%
+            \numlist[list-final-separator={, }]{0.3375;0.675;1.25},\\
+            \SIlist{2.5;5.0}{\second}
+        } & \\
         \( t_\text{p} \) & \SI{300}{\milli\second} & Processing delay \\
@@
     \end{tabular}
-\end{table*}
+\end{table}
 
@@
 Experimental parameters are detailed in \cref{tab:params}.
-The execution times for each step are generated by a realistic model without any distribution fitting on the data.
-
- (this timing model is separate from the internal timing models inside the sampling schemes).
-We set \( R_\text{min} \) to \SI{0.5}{\hertz} and \( \alpha \) to \num{3.0}.
-The \ac{RTT} of the system is set to one of \SIlist[list-final-separator={, or}]{150;300;600}{\milli\second}, constant for all steps of the task.
-For each combination of sampling scheme configuration (original adaptive by \textcite{Wang2019Towards}, adaptive with \edgedroid{} model, greedy) and \ac{RTT}, we run \num{100} repetitions of the task for good statistical significance.
+The execution times for each step are generated by a timing model without any distribution fitting on the data.
+For each combination of sampling scheme configuration, \ac{RTT}, and execution time model neuroticism (low or high), we run \num{100} repetitions of the task for good statistical significance.
+Note that the embedded timing model in our sampling-count-optimized aperiodic sampling scheme is always parameterized with a neuroticism matching the neuroticism of the external execution time model.
 
@@
     \centering
-    \begin{subtable}[t]{.25\textwidth}
+    \begin{subfigure}[t]{\textwidth}
         \centering
-        \caption{Parameters used for \textcite{Wang2019Towards}'s adaptive sampling scheme.}
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
+        \caption{\todo[inline]{Note logarithmic yscale.}}
+    \end{subfigure}\\
     \medskip
-        \begin{tabular}{lr}
-            \toprule
-            \( R_\text{min} \)  & \SI{0.5}{\hertz} \\
-            \( \alpha \)        & \num{3.0} \\
-            \ac{RTT}            & \SIlist{150;300;600}{\milli\second} \\
-            \bottomrule
-        \end{tabular}
-    \end{subtable}%
-    \hspace{.05\textwidth}%
-    \begin{subtable}[t]{.65\textwidth}
+    \begin{subfigure}[t]{\textwidth}
         \centering
-        \caption{Mean number of captured samples per sampling scheme at the three different \acp{RTT}.}
-        \medskip
-        \begin{tabular}{llrrr}
-            \toprule 
-            Timing model neuro. & Sampling scheme & \SI{150}{\milli\second} & \SI{300}{\milli\second} & \SI{600}{\milli\second} \\
-           \midrule
-           \multirow[c]{2}{*}{Low} & Greedy & \num{36.53} & \num{18.88} & \num{10.52} \\
-            & Adaptive Gaussian fit & \num{19.97} & \num{10.43} & \num{6.39} \\
-           \multirow[c]{2}{*}{High} & Greedy & \num{37.05} & \num{19.04} & \num{10.89} \\
-            & Adaptive Gaussian fit & \num{20.29} & \num{10.79} & \num{6.67} \\
-            \bottomrule
-        \end{tabular}
-    \end{subtable}\\
-    \bigskip
-    \begin{subfigure}[]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/diff_sampling_junjue_both.png}
-        \caption{%
-            Percentage reduction in mean number of captured samples per step (higher is better).
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
-            \todo[inline]{%
-                % Vishnu: minimize number of samples.
-                I don't like the two plots, I think the message gets diluted and confusing.
-                I suggest only keeping one ``ground truth'', high neuroticism.
-                Otherwise, every plot and table needs to be duplicated.
-            }
-        }
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
+        \caption{}
     \end{subfigure}
-    \caption{%
-        Results of parameterizing \textcite{Wang2019Towards}'s adaptive sampling scheme using the realistic timing model.
-        \emph{ED2} indicates a sampling scheme using the \edgedroid{} model for \ac{CDF} estimation.
-        \emph{HN} and \emph{LN} indicate high or low neuroticism, respectively, and \emph{ExG fit} indicates a timing model which fits an \ac{exGaussian} distribution to the execution time data for each step.
-    }\label{fig:samples}
+    \caption{Sampling optimization}\label{fig:optimization:samples}
 \end{figure*}
 
-The results of this investigation are presented in \cref{fig:samples}
+The results of this investigation are presented in \cref{fig:optimization:samples}
 They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
@@
 
-\subsubsection{Minimizing number of samples }\label{sec:aprxSol_Energy_samples}
-
-\subsubsection{Evaluating the impact of the realistic timing model on the energy-optimal sampling scheme}
-
-\begin{figure*}
-    \centering
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
-        \caption{\todo[inline]{Note logarithmic yscale.}}
-    \end{subfigure}\\
-    \medskip
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
-        \caption{}
-    \end{subfigure}
-    \caption{Sampling optimization}
-\end{figure*}
-
 \begin{figure*}

### 2022-11-06 implications_optimization.tex
working on sampling still

@@
 We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
-That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfies the condition
+That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfy the condition
 \begin{alignat}{1}\label{eq:optimization:sampling}
@@
         \( P_\text{c} \) & \SI{45}{\milli\watt} & Communication power \\
+        \( \alpha_\text{samples} \) & \( 1.9 \sigma^{-1} \) & For sample-count optimization. \\
+        \( \beta_\text{samples} \) & \num{1.0} & For sample-count optimization. \\
+        \( \alpha_\text{energy} \) & \( t_\text{c}(P_\text{c} - P_0) \) & For energy optimization. \\
+        \( \beta_\text{energy} \) & \( P_0 \) & For energy optimization. \\
         \bottomrule
@@
 
-We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks.
+We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks with varying constant \acp{RTT}.
 Experimental parameters are detailed in \cref{tab:params}.
+We set the maximum expected wait time \( w_0 \) and \( \beta \) factors of our sample-count-optimized scheme to \SI{1.0}{\second} and \num{1.0}, respectively, for mathematical simplicity, and derive \( \alpha = 1.9 \sigma^{-1} \).
+Next, we derive \( R_\text{min} \) for \textcite{Wang2019Towards}'s scheme from \( w_0 \).
+As discussed in \cref{sec:background}, we assume that wait times are uniformly distributed between \SI{0}{\second} and sampling interval of each step.
+A maximum expected wait time \( w_0 = 1.0\,\si{\second} \) thus translates into a maximum expected sampling interval of \SI{2.0}{\second}, yielding a minimum sampling rate \( R_\text{min} = {2.0\,\si{\second}}^{-1} = 0.5\,\si{\hertz} \).
+
 The execution times for each step are generated by a timing model without any distribution fitting on the data.

### 2022-11-06 implications_optimization.tex
energy

@@
 
-\subsection{Optimizing for mean number of samples per step}
+\subsection{Optimizing for mean number of samples per step}\label{ssec:optimization:samples}
 
@@
 In the following, we will show the effects of our optimization approach combined with our timing models compared to \textcite{Wang2019Towards}'s state-of-the-art approach.
-\todo[inline]{Sounds clunky}
 For this we implement these sampling schemes in Python and run a number of simulations with them.
@@
         } & \\
-        \( t_\text{p} \) & \SI{300}{\milli\second} & Processing delay \\
-        \( t_\text{c} \) & \( \text{\ac{RTT}} - t_\text{p} \) & Communication delay \\
+        \( \tau_\text{p} \) & \SI{300}{\milli\second} & Processing delay \\
+        \( \tau_\text{c} \) & \( \text{\ac{RTT}} - \tau_\text{p} \) & Communication delay \\
         \( w_0 \) & \SI{1.0}{\second} & \\
@@
         \( \beta_\text{samples} \) & \num{1.0} & For sample-count optimization. \\
-        \( \alpha_\text{energy} \) & \( t_\text{c}(P_\text{c} - P_0) \) & For energy optimization. \\
+        \( \alpha_\text{energy} \) & \( \tau_\text{c}(P_\text{c} - P_0) \) & For energy optimization. \\
         \( \beta_\text{energy} \) & \( P_0 \) & For energy optimization. \\
@@
 We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks with varying constant \acp{RTT}.
-Experimental parameters are detailed in \cref{tab:params}.
-We set the maximum expected wait time \( w_0 \) and \( \beta \) factors of our sample-count-optimized scheme to \SI{1.0}{\second} and \num{1.0}, respectively, for mathematical simplicity, and derive \( \alpha = 1.9 \sigma^{-1} \).
+Experimental parameters are summarized in \cref{tab:params}.
+We set the \( w_0 \) and \( \beta \) factors of our sample-count-optimized scheme to \SI{1.0}{\second} and \num{1.0}, respectively, for mathematical simplicity, and derive \( \alpha = 1.9 \sigma^{-1} \).
 Next, we derive \( R_\text{min} \) for \textcite{Wang2019Towards}'s scheme from \( w_0 \).
@@
 
-The results of this investigation are presented in \cref{fig:optimization:samples}
-They clearly show the potentially dramatic effects of more accurate execution time estimation when optimizing sampling rates.
-When using the realistic model for execution time estimation, the sampling schemes achieve between \SI{14}{\percent} and \textasciitilde\SI{27}{\percent} better average performance, depending on the \edgedroid{} model parameterization and the system \ac{RTT}.
-\todo[inline]{Update with new ground truths.}
-This is true even for configurations with low neuroticism which do not exactly match the parameterization of the external timing model driving the execution times for the steps, showing thus the importance of even less-accurate approximations.
+The results of this investigation are presented in \cref{fig:optimization:samples}.
+\todo[inline]{Discuss results.}
 
@@
 
-Energy consumption is another yet unexplored dimension of \ac{WCA} with great opportunities for optimization.
-To explore the potential implications of our realistic model of human timings for energy consumption, we will introduce here a novel scheme for sampling in step-based \ac{WCA} systems.
+Next we will explore the implications of combining our timing models with \cref{eq:epsilon_terminal} when directly optimizing for energy consumption in \ac{WCA}.
+Although, as mentioned above, minimizing the number of samples captured during a step can \emph{potentially} translate into a reduction in the energy consumption, this is not a given.
+Energy consumption depends on multiple other factors other than number of samples captured, such as idle versus communication power and delays, and thus cannot be optimized by simply minimizing the number of samples taken.
 
-\subsubsection{Energy minimization}\label{sec:aprxSol_Energy}
-
-With the general solution available, minimizing energy $E$ boils down to finding the appropriate $\alpha$ and $\beta$ that $\mathcal{E}=E$ in \cref{eq:epsilon_terminal}.
-For this, we directly take the modeling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
+In the following we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
+We directly take the modeling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
 With feedback given even to the discarded samples in our model, we have, 
@@
     =&\;2\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +2\tau_{\text{c}}) P_0\nonumber\\
-&\Rightarrow \alpha=2\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0.\nonumber
+&\Rightarrow \alpha=2\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0 \label{eq:optimization:energy}
 \end{alignat}
+\( \tau_\text{p} \) and \( \tau_\text{c} \) correspond to the processing and communication delay for each sample.
+\( P_\text{c} \) and \( P_0 \) correspond to the communication and idle power, respectively, of the \ac{WCA} client device.
 
-\todo[inline]{is communication delay and processing delay defined already?}
+We proceed to repeat the experiment detailed in \cref{ssec:optimization:samples}, replacing the sample-count-optimized sampling scheme with a new implementation instead minimizing energy, using \cref{eq:optimization:energy} for the calculation of \( \alpha \) and \( \beta \).
+We refer to this scheme as the \emph{energy-optimized aperiodic} sampling scheme.
+For the power constants, we reuse the values estimated by the authors in \textcite{TMCperiodic}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
+On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 300\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
 
@@
     \end{subfigure}
-    \caption{Energy optimization}
+    \caption{Energy optimization}\label{fig:optimization:energy}
 \end{figure*}
+
+Again, we include the reference greedy and offline optimum schemes, as well as \textcite{Wang2019Towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \ac{RTT}.
+The results of these experiments are presented in \cref{fig:optimization:energy}.
+
+\todo[inline]{Discussss}

### 2022-11-06 implications_optimization.tex
fix plot
new plots, parameters

@@
 
-Again, our optimization approaches focus on individual steps.
+Again, our optimization approach focuses on individual steps.
 Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
@@
 
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum aperiodic sampling interval.
+In this work, we use the modeling from~\cite{secAperiodic} to find the optimum \emph{aperiodic} sampling interval.
 However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
@@
 \( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
-\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually, and \( \varphi \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
+\( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually.
+\( \varphi \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
 Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
@@
 The first scheme uses our approach and \cref{eq:epsilon_terminal,eq:optimization:sampling} to determine the optimum sampling instants and the values for \( \alpha \) and \( \beta \) at each step.
-It includes an embedded timing model (without any distribution fitting) to provide updated estimates of \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
+It includes an embedded timing model (without any distribution fitting) to provide updated estimates of the mean execution time \( \mu \) and \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
 We will refer to this scheme as the \emph{sample-count-optimized aperiodic} sampling scheme.
@@
         Repetitions & \num{100} & \\
-        \acp{RTT} & \makecell[cr]{%
-            \numlist[list-final-separator={, }]{0.3375;0.675;1.25},\\
-            \SIlist{2.5;5.0}{\second}
-        } & \\
-        \( \tau_\text{p} \) & \SI{300}{\milli\second} & Processing delay \\
+        \acp{RTT} & \( \left\{ 0.3, 0.6,\ldots,4.2 \right\} \) & \\
+        \( \tau_\text{p} \) & \SI{250}{\milli\second} & Processing delay \\
         \( \tau_\text{c} \) & \( \text{\ac{RTT}} - \tau_\text{p} \) & Communication delay \\
@@
         \( R_\text{min} \) & \SI{0.5}{\hertz} & Derived as \( R_\text{min} = {(2 w_0)}^{-1} \) \\
-        \( \varphi \) & 3.0 & Scaling factor, \textcite{Wang2019Towards}\\
+        \( \varphi \) & \num{1.5} & Scaling factor, \textcite{Wang2019Towards}\\
         \( P_0 \) & \SI{15}{\milli\watt} & Idle power \\
@@
         \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
-        \caption{\todo[inline]{Note logarithmic yscale.}}
+        \caption{%
+            Round-trip time versus mean number of captured samples per step, averaged over all \num{100} repetitions of the experiment.
+            Note the logarithmic scale on the vertical axis.
+            Error bars indicate \SI{95}{\percent} \acp{CI}.
+        }
     \end{subfigure}\\
@@
         \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
-        \caption{}
+        \caption{Number of samples captured per step }
     \end{subfigure}
     \caption{Sampling optimization}\label{fig:optimization:samples}
+    \todo[inline]{Second subplot seems... redundant.}
 \end{figure*}
@@
 We proceed to repeat the experiment detailed in \cref{ssec:optimization:samples}, replacing the sample-count-optimized sampling scheme with a new implementation instead minimizing energy, using \cref{eq:optimization:energy} for the calculation of \( \alpha \) and \( \beta \).
+Once again, we embed a timing model into the sampling scheme to provide updated estimates of the mean execution time \( \mu \) at each step.
 We refer to this scheme as the \emph{energy-optimized aperiodic} sampling scheme.
 For the power constants, we reuse the values estimated by the authors in \textcite{TMCperiodic}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
-On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 300\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
+On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
 

### 2022-11-07 appendices.tex
fixed appendix

@@
 \appendix{}\label[appendixwithoutnumber]{appx1}
-\subsection*{Sampling optimisation}
-In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in section \ref{sec:aprxSol}.
-We start with the general solution where the objective function to be minimised is given by \eqref{eq:epsilon_terminal}.
+% \subsection*{Sampling optimisation}
+
+In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in section \cref{ssec:optframework}.
+We start with the general solution, where the objective function to be minimized is given by \cref{eq:epsilon_terminal}:
 \begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C\nonumber
 \end{alignat}
-To proceed with the optimization, we first borrow the idea of checkpointing density from \textcite{CP_C,CP_C} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+
+We first borrow the idea of checkpointing density from \textcite{Satoshi1992Optimal} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
@@
 \end{alignat}
-Note that, for periodic sampling, this function is a constant, which is equal to the sampling frequency. In the aperiodic case, we find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
-Note that, by construction, the number of samples taken up to any time instant can be computed directly by finding the area under $r(t)$.
-Thus, we get the expected number of samples $\mathbb{E}[\mathcal{S}]$ as
+Note that, for periodic sampling, this function is a constant, equal to the sampling frequency.
+In the aperiodic case, we find $r^*(t)$, the $r(t)$ that minimizes $\mathcal{E}$.
+By construction, the number of samples taken up to any time instant can be computed directly by computing the area under $r(t)$.
+Thus, we obtain the expected number of samples $\mathbb{E}[\mathcal{S}]$ as
 \begin{alignat}{1}
@@
 \end{alignat}
+
 To find 
@@
 $\mathbb{E}[\mathcal{W}]$, we use the conditional \ac{CDF} of the execution time.
-We get,
 \begin{multline*}
@@
 Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ correspond to the \ac{CDF}, \ac{CCDF} and \ac{PDF} of the execution time distribution.
-\begin{alignat*}{1}
-\!\!\!\Rightarrow\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&=\dfrac{\mathbb{P}(t_n-t\leq\mathcal{T}\leq t_n)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\\
-&\approx\dfrac{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n}-t)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\tag{$A_1$}\label{Apx1}
-\end{alignat*}
+\begin{alignat}{1}
+\!\!\!\Rightarrow\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&=\dfrac{\mathbb{P}(t_n-t\leq\mathcal{T}\leq t_n)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\nonumber\\
+&\approx\dfrac{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n}-t)}{F_\mathcal{T}(t_n)-F_\mathcal{T}(t_{n-1})}\label{Apx1}
+\end{alignat}
 Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
-Expanding the \ac{CDF} using Taylor series, 
+Expanding the \ac{CDF} using Taylor series gives
 \begin{multline*}
@@
 \end{multline*}
-Simplifying and using the first order approximation, we get
-\begin{alignat*}{1}
+Simplifying and using the first order approximation, we arrive at
+\begin{alignat}{1}
 % &=\dfrac{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(-t)+f'_\mathcal{T}(t_n)(-t)^2/2!+\dots\big)}{F_\mathcal{T}(t_n)-\big(F_\mathcal{T}(t_n)+f_\mathcal{T}(t_n)(t_{n-1}-t_n)+f'_\mathcal{T}(t_n)(t_{n-1}-t_n)^2/2!+\dots\big)}\\&
-\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&\approx\dfrac{tf_\mathcal{T}(t_n)}{(t_{n}-t_{n-1})f_\mathcal{T}(t_n)}\tag{$A_2$}\label{Apx2}\\
-\Rightarrow\mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&= 1-\dfrac{t}{(t_{n}-t_{n-1})}.
-\end{alignat*}
-Using this \ac{CCDF}, we find the conditional expectation of $\mathcal{W}$.
+\mathbb{P}(\mathcal{W}\leq t\,\big\vert\,t_{n-1}<\mathcal{T}\leq t_n)&\approx\dfrac{tf_\mathcal{T}(t_n)}{(t_{n}-t_{n-1})f_\mathcal{T}(t_n)}\label{Apx2}\\
+\Rightarrow\mathbb{P}(\mathcal{W}> t\,|\,t_{n-1}<\mathcal{T}\leq t_n)&= 1-\dfrac{t}{(t_{n}-t_{n-1})}\nonumber
+\end{alignat}
+
+Next, using the above \ac{CCDF}, we find the conditional expectation of $\mathcal{W}$.
 \begin{alignat*}{1}
@@
 That is,
-If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $t_n\!-\!t_{n-1}$ as,
+If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $(t_n\!-\!t_{n-1})$ as
 \begin{alignat}{1}
- (t_n-t_{n-1})&\approx\dfrac{1}{r(t)},\;\forall t,n:t_{n-1}\!<\!t\!\leq\!t_n.\tag{$A_3$}\label{Apx3}\\
+ (t_n-t_{n-1})&\approx\dfrac{1}{r(t)},\;\forall t,n:t_{n-1}\!<\!t\!\leq\!t_n.\label{Apx3}\\
 % \end{alignat}
@@
 \mathcal{E}&=\int_{0}^{\infty}\Big(\alpha\int_{x=0}^{t}r(x)\,\mathrm{d}x+\beta\dfrac{1}{2r(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.\label{epsilon_eulerForm}\\
-\intertext{Let $g(t)=\int_{0}^{t}r(x)\dif x$. Then  $g'(t)=\tfrac{\mathrm{d}}{\mathrm{d}t}g(t)=r(t)$. That is,}
-\vspace{-4cm}\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha g(t)+\dfrac{\beta}{2g'(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t.
+\intertext{%
+    Let $g(t)=\int_{0}^{t}r(x)\dif x$.
+    Then  $g'(t)=\tfrac{\mathrm{d}}{\mathrm{d}t}g(t)=r(t)$.
+    That is,
+}
+\mathcal{E}&=\int_{0}^{\infty}\Big(\alpha g(t)+\dfrac{\beta}{2g'(t)}\Big)f_{\mathcal{T}}(t)\,\mathrm{d}t
 \end{alignat*}
-As per the Euler-Lagrange equation from the calculus of variations~\cite{calcVariation2,calcVariation1}, the extreme value of $\mathcal{E}$ is obtained at 
+
+As per the Euler-Lagrange equation from the calculus of variations~\cite{Bellman1954Dynamic,Arfken2015Calculus}, the extreme value of $\mathcal{E}$ is obtained at 
 \begin{alignat}{1}
@@
 
-By making use of this general solution, we can also prove the results given in section \ref{sec:aprxSol_Energy_samples}, where we modify the problem and find the optimum sampling instants that minimises the expected number of samples for a given upper bound $w_0$ for the expected wait time.
-First note that the general solution has constants $\alpha$ and $\beta$ that corresponds to the weights given to the cost of sampling and cost of waiting, respectively.
-The optimisation criteria changes from minimising wait time to minimising the number of samples, when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
+By making use of this general solution, we can also prove the results given in section \cref{ssec:optimization:samples}, where we modify the problem and find the optimum sampling instants that minimize the expected number of samples for a given upper bound $w_0$ for the expected wait time.
+First note that the general solution has constants $\alpha$ and $\beta$ that correspond to the weights given to the cost of sampling and cost of waiting, respectively.
+The optimization criteria changes from minimizing wait time to minimizing the number of samples when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.
 Since any positive real value is valid for this ratio, one can achieve any valid point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via simply by varying the ratio $\frac{\alpha}{\beta}$.
 Furthermore, since the sampling instants are aperiodic and can take any positive real values, the bound will be tight at the optimum.
-Hence, to solve for the modified optimisation problem explained in section \ref{sec:aprxSol_Energy_samples} that minimises $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $(\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnRayleigh}.
+Hence, to solve for the modified optimization problem explained in section \cref{ssec:optimization:samples} that minimizes $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnRayleigh}.
 \begin{alignat*}{1}
@@
 % &\approx 0.728637\sqrt{\frac{\alpha\sigma}{\beta}}\\
-\intertext{where $\mathlarger{\Gamma(x)}$ is the gamma function. Thus, when the upper bound $w_0$ is tight,}
+\intertext{%
+    where $\mathlarger{\Gamma(x)}$ is the gamma function.
+    Thus, when the upper bound $w_0$ is tight,
+}
 \mathbb{E}[\mathcal{W}]&=w_0=\sqrt{\frac{\alpha\sigma}{2\sqrt{2}\beta}}\mathlarger{\Gamma}(\tfrac{3}{4})\\

### 2022-11-07 implications_optimization.tex
fixed appendix

@@
 
-\subsection{Optimization framework}
+\subsection{Optimization framework}\label{ssec:optframework}
 
@@
         \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
-        \caption{Number of samples captured per step }
+        \caption{%
+            Percentage difference in mean number of captured samples per step with respect to the greedy sampling scheme.
+            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
+        }
     \end{subfigure}
-    \caption{Sampling optimization}\label{fig:optimization:samples}
-    \todo[inline]{Second subplot seems... redundant.}
+    \caption{%
+        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \ac{CDF}-based approach.
+    }\label{fig:optimization:samples}
 \end{figure*}
@@
 
-\subsection{Optimizing for energy consumption}
+\subsection{Optimizing for energy consumption}\label{ssec:optimization:energy}
 

### 2022-11-07 background.tex
tweaking

--- 2022-11-03 background.tex
@@
 
-Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{greedy} sampling.
-In this scheme, a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
-This means that the interval between these sampling instants is not necessarily constant, as it is subject to fluctuations due to resource contention on both the network and compute side.
-It also has as a consequence that acknowledgements must be provided even for those samples which do not cause the generation of a new instruction --- these acknowledgements are simply used for flow control and do not generate user-noticeable feedback.
+As expressed above, \acp{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
+\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.\ which captures the final state of step \( M \)) is taken.
+\( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
+We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \) (once again, refer to \cref{fig:step} for a visual breakdown of these values).
+The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\ac{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \ac{WCA}.
 
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling intervals (following a greedy scheme) at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
-\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.
-which captures the final state of step \( M \)) is taken.
-\( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
-If we define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), it must follow that (see also~\cite{olguinmunoz:impact2021})\footnote{\( U(a, b) \) represents the continuous uniform distribution in the open interval \( (a, b) \).}
-% \todo[inline]{I think it's an assumption rather than a must. Specifically, it is not uniform if the general execution time distribution is Rayleigh/ExpGaussian -vishnu}
-\begin{align}\label{eq:tc}
-    t_c &\thicksim U(t_{n - 1}, t_n)
-\end{align}
+% It should be noted 
+% it must follow that (see also~\cite{olguinmunoz:impact2021})\footnote{\( U(a, b) \) represents the continuous uniform distribution in the open interval \( (a, b) \).}
+% % \todo[inline]{I think it's an assumption rather than a must. Specifically, it is not uniform if the general execution time distribution is Rayleigh/ExpGaussian -vishnu}
+% \begin{align}\label{eq:tc}
+%     t_c &\thicksim U(t_{n - 1}, t_n)
+% \end{align}
 
-We define the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
-The sum of the latter two values (i.e. the interval \( t_{n + 1} - t_c \)) we call \emph{\ac{TTF}}.
-Once again, refer to \cref{fig:step} for a visual breakdown of these values.
+% Current implementations, such as those in~\cite{Chen2015LEGO,Chen2018application}, employ \emph{greedy} sampling.
+% In this scheme, a new sample is immediately taken as soon as an acknowledgement is received for the previous sample.
+% This means that the interval between these sampling instants is not necessarily constant, as it is subject to fluctuations due to resource contention on both the network and compute side.
+% It also has as a consequence that acknowledgements must be provided even for those samples which do not cause the generation of a new instruction --- these acknowledgements are simply used for flow control and do not generate user-noticeable feedback.
 
@@
 For this work, we used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
-\emph{Delay} corresponded to a temporarily fixed time duration of the processing of each frame was padded to --- i.e.
-if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
+\emph{Delay} corresponded to a temporarily fixed time duration of the processing of each frame was padded to --- i.e.\ if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
 The correlation between this variable and execution time was then studied.
@@
 \begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
-    \item \cref{eq:tc} above
+    \item that \( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality~\cite{olguinmunoz:impact2021}
     \item that for a step subject to delay \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
 \end{enumerate*},
-then it follows that, \emph{on average}, for \emph{all} steps subject to a delay \( D \):
+then it follows that the expected \ac{TTF} for steps subject to a delay \( D \) can be expressed as
 
-\begin{align}\label{eq:ttf}
-    \text{\emph{\ac{TTF}}} = \frac{3}{2}D
+\begin{align}
+    \mathbb{E}\left[\text{\ac{TTF}}\right] &= \mathbb{E}\left[\mathcal{W}\right] + D
+    = \mathbb{E}\left[U(t_{n - 1}, t_n)\right] + D\nonumber\\
+    &= \mathbb{E}\left[U(0, D)\right] + D
+    = \frac{1}{2}D + D = \frac{3}{2}D\label{eq:ttf}
 \end{align}

### 2022-11-07 implications_footprint.tex
tweaking

@@
 \section{Implications for the study of \ac{WCA} footprints}\label{sec:implications:footprint}
-\todo[inline]{Two subsections (simulation and real system.)}
-\todo[inline]{Leave it for now.}
-% \todo[inline]{Introduction to this section?}
 
-\subsection{Application lifetimes}
+% \subsection{Application lifetimes}
 
-We begin by studying the implications of such a model on the estimation of application lifetimes.
+We begin by studying the implications of such a model on the estimation of application footprints as described by their lifetimes.
 In the context of \ac{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
@@
 Our choice of wireless standard is simply motivated by a desire to amplify the potential effects of network congestion.
-\todo[inline]{More justification?}
 
@@
 
-\todo[inline]{%
-    Ergo we need a second order model. Everything that comes afterwards is only with second order model. Play with RTT.
-}
+% \todo[inline]{%
+%     Ergo we need a second order model. Everything that comes afterwards is only with second order model. Play with RTT.
+% }
 

### 2022-11-07 implications_optimization.tex
tweaking

@@
 
-\todo[inline]{Finish}
+% \todo[inline]{Finish}
 
@@
 
-Again, our optimization approach focuses on individual steps.
+Our optimization approach focuses on the individual steps which comprise a complete \ac{WCA} task.
 Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
-In \textcite{ICCperiodic1,TMCperiodic}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
+In \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
 The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
@@
 With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in \textcite{secAperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+Next, in \textcite{Moothedath2022Aperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
 
-In this work, we use the modeling from~\cite{secAperiodic} to find the optimum \emph{aperiodic} sampling interval.
+In this work, we use the modeling from~\cite{Moothedath2022Aperiodic} to find the optimum \emph{aperiodic} sampling interval.
 However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
@@
 \end{alignat}
-\todo[inline]{
-Optimization is own section.    
-First thing in optimization.
-Represents a lot of tradeoffs in these systems.
-For two particular metrics, we discuss solutions of this equations for different optimization purposes.
-We compare to other approaches.
-State of the art and offline optimum.
-Optimization business should go into own section.}
-Here, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
-For instance, in the modeling used by \textcite{ICCperiodic1,TMCperiodic,secAperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
+
+Above, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
+For instance, in the modeling used by \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
 The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
@@
 
-Define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+We first define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
-{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
+{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1\label{rt}
 \end{alignat}
@@
 % \end{alignat}
-Now, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
+Next, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
 \begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}.\nonumber\\
-\intertext{Thus, for a Rayleigh distributed $\mathcal{T}$ with parameter $\sigma$,}
+r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}\nonumber\\
+\intertext{in turn resulting in, for a $\mathcal{T}$ distributed according to a Rayleigh distribution with parameter $\sigma$,}
 r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
@@
 In the following we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
-We directly take the modeling from~\cite{secAperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
+We directly take the modeling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
 With feedback given even to the discarded samples in our model, we have, 
@@
 We refer to this scheme as the \emph{energy-optimized aperiodic} sampling scheme.
-For the power constants, we reuse the values estimated by the authors in \textcite{TMCperiodic}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
+For the power constants, we reuse the values estimated by the authors in \textcite{Moothedath2022EnergyEfficient}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
 On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
@@
         \includegraphics[width=\textwidth]{figs/new_model/energy_optimization.png}
-        \caption{}
+        \caption{%
+            Round-trip time versus mean per step energy consumption, averaged over all \num{100} repetitions of the experiment.
+            Error bars indicate \SI{95}{\percent} \acp{CI}.
+        }
     \end{subfigure}\\
@@
         \includegraphics[width=\textwidth]{figs/new_model/energy_optimization_diff.png}
-        \caption{}
+        \caption{%
+            Percentage difference in mean per step energy consumption with respect to the greedy sampling scheme.
+            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
+        }
     \end{subfigure}
-    \caption{Energy optimization}\label{fig:optimization:energy}
+    \caption{%
+        Summary of results for experiment comparing the energy-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \ac{CDF}-based approach.
+    }\label{fig:optimization:energy}
 \end{figure*}

### 2022-11-07 intro.tex
tweaking

@@
     Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
-    \item\label{item:contrib:frames} We introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-    \item Finally, using \cref{item:contrib:model,item:contrib:frames}, we explore the potential for optimization in these systems, and propose a novel strategy for the optimum sampling of \ac{WCA} applications with respect to the number of video frames captured per step.
+    We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    \item Using \cref{item:contrib:model}, we explore the potential for optimization in these systems, and propose a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}.
+    This framework is applicable to a number of optimization purposes in \ac{WCA}, and we demonstrate its use for two relevant ones: number of samples captured per step and energy consumed per step.
 \end{enumerate}
@@
 \todo[inline]{%
-Nice intro.
-Perhaps too long, and a concise presentation - preferably as bullet list - of the contributions is missing.
-Try to shorten in particular for the first three paragraphs.
-Another question would be if there is no other related work.
-For instance, Zorzi recently published a model for AR and VR - trace-based - that would be a good reference to have.
+Shorten.
+Include Zorzi et al. AR VR
 }

### 2022-11-07 conclusion.tex
tweaking
too tired to continue

@@
 Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming and prohibitively expensive for many research groups.
-In this paper, we have presented the \edgedroid{} model of human timing behavior for \ac{WCA}
+
+In the first half of this paper, we have introduced the \edgedroid{} model of human timing behavior for \ac{WCA}, the first data-driven model for human timings in \ac{WCA} applications.
+This model represents a stochastic approach to execution time modeling which builds upon the data collected for \textcite{olguinmunoz:impact2021}.
+Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+
+In the second half, we have explored the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
+We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}.
+We have shown that this framework is applicable to a myriad of parameters in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
+Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
+
+\todo[inline]{Finish}

### 2022-11-07 implications_optimization.tex
too tired to continue

@@
 For this we implement these sampling schemes in Python and run a number of simulations with them.
-The first scheme uses our approach and \cref{eq:epsilon_terminal,eq:optimization:sampling} to determine the optimum sampling instants and the values for \( \alpha \) and \( \beta \) at each step.
+The first scheme uses our approach, \cref{eq:epsilon_terminal,eq:optimization:sampling}, to determine the optimum sampling instants and the values for \( \alpha \) and \( \beta \) at each step.
 It includes an embedded timing model (without any distribution fitting) to provide updated estimates of the mean execution time \( \mu \) and \( \sigma \) at every step as well, allowing it to adapt to the state of the system.
@@
 A maximum expected wait time \( w_0 = 1.0\,\si{\second} \) thus translates into a maximum expected sampling interval of \SI{2.0}{\second}, yielding a minimum sampling rate \( R_\text{min} = {2.0\,\si{\second}}^{-1} = 0.5\,\si{\hertz} \).
+It should also be noted that in both aperiodic schemes, our sampling-count-optimized approach and the \ac{CDF}-based approach, there exists the possibility for sampling instants to be \emph{missed} due to the actual \ac{RTT} of the system being higher than the parameterization of the schemes.
+In these cases, both schemes will degrade into greedy sampling as they attempt to sample as fast as possible.
 
@@
         \caption{%
-            Percentage difference in mean number of captured samples per step with respect to the greedy sampling scheme.
+            Percentage difference in mean number of captured samples per step by the sample-count-optimized approach with respect to the three reference schemes.
+            In other words, curves represent the relative performance of the sample-count-optimized scheme when using the corresponding reference scheme as baseline.
             Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
@@
 
-The results of this investigation are presented in \cref{fig:optimization:samples}.
-\todo[inline]{Discuss results.}
+The results of this investigation are presented in \cref{fig:optimization:samples}, and clearly show the advantages of using the sample-count-optimized scheme over the current state-of-the-art.
+The performance of \textcite{Wang2019Towards}'s approach appears to degrade with lower \acp{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
+On the other hand our sampling scheme exhibits consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
+This makes the sample-count-optimized approach more predictable, reliable, and consistent, and thus more suitable for real-world deployments.
+
+It should be noted that although both schemes seem to tend towards \num{2.0} samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
+As \acp{RTT} increase way above the expected wait time, the probability of the sum of the first sampling interval and the \ac{RTT} being larger than the execution time of the step tends towards \num{1.0}.
+This leads to these sampling schemes consistently sampling only twice each step: a first sample which is taken before the execution time of the step, and a second one \ac{RTT} seconds later, after the execution time has been reached.
 
@@
         \caption{%
-            Percentage difference in mean per step energy consumption with respect to the greedy sampling scheme.
+            Percentage difference in mean per step energy consumption by the energy-optimized sampling scheme with respect to the three reference schemes.
             Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
@@
 Again, we include the reference greedy and offline optimum schemes, as well as \textcite{Wang2019Towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \ac{RTT}.
-The results of these experiments are presented in \cref{fig:optimization:energy}.
 
-\todo[inline]{Discussss}
+The results of these experiments are presented in \cref{fig:optimization:energy}, and they clearly illustrate the advantages of the integration of our timing models with an adaptive, energy-optimized sampling scheme when compared to unoptimized and state-of-the-art sampling schemes.
+Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{Wang2019Towards}'s state-of-the-art, and is up to \SI{40}{\percent} more efficient than the greedy scheme.
+Furthermore, once again the behavior of our approach is more consistent and reliable than the competition, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.

### 2022-11-07 implications_optimization.tex
Update on Overleaf.

@@
 
-In the following we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
-We directly take the modeling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the assumption of one-way communication in all but the final sample.
-With feedback given even to the discarded samples in our model, we have, 
+In the following, we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
+We directly take the modelling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the  assumption of one-way communication in all but the final sample.
+With feedback given even to the discarded samples in our model and communication delay defined as the total delay in either direction, we have, 
 \begin{alignat}{2}
-    \mathrm{E}=&\;2\mathcal{S}\tau_cP_c+(\mathcal{T}+\mathcal{W}+\tau_\mathrm{p}+2\tau_\mathrm{c}-2\mathcal{S}\tau_c)P_0\nonumber\\
+    \mathrm{E}=&\;\mathcal{S}\tau_cP_c+(\mathcal{T}+\mathcal{W}+\tau_\mathrm{p}+\tau_\mathrm{c}-\mathcal{S}\tau_c)P_0\nonumber\\
     % &=s\tau_c(P_c-P_0^{(t)})+wP_0^{(t)}+(\tau+\tau_c+\tau_p) P_0^{(t)}+\tau_cP_c\\
-    =&\;2\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +2\tau_{\text{c}}) P_0\nonumber\\
-&\Rightarrow \alpha=2\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0 \label{eq:optimization:energy}
+    =&\;\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +\tau_{\text{c}}) P_0\nonumber\\
+&\Rightarrow \alpha=\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0 \label{eq:optimization:energy}
 \end{alignat}
-\( \tau_\text{p} \) and \( \tau_\text{c} \) correspond to the processing and communication delay for each sample.
+\( \tau_\text{p} \) and \( \tau_\text{c} \) correspond to the processing and two-way communication delay for each sample.
 \( P_\text{c} \) and \( P_0 \) correspond to the communication and idle power, respectively, of the \ac{WCA} client device.

### 2022-11-08 intro.tex
Update on Overleaf.

@@
 We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
-Finally, in \textcite{ICCperiodic1,TMCperiodic} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
-This is further extended into an aperiodic strategy in \textcite{secAperiodic}.
+Finally, in \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
+This is further extended into an aperiodic strategy in \textcite{Moothedath2022Aperiodic}.
 

### 2022-11-11 intro.tex
intro

--- 2022-11-07 intro.tex
@@
     We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-    \item Using \cref{item:contrib:model}, we explore the potential for optimization in these systems, and propose a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}.
-    This framework is applicable to a number of optimization purposes in \ac{WCA}, and we demonstrate its use for two relevant ones: number of samples captured per step and energy consumed per step.
+    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
+    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
+    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we use to study two potential avenues for optimization in \ac{WCA}; number of processed samples and energy consumption per step.
+    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed per step while maintaining comparable levels of responsiveness, and an average of \SI{20}{\percent} less energy consumed.
 \end{enumerate}
 
-Our contributions build on previous work on the characterization of human perception of delay in \ac{WCA} systems.
-We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.
-
-Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}, as well as on optimum sampling of edge-bound interactive systems.
-In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
-We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
-In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
-In \textcite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
-We found that humans generally pace themselves according to the perceived system responsiveness.
-When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.
-We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
-Finally, in \textcite{ICCperiodic1,TMCperiodic} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
-This is further extended into an aperiodic strategy in \textcite{secAperiodic}.
+\medskip
 
 This paper is structured as follows.
+\cref{sec:relwork} discusses related work in the field of \acl{WCA} modeling.
 In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
@@
 Next in \cref{sec:implications:footprint} we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
-This is followed by a more in-depth investigation in\cref{sec:implications:optimization} on the potential consequences of such a model on the general optimization of \ac{WCA} systems.
-Finally, in \cref{sec:conclusion} we summarize and conclude this paper.
-
-\todo[inline]{%
-Shorten.
-Include Zorzi et al. AR VR
-}
+This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \ac{WCA} systems.
+In this section, we also introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
+Finally, in \cref{sec:conclusion} we summarize and conclude this paper, as well as briefly discuss potential avenues for future work.

### 2022-11-11 related_work.tex
intro

\section{Related work}\label{sec:relwork}

Our contributions build on previous work on the characterization of human perception of delay in \ac{WCA} systems.
We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.

Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}, as well as on optimum sampling of edge-bound interactive systems.
In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
In \textcite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
We found that humans generally pace themselves according to the perceived system responsiveness.
When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.
We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
Finally, in \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
This is further extended into an aperiodic strategy in \textcite{Moothedath2022Aperiodic}.

\todo[inline]{%
Include Zorzi et al. AR VR
}

### 2022-11-12 intro.tex
relwork

@@
 They do so by capturing user and environment information, and leveraging compute-intensive algorithms to analyze the data and produce timely context-dependent feedback.
-\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{Ha2014towards,Satya2019augmenting}, and have since then expanded to a greater range of use cases.
-In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical}.
+\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since then expanded to a greater range of use cases.
+In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
 
@@
     We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach in \textcite{olguin2018scaling,olguin2019edgedroid}.
     \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.

### 2022-11-12 related_work.tex
lecci
relwork

@@
 
-Our contributions build on previous work on the characterization of human perception of delay in \ac{WCA} systems.
-We expand upon the findings of \textcite{Ha2014towards,Chen2017Empirical}, who identified the need and opportunities for edge offloading of \ac{WCA} and characterized latency bounds in these systems.
+There exist already in literature a number of approaches to the characterization and benchmarking of mobile \ac{XR}, and by extension \ac{WCA}, deployments.
+\emph{OpenRTiST}~\cite{george2020openrtist} is a tool which uses a compute-heavy, yet latency-sensitive workload --- image style transfer~\cite{jing2019neural} --- to load and benchmark edge computing deployments.
+It acts as an end-to-end realistic \ac{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
+\textcite{Chen2017Empirical} use a number of prototype \ac{WCA} implementation to study and characterize real-world latency bounds in these systems.
+In \textcite{lecci2021open}, the authors collect and analyze large amounts of \ac{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
+\textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for for the benchmarking of mobile hardware in the context of \ac{AR}.
+The toolkit incorporates a series of workloads which stress different components of the mobile device and calculates a score for each workload.
+\emph{Yahoo Cloud Serving Benchmark}~\cite{cooper2010benchmarking} and \emph{DCBench}~\cite{jia2013characterizing} are workload sets intended for benchmarking cloud services which have further been used in the context of benchmarking edge computing infrastructure for \ac{AR} suitability.
+\emph{Edgebench}~\cite{das2018edgebench} and \emph{Defog}~\cite{mcchesney2019defog} benchmark workload performance on the edge versus on the cloud.
 
-Furthermore, we directly build upon and expand on our previous work on the modeling of human behavior and characterization of human timings in \ac{WCA}, as well as on optimum sampling of edge-bound interactive systems.
+We contribute to this body of work by providing the first tool for the benchmarking of \ac{WCA} that dynamically considers the effects of human behavior.
+In contrast to the above described works, which generate loads with relatively static profiles, our model is able to react to changes in system responsiveness like a human would.
+We achieve this by building upon our own previous contributions to this field.
 In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
@@
 When interacting with a highly responsive system, humans tended to speed up with each step; conversely, humans tended to slow down in highly unresponsive states.
-We name the new model presented in this paper \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach.
-Finally, in \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
-This is further extended into an aperiodic strategy in \textcite{Moothedath2022Aperiodic}.
 
-\todo[inline]{%
-Include Zorzi et al. AR VR
-}
+The characterization and optimization, of mobile \ac{XR} has also long been a topic of research.
+In \textcite{srinivasan2009performance}, the authors analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
+They proceed to use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
+The authors of \textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
+In \textcite{huang2021proactive}, the authors present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
+\textcite{al2017energy} propose an resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
+As above, we have also previously directly contributed to this field.
+In \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
+This is extended into an aperiodic strategy in \textcite{Moothedath2022Aperiodic}.
+We further develop this approach in this work by deriving a more practical approximate solution to the aperiodic strategy, which we then use to implement a novel framework for the generic optimization of responsiveness-resource consumption trade-offs.

### 2022-11-13 implications_optimization.tex
Update on Overleaf.
eq

@@
 For instance, in the modeling used by \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
-The outline of our solution that minimizes \cref{eq:epsilon_terminal} is provided below with the complete mathematical derivation in \cref{appx1}.
+Our solution which minimizes \cref{eq:epsilon_terminal} is provided in \cref{tn_approx_rayleigh}.
+It assumes a $\mathcal{T}$ distributed according to a Rayleigh distribution with parameter $\sigma$.
+The complete mathematical derivation that leads to this formula is detailed in \cref{appx1}.
 
-\bigskip
-
-We first define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
-{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1\label{rt}
-\end{alignat}
-For instance, for periodic sampling, the rate function is a constant equal to the sampling frequency. 
-% Using this rate function, we rewrite the expected number of samples as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{S}]&=\int_{t=0}^{\infty}\bigg(\!\int_{x=0}^{t}\!\!\!\!r(x)\,\mathrm{d}x\bigg)f_{\mathcal{T}}(t)\,\mathrm{d}t,\nonumber
-% \end{alignat}
-% and the approximate the expected wait time as
-% \begin{alignat}{1}
-% \mathbb{E}[\mathcal{W}]&=\int_{0}^{\infty}\dfrac{1}{2r(t)}f_\mathcal{T}(t)\,\mathrm{d}t.\nonumber
-% \end{alignat}
-Next, rewriting \cref{eq:epsilon_terminal} in terms of $r(t)$ and minimizing it gives as an optimum rate function $r^*(t)$ as
-\begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta f_\mathcal{T}(t)}{2\alpha\bar{F}_\mathcal{T}(t)}}\nonumber\\
-\intertext{in turn resulting in, for a $\mathcal{T}$ distributed according to a Rayleigh distribution with parameter $\sigma$,}
 r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
-% \Rightarrow &\int_{t_n}^{t_{n+1}}\!\!\!\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\,\mathrm{d}t=1,\;\forall n\geq1\nonumber\tag{from \eqref{rt}}\\
-% \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
 \Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tn_approx_rayleigh}
@@
 However, an unconstrained optimization of the number of samples is trivial and meaningless as the solution points to a single sample at $t\!\rightarrow\!\infty$, which also takes the wait time to infinity.
-We look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
+Thus, we look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
 That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfy the condition

### 2022-11-13 implications_optimization.tex
opti intro

@@
 At the same time, imposing too stringent requirements on resource consumption can adversely affect application responsiveness, which in turn has the potential to drastically reduce quality of experience for users.
-Optimizing the deployment of these applications thus intrinsically involves achieving a balance between the reduction of resource consumption while maintaining an acceptable level of system responsiveness.
+Optimizing the deployment of these applications thus inherently involves achieving a balance between the reduction of resource consumption while maintaining an acceptable level of system responsiveness.
 
-In the following section we will study the optimization of resource consumption-responsiveness trade-offs in two dimensions of \ac{WCA}, number of samples captured per step and energy consumption per step.
-
-% \todo[inline]{Finish}
+A question then arises as to what extent accurate modeling of human behavior plays a role in the optimization of \ac{WCA} applications and systems.
+We have already seen from the results in \cref{sec:implications:footprint} that changes in responsiveness can have drastic effects on human execution times and subsequently on overall application lifetimes.
+This points towards parameters with dependencies on step timings as prime candidates for evaluating the effects of human behavior on their potential for optimization of different dimensions of \ac{WCA}.
+One such parameter of particular interest to us is sampling strategy, which intrinsically hinges on a stochastic understanding of human timings in a step.
+In the following section we will study the potential advantages of using the realistic model presented in \cref{sec:model} to estimate the execution time distributions of steps to adapt the sampling strategy at runtime.
+We will first introduce a generic optimization framework which produces an aperiodic sampling strategy that optimizes for arbitrary responsiveness-resource consumption trade-offs in \ac{WCA}.
+We will then employ this framework for the minimization of two different parameters: number of samples and total energy consumption per step.
+In order to illustrate the advantages of our framework, we will compare our results with a state-of-the-art baseline which does not dynamically adapt.
 

### 2022-11-13 abstract.tex
pre-shorten draft

\begin{abstract}
    \acf*{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
    Recruiting humans is cumbersome and expensive, and human-subject research is often hard to repeat and replicate.
    Considering the intended mass-scale deployment of these applications in the future, there exists a real need for tools enabling human-independent benchmarking.

    We present in this paper the first data-driven model which allows for the complete end-to-end emulation of humans in \acs*{WCA}.
    We demonstrate the utility of this model by employing it to study application lifetime footprints, where we find an up to \SI{12}{\percent} difference with respect to first order approximations.
    Further, we use this model to develop a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \acs*{WCA}.
    We show that our framework can provide up to a \SI{50}{\percent} reduction in processed frame samples and up to a \SI{20}{\percent} reduction in energy consumption when compared to the state-of-the-art.
\end{abstract}

### 2022-11-13 acks.tex
pre-shorten draft

\section*{Acknowledgements}\label{sec:acks}

\todo[inline]{%
    TODO: Thanks

    TODO: Check if funding is correct.
}

This work has been partially funded by the \ac{SSF} (grant number \verb|ITM17-0246| (ExPECA)), and the \ac{NSF} (grant number \verb|CNS-1518865|).
The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or funding sources.

### 2022-11-13 conclusion.tex
pre-shorten draft

@@
 
-In the second half, we have explored the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
-We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}.
+Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
+We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant over- or underestimations of application footprint.
+
+Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
+We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy. 
 We have shown that this framework is applicable to a myriad of parameters in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
@@
 
-\todo[inline]{Finish}
+This work serves as an important, but yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
+Many directions remain open to additional exploration in this space.
+For instance, our current model only targets a particular class of step-based \ac{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \ac{AR} and \ac{XR}, is on our roadmap.
+Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the US.\@
+An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.
+
+In terms of the optimization framework, our current approach builds upon the assumption that execution times are Rayleigh-distributed.
+Although this distribution fits the data well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
+Our future efforts consider then developing this framework towards these more accurate distributions.
+Finally, our current approach only considers a single client, which will be far from the case in a real-world \ac{WCA} deployment.
+As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.

### 2022-11-14 background.tex
trim

@@
 \acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \ac{AR} applications.
-In this work we focus on a particular category of these applications --- denoted ``step-based'' \ac{WCA} --- which have as their ultimate goal the guiding of a user through a sequential task.
+In this work we focus on a particular category --- denoted ``step-based'' \ac{WCA} --- which have as their goal the guiding of a user through a sequential task.
 Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
@@
 
-In order to discuss these applications with precision and fully understand them, we provide some definitions and explanations relating to their operation.
+In order to discuss these applications with precision we provide some definitions relating to their operation.
 First of all, a \emph{step} is formally understood as a specific action to be performed by the user, described by a single instruction, and a \emph{task} consists of a series of steps to be executed in sequence (see \cref{fig:task}).
@@
 
-As expressed above, \acp{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
+\acp{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
 Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
@@
 \( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
-We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \) (once again, refer to \cref{fig:step} for a visual breakdown of these values).
+We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
 The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\ac{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \ac{WCA}.
@@
 
-In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
-We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA} devised by \textcite{Chen2015LEGO}.
+In \textcite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
+We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA}~\cite{Chen2015LEGO}.
 Subjects interacted with the assistant while we altered the responsiveness in realtime and captured key application and task performance metrics.
-Additionally, we also employed questionnaires to evaluate key personality traits of the participants, and correlated these results to the task performance metrics collected.
+Additionally, we also employed questionnaires to evaluate personality traits of the participants, and correlated these results to the task performance metrics collected.
 
-For this work, we used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
+We used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
 \emph{Delay} corresponded to a temporarily fixed time duration of the processing of each frame was padded to --- i.e.\ if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
@@
 
-Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5} without loss of accuracy.
+Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5}.
 
-With this in mind, we can reformulate the conclusions of this previous work as follows:
+We can then reformulate our previous conclusions as:
 \begin{enumerate}
@@
     \item The above effects are modulated by an individual personality trait.
-    We will discuss this in detail below in \cref{ssec:moderationeffects}
+    % We will discuss this in detail below in \cref{ssec:moderationeffects}
 \end{enumerate}
@@
         \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \ac{TTF} in \textcite{olguinmunoz:impact2021}.
-        Fits were performed using the \ac{MLE} method.
+        Distributions were fitted using \ac{MLE}.
     }\label{fig:muexgaussian}

### 2022-11-14 conclusion.tex
trim

@@
 Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
-We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant over- or underestimations of application footprint.
+We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
 
@@
 We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy. 
-We have shown that this framework is applicable to a myriad of parameters in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
+We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
 Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
@@
 Although this distribution fits the data well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
-Our future efforts consider then developing this framework towards these more accurate distributions.
+Our future efforts consider developing this framework towards these more accurate distributions.
 Finally, our current approach only considers a single client, which will be far from the case in a real-world \ac{WCA} deployment.

### 2022-11-14 implications_footprint.tex
trim

@@
 In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This new model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
-This distribution is then randomly sampled at runtime to obtain execution times for each step, without any sort of adjustment to the current state of the system.
+This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
+This distribution is then randomly sampled at runtime to obtain execution times for each step, without any adjustment to the current state of the system.
 
@@
 The testbed configurations include setups with \numlist{1;4;7;10} clients.
-Due to having only limited time, each combination of testbed and timing model configuration is only repeated \num{10} times.
+Due to having limited time, each combination of testbed and timing model configuration is only repeated \num{10} times.
 The results are presented in \cref{fig:testbed_lifetimes}.

### 2022-11-14 implications_optimization.tex
trim

@@
 We will first introduce a generic optimization framework which produces an aperiodic sampling strategy that optimizes for arbitrary responsiveness-resource consumption trade-offs in \ac{WCA}.
-We will then employ this framework for the minimization of two different parameters: number of samples and total energy consumption per step.
+We will then employ this framework for the minimization of two different metrics: number of samples and total energy consumption per step.
 In order to illustrate the advantages of our framework, we will compare our results with a state-of-the-art baseline which does not dynamically adapt.
@@
 This represents a completely unoptimized design with no considerations for resource-consumption trade-offs; it simply attempts to maximize the number of captured samples per step.
-This is an interesting approach to include given that it corresponds to the default sampling strategy used in most existing \ac{WCA} prototypes.
+This is an interesting approach to include as it corresponds to the sampling strategy used in most existing \ac{WCA} prototypes.
 
@@
 It should also be noted that in both aperiodic schemes, our sampling-count-optimized approach and the \ac{CDF}-based approach, there exists the possibility for sampling instants to be \emph{missed} due to the actual \ac{RTT} of the system being higher than the parameterization of the schemes.
-In these cases, both schemes will degrade into greedy sampling as they attempt to sample as fast as possible.
+In these cases, both schemes will degrade into greedy sampling.
 
@@
 The performance of \textcite{Wang2019Towards}'s approach appears to degrade with lower \acp{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
-On the other hand our sampling scheme exhibits consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
+On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
 This makes the sample-count-optimized approach more predictable, reliable, and consistent, and thus more suitable for real-world deployments.

### 2022-11-14 intro.tex
trim

@@
 \acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
-These systems aim to seamlessly integrate into the day-to-day of users to provide real-time, context-aware information and feedback.
-They do so by capturing user and environment information, and leveraging compute-intensive algorithms to analyze the data and produce timely context-dependent feedback.
-\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since then expanded to a greater range of use cases.
+These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
+\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
 In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
 
-A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses --- they are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers by tracking their progress and providing feedback and instructions at appropriate times.
-Moreover, as their name suggests, another key attribute of these applications is their wearability, which implies the use of lightweight, battery-powered, low-energy consumption devices.
-On the other hand, their context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs such as real-time video feeds which require intensive compute capabilities to process.
-Furthermore, as with any \ac{AR} application, \ac{WCA} systems are sensitive to latency.
+A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses.
+They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
+This context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video feeds, which require intensive compute capabilities to process.
+This also translates into latency-sensitivity, as with any \ac{AR} application.
 Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
+On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.
 
-The combination of these characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+The combination of these opposing characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
 However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
@@
 
-With that context, we can express the contributions of this paper as follows:
+In that context, the contributions of this paper are as follows:
 \begin{enumerate}
@@
     We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-    We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach in \textcite{olguin2018scaling,olguin2019edgedroid}.
+    We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
     \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
@@
     \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we use to study two potential avenues for optimization in \ac{WCA}; number of processed samples and energy consumption per step.
-    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed per step while maintaining comparable levels of responsiveness, and an average of \SI{20}{\percent} less energy consumed.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
 \end{enumerate}

### 2022-11-14 model.tex
add James comments as todonotes
Update on Overleaf.
fix
tweaking
pre-shorten draft
trim

@@
 And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}.
-Finally, in \cref{ssec:model:obtaining} we briefly discuss where implementations of this model can be obtained.
+% Finally, in \cref{ssec:model:obtaining} we briefly discuss where implementations of this model can be obtained.
 
@@
 
-As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
-For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-We group the data previously collected according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
+% As expressed in \cref{ssec:plos}, for \textcite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
+% For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
+The processing of the data from \textcite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
 The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
@@
 
-We employ a cleaned and re-parameterized copy of the aforementioned timing data.
+We employ a cleaned and re-parameterized copy of the timing data.
 The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
@@
 This weighing ensures that drastic changes in \ac{TTF} are significantly remembered by the model for at least \num{4} steps --- in line with our previous findings on human behavior --- and are subsequently quickly forgotten.
-
 \begin{equation}\label{eq:weights}
@@
     \left\{ \begin{array}{ll}
-        \frac{e^{0.7 i}}{\sum\limits^{12}_{j=1} e^{0.7 j}} & 1 \leq i \leq 12 \\
+        \frac{e^{-0.7 i}}{\sum\limits^{12}_{j=1} e^{-0.7 j}} & 1 \leq i \leq 12 \\
         & \\
@@
 The resulting weighted \acp{TTF} are then binned into continuous ranges by splitting the data on the \num{7}-quantiles.
-The exact resulting bins of this operation on the data can be seen in \cref{tab:defaultbins}.
+The exact resulting bins of this operation on the data are presented in \cref{tab:defaultbins}.
 
 Next, the data is further tagged according to its associated level of normalized neuroticism.
-For this work, we use two levels, \emph{low} and \emph{high}, the exact values for which can be seen in \cref{tab:defaultbins}.
+For this work, we use two levels, \emph{low} and \emph{high}, the exact values for which can also be seen in \cref{tab:defaultbins}.
 
 After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \acp{TTF}.
-We devise two variants of a model which uses the above-described pre-processed data to generate realistic timings for \ac{WCA}.
-These models operate as follows:
+We devise two variants of such a model, which operate as follows:
 
@@
     \item At the beginning of each step, the model is fed the measured \ac{TTF} for the previous step.
-    \item The model calculates a weighted average over the latest \num{12} steps recorded using the weights defined in \cref{eq:weights}.
+    \item The model calculates a weighted average over the latest \num{12} steps using the weights defined in \cref{eq:weights}.
     \item The resulting weighted \ac{TTF} is binned into the corresponding range (\cref{tab:defaultbins}).
@@
     \begin{itemize}
-        \item directly sampling the empirical execution time values;
-        \item or, first fitting a distribution to the execution time samples and then sampling the distribution instead.
+        \item directly sampling the execution time values;
+        \item or, using \ac{MLE} to fit a distribution to the execution time samples and then sampling the distribution instead.
     \end{itemize}
@@
 The distribution chosen for the second variant described above corresponds to the \ac{exGaussian} distribution, previously discussed in \cref{ssec:moderationeffects}.
-Fitting is done using \ac{MLE}.
 
@@
 Additionally, we can observe the effects of neuroticism on generated execution times, as specified in \cref{it:neuro}.
-At low neuroticism, the average difference between execution times at low versus high \acp{TTF} was of roughly \SI{6.2}{\percent}, compared to \SI{12.5}{\percent} at high neuroticism, clearly putting into evidence the modulating effect of this variable.
+At low neuroticism, the average difference between execution times at low versus high \acp{TTF} was of roughly \SI{6.2}{\percent}, compared to \SI{12.5}{\percent} at high neuroticism.
 
@@
 Next, \cref{fig:exectimeduration} shows the evolution of generated execution times while the model is subject to a fixed \ac{TTF}, either \emph{low} (\SI{0}{\second}) or \emph{high} (\SI{5}{\second}).
-These results were generated by first warming up the model with \num{25} random \acp{TTF}, and recording the generated execution times over a sequence of \num{12} steps at a fixed \ac{TTF}; this procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
+These results were generated by first warming up the model with \num{25} random \acp{TTF}, and then recording the generated execution times over a sequence of \num{12} steps at a fixed \ac{TTF}; this procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
 Once again, we see here behavior matching what is expected of the model, in particular with respect to \cref{it:duration}.
@@
 As before, we run \num{600} repetitions of this procedure for each combination of model configuration, origin \ac{TTF}, and destination \ac{TTF}.
-Execution times generated immediately after a transition from a higher \ac{TTF} into a lower one are consistently higher than execution times generated without an preceding change in \acp{TTF}.
+Execution times generated immediately after a transition from a higher \ac{TTF} into a lower one are consistently higher than execution times generated without a preceding change in \acp{TTF}.
 Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \ac{TTF} into a higher one.
@@
 The tags assigned corresponded to:
-\begin{description}[font={\bfseries\ttfamily}]
-    \item[SUCCESS:] frames which triggered a transition to a new step in the logical task model of the \ac{WCA}, and thus cause the generation of feedback to the user.
-        This tag is also used for frames which corrected a previous mistake.
-    \item[REPEAT:] frames which contained the same board state as the previous successful frame, and thus produced no feedback.
+\begin{description}[font={\bfseries\ttfamily}, wide]
+    \item[SUCCESS:] frames which triggered a transition to a new step (or the correction of a previous mistake) in the logical task model of the \ac{WCA}, and thus cause the generation of feedback to the user.
+    \item[REPEAT:] frames which captured the same board state as the previous successful frame, and thus produced no feedback.
     \item[LOW\_CONFIDENCE] frames for which the image recognition algorithm in the \ac{WCA} did not reach the necessary confidence threshold to interpret it as a valid board state.
@@
 We correlate this frame data with the step timing data described in \cref{ssec:model:exectimes} to match frames with their corresponding step execution times.
-We assign to each frame a normalized instant value corresponding to its capture instant (expressed in seconds since the start of the step) divided by the total execution time of the step.
-To exemplify, for a step with execution time \( t_\text{exec} = \SI{10}{\second} \), a frame captured at time \( \tau = \SI{3}{\second} \) from the start of the step, will have a normalized instant value \( t_\text{norm} \):
+We assign to each frame a normalized instant value \( t_\text{norm} \) corresponding to its capture instant \( \tau \) (expressed in seconds since the start of the step) divided by the total execution time \( t_\text{exec} \) of the step:
+
 \begin{align}
-    t_\text{norm} = \frac{\tau}{t_\text{exec}} = \frac{\SI{3}{\second}}{\SI{10}{\second}} = 0.3
+    \left( t_\text{norm} = \frac{\tau}{t_\text{exec}} \right) \in [0, 1]\label{eq:tnorm}
 \end{align}
 
-This allows us to then analyze the distribution of frame tag probabilities as a step progresses, independently of execution times.
+% To exemplify, for a step with execution time \( t_\text{exec} = \SI{10}{\second} \), a frame captured at time \( \tau = \SI{3}{\second} \) from the start of the step, will have a normalized instant value \( t_\text{norm} \):
+% \begin{align}
+%     t_\text{norm} = \frac{\tau}{t_\text{exec}} = \frac{\SI{3}{\second}}{\SI{10}{\second}} = 0.3
+% \end{align}
+
+This allows us to analyze the distribution of frame tag probabilities as a step progresses, independently of execution times.
 This is illustrated in \cref{fig:frameprobs}.
@@
 As the user starts moving and performing actions, \texttt{BLANK} frames start to dominate, as this activity prevents the \ac{WCA} from capturing ``clean'' frames.
-Finally, it must be noted that \texttt{SUCCESS} frames are not included in this probability density plot, as, by definition (\( t_\text{norm} \) represents the normalized instant value):
+Finally, it must be noted that \texttt{SUCCESS} frames are not included in this probability density plot, as, by definition:
 \begin{equation}
@@
 \end{equation}
+
 That is, any frame captured immediately at or after the execution time has been reached will contain the finished board state and thus correspond to a \texttt{SUCCESS} frame.
@@
 
-Using the above insights, together with the corresponding recorded video frames, we devise an scheme for the procedural generation of a synthetic trace for any step in \ac{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
+Using the above insights, together with the corresponding recorded video frames, we devise a scheme for the procedural generation of a synthetic trace for any step in \ac{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
 We first prepare a discretized representation of the probability density map in \cref{fig:frameprobs}.
@@
 \begin{enumerate}
-    \item We calculate the normalized instant value \( t_\text{norm} = \frac{\tau}{t_\text{exec}} \).
-    \item If \( t_\text{norm} \ge 1.0 \), we select the \texttt{SUCCESS} frame and finish the trace.
-    \item If instead \( t_\text{norm} < 1.0 \), we find the appropriate bin for \( t_\text{norm} \) and then select a frame by performing a weighted random sampling of the frame categories in the normalized instant bin using the relative proportions for each category.
+    \item We calculate \( t_\text{norm} \) according to \cref{eq:tnorm}.
+    \item If \( t_\text{norm} \ge 1.0 \), we select the \texttt{SUCCESS} frame and stop sampling.
+    \item If instead \( t_\text{norm} < 1.0 \), we find the appropriate bin for \( t_\text{norm} \) and then select a frame by performing a weighted random sampling of the frame categories in the normalized instant bin.
 \end{enumerate}

### 2022-11-18 body/part1/futurework.tex
structure

\chapter{Open questions \& future work}

Extending methodology to other classes of \ac{WCA}.
Ping-pong is the typical example --- look for other classes too.

Extending methodology to other systems.
\ac{AR}, \ac{XR} in general?
Closed-loop control? (Although we already started with CLEAVE).

Further optimization of \ac{WCA} systems.
In this thesis we touch upon sampling, but other dimensions can also be optimized.
Resource allocation, scheduling, user identification and classification, QoS/QoE guarantees?

### 2022-11-18 body/part1/introduction.tex
structure

\chapter{Introduction}

\section{Background \& Related Work}

Deep focus on theoretical background.

\subsection{Theoretical background}
\textcite{Satya2009Case}.
Other papers detailing the advantages of Edge computing.

Next, what applications benefit from the Edge?
\ac{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
Main points: prototypes and characterization of system latencies and load.

\section{Presentation of research questions}

How will human-in-the-loop/\ac{WCA} systems look in production?

How will they look in large scale deployments?

How many applications can given systems support?

How can these dimensions be studied?
Here's where modeling and empirical data-driven research comes in.

Is there potential for optimization?
Yes, trading of system responsiveness for decrease in system load.
Underlying principle under aperiodic, adaptive sampling.
The potential for optimization is novel and stems from \textcite{Olguin:ImpactWCA2021}.

### 2022-11-18 body/part1/overview.tex
structure

\chapter{Overview of this thesis' contributions}

\section{Methodology}

Study by example, human subject research, and modeling.

Empirical research.

Include testbeds!

\section{Contributions of this thesis}

Here's where we ``present'' the papers.
However, we do this by presenting a ``red thread'' of contributions and key results.
Drive towards one big picture.

Add a hardened compilation of key results, conclusions, figures, and tables.
Synthesize and summarize the research questions and conclusions to find the red thread.

EdgeDroid (first demo paper at SEC, then HotMobile paper).

Tie in CLEAVE paper here as an extension of the approach towards other latency-sensitive applications.

Potentially include Ainur paper as a further extension of the methodology?

PLOS ONE paper as a stepping stone towards the final goal of a more realistic model?

Finally, EdgeDroid2 paper:
Presents the final model of human behavior
Applies it to a real optimization problem, employing theoretical results proved by Vishnu

### 2022-11-19 body/part1/overview.tex
adding intros and conclusions to build thread

@@
 
+\todo[inline]{
     Study by example, human subject research, and modeling.
@@
     Include testbeds!
+}
 
@@
 
+\todo[inline]{
     Here's where we ``present'' the papers.
@@
     Applies it to a real optimization problem, employing theoretical results proved by Vishnu
+}
+
+\todo[inline]{Intro and conclusions from EdgeDroid 1.0 paper}
+
+There is increasing interest from academia and industry in novel applications such as immersive augmented reality (AR) or wearable cognitive assistance (WCA)~\cite{Chatzopoulos:Hyperion,Ha:TowardsWearableCogAssist}, also known as human-in-the-loop applications.
+These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
+Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as Deep Neural Networks (DNN)~\cite{Ha:TowardsWearableCogAssist}.
+\Cref{fig:pipeline2} depicts such a system.
+These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
+Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{Chen:AnEmpiricalStudyOfLatency}.
+
+In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
+There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{Bittmann_Edge,flinn2012cyber,Chen:AnEmpiricalStudyOfLatency,Ha:JITProvisioning}.
+However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
+A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
+What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
+There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
+Finally, in production use these applications will most likely run concurrently with others.
+How does this, together with other best-effort applications, impact the latencies perceived by the human user?
+Existing studies~\cite{Ha:TowardsWearableCogAssist, Chen:EarlyImplementation, satya2009case,Chatzopoulos:Hyperion} of this class of applications have only lightly touched upon these issues~\cite{Chen:AnEmpiricalStudyOfLatency}.
+On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{Zubaidy15,Schiessl17} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
+We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
+
+The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
+We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
+
+Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
+This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
+EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
+This sensory information is then processed by the original compute process at the backend, generating feedback.
+However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
+Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
+Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
+Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
+It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
+To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications
+
+Benchmarking human-in-the-loop applications is hard, given their tight interaction with human users who complicate the scaling and repeatability of experiments.
+In this paper, we have presented a benchmarking suite for this type of applications, called EdgeDroid 1.0, capable of cutting out the need for users in performance evaluations.
+We achieve this by employing pre-recorded sensory input traces which we play over the network to the real application backend, employing a parameterized user model to react to feedback.
+We demonstrate its utility through a series of use case scenarios, from which we are able to extract metrics regarding latency both in regards to the application itself and the hardware stack.
+We believe the EdgeDroid 1.0 suite thus represents an important first step towards enabling inexpensive and low-complexity large-scale research on the scaling limits of this type of applications, a requirement for wide adoption of the technology.
+
+Nonetheless, there is still future work to be done.
+
+The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with WCA applications in order to present a more complete model in the future.
+As mentioned in \Cref{sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
+Rather, EdgeDroid 1.0 models a perfectly stoic user who is like an automaton and responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+Of course, no real human user is an automaton.
+In the future, we envision creating  many versions of EdgeDroid (i.e., EdgeDroid 2.0, EdgeDroid 3.0, etc.) that embody more human-like user models that more accurately emulate attributes such as those mentioned above.
+Experimental validation of these human-like user models via user studies will be an important part of our future work.
+
+We are also working on expanding the benchmarking suite to also work first with other types of Wearable Cognitive Assistance, and later with other categories of human-in-the-loop applications.
+Other types of WCA we will consider in future iterations of the tool include real-time task-assistance WCA applications (such as the Ping-Pong application described in~\cite{Chen:EarlyImplementation}), which don't have a linear task model like task-guidance WCA and have tighter latency bounds and context- and information-providing WCA applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
+The latter also do not have a linear task model, but present more lax latency bounds.
+
+\todo[inline]{Intro and conclusions from CLEAVE paper}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
+    \caption{
+        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
+    }\label{fig:cleave:ncs:struct}
+\end{figure*}
+
+The number and applications of \acp{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
+However, this rapid increase in adoption has mostly been limited to industrial contexts.
+Although \acp{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
+This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
+These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \ac{CPS} in the future.
+
+\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \ac{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
+This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
+
+%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \acp{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
+
+One the one hand, related literature in \acp{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level. 
+%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
+On the other, there exist several approaches when considering experimental methodologies. 
+%A number of works concerning \acp{NCS} deal with experimental studies.
+%\acp{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
+One such approach uses setups in which the complete system is built on top of real hardware.
+This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
+Conversely, other studies choose to instead use completely \emph{simulated} \ac{NCS} setups.
+The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
+These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \acp{NCS}.
+Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
+    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
+    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
+\end{enumerate*}
+
+As evidenced above, experimental research in \acp{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
+This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
+Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \ac{NCS}.
+This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
+
+Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
+In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\ 
+Their methodology utilizes joint knowledge of control, computation, and communication. 
+In their work various architectural elements and the corresponding delays associated with the NCS are modelled. 
+Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation. 
+This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
+Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
+
+We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
+In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
+As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
+``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
+While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
+
+Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
+It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
+The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
+These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
+
+\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
+It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+
+The rest of this paper is structured as follows.
+\cref{sec:approach} presents the design principles and implementation of the framework.
+In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \ac{CLEAVE}.
+Finally, in \cref{sec:conclusion} we conclude this work.
+
+The issue of repeatable and scalable benchmarks has been largely glossed over in \ac{NCS} literature, as existing experimental research studies tend to implement \emph{ad-hoc} solutions.
+
+In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
+We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
+\begin{enumerate}
+    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
+    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
+    Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
+    In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
+    Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
+    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
+    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
+    The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
+\end{enumerate}
+
+We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
+We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
+    \item similar setups exist in real-world industrial use
+    \item video analytics has long been proposed as a ``killer app'' for edge computing.
+\end{enumerate*}
+Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
+We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+
+There is still, however, work to be done.
+We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \acp{NCS} to share with the community.
+At the moment, the interactions of \ac{CLEAVE} and tools such as Docker are still quite superficial.
+Our goal is to achieve a much tighter integration, e.g.\ by providing the toolkit as pre-packaged container images.
+Finally, the validity of the results obtained by the framework will have to be verified through more thorough, realistic scenarios than what we have been able to show in this work.
+In particular, we intend to perform large-scale experimentation targetting 5G cellular deployments, as this technology is set to become the backbone of edge networks in the near future.
+
+\todo[inline]{Intro and conclusions from Ainur paper}
+
+Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications.
+On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
+The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.
+
+Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the 
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item \acs*{COSMOS}
+    \item \acs*{POWDER}
+    \item \acs*{ARA}
+    \item Drexel Grid \ac{SDR}
+\end{enumerate*} testbeds.
+\acs{COSMOS}~(\emph{\acl{COSMOS}})\acused{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \acp{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
+It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
+\ac{COSMOS} relies on the \ac{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \ac{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
+
+\acs{POWDER}~(\emph{\acl{POWDER}})\acused{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
+The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
+\ac{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \ac{VM} images containing the necessary software and configurations.
+These profiles are defined through using \ac{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
+
+The \ac{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
+Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
+It includes a broad range of wireless platforms ranging from low-\ac{UHF} massive \ac{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \ac{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+\ac{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \ac{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
+This affords the \ac{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
+
+Finally, the \emph{Drexel Grid \ac{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
+Primarily intended for \ac{SDR}-centric research, it does not integrate any core, cloud or edge components.
+However, the testbed extensively employs the \ac{LXC} runtime for the deployment of both experimental code and \ac{SDR} software, which affords users great freedom when it comes to the development of experiments.
+
+Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
+Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
+Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
+This is, for instance, the case with \ac{COSMOS} and \ac{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
+These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.
+
+To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
+CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \acp{VM} containing the necessary software for experiments.
+Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
+Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \iac{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.
+
+In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
+Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
+The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
+It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
+Furthermore, we make Ainur available to the community as \ac{FOSS}.
+It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
+
+The rest of this paper is structured as follows.
+\cref{sec:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
+This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
+Next, \cref{sec:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
+Finally, in \cref{sec:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
+
+In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
+Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
+    \item provisioning of and connection to remote cloud instances
+    \item initialization of \ac{IP} layer connectivity between hosts
+    \item collection of logs and data
+    \item deployment, scaling, and lifecycle management of containerized processes
+\end{enumerate*}.
+We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
+We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
+We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
+It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
+
+In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \ac{CHI}~\cite{keahey2020lessons}.
+While we believe Ainur to be unique and valuable in its focus on fully-automated end-to-end wireless edge computing experimentation, integrating with the above software stacks will provide a number of features and functionalities which will greatly expand Ainur's potential.
+Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \ac{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
+Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\ac{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.
+
+\todo[inline]{Intro and conclusions from PLOS paper}
+
+Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
+They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
+Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
+Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
+An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
+
+WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
+More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
+Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
+Detethering this assistance from its current fixed location will surely make it available to many more fields.
+
+Based on these use cases, we identify three main requirements for WCA:\@
+\begin{enumerate}
+    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
+
+    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
+    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+
+    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
+    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
+    This means that users will have expectations of constant, immediate feedback as they progress through the task.
+    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
+    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
+\end{enumerate}
+
+\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
+\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
+The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
+Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
+However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
+\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.
+
+% However, additional challenges remain in the path toward widespread adoption.
+\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
+We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
+\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
+% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
+Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
+It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}.
+Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
+A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
+% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.
+
+This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
+We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
+While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
+We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
+This is an important step toward a more systematic understanding of human behavior in this domain.
+
+We present in this paper the design and elaboration of an experimental WCA test-bed.
+This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
+Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
+The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
+
+Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
+
+\begin{itemize}
+  \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}
+
+  In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
+  The extent or form of these changes were however unknown.
+  We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.
+
+  Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
+  While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
+  To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
+  Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.
+
+  \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}
+
+  We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.
+
+  The results we present, however, seem to refute this hypothesis. 
+  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.
+
+  \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}
+
+  In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
+  In particular, we expected emotional effects to be correlated with the strength of the added delay.
+
+  The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
+  This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
+
+  \item \emph{Are these effects mediated by personality indicators in any way?}
+
+  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
+  We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
+
+  The results obtained agree with our hypothesis. 
+  We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.
+
+\end{itemize}
+
+% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
+% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
+% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
+% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
+% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
+
+We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
+One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and direct implications for resource and power consumption.
+Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished --- this has potential consequences for resource allocation strategies.
+Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource allocation across applications potentially looks different from what could be considered ``fair''.
+
+Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
+These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
+Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
+Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
+
+The structure of this paper is as follows.
+We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
+\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
+Then, in \cref{sec:results} we detail the results of our experiment.
+Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
+
+In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
+Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.
+
+We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
+The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
+When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
+This was evidenced by an increase in task execution times, even after accounting for the artificially introduced delays, as well as accelerometer data from participants wrists.
+Additionally, we found that the strength of this effect is modulated by individual differences between subjects.
+
+These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
+In this context, we believe there are two direct and important next steps to be performed.
+The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
+We believe these need to be implemented, tested and validated in real setups.
+In particular, we identify two of these implications to be prime candidates for their own experimental studies.
+One, our postulation that in cases where an impaired application is close to finishing, diverting resources to it might not be the most optimal course of action; and two, the possibility that ``fair'' degradation of system responsiveness across applications may be, in some cases, undesirable.
+Both of these questions could be answered with straightforward setups.
+
+The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
+These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
+However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
+Incorporating the results presented here would allow for much more realistic workloads.
+For instance, our findings relating to the effects of delay on execution times could be directly adapted to modulate timings in the input stream.
+Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
+Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.
+
+\todo[inline]{Intro and conclusions from EdgeDroid 2.0}
+
+\acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
+These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
+\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
+In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
+
+A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses.
+They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
+This context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video feeds, which require intensive compute capabilities to process.
+This also translates into latency-sensitivity, as with any \ac{AR} application.
+Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
+On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.
+
+The combination of these opposing characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
+One key gap in knowledge pertains to the current lack of tools and methodologies for scalable and repeatable study of \ac{WCA} application performance and resource utilization in realistic deployments.
+Due to their human-in-the-loop nature, these applications present a challenge to benchmark and characterize, in particular in real-scale deployments where dozens or even hundreds of users might concurrently use the system.
+Recruiting a large enough cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers.
+Furthermore, humans are unpredictable, and human-subject research is thus often hard to repeat and replicate.
+There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do away with the human-the-loop.
+
+\medskip
+
+In that context, the contributions of this paper are as follows:
+\begin{enumerate}
+    \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+    Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
+    We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
+    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
+    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
+    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
+\end{enumerate}
+
+\medskip
+
+This paper is structured as follows.
+\cref{sec:relwork} discusses related work in the field of \acl{WCA} modeling.
+In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
+In \cref{sec:model} we detail our model for the generation of realistic human timings in \ac{WCA}.
+We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
+Next in \cref{sec:implications:footprint} we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
+This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \ac{WCA} systems.
+In this section, we also introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
+Finally, in \cref{sec:conclusion} we summarize and conclude this paper, as well as briefly discuss potential avenues for future work.
+
+The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic human-in-the-loop nature.
+Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
+Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming and prohibitively expensive for many research groups.
+
+In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \ac{WCA}, the first data-driven model for human timings in \ac{WCA} applications.
+This model represents a stochastic approach to execution time modeling which builds upon the data collected for \textcite{olguinmunoz:impact2021}.
+Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+
+Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
+We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
+
+Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
+We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy. 
+We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
+Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
+
+This work serves as an important, but yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
+Many directions remain open to additional exploration in this space.
+For instance, our current model only targets a particular class of step-based \ac{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \ac{AR} and \ac{XR}, is on our roadmap.
+Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the US.\@
+An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.
+
+In terms of the optimization framework, our current approach builds upon the assumption that execution times are Rayleigh-distributed.
+Although this distribution fits the data well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
+Our future efforts consider developing this framework towards these more accurate distributions.
+Finally, our current approach only considers a single client, which will be far from the case in a real-world \ac{WCA} deployment.
+As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.

### 2022-11-19 body/part1/overview.tex
contrib

@@
 }
+
+This thesis presents seven key contributions to the literature; these are summarized below:
+
+\begin{enumerate}
+    \item\label{item:contrib:methodology} A methodological approach to studying system responsiveness versus resource consumption trade-offs in \ac{WCA}.
+    This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
+    \item In conjunction with the aforementioned methodology, a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
+    This tool implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
+    \item A characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
+    In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
+    \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.
+    \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \ac{NCS}.
+    \item A novel tool, CLEAVE, for the emulation of \ac{NCS} on edge infrastructure.
+    This tool has been designed from the ground-up to be edge-native, implemented using widely-adopted technologies in the disciplines of cloud and edge computing.
+    \item Finally, a software framework for the management and automation of the testbeds required for the methodology described in \cref{item:contrib:methodology,item:contrib:methodology:ncs}.
+\end{enumerate}
 

### 2022-11-19 body/part1/overview.tex
struct

@@
 
-This thesis presents seven key contributions to the literature; these are summarized below:
+This thesis presents seven main key contributions to the literature, which we summarize below:
 
@@
     \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.
-    \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \ac{NCS}.
-    \item A novel tool, CLEAVE, for the emulation of \ac{NCS} on edge infrastructure.
+    \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \acp{NCS}.
+    \item A novel tool, CLEAVE, for the emulation of \acp{NCS} on edge infrastructure.
     This tool has been designed from the ground-up to be edge-native, implemented using widely-adopted technologies in the disciplines of cloud and edge computing.
@@
 \end{enumerate}
+
+\section{A methodology for the study of latency trade-offs in \acs{WCA}}
+
+\subsection{The EdgeDroid model of human behavior}
+
+\subsubsection{Characterizing human reaction to delays in \ac{WCA}}
+
+\subsubsection{Implementing the model}
+
+\subsubsection{Implications of the EdgeDroid model for \ac{WCA} optimization}
+
+\section{A methodology for the study of latency trade-offs in \acsp{NCS}}
+
+\subsection{The CLEAVE emulation framework}
+
+\section{The Ainur software framework for Edge Computing testbeds}
 

### 2022-11-20 body/part1/overview.tex
structuring

@@
 
-This thesis presents seven main key contributions to the literature, which we summarize below:
+We summarize the seven key contributions of this thesis below:
 
@@
 
-\subsection{The EdgeDroid model of human behavior}
+\todo[inline]{
+    \cref{paper:EdgeDroid1.0} introduces our methodology, and validates its use in studying WCA.
+    Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
+}
 
-\subsubsection{Characterizing human reaction to delays in \ac{WCA}}
-
-\subsubsection{Implementing the model}
-
-\subsubsection{Implications of the EdgeDroid model for \ac{WCA} optimization}
-
-\section{A methodology for the study of latency trade-offs in \acsp{NCS}}
-
-\subsection{The CLEAVE emulation framework}
-
-\section{The Ainur software framework for Edge Computing testbeds}
-
-\todo[inline]{Intro and conclusions from EdgeDroid 1.0 paper}
-
-There is increasing interest from academia and industry in novel applications such as immersive augmented reality (AR) or wearable cognitive assistance (WCA)~\cite{Chatzopoulos:Hyperion,Ha:TowardsWearableCogAssist}, also known as human-in-the-loop applications.
-These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
-Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as Deep Neural Networks (DNN)~\cite{Ha:TowardsWearableCogAssist}.
-\Cref{fig:pipeline2} depicts such a system.
-These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
-Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{Chen:AnEmpiricalStudyOfLatency}.
-
-In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
-There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{Bittmann_Edge,flinn2012cyber,Chen:AnEmpiricalStudyOfLatency,Ha:JITProvisioning}.
-However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
-A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
-What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
-There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
-Finally, in production use these applications will most likely run concurrently with others.
-How does this, together with other best-effort applications, impact the latencies perceived by the human user?
-Existing studies~\cite{Ha:TowardsWearableCogAssist, Chen:EarlyImplementation, satya2009case,Chatzopoulos:Hyperion} of this class of applications have only lightly touched upon these issues~\cite{Chen:AnEmpiricalStudyOfLatency}.
-On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{Zubaidy15,Schiessl17} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
-We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
+\begin{figure}
+    \centering%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
+    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
+    \label{fig:comparison:rtt}%
+\end{figure}%
+\begin{figure}
+    \centering%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
+    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
+    \label{fig:comparison:feedback}%
+\end{figure}%
+\begin{figure}%
+    \centering%
+    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
+    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
+    \label{fig:comparison:tasksteps}
+\end{figure}%
 
@@
 
-Benchmarking human-in-the-loop applications is hard, given their tight interaction with human users who complicate the scaling and repeatability of experiments.
-In this paper, we have presented a benchmarking suite for this type of applications, called EdgeDroid 1.0, capable of cutting out the need for users in performance evaluations.
-We achieve this by employing pre-recorded sensory input traces which we play over the network to the real application backend, employing a parameterized user model to react to feedback.
-We demonstrate its utility through a series of use case scenarios, from which we are able to extract metrics regarding latency both in regards to the application itself and the hardware stack.
-We believe the EdgeDroid 1.0 suite thus represents an important first step towards enabling inexpensive and low-complexity large-scale research on the scaling limits of this type of applications, a requirement for wide adoption of the technology.
+\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
+Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
+We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
+Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
+These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
+Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.
 
-Nonetheless, there is still future work to be done.
+In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.
 
-The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with WCA applications in order to present a more complete model in the future.
-As mentioned in \Cref{sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
-Rather, EdgeDroid 1.0 models a perfectly stoic user who is like an automaton and responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-Of course, no real human user is an automaton.
-In the future, we envision creating  many versions of EdgeDroid (i.e., EdgeDroid 2.0, EdgeDroid 3.0, etc.) that embody more human-like user models that more accurately emulate attributes such as those mentioned above.
-Experimental validation of these human-like user models via user studies will be an important part of our future work.
+For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
+This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.
 
-We are also working on expanding the benchmarking suite to also work first with other types of Wearable Cognitive Assistance, and later with other categories of human-in-the-loop applications.
-Other types of WCA we will consider in future iterations of the tool include real-time task-assistance WCA applications (such as the Ping-Pong application described in~\cite{Chen:EarlyImplementation}), which don't have a linear task model like task-guidance WCA and have tighter latency bounds and context- and information-providing WCA applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
-The latter also do not have a linear task model, but present more lax latency bounds.
+We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
+These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
+It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.
 
-\todo[inline]{Intro and conclusions from CLEAVE paper}
+We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
+One, scaling behavior for the application seems to be linear with respect to the number of clients.
+Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
+It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
+Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.
 
-The number and applications of \acp{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
-However, this rapid increase in adoption has mostly been limited to industrial contexts.
-Although \acp{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
-This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
-These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \ac{CPS} in the future.
+EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
+From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
+Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
+Here, the downlink time sometimes almost reaches the ideal bound by itself.
+A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.
 
-\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \ac{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
-This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
+Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
+The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.
 
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \acp{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
+\subsection{The EdgeDroid model of human behavior}
 
-One the one hand, related literature in \acp{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level. 
-%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
-On the other, there exist several approaches when considering experimental methodologies. 
-%A number of works concerning \acp{NCS} deal with experimental studies.
-%\acp{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
-One such approach uses setups in which the complete system is built on top of real hardware.
-This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-Conversely, other studies choose to instead use completely \emph{simulated} \ac{NCS} setups.
-The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \acp{NCS}.
-Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
-    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-\end{enumerate*}
+\todo[inline]{%
+    Introduce as follow-up to methodology discussion.
+    Natural implication of that is that we need a more detailed model of human behavior.
+    EdgeDroid 1.0 was an initial, very coarse approximation.
+    Here we present our final, polished model based on empirical data.
+}
 
-As evidenced above, experimental research in \acp{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
-This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \ac{NCS}.
-This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
+\subsubsection{Characterizing human reaction to delays in \ac{WCA}}
 
-Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
-In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\ 
-Their methodology utilizes joint knowledge of control, computation, and communication. 
-In their work various architectural elements and the corresponding delays associated with the NCS are modelled. 
-Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation. 
-This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
-Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
+\todo[inline]{PLOS paper describes the acquisition of the necessary data and insights for the models.}
 
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
-
-The rest of this paper is structured as follows.
-\cref{sec:approach} presents the design principles and implementation of the framework.
-In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \ac{CLEAVE}.
-Finally, in \cref{sec:conclusion} we conclude this work.
-
-The issue of repeatable and scalable benchmarks has been largely glossed over in \ac{NCS} literature, as existing experimental research studies tend to implement \emph{ad-hoc} solutions.
-
-In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
-We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
-\begin{enumerate}
-    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
-    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
-    Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
-    In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
-    Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
-    The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
-\end{enumerate}
-
-We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
-We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
-    \item similar setups exist in real-world industrial use
-    \item video analytics has long been proposed as a ``killer app'' for edge computing.
-\end{enumerate*}
-Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
-
-There is still, however, work to be done.
-We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \acp{NCS} to share with the community.
-At the moment, the interactions of \ac{CLEAVE} and tools such as Docker are still quite superficial.
-Our goal is to achieve a much tighter integration, e.g.\ by providing the toolkit as pre-packaged container images.
-Finally, the validity of the results obtained by the framework will have to be verified through more thorough, realistic scenarios than what we have been able to show in this work.
-In particular, we intend to perform large-scale experimentation targetting 5G cellular deployments, as this technology is set to become the backbone of edge networks in the near future.
-
-\todo[inline]{Intro and conclusions from Ainur paper}
-
-Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications.
-On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
-The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.
-
-Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the 
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
-    \item \acs*{COSMOS}
-    \item \acs*{POWDER}
-    \item \acs*{ARA}
-    \item Drexel Grid \ac{SDR}
-\end{enumerate*} testbeds.
-\acs{COSMOS}~(\emph{\acl{COSMOS}})\acused{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \acp{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
-It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
-\ac{COSMOS} relies on the \ac{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \ac{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
-
-\acs{POWDER}~(\emph{\acl{POWDER}})\acused{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
-The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
-\ac{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \ac{VM} images containing the necessary software and configurations.
-These profiles are defined through using \ac{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
-
-The \ac{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
-It includes a broad range of wireless platforms ranging from low-\ac{UHF} massive \ac{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \ac{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-\ac{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \ac{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
-This affords the \ac{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
-
-Finally, the \emph{Drexel Grid \ac{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
-Primarily intended for \ac{SDR}-centric research, it does not integrate any core, cloud or edge components.
-However, the testbed extensively employs the \ac{LXC} runtime for the deployment of both experimental code and \ac{SDR} software, which affords users great freedom when it comes to the development of experiments.
-
-Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
-Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
-Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
-This is, for instance, the case with \ac{COSMOS} and \ac{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
-These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.
-
-To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
-CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \acp{VM} containing the necessary software for experiments.
-Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
-Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \iac{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.
-
-In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
-Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
-The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
-It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \ac{FOSS}.
-It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
-
-The rest of this paper is structured as follows.
-\cref{sec:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
-This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
-Next, \cref{sec:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
-Finally, in \cref{sec:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
-
-In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
-Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
-    \item provisioning of and connection to remote cloud instances
-    \item initialization of \ac{IP} layer connectivity between hosts
-    \item collection of logs and data
-    \item deployment, scaling, and lifecycle management of containerized processes
-\end{enumerate*}.
-We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
-We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
-We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
-
-In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \ac{CHI}~\cite{keahey2020lessons}.
-While we believe Ainur to be unique and valuable in its focus on fully-automated end-to-end wireless edge computing experimentation, integrating with the above software stacks will provide a number of features and functionalities which will greatly expand Ainur's potential.
-Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \ac{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
-Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\ac{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.
-
-\todo[inline]{Intro and conclusions from PLOS paper}
-
-Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
-They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
-Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
-Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
-An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
-
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
-Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
-Detethering this assistance from its current fixed location will surely make it available to many more fields.
-
-Based on these use cases, we identify three main requirements for WCA:\@
-\begin{enumerate}
-    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
-
-    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
-    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
-
-    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
-    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
-    This means that users will have expectations of constant, immediate feedback as they progress through the task.
-    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
-    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
-\end{enumerate}
-
-\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
-\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
-Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
-However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
-\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.
-
-% However, additional challenges remain in the path toward widespread adoption.
-\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
 We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
@@
 A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
-% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.
 
@@
 
-The structure of this paper is as follows.
-We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
-\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
-Then, in \cref{sec:results} we detail the results of our experiment.
-Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
+We start our discussion with the main results of our experimentation presented in the previous section:
 
-In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
-Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.
+\begin{itemize}
+  \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional} behavioral slow-down.
+  That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their execution of the task.
+  This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
+  Moreover, there is a temporal component to this effect; users become progressively slower the more time passes with reduced system responsiveness.
 
-We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
-The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
-When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
-This was evidenced by an increase in task execution times, even after accounting for the artificially introduced delays, as well as accelerometer data from participants wrists.
-Additionally, we found that the strength of this effect is modulated by individual differences between subjects.
+  \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
+  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
+  The question of whether any lingering effect can be measured after these four steps remains open.
 
-These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
-In this context, we believe there are two direct and important next steps to be performed.
-The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
-We believe these need to be implemented, tested and validated in real setups.
-In particular, we identify two of these implications to be prime candidates for their own experimental studies.
-One, our postulation that in cases where an impaired application is close to finishing, diverting resources to it might not be the most optimal course of action; and two, the possibility that ``fair'' degradation of system responsiveness across applications may be, in some cases, undesirable.
-Both of these questions could be answered with straightforward setups.
+  \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at performing steps as the task progresses.
+  However, the strength of this effect decreases as delay increases.
+  Whereas for blocks without delay users performed the last four steps of a 12-step block on average \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.
 
-The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
-These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
-However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
-Incorporating the results presented here would allow for much more realistic workloads.
-For instance, our findings relating to the effects of delay on execution times could be directly adapted to modulate timings in the input stream.
-Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
-Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.
+  \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay. 
+  The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
+  Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables related to attentiveness, respectively.
+  In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
+  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
+  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
+\end{itemize}
 
-\todo[inline]{Intro and conclusions from EdgeDroid 2.0}
+A central question therefore arises: to which physio- and psychological mechanisms can these findings, most importantly the substantial slow-down in task execution, be attributed?
 
-\acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
-These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
-\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
-In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
+In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional reactions.
+These could in turn elicit generalized arousal.
+We also postulated that adapting to delay might progressively deplete cognitive resources in users.
+However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
+Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+The acceleration data further did not indicate that extended delay significantly increases erratic movement.
+% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+% The acceleration data further do not indicate that extended delay increases erratic movement.
+To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
+That is, users fail to capitalize on the new information as quickly as they could.
+Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.
+We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
+As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
+Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
+This also provides a possible explanation for the lingering effects of delay after an acceptable system responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.
 
-A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses.
-They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
-This context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video feeds, which require intensive compute capabilities to process.
-This also translates into latency-sensitivity, as with any \ac{AR} application.
-Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
-On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.
+As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
+However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.
 
-The combination of these opposing characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
-However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
-One key gap in knowledge pertains to the current lack of tools and methodologies for scalable and repeatable study of \ac{WCA} application performance and resource utilization in realistic deployments.
-Due to their human-in-the-loop nature, these applications present a challenge to benchmark and characterize, in particular in real-scale deployments where dozens or even hundreds of users might concurrently use the system.
-Recruiting a large enough cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers.
-Furthermore, humans are unpredictable, and human-subject research is thus often hard to repeat and replicate.
-There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do away with the human-the-loop.
+The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.
 
-\medskip
+\begin{itemize}
+
+  \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to significantly extended application lifetimes.
+  This has direct implications for resource and power consumption.
+
+  \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a high-responsive state could have unconventional consequences for resource allocation.
+  This is of particular importance, for instance, for cases where the user may be able to finish the task before these effects subside.
+  In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired application.
+
+  \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of system responsiveness across applications may not ultimately be beneficial to the system as a whole.
+  Take for instance two applications on the same system which negatively interfere with each other. 
+  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop. 
+  In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.
+
+  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
+  This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.
+  However, this remains an open challenge.
+
+\end{itemize}
+
+To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of WCA deployments.
+Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable tool to tackle inefficiencies in these systems.
+Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the relationship between application responsiveness and human behavior.
+More research in this area will surely uncover more complex and interesting behaviors.
+Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
+These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
+
+\subsubsection{Implementing the model}
 
@@
 
-\medskip
-
-This paper is structured as follows.
-\cref{sec:relwork} discusses related work in the field of \acl{WCA} modeling.
-In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
-In \cref{sec:model} we detail our model for the generation of realistic human timings in \ac{WCA}.
-We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
-Next in \cref{sec:implications:footprint} we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
-This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \ac{WCA} systems.
-In this section, we also introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
-Finally, in \cref{sec:conclusion} we summarize and conclude this paper, as well as briefly discuss potential avenues for future work.
-
 The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic human-in-the-loop nature.
@@
 Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+
+\subsubsection{Implications of the EdgeDroid model for \ac{WCA} optimization}
 
@@
 
-This work serves as an important, but yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
-Many directions remain open to additional exploration in this space.
-For instance, our current model only targets a particular class of step-based \ac{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \ac{AR} and \ac{XR}, is on our roadmap.
-Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the US.\@
-An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.
+\section{A methodology for the study of latency trade-offs in \acsp{NCS}}
 
-In terms of the optimization framework, our current approach builds upon the assumption that execution times are Rayleigh-distributed.
-Although this distribution fits the data well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
-Our future efforts consider developing this framework towards these more accurate distributions.
-Finally, our current approach only considers a single client, which will be far from the case in a real-world \ac{WCA} deployment.
-As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.
+\subsection{The CLEAVE emulation framework}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
+    \caption{
+        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
+    }\label{fig:cleave:ncs:struct}
+\end{figure*}
+
+We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
+In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
+As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
+``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
+While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
+
+Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
+It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
+The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
+These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
+
+\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
+It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+
+In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
+We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
+\begin{enumerate}
+    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
+    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
+    Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
+    In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
+    Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
+    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
+    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
+    The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
+\end{enumerate}
+
+We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
+We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
+    \item similar setups exist in real-world industrial use
+    \item video analytics has long been proposed as a ``killer app'' for edge computing.
+\end{enumerate*}
+Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
+We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+
+\section{The Ainur software framework for Edge Computing testbeds}
+
+In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
+Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
+The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
+It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
+Furthermore, we make Ainur available to the community as \ac{FOSS}.
+It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
+
+In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
+Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
+    \item provisioning of and connection to remote cloud instances
+    \item initialization of \ac{IP} layer connectivity between hosts
+    \item collection of logs and data
+    \item deployment, scaling, and lifecycle management of containerized processes
+\end{enumerate*}.
+We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
+We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
+We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
+It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\

### 2022-11-22 body/part1/overview.tex
edits

@@
 
-\section{Methodology}
+\todo[inline]{
+    Methodology
 
-\todo[inline]{
     Study by example, human subject research, and modeling.
@@
     Synthesize and summarize the research questions and conclusions to find the red thread.
-
-    EdgeDroid (first demo paper at SEC, then HotMobile paper).
-
-    Tie in CLEAVE paper here as an extension of the approach towards other latency-sensitive applications.
-
-    Potentially include Ainur paper as a further extension of the methodology?
-
-    PLOS ONE paper as a stepping stone towards the final goal of a more realistic model?
-
-    Finally, EdgeDroid2 paper:
-    Presents the final model of human behavior
-    Applies it to a real optimization problem, employing theoretical results proved by Vishnu
 }
 
-We summarize the seven key contributions of this thesis below:
-
-\begin{enumerate}
+\begin{contributions}
+    \begin{enumerate}[
+            before = \leavevmode%\vspace{-.5\baselineskip}
+        ]
         \item\label{item:contrib:methodology} A methodological approach to studying system responsiveness versus resource consumption trade-offs in \ac{WCA}.
@@
     \end{enumerate}
+\end{contributions}
 
@@
 \todo[inline]{
-    \cref{paper:EdgeDroid1.0} introduces our methodology, and validates its use in studying WCA.
+    \cref{paper:Olguin:DemoScaling2018,paper:Olguin:EdgeDroid2019} introduce our methodology, and validate its use in studying WCA.
     Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
@@
 
-\todo[inline]{PLOS paper describes the acquisition of the necessary data and insights for the models.}
+\todo[inline]{\cref{paper:Olguin:ImpactWCA2021} describes the acquisition of the necessary data and insights for the models.}
 
-We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
-\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
-% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
-Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}.
-Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
-A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
+\todo[inline]{From introduction}
 
-This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
-We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
-While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
-We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
-This is an important step toward a more systematic understanding of human behavior in this domain.
-
-We present in this paper the design and elaboration of an experimental WCA test-bed.
-This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
-Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
-The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
-
-Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
+We intended to answer four core research questions relating to human responses to decreased application responsiveness.
 
@@
           The results we present, however, seem to refute this hypothesis.
-  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.
+          We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as system responsiveness was altered.
 
@@
 Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
+
+\todo[inline]{From discussion}
 
@@
 
+\todo[inline]{Some diagrams, plots, and results}
+
+\begin{figure}[h]
+    \centering
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
+        \caption{Structure of a step in a generic cognitive assistance application.
+            The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
+            % by the assistant, as they do not contain relevant information.
+            Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
+            % which will subsequently be provided to the user.}%
+        }
+        \label{fig:cogassist:step}
+    \end{subfigure}%
+    % \medskip%
+    \hfill%
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
+        \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
+        \label{fig:cogassist:step:delay}
+    \end{subfigure}
+    \medskip%
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
+        \caption{Structure of a block in the experimental task.}%
+        \label{fig:cogassist:block}
+    \end{subfigure}%
+    % \medskip%
+    \hfill%
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
+        \caption{Visualization of the execution time of a step.}%
+        \label{fig:exectime:diagram}
+    \end{subfigure}%
+    % \medskip%
+    \caption{Components of the cognitive assistance task.}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
+    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
+    \caption{Mean per-step execution time vs.\ delay, by step slice.
+        Error bars indicate S.E.M.}
+    \label{fig:exectime:delay:slice}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
+    \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
+\end{figure}
+
+\begin{table}[h]
+    \centering
+    \caption{Principal Component Analysis}\label{tab:pca}
+    \begin{subtable}[h]{\textwidth}
+        \centering
+        \caption{Main components identified.}
+        \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
+        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
+            \toprule
+            \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} \\
+            \midrule
+            BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
+            BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
+            ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
+            ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
+            Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
+            Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
+            \bottomrule
+        \end{tabular*}
+    \end{subtable}
+    \newline
+    \medskip
+    \newline
+    \begin{subtable}[h]{\textwidth}
+        \centering
+        \caption{Percentages of variance explained by the components.}
+        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
+            \toprule
+            {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total} \\
+            \midrule
+            Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent} & \SI{73.13}{\percent} \\
+            \bottomrule
+        \end{tabular*}
+    \end{subtable}
+\end{table}
+
+\FloatBarrier%
 \subsubsection{Implementing the model}
 
-In that context, the contributions of this paper are as follows:
-\begin{enumerate}
-    \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+\todo[inline]{Present the implementation of the model from \cref{paper:EdgeDroid2}}
+
+We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
 Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
@@
 We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
-    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
-    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
-    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
-\end{enumerate}
 
@@
 
-\subsubsection{Implications of the EdgeDroid model for \ac{WCA} optimization}
+\todo[inline]{TODO: Some relevant results. Need to add paper to thesis.}
+
+\subsection{Implications of the methodology for \ac{WCA} optimization}
+
+\todo[inline]{TODO, same as above}
+\todo[inline]{Include discussion from EdgeDroid 1.0?}
+
+\begin{enumerate}
+    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
+    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
+    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
+\end{enumerate}
 
@@
 
+\FloatBarrier%
 \section{A methodology for the study of latency trade-offs in \acsp{NCS}}
+
+\todo[inline]{\Cref{paper:CLEAVE2022} implements both the methodology and the tool.}
 
@@
 
-\section{The Ainur software framework for Edge Computing testbeds}
+\todo[inline]{Some plots and results}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
+    \caption{
+        The 2D inverted pendulum system.
+    }\label{fig:invpend}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
+        \caption{Toppled plants.}\label{fig:video:toppled}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
+        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
+    \end{subfigure}\\
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
+        \caption{Packet losses.}\label{fig:video:drop}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
+        \caption{\acsp{RTT}.}\label{fig:video:rtt}
+    \end{subfigure}%
+    \caption{
+        Multi-loop, resource-constrained setup results.
+        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
+    }\label{fig:video:results}
+\end{figure}
+
+\FloatBarrier%
+\section{A software framework for Edge-computing testbeds}
+
+\todo[inline]{\cref{paper:Ainur2022} describes a software framework for testbeds used for the methodology described in this thesis}
 
@@
 It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
+
+\todo[inline]{Some plots}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
+    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
+\end{figure}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
+    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
+    \caption{Network structure assumed by Ainur.}\label{fig:network}
+\end{figure*}

### 2022-11-26 body/part1/overview.tex
scope

@@
 \todo[inline]{
-    Methodology
-
-    Study by example, human subject research, and modeling.
-
-    Empirical research.
-
-    Include testbeds!
+    Methodology: a system of methods used in a particular area of study or activity.
 }
 
-\section{Contributions of this thesis}
+\section{Scope of this thesis}
+
+We study in this thesis a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \ac{WCA} and \ac{NCS}.
+This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
+We validate this methodology by example, presenting a deep exploration of its implementation and applications in the context of \acl{WCA}, and a initial, more shallow foray into its applications in the context of \ac{NCS}.
+
+More specifically, in \cref{paper:Olguin:DemoScaling2018,paper:Olguin:EdgeDroid2019} we introduce this methodology for \ac{WCA}.
+\cref{paper:Olguin:DemoScaling2018} presents a short, high level overview of this methodological approach, in particular making very straightforward assumptions about human behavior in \ac{WCA} applications.
+\cref{paper:Olguin:EdgeDroid2019} extends the discussion from \cref{paper:Olguin:DemoScaling2018}, providing more details on the implementation of the necessary measurement framework and the tooling employed, as well as on the assumptions that were made about human behavior and reactions.
+We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+This results in a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+\cref{paper:Olguin:EdgeDroid2019} furthermore presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
+
+In that context, we present several ancillary sub-contributions.
+The first of these corresponds to  a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
+This tool corresponds to a key enabler of our main contribution, as it implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
+Furthermore, as part of the ground-laying work for this model, we present a first-ever characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
+In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
 
@@
         ]
-        \item\label{item:contrib:methodology} A methodological approach to studying system responsiveness versus resource consumption trade-offs in \ac{WCA}.
-        This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-        \item In conjunction with the aforementioned methodology, a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
-              This tool implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
-        \item A characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
-              In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
+        \item\label{item:contrib:methodology} 
+        \item In conjunction with the aforementioned methodology,
+        \item A 
         \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.

### 2022-11-27 body/part1/overview.tex
reworking references, overview

@@
 
-More specifically, in \cref{paper:Olguin:DemoScaling2018,paper:Olguin:EdgeDroid2019} we introduce this methodology for \ac{WCA}.
-\cref{paper:Olguin:DemoScaling2018} presents a short, high level overview of this methodological approach, in particular making very straightforward assumptions about human behavior in \ac{WCA} applications.
-\cref{paper:Olguin:EdgeDroid2019} extends the discussion from \cref{paper:Olguin:DemoScaling2018}, providing more details on the implementation of the necessary measurement framework and the tooling employed, as well as on the assumptions that were made about human behavior and reactions.
+More specifically, in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we introduce this methodology for \ac{WCA}.
+\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach, in particular making very straightforward assumptions about human behavior in \ac{WCA} applications.
+\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed, as well as on the assumptions that were made about human behavior and reactions.
 We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
 This results in a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-\cref{paper:Olguin:EdgeDroid2019} furthermore presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
+\cref{paper:olguinmunoz2019edgedroid} furthermore presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
 
-In that context, we present several ancillary sub-contributions.
-The first of these corresponds to  a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
-This tool corresponds to a key enabler of our main contribution, as it implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
-Furthermore, as part of the ground-laying work for this model, we present a first-ever characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
-In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
+Next, in \cref{paper:olguinmunoz2021impact} we present a deep characterization of human behavior in \ac{WCA}.
+This represents a critical prerequisite for an extension of the human model presented in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} towards a more advanced realistic iteration.
+We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
+System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
+Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
 
-\todo[inline]{
-    Here's where we ``present'' the papers.
-    However, we do this by presenting a ``red thread'' of contributions and key results.
-    Drive towards one big picture.
+\cref{paper:olguinmunoz2023realistic} then builds upon our work in \cref{paper:olguinmunoz2021impact}.
+We employ the insights obtained on human reactions to changes in system responsiveness to develop the first ever data-driven model of human timings for \ac{WCA}.
+The model is further validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+Finally, the paper explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
 
-    Add a hardened compilation of key results, conclusions, figures, and tables.
-    Synthesize and summarize the research questions and conclusions to find the red thread.
-}
+On the other hand, \cref{paper:olguinmunoz2022cleave} presents an extension to the methodology introduced and developed in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} to the field of \aclp{NCS}.
+We implement a tool for the emulation of these systems on edge computing infrastructure, and deploy it on an edge computing testbed for validation.
 
-\begin{contributions}
-    \begin{enumerate}[
-            before = \leavevmode%\vspace{-.5\baselineskip}
-        ]
-        \item\label{item:contrib:methodology} 
-        \item In conjunction with the aforementioned methodology,
-        \item A 
-        \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.
-              \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \acp{NCS}.
-        \item A novel tool, CLEAVE, for the emulation of \acp{NCS} on edge infrastructure.
-              This tool has been designed from the ground-up to be edge-native, implemented using widely-adopted technologies in the disciplines of cloud and edge computing.
-        \item Finally, a software framework for the management and automation of the testbeds required for the methodology described in \cref{item:contrib:methodology,item:contrib:methodology:ncs}.
-    \end{enumerate}
-\end{contributions}
+Finally, \cref{paper:olguinmunoz2022ainur} presents the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
+This framework represents an ancillary contribution, tangential to yet crucial for the research presented in these works.
 
-\section{A methodology for the study of latency trade-offs in \acs{WCA}}
+\medskip%
 
-\todo[inline]{
-    \cref{paper:Olguin:DemoScaling2018,paper:Olguin:EdgeDroid2019} introduce our methodology, and validate its use in studying WCA.
-    Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
-}
+In the following sections, we will summarize in detail the key contributions of each of the above-mentioned works.
+\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} are discussed in detail in \cref{summary:2018demo,summary:2019edgedroid}.
+Next, in \cref{summary:model} we present \cref{paper:olguinmunoz2021impact} (\cref{summary:2021impact}) and \cref{paper:olguinmunoz2023realistic} (\cref{summary:2023realistic}).
+\cref{paper:olguinmunoz2022cleave} is summarized in \cref{summary:2022cleave}, and we finalize this chapter with a summary of \cref{paper:olguinmunoz2022ainur} in \cref{summary:2022ainur}.
+
+\section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}
+
+\subsection{\cref{paper:olguinmunoz2018demoscaling}}\label{summary:2018demo}
+
+As discussed previously, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
+This short papers' main contribution is the introduction of our methodological framework for tackling this challenge.
+
+\subsection{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
 
@@
 
-\subsection{The EdgeDroid model of human behavior}
+\section{A realistic model of human behavior in \acs{WCA}}\label{summary:model}
 
-\todo[inline]{%
-    Introduce as follow-up to methodology discussion.
-    Natural implication of that is that we need a more detailed model of human behavior.
-    EdgeDroid 1.0 was an initial, very coarse approximation.
-    Here we present our final, polished model based on empirical data.
-}
+\subsection{\cref{paper:olguinmunoz2021impact}}\label{summary:2021impact}
 
-\subsubsection{Characterizing human reaction to delays in \ac{WCA}}
-
-\todo[inline]{\cref{paper:Olguin:ImpactWCA2021} describes the acquisition of the necessary data and insights for the models.}
+\todo[inline]{\cref{paper:olguinmunoz2021impact} describes the acquisition of the necessary data and insights for the models.}
 
@@
 \FloatBarrier%
-\subsubsection{Implementing the model}
+\subsection{\cref{paper:olguinmunoz2023realistic}}\label{summary:2023realistic}
 
@@
 
-\subsection{Implications of the methodology for \ac{WCA} optimization}
-
-\todo[inline]{TODO, same as above}
-\todo[inline]{Include discussion from EdgeDroid 1.0?}
+\subsubsection{Implications of the methodology for \ac{WCA} optimization}
 
@@
 \FloatBarrier%
-\section{A methodology for the study of latency trade-offs in \acsp{NCS}}
+\section{Extending the methodology to \acsp{NCS}}
 
-\todo[inline]{\Cref{paper:CLEAVE2022} implements both the methodology and the tool.}
+\subsection{\cref{paper:olguinmunoz2022cleave}}\label{summary:2022cleave}
+\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
 
@@
 \FloatBarrier%
-\section{A software framework for Edge-computing testbeds}
+\section{Orchestrating edge computing testbeds}
 
-\todo[inline]{\cref{paper:Ainur2022} describes a software framework for testbeds used for the methodology described in this thesis}
+\subsection{\cref{paper:olguinmunoz2022ainur}}\label{summary:2022ainur}
+
+\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
 

### 2022-11-30 body/part1/overview.tex
scope

@@
 More specifically, in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we introduce this methodology for \ac{WCA}.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach, in particular making very straightforward assumptions about human behavior in \ac{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed, as well as on the assumptions that were made about human behavior and reactions.
+\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach, making very straightforward assumptions about human behavior in \ac{WCA} applications.
+\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
+This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
+Finally, \cref{paper:olguinmunoz2019edgedroid} discusses in detail the assumptions that were made about human behavior and reactions.
 We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-This results in a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-\cref{paper:olguinmunoz2019edgedroid} furthermore presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
+The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
 
-Next, in \cref{paper:olguinmunoz2021impact} we present a deep characterization of human behavior in \ac{WCA}.
-This represents a critical prerequisite for an extension of the human model presented in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} towards a more advanced realistic iteration.
+Although the above approach to human behavior in \ac{WCA} represents a useful initial approximation, it is nonetheless not a realistic model of it.
+In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
 We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
@@
 
-\cref{paper:olguinmunoz2023realistic} then builds upon our work in \cref{paper:olguinmunoz2021impact}.
-We employ the insights obtained on human reactions to changes in system responsiveness to develop the first ever data-driven model of human timings for \ac{WCA}.
-The model is further validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-Finally, the paper explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \ac{WCA}.
+The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+This work also explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
 
@@
 We implement a tool for the emulation of these systems on edge computing infrastructure, and deploy it on an edge computing testbed for validation.
+In particular, this work showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.
 
@@
 This framework represents an ancillary contribution, tangential to yet crucial for the research presented in these works.
-
-\medskip%
+Without it, the experimental approach described in these works would not have been feasible.
 
@@
 This short papers' main contribution is the introduction of our methodological framework for tackling this challenge.
+
+As previously explained, our methodology is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
+Two main arguments drive this approach.
+One, it maintains the realism and accuracy of the system.
+This is particularly important in wireless and mobile edge computing systems, which present complex behaviors and interactions intrinsic to their networking technologies.
+And two, it allows for ease of scaling.
+The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \ac{WCA}) or due its cyber-physical nature (in \acp{NCS}).
+Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
+Raspberry Pis, Beaglebones, etc.).
+
+This work presents the first approach to this concept for \ac{WCA}.
+To establish a reproducible and comparable workload, we use a trace-based design for the client-side emulation.
+We first record a trace of the sensory input data while having a human user operate the target application.
+The collected data consists of the sensory inputs provided to the system at runtime, for instance --- in the case of a video-based application --- the video feed from the camera.
+
+To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
+This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
+Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
+
+\begin{figure}
+    % Define block styles
+    %         \tikzstyle{block} = [circle, draw, fill=red!20, 
+    %         text width=5em, text centered, rounded corners, minimum size=4em]
+    %         \tikzstyle{decision} = [diamond, aspect=2, draw, fill=yellow!20,
+    %         text width=5em, text centered, rounded corners, minimum size=4em]
+    %        \tikzstyle{line} = [draw, -{Latex[length=3mm]}]
+    \centering
+    \adjustbox{scale=0.75}{
+        \begin{tikzpicture}[align=center,
+        node distance=.5cm and 1.5cm,
+        every initial by arrow/.style={-{Latex[length=2mm]}}]
+        % Place nodes              
+        \node [initial, state, minimum size=6em, initial text=] (play) {Play};               
+        \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
+        \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
+        \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};
+
+        % Draw edges
+        \path[draw, -{Latex[length=2mm]}]
+        (play) edge [bend right=20] node[left] {Step done} (rewind)
+        edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
+        edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)
+
+        (change) edge [bend left=20] node[right] {Step changed} (play)
+        edge [bend left=20] node[right] {All steps done} (shutdown)
+
+        (rewind) edge [bend right=20] node[right] {Rewound} (play)
+        edge [bend right=20] node[right] {Too many rewinds} (shutdown);
+
+        \end{tikzpicture}
+    }
+    \caption{State diagram of the user model}
+    \label{fig:usermodel}
+\end{figure}
+
+However, merely replaying the trace is not enough, as disparities between system responsiveness at recording versus replay time can cause the trace so drift out of synchronization with the current state of the system.
+Thus, we introduce here the concept of a human user model which is able to adapt its behavior in a realistic manner based on the current responsiveness of the system by systematically controlling the replay of the trace.
+The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
+The trace is preprocessed into segments that correspond to individual steps of the \ac{WCA} application; this is performed manually as a preparation step.
+The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \ac{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
+On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
+To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.
+
+\todo[inline]{EdgeDroid paper}
 

### 2022-12-01 appendices.tex
tweaking
Update on Overleaf.

@@
 
-In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in section \cref{ssec:optframework}.
+In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in \cref{ssec:optframework}.
 We start with the general solution, where the objective function to be minimized is given by \cref{eq:epsilon_terminal}:
@@
 \begin{alignat}{1}
-{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1.\label{rt}
+{\int_{t_{n-1}}^{t_n}}r(t)\dif t=1,\;\forall n\geq1\label{rt}
 \end{alignat}
@@
 Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
-Expanding the \ac{CDF} using Taylor series gives
+We expand $F_\mathcal{T}(t_{n}-t)$ and $F_\mathcal{T}(t_{n-1})$ using Taylor series. Thus we can write the conditionsl \ac{CDF} as
 \begin{multline*}
@@
 \end{multline*}
-Simplifying and using the first order approximation, we arrive at
+Simplifying and approximating by ignoring the higher order terms, we arrive at
 \begin{alignat}{1}
@@
 That is,
-If $r(t)$ is varying slowly between two consecutive sampling instants, we can approximate the sampling interval $(t_n\!-\!t_{n-1})$ as
+If $r(t)$ is varying slowly between two consecutive sampling instants due to the closeness of two sampling intervals, we can approximate the sampling interval $(t_n\!-\!t_{n-1})$ as
 \begin{alignat}{1}
@@
 \Rightarrow &\;t_{n+1}^{\frac{3}{2}}-t_{n}^{\frac{3}{2}}=3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\nonumber\\
-\Rightarrow &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tnRayleigh}.
+\intertext{We have $t_0=0$. Substituting $n=1,2,\dots$, in order, in the above equation provides us our final result}
+&\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tnRayleigh}.
 \end{alignat}
+% Note that throughout the paper we use a exGaussian distribution but not Rayleigh. This is because, 
+Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an exponentially modified Gaussian distribution as well as a Rayleigh distribution, with the former attracting more attention from works like \cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+This is the reason why the above results are applicable in this work where we have predominantly considered exGaussian distribution.
+Furthermore, we have also verified the closeness of the results as well as the validity of the approximations made in the proofs using distribution fitting and simulations.
 
-By making use of this general solution, we can also prove the results given in section \cref{ssec:optimization:samples}, where we modify the problem and find the optimum sampling instants that minimize the expected number of samples for a given upper bound $w_0$ for the expected wait time.
+By making use of this general solution, we can also prove the results given in \cref{ssec:optimization:samples}, where we modify the problem and find the optimum sampling instants that minimize the expected number of samples for a given upper bound $w_0$ for the expected wait time.
 First note that the general solution has constants $\alpha$ and $\beta$ that correspond to the weights given to the cost of sampling and cost of waiting, respectively.
@@
 Furthermore, since the sampling instants are aperiodic and can take any positive real values, the bound will be tight at the optimum. 
-Hence, to solve for the modified optimization problem explained in section \cref{ssec:optimization:samples} that minimizes $(\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnRayleigh}.
+% This is because (1) a higher bound on $\mathbb{\mathcal{W}}$ is always associated with a lower optimum on $\mathbb{\mathcal{S}}$ and (2) any $\mathbb{\mathcal{W}}$ can be achieved, for instance by a scaling all the sampling instance of a given policy by an appropriate constant.
+Hence, to solve for the modified optimization problem explained in \cref{ssec:optimization:samples} that minimizes $\mathbb{E}[\mathcal{S}]$ with an upper bound $w_0$ on $\mathbb{E}[\mathcal{W}]$, we equate \cref{Ew} to $w_0$, find the corresponding $\frac{\alpha}{\beta}$, and find the optimum set of sampling instants by plugging this ratio into \cref{tnRayleigh}.
 \begin{alignat*}{1}
@@
 % &=1.88355\frac{w_0^2}{\sigma}
-&\approx1.9\frac{w_0^2}{\sigma}
+&\approx1.9\frac{w_0^2}{\sigma}.
 \end{alignat*}

### 2022-12-02 body/part1/background.tex
working

--- 2022-11-18 body/part1/introduction.tex
@@
-\chapter{Introduction}
-
-\section{Background \& Related Work}
+\chapter{Background \& Related Work}
 
@@
 
-\subsection{Theoretical background}
+\section{Theoretical background}
 \textcite{Satya2009Case}.
@@
 Main points: prototypes and characterization of system latencies and load.
-
-\section{Presentation of research questions}
-
-How will human-in-the-loop/\ac{WCA} systems look in production?
-
-How will they look in large scale deployments?
-
-How many applications can given systems support?
-
-How can these dimensions be studied?
-Here's where modeling and empirical data-driven research comes in.
-
-Is there potential for optimization?
-Yes, trading of system responsiveness for decrease in system load.
-Underlying principle under aperiodic, adaptive sampling.
-The potential for optimization is novel and stems from \textcite{Olguin:ImpactWCA2021}.

### 2022-12-02 body/part1/introduction.tex
struct
working

@@
-\chapter{Introduction}
+\chapter{Introduction}\label{chap:introduction}
 
-\section{Background \& Related Work}
+\todo[inline]{
+    Introduction here. Why do we need this work?
+}
 
-Deep focus on theoretical background.
+\todo[inline]{
+    Reorganize EdgeDroid, Cleave, Ainur, together.
 
-\subsection{Theoretical background}
-\textcite{Satya2009Case}.
-Other papers detailing the advantages of Edge computing.
+    Pull in multi-loop results for EdgeDroid?
 
-Next, what applications benefit from the Edge?
-\ac{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
-Main points: prototypes and characterization of system latencies and load.
+    Think about experiments.
+}
 
-\section{Presentation of research questions}
+\section{Summary of the key contributions of this thesis}
 
-How will human-in-the-loop/\ac{WCA} systems look in production?
+This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
+Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \ac{WCA} and \ac{NCS}.
+This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
+We validate this methodology by example, presenting implementations of its application in the context of \ac{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
 
-How will they look in large scale deployments?
+Secondly, we present a deeper exploration of this methodology for \ac{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \ac{WCA}.
+This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
 
-How many applications can given systems support?
+Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \ac{WCA} deployments.
+In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
+We conclude that significant improvements can be achieved through the use of our human user model.
 
-How can these dimensions be studied?
-Here's where modeling and empirical data-driven research comes in.
+\subsection{Overview of each included paper}
 
-Is there potential for optimization?
-Yes, trading of system responsiveness for decrease in system load.
-Underlying principle under aperiodic, adaptive sampling.
-The potential for optimization is novel and stems from \textcite{Olguin:ImpactWCA2021}.
+\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
+\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \ac{WCA}, making very straightforward assumptions about human behavior in \ac{WCA} applications.
+\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
+This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
+\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
+We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+
+\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \acp{NCS}.
+As discussed above, these applications are similarly difficult as \ac{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
+\todo[inline]{More argumentation or refer to previous section.}
+We apply thus in this work our methodology to \ac{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
+The software acts as a middleware which abstracts away the network from the development of these workloads, allowing for quick prototyping and deployment.
+This work also showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.
+
+\cref{paper:olguinmunoz2022ainur} presents a tangential contribution.
+It introduces the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
+This framework represents an ancillary contribution, crucial for the research presented in these works.
+Without it, the experimental approach described in these works would not have been feasible.
+
+Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \ac{WCA}, as well as study its implications for optimization of these systems.
+Although the above approach to human behavior in \ac{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
+In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
+We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
+System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
+Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
+
+\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \ac{WCA}.
+The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+This work also explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+
+\section{Structure of this thesis}
+
+\todo[inline]{This thesis is structured as follows.}

### 2022-12-02 body/part1/overview.tex
working

@@
-\chapter{Overview of this thesis' contributions}
+\chapter{Key contributions and results}\label{chap:contributions}
 
-\todo[inline]{
-    Methodology: a system of methods used in a particular area of study or activity.
-}
-
-\section{Scope of this thesis}
-
-We study in this thesis a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \ac{WCA} and \ac{NCS}.
-This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting a deep exploration of its implementation and applications in the context of \acl{WCA}, and a initial, more shallow foray into its applications in the context of \ac{NCS}.
-
-More specifically, in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we introduce this methodology for \ac{WCA}.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach, making very straightforward assumptions about human behavior in \ac{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
-This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
-Finally, \cref{paper:olguinmunoz2019edgedroid} discusses in detail the assumptions that were made about human behavior and reactions.
-We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-
-Although the above approach to human behavior in \ac{WCA} represents a useful initial approximation, it is nonetheless not a realistic model of it.
-In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \ac{WCA}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
-
-On the other hand, \cref{paper:olguinmunoz2022cleave} presents an extension to the methodology introduced and developed in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} to the field of \aclp{NCS}.
-We implement a tool for the emulation of these systems on edge computing infrastructure, and deploy it on an edge computing testbed for validation.
-In particular, this work showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.
-
-Finally, \cref{paper:olguinmunoz2022ainur} presents the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
-This framework represents an ancillary contribution, tangential to yet crucial for the research presented in these works.
-Without it, the experimental approach described in these works would not have been feasible.
-
-In the following sections, we will summarize in detail the key contributions of each of the above-mentioned works.
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} are discussed in detail in \cref{summary:2018demo,summary:2019edgedroid}.
-Next, in \cref{summary:model} we present \cref{paper:olguinmunoz2021impact} (\cref{summary:2021impact}) and \cref{paper:olguinmunoz2023realistic} (\cref{summary:2023realistic}).
-\cref{paper:olguinmunoz2022cleave} is summarized in \cref{summary:2022cleave}, and we finalize this chapter with a summary of \cref{paper:olguinmunoz2022ainur} in \cref{summary:2022ainur}.
+\todo[inline]{Reorganize}
 

### 2022-12-03 body/part1/overview.tex
restructuring again

@@
 
-\subsection{\cref{paper:olguinmunoz2018demoscaling}}\label{summary:2018demo}
+\subsection{Applying the methodology to \acs{WCA}}
+
+\todo[inline]{\cref{paper:olguinmunoz2018demoscaling}}\label{summary:2018demo}
 
@@
 
-\subsection{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
-
 \begin{figure}
@@
 
-\section{A realistic model of human behavior in \acs{WCA}}\label{summary:model}
+\todo[inline]{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
 
-\subsection{\cref{paper:olguinmunoz2021impact}}\label{summary:2021impact}
+\subsection{Applying the methodology to \acsp{NCS}}
+\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
+    \caption{
+        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
+    }\label{fig:cleave:ncs:struct}
+\end{figure*}
+
+We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
+In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
+As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
+``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
+While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
+
+Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
+It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
+The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
+These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
+
+\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
+It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+
+In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
+We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
+\begin{enumerate}
+    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
+          The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
+          Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
+          In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
+          Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
+    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
+    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
+          The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
+\end{enumerate}
+
+We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
+We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
+    \item similar setups exist in real-world industrial use
+    \item video analytics has long been proposed as a ``killer app'' for edge computing.
+\end{enumerate*}
+Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
+We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+
+\todo[inline]{Some plots and results}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
+    \caption{
+        The 2D inverted pendulum system.
+    }\label{fig:invpend}
+\end{figure}
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
+        \caption{Toppled plants.}\label{fig:video:toppled}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
+        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
+    \end{subfigure}\\
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
+        \caption{Packet losses.}\label{fig:video:drop}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[h]{.22\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
+        \caption{\acsp{RTT}.}\label{fig:video:rtt}
+    \end{subfigure}%
+    \caption{
+        Multi-loop, resource-constrained setup results.
+        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
+    }\label{fig:video:results}
+\end{figure}
+
+\FloatBarrier%
+
+\subsection{Orchestrating experimentation in edge computing testbeds}
+
+\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
+
+In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
+Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
+The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
+It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
+Furthermore, we make Ainur available to the community as \ac{FOSS}.
+It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
+
+In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
+Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
+    \item provisioning of and connection to remote cloud instances
+    \item initialization of \ac{IP} layer connectivity between hosts
+    \item collection of logs and data
+    \item deployment, scaling, and lifecycle management of containerized processes
+\end{enumerate*}.
+We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
+We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
+We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
+It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
+
+\todo[inline]{Some plots}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
+    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
+\end{figure}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
+    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
+\end{figure*}
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
+    \caption{Network structure assumed by Ainur.}\label{fig:network}
+\end{figure*}
+
+\section{Extending the methodology in the context of \acs{WCA}}
+
+\subsection{Characterizing human behavior}
 
@@
 
-\FloatBarrier%
-\subsection{\cref{paper:olguinmunoz2023realistic}}\label{summary:2023realistic}
-
-\todo[inline]{Present the implementation of the model from \cref{paper:EdgeDroid2}}
+\subsection{Emulating human behavior}
 
@@
 
-\subsubsection{Implications of the methodology for \ac{WCA} optimization}
+\section{Implications for the optimization of \ac{WCA}}
 
@@
 Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
-
-\FloatBarrier%
-\section{Extending the methodology to \acsp{NCS}}
-
-\subsection{\cref{paper:olguinmunoz2022cleave}}\label{summary:2022cleave}
-\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
-
-\subsection{The CLEAVE emulation framework}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
-
-In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
-We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
-\begin{enumerate}
-    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
-          The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
-          Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
-          In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
-          Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
-          The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
-\end{enumerate}
-
-We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
-We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
-    \item similar setups exist in real-world industrial use
-    \item video analytics has long been proposed as a ``killer app'' for edge computing.
-\end{enumerate*}
-Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
-
-\todo[inline]{Some plots and results}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
-    \caption{
-        The 2D inverted pendulum system.
-    }\label{fig:invpend}
-\end{figure}
-
-\begin{figure}[t]
-    \centering
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
-        \caption{Toppled plants.}\label{fig:video:toppled}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
-    \end{subfigure}\\
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
-        \caption{Packet losses.}\label{fig:video:drop}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\acsp{RTT}.}\label{fig:video:rtt}
-    \end{subfigure}%
-    \caption{
-        Multi-loop, resource-constrained setup results.
-        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
-    }\label{fig:video:results}
-\end{figure}
-
-\FloatBarrier%
-\section{Orchestrating edge computing testbeds}
-
-\subsection{\cref{paper:olguinmunoz2022ainur}}\label{summary:2022ainur}
-
-\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
-
-In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
-Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
-The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
-It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \ac{FOSS}.
-It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
-
-In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
-Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
-    \item provisioning of and connection to remote cloud instances
-    \item initialization of \ac{IP} layer connectivity between hosts
-    \item collection of logs and data
-    \item deployment, scaling, and lifecycle management of containerized processes
-\end{enumerate*}.
-We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
-We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
-We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
-
-\todo[inline]{Some plots}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
-    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
-    \caption{Network structure assumed by Ainur.}\label{fig:network}
-\end{figure*}

### 2022-12-04 abstract.tex
Satya & Bobby comments

@@
     \acf*{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
-    Recruiting humans is cumbersome and expensive, and human-subject research is often hard to repeat and replicate.
-    Considering the intended mass-scale deployment of these applications in the future, there exists a real need for tools enabling human-independent benchmarking.
+    Employing user testing to optimize system parameters is generally not feasible, given the scope of the problem and the number of observations needed for to detect small but important effects in controlled experiments.
+    Considering the intended mass-scale deployment of \ac{WCA} applications in the future, there exists a need for tools enabling human-independent benchmarking.
 
     We present in this paper the first data-driven model which allows for the complete end-to-end emulation of humans in \acs*{WCA}.
-    We demonstrate the utility of this model by employing it to study application lifetime footprints, where we find an up to \SI{12}{\percent} difference with respect to first order approximations.
+    We demonstrate the utility of this model by employing it to study application task durations, where we find an up to \SI{12}{\percent} difference with respect to first order approximations.
     Further, we use this model to develop a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \acs*{WCA}.

### 2022-12-04 acks.tex
Satya & Bobby comments

@@
 
-\todo[inline]{%
-    TODO: Thanks
-
-    TODO: Check if funding is correct.
-}
-
-This work has been partially funded by the \ac{SSF} (grant number \verb|ITM17-0246| (ExPECA)), and the \ac{NSF} (grant number \verb|CNS-1518865|).
+This work has been partially funded by the \ac{SSF} (grant number \verb|ITM17-0246| (ExPECA)), and the \ac{NSF} (grant number \verb|CNS-2106862|).
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

### 2022-12-04 appendices.tex
Satya & Bobby comments

@@
 
-In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval we discussed in \cref{ssec:optframework}.
+In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval discussed in \cref{ssec:optframework}.
 We start with the general solution, where the objective function to be minimized is given by \cref{eq:epsilon_terminal}:
@@
 Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
-We expand $F_\mathcal{T}(t_{n}-t)$ and $F_\mathcal{T}(t_{n-1})$ using Taylor series. Thus we can write the conditionsl \ac{CDF} as
+We expand $F_\mathcal{T}(t_{n}-t)$ and $F_\mathcal{T}(t_{n-1})$ using Taylor series. Thus we can write the \ac{CCDF} as
 \begin{multline*}
@@
 &\;t_n=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tnRayleigh}.
-\end{alignat}
+x\end{alignat}
 % Note that throughout the paper we use a exGaussian distribution but not Rayleigh. This is because, 
@@
 Since any positive real value is valid for this ratio, one can achieve any valid point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via simply by varying the ratio $\frac{\alpha}{\beta}$.
+\newpage%
 Furthermore, since the sampling instants are aperiodic and can take any positive real values, the bound will be tight at the optimum. 

### 2022-12-04 background.tex
Satya & Bobby comments

@@
-\section{A background on \acs*{WCA}}\label{sec:background}
+\section{Background of \acs*{WCA}}\label{sec:background}
 
@@
 \acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \ac{AR} applications.
-In this work we focus on a particular category --- denoted ``step-based'' \ac{WCA} --- which have as their goal the guiding of a user through a sequential task.
+In this work we focus on a particular category --- denoted ``step-based'' \ac{WCA} --- that have as their goal the guiding of a user through a sequential task.
 Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
@@
 
-In \textcite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} through human-subject trials.
+In \textcite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} in a controlled experiment.
 We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA}~\cite{Chen2015LEGO}.
 Subjects interacted with the assistant while we altered the responsiveness in realtime and captured key application and task performance metrics.
-Additionally, we also employed questionnaires to evaluate personality traits of the participants, and correlated these results to the task performance metrics collected.
+Additionally, we also employed questionnaires to evaluate personality traits of the participants, and correlated these results with the task performance metrics.
 
 We used \emph{delay} and execution time as the variables for system responsiveness and human behavior, respectively.
-\emph{Delay} corresponded to a temporarily fixed time duration of the processing of each frame was padded to --- i.e.\ if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user at exactly \( D \) seconds after frame capture.
+\emph{Delay} corresponded to a temporarily fixed time duration for the processing interval of each frame.
+That is, if during a series of steps delay was set to \( D \) seconds, the feedback for each frame was provided to the user \( D \) seconds after frame capture.
 The correlation between this variable and execution time was then studied.
-Delay, however, was merely an experimental tool, and if we consider
-\begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
-    \item that \( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality~\cite{olguinmunoz:impact2021}
-    \item that for a step subject to delay \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
-\end{enumerate*},
-then it follows that the expected \ac{TTF} for steps subject to a delay \( D \) can be expressed as
+To compensate for the distribution of \( t_c \)\footnote{\( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality.}~\cite{olguinmunoz:impact2021}, in the present work we adjust the nominal delay \( D \) by a factor \num{1.5}.
+We will refer to this adjusted delay as \emph{\acf{TTF}}.
 
-\begin{align}
-    \mathbb{E}\left[\text{\ac{TTF}}\right] &= \mathbb{E}\left[\mathcal{W}\right] + D
-    = \mathbb{E}\left[U(t_{n - 1}, t_n)\right] + D\nonumber\\
-    &= \mathbb{E}\left[U(0, D)\right] + D
-    = \frac{1}{2}D + D = \frac{3}{2}D\label{eq:ttf}
-\end{align}
+% Delay, however, was merely an experimental tool, and if we consider
+% \begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
+%     \item that \( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality~\cite{olguinmunoz:impact2021}
+%     \item that for a step subject to delay \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
+% \end{enumerate*},
+% then it follows that the expected \ac{TTF} for steps subject to a delay \( D \) can be expressed as
 
-Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5}.
+% \begin{align}
+%     \mathbb{E}\left[\text{\ac{TTF}}\right] &= \mathbb{E}\left[\mathcal{W}\right] + D
+%     = \mathbb{E}\left[U(t_{n - 1}, t_n)\right] + D\nonumber\\
+%     &= \mathbb{E}\left[U(0, D)\right] + D
+%     = \frac{1}{2}D + D = \frac{3}{2}D\label{eq:ttf}
+% \end{align}
 
-We can then reformulate our previous conclusions as:
+% Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5}.
+
+Our findings can be summarized as follows:
 \begin{enumerate}
@@
 
-    \item The above effects are modulated by an individual personality trait.
+    \item The above effects are modulated by measures of individual levels of personality characteristics and focus.
     % We will discuss this in detail below in \cref{ssec:moderationeffects}
@@
 
-We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}, introduced by \textcite{oliver:bfi1999}, and the \acf{ITQ}, described by \textcite{witmer1998measuring}.
-Out of the collected individual-difference variables, the most salient one corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
+We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}~\cite{oliver:bfi1999},and the \acf{ITQ}~\textcite{witmer1998measuring}.
+Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
 Delay discounting is the tendency to devalue rewards for which one must wait; high rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
-    \caption{%
-        Correlation between neuroticism and the mean execution time of the last four steps in segments of \num{12} steps subject to the same \ac{TTF} in~\textcite{olguinmunoz:impact2021}.
-        Pearson correlation coefficient \( \rho = 0.418 \), 2-tailed \( p < 0.05 \).
-    }\label{fig:neuroexectimecorrel}
-\end{figure}
+% Bobby suggested we remove this figure and just give the pearson coeff.
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
+%     \caption{%
+%         Correlation between neuroticism and the mean execution time of the last four steps in segments of \num{12} steps subject to the same \ac{TTF} in~\textcite{olguinmunoz:impact2021}.
+%         Pearson correlation coefficient \( \rho = 0.418 \), 2-tailed \( p < 0.05 \).
+%     }\label{fig:neuroexectimecorrel}
+% \end{figure}
 
-Neuroticism was identified as a modulating factor for the pacing effects through a \ac{PCA}.
+Linear regression showed a significant correlation between individual neuroticism scores and the execution time late in a series of high-delay steps, \( \rho = 0.418 \), \num{2}-tailed \( p < 0.05 \)
+Neuroticism was further identified as a modulating factor for the pacing effects through a \ac{PCA}.
 Out of the three identified components, which cumulatively accounted for \SI{73.13}{\percent} of the variance in the results, neuroticism was included in the first two.
-The effect of neuroticism was transversal to all \acp{TTF} and impairment durations in the tasks --- we show an example of the correlation between execution time and neuroticism in \cref{fig:neuroexectimecorrel}
+The effect of neuroticism was observed across all \acp{TTF} and impairment durations in the tasks.
 
@@
 
-Furthermore, we found that execution times, when grouping by experimental variables such as neuroticism, \ac{TTF}, and continuous segments of steps subject to the same \ac{TTF}, were well-fit by an \ac{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
+\todo[inline]{%
+    Comment by Bobby: should we report the values for our goodness-of-fit tests? I think this is important.
+
+    Manuel: I could, but we would have to find a way to report them for all the possible groups that the model forms.
+    Maybe report the best fit? Or the average fit?
+}
+Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \ac{TTF}, and continuous segments of steps subject to the same \ac{TTF}, were well-fit by an \ac{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
 This distribution has an ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.

### 2022-12-04 conclusion.tex
Satya & Bobby comments

@@
 
-The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic human-in-the-loop nature.
-Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
-Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming and prohibitively expensive for many research groups.
+This paper addresses the difficulty of benchmarking \ac{WCA} by offering a model-based alternative to the extensive human-user studies that would otherwise be required.
+It first introduces the \edgedroid{} model of human timing behavior for \ac{WCA}, the first data-driven model for human timing in \ac{WCA} applications.
+This model represents a stochastic approach to execution time modeling which builds upon prior data~\cite{olguinmunoz:impact2021}.
+It further introduces a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
 
-In the first half of this paper, we have introduced the \edgedroid{} model of human timing behavior for \ac{WCA}, the first data-driven model for human timings in \ac{WCA} applications.
-This model represents a stochastic approach to execution time modeling which builds upon the data collected for \textcite{olguinmunoz:impact2021}.
-Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+The paper then explores the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
+It shows that less realistic modeling approachesm which do not take into account higher-order effects on execution time distributions, can potentially lead to substantial mis-estimations of the application footprint.
 
-Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
-We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
+Finally, the paper delves into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
+It proposes a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}, which results in an adaptive sampling strategy. 
+We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
+Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, thus proving the value of such frameworks for the design of \ac{WCA} applications.
 
-Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
-We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy. 
-We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
-Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
-
-This work serves as an important, but yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
+This work serves as an important, yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
 Many directions remain open to additional exploration in this space.
 For instance, our current model only targets a particular class of step-based \ac{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \ac{AR} and \ac{XR}, is on our roadmap.
-Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the US.\@
+Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the United States.
 An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.
+Corresponding research is needed to determine individual difference factors that would substantially impact response to \ac{WCA} in novel populations.
 
-In terms of the optimization framework, our current approach builds upon the assumption that execution times are Rayleigh-distributed.
-Although this distribution fits the data well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
+In terms of the optimization framework, our current model adopts the assumption that execution times are Rayleigh-distributed.
+Although this distribution fits the data reasonably well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
 Our future efforts consider developing this framework towards these more accurate distributions.
@@
 As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.
+
+\todo[inline]{Comment from Bobby: Reviewers might wonder why we stuck with the Rayleigh distribution.}

### 2022-12-04 implications_footprint.tex
Satya & Bobby comments

@@
 
-In order to illustrate the consequences of using a less realistic model which does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
+In order to illustrate the consequences of using a less realistic model that does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
 This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.

### 2022-12-04 implications_optimization.tex
Satya & Bobby comments

@@
 On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
-This makes the sample-count-optimized approach more predictable, reliable, and consistent, and thus more suitable for real-world deployments.
+As mentioned above, the future feasibility and mass adoption of \ac{WCA} depends on these applications not hogging the available resources.
+Our work advances this goal by being consistently more efficient than existing alternatives at minimizing the number of samples per step, and thus reducing network and processing load.
 
 It should be noted that although both schemes seem to tend towards \num{2.0} samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
-As \acp{RTT} increase way above the expected wait time, the probability of the sum of the first sampling interval and the \ac{RTT} being larger than the execution time of the step tends towards \num{1.0}.
+As \acp{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \ac{RTT} being larger than the execution time of the step tends towards \num{1.0}.
 This leads to these sampling schemes consistently sampling only twice each step: a first sample which is taken before the execution time of the step, and a second one \ac{RTT} seconds later, after the execution time has been reached.

### 2022-12-04 intro.tex
Satya & Bobby comments

@@
 This also translates into latency-sensitivity, as with any \ac{AR} application.
-Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
+Delays and jitter can be jarring to the user, causing them discomfort, leading them to make mistakes and potentially even to abandoning the application altogether.
 On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.
@@
 Due to their human-in-the-loop nature, these applications present a challenge to benchmark and characterize, in particular in real-scale deployments where dozens or even hundreds of users might concurrently use the system.
-Recruiting a large enough cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers.
-Furthermore, humans are unpredictable, and human-subject research is thus often hard to repeat and replicate.
-There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do away with the human-the-loop.
+Accordingly, recruiting a  cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers
+There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do not rely on direct testing of the human-in-the-loop.
 
@@
     We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
-    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
+    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \ac{WCA}, understanding this term as the duration of a specific complete execution of the application task.
+    In accordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
     \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
     We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    First, we study the potential for reducing the number of samples captured per step.
+    This is a valuable endeavor, as reducing the number of samples captured and subsequently processed directly translates in lower bandwidth demand on the wireless network and processor time demand on the cloudlet.
+    Next, we explore the potential for direct optimization of energy consumption.
+    The economic feasibility of WCA, and hence its likelihood of commercial adoption depends on it not being a resource hog, and this work furthers that effort.
     Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
@@
 In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
-In \cref{sec:model} we detail our model for the generation of realistic human timings in \ac{WCA}.
+In \cref{sec:model} we detail our model for the generation of realistic human timing in \ac{WCA}.
 We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.

### 2022-12-04 model.tex
Satya & Bobby comments

@@
 
-In the following, we employ the above insights together with the actual data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+In the following, we employ the above insights together with the data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
 We detail the construction of a model which uses the data collected to generate, at runtime, realistic execution times. 
 In order to accurately emulate the behavior of a human, such a model needs to implement two main behaviors.
-One, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system.
+First, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system, as well as salient individual-difference measures.
 We detail this in \cref{ssec:model:exectimes}.
-And two, it needs to produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}.
+Neuroticism is incorporated in this aspect of the model, as it is the most salient individual difference in the data; however, other factors such as immersive tendency could be treated similarly.
+Second, the model must produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}. 
 % Finally, in \cref{ssec:model:obtaining} we briefly discuss where implementations of this model can be obtained.

### 2022-12-04 related_work.tex
Satya & Bobby comments

@@
 
-There exist already in literature a number of approaches to the characterization and benchmarking of mobile \ac{XR}, and by extension \ac{WCA}, deployments.
+There exist a number of approaches to the characterization and benchmarking of mobile \ac{XR}, and by extension \ac{WCA}, deployments.
 \emph{OpenRTiST}~\cite{george2020openrtist} is a tool which uses a compute-heavy, yet latency-sensitive workload --- image style transfer~\cite{jing2019neural} --- to load and benchmark edge computing deployments.
@@
 In \textcite{srinivasan2009performance}, the authors analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
-They proceed to use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
+They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
 The authors of \textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
 In \textcite{huang2021proactive}, the authors present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
-\textcite{al2017energy} propose an resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
+\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
 As above, we have also previously directly contributed to this field.

### 2022-12-05 appendices.tex
jaya

@@
 
-We first borrow the idea of checkpointing density from \textcite{Satoshi1992Optimal} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+We first borrow the idea of checkpointing density from~\cite{Satoshi1992Optimal} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
@@
 % Note that throughout the paper we use a exGaussian distribution but not Rayleigh. This is because, 
-Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an exponentially modified Gaussian distribution as well as a Rayleigh distribution, with the former attracting more attention from works like \cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
-This is the reason why the above results are applicable in this work where we have predominantly considered exGaussian distribution.
+Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \acl{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+This is the reason why the above results are applicable in this work where we have predominantly considered \ac{exGaussian} distribution.
 Furthermore, we have also verified the closeness of the results as well as the validity of the approximations made in the proofs using distribution fitting and simulations.

### 2022-12-05 background.tex
jaya

@@
 \acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \ac{AR} applications.
-In this work we focus on a particular category --- denoted ``step-based'' \ac{WCA} --- that have as their goal the guiding of a user through a sequential task.
+In this work we focus on a particular category --- ``step-based'' \ac{WCA} --- that have as their goal the guiding of a user through a sequential task.
 Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
@@
 The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
-Whenever the assistant application detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
-The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.
-because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
-Herein lies one of key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
+Whenever the assistant detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.~because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
+Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
 
@@
 
-In \textcite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} in a controlled experiment.
+In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} in a controlled experiment.
 We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA}~\cite{Chen2015LEGO}.
-Subjects interacted with the assistant while we altered the responsiveness in realtime and captured key application and task performance metrics.
+\num{40} subjects interacted with the assistant, performing a \num{169}-step task while we altered the responsiveness in realtime and captured key application and task performance metrics.
 Additionally, we also employed questionnaires to evaluate personality traits of the participants, and correlated these results with the task performance metrics.
@@
 
-We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}~\cite{oliver:bfi1999},and the \acf{ITQ}~\textcite{witmer1998measuring}.
+We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}~\cite{oliver:bfi1999}, and the \acf{ITQ}~\cite{witmer1998measuring}.
 Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
@@
 
+In this work, we will use a normalized scale to describe neuroticism, derived from the minimum and maximum obtainable values for this variable in the \ac{BFI}~\cite{oliver:bfi1999}.
+\emph{Low} and \emph{high} neuroticism will refer to the \( [0.0, 0.5) \) and \( [0.5, 1.0] \) ranges respectively.
+
 Linear regression showed a significant correlation between individual neuroticism scores and the execution time late in a series of high-delay steps, \( \rho = 0.418 \), \num{2}-tailed \( p < 0.05 \)
@@
     \caption{%
-        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \ac{TTF} in \textcite{olguinmunoz:impact2021}.
+        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \ac{TTF} in~\cite{olguinmunoz:impact2021}.
         Distributions were fitted using \ac{MLE}.

### 2022-12-05 conclusion.tex
jaya

@@
 The paper then explores the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
-It shows that less realistic modeling approachesm which do not take into account higher-order effects on execution time distributions, can potentially lead to substantial mis-estimations of the application footprint.
+It shows that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions, can potentially lead to substantial mis-estimations of the application footprint.
 
@@
 It proposes a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}, which results in an adaptive sampling strategy. 
-We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
+We have shown that this framework is applicable to both the minimization of number of samples and total energy consumption per step by showcasing experimental results.
 Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, thus proving the value of such frameworks for the design of \ac{WCA} applications.
@@
 
-\todo[inline]{Comment from Bobby: Reviewers might wonder why we stuck with the Rayleigh distribution.}
+\todo[inline]{Comment from Bobby: Reviewers might wonder why we stuck with the Rayleigh distribution.
+Manuel: I think Vishnu already fixed this in the appendix.}

### 2022-12-05 implications_footprint.tex
jaya

@@
 In order to illustrate the consequences of using a less realistic model that does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for \textcite{olguinmunoz:impact2021}.
+This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for~\cite{olguinmunoz:impact2021}.
 This distribution is then randomly sampled at runtime to obtain execution times for each step, without any adjustment to the current state of the system.
@@
 
-We deploy configurations running \num{45}-step versions of the LEGO task described in \textcite{olguinmunoz:impact2021}.
+We deploy configurations running \num{45}-step versions of the LEGO task described in~\cite{olguinmunoz:impact2021}.
 The testbed configurations include setups with \numlist{1;4;7;10} clients.
@@
 The results are presented in \cref{fig:testbed_lifetimes}.
-Owing to the low number of samples, we opt here for the geometric rather than arithmetic mean to represent our results.
+Owing to the low number of samples, specifically to minimize the effects of potential outliers, we opt here for the geometric rather than arithmetic mean to represent our results.
 The results are nonetheless clear, and follow the same pattern as the previously discussed results under ideal, controlled circumstances.

### 2022-12-05 implications_optimization.tex
jaya

@@
 With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in \textcite{Moothedath2022Aperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+Next, in~\cite{Moothedath2022Aperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
 
@@
 However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
-Furthermore, instead of directly optimizing for energy, we start by noting that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
-% in a format similar to that of the energy, 
+Furthermore, instead of directly optimizing for energy, we start by noting that, in general, any objective metric in these applications which relates to sampling and the responsiveness of the system will present itself
+% that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves 
+% % in a format similar to that of the energy, 
 as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
@@
 \begin{alignat}{1}
-    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C,\;\label{eq:epsilon_terminal}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C\;\label{eq:epsilon_terminal}
 \end{alignat}
 
-Above, $\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
-For instance, in the modeling used by \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
+\todo[inline]{%
+    Comment by Jaya:
+    The presentation of the optimization for energy metric in (5) and later in Section 6.2 presenting the optimization for minimizing number of samples subject to weight penalty is done in a non-trivial way.
+A natural way to present it is to state the optimization problem in 6.2 first and state that the function in (5) is a Legrangian function of this optimization with alpha/beta being the Lagrangian parameter.
+People are well aware of the Lagrangian formulation, so at least mention that (5) is a Lagrangian function and alpha/beta is a Lagrangian parameter.
+
+    Manuel: I've added a mention to the Lagrangian function below. Do we need anything else?
+}
+
+This corresponds to a Lagrangian function of this optimization, with \ensuremath{\frac{\alpha}{\beta}} representing the Lagrangian parameter.
+$\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
+For instance, in the modeling used by~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
 Our solution which minimizes \cref{eq:epsilon_terminal} is provided in \cref{tn_approx_rayleigh}.
@@
 \begin{alignat}{1}
-r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
-\Rightarrow t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tn_approx_rayleigh}
+% r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
+% \Rightarrow
+t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tn_approx_rayleigh}
 \end{alignat}
@@
 Thus, we look at the constrained optimization of the expected number of samples with an upper bound $w_0$ for the expected wait.
-That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$. We show in \cref{appx1} that the appropriate $\alpha$ and $\beta$ for this problem satisfy the condition
+That is, $\mathbb{E}[\mathcal{W}]\!\leq\!w_0$.
+We show in \cref{appx1} that we can find appropriate $\alpha$ and $\beta$ for this problem by satisfying the condition
 \begin{alignat}{1}\label{eq:optimization:sampling}
@@
 \end{alignat}
-where, $\mathlarger{\Gamma}(x)$ is the Gamma function.
+where $\mathlarger{\Gamma}(x)$ is the Gamma function.
 
@@
 
-In the following, we will show the effects of our optimization approach combined with our timing models compared to \textcite{Wang2019Towards}'s state-of-the-art approach.
+In the following, we will show the effects of our optimization approach combined with our timing models compared to the state-of-the-art approach~\cite{Wang2019Towards}.
 For this we implement these sampling schemes in Python and run a number of simulations with them.
@@
 
-It should be noted that although both schemes seem to tend towards \num{2.0} samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
+It should be noted that although both schemes seem to tend towards two samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
 As \acp{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \ac{RTT} being larger than the execution time of the step tends towards \num{1.0}.
@@
 In the following, we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
-We directly take the modelling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the  assumption of one-way communication in all but the final sample.
+We directly take the modeling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the  assumption of one-way communication in all but the final sample.
 With feedback given even to the discarded samples in our model and communication delay defined as the total delay in either direction, we have, 
@@
 We refer to this scheme as the \emph{energy-optimized aperiodic} sampling scheme.
-For the power constants, we reuse the values estimated by the authors in \textcite{Moothedath2022EnergyEfficient}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
+For the power constants, we reuse the values estimated by the authors in~\cite{Moothedath2022EnergyEfficient}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
 On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
@@
 The results of these experiments are presented in \cref{fig:optimization:energy}, and they clearly illustrate the advantages of the integration of our timing models with an adaptive, energy-optimized sampling scheme when compared to unoptimized and state-of-the-art sampling schemes.
-Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{Wang2019Towards}'s state-of-the-art, and is up to \SI{40}{\percent} more efficient than the greedy scheme.
+Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{Wang2019Towards}'s state-of-the-art, and is up to \SI{40}{\percent} more energy efficient than the greedy scheme.
 Furthermore, once again the behavior of our approach is more consistent and reliable than the competition, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.

### 2022-12-05 intro.tex
jaya

@@
     \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
-    Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
+    Using the data collected for~\cite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
     We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.

### 2022-12-05 model.tex
jaya

@@
 % For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-The processing of the data from \textcite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
+The processing of the data from~\cite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
 The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
@@
 We employ a cleaned and re-parameterized copy of the timing data.
-The \num{6760} data points are arranged in a table together with identifiers for the subjects, their neuroticism score, and a sequence number for each step.
+The \num{6760} data points are arranged in a table together with identifiers for the subjects, their normalized neuroticism score, and a sequence number for each step.
 
@@
         Neuroticism        & \( [0, 0.5) \) & & & & & & \( [0.5, 1.0] \)      \\
-        Weighted TTF & \((-\infty, 0.82]\) & \((0.82, 1.53]\) & \((1.53, 2.08]\) & \((2.08, 2.67]\) & \((2.67, 3.45]\) & \((3.45, 4.13]\) & \((4.13, \infty]\)
+        Weighted TTF & \([0.0], 0.82]\) & \((0.82, 1.53]\) & \((1.53, 2.08]\) & \((2.08, 2.67]\) & \((2.67, 3.45]\) & \((3.45, 4.13]\) & \((4.13, \infty]\)
     \end{tabular}
@@
     At low (\SI{0}{\second}) \ac{TTF}, the models speed-up over time; conversely at high (\SI{5.0}{\second}) \ac{TTF}, the models either present no change or drastically increase their generated execution times, depending on the level of neuroticism.
-    Error bars indicate \SI{95}{\percent} \ac{CI}.
+    Error bands indicate \SI{95}{\percent} \ac{CI}.
     }\label{fig:exectimeduration}

### 2022-12-05 related_work.tex
jaya

@@
 \textcite{Chen2017Empirical} use a number of prototype \ac{WCA} implementation to study and characterize real-world latency bounds in these systems.
-In \textcite{lecci2021open}, the authors collect and analyze large amounts of \ac{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
+In~\cite{lecci2021open}, the authors collect and analyze large amounts of \ac{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
 \textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for for the benchmarking of mobile hardware in the context of \ac{AR}.
@@
 We achieve this by building upon our own previous contributions to this field.
-In \textcite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
+In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
 We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
 In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
-In \textcite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
+In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
 We found that humans generally pace themselves according to the perceived system responsiveness.
@@
 
-The characterization and optimization, of mobile \ac{XR} has also long been a topic of research.
-In \textcite{srinivasan2009performance}, the authors analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
+The characterization and optimization of mobile \ac{XR} has also long been a topic of research.
+\textcite{srinivasan2009performance} analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
 They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
-The authors of \textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
-In \textcite{huang2021proactive}, the authors present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
+\textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
+\textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
 \textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
 As above, we have also previously directly contributed to this field.
-In \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
-This is extended into an aperiodic strategy in \textcite{Moothedath2022Aperiodic}.
+In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
+This is extended into an aperiodic strategy in~\cite{Moothedath2022Aperiodic}.
 We further develop this approach in this work by deriving a more practical approximate solution to the aperiodic strategy, which we then use to implement a novel framework for the generic optimization of responsiveness-resource consumption trade-offs.

### 2022-12-05 background.tex
vishnu

@@
     \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
-    Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \ac{TTF} of \SI{2.475}{\second}, and a further \num{14} percentage points slower at a mean \ac{TTF} of \SI{4.5}{\second}.
+    Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \ac{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \ac{TTF} of \SI{4.5}{\second}.
 

### 2022-12-09 body/part1/overview.tex
methodology figure
writing

@@
 \chapter{Key contributions and results}\label{chap:contributions}
-
-\todo[inline]{Reorganize}
 
@@
 
-\subsection{Applying the methodology to \acs{WCA}}
+\begin{figure}
+    \centering
+    \includegraphics[width=.8\textwidth]{Figs/methodology.png}
+    \caption{%
+        Overview of the proposed methodology for the study of feedback-loop systems on edge computing infrastructure.
+        The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
+        The backend/server side of the system along with the network remain unchanged.
+    }\label{fig:methodology}
+\end{figure}
 
-\todo[inline]{\cref{paper:olguinmunoz2018demoscaling}}\label{summary:2018demo}
+As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
+In the following, we introduce our methodological framework for tackling this challenge.
 
-As discussed previously, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
-This short papers' main contribution is the introduction of our methodological framework for tackling this challenge.
+A general overview of our methodology is illustrated in \cref{fig:methodology}.
+It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
+We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \ac{COTS} general-purpose computing devices.
+The backend side and the network are not changed.
 
-As previously explained, our methodology is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
 Two main arguments drive this approach.
-One, it maintains the realism and accuracy of the system.
-This is particularly important in wireless and mobile edge computing systems, which present complex behaviors and interactions intrinsic to their networking technologies.
-And two, it allows for ease of scaling.
+One, the difficulty of scaling the number of clients in these systems in a research context.
+The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
 The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \ac{WCA}) or due its cyber-physical nature (in \acp{NCS}).
@@
 
-This work presents the first approach to this concept for \ac{WCA}.
-To establish a reproducible and comparable workload, we use a trace-based design for the client-side emulation.
-We first record a trace of the sensory input data while having a human user operate the target application.
-The collected data consists of the sensory inputs provided to the system at runtime, for instance --- in the case of a video-based application --- the video feed from the camera.
+And two, it maintains the realism of the most complex parts of the system.
 
-To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
-This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
+\todo[inline]{Fix and finish}
+
+In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
+We first discuss the methodology in the context of a human-in-the-loop application, \acl{WCA}, in section \cref{ssec:methodology:wca}.
+Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \acf{NCS}.
+
+\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
 
 \begin{figure}
-    % Define block styles
-    %         \tikzstyle{block} = [circle, draw, fill=red!20, 
-    %         text width=5em, text centered, rounded corners, minimum size=4em]
-    %         \tikzstyle{decision} = [diamond, aspect=2, draw, fill=yellow!20,
-    %         text width=5em, text centered, rounded corners, minimum size=4em]
-    %        \tikzstyle{line} = [draw, -{Latex[length=3mm]}]
+    \centering
+    \begin{subfigure}[t]{\textwidth}
+        \centering
+        \includegraphics[width=.8\textwidth]{Figs/trace_edgedroid.png}
+        \caption{%
+            High level conceptual design of our methodology for \ac{WCA}.
+            We replace the human by a trace-driven emulation.
+            This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
+        }
+    \end{subfigure}
+    \medskip
+    \begin{subfigure}[t]{\textwidth}
         \centering
@@
         }
-    \caption{State diagram of the user model}
-    \label{fig:usermodel}
+        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.}\label{fig:usermodel}
+    \end{subfigure}
+    \caption{%
+        Overview of our first implementation of the methodology for \ac{WCA}.
+    }\label{fig:edgedroid1:trace}
 \end{figure}
+
+We first introduce our methodology in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
+The former corresponds to an extended abstract paper which discusses a high level overview of our approach; the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
+
+The overall goal of these papers is to explore this methodological approach as applied to \ac{WCA}.
+For this purpose, we introduce a tool, \emph{EdgeDroid 1.0}, which mimics the behavior of a human interacting with a \ac{WCA} client application.
+
+To establish a reproducible and comparable workload, we use a trace-based design for the client-side emulation.
+We first record a trace of the sensory input data while having a human user operate the target application.
+The collected data consists of the sensory inputs provided to the system at runtime, for instance --- in the case of a video-based application --- the video feed from the camera.
+
+To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
+This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
+Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
 
@@
 
-\subsection{Applying the methodology to \acsp{NCS}}
+\subsection{Applying the methodology to \acsp{NCS}}\label{ssec:methodology:ncs}
+
 \todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}

### 2022-12-11 body/part1/overview.tex
writing

@@
     \centering
-    \begin{subfigure}[t]{\textwidth}
+    \begin{subfigure}[b]{.47\textwidth}
         \centering
-        \includegraphics[width=.8\textwidth]{Figs/trace_edgedroid.png}
+        \includegraphics[width=\textwidth]{Figs/trace_edgedroid.png}
         \caption{%
@@
     \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{.47\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
+        \caption{%
+            Architectural overview of the EdgeDroid \num{1.0} tool.
+            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \ac{WCA} backend running on a cloudlet.%
+            }
+    \end{subfigure}\\
     \medskip
@@
         \centering
-        \adjustbox{scale=0.75}{
+        \adjustbox{scale=0.7}{
             \begin{tikzpicture}[align=center,
@@
         }
-        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.}\label{fig:usermodel}
+        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.
+This model approximates the behavior of an ``ideal'' human, one that is patient and makes no mistakes.}\label{fig:usermodel}
     \end{subfigure}
@@
 The overall goal of these papers is to explore this methodological approach as applied to \ac{WCA}.
-For this purpose, we introduce a tool, \emph{EdgeDroid 1.0}, which mimics the behavior of a human interacting with a \ac{WCA} client application.
+As discussed in \cref{chap:introduction}, the main challenge to benchmarking \ac{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
+Humans are unreliable, and, perhaps more importantly, hard to scale.
+\todo[inline]{More reasons}
 
-To establish a reproducible and comparable workload, we use a trace-based design for the client-side emulation.
-We first record a trace of the sensory input data while having a human user operate the target application.
-The collected data consists of the sensory inputs provided to the system at runtime, for instance --- in the case of a video-based application --- the video feed from the camera.
+The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
+We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \ac{WCA} backend.
+The tool is implemented in Android, allowing for its easy deployment on the same kind of \ac{COTS} mobile devices on which real \ac{WCA} applications are intended to run in the future.
+Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
+On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \ac{WCA} by a human volunteer.
+The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
+Once recorded, the trace is manually segmented into the logical component steps of the task, and these are replayed in sequence back to the backend by the EdgeDroid tool.
+
+Utilizing a trace for the emulation of human behavior in \ac{WCA} has multiple advantages.
+The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
+Using a trace ensures that the same computation is performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
+The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
+The third is extensibility.
+
+\todo[]{}
+
+reproducible and comparable workload, we use a trace-based design for the client-side emulation.
+.
 
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
+    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
     \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
+    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
     \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
@@
     \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
+    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
     \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
@@
     \centering
-    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
+    \includegraphics[width=.6\textwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
     \caption{
@@
         \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
+        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
             \toprule
@@
         \caption{Percentages of variance explained by the components.}
-        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
+        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
             \toprule

### 2022-12-12 abstract.tex
jaya
comments james

@@
 \begin{abstract}
-    \acf*{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
-    Employing user testing to optimize system parameters is generally not feasible, given the scope of the problem and the number of observations needed for to detect small but important effects in controlled experiments.
+    \ac{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
+    Employing user testing to optimize system parameters is generally not feasible, given the scope of the problem and the number of observations needed to detect small but important effects in controlled experiments.
     Considering the intended mass-scale deployment of \ac{WCA} applications in the future, there exists a need for tools enabling human-independent benchmarking.
 
-    We present in this paper the first data-driven model which allows for the complete end-to-end emulation of humans in \acs*{WCA}.
-    We demonstrate the utility of this model by employing it to study application task durations, where we find an up to \SI{12}{\percent} difference with respect to first order approximations.
-    Further, we use this model to develop a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \acs*{WCA}.
-    We show that our framework can provide up to a \SI{50}{\percent} reduction in processed frame samples and up to a \SI{20}{\percent} reduction in energy consumption when compared to the state-of-the-art.
+    We present in this paper the first model for the complete end-to-end emulation of humans in \ac{WCA}.
+    We build this model through statistical analysis of data collected from previous work in this field, and demonstrate its utility by studying application task durations.
+    Compared to first-order approximations, our model shows a \textasciitilde\SI{36}{\percent} larger gap between step execution times at high system impairment versus low.
+    We further introduce a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \ac{WCA}, and show that by combining this framework with our realistic model of human behavior, significant reductions of up to \SI{50}{\percent} in number processed frame samples and \SI{20}{\percent} in energy consumption can be achieved with respect to the state-of-the-art.
 \end{abstract}
+\acresetall%

### 2022-12-12 background.tex
comments james

@@
 
-\todo[inline]{%
-    Comment by Bobby: should we report the values for our goodness-of-fit tests? I think this is important.
-
-    Manuel: I could, but we would have to find a way to report them for all the possible groups that the model forms.
-    Maybe report the best fit? Or the average fit?
-}
 Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \ac{TTF}, and continuous segments of steps subject to the same \ac{TTF}, were well-fit by an \ac{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
+When grouping by level of neuroticism, \ac{TTF}, and \emph{slice}\footnote{%
+In~\cite{olguinmunoz:impact2021} the \emph{slice} to which a step belongs to refers to whether the step occurred in the first, second, or third four-step segment of a sequence of steps subject to the same \ac{TTF}.
+}, the best fit statistic was \ensuremath{0.028} (\ensuremath{p = 0.999}).
 This distribution has an ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.

### 2022-12-12 conclusion.tex
comments james

@@
 This paper addresses the difficulty of benchmarking \ac{WCA} by offering a model-based alternative to the extensive human-user studies that would otherwise be required.
-It first introduces the \edgedroid{} model of human timing behavior for \ac{WCA}, the first data-driven model for human timing in \ac{WCA} applications.
+It first introduces the \edgedroid{} model of human timing behavior for \ac{WCA}.
 This model represents a stochastic approach to execution time modeling which builds upon prior data~\cite{olguinmunoz:impact2021}.
@@
 As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.
-
-\todo[inline]{Comment from Bobby: Reviewers might wonder why we stuck with the Rayleigh distribution.
-Manuel: I think Vishnu already fixed this in the appendix.}

### 2022-12-12 implications_optimization.tex
comments james

@@
 With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in~\cite{Moothedath2022Aperiodic}, the author retains the model, removes the constraint of periodicity, and finds the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+Next, in~\cite{Moothedath2022Aperiodic}, the author retains the model and removes the constraint of periodicity to find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
 
@@
 \end{alignat}
-
-\todo[inline]{%
-    Comment by Jaya:
-    The presentation of the optimization for energy metric in (5) and later in Section 6.2 presenting the optimization for minimizing number of samples subject to weight penalty is done in a non-trivial way.
-A natural way to present it is to state the optimization problem in 6.2 first and state that the function in (5) is a Legrangian function of this optimization with alpha/beta being the Lagrangian parameter.
-People are well aware of the Lagrangian formulation, so at least mention that (5) is a Lagrangian function and alpha/beta is a Lagrangian parameter.
-
-    Manuel: I've added a mention to the Lagrangian function below. Do we need anything else?
-}
 

### 2022-12-12 intro.tex
comments james

@@
 They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
-This context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video feeds, which require intensive compute capabilities to process.
+This context-sensitivity and proactivity in providing user feedback translate into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video, which require intensive compute capabilities to process.
 This also translates into latency-sensitivity, as with any \ac{AR} application.
-Delays and jitter can be jarring to the user, causing them discomfort, leading them to make mistakes and potentially even to abandoning the application altogether.
+Delays and jitter can be jarring to the user, causing discomfort and leading them to make mistakes and potentially even abandoning the application altogether.
 On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.
 
-The combination of these opposing characteristic has led to \ac{WCA} applications being identified in research literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+The combination of these opposing characteristics has led to \ac{WCA} applications being identified in the literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
 However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
@@
 
-In that context, the contributions of this paper are as follows:
+In that context, the contributions of this paper are:
 \begin{enumerate}
-    \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+    \item\label{item:contrib:model} We introduce the first, to our knowledge, stochastic model for human timings in \ac{WCA} applications.
     Using the data collected for~\cite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
@@
     We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \ac{WCA}, understanding this term as the duration of a specific complete execution of the application task.
+    \item\label{item:contrib:footprint} Using this model, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \ac{WCA}, understanding this term as the duration of a specific complete execution of the application task.
     In accordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
@@
     The economic feasibility of WCA, and hence its likelihood of commercial adoption depends on it not being a resource hog, and this work furthers that effort.
-    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
+    Leveraging our more involved user model, combined with the introduced optimization approach, we achieve a \textasciitilde\SI{60}{\percent} reduction in the number of samples processed per step.
+    We also achieve an improvement of \SI{20}{\percent} in energy consumption, all while maintaining comparable levels of system responsiveness.
 \end{enumerate}
@@
 We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
-Next in \cref{sec:implications:footprint} we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
+Next, in \cref{sec:implications:footprint}, we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
 This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \ac{WCA} systems.
-In this section, we also introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
+In the same section, we introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
 Finally, in \cref{sec:conclusion} we summarize and conclude this paper, as well as briefly discuss potential avenues for future work.

### 2022-12-12 model.tex
comments james

@@
 
-After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \acp{TTF}.
-We devise two variants of such a model, which operate as follows:
+After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \acp{TTF}:
 
@@
 
-We implement these models in Python 3.10, and verify their correct behavior in the following.
+We implement these two variants of the model in Python 3.10, and verify their correct behavior in the following.
 
@@
     \item\label{it:ttftoexectime} Higher \acp{TTF} should result in higher execution times.
-    \item\label{it:duration} When subject to a series of steps at the same level of impairment, desired model behavior depends on the impairment:
+    \item\label{it:duration} When subject to a series of steps at the same level of impairment, desired behavior depends on the level:
     \begin{enumerate}
@@
         \item At medium \acp{TTF}, execution times should remain more or less the same.
-        \item At high \acp{TTF}, execution times should progressively increase.
+        \item At high \acp{TTF}, execution times should increase.
     \end{enumerate}
-    \item The effects on execution times due to past reductions or improvements in system responsiveness should linger for at least a few steps whenever system responsiveness changes.
+    \item The effects on execution times due to past changes in system responsiveness should linger for at least a few steps whenever system responsiveness changes anew.
     \item\label{it:neuro} Finally, neuroticism should act as a modulating factor for the above effects.

### 2022-12-12 implications_optimization.tex
Update on Overleaf.

@@
 
-This corresponds to a Lagrangian function of this optimization, with \ensuremath{\frac{\alpha}{\beta}} representing the Lagrangian parameter.
-$\alpha, \beta$ and $C$ are constants which are responsible for modifying the objective function from one metric to another.
+% This corresponds to a Lagrangian function of this optimization, with \ensuremath{\frac{\alpha}{\beta}} representing the Lagrangian parameter.
+Here, $\alpha, \beta$ and $C$ are constants responsible for modifying the objective function from one metric to another.
 For instance, in the modeling used by~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.

### 2022-12-13 appendices.tex
fixes for IEEE ToMC

@@
+\clearpage%
+\pagestyle{empty}%
 \appendix{}\label[appendixwithoutnumber]{appx1}
@@
 By making use of this general solution, we can also prove the results given in \cref{ssec:optimization:samples}, where we modify the problem and find the optimum sampling instants that minimize the expected number of samples for a given upper bound $w_0$ for the expected wait time.
-First note that the general solution has constants $\alpha$ and $\beta$ that correspond to the weights given to the cost of sampling and cost of waiting, respectively.
+First note that the general solution has constants $\alpha$ and $\beta$ corresponding to the weights given to the cost of sampling and waiting, respectively.
 The optimization criteria changes from minimizing wait time to minimizing the number of samples when the ratio $\frac{\alpha}{\beta}$ goes from zero to infinity.

### 2022-12-13 biographies.tex
fixes for IEEE ToMC

\newpage%
\normalfont%
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/manuel.jpeg}}]%
    Manuel Olguín~Muñoz obtained in 2017 his B.Sc.Eng.\ in Computing and Engineer's Degree in Computing from Universidad de Chile.
    He is currently a final-year Ph.D. candidate at the division of Information Science and Engineering, School of Electrical Engineering and Computer Science (EECS), at KTH Royal Institute of Technology, Sweden.
    His research interests lie in the benchmarking and performance analysis of Edge Computing infrastructures and the applications deployed on them, in particular with respect to human-in-the-loop applications and cyber-physical systems.
    He also has an active role in designing and implementing the first 5G-enabled Edge Computing testbed at KTH and northern Europe.
\end{IEEEbiography}
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/vishnu.jpeg}}]%
    Vishnu Narayanan Moothedath is a doctoral student in the department of Intelligent Systems under the School of Electrical Engineering and Computer Science (EECS) at KTH Royal Institute of Technology from 2021.
    He completed his master's degree in Communication Systems from the Indian Institute of Technology (IIT) Madras in 2016 and his bachelor's degree in Electrics and Communication Engineering from the National Institute of Technology (NIT) Calicut in 2012.
    He has also worked for five years in the cellular industry with Intel India Pvt.\ Ltd.\ and Apple India Pvt.\ Ltd.\ in their respective LTE/5G-NR base-band modem group.
    His current research is in the area of edge computing and performance optimization with a specific focus on improving the energy efficiency and responsiveness of edge computing systems through optimized sampling.
\end{IEEEbiography}
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/jaya.jpeg}}]%
    Jaya Prakash Champati (IEEE member) received his bachelor of technology degree from the National Institute of Technology Warangal, India in 2008, and master of technology degree from the Indian Institute of Technology (IIT) Bombay, India in 2010.
    He received his Ph.D. in Electrical and Computer Engineering from the University of Toronto, Canada in 2017.
    From 2017--2020, he was a post-doctoral researcher with the division of Information Science and Engineering, EECS, KTH Royal Institute of Technology,Sweden.
    He is currently a Research Assistant Professor at IMDEA Networks Institute, Madrid, Spain.
    His general research interest is in the design and analysis of algorithms for scheduling problems that arise in networking and information systems.
    Currently, his research focus is in Edge Computing/Intelligence, Age of Information, Cyber-Physical Systems (CPS), and Internet of Things (IoT).
    Prior to joining PhD he worked at Broadcom Communications, where he was involved in developing the LTE MAC layer.
    He was a recipient of the best paper award at IEEE National Conference on Communications, India, 2011.
\end{IEEEbiography}
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/bobby.jpeg}}]%
    Roberta L. Klatzky (Fellow, IEEE) received the Ph.D. degree in cognitive psychology from Stanford University, Stanford, CA, USA.
    She is currently the Charles J. Queenan, Jr. University Professor with Carnegie Mellon University, Pittsburgh, PA, USA, where she is a Member of the Department of Psychology, Human-Computer Interaction Institute, and Neuroscience Institute.
    Her research interests include human perception and cognition, with special emphasis on spatial cognition and haptic perception.
\end{IEEEbiography}
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/satya.jpeg}}]%
    Mahadev ``Satya'' Satyanarayanan's multi-decade research career has focused on the challenges of performance, scalability, availability and trust in information systems that reach from the cloud to the mobile edge of the Internet.
    In the course of this work, he has pioneered many advances in distributed systems, mobile computing, pervasive computing, and the Internet of Things (IoT).
    Most recently, his seminal 2009 publication ``The Case for VM-based Cloudlets in Mobile Computing'' and the ensuing research has  led to the emergence of Edge Computing (also known as ``Fog Computing'').
    Satya is the Carnegie Group Professor of Computer Science at Carnegie Mellon University.
    He received the Ph.D. in Computer Science from Carnegie Mellon, after Bachelor's and Master's degrees from the Indian Institute of Technology, Madras.
    He is a Fellow of the ACM and the IEEE.
\end{IEEEbiography}
\newpage%
% \vskip -2\baselineskip plus -1fil%
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bios/james.jpeg}}]%
    James Gross received his Ph.D. degree from TU Berlin in 2006.
    From 2008--2012 he was with RWTH Aachen University as assistant professor and research associate of RWTH's center of excellence on Ultra-high speed Mobile Information and Communication (UMIC).
    Since November 2012, he has been with the Electrical Engineering and Computer Science School, KTH Royal Institute of Technology, Stockholm, where he is professor for machine-to-machine communications.
    At KTH, James served as director for the ACCESS Linnaeus Centre from 2016 to 2019, while he is currently associate director in the newly formed KTH Digital Futures research center, as well as co-director in the newly formed VINNOVA competence center on Trustworthy Edge Computing Systems and Applications (TECoSA).
    His research interests are in the area of mobile systems and networks, with a focus on critical machine-to-machine communications, edge computing, resource allocation, as well as performance evaluation.
    He has authored over 150 (peer-reviewed) papers in international journals and conferences.
    His work has been awarded multiple times, including the best paper awards at ACM MSWiM 2015, IEEE WoWMoM 2009 and European Wireless 2009.
    In 2007, he was the recipient of the ITG/KuVS dissertation award for his Ph.D. thesis.
\end{IEEEbiography}

### 2022-12-13 cover_letter.tex
fixes for IEEE ToMC

\documentclass[10pt, letterpaper]{letter}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[]{todonotes}
\usepackage[]{geometry}
\usepackage[inline]{enumitem}
\usepackage[nolist]{acronym}
\usepackage[]{siunitx}
\usepackage[
  style=numeric-comp,
  sorting=none,
  sortcites,
  hyperref,
  mincitenames=1,
  maxcitenames=2,
  maxbibnames=2,
  minbibnames=1,
  citestyle=numeric-comp, % for [1, 2] instead of [1], [2]
  backend=bibtex
]{biblatex}
\bibliography{bibliography.bib}
% \AtBeginBibliography{\small}
% \AtEveryBibitem{\clearfield{day}}
% \AtEveryBibitem{\clearfield{isbn}}
% % \AtEveryBibitem{\clearfield{url}}
% \AtEveryBibitem{\clearfield{series}}
% \AtEveryBibitem{\clearlist{location}}
% \AtEveryBibitem{\clearfield{doi}}
\input{acronyms.tex}

\signature{
    Manuel {Olguín Muñoz}\\
    Division of Information~Science and Engineering, EECS\\
    KTH Royal Institute of Technology\\
    Malvinas väg 10, 7th floor\\
    100 44 Stockholm, Sweden\\
    \href{mailto:molguin@kth.se}{\url{molguin@kth.se}}
}
% \address{}
\begin{document}

\begin{letter}{
    Qian Zhang, Ph.D.\\
    Editor-in-Chief\\
    IEEE Transactions on Mobile Computing
}
\opening{Dear Dr.\ Zhang:}

Please find enclosed a research article submission to IEEE Transactions on Mobile computing entitled \emph{``Realistic Modeling of Human Timings for Wearable Cognitive Assistance''}.
% We confirm that this work is original and has not been published elsewhere, nor is it currently under consideration for publication elsewhere.

This manuscript presents two main contributions.
We first introduce a stochastic model for human behavior in the context of \ac{WCA}.
This model is built from data collected for our previous work in ``\citefield{olguinmunoz:impact2021}{title}'' (\citeauthor{olguinmunoz:impact2021}, \citefield{olguinmunoz:impact2021}{year}), and represents the first realistic end-to-end modeling approach for \ac{WCA}.
Given the difficulty of scalably and repeatably studying systems such as \ac{WCA} which include a human-in-the-loop, we believe this model to be an important contribution to the development and expansion of this field of research.

Furthermore, we introduce a novel optimization framework for resource consumption-responsiveness trade-offs in these systems.
We show that by combining these two contributions, we are able to obtain significant reductions --- up to \SI{50}{\percent} --- in relevant metrics relating to system load and resource consumption, such as number of processed samples and raw energy consumption per logical application step.
We believe these results provide key insights for developers and researchers working on developing and benchmarking \ac{WCA} applications as well as on cost-efficient strategies for resource allocation in these systems.

We do not have any opposed reviewers or editors.

I will be serving as the corresponding author for the manuscript, assuming all the responsibilities this entails. 
All of the authors listed have agreed to the byline order and to submission of the manuscript, and understand that, if accepted for publication, a certification of authorship form will be required that will be signed by all of us.

\closing{Sincerely,}
\end{letter}
\end{document}

### 2022-12-15 background.tex
modifications to work with thesis

@@
         \centering
-        \includegraphics[width=\columnwidth]{figs/task.png}
+        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/task.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\columnwidth]{figs/step_time.png}
+        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/step_time.png}
         \caption{%
@@
     \centering
-    \includegraphics[width=\columnwidth]{figs/new_model/mu_fits_exgaussian_slice0.png}
+    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/new_model/dist_fits_neuro.png}
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/dist_fits_neuro.png}
     \caption{%

### 2022-12-15 implications_footprint.tex
modifications to work with thesis

@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_all_ttfs.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_all_ttfs.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_diff.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_diff.png}
         \caption{%
@@
     \centering
-    \includegraphics[width=\columnwidth]{figs/EdgeDroid2ExperimentalSetup.png}
+    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/EdgeDroid2ExperimentalSetup.png}
     \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed_diff.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed_diff.png}
         \caption{%

### 2022-12-15 implications_optimization.tex
modifications to work with thesis

@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/sampling_optimization.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/sampling_optimization_diff.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/energy_optimization.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization_diff.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/energy_optimization_diff.png}
         \caption{%

### 2022-12-15 model.tex
modifications to work with thesis

@@
     \centering
-    \includegraphics[width=\columnwidth]{figs/new_model/ttf_to_exectime.png}
+    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/ttf_to_exectime.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=\columnwidth]{figs/new_model/exectime_over_steps.png}
+    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/exectime_over_steps.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{figs/new_model/transitions.png}
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/transitions.png}
     \caption{%
@@
     \centering
-    \includegraphics[width=\columnwidth]{model_data/frame_probabilities.png}
+    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/model_data/frame_probabilities.png}
     \caption{%

### 2022-12-15 body/part1/overview.tex
edgedroid 2 paper!

@@
 
-Utilizing a trace for the emulation of human behavior in \ac{WCA} has multiple advantages.
+We opt for a trace-based approach for our first implementation of the methodology for \ac{WCA} for three key reasons.
 The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
 Using a trace ensures that the same computation is performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
+
 The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
-The third is extensibility.
+Compatible traces can be easily obtained simply by instrumenting existing \ac{WCA} client applications to record all captured inputs.
+
+The final advantage corresponds to extensibility.
+As long as the tasks belong to the same category of \ac{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
 

### 2022-12-18 bib-to-titlecase.py
tweaking references

@@ -1,0 +1,62 @@
+#!/usr/bin/env python
+#
+# Original by Daniel L. Greenwald
+# http://dlgreenwald.weebly.com/blog/capitalizing-titles-in-bibtex
+# Modified by Garrett Dash Nelson
+# https://gist.github.com/garrettdashnelson/af0f8307393da37c6f94eda8c4613a4f
+#
+# Modified by Manuel Olguín Muñoz
+import io
+import re
+from collections import deque
+from typing import Sequence
+
+import click as click
+from titlecase import titlecase
+
+# Match title, Title, booktitle, Booktitle fields
+pattern = re.compile(r'(\W*)([Bb]ook)?([Tt]itle = {+)(.*)(}+,)')
+
+
+class LineError(Exception):
+    def __init__(self, num: int, line: str):
+        super().__init__(f"Error in line num {num:d} (line: \"{line}\")")
+
+
+def to_titlecase(lines: Sequence[str]) -> Sequence[str]:
+    # Search for title strings and replace with titlecase
+    new_lines = deque()
+    for line_num, line in enumerate(lines):
+        # Check if line contains title
+        match_obj = pattern.match(line)
+        if match_obj is not None:
+            try:
+                # Need to "escape" any special chars to avoid misinterpreting them in the regular expression.
+                old_title = re.escape(match_obj.group(4))
+
+                # Apply titlecase to get the correct title.
+                new_title = titlecase(match_obj.group(4))
+
+                # Replace and add to list
+                p_title = re.compile(old_title)
+                newline = p_title.sub(new_title, line)
+            except Exception as e:
+                raise LineError(line_num, line) from e
+
+            new_lines.append(newline)
+        else:
+            # If not title, add as is.
+            new_lines.append(line)
+
+    return new_lines
+
+
+@click.command
+@click.argument("input-file", type=click.File())
+@click.option("-o", "--output-file", type=click.File(mode="w", lazy=True, atomic=True), default="-", show_default=True)
+def main(input_file: io.StringIO, output_file: io.StringIO):
+    output_file.writelines(to_titlecase(input_file.readlines()))
+
+
+if __name__ == '__main__':
+    main()

### 2022-12-19 body/part1/overview.tex
fix biblatex

--- 2022-12-09 body/part1/overview.tex
@@
 \chapter{Key contributions and results}\label{chap:contributions}
-
-\section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.8\textwidth]{Figs/methodology.png}
-    \caption{%
-        Overview of the proposed methodology for the study of feedback-loop systems on edge computing infrastructure.
-        The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
-        The backend/server side of the system along with the network remain unchanged.
-    }\label{fig:methodology}
-\end{figure}
-
-As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
-In the following, we introduce our methodological framework for tackling this challenge.
-
-A general overview of our methodology is illustrated in \cref{fig:methodology}.
-It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
-We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \ac{COTS} general-purpose computing devices.
-The backend side and the network are not changed.
-
-Two main arguments drive this approach.
-One, the difficulty of scaling the number of clients in these systems in a research context.
-The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \ac{WCA}) or due its cyber-physical nature (in \acp{NCS}).
-Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
-Raspberry Pis, Beaglebones, etc.).
-
-And two, it maintains the realism of the most complex parts of the system.
-
-\todo[inline]{Fix and finish}
-
-In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
-We first discuss the methodology in the context of a human-in-the-loop application, \acl{WCA}, in section \cref{ssec:methodology:wca}.
-Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \acf{NCS}.
-
-\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \includegraphics[width=.8\textwidth]{Figs/trace_edgedroid.png}
-        \caption{%
-            High level conceptual design of our methodology for \ac{WCA}.
-            We replace the human by a trace-driven emulation.
-            This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
-        }
-    \end{subfigure}
-    \medskip
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \adjustbox{scale=0.75}{
-            \begin{tikzpicture}[align=center,
-            node distance=.5cm and 1.5cm,
-            every initial by arrow/.style={-{Latex[length=2mm]}}]
-            % Place nodes              
-            \node [initial, state, minimum size=6em, initial text=] (play) {Play};               
-            \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
-            \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
-            \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};
-
-            % Draw edges
-            \path[draw, -{Latex[length=2mm]}]
-            (play) edge [bend right=20] node[left] {Step done} (rewind)
-            edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
-            edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)
-
-            (change) edge [bend left=20] node[right] {Step changed} (play)
-            edge [bend left=20] node[right] {All steps done} (shutdown)
-
-            (rewind) edge [bend right=20] node[right] {Rewound} (play)
-            edge [bend right=20] node[right] {Too many rewinds} (shutdown);
-
-            \end{tikzpicture}
-        }
-        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.}\label{fig:usermodel}
-    \end{subfigure}
-    \caption{%
-        Overview of our first implementation of the methodology for \ac{WCA}.
-    }\label{fig:edgedroid1:trace}
-\end{figure}
-
-We first introduce our methodology in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
-The former corresponds to an extended abstract paper which discusses a high level overview of our approach; the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
-
-The overall goal of these papers is to explore this methodological approach as applied to \ac{WCA}.
-For this purpose, we introduce a tool, \emph{EdgeDroid 1.0}, which mimics the behavior of a human interacting with a \ac{WCA} client application.
-
-To establish a reproducible and comparable workload, we use a trace-based design for the client-side emulation.
-We first record a trace of the sensory input data while having a human user operate the target application.
-The collected data consists of the sensory inputs provided to the system at runtime, for instance --- in the case of a video-based application --- the video feed from the camera.
-
-To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
-This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
-
-However, merely replaying the trace is not enough, as disparities between system responsiveness at recording versus replay time can cause the trace so drift out of synchronization with the current state of the system.
-Thus, we introduce here the concept of a human user model which is able to adapt its behavior in a realistic manner based on the current responsiveness of the system by systematically controlling the replay of the trace.
-The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
-The trace is preprocessed into segments that correspond to individual steps of the \ac{WCA} application; this is performed manually as a preparation step.
-The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \ac{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
-On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
-To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.
-
-\todo[inline]{EdgeDroid paper}
-
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
-    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
-    \label{fig:comparison:rtt}%
-\end{figure}%
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
-    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
-    \label{fig:comparison:feedback}%
-\end{figure}%
-\begin{figure}%
-    \centering%
-    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
-    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
-    \label{fig:comparison:tasksteps}
-\end{figure}%
-
-The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
-We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
-
-Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
-This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
-EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
-This sensory information is then processed by the original compute process at the backend, generating feedback.
-However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
-Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
-Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
-Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
-It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
-To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications
-
-\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
-Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
-We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
-Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
-These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
-Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.
-
-In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.
-
-For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
-This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.
-
-We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
-These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
-It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.
-
-We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
-One, scaling behavior for the application seems to be linear with respect to the number of clients.
-Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
-
-It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
-Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.
-
-EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
-From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
-Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
-Here, the downlink time sometimes almost reaches the ideal bound by itself.
-A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.
-
-Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
-The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.
-
-\todo[inline]{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
-
-\subsection{Applying the methodology to \acsp{NCS}}\label{ssec:methodology:ncs}
-
-\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
-
-In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
-We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
-\begin{enumerate}
-    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
-          The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
-          Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
-          In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
-          Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
-          The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
-\end{enumerate}
-
-We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
-We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
-    \item similar setups exist in real-world industrial use
-    \item video analytics has long been proposed as a ``killer app'' for edge computing.
-\end{enumerate*}
-Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
-
-\todo[inline]{Some plots and results}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
-    \caption{
-        The 2D inverted pendulum system.
-    }\label{fig:invpend}
-\end{figure}
-
-\begin{figure}[t]
-    \centering
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
-        \caption{Toppled plants.}\label{fig:video:toppled}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
-    \end{subfigure}\\
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
-        \caption{Packet losses.}\label{fig:video:drop}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\acsp{RTT}.}\label{fig:video:rtt}
-    \end{subfigure}%
-    \caption{
-        Multi-loop, resource-constrained setup results.
-        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
-    }\label{fig:video:results}
-\end{figure}
-
-\FloatBarrier%
-
-\subsection{Orchestrating experimentation in edge computing testbeds}
-
-\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
-
-In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
-Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
-The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
-It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \ac{FOSS}.
-It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
-
-In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
-Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
-    \item provisioning of and connection to remote cloud instances
-    \item initialization of \ac{IP} layer connectivity between hosts
-    \item collection of logs and data
-    \item deployment, scaling, and lifecycle management of containerized processes
-\end{enumerate*}.
-We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
-We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
-We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
-
-\todo[inline]{Some plots}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
-    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
-    \caption{Network structure assumed by Ainur.}\label{fig:network}
-\end{figure*}
-
-\section{Extending the methodology in the context of \acs{WCA}}
-
-\subsection{Characterizing human behavior}
-
-\todo[inline]{\cref{paper:olguinmunoz2021impact} describes the acquisition of the necessary data and insights for the models.}
-
-\todo[inline]{From introduction}
-
-We intended to answer four core research questions relating to human responses to decreased application responsiveness.
-
-\begin{itemize}
-    \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}
-
-          In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
-          The extent or form of these changes were however unknown.
-          We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.
-
-          Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
-          While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
-          To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-          Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.
-
-    \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}
-
-          We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.
-
-          The results we present, however, seem to refute this hypothesis.
-          We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as system responsiveness was altered.
-
-    \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}
-
-          In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
-          In particular, we expected emotional effects to be correlated with the strength of the added delay.
-
-          The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
-          This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
-
-    \item \emph{Are these effects mediated by personality indicators in any way?}
-
-          Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
-          We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
-
-          The results obtained agree with our hypothesis.
-          We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.
-
-\end{itemize}
-
-% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
-% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
-% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
-% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
-
-We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
-One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and direct implications for resource and power consumption.
-Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished --- this has potential consequences for resource allocation strategies.
-Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource allocation across applications potentially looks different from what could be considered ``fair''.
-
-Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
-These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
-Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
-Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
-
-\todo[inline]{From discussion}
-
-We start our discussion with the main results of our experimentation presented in the previous section:
-
-\begin{itemize}
-    \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional} behavioral slow-down.
-          That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their execution of the task.
-          This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
-          Moreover, there is a temporal component to this effect; users become progressively slower the more time passes with reduced system responsiveness.
-
-    \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
-          This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
-          The question of whether any lingering effect can be measured after these four steps remains open.
-
-    \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at performing steps as the task progresses.
-          However, the strength of this effect decreases as delay increases.
-          Whereas for blocks without delay users performed the last four steps of a 12-step block on average \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.
-
-    \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay.
-          The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
-          Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables related to attentiveness, respectively.
-          In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
-          The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
-          These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
-\end{itemize}
-
-A central question therefore arises: to which physio- and psychological mechanisms can these findings, most importantly the substantial slow-down in task execution, be attributed?
-
-In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional reactions.
-These could in turn elicit generalized arousal.
-We also postulated that adapting to delay might progressively deplete cognitive resources in users.
-However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
-Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-The acceleration data further did not indicate that extended delay significantly increases erratic movement.
-% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-% The acceleration data further do not indicate that extended delay increases erratic movement.
-To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
-That is, users fail to capitalize on the new information as quickly as they could.
-Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.
-We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
-As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
-Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
-This also provides a possible explanation for the lingering effects of delay after an acceptable system responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.
-
-As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
-However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.
-
-The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.
-
-\begin{itemize}
-
-    \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to significantly extended application lifetimes.
-          This has direct implications for resource and power consumption.
-
-    \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a high-responsive state could have unconventional consequences for resource allocation.
-          This is of particular importance, for instance, for cases where the user may be able to finish the task before these effects subside.
-          In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired application.
-
-    \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of system responsiveness across applications may not ultimately be beneficial to the system as a whole.
-          Take for instance two applications on the same system which negatively interfere with each other.
-          The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop.
-          In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.
-
-    \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
-          This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.
-          However, this remains an open challenge.
-
-\end{itemize}
-
-To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of WCA deployments.
-Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable tool to tackle inefficiencies in these systems.
-Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the relationship between application responsiveness and human behavior.
-More research in this area will surely uncover more complex and interesting behaviors.
-Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
-These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
-
-\todo[inline]{Some diagrams, plots, and results}
-
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
-        \caption{Structure of a step in a generic cognitive assistance application.
-            The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
-            % by the assistant, as they do not contain relevant information.
-            Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
-            % which will subsequently be provided to the user.}%
-        }
-        \label{fig:cogassist:step}
-    \end{subfigure}%
-    % \medskip%
-    \hfill%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
-        \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
-        \label{fig:cogassist:step:delay}
-    \end{subfigure}
-    \medskip%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
-        \caption{Structure of a block in the experimental task.}%
-        \label{fig:cogassist:block}
-    \end{subfigure}%
-    % \medskip%
-    \hfill%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
-        \caption{Visualization of the execution time of a step.}%
-        \label{fig:exectime:diagram}
-    \end{subfigure}%
-    % \medskip%
-    \caption{Components of the cognitive assistance task.}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
-    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
-    \caption{Mean per-step execution time vs.\ delay, by step slice.
-        Error bars indicate S.E.M.}
-    \label{fig:exectime:delay:slice}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
-    \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
-\end{figure}
-
-\begin{table}[h]
-    \centering
-    \caption{Principal Component Analysis}\label{tab:pca}
-    \begin{subtable}[h]{\textwidth}
-        \centering
-        \caption{Main components identified.}
-        \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-            \toprule
-            \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} \\
-            \midrule
-            BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
-            BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
-            ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
-            ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
-            Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
-            Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
-            \bottomrule
-        \end{tabular*}
-    \end{subtable}
-    \newline
-    \medskip
-    \newline
-    \begin{subtable}[h]{\textwidth}
-        \centering
-        \caption{Percentages of variance explained by the components.}
-        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
-            \toprule
-            {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total} \\
-            \midrule
-            Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent} & \SI{73.13}{\percent} \\
-            \bottomrule
-        \end{tabular*}
-    \end{subtable}
-\end{table}
-
-\subsection{Emulating human behavior}
-
-We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
-Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
-We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-
-The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic human-in-the-loop nature.
-Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
-Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming and prohibitively expensive for many research groups.
-
-In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \ac{WCA}, the first data-driven model for human timings in \ac{WCA} applications.
-This model represents a stochastic approach to execution time modeling which builds upon the data collected for \textcite{olguinmunoz:impact2021}.
-Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
-
-\todo[inline]{TODO: Some relevant results. Need to add paper to thesis.}
-
-\section{Implications for the optimization of \ac{WCA}}
-
-\begin{enumerate}
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
-    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
-    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
-    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
-\end{enumerate}
-
-Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
-We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
-
-Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
-We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy.
-We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
-Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
+\input{body/part1/overview/methodology}
+\input{body/part1/overview/extending-wca}
+\input{body/part1/overview/implications-optimization}

### 2022-12-19 body/part1/overview/extending-wca.tex
fix biblatex

--- 2022-08-31 publications/2021ImpactDelayedResponse/body.tex
@@
-\section*{Abstract}
+\section{Extending the methodology in the context of \acs{WCA}}
 
-Wearable Cognitive Assistants (WCA) are anticipated to become a widely-used application class, in conjunction with emerging network infrastructures like 5G that incorporate edge computing capabilities.
-While prototypical studies of such applications exist today, the relationship between infrastructure service provisioning and its implication for WCA usability is largely unexplored despite the relevance that these applications have for future networks.
-This paper presents an experimental study assessing how WCA users react to varying end-to-end delays induced by the application pipeline or infrastructure.
-Participants interacted directly with an instrumented task-guidance WCA as delays were introduced into the system in a controllable fashion.
-System and task state were tracked in real time, and biometric data from wearable sensors on the participants were recorded.
-Our results show that periods of extended system delay cause users to correspondingly (and substantially) slow down in their guided task execution, an effect that persists for a time after the system returns to a more responsive state.
-Furthermore, the slow-down in task execution is correlated with a personality trait, neuroticism, associated with intolerance for time delays.
-We show that our results implicate impaired cognitive planning, as contrasted with resource depletion or emotional arousal, as the reason for slowed user task executions under system delay.
-The findings have several implications for the design and operation of WCA applications as well as computational and communication infrastructure, and additionally for the development of performance analysis tools for WCA.\@
+\todo[inline]{%
+    Why do we need to extend the methodology? What was wrong with the approach in the previous section?
+}
 
-\section{Introduction}
+\subsection{Characterizing human behavior}
 
-Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
-They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
-Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
-Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
-An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
+\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on
+human behavior in step-based \ac{WCA}.
+This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \ac{WCA}, first introduced
+in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human
+behavior.
 
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
-Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
-Detethering this assistance from its current fixed location will surely make it available to many more fields.
+We employed a modified version of the step-based LEGO cognitive assistant first introduced by \citeauthor{chen2015early}
 
-Based on these use cases, we identify three main requirements for WCA:\@
-\begin{enumerate}
-    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
+\todo[inline]{\cref{paper:olguinmunoz2021impact} describes the acquisition of the necessary data and insights for the
+models.}
 
-    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
-    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+\todo[inline]{From introduction}
 
-    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
-    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
-    This means that users will have expectations of constant, immediate feedback as they progress through the task.
-    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
-    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
-\end{enumerate}
-
-\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
-\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
-Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
-However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
-\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.
-
-% However, additional challenges remain in the path toward widespread adoption.
-\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
-We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
-\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
-% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
-Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}.
-Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
-A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
-% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.
-
-This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
-We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
-While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
-We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
-This is an important step toward a more systematic understanding of human behavior in this domain.
-
-We present in this paper the design and elaboration of an experimental WCA test-bed.
-This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
-Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
-The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
-
-Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
+We intended to answer four core research questions relating to human responses to decreased application responsiveness.
 
@@
 
-  In line with previous research in this area, we expected subjects to change their temporal profiles as system responsiveness decreased.
+    In line with previous research in this area, we expected subjects to change their temporal profiles as system
+    \item responsiveness decreased.
     The extent or form of these changes were however unknown.
-  We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by subjects.
+    We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by
+    \item subjects.
 
     Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
-  While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
-  To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-  Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system responsiveness improves.
+    While it would seem self-evident that users take longer to complete a task while using a system with low
+    \item responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down
+    \item represents a source of substantial additional delay.
+    To be more precise, the data indicate that users slow down not only because they have to wait for the system to
+    \item catch up, but that their reactions to new instructions is also delayed.
+    Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system
+    \item responsiveness improves.
 
@@
 
-  We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added annoyance of dealing with an unresponsive system.
+    We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added
+    \item annoyance of dealing with an unresponsive system.
 
     The results we present, however, seem to refute this hypothesis.
-  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.
+    We were not able to detect any significant effects on the physiological signals obtained from the biometric
+    \item sensors as system responsiveness was altered.
 
@@
 
-  The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
+    The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to
+    \item a disruption of participants' cognitive plan for the task and not to an emotional response.
     This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
@@
 
-  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
+    Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in
+    \item mediating these effects, as it has previously been connected to intolerance for time
+    \item delay~\cite{hirsh2008delay}.
     We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
@@
     The results obtained agree with our hypothesis.
-  We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were found to play a role through factor analysis.
+    We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were
+    \item found to play a role through factor analysis.
 
@@
 
-% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
+% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive
+% plan for the task.
 % This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
-% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
-% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
-% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
+% While it would seem self-evident that users take longer to complete a task while using a system with low
+% responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down
+% represents a source of substantial additional delay.
+% To be more precise, the data indicate that users slow down not only because they have to wait for the system to
+% catch up, but that their reactions to new instructions is also delayed.
+% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic
+% personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
 
 We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
-One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and direct implications for resource and power consumption.
-Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished --- this has potential consequences for resource allocation strategies.
-Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource allocation across applications potentially looks different from what could be considered ``fair''.
+One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and
+direct implications for resource and power consumption.
+Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished ---
+this has potential consequences for resource allocation strategies.
+Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource
+allocation across applications potentially looks different from what could be considered ``fair''.
 
-Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
-These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
-Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
-Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
+Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of
+WCA.\@
+These results represent unexpected, valuable findings, which can be employed to model and understand how users
+interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
+Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation
+tools such as our previous work in \cite{olguin:2018, olguin:2019}.
+Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of
+realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
 
-The structure of this paper is as follows.
-We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
-\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
-Then, in \cref{sec:results} we detail the results of our experiment.
-Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
+\todo[inline]{From discussion}
 
-\section{Background and Related Work}\label{sec:background}
+We start our discussion with the main results of our experimentation presented in the previous section:
 
-% Overarching theo construct: people perform temp seq task in the framework of cognition and emotion and individual differences mediate response to delay.
-In particular we view cognition as described by models such as ACT-R as a seq of procedures.
+\begin{itemize}
+    \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional}
+    \item behavioral slow-down.
+    That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their
+    \item execution of the task.
+    This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on
+    \item average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
+    Moreover, there is a temporal component to this effect; users become progressively slower the more time passes
+    \item with reduced system responsiveness.
 
-% Section 2.
-% Accordingly we review time perception...
+    \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness
+    \item \emph{remain} for at least a few steps after system responsiveness improves.
+    This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following
+    \item a high-delay block, as pictured in \cref{fig:exectime:transition}.
+    The question of whether any lingering effect can be measured after these four steps remains open.
 
-% Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
-% Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
-% In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
+    \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at
+    \item performing steps as the task progresses.
+    However, the strength of this effect decreases as delay increases.
+    Whereas for blocks without delay users performed the last four steps of a 12-step block on average
+    \item \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.
 
-Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
-In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
-Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
-
-Accordingly, in the following we review background work in the areas of 
-\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-  \item[(\cref{ssec:timeperception})] time perception
-  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
-\end{enumerate*}
-
-\subsection{Time perception in computing systems}\label{ssec:timeperception}
-
-The question of how people respond to delay in a computer system is grounded in how people perceive time.
-Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-The estimation of an event's duration varies with context of various types
-\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-    \item arousal can expand durations~\cite{droit2011emotion}.
-\end{enumerate*}
-
-Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{Loewenstein1992anomaliesintertemporalchoice}.
-Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-Violations of expectancy have a strong impact on the acceptability of computer systems.
-Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-In immersive systems like WCA, which aim to provide seamless interaction, delays are particularly noticeable.
-
-It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowsky:2011:40years}.
-However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.
-
-For WCA the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-They then proposed a latency tolerance zone according to the task demands.
-For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
-Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowsky:2011:40years}.
-
-\subsection{Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}
-
-While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood. 
-These mechanisms could be cognitive or emotional in origin.
-
-A first possible explanation comes from research on cognitive and motor planning. 
-Delay may move users from relatively automatic to more attention-demanding processing.
-Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-
-Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-
-A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-
-Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-
-The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-
-Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of WCA, which are intended for relatively inexperienced users of an application.
-These and other measures of individual variation were considered here.
-
-\section{Experimental Design}\label{sec:experimentaldesign}
-
-The core elements of our experiment are shown in \cref{fig:experimentaltestbed}:
-\begin{itemize}
-    \item Subjects interact with a WCA while wearing an array of biometric sensors.
-    \item The \emph{responsiveness} of the application, i.e.\ the interval of time between an input being provided to the system and the associated output returned to the user, is manipulated in real time.
-    The effects of these manipulations on the subjects are recorded and subsequently analyzed.    
+    \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response
+    \item to delay.
+    The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of
+    \item neuroticism and both measures of immersion: focus and involvement.
+    Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables
+    \item related to attentiveness, respectively.
+    In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive
+    \item and involved in the task.
+    The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures
+    \item did not correlate with outcomes in other computer-intensive environments such as immersive
+    \item VR~\cite{quesnel2018you}.
+    These correlations are also consistent with previous findings indicating that individuals scoring high in
+    \item neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
 \end{itemize}
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig1.eps}
-  \caption{Experimental test-bed.
-  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
-  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
-  \label{fig:experimentaltestbed}
-\end{figure}
+A central question therefore arises: to which physio- and psychological mechanisms can these findings, most
+importantly the substantial slow-down in task execution, be attributed?
 
-This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board under record number \texttt{STUDY2019\_00000247}. 
-Written consent was obtained.
+In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional
+reactions.
+These could in turn elicit generalized arousal.
+We also postulated that adapting to delay might progressively deplete cognitive resources in users.
+However, the present data provide relatively little support for these alternatives, in that the predicted measures
+did not produce the expected statistical trends.
+Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\
+short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+The acceleration data further did not indicate that extended delay significantly increases erratic movement.
+% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays,
+% and speed-induced errors and non-completions predicted by resource depletion were not observed.
+% The acceleration data further do not indicate that extended delay increases erratic movement.
+To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
+That is, users fail to capitalize on the new information as quickly as they could.
+Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control
+mechanisms, and not from emotion or resource depletion.
+We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
+As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been
+modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
+Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based
+control at the step-by-step level that is easily diverted.
+This also provides a possible explanation for the lingering effects of delay after an acceptable system
+responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.
 
-Subjects were recruited from a pool of undergraduate students at Carnegie Mellon University.
-Students enrolled in an introductory-level psychology course fulfill a research requirement as part of the plan of study.
-This requirement may be fulfilled either through the elaboration of a written essay or by volunteering as a participant in a small number of research experiments.
+As to the applicability of our findings to other applications, it must be noted that these results pertain to a
+specific class of applications, namely step-based task-guidance WCA.\@
+However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of
+seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance
+the state.
 
-No particular exclusion criteria were applied, and as specified by our approved data-collection protocol, no gender- or sex-related statistics were collected.
-In total, 40 participants were recruited, all of them of college student age (\numrange{18}{25} years old).
+The results here presented provide a number of possible implications for WCA system design and optimization, both for
+single and multi-application flows.
 
-\subsection{The Cognitive Assistance Application}
+\begin{itemize}
 
-We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{Chen:EarlyImplementation}.
-This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
-% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
-The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
-Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
-In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
-% The results of the user performing these instructions are provided as inputs to the system.
-% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.
+    \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to
+    \item significantly extended application lifetimes.
+    This has direct implications for resource and power consumption.
 
-% In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
-In the following, we will provide some formal definitions relating to the WCA application.
-% \theoremstyle{definition}
-% \begin{definition}[Task and subtask]
-A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
-\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{fig:task:steps}).
-% We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
-A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
-At that point, the assistant either provides a new instruction to the user, or the task ends.
-  % See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
-% \end{definition}
+    \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a
+    \item high-responsive state could have unconventional consequences for resource allocation.
+    This is of particular importance, for instance, for cases where the user may be able to finish the task before
+    \item these effects subside.
+    In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired
+    \item application.
 
-\begin{figure*}[h]
-  \centering
-  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig2.eps}
-  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
-\end{figure*}
+    \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of
+    \item system responsiveness across applications may not ultimately be beneficial to the system as a whole.
+    Take for instance two applications on the same system which negatively interfere with each other.
+    The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn
+    \item causes them to interfere even longer, potentially entering a positive feedback loop.
+    In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead
+    \item to resources being freed up faster system-wide.
 
-% \begin{definition}[Step]
+    \item Based on our findings relating individual differences between users and their sensitivity to delays, it
+    \item might also be possible to extrapolate user characteristics from measured execution times.
+    This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users
+    \item with a higher sensitivity to system-state degradation.
+    However, this remains an open challenge.
 
-% While the subtask remains unfinished, the assistant discards results of the processing of the sampled inputs.
-% At the beginning of a step, an instruction is given to the user.
-% The user then proceeds to perform the subtask specified by the instruction, while the cognitive assistant is continuously sampling the subtask state at specified intervals.
+\end{itemize}
 
-% Once the user finishes the required action, the next sample taken will contain a valid input, and thus the cognitive assistant will provide a new instruction. 
-% This finishes the current step and potentially begins a new one if there remain instructions to be performed in the task.
-% \end{definition}
+To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of
+WCA deployments.
+Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable
+tool to tackle inefficiencies in these systems.
+Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the
+relationship between application responsiveness and human behavior.
+More research in this area will surely uncover more complex and interesting behaviors.
+Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that
+might be constructed under existing architectures like ACT-R.
+These same parameters could be used to modulate the timing and generation of inputs in trace-based workload
+generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same
+trace to generate workloads for a multitude of different user profiles.
 
-In the base LEGO Assistant, the task consists of the assembly of a LEGO model; each step requires the user to append a LEGO brick to the model at a specific location and orientation.
-The system monitors progress through a video feed and provides timely feedback in the form of visual and textual instructions to guide the user.
-
-The LEGO assistant has features that make it a good target for assessment of delay effects in a cognitive assistant.
-In particular
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-    \item it is easily understood by users and requires essentially no training
-    \item its step-by-step nature simplifies the isolation of the relevant experimental variables and the effects of the delay
-    % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
-    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
-\end{enumerate*}
-
-\begin{figure*}[h]
-  \centering
-  % \input{plots/lego_hierarchy_diagram.tex}
-  \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig3.eps}
-  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
-  \label{fig:lego:hierarchical}
-\end{figure*}
-
-% The original design of the LEGO assistance application was based on a client-server model communicating over a wireless network, with the client software running on a wearable device and the server software deployed on a cloudlet.
-For the purposes of this study, the original client-server design of the LEGO Assistant was altered to be locally executable.
-This was done in order to eliminate the stochastic effects of jitter and latency from the network link, and greatly simplify instrumentation.
-% By this means, we achieve much more fine-grained control over the latencies to which the system is subject.
-% Additionally, this greatly simplified the instrumentation of the application.
-Instructions were output in image and text form to a computer display situated on a table directly in front of the participants.
-Participants performed the instructions on the table, these actions being captured by a high-definition camera located on top of the display.
-
-% Finally, a data collection and experimental layer was implemented between the user interface and video capture and the core processing component of the LEGO Assistant.
-% This layer controlled the manipulated experimental variables and recorded the measured biometric and task-related effects.
-
-\subsection{The Experimental LEGO Task}
-
-For the experimental LEGO task, a key modification was made to the structure of the steps.
-After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
-The length of this delay is one of two independent variables we manipulated for the experiment.
-Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
-% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
-% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
-Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
-% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.
-
-In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
-% \begin{definition}[Block]
-%   A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
-The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
-We used values of \numlist{4;8;12} steps, chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
-Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block.
-% ; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
-% \end{definition}
-
-To assign tasks to the participants, a pseudo-random permutation of the combinations of block length and delay was first generated, and a unique sequence of steps assigned to each of these combinations to create 21 unique blocks.
-A \emph{Latin~square}-type design was then used to reorder this initial permutation in order to generate a task for each participant.
-This ensures a counterbalance of the order of the blocks across participants and avoids systematic learning effects. 
-% The design rotates the block types (as defined by length and delay) across participants so that each type is tested in each ordinal position, but it only coarsely samples from the \( 21 \times 20 \) possible sequences from one block type to another.
-Note that unlike the base LEGO assistant task, in which instructions led participants through the assembly of a specific model, the experimental LEGO task consisted of a sequence of instructions to either add or remove pieces with no evident goal.
-% Users were directed to either add a piece to the ongoing model or to remove a piece, and blocks were designed in such as way so as to ensure that the transitions between them were invisible to the user.
-
-% In this paper we will thus consider blocks to be our basic element of study, and most aggregations will be done at this level (with a few exceptions).
-% For this, we will need additional definitions:
-
-% \begin{definition}[Delay associated with a block]
-%   We will refer to the delay of a block as the delay applied to every frame of every step in that block.
-% \end{definition}
-
-% \begin{definition}[Execution time]
-
-Given the variability in the system latency to detect step completion, correcting step-completion time for system response time by a fixed amount would not enable sufficient precision.
-We therefore define the \emph{execution time} empirically for an individual step as the total time between the user receiving the instruction for the step and the time instant of the sample capturing the completed step.
-% their presenting the completed subtask to the system.
-
-% We define the \emph{execution time} of a step as the time between the user receiving an instruction for a subtask and them presenting the completed subtask to the system.
-% This, however, could not be measured directly and has to be inferred from the records of inputs processed by the system.
-
-For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{fig:exectime:diagram}).
-\( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
-If we define \( t_{c} \) as the instant marking when the user presented the finished step to the system, it must follow that \( t_{n-1} < t_{c} < t_{n} \).
-Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
-However, this introduces no bias in the results, as we can assume that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
-We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
-% This is also aggregated into an average at the block level.
-
-% \begin{figure}
-%   \centering
-%   \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
-%   \caption{Visualization of the execution time of a step.}%
-%   \label{fig:exectime:diagram}
-% \end{figure}
-
-% \end{definition}
+\todo[inline]{Some diagrams, plots, and results}
 
@@
         \caption{Structure of a step in a generic cognitive assistance application.
-    The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
+        The assistant provides an instruction to the user and continuously samples the step state; inputs captured
+        while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new
+        instruction).
         % by the assistant, as they do not contain relevant information.
-    Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
+        Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new
+        instruction.
         % which will subsequently be provided to the user.}%
@@
         \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
-    \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
+        \caption{In the experimental task, an additional variable segment of time is introduced immediately following
+        the processing of the input frame in order to extend the perceived processing time of the input to a specific
+        target delay.}%
         \label{fig:cogassist:step:delay}
@@
 
-\subsection{Collected Data}
-
-The collected data from the experiments fall into four categories: behavioral and personality indicators, \emph{frame-to-frame} metrics, biometric data, and video recordings.
-
-\subsubsection{Behavioral and Personality Indicators}
-
-Before beginning the experimental procedure, participants were asked to fill out two questionnaires.
-
-The first of these, the Big Five Inventory of Personality (\emph{BFI}~\cite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
-Of these, extroversion and neuroticism have been related to tolerance for delay~\cite{hirsh2008delay}.
-
-The second survey, the Immersive Tendencies Questionnaire (\emph{ITQ}~\cite{witmer1998:itq}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
-These questions use a 7-point horizontal scale  with opposing descriptors anchoring  the ends.
-Participants were asked to mark the appropriate point in the scale, and these responses were converted to a numerical value between 1 and 7 for processing.
-
-In post-processing, the obtained scores for both questionnaires were normalized to fall in the \( [0, 1] \) range for ease of interpretation. 
-See \cref{tab:qscores} for their means and standard deviations.
-
-\begin{table}[h]
-  \centering
-  \caption{Means and standard deviations of normalized questionnaire scores.}\label{tab:qscores}
-  \setlength{\tabcolsep}{0pt} 
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}clrr@{}}
-    \toprule
-    \textbf{Questionnaire} & \textbf{Metric} & \textbf{Mean} & \textbf{SD}\\
-    \midrule
-    \multirow{5}{*}{BFI} 
-        & Agreeableness &               0.705 &                  0.136 \\
-        & Conscientiousness &               0.562 &                  0.149 \\
-        & Extroversion &               0.491 &                  0.180 \\
-        & Neuroticism &               0.524 &                  0.230 \\
-        & Openness &               0.677 &                  0.152 \\
-    \cline{1-4}
-    \multirow{4}{*}{ITQ} 
-        & Focus &               0.626 &                  0.126 \\
-        & Games &               0.463 &                  0.261 \\
-        & Involvement &               0.568 &                  0.175 \\
-        & Total &               0.540 &                  0.092 \\
-    \bottomrule
-  \end{tabular*}
-\end{table}
-
-\subsubsection{Frame-to-frame metrics}
-During the execution of the task, we logged every event occurring in the application pipeline.
-Each incoming frame (including frames discarded by the assistant), as well as its associated outputs, was logged at multiple points in the process along with associated metadata such as currently implemented delay as specified by the experimental design.
-This allowed us to extract metrics relating to the performance of the task, such as the time spent by participants on particular steps, any mistakes made, etc.
-In particular, it made possible the easy segmentation of the other time-series data we collected into our main unit of analysis, the aforementioned \emph{block}.
-
-\subsubsection{Biometric Data}
-The participants wore devices to acquire four physiological measures:
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-  \item galvanic skin response (\emph{GSR})
-  \item accelerometer data from the dominant wrist
-  \item brain activity in the form of electroencephalography (\emph{EEG})
-  \item heart rate.
-\end{enumerate*}
-
-These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\cite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
-GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\cite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
-EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\cite{Antonenko2010,Grimes2008,kumar2016measurement}.
-
-Wrist acceleration, GSR and heart rate data were obtained using the Empatica E4~\cite{empatica:e4} biosensing wristband.
-Accelerometer data was sampled at \SI{32}{\hertz},
-GSR was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} (BVP) signal sampled at \SI{64}{\hertz}.
-Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach a stable equilibrium and establish a baseline for the signals.
-
-The E4 wristband was chosen due to its small, non-invasive, and wireless form factor (samples were streamed to the system over Bluetooth LE) and for the fact that its use in research has been experimentally validated in previous studies~\cite{ragot2017emotion, mccarthy2016validation}.
-The E4 also includes a skin temperature thermometer, however the measure was not used for the present study.
-
-For the EEG data we employed the OpenBCI EEG Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
-It provides a quick and non-invasive way of obtaining EEG signals from participants.
-Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
-The signals were sampled at \SI{200}{\hertz} and postprocessed to 
-\begin{enumerate*}[itemjoin={{, }},
-                  itemjoin*={{, and }},
-                  label={{(\arabic*)}}]
-  \item remove frequencies outside the \SIrange{0.1}{40}{\hertz} range
-  \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
-\end{enumerate*}
-
-\subsubsection{Video recordings}
-
-During the task, participants were recorded by two separate cameras.
-One camera was angled downwards, towards the table, the LEGO board, and the participant's hands.
-This camera was used to capture the necessary inputs for the LEGO assembly task as well as to record the actions performed by the user.
-The second camera was angled horizontally, towards the participant, in order to record their facial expressions during the execution of the task.
-
-Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
-Examples of the captured frames can be seen in \cref{fig:sampleframes}.
-The video feeds were not used for the present study, but may be utilized in future analysis.
-
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}[t]{.35\textwidth}
-    \centering
-    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5a.eps}
-    \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
-  \end{subfigure}%
-  % \par\bigskip
-  % \hspace{.04\textwidth}%
-  \hfill%
-  \begin{subfigure}[t]{.60\textwidth}
-    \centering
-    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5b.eps}
-    \caption{Frame from the board recording of the same participant.}
-  \end{subfigure}
-  \caption{Sample frames captured during the experiment.}\label{fig:sampleframes}
-\end{figure}
-
-\section{Results}\label{sec:results}
-
-The results consist of execution time per step. and the outcome of physiological variables: heart rate, GSR, and EEG.\@
-Each will be discussed in turn.
-%The temperature data could not be analyzed due to a progressive increase in the sensor data over the course of the session, which obscured effects of the experimental variables.
-
-\subsection{Execution Time}
-
-% Before describing the analysis and results relating to execution time, it should be noted that accuracy for this metric was essentially perfect.
-% No mistakes were made by participants during the execution of the task and no major outliers could be identified in the samples.
-
-Before describing the analysis and results relating to execution time, it should be noted that participants' performance during the execution of the task was error free; all steps were completed as instructed.
-
-\cref{fig:exectime} shows the mean per-step execution time per block, averaged over block length (number of steps) and artificial delay.
-We can clearly see a trend for the execution time to increase with the delay, increasingly so for longer blocks. 
-Since the per-step execution time  compensates for the added delay in the measure \emph{per se}, this trend must result from the participants' behavioral adjustment to the delay.
-This leads to one of the key outcomes from this study: 
-\emph{participants tend to act more slowly on steps affected by longer delays} --- i.e.\ there is evidence of a pacing effect in users' behavior with respect to the responsiveness of the system.
-
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
-% An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
-% The degrees of freedom represent the number of observations going into each of these variability measures.
-% As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
-% The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
-% Conventionally \( p < .05 \) is the criterion for significance.
-% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
-
-% \begin{figure}[h]
-%     \centering
-%     \includegraphics[width=.8\textwidth]{plots/exec_time_delay_length.eps}
-%     \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
-% \end{figure}
-
 \begin{figure}[h]
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
-    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
+    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean
+        (S.E.M.)}\label{fig:exectime}%
 \end{figure}
-
-\begin{table}[h]
-  \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{tab:anova:exectime}
-  \setlength{\tabcolsep}{0pt}%
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-    \toprule
-    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
-    \midrule
-    length         &    \(F(2, 78) = 9.59\) &   \(< 0.001\) &           \(0.20\) \\
-    delay          &  \(F(6, 234) = 15.52\) &  \(< 0.0001\) &           \(0.28\) \\
-    length \(\times\) delay &  \(F(12, 468) = 3.84\) &  \(< 0.0001\) &           \(0.09\) \\
-    \bottomrule
-  \end{tabular*}
-\end{table}
-
-Further analysis focused on the progressive effect on per-step execution time as additional steps occurred at a constant delay.
-For this purpose, the steps within a block were aggregated over sequences of 4, constituting a \emph{slice}.
-Note that the first slice within a block is procedurally identical for all block lengths, in the sense that a participant currently performing a step in the first slice of a block has no way of predicting if the block ends after step 4 or not.
-The same logic can be applied to slice 2 for blocks of length 8 and 12.
-Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
-An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{tab:anova:exectime:slice}.
-As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
-% We refer to the measure at slice 3 as a percentage of the measure at slice 1 as the ``change factor''.
-With the longest delays, where the execution time per step was longest, the slow-down persisted; that is, the system response time hindered the participant's execution throughout the course of the block.
-
-\begin{table}[h]
-  \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{tab:anova:exectime:slice}
-  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-    \toprule
-    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
-    \midrule
-    slice         &   \( F(2, 78) = 88.79 \) &  \( < 0.0001 \) &         \( 0.69 \) \\
-    delay         &  \( F(6, 234) = 14.13 \) &  \( < 0.0001 \) &         \( 0.27 \) \\
-    slice \( \times \) delay &  \( F(12, 468) = 2.49 \) &    \( < 0.01 \) &         \( 0.06 \) \\
-    \bottomrule
-  \end{tabular*}%
-\end{table}
 
@@
 
-The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
-Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
-Note that use of the first four steps controls for block length.
-We performed analyses where participants transitioned from either a short- or long-delay block into a
-\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-  \item no delay block (36 subjects)
-  \item \SI{1.65}{\second} delay block (40 subjects)
-  \item \SI{2.7}{\second} delay block (40 subjects)
-  \item \SI{3.0}{\second} delay block (37 subjects).
-\end{enumerate*}
-The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
-We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).
-
 \begin{figure}[h]
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
-  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
+    \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to
+        \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
 \end{figure}
-
-In summary, we observe two direct effects on execution time due to added delays.
-Firstly, we see a clear hampering of the improvement of execution time across steps that is otherwise evident across blocks.
-Secondly, we notice that this effect lingers on even after the delay is removed, affecting subsequent blocks in the task.
-
-\subsection{Acceleration}
-
-Acceleration data from the E4 wristband were taken in 3 axes defined relative to the device.
-As our interest was in the amount of movement rather than the direction in space, we calculated a ``movement score'' for each block which we defined as:
-\begin{equation}
-    M_B = \frac{ \sum_{A_B} |\: \overrightarrow{\alpha_j} \:| }{\Delta T_{B}}
-\end{equation}
-
-where \( B \) is an arbitrary block, and \(A_{B} = \{ \overrightarrow{\alpha_0}, \overrightarrow{\alpha_1}, \ldots, \overrightarrow{\alpha_k} \} \) represents the set of acceleration vector samples collected for said block.
-This score would include the time imposed by the delay and the time to decode the instructions and respond by moving the pieces so that the next state was recognized.
-It is only during the last part of the step that explicit movement is required, so any additional accelerations would derive from arbitrary movements while processing and waiting.
-We normalized the sum of the accelerations by the duration of the step to correct for differences in delay and the execution time, which tends to increase with delay as described above.
-An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{tab:anova:acc} and \cref{fig:acc:delaylength}.
-There was also a significant delay by length interaction, reflecting that delay effects occurred particularly for the longer lengths.
-
-\begin{table}[h]
-  \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and block length.}\label{tab:anova:acc}
-  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-    \toprule
-    \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\(\eta^{2}_{p}\)} \\ 
-    \midrule
-    delay & \(F(6, 234) = 18.56\) & \(<0.001\) & \(0.32\) \\
-    length \( \times \) delay & \(F(12, 468) = 3.04\) & \(<0.001\) & \(0.07\) \\ 
-    \bottomrule
-  \end{tabular*}%
-\end{table}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig9.eps}
-    \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{fig:acc:delaylength}
-\end{figure}
-
-We next conducted the same analysis as for execution time, dividing blocks into three ``slices'' of four steps each, such that slice 1 comprised the first four steps of all block lengths, slice 2 the second four steps of blocks of length 8 and 12, and slice 3 the last four steps of blocks 12 steps long.
-As shown in \cref{fig:acc:delayslice} the effect of delay was attributable only to the later slices, yielding main effect of slice number and delay and an interaction (\cref{tab:anova:acc:slice}).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
-  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
-\end{figure}
-
-\begin{table}[h]
-  \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and slice number.}\label{tab:anova:acc:slice}
-  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-    \toprule
-    \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\(\eta^{2}_{p}\)} \\
-    \midrule 
-    slice & \(F(2, 78) = 8.39\) & \(<0.01\) & \(0.18\) \\
-    delay & \(F(6, 234) = 26.92\) & \(<0.001\) & \(0.41\) \\
-    delay \(\times\) slice & \(F(12, 468) = 4.08\) & \(<0.001\) & \(0.1\) \\
-    \bottomrule
-  \end{tabular*}%
-\end{table}
-
-These results present further evidence of the aforementioned pacing effect.
-As a sequence of steps with a long delay unfolds, more of the step duration is spent without moving.
-This can be interpreted in the context of the increased execution time at long delays demonstrated in the previous analysis.
-Assuming that adding or deleting a LEGO block takes essentially the same amount of time and accelerates the wrist similarly at any one step, it appears that the participant simply remains stationary during the extra time that is induced by a series of long delays.
-Accordingly, the acceleration per unit time, our movement score, is reduced.
-
-\subsection{EEG}
-Analyses were conducted on the log EEG power in the alpha band, beta band, and total of all bands measured.
-Readings from the two frontally placed poles were highly correlated and were pooled into an average.
-Logs were taken for analysis because the EEG distribution tended to have a rightward tail.
-Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 s.d.s.\ from the mean of all participants).
-
-The analyses then comprised 28 participants.\@
-Omnibus ANOVAs were conducted with delay and block length (number of steps) as factors, on the EEG data from the alpha and beta band.
-The analysis of alpha EEG found no significant effects.
-Beta EEG showed only an effect of block length, \( F(2,54) = 3.56 \), \( p = .035 \), \( \eta^{2}_{p} = .12 \), reflecting a tendency for the 4-step length to produce lower log power (mean 3.9 vs.
-4.1 for lengths of 8 and 12).
-However, this effect was small and not consistent across delays.
-
-Again the analysis dividing blocks into 4-step slices was conducted, with delay and slice number as factors.  
-For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{tab:anova:egg:alphabeta}.
-Both bands showed the same tendency:  EEG declined as a sequence of steps with the same delay progressed.
-These effects are shown in \cref{fig:logeegpow:slices}.
-Thus, EEG taken from frontal locations mimics the execution time data in showing a decline over the course of a block, but unlike the execution time, there was no tendency for the decline in EEG to be reduced at longer delays.
-
-\begin{table}[h]
-  \centering
-  \caption{Significant effects on log EEG power from ANOVA on factors delay and block slice.}\label{tab:anova:egg:alphabeta}
-  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}llrrr}
-    \toprule
-    \textbf{Band} & \textbf{Factor} & \textbf{F-test} & \textbf{\(p\)} & \textbf{\( \eta^{2}_{p} \)} \\ \midrule
-    alpha   & slice & \( F(2, 78) = 28.26 \) & \( <0.001 \) & \( 0.51 \) \\ %\midrule
-    beta    & slice & \( F(2, 78) = 29.18 \) & \( <0.001 \) & \( 0.52 \) \\ 
-    \bottomrule
-  \end{tabular*}
-\end{table}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig11.eps}
-    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
-\end{figure}
-
-\subsection{Galvanic Skin Response (GSR) and Heart Rate}
-The measure of GSR was the log of total amplitude in the signal.
-To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, GSR data were normalized by the temporal duration of the block.
-Due to sensor failure and the elimination of one subject with extreme values of GSR amplitude using the same rule as for EEG, the analysis comprised 34 participants.
-No systematic trend due to delay or block length was observed.
-In addition, the same analysis of slice position within block length as was performed for the EEG revealed no effects, indicating that GSR was stable across step positions within a block.
-
-Heart rate, measured in beats per minute, averaged \num{90.2}~BPM and showed no trend related to the experimental variables of delay and block length.
-The same was true of the variability in beats per minute (average standard deviation across conditions was \num{20.3}).
-
-The absence of systematic trends in both these results is interesting in the context of our initial suggestions of potential mechanisms relating delay in the application to human behavior.
-In section \cref{ssec:potentialmechs} we proposed that emotional arousal during the execution of the task was a potential explanation for the effects observed in the human.
-However, these results seem to refute this hypothesis, and will be further discussed in \cref{sec:discussion}.
-
-\subsection{Individual Difference Analysis}
-
-The Big 5 personality inventory and Immersive Tendencies Questionnaires (ITQ) were combined with outcome variables in an analysis of individual differences.
-Given the previous results, we initially considered the following outcome variables:
-\begin{itemize}
-    \item execution time in the most demanding block (length 12, delay \SI{3.0}{\second}); 
-    \item total of alpha and beta bands EEG in the first four steps at all lengths (``Slice 1'');
-    \item average heart rate;
-    \item and average log GSR.\@
-\end{itemize}
-
-Although the heart rate and GSR data had produced no significant effects in the analysis of experimental variables, they could in principle correlate with experimental outcomes across individuals. 
-A principal-components analysis (\emph{PCA}) on a subset of these variables, shown in \cref{tab:pca}, was ultimately conducted on the 28 participants for which there was EEG data. 
-Three of the personality measures were excluded after initial analyses indicated significant correlations among neuroticism, openness, agreeableness, and extroversion.
-Neuroticism (with poles of sensitivity and security) was selected as most relevant to the issue of response to system delay and was included along with conscientiousness.
-GSR was used in a subsidiary analysis, because only 25 participants had both EEG and GSR measures that were reliable.
 
@@
         \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-    \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
+        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
             \toprule
-      \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} \\
+            \textbf{Factor}                            & \textbf{Comp. 1}                    & \textbf{Comp. 2}
+            & \textbf{Comp. 3}                    \\
             \midrule
-      BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
-      BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
-      ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
-      ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
-      Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
-      Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
+            BFI Conscientiousness                      & \textcolor{lightgray}{\( -0.022 \)} & \( 0.668 \)
+            & \textcolor{lightgray}{\( -0.481 \)} \\
+            BFI Neuroticism                            & \( 0.600 \)                         & \( -0.678 \)
+            & \textcolor{lightgray}{\( -0.118 \)} \\
+            ITQ Focus                                  & \( 0.678 \)                         &
+            \textcolor{lightgray}{\( 0.203 \)}  & \( 0.504 \)                         \\
+            ITQ Involvement                            & \( 0.573 \)                         & \( 0.540 \)
+            & \textcolor{lightgray}{\( 0.417 \)}  \\
+            Exec. Time (delay 3.0 s, length 12)        & \( 0.758 \)                         &
+            \textcolor{lightgray}{\( -0.178 \)}
+            & \textcolor{lightgray}{\( -0.348 \)}
+            \\
+            Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} &
+            \textcolor{lightgray}{\( -0.251 \)}
+            & \( 0.589 \)
+            \\
             \bottomrule
@@
         \caption{Percentages of variance explained by the components.}
-    \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
+        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
             \toprule
-      {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total} \\
+            {}                 & \textbf{Comp. 1}     & \textbf{Comp. 2}     & \textbf{Comp. 3}     & \textbf{Total}
+            \\
             \midrule
-      Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent} & \SI{73.13}{\percent} \\
+            Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent}
+            & \SI{73.13}{\percent}
+            \\
             \bottomrule
@@
 
-The PCA produced three components that accounted for \SI{73.13}{\percent} of the variance in the six factors considered.
-Component 1 included neuroticism, both ITQ scores, and execution time, indicating that more sensitive and immersed individuals tended to slow their responses under extended delay.
-Component 2 included conscientiousness and the absence of neuroticism along with immersive involvement, indicating that efficient and secure participants tended to be more involved.
-The third component had positive loadings on EEG in Slice 1 (first four steps of a block) and the focus component of the ITQ.\@
-An additional analysis including heart rate added a component but improved the PCA fit by only \SI{4.5}{\percent}, indicating that any variability in this measure across individuals is unrelated to personality, immersiveness, or outcome.
-A further analysis including GSR produced a solution in which GSR loaded with the first component, along with execution time.
-% A further analysis including GSR again produced a three-component solution, in which GSR loaded with the first component (execution time, sensitivity, immersion).
+\subsection{Emulating human behavior}
 
-\cref{fig:neuro:exectime:reg} shows the correlation between neuroticism and execution time for the block with extreme values of length (12 steps) and delay (\SI{3.0}{\second}), for the full set of 40 participants.
-It confirms the relationship indicated by the first component in the PCA using the smaller sample of 28 participants; that is, higher neuroticism is associated with more responsiveness to delay.
+We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as
+input past measurements of system responsiveness and produces realistic step execution times.
+We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined
+with the timing model for a full end-to-end emulation of a human.
+We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid
+approach~\cite{olguin2018scaling,olguin2019edgedroid}.
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig12.eps}
-    \caption{Correlation between neuroticism score of participants and their execution time in the longest block at the longest delay.
-    Pearson correlation coefficient \( r = .40 \); 2-tailed \( p = .01 \).}
-    \label{fig:neuro:exectime:reg}
-\end{figure}
+The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic
+human-in-the-loop nature.
+Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
+Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming
+and prohibitively expensive for many research groups.
 
-On the whole, these results suggest that individual differences in widely accepted personality variables and immersive tendencies moderate the response to delay.
-This fact could have practical implications in the future.
-It could, for instance, provide a tunable parameter for eventual models aiming to emulate human interaction with a WCA.\@
-In addition, physiological measures of heart rate and EEG appear not to be direct indicators of  behavioral response to delay, although GSR may be more promising in this regard.
+In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \ac{WCA},
+the first data-driven model for human timings in \ac{WCA} applications.
+This model represents a stochastic approach to execution time modeling which builds upon the data collected for
+\textcite{olguinmunoz:impact2021}.
+Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
 
-\section{Discussion}\label{sec:discussion}
-
-% First summarize the main findings
-We start our discussion with the main results of our experimentation presented in the previous section:
-
-\begin{itemize}
-  \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional} behavioral slow-down.
-  That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their execution of the task.
-  This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
-  Moreover, there is a temporal component to this effect; users become progressively slower the more time passes with reduced system responsiveness.
-
-  \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
-  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
-  The question of whether any lingering effect can be measured after these four steps remains open.
-
-  \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at performing steps as the task progresses.
-  However, the strength of this effect decreases as delay increases.
-  Whereas for blocks without delay users performed the last four steps of a 12-step block on average \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.
-
-  \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay. 
-  The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
-  Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables related to attentiveness, respectively.
-  In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
-  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
-  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
-\end{itemize}
-
-A central question therefore arises: to which physio- and psychological mechanisms can these findings, most importantly the substantial slow-down in task execution, be attributed?
-
-In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional reactions.
-These could in turn elicit generalized arousal.
-We also postulated that adapting to delay might progressively deplete cognitive resources in users.
-However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
-Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-The acceleration data further did not indicate that extended delay significantly increases erratic movement.
-% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-% The acceleration data further do not indicate that extended delay increases erratic movement.
-To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
-That is, users fail to capitalize on the new information as quickly as they could.
-Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.
-We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
-As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
-Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
-This also provides a possible explanation for the lingering effects of delay after an acceptable system responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.
-
-As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
-However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.
-
-The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.
-
-\begin{itemize}
-
-  \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to significantly extended application lifetimes.
-  This has direct implications for resource and power consumption.
-
-  \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a high-responsive state could have unconventional consequences for resource allocation.
-  This is of particular importance, for instance, for cases where the user may be able to finish the task before these effects subside.
-  In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired application.
-
-  \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of system responsiveness across applications may not ultimately be beneficial to the system as a whole.
-  Take for instance two applications on the same system which negatively interfere with each other. 
-  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop. 
-  In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.
-
-  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
-  This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.
-  However, this remains an open challenge.
-
-\end{itemize}
-
-To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of WCA deployments.
-Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable tool to tackle inefficiencies in these systems.
-Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the relationship between application responsiveness and human behavior.
-More research in this area will surely uncover more complex and interesting behaviors.
-Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
-These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
-
-\section{Conclusion}\label{sec:conclusion}
-
-In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
-Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.
-
-We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
-The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
-When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
-This was evidenced by an increase in task execution times, even after accounting for the artificially introduced delays, as well as accelerometer data from participants wrists.
-Additionally, we found that the strength of this effect is modulated by individual differences between subjects.
-
-These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
-In this context, we believe there are two direct and important next steps to be performed.
-The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
-We believe these need to be implemented, tested and validated in real setups.
-In particular, we identify two of these implications to be prime candidates for their own experimental studies.
-One, our postulation that in cases where an impaired application is close to finishing, diverting resources to it might not be the most optimal course of action; and two, the possibility that ``fair'' degradation of system responsiveness across applications may be, in some cases, undesirable.
-Both of these questions could be answered with straightforward setups.
-
-The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
-These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
-However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
-Incorporating the results presented here would allow for much more realistic workloads.
-For instance, our findings relating to the effects of delay on execution times could be directly adapted to modulate timings in the input stream.
-Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
-Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.
-
-\begin{acks}
-  This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
-  Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or funding sources.
-\end{acks}
+\todo[inline]{TODO: Some relevant results. Need to add paper to thesis.}

### 2022-12-19 body/part1/overview/implications-optimization.tex
fix biblatex

\section{Implications for the optimization of \ac{WCA}}

\begin{enumerate}
    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
\end{enumerate}

Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.

Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy.
We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.

### 2022-12-19 body/part1/overview/methodology.tex
fix biblatex

\section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}

\begin{figure}
    \centering
    \includegraphics[width=.8\textwidth]{Figs/methodology.png}
    \caption{%
        Overview of the proposed methodology for the study of feedback-loop systems on edge computing infrastructure.
        The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
        The backend/server side of the system along with the network remain unchanged.
    }\label{fig:methodology}
\end{figure}

As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
In the following, we introduce our methodological framework for tackling this challenge.

A general overview of our methodology is illustrated in \cref{fig:methodology}.
It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \ac{COTS} general-purpose computing devices.
The backend side and the network are not changed.

Two main arguments drive this approach.
One, the difficulty of scaling the number of clients in these systems in a research context.
The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \ac{WCA}) or due its cyber-physical nature (in \acp{NCS}).
Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g. Raspberry Pis, Beaglebones, etc.).

And two, it maintains the realism of the most complex parts of the system.

\todo[inline]{Fix and finish}

In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
We first discuss the methodology in the context of a human-in-the-loop application, \acl{WCA}, in section \cref{ssec:methodology:wca}.
Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \acf{NCS}.

\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}

\begin{figure}
    \centering
    \begin{subfigure}[b]{.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{Figs/trace_edgedroid.png}
        \caption{%
            High level conceptual design of our methodology for \ac{WCA}.
            We replace the human by a trace-driven emulation.
            This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
        }
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{.47\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
        \caption{%
            Architectural overview of the EdgeDroid \num{1.0} tool.
            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \ac{WCA} backend running on a cloudlet.%
        }
    \end{subfigure}\\
    \medskip
    \begin{subfigure}[t]{\textwidth}
        \centering
        \adjustbox{scale=0.7}{
            \begin{tikzpicture}[align=center,
                node distance=.5cm and 1.5cm,
                every initial by arrow/.style={-{Latex[length=2mm]}}]
                % Place nodes
                \node [initial, state, minimum size=6em, initial text=] (play) {Play};
                \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
                \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
                \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};

                % Draw edges
                \path[draw, -{Latex[length=2mm]}]
                (play) edge [bend right=20] node[left] {Step done} (rewind)
                edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
                edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)

                (change) edge [bend left=20] node[right] {Step changed} (play)
                edge [bend left=20] node[right] {All steps done} (shutdown)

                (rewind) edge [bend right=20] node[right] {Rewound} (play)
                edge [bend right=20] node[right] {Too many rewinds} (shutdown);

            \end{tikzpicture}
        }
        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime. This model approximates the behavior of an ``ideal'' human, one that is patient and makes no mistakes.}\label{fig:usermodel}
    \end{subfigure}
    \caption{%
        Overview of our first implementation of the methodology for \ac{WCA}.
    }\label{fig:edgedroid1:trace}
\end{figure}

We first introduce our methodology in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
The former corresponds to an extended abstract paper which discusses a high level overview of our approach; the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.

The overall goal of these papers is to explore this methodological approach as applied to \ac{WCA}.
As discussed in \cref{chap:introduction}, the main challenge to benchmarking \ac{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
Humans are unreliable, and, perhaps more importantly, hard to scale.
\todo[inline]{More reasons}

The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \ac{WCA} backend.
The tool is implemented in Android, allowing for its easy deployment on the same kind of \ac{COTS} mobile devices on which real \ac{WCA} applications are intended to run in the future.
Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \ac{WCA} by a human volunteer.
The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
Once recorded, the trace is manually segmented into the logical component steps of the task, and these are replayed in sequence back to the backend by the EdgeDroid tool.

We opt for a trace-based approach for our first implementation of the methodology for \ac{WCA} for three key reasons.
The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
Using a trace ensures that the same computation is performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.

The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
Compatible traces can be easily obtained simply by instrumenting existing \ac{WCA} client applications to record all captured inputs.

The final advantage corresponds to extensibility.
As long as the tasks belong to the same category of \ac{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.

\todo[]{}

reproducible and comparable workload, we use a trace-based design for the client-side emulation.
.

To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.

However, merely replaying the trace is not enough, as disparities between system responsiveness at recording versus replay time can cause the trace so drift out of synchronization with the current state of the system.
Thus, we introduce here the concept of a human user model which is able to adapt its behavior in a realistic manner based on the current responsiveness of the system by systematically controlling the replay of the trace.
The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
The trace is preprocessed into segments that correspond to individual steps of the \ac{WCA} application; this is performed manually as a preparation step.
The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \ac{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.

\todo[inline]{EdgeDroid paper}

\begin{figure}
    \centering%
    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
    \label{fig:comparison:rtt}%
\end{figure}%
\begin{figure}
    \centering%
    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
    \label{fig:comparison:feedback}%
\end{figure}%
\begin{figure}%
    \centering%
    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
    \label{fig:comparison:tasksteps}
\end{figure}%

The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.

Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
This sensory information is then processed by the original compute process at the backend, generating feedback.
However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications

\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.

In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.

For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.

We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.

We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
One, scaling behavior for the application seems to be linear with respect to the number of clients.
Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.

It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.

EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
Here, the downlink time sometimes almost reaches the ideal bound by itself.
A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.

Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.

\todo[inline]{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}

\subsection{Applying the methodology to \acsp{NCS}}\label{ssec:methodology:ncs}

\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
    \caption{
        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
    }\label{fig:cleave:ncs:struct}
\end{figure*}

We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.

Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.

\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.

In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
\begin{enumerate}
    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
    Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
    In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
    Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
    The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
\end{enumerate}

We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
    \item similar setups exist in real-world industrial use
    \item video analytics has long been proposed as a ``killer app'' for edge computing.
\end{enumerate*}
Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.

\todo[inline]{Some plots and results}

\begin{figure}
    \centering
    \includegraphics[width=.6\textwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
    \caption{
        The 2D inverted pendulum system.
    }\label{fig:invpend}
\end{figure}

\begin{figure}[t]
    \centering
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
        \caption{Toppled plants.}\label{fig:video:toppled}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
    \end{subfigure}\\
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
        \caption{Packet losses.}\label{fig:video:drop}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
        \caption{\acsp{RTT}.}\label{fig:video:rtt}
    \end{subfigure}%
    \caption{
        Multi-loop, resource-constrained setup results.
        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
    }\label{fig:video:results}
\end{figure}

\FloatBarrier%

\subsection{Orchestrating experimentation in edge computing testbeds}

\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}

In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
Furthermore, we make Ainur available to the community as \ac{FOSS}.
It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.

In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
    \item provisioning of and connection to remote cloud instances
    \item initialization of \ac{IP} layer connectivity between hosts
    \item collection of logs and data
    \item deployment, scaling, and lifecycle management of containerized processes
\end{enumerate*}.
We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\

\todo[inline]{Some plots}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
    \caption{Network structure assumed by Ainur.}\label{fig:network}
\end{figure*}

### 2022-12-19 body/part1/overview/extending-wca.tex
extending wca

@@
 
-\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on
-human behavior in step-based \ac{WCA}.
-This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \ac{WCA}, first introduced
-in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human
-behavior.
+\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \ac{WCA}.
+This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \ac{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
 
-We employed a modified version of the step-based LEGO cognitive assistant first introduced by \citeauthor{chen2015early}
+We present in this paper the design and execution of a study in which undergraduate students of diverse fields of
+study\footnote{%
+    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \ac{CMU}.
+}, were asked to interact with and follow the instructions given to them by a \ac{WCA} application.
+Unbeknownst to the participants, the system responsiveness of the \ac{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.
+
+With this study, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
+One, do subjects change the temporal profile of their actions in relation to system latency?
+We expected this to be the case, although the extent or form of these changes were unknown.
+Two, do users show signals of arousal in physiological responses to changes in system latency?
+As latency increased and due to the added annoyance of dealing with a less responsive system, we expected users to begin showing signs of stress and frustration.
+Three, are responses to delay effects in subjects mediated by cognition and/or emotion?
+We expected delay effects in subjects to be mediated primarily by emotion, and in particular expected these effects to be correlated with the level of the reduction in system responsiveness.
+And four, are these effects mediated by personality indicators in any way?
+It was our hypothesis that, among others, the individual trait of \emph{neuroticism}~\cite{john1999big} would play a particularly important role in mediating effects of reduced system responsiveness on subjects.
+
+\begin{figure}[tb]
+    \centering
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a}
+        \caption{}\label{sfig:regularwcaexec}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b}
+        \caption{}\label{sfig:delaywcaexec}
+    \end{subfigure}
+    \caption{%
+        Comparison between normal task execution in the LEGO \ac{WCA}~\cite{chen2015early}~(\labelcref{sfig:regularwcaexec}) and modified task execution with delay in the experimental setup of \cref{paper:olguinmunoz2021impact}~(\labelcref{sfig:delaywcaexec}).
+        In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target interval of time denoted \emph{delay}.
+    }\label{fig:regularwca-vs-delaywca}
+\end{figure}
+
+We employed a version of the step-based LEGO cognitive assistant first introduced by \citeauthor{chen2015early}~\cite{chen2015early}.
+The software was modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
+This is illustrated in \cref{fig:regularwca-vs-delaywca}.
+If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly \ensuremath{\mathbb{D}} seconds after frame capture.
+Furthermore, the assistant was instrumented to capture key application and task performance metrics in real time, in particular step execution time.
+\todo[inline]{Task, latin square?}
+
+In order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
+This array included devices to capture four physiological measures:
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item \ac{GSR} (also known as electrodermal activity)
+    \item accelerometer data from the dominant wrist
+    \item brain activity in the form of \ac{EEG}
+    \item heart rate
+\end{enumerate*}.
+Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \ac{BFI} and the \ac{ITQ}.
 

### 2022-12-19 body/glossary.tex
adding glossary

\newabbreviation{WCA}{WCA}{Wearable Cognitive Assistance}
\newabbreviation{AR}{AR}{Augmented Reality}
\newabbreviation{VR}{VR}{Virtual Reality}
\newabbreviation{XR}{XR}{eXtended Reality}
\newabbreviation{GPU}{GPU}{Graphical Processing Unit}
\newabbreviation{FPS}{FPS}{Frames Per Second}
\newabbreviation{DNN}{DNN}{Deep Neural Network}
\newabbreviation{CPS}{CPS}{Cyber-Physical System}
\newabbreviation{NCS}{NCS}{Networked Control System}
%\acroindefinite{NCS}{an}{a}

\newabbreviation{LAN}{LAN}{Local-Area Network}
\newabbreviation{WLAN}{WLAN}{Wireless Local-Area Network}
\newabbreviation{KPI}{KPI}{Key Performance Indicator}
\newabbreviation{WPAN}{WPAN}{Wireless Personal Area Network}
\newabbreviation{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArkin`g` serVice on the Edge}
\newabbreviation{OS}{OS}{Operating System}
\newabbreviation{UDP}{UDP}{User Datagram Protocol}
\newabbreviation{TCP}{TCP}{Transmission Control Protocol}
\newabbreviation{RMS}{RMS}{Root Mean Square}
\newabbreviation{RTT}{RTT}{Round-Trip Time}
\newabbreviation{CI}{CI}{Confidence Interval}
\newabbreviation{AP}{AP}{Access Point}
\newabbreviation{API}{API}{Application Programming Interface}
\newabbreviation{SSF}{SSF}{Swedish Foundation for Strategic Research}
\newabbreviation{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
\newabbreviation{ABC}{ABC}{Abstract Base Class}
%\acroplural{ABC}[ABCs]{Abstract Base Classes}

\newabbreviation{URL}{URL}{Uniform Record Locator}
\newabbreviation{AWS}{AWS}{Amazon Web Services}
\newabbreviation{EC2}{EC2}{Elastic Compute 2}
\newabbreviation{AMI}{AMI}{Amazon Machine Image}
\newabbreviation{SSH}{SSH}{Secure Shell}
\newabbreviation{IP}{IP}{Internet Protocol}
\newabbreviation{VPN}{VPN}{Virtual Private Network}
\newabbreviation{NAT}{NAT}{Network Address Translation}
\newabbreviation{NTP}{NTP}{Network Time Protocol}
%\acroindefinite{NTP}{an}{a}

\newabbreviation{SDR}{SDR}{Software-Defined Radio}
%\acroindefinite{SDR}{an}{a}

\newabbreviation{VLAN}{VLAN}{Virtual Local-Area Network}
\newabbreviation{APT}{APT}{Advaced Packaging Tool}
\newabbreviation{LTE}{LTE}{Long-Term Evolution}
\newabbreviation{SSID}{SSID}{Service Set Identifier}
\newabbreviation{EPC}{EPC}{Evolved Packet Core}
\newabbreviation{UE}{UE}{User Equipment}
\newabbreviation{eNodeB}{eNodeB}{Evolved Node B}
\newabbreviation{DNS}{DNS}{Domain Name System}
\newabbreviation{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
\newabbreviation{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
%\acroindefinite{HTTP}{an}{a}

\newabbreviation{YAML}{YAML}{YAML Ain't Markup Language}
\newabbreviation{O-RAN}{O-RAN}{Open Radio Access Network}
\newabbreviation{FOSS}{FOSS}{Free and Open Source Software}
\newabbreviation{COSMOS}{COSMOS}{Cloud enhanced Open Software defined MObile wireless testbed for city-Scale deployment}
\newabbreviation{OMF}{OMF}{ORBIT Management Framework}
\newabbreviation{POWDER}{POWDER}{Platform for Open Wireless Data-driven Experimental Research}
\newabbreviation{COTS}{COTS}{Commercial Off-The-Shelves}
\newabbreviation{RF}{RF}{Radio-Frequency}
\newabbreviation{VM}{VM}{Virtual Machine}
\newabbreviation{REST}{REST}{REpresentational Sate Transfer}
\newabbreviation{MaaS}{MaaS}{Metal-as-a-Service}
\newabbreviation{OEDL}{OEDL}{\ac{OMF} Experiment Description Language}
\newabbreviation{RSpec}{RSpec}{Resource Specification}
\newabbreviation{LXC}{LXC}{LinuX Containers}
\newabbreviation{MEC}{MEC}{Mobile Edge Computing}
\newabbreviation{USRP}{USRP}{Universal Software Radio Peripheral}
\newabbreviation{OAI}{OAI}{OpenAirInterface}
\newabbreviation{NR}{NR}{New Radio}
\newabbreviation{FPGA}{FPGA}{Field-Programmable Gate Array}
\newabbreviation{ARA}{ARA}{Agricultural and ruRAl communities}
\newabbreviation{UHF}{UHF}{Ultra-High Frequency}
\newabbreviation{MIMO}{MIMO}{Multi-Input Multi-Output}
\newabbreviation{CHI}{CHI}{CHameleon Infrastructure}
\newabbreviation{CMU}{CMU}{Carnegie Mellon University}
\newabbreviation{GSR}{GSR}{Galvanic Skin Response}
\newabbreviation{EEG}{EEG}{electroencephalography}
\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits~\cite{john1999big}}
\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}

### 2022-12-19 abstract.tex
glossaries

@@
 \begin{abstract}
-    \ac{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
+    \gls{WCA} applications present a challenge to benchmark and characterize due to their human-in-the-loop nature.
     Employing user testing to optimize system parameters is generally not feasible, given the scope of the problem and the number of observations needed to detect small but important effects in controlled experiments.
-    Considering the intended mass-scale deployment of \ac{WCA} applications in the future, there exists a need for tools enabling human-independent benchmarking.
+    Considering the intended mass-scale deployment of \gls{WCA} applications in the future, there exists a need for tools enabling human-independent benchmarking.
 
-    We present in this paper the first model for the complete end-to-end emulation of humans in \ac{WCA}.
+    We present in this paper the first model for the complete end-to-end emulation of humans in \gls{WCA}.
     We build this model through statistical analysis of data collected from previous work in this field, and demonstrate its utility by studying application task durations.
     Compared to first-order approximations, our model shows a \textasciitilde\SI{36}{\percent} larger gap between step execution times at high system impairment versus low.
-    We further introduce a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \ac{WCA}, and show that by combining this framework with our realistic model of human behavior, significant reductions of up to \SI{50}{\percent} in number processed frame samples and \SI{20}{\percent} in energy consumption can be achieved with respect to the state-of-the-art.
+    We further introduce a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \gls{WCA}, and show that by combining this framework with our realistic model of human behavior, significant reductions of up to \SI{50}{\percent} in number processed frame samples and \SI{20}{\percent} in energy consumption can be achieved with respect to the state-of-the-art.
 \end{abstract}

### 2022-12-19 appendices.tex
Revert "fixes for IEEE ToMC"
This reverts commit 18f0732eab5bbfdf6d172046f0725d1b21ce31c6.
glossaries

@@
-\clearpage%
-\pagestyle{empty}%
 \appendix{}\label[appendixwithoutnumber]{appx1}
@@
 % the expected wait 
-$\mathbb{E}[\mathcal{W}]$, we use the conditional \ac{CDF} of the execution time.
+$\mathbb{E}[\mathcal{W}]$, we use the conditional \gls{CDF} of the execution time.
 \begin{multline*}
@@
 Thus, we are only interested in $0\!\leq\!t\!\leq\! (t_n-t_{n-1})$.
-Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ correspond to the \ac{CDF}, \ac{CCDF} and \ac{PDF} of the execution time distribution.
+Let $F_{\mathcal{T}}$, $\bar{F}_{\mathcal{T}}$ and $f_{\mathcal{T}}$ correspond to the \gls{CDF}, \gls{CCDF} and \gls{PDF} of the execution time distribution.
 \begin{alignat}{1}
@@
 Here, \cref{Apx1} is an approximation merely for mathematical maturity due to the slackness of the first inequality in the numerator.
-We expand $F_\mathcal{T}(t_{n}-t)$ and $F_\mathcal{T}(t_{n-1})$ using Taylor series. Thus we can write the \ac{CCDF} as
+We expand $F_\mathcal{T}(t_{n}-t)$ and $F_\mathcal{T}(t_{n-1})$ using Taylor series. Thus we can write the \gls{CCDF} as
 \begin{multline*}
@@
 
-Next, using the above \ac{CCDF}, we find the conditional expectation of $\mathcal{W}$.
+Next, using the above \gls{CCDF}, we find the conditional expectation of $\mathcal{W}$.
 \begin{alignat*}{1}
@@
 Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \acl{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
-This is the reason why the above results are applicable in this work where we have predominantly considered \ac{exGaussian} distribution.
+This is the reason why the above results are applicable in this work where we have predominantly considered \gls{exGaussian} distribution.
 Furthermore, we have also verified the closeness of the results as well as the validity of the approximations made in the proofs using distribution fitting and simulations.

### 2022-12-19 background.tex
glossaries

@@
         \caption{%
-            Overview of a task in a \ac{WCA}, composed of a series of steps.
+            Overview of a task in a \gls{WCA}, composed of a series of steps.
             Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
-            The \ac{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
+            The \gls{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
         }\label{fig:task}
@@
             The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
-            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \ac{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
+            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \gls{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
         }\label{fig:step}
@@
 
-\acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \ac{AR} applications.
-In this work we focus on a particular category --- ``step-based'' \ac{WCA} --- that have as their goal the guiding of a user through a sequential task.
+\acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \gls{AR} applications.
+In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
 Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
 
-Step-based \acp{WCA} operate analogously to how \ac{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
+Step-based \acp{WCA} operate analogously to how \gls{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
 The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
@@
 \acp{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \ac{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \gls{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
 \( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.\ which captures the final state of step \( M \)) is taken.
 \( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
-We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \ac{RTT}}, respectively, of a step \( M \).
-The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\ac{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \ac{WCA}.
+We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \gls{RTT}}, respectively, of a step \( M \).
+The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\gls{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \gls{WCA}.
 
@@
 
-In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \ac{WCA} in a controlled experiment.
-We employed a modified and instrumented version of the above-mentioned, step-based LEGO \ac{WCA}~\cite{Chen2015LEGO}.
+In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \gls{WCA} in a controlled experiment.
+We employed a modified and instrumented version of the above-mentioned, step-based LEGO \gls{WCA}~\cite{Chen2015LEGO}.
 \num{40} subjects interacted with the assistant, performing a \num{169}-step task while we altered the responsiveness in realtime and captured key application and task performance metrics.
@@
 % \end{enumerate*},
-% then it follows that the expected \ac{TTF} for steps subject to a delay \( D \) can be expressed as
+% then it follows that the expected \gls{TTF} for steps subject to a delay \( D \) can be expressed as
 
 % \begin{align}
-%     \mathbb{E}\left[\text{\ac{TTF}}\right] &= \mathbb{E}\left[\mathcal{W}\right] + D
+%     \mathbb{E}\left[\text{\gls{TTF}}\right] &= \mathbb{E}\left[\mathcal{W}\right] + D
 %     = \mathbb{E}\left[U(t_{n - 1}, t_n)\right] + D\nonumber\\
@@
 
-% Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \ac{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5}.
+% Ergo, we can directly re-parameterize the data from~\cite{olguinmunoz:impact2021} to use \gls{TTF} instead of delay by simply multiplying the delay value of each step by \num{1.5}.
 
@@
     \item System slow-down induces \emph{additional} behavioral slow-down which scales with the decrease in system responsiveness.
-    Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \ac{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \ac{TTF} of \SI{4.5}{\second}.
+    Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
 
@@
 We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}~\cite{oliver:bfi1999}, and the \acf{ITQ}~\cite{witmer1998measuring}.
-Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \ac{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
+Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \gls{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
 Delay discounting is the tendency to devalue rewards for which one must wait; high rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
@@
 %     \caption{%
-%         Correlation between neuroticism and the mean execution time of the last four steps in segments of \num{12} steps subject to the same \ac{TTF} in~\textcite{olguinmunoz:impact2021}.
+%         Correlation between neuroticism and the mean execution time of the last four steps in segments of \num{12} steps subject to the same \gls{TTF} in~\textcite{olguinmunoz:impact2021}.
 %         Pearson correlation coefficient \( \rho = 0.418 \), 2-tailed \( p < 0.05 \).
@@
 
-In this work, we will use a normalized scale to describe neuroticism, derived from the minimum and maximum obtainable values for this variable in the \ac{BFI}~\cite{oliver:bfi1999}.
+In this work, we will use a normalized scale to describe neuroticism, derived from the minimum and maximum obtainable values for this variable in the \gls{BFI}~\cite{oliver:bfi1999}.
 \emph{Low} and \emph{high} neuroticism will refer to the \( [0.0, 0.5) \) and \( [0.5, 1.0] \) ranges respectively.
@@
 Linear regression showed a significant correlation between individual neuroticism scores and the execution time late in a series of high-delay steps, \( \rho = 0.418 \), \num{2}-tailed \( p < 0.05 \)
-Neuroticism was further identified as a modulating factor for the pacing effects through a \ac{PCA}.
+Neuroticism was further identified as a modulating factor for the pacing effects through a \gls{PCA}.
 Out of the three identified components, which cumulatively accounted for \SI{73.13}{\percent} of the variance in the results, neuroticism was included in the first two.
@@
     \caption{%
-        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \ac{TTF} in~\cite{olguinmunoz:impact2021}.
-        Distributions were fitted using \ac{MLE}.
+        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \gls{TTF} in~\cite{olguinmunoz:impact2021}.
+        Distributions were fitted using \gls{MLE}.
     }\label{fig:muexgaussian}
@@
     \caption{%
-        Example \ac{exGaussian} fits on execution times from steps \numrange{4}{8} in a segment of steps at the maximum experimental \ac{TTF}.
+        Example \gls{exGaussian} fits on execution times from steps \numrange{4}{8} in a segment of steps at the maximum experimental \gls{TTF}.
         The effects of neuroticism are clearly visible in the tail and the mean of the distributions.
@@
 
-Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \ac{TTF}, and continuous segments of steps subject to the same \ac{TTF}, were well-fit by an \ac{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
-When grouping by level of neuroticism, \ac{TTF}, and \emph{slice}\footnote{%
-In~\cite{olguinmunoz:impact2021} the \emph{slice} to which a step belongs to refers to whether the step occurred in the first, second, or third four-step segment of a sequence of steps subject to the same \ac{TTF}.
+Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \gls{TTF}, and continuous segments of steps subject to the same \gls{TTF}, were well-fit by an \gls{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
+When grouping by level of neuroticism, \gls{TTF}, and \emph{slice}\footnote{%
+In~\cite{olguinmunoz:impact2021} the \emph{slice} to which a step belongs to refers to whether the step occurred in the first, second, or third four-step segment of a sequence of steps subject to the same \gls{TTF}.
 }, the best fit statistic was \ensuremath{0.028} (\ensuremath{p = 0.999}).
@@
 We found that the effects of neuroticism on execution times were clearly identifiable in the fitted distributions, in particular in their means and tails.
-\cref{fig:muexgaussian} shows an example of this modulating effect, illustrating the behavior of the mean (\( \mu \) parameter) of \ac{exGaussian} distributions fitted to the execution times of the first four steps of segments of steps subject to the same \ac{TTF}.
-Finally, \cref{fig:fitsneuro} shows an example of the effects of neuroticism on the fitted \ac{exGaussian} distributions for a specific group of execution times.
+\cref{fig:muexgaussian} shows an example of this modulating effect, illustrating the behavior of the mean (\( \mu \) parameter) of \gls{exGaussian} distributions fitted to the execution times of the first four steps of segments of steps subject to the same \gls{TTF}.
+Finally, \cref{fig:fitsneuro} shows an example of the effects of neuroticism on the fitted \gls{exGaussian} distributions for a specific group of execution times.
 Higher neuroticism directly translates into a higher mean and longer tail.

### 2022-12-19 conclusion.tex
glossaries

@@
 
-This paper addresses the difficulty of benchmarking \ac{WCA} by offering a model-based alternative to the extensive human-user studies that would otherwise be required.
-It first introduces the \edgedroid{} model of human timing behavior for \ac{WCA}.
+This paper addresses the difficulty of benchmarking \gls{WCA} by offering a model-based alternative to the extensive human-user studies that would otherwise be required.
+It first introduces the \edgedroid{} model of human timing behavior for \gls{WCA}.
 This model represents a stochastic approach to execution time modeling which builds upon prior data~\cite{olguinmunoz:impact2021}.
-It further introduces a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+It further introduces a novel procedure for the generation of synthetic traces of frames in step-based \gls{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
 
-The paper then explores the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
+The paper then explores the impact of such a realistic model on the application lifetime footprint of \gls{WCA} applications.
 It shows that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions, can potentially lead to substantial mis-estimations of the application footprint.
 
-Finally, the paper delves into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
-It proposes a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA}, which results in an adaptive sampling strategy. 
+Finally, the paper delves into the potential for optimization in \gls{WCA} systems using the previously discussed timing models.
+It proposes a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \gls{WCA}, which results in an adaptive sampling strategy.
 We have shown that this framework is applicable to both the minimization of number of samples and total energy consumption per step by showcasing experimental results.
-Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, thus proving the value of such frameworks for the design of \ac{WCA} applications.
+Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, thus proving the value of such frameworks for the design of \gls{WCA} applications.
 
-This work serves as an important, yet initial step towards realistic modeling of human behavior in \ac{WCA}, and more generally \ac{AR}.
+This work serves as an important, yet initial step towards realistic modeling of human behavior in \gls{WCA}, and more generally \gls{AR}.
 Many directions remain open to additional exploration in this space.
-For instance, our current model only targets a particular class of step-based \ac{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \ac{AR} and \ac{XR}, is on our roadmap.
+For instance, our current model only targets a particular class of step-based \gls{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \gls{AR} and \gls{XR}, is on our roadmap.
 Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the United States.
 An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.
-Corresponding research is needed to determine individual difference factors that would substantially impact response to \ac{WCA} in novel populations.
+Corresponding research is needed to determine individual difference factors that would substantially impact response to \gls{WCA} in novel populations.
 
@@
 Our future efforts consider developing this framework towards these more accurate distributions.
-Finally, our current approach only considers a single client, which will be far from the case in a real-world \ac{WCA} deployment.
+Finally, our current approach only considers a single client, which will be far from the case in a real-world \gls{WCA} deployment.
 As such, another milestone in this context could be the exploration for a potential extension towards a collaborative and/or distributed solution.

### 2022-12-19 implications_footprint.tex
glossaries

@@
-\section{Implications for the study of \ac{WCA} footprints}\label{sec:implications:footprint}
+\section{Implications for the study of \gls{WCA} footprints}\label{sec:implications:footprint}
 
@@
 We begin by studying the implications of such a model on the estimation of application footprints as described by their lifetimes.
-In the context of \ac{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
-This is an important metric for \ac{WCA} optimization, as it directly relates to system resource utilization and contention, and to energy consumption.
+In the context of \gls{WCA}, we will understand \emph{application lifetime} as the time it takes a user to complete a specified task.
+This is an important metric for \gls{WCA} optimization, as it directly relates to system resource utilization and contention, and to energy consumption.
 
 In order to illustrate the consequences of using a less realistic model that does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \ac{exGaussian} distribution fitted to all execution time samples collected for~\cite{olguinmunoz:impact2021}.
+This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \gls{exGaussian} distribution fitted to all execution time samples collected for~\cite{olguinmunoz:impact2021}.
 This distribution is then randomly sampled at runtime to obtain execution times for each step, without any adjustment to the current state of the system.
@@
 We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
-These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \ac{TTF} and record the generated execution time.
+These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \gls{TTF} and record the generated execution time.
 We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
-This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \ac{TTF}.
+This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \gls{TTF}.
 
@@
     \caption{%
-        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
+        Experimental setup used to study the implications of the realistic models of human behavior for \gls{WCA}.
         We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \acs*{COTS} \acs{IEEE} \num{802.11}b/g access point.
@@
 The models, reference and realistic, are deployed on the Raspberry Pi clients of the testbed depicted in \cref{fig:expsetup}.
-For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \ac{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
+For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \gls{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
 Clients and cloudlet communicate over an \acs{IEEE} \num{802.11}b/g wireless network.
@@
             \# clients & 1 & 4 & 7 & 10 \\
-            \ac{RTT} & \SI{0.42}{\second} & \SI{1.12}{\second} & \SI{1.92}{\second} & \SI{2.68}{\second} \\
+            \gls{RTT} & \SI{0.42}{\second} & \SI{1.12}{\second} & \SI{1.92}{\second} & \SI{2.68}{\second} \\
             \bottomrule
@@
 The results are nonetheless clear, and follow the same pattern as the previously discussed results under ideal, controlled circumstances.
-With just a single client and mean \ac{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
+With just a single client and mean \gls{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
 At \num{10} clients, the results mimic those at higher \acp{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
 
-These results highlight the importance of accurate execution time modeling when studying \ac{WCA} application lifetimes.
+These results highlight the importance of accurate execution time modeling when studying \gls{WCA} application lifetimes.
 Not only do we see considerable differences in lifetimes at relatively moderate levels of system congestion, but the sign of these differences depends on the load placed on the system.
-Imagine thus a system designer studying resource consumption optimization in a \ac{WCA}.
+Imagine thus a system designer studying resource consumption optimization in a \gls{WCA}.
 If they were to employ the reference model for their study on an unimpaired system, it could lead them to significantly underestimate the potential for optimization of resource consumption, leaving performance (and, potentially, monetary) gains on the table.
@@
 % .as well as an additional setup with \num{10} clients during which an instance of the \emph{iperf3}\footnote{\href{https://iperf.fr/}{https://iperf.fr/}} network load generator is also running on each client.
-% The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \ac{TCP} traffic, further congesting the wireless network.
+% The \emph{iperf3} instances each generate \SI{1.5}{Mbps} of upstream \gls{TCP} traffic, further congesting the wireless network.

### 2022-12-19 implications_optimization.tex
glossaries

@@
-\section{Implications for the optimization of resource consumption and responsiveness trade-offs in \ac{WCA}}\label{sec:implications:optimization}
+\section{Implications for the optimization of resource consumption and responsiveness trade-offs in \gls{WCA}}\label{sec:implications:optimization}
 
-As with any application intended for mass deployment on multi-tenant systems such as the edge, \ac{WCA} applications will have to be configured and designed to minimize resource consumption at runtime.
+As with any application intended for mass deployment on multi-tenant systems such as the edge, \gls{WCA} applications will have to be configured and designed to minimize resource consumption at runtime.
 At the same time, imposing too stringent requirements on resource consumption can adversely affect application responsiveness, which in turn has the potential to drastically reduce quality of experience for users.
@@
 
-A question then arises as to what extent accurate modeling of human behavior plays a role in the optimization of \ac{WCA} applications and systems.
+A question then arises as to what extent accurate modeling of human behavior plays a role in the optimization of \gls{WCA} applications and systems.
 We have already seen from the results in \cref{sec:implications:footprint} that changes in responsiveness can have drastic effects on human execution times and subsequently on overall application lifetimes.
-This points towards parameters with dependencies on step timings as prime candidates for evaluating the effects of human behavior on their potential for optimization of different dimensions of \ac{WCA}.
+This points towards parameters with dependencies on step timings as prime candidates for evaluating the effects of human behavior on their potential for optimization of different dimensions of \gls{WCA}.
 One such parameter of particular interest to us is sampling strategy, which intrinsically hinges on a stochastic understanding of human timings in a step.
 In the following section we will study the potential advantages of using the realistic model presented in \cref{sec:model} to estimate the execution time distributions of steps to adapt the sampling strategy at runtime.
-We will first introduce a generic optimization framework which produces an aperiodic sampling strategy that optimizes for arbitrary responsiveness-resource consumption trade-offs in \ac{WCA}.
+We will first introduce a generic optimization framework which produces an aperiodic sampling strategy that optimizes for arbitrary responsiveness-resource consumption trade-offs in \gls{WCA}.
 We will then employ this framework for the minimization of two different metrics: number of samples and total energy consumption per step.
@@
 
-Our optimization approach focuses on the individual steps which comprise a complete \ac{WCA} task.
+Our optimization approach focuses on the individual steps which comprise a complete \gls{WCA} task.
 Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
@@
 % In this optimization, the constant parts are excluded as it does not affect the process.
-With a value much smaller than the execution time of the event, the \ac{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
+With a value much smaller than the execution time of the event, the \gls{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
 Next, in~\cite{Moothedath2022Aperiodic}, the author retains the model and removes the constraint of periodicity to find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
@@
 On the other, the relationship between number of samples captured and sent and network congestion is exponential, and too high sampling rates quickly lead to bottlenecks on the network, particularly in multi-tenant environments.
-Additionally, the energy cost of capturing a sample on a \ac{WCA} client device is often much higher than remaining in an idle state, and thus excessive sampling leads to drastically increased energy consumption.
-Optimizing the number of samples captured per step can thus be a straightforward way of reducing resource consumption and contention in \ac{WCA} applications.
+Additionally, the energy cost of capturing a sample on a \gls{WCA} client device is often much higher than remaining in an idle state, and thus excessive sampling leads to drastically increased energy consumption.
+Optimizing the number of samples captured per step can thus be a straightforward way of reducing resource consumption and contention in \gls{WCA} applications.
 
@@
 We introduce here a reference scheme to which we will compare our approach.
-In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \ac{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
+In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \gls{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
 At every sampling instant \( t \), the scheme adapts the sampling rate \( R(t) \) of the system according to the estimated likelihood of the user having finished the step,
@@
 \( R_\text{max} \) and \( R_\text{min} \) correspond to the maximum and minimum sampling rates of the system, respectively.
-\( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\ac{RTT}}_\mu \), where \( \text{\ac{RTT}}_\mu \) corresponds to the mean \ac{RTT} of the system.
+\( R_\text{max} \) can directly be assumed to correspond to \( 1 / \text{\gls{RTT}}_\mu \), where \( \text{\gls{RTT}}_\mu \) corresponds to the mean \gls{RTT} of the system.
 \( R_\text{min} \) needs to either be calculated according to the latency bounds of the system or specified manually.
 \( \varphi \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
-Finally, \( CDF \) corresponds to the \ac{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
+Finally, \( CDF \) corresponds to the \gls{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
 
@@
 
-We implement \citeauthor{Wang2019Towards}'s original design using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \ac{CDF} calculation.
-This scheme does not include an embedded timing model, and uses the same \ac{CDF} for every step.
+We implement \citeauthor{Wang2019Towards}'s original design using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \gls{CDF} calculation.
+This scheme does not include an embedded timing model, and uses the same \gls{CDF} for every step.
 
@@
 This represents a completely unoptimized design with no considerations for resource-consumption trade-offs; it simply attempts to maximize the number of captured samples per step.
-This is an interesting approach to include as it corresponds to the sampling strategy used in most existing \ac{WCA} prototypes.
+This is an interesting approach to include as it corresponds to the sampling strategy used in most existing \gls{WCA} prototypes.
 
@@
         \( \tau_\text{p} \) & \SI{250}{\milli\second} & Processing delay \\
-        \( \tau_\text{c} \) & \( \text{\ac{RTT}} - \tau_\text{p} \) & Communication delay \\
+        \( \tau_\text{c} \) & \( \text{\gls{RTT}} - \tau_\text{p} \) & Communication delay \\
         \( w_0 \) & \SI{1.0}{\second} & \\
@@
 A maximum expected wait time \( w_0 = 1.0\,\si{\second} \) thus translates into a maximum expected sampling interval of \SI{2.0}{\second}, yielding a minimum sampling rate \( R_\text{min} = {2.0\,\si{\second}}^{-1} = 0.5\,\si{\hertz} \).
-It should also be noted that in both aperiodic schemes, our sampling-count-optimized approach and the \ac{CDF}-based approach, there exists the possibility for sampling instants to be \emph{missed} due to the actual \ac{RTT} of the system being higher than the parameterization of the schemes.
+It should also be noted that in both aperiodic schemes, our sampling-count-optimized approach and the \gls{CDF}-based approach, there exists the possibility for sampling instants to be \emph{missed} due to the actual \gls{RTT} of the system being higher than the parameterization of the schemes.
 In these cases, both schemes will degrade into greedy sampling.
@@
 The execution times for each step are generated by a timing model without any distribution fitting on the data.
-For each combination of sampling scheme configuration, \ac{RTT}, and execution time model neuroticism (low or high), we run \num{100} repetitions of the task for good statistical significance.
+For each combination of sampling scheme configuration, \gls{RTT}, and execution time model neuroticism (low or high), we run \num{100} repetitions of the task for good statistical significance.
 Note that the embedded timing model in our sampling-count-optimized aperiodic sampling scheme is always parameterized with a neuroticism matching the neuroticism of the external execution time model.
@@
     \caption{%
-        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \ac{CDF}-based approach.
+        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \gls{CDF}-based approach.
     }\label{fig:optimization:samples}
@@
 On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
-As mentioned above, the future feasibility and mass adoption of \ac{WCA} depends on these applications not hogging the available resources.
+As mentioned above, the future feasibility and mass adoption of \gls{WCA} depends on these applications not hogging the available resources.
 Our work advances this goal by being consistently more efficient than existing alternatives at minimizing the number of samples per step, and thus reducing network and processing load.
@@
 It should be noted that although both schemes seem to tend towards two samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
-As \acp{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \ac{RTT} being larger than the execution time of the step tends towards \num{1.0}.
-This leads to these sampling schemes consistently sampling only twice each step: a first sample which is taken before the execution time of the step, and a second one \ac{RTT} seconds later, after the execution time has been reached.
+As \acp{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \gls{RTT} being larger than the execution time of the step tends towards \num{1.0}.
+This leads to these sampling schemes consistently sampling only twice each step: a first sample which is taken before the execution time of the step, and a second one \gls{RTT} seconds later, after the execution time has been reached.
 
@@
 
-Next we will explore the implications of combining our timing models with \cref{eq:epsilon_terminal} when directly optimizing for energy consumption in \ac{WCA}.
+Next we will explore the implications of combining our timing models with \cref{eq:epsilon_terminal} when directly optimizing for energy consumption in \gls{WCA}.
 Although, as mentioned above, minimizing the number of samples captured during a step can \emph{potentially} translate into a reduction in the energy consumption, this is not a given.
@@
 \( \tau_\text{p} \) and \( \tau_\text{c} \) correspond to the processing and two-way communication delay for each sample.
-\( P_\text{c} \) and \( P_0 \) correspond to the communication and idle power, respectively, of the \ac{WCA} client device.
+\( P_\text{c} \) and \( P_0 \) correspond to the communication and idle power, respectively, of the \gls{WCA} client device.
 
@@
 For the power constants, we reuse the values estimated by the authors in~\cite{Moothedath2022EnergyEfficient}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
-On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \ac{RTT} for the task, we set \( \tau_\text{c} = \text{\ac{RTT}} - \tau_\text{p} \).
+On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \gls{RTT} for the task, we set \( \tau_\text{c} = \text{\gls{RTT}} - \tau_\text{p} \).
 
@@
     \caption{%
-        Summary of results for experiment comparing the energy-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \ac{CDF}-based approach.
+        Summary of results for experiment comparing the energy-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \gls{CDF}-based approach.
     }\label{fig:optimization:energy}
@@
 
-Again, we include the reference greedy and offline optimum schemes, as well as \textcite{Wang2019Towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \ac{RTT}.
+Again, we include the reference greedy and offline optimum schemes, as well as \textcite{Wang2019Towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \gls{RTT}.
 

### 2022-12-19 intro.tex
glossaries

@@
 These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
-\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
-In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
+\gls{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
+In particular, following the success of non-wearable \gls{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \gls{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
 
 A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses.
-They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
+They are intended to operate as autonomous guides, much akin to how \gls{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
 This context-sensitivity and proactivity in providing user feedback translate into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video, which require intensive compute capabilities to process.
-This also translates into latency-sensitivity, as with any \ac{AR} application.
+This also translates into latency-sensitivity, as with any \gls{AR} application.
 Delays and jitter can be jarring to the user, causing discomfort and leading them to make mistakes and potentially even abandoning the application altogether.
@@
 
-The combination of these opposing characteristics has led to \ac{WCA} applications being identified in the literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+The combination of these opposing characteristics has led to \gls{WCA} applications being identified in the literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
 However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
-One key gap in knowledge pertains to the current lack of tools and methodologies for scalable and repeatable study of \ac{WCA} application performance and resource utilization in realistic deployments.
+One key gap in knowledge pertains to the current lack of tools and methodologies for scalable and repeatable study of \gls{WCA} application performance and resource utilization in realistic deployments.
 Due to their human-in-the-loop nature, these applications present a challenge to benchmark and characterize, in particular in real-scale deployments where dozens or even hundreds of users might concurrently use the system.
-Accordingly, recruiting a  cohort of subjects for realistic benchmarking and study of \ac{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers
-There exists therefore a real need for scalable tools for \ac{WCA} benchmarking which do not rely on direct testing of the human-in-the-loop.
+Accordingly, recruiting a  cohort of subjects for realistic benchmarking and study of \gls{WCA} systems can be prohibitively cumbersome and expensive for many research groups and system designers
+There exists therefore a real need for scalable tools for \gls{WCA} benchmarking which do not rely on direct testing of the human-in-the-loop.
 
@@
 \begin{enumerate}
-    \item\label{item:contrib:model} We introduce the first, to our knowledge, stochastic model for human timings in \ac{WCA} applications.
+    \item\label{item:contrib:model} We introduce the first, to our knowledge, stochastic model for human timings in \gls{WCA} applications.
     Using the data collected for~\cite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
-    We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
+    We also introduce a novel way to generate dynamic traces of frames for \gls{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
     We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-    \item\label{item:contrib:footprint} Using this model, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \ac{WCA}, understanding this term as the duration of a specific complete execution of the application task.
+    \item\label{item:contrib:footprint} Using this model, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \gls{WCA}, understanding this term as the duration of a specific complete execution of the application task.
     In accordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
-    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \gls{WCA} when considering human behavior using our model.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \gls{WCA} optimization; number of processed samples and energy consumption per step.
     First, we study the potential for reducing the number of samples captured per step.
@@
 \cref{sec:relwork} discusses related work in the field of \acl{WCA} modeling.
-In \cref{sec:background} we define and discuss key concepts in \ac{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
-In \cref{sec:model} we detail our model for the generation of realistic human timing in \ac{WCA}.
+In \cref{sec:background} we define and discuss key concepts in \gls{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
+In \cref{sec:model} we detail our model for the generation of realistic human timing in \gls{WCA}.
 We present its design, verify its expected behavior with respect to our previous results, and introduce the dynamic trace generation for full end-to-end emulation of human behavior.
 Next, in \cref{sec:implications:footprint}, we discuss the potential implications of such a model on application lifetimes by studying a small series of representative scenarios.
-This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \ac{WCA} systems.
-In the same section, we introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \ac{WCA}.
+This is followed by a more in-depth investigation in \cref{sec:implications:optimization} on the potential consequences of a human timing model on the general optimization of \gls{WCA} systems.
+In the same section, we introduce our generic optimization framework for resource consumption versus responsiveness trade-offs in \gls{WCA}.
 Finally, in \cref{sec:conclusion} we summarize and conclude this paper, as well as briefly discuss potential avenues for future work.

### 2022-12-19 model.tex
glossaries

@@
-\section{A model of human behavior for \ac{WCA}}\label{sec:model}
+\section{A model of human behavior for \gls{WCA}}\label{sec:model}
 
-In the following, we employ the above insights together with the data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \ac{WCA}.
+In the following, we employ the above insights together with the data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \gls{WCA}.
 We detail the construction of a model which uses the data collected to generate, at runtime, realistic execution times. 
 In order to accurately emulate the behavior of a human, such a model needs to implement two main behaviors.
-First, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \ac{WCA} system, as well as salient individual-difference measures.
+First, it needs to generate realistic execution times for each step in the task, considering the current and historical impairment of the \gls{WCA} system, as well as salient individual-difference measures.
 We detail this in \cref{ssec:model:exectimes}.
@@
 
-% As expressed in \cref{ssec:plos}, for \textcite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
+% As expressed in \cref{ssec:plos}, for \textcite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \gls{WCA}, as well as personality trait scores for each participant.
 % For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-The processing of the data from~\cite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \ac{TTF}.
+The processing of the data from~\cite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \gls{TTF}.
 The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
@@
 We begin by calculating rolling weighted averages of the \acp{TTF} of each of the \num{40} individual repetitions of the data.
-We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \ac{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \textasciitilde\SI{25}{\percent}, the third-most \textasciitilde\SI{12}{\percent}, and so on.
-We additionally pad the data for each run with \num{12} copies of the first \ac{TTF} in order to ensure sensible values for the first twelve steps; this padding is removed after the weighted values have been calculated.
-This weighing ensures that drastic changes in \ac{TTF} are significantly remembered by the model for at least \num{4} steps --- in line with our previous findings on human behavior --- and are subsequently quickly forgotten.
+We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \gls{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \textasciitilde\SI{25}{\percent}, the third-most \textasciitilde\SI{12}{\percent}, and so on.
+We additionally pad the data for each run with \num{12} copies of the first \gls{TTF} in order to ensure sensible values for the first twelve steps; this padding is removed after the weighted values have been calculated.
+This weighing ensures that drastic changes in \gls{TTF} are significantly remembered by the model for at least \num{4} steps --- in line with our previous findings on human behavior --- and are subsequently quickly forgotten.
 \begin{equation}\label{eq:weights}
@@
 \begin{enumerate}
-    \item At the beginning of each step, the model is fed the measured \ac{TTF} for the previous step.
+    \item At the beginning of each step, the model is fed the measured \gls{TTF} for the previous step.
     \item The model calculates a weighted average over the latest \num{12} steps using the weights defined in \cref{eq:weights}.
-    \item The resulting weighted \ac{TTF} is binned into the corresponding range (\cref{tab:defaultbins}).
-    \item The model then filters the pre-processed data to find execution time samples associated with this same discretized weighted \ac{TTF}.
+    \item The resulting weighted \gls{TTF} is binned into the corresponding range (\cref{tab:defaultbins}).
+    \item The model then filters the pre-processed data to find execution time samples associated with this same discretized weighted \gls{TTF}.
     \item The selected samples are further filtered to match the desired level of neuroticism.
@@
         \item directly sampling the execution time values;
-        \item or, using \ac{MLE} to fit a distribution to the execution time samples and then sampling the distribution instead.
+        \item or, using \gls{MLE} to fit a distribution to the execution time samples and then sampling the distribution instead.
     \end{itemize}
@@
 
-The distribution chosen for the second variant described above corresponds to the \ac{exGaussian} distribution, previously discussed in \cref{ssec:moderationeffects}.
+The distribution chosen for the second variant described above corresponds to the \gls{exGaussian} distribution, previously discussed in \cref{ssec:moderationeffects}.
 
@@
         Higher \acp{TTF} directly lead to higher execution times.
-        Error bars indicate the \SI{95}{\percent} \ac{CI}.
+        Error bars indicate the \SI{95}{\percent} \gls{CI}.
     }\label{fig:ttf_to_exectime}
@@
 
-\cref{fig:ttf_to_exectime} shows the mean execution time outputted by the model when fed three different levels of \ac{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}).
-These results were generated by first warming up the model by feeding it \num{25} \acp{TTF} selected at random from the data before feeding it the desired input \ac{TTF}, and recording the generated execution time.
-This procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
+\cref{fig:ttf_to_exectime} shows the mean execution time outputted by the model when fed three different levels of \gls{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}).
+These results were generated by first warming up the model by feeding it \num{25} \acp{TTF} selected at random from the data before feeding it the desired input \gls{TTF}, and recording the generated execution time.
+This procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
 The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \acp{TTF} reaching \SI{14}{\percent} (\textasciitilde\SI{5.6}{\second} to \textasciitilde\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
@@
     Effects of prolonged exposure to constant levels of system impairment on the model.
-    At low (\SI{0}{\second}) \ac{TTF}, the models speed-up over time; conversely at high (\SI{5.0}{\second}) \ac{TTF}, the models either present no change or drastically increase their generated execution times, depending on the level of neuroticism.
-    Error bands indicate \SI{95}{\percent} \ac{CI}.
+    At low (\SI{0}{\second}) \gls{TTF}, the models speed-up over time; conversely at high (\SI{5.0}{\second}) \gls{TTF}, the models either present no change or drastically increase their generated execution times, depending on the level of neuroticism.
+    Error bands indicate \SI{95}{\percent} \gls{CI}.
     }\label{fig:exectimeduration}
@@
 
-Next, \cref{fig:exectimeduration} shows the evolution of generated execution times while the model is subject to a fixed \ac{TTF}, either \emph{low} (\SI{0}{\second}) or \emph{high} (\SI{5}{\second}).
-These results were generated by first warming up the model with \num{25} random \acp{TTF}, and then recording the generated execution times over a sequence of \num{12} steps at a fixed \ac{TTF}; this procedure is repeated \num{600} times for each configuration and target \ac{TTF}.
+Next, \cref{fig:exectimeduration} shows the evolution of generated execution times while the model is subject to a fixed \gls{TTF}, either \emph{low} (\SI{0}{\second}) or \emph{high} (\SI{5}{\second}).
+These results were generated by first warming up the model with \num{25} random \acp{TTF}, and then recording the generated execution times over a sequence of \num{12} steps at a fixed \gls{TTF}; this procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
 Once again, we see here behavior matching what is expected of the model, in particular with respect to \cref{it:duration}.
@@
         Effects of changes in system impairment on subsequently generated execution times.
-        These effects linger on after the change, and thus execution times immediately after a transition are either consistently lower or higher than otherwise at the new \ac{TTF}, depending on the old \ac{TTF}.
-        Error bars indicate \SI{95}{\percent} \ac{CI}.
+        These effects linger on after the change, and thus execution times immediately after a transition are either consistently lower or higher than otherwise at the new \gls{TTF}, depending on the old \gls{TTF}.
+        Error bars indicate \SI{95}{\percent} \gls{CI}.
     }\label{fig:transitions}
@@
 Finally, in \cref{fig:transitions} we showcase the behavior of the model when comparing execution times generated immediately after a change in system responsiveness.
-We generate these results by first warming up the model by feeding it a fixed \ac{TTF} (which we will refer to as the \emph{origin} \ac{TTF}) \num{25} times.
-Next, another \ac{TTF} value (the \emph{destination} \ac{TTF}) is fed to the model, and we record the output execution time.
-Each sample is tagged according to the relation between origin and destination \ac{TTF}, either lower to higher, higher to lower, or equal.
-As before, we run \num{600} repetitions of this procedure for each combination of model configuration, origin \ac{TTF}, and destination \ac{TTF}.
-Execution times generated immediately after a transition from a higher \ac{TTF} into a lower one are consistently higher than execution times generated without a preceding change in \acp{TTF}.
-Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \ac{TTF} into a higher one.
+We generate these results by first warming up the model by feeding it a fixed \gls{TTF} (which we will refer to as the \emph{origin} \gls{TTF}) \num{25} times.
+Next, another \gls{TTF} value (the \emph{destination} \gls{TTF}) is fed to the model, and we record the output execution time.
+Each sample is tagged according to the relation between origin and destination \gls{TTF}, either lower to higher, higher to lower, or equal.
+As before, we run \num{600} repetitions of this procedure for each combination of model configuration, origin \gls{TTF}, and destination \gls{TTF}.
+Execution times generated immediately after a transition from a higher \gls{TTF} into a lower one are consistently higher than execution times generated without a preceding change in \acp{TTF}.
+Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \gls{TTF} into a higher one.
 These results are once again in line with our findings in~\cite{olguinmunoz:impact2021}, in which we found lingering effects of transitions between levels of system impairment on human execution times.
@@
 Apart from the aforementioned timing and performance data, for~\cite{olguinmunoz:impact2021} we also recorded all collected video frame samples together with matching metadata.
-For each video frame submitted to the \ac{WCA} during the tasks, we recorded
+For each video frame submitted to the \gls{WCA} during the tasks, we recorded
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
@@
     \item sample submission timestamp
-    \item \ac{WCA} processing completed timestamp
+    \item \gls{WCA} processing completed timestamp
     \item result or acknowledgement returned timestamp
-    \item a tag representing the result of the \ac{WCA} processing
+    \item a tag representing the result of the \gls{WCA} processing
 \end{enumerate*}.
@@
 \begin{description}[font={\bfseries\ttfamily}, wide]
-    \item[SUCCESS:] frames which triggered a transition to a new step (or the correction of a previous mistake) in the logical task model of the \ac{WCA}, and thus cause the generation of feedback to the user.
+    \item[SUCCESS:] frames which triggered a transition to a new step (or the correction of a previous mistake) in the logical task model of the \gls{WCA}, and thus cause the generation of feedback to the user.
     \item[REPEAT:] frames which captured the same board state as the previous successful frame, and thus produced no feedback.
-    \item[LOW\_CONFIDENCE] frames for which the image recognition algorithm in the \ac{WCA} did not reach the necessary confidence threshold to interpret it as a valid board state.
+    \item[LOW\_CONFIDENCE] frames for which the image recognition algorithm in the \gls{WCA} did not reach the necessary confidence threshold to interpret it as a valid board state.
         These frames also produce no feedback.
@@
 This is illustrated in \cref{fig:frameprobs}.
-Intuitively, \texttt{REPEAT} frames dominate the early instants after a step transition, as the user has not had time to start performing the new instruction and thus the \ac{WCA} keeps capturing frames representing the previous state of the board.
-As the user starts moving and performing actions, \texttt{BLANK} frames start to dominate, as this activity prevents the \ac{WCA} from capturing ``clean'' frames.
+Intuitively, \texttt{REPEAT} frames dominate the early instants after a step transition, as the user has not had time to start performing the new instruction and thus the \gls{WCA} keeps capturing frames representing the previous state of the board.
+As the user starts moving and performing actions, \texttt{BLANK} frames start to dominate, as this activity prevents the \gls{WCA} from capturing ``clean'' frames.
 Finally, it must be noted that \texttt{SUCCESS} frames are not included in this probability density plot, as, by definition:
@@
 
-Using the above insights, together with the corresponding recorded video frames, we devise a scheme for the procedural generation of a synthetic trace for any step in \ac{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
+Using the above insights, together with the corresponding recorded video frames, we devise a scheme for the procedural generation of a synthetic trace for any step in \gls{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
 We first prepare a discretized representation of the probability density map in \cref{fig:frameprobs}.
@@
 
-We provide the model implementations in Python~\num{3.10} as well as the base data to the community as \ac{FOSS}.
+We provide the model implementations in Python~\num{3.10} as well as the base data to the community as \gls{FOSS}.
 All of these are published on the \href{https://github.com/KTH-EXPECA/EdgeDroid2}{\texttt{KTH-EXPECA/EdgeDroid2}} repository on GitHub under a permissive Apache version 2 license.

### 2022-12-19 old_model.tex
glossaries

@@
 
-As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \ac{WCA}, as well as personality trait scores for each participant.
+As expressed in \cref{ssec:plos}, for ~\cite{olguinmunoz:impact2021} we collected timing and task performance data for \num{40} participants performing a \num{169}-step task, guided by a \gls{WCA}, as well as personality trait scores for each participant.
 For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
@@
 
-We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \ac{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \ac{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
+We begin by discretizing normalized neuroticism into \emph{low} and \emph{high} ranges, and \gls{TTF} into three ranges by splitting the data on the 3-quantiles; we will refer to these ranges of \gls{TTF} as \emph{low}, \emph{medium}, and \emph{high impairment}.
 \cref{tab:defaultbins} shows the exact values used for these ranges.
@@
 The next steps of processing are slightly more complex.
-For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \ac{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
+For the data of each subject (which we will refer to as a \emph{run}), we shift the execution times by one sample such that each \gls{TTF} measurement (and therefore, impairment level) is associated with the execution time for the next step.
 This comes directly from our discussions in \cref{sec:background} and~\cite{olguinmunoz:impact2021}.
@@
 The first of these samples execution times directly from the empirical data from~\cite{olguinmunoz:impact2021}; we will refer to it hereafter as the \emph{empirical model}.
-The second corresponds to one which first fits a theoretical \ac{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
-The \ac{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+The second corresponds to one which first fits a theoretical \gls{exGaussian} distribution to the data before sampling; we will refer to this one as the \emph{theoretical model}.
+The \gls{exGaussian} distribution was chosen due to the vast body of research supporting its suitability for the modeling of the timing of human actions~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
 Collectively, we will refer to both these models interchangeably as the \emph{realistic models} or \emph{second-order models}.
@@
     \item a model which randomly samples from the raw data without any grouping
-    \item a model which fits an \ac{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it  
+    \item a model which fits an \gls{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it
 \end{enumerate*}.
@@
     \toprule
-    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\ac{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\ac{TTF} \( = \) \SI{5}{\second}} \\
+    {} & {} & \multicolumn{2}{>{\columncolor[HTML]{EAEAEA}}c}{\gls{TTF} \( = \) \SI{0}{\second}} & \multicolumn{2}{c}{\gls{TTF} \( = \) \SI{5}{\second}} \\
     {Model} & {Neuroticism} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} & {Total Duration} & {Diff.\ w.r.t.\ fitted-naive} \\
@@
     \caption{
-        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \ac{TTF} setup.
+        Comparison of the total duration of \num{100}-step tasks for different models in the controlled \gls{TTF} setup.
         The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
@@
 We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
-For each \ac{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.
+For each \gls{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.
 
@@
 Two interesting behaviors of the realistic models can be noticed in this plot.
-One, all realistic models are faster than the reference models at the lowest \ac{TTF}.
+One, all realistic models are faster than the reference models at the lowest \gls{TTF}.
 When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.
 
-Two, at the highest \ac{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
-For \ac{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
+Two, at the highest \gls{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
+For \gls{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
 
@@
     \caption{%
-        Experimental setup used to study the implications of the realistic models of human behavior for \ac{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \ac{COTS} \acs{IEEE} \num{802.11}n access point.
+        Experimental setup used to study the implications of the realistic models of human behavior for \gls{WCA}.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \gls{COTS} \acs{IEEE} \num{802.11}n access point.
     }\label{fig:expsetup}
@@
 
-These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \ac{WCA} backend running variants of the LEGO task.
+These models are then integrated into client software for communicating with an adapted and instrumented version of the Gabriel~\cite{Chen2015LEGO,Chen2018application} \gls{WCA} backend running variants of the LEGO task.
 In the following we use these client-server loops together with the testbed show in \cref{fig:expsetup} to illustrate the implications of the realistic timing models.
@@
     \end{subfigure}%
-    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \ac{CI}.}
+    \caption{Results for both scenarios described in \cref{ssec:model:verification}. Error bars indicate the \SI{95}{\percent} \gls{CI}.}
 \end{figure*}
@@
 
-The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \ac{WCA} applications.
+The results for scenarios \labelcref{verification:scaling,verification:neuro} showcase the importance and potential impact of realistic models for the study of \gls{WCA} applications.
 

### 2022-12-19 related_work.tex
vishnu
glossaries

@@
 
-There exist a number of approaches to the characterization and benchmarking of mobile \ac{XR}, and by extension \ac{WCA}, deployments.
+There exist a number of approaches to the characterization and benchmarking of mobile \gls{XR}, and by extension \gls{WCA}, deployments.
 \emph{OpenRTiST}~\cite{george2020openrtist} is a tool which uses a compute-heavy, yet latency-sensitive workload --- image style transfer~\cite{jing2019neural} --- to load and benchmark edge computing deployments.
-It acts as an end-to-end realistic \ac{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
-\textcite{Chen2017Empirical} use a number of prototype \ac{WCA} implementation to study and characterize real-world latency bounds in these systems.
-In~\cite{lecci2021open}, the authors collect and analyze large amounts of \ac{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
-\textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for for the benchmarking of mobile hardware in the context of \ac{AR}.
+It acts as an end-to-end realistic \gls{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
+\textcite{Chen2017Empirical} use a number of prototype \gls{WCA} implementation to study and characterize real-world latency bounds in these systems.
+In~\cite{lecci2021open}, the authors collect and analyze large amounts of \gls{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
+\textcite{chetoui2022arbench} propose \emph{ARBench}, a toolkit for the benchmarking of mobile hardware in the context of \gls{AR}.
 The toolkit incorporates a series of workloads which stress different components of the mobile device and calculates a score for each workload.
-\emph{Yahoo Cloud Serving Benchmark}~\cite{cooper2010benchmarking} and \emph{DCBench}~\cite{jia2013characterizing} are workload sets intended for benchmarking cloud services which have further been used in the context of benchmarking edge computing infrastructure for \ac{AR} suitability.
+\emph{Yahoo Cloud Serving Benchmark}~\cite{cooper2010benchmarking} and \emph{DCBench}~\cite{jia2013characterizing} are workload sets intended for benchmarking cloud services which have further been used in the context of benchmarking edge computing infrastructure for \gls{AR} suitability.
 \emph{Edgebench}~\cite{das2018edgebench} and \emph{Defog}~\cite{mcchesney2019defog} benchmark workload performance on the edge versus on the cloud.
 
-We contribute to this body of work by providing the first tool for the benchmarking of \ac{WCA} that dynamically considers the effects of human behavior.
+We contribute to this body of work by providing the first tool for the benchmarking of \gls{WCA} that dynamically considers the effects of human behavior.
 In contrast to the above described works, which generate loads with relatively static profiles, our model is able to react to changes in system responsiveness like a human would.
 We achieve this by building upon our own previous contributions to this field.
-In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \ac{WCA} we called \emph{EdgeDroid}.
-We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \ac{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
-In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \ac{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
-In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \ac{WCA} applications through human-subject studies.
+In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \gls{WCA} we called \emph{EdgeDroid}.
+We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \gls{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
+In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \gls{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
+In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \gls{WCA} applications through human-subject studies.
 We found that humans generally pace themselves according to the perceived system responsiveness.
@@
 
-The characterization and optimization of mobile \ac{XR} has also long been a topic of research.
-\textcite{srinivasan2009performance} analyze a \ac{MAR} workload and identify and characterize the computational bottlenecks in the application.
+The characterization and optimization of mobile \gls{XR} has also long been a topic of research.
+\textcite{srinivasan2009performance} analyze a \gls{MAR} workload and identify and characterize the computational bottlenecks in the application.
 They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
-\textcite{Wang2019Towards} develop a taxonomy for \ac{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
-\textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \ac{AR}.
-\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \ac{AR} applications to minimize energy consumption.
+\textcite{Wang2019Towards} develop a taxonomy for \gls{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
+\textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \gls{AR}.
+\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \gls{AR} applications to minimize energy consumption.
 As above, we have also previously directly contributed to this field.
-In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \ac{WCA}.
+In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \gls{WCA}.
 This is extended into an aperiodic strategy in~\cite{Moothedath2022Aperiodic}.

### 2022-12-19 body/glossary.tex
glossaries

@@
 \newabbreviation{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
-\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArkin`g` serVice on the Edge}
 \newabbreviation{OS}{OS}{Operating System}
@@
 %\acroplural{ABC}[ABCs]{Abstract Base Classes}
-
 \newabbreviation{URL}{URL}{Uniform Record Locator}
@@
 %\acroindefinite{NTP}{an}{a}
-
 \newabbreviation{SDR}{SDR}{Software-Defined Radio}
 %\acroindefinite{SDR}{an}{a}
-
 \newabbreviation{VLAN}{VLAN}{Virtual Local-Area Network}
@@
 %\acroindefinite{HTTP}{an}{a}
-
 \newabbreviation{YAML}{YAML}{YAML Ain't Markup Language}
@@
 \newabbreviation{MaaS}{MaaS}{Metal-as-a-Service}
-\newabbreviation{OEDL}{OEDL}{\ac{OMF} Experiment Description Language}
+\newabbreviation{OEDL}{OEDL}{\gls{OMF} Experiment Description Language}
 \newabbreviation{RSpec}{RSpec}{Resource Specification}
@@
 \newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}
+\newabbreviation{GPS}{GPS}{Global Positioning System}
+\newabbreviation{FSM}{FSM}{Finite State Machine}
+%\acroindefinite{NCS}{an}{a}
+\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}
+\newabbreviation{TTF}{TTF}{Time-to-Feedback}
+\newabbreviation{MAR}{MAR}{Mobile \gls{AR}}
+\newabbreviation{QoE}{QoE}{Quality of Experience}
+\newabbreviation{QoS}{QoS}{Quality of Service}
+\newabbreviation{PCA}{PCA}{Principal Component Analysis}
+\newabbreviation{exGaussian}{Ex-Gaussian}{Exponentially Modified Gaussian}
+\newabbreviation{IEEE}{IEEE}{Institute of Electrical and Electronics Engineers}
+\newabbreviation{CPU}{CPU}{Central Processing Unit}
+\newabbreviation{N/A}{N/A}{Not Applicable}
+\newabbreviation{CIO}{CIO}{Confidence Interval}
+\newabbreviation{MLE}{MLE}{Maximum Likelihood Estimation}
+\newabbreviation{CDF}{CDF}{Cumulative Distribution Function}
+\newabbreviation{CCDF}{CCDF}{Complementary \gls{CDF}}
+\newabbreviation{PDF}{PDF}{Probability Density Function}
+\newabbreviation{ECDF}{ECDF}{Empirical \gls{CDF}}
+\newabbreviation{NSF}{NSF}{United States National Science Foundation}

### 2022-12-19 body/part1/introduction.tex
glossaries

@@
 This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
-Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \ac{WCA} and \ac{NCS}.
+Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \gls{WCA} and \gls{NCS}.
 This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting implementations of its application in the context of \ac{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
+We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
 
-Secondly, we present a deeper exploration of this methodology for \ac{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \ac{WCA}.
+Secondly, we present a deeper exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
 This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
 
-Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \ac{WCA} deployments.
+Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
 In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
@@
 \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \ac{WCA}, making very straightforward assumptions about human behavior in \ac{WCA} applications.
+\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
 \cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
@@
 \cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \acp{NCS}.
-As discussed above, these applications are similarly difficult as \ac{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
+As discussed above, these applications are similarly difficult as \gls{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
 \todo[inline]{More argumentation or refer to previous section.}
-We apply thus in this work our methodology to \ac{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+We apply thus in this work our methodology to \gls{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
 This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
@@
 
-Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \ac{WCA}, as well as study its implications for optimization of these systems.
-Although the above approach to human behavior in \ac{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
+Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
+Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
 In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
+We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
 System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
@@
 
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \ac{WCA}.
+\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
 The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
 

### 2022-12-19 body/part1/overview/extending-wca.tex
glossaries

@@
 
-\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \ac{WCA}.
-This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \ac{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
+\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \gls{WCA}.
+This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
 
@@
 study\footnote{%
-    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \ac{CMU}.
-}, were asked to interact with and follow the instructions given to them by a \ac{WCA} application.
-Unbeknownst to the participants, the system responsiveness of the \ac{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.
+    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \gls{CMU}.
+}, were asked to interact with and follow the instructions given to them by a \gls{WCA} application.
+Unbeknownst to the participants, the system responsiveness of the \gls{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.
 
@@
     \caption{%
-        Comparison between normal task execution in the LEGO \ac{WCA}~\cite{chen2015early}~(\labelcref{sfig:regularwcaexec}) and modified task execution with delay in the experimental setup of \cref{paper:olguinmunoz2021impact}~(\labelcref{sfig:delaywcaexec}).
+        Comparison between normal task execution in the LEGO \gls{WCA}~\cite{chen2015early}~(\labelcref{sfig:regularwcaexec}) and modified task execution with delay in the experimental setup of \cref{paper:olguinmunoz2021impact}~(\labelcref{sfig:delaywcaexec}).
         In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target interval of time denoted \emph{delay}.
@@
 \begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
-    \item \ac{GSR} (also known as electrodermal activity)
+    \item \gls{GSR} (also known as electrodermal activity)
     \item accelerometer data from the dominant wrist
-    \item brain activity in the form of \ac{EEG}
+    \item brain activity in the form of \gls{EEG}
     \item heart rate
 \end{enumerate*}.
-Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \ac{BFI} and the \ac{ITQ}.
+Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI} and the \gls{ITQ}.
 
@@
 
-We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
+We introduce the first, to our knowledge, data-driven model for human timings in \gls{WCA} applications.
 Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as
 input past measurements of system responsiveness and produces realistic step execution times.
-We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined
+We also introduce a novel way to generate dynamic traces of frames for \gls{WCA} applications which can be combined
 with the timing model for a full end-to-end emulation of a human.
@@
 
-The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic
+The study and benchmarking of \gls{WCA} applications is a challenging discipline due to these application's intrinsic
 human-in-the-loop nature.
@@
 
-In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \ac{WCA},
-the first data-driven model for human timings in \ac{WCA} applications.
+In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \gls{WCA},
+the first data-driven model for human timings in \gls{WCA} applications.
 This model represents a stochastic approach to execution time modeling which builds upon the data collected for
 \textcite{olguinmunoz:impact2021}.
-Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
+Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \gls{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
 

### 2022-12-19 body/part1/overview/implications-optimization.tex
glossaries

@@
-\section{Implications for the optimization of \ac{WCA}}
+\section{Implications for the optimization of \gls{WCA}}
 
 \begin{enumerate}
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
+    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \gls{WCA}.
     In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
-    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
+    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \gls{WCA} when considering human behavior using our model.
+    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \gls{WCA} optimization; number of processed samples and energy consumption per step.
     Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
@@
 
-Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
+Next, we have explored the impact of such a realistic model on the application lifetime footprint of \gls{WCA} applications.
 We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
 
-Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
-We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy.
+Finally, we have delved into the potential for optimization in \gls{WCA} systems using the previously discussed timing models.
+We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \gls{WCA} which results in an adaptive sampling strategy.
 We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
-Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.
+Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \gls{WCA} applications.

### 2022-12-19 body/part1/overview/methodology.tex
glossaries

@@
 
-As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \ac{NCS}) or with human involvement (such as \ac{WCA}).
+As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \gls{NCS}) or with human involvement (such as \gls{WCA}).
 In the following, we introduce our methodological framework for tackling this challenge.
@@
 It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
-We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \ac{COTS} general-purpose computing devices.
+We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \gls{COTS} general-purpose computing devices.
 The backend side and the network are not changed.
@@
 The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \ac{WCA}) or due its cyber-physical nature (in \acp{NCS}).
+The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \gls{WCA}) or due its cyber-physical nature (in \acp{NCS}).
 Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
@@
         \caption{%
-            High level conceptual design of our methodology for \ac{WCA}.
+            High level conceptual design of our methodology for \gls{WCA}.
             We replace the human by a trace-driven emulation.
@@
             Architectural overview of the EdgeDroid \num{1.0} tool.
-            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \ac{WCA} backend running on a cloudlet.%
+            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \gls{WCA} backend running on a cloudlet.%
         }
@@
     \caption{%
-        Overview of our first implementation of the methodology for \ac{WCA}.
+        Overview of our first implementation of the methodology for \gls{WCA}.
     }\label{fig:edgedroid1:trace}
@@
 
-The overall goal of these papers is to explore this methodological approach as applied to \ac{WCA}.
-As discussed in \cref{chap:introduction}, the main challenge to benchmarking \ac{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
+The overall goal of these papers is to explore this methodological approach as applied to \gls{WCA}.
+As discussed in \cref{chap:introduction}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
 Humans are unreliable, and, perhaps more importantly, hard to scale.
@@
 The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
-We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \ac{WCA} backend.
-The tool is implemented in Android, allowing for its easy deployment on the same kind of \ac{COTS} mobile devices on which real \ac{WCA} applications are intended to run in the future.
+We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \gls{WCA} backend.
+The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile devices on which real \gls{WCA} applications are intended to run in the future.
 Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
-On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \ac{WCA} by a human volunteer.
+On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer.
 The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
@@
 
-We opt for a trace-based approach for our first implementation of the methodology for \ac{WCA} for three key reasons.
+We opt for a trace-based approach for our first implementation of the methodology for \gls{WCA} for three key reasons.
 The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
@@
 The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
-Compatible traces can be easily obtained simply by instrumenting existing \ac{WCA} client applications to record all captured inputs.
+Compatible traces can be easily obtained simply by instrumenting existing \gls{WCA} client applications to record all captured inputs.
 
 The final advantage corresponds to extensibility.
-As long as the tasks belong to the same category of \ac{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
+As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
 
@@
 The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
-The trace is preprocessed into segments that correspond to individual steps of the \ac{WCA} application; this is performed manually as a preparation step.
-The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \ac{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
+The trace is preprocessed into segments that correspond to individual steps of the \gls{WCA} application; this is performed manually as a preparation step.
+The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \gls{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
 On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
@@
 We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \ac{NCS}.
+In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \gls{NCS}.
 As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
@@
 
-Our framework, \ac{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
+Our framework, \gls{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
 It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
@@
 
-\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
+\gls{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
 It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
 
-In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \ac{NCS} benchmarking with a particular focus on edge deployment.
-We argue our approach, \ac{CLEAVE}, embodies a better solution than previous work for a number of reasons:
+In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \gls{NCS} benchmarking with a particular focus on edge deployment.
+We argue our approach, \gls{CLEAVE}, embodies a better solution than previous work for a number of reasons:
 \begin{enumerate}
@@
     Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \ac{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \ac{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
+    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \gls{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
+    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \gls{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
     The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
@@
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the inverted pendulum plant is ubiquitous in \ac{NCS} research
+    \item the inverted pendulum plant is ubiquitous in \gls{NCS} research
     \item similar setups exist in real-world industrial use
@@
 Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \ac{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
 
@@
         \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
+        \caption{Angle \gls{RMS}.}\label{fig:video:rms}
     \end{subfigure}\\
@@
 It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \ac{FOSS}.
+Furthermore, we make Ainur available to the community as \gls{FOSS}.
 It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
@@
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \ac{LTE} and 5G
+    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \gls{LTE} and 5G
     \item provisioning of and connection to remote cloud instances
-    \item initialization of \ac{IP} layer connectivity between hosts
+    \item initialization of \gls{IP} layer connectivity between hosts
     \item collection of logs and data
@@
 We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
+It has been released as \gls{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
 

### 2022-12-19 publications/2022Ainur/ainur.tex
glossaries

@@
 The core goal of the framework is to facilitate the creation of the desired communication and computation elements to discover their implications to the performance of end-to-end distributed applications.
-Ainur achieves this by offering a Python \ac{API} which is used to describe an end-to-end experiment in a procedural manner, and we plan to eventually provide toolkits for declarative configuration of experiments.
+Ainur achieves this by offering a Python \gls{API} which is used to describe an end-to-end experiment in a procedural manner, and we plan to eventually provide toolkits for declarative configuration of experiments.
 
@@
 
-Hosts correspond to either bare-metal machines or cloud instances on \ac{AWS} \ac{EC2}.
+Hosts correspond to either bare-metal machines or cloud instances on \gls{AWS} \gls{EC2}.
 Users are free to combine and interconnect these in any configuration.
-Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including WiFi, 4G \ac{LTE}, and 5G.
-To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \ac{USRP} \ac{SDR}s and some computation hosts dedicated to signal processing and/or radio management.
+Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including WiFi, 4G \gls{LTE}, and 5G.
+To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \gls{USRP} \gls{SDR}s and some computation hosts dedicated to signal processing and/or radio management.
 
@@
 
-    \item[\ac{AWS} and \texttt{boto3}:]
-    Ainur employs the Amazon \texttt{boto3} library for Python to directly interface with \ac{AWS} \ac{EC2} and deploy, configure, and manage remote cloud instances.
+    \item[\gls{AWS} and \texttt{boto3}:]
+    Ainur employs the Amazon \texttt{boto3} library for Python to directly interface with \gls{AWS} \gls{EC2} and deploy, configure, and manage remote cloud instances.
 
-    \item[\ac{OAI}:]
-    \ac{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \acp{SDR} like the \ac{USRP}~\cite{KALTENBERGER2020107284}. 
-    Ainur can deploy, configure, and manage \ac{OAI} software components that implement 4G \ac{LTE} and 5G \ac{NR}.
+    \item[\gls{OAI}:]
+    \gls{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \acp{SDR} like the \gls{USRP}~\cite{KALTENBERGER2020107284}.
+    Ainur can deploy, configure, and manage \gls{OAI} software components that implement 4G \gls{LTE} and 5G \gls{NR}.
 
@@
     Finally, the Mango Communications project implements real-time 802.11 (WiFi) MAC and PHY in Xilinx \acp{FPGA}.
-    It can be used on a variety of hardware platforms including \ac{USRP} \acp{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
+    It can be used on a variety of hardware platforms including \gls{USRP} \acp{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
 \end{description}
@@
 Ainur follows a layered architecture which closely mimics the conceptual layers of the \acs{TCP}/\acs{IP} stack.
-Components are deployed in bottom-up order, starting with the establishment of physical links, through the establishment of \ac{IP} connectivity and deployment of links to the cloud, and ending with the distribution and initialization of workloads on top of a container orchestration layer.
+Components are deployed in bottom-up order, starting with the establishment of physical links, through the establishment of \gls{IP} connectivity and deployment of links to the cloud, and ending with the distribution and initialization of workloads on top of a container orchestration layer.
 The architectural modules of the framework can broadly be classified according to the below categories:
@@
     This layer interacts with hardware such as managed switches and \acp{SDR} (and their associated computation hosts) to create links between the desired workload hosts.
-    These links correspond to ethernet, WiFi, and even an entire software-defined 4G \ac{LTE} and 5G networks.
+    These links correspond to ethernet, WiFi, and even an entire software-defined 4G \gls{LTE} and 5G networks.
 
     \item[Cloud Layer:]
-    Handles integration with cloud services (currently, \ac{AWS} \ac{EC2}). 
+    Handles integration with cloud services (currently, \gls{AWS} \gls{EC2}).
     Only deployed in the case of an experiment requiring cloud instances, it manages the instantiation and configuration of remote cloud resources.
 
-    \item[\ac{IP} Layer:]
-    Configures and establishes \ac{IP} layer connectivity of packets between hosts in the experimental setup, both local and cloud.
-    This includes assigning valid \ac{IP} addresses to hosts, establishing \ac{VPN} routes to cloud instances through a pair of pre-configured \ac{VPN} gateways, and configuring routing tables to ensure any two hosts in the workload network can communicate with each other.
+    \item[\gls{IP} Layer:]
+    Configures and establishes \gls{IP} layer connectivity of packets between hosts in the experimental setup, both local and cloud.
+    This includes assigning valid \gls{IP} addresses to hosts, establishing \gls{VPN} routes to cloud instances through a pair of pre-configured \gls{VPN} gateways, and configuring routing tables to ensure any two hosts in the workload network can communicate with each other.
 
@@
 They allow for easy packaging of software in predefined, consistent, and conflict-free execution environments, including all necessary dependencies.
-Containers are lightweight and portable compared to \ac{VM}-based virtualization, making them an ideal solution for the distribution, deployment, and orchestration of software in distributed computing environments.
-They deploy quickly and are very configurable, while abstracting away the complexities and delays that come with in creating, managing, and moving (potentially huge) \ac{VM} disk images.
+Containers are lightweight and portable compared to \gls{VM}-based virtualization, making them an ideal solution for the distribution, deployment, and orchestration of software in distributed computing environments.
+They deploy quickly and are very configurable, while abstracting away the complexities and delays that come with in creating, managing, and moving (potentially huge) \gls{VM} disk images.
 
@@
 As expressed above, Ainur also leverages containers for the communication stack.
-With the advent of \ac{O-RAN} and \acp{SDR}, all components of the wireless network stack can run on general-purpose processors.
+With the advent of \gls{O-RAN} and \acp{SDR}, all components of the wireless network stack can run on general-purpose processors.
 They can thus be containerized and distributed across multiple hosts.
@@
 
-A full-featured, and rather complex deployment is depicted in \cref{fig:network}, including a number of optional testbed features such as offload to the cloud through \ac{VPN} gateways.
+A full-featured, and rather complex deployment is depicted in \cref{fig:network}, including a number of optional testbed features such as offload to the cloud through \gls{VPN} gateways.
 However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{fig:network:minimal}.
@@
 
-% Also assumed is the existence of \ac{VPN} gateways to the cloud, time synchronization, and configured hostname lookups (possibly using a \ac{DNS} service).
+% Also assumed is the existence of \gls{VPN} gateways to the cloud, time synchronization, and configured hostname lookups (possibly using a \gls{DNS} service).

### 2022-12-19 publications/2022Ainur/assumptions.tex
glossaries

@@
 \subsubsection{Time synchronization}
-Hosts on the testbed synchronize their clocks to a central \ac{NTP} server.
-This is however only required for consistent time measurements \emph{across} devices; measurements with respect to the clock of a single device (e.g.\ \ac{RTT} measurements) do not require synchronization.
+Hosts on the testbed synchronize their clocks to a central \gls{NTP} server.
+This is however only required for consistent time measurements \emph{across} devices; measurements with respect to the clock of a single device (e.g.\ \gls{RTT} measurements) do not require synchronization.
 
@@
     The control network, whose links are represented in orange in \cref{fig:network}, carries control plane data between the control host and the rest of the hosts on the testbed.
-    It also provides internet access through a gateway router, and control plane access to the cloud instances through the control \ac{VPN} gateway (more details on this below).
+    It also provides internet access through a gateway router, and control plane access to the cloud instances through the control \gls{VPN} gateway (more details on this below).
 
@@
     This network carries data generated by the workloads deployed on the testbed, and is represented by blue links in \cref{fig:network}
-    It also provides access to the workload data \ac{VPN} link to cloud instances through the workload data gateway.
-    Although the physical links between devices are always present, interfaces on this network by default have no \ac{IP} addresses --- these are managed dynamically by Ainur.
+    It also provides access to the workload data \gls{VPN} link to cloud instances through the workload data gateway.
+    Although the physical links between devices are always present, interfaces on this network by default have no \gls{IP} addresses --- these are managed dynamically by Ainur.
 
@@
     In order to manage it through software, the managed switch is also connected to the control network through a single link.
-    The associated interface is on a separate, exclusive \ac{VLAN}, and no data from the workload data network is allowed to enter the control network.
+    The associated interface is on a separate, exclusive \gls{VLAN}, and no data from the workload data network is allowed to enter the control network.
 
-    \item[\ac{VPN} gateways]
+    \item[\gls{VPN} gateways]
 
-    Two \ac{VPN} gateways are present on the testbed; one for control traffic, and one for workload data traffic to \ac{AWS} cloud instances.
+    Two \gls{VPN} gateways are present on the testbed; one for control traffic, and one for workload data traffic to \gls{AWS} cloud instances.
     More details on these gateways are given in \cref{sec:layer3}.
@@
 \begin{description}[style=unboxed]
-    % \item[\ac{APT}]
+    % \item[\gls{APT}]
 
-    % Hosts use \ac{APT} to manage installed packages.
+    % Hosts use \gls{APT} to manage installed packages.
 
@@
     Docker\footnote{\url{https://docs.docker.com/}} is installed on the hosts.
-    It is configured to allow regular users to run the Docker client and the server is set up to listen on unencrypted \ac{TCP} port \num{2375} on all interfaces, in addition to the default local UNIX socket.
+    It is configured to allow regular users to run the Docker client and the server is set up to listen on unencrypted \gls{TCP} port \num{2375} on all interfaces, in addition to the default local UNIX socket.
 \end{description}
@@
 
-    The \ac{SSH} client on the control host and the \ac{SSH} servers on the controlled hosts (i.e.\ workload hosts, \ac{VPN} gateways) have been configured to allow password-less, public-key based authentication.
+    The \gls{SSH} client on the control host and the \gls{SSH} servers on the controlled hosts (i.e.\ workload hosts, \gls{VPN} gateways) have been configured to allow password-less, public-key based authentication.
 
@@
 
-Cloud integration in Ainur is done in \ac{AWS}\footnote{\url{https://aws.amazon.com/}} and the associated \texttt{boto3}\footnote{\url{https://aws.amazon.com/sdk-for-python/}} Python library.
+Cloud integration in Ainur is done in \gls{AWS}\footnote{\url{https://aws.amazon.com/}} and the associated \texttt{boto3}\footnote{\url{https://aws.amazon.com/sdk-for-python/}} Python library.
 
 \begin{description}[style=unboxed]
-    \item[Authentication to \ac{AWS}]
+    \item[Authentication to \gls{AWS}]
 
-    Credentials to \ac{AWS} exist in the execution environment of Ainur on the control hosts.
+    Credentials to \gls{AWS} exist in the execution environment of Ainur on the control hosts.
     Specifically, these are credentials that allow Ainur to fully manage:
     \begin{itemize}
-        \item \ac{EC2} instances, to create and terminate required compute nodes on-the-fly;
-        \item security groups, to allow remote access over \ac{SSH} and \ac{VPN} to the compute nodes;
-        \item and private access keys, to securely log in to the nodes over \ac{SSH}.
+        \item \gls{EC2} instances, to create and terminate required compute nodes on-the-fly;
+        \item security groups, to allow remote access over \gls{SSH} and \gls{VPN} to the compute nodes;
+        \item and private access keys, to securely log in to the nodes over \gls{SSH}.
     \end{itemize}
@@
 
-    \Ac{EC2} instances deployed by Ainur are expected to be based on a Linux \ac{AMI} equipped with the similar software as the local hosts.
+    \gls{EC2} instances deployed by Ainur are expected to be based on a Linux \gls{AMI} equipped with the similar software as the local hosts.
     Specifically:
@@
     \begin{itemize}
-        \item Docker is installed and configured for regular users and over-the-network access on \ac{TCP} port \num{2375}.
+        \item Docker is installed and configured for regular users and over-the-network access on \gls{TCP} port \num{2375}.
         \item Netplan and systemd are present and in use.
@@
 
-    \item[\ac{VPN} gateways]
+    \item[\gls{VPN} gateways]
 
@@
 
-        Two gateways are present on the testbed networks, as illustrated in \cref{fig:network}; one for the control traffic \ac{VPN} link, and another for the workload data traffic \ac{VPN} link.
+        Two gateways are present on the testbed networks, as illustrated in \cref{fig:network}; one for the control traffic \gls{VPN} link, and another for the workload data traffic \gls{VPN} link.
 
-        \item[Public \ac{IP} address]
+        \item[Public \gls{IP} address]
 
-        Both gateways are reachable on the same public \ac{IP} address of the gateway router.
-        The gateway router listens on two different \ac{UDP} ports and forwards data to the corresponding \ac{VPN} gateway.
+        Both gateways are reachable on the same public \gls{IP} address of the gateway router.
+        The gateway router listens on two different \gls{UDP} ports and forwards data to the corresponding \gls{VPN} gateway.
 
@@
 
-        Static routes to the control \ac{VPN} link through the corresponding gateway are configured on all devices attached to the control network.
+        Static routes to the control \gls{VPN} link through the corresponding gateway are configured on all devices attached to the control network.
 

### 2022-12-19 publications/2022Ainur/conclusion.tex
glossaries

@@
 We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \ac{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
+It has been released as \gls{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
 
-In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \ac{CHI}~\cite{keahey2020lessons}.
+In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \gls{CHI}~\cite{keahey2020lessons}.
 While we believe Ainur to be unique and valuable in its focus on fully-automated end-to-end wireless edge computing experimentation, integrating with the above software stacks will provide a number of features and functionalities which will greatly expand Ainur's potential.
-Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \ac{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
-Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\ac{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.
+Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
+Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\gls{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.

### 2022-12-19 publications/2022Ainur/demo.tex
glossaries

@@
         \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-            \item \ac{SDR} processing hosts 
-            \item additional base-station host for the Core Network and \ac{eNodeB}
+            \item \gls{SDR} processing hosts
+            \item additional base-station host for the Core Network and \gls{eNodeB}
         \end{enumerate*}.
@@
 \Cref{fig:democonfigs} illustrates the possible configurations for the testbed.
-Workloads are deployed on client hosts, and computation is offloaded either to an edge server (cloudlet), or to \ac{AWS} \ac{EC2} instances on the cloud.
+Workloads are deployed on client hosts, and computation is offloaded either to an edge server (cloudlet), or to \gls{AWS} \gls{EC2} instances on the cloud.
 Communication between the client- and server-sides of the workload will occur over one of three possible physical link layer setups:
@@
     \item[WiFi:] \iac{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board WiFi.
-    \item[4G \ac{LTE}:] \iac{SDR} is configured as an 4G \ac{LTE} base station, and another is configured as an 4G \ac{LTE} \ac{UE}.
-    Together these radios act as an \ac{LTE} bridge between the client-side and the server-side of the network, and all client-server traffic is routed through them.
+    \item[4G \gls{LTE}:] \iac{SDR} is configured as an 4G \gls{LTE} base station, and another is configured as an 4G \gls{LTE} \gls{UE}.
+    Together these radios act as an \gls{LTE} bridge between the client-side and the server-side of the network, and all client-server traffic is routed through them.
     \item[Ethernet:] connects everything through plain ethernet.
@@
 The number of workload instances (and therefore client hosts) deployed in each execution of this demonstration will range from \numrange[]{1}{10}.
-Workloads deployed to the cloud will be able to target any of \ac{AWS}'s datacenters.
+Workloads deployed to the cloud will be able to target any of \gls{AWS}'s datacenters.
 
 We will employ two different workloads for this demonstration, illustrated in \cref{fig:wkld}.
-The first of these (\cref{fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \ac{MNIST} dataset~\cite{mnist}.
+The first of these (\cref{fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{mnist}.
 The application consists of a client with a web interface, and \iac{HTTP} server which responds to requests from the client and performs the actual image recognition.
@@
 It consists of an emulation of the physical inverted pendulum system plant and a software-implemented proportional-differential controller.
-These components communicate with each other over the \ac{UDP}.
+These components communicate with each other over the \gls{UDP}.
 This workload will be used to showcase the utility of Ainur for automating the execution of batches of experiments potentially including multiple different clients, servers, and physical layers.
@@
     \item \num{10} Raspberry Pi 4 Model B boards, acting as client-side workload hosts
-    \item a Raspberry Pi 4 Model B acting as \ac{VPN} gateway for the workload network
-    \item a Raspberry Pi 4 Model B acting as \ac{VPN} gateway for the management network, as well as hosting the necessary \ac{NTP} and \ac{DNS} server software
+    \item a Raspberry Pi 4 Model B acting as \gls{VPN} gateway for the workload network
+    \item a Raspberry Pi 4 Model B acting as \gls{VPN} gateway for the management network, as well as hosting the necessary \gls{NTP} and \gls{DNS} server software
     \item a \texttt{i386} workstation, acting as an edge-side workload host
-    \item a configurable number of  \ac{AWS} \ac{EC2} cloud instances acting as cloud servers
+    \item a configurable number of  \gls{AWS} \gls{EC2} cloud instances acting as cloud servers
     \item a separate \texttt{i386} workstation hosting the Fluent server and on which Ainur is deployed as well.
 \end{enumerate*}
-These nodes are interconnected using a combination of managed switches, \acp{SDR}, and \ac{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.
+These nodes are interconnected using a combination of managed switches, \acp{SDR}, and \gls{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.
 
@@
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item a location to offload computation to (local (i.e.\ no offloading), edge, or any \ac{AWS} datacenter)
-    \item a physical layer for the first hop of the network (WiFi, 4G \ac{LTE}, or ethernet). 
+    \item a location to offload computation to (local (i.e.\ no offloading), edge, or any \gls{AWS} datacenter)
+    \item a physical layer for the first hop of the network (WiFi, 4G \gls{LTE}, or ethernet).
 \end{enumerate*}
 Through the Ainur command line, we will deploy a single client-server pair according to the specifications.
-Next, participants will be able to access the client web interface and interact with the server by requesting classification of images from the \ac{MNIST} database.
+Next, participants will be able to access the client web interface and interact with the server by requesting classification of images from the \gls{MNIST} database.
 The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{fig:mockup}.
@@
 
-Participants will be asked to specify the full, arbitrarily complex, experimental scenario, including the number of clients, physical layers (single or multiple, and which clients are on each physical link), and offloading configuration (number of instances on the edge vs.\ on the cloud, which \ac{AWS} datacenters to deploy to).
+Participants will be asked to specify the full, arbitrarily complex, experimental scenario, including the number of clients, physical layers (single or multiple, and which clients are on each physical link), and offloading configuration (number of instances on the edge vs.\ on the cloud, which \gls{AWS} datacenters to deploy to).
 Configuration will be specified through a \acs{YAML} file, which will then be parsed by Ainur for the automatic execution of the scenario.

### 2022-12-19 publications/2022Ainur/futurework.tex
glossaries

@@
         This also requires quite a lot of synchronization between people working on different parts of the code.
-        \item It makes experimentation synchronous, and when running things over \ac{SSH}, it requires the use of tools such as \texttt{screen} or \texttt{tmux} to be able to run long experiments without having to keep the connection open at all times.
+        \item It makes experimentation synchronous, and when running things over \gls{SSH}, it requires the use of tools such as \texttt{screen} or \texttt{tmux} to be able to run long experiments without having to keep the connection open at all times.
         \item We currently have no way of ensuring mutual exclusion of multiple instances of Ainur on the testbed, and implementing such functionality would probably require some sort of centralized instance tracking mechanism.
@@
     In other words, we would have several smaller server applications running on potentially different hosts on the network, each handling a specific part of the functionality we need.
-    For instance, we could have a ``Layer 3'' service which handles setting up \ac{IP}-layer connectivity, and another ``SDR'' service which manages the \acp{SDR}.
+    For instance, we could have a ``Layer 3'' service which handles setting up \gls{IP}-layer connectivity, and another ``SDR'' service which manages the \acp{SDR}.
     Services could communicate with each other using simple HTTP \acp{API} for easy debugging, and we would have an ``Ainur'' client which orchestrates the services into an experimental run.
@@
         \item It decouples the implementation details for each service.
-        As long as the \ac{API} is well-defined, it doesn't matter how the service is implemented - we could even use different languages for different functionalities.
+        As long as the \gls{API} is well-defined, it doesn't matter how the service is implemented - we could even use different languages for different functionalities.
         \item Adding new functionality becomes simply a matter of writing a new microservice and plugging it into the existing setup.

### 2022-12-19 publications/2022Ainur/introduction.tex
glossaries

@@
     \item \acs*{ARA}
-    \item Drexel Grid \ac{SDR}
+    \item Drexel Grid \gls{SDR}
 \end{enumerate*} testbeds.
@@
 It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
-\ac{COSMOS} relies on the \ac{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \ac{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
+\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
 
@@
 The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
-\ac{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \ac{VM} images containing the necessary software and configurations.
-These profiles are defined through using \ac{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
+\gls{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \gls{VM} images containing the necessary software and configurations.
+These profiles are defined through using \gls{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
 
-The \ac{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
+The \gls{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
 Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
-It includes a broad range of wireless platforms ranging from low-\ac{UHF} massive \ac{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \ac{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-\ac{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \ac{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
-This affords the \ac{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
+It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+\gls{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \gls{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
+This affords the \gls{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
 
-Finally, the \emph{Drexel Grid \ac{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
-Primarily intended for \ac{SDR}-centric research, it does not integrate any core, cloud or edge components.
-However, the testbed extensively employs the \ac{LXC} runtime for the deployment of both experimental code and \ac{SDR} software, which affords users great freedom when it comes to the development of experiments.
+Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
+Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
+However, the testbed extensively employs the \gls{LXC} runtime for the deployment of both experimental code and \gls{SDR} software, which affords users great freedom when it comes to the development of experiments.
 
@@
 Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
-This is, for instance, the case with \ac{COSMOS} and \ac{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
+This is, for instance, the case with \gls{COSMOS} and \gls{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
 These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.
@@
 % \todo[inline]{
-%     COSMOS: Experimenters work on these resources using the \ac*{OMF}~\cite{omf} tool, originally leveraged for ORBIT~\cite{orbit}, and \ac{SSH} sessions, thus allowing for free-form scripting of experimental procedures.
+%     COSMOS: Experimenters work on these resources using the \ac*{OMF}~\cite{omf} tool, originally leveraged for ORBIT~\cite{orbit}, and \gls{SSH} sessions, thus allowing for free-form scripting of experimental procedures.
 
@@
 
-%     POWDER: To abstract and redistribute resources and \acp{API}, \ac{POWDER} leverages the concept of \emph{profiles}, corresponding shareable packages containing experimental configurations and descriptions\cite{emulab}.
-%     Automation and web interface control are limited to some test experiments, and thus regular users must interact with the platform through \ac{SSH}.
+%     POWDER: To abstract and redistribute resources and \acp{API}, \gls{POWDER} leverages the concept of \emph{profiles}, corresponding shareable packages containing experimental configurations and descriptions\cite{emulab}.
+%     Automation and web interface control are limited to some test experiments, and thus regular users must interact with the platform through \gls{SSH}.
 
@@
 %     Drexel: Orchestration in this testbed is facilitated through a container-based approach, leveraging LXC containers, while deployment and configuration is facilitated through Ansible.
-%     Container deployment is requested by users through a web-based reservation system or a \ac{REST} web \ac{API}.
-%     \ac{OS} deployment on host computers is facilitated through Canonical \ac{MaaS}.
+%     Container deployment is requested by users through a web-based reservation system or a \gls{REST} web \gls{API}.
+%     \gls{OS} deployment on host computers is facilitated through Canonical \gls{MaaS}.
 

### 2022-12-19 publications/2022Ainur/software_layers/cloud.tex
glossaries

@@
 
-Integration with \ac{AWS} is handled by the \mintinline{python}{CloudInstances} class.
-The implementation of this class uses \verb|boto3| for interaction with the \ac{AWS}, and thus needs access to appropriate credentials\footnote{%
+Integration with \gls{AWS} is handled by the \mintinline{python}{CloudInstances} class.
+The implementation of this class uses \verb|boto3| for interaction with the \gls{AWS}, and thus needs access to appropriate credentials\footnote{%
     \url{https://boto3.amazonaws.com/v1/documentation/api/latest/guide/credentials.html}%
@@
 
-Each instance of this class is associated with a specific \ac{AWS} region, and manages the end-to-end lifecycle of an arbitrary number of \ac{AWS} \ac{EC2} instances on said region.
+Each instance of this class is associated with a specific \gls{AWS} region, and manages the end-to-end lifecycle of an arbitrary number of \gls{AWS} \gls{EC2} instances on said region.
 Specifically, \mintinline{python}{CloudInstances} handles
@@
 
-By default, \mintinline{python}{CloudInstances} spawns instances of type \verb|t3.micro|, with at least one security group allowing inbound \ac{SSH} traffic and unrestricted traffic between instances.
+By default, \mintinline{python}{CloudInstances} spawns instances of type \verb|t3.micro|, with at least one security group allowing inbound \gls{SSH} traffic and unrestricted traffic between instances.
 The type of instances spawned can be modified at runtime, and additional security groups can be added at will.
 
-The spawned \ac{EC2} instances are created from a preconfigured \ac{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
-This \ac{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \ac{AMI} --- or a custom \ac{AMI} which has been preconfigured by the user.
-Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \ac{AMI} used for deploying (with the exception that it needs to be a Linux-based \ac{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.
+The spawned \gls{EC2} instances are created from a preconfigured \gls{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
+This \gls{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \gls{AMI} --- or a custom \gls{AMI} which has been preconfigured by the user.
+Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \gls{AMI} used for deploying (with the exception that it needs to be a Linux-based \gls{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.

### 2022-12-19 publications/2022Ainur/software_layers/flow.tex
glossaries

@@
 Immediately after start, Ainur reads and parses the deployment configuration and parses it into a set of various dataclasses and Python dictionaries.
-Next, if deploying anything to the cloud, Ainur spawns the required number of \ac{AWS} \ac{EC2} instance.
+Next, if deploying anything to the cloud, Ainur spawns the required number of \gls{AWS} \gls{EC2} instance.
 Execution then goes on to configure the physical links, connecting hosts with each other through clever management of \acp{VLAN} and \acp{SDR}.
-Once physical and/or \ac{VLAN} connections are configure, \ac{IP} addresses are assigned to devices and static routes between them are configured.
-If cloud instances are used, \ac{VPN} connections are established to them.
+Once physical and/or \gls{VLAN} connections are configure, \gls{IP} addresses are assigned to devices and static routes between them are configured.
+If cloud instances are used, \gls{VPN} connections are established to them.
 At this stage, the testbed network is fully configured, and bidirectional traffic between any two hosts is possible over the workload data network.

### 2022-12-19 publications/2022Ainur/software_layers/ip_layer.tex
glossaries

@@
-\subsubsection{\ac{IP} Layer Connectivity}\label{sec:layer3}
+\subsubsection{\gls{IP} Layer Connectivity}\label{sec:layer3}
 
-Layer 3 connectivity, that is, \ac{IP} addressing and routing, is handled by two main components in Ainur: \mintinline{python}{LANLayer} and \mintinline{python}{VPNCloudMesh}.
+Layer 3 connectivity, that is, \gls{IP} addressing and routing, is handled by two main components in Ainur: \mintinline{python}{LANLayer} and \mintinline{python}{VPNCloudMesh}.
 
@@
 It can be used as a context manager for automatic network teardown, as well as an iterator over the hosts it manages.
-This class exposes two main \ac{API} methods, \mintinline{python}{self.add_hosts()} and \mintinline{python}{self.tear_down()}:
+This class exposes two main \gls{API} methods, \mintinline{python}{self.add_hosts()} and \mintinline{python}{self.tear_down()}:
 \begin{description}[]
     \item[\mintinline{python}{LANLayer.add_hosts(self, layer2)}]
-    This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \ac{IP} layer connectivity on the hosts within.
-    This includes assigning \ac{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
+    This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \gls{IP} layer connectivity on the hosts within.
+    This includes assigning \gls{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
     The underlying implementation uses Ansible to write Netplan YAML configuration files on the target hosts and subsequently apply the network configuration.
@@
     \item[\mintinline{python}{LANLayer.tear_down(self)}]
-    Performs a clean teardown of \ac{IP} layer connectivity of the hosts managed by the instance, removing the created Netplan configuration files and reverting host network configuration to its initial state.
+    Performs a clean teardown of \gls{IP} layer connectivity of the hosts managed by the instance, removing the created Netplan configuration files and reverting host network configuration to its initial state.
     When \mintinline{python}{LANLayer} is used as a context manager, this method is automatically called the moment execution leaves the \mintinline{python}{with} block.
@@
 
-\mintinline{python}{VPNCloudMesh}\label{sec:vpn} handles \ac{VPN} connectivity to running cloud instances.
-It makes some strong assumptions about the presence of two preconfigured \ac{VPN} gateways on the testbed (some of these have been mentioned above, but we will repeat them here for clarity):
+\mintinline{python}{VPNCloudMesh}\label{sec:vpn} handles \gls{VPN} connectivity to running cloud instances.
+It makes some strong assumptions about the presence of two preconfigured \gls{VPN} gateways on the testbed (some of these have been mentioned above, but we will repeat them here for clarity):
 
@@
     \item One of these gateways is additionally connected to the workload data network through a separate interface.
-    \item These gateways are both listening on separate \ac{UDP} ports on a single public IP address.
-    In other words, both gateways access the public internet through a single \ac{NAT} gateway, on which appropriate \ac{UDP} port forwarding rules are configured to allow \ac{VPN} traffic back into the control network.
+    \item These gateways are both listening on separate \gls{UDP} ports on a single public IP address.
+    In other words, both gateways access the public internet through a single \gls{NAT} gateway, on which appropriate \gls{UDP} port forwarding rules are configured to allow \gls{VPN} traffic back into the control network.
     \item The VPNCloud connections configured on the the gateways share the same pre-shared key.
     \item The VPNCloud links belong to mutually \emph{disjoint} \acs{IP}v4 address ranges.
-    % \item The gateways are assigned the first host address in the available address space of their respective \ac{VPN} links.
-    % For example, if the gateway \ac{VPN} link address space is \texttt{10.0.0.0/24}, the gateway has address \texttt{10.0.0.1} on the \ac{VPN} link.
-    \item Devices on the control network have been preconfigured with appropriate \ac{IP} routes to reach the control gateway and its associated \ac{VPN} link.
+    % \item The gateways are assigned the first host address in the available address space of their respective \gls{VPN} links.
+    % For example, if the gateway \gls{VPN} link address space is \texttt{10.0.0.0/24}, the gateway has address \texttt{10.0.0.1} on the \gls{VPN} link.
+    \item Devices on the control network have been preconfigured with appropriate \gls{IP} routes to reach the control gateway and its associated \gls{VPN} link.
     \item Finally, cloud instances intended to be used with this class have VPNCloud installed on them.
@@
 
-The \ac{API} of this class exposes the below relevant methods:
+The \gls{API} of this class exposes the below relevant methods:
 
@@
     \begin{description}[font=\normalfont\ttfamily\underline]
-        \item[gateway\_ip] An \texttt{IPv4Address} object corresponding to the shared public \ac{IP}v4 address of the \ac{VPN} gateways (i.e.\ the public \ac{IP} address of the \ac{NAT} gateway.)
+        \item[gateway\_ip] An \texttt{IPv4Address} object corresponding to the shared public \gls{IP}v4 address of the \gls{VPN} gateways (i.e.\ the public \gls{IP} address of the \gls{NAT} gateway.)
 
         \item[vpn\_psk]
-        The pre-shared key used to authenticate and encrypt traffic on the \ac{VPN} links.
+        The pre-shared key used to authenticate and encrypt traffic on the \gls{VPN} links.
 
@@
         \item[gw\_mgmt\_ip~\normalfont{and} gw\_wkld\_ip]
-        Full \ac{IP} addresses of the \ac{VPN} links for the control and workload data networks, respectively, including subnet masks.
+        Full \gls{IP} addresses of the \gls{VPN} links for the control and workload data networks, respectively, including subnet masks.
         These values are passed as \mintinline[breakafter=4]{python}{IPv4Interface} objects, and are used to set up correct routing tables to the cloud instances.
@@
         \item[mgmt\_port~\normalfont{and} wkld\_port]
-        \ac{UDP} ports on the \ac{NAT} gateway forwarded to the control and workload gateways, respectively.
+        \gls{UDP} ports on the \gls{NAT} gateway forwarded to the control and workload gateways, respectively.
     \end{description}
@@
     The \texttt{cloud\_layer} argument corresponds to a \mintinline{python}{CloudInstances} object instance, and the \texttt{host\_configs} to a collection of host configuration dataclasses.
-    This method matches the host configurations to cloud instances in the \mintinline{python}{CloudInstances} object and connects them to to the local control and workload data networks through the two aforementioned separate \ac{VPN} links.
-    The underlying implementation uses a combination of the \texttt{boto3} library and Ansible to configure appropriate security groups on the instances and subsequently initialize and connect the \ac{VPN} links.
+    This method matches the host configurations to cloud instances in the \mintinline{python}{CloudInstances} object and connects them to to the local control and workload data networks through the two aforementioned separate \gls{VPN} links.
+    The underlying implementation uses a combination of the \texttt{boto3} library and Ansible to configure appropriate security groups on the instances and subsequently initialize and connect the \gls{VPN} links.
 
     \item[\mintinline{python}{VPNCloudMesh.tear_down(self)}]
-    This method tears down the \ac{VPN} links, disconnecting the remote instances from the gateways and removing any created routes to the cloud from local devices.
+    This method tears down the \gls{VPN} links, disconnecting the remote instances from the gateways and removing any created routes to the cloud from local devices.
     When an instance of this class is used as a context manager, this method is automatically called at the end of the \mintinline{python}{with} block, ensuring automatic cleanup of resources.

### 2022-12-19 publications/2022Ainur/software_layers/physical.tex
glossaries

@@
 This layer creates and deploy the underlying physical connections of the data network and is implemented in the \mintinline{python}{PhysicalLayer} class. 
-Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \ac{LTE} network.
+Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \gls{LTE} network.
 
@@
 
-Currently, WiFi and \ac{LTE} implementations are available. 
+Currently, WiFi and \gls{LTE} implementations are available.
 User-defined implementation for other wireless connections are possible and encouraged.
@@
     \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station WiFi nodes.
-    It searches for an WiFi \ac{SSID} among the networks that are being advertised and attempts for establishing the connection.
+    It searches for an WiFi \gls{SSID} among the networks that are being advertised and attempts for establishing the connection.
 
@@
 
-    \item[\ac{SDR}-based \ac{LTE} Network]
-    \mintinline{python}{SDRLTE} class can configure and deploy a complete software-defined \ac{LTE} network.
+    \item[\gls{SDR}-based \gls{LTE} Network]
+    \mintinline{python}{SDRLTE} class can configure and deploy a complete software-defined \gls{LTE} network.
     The constructor must be provided with an \mintinline{python}{EPC} class, a set of \mintinline{python}{ENodeB} classes, and a set of \mintinline{python}{LTEUE} classes from the \emph{autoran} Python package\footnote{See https://github.com/samiemostafavi/autoran}.
-    This package exploits the \emph{docker-py} framework to run containerized OpenAirInterface \ac{LTE} functions on remote hosts.
+    This package exploits the \emph{docker-py} framework to run containerized OpenAirInterface \gls{LTE} functions on remote hosts.
     Network configurations are needed in the form of Python dictionaries for the construction of \emph{autoran} classes.
 
-    \ac{LTE} \ac{EPC} consists of HSS, MME, SPGW-U, and SPGW-C services which can be run on the same host.
-    For each \ac{eNodeB} or \ac{UE}, a software-defined radio and a host for signal processing must be specified additionally.
-    The resulting topology of an \ac{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
-    Unlike WiFi wireless connection, due to the fact that \ac{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.
+    \gls{LTE} \gls{EPC} consists of HSS, MME, SPGW-U, and SPGW-C services which can be run on the same host.
+    For each \gls{eNodeB} or \gls{UE}, a software-defined radio and a host for signal processing must be specified additionally.
+    The resulting topology of an \gls{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
+    Unlike WiFi wireless connection, due to the fact that \gls{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.
 
@@
         \includegraphics[width=0.8\linewidth]{figures/lte.png}
-        \caption{An example of \ac{LTE} data network topology in data network and its corresponding \acp{VLAN}}
+        \caption{An example of \gls{LTE} data network topology in data network and its corresponding \acp{VLAN}}
         \label{fig:lte}

### 2022-12-19 publications/2022Ainur/software_layers/storage_logging.tex
glossaries

@@
 Dynamic, distributed storage shared across testbed hosts is managed by the \mintinline{python}{ExperimentStorage} class.
-This class exposes a very simple \ac{API}:
+This class exposes a very simple \gls{API}:
 \begin{itemize}
@@
 At deployment time, Ainur configures a FluentBit client container in each workload host which captures logs produced by the workload.
-These FluentBit clients are configured to listen on a \ac{TCP} port to which workload logs are forwarded by Dockers built-in FluentD logging driver\footnote{\url{https://docs.docker.com/config/containers/logging/fluentd/}}.
+These FluentBit clients are configured to listen on a \gls{TCP} port to which workload logs are forwarded by Dockers built-in FluentD logging driver\footnote{\url{https://docs.docker.com/config/containers/logging/fluentd/}}.
 This means that logs are automatically captured by Docker from the \emph{standard output} of the containerized workload process.

### 2022-12-19 publications/2022Ainur/software_layers/swarm.tex
glossaries

@@
 
-A simple example of the usage of this class and its basic \ac{API} can be observed in \cref{lst:swarm:example}, and in the following we will detail the major steps in operating this class:
+A simple example of the usage of this class and its basic \gls{API} can be observed in \cref{lst:swarm:example}, and in the following we will detail the major steps in operating this class:
 \begin{description}[]

### 2022-12-19 publications/2022CLEAVE/acks.tex
glossaries

\section*{Acknowledgements}\label{sec:acks}

This research has been partially funded by
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
    \item the VINNOVA Competence Center for \gls{TECoSA} at KTH Royal Institute of Technology
    \item the \gls{SSF}, through grant number ITM17--0246.
\end{enumerate*}

We also wish to thank S.~S.~Mostafavi, V.~N.~Moothedath, M.~Alsakati, O.~Ferm, S.~Vojcic, and A.~Bhattacharya for their support and valuable contributions to this work.

Finally, we would like to thank the reviewers for their insightful comments and suggestions, which greatly helped us improve this work.

### 2022-12-19 publications/2022CLEAVE/cleave.tex
glossaries

@@
-\section{The \ac{CLEAVE} framework}\label{sec:approach}
+\section{The \gls{CLEAVE} framework}\label{sec:approach}
 
-In this section, we detail our \ac{CLEAVE} framework for the performance evaluation of \aclp{NCS}, with an emphasis on edge deployments.
+In this section, we detail our \gls{CLEAVE} framework for the performance evaluation of \aclp{NCS}, with an emphasis on edge deployments.
 Our design follows a virtualized emulation approach; providing a framework for the real-time \emph{emulation} of physical control systems and the easy implementation of softwarized controllers interacting with \emph{real} networks.
-\ac{CLEAVE} is thus a platform with an easy-to-use \ac{API} on which a multitude of \ac{NCS} types can be emulated.
-This allows for great flexibility, as the core components of the \ac{NCS} can be switched out while maintaining the realism of the network, the most complex and limiting component of edge \acp{NCS}.
+\gls{CLEAVE} is thus a platform with an easy-to-use \gls{API} on which a multitude of \gls{NCS} types can be emulated.
+This allows for great flexibility, as the core components of the \gls{NCS} can be switched out while maintaining the realism of the network, the most complex and limiting component of edge \acp{NCS}.
 
 The framework is distributed as \emph{free and open-source} software through the \emph{GitHub} organization of the \emph{KTH ExPECA} research group\footnote{\url{https://github.com/KTH-EXPECA/CLEAVE}}.
-We also plan to provide a library of \ac{NCS} implementations, making the tool even more accessible.
+We also plan to provide a library of \gls{NCS} implementations, making the tool even more accessible.
 
@@
 
-The design of \ac{CLEAVE} attempts to target a broad a set of different control systems as possible, and tries to minimize the number of assumptions made about the systems implemented on top of the framework.
+The design of \gls{CLEAVE} attempts to target a broad a set of different control systems as possible, and tries to minimize the number of assumptions made about the systems implemented on top of the framework.
 These are:
@@
 
-\cref{fig:cleave:ncs:struct} shows an overview of the conceptual structure and operation of \ac{CLEAVE}.
+\cref{fig:cleave:ncs:struct} shows an overview of the conceptual structure and operation of \gls{CLEAVE}.
 The \emph{State} implements the discrete-time behavior of the Plant.
@@
 
-Implementation-wise, \ac{CLEAVE} is built on top of the well-established and stable \emph{Twisted}\footnote{\url{https://twistedmatrix.com}} asynchronous programming and networking framework.
+Implementation-wise, \gls{CLEAVE} is built on top of the well-established and stable \emph{Twisted}\footnote{\url{https://twistedmatrix.com}} asynchronous programming and networking framework.
 The aforementioned core components (\emph{State}, \emph{Controller}, \emph{Sensors}, and \emph{Actuators}) are provided as \acp{ABC} with core functionality that users must extend for their specific use cases and \acp{NCS}.
@@
 Users are therefore free to implement arbitrarily complex functionality in these components.
-On the other hand, \ac{CLEAVE} handles network communication and metric collection autonomously; users only need to select the desired transport protocol (we do however provide an \ac{API} for extending the available transports).
+On the other hand, \gls{CLEAVE} handles network communication and metric collection autonomously; users only need to select the desired transport protocol (we do however provide an \gls{API} for extending the available transports).
 
@@
 \textbf{State Base Class}:
-\ac{ABC} with an abstract method
+\gls{ABC} with an abstract method
 \mintinline{python}{advance()} which must be extended with the implementation of discrete-time emulation of the plant.
@@
 Defines a single abstract method \mintinline{python}{process()}, taking a mapping of names to values corresponding to the sensed variables of the \emph{State}, and returning a similar mapping corresponding to the actuated variables.
-This method is called by \ac{CLEAVE} in an event-based manner whenever samples are received from the plant side; however, we intend to extend this to time-interval-based approaches.
+This method is called by \gls{CLEAVE} in an event-based manner whenever samples are received from the plant side; however, we intend to extend this to time-interval-based approaches.
 
@@
 These Python configuration files are then provided as arguments to the \texttt{cleave} command line provided by the framework.
-\ac{CLEAVE} imports and parses them, and then sets up and executes the emulation.
+\gls{CLEAVE} imports and parses them, and then sets up and executes the emulation.
 
-Although the core design of \ac{CLEAVE} follows a single-loop approach, multi-loop scenarios can be easily implemented through the combination of multiple \ac{CLEAVE} instances.
+Although the core design of \gls{CLEAVE} follows a single-loop approach, multi-loop scenarios can be easily implemented through the combination of multiple \gls{CLEAVE} instances.
 The framework does not make any assumptions about the individual components other than the signatures of methods on the \acp{ABC} that must be extended and implemented.
@@
 
-For the purpose of this work, we have used \ac{CLEAVE} to implement an emulated \emph{inverted pendulum} control loop (see \cref{fig:invpend}), the goal of which is to balance the vertically free-swinging pendulum by applying horizontal forces on the cart.
+For the purpose of this work, we have used \gls{CLEAVE} to implement an emulated \emph{inverted pendulum} control loop (see \cref{fig:invpend}), the goal of which is to balance the vertically free-swinging pendulum by applying horizontal forces on the cart.
 We have chosen this system as an initial benchmark for its relative simplicity as well as prevalence in the field of automatic control as one of the fundamental examples of linear control.
@@
 }
-The inverted pendulum plant is merely a proof-of-concept, and \ac{CLEAVE} could be used to implement much more complex physical systems with even more stringent latency requirements.
+The inverted pendulum plant is merely a proof-of-concept, and \gls{CLEAVE} could be used to implement much more complex physical systems with even more stringent latency requirements.
 For instance, the flexibility afforded by the choice of programming language and the abstractions detailed in \cref{sec:design}, allow for relatively straightforward emulation of autonomous drone dynamics.

### 2022-12-19 publications/2022CLEAVE/conclusion.tex
glossaries

--- 2022-12-19 body/part1/overview/methodology.tex
@@
-\section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}
+\section{Conclusion}\label{sec:conclusion}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=.8\textwidth]{Figs/methodology.png}
-    \caption{%
-        Overview of the proposed methodology for the study of feedback-loop systems on edge computing infrastructure.
-        The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
-        The backend/server side of the system along with the network remain unchanged.
-    }\label{fig:methodology}
-\end{figure}
-
-As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \gls{NCS}) or with human involvement (such as \gls{WCA}).
-In the following, we introduce our methodological framework for tackling this challenge.
-
-A general overview of our methodology is illustrated in \cref{fig:methodology}.
-It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
-We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \gls{COTS} general-purpose computing devices.
-The backend side and the network are not changed.
-
-Two main arguments drive this approach.
-One, the difficulty of scaling the number of clients in these systems in a research context.
-The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \gls{WCA}) or due its cyber-physical nature (in \acp{NCS}).
-Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
-Raspberry Pis, Beaglebones, etc.).
-
-And two, it maintains the realism of the most complex parts of the system.
-
-\todo[inline]{Fix and finish}
-
-In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
-We first discuss the methodology in the context of a human-in-the-loop application, \acl{WCA}, in section \cref{ssec:methodology:wca}.
-Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \acf{NCS}.
-
-\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{.47\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/trace_edgedroid.png}
-        \caption{%
-            High level conceptual design of our methodology for \gls{WCA}.
-            We replace the human by a trace-driven emulation.
-            This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
-        }
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{.47\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
-        \caption{%
-            Architectural overview of the EdgeDroid \num{1.0} tool.
-            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \gls{WCA} backend running on a cloudlet.%
-        }
-    \end{subfigure}\\
-    \medskip
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \adjustbox{scale=0.7}{
-            \begin{tikzpicture}[align=center,
-                node distance=.5cm and 1.5cm,
-                every initial by arrow/.style={-{Latex[length=2mm]}}]
-                % Place nodes
-                \node [initial, state, minimum size=6em, initial text=] (play) {Play};
-                \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
-                \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
-                \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};
-
-                % Draw edges
-                \path[draw, -{Latex[length=2mm]}]
-                (play) edge [bend right=20] node[left] {Step done} (rewind)
-                edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
-                edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)
-
-                (change) edge [bend left=20] node[right] {Step changed} (play)
-                edge [bend left=20] node[right] {All steps done} (shutdown)
-
-                (rewind) edge [bend right=20] node[right] {Rewound} (play)
-                edge [bend right=20] node[right] {Too many rewinds} (shutdown);
-
-            \end{tikzpicture}
-        }
-        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.
-This model approximates the behavior of an ``ideal'' human, one that is patient and makes no mistakes.}\label{fig:usermodel}
-    \end{subfigure}
-    \caption{%
-        Overview of our first implementation of the methodology for \gls{WCA}.
-    }\label{fig:edgedroid1:trace}
-\end{figure}
-
-We first introduce our methodology in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
-The former corresponds to an extended abstract paper which discusses a high level overview of our approach; the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
-
-The overall goal of these papers is to explore this methodological approach as applied to \gls{WCA}.
-As discussed in \cref{chap:introduction}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
-Humans are unreliable, and, perhaps more importantly, hard to scale.
-\todo[inline]{More reasons}
-
-The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
-We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \gls{WCA} backend.
-The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile devices on which real \gls{WCA} applications are intended to run in the future.
-Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
-On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer.
-The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
-Once recorded, the trace is manually segmented into the logical component steps of the task, and these are replayed in sequence back to the backend by the EdgeDroid tool.
-
-We opt for a trace-based approach for our first implementation of the methodology for \gls{WCA} for three key reasons.
-The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
-Using a trace ensures that the same computation is performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-
-The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
-Compatible traces can be easily obtained simply by instrumenting existing \gls{WCA} client applications to record all captured inputs.
-
-The final advantage corresponds to extensibility.
-As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
-
-\todo[]{}
-
-reproducible and comparable workload, we use a trace-based design for the client-side emulation.
-.
-
-To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
-This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
-
-However, merely replaying the trace is not enough, as disparities between system responsiveness at recording versus replay time can cause the trace so drift out of synchronization with the current state of the system.
-Thus, we introduce here the concept of a human user model which is able to adapt its behavior in a realistic manner based on the current responsiveness of the system by systematically controlling the replay of the trace.
-The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
-The trace is preprocessed into segments that correspond to individual steps of the \gls{WCA} application; this is performed manually as a preparation step.
-The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \gls{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
-On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
-To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.
-
-\todo[inline]{EdgeDroid paper}
-
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
-    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
-    \label{fig:comparison:rtt}%
-\end{figure}%
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
-    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
-    \label{fig:comparison:feedback}%
-\end{figure}%
-\begin{figure}%
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
-    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
-    \label{fig:comparison:tasksteps}
-\end{figure}%
-
-The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
-We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
-
-Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
-This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
-EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
-This sensory information is then processed by the original compute process at the backend, generating feedback.
-However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
-Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
-Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
-Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
-It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
-To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications
-
-\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
-Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
-We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
-Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
-These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
-Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.
-
-In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.
-
-For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
-This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.
-
-We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
-These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
-It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.
-
-We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
-One, scaling behavior for the application seems to be linear with respect to the number of clients.
-Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
-
-It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
-Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.
-
-EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
-From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
-Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
-Here, the downlink time sometimes almost reaches the ideal bound by itself.
-A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.
-
-Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
-The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.
-
-\todo[inline]{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
-
-\subsection{Applying the methodology to \acsp{NCS}}\label{ssec:methodology:ncs}
-
-\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \gls{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \gls{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\gls{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+The issue of repeatable and scalable benchmarks has been largely glossed over in \gls{NCS} literature, as existing experimental research studies tend to implement \emph{ad-hoc} solutions.
 
@@
 
-\todo[inline]{Some plots and results}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.6\textwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
-    \caption{
-        The 2D inverted pendulum system.
-    }\label{fig:invpend}
-\end{figure}
-
-\begin{figure}[t]
-    \centering
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
-        \caption{Toppled plants.}\label{fig:video:toppled}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \gls{RMS}.}\label{fig:video:rms}
-    \end{subfigure}\\
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
-        \caption{Packet losses.}\label{fig:video:drop}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\acsp{RTT}.}\label{fig:video:rtt}
-    \end{subfigure}%
-    \caption{
-        Multi-loop, resource-constrained setup results.
-        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
-    }\label{fig:video:results}
-\end{figure}
-
-\FloatBarrier%
-
-\subsection{Orchestrating experimentation in edge computing testbeds}
-
-\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
-
-In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
-Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
-The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
-It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \gls{FOSS}.
-It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
-
-In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
-Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \gls{LTE} and 5G
-    \item provisioning of and connection to remote cloud instances
-    \item initialization of \gls{IP} layer connectivity between hosts
-    \item collection of logs and data
-    \item deployment, scaling, and lifecycle management of containerized processes
-\end{enumerate*}.
-We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
-We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
-We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \gls{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
-
-\todo[inline]{Some plots}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
-    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
-    \caption{Network structure assumed by Ainur.}\label{fig:network}
-\end{figure*}
+There is still, however, work to be done.
+We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \acp{NCS} to share with the community.
+At the moment, the interactions of \gls{CLEAVE} and tools such as Docker are still quite superficial.
+Our goal is to achieve a much tighter integration, e.g.\ by providing the toolkit as pre-packaged container images.
+Finally, the validity of the results obtained by the framework will have to be verified through more thorough, realistic scenarios than what we have been able to show in this work.
+In particular, we intend to perform large-scale experimentation targetting 5G cellular deployments, as this technology is set to become the backbone of edge networks in the near future.

### 2022-12-19 publications/2022CLEAVE/experiments.tex
glossaries

@@
 
-In this section, we demonstrate the utility of \ac{CLEAVE} by walking readers through an example use case.
+In this section, we demonstrate the utility of \gls{CLEAVE} by walking readers through an example use case.
 With this, we aim to showcase the ability of our framework to provide accurate and repeatable measurements of the performance of \acp{NCS} deployed on edge computing infrastructure.
@@
 We present the following use case scenario.
-A simple \ac{NCS} consisting of an \emph{inverted pendulum} plant controlled by a \emph{proportional-differential} controller is to be deployed on an Edge server.
+A simple \gls{NCS} consisting of an \emph{inverted pendulum} plant controlled by a \emph{proportional-differential} controller is to be deployed on an Edge server.
 The connection between plant and server is over a WiFi (IEEE 802.11n) link.
@@
 Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \acp{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
-4G, for instance, has average \acp{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \ac{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
+4G, for instance, has average \acp{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \gls{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
 Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \acp{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
-Finally, video analytics is one of the main proposed use cases for edge computing~\cite{Ananthanarayanan2017Analytics,Yi2017Analytics,Wang2018Analytics}, and thus we foresee edge \ac{NCS} deployments being deployed in parallel with such applications in the future.
+Finally, video analytics is one of the main proposed use cases for edge computing~\cite{Ananthanarayanan2017Analytics,Yi2017Analytics,Wang2018Analytics}, and thus we foresee edge \gls{NCS} deployments being deployed in parallel with such applications in the future.
 However, before deployment can become a reality, key questions such as: 
-``what are the baseline requirements of the \ac{NCS}?''; ``what are the achievable best-case end-to-end latencies in the system?''; and
-``how does the concurrent video-streaming traffic affect \ac{NCS} stability?'' need to be answered.
+``what are the baseline requirements of the \gls{NCS}?''; ``what are the achievable best-case end-to-end latencies in the system?''; and
+``how does the concurrent video-streaming traffic affect \gls{NCS} stability?'' need to be answered.
 
 We attempt to study these questions using the experimental setup in \cref{fig:cleave:expsetup}.
-\ac{CLEAVE} is deployed on a testbed consisting of \num{10} Raspberry Pi 4B  clients connected wirelessly to an IEEE 802.11n \ac{AP}.
-Connected to the Ethernet backbone of this \ac{AP} is a general-purpose \verb|x86_64| desktop computer acting as a Cloudlet/Edge server.
+\gls{CLEAVE} is deployed on a testbed consisting of \num{10} Raspberry Pi 4B  clients connected wirelessly to an IEEE 802.11n \gls{AP}.
+Connected to the Ethernet backbone of this \gls{AP} is a general-purpose \verb|x86_64| desktop computer acting as a Cloudlet/Edge server.
 
@@
         The setup used for our experimentation. 
-        % Containerized versions of the core CLEAVE emulation components are deployed inside a Docker Swarm Overlay Network spanning \num{10} Raspberry Pi 4B clients connected to a single Cloudlet over an IEEE 802.11n \ac{AP}.
+        % Containerized versions of the core CLEAVE emulation components are deployed inside a Docker Swarm Overlay Network spanning \num{10} Raspberry Pi 4B clients connected to a single Cloudlet over an IEEE 802.11n \gls{AP}.
     }\label{fig:cleave:expsetup}
@@
 The first step in our case study is to evaluate the baseline performance of the control system, both locally and over the network.
-We configure \ac{CLEAVE} to run a single loop with both plant and controller co-located on the Edge server, and we also configure a separate setup with an identical loop executed over the wireless link.
+We configure \gls{CLEAVE} to run a single loop with both plant and controller co-located on the Edge server, and we also configure a separate setup with an identical loop executed over the wireless link.
 For both of these setups, we execute a series of scenarios varying:
@@
 
-This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \ac{CLEAVE}'s design.
+This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \gls{CLEAVE}'s design.
 Scenarios are executed automatically in batches using a simple Python script which interacts with Docker through the widely adopted \emph{docker-py}\footnote{Docker SDK for Python: \url{https://docker-py.readthedocs.io/en/stable/}} library.
-This is \ac{CLEAVE}'s first advantage over existing frameworks; it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.
+This is \gls{CLEAVE}'s first advantage over existing frameworks; it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.
 
@@
 \cref{fig:single} shows a summary of the results relating to the single-loop scenarios:
-\cref{fig:single:topple} shows the fraction of plants that toppled in each scenario, and \cref{fig:single:rms} shows the average \ac{RMS} for the absolute pendula angles.
+\cref{fig:single:topple} shows the fraction of plants that toppled in each scenario, and \cref{fig:single:rms} shows the average \gls{RMS} for the absolute pendula angles.
 \cref{fig:single:rtt} shows average latency due to network and processing (excluding synthetic delays) for both single-loop scenarios.
@@
 As expected, higher sampling rates tend to correlate with better quality of control; at higher sampling rates the system was able to reach stability at higher \acp{RTT}.
-These initial results already hint at interesting consequences for such an edge-bound \ac{NCS} deployment.
+These initial results already hint at interesting consequences for such an edge-bound \gls{NCS} deployment.
 For instance, it is clear from \cref{fig:single:topple,fig:single:rms} that network delays can, to a certain extent, be compensated for by increasing the sampling rate of the system.
@@
 We deploy \num{6} control loops on the experimental setup depicted in \cref{fig:cleave:expsetup}.
-On the remaining \num{4} clients we run the \emph{iperf3}\footnote{iperf3: \url{https://iperf.fr/}} traffic load generator, each generating \SI[per-mode=symbol]{6.5}{\mega\bit\per\second} of uplink \ac{UDP} traffic.
+On the remaining \num{4} clients we run the \emph{iperf3}\footnote{iperf3: \url{https://iperf.fr/}} traffic load generator, each generating \SI[per-mode=symbol]{6.5}{\mega\bit\per\second} of uplink \gls{UDP} traffic.
 This emulates the load generated on the network by \SI{1080}{p} Full-HD video streaming, originating from the clients and terminating in the cloudlet.
-Based on the baseline results, we execute setups with \ac{NCS} plant sampling rates of \SIlist{20;40;60}{\hertz}.
+Based on the baseline results, we execute setups with \gls{NCS} plant sampling rates of \SIlist{20;40;60}{\hertz}.
 Each sampling rate configuration is run for \SI{5}{\minute}, and then repeated \num{30} times to obtain statistical significance.
@@
 \cref{fig:video:toppled} shows the fraction of repetitions of each setup in which \emph{at least} one plant failed to maintain stability and toppled.
-\cref{fig:video:rms} shows the \ac{RMS} for the pendulum angles for each setup, only considering data from plants that did not topple.
-\cref{fig:video:drop} shows the fraction of \ac{UDP} datagrams dropped, averaged over all plants and repetitions per setup.
-\cref{fig:video:rtt} shows the measured end-to-end plant-side \ac{RTT}, averaged over all plants and repetitions per setup.
+\cref{fig:video:rms} shows the \gls{RMS} for the pendulum angles for each setup, only considering data from plants that did not topple.
+\cref{fig:video:drop} shows the fraction of \gls{UDP} datagrams dropped, averaged over all plants and repetitions per setup.
+\cref{fig:video:rtt} shows the measured end-to-end plant-side \gls{RTT}, averaged over all plants and repetitions per setup.
 
 These results are interesting in their counter-intuitiveness compared to the single-loop baseline results.
-While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, \cref{fig:video:toppled,fig:video:rms} show this not to be the case for \ac{NCS} in resource-constrained scenarios.
-\SI{60}{\hertz} was the least stable configuration, with at least one pendulum toppling in around \SI{16}{\percent} of the repetitions, and mean pendulum angle \ac{RMS} circa \num{3} times that of the \SI{40}{\hertz} scenario.
-\SI{40}{\hertz} was in turn the second worst configuration --- although it presented no toppled pendula, average angle \ac{RMS} doubled that of the \SI{20}{\hertz} setup.
+While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, \cref{fig:video:toppled,fig:video:rms} show this not to be the case for \gls{NCS} in resource-constrained scenarios.
+\SI{60}{\hertz} was the least stable configuration, with at least one pendulum toppling in around \SI{16}{\percent} of the repetitions, and mean pendulum angle \gls{RMS} circa \num{3} times that of the \SI{40}{\hertz} scenario.
+\SI{40}{\hertz} was in turn the second worst configuration --- although it presented no toppled pendula, average angle \gls{RMS} doubled that of the \SI{20}{\hertz} setup.
 
@@
 The Edge will be \emph{multi-tenant} and \emph{multi-instance}. 
-\ac{NCS} deployments will have to be designed with shared resources in mind, and given the complexity of these systems, experimental tools like \ac{CLEAVE} will be key for their succesful adoption and massification.
+\gls{NCS} deployments will have to be designed with shared resources in mind, and given the complexity of these systems, experimental tools like \gls{CLEAVE} will be key for their succesful adoption and massification.

### 2022-12-19 publications/2022CLEAVE/intro.tex
glossaries

@@
 This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
-These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \ac{CPS} in the future.
+These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
 
-\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \ac{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
 Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
@@
 This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-Conversely, other studies choose to instead use completely \emph{simulated} \ac{NCS} setups.
+Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
 The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
@@
 This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \ac{NCS}.
+Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
 This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
@@
 
-\ac{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
+\gls{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
 It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
@@
 \cref{sec:approach} presents the design principles and implementation of the framework.
-In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \ac{CLEAVE}.
+In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \gls{CLEAVE}.
 Finally, in \cref{sec:conclusion} we conclude this work.

### 2022-12-19 body/glossary.tex
glossary fixes

@@
 \newabbreviation{CPS}{CPS}{Cyber-Physical System}
-\newabbreviation{NCS}{NCS}{Networked Control System}
+\newabbreviation[prefixfirst={a\ }, prefix={an\ }]{NCS}{NCS}{Networked Control System}
 %\acroindefinite{NCS}{an}{a}
@@
 \newabbreviation{NAT}{NAT}{Network Address Translation}
-\newabbreviation{NTP}{NTP}{Network Time Protocol}
+\newabbreviation[prefixfirst={a\ }, prefix={an\ }]{NTP}{NTP}{Network Time Protocol}
 %\acroindefinite{NTP}{an}{a}
-\newabbreviation{SDR}{SDR}{Software-Defined Radio}
+\newabbreviation[prefixfirst={a\ }, prefix={an\ }]{SDR}{SDR}{Software-Defined Radio}
 %\acroindefinite{SDR}{an}{a}
@@
 \newabbreviation{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
-\newabbreviation{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
+\newabbreviation[prefixfirst={a\ }, prefix={an\ }]{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
 %\acroindefinite{HTTP}{an}{a}
@@
 \newabbreviation{FSM}{FSM}{Finite State Machine}
-%\acroindefinite{NCS}{an}{a}
 \newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}

### 2022-12-19 body/part1/introduction.tex
glossary fixes

@@
 This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
+We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
 
@@
 
-\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \acp{NCS}.
+\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \glspl{NCS}.
 As discussed above, these applications are similarly difficult as \gls{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.

### 2022-12-19 body/part1/overview/methodology.tex
glossary fixes

@@
 The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \gls{WCA}) or due its cyber-physical nature (in \acp{NCS}).
+The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \gls{WCA}) or due its cyber-physical nature (in \glspl{NCS}).
 Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
@@
 In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
-We first discuss the methodology in the context of a human-in-the-loop application, \acl{WCA}, in section \cref{ssec:methodology:wca}.
-Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \acf{NCS}.
+We first discuss the methodology in the context of a human-in-the-loop application, \gls{WCA}, in section \cref{ssec:methodology:wca}.
+Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \glspl{NCS}.
 
-\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
+\subsection{Applying the methodology to \gls{WCA}}\label{ssec:methodology:wca}
 
@@
 
-\subsection{Applying the methodology to \acsp{NCS}}\label{ssec:methodology:ncs}
+\subsection{Applying the methodology to \glspl{NCS}}\label{ssec:methodology:ncs}
 
@@
     \caption{
-        Structure of an emulated \acl*{NCS} in \acs*{CLEAVE}.
+        Structure of an emulated \glspl{NCS} in \gls{CLEAVE}.
     }\label{fig:cleave:ncs:struct}
@@
 ``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \acp{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
+While the core idea of edge computing is widely accepted as fundamental for pervasive \glspl{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
 
@@
 The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \acp{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
+These virtual \glspl{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
 
@@
 Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \glspl{NCS}.
 
@@
         \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\acsp{RTT}.}\label{fig:video:rtt}
+        \caption{\glspl{RTT}.}\label{fig:video:rtt}
     \end{subfigure}%
@@
         Multi-loop, resource-constrained setup results.
-        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
+        Error bars indicate \SI{95}{\percent} \glspl{CI} in all plots.
     }\label{fig:video:results}

### 2022-12-19 publications/2018DemoScalingOnTheEdge/paper_body.tex
glossaries
glossary fixes

--- 2022-09-24 publications/2018DemoScalingOnTheEdge/paper_body.tex
@@
 
-    \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly Deep Neural Networks (DNNs).
+    \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly \glspl{DNN}.
     Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{Flinn2012Cyber}. 
-    While state-of-the-art DNN object detectors can run at more than 20 FPS on a server GPU, their performances are much worse on mobile GPUs - some models cannot even be loaded due to memory constraints.
+    While state-of-the-art DNN object detectors can run at more than 20 \gls{FPS} on a server \gls{GPU}, their performances are much worse on mobile \glspl{GPU} - some models cannot even be loaded due to memory constraints.
     Cloudlet-based applications overcome these challenges by offloading the computation. 
@@
 The tool needs to be extended to work on a much a broader spectrum of applications, in particular those that do not have a clear task model.
-Furthermore, the current implementation does not support hardware accelerators (e.g. GPUs) that are commonly used for DNN inference.
+Furthermore, the current implementation does not support hardware accelerators (e.g. \glspl{GPU}) that are commonly used for \gls{DNN} inference.
 

### 2022-12-19 publications/2022Ainur/abstract.tex
glossary fixes

@@
   Experimental research on wireless networking in combination with edge and cloud computing has been the subject of explosive interest in the last decade.
-  This development has been driven by the increasing complexity of modern wireless technologies and the  extensive softwarization of these through projects such as a \acf*{O-RAN}.
+  This development has been driven by the increasing complexity of modern wireless technologies and the  extensive softwarization of these through projects such as a \gls{O-RAN}.
   In this context, a number of small- to mid-scale testbeds have emerged, employing a variety of technologies to target a wide array of use-cases and scenarios in the context of novel mobile communication technologies such as 5G and beyond-5G.
@@
   It is designed to configure, manage, orchestrate, and deploy workloads from an end-to-end perspective.
-  Ainur is built on top of cloud-native technologies such as Docker, and is provided as \acs*{FOSS} to the community through the {KTH-EXPECA/Ainur} repository on GitHub.
+  Ainur is built on top of cloud-native technologies such as Docker, and is provided as \gls{FOSS} to the community through the {KTH-EXPECA/Ainur} repository on GitHub.
   We demonstrate the utility of the platform with a series of scenarios, showcasing in particular its flexibility with respect to physical link definition, computation placement, and automation of arbitrarily complex experimental scenarios.

### 2022-12-19 publications/2022Ainur/ainur.tex
glossary fixes

@@
     \item[\gls{OAI}:]
-    \gls{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \acp{SDR} like the \gls{USRP}~\cite{KALTENBERGER2020107284}.
+    \gls{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \glspl{SDR} like the \gls{USRP}~\cite{KALTENBERGER2020107284}.
     Ainur can deploy, configure, and manage \gls{OAI} software components that implement 4G \gls{LTE} and 5G \gls{NR}.
@@
     \item[Mango Communications 802.11:]
-    Finally, the Mango Communications project implements real-time 802.11 (WiFi) MAC and PHY in Xilinx \acp{FPGA}.
-    It can be used on a variety of hardware platforms including \gls{USRP} \acp{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
+    Finally, the Mango Communications project implements real-time 802.11 (WiFi) MAC and PHY in Xilinx \glspl{FPGA}.
+    It can be used on a variety of hardware platforms including \gls{USRP} \glspl{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
 \end{description}
@@
 
-Ainur follows a layered architecture which closely mimics the conceptual layers of the \acs{TCP}/\acs{IP} stack.
+Ainur follows a layered architecture which closely mimics the conceptual layers of the \gls{TCP}/\gls{IP} stack.
 Components are deployed in bottom-up order, starting with the establishment of physical links, through the establishment of \gls{IP} connectivity and deployment of links to the cloud, and ending with the distribution and initialization of workloads on top of a container orchestration layer.
@@
     Creates and deploys the underlying physical connections of the workload data network.
-    This layer interacts with hardware such as managed switches and \acp{SDR} (and their associated computation hosts) to create links between the desired workload hosts.
+    This layer interacts with hardware such as managed switches and \glspl{SDR} (and their associated computation hosts) to create links between the desired workload hosts.
     These links correspond to ethernet, WiFi, and even an entire software-defined 4G \gls{LTE} and 5G networks.
@@
 As expressed above, Ainur also leverages containers for the communication stack.
-With the advent of \gls{O-RAN} and \acp{SDR}, all components of the wireless network stack can run on general-purpose processors.
+With the advent of \gls{O-RAN} and \glspl{SDR}, all components of the wireless network stack can run on general-purpose processors.
 They can thus be containerized and distributed across multiple hosts.
@@
 
-The framework is released as \acl{FOSS} under a permissive Apache license.
+The framework is released as \gls{FOSS} under a permissive Apache license.
 It can be downloaded from the Ainur repository~\cite{ainur:github} of the {KTH-EXPECA} organization on GitHub.

### 2022-12-19 publications/2022Ainur/assumptions.tex
glossary fixes

@@
     \item be able to dynamically reconfigure, segment, and route the workload data network for different experiments
-    \item ensure consistent and accurate network traffic measurements (\acp{RTT}, bandwidth, etc.) on the workload data network, uncontaminated by control traffic.
+    \item ensure consistent and accurate network traffic measurements (\glspl{RTT}, bandwidth, etc.) on the workload data network, uncontaminated by control traffic.
 \end{enumerate*}
@@
     Devices on the workload data network are interconnected through a managed switch.
-    This switch is used by Ainur to change the virtual topology of the workload data network at runtime using \acp{VLAN}, for instance separating the network into front-end and back-end networks.
+    This switch is used by Ainur to change the virtual topology of the workload data network at runtime using \glspl{VLAN}, for instance separating the network into front-end and back-end networks.
 
@@
 
-    \item[\acp{AMI}]
+    \item[\glspl{AMI}]
 

### 2022-12-19 publications/2022Ainur/demo.tex
glossary fixes

@@
         \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_1.png}
-        \caption{Workload 1: \acs*{MNIST} image classifier}\label{fig:wkld:mnist}
+        \caption{Workload 1: \gls{MNIST} image classifier}\label{fig:wkld:mnist}
     \end{subfigure}%
@@
         \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_2.png}
-        \caption{Workload 2: inverted pendulum \acs*{NCS}}\label{fig:wkld:ncs}
+        \caption{Workload 2: inverted pendulum \gls{NCS}}\label{fig:wkld:ncs}
     \end{subfigure}
@@
 \begin{description}
-    \item[WiFi:] \iac{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board WiFi.
-    \item[4G \gls{LTE}:] \iac{SDR} is configured as an 4G \gls{LTE} base station, and another is configured as an 4G \gls{LTE} \gls{UE}.
+    \item[WiFi:] \pgls{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board WiFi.
+    \item[4G \gls{LTE}:] \pgls{SDR} is configured as an 4G \gls{LTE} base station, and another is configured as an 4G \gls{LTE} \gls{UE}.
     Together these radios act as an \gls{LTE} bridge between the client-side and the server-side of the network, and all client-server traffic is routed through them.
@@
 The first of these (\cref{fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{mnist}.
-The application consists of a client with a web interface, and \iac{HTTP} server which responds to requests from the client and performs the actual image recognition.
+The application consists of a client with a web interface, and \pgls{HTTP} server which responds to requests from the client and performs the actual image recognition.
 This workload is intended to showcase in an interactive manner the effects of placing computation at different points of the network, and how Ainur simplifies these deployments.
 
-The second workload (\cref{fig:wkld:ncs}) corresponds to \iac{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \acp{NCS} using cloud-native technologies~\cite{cleave}.
+The second workload (\cref{fig:wkld:ncs}) corresponds to \pgls{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \glspl{NCS} using cloud-native technologies~\cite{cleave}.
 It consists of an emulation of the physical inverted pendulum system plant and a software-implemented proportional-differential controller.
@@
 \end{enumerate*}
-These nodes are interconnected using a combination of managed switches, \acp{SDR}, and \gls{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.
+These nodes are interconnected using a combination of managed switches, \glspl{SDR}, and \gls{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.
 
@@
 Participants will be asked to specify the full, arbitrarily complex, experimental scenario, including the number of clients, physical layers (single or multiple, and which clients are on each physical link), and offloading configuration (number of instances on the edge vs.\ on the cloud, which \gls{AWS} datacenters to deploy to).
-Configuration will be specified through a \acs{YAML} file, which will then be parsed by Ainur for the automatic execution of the scenario.
+Configuration will be specified through a \gls{YAML} file, which will then be parsed by Ainur for the automatic execution of the scenario.

### 2022-12-19 publications/2022Ainur/futurework.tex
glossary fixes

@@
     In other words, we would have several smaller server applications running on potentially different hosts on the network, each handling a specific part of the functionality we need.
-    For instance, we could have a ``Layer 3'' service which handles setting up \gls{IP}-layer connectivity, and another ``SDR'' service which manages the \acp{SDR}.
-    Services could communicate with each other using simple HTTP \acp{API} for easy debugging, and we would have an ``Ainur'' client which orchestrates the services into an experimental run.
+    For instance, we could have a ``Layer 3'' service which handles setting up \gls{IP}-layer connectivity, and another ``SDR'' service which manages the \glspl{SDR}.
+    Services could communicate with each other using simple HTTP \glspl{API} for easy debugging, and we would have an ``Ainur'' client which orchestrates the services into an experimental run.
 

### 2022-12-19 publications/2022Ainur/introduction.tex
glossary fixes

@@
 \begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
-    \item \acs*{COSMOS}
-    \item \acs*{POWDER}
-    \item \acs*{ARA}
+    \item \gls{COSMOS}
+    \item \gls{POWDER}
+    \item \gls{ARA}
     \item Drexel Grid \gls{SDR}
 \end{enumerate*} testbeds.
-\acs{COSMOS}~(\emph{\acl{COSMOS}})\acused{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \acp{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
+\gls{COSMOS}~(\emph{\gls{COSMOS}}) is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
 It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
@@
 
-\acs{POWDER}~(\emph{\acl{POWDER}})\acused{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
-The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
+\gls{POWDER}~(\emph{\gls{POWDER}}) promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
+The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
 \gls{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \gls{VM} images containing the necessary software and configurations.
@@
 Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
-It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
 \gls{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \gls{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
@@
 
-Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
+Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
 Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
@@
 This is, for instance, the case with \gls{COSMOS} and \gls{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
-These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.
+These testbeds further leverage virtualization technology based on \glspl{VM} instead of more lightweight and edge-compatible solutions such as containers.
 
@@
 CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \acp{VM} containing the necessary software for experiments.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \glspl{VM} containing the necessary software for experiments.
 Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
-Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \iac{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.
+Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \pgls{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.
 
@@
 % \todo[inline]{
-%     COSMOS: Experimenters work on these resources using the \ac*{OMF}~\cite{omf} tool, originally leveraged for ORBIT~\cite{orbit}, and \gls{SSH} sessions, thus allowing for free-form scripting of experimental procedures.
+%     COSMOS: Experimenters work on these resources using the \gls{OMF}~\cite{omf} tool, originally leveraged for ORBIT~\cite{orbit}, and \gls{SSH} sessions, thus allowing for free-form scripting of experimental procedures.
 
@@
 
-%     POWDER: To abstract and redistribute resources and \acp{API}, \gls{POWDER} leverages the concept of \emph{profiles}, corresponding shareable packages containing experimental configurations and descriptions\cite{emulab}.
+%     POWDER: To abstract and redistribute resources and \glspl{API}, \gls{POWDER} leverages the concept of \emph{profiles}, corresponding shareable packages containing experimental configurations and descriptions\cite{emulab}.
 %     Automation and web interface control are limited to some test experiments, and thus regular users must interact with the platform through \gls{SSH}.

### 2022-12-19 publications/2022Ainur/software_layers/cloud.tex
glossary fixes

@@
 The spawned \gls{EC2} instances are created from a preconfigured \gls{AMI}, the ID of which must be passed to the \mintinline{python}{CloudInstances} instance at runtime.
-This \gls{AMI} can be either be one of the default \acp{AMI} provided by \acp{AWS} --- e.g.\ a plain Ubuntu or CentOS \gls{AMI} --- or a custom \gls{AMI} which has been preconfigured by the user.
+This \gls{AMI} can be either be one of the default \glspl{AMI} provided by \glspl{AWS} --- e.g.\ a plain Ubuntu or CentOS \gls{AMI} --- or a custom \gls{AMI} which has been preconfigured by the user.
 Note that although the \mintinline{python}{CloudInstances} instance places no restrictions on the \gls{AMI} used for deploying (with the exception that it needs to be a Linux-based \gls{AMI}), components such as the VPN layer (see \cref{sec:vpn}) depending on these cloud instances might have requirements of their own.

### 2022-12-19 publications/2022Ainur/software_layers/physical.tex
glossary fixes

@@
 Therefore, we use a managed switch at the heart of the physical layer and connect all the hosts and radios to it.
-According to the data network description, the physical layer then creates the required \acp{VLAN} in the switch.
+According to the data network description, the physical layer then creates the required \glspl{VLAN} in the switch.
 
-At instantiation time, this class first configures ethernet connections by creating \acp{VLAN} in the workload data network switch.
+At instantiation time, this class first configures ethernet connections by creating \glspl{VLAN} in the workload data network switch.
 It then triggers the deployment of potential wireless connections using \mintinline{python}{WirelessConnection} objects.
@@
 A set of \mintinline{python}{WirelessConnection} instances must be provided to the constructor for deployment.
-This class can be used as a context manager in a \mintinline{python}{with} statement; at the end of the block, it tears down all connections and cleans the added \acp{VLAN} from the switch.
+This class can be used as a context manager in a \mintinline{python}{with} statement; at the end of the block, it tears down all connections and cleans the added \glspl{VLAN} from the switch.
 
@@
     The \mintinline{python}{SDRWifi} class handles the configuration and deployment of a software-defined IEEE 802.11 network using the Mangocomm software package and its supported software-defined radios \footnote{See https://mangocomm.com/802.11-mac-phy/}.
-    It lunches a Docker container on the same machine to communicate with the \acp{SDR} over the control network.
+    It lunches a Docker container on the same machine to communicate with the \glspl{SDR} over the control network.
     This communication is for the configuration and monitoring the WiFi nodes.
@@
         \includegraphics[width=0.8\linewidth]{figures/wifi.png}
-        \caption{An example of WiFi network topology in data network and its corresponding \acp{VLAN}}
+        \caption{An example of WiFi network topology in data network and its corresponding \glspl{VLAN}}
         \label{fig:wifi}
@@
         \includegraphics[width=0.8\linewidth]{figures/lte.png}
-        \caption{An example of \gls{LTE} data network topology in data network and its corresponding \acp{VLAN}}
+        \caption{An example of \gls{LTE} data network topology in data network and its corresponding \glspl{VLAN}}
         \label{fig:lte}

### 2022-12-19 publications/2022CLEAVE/abstract.tex
glossary fixes

\begin{abstract}
    As the number of cyber-physical systems rises, it becomes increasingly crucial to study \glspl{NCS} combining control communication co-design. This nature of \glspl{NCS} has led to task-specific approaches to research, creating a dearth of generalizable, repeatable, and scalable experimentation. Further, with the advent of edge computing solutions, it is of paramount importance to explore its relevance in such applications. In this work, we present \gls{CLEAVE}, a novel, completely software-based framework for repeatable and scalable experimentation in edge native \glspl{NCS}.
    Our approach is based on the emulation of physical plants communicating over a real network with software-based controllers.
    \gls{CLEAVE} is designed and built for the edge, using Python3 and with full compatibility with industry-standard containerization solutions.
    Although designed for single-loop emulations, the flexibility afforded by the aforementioned characteristics allow our framework to be adapted to a multitude of complex scenarios.

    We validate \gls{CLEAVE} using an initial implementation of an inverted pendulum \gls{NCS}.
    Our results showcase the utility of the tool as a repeatable, extensible, and scalable solution to \gls{NCS} performance evaluation and benchmarking on the Edge.
\end{abstract}

### 2022-12-19 publications/2022CLEAVE/cleave.tex
glossary fixes

@@
 
-In this section, we detail our \gls{CLEAVE} framework for the performance evaluation of \aclp{NCS}, with an emphasis on edge deployments.
+In this section, we detail our \gls{CLEAVE} framework for the performance evaluation of \glspl{NCS}, with an emphasis on edge deployments.
 Our design follows a virtualized emulation approach; providing a framework for the real-time \emph{emulation} of physical control systems and the easy implementation of softwarized controllers interacting with \emph{real} networks.
 \gls{CLEAVE} is thus a platform with an easy-to-use \gls{API} on which a multitude of \gls{NCS} types can be emulated.
-This allows for great flexibility, as the core components of the \gls{NCS} can be switched out while maintaining the realism of the network, the most complex and limiting component of edge \acp{NCS}.
+This allows for great flexibility, as the core components of the \gls{NCS} can be switched out while maintaining the realism of the network, the most complex and limiting component of edge \glspl{NCS}.
 
@@
 Implementation-wise, \gls{CLEAVE} is built on top of the well-established and stable \emph{Twisted}\footnote{\url{https://twistedmatrix.com}} asynchronous programming and networking framework.
-The aforementioned core components (\emph{State}, \emph{Controller}, \emph{Sensors}, and \emph{Actuators}) are provided as \acp{ABC} with core functionality that users must extend for their specific use cases and \acp{NCS}.
+The aforementioned core components (\emph{State}, \emph{Controller}, \emph{Sensors}, and \emph{Actuators}) are provided as \glspl{ABC} with core functionality that users must extend for their specific use cases and \glspl{NCS}.
 The framework makes no assumptions about the inner workings of the user implementations of these classes other than the interfaces they expose.
@@
 
-\subsection{Configuring and Executing \iac{NCS}}
+\subsection{Configuring and Executing \pgls{NCS}}
 
@@
 Although the core design of \gls{CLEAVE} follows a single-loop approach, multi-loop scenarios can be easily implemented through the combination of multiple \gls{CLEAVE} instances.
-The framework does not make any assumptions about the individual components other than the signatures of methods on the \acp{ABC} that must be extended and implemented.
+The framework does not make any assumptions about the individual components other than the signatures of methods on the \glspl{ABC} that must be extended and implemented.
 Combined with the Python emulation configuration scripts, this makes development of more complex scenarios merely a matter of implementing the desired functionality within the framework itself.

### 2022-12-19 publications/2022CLEAVE/conclusion.tex
glossary fixes

@@
 Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \acp{NCS}.
+We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \glspl{NCS}.
 
 There is still, however, work to be done.
-We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \acp{NCS} to share with the community.
+We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \glspl{NCS} to share with the community.
 At the moment, the interactions of \gls{CLEAVE} and tools such as Docker are still quite superficial.

### 2022-12-19 publications/2022CLEAVE/experiments.tex
glossary fixes

@@
 In this section, we demonstrate the utility of \gls{CLEAVE} by walking readers through an example use case.
-With this, we aim to showcase the ability of our framework to provide accurate and repeatable measurements of the performance of \acp{NCS} deployed on edge computing infrastructure.
+With this, we aim to showcase the ability of our framework to provide accurate and repeatable measurements of the performance of \glspl{NCS} deployed on edge computing infrastructure.
 
@@
 Secondly, similar, albeit non-Edge-enabled, setups have been a reality in industry for almost two decades, as evidenced by\ \cite{Gupta2010NCSOverview}.
-Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \acp{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
-4G, for instance, has average \acp{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \gls{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
-Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \acp{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
+Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \glspl{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
+4G, for instance, has average \glspl{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \gls{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
+Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \glspl{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
 Finally, video analytics is one of the main proposed use cases for edge computing~\cite{Ananthanarayanan2017Analytics,Yi2017Analytics,Wang2018Analytics}, and thus we foresee edge \gls{NCS} deployments being deployed in parallel with such applications in the future.
@@
         \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_rms}
-        \caption{Mean angle \acs*{RMS}; red line indicates \( y = 180 \).}\label{fig:single:rms}
+        \caption{Mean angle \gls{RMS}; red line indicates \( y = 180 \).}\label{fig:single:rms}
     \end{subfigure}\\
@@
         \caption{
-            Mean \acsp*{RTT}.
+            Mean \glspl{RTT}.
         }\label{fig:single:rtt}
@@
         Baseline results.
-        Error bars indicate \SI{95}{\percent} \acp{CI}.
+        Error bars indicate \SI{95}{\percent} \glspl{CI}.
         }%
@@
 
-As expected, higher sampling rates tend to correlate with better quality of control; at higher sampling rates the system was able to reach stability at higher \acp{RTT}.
+As expected, higher sampling rates tend to correlate with better quality of control; at higher sampling rates the system was able to reach stability at higher \glspl{RTT}.
 These initial results already hint at interesting consequences for such an edge-bound \gls{NCS} deployment.
 For instance, it is clear from \cref{fig:single:topple,fig:single:rms} that network delays can, to a certain extent, be compensated for by increasing the sampling rate of the system.
-A corollary of this is, conversely, that at lower network latencies \acp{NCS} are able to stabilize at lower sampling rates.
+A corollary of this is, conversely, that at lower network latencies \glspl{NCS} are able to stabilize at lower sampling rates.
 Adaptive sampling might thus be a viable method for optimizing resource utilization.
 
-Once baselines for single loops in the use case scenario have been established, we can proceed to studying the interaction between the \acp{NCS} and the video-streaming applications.
+Once baselines for single loops in the use case scenario have been established, we can proceed to studying the interaction between the \glspl{NCS} and the video-streaming applications.
 We deploy \num{6} control loops on the experimental setup depicted in \cref{fig:cleave:expsetup}.
@@
 Whereas both the \SIlist{20;40}{\hertz} setups show losses well below \SI{5}{\percent}, the \SI{60}{\hertz} scenario shows an average of around \SI{13}{\percent} of datagrams lost.
-The differences in \acp{RTT} results are equally telling; \acp{RTT} for the \SI{60}{\hertz} setup were on average approx.\ \num{3} times those for the \SIlist{20;40}{\hertz} setups.
-These results stem from the contention for network resources, and hint at important trade-offs system designers will have to take into consideration when designing and developing \acp{NCS} for deployment on the Edge.
+The differences in \glspl{RTT} results are equally telling; \glspl{RTT} for the \SI{60}{\hertz} setup were on average approx.\ \num{3} times those for the \SIlist{20;40}{\hertz} setups.
+These results stem from the contention for network resources, and hint at important trade-offs system designers will have to take into consideration when designing and developing \glspl{NCS} for deployment on the Edge.
 The Edge will be \emph{multi-tenant} and \emph{multi-instance}. 

### 2022-12-19 publications/2022CLEAVE/intro.tex
glossary fixes

@@
 
-The number and applications of \acp{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
+The number and applications of \glspl{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
 However, this rapid increase in adoption has mostly been limited to industrial contexts.
-Although \acp{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
-This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
-These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
+Although \glspl{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
+This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \glspl{CPS} will be made possible.
+These technologies meet two key requirements of \glspl{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
 
-\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
+\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
 This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
 
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \acp{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
+%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
 
-One the one hand, related literature in \acp{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level. 
+One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
 %follows a theoretical approach, and only a small fraction of it deals with experimental studies.
 On the other, there exist several approaches when considering experimental methodologies. 
-%A number of works concerning \acp{NCS} deal with experimental studies.
-%\acp{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
+%A number of works concerning \glspl{NCS} deal with experimental studies.
+%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
 One such approach uses setups in which the complete system is built on top of real hardware.
@@
 The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \acp{NCS}.
+These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
 Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
@@
 
-As evidenced above, experimental research in \acp{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
+As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
 This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.

### 2022-12-19 appendices.tex
glossary fixes

@@
 % Note that throughout the paper we use a exGaussian distribution but not Rayleigh. This is because, 
-Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \acl{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \gls{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
 This is the reason why the above results are applicable in this work where we have predominantly considered \gls{exGaussian} distribution.

### 2022-12-19 background.tex
glossary fixes

@@
-\section{Background of \acs*{WCA}}\label{sec:background}
+\section{Background of \gls{WCA}}\label{sec:background}
 
@@
     \end{subfigure}
-    \caption{Key concepts in \acl{WCA}}
+    \caption{Key concepts in \gls{WCA}}
 \end{figure}
 
-\acf{WCA} applications represent a category of novel, context-sensitive and highly-interactive \gls{AR} applications.
+\glspl{WCA} applications represent a category of novel, context-sensitive and highly-interactive \gls{AR} applications.
 In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
@@
 
-Step-based \acp{WCA} operate analogously to how \gls{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
+Step-based \glspl{WCA} operate analogously to how \gls{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
 The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
@@
 
-\acp{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
+\glspl{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
 Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \gls{WCA} captures the state of the physical system, as illustrated in \cref{fig:step}.
@@
 To compensate for the distribution of \( t_c \)\footnote{\( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality.}~\cite{olguinmunoz:impact2021}, in the present work we adjust the nominal delay \( D \) by a factor \num{1.5}.
-We will refer to this adjusted delay as \emph{\acf{TTF}}.
+We will refer to this adjusted delay as \emph{\glspl{TTF}}.
 
@@
 
-    \item\label{item:speedup} At lower \acp{TTF}, humans get faster at performing steps as the task progresses, 
+    \item\label{item:speedup} At lower \glspl{TTF}, humans get faster at performing steps as the task progresses,
     For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.
@@
 
-We recorded variables related to well-known individual differences, encompassing both the \acf{BFI}~\cite{oliver:bfi1999}, and the \acf{ITQ}~\cite{witmer1998measuring}.
+We recorded variables related to well-known individual differences, encompassing both the \glspl{BFI}~\cite{oliver:bfi1999}, and the \glspl{ITQ}~\cite{witmer1998measuring}.
 Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \gls{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
@@
 Out of the three identified components, which cumulatively accounted for \SI{73.13}{\percent} of the variance in the results, neuroticism was included in the first two.
-The effect of neuroticism was observed across all \acp{TTF} and impairment durations in the tasks.
+The effect of neuroticism was observed across all \glspl{TTF} and impairment durations in the tasks.
 
@@
     \caption{%
-        \( \mu \) parameter of \acs*{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \gls{TTF} in~\cite{olguinmunoz:impact2021}.
+        \( \mu \) parameter of \gls{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \gls{TTF} in~\cite{olguinmunoz:impact2021}.
         Distributions were fitted using \gls{MLE}.

### 2022-12-19 conclusion.tex
glossary fixes

@@
 In terms of the optimization framework, our current model adopts the assumption that execution times are Rayleigh-distributed.
-Although this distribution fits the data reasonably well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \acl{exGaussian}).
+Although this distribution fits the data reasonably well, there are other which more accurately describe the behavior of human execution times (e.g.\ the \gls{exGaussian}).
 Our future efforts consider developing this framework towards these more accurate distributions.

### 2022-12-19 implications_footprint.tex
glossary fixes

@@
 
-We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \acp{TTF}.
+We start by studying application lifetimes in a controlled, ideal setup by using the timing models to generate execution times for sequences of \num{100} steps subject to constant \glspl{TTF}.
 These runs are completely simulated and no sampling of video frames is performed; for each step, we simply feed the models a predefined \gls{TTF} and record the generated execution time.
-We use the combination of \acp{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
-This is done for \num{25} linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \gls{TTF}.
+We use the combination of \glspl{TTF} and execution times to calculate theoretical step duration times and subsequent total application lifetimes.
+This is done for \num{25} linearly distributed \glspl{TTF} in the \SIrange[]{0}{5}{\second} range; \num{45} independent repetitions for each combination of model configuration and \gls{TTF}.
 
@@
         \caption{%
-            Evolution of mean application lifetimes as \acp{TTF} increase.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
+            Evolution of mean application lifetimes as \glspl{TTF} increase.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}.
         }
@@
         \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model at select \acp{TTF}.
-            Error bars indicate the \SI{95}{\percent} \acp{CI}, calculated using a two-sided T-test.
+            Percentage difference in mean application lifetimes with respect to the reference model at select \glspl{TTF}.
+            Error bars indicate the \SI{95}{\percent} \glspl{CI}, calculated using a two-sided T-test.
         }
     \end{subfigure}
-    \caption{\acl{TTF} versus application lifetime.}\label{fig:lifetimes}
+    \caption{\gls{TTF} versus application lifetime.}\label{fig:lifetimes}
 \end{figure}
@@
 The results of this investigation are presented in \cref{fig:lifetimes}.
-Compared to the reference model, our realistic model is, on average, roughly \SI{11}{\percent} faster when subject to low \acp{TTF}.
-At higher \acp{TTF}, the behavior of the model depends on its level of normalized neuroticism.
+Compared to the reference model, our realistic model is, on average, roughly \SI{11}{\percent} faster when subject to low \glspl{TTF}.
+At higher \glspl{TTF}, the behavior of the model depends on its level of normalized neuroticism.
 At low neuroticism, the behavior of the realistic model results in total task durations that are basically indistinguishable from the reference model.
@@
         Experimental setup used to study the implications of the realistic models of human behavior for \gls{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \acs*{COTS} \acs{IEEE} \num{802.11}b/g access point.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \gls{COTS} \gls{IEEE} \num{802.11}b/g access point.
     }\label{fig:expsetup}
@@
 For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \gls{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
-Clients and cloudlet communicate over an \acs{IEEE} \num{802.11}b/g wireless network.
+Clients and cloudlet communicate over an \gls{IEEE} \num{802.11}b/g wireless network.
 Our choice of wireless standard is simply motivated by a desire to amplify the potential effects of network congestion.
@@
             Note that due to the low number of samples, means have been calculated using the geometric instead of arithmetic average. 
-            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using bootstrapping.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}, calculated using bootstrapping.
         }
@@
         \end{tabular}
-        \caption{Mean measured \aclp{RTT} for each testbed configuration.}
+        \caption{Mean measured \glspl{RTT} for each testbed configuration.}
     \end{subtable}
@@
 With just a single client and mean \gls{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
-At \num{10} clients, the results mimic those at higher \acp{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
+At \num{10} clients, the results mimic those at higher \glspl{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
 

### 2022-12-19 implications_optimization.tex
glossary fixes

@@
         Repetitions & \num{100} & \\
-        \acp{RTT} & \( \left\{ 0.3, 0.6,\ldots,4.2 \right\} \) & \\
+        \glspl{RTT} & \( \left\{ 0.3, 0.6,\ldots,4.2 \right\} \) & \\
         \( \tau_\text{p} \) & \SI{250}{\milli\second} & Processing delay \\
@@
 
-We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks with varying constant \acp{RTT}.
+We proceed to set up an experiment where these sampling approaches are deployed on identical, simulated, tasks with varying constant \glspl{RTT}.
 Experimental parameters are summarized in \cref{tab:params}.
@@
             Note the logarithmic scale on the vertical axis.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}.
         }
@@
             In other words, curves represent the relative performance of the sample-count-optimized scheme when using the corresponding reference scheme as baseline.
-            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}, calculated using a two-sided t-test.
         }
@@
 The results of this investigation are presented in \cref{fig:optimization:samples}, and clearly show the advantages of using the sample-count-optimized scheme over the current state-of-the-art.
-The performance of \textcite{Wang2019Towards}'s approach appears to degrade with lower \acp{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
-On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \acp{RTT}.
+The performance of \textcite{Wang2019Towards}'s approach appears to degrade with lower \glspl{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
+On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \glspl{RTT}.
 As mentioned above, the future feasibility and mass adoption of \gls{WCA} depends on these applications not hogging the available resources.
@@
 
-It should be noted that although both schemes seem to tend towards two samples per step as \acp{RTT} increase, this is simply an artifact of our experimental setup.
-As \acp{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \gls{RTT} being larger than the execution time of the step tends towards \num{1.0}.
+It should be noted that although both schemes seem to tend towards two samples per step as \glspl{RTT} increase, this is simply an artifact of our experimental setup.
+As \glspl{RTT} increase above the expected wait time, the probability of the sum of the first sampling interval and the \gls{RTT} being larger than the execution time of the step tends towards \num{1.0}.
 This leads to these sampling schemes consistently sampling only twice each step: a first sample which is taken before the execution time of the step, and a second one \gls{RTT} seconds later, after the execution time has been reached.
@@
             Round-trip time versus mean per step energy consumption, averaged over all \num{100} repetitions of the experiment.
-            Error bars indicate \SI{95}{\percent} \acp{CI}.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}.
         }
@@
             Percentage difference in mean per step energy consumption by the energy-optimized sampling scheme with respect to the three reference schemes.
-            Error bars indicate \SI{95}{\percent} \acp{CI}, calculated using a two-sided t-test.
+            Error bars indicate \SI{95}{\percent} \glspl{CI}, calculated using a two-sided t-test.
         }

### 2022-12-19 intro.tex
glossary fixes

@@
 
-\acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
+\glspl{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
 These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
@@
 This paper is structured as follows.
-\cref{sec:relwork} discusses related work in the field of \acl{WCA} modeling.
+\cref{sec:relwork} discusses related work in the field of \gls{WCA} modeling.
 In \cref{sec:background} we define and discuss key concepts in \gls{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.

### 2022-12-19 model.tex
glossary fixes

@@
 
-We begin by calculating rolling weighted averages of the \acp{TTF} of each of the \num{40} individual repetitions of the data.
+We begin by calculating rolling weighted averages of the \glspl{TTF} of each of the \num{40} individual repetitions of the data.
 We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \gls{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \textasciitilde\SI{25}{\percent}, the third-most \textasciitilde\SI{12}{\percent}, and so on.
@@
 
-The resulting weighted \acp{TTF} are then binned into continuous ranges by splitting the data on the \num{7}-quantiles.
+The resulting weighted \glspl{TTF} are then binned into continuous ranges by splitting the data on the \num{7}-quantiles.
 The exact resulting bins of this operation on the data are presented in \cref{tab:defaultbins}.
@@
 
-After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \acp{TTF}:
+After this preprocessing has been finished, the data is ready to be used for the generation of execution times by applying the above steps in real-time to measured \glspl{TTF}:
 
@@
 \begin{enumerate}
-    \item\label{it:ttftoexectime} Higher \acp{TTF} should result in higher execution times.
+    \item\label{it:ttftoexectime} Higher \glspl{TTF} should result in higher execution times.
     \item\label{it:duration} When subject to a series of steps at the same level of impairment, desired behavior depends on the level:
     \begin{enumerate}
-        \item At low \acp{TTF}, the model should speed up; i.e.\ execution times should decrease.
-        \item At medium \acp{TTF}, execution times should remain more or less the same.
-        \item At high \acp{TTF}, execution times should increase.
+        \item At low \glspl{TTF}, the model should speed up; i.e.\ execution times should decrease.
+        \item At medium \glspl{TTF}, execution times should remain more or less the same.
+        \item At high \glspl{TTF}, execution times should increase.
     \end{enumerate}
@@
     \caption{%
-        Effects of feeding three different \acp{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}) into the model on the generated execution times.
-        Higher \acp{TTF} directly lead to higher execution times.
+        Effects of feeding three different \glspl{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}) into the model on the generated execution times.
+        Higher \glspl{TTF} directly lead to higher execution times.
         Error bars indicate the \SI{95}{\percent} \gls{CI}.
@@
 \cref{fig:ttf_to_exectime} shows the mean execution time outputted by the model when fed three different levels of \gls{TTF} (\emph{low}, \SI{0}{\second}, \emph{medium}, \SI{2.5}{\second}, or \emph{high}, \SI{5}{\second}).
-These results were generated by first warming up the model by feeding it \num{25} \acp{TTF} selected at random from the data before feeding it the desired input \gls{TTF}, and recording the generated execution time.
+These results were generated by first warming up the model by feeding it \num{25} \glspl{TTF} selected at random from the data before feeding it the desired input \gls{TTF}, and recording the generated execution time.
 This procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
-The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \acp{TTF} reaching \SI{14}{\percent} (\textasciitilde\SI{5.6}{\second} to \textasciitilde\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
+The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \glspl{TTF} reaching \SI{14}{\percent} (\textasciitilde\SI{5.6}{\second} to \textasciitilde\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
 Additionally, we can observe the effects of neuroticism on generated execution times, as specified in \cref{it:neuro}.
-At low neuroticism, the average difference between execution times at low versus high \acp{TTF} was of roughly \SI{6.2}{\percent}, compared to \SI{12.5}{\percent} at high neuroticism.
+At low neuroticism, the average difference between execution times at low versus high \glspl{TTF} was of roughly \SI{6.2}{\percent}, compared to \SI{12.5}{\percent} at high neuroticism.
 
@@
 Next, \cref{fig:exectimeduration} shows the evolution of generated execution times while the model is subject to a fixed \gls{TTF}, either \emph{low} (\SI{0}{\second}) or \emph{high} (\SI{5}{\second}).
-These results were generated by first warming up the model with \num{25} random \acp{TTF}, and then recording the generated execution times over a sequence of \num{12} steps at a fixed \gls{TTF}; this procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
+These results were generated by first warming up the model with \num{25} random \glspl{TTF}, and then recording the generated execution times over a sequence of \num{12} steps at a fixed \gls{TTF}; this procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
 Once again, we see here behavior matching what is expected of the model, in particular with respect to \cref{it:duration}.
-At low \acp{TTF}, the model is on average, across all configurations, \SI{8.2}{\percent} faster at step \num{12} when compared to step \num{1}.
-At high \acp{TTF}, the behavior changes depending on the level of neuroticism of the model.
+At low \glspl{TTF}, the model is on average, across all configurations, \SI{8.2}{\percent} faster at step \num{12} when compared to step \num{1}.
+At high \glspl{TTF}, the behavior changes depending on the level of neuroticism of the model.
 Low neuroticism models basically do not change their execution times, whereas high neuroticism configurations are on average \SI{10}{\percent} slower after \num{12} steps.
@@
 As before, we run \num{600} repetitions of this procedure for each combination of model configuration, origin \gls{TTF}, and destination \gls{TTF}.
-Execution times generated immediately after a transition from a higher \gls{TTF} into a lower one are consistently higher than execution times generated without a preceding change in \acp{TTF}.
+Execution times generated immediately after a transition from a higher \gls{TTF} into a lower one are consistently higher than execution times generated without a preceding change in \glspl{TTF}.
 Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \gls{TTF} into a higher one.

### 2022-12-19 old_model.tex
glossary fixes

@@
 For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-The data is pre-processed to group execution time values by level of neuroticism, previous \acp{TTF}, and duration of impairment.
+The data is pre-processed to group execution time values by level of neuroticism, previous \glspl{TTF}, and duration of impairment.
 The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
@@
         Comparison of the total duration of \num{100}-step tasks for different models in the controlled \gls{TTF} setup.
-        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \acp{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
+        The table shows a detailed breakdown of durations for the fitted-naive, empirical, and theoretical models at the lowest and highest \glspl{TTF}, as well as relative differences with respect to the corresponding fitted-naive baseline for each case.
     }\label{fig:ctrlttf}
@@
 
-We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \acp{TTF} in the \SIrange[]{0}{5}{\second} range.
+We first compare these models using a controlled, ideal setup with \num{25} constant, linearly distributed \glspl{TTF} in the \SIrange[]{0}{5}{\second} range.
 For each \gls{TTF}, the models execute \num{30} independent repetitions of a \num{100} step task.
@@
 These behaviors clearly stem from the effects identified in \textcite{olguinmunoz:impact2021}, described in \cref{ssec:plos}.
-At low levels of impairments, i.e.\ small \acp{TTF}, the speed-up effect dominates, bringing execution times to a minimum and consequently resulting in shorter total durations.
+At low levels of impairments, i.e.\ small \glspl{TTF}, the speed-up effect dominates, bringing execution times to a minimum and consequently resulting in shorter total durations.
 On the other hand, at high levels of impairment, the slow-down (pacing) effect starts to take over, pushing step execution times and total durations up.
@@
         Experimental setup used to study the implications of the realistic models of human behavior for \gls{WCA}.
-        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \gls{COTS} \acs{IEEE} \num{802.11}n access point.
+        We deploy containerized instances of the client-server loop running the models on a testbed consisting of \num{10} Raspberry Pi clients connected to a cloudlet over a \gls{COTS} \gls{IEEE} \num{802.11}n access point.
     }\label{fig:expsetup}
@@
             At lower levels of resource contention, the realistic models outperform the naive one due to the speed-up effect described in \cref{sec:background}.
-            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \acp{TTF} takes over.
+            However, at higher levels of contention, the behavioral slow-down of the realistic models due to higher \glspl{TTF} takes over.
         }\label{fig:scaling_duration}
@@
 
-On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \acp{TTF} in this scenario reach the medium impairment level and practically no \acp{TTF} pass threshold for the highest level of impairment.
+On the other, the realistic models experience a mean slowdown of about \SI{5.13}{\percent} at the highest level of resource contention (\num{10} concurrent loops) when compared to the naive model, even though only about \SI{22}{\percent} of \glspl{TTF} in this scenario reach the medium impairment level and practically no \glspl{TTF} pass threshold for the highest level of impairment.
 This showcases that the additional user slow-down due to reduced system responsiveness, originally described in \textcite{olguinmunoz:impact2021}, causes a noticeable and significant extension of system utilization time, even at lower and medium levels of impairment.
@@
 The realistic models were on average \SI{3.8}{\percent} slower than the reference naive model when configured with a low level of neuroticism, and \SI{7.3}{\percent} slower when configured with a high neuroticism.
-In this scenario, an average of \SI{50}{\percent} of all recorded \acp{TTF} fell within the medium impairment range, and about \SI{26.6}{\percent} within the high impairment range, showing that the effects of behavioral slow-down are considerable even in moderately impaired systems.
+In this scenario, an average of \SI{50}{\percent} of all recorded \glspl{TTF} fell within the medium impairment range, and about \SI{26.6}{\percent} within the high impairment range, showing that the effects of behavioral slow-down are considerable even in moderately impaired systems.
 

### 2022-12-19 body/part1/overview/extending-wca.tex
glossary fixes
more glossary fixes

@@
-\section{Extending the methodology in the context of \acs{WCA}}
+\section{Extending the methodology in the context of \gls{WCA}}
 
@@
 \end{enumerate*}.
-Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI} and the \gls{ITQ}.
+Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
 

### 2022-12-19 publications/2022Ainur/introduction.tex
more glossary fixes

@@
 \end{enumerate*} testbeds.
-\gls{COSMOS}~(\emph{\gls{COSMOS}}) is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
+\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
 It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
@@
 
-\gls{POWDER}~(\emph{\gls{POWDER}}) promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
+\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
 The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.

### 2022-12-19 body/part1/overview/extending-wca.tex
extending wca
writing

@@
 
-We present in this paper the design and execution of a study in which undergraduate students of diverse fields of
+We present in this paper the design and execution of a study in which \num{40} undergraduate students of diverse fields of
 study\footnote{%
@@
 
-We employed a version of the step-based LEGO cognitive assistant first introduced by \citeauthor{chen2015early}~\cite{chen2015early}.
-The software was modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
+We employed a version of the step-based \emph{Gabriel} cognitive assistant framework first introduced by \citeauthor{chen2018application}~\cite{chen2018application}.
+The assistant was instrumented to capture key application and task performance metrics in real time, in particular step execution time.
+Furthermore, the software was modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
 This is illustrated in \cref{fig:regularwca-vs-delaywca}.
-If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly \ensuremath{\mathbb{D}} seconds after frame capture.
-Furthermore, the assistant was instrumented to capture key application and task performance metrics in real time, in particular step execution time.
-\todo[inline]{Task, latin square?}
+If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
+Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as  \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=.6\textwidth]{publications/2021ImpactDelayedResponse/Fig4c}
+    \caption{Structure of a \emph{block} in the experimental LEGO task.}\label{fig:stepblock}
+    \todo[inline]{This figure may need reworking for clarity.}
+\end{figure}
+
+The actual \gls{WCA} task employed for this study corresponded to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
+Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consisted of a pseudo-random sequence of steps instructing subjects to add or remove pieces.
+We introduce an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
+To study the effects of a delay applied across multiple steps, we manipulate the length of blocks during the task, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
+We then generate a pseudo-random permutation of the combinations of block length and delay values, and assign a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
+Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reorder the initial permutation to generate a task for each of the \num{40} participants.
 
@@
 This array included devices to capture four physiological measures:
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
     \item \gls{GSR} (also known as electrodermal activity)
@@
 Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
+The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
+
+\subsubsection{Results}
+
+The results of this study can be summarized as follows:
+
+\begin{itemize}
+    \item Reduced system responsiveness induces \emph{additional} behavioral slow-down in users of \gls{WCA}.
+    This additional slow-down moreover scales with the strength of the reduction in system responsiveness.
+
+    \item At higher levels of system responsiveness, humans speed up as the task progresses.
+    However, as system responsiveness decreases, this effect is dampened.
+    Eventually, at the highest levels of system impairment, this effect inverts and humans actually progressively slow-down the longer they spend in a degraded system state.
+
+    \item The effects of reduced system responsiveness on users linger for at least \num{4} steps after system responsiveness improves.
+    \item Measures of individual levels of personality modulate the above effects.
+\end{itemize}
+
+In the following, we will discuss in more detail each of the above conclusions.
+
+\paragraph{Reduced system responsiveness leads to additional behavioral slow-down in users.}
 

### 2022-12-20 body/part1/overview/extending-wca.tex
writing

@@
 \paragraph{Reduced system responsiveness leads to additional behavioral slow-down in users.}
-
-\todo[inline]{\cref{paper:olguinmunoz2021impact} describes the acquisition of the necessary data and insights for the
-models.}
-
-\todo[inline]{From introduction}
-
-We intended to answer four core research questions relating to human responses to decreased application responsiveness.
-
-\begin{itemize}
-    \item \emph{Do subjects change the temporal profile of their actions in relation to system latency?}
-
-    In line with previous research in this area, we expected subjects to change their temporal profiles as system
-    \item responsiveness decreased.
-    The extent or form of these changes were however unknown.
-    We also hypothesized that large enough drops in responsiveness could lead to complete abandonment of the task by
-    \item subjects.
-
-    Our results show an emergent pacing effect on user actions as system responsiveness is reduced.
-    While it would seem self-evident that users take longer to complete a task while using a system with low
-    \item responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down
-    \item represents a source of substantial additional delay.
-    To be more precise, the data indicate that users slow down not only because they have to wait for the system to
-    \item catch up, but that their reactions to new instructions is also delayed.
-    Moreover, this effect scales with the decrease in responsiveness and remains for a while, even after system
-    \item responsiveness improves.
-
-    \item \emph{Do subjects show signals of arousal in physiological responses to changes in system latency?}
-
-    We hypothesized subjects would show signs of stress and frustration as system latency increased, due to the added
-    \item annoyance of dealing with an unresponsive system.
-
-    The results we present, however, seem to refute this hypothesis.
-    We were not able to detect any significant effects on the physiological signals obtained from the biometric
-    \item sensors as system responsiveness was altered.
-
-    \item \emph{Are responses to delay effects in subjects mediated by cognition and/or emotion?}
-
-    In line with previous items, we expected delay effects in subjects to be mediated primarily by emotion.
-    In particular, we expected emotional effects to be correlated with the strength of the added delay.
-
-    The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to
-    \item a disruption of participants' cognitive plan for the task and not to an emotional response.
-    This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
-
-    \item \emph{Are these effects mediated by personality indicators in any way?}
-
-    Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in
-    \item mediating these effects, as it has previously been connected to intolerance for time
-    \item delay~\cite{hirsh2008delay}.
-    We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
-
-    The results obtained agree with our hypothesis.
-    We found significant effects of neuroticism on the responses exhibited by subjects, and all three traits were
-    \item found to play a role through factor analysis.
-
-\end{itemize}
-
-% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive
-% plan for the task.
-% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
-% While it would seem self-evident that users take longer to complete a task while using a system with low
-% responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down
-% represents a source of substantial additional delay.
-% To be more precise, the data indicate that users slow down not only because they have to wait for the system to
-% catch up, but that their reactions to new instructions is also delayed.
-% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic
-% personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.
-
-We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
-One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and
-direct implications for resource and power consumption.
-Another is the fact that the adverse effects of delay on users do not immediately subside as delay is diminished ---
-this has potential consequences for resource allocation strategies.
-Moreover, in multi-user scenarios, the dependency of user slow-down effects on delay mean efficient resource
-allocation across applications potentially looks different from what could be considered ``fair''.
-
-Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of
-WCA.\@
-These results represent unexpected, valuable findings, which can be employed to model and understand how users
-interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
-Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation
-tools such as our previous work in \cite{olguin:2018, olguin:2019}.
-Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of
-realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
-
-\todo[inline]{From discussion}
-
-We start our discussion with the main results of our experimentation presented in the previous section:
-
-\begin{itemize}
-    \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional}
-    \item behavioral slow-down.
-    That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their
-    \item execution of the task.
-    This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on
-    \item average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
-    Moreover, there is a temporal component to this effect; users become progressively slower the more time passes
-    \item with reduced system responsiveness.
-
-    \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness
-    \item \emph{remain} for at least a few steps after system responsiveness improves.
-    This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following
-    \item a high-delay block, as pictured in \cref{fig:exectime:transition}.
-    The question of whether any lingering effect can be measured after these four steps remains open.
-
-    \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at
-    \item performing steps as the task progresses.
-    However, the strength of this effect decreases as delay increases.
-    Whereas for blocks without delay users performed the last four steps of a 12-step block on average
-    \item \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.
-
-    \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response
-    \item to delay.
-    The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of
-    \item neuroticism and both measures of immersion: focus and involvement.
-    Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables
-    \item related to attentiveness, respectively.
-    In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive
-    \item and involved in the task.
-    The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures
-    \item did not correlate with outcomes in other computer-intensive environments such as immersive
-    \item VR~\cite{quesnel2018you}.
-    These correlations are also consistent with previous findings indicating that individuals scoring high in
-    \item neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
-\end{itemize}
-
-A central question therefore arises: to which physio- and psychological mechanisms can these findings, most
-importantly the substantial slow-down in task execution, be attributed?
-
-In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional
-reactions.
-These could in turn elicit generalized arousal.
-We also postulated that adapting to delay might progressively deplete cognitive resources in users.
-However, the present data provide relatively little support for these alternatives, in that the predicted measures
-did not produce the expected statistical trends.
-Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\
-short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
-The acceleration data further did not indicate that extended delay significantly increases erratic movement.
-% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays,
-% and speed-induced errors and non-completions predicted by resource depletion were not observed.
-% The acceleration data further do not indicate that extended delay increases erratic movement.
-To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
-That is, users fail to capitalize on the new information as quickly as they could.
-Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control
-mechanisms, and not from emotion or resource depletion.
-We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
-As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been
-modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
-Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based
-control at the step-by-step level that is easily diverted.
-This also provides a possible explanation for the lingering effects of delay after an acceptable system
-responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.
-
-As to the applicability of our findings to other applications, it must be noted that these results pertain to a
-specific class of applications, namely step-based task-guidance WCA.\@
-However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of
-seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance
-the state.
-
-The results here presented provide a number of possible implications for WCA system design and optimization, both for
-single and multi-application flows.
-
-\begin{itemize}
-
-    \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to
-    \item significantly extended application lifetimes.
-    This has direct implications for resource and power consumption.
-
-    \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a
-    \item high-responsive state could have unconventional consequences for resource allocation.
-    This is of particular importance, for instance, for cases where the user may be able to finish the task before
-    \item these effects subside.
-    In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired
-    \item application.
-
-    \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of
-    \item system responsiveness across applications may not ultimately be beneficial to the system as a whole.
-    Take for instance two applications on the same system which negatively interfere with each other.
-    The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn
-    \item causes them to interfere even longer, potentially entering a positive feedback loop.
-    In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead
-    \item to resources being freed up faster system-wide.
-
-    \item Based on our findings relating individual differences between users and their sensitivity to delays, it
-    \item might also be possible to extrapolate user characteristics from measured execution times.
-    This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users
-    \item with a higher sensitivity to system-state degradation.
-    However, this remains an open challenge.
-
-\end{itemize}
-
-To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of
-WCA deployments.
-Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable
-tool to tackle inefficiencies in these systems.
-Moreover, we also argue these findings represent a first step towards a full-fledged understanding of the
-relationship between application responsiveness and human behavior.
-More research in this area will surely uncover more complex and interesting behaviors.
-Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that
-might be constructed under existing architectures like ACT-R.
-These same parameters could be used to modulate the timing and generation of inputs in trace-based workload
-generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same
-trace to generate workloads for a multitude of different user profiles.
-
-\todo[inline]{Some diagrams, plots, and results}
-
-\begin{figure}[h]
-    \centering
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
-        \caption{Structure of a step in a generic cognitive assistance application.
-        The assistant provides an instruction to the user and continuously samples the step state; inputs captured
-        while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new
-        instruction).
-        % by the assistant, as they do not contain relevant information.
-        Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new
-        instruction.
-        % which will subsequently be provided to the user.}%
-        }
-        \label{fig:cogassist:step}
-    \end{subfigure}%
-    % \medskip%
-    \hfill%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
-        \caption{In the experimental task, an additional variable segment of time is introduced immediately following
-        the processing of the input frame in order to extend the perceived processing time of the input to a specific
-        target delay.}%
-        \label{fig:cogassist:step:delay}
-    \end{subfigure}
-    \medskip%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
-        \caption{Structure of a block in the experimental task.}%
-        \label{fig:cogassist:block}
-    \end{subfigure}%
-    % \medskip%
-    \hfill%
-    \begin{subfigure}[t]{.49\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
-        \caption{Visualization of the execution time of a step.}%
-        \label{fig:exectime:diagram}
-    \end{subfigure}%
-    % \medskip%
-    \caption{Components of the cognitive assistance task.}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
-    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean
-        (S.E.M.)}\label{fig:exectime}%
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
-    \caption{Mean per-step execution time vs.\ delay, by step slice.
-    Error bars indicate S.E.M.}
-    \label{fig:exectime:delay:slice}
-\end{figure}
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
-    \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to
-        \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
-\end{figure}
-
-\begin{table}[h]
-    \centering
-    \caption{Principal Component Analysis}\label{tab:pca}
-    \begin{subtable}[h]{\textwidth}
-        \centering
-        \caption{Main components identified.}
-        \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
-        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
-            \toprule
-            \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2}
-            & \textbf{Comp. 3} \\
-            \midrule
-            BFI Conscientiousness & \textcolor{lightgray}{\( -0.022 \)} & \( 0.668 \)
-            & \textcolor{lightgray}{\( -0.481 \)} \\
-            BFI Neuroticism & \( 0.600 \) & \( -0.678 \)
-            & \textcolor{lightgray}{\( -0.118 \)} \\
-            ITQ Focus & \( 0.678 \) &
-            \textcolor{lightgray}{\( 0.203 \)} & \( 0.504 \) \\
-            ITQ Involvement & \( 0.573 \) & \( 0.540 \)
-            & \textcolor{lightgray}{\( 0.417 \)} \\
-            Exec. Time (delay 3.0 s, length 12) & \( 0.758 \) &
-            \textcolor{lightgray}{\( -0.178 \)}
-            & \textcolor{lightgray}{\( -0.348 \)}
-            \\
-            Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} &
-            \textcolor{lightgray}{\( -0.251 \)}
-            & \( 0.589 \)
-            \\
-            \bottomrule
-        \end{tabular*}
-    \end{subtable}
-    \newline
-    \medskip
-    \newline
-    \begin{subtable}[h]{\textwidth}
-        \centering
-        \caption{Percentages of variance explained by the components.}
-        \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
-            \toprule
-            {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total}
-            \\
-            \midrule
-            Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent}
-            & \SI{73.13}{\percent}
-            \\
-            \bottomrule
-        \end{tabular*}
-    \end{subtable}
-\end{table}
-
-\subsection{Emulating human behavior}
-
-We introduce the first, to our knowledge, data-driven model for human timings in \gls{WCA} applications.
-Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as
-input past measurements of system responsiveness and produces realistic step execution times.
-We also introduce a novel way to generate dynamic traces of frames for \gls{WCA} applications which can be combined
-with the timing model for a full end-to-end emulation of a human.
-We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid
-approach~\cite{olguin2018scaling,olguin2019edgedroid}.
-
-The study and benchmarking of \gls{WCA} applications is a challenging discipline due to these application's intrinsic
-human-in-the-loop nature.
-Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
-Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming
-and prohibitively expensive for many research groups.
-
-In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \gls{WCA},
-the first data-driven model for human timings in \gls{WCA} applications.
-This model represents a stochastic approach to execution time modeling which builds upon the data collected for
-\textcite{olguinmunoz:impact2021}.
-Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \gls{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
-
-\todo[inline]{TODO: Some relevant results. Need to add paper to thesis.}

### 2022-12-20 body/part1/overview/implications-optimization.tex
writing

@@
 \section{Implications for the optimization of \gls{WCA}}
-
-\begin{enumerate}
-    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \gls{WCA}.
-    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
-    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \gls{WCA} when considering human behavior using our model.
-    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \gls{WCA} optimization; number of processed samples and energy consumption per step.
-    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
-\end{enumerate}
-
-Next, we have explored the impact of such a realistic model on the application lifetime footprint of \gls{WCA} applications.
-We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.
-
-Finally, we have delved into the potential for optimization in \gls{WCA} systems using the previously discussed timing models.
-We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \gls{WCA} which results in an adaptive sampling strategy.
-We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
-Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \gls{WCA} applications.

### 2022-12-20 body/part1/overview/methodology.tex
writing

@@
 \todo[]{}
-
-reproducible and comparable workload, we use a trace-based design for the client-side emulation.
-.
-
-To then use the trace for reproducible experiments, we introduce a benchmarking suite which can replay the trace to the original application.
-This results in the same computation to be performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-Additionally, the suite is instrumented to collect system level metrics at runtime for posterior analysis.
-
-However, merely replaying the trace is not enough, as disparities between system responsiveness at recording versus replay time can cause the trace so drift out of synchronization with the current state of the system.
-Thus, we introduce here the concept of a human user model which is able to adapt its behavior in a realistic manner based on the current responsiveness of the system by systematically controlling the replay of the trace.
-The initial approximation we introduce, illustrated in \cref{fig:usermodel}, here is that of a human that is patient and does not make mistakes, and any error message received from the application backend is ignored.
-The trace is preprocessed into segments that correspond to individual steps of the \gls{WCA} application; this is performed manually as a preparation step.
-The user model then replays each segment in order; whenever feedback corresponding to a change of step is received from the \gls{WCA} backend, the user model transitions to the next segment in the trace, no matter whether the current segment has been replayed out completely or not.
-On the other hand, if no step transition feedback has been received at the end of the current trace segment, the model rewinds the trace a $\tau$ seconds and tries again.
-To avoid infinite loops, where the application is stuck on a step forever, we have a maximum number of possible rewinds, after which the application shuts down.
-
-\todo[inline]{EdgeDroid paper}
-
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
-    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
-    \label{fig:comparison:rtt}%
-\end{figure}%
-\begin{figure}
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
-    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
-    \label{fig:comparison:feedback}%
-\end{figure}%
-\begin{figure}%
-    \centering%
-    \includegraphics[width=.85\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
-    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
-    \label{fig:comparison:tasksteps}
-\end{figure}%
-
-The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
-We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
-
-Due to the complex nature of the applications and the infrastructure, we opt for experimentally studying the trade-offs in a repeatable and controllable fashion.
-This is difficult mainly due to the unpredictable reaction of human users to the feedback from the backend --- a user might very well misinterpret the feedback handed to them.
-EdgeDroid 1.0 mimics the operation of human-in-the-loop applications by replaying recorded traces of sensory input.
-This sensory information is then processed by the original compute process at the backend, generating feedback.
-However, this feedback is not processed by humans, but by a parameterizable model of human reaction.
-Through synchronized time tracking at the different processing points of the application, EdgeDroid 1.0 allows for accurate measurements of key performance metrics such as the distribution of delays across the application pipeline.
-Analysis of these metrics can be performed down to the individual input sample, allowing us to zoom into the internal model of the application under consideration.
-Thus, EdgeDroid 1.0 allows us to illuminate the many latency trade-offs existing at the level of the infrastructure, as well as the level of the application.
-It can also be used for debugging and validation, by comparing the expected execution flow of a particular trace with the actual flow during the benchmarking.
-To the best of our knowledge, this experimentally-driven benchmarking approach is the first one towards experimental performance characterization and potential optimization of human-in-the-loop applications
-
-\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
-Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
-We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
-Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
-These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
-Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.
-
-In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.
-
-For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
-This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.
-
-We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
-These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
-It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.
-
-We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
-One, scaling behavior for the application seems to be linear with respect to the number of clients.
-Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
-
-It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
-Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.
-
-EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
-From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
-Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
-Here, the downlink time sometimes almost reaches the ideal bound by itself.
-A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.
-
-Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
-The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.
-
-\todo[inline]{\cref{paper:olguinmunoz2019edgedroid}}\label{summary:2019edgedroid}
-
-\subsection{Applying the methodology to \glspl{NCS}}\label{ssec:methodology:ncs}
-
-\todo[inline]{\Cref{paper:olguinmunoz2022cleave} implements both the methodology and the tool.}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \glspl{NCS} in \gls{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \gls{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \glspl{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \gls{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \glspl{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\gls{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
-
-In this work, we aimed to tackle this issue through a fully software-based framework for repeatable, reproducible, and easily scalable \gls{NCS} benchmarking with a particular focus on edge deployment.
-We argue our approach, \gls{CLEAVE}, embodies a better solution than previous work for a number of reasons:
-\begin{enumerate}
-    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
-    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
-    Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
-    In contrast, our approach requires only general-purpose computing platforms, and can be employed by basically anyone with access to a computer.
-    Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \gls{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \gls{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
-    The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.
-\end{enumerate}
-
-We validate the utility of this tool through an example use case approximating a proposed edge deployment of inverted pendula control loops co-located with video analytics services.
-We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item the inverted pendulum plant is ubiquitous in \gls{NCS} research
-    \item similar setups exist in real-world industrial use
-    \item video analytics has long been proposed as a ``killer app'' for edge computing.
-\end{enumerate*}
-Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.
-We believe \gls{CLEAVE} represents an important step towards enabling inexpensive and low-complexity scalable research for real-world deployment of edge-bound \glspl{NCS}.
-
-\todo[inline]{Some plots and results}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.6\textwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
-    \caption{
-        The 2D inverted pendulum system.
-    }\label{fig:invpend}
-\end{figure}
-
-\begin{figure}[t]
-    \centering
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
-        \caption{Toppled plants.}\label{fig:video:toppled}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \gls{RMS}.}\label{fig:video:rms}
-    \end{subfigure}\\
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
-        \caption{Packet losses.}\label{fig:video:drop}
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\glspl{RTT}.}\label{fig:video:rtt}
-    \end{subfigure}%
-    \caption{
-        Multi-loop, resource-constrained setup results.
-        Error bars indicate \SI{95}{\percent} \glspl{CI} in all plots.
-    }\label{fig:video:results}
-\end{figure}
-
-\FloatBarrier%
-
-\subsection{Orchestrating experimentation in edge computing testbeds}
-
-\todo[inline]{\cref{paper:olguinmunoz2022ainur} describes a software framework for testbeds used for the methodology described in this thesis}
-
-In this work, we present our solution to the challenge of testbed automation: Ainur, a framework for wireless testbed automation with a specific focus on end-to-end experimental research in the context of edge-computing using cloud- and edge-native technologies.
-Ainur is designed to deploy experimental runs from a workload perspective by configuring the physical testbed, initializing all involved software components, deploying and executing the experimental workload, collecting logs and data, and finally gracefully degrading the system.
-The framework allows for dynamic, software-definition of physical and logical links, network topology, cloud and edge computing resources, as well as experimental workload deployment and orchestration.
-It heavily leverages cloud-native technologies, such as Docker containers, in order to support a wide variety of different testbed hardware setups and experimental configurations and workloads, as well as to be as easily extendable as possible.
-Furthermore, we make Ainur available to the community as \gls{FOSS}.
-It can be obtained from the {KTH-EXPECA/Ainur} repository on GitHub~\cite{ainur:github}, released under an Apache version \num{2.0} license.
-
-In this paper, we have introduced a framework for repeatable end-to-end testbed automation in the context of wireless networking and edge computing research.
-Named Ainur, it simplifies the execution and verification of end-to-end experimentation by automating the
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
-    \item establishment of physical links between hosts, including the configuration of complex wireless systems such as 4G \gls{LTE} and 5G
-    \item provisioning of and connection to remote cloud instances
-    \item initialization of \gls{IP} layer connectivity between hosts
-    \item collection of logs and data
-    \item deployment, scaling, and lifecycle management of containerized processes
-\end{enumerate*}.
-We have described its general architecture, which follows a layered design mimicking the network stack layers the framework directly interacts with, as well as the underlying assumptions about its deployment environment and specific requirements for its deployment.
-We have also outlined a demonstration which showcases the flexibility and power of the framework by deploying two different workloads to our testbed.
-We believe our framework represents an important step towards repeatable, replicable, yet low-access barrier end-to-end wireless testbed experimentation.
-It has been released as \gls{FOSS} and can be found on GitHub~\cite{ainur:github}.\\
-
-\todo[inline]{Some plots}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
-    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
-\end{figure}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
-\end{figure*}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
-    \caption{Network structure assumed by Ainur.}\label{fig:network}
-\end{figure*}

### 2023-01-06 body/glossary.tex
more glossary fixes
delays to ttf

@@
 \newabbreviation{WCA}{WCA}{Wearable Cognitive Assistance}
-\newabbreviation{AR}{AR}{Augmented Reality}
-\newabbreviation{VR}{VR}{Virtual Reality}
-\newabbreviation{XR}{XR}{eXtended Reality}
+\newabbreviation[longplural={Augmented Realities}]{AR}{AR}{Augmented Reality}
+\newabbreviation[longplural={Virtual Realities}]{VR}{VR}{Virtual Reality}
+\newabbreviation[longplural={eXtended Realities}]{XR}{XR}{eXtended Reality}
 \newabbreviation{GPU}{GPU}{Graphical Processing Unit}
@@
 \newabbreviation{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
-\newabbreviation{ABC}{ABC}{Abstract Base Class}
+\newabbreviation[longplural={Abstract Base Classes}]{ABC}{ABC}{Abstract Base Class}
 %\acroplural{ABC}[ABCs]{Abstract Base Classes}
@@
 \newabbreviation{EEG}{EEG}{electroencephalography}
-\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits~\cite{john1999big}}
-\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}
+\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits}
+\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire}
 \newabbreviation{GPS}{GPS}{Global Positioning System}
@@
 \newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}
-\newabbreviation{TTF}{TTF}{Time-to-Feedback}
+\newabbreviation[longplural={Times-to-Feedback}]{TTF}{TTF}{Time-to-Feedback}
 \newabbreviation{MAR}{MAR}{Mobile \gls{AR}}

### 2023-01-06 body/part1/overview/extending-wca.tex
delays to ttf

@@
 If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
+That is,
+\begin{align}
+    t_k - t_{k - 1} = \mathbb{D} \qquad \forall k \in [2, n + 1]
+\end{align}
 Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
+In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note that the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
+Thus the average \gls{TTF} for a step can be expressed as
+\begin{align*}
+    TTF = (t_{n + 1} - t_{n}) + \frac{1}{2}(t_n - t_{n - 1})
+\end{align*}
+which, for a step with a given delay \ensuremath{\mathbb{D}}, works out to\footnote{%
+    The \num{0}/no-delay case was handled by calculating the empirical average frame \gls{RTT} for these steps, and using this value as the delay \ensuremath{\mathbb{D}}.
+    This yielded an average \gls{TTF} of \SI{0.57}{\second} for these steps.
+}
+\begin{align}
+    TTF &= \mathbb{D} + \frac{1}{2}\mathbb{D}\nonumber\\
+        &= \frac{3}{2}\mathbb{D}
+\end{align}
+In the following, we will forgo the delay values and present results with respect to the adjusted \glspl{TTF}.
 
@@
 \paragraph{Reduced system responsiveness leads to additional behavioral slow-down in users.}
+
+Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.

### 2023-01-09 body/part1/overview/extending-wca.tex
impact results

@@
 It was our hypothesis that, among others, the individual trait of \emph{neuroticism}~\cite{john1999big} would play a particularly important role in mediating effects of reduced system responsiveness on subjects.
+
+Our results partially agree with our initial hypotheses.
+We found that reduced system responsiveness induces \emph{additional} behavioral slow-down in users of \gls{WCA}.
+This additional slow-down moreover scales with the strength of the reduction in system responsiveness.
+Furthermore, we note that humans tend to speed up as the task progresses.
+However, as system responsiveness decreases, this effect is strongly dampened.
+We also found that the effects of reduced system responsiveness on users linger for at least \num{4} steps after system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
+
+Finally, we also found that the above effects are modulated by measures of individual levels of personality.
+In particular, the trait of \emph{neuroticism} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
+
+\todo[inline]{Discuss implications?}
+
+\subsubsection{Methodology}
 
@@
 Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consisted of a pseudo-random sequence of steps instructing subjects to add or remove pieces.
-We introduce an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
-To study the effects of a delay applied across multiple steps, we manipulate the length of blocks during the task, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
-We then generate a pseudo-random permutation of the combinations of block length and delay values, and assign a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
-Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reorder the initial permutation to generate a task for each of the \num{40} participants.
 
-In order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
+We introduced an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
+To study the effects of a delay applied across multiple steps, we manipulated the length of blocks during the task, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
+Furthermore, we defined a \emph{slice} as a continuous segment of \num{4} steps within a block.
+In other words, \num{4}-step blocks were composed of a single slice, whereas \num{8}- and \num{12}-step blocks were composed respectively of \num{2} and \num{3} consecutive slices.
+
+Using the above experimental design components, we then generated a task for each participant.
+We generated a pseudo-random permutation of the combinations of block length and delay values, and assigned a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
+Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reordered the initial permutation to generate \num{40} tasks.
+
+To measure the extent to which individual differences in each subject affected obtained task-performance results, participants were asked to complete two questionnaires previous to beginning the task;
+the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
+These correspond to well-established tools for the measurement of individual personality traits in humans.
+The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item agreeableness
+    \item conscientiousness
+    \item extroversion
+    \item neuroticism
+    \item openness
+\end{enumerate*}.
+The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
+The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
+
+Finally, in order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
 This array included devices to capture four physiological measures:
@@
 \end{enumerate*}.
-Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
-The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
+However, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
+Please refer to \cref{paper:olguinmunoz2021impact} for details.
 
@@
 
-The results of this study can be summarized as follows:
+\begin{figure*}[tb]
+    \centering
+    \begin{subfigure}[t]{0.45\textwidth}
+        \includegraphics[height=13em]{./Figs/2021Impact/ttf_vs_exectime_per_block}
+        \caption{%
+            \gls{TTF} vs average execution time, grouped by block length.
+            Transparent bands indicate \SI{95}{\percent} \glspl{CI}.
+        }\label{fig:ttfvsexectime:block}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{0.45\textwidth}
+        \includegraphics[height=13em]{./Figs/2021Impact/ttf_vs_exectime_per_slice}
+        \caption{%
+            \gls{TTF} vs average execution time, grouped by step slice in a block.
+            Slice \num{1} corresponds to steps \numrange{1}{4}, \num{2} to steps \numrange{5}{8}, and \num{3} to steps \numrange{9}{12} in a block.
+            Blocks with \num{4} steps only contribute to slice \num{1}, and blocks with \num{8} steps to slices \num{1} and \num{2}.
+            Transparent bands indicate \SI{95}{\percent} \glspl{CI}.
+        }\label{fig:ttfvsexectime:slice}
+    \end{subfigure}%
+    \caption{Effects of perceived \gls{TTF} on measured step execution times by participants.}\label{fig:ttfvsexectime}
+\end{figure*}
 
-\begin{itemize}
-    \item Reduced system responsiveness induces \emph{additional} behavioral slow-down in users of \gls{WCA}.
-    This additional slow-down moreover scales with the strength of the reduction in system responsiveness.
+We found that as system responsiveness decreased, the speed at which participants completed instructions tended to slow down.
+Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
+\Cref{fig:ttfvsexectime:block} illustrates this effect clearly;
+for all block lengths, as \gls{TTF} grows we see an accompanying increase in execution times.
+Note that by measuring \emph{step execution time} we are \emph{per se} compensating for the added delay in the measure, and thus this trend must result from the participants' behavioral slow-down as a response to the perceived increase in \gls{TTF}.
+This was confirmed through an \gls{ANOVA}~\cite{fujikoshi1993two} including factors of block length and \gls{TTF}; significant effects were found for both factors and their interaction (\ensuremath{p < 0.01}).
 
-    \item At higher levels of system responsiveness, humans speed up as the task progresses.
-    However, as system responsiveness decreases, this effect is dampened.
-    Eventually, at the highest levels of system impairment, this effect inverts and humans actually progressively slow-down the longer they spend in a degraded system state.
-
-    \item The effects of reduced system responsiveness on users linger for at least \num{4} steps after system responsiveness improves.
-    \item Measures of individual levels of personality modulate the above effects.
-\end{itemize}
-
-In the following, we will discuss in more detail each of the above conclusions.
-
-\paragraph{Reduced system responsiveness leads to additional behavioral slow-down in users.}
-
-Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
+\todo[inline]{Write about slice effects.}

### 2023-01-11 body/part1/overview/extending-wca.tex
moar results

@@
 
-We found that as system responsiveness decreased, the speed at which participants completed instructions tended to slow down.
+\Cref{fig:ttfvsexectime} shows effects on the measured execution times of participants as perceived \gls{TTF} increased.
+These results have been grouped by block length (\cref{fig:ttfvsexectime:block}), showing the cumulative effects of delay over series of steps, and slice number (\cref{fig:ttfvsexectime:slice}), showing the evolution of these effects within a series of steps at the same level of impairment.
+
+\cref{fig:ttfvsexectime:block} shows that as system responsiveness decreased, the speed at which participants completed instructions tended to slow down.
 Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
@@
 for all block lengths, as \gls{TTF} grows we see an accompanying increase in execution times.
+Moreover, this effect seemed stronger for longer blocks at higher levels of impairment, indicating a slow-down which compounds with the number of steps.
 Note that by measuring \emph{step execution time} we are \emph{per se} compensating for the added delay in the measure, and thus this trend must result from the participants' behavioral slow-down as a response to the perceived increase in \gls{TTF}.
@@
 
-\todo[inline]{Write about slice effects.}
+On the other hand, in \cref{fig:ttfvsexectime:slice} we observe the evolution of execution times over \emph{slices} of \num{4} steps within blocks.
+This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
+At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \textasciitilde\SI{26}{\percent} lower than for steps \num{1} to \num{4}.
+However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
+This was also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.
+
+As mentioned before, we also found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
+On average, the mean execution time of the first four steps of blocks were roughly \SI{10}{\percent} lower when the previous block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the previous block was at a \ensuremath{TTF > \SI{2.475}{\second}}.
+
+Finally, analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
+This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
+Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \textasciitilde\SI{73}{\percent} of the variance in the execution time results, neuroticism was present in the first two.
+Furthermore, individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, \num{2}-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.

### 2023-01-12 body/part1/overview/extending-wca.tex
writing

@@
 
-\subsubsection{Results}
+\subsubsection{Results}\label{impact:results}
 
@@
 
+\begin{wrapfigure}{l}{.5\textwidth}
+    \centering
+    \includegraphics[height=13em]{figs/2021Impact/previousblock_vs_exectime}
+    \caption{}\label{fig:prevblockvsexectime}
+\end{wrapfigure}
+
 As mentioned before, we also found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
-On average, the mean execution time of the first four steps of blocks were roughly \SI{10}{\percent} lower when the previous block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the previous block was at a \ensuremath{TTF > \SI{2.475}{\second}}.
+\Cref{fig:prevblockvsexectime} illustrates this effect.
+On average, the mean execution time of the first four steps of blocks was roughly \SI{7}{\percent} lower when the previous block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the previous block was at a \ensuremath{TTF > \SI{2.475}{\second}} (single-tailed \ensuremath{p < 0.001}).
 
@@
 Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \textasciitilde\SI{73}{\percent} of the variance in the execution time results, neuroticism was present in the first two.
-Furthermore, individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, \num{2}-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+Furthermore, individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+
+\subsection{Extending the EdgeDroid tool with a realistic model of human behavior}
+
+\cref{paper:olguinmunoz2023realistic} discusses the elaboration of a model of human behavior for \gls{WCA}.
+Built upon the data and results from \cref{paper:olguinmunoz2021impact}, our model is capable of dynamically generating realistic predictions of execution times for steps in a \gls{WCA} based on historical system state data and a parameterized level of neuroticism.
+Furthermore, we implement a new version of the EdgeDroid tool capable of using this new model, together with a novel procedure for the generation of realistic synthetic traces of frames for \gls{WCA}, to fully emulate a human in the context of these systems.
+
+\subsubsection{Designing a realistic model for human timing ins \gls{WCA}}
+
+As evidenced from the results of \cref{paper:olguinmunoz2021impact}, a model capable of generating realistic execution times for \gls{WCA} must take several things in consideration.
+As discussed in \cref{impact:results}, the execution time for a step depends not only on the most recent perceived \gls{TTF}, but also on how long the system has been in the current level of responsiveness.
+Additionally, as shown in \cref{fig:prevblockvsexectime}

### 2023-01-15 body/part1/overview/extending-wca.tex
figs -> Figs
more writing

@@
 
-\begin{wrapfigure}{l}{.5\textwidth}
+\begin{figure}
+    \begin{minipage}[t]{.45\textwidth}
         \centering
-    \includegraphics[height=13em]{figs/2021Impact/previousblock_vs_exectime}
-    \caption{}\label{fig:prevblockvsexectime}
-\end{wrapfigure}
+        \includegraphics[height=13em]{Figs/2021Impact/previousblock_vs_exectime}
+        \caption{Mean execution time of the first slice of steps in a block, grouped by \acs{TTF} of the block immediately preceding the current one.}\label{fig:prevblockvsexectime}
+    \end{minipage}%
+    \hfill%
+    \begin{minipage}[t]{.45\textwidth}
+        \centering
+        \includegraphics[height=13em]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0}
+        \caption{%
+            Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \acs{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
+            Fits were achieved through \gls{MLE}.
+        }\label{fig:ttfvsexgaussianmu}
+    \end{minipage}
+\end{figure}
 
-As mentioned before, we also found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
+As mentioned before, we found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
 \Cref{fig:prevblockvsexectime} illustrates this effect.
@@
 
-Finally, analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
+Analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
 This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
 Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \textasciitilde\SI{73}{\percent} of the variance in the execution time results, neuroticism was present in the first two.
-Furthermore, individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+
+Finally, we found that execution times were well-fit by \gls{exGaussian} distributions.
+Previous research has established strong arguments for the use of this distribution for the modeling of human reaction times and the timings of human actions~\cite{rohrer1994analysis,palmer2011what,marmolejo_ramos2022generalised}.
+When grouping execution times by low/high neuroticism\footnote{%
+    \emph{Low} refers to values less than the middle value obtainable for neuroticism in the \gls{BFI}; \emph{high} to values greater than or equal to it.
+}, \gls{TTF}, and slice, the best fit statistic obtained using a Kolmogorov-Smirnov goodness-of-fit test~\cite{massey_jr1951kolmogorov} was \num{0.028}, with a \ensuremath{p}-value of \num{0.999}.
+The effects of neuroticism were most easily observed in the \ensuremath{\mu} parameter and the tails of the fitted distribution;
+an example of this can be observed in \cref{fig:ttfvsexgaussianmu}.
+There is a clear and direct correlation between higher neuroticism and higher \ensuremath{\mu} values and longer tails.
 
@@
 
-\subsubsection{Designing a realistic model for human timing ins \gls{WCA}}
+\subsubsection{Designing a realistic model for human timings in \gls{WCA}}
 
-As evidenced from the results of \cref{paper:olguinmunoz2021impact}, a model capable of generating realistic execution times for \gls{WCA} must take several things in consideration.
 As discussed in \cref{impact:results}, the execution time for a step depends not only on the most recent perceived \gls{TTF}, but also on how long the system has been in the current level of responsiveness.
-Additionally, as shown in \cref{fig:prevblockvsexectime}
+Moreover, as shown in \cref{fig:prevblockvsexectime}, the effects of reduced system responsiveness on generated execution times must be taken into consideration even after the system transitions into a more responsive state.
+All of these effects are also modulated by neuroticism, adding both another layer of complexity and a means of parameterizing the model.
+
+Our timing model produces realistic execution times for each step, according to the above effects.
+We achieve this through the following design:
+
+\begin{enumerate}
+    \item Our model takes two inputs, the measured \gls{TTF} for the most recent step and a level of normalized neuroticism.
+    We use two levels of neuroticism: \emph{low}, for normalized values in the range \ensuremath{[0, 0.5)}; and \emph{high}, for the range \ensuremath{[0.5, 1.0]}.
+    \item The model is stateful, keeping the values of the \num{12} most recent \glspl{TTF} in memory.
+    \item Whenever a new \gls{TTF} is received, the thirteenth most recent \gls{TTF} is dropped, and a weighted average is calculated using the values in memory.
+    This value is obtained using the formula
+    \begin{equation}\label{eq:modelweights}
+    w_{n - i} =
+    \left\{ \begin{array}{ll}
+                \frac{e^{-0.7 i}}{\sum\limits^{12}_{j=1} e^{-0.7 j}} & 1 \leq i \leq 12 \\
+                & \\
+                0 & i > 12
+    \end{array} \right.
+    \end{equation}
+    This results in the \gls{TTF} for the most recent step accounting for approximately \SI{50}{\percent} of the weighted result, with this percentage decaying by a factor of about \ensuremath{\frac{1}{2}} for each step further back in the sequence.
+
+    \item The resulting weighted rolling \gls{TTF} is used to look-up potential values for the execution time in the original data from \cref{paper:olguinmunoz2021impact}.
+    We do this by pre-processing the original data from the \num{40} independent repetitions of the LEGO task with this same rolling weighted average formula, and then binning samples according to the resulting \num{7}-quantiles for weighted \glspl{TTF}.
+    The weighted \gls{TTF} calculated during runtime is then matched to the corresponding bin of samples.
+
+    \item The selected samples are further filtered according to the selected level of neuroticism for the model.
+    \item Finally, a realistic execution time for the step is generated either by directly sampling the execution time values from the selected samples, or first fitting a theoretical \gls{exGaussian} distribution to these values and then sampling the distribution instead.
+\end{enumerate}

### 2023-01-15 body/part1/overview/extending-wca.tex
exectime model

@@
 
-\subsubsection{Designing a realistic model for human timings in \gls{WCA}}
+\subsubsection{Design of a realistic model for human timings in \gls{WCA}}
+
+\todo[inline]{Do we need a discussion on why an improved model is necessary?}
 
@@
 
-    \item The resulting weighted rolling \gls{TTF} is used to look-up potential values for the execution time in the original data from \cref{paper:olguinmunoz2021impact}.
+    \item The resulting weighted rolling \gls{TTF} is used to look up potential values for the execution time in the original data from \cref{paper:olguinmunoz2021impact}.
     We do this by pre-processing the original data from the \num{40} independent repetitions of the LEGO task with this same rolling weighted average formula, and then binning samples according to the resulting \num{7}-quantiles for weighted \glspl{TTF}.
@@
 \end{enumerate}
+
+\begin{figure}[tb]
+    \centering
+    \includegraphics[width=\textwidth]{Figs/2023EdgeDroid2/model_exectime_over_steps}
+    \caption{
+        Illustration of the execution times output by the model at fixed values of \glspl{TTF}, over a series of steps, parameterized by levels of neuroticism.
+        Shaded bands indicate \SI{95}{\percent} \glspl{CI}.
+    }\label{fig:model:exectimesoversteps}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[height=13em]{Figs/2023EdgeDroid2/model_exectimes_transition}
+    \caption{%
+        Mean execution times generated by the model after a single step at a \gls{TTF} of \SI{2.5}{\second} preceded by a sequence of \num{25} steps at fixed lower (instantaneous feedback) or higher (\SI{5}{\second}) \gls{TTF}.
+        Errorbars indicate \SI{95}{\percent} \glspl{CI}.
+    }\label{fig:model:exectimestransitions}
+\end{figure}
+
+\Cref{fig:model:exectimesoversteps,fig:model:exectimestransitions} illustrate the adherence of the model to the conclusions obtained from our work in \cref{paper:olguinmunoz2021impact}.
+
+\Cref{fig:model:exectimesoversteps} shows the output of the execution time model described above over a series of steps at three levels of \gls{TTF};
+\emph{low} (instantaneous feedback), \emph{medium} (\SI{2.5}{\second}), and \emph{high} (\SI{5}{\second}).
+When the system is highly responsive (low \glspl{TTF}), the model tends to speed up as the task progresses.
+Conversely, at high levels of impairment, execution times depend on strongly on the parameterized level of neuroticism;
+as concluded from our previous discussion, higher neuroticism leads to a stronger reaction to reduced responsiveness and subsequently to longer execution times.
+
+In \cref{fig:model:exectimestransitions}, moreover, we present the lingering effects of reduced system responsiveness on \glspl{TTF} generated after the system returns to a more responsive state.
+These results were generated by first feeding a sequence of \num{25} identical \glspl{TTF} to the model, with values of either \SI{0}{\second} (i.e.\ instantaneous feedback) or \SI{5}{\second}.
+Next, a single \gls{TTF} of \SI{2.5}{\second} was fed to the model, and the generated execution time was recorded.
+This procedure was repeated \num{600} times.
+As expected from our discussion in \cref{impact:results}, transitions from a higher (\SI{5}{\second}) \gls{TTF} resulted in noticeably inflated execution times (\ensuremath{+\SI{8.5}{\percent}} at a high level of neuroticism) for the first step at \SI{2.5}{\second} \gls{TTF} compared to when the transition was from a lower \gls{TTF}.

### 2023-01-16 body/part1/overview/extending-wca.tex
more writing

@@
 
-Finally, in order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
+In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
 This array included devices to capture four physiological measures:
@@
 \end{enumerate*}.
-However, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
+Unfortunately, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
 Please refer to \cref{paper:olguinmunoz2021impact} for details.
+
+Finally, we collected video recordings of participants' faces and as well as of the live feed of frames provided as input to the \gls{WCA}.
+The facial recordings will also not be discussed in this section, as they were not employed in any analyses.
+However, the recording of the live feed was employed later on in the implementation of a procedure for the generation of dynamic synthetic traces for the LEGO task in the context of the EdgeDroid load generator for \gls{WCA}.
+This will be discussed in \cref{model:traces}.
 
@@
 
-\todo[inline]{Do we need a discussion on why an improved model is necessary?}
+\todo[inline]{Probably needs to repeat a discussion on why the new model is necessary.
+This discussion should first be presented at the beginning of this section however.}
 
@@
 As expected from our discussion in \cref{impact:results}, transitions from a higher (\SI{5}{\second}) \gls{TTF} resulted in noticeably inflated execution times (\ensuremath{+\SI{8.5}{\percent}} at a high level of neuroticism) for the first step at \SI{2.5}{\second} \gls{TTF} compared to when the transition was from a lower \gls{TTF}.
+
+\subsubsection{Procedural generation of dynamic traces of video frames for \gls{WCA}}\label{model:traces}

### 2023-01-16 body/part1/overview/extending-wca.tex
writing

@@
 \subsubsection{Procedural generation of dynamic traces of video frames for \gls{WCA}}\label{model:traces}
+
+\begin{figure}
+    \centering
+    \includegraphics[height=13em]{publications/2023EdgeDroid2/model_data/frame_probabilities}
+    \caption{%
+        Distribution of processing results for frames captured during the execution of steps in the LEGO task.
+        Note that \emph{Success} results are not included since by definition \ensuremath{P(\text{Success}|t_i) = 1.0} for all normalized instants \ensuremath{t_i \geq 1.0}.
+    }
+    \label{fig:modelframes}
+\end{figure}
+
+Our design for a procedure for the generation of synthetic yet realistic traces of video inputs for \gls{WCA} is based on a stochastic approach.
+Analysis of the distribution of processing results of frames during the execution of steps in the LEGO task highlighted a pattern that was independent from execution times, illustrated in

### 2023-01-16 body/part1/overview/extending-wca.tex
more writing yay

--- 2023-01-16 body/part1/overview/extending-wca.tex
@@
 This array included devices to capture four physiological measures:
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
+\begin{enumerate*}
+    [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
     \item \gls{GSR} (also known as electrodermal activity)
@@
     This value is obtained using the formula
-    \begin{equation}\label{eq:modelweights}
+    \begin{equation}
+        \label{eq:modelweights}
         w_{n - i} =
@@
 \begin{figure}
+    \begin{minipage}[t]{.45\textwidth}
         \centering
-    \includegraphics[height=13em]{Figs/2023EdgeDroid2/model_exectimes_transition}
+        \includegraphics[width=\textwidth]{Figs/2023EdgeDroid2/model_exectimes_transition}
         \caption{%
@@
         }\label{fig:model:exectimestransitions}
+    \end{minipage}%
+    \hfill%
+    \begin{minipage}[t]{.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/model_data/frame_probabilities}
+        \caption{%
+            Distribution of processing results for frames captured during the execution of steps in the LEGO task.
+            Note that \emph{Success} results are not included since by definition \ensuremath{P(\text{Success}|t_\text{norm}) = 1.0} for all normalized instants \ensuremath{t_\text{norm} \geq 1.0}.
+        }\label{fig:modelframes}
+    \end{minipage}
 \end{figure}
@@
 \subsubsection{Procedural generation of dynamic traces of video frames for \gls{WCA}}\label{model:traces}
+
+Our design for a procedure for the generation of synthetic yet realistic traces of video inputs for \gls{WCA} follows a stochastic approach.
+Analysis of the evolution of the distribution of processing results of frames over their normalized capture instant with respect to step execution time highlighted the pattern illustrated in \cref{fig:modelframes}.
+We define the normalized capture instant \ensuremath{t_\text{norm}} for each frame as the timestamp at which it was captured with respect to the beginning of the step, normalized by the total execution time of the same step.
+The distribution of the four possible categories of outputs of the \gls{CV} algorithm in the Gabriel LEGO Task~\cite{chen2018application} for each frame (\emph{blank}, i.e.\ no discernible board state, noise; \emph{low confidence}; \emph{success}, a frame which triggers a step transition;
+and \emph{repeat}, that is, no change since the last \emph{success} frame) were plotted against normalized capture instants.
+
+These results are then used together with a repository of tagged frames for each step to generate a synthetic trace at runtime by:
+\begin{enumerate}
+    \item At each sampling instant, calculate the current \ensuremath{t_\text{norm}} with respect to the next desired execution time.
+    \item If \ensuremath{t_\text{norm} >= 1}, send a pre-selected \emph{success}-tagged frame for the current step to the \gls{WCA} backend.
+    \item Otherwise, sample the distribution of result categories at the selected \ensuremath{t_\text{norm}} and return a pre-selected frame matching the selected category.
+\end{enumerate}
+
+This procedure results in a synthetic stream of frames which generates the same processing load on the backend as a real stream would.
+The use of a repository of frames collected from real executions of the task by humans ensures that the load on the network also approximates reality.
+More details and a longer discussion on this approach can be found in \cref{paper:olguinmunoz2023realistic}, \cref{ssec:model:frames}.

### 2023-01-17 body/part1/overview/extending-wca.tex
wrap

@@
-\section{Extending the methodology in the context of \gls{WCA}}
+\section{Extending the methodology in the context of \acs{WCA}}\label{sec:extendingmethodology}
 
@@
 In particular, the trait of \emph{neuroticism} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
-
-\todo[inline]{Discuss implications?}
 
@@
 
+\subsubsection{Implications}
+
+The results summarized above lead to interesting implications for the design and optimization of \gls{WCA} application systems.
+The behavioral slow-down in users as a response to reductions in system responsiveness translates into noticeably extended application lifetimes, even for short durations of impairment.
+Combined with the permanence of these effects for a period even after the system returns to an improved level of responsiveness, this could lead to unconventional consequences for resource allocation, in particular in multi-tenant deployment and in cases where a user may be able to finish the task before the effects of slow-down subside.
+In such cases, it might be more efficient to not divert resources to the impaired application, as fair degradation may not ultimately be beneficial to the system as a whole.
+It might even be more beneficial to divert resources \emph{away} from the impaired application, to ensure optimal system responsiveness for other tenants in the system.
+
+These implications, however, are merely speculative.
+Moreover and as previously discussed, due to the complexity of these systems, accurate study of the actual consequences of the results discussed in the present section can only be achieved through an experimental approach.
+\todo[inline]{Does this need a better transition?}
+
 \subsection{Extending the EdgeDroid tool with a realistic model of human behavior}
@@
 
-\todo[inline]{Probably needs to repeat a discussion on why the new model is necessary.
+\todo[inline]{Probably needs to repeat a discussion on why the new model is necessary, in a shorter summarized form.
 This discussion should first be presented at the beginning of this section however.}
@@
 
-\subsubsection{Procedural generation of dynamic traces of video frames for \gls{WCA}}\label{model:traces}
+\subsubsection{Procedural generation of dynamic traces of video frames for \acs{WCA}}\label{model:traces}
 

### 2023-01-17 body/part1/overview/implications-optimization.tex
wrap

@@
-\section{Implications for the optimization of \gls{WCA}}
+\section{Implications for the optimization of \acs{WCA}}
+
+In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
+The version of this discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
+
+\todo[inline]{Do we need more experiments?}
+
+\subsection{Implications with respect to application lifetime footprints}
+
+\subsection{Implications with respect to number of samples per step}
+
+\subsection{Implications with respect to raw energy consumption}

### 2023-01-19 body/part1/overview/implications-optimization.tex
duration

@@
 In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
-The version of this discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
+The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
 
@@
 
-\subsection{Implications with respect to application lifetime footprints}
+\subsection{Implications with respect to task completion times}
+
+The term \emph{task completion time}, in the context of \gls{WCA}, will refer to the time it takes a user to work their way through the task presented by an assistant application.
+In \cref{paper:olguinmunoz2023realistic} we refer interchangeably to task completion times as application lifetimes.
+This metric directly relates to system resource and energy consumption --- longer task completion times mean that the system is occupied for longer and thus uses more resources.
+It thus presents a relatively straightforward yet interesting starting point for the study of the implications of the model on the optimization of \gls{WCA} systems.
+
+The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
+Specifically, we aim to quantify the difference in task completion times when using the realistic compared to a baseline model which does not take into consideration higher order effects on execution times.
+Our reference baseline corresponds to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping whatsoever) which is randomly sampled to obtain execution time values.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=13em]{figs/2023EdgeDroid2/task_durations_diff}
+    \caption{Difference in task completion times of the realistic model with respect to the reference baseline.
+    Bars indicate \SI{95}{\percent} \glspl{CI}.}\label{fig:taskcompletiontimesdiff}
+\end{figure}
+
+In \cref{fig:taskcompletiontimesdiff} we present the results of an initial, simulated, exploration of the difference in task execution times generated by the model with respect to the baseline.
+A \num{100}-step task is simulated at five different, constant \glspl{TTF}; baseline total duration was, on average, \SI{834.87}{\second}.
 

### 2023-01-21 body/part1/overview/implications-optimization.tex
fix
writing

@@
 
-The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
-Specifically, we aim to quantify the difference in task completion times when using the realistic compared to a baseline model which does not take into consideration higher order effects on execution times.
-Our reference baseline corresponds to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping whatsoever) which is randomly sampled to obtain execution time values.
-
 \begin{figure}
     \centering
-    \includegraphics[height=13em]{figs/2023EdgeDroid2/task_durations_diff}
+    \includegraphics[height=13em]{Figs/2023EdgeDroid2/task_durations_diff}
     \caption{Difference in task completion times of the realistic model with respect to the reference baseline.
@@
 
-In \cref{fig:taskcompletiontimesdiff} we present the results of an initial, simulated, exploration of the difference in task execution times generated by the model with respect to the baseline.
-A \num{100}-step task is simulated at five different, constant \glspl{TTF}; baseline total duration was, on average, \SI{834.87}{\second}.
+The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
+Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
+In \cref{fig:taskcompletiontimesdiff} we present the results of an initial exploration of this difference, comparing the total task durations of simulated \num{100}-step tasks.
+Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping whatsoever) which is randomly sampled to obtain execution time values.
+
+These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low \glspl{TTF} for both levels of neuroticism.
+At higher \glspl{TTF}, the high neuroticism parameterization also results in task durations more than \SI{11}{\percent} longer than the baseline.
 

### 2023-01-22 body/part1/overview/implications-optimization.tex
more writing yay!

@@
     \caption{Difference in task completion times of the realistic model with respect to the reference baseline.
-    Bars indicate \SI{95}{\percent} \glspl{CI}.}\label{fig:taskcompletiontimesdiff}
+    Geometric averages over \num{10} over ten total repetitions per configuration.}\label{fig:taskcompletiontimesdiff}
 \end{figure}
@@
 Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
-In \cref{fig:taskcompletiontimesdiff} we present the results of an initial exploration of this difference, comparing the total task durations of simulated \num{100}-step tasks.
-Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping whatsoever) which is randomly sampled to obtain execution time values.
+In \cref{fig:taskcompletiontimesdiff} we present the results of an experimental exploration of this difference.
+We compare the total task durations of emulated \num{45}-step tasks using the EdgeDroid benchmarking tool with the improved model of human behavior to a baseline model.
+Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping) which is randomly sampled to obtain execution time values.
+We perform this experiment on the edge computing testbed discussed in \cref{sec:testbed};
+for simplicity and to magnify the effects of latency, in this setup we use an \acs{IEEE} \num{802.11}b/g physical layer.
 
-These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low \glspl{TTF} for both levels of neuroticism.
-At higher \glspl{TTF}, the high neuroticism parameterization also results in task durations more than \SI{11}{\percent} longer than the baseline.
+These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low levels of resource contention for both levels of neuroticism.
+At higher degrees of resource contention, the high neuroticism parameterization also results in task durations more than \SI{8}{\percent} longer than the baseline.
+
+\todo[inline]{Briefly discuss implications of the above?}
 
@@
 
-\subsection{Implications with respect to raw energy consumption}
+In the present section and the following we briefly discuss the implications of our methodology on the optimization potential of number of samples and energy consumption per step, respectively.
+Both these analyses depend on an understanding of the tradeoffs between resource consumption and responsiveness in \gls{WCA};
+fewer samples per step translates into longer sampling periods, and lower energy consumption tends to be made possible by processing fewer bits of data per step.
+As such, we employ in these sections an optimization framework for the optimization of said trade-offs based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
+We note that objective metrics relating to sampling and the responsiveness of a \gls{WCA} system will present itself as a linear combination of the number of expected samples and expected wait time per step plus some constant term:
+\begin{align}
+    \mathcal{E} = \alpha\mathbb{E}[\mathcal{S}] + \beta\mathbb{E}[\mathcal{W}] + C\label{eq:genericopt}
+\end{align}
+The above objective function is generic, and can be adapted to any metric which relates to the sampling and responsiveness of \gls{WCA} by modifying the factors \ensuremath{\alpha}, \ensuremath{\beta}, and \ensuremath{C}.
+
+For the analyses in this subsection and \cref{ssec:implications:energy}, we use the solution
+\begin{align}
+    t_n = \left( 3\sigma\sqrt {\frac{\alpha}{2\beta}} \right)^{\frac{2}{3}} n^{\frac{2}{3}}
+\end{align}
+where \ensuremath{t_n} once again corresponds to the \ensuremath{n}-th sampling instant in a step.
+Note that any positive real value is valid for \ensuremath{\frac{\alpha}{2\beta}}, and this ratio controls the optimization criteria --- as it goes to infinity, the criteria changes from minimizing wait time (i.e.\ maximizing responsiveness) to minimizing the number of samples per step (i.e.\ minimizing energy consumption).
+
+The full derivation and explanation of this framework is out of the scope of this summary.
+Please refer to \cref{paper:olguinmunoz2023realistic} and its appendix for a detailed discussion on this optimization approach.
+\todo{Include ref to section?}
+
+\subsection{Implications with respect to raw energy consumption}\label{ssec:implications:energy}

### 2023-01-23 body/part1/overview/implications-optimization.tex
fixing papers

@@
 \todo[inline]{Do we need more experiments?}
+\todo[inline]{Include discussions on the metrics here (grab inspiration from EdgeDroid2 paper.)}
 
@@
 where \ensuremath{t_n} once again corresponds to the \ensuremath{n}-th sampling instant in a step.
-Note that any positive real value is valid for \ensuremath{\frac{\alpha}{2\beta}}, and this ratio controls the optimization criteria --- as it goes to infinity, the criteria changes from minimizing wait time (i.e.\ maximizing responsiveness) to minimizing the number of samples per step (i.e.\ minimizing energy consumption).
+Note that any positive real value is valid for \ensuremath{\frac{\alpha}{2\beta}}.
+This ratio controls the optimization criteria --- as it goes to infinity, the criteria changes from minimizing wait time (i.e.\ maximizing responsiveness) to minimizing the number of samples per step (i.e.\ minimizing energy consumption).
 
@@
 
+\medskip
+
 \subsection{Implications with respect to raw energy consumption}\label{ssec:implications:energy}

### 2023-01-23 publications/2018DemoScalingOnTheEdge/paper_body.tex
fixing papers

@@
 
-Previous works on cloudlets~\cite{Satya2009Case,Ha2013Impact}, one of the earliest incarnation of edge computing, enable small data-centers at the edge of the Internet. 
+Previous works on cloudlets~\cite{satyanarayanan2009case,ha2013impact}, one of the earliest incarnation of edge computing, enable small data-centers at the edge of the Internet.
 Many futuristic applications become viable with these clusters that are only one wireless hop away. 
-One of the most promising genres of these emerging applications is human-in-the-loop applications such as wearable cognitive assistance~\cite{Ha2014Towards}.
+One of the most promising genres of these emerging applications is human-in-the-loop applications such as wearable cognitive assistance~\cite{ha2014towards}.
 
 In these applications, sensor data, for example video and audio, are continuously streamed to a cloudlet, where they are analyzed in realtime in order to assist users to complete a particular task. 
-Researchers have built prototypes of these applications to help users assemble LEGO models and IKEA furniture, and even learn how to play ping-pong~\cite{Satya2009Case,Chen2015Early}.
+Researchers have built prototypes of these applications to help users assemble LEGO models and IKEA furniture, and even learn how to play ping-pong~\cite{satyanarayanan2009case,chen2015early}.
 % In general, all human-in-the-loop applications capture in some form user and environment information and typically leverage compute-intensive algorithms to analyze the data in order to provide feedback to the user.
@@
 	\item[Latency Sensitive:] Given their tight interaction with the physical world, the quality of human-in-the-loop applications is determined by the latencies experienced by users. 
-    These applications are different from conventional mobile applications by the low latency requirements inherent to the applications themselves~\cite{Suzuki2016Vehicle,Chen2017Empirical}. 
+    These applications are different from conventional mobile applications by the low latency requirements inherent to the applications themselves~\cite{sasaki2016vehicle,chen2017empirical}.
     For example, consider a Ping-Pong Assistance application that instructs a user where to hit a ball -- any instruction delivered after the user has made a hit is useless.
@@
     \item[Compute Intensive:] Cognitive Assistance applications aim to enhance the cognitive capabilities of users, and are thus compute intensive as well due to widespread use of the state-of-art computer vision and machine learning algorithms, particularly \glspl{DNN}.
-    Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{Flinn2012Cyber}. 
+    Although mobile devices are becoming increasingly powerful, the gap between mobile and static elements continues to exist~\cite{flinn2012cyber}.
     While state-of-the-art DNN object detectors can run at more than 20 \gls{FPS} on a server \gls{GPU}, their performances are much worse on mobile \glspl{GPU} - some models cannot even be loaded due to memory constraints.
@@
 
-\section{The Measurement Framework}\label{sec:framework}
+\section{The Measurement Framework}\label{paper:olguinmunoz2018demo:sec:framework}
 
@@
 %         \captionof{figure}{General architecture of the benchmarking suite.}
-% 		\label{fig:TraceReplayArch}
+% 		\label{paper:olguinmunoz2018demo:fig:TraceReplayArch}
 % 	\end{minipage}
@@
 
-In order to imitate a human behavior as close as possible, we propose the following user model as shown in Figure~\ref{fig:usermodel}.
+In order to imitate a human behavior as close as possible, we propose the following user model as shown in \cref{paper:olguinmunoz2018demo:fig:usermodel}.
 We assume a user that is patient and does not make mistakes; any error message received from the application backend is ignored.
@@
     }
-    \caption{State diagram of the user model}
-    \label{fig:usermodel}
+    \caption{State diagram of the user model}\label{paper:olguinmunoz2018demo:fig:usermodel}
 \end{figure}
 
-%A state diagram for the behavior of this user can be seen in Figure \ref{fig:usermodel}.
+%A state diagram for the behavior of this user can be seen in \cref{paper:olguinmunoz2018demo:fig:usermodel}.
 
@@
 
-The suite has three elements, as shown in Figure~\ref{fig:TraceReplayArch}:
+The suite has three elements, as shown in \cref{paper:olguinmunoz2018demo:fig:TraceReplayArch}:
 % the \emph{application backend}, the \emph{client emulator}, and the \emph{control backend} or \emph{server}, which interact in the following ways:
@@
   \includegraphics[width=.7\columnwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
-  \captionof{figure}{General architecture of the benchmarking suite.}
-  \label{fig:TraceReplayArch}
+  \captionof{figure}{General architecture of the benchmarking suite.}\label{paper:olguinmunoz2018demo:fig:TraceReplayArch}
 \end{figure}
@@
 
-We will demonstrate the benchmarking framework using the \emph{LEGO Task Guidance} application previously developed by \textcite{Chen2015Early}.
+We will demonstrate the benchmarking framework using the \emph{LEGO Task Guidance} application previously developed by \textcite{chen2015early}.
 This application guides a user step by step in the assembly of a LEGO model; input consists exclusively of video feed of the current state of the assembly, whereas feedback includes visual and auditory components in the form of animations and speech, respectively.
 
-The benchmarking tool will be employed to extract key real-time metrics from this application, \emph{e.g.} average computation and network times, as well as the estimated \emph{user experience} level of the system as a whole (\emph{flawless}, \emph{impaired} or \emph{unusable}, based on the categorization in \cite{Chen2017Empirical}).
+The benchmarking tool will be employed to extract key real-time metrics from this application, \emph{e.g.} average computation and network times, as well as the estimated \emph{user experience} level of the system as a whole (\emph{flawless}, \emph{impaired} or \emph{unusable}, based on the categorization in~\cite{chen2017empirical}).
 
@@
 In addition, our current user model is simplistic. 
-This model could be expanded to emulate a human user more accurately and realistically, e.g. making mistakes and responding to feedback to correct them.
+This model could be expanded to emulate a human user more accurately and realistically, e.g.\ making mistakes and responding to feedback to correct them.
 

### 2023-01-23 publications/2019EdgeDroid/paper_body.tex
fixing

@@
-\section{Introduction}\label{sec:intro}
+\section{Introduction}\label{paper:olguinmunoz2019edgedroid:sec:intro}
 
-There is increasing interest from academia and industry in novel applications such as immersive augmented reality (AR) or wearable cognitive assistance (WCA)~\cite{Chatzopoulos:Hyperion,Ha:TowardsWearableCogAssist}, also known as human-in-the-loop applications.
+There is increasing interest from academia and industry in novel applications such as immersive \gls{AR} or \gls{WCA}~\cite{chatzopoulos2017hyperion,ha2014towards}, also known as human-in-the-loop applications.
 These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
-Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as Deep Neural Networks (DNN)~\cite{Ha:TowardsWearableCogAssist}.
-\Cref{fig:pipeline2} depicts such a system.
+Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as \glspl{DNN}~\cite{ha2014towards}.
+\Cref{paper:olguinmunoz2019edgedroid:fig:pipeline2} depicts such a system.
 These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
-Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{Chen:AnEmpiricalStudyOfLatency}.
+Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{chen2017empirical}.
 
 In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
-There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{Bittmann_Edge,flinn2012cyber,Chen:AnEmpiricalStudyOfLatency,Ha:JITProvisioning}.
+There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
 However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
@@
 How does this, together with other best-effort applications, impact the latencies perceived by the human user?
-Existing studies~\cite{Ha:TowardsWearableCogAssist, Chen:EarlyImplementation, satya2009case,Chatzopoulos:Hyperion} of this class of applications have only lightly touched upon these issues~\cite{Chen:AnEmpiricalStudyOfLatency}.
-On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{Zubaidy15,Schiessl17} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
+Existing studies~\cite{ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion} of this class of applications have only lightly touched upon these issues~\cite{chen2017empirical}.
+On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{al_zubaidy2015performance,schiessl2017finite} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
 We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
@@
 The rest of the paper is structured as follows:
-\Cref{sec:background} presents some background on human-in-the-loop applications.
-\Cref{sec:approach} discusses the general approach taken with EdgeDroid, while \Cref{sec:implementation} exposes the implementation details for EdgeDroid 1.0.
-We show the value of the tool through use case analysis in \Cref{sec:usecases,sec:results}, before discussing future work and concluding in \Cref{sec:conclusions}.
+\cref{paper:olguinmunoz2019edgedroid:sec:background} presents some background on human-in-the-loop applications.
+\cref{paper:olguinmunoz2019edgedroid:sec:approach} discusses the general approach taken with EdgeDroid, while \cref{paper:olguinmunoz2019edgedroid:sec:implementation} exposes the implementation details for EdgeDroid 1.0.
+We show the value of the tool through use case analysis in \Cref{paper:olguinmunoz2019edgedroid:sec:usecases,paper:olguinmunoz2019edgedroid:sec:results}, before discussing future work and concluding in \cref{paper:olguinmunoz2019edgedroid:sec:conclusions}.
 
 %\section{Design \& Implementation}\label{sec:design_impl}
-\section{Background}\label{sec:background}
+\section{Background}\label{paper:olguinmunoz2019edgedroid:sec:background}
 
@@
 
-Given the wide range of applications which fall under this concept, we have chosen to focus our initial efforts on one particular category: task-guidance wearable cognitive assistance (WCA)~\cite{Ha:TowardsWearableCogAssist}.
-We have chosen this type of application for two reasons: one, their relative simplicity in terms of execution flow, and two, the already well-established existing body of work~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation,Chen:AnEmpiricalStudyOfLatency}.
-Task-guidance WCA applications aim to guide users in the execution of a task by monitoring user actions and providing real-time instructions, usually through wearable sensors and gadgets such as the Google Glass platform.
+Given the wide range of applications which fall under this concept, we have chosen to focus our initial efforts on one particular category: task-guidance \gls{WCA}~\cite{ha2014towards}.
+We have chosen this type of application for two reasons: one, their relative simplicity in terms of execution flow, and two, the already well-established existing body of work~\cite{ha2014towards,chen2015early,chen2017empirical}.
+Task-guidance \gls{WCA} applications aim to guide users in the execution of a task by monitoring user actions and providing real-time instructions, usually through wearable sensors and gadgets such as the Google Glass platform.
 They have many potential use cases including training and assistance for professionals performing complex tasks.
 Imagine for instance a specialized technician performing field repairs on a complex piece of machinery.
-A task-guidance WCA application could offer real-time guidance in this task, by analyzing a real-time video feed from the technician's head-mounted camera and providing step-by-step repair instructions.
+A task-guidance \gls{WCA} application could offer real-time guidance in this task, by analyzing a real-time video feed from the technician's head-mounted camera and providing step-by-step repair instructions.
 
-Broadly speaking, task-guidance WCA applications (and human-in-the-loop in general) process a multitude of inputs in parallel, which are also usually continuous, multidimensional and time-sensitive, e.g.\ video and audio feeds.
-These inputs are passed to the computation backend, where they are processed in the pipeline depicted in \cref{fig:pipeline2}.
+Broadly speaking, task-guidance \gls{WCA} applications (and human-in-the-loop in general) process a multitude of inputs in parallel, which are also usually continuous, multidimensional and time-sensitive, e.g.\ video and audio feeds.
+These inputs are passed to the computation backend, where they are processed in the pipeline depicted in \cref{paper:olguinmunoz2019edgedroid:fig:pipeline2}.
 The first step of the backend processing is detection, which acts as a filter for irrelevant inputs.
@@
     \input{publications/2019EdgeDroid/img/pipeline.tex}
-    \caption{Pipeline of human-in-the-loop applications.}\label{fig:pipeline2}
+    \caption{Pipeline of human-in-the-loop applications.}\label{paper:olguinmunoz2019edgedroid:fig:pipeline2}
 \end{figure}
 
-The task model of a task-guidance WCA is a parameterized representation of the task in question and the steps required to complete it.
+The task model of a task-guidance \gls{WCA} is a parameterized representation of the task in question and the steps required to complete it.
 It can be represented as a Finite State Machine (FSM) \( M\{S, E\} \), where \(S\) is a set of states and \(E\) is a set of edges connecting these states.
@@
     \input{publications/2019EdgeDroid/img/taskmodel.tex}
-    \captionof{figure}{Segment of the internal task model of a linear task-guidance WCA.}%
-    \label{fig:taskmodel}
+    \captionof{figure}{Segment of the internal task model of a linear task-guidance \acs{WCA}.}%
+    \label{paper:olguinmunoz2019edgedroid:fig:taskmodel}
 \end{figure}
 
-These formalisms are exemplified in \cref{fig:taskmodel}, which represents an arbitrary segment of a linear task-guidance WCA application.
+These formalisms are exemplified in \cref{paper:olguinmunoz2019edgedroid:fig:taskmodel}, which represents an arbitrary segment of a linear task-guidance \gls{WCA} application.
 \(s_{i}, s_{j} \in S\) represent sequential states in the execution of the task, with the edge \(e_{i,j}\) symbolizing the unique correct way of transitioning between them.
@@
 
-\section{Approach of EdgeDroid}\label{sec:approach}
+\section{Approach of EdgeDroid}\label{paper:olguinmunoz2019edgedroid:sec:approach}
 
@@
 
-\section{Implementation of EdgeDroid}\label{sec:implementation}
+\section{Implementation of EdgeDroid}\label{paper:olguinmunoz2019edgedroid:sec:implementation}
 
@@
     \includegraphics[width=.7\columnwidth]{publications/2019EdgeDroid/img/TraceReplay_GenArch}
-    \caption{Architecture of the benchmarking suite.}\label{fig:TraceReplayArch}
+    \caption{Architecture of the benchmarking suite.}\label{paper:olguinmunoz2019edgedroid:fig:TraceReplayArch}
 \end{figure}
@@
 
-The system architecture is composed of a \emph{cloudlet}, where one or more instances of the target application run, and one or more client devices, as pictured in \cref{fig:TraceReplayArch}.
+The system architecture is composed of a \emph{cloudlet}, where one or more instances of the target application run, and one or more client devices, as pictured in \cref{paper:olguinmunoz2019edgedroid:fig:TraceReplayArch}.
 These correspond to completely unaltered instances of the target application, running inside containers~\cite{docker}.
@@
 The emulators achieve this by mimicking the operation of real clients, but instead of obtaining the input data from the on-board sensors, they extract it from the previously recorded trace.
-They replay this data over the network to the \emph{application instances} employing, as mentioned in \Cref{sec:background}, a \emph{user model}, while simultaneously collecting client-side statistics and measurements.
+They replay this data over the network to the \emph{application instances} employing, as mentioned in \cref{paper:olguinmunoz2019edgedroid:sec:background}, a \emph{user model}, while simultaneously collecting client-side statistics and measurements.
 
@@
 
-For our initial iteration of the EdgeDroid 1.0 benchmarking suite, we implement a preliminary user model, depicted in \Cref{fig:usermodel}, to run on the \emph{client emulators}, designed with a linear task model in mind.
+For our initial iteration of the EdgeDroid 1.0 benchmarking suite, we implement a preliminary user model, depicted in \cref{paper:olguinmunoz2019edgedroid:fig:usermodel}, to run on the \emph{client emulators}, designed with a linear task model in mind.
 In EdgeDroid 1.0, our model is that of a user who is totally impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
@@
     \input{publications/2019EdgeDroid/img/usermodel.tex}
-    \captionof{figure}{Preliminary user model.}\label{fig:usermodel}
+    \captionof{figure}{Preliminary user model.}\label{paper:olguinmunoz2019edgedroid:fig:usermodel}
 \end{figure}
 
-\section{Use Cases}\label{sec:usecases}
+\section{Use Cases}\label{paper:olguinmunoz2019edgedroid:sec:usecases}
 
@@
     \caption{Hardware used in the experiments.}%
-    \label{tbl:cloudletclienthardware}%
+    \label{paper:olguinmunoz2019edgedroid:tbl:cloudletclienthardware}%
     \begin{tabular}{@{}l@{\qquad}lclcl@{}}
@@
 
-We chose the open source \emph{Gabriel} platform~\cite{Ha:TowardsWearableCogAssist} running the LEGO assistance application~\cite{Chen:AnEmpiricalStudyOfLatency} for our experimentation.
+We chose the open source \emph{Gabriel} platform~\cite{ha2014towards} running the LEGO assistance application~\cite{chen2017empirical} for our experimentation.
 This application guides a user through the assembly of a 2-dimensional LEGO model with visual and auditory instructions.
@@
 We chose it due to its maturity and stability compared to other applications which run on the \emph{Gabriel} platform, as well as the relative simplicity of the task it performs.
-The application employs a straightforward, linear task model as the one described in \Cref{sec:background}.
+The application employs a straightforward, linear task model as the one described in \cref{paper:olguinmunoz2019edgedroid:sec:background}.
 Each state only allows for one specific correct action by the user, and any other action triggers negative feedback until the application is reset to the previous state.
@@
 
-\Cref{tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
+\cref{paper:olguinmunoz2019edgedroid:tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
 The components communicated through a single consumer-grade WiFi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
@@
 
-\section{Results}\label{sec:results}
+\section{Results}\label{paper:olguinmunoz2019edgedroid:sec:results}
 
@@
     \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
-    \label{fig:comparison:rtt}%
+    \label{paper:olguinmunoz2019edgedroid:fig:comparison:rtt}%
 \end{figure}%
@@
     \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
-    \label{fig:comparison:feedback}%
+    \label{paper:olguinmunoz2019edgedroid:fig:comparison:feedback}%
 \end{figure}%
@@
     \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
-    \label{fig:comparison:tasksteps}
+    \label{paper:olguinmunoz2019edgedroid:fig:comparison:tasksteps}
 \end{figure}%
@@
 
-\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
-Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
+\cref{paper:olguinmunoz2019edgedroid:fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
+Next, \cref{paper:olguinmunoz2019edgedroid:fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
 We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
-Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
+Finally, \cref{paper:olguinmunoz2019edgedroid:fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
 These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
@@
 
-For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
+For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{chen2017empirical}.
 This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.
 
-We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
+We will begin our analysis of the experiment results with \cref{paper:olguinmunoz2019edgedroid:fig:comparison:rtt}.
 These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
@@
 
-We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
+We can identify a pair of interesting effects in the scaling of the task-guidance \gls{WCA} application.
 One, scaling behavior for the application seems to be linear with respect to the number of clients.
@@
 
-EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
+EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{paper:olguinmunoz2019edgedroid:fig:comparison:feedback}.
 From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
@@
 
-Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
+Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance \gls{WCA}. Consider \cref{paper:olguinmunoz2019edgedroid:fig:comparison:tasksteps}.
 The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.
 
-\section{Conclusions \& Future Work}\label{sec:conclusions}
+\section{Conclusions \& Future Work}\label{paper:olguinmunoz2019edgedroid:sec:conclusions}
 
@@
 
-The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with WCA applications in order to present a more complete model in the future.
-As mentioned in \Cref{sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
+The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with \gls{WCA} applications in order to present a more complete model in the future.
+As mentioned in \cref{paper:olguinmunoz2019edgedroid:sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
 Rather, EdgeDroid 1.0 models a perfectly stoic user who is like an automaton and responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
@@
 We are also working on expanding the benchmarking suite to also work first with other types of Wearable Cognitive Assistance, and later with other categories of human-in-the-loop applications.
-Other types of WCA we will consider in future iterations of the tool include real-time task-assistance WCA applications (such as the Ping-Pong application described in~\cite{Chen:EarlyImplementation}), which don't have a linear task model like task-guidance WCA and have tighter latency bounds and context- and information-providing WCA applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
+Other types of \glspl{WCA} we will consider in future iterations of the tool include real-time task-assistance \gls{WCA} applications (such as the Ping-Pong application described in~\cite{chen2017empirical}), which don't have a linear task model like task-guidance \gls{WCA} and have tighter latency bounds and context- and information-providing \gls{WCA} applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
 The latter also do not have a linear task model, but present more lax latency bounds.
@@
     We also thank our shepherd, Wenjun Hu, and the anonymous reviewers for helping us improve the paper.
-    This research was supported in part by the National Science Foundation (NSF), grant number CNS-1518865, the VINNOVA grant MERIT (2017--05232).
+    This research was supported in part by the \gls{NSF}, grant number \verb|CNS-1518865|, the VINNOVA grant \verb|MERIT (2017--05232)|.
     Additional support was provided by Intel,\ Vodafone,\ Deutsche~Telekom,\ Verizon,\ Crown~Castle,\ NTT,\ and the Conklin~Kistler family fund.

### 2023-01-23 publications/2021ImpactDelayedResponse/body.tex
fixing paper c

@@
 \begin{abstract}
-  Wearable Cognitive Assistants (WCA) are anticipated to become a widely-used application class, in conjunction with emerging network infrastructures like 5G that incorporate edge computing capabilities.
-  While prototypical studies of such applications exist today, the relationship between infrastructure service provisioning and its implication for WCA usability is largely unexplored despite the relevance that these applications have for future networks.
-  This paper presents an experimental study assessing how WCA users react to varying end-to-end delays induced by the application pipeline or infrastructure.
-  Participants interacted directly with an instrumented task-guidance WCA as delays were introduced into the system in a controllable fashion.
+    \glsunset{WCA}
+  Wearable Cognitive Assistants (\gls{WCA}) are anticipated to become a widely-used application class, in conjunction with emerging network infrastructures like 5G that incorporate edge computing capabilities.
+  While prototypical studies of such applications exist today, the relationship between infrastructure service provisioning and its implication for \gls{WCA} usability is largely unexplored despite the relevance that these applications have for future networks.
+  This paper presents an experimental study assessing how \gls{WCA} users react to varying end-to-end delays induced by the application pipeline or infrastructure.
+  Participants interacted directly with an instrumented task-guidance \gls{WCA} as delays were introduced into the system in a controllable fashion.
   System and task state were tracked in real time, and biometric data from wearable sensors on the participants were recorded.
@@
   We show that our results implicate impaired cognitive planning, as contrasted with resource depletion or emotional arousal, as the reason for slowed user task executions under system delay.
-  The findings have several implications for the design and operation of WCA applications as well as computational and communication infrastructure, and additionally for the development of performance analysis tools for WCA.\@
+  The findings have several implications for the design and operation of \gls{WCA} applications as well as computational and communication infrastructure, and additionally for the development of performance analysis tools for \gls{WCA}.\@
 \end{abstract}
@@
 
-Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
+Wearable Cognitive Assistants, \gls{WCA} for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
 They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
@@
 
-Based on these use cases, we identify three main requirements for WCA:\@
+Based on these use cases, we identify three main requirements for \gls{WCA}:\@
 \begin{enumerate}
-    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
+    \item\label{item:pervasive} \gls{WCA} systems should be available whenever the user requires them, without being tethered to a particular physical location.
+Assistants need to be pervasive and mobile.
 
     \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
-    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+    In this sense, \gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
 
@@
 However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
-\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.
+\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
 
@@
 Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
-A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
+A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of \gls{WCA}.\@
 % Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.
 
-This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
-We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
+This paper builds upon preliminary work in the field of time perception and delay characterization of \gls{WCA}.\@
+We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in \gls{WCA}, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
 While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
@@
 
-We present in this paper the design and elaboration of an experimental WCA test-bed.
+We present in this paper the design and elaboration of an experimental \gls{WCA} test-bed.
 This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
@@
 
-  The results point in a different direction though, indicating that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
+  The results point in a different direction though, indicating that reduced responsiveness in \gls{WCA} systems leads to a disruption of participants' cognitive plan for the task and not to an emotional response.
   This is evidenced by the previously discussed pacing effect and the lack of significant physiological responses.
@@
 
-% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
+% Our results indicate that reduced responsiveness in \gls{WCA} systems leads to a disruption of participants' cognitive plan for the task.
 % This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
@@
 
-We believe that these results provide concrete and relevant implications for WCA design, deployment, and optimization.
+We believe that these results provide concrete and relevant implications for \gls{WCA} design, deployment, and optimization.
 One example is the behavioral slow-down, as it extends application runtime significantly, and thus has clear and direct implications for resource and power consumption.
@@
 
-Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
+Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of \gls{WCA}.\@
 These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
 Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
-Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@
+Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of \gls{WCA}.\@
 
@@
 Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-In immersive systems like WCA, which aim to provide seamless interaction, delays are particularly noticeable.
+In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
 
@@
 However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.
+Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
 
-For WCA the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
 They then proposed a latency tolerance zone according to the task demands.
@@
 
-Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
+Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
 Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
@@
 Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of WCA, which are intended for relatively inexperienced users of an application.
+Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
 These and other measures of individual variation were considered here.
@@
 \begin{itemize}
-    \item Subjects interact with a WCA while wearing an array of biometric sensors.
+    \item Subjects interact with a \gls{WCA} while wearing an array of biometric sensors.
     \item The \emph{responsiveness} of the application, i.e.\ the interval of time between an input being provided to the system and the associated output returned to the user, is manipulated in real time.
@@
 We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{Chen:EarlyImplementation}.
-This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
+This application belongs to a category of \gls{WCA}s designed to guide users through the execution of a sequential task.
 % Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
@@
 % In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
-In the following, we will provide some formal definitions relating to the WCA application.
+In the following, we will provide some formal definitions relating to the \gls{WCA} application.
 % \theoremstyle{definition}
@@
 This fact could have practical implications in the future.
-It could, for instance, provide a tunable parameter for eventual models aiming to emulate human interaction with a WCA.\@
+It could, for instance, provide a tunable parameter for eventual models aiming to emulate human interaction with a \gls{WCA}.\@
 In addition, physiological measures of heart rate and EEG appear not to be direct indicators of  behavioral response to delay, although GSR may be more promising in this regard.
@@
 
-As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
+As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance \gls{WCA}.\@
 However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.
 
-The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.
+The results here presented provide a number of possible implications for \gls{WCA} system design and optimization, both for single and multi-application flows.
 
@@
 
-To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of WCA deployments.
+To wrap up, we believe the present data provide novel and unexpected insights for the understanding and optimizing of \gls{WCA} deployments.
 Although more subtle than expected, and in some cases somewhat counterintuitive, these insights represent a valuable tool to tackle inefficiencies in these systems.
@@
 More research in this area will surely uncover more complex and interesting behaviors.
-Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of WCA that might be constructed under existing architectures like ACT-R.
+Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of \gls{WCA} that might be constructed under existing architectures like ACT-R.
 These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
@@
 
-In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
+In this paper, we presented the results of a study on the physiological and behavioral reactions of users of \gls{WCA} to delays in the application pipeline.
 Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.
 
-We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
-The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
+We approached this in an experimental manner, by having participants interact with an instrumented \gls{WCA} setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
+The results show that user interactions in a \gls{WCA} slow down in the presence of delays in the application pipeline.
 When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
@@
 
-These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
+These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of \gls{WCA} systems.
 In this context, we believe there are two direct and important next steps to be performed.
@@
 
-The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
-These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
+The second step corresponds to the extension of existing tools for \gls{WCA} benchmarking with the findings presented in this paper.
+These tools are of great usefulness for the study of \gls{WCA} systems, as they allow for automated large-scale testing without having to resort to human users.
 However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
@@
 Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
-Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.
+Extending and perfecting these tools will allow for much more realistic benchmarking and testing of \gls{WCA} systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.
 

### 2023-01-23 publications/2021ImpactDelayedResponse/body.tex
still fixing paper c

@@
 \cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
+The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
 Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
@@
 The structure of this paper is as follows.
-We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
+We describe the existing body of research around time perception and the effects of delay on human performance in \cref{paper:olguinmunoz2021impact:sec:background}.
 \cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
 Then, in \cref{sec:results} we detail the results of our experiment.
-Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
+Implications for further modeling of the effects of delay are presented in \cref{paper:olguinmunoz2021impact:sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
 
-\section{Background and Related Work}\label{sec:background}
+\section{Background and Related Work}\label{paper:olguinmunoz2021impact:sec:background}
 
@@
 
-\subsection{Time perception in computing systems}\label{ssec:timeperception}
+\subsection{Time perception in computing systems}\label{paper:olguinmunoz2021impact:ssec:timeperception}
 
@@
 
-\subsection{Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}
+\subsection{Mechanisms Relating Delay to Human Performance}\label{paper:olguinmunoz2021impact:ssec:potentialmechs}
 
@@
 
-\section{Experimental Design}\label{sec:experimentaldesign}
+\section{Experimental Design}\label{paper:olguinmunoz2021impact:sec:experimentaldesign}
 
-The core elements of our experiment are shown in \cref{fig:experimentaltestbed}:
+The core elements of our experiment are shown in \cref{paper:olguinmunoz2021impact:fig:experimentaltestbed}:
 \begin{itemize}
@@
   The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
-  \label{fig:experimentaltestbed}
+  \label{paper:olguinmunoz2021impact:fig:experimentaltestbed}
 \end{figure}
@@
 A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
-\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{fig:task:steps}).
+\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g.
+\emph{put a green \(1 \times 3\) brick on the board} (see \cref{paper:olguinmunoz2021impact:fig:task:steps}).
 % We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
@@
   \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig2.eps}
-  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
+  \caption{Example of a Cognitive Assistance task and its component steps.}\label{paper:olguinmunoz2021impact:fig:task:steps}
 \end{figure*}
@@
   \caption{Hierarchical cognitive structure of a step in the LEGO task.}
-  \label{fig:lego:hierarchical}
+  \label{paper:olguinmunoz2021impact:fig:lego:hierarchical}
 \end{figure*}
@@
 For the experimental LEGO task, a key modification was made to the structure of the steps.
-After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
+After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{paper:olguinmunoz2021impact:fig:cogassist:step,fig:cogassist:step:delay}.
 The length of this delay is one of two independent variables we manipulated for the experiment.
@@
 
-In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
+In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{paper:olguinmunoz2021impact:fig:cogassist:block}.
 % \begin{definition}[Block]
-%   A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
+%   A sequence of consecutive steps within a task subject to the same delay; see \cref{paper:olguinmunoz2021impact:fig:cogassist:block}.
 The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
@@
 
-For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{fig:exectime:diagram}).
+For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{paper:olguinmunoz2021impact:fig:exectime:diagram}).
 \( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
@@
 %   \caption{Visualization of the execution time of a step.}%
-%   \label{fig:exectime:diagram}
+%   \label{paper:olguinmunoz2021impact:fig:exectime:diagram}
 % \end{figure}
@@
     }
-    \label{fig:cogassist:step}
+    \label{paper:olguinmunoz2021impact:fig:cogassist:step}
   \end{subfigure}%
@@
     \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
-    \label{fig:cogassist:step:delay}
+    \label{paper:olguinmunoz2021impact:fig:cogassist:step:delay}
   \end{subfigure}
@@
     \caption{Structure of a block in the experimental task.}%
-    \label{fig:cogassist:block}
+    \label{paper:olguinmunoz2021impact:fig:cogassist:block}
   \end{subfigure}%
@@
     \caption{Visualization of the execution time of a step.}%
-    \label{fig:exectime:diagram}
+    \label{paper:olguinmunoz2021impact:fig:exectime:diagram}
   \end{subfigure}%
@@
   \centering
-  \caption{Means and standard deviations of normalized questionnaire scores.}\label{tab:qscores}
+  \caption{Means and standard deviations of normalized questionnaire scores.}\label{paper:olguinmunoz2021impact:tab:qscores}
   \setlength{\tabcolsep}{0pt} 
@@
 Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
-Examples of the captured frames can be seen in \cref{fig:sampleframes}.
+Examples of the captured frames can be seen in \cref{paper:olguinmunoz2021impact:fig:sampleframes}.
 The video feeds were not used for the present study, but may be utilized in future analysis.
@@
   \end{subfigure}
-  \caption{Sample frames captured during the experiment.}\label{fig:sampleframes}
+  \caption{Sample frames captured during the experiment.}\label{paper:olguinmunoz2021impact:fig:sampleframes}
 \end{figure}
 
-\section{Results}\label{sec:results}
+\section{Results}\label{paper:olguinmunoz2021impact:sec:results}
 
@@
 
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
+We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
 % An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
@@
 % Conventionally \( p < .05 \) is the criterion for significance.
-% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
+% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
 
@@
   \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{tab:anova:exectime}
+  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:exectime}
   \setlength{\tabcolsep}{0pt}%
@@
 Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
-An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{tab:anova:exectime:slice}.
+An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{paper:olguinmunoz2021impact:tab:anova:exectime:slice}.
 As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
@@
   \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{tab:anova:exectime:slice}
+  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{paper:olguinmunoz2021impact:tab:anova:exectime:slice}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     Error bars indicate S.E.M.}
-    \label{fig:exectime:delay:slice}
+    \label{paper:olguinmunoz2021impact:fig:exectime:delay:slice}
 \end{figure}
@@
 \end{enumerate*}
-The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
+The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
 We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).
@@
   \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
-  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
+  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \).
+Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:exectime:transition}%
 \end{figure}
@@
 We normalized the sum of the accelerations by the duration of the step to correct for differences in delay and the execution time, which tends to increase with delay as described above.
-An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{tab:anova:acc} and \cref{fig:acc:delaylength}.
+An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{paper:olguinmunoz2021impact:tab:anova:acc} and \cref{fig:acc:delaylength}.
 There was also a significant delay by length interaction, reflecting that delay effects occurred particularly for the longer lengths.
@@
   \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and block length.}\label{tab:anova:acc}
+  \caption{Significant effects on accelerometer data from ANOVA on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:acc}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig9.eps}
-    \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{fig:acc:delaylength}
+    \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:acc:delaylength}
 \end{figure}
@@
   \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
-  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
+  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:acc:delayslice}
 \end{figure}
@@
   \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and slice number.}\label{tab:anova:acc:slice}
+  \caption{Significant effects on accelerometer data from ANOVA on factors delay and slice number.}\label{paper:olguinmunoz2021impact:tab:anova:acc:slice}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
 Again the analysis dividing blocks into 4-step slices was conducted, with delay and slice number as factors.  
-For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{tab:anova:egg:alphabeta}.
+For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}.
 Both bands showed the same tendency:  EEG declined as a sequence of steps with the same delay progressed.
-These effects are shown in \cref{fig:logeegpow:slices}.
+These effects are shown in \cref{paper:olguinmunoz2021impact:fig:logeegpow:slices}.
 Thus, EEG taken from frontal locations mimics the execution time data in showing a decline over the course of a block, but unlike the execution time, there was no tendency for the decline in EEG to be reduced at longer delays.
@@
   \centering
-  \caption{Significant effects on log EEG power from ANOVA on factors delay and block slice.}\label{tab:anova:egg:alphabeta}
+  \caption{Significant effects on log EEG power from ANOVA on factors delay and block slice.}\label{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig11.eps}
-    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{fig:logeegpow:slices}
+    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:logeegpow:slices}
 \end{figure}
@@
 In section \cref{ssec:potentialmechs} we proposed that emotional arousal during the execution of the task was a potential explanation for the effects observed in the human.
-However, these results seem to refute this hypothesis, and will be further discussed in \cref{sec:discussion}.
+However, these results seem to refute this hypothesis, and will be further discussed in \cref{paper:olguinmunoz2021impact:sec:discussion}.
 
@@
   \centering
-  \caption{Principal Component Analysis}\label{tab:pca}
+  \caption{Principal Component Analysis}\label{paper:olguinmunoz2021impact:tab:pca}
   \begin{subtable}[h]{\textwidth}
@@
     Pearson correlation coefficient \( r = .40 \); 2-tailed \( p = .01 \).}
-    \label{fig:neuro:exectime:reg}
+    \label{paper:olguinmunoz2021impact:fig:neuro:exectime:reg}
 \end{figure}
@@
 
-\section{Discussion}\label{sec:discussion}
+\section{Discussion}\label{paper:olguinmunoz2021impact:sec:discussion}
 
@@
   \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
-  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
+  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
   The question of whether any lingering effect can be measured after these four steps remains open.
@@
 
-\section{Conclusion}\label{sec:conclusion}
+\section{Conclusion}\label{paper:olguinmunoz2021impact:sec:conclusion}
 
@@
 In this context, we believe there are two direct and important next steps to be performed.
-The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
+The first of these relates to the implications for system optimization and resource allocation discussed in \cref{paper:olguinmunoz2021impact:sec:discussion}.
 We believe these need to be implemented, tested and validated in real setups.

### 2023-01-24 body/part1/overview/extending-wca.tex
fixing papers
restructuring for easier following of the red thread

@@
-\section{Extending the methodology in the context of \acs{WCA}}\label{sec:extendingmethodology}
+\section{Implications of the methodology in the context of \acs{WCA}}\label{sec:extendingmethodology}
+
+\todo[inline]{%
+    Rework this section.
+    First, discuss that we wish a more detailed study on the implications of the methodology for the optimization of WCA.
+    However, to achieve that we must first improve upon our model of human behavior.
+    This leads to us first characterizing human behavior, to then building an improved model.
+}
 
@@
 
-\subsection{Extending the EdgeDroid tool with a realistic model of human behavior}
+\subsection{A new model of human behavior in \acs{WCA}}
 
@@
     \item At each sampling instant, calculate the current \ensuremath{t_\text{norm}} with respect to the next desired execution time.
-    \item If \ensuremath{t_\text{norm} >= 1}, send a pre-selected \emph{success}-tagged frame for the current step to the \gls{WCA} backend.
+    \item If \ensuremath{t_\text{norm} \geq 1}, send a pre-selected \emph{success}-tagged frame for the current step to the \gls{WCA} backend.
     \item Otherwise, sample the distribution of result categories at the selected \ensuremath{t_\text{norm}} and return a pre-selected frame matching the selected category.
@@
 More details and a longer discussion on this approach can be found in \cref{paper:olguinmunoz2023realistic}, \cref{ssec:model:frames}.
+
+\subsection{Implications with respect to task completion times}
+
+In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
+The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
+
+\todo[inline]{Do we need more experiments?}
+\todo[inline]{Include discussions on the metrics here (grab inspiration from EdgeDroid2 paper.)}
+
+The term \emph{task completion time}, in the context of \gls{WCA}, will refer to the time it takes a user to work their way through the task presented by an assistant application.
+In \cref{paper:olguinmunoz2023realistic} we refer interchangeably to task completion times as application lifetimes.
+This metric directly relates to system resource and energy consumption --- longer task completion times mean that the system is occupied for longer and thus uses more resources.
+It thus presents a relatively straightforward yet interesting starting point for the study of the implications of the model on the optimization of \gls{WCA} systems.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=13em]{Figs/2023EdgeDroid2/task_durations_diff}
+    \caption{Difference in task completion times of the realistic model with respect to the reference baseline.
+    Geometric averages over \num{10} over ten total repetitions per configuration.}\label{fig:taskcompletiontimesdiff}
+\end{figure}
+
+The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
+Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
+In \cref{fig:taskcompletiontimesdiff} we present the results of an experimental exploration of this difference.
+We compare the total task durations of emulated \num{45}-step tasks using the EdgeDroid benchmarking tool with the improved model of human behavior to a baseline model.
+Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping) which is randomly sampled to obtain execution time values.
+We perform this experiment on the edge computing testbed discussed in \cref{sec:testbed};
+for simplicity and to magnify the effects of latency, in this setup we use an \acs{IEEE} \num{802.11}b/g physical layer.
+
+These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low levels of resource contention for both levels of neuroticism.
+At higher degrees of resource contention, the high neuroticism parameterization also results in task durations more than \SI{8}{\percent} longer than the baseline.
+
+\todo[inline]{Briefly discuss implications of the above?}
+
+\subsection{Implications with respect to number of samples per step}
+
+In the present section and the following we briefly discuss the implications of our methodology on the optimization potential of number of samples and energy consumption per step, respectively.
+Both these analyses depend on an understanding of the tradeoffs between resource consumption and responsiveness in \gls{WCA};
+fewer samples per step translates into longer sampling periods, and lower energy consumption tends to be made possible by processing fewer bits of data per step.
+As such, we employ in these sections an optimization framework for the optimization of said trade-offs based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
+We note that objective metrics relating to sampling and the responsiveness of a \gls{WCA} system will present itself as a linear combination of the number of expected samples and expected wait time per step plus some constant term:
+\begin{align}
+    \mathcal{E} = \alpha\mathbb{E}[\mathcal{S}] + \beta\mathbb{E}[\mathcal{W}] + C\label{eq:genericopt}
+\end{align}
+The above objective function is generic, and can be adapted to any metric which relates to the sampling and responsiveness of \gls{WCA} by modifying the factors \ensuremath{\alpha}, \ensuremath{\beta}, and \ensuremath{C}.
+
+For the analyses in this subsection and \cref{ssec:implications:energy}, we use the solution
+\begin{align}
+    t_n = \left( 3\sigma\sqrt {\frac{\alpha}{2\beta}} \right)^{\frac{2}{3}} n^{\frac{2}{3}}
+\end{align}
+where \ensuremath{t_n} once again corresponds to the \ensuremath{n}-th sampling instant in a step.
+Note that any positive real value is valid for \ensuremath{\frac{\alpha}{2\beta}}.
+This ratio controls the optimization criteria --- as it goes to infinity, the criteria changes from minimizing wait time (i.e.\ maximizing responsiveness) to minimizing the number of samples per step (i.e.\ minimizing energy consumption).
+
+The full derivation and explanation of this framework is out of the scope of this summary.
+Please refer to \cref{paper:olguinmunoz2023realistic} and its appendix for a detailed discussion on this optimization approach.
+\todo{Include ref to section?}
+
+\medskip
+
+\subsection{Implications with respect to raw energy consumption}\label{ssec:implications:energy}

### 2023-01-24 body/part1/overview/implications-optimization.tex
restructuring for easier following of the red thread

--- 2023-01-21 body/part1/overview/implications-optimization.tex
@@
 \section{Implications for the optimization of \acs{WCA}}
-
-In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
-The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
-
-\todo[inline]{Do we need more experiments?}
-
-\subsection{Implications with respect to task completion times}
-
-The term \emph{task completion time}, in the context of \gls{WCA}, will refer to the time it takes a user to work their way through the task presented by an assistant application.
-In \cref{paper:olguinmunoz2023realistic} we refer interchangeably to task completion times as application lifetimes.
-This metric directly relates to system resource and energy consumption --- longer task completion times mean that the system is occupied for longer and thus uses more resources.
-It thus presents a relatively straightforward yet interesting starting point for the study of the implications of the model on the optimization of \gls{WCA} systems.
-
-\begin{figure}
-    \centering
-    \includegraphics[height=13em]{Figs/2023EdgeDroid2/task_durations_diff}
-    \caption{Difference in task completion times of the realistic model with respect to the reference baseline.
-    Bars indicate \SI{95}{\percent} \glspl{CI}.}\label{fig:taskcompletiontimesdiff}
-\end{figure}
-
-The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
-Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
-In \cref{fig:taskcompletiontimesdiff} we present the results of an initial exploration of this difference, comparing the total task durations of simulated \num{100}-step tasks.
-Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping whatsoever) which is randomly sampled to obtain execution time values.
-
-These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low \glspl{TTF} for both levels of neuroticism.
-At higher \glspl{TTF}, the high neuroticism parameterization also results in task durations more than \SI{11}{\percent} longer than the baseline.
-
-\subsection{Implications with respect to number of samples per step}
-
-\subsection{Implications with respect to raw energy consumption}

### 2023-01-24 body/part1/overview/methodology.tex
wrap
more writing yay!
restructuring for easier following of the red thread

@@
 \section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}
+
+\todo[inline]{%
+    Discuss the methodology in general terms.
+    Avoid going too deeply into WCA or NCSs here, just talk about the advantages of our approach.
+    Refer to first two EdgeDroid papers as well as CLEAVE paper for examples.
+    Intertwine them.
+}
 
@@
 
-\subsection{Applying the methodology to \gls{WCA}}\label{ssec:methodology:wca}
+%\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
 
@@
 
-\todo[]{}
+\todo[inline]{}
+
+%\subsection{Applying the methodology to \acsp{NCS}}
+
+\todo[inline]{}
+
+\subsection{Construction of a testbed for the implementation of the methodology}\label{sec:testbed}
+
+\todo[inline]{}

### 2023-01-24 publications/2021ImpactDelayedResponse/body.tex
fixing paper and strucutre

@@
   While prototypical studies of such applications exist today, the relationship between infrastructure service provisioning and its implication for \gls{WCA} usability is largely unexplored despite the relevance that these applications have for future networks.
+
   This paper presents an experimental study assessing how \gls{WCA} users react to varying end-to-end delays induced by the application pipeline or infrastructure.
@@
 
-Wearable Cognitive Assistants, \gls{WCA} for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
+Wearable Cognitive Assistants, \gls{WCA} for short, have recently started to garner attention from the research community~\cite{ha2014towards,chen2015early}.
 They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
-Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
+Their mode of operation is analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
 Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
@@
 
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
-Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
+WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{ha2014towards,satyanarayanan2019augmenting}.
+More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{chen2017empirical}.
+Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
 Detethering this assistance from its current fixed location will surely make it available to many more fields.
@@
 \begin{enumerate}
-    \item\label{item:pervasive} \gls{WCA} systems should be available whenever the user requires them, without being tethered to a particular physical location.
+    \item\label{paper:olguinmunoz2021impact:item:pervasive} \gls{WCA} systems should be available whenever the user requires them, without being tethered to a particular physical location.
 Assistants need to be pervasive and mobile.
 
-    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
+    \item\label{paper:olguinmunoz2021impact:item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
     In this sense, \gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
 
-    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
+    \item\label{paper:olguinmunoz2021impact:item:lowlatency} Feedback should be ``quick'', relative to the task at hand.
     This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
@@
     In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
-    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
+    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant,chen2015early} late guidance is at best a nuisance and at worst a severe handicap.
 \end{enumerate}
 
-\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
-\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
+\cref{paper:olguinmunoz2021impact:item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
+\cref{paper:olguinmunoz2021impact:item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
+The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{paper:olguinmunoz2021impact:item:pervasive}.
 Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
 However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
-\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
+\cref{paper:olguinmunoz2021impact:item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
 
 % However, additional challenges remain in the path toward widespread adoption.
-\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
+\cref{paper:olguinmunoz2021impact:item:lowlatency} forms the base motivation for the research presented in this paper.
 We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
@@
 Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}.
+It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowski201140}.
 Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
@@
 This paper builds upon preliminary work in the field of time perception and delay characterization of \gls{WCA}.\@
-We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in \gls{WCA}, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
+We expand upon the findings of Ha et al.~\cite{ha2014towards}, who identified the need for low-latency offloading in \gls{WCA}, and of Chen et al.~\cite{chen2017empirical}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
 While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
@@
 Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
-The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@
+The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and \gls{EEG}.\@
 
@@
 
-  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
-  We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.
+  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999big} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
+  We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998measuring} to play roles in this.
 
@@
 These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
-Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
+Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid}.
 Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of \gls{WCA}.\@
@@
 We describe the existing body of research around time perception and the effects of delay on human performance in \cref{paper:olguinmunoz2021impact:sec:background}.
-\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
-Then, in \cref{sec:results} we detail the results of our experiment.
-Implications for further modeling of the effects of delay are presented in \cref{paper:olguinmunoz2021impact:sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.
+\cref{paper:olguinmunoz2021impact:sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
+Then, in \cref{paper:olguinmunoz2021impact:sec:results} we detail the results of our experiment.
+Implications for further modeling of the effects of delay are presented in \cref{paper:olguinmunoz2021impact:sec:discussion} before finally concluding the paper in \cref{paper:olguinmunoz2021impact:sec:conclusion}.
 
@@
 \begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-  \item[(\cref{ssec:timeperception})] time perception
-  \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
+  \item[(\cref{paper:olguinmunoz2021impact:ssec:timeperception})] time perception
+  \item[(\cref{paper:olguinmunoz2021impact:ssec:potentialmechs})] mechanisms relating human performance to delay.
 \end{enumerate*}
@@
     \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-    \item arousal can expand durations~\cite{droit2011emotion}.
+    \item arousal can expand durations~\cite{droit_volet2011emotion}.
 \end{enumerate*}
@@
 Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{Loewenstein1992anomaliesintertemporalchoice}.
+It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
 Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
@@
 
-It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowsky:2011:40years}.
+It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
 However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
@@
 
-For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
 They then proposed a latency tolerance zone according to the task demands.
 For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
-Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowsky:2011:40years}.
+Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
 
@@
 
-We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{Chen:EarlyImplementation}.
+We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{chen2015early}.
 This application belongs to a category of \gls{WCA}s designed to guide users through the execution of a sequential task.
@@
     % \item The states of the display can be recognized by simple image processing algorithms and do not require extensive training as for machine-learning-based applications.
-    \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
+    \item each step has an intrinsic hierarchical cognitive structure (see \cref{paper:olguinmunoz2021impact:fig:lego:hierarchical}), affording multiple levels of cognitive control.
 \end{enumerate*}
@@
 For the experimental LEGO task, a key modification was made to the structure of the steps.
-After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{paper:olguinmunoz2021impact:fig:cogassist:step,fig:cogassist:step:delay}.
+After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{paper:olguinmunoz2021impact:fig:cogassist:step,paper:olguinmunoz2021impact:fig:cogassist:step:delay}.
 The length of this delay is one of two independent variables we manipulated for the experiment.
-Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
+Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{chen2017empirical}.
 % A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
@@
 
-The first of these, the Big Five Inventory of Personality (\emph{BFI}~\cite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
+The first of these, the \gls{BFI}~\cite{john1999big}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
 Of these, extroversion and neuroticism have been related to tolerance for delay~\cite{hirsh2008delay}.
 
-The second survey, the Immersive Tendencies Questionnaire (\emph{ITQ}~\cite{witmer1998:itq}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
+The second survey, the \gls{ITQ}~\cite{witmer1998measuring}), comprises 29 questions, 28 of which were used for the study (one categorical question was disregarded), assessing the sub-scales of \emph{involvement}, the tendency to become involved in activities; \emph{focus}, the tendency to maintain focus on current activities, and \emph{games}, the tendency to play games.
 These questions use a 7-point horizontal scale  with opposing descriptors anchoring  the ends.
@@
 In post-processing, the obtained scores for both questionnaires were normalized to fall in the \( [0, 1] \) range for ease of interpretation. 
-See \cref{tab:qscores} for their means and standard deviations.
+See \cref{paper:olguinmunoz2021impact:tab:qscores} for their means and standard deviations.
 
@@
     \midrule
-    \multirow{5}{*}{BFI} 
+    \multirow{5}{*}{\gls{BFI}}
         & Agreeableness &               0.705 &                  0.136 \\
@@
     \cline{1-4}
-    \multirow{4}{*}{ITQ} 
+    \multirow{4}{*}{\gls{ITQ}}
         & Focus &               0.626 &                  0.126 \\
@@
                   label={{(\arabic*)}}]
-  \item galvanic skin response (\emph{GSR})
+  \item \glsreset{GSR}\gls{GSR}
   \item accelerometer data from the dominant wrist
-  \item brain activity in the form of electroencephalography (\emph{EEG})
+  \item brain activity in the form of \glsreset{EEG}\gls{EEG}
   \item heart rate.
@@
 
-These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\cite{khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload}.
-GSR (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\cite{peterson1907psycho,Healey2005,Son2010,khawadi2015:usinggsrtrust,kuikkaniemi2010:biofeedback,solovey2014:classifyingdriverworkload,}.
-EEG on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\cite{Antonenko2010,Grimes2008,kumar2016measurement}.
+These metrics have been used as indicators of stress and cognitive load by an ample body of previous research~\cite{khawaji2015using,kuikkaniemi2010influence,solovey2014classifying}.
+\gls{GSR} (also known as electrodermal activity) can be interpreted as an indicator of physiological arousal and has long been a widely used metric in studies seeking to characterize mental workload~\cite{peterson1907psycho,healey2005detecting,son2010age,khawaji2015using,kuikkaniemi2010influence,solovey2014classifying}.
+\gls{EEG} on the other hand has previously been used to measure cognitive load in the context of human-computer interactions~\cite{antonenko2010using,grimes2008feasibility,kumar2016measurement}.
 
-Wrist acceleration, GSR and heart rate data were obtained using the Empatica E4~\cite{empatica:e4} biosensing wristband.
+Wrist acceleration, \gls{GSR} and heart rate data were obtained using the Empatica E4~\cite{empatica:e4} biosensing wristband.
 Accelerometer data was sampled at \SI{32}{\hertz},
-GSR was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} (BVP) signal sampled at \SI{64}{\hertz}.
+\gls{GSR} was sampled at \SI{4}{\hertz}, and instantaneous heart rate was calculated from a \emph{blood volume pulse} signal sampled at \SI{64}{\hertz}.
 Participants were asked to wear the device for approximately 10--15 minutes before starting the experiment, in order to allow the sensors to reach a stable equilibrium and establish a baseline for the signals.
@@
 
-For the EEG data we employed the OpenBCI EEG Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
-It provides a quick and non-invasive way of obtaining EEG signals from participants.
+For the \gls{EEG} data we employed the OpenBCI \gls{EEG} Headband Kit~\cite{openbci:headbandkit} consisting of a number of dry electrodes fastened to a Velcro headband.
+It provides a quick and non-invasive way of obtaining \gls{EEG} signals from participants.
 Electrodes were placed according to the \emph{10--20 Electrode System}~\cite{eeg1020system:1961} on the \emph{Fp1} and \emph{Fp2} points, in order to capture brain activity in the frontal lobe, with ground and reference electrodes on the right and left earlobes, respectively.
@@
     \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5a.eps}
-    \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
+    \caption{Frame from the face recording of a random participant, clearly showing the locations of the \gls{EEG} electrodes.}
   \end{subfigure}%
@@
 
-The results consist of execution time per step. and the outcome of physiological variables: heart rate, GSR, and EEG.\@
+The results consist of execution time per step. and the outcome of physiological variables: heart rate, \gls{GSR}, and \gls{EEG}.\@
 Each will be discussed in turn.
@@
 
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
+We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{fujikoshi1993two}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
 % An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
@@
 An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{paper:olguinmunoz2021impact:tab:anova:exectime:slice}.
-As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
+As shown in \cref{paper:olguinmunoz2021impact:fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
 % We refer to the measure at slice 3 as a percentage of the measure at slice 1 as the ``change factor''.
@@
 We normalized the sum of the accelerations by the duration of the step to correct for differences in delay and the execution time, which tends to increase with delay as described above.
-An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{paper:olguinmunoz2021impact:tab:anova:acc} and \cref{fig:acc:delaylength}.
+An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{paper:olguinmunoz2021impact:tab:anova:acc} and \cref{paper:olguinmunoz2021impact:fig:acc:delaylength}.
 There was also a significant delay by length interaction, reflecting that delay effects occurred particularly for the longer lengths.
@@
 We next conducted the same analysis as for execution time, dividing blocks into three ``slices'' of four steps each, such that slice 1 comprised the first four steps of all block lengths, slice 2 the second four steps of blocks of length 8 and 12, and slice 3 the last four steps of blocks 12 steps long.
-As shown in \cref{fig:acc:delayslice} the effect of delay was attributable only to the later slices, yielding main effect of slice number and delay and an interaction (\cref{tab:anova:acc:slice}).
+As shown in \cref{paper:olguinmunoz2021impact:fig:acc:delayslice} the effect of delay was attributable only to the later slices, yielding main effect of slice number and delay and an interaction (\cref{paper:olguinmunoz2021impact:tab:anova:acc:slice}).
 
@@
 
-\subsection{EEG}
-Analyses were conducted on the log EEG power in the alpha band, beta band, and total of all bands measured.
+\subsection{\acs{EEG}}
+Analyses were conducted on the log \gls{EEG} power in the alpha band, beta band, and total of all bands measured.
 Readings from the two frontally placed poles were highly correlated and were pooled into an average.
-Logs were taken for analysis because the EEG distribution tended to have a rightward tail.
+Logs were taken for analysis because the \gls{EEG} distribution tended to have a rightward tail.
 Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 s.d.s.\ from the mean of all participants).
@@
 The analyses then comprised 28 participants.\@
-Omnibus ANOVAs were conducted with delay and block length (number of steps) as factors, on the EEG data from the alpha and beta band.
-The analysis of alpha EEG found no significant effects.
-Beta EEG showed only an effect of block length, \( F(2,54) = 3.56 \), \( p = .035 \), \( \eta^{2}_{p} = .12 \), reflecting a tendency for the 4-step length to produce lower log power (mean 3.9 vs.
+Omnibus ANOVAs were conducted with delay and block length (number of steps) as factors, on the \gls{EEG} data from the alpha and beta band.
+The analysis of alpha \gls{EEG} found no significant effects.
+Beta \gls{EEG} showed only an effect of block length, \( F(2,54) = 3.56 \), \( p = .035 \), \( \eta^{2}_{p} = .12 \), reflecting a tendency for the 4-step length to produce lower log power (mean 3.9 vs.
 4.1 for lengths of 8 and 12).
@@
 For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}.
-Both bands showed the same tendency:  EEG declined as a sequence of steps with the same delay progressed.
+Both bands showed the same tendency:  \gls{EEG} declined as a sequence of steps with the same delay progressed.
 These effects are shown in \cref{paper:olguinmunoz2021impact:fig:logeegpow:slices}.
-Thus, EEG taken from frontal locations mimics the execution time data in showing a decline over the course of a block, but unlike the execution time, there was no tendency for the decline in EEG to be reduced at longer delays.
+Thus, \gls{EEG} taken from frontal locations mimics the execution time data in showing a decline over the course of a block, but unlike the execution time, there was no tendency for the decline in \gls{EEG} to be reduced at longer delays.
 
@@
   \centering
-  \caption{Significant effects on log EEG power from ANOVA on factors delay and block slice.}\label{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}
+  \caption{Significant effects on log \gls{EEG} power from ANOVA on factors delay and block slice.}\label{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig11.eps}
-    \caption{Log of the average EEG Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:logeegpow:slices}
+    \caption{Log of the average \gls{EEG} Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:logeegpow:slices}
 \end{figure}
 
-\subsection{Galvanic Skin Response (GSR) and Heart Rate}
-The measure of GSR was the log of total amplitude in the signal.
-To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, GSR data were normalized by the temporal duration of the block.
-Due to sensor failure and the elimination of one subject with extreme values of GSR amplitude using the same rule as for EEG, the analysis comprised 34 participants.
+\subsection{\acs{GSR} and Heart Rate}
+The measure of \gls{GSR} was the log of total amplitude in the signal.
+To control for effects of block length (steps~\( \times \)~delay) and the slow-down in execution time at longer delays, \gls{GSR} data were normalized by the temporal duration of the block.
+Due to sensor failure and the elimination of one subject with extreme values of \gls{GSR} amplitude using the same rule as for \gls{EEG}, the analysis comprised 34 participants.
 No systematic trend due to delay or block length was observed.
-In addition, the same analysis of slice position within block length as was performed for the EEG revealed no effects, indicating that GSR was stable across step positions within a block.
+In addition, the same analysis of slice position within block length as was performed for the \gls{EEG} revealed no effects, indicating that \gls{GSR} was stable across step positions within a block.
 
@@
 The absence of systematic trends in both these results is interesting in the context of our initial suggestions of potential mechanisms relating delay in the application to human behavior.
-In section \cref{ssec:potentialmechs} we proposed that emotional arousal during the execution of the task was a potential explanation for the effects observed in the human.
+In section \cref{paper:olguinmunoz2021impact:ssec:potentialmechs} we proposed that emotional arousal during the execution of the task was a potential explanation for the effects observed in the human.
 However, these results seem to refute this hypothesis, and will be further discussed in \cref{paper:olguinmunoz2021impact:sec:discussion}.
@@
 
-The Big 5 personality inventory and Immersive Tendencies Questionnaires (ITQ) were combined with outcome variables in an analysis of individual differences.
+The \gls{BFI} and \gls{ITQ} were combined with outcome variables in an analysis of individual differences.
 Given the previous results, we initially considered the following outcome variables:
@@
     \item execution time in the most demanding block (length 12, delay \SI{3.0}{\second}); 
-    \item total of alpha and beta bands EEG in the first four steps at all lengths (``Slice 1'');
+    \item total of alpha and beta bands \gls{EEG} in the first four steps at all lengths (``Slice 1'');
     \item average heart rate;
-    \item and average log GSR.\@
+    \item and average log \gls{GSR}.\@
 \end{itemize}
 
-Although the heart rate and GSR data had produced no significant effects in the analysis of experimental variables, they could in principle correlate with experimental outcomes across individuals. 
-A principal-components analysis (\emph{PCA}) on a subset of these variables, shown in \cref{tab:pca}, was ultimately conducted on the 28 participants for which there was EEG data. 
+Although the heart rate and \gls{GSR} data had produced no significant effects in the analysis of experimental variables, they could in principle correlate with experimental outcomes across individuals.
+A \gls{PCA} on a subset of these variables, shown in \cref{paper:olguinmunoz2021impact:tab:pca}, was ultimately conducted on the 28 participants for which there was \gls{EEG} data.
 Three of the personality measures were excluded after initial analyses indicated significant correlations among neuroticism, openness, agreeableness, and extroversion.
 Neuroticism (with poles of sensitivity and security) was selected as most relevant to the issue of response to system delay and was included along with conscientiousness.
-GSR was used in a subsidiary analysis, because only 25 participants had both EEG and GSR measures that were reliable.
+\gls{GSR} was used in a subsidiary analysis, because only 25 participants had both \gls{EEG} and \gls{GSR} measures that were reliable.
 
@@
       \midrule
-      BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
-      BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
-      ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
-      ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
+      \gls{BFI} Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
+      \gls{BFI} Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
+      \gls{ITQ} Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
+      \gls{ITQ} Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
       Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
-      Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
+      Log \gls{EEG} power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
       \bottomrule
@@
 
-The PCA produced three components that accounted for \SI{73.13}{\percent} of the variance in the six factors considered.
-Component 1 included neuroticism, both ITQ scores, and execution time, indicating that more sensitive and immersed individuals tended to slow their responses under extended delay.
+The \gls{PCA} produced three components that accounted for \SI{73.13}{\percent} of the variance in the six factors considered.
+Component 1 included neuroticism, both \gls{ITQ} scores, and execution time, indicating that more sensitive and immersed individuals tended to slow their responses under extended delay.
 Component 2 included conscientiousness and the absence of neuroticism along with immersive involvement, indicating that efficient and secure participants tended to be more involved.
-The third component had positive loadings on EEG in Slice 1 (first four steps of a block) and the focus component of the ITQ.\@
-An additional analysis including heart rate added a component but improved the PCA fit by only \SI{4.5}{\percent}, indicating that any variability in this measure across individuals is unrelated to personality, immersiveness, or outcome.
-A further analysis including GSR produced a solution in which GSR loaded with the first component, along with execution time.
-% A further analysis including GSR again produced a three-component solution, in which GSR loaded with the first component (execution time, sensitivity, immersion).
+The third component had positive loadings on \gls{EEG} in Slice 1 (first four steps of a block) and the focus component of the \gls{ITQ}.\@
+An additional analysis including heart rate added a component but improved the \gls{PCA} fit by only \SI{4.5}{\percent}, indicating that any variability in this measure across individuals is unrelated to personality, immersiveness, or outcome.
+A further analysis including \gls{GSR} produced a solution in which \gls{GSR} loaded with the first component, along with execution time.
+% A further analysis including \gls{GSR} again produced a three-component solution, in which \gls{GSR} loaded with the first component (execution time, sensitivity, immersion).
 
-\cref{fig:neuro:exectime:reg} shows the correlation between neuroticism and execution time for the block with extreme values of length (12 steps) and delay (\SI{3.0}{\second}), for the full set of 40 participants.
-It confirms the relationship indicated by the first component in the PCA using the smaller sample of 28 participants; that is, higher neuroticism is associated with more responsiveness to delay.
+\cref{paper:olguinmunoz2021impact:fig:neuro:exectime:reg} shows the correlation between neuroticism and execution time for the block with extreme values of length (12 steps) and delay (\SI{3.0}{\second}), for the full set of 40 participants.
+It confirms the relationship indicated by the first component in the \gls{PCA} using the smaller sample of 28 participants; that is, higher neuroticism is associated with more responsiveness to delay.
 
@@
 It could, for instance, provide a tunable parameter for eventual models aiming to emulate human interaction with a \gls{WCA}.\@
-In addition, physiological measures of heart rate and EEG appear not to be direct indicators of  behavioral response to delay, although GSR may be more promising in this regard.
+In addition, physiological measures of heart rate and \gls{EEG} appear not to be direct indicators of  behavioral response to delay, although \gls{GSR} may be more promising in this regard.
 
@@
 
-  \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay. 
+  \item Fourthly, in terms of inter-subject differences, \gls{PCA} revealed three main factors governing users' response to delay.
   The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
@@
   In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
-  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
+  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018are}.
   These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
@@
 However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
-Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+Specifically, physiological measures of \gls{GSR} and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
 The acceleration data further did not indicate that extended delay significantly increases erratic movement.
-% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+% physiological measures of \gls{GSR} and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
 % The acceleration data further do not indicate that extended delay increases erratic movement.
@@
 We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
-As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
+As was described in \cref{paper:olguinmunoz2021impact:sec:experimentaldesign,paper:olguinmunoz2021impact:fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
 Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
@@
 Finally, we believe the data provide parameters that can usefully be integrated into cognitive models of \gls{WCA} that might be constructed under existing architectures like ACT-R.
-These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguin:2018,olguin:2019}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
+These same parameters could be used to modulate the timing and generation of inputs in trace-based workload generation tools such as the EdgeDroid platform~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid}, allowing the tool to use the same trace to generate workloads for a multitude of different user profiles.
 
@@
 \begin{acks}
-  This research was supported in part by the Swedish Foundation for Strategic Research (SSF) under grant \texttt{ITM17-0246} (ExPECA) and the United States National Science Foundation (NSF) under grant number \texttt{CNS-1518865}.
+  This research was supported in part by the \gls{SSF} under grant \texttt{ITM17-0246} (ExPECA) and the \gls{NSF} under grant number \texttt{CNS-1518865}.
   Any opinions, findings, conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the view(s) of their employers or funding sources.

### 2023-01-25 body/glossary.tex
impact results
more writing yay
duration
fixing papers
fixing paper and strucutre
restructuring

--- 2022-12-19 body/glossary.tex
@@
-\newabbreviation{WCA}{WCA}{Wearable Cognitive Assistance}
-\newabbreviation{AR}{AR}{Augmented Reality}
-\newabbreviation{VR}{VR}{Virtual Reality}
-\newabbreviation{XR}{XR}{eXtended Reality}
-\newabbreviation{GPU}{GPU}{Graphical Processing Unit}
-\newabbreviation{FPS}{FPS}{Frames Per Second}
-\newabbreviation{DNN}{DNN}{Deep Neural Network}
-\newabbreviation{CPS}{CPS}{Cyber-Physical System}
-\newabbreviation{NCS}{NCS}{Networked Control System}
+%! suppress = MissingImport
+\newacronym[longplural={Wearable Cognitive Assistants}]{WCA}{WCA}{Wearable Cognitive Assistance}
+\newacronym[longplural={Augmented Realities}]{AR}{AR}{Augmented Reality}
+\newacronym[longplural={Virtual Realities}]{VR}{VR}{Virtual Reality}
+\newacronym[longplural={eXtended Realities}]{XR}{XR}{eXtended Reality}
+\newacronym{GPU}{GPU}{Graphical Processing Unit}
+\newacronym{FPS}{FPS}{Frames Per Second}
+\newacronym{DNN}{DNN}{Deep Neural Network}
+\newacronym{CPS}{CPS}{Cyber-Physical System}
+\newacronym[prefixfirst={a\ }, prefix={an\ }]{NCS}{NCS}{Networked Control System}
 %\acroindefinite{NCS}{an}{a}
 
-\newabbreviation{LAN}{LAN}{Local-Area Network}
-\newabbreviation{WLAN}{WLAN}{Wireless Local-Area Network}
-\newabbreviation{KPI}{KPI}{Key Performance Indicator}
-\newabbreviation{WPAN}{WPAN}{Wireless Personal Area Network}
-\newabbreviation{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
-\newabbreviation{OS}{OS}{Operating System}
-\newabbreviation{UDP}{UDP}{User Datagram Protocol}
-\newabbreviation{TCP}{TCP}{Transmission Control Protocol}
-\newabbreviation{RMS}{RMS}{Root Mean Square}
-\newabbreviation{RTT}{RTT}{Round-Trip Time}
-\newabbreviation{CI}{CI}{Confidence Interval}
-\newabbreviation{AP}{AP}{Access Point}
-\newabbreviation{API}{API}{Application Programming Interface}
-\newabbreviation{SSF}{SSF}{Swedish Foundation for Strategic Research}
-\newabbreviation{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
-\newabbreviation{ABC}{ABC}{Abstract Base Class}
+\newacronym{LAN}{LAN}{Local-Area Network}
+\newacronym{WLAN}{WLAN}{Wireless Local-Area Network}
+\newacronym{KPI}{KPI}{Key Performance Indicator}
+\newacronym{WPAN}{WPAN}{Wireless Personal Area Network}
+\newacronym{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
+\newacronym{OS}{OS}{Operating System}
+\newacronym{UDP}{UDP}{User Datagram Protocol}
+\newacronym{TCP}{TCP}{Transmission Control Protocol}
+\newacronym{RMS}{RMS}{Root Mean Square}
+\newacronym{RTT}{RTT}{Round-Trip Time}
+\newacronym{CI}{CI}{Confidence Interval}
+\newacronym{AP}{AP}{Access Point}
+\newacronym{API}{API}{Application Programming Interface}
+\newacronym[user1={Stiftelsen för Strategisk Forskning}]{SSF}{SSF}{Swedish Foundation for Strategic Research}
+\newacronym{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
+\newacronym[longplural={Abstract Base Classes}]{ABC}{ABC}{Abstract Base Class}
 %\acroplural{ABC}[ABCs]{Abstract Base Classes}
-\newabbreviation{URL}{URL}{Uniform Record Locator}
-\newabbreviation{AWS}{AWS}{Amazon Web Services}
-\newabbreviation{EC2}{EC2}{Elastic Compute 2}
-\newabbreviation{AMI}{AMI}{Amazon Machine Image}
-\newabbreviation{SSH}{SSH}{Secure Shell}
-\newabbreviation{IP}{IP}{Internet Protocol}
-\newabbreviation{VPN}{VPN}{Virtual Private Network}
-\newabbreviation{NAT}{NAT}{Network Address Translation}
-\newabbreviation{NTP}{NTP}{Network Time Protocol}
+\newacronym{URL}{URL}{Uniform Record Locator}
+\newacronym{AWS}{AWS}{Amazon Web Services}
+\newacronym{EC2}{EC2}{Elastic Compute 2}
+\newacronym{AMI}{AMI}{Amazon Machine Image}
+\newacronym{SSH}{SSH}{Secure Shell}
+\newacronym{IP}{IP}{Internet Protocol}
+\newacronym{VPN}{VPN}{Virtual Private Network}
+\newacronym{NAT}{NAT}{Network Address Translation}
+\newacronym[prefixfirst={a\ }, prefix={an\ }]{NTP}{NTP}{Network Time Protocol}
 %\acroindefinite{NTP}{an}{a}
-\newabbreviation{SDR}{SDR}{Software-Defined Radio}
+\newacronym[prefixfirst={a\ }, prefix={an\ }]{SDR}{SDR}{Software-Defined Radio}
 %\acroindefinite{SDR}{an}{a}
-\newabbreviation{VLAN}{VLAN}{Virtual Local-Area Network}
-\newabbreviation{APT}{APT}{Advaced Packaging Tool}
-\newabbreviation{LTE}{LTE}{Long-Term Evolution}
-\newabbreviation{SSID}{SSID}{Service Set Identifier}
-\newabbreviation{EPC}{EPC}{Evolved Packet Core}
-\newabbreviation{UE}{UE}{User Equipment}
-\newabbreviation{eNodeB}{eNodeB}{Evolved Node B}
-\newabbreviation{DNS}{DNS}{Domain Name System}
-\newabbreviation{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
-\newabbreviation{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
+\newacronym{VLAN}{VLAN}{Virtual Local-Area Network}
+\newacronym{APT}{APT}{Advaced Packaging Tool}
+\newacronym{LTE}{LTE}{Long-Term Evolution}
+\newacronym{SSID}{SSID}{Service Set Identifier}
+\newacronym{EPC}{EPC}{Evolved Packet Core}
+\newacronym{UE}{UE}{User Equipment}
+\newacronym{eNodeB}{eNodeB}{Evolved Node B}
+\newacronym{DNS}{DNS}{Domain Name System}
+\newacronym{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
+\newacronym[prefixfirst={a\ }, prefix={an\ }]{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
 %\acroindefinite{HTTP}{an}{a}
-\newabbreviation{YAML}{YAML}{YAML Ain't Markup Language}
-\newabbreviation{O-RAN}{O-RAN}{Open Radio Access Network}
-\newabbreviation{FOSS}{FOSS}{Free and Open Source Software}
-\newabbreviation{COSMOS}{COSMOS}{Cloud enhanced Open Software defined MObile wireless testbed for city-Scale deployment}
-\newabbreviation{OMF}{OMF}{ORBIT Management Framework}
-\newabbreviation{POWDER}{POWDER}{Platform for Open Wireless Data-driven Experimental Research}
-\newabbreviation{COTS}{COTS}{Commercial Off-The-Shelves}
-\newabbreviation{RF}{RF}{Radio-Frequency}
-\newabbreviation{VM}{VM}{Virtual Machine}
-\newabbreviation{REST}{REST}{REpresentational Sate Transfer}
-\newabbreviation{MaaS}{MaaS}{Metal-as-a-Service}
-\newabbreviation{OEDL}{OEDL}{\gls{OMF} Experiment Description Language}
-\newabbreviation{RSpec}{RSpec}{Resource Specification}
-\newabbreviation{LXC}{LXC}{LinuX Containers}
-\newabbreviation{MEC}{MEC}{Mobile Edge Computing}
-\newabbreviation{USRP}{USRP}{Universal Software Radio Peripheral}
-\newabbreviation{OAI}{OAI}{OpenAirInterface}
-\newabbreviation{NR}{NR}{New Radio}
-\newabbreviation{FPGA}{FPGA}{Field-Programmable Gate Array}
-\newabbreviation{ARA}{ARA}{Agricultural and ruRAl communities}
-\newabbreviation{UHF}{UHF}{Ultra-High Frequency}
-\newabbreviation{MIMO}{MIMO}{Multi-Input Multi-Output}
-\newabbreviation{CHI}{CHI}{CHameleon Infrastructure}
-\newabbreviation{CMU}{CMU}{Carnegie Mellon University}
-\newabbreviation{GSR}{GSR}{Galvanic Skin Response}
-\newabbreviation{EEG}{EEG}{electroencephalography}
-\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits~\cite{john1999big}}
-\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}
-\newabbreviation{GPS}{GPS}{Global Positioning System}
-\newabbreviation{FSM}{FSM}{Finite State Machine}
-%\acroindefinite{NCS}{an}{a}
-\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}
-\newabbreviation{TTF}{TTF}{Time-to-Feedback}
-\newabbreviation{MAR}{MAR}{Mobile \gls{AR}}
-\newabbreviation{QoE}{QoE}{Quality of Experience}
-\newabbreviation{QoS}{QoS}{Quality of Service}
-\newabbreviation{PCA}{PCA}{Principal Component Analysis}
-\newabbreviation{exGaussian}{Ex-Gaussian}{Exponentially Modified Gaussian}
-\newabbreviation{IEEE}{IEEE}{Institute of Electrical and Electronics Engineers}
-\newabbreviation{CPU}{CPU}{Central Processing Unit}
-\newabbreviation{N/A}{N/A}{Not Applicable}
-\newabbreviation{CIO}{CIO}{Confidence Interval}
-\newabbreviation{MLE}{MLE}{Maximum Likelihood Estimation}
-\newabbreviation{CDF}{CDF}{Cumulative Distribution Function}
-\newabbreviation{CCDF}{CCDF}{Complementary \gls{CDF}}
-\newabbreviation{PDF}{PDF}{Probability Density Function}
-\newabbreviation{ECDF}{ECDF}{Empirical \gls{CDF}}
-\newabbreviation{NSF}{NSF}{United States National Science Foundation}
+\newacronym{YAML}{YAML}{YAML Ain't Markup Language}
+\newacronym{O-RAN}{O-RAN}{Open Radio Access Network}
+\newacronym{FOSS}{FOSS}{Free and Open Source Software}
+\newacronym{COSMOS}{COSMOS}{Cloud enhanced Open Software defined MObile wireless testbed for city-Scale deployment}
+\newacronym{OMF}{OMF}{ORBIT Management Framework}
+\newacronym{POWDER}{POWDER}{Platform for Open Wireless Data-driven Experimental Research}
+\newacronym{COTS}{COTS}{Commercial Off-The-Shelves}
+\newacronym{RF}{RF}{Radio-Frequency}
+\newacronym{VM}{VM}{Virtual Machine}
+\newacronym{REST}{REST}{REpresentational Sate Transfer}
+\newacronym{MaaS}{MaaS}{Metal-as-a-Service}
+\newacronym{OEDL}{OEDL}{\glsfmtshort{OMF} Experiment Description Language}
+\newacronym{RSpec}{RSpec}{Resource Specification}
+\newacronym{LXC}{LXC}{LinuX Containers}
+\newacronym{MEC}{MEC}{Mobile Edge Computing}
+\newacronym{USRP}{USRP}{Universal Software Radio Peripheral}
+\newacronym{OAI}{OAI}{OpenAirInterface}
+\newacronym{NR}{NR}{New Radio}
+\newacronym{FPGA}{FPGA}{Field-Programmable Gate Array}
+\newacronym{ARA}{ARA}{Agricultural and ruRAl communities}
+\newacronym{UHF}{UHF}{Ultra-High Frequency}
+\newacronym{MIMO}{MIMO}{Multi-Input Multi-Output}
+\newacronym{CHI}{CHI}{CHameleon Infrastructure}
+\newacronym{CMU}{CMU}{Carnegie Mellon University}
+\newacronym{GSR}{GSR}{Galvanic Skin Response}
+\newacronym{EEG}{EEG}{ElectroEncephaloGraphy}
+\newacronym{BFI}{BFI}{Big Five Inventory of personality traits}
+\newacronym{ITQ}{ITQ}{Immersive Tendencies Questionnaire}
+\newacronym{GPS}{GPS}{Global Positioning System}
+\newacronym{FSM}{FSM}{Finite State Machine}
+\newacronym{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}
+\newacronym[longplural={Times-to-Feedback}]{TTF}{TTF}{Time-to-Feedback}
+\newacronym{MAR}{MAR}{Mobile \glsfmtshort{AR}}
+\newacronym{QoE}{QoE}{Quality of Experience}
+\newacronym{QoS}{QoS}{Quality of Service}
+\newacronym{PCA}{PCA}{Principal Component Analysis}
+\newacronym{exGaussian}{Ex-Gaussian}{Exponentially Modified Gaussian}
+\newacronym{IEEE}{IEEE}{Institute of Electrical and Electronics Engineers}
+\newacronym{CPU}{CPU}{Central Processing Unit}
+\newacronym{N/A}{N/A}{Not Applicable}
+\newacronym{CIO}{CIO}{Confidence Interval}
+\newacronym{MLE}{MLE}{Maximum Likelihood Estimation}
+\newacronym{CDF}{CDF}{Cumulative Distribution Function}
+\newacronym{CCDF}{CCDF}{Complementary \glsfmtshort{CDF}}
+\newacronym{PDF}{PDF}{Probability Density Function}
+\newacronym{ECDF}{ECDF}{Empirical \glsfmtshort{CDF}}
+\newacronym{NSF}{NSF}{United States National Science Foundation}
+\newacronym{ANOVA}{ANOVA}{ANalysis Of VAriance}
+\newacronym{CV}{CV}{Computer Vision}
+\newacronym{SEM}{SEM}{Standard Error of the Mean}
+
+\glsaddall%

### 2023-01-25 body/part1/overview/extending-wca.tex
restructuring

@@
-\section{Implications of the methodology in the context of \acs{WCA}}\label{sec:extendingmethodology}
+\section{Implications of the methodology in the context of \glsfmtshort{WCA}}\label{sec:extendingmethodology}
 
@@
         \includegraphics[height=13em]{Figs/2021Impact/previousblock_vs_exectime}
-        \caption{Mean execution time of the first slice of steps in a block, grouped by \acs{TTF} of the block immediately preceding the current one.}\label{fig:prevblockvsexectime}
+        \caption{Mean execution time of the first slice of steps in a block, grouped by \glsfmtshort{TTF} of the block immediately preceding the current one.}\label{fig:prevblockvsexectime}
     \end{minipage}%
@@
         \caption{%
-            Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \acs{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
+            Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \glsfmtshort{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
             Fits were achieved through \gls{MLE}.
@@
 
-\subsection{A new model of human behavior in \acs{WCA}}
+\subsection{A new model of human behavior in \glsfmtshort{WCA}}
 
@@
 
-\subsubsection{Procedural generation of dynamic traces of video frames for \acs{WCA}}\label{model:traces}
+\subsubsection{Procedural generation of dynamic traces of video frames for \glsfmtshort{WCA}}\label{model:traces}
 
@@
 We perform this experiment on the edge computing testbed discussed in \cref{sec:testbed};
-for simplicity and to magnify the effects of latency, in this setup we use an \acs{IEEE} \num{802.11}b/g physical layer.
+for simplicity and to magnify the effects of latency, in this setup we use an \glsfmtshort{IEEE} \num{802.11}b/g physical layer.
 

### 2023-01-25 body/part1/overview/methodology.tex
restructuring

@@
     Intertwine them.
+}
+
+\subsection{Use case: \glsfmtshort{WCA}}\glsreset{WCA}
+
+\todo[inline]{Discuss, at high level, the advantages of this methodology for WCA.
+Challenges, opportunities. Maybe refer to the ``use cases'' of the first edgedroid paper.}
+
+\subsection{Use case: \glsfmtshortpl{NCS}}\glsreset{NCS}
+
+\todo[inline]{Discuss, at high level, the advantages ofthis methodology for NCSs.
+Refer back to related work. }
+
+\subsection{Tools for the application of the methodology}
+
+\todo[inline]{Here we actually present the tools}
+
+\subsubsection{\glsfmtshort{WCA}: EdgeDroid}
+
+\subsubsection{\glsfmtshortpl{NCS}: CLEAVE}
+
+\subsubsection{Testbeds: Ainur}
+
+\todo[inline]{%
+Everything below is ``old'' stuff that needs to be re-edited.%
 }
@@
 
-%\subsection{Applying the methodology to \acs{WCA}}\label{ssec:methodology:wca}
+%\subsection{Applying the methodology to \glsfmtshort{WCA}}\label{ssec:methodology:wca}
 
@@
 As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
-
-\todo[inline]{}
-
-%\subsection{Applying the methodology to \acsp{NCS}}
-
-\todo[inline]{}
-
-\subsection{Construction of a testbed for the implementation of the methodology}\label{sec:testbed}
-
-\todo[inline]{}

### 2023-01-25 publications/2021ImpactDelayedResponse/body.tex
restructuring

@@
 
-\subsection{\acs{EEG}}
+\subsection{\glsfmtshort{EEG}}
 Analyses were conducted on the log \gls{EEG} power in the alpha band, beta band, and total of all bands measured.
@@
 
-\subsection{\acs{GSR} and Heart Rate}
+\subsection{\glsfmtshort{GSR} and Heart Rate}
 The measure of \gls{GSR} was the log of total amplitude in the signal.

### 2023-01-26 publications/2021ImpactDelayedResponse/body.tex
fix impact paper

@@
 
-Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
+Both video feeds were captured at a rate of \num{24}~\gls{FPS} in parallel processes to ensure a constant rate of capture.
 Examples of the captured frames can be seen in \cref{paper:olguinmunoz2021impact:fig:sampleframes}.
@@
 
-We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{fujikoshi1993two}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
+We confirmed this effect through an \gls{ANOVA}~\cite{fujikoshi1993two} with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
 % An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
-    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
+    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the \gls{SEM}}\label{fig:exectime}%
 \end{figure}
@@
   \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:exectime}
+  \caption{Significant effects on per-step execution time from \gls{ANOVA} on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:exectime}
   \setlength{\tabcolsep}{0pt}%
@@
 Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
-An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{paper:olguinmunoz2021impact:tab:anova:exectime:slice}.
+An \gls{ANOVA} on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{paper:olguinmunoz2021impact:tab:anova:exectime:slice}.
 As shown in \cref{paper:olguinmunoz2021impact:fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
@@
   \centering
-  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{paper:olguinmunoz2021impact:tab:anova:exectime:slice}
+  \caption{Significant effects on per-step execution time from \gls{ANOVA} on factors block slice and delay.}\label{paper:olguinmunoz2021impact:tab:anova:exectime:slice}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \caption{Mean per-step execution time vs.\ delay, by step slice.
-    Error bars indicate S.E.M.}
+    Error bars indicate \gls{SEM}}
     \label{paper:olguinmunoz2021impact:fig:exectime:delay:slice}
@@
   \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \).
-Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:exectime:transition}%
+Error bars indicate \gls{SEM}}\label{paper:olguinmunoz2021impact:fig:exectime:transition}%
 \end{figure}
@@
 We normalized the sum of the accelerations by the duration of the step to correct for differences in delay and the execution time, which tends to increase with delay as described above.
-An ANOVA on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{paper:olguinmunoz2021impact:tab:anova:acc} and \cref{paper:olguinmunoz2021impact:fig:acc:delaylength}.
+An \gls{ANOVA} on block length and delay showed a significant effect such that movement score decreased with delay, as shown in \cref{paper:olguinmunoz2021impact:tab:anova:acc} and \cref{paper:olguinmunoz2021impact:fig:acc:delaylength}.
 There was also a significant delay by length interaction, reflecting that delay effects occurred particularly for the longer lengths.
@@
   \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:acc}
+  \caption{Significant effects on accelerometer data from \gls{ANOVA} on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:acc}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig9.eps}
-    \caption{Movement score vs.\ delay, per block length. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:acc:delaylength}
+    \caption{Movement score vs.\ delay, per block length. Error bars indicate \gls{SEM}}\label{paper:olguinmunoz2021impact:fig:acc:delaylength}
 \end{figure}
@@
   \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
-  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:acc:delayslice}
+  \caption{Movement score vs.\ delay, per block slice. Error bars indicate \gls{SEM}}\label{paper:olguinmunoz2021impact:fig:acc:delayslice}
 \end{figure}
@@
   \centering
-  \caption{Significant effects on accelerometer data from ANOVA on factors delay and slice number.}\label{paper:olguinmunoz2021impact:tab:anova:acc:slice}
+  \caption{Significant effects on accelerometer data from \gls{ANOVA} on factors delay and slice number.}\label{paper:olguinmunoz2021impact:tab:anova:acc:slice}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
 Logs were taken for analysis because the \gls{EEG} distribution tended to have a rightward tail.
-Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 s.d.s.\ from the mean of all participants).
+Twelve participants were excluded from the analysis, 9 due to device failure and 3 because of extreme values (i.e., the participant mean of log total power was greater than 3 standard deviations from the mean of all participants).
 
 The analyses then comprised 28 participants.\@
-Omnibus ANOVAs were conducted with delay and block length (number of steps) as factors, on the \gls{EEG} data from the alpha and beta band.
+Omnibus \glspl{ANOVA} were conducted with delay and block length (number of steps) as factors, on the \gls{EEG} data from the alpha and beta band.
 The analysis of alpha \gls{EEG} found no significant effects.
@@
 Again the analysis dividing blocks into 4-step slices was conducted, with delay and slice number as factors.  
-For both the alpha and beta bands, ANOVAs  yielded effects only of slice number --- see \cref{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}.
+For both the alpha and beta bands, \glspl{ANOVA}  yielded effects only of slice number --- see \cref{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}.
 Both bands showed the same tendency:  \gls{EEG} declined as a sequence of steps with the same delay progressed.
@@
   \centering
-  \caption{Significant effects on log \gls{EEG} power from ANOVA on factors delay and block slice.}\label{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}
+  \caption{Significant effects on log \gls{EEG} power from \gls{ANOVA} on factors delay and block slice.}\label{paper:olguinmunoz2021impact:tab:anova:egg:alphabeta}
   \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
@@
     \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig11.eps}
-    \caption{Log of the average \gls{EEG} Power for alpha and beta bands per step slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:logeegpow:slices}
+    \caption{Log of the average \gls{EEG} Power for alpha and beta bands per step slice. Error bars indicate \gls{SEM}.}\label{paper:olguinmunoz2021impact:fig:logeegpow:slices}
 \end{figure}
@@
 
-Heart rate, measured in beats per minute, averaged \num{90.2}~BPM and showed no trend related to the experimental variables of delay and block length.
-The same was true of the variability in beats per minute (average standard deviation across conditions was \num{20.3}).
+Heart rate averaged \num{90.2}~\gls{BPM} and showed no trend related to the experimental variables of delay and block length.
+The same was true of the variability in \gls{BPM} (average standard deviation across conditions was \num{20.3}).
 
@@
 However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
-Specifically, physiological measures of \gls{GSR} and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
+Specifically, physiological measures of \gls{GSR} and heart rate failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
 The acceleration data further did not indicate that extended delay significantly increases erratic movement.

### 2023-01-28 body/part1/background.tex
glossaries
restructuring for easier following of the red thread
fixing paper and strucutre
restructuring
relwork

@@
-\chapter{Background \& Related Work}
+\section{Background \& related work}
+\todo[inline]{%
+    \textcite{satyanarayanan2009case}
 
-Deep focus on theoretical background.
-
-\section{Theoretical background}
-\textcite{Satya2009Case}.
     Other papers detailing the advantages of Edge computing.
@@
     Next, what applications benefit from the Edge?
-\ac{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
+    \gls{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
     Main points: prototypes and characterization of system latencies and load.
+}
+
+\subsection{Edge computing}
+
+Edge computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
+Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
+These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
+Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters~\citationeeded.
+These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
+
+Through this design, cloud computing affords significant advantages to users.
+As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
+Thanks the economies of scale and virtualization technologies, the cloud is highly scalable;
+services can be scaled simply by spawning more \glspl{VM}.
+The specialized nature of cloud providers, the scale of modern datacenters, and the use of virtualization also make the cloud highly reliable.
+When hardware fails, recovering is simply a matter of migrating the service container or \gls{VM} to an available compute node~\cite{endo2016high}.
+
+However, the cloud is not suitable for all applications, and presents important drawbacks and challenges for bandwidth-hungry, privacy-requiring, and/or latency-sensitive applications.
+Current cloud infrastructures are ill-suited to deal with the tremendous amounts of data generated by \gls{IOT} sensor networks.
+The centralized nature of cloud data centers, which are often placed a large distance away from potential users, lead to prohibitively high latencies for highly interactive immersive applications such as mobile \gls{XR}\citationeeded.
+Furthermore, this same centralized nature has led to concerns about the security of cloud computing\todo{needs to be expanded a bit}.
+
+Edge computing emerges as an answer to these challenges~\cite{satyanarayanan2009case,satyanarayanan2017emergence}.
+This term refers to a novel distributed computing paradigm, complementary to cloud computing, which tackles the challenges of the cloud by moving computation closer to where it is needed.
+Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter --- a \emph{cloudlet}~\cite{satyanarayanan2009case} --- at the \emph{edge} of the network, one or two hops away from the user.
+This design leads to several advantages.
+One, edge computing is uniquely suited for highly latency-sensitive yet resource-intensive applications such as the aforementioned mobile \gls{XR} as well as \gls{CPS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
+
+\subsection{\glsfmtshort{WCA} and other \glsfmtshort{XR} applications}
+
+There is increasing interest from academia and industry in novel applications such as immersive \gls{AR} or \gls{WCA}~\cite{chatzopoulos2017hyperion,ha2014towards}, also known as human-in-the-loop applications.
+These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
+Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as \glspl{DNN}~\cite{ha2014towards}.
+\Cref{paper:olguinmunoz2019edgedroid:fig:pipeline2} depicts such a system.
+These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
+Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{chen2017empirical}.
+
+In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
+There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
+However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
+A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
+What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
+There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
+Finally, in production use these applications will most likely run concurrently with others.
+How does this, together with other best-effort applications, impact the latencies perceived by the human user?
+Existing studies~\cite{ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion} of this class of applications have only lightly touched upon these issues~\cite{chen2017empirical}.
+On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{al_zubaidy2015performance,schiessl2017finite} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
+We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
+
+The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
+We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
+
+Wearable Cognitive Assistants, \gls{WCA} for short, have recently started to garner attention from the research community~\cite{ha2014towards,chen2015early}.
+They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
+Their mode of operation is analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
+Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
+An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
+
+WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{ha2014towards,satyanarayanan2019augmenting}.
+More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{chen2017empirical}.
+Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
+Detethering this assistance from its current fixed location will surely make it available to many more fields.
+
+Based on these use cases, we identify three main requirements for \gls{WCA}:\@
+\begin{enumerate}
+    \item \gls{WCA} systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
+
+    \item Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
+    In this sense, \gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+
+    \item Feedback should be ``quick'', relative to the task at hand.
+    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
+    This means that users will have expectations of constant, immediate feedback as they progress through the task.
+    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
+    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant,chen2015early} late guidance is at best a nuisance and at worst a severe handicap.
+\end{enumerate}
+
+\cref{paper:olguinmunoz2021impact:item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
+\cref{paper:olguinmunoz2021impact:item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
+The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{paper:olguinmunoz2021impact:item:pervasive}.
+Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
+However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
+\cref{paper:olguinmunoz2021impact:item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
+
+% However, additional challenges remain in the path toward widespread adoption.
+\cref{paper:olguinmunoz2021impact:item:lowlatency} forms the base motivation for the research presented in this paper.
+We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
+\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
+% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
+Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
+It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowski201140}.
+Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
+A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of \gls{WCA}.\@
+
+\begin{figure}
+    \centering
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/task.png}
+        \caption{%
+            Overview of a task in a \gls{WCA}, composed of a series of steps.
+            Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
+            The \gls{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
+        }\label{background:fig:wcatask}
+    \end{subfigure}\\
+    \begin{subfigure}{\columnwidth}
+        \centering
+        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/step_time.png}
+        \caption{%
+            Breakdown of a step into its timing components.
+            The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
+            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \gls{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
+        }\label{background:fig:wcastep}
+    \end{subfigure}
+    \caption{Key concepts in \gls{WCA}}
+\end{figure}
+
+\glspl{WCA} applications represent a category of novel, context-sensitive and highly-interactive \gls{AR} applications.
+In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
+Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
+
+Step-based \glspl{WCA} operate analogously to how \gls{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
+The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
+Whenever the assistant detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
+The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.~because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
+Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
+
+In order to discuss these applications with precision we provide some definitions relating to their operation.
+First of all, a \emph{step} is formally understood as a specific action to be performed by the user, described by a single instruction, and a \emph{task} consists of a series of steps to be executed in sequence (see \cref{background:fig:wcatask}).
+A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided; we call the time interval between these two events the \emph{step duration}.
+
+\glspl{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
+Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \gls{WCA} captures the state of the physical system, as illustrated in \cref{background:fig:wcastep}.
+\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.\ which captures the final state of step \( M \)) is taken.
+\( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
+We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \gls{RTT}}, respectively, of a step \( M \).
+The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\gls{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \gls{WCA}.
+
+\subsection{Mechanisms relating human behavior to system responsiveness}
+
+The question of how people respond to delay in a computer system is grounded in how people perceive time.
+Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
+More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
+The estimation of an event's duration varies with context of various types
+\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
+\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
+\item arousal can expand durations~\cite{droit_volet2011emotion}.
+\end{enumerate*}
+
+Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
+It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
+Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
+Violations of expectancy have a strong impact on the acceptability of computer systems.
+Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
+In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
+
+It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
+However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
+Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
+
+For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+They then proposed a latency tolerance zone according to the task demands.
+For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
+Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
+
+While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
+These mechanisms could be cognitive or emotional in origin.
+
+A first possible explanation comes from research on cognitive and motor planning.
+Delay may move users from relatively automatic to more attention-demanding processing.
+Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
+As competency in a task increases, execution of the hierarchy becomes increasingly automated.
+Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
+Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
+
+Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
+Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
+
+A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
+As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
+
+Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
+
+The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
+If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
+Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
+Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
+Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
+
+Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
+Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
+High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
+Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
+Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
+Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
+These and other measures of individual variation were considered here.
+
+\subsection{\glsfmtshortpl{NCS} and other \glsfmtshortpl{CPS}}
+
+The number and applications of \glspl{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
+However, this rapid increase in adoption has mostly been limited to industrial contexts.
+Although \glspl{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
+This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \glspl{CPS} will be made possible.
+These technologies meet two key requirements of \glspl{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
+
+\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
+This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
+
+%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
+
+One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
+%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
+On the other, there exist several approaches when considering experimental methodologies.
+%A number of works concerning \glspl{NCS} deal with experimental studies.
+%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
+One such approach uses setups in which the complete system is built on top of real hardware.
+This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
+Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
+The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
+These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
+Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
+    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
+    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
+\end{enumerate*}
+
+As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
+This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
+Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
+This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
+
+Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
+In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
+Their methodology utilizes joint knowledge of control, computation, and communication.
+In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
+Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
+This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
+Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
+
+\subsubsection{Benchmarking tools for edge computing}
+
+Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications.
+On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
+The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.
+
+Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
+\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+    \item \gls{COSMOS}
+    \item \gls{POWDER}
+    \item \gls{ARA}
+    \item Drexel Grid \gls{SDR}
+\end{enumerate*} testbeds.
+\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
+It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
+\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
+
+\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
+The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
+\gls{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \gls{VM} images containing the necessary software and configurations.
+These profiles are defined through using \gls{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
+
+The \gls{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
+Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
+It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+\gls{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \gls{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
+This affords the \gls{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
+
+Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
+Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
+However, the testbed extensively employs the \gls{LXC} runtime for the deployment of both experimental code and \gls{SDR} software, which affords users great freedom when it comes to the development of experiments.
+
+Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
+Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
+Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
+This is, for instance, the case with \gls{COSMOS} and \gls{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
+These testbeds further leverage virtualization technology based on \glspl{VM} instead of more lightweight and edge-compatible solutions such as containers.
+
+To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
+CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \glspl{VM} containing the necessary software for experiments.
+Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
+Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \pgls{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.

### 2023-01-28 body/part1/background.tex
editing

@@
 However, the cloud is not suitable for all applications, and presents important drawbacks and challenges for bandwidth-hungry, privacy-requiring, and/or latency-sensitive applications.
-Current cloud infrastructures are ill-suited to deal with the tremendous amounts of data generated by \gls{IOT} sensor networks.
-The centralized nature of cloud data centers, which are often placed a large distance away from potential users, lead to prohibitively high latencies for highly interactive immersive applications such as mobile \gls{XR}\citationeeded.
+Current cloud infrastructures are ill-suited to deal with the tremendous amounts of data generated by \gls{IOT} sensor networks~\cite{shi2016edge}.
+The centralized nature of cloud data centers, which are often placed a large distance away from potential users, lead to prohibitively high latencies for highly interactive immersive applications such as mobile \gls{XR}~\cite{satyanarayanan2009case,varghese2016challenges,shi2016promise}.
 Furthermore, this same centralized nature has led to concerns about the security of cloud computing\todo{needs to be expanded a bit}.
 
-Edge computing emerges as an answer to these challenges~\cite{satyanarayanan2009case,satyanarayanan2017emergence}.
+Edge computing emerges as an answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence}.
 This term refers to a novel distributed computing paradigm, complementary to cloud computing, which tackles the challenges of the cloud by moving computation closer to where it is needed.
@@
 One, edge computing is uniquely suited for highly latency-sensitive yet resource-intensive applications such as the aforementioned mobile \gls{XR} as well as \gls{CPS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
+
+\todo[inline]{Include diagram edge versus fog versus cloud.}
 

### 2023-01-30 body/part1/background.tex
edge computing diagram
edge

--- 2023-01-28 body/part1/background.tex
@@
 As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
-Thanks the economies of scale and virtualization technologies, the cloud is highly scalable;
+Thanks to economies of scale and virtualization technologies, the cloud is highly scalable;
 services can be scaled simply by spawning more \glspl{VM}.
@@
 
-However, the cloud is not suitable for all applications, and presents important drawbacks and challenges for bandwidth-hungry, privacy-requiring, and/or latency-sensitive applications.
-Current cloud infrastructures are ill-suited to deal with the tremendous amounts of data generated by \gls{IOT} sensor networks.
-The centralized nature of cloud data centers, which are often placed a large distance away from potential users, lead to prohibitively high latencies for highly interactive immersive applications such as mobile \gls{XR}\citationeeded.
-Furthermore, this same centralized nature has led to concerns about the security of cloud computing\todo{needs to be expanded a bit}.
+However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive applications.
+In order to achieve the necessary economies of scale, cloud datacenters are designed to serve users distributed across vast geographical areas.
+These installations are thus often located ``far'' from potential users;
+for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
+This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{satyanarayanan2009case,varghese2016challenges,shi2016promise}.
+The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
+Such latencies are unfeasible to consistently achieve with cloud computing~\cite{dang2021cloudy}.
 
-Edge computing emerges as an answer to these challenges~\cite{satyanarayanan2009case,satyanarayanan2017emergence}.
+Current cloud infrastructures are also ill-suited to deal with the bandwidth requirements of the \gls{IOT} and of mobile computing.
+As smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face significant challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=30em]{Figs/edgecomputing}
+    \caption{%
+        Conceptual design of edge computing.
+        Micro-clouds (known as ``cloudlets''~\cite{satyanarayanan2009case}) are placed a at the edge of the network, a few hops away from end users.
+        In other words, these cloudlets are located between users and the internet and the cloud.
+    }\label{fig:edgecomputing}
+\end{figure}
+
+Edge computing, emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,wang2019towards}.
 This term refers to a novel distributed computing paradigm, complementary to cloud computing, which tackles the challenges of the cloud by moving computation closer to where it is needed.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter --- a \emph{cloudlet}~\cite{satyanarayanan2009case} --- at the \emph{edge} of the network, one or two hops away from the user.
-This design leads to several advantages.
-One, edge computing is uniquely suited for highly latency-sensitive yet resource-intensive applications such as the aforementioned mobile \gls{XR} as well as \gls{CPS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
+Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter --- a \emph{cloudlet}~\cite{satyanarayanan2009case} --- at the \emph{edge} of the network.
+These coudlets feature most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, one or two hops away from the user (see \cref{fig:edgecomputing})
+
+The architecture of edge computing offers several advantages.
+Thanks to their proximity to users, cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
+Mobile \gls{XR}, such as \gls{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application}, and 5G-enabled \glspl{CPS} and \glspl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient} are uniquely poised to take advantage of this characteristic of edge computing.
 

### 2023-01-30 body/part1/background.tex
edge computing

@@
 Thanks to their proximity to users, cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Mobile \gls{XR}, such as \gls{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application}, and 5G-enabled \glspl{CPS} and \glspl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient} are uniquely poised to take advantage of this characteristic of edge computing.
+Mobile \gls{XR}, such as \gls{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application}, and 5G-enabled \glspl{CPS} and \glspl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient} are uniquely poised to take advantage of these characteristic of edge computing.
+By processing and aggregating data closer to where its needed, cloudlets can also drastically reduce the amount of traffic going to cloud datacenters.
+Cloudlets also present opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
 

### 2023-01-30 body/part1/background.tex
wca

@@
 
-Edge computing, emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,wang2019towards}.
+Edge computing, emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
 This term refers to a novel distributed computing paradigm, complementary to cloud computing, which tackles the challenges of the cloud by moving computation closer to where it is needed.
@@
 
-\subsection{\glsfmtshort{WCA} and other \glsfmtshort{XR} applications}
+\todo[inline]{Transition to talk about research and tools for Edge computing research.}
 
-There is increasing interest from academia and industry in novel applications such as immersive \gls{AR} or \gls{WCA}~\cite{chatzopoulos2017hyperion,ha2014towards}, also known as human-in-the-loop applications.
-These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
-Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as \glspl{DNN}~\cite{ha2014towards}.
-\Cref{paper:olguinmunoz2019edgedroid:fig:pipeline2} depicts such a system.
-These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
-Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{chen2017empirical}.
+\subsection{\glsfmtlong{WCA}}
 
-In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
-There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
-However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
-A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
-What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
-There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
-Finally, in production use these applications will most likely run concurrently with others.
-How does this, together with other best-effort applications, impact the latencies perceived by the human user?
-Existing studies~\cite{ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion} of this class of applications have only lightly touched upon these issues~\cite{chen2017empirical}.
-On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{al_zubaidy2015performance,schiessl2017finite} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
-We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
+There is increasing interest from academia and industry in a novel category of wearable and mobile immersive applications called \gls{WCA}~\cite{ha2014towards,chen2018application,chen2015early,belletier2021wearable,chen2017empirical,wang2020scaling,chatzopoulos2017hyperion}
+They represent a particular class of \gls{AR} applications, aiming amplify human cognition in both day-to-day tasks and professional settings.
+\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
+Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
+The input is then parsed to an internal symbolic representation of the state of the task, and whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
+Feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
+If no change in state is detected, the application remains silent and out-of-the-way of the user;
+in other words, samples which capture an intermediate or unfinished state, or simply noise are silently discarded.
+\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
 
-The goal of this paper is to provide a methodological approach to studying these latency trade-offs, along with a tool, EdgeDroid 1.0\footnote{We plan to make the EdgeDroid 1.0 benchmarking suite available as Free and Open Source Software and the recorded traces under a Creative Commons License.}, that simplifies the benchmarking of human-in-the-loop applications.
-We view EdgeDroid 1.0 to be the very first, and simplest, of a family of tools that will embody increasingly sophisticated and accurate models of user behavior.
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{Figs/wca_state}
+    \caption{%
+        Mode of operation of \gls{WCA} applications.
+        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
+        Samples that do not result in feedback are silently discarded.
+    }\label{fig:wca}
+\end{figure}
 
-Wearable Cognitive Assistants, \gls{WCA} for short, have recently started to garner attention from the research community~\cite{ha2014towards,chen2015early}.
-They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
-Their mode of operation is analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
 Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
+Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
 An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
@@
 WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{ha2014towards,satyanarayanan2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{chen2017empirical}.
+More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{chen2017empirical,chen2018application}.
 Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
 Detethering this assistance from its current fixed location will surely make it available to many more fields.
+
+However, these applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
+Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{chen2017empirical}.
+There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
 
@@
 \cref{paper:olguinmunoz2021impact:item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
-
-% However, additional challenges remain in the path toward widespread adoption.
-\cref{paper:olguinmunoz2021impact:item:lowlatency} forms the base motivation for the research presented in this paper.
-We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
-\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
-% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
-Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
-It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowski201140}.
-Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
-A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of \gls{WCA}.\@
 
@@
 
-\glspl{WCA} applications represent a category of novel, context-sensitive and highly-interactive \gls{AR} applications.
 In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
-Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
-
-Step-based \glspl{WCA} operate analogously to how \gls{GPS} navigation assistants guide users, by seamlessly and continuously monitoring the progress of the user and autonomously providing relevant instructions and feedback.
-The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video frames.
-Whenever the assistant detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
-The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.~because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
-Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
+Examples of such applications are the LEGO and IKEA assistants~\cite{chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
 
@@
 
-\subsection{Mechanisms relating human behavior to system responsiveness}
+\todo[inline]{Talk about existing research. Prototypes and benchmarking.}
 
-The question of how people respond to delay in a computer system is grounded in how people perceive time.
-Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-The estimation of an event's duration varies with context of various types
-\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-\item arousal can expand durations~\cite{droit_volet2011emotion}.
-\end{enumerate*}
+In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
+There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
+However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
+A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
+What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
+There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
+Finally, in production use these applications will most likely run concurrently with others.
+How does this, together with other best-effort applications, impact the latencies perceived by the human user?
+Existing studies~\cite{ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion} of this class of applications have only lightly touched upon these issues~\cite{chen2017empirical}.
+On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{al_zubaidy2015performance,schiessl2017finite} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
+We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
 
-Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
-Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-Violations of expectancy have a strong impact on the acceptability of computer systems.
-Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
-
-It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
-However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
-
-For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-They then proposed a latency tolerance zone according to the task demands.
-For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
-Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
-
-While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
-These mechanisms could be cognitive or emotional in origin.
-
-A first possible explanation comes from research on cognitive and motor planning.
-Delay may move users from relatively automatic to more attention-demanding processing.
-Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-
-Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-
-A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-
-Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-
-The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-
-Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
+%\subsection{Mechanisms relating human behavior to system responsiveness}
+%
+%The question of how people respond to delay in a computer system is grounded in how people perceive time.
+%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
+%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
+%The estimation of an event's duration varies with context of various types
+%\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
+%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
+%\item arousal can expand durations~\cite{droit_volet2011emotion}.
+%\end{enumerate*}
+%
+%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
+%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
+%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
+%Violations of expectancy have a strong impact on the acceptability of computer systems.
+%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
+%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
+%
+%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
+%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
+%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
+%
+%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+%They then proposed a latency tolerance zone according to the task demands.
+%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
+%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
+%
+%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
+%These mechanisms could be cognitive or emotional in origin.
+%
+%A first possible explanation comes from research on cognitive and motor planning.
+%Delay may move users from relatively automatic to more attention-demanding processing.
+%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
+%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
+%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
+%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
+%
+%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
+%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
+%
+%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
+%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
+%
+%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
+%
+%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
+%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
+%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
+%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
+%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
+%
+%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
+%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
+%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
+%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
+%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
 Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
-These and other measures of individual variation were considered here.
+%These and other measures of individual variation were considered here.
 

### 2023-01-31 body/part1/background.tex
edge

@@
 \subsection{\glsfmtlong{WCA}}
+\glsreset{WCA}
 
@@
 They represent a particular class of \gls{AR} applications, aiming amplify human cognition in both day-to-day tasks and professional settings.
-\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
+\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
+These systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
+More recently, their scope has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards,ha2014towards}.
+Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}, and it is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
+\todo[inline]{Develop a bit further. Talk about more reactive setups like ping pong too.}
+
+We identify a number of defining characteristics of \gls{WCA}.
+One, and as the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
+they are pervasive and mobile~\cite{ha2014towards}.
+
+Next, interaction with \glspl{WCA} is immersive and seamless.
+As mentioned above, \glspl{WCA} operate in a manner similar to \gls{GPS} guidance systems.
 Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
@@
 \Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
+Assistants should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
+\gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+
+Note that this mode interaction implies seamless integration of the application with the context of the user and the real world.
+The user does not trigger an update explicitly at any point, leading to one of the key characteristics of \gls{WCA}.
+The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
+An important corollary of this is that these are therefore the \emph{only} points in time at which users can notice changes in system responsiveness.
+
+Finally, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand) to the user.
+This goes hand in hand with the previous characteristic of seamless interaction, as users will have expectations of constant and immediate feedback as they interact with the system.
+As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
+Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).
+
+The above characteristics together lead to interesting consequences for \gls{WCA} system design.
+Their wearable nature directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
+The requirement of immersive and seamless interaction suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
+This kind of compute- and energy-intensive processing is unfeasible to be performed on wearable devices, and thus these applications necessarily make use of compute offloading.
+As explained above, the cloud is unsuitable as an offloading point for these applications, given the high latencies incurred, and there is thus a broad understanding today that with the advent of edge computing, \glspl{WCA} will become a practical reality~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just,wang2020scaling,chen2018application,ha2014towards}.
+
+There's more to end-to-end latency than merely the question of where the compute backend is placed, however.
+The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
+These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
+Existing studies of this class of applications have only recently started to delve more deeply upon these issues~\cite{chen2017empirical,wang2019towards,wang2020scaling,ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion}.
+Recently published models for end-to-end latency of edge computing architectures, on the other hand, are quite complex, while not accounting for the specifics of human-in-the-loop applications and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite,moothedath2022energy2,moothedath2022energy1,moothedath2021energy}.
+The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
+
+\subsubsection{Characterization of \glsfmtshort{WCA}}
+
+\todo[inline]{%
+    Needs reformulating. Right now is just copy of EdgeDroid2 paper section.
+}
 
@@
 \end{figure}
-
-Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
-Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
-An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.
-
-WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{ha2014towards,satyanarayanan2019augmenting}.
-More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{chen2017empirical,chen2018application}.
-Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
-Detethering this assistance from its current fixed location will surely make it available to many more fields.
-
-However, these applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
-Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{chen2017empirical}.
-There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
-
-Based on these use cases, we identify three main requirements for \gls{WCA}:\@
-\begin{enumerate}
-    \item \gls{WCA} systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.
-
-    \item Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
-    In this sense, \gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
-
-    \item Feedback should be ``quick'', relative to the task at hand.
-    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
-    This means that users will have expectations of constant, immediate feedback as they progress through the task.
-    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
-    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant,chen2015early} late guidance is at best a nuisance and at worst a severe handicap.
-\end{enumerate}
-
-\cref{paper:olguinmunoz2021impact:item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
-\cref{paper:olguinmunoz2021impact:item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{paper:olguinmunoz2021impact:item:pervasive}.
-Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
-However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
-\cref{paper:olguinmunoz2021impact:item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.
 
@@
 The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\gls{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \gls{WCA}.
-
-\todo[inline]{Talk about existing research. Prototypes and benchmarking.}
-
-In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
-There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just}.
-However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
-A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
-What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
-There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
-Finally, in production use these applications will most likely run concurrently with others.
-How does this, together with other best-effort applications, impact the latencies perceived by the human user?
-Existing studies~\cite{ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion} of this class of applications have only lightly touched upon these issues~\cite{chen2017empirical}.
-On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{al_zubaidy2015performance,schiessl2017finite} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
-We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.
 

### 2023-02-02 body/part1/background.tex
wca

@@
 There is increasing interest from academia and industry in a novel category of wearable and mobile immersive applications called \gls{WCA}~\cite{ha2014towards,chen2018application,chen2015early,belletier2021wearable,chen2017empirical,wang2020scaling,chatzopoulos2017hyperion}
-They represent a particular class of \gls{AR} applications, aiming amplify human cognition in both day-to-day tasks and professional settings.
+They represent a particular class of \gls{AR} applications, aiming to amplify human cognition in both day-to-day tasks and professional settings.
 \glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
@@
 \todo[inline]{Develop a bit further. Talk about more reactive setups like ping pong too.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{Figs/wca_state}
+    \caption{%
+        Mode of operation of \gls{WCA} applications.
+        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
+        Samples that do not result in feedback are silently discarded.
+    }\label{fig:wca}
+\end{figure}
 
@@
 
-\todo[inline]{%
-    Needs reformulating. Right now is just copy of EdgeDroid2 paper section.
-}
-
 \begin{figure}
     \centering
-    \includegraphics[width=.9\textwidth]{Figs/wca_state}
-    \caption{%
-        Mode of operation of \gls{WCA} applications.
-        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
-        Samples that do not result in feedback are silently discarded.
-    }\label{fig:wca}
+    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
+    \caption[]{%
+        Breakdown of a step in a \gls{WCA} into its timing components.
+        \ensuremath{t_k | k \in \{1, \ldots, n \}} correspond to sampling instants.
+        \ensuremath{t_0} indicates the instant at which the instruction for step \ensuremath{M} is provided to the user and the first sample for said step is captured.
+        The instruction for step \ensuremath{M + 1} is provided at \ensuremath{t_{n+1}}.
+        Finally, \ensuremath{t_c} marks the instant at which the user finishes the instruction, and \ensuremath{t_n} the instant of capture of the first sample after the instruction has been completed;
+        this sample therefore also corresponds to the final sample of step \ensuremath{M}.
+        Figure originally included in \cref{paper:olguinmunoz2023realistic}.
+    }\label{fig:wcastep}
 \end{figure}
 
-\begin{figure}
-    \centering
-    \begin{subfigure}{\columnwidth}
-        \centering
-        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/task.png}
-        \caption{%
-            Overview of a task in a \gls{WCA}, composed of a series of steps.
-            Each steps starts with an instruction being provided to the user and ends with the instruction for the next step.
-            The \gls{WCA} continuously samples the task state, automatically triggering transitions between steps as correct (or incorrect) states are recognized.
-        }\label{background:fig:wcatask}
-    \end{subfigure}\\
-    \begin{subfigure}{\columnwidth}
-        \centering
-        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/step_time.png}
-        \caption{%
-            Breakdown of a step into its timing components.
-            The instruction for step \( M \) and \( M + 1 \) are provided to the user at \( t_0 \) and \( t_{n+1} \), respectively.
-            \( t_k | k \in \{1, \ldots, n \} \) correspond to the \gls{WCA} sampling instants for step \( M \), and \( t_c \) marks the instant at which the user finishes performing the instruction.
-        }\label{background:fig:wcastep}
-    \end{subfigure}
-    \caption{Key concepts in \gls{WCA}}
-\end{figure}
+An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
+We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
+In this section, we will provide some definitions relating to the operation of these applications.
 
-In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
-Examples of such applications are the LEGO and IKEA assistants~\cite{chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
+As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
+In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
+We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
+A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
+the time interval between these two events will be referred to as the \emph{step duration}.
 
-In order to discuss these applications with precision we provide some definitions relating to their operation.
-First of all, a \emph{step} is formally understood as a specific action to be performed by the user, described by a single instruction, and a \emph{task} consists of a series of steps to be executed in sequence (see \cref{background:fig:wcatask}).
-A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided; we call the time interval between these two events the \emph{step duration}.
+In the following, please refer to \cref{fig:wcastep}.
+\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
+\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
+this sample is the first sample to capture the finished state of the step.
+At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
+We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
 
-\glspl{WCA} employ sampling, most commonly of video feeds, to keep track of the state of the real world.
-Take \( \{ t_0, t_1, \ldots, t_{n + 1} \} \) a series of discrete and sequential sampling instants at which the \gls{WCA} captures the state of the physical system, as illustrated in \cref{background:fig:wcastep}.
-\( t_0 \) corresponds to the instant at which the instruction for step \( M \) is provided and the first sample is taken, and \( t_n \) to the instant at which the final sample (i.e.\ which captures the final state of step \( M \)) is taken.
-\( t_{n + 1} \) is then the instant at which the result for sample \( t_n \) is returned and the instruction for step \( M + 1 \) is provided.
-We define \( t_c \) as the point in time at which the user finishes performing the instruction for step \( M \), and the intervals \( t_c - t_0 \), \( t_n - t_c \), and \( t_{n + 1} - t_n \) as the \emph{execution time}, \emph{wait time}, and \emph{last sample \gls{RTT}}, respectively, of a step \( M \).
-The sum of the latter two values (i.e.\ the interval \( t_{n + 1} - t_c \)) we call \emph{\gls{TTF}}, a metric which we will repeatedly refer to in this paper as it directly describes the responsiveness of a \gls{WCA}.
+\glsreset{TTF}%
+From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
+If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
+\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}, label={(\arabic*)}]
+    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
+    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
+    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
+\end{enumerate*}
+These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
+We will refer to them extensively in \cref{chap:contributions}.
 

### 2023-02-02 body/part1/background.tex
comment

@@
 Recently published models for end-to-end latency of edge computing architectures, on the other hand, are quite complex, while not accounting for the specifics of human-in-the-loop applications and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite,moothedath2022energy2,moothedath2022energy1,moothedath2021energy}.
+\todo[inline]{%
+    Maybe also present ``other'' research on WCA here? E.g.\ other research in WCA has focused on... So-and-so have... blah blah.
+}
 The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.

### 2023-02-02 publications/2022CLEAVE/cleave.tex
fixing cleave paper

@@
-\section{The \gls{CLEAVE} framework}\label{sec:approach}
+\section{The \gls{CLEAVE} framework}\label{paper:olguinmunoz2022cleave:approach}
 
@@
 
-\subsection{Design and Implementation}\label{sec:design}
+\subsection{Design and Implementation}\label{paper:olguinmunoz2022cleave:design}
 
@@
 
-\cref{fig:cleave:ncs:struct} shows an overview of the conceptual structure and operation of \gls{CLEAVE}.
+\begin{figure*}
+    \centering
+    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
+    \caption{
+        Structure of an emulated \glspl{NCS} in \gls{CLEAVE}.
+    }\label{paper:olguinmunoz2022cleave:fig:cleave:ncs:struct}
+\end{figure*}
+
+\cref{paper:olguinmunoz2022cleave:fig:cleave:ncs:struct} shows an overview of the conceptual structure and operation of \gls{CLEAVE}.
 The \emph{State} implements the discrete-time behavior of the Plant.
@@
     \centering
-    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
+    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum}
     \caption{
         The 2D inverted pendulum system.
-    }\label{fig:invpend}
+    }\label{paper:olguinmunoz2022cleave:fig:invpend}
 \end{figure}
 
-For the purpose of this work, we have used \gls{CLEAVE} to implement an emulated \emph{inverted pendulum} control loop (see \cref{fig:invpend}), the goal of which is to balance the vertically free-swinging pendulum by applying horizontal forces on the cart.
+For the purpose of this work, we have used \gls{CLEAVE} to implement an emulated \emph{inverted pendulum} control loop (see \cref{paper:olguinmunoz2022cleave:fig:invpend}), the goal of which is to balance the vertically free-swinging pendulum by applying horizontal forces on the cart.
 We have chosen this system as an initial benchmark for its relative simplicity as well as prevalence in the field of automatic control as one of the fundamental examples of linear control.
@@
 The inverted pendulum plant is merely a proof-of-concept, and \gls{CLEAVE} could be used to implement much more complex physical systems with even more stringent latency requirements.
-For instance, the flexibility afforded by the choice of programming language and the abstractions detailed in \cref{sec:design}, allow for relatively straightforward emulation of autonomous drone dynamics.
+For instance, the flexibility afforded by the choice of programming language and the abstractions detailed in \cref{paper:olguinmunoz2022cleave:design}, allow for relatively straightforward emulation of autonomous drone dynamics.
 Propeller dynamics could be implemented by \emph{Actuator} objects, gyroscopes by \emph{Sensors}, and the physical behavior of the plant could be modeled using a general purpose 3D physics engine (\emph{e.g.} the Panda3D\footnote{\url{https://docs.panda3d.org/1.10/python/programming/physics/builtin/index}{https://docs.panda3d.org/1.10/python/programming/physics/builtin/index}} engine).

### 2023-02-02 publications/2022CLEAVE/conclusion.tex
fixing cleave paper

@@
-\section{Conclusion}\label{sec:conclusion}
+\section{Conclusion}\label{paper:olguinmunoz2022cleave:conclusion}
 
@@
 \begin{enumerate}
-    \item Compared to fully physical approaches, such as those used in\ \cite{Baumann2018LowPower} and\ \cite{Cuenca2019UAV}, our approach allows for greater flexibility and scalability.
-    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as Zoppi \emph{et al.'s} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
+    \item Compared to fully physical approaches, such as those used in\ \cite{baumann2018evaluating} and\ \cite{cuenca2019periodic}, our approach allows for greater flexibility and scalability.
+    The aforementioned approaches rely on specialized and sometimes entirely custom-built physical platforms, and although flexible and cheap approaches such as\ \cite{zoppi2020ncsbench} --- which uses a LEGO-based physical plant --- exist, these still do not reach the level of flexibility afforded by a fully software-based framework.
     Experimenters still need copies of the hardware, making anything other than small-scale setups unfeasible.
@@
     Scalable deployments can in turn easily and cheaply be set up using single-board computers and/or virtual cloud instances.
-    \item When compared to simulated approaches such as\ \cite{Ma2019DynamicSched}, \gls{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
-    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{Wang2020VoltageControl}, \gls{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
+    \item When compared to simulated approaches such as\ \cite{ma2019optimal}, \gls{CLEAVE} provides a higher level of realism, in particular with regards to the network segment of the system.
+    \item Finally, although it shares much in common with previous emulated approaches such as the one employed in\ \cite{wang2020inverter}, \gls{CLEAVE} has an advantage by specifically targeting a general-purpose approach using industry-standard, cloud- and edge-native tools and software.
     The tool can easily be deployed and scaled using widely-used frameworks such as Docker Swarm and Kubernetes.

### 2023-02-02 publications/2022CLEAVE/experiments.tex
fixing cleave paper

@@
-\section{Use case validation}\label{sec:experiments}
+\section{Use case validation}\label{paper:olguinmunoz2022cleave:experiments}
 
@@
 We believe this to be an interesting and representative use-case scenario for a number of reasons.
-Firstly, the inverted pendulum is broadly used as a benchmark in control-system literature, see for instance\ \cite{Baumann2018LowPower} and\ \cite{Natale2004InvPendEthernet}.
+Firstly, the inverted pendulum is broadly used as a benchmark in control-system literature, see for instance\ \cite{baumann2018evaluating} and\ \cite{natale2004inverted}.
 A key reason for this is that such a relatively simplistic system allows for straightforward reparametrization to obtain varied system dynamics, which in turn makes a broader range of experiments possible.
-Secondly, similar, albeit non-Edge-enabled, setups have been a reality in industry for almost two decades, as evidenced by\ \cite{Gupta2010NCSOverview}.
+Secondly, similar, albeit non-Edge-enabled, setups have been a reality in industry for almost two decades, as evidenced by\ \cite{gupta2010networked}.
 Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \glspl{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
@@
 Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \glspl{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
-Finally, video analytics is one of the main proposed use cases for edge computing~\cite{Ananthanarayanan2017Analytics,Yi2017Analytics,Wang2018Analytics}, and thus we foresee edge \gls{NCS} deployments being deployed in parallel with such applications in the future.
+Finally, video analytics is one of the main proposed use cases for edge computing~\cite{ananthanarayanan2017real,yi2017lavea,wang2018bandwidth}, and thus we foresee edge \gls{NCS} deployments being deployed in parallel with such applications in the future.
 However, before deployment can become a reality, key questions such as: 
-``what are the baseline requirements of the \gls{NCS}?''; ``what are the achievable best-case end-to-end latencies in the system?''; and
-``how does the concurrent video-streaming traffic affect \gls{NCS} stability?'' need to be answered.
+``what are the baseline requirements of the \gls{NCS}?'';
+``what are the achievable best-case end-to-end latencies in the system?'';
+and ``how does the concurrent video-streaming traffic affect \gls{NCS} stability?'' need to be answered.
 
-We attempt to study these questions using the experimental setup in \cref{fig:cleave:expsetup}.
+We attempt to study these questions using the experimental setup in \cref{paper:olguinmunoz2022cleave:fig:cleave:expsetup}.
 \gls{CLEAVE} is deployed on a testbed consisting of \num{10} Raspberry Pi 4B  clients connected wirelessly to an IEEE 802.11n \gls{AP}.
@@
     \centering
-    \includegraphics[width=.95\columnwidth]{publications/2022CLEAVE/images/CLEAVE_experiment_setup}
+    \includegraphics[width=.95\textwidth]{publications/2022CLEAVE/images/CLEAVE_experiment_setup}
     \caption{
@@
         % Containerized versions of the core CLEAVE emulation components are deployed inside a Docker Swarm Overlay Network spanning \num{10} Raspberry Pi 4B clients connected to a single Cloudlet over an IEEE 802.11n \gls{AP}.
-    }\label{fig:cleave:expsetup}
+    }\label{paper:olguinmunoz2022cleave:fig:cleave:expsetup}
 \end{figure}
@@
 
-This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \gls{CLEAVE}'s design.
+This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \gls{CLEAVE}s design.
 Scenarios are executed automatically in batches using a simple Python script which interacts with Docker through the widely adopted \emph{docker-py}\footnote{Docker SDK for Python: \url{https://docker-py.readthedocs.io/en/stable/}} library.
-This is \gls{CLEAVE}'s first advantage over existing frameworks; it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.
+This is \gls{CLEAVE}s first advantage over existing frameworks;
+it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.
 
@@
     \centering
-    \begin{subfigure}[t]{\columnwidth}
+    \begin{subfigure}[t]{.7\textwidth}
         \centering
         \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_toppled}
-        \caption{Fraction of toppled plants.}\label{fig:single:topple}
+        \caption{Fraction of toppled plants.}\label{paper:olguinmunoz2022cleave:fig:single:topple}
     \end{subfigure}\\
-    \begin{subfigure}[t]{\columnwidth}
+    \begin{subfigure}[t]{.7\textwidth}
         \centering
         \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_single_loop_rms}
-        \caption{Mean angle \gls{RMS}; red line indicates \( y = 180 \).}\label{fig:single:rms}
+        \caption{Mean angle \gls{RMS}; red line indicates \( y = 180 \).}\label{paper:olguinmunoz2022cleave:fig:single:rms}
     \end{subfigure}\\
-    \begin{subfigure}[t]{\columnwidth}
+    \begin{subfigure}[t]{.7\textwidth}
         \centering
@@
         }%
-    \label{fig:single}
+    \label{paper:olguinmunoz2022cleave:fig:single}
 \end{figure}
 
-\cref{fig:single} shows a summary of the results relating to the single-loop scenarios:
-\cref{fig:single:topple} shows the fraction of plants that toppled in each scenario, and \cref{fig:single:rms} shows the average \gls{RMS} for the absolute pendula angles.
+\cref{paper:olguinmunoz2022cleave:fig:single} shows a summary of the results relating to the single-loop scenarios:
+\cref{paper:olguinmunoz2022cleave:fig:single:topple} shows the fraction of plants that toppled in each scenario, and \cref{paper:olguinmunoz2022cleave:fig:single:rms} shows the average \gls{RMS} for the absolute pendula angles.
 \cref{fig:single:rtt} shows average latency due to network and processing (excluding synthetic delays) for both single-loop scenarios.
@@
 
-As expected, higher sampling rates tend to correlate with better quality of control; at higher sampling rates the system was able to reach stability at higher \glspl{RTT}.
+As expected, higher sampling rates tend to correlate with better quality of control;
+at higher sampling rates the system was able to reach stability at higher \glspl{RTT}.
 These initial results already hint at interesting consequences for such an edge-bound \gls{NCS} deployment.
-For instance, it is clear from \cref{fig:single:topple,fig:single:rms} that network delays can, to a certain extent, be compensated for by increasing the sampling rate of the system.
+For instance, it is clear from \cref{paper:olguinmunoz2022cleave:fig:single:topple,paper:olguinmunoz2022cleave:fig:single:rms} that network delays can, to a certain extent, be compensated for by increasing the sampling rate of the system.
 A corollary of this is, conversely, that at lower network latencies \glspl{NCS} are able to stabilize at lower sampling rates.
@@
 Once baselines for single loops in the use case scenario have been established, we can proceed to studying the interaction between the \glspl{NCS} and the video-streaming applications.
-We deploy \num{6} control loops on the experimental setup depicted in \cref{fig:cleave:expsetup}.
+We deploy \num{6} control loops on the experimental setup depicted in \cref{paper:olguinmunoz2022cleave:fig:cleave:expsetup}.
 On the remaining \num{4} clients we run the \emph{iperf3}\footnote{iperf3: \url{https://iperf.fr/}} traffic load generator, each generating \SI[per-mode=symbol]{6.5}{\mega\bit\per\second} of uplink \gls{UDP} traffic.
@@
     \centering
-    \begin{subfigure}[h]{.22\textwidth}
+    \begin{subfigure}[h]{.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
-        \caption{Toppled plants.}\label{fig:video:toppled}
+        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
+        \caption{Toppled plants.}\label{paper:olguinmunoz2022cleave:fig:video:toppled}
     \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
+%    \hfill%
+    \begin{subfigure}[h]{.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
-        \caption{Angle \gls{RMS}.}\label{fig:video:rms}
+        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
+        \caption{Angle \gls{RMS}.}\label{paper:olguinmunoz2022cleave:fig:video:rms}
     \end{subfigure}\\
-    \begin{subfigure}[h]{.22\textwidth}
+    \begin{subfigure}[h]{.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
-        \caption{Packet losses.}\label{fig:video:drop}
+        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
+        \caption{Packet losses.}\label{paper:olguinmunoz2022cleave:fig:video:drop}
     \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[h]{.22\textwidth}
+%    \hfill%
+    \begin{subfigure}[h]{.5\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
-        \caption{\glspl{RTT}.}\label{fig:video:rtt}
+        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
+        \caption{\glspl{RTT}.}\label{paper:olguinmunoz2022cleave:fig:video:rtt}
     \end{subfigure}%
@@
         Error bars indicate \SI{95}{\percent} \glspl{CI} in all plots.
-    }\label{fig:video:results}
+    }\label{paper:olguinmunoz2022cleave:fig:video:results}
 \end{figure}
 
-\cref{fig:video:results} shows a summary of the results obtained.
-\cref{fig:video:toppled} shows the fraction of repetitions of each setup in which \emph{at least} one plant failed to maintain stability and toppled.
-\cref{fig:video:rms} shows the \gls{RMS} for the pendulum angles for each setup, only considering data from plants that did not topple.
-\cref{fig:video:drop} shows the fraction of \gls{UDP} datagrams dropped, averaged over all plants and repetitions per setup.
-\cref{fig:video:rtt} shows the measured end-to-end plant-side \gls{RTT}, averaged over all plants and repetitions per setup.
+\cref{paper:olguinmunoz2022cleave:fig:video:results} shows a summary of the results obtained.
+\cref{paper:olguinmunoz2022cleave:fig:video:toppled} shows the fraction of repetitions of each setup in which \emph{at least} one plant failed to maintain stability and toppled.
+\cref{paper:olguinmunoz2022cleave:fig:video:rms} shows the \gls{RMS} for the pendulum angles for each setup, only considering data from plants that did not topple.
+\cref{paper:olguinmunoz2022cleave:fig:video:drop} shows the fraction of \gls{UDP} datagrams dropped, averaged over all plants and repetitions per setup.
+\cref{paper:olguinmunoz2022cleave:fig:video:rtt} shows the measured end-to-end plant-side \gls{RTT}, averaged over all plants and repetitions per setup.
 
 These results are interesting in their counter-intuitiveness compared to the single-loop baseline results.
-While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, \cref{fig:video:toppled,fig:video:rms} show this not to be the case for \gls{NCS} in resource-constrained scenarios.
+While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, \cref{paper:olguinmunoz2022cleave:fig:video:toppled,paper:olguinmunoz2022cleave:fig:video:rms} show this not to be the case for \gls{NCS} in resource-constrained scenarios.
 \SI{60}{\hertz} was the least stable configuration, with at least one pendulum toppling in around \SI{16}{\percent} of the repetitions, and mean pendulum angle \gls{RMS} circa \num{3} times that of the \SI{40}{\hertz} scenario.
@@
 
-\cref{fig:video:drop,fig:video:rtt} explains these behaviors.
+\cref{paper:olguinmunoz2022cleave:fig:video:drop,paper:olguinmunoz2022cleave:fig:video:rtt} explain these behaviors.
 Whereas both the \SIlist{20;40}{\hertz} setups show losses well below \SI{5}{\percent}, the \SI{60}{\hertz} scenario shows an average of around \SI{13}{\percent} of datagrams lost.

### 2023-02-02 publications/2022CLEAVE/intro.tex
fixing cleave paper

@@
-\section{Introduction}\label{sec:intro}
+\section{Introduction}\label{paper:olguinmunoz2022cleave:intro}
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \glspl{NCS} in \gls{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-The number and applications of \glspl{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
+The number and applications of \glspl{CPS}~\cite{rajkumar2010cyber} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
 However, this rapid increase in adoption has mostly been limited to industrial contexts.
 Although \glspl{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
-This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \glspl{CPS} will be made possible.
+This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{satyanarayanan2017emergence}, consumer-grade \glspl{CPS} will be made possible.
 These technologies meet two key requirements of \glspl{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
 
-\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
+\glspl{NCS}~\cite{gupta2010networked}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{wan2020efficient}.
 This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
@@
 One such approach uses setups in which the complete system is built on top of real hardware.
-This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
+This approach is employed in the works of Baumann \emph{et al.}~\cite{baumann2018evaluating} and Cuenca \emph{et al.}~\cite{cuenca2019periodic}; in both of these, the authors implement their approach on physical testbeds.
 Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
-The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
+The authors in\ \cite{ma2019optimal} have opted for such an approach.
 These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
@@
 \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
-    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
+    \item a real network interacts with a simulated or emulated control system~\cite{wang2020inverter}
+    \item a simulated network interacts with a real control system~\cite{natale2004inverted}.
 \end{enumerate*}
@@
 
-Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
+Zoppi \emph{et al.}~\cite{zoppi2020ncsbench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
 In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\ 
@@
 Our framework, \gls{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
+It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{olguinmunoz2019edgedroid}.
 The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
@@
 The rest of this paper is structured as follows.
-\cref{sec:approach} presents the design principles and implementation of the framework.
-In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \gls{CLEAVE}.
-Finally, in \cref{sec:conclusion} we conclude this work.
+\cref{paper:olguinmunoz2022cleave:approach} presents the design principles and implementation of the framework.
+In \cref{paper:olguinmunoz2022cleave:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \gls{CLEAVE}.
+Finally, in \cref{paper:olguinmunoz2022cleave:conclusion} we conclude this work.

### 2023-02-03 publications/2022Ainur/abstract.tex
fixing ainur paper

@@
   It is designed to configure, manage, orchestrate, and deploy workloads from an end-to-end perspective.
-  Ainur is built on top of cloud-native technologies such as Docker, and is provided as \gls{FOSS} to the community through the {KTH-EXPECA/Ainur} repository on GitHub.
+  Ainur is built on top of cloud-native technologies such as Docker, and is provided as \gls{FOSS} to the community through the \href{https://github.com/KTH-EXPECA/Ainur}{KTH-EXPECA/Ainur} repository on GitHub.
   We demonstrate the utility of the platform with a series of scenarios, showcasing in particular its flexibility with respect to physical link definition, computation placement, and automation of arbitrarily complex experimental scenarios.

### 2023-02-03 publications/2022Ainur/ainur.tex
fixing ainur paper

@@
-\section{The Ainur Framework}\label{sec:ainur}
+\section{The Ainur Framework}\label{paper:olguinmunoz2022airnur:ainur}
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
-    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
+    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/overview}
+    \caption{Layered structure of an Ainur experiment}\label{paper:olguinmunoz2022airnur:fig:overview}
 \end{figure}
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
+    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/flow2}
+    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{paper:olguinmunoz2022airnur:fig:flow}
 \end{figure*}
@@
     \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
-    \caption{Network structure assumed by Ainur.}\label{fig:network}
+    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/network}
+    \caption{Network structure assumed by Ainur.}\label{paper:olguinmunoz2022airnur:fig:network}
 \end{figure*}
@@
 \end{enumerate*}.
-In \cref{fig:overview}, the workload is represented by a client-server process pair.
+In \cref{paper:olguinmunoz2022airnur:fig:overview}, the workload is represented by a client-server process pair.
 These could correspond, for instance, to the emulation of an inverted pendulum and a matching controller.
@@
 Users are free to combine and interconnect these in any configuration.
-Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including WiFi, 4G \gls{LTE}, and 5G.
+Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including Wi-Fi, 4G \gls{LTE}, and 5G.
 To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \gls{USRP} \gls{SDR}s and some computation hosts dedicated to signal processing and/or radio management.
@@
     We leverage Docker containers to great extent for the virtualization and orchestration of workloads, as well as the encapsulation and management of complex network configurations.
-    See \cref{ssec:containers} for more details.
+    See \cref{paper:olguinmunoz2022airnur:containers} for more details.
 
@@
     \item[\gls{OAI}:]
-    \gls{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \glspl{SDR} like the \gls{USRP}~\cite{KALTENBERGER2020107284}.
+    \gls{OAI} is an open-source project that implements 3GPP technology on general purpose \texttt{x86} computing hardware and off-the-shelf \glspl{SDR} like the \gls{USRP}~\cite{kaltenberger2020openairinterface}.
     Ainur can deploy, configure, and manage \gls{OAI} software components that implement 4G \gls{LTE} and 5G \gls{NR}.
@@
     \item[Mango Communications 802.11:]
-    Finally, the Mango Communications project implements real-time 802.11 (WiFi) MAC and PHY in Xilinx \glspl{FPGA}.
-    It can be used on a variety of hardware platforms including \gls{USRP} \glspl{SDR}, and is employed in Ainur to provision end-to-end WiFi links.
+    Finally, the Mango Communications project implements real-time 802.11 (Wi-Fi) MAC and PHY in Xilinx\\\glspl{FPGA}.
+    It can be used on a variety of hardware platforms including \gls{USRP} \glspl{SDR}, and is employed in Ainur to provision end-to-end Wi-Fi links.
 \end{description}
@@
     \item[Configuration Layer:]
-    The lowest layer of Ainur. It handles the parsing of configuration files describing experimental scenarios.
+    The lowest layer of Ainur.
+    It handles the parsing of configuration files describing experimental scenarios.
     Optional, as it is only needed when Ainur is running experiments described in a declarative manner.
@@
     This layer interacts with hardware such as managed switches and \glspl{SDR} (and their associated computation hosts) to create links between the desired workload hosts.
-    These links correspond to ethernet, WiFi, and even an entire software-defined 4G \gls{LTE} and 5G networks.
+    These links correspond to ethernet, Wi-Fi, and even an entire software-defined 4G \gls{LTE} and 5G networks.
 
@@
 
-\subsection{Containerization}\label{ssec:containers}
+\subsection{Containerization}\label{paper:olguinmunoz2022airnur:containers}
 
@@
     \centering
-    \includegraphics[width=.8\columnwidth]{publications/2022Ainur/figures/network-minimal.png}
-    \caption{Minimal testbed setup on which Ainur can be deployed.}\label{fig:network:minimal}
+    \includegraphics[height=25em]{publications/2022Ainur/figures/network-minimal}
+    \caption{Minimal testbed setup on which Ainur can be deployed.}\label{paper:olguinmunoz2022airnur:fig:network:minimal}
 \end{figure}
@@
 
-A full-featured, and rather complex deployment is depicted in \cref{fig:network}, including a number of optional testbed features such as offload to the cloud through \gls{VPN} gateways.
-However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{fig:network:minimal}.
+A full-featured, and rather complex deployment is depicted in \cref{paper:olguinmunoz2022airnur:fig:network}, including a number of optional testbed features such as offload to the cloud through \gls{VPN} gateways.
+However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{paper:olguinmunoz2022airnur:fig:network:minimal}.
 The only requirements for the aforementioned hosts are that a general purpose Linux distribution such as Ubuntu can be deployed on them, and --- in the case of the workload hosts --- the presence of two network interfaces.

### 2023-02-03 publications/2022Ainur/demo.tex
fixing ainur paper

@@
     \centering
-    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/demo_hardware.png}
+    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/demo_hardware}
     \caption[caption]{%
@@
         \end{enumerate*}.
-    }\label{fig:democonfigs}
+    }\label{paper:olguinmunoz2022airnur:fig:democonfigs}
 \end{figure*}
@@
     \centering
-    \begin{subfigure}{\columnwidth}
+    \begin{subfigure}{.45\textwidth}
         \centering
-        \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_1.png}
-        \caption{Workload 1: \gls{MNIST} image classifier}\label{fig:wkld:mnist}
+        \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_1}
+        \caption{Workload 1: \acs{MNIST} image classifier}\label{paper:olguinmunoz2022airnur:fig:wkld:mnist}
     \end{subfigure}%
-    \vspace{1em}
-    \begin{subfigure}{\columnwidth}
+    \hfill%
+%    \vspace{1em}
+    \begin{subfigure}{.45\textwidth}
         \centering
-        \includegraphics[height=8em]{publications/2022Ainur/figures/demo_workload_2.png}
-        \caption{Workload 2: inverted pendulum \gls{NCS}}\label{fig:wkld:ncs}
+        \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_2}
+        \caption{Workload 2: inverted pendulum \acs{NCS}}\label{paper:olguinmunoz2022airnur:fig:wkld:ncs}
     \end{subfigure}
-    \caption{Demonstration workloads}\label{fig:wkld}
+    \caption{Demonstration workloads}\label{paper:olguinmunoz2022airnur:fig:wkld}
 \end{figure}
 
-\section{Demo}\label{sec:demo}
+\section{Demo}\label{paper:olguinmunoz2022airnur:demo}
 
@@
 
-\Cref{fig:democonfigs} illustrates the possible configurations for the testbed.
+\Cref{paper:olguinmunoz2022airnur:fig:democonfigs} illustrates the possible configurations for the testbed.
 Workloads are deployed on client hosts, and computation is offloaded either to an edge server (cloudlet), or to \gls{AWS} \gls{EC2} instances on the cloud.
@@
 \begin{description}
-    \item[WiFi:] \pgls{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board WiFi.
+    \item[WiFi:] \pgls{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board Wi-Fi.
     \item[4G \gls{LTE}:] \pgls{SDR} is configured as an 4G \gls{LTE} base station, and another is configured as an 4G \gls{LTE} \gls{UE}.
@@
 The number of workload instances (and therefore client hosts) deployed in each execution of this demonstration will range from \numrange[]{1}{10}.
-Workloads deployed to the cloud will be able to target any of \gls{AWS}'s datacenters.
+Workloads deployed to the cloud will be able to target any of \gls{AWS}s datacenters.
 
-We will employ two different workloads for this demonstration, illustrated in \cref{fig:wkld}.
-The first of these (\cref{fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{mnist}.
+We will employ two different workloads for this demonstration, illustrated in \cref{paper:olguinmunoz2022airnur:fig:wkld}.
+The first of these (\cref{paper:olguinmunoz2022airnur:fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{deng2012mnist}.
 The application consists of a client with a web interface, and \pgls{HTTP} server which responds to requests from the client and performs the actual image recognition.
@@
 
-The second workload (\cref{fig:wkld:ncs}) corresponds to \pgls{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \glspl{NCS} using cloud-native technologies~\cite{cleave}.
+The second workload (\cref{paper:olguinmunoz2022airnur:fig:wkld:ncs}) corresponds to \pgls{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \glspl{NCS} using cloud-native technologies~\cite{olguinmunoz2022cleave}.
 It consists of an emulation of the physical inverted pendulum system plant and a software-implemented proportional-differential controller.
@@
 \end{enumerate*}
-These nodes are interconnected using a combination of managed switches, \glspl{SDR}, and \gls{VPN} gateways; please refer to \cref{fig:network} for an architectural overview of this setup.
+These nodes are interconnected using a combination of managed switches, \glspl{SDR}, and \gls{VPN} gateways; please refer to \cref{paper:olguinmunoz2022airnur:fig:network} for an architectural overview of this setup.
 
@@
     \centering
-    \includegraphics[width=.65\columnwidth]{publications/2022Ainur/figures/demo_mockup.png}
+    \includegraphics[height=25em]{publications/2022Ainur/figures/demo_mockup}
     \caption{Screenshot of the user interface of demo workload 1.
-After deciding on the placement of the compute backend and the type of physical layer connecting client and backend, participants will use this interface to interact with the system.}\label{fig:mockup}
+After deciding on the placement of the compute backend and the type of physical layer connecting client and backend, participants will use this interface to interact with the system.}\label{paper:olguinmunoz2022airnur:fig:mockup}
 \end{figure}
@@
     \item a location to offload computation to (local (i.e.\ no offloading), edge, or any \gls{AWS} datacenter)
-    \item a physical layer for the first hop of the network (WiFi, 4G \gls{LTE}, or ethernet).
+    \item a physical layer for the first hop of the network (Wi-Fi, 4G \gls{LTE}, or ethernet).
 \end{enumerate*}
@@
 Next, participants will be able to access the client web interface and interact with the server by requesting classification of images from the \gls{MNIST} database.
-The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{fig:mockup}.
+The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{paper:olguinmunoz2022airnur:fig:mockup}.
 

### 2023-02-03 publications/2022Ainur/introduction.tex
fixing ainur paper

@@
 \end{enumerate*} testbeds.
-\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
+\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
 It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
-\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
+\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{ott2005orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
 
-\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
+\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{breen2020powder}.
 The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
@@
 
-Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
+Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{dandekar2019grid}.
 Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
@@
 
-To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
+To the best of our knowledge, CloudRAFT~\cite{moorthy2022cloudraft} is the only work to tackle (to a certain extent) this challenge.
 CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \glspl{VM} containing the necessary software for experiments.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds;
+and employs pre-built \glspl{VM} containing the necessary software for experiments.
 Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
@@
 The rest of this paper is structured as follows.
-\cref{sec:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
+\cref{paper:olguinmunoz2022airnur:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
 This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
-Next, \cref{sec:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
-Finally, in \cref{sec:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
+Next, \cref{paper:olguinmunoz2022airnur:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
+Finally, in \cref{paper:olguinmunoz2022airnur:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
 

### 2023-02-03 body/part1/background.tex
fixing formatting

@@
 
-\subsection{\glsfmtlong{WCA}}
+\subsection{\acl{WCA}}
 \glsreset{WCA}
@@
 
-\subsubsection{Characterization of \glsfmtshort{WCA}}
+\subsubsection{Characterization of \acs{WCA}}
 
@@
 If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}, label={(\arabic*)}]
+\begin{inlineenum}
     \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
@@
     \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
-\end{enumerate*}
+\end{inlineenum}
 These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
@@
 %The estimation of an event's duration varies with context of various types
-%\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
 %\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
@@
 %\item arousal can expand durations~\cite{droit_volet2011emotion}.
-%\end{enumerate*}
+%\end{inlineenum}
 %
@@
 
-\subsection{\glsfmtshortpl{NCS} and other \glsfmtshortpl{CPS}}
+\subsection{\acsp{NCS} and other \acsp{CPS}}
 
@@
 Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
+\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
     \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
     \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{inlineenum}
     \item \gls{COSMOS}
@@
     \item Drexel Grid \gls{SDR}
-\end{enumerate*} testbeds.
+\end{inlineenum} testbeds.
 \gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.

### 2023-02-03 publications/2019EdgeDroid/paper_body.tex
restructuring
fixing formatting

@@
 \cref{paper:olguinmunoz2019edgedroid:tbl:cloudletclienthardware} shows the hardware and software specifications for the cloudlet and the clients.
-The components communicated through a single consumer-grade WiFi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
+The components communicated through a single consumer-grade Wi-Fi (IEEE 802.11n, \SI{2.4}{\giga\hertz}) access point.
 We considered exclusively off-the-shelves hardware and software.
@@
 Our measurements were obtained from series of scenarios, repeated 100 times each.
-We performed three \emph{optimal} scenarios with 1, 5 and 10 clients each, the signal strength of the WiFi link being an excellent \SI{-40}{\dBm}.
+We performed three \emph{optimal} scenarios with 1, 5 and 10 clients each, the signal strength of the Wi-Fi link being an excellent \SI{-40}{\dBm}.
 Next, a scenario where the 10 clients were moved to another room roughly \SI{10}{\meter} away, thus degrading the network link to an average measured strength of \SI{-73}{\dBm}.
@@
 One, scaling behavior for the application seems to be linear with respect to the number of clients.
-Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
+Two, in the case of the impaired Wi-Fi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.
 

### 2023-02-03 publications/2021ImpactDelayedResponse/Manuscript.tex
fixing formatting

@@
 Accordingly, in the following we review background work in the areas of 
-\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
   \item[(\cref{ssec:timeperception})] time perception
   \item[(\cref{ssec:potentialmechs})] mechanisms relating human performance to delay.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 The estimation of an event's duration varies with context of various types
-\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
     \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
@@
     \item arousal can expand durations~\cite{droit2011emotion}.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 In particular
-\begin{enumerate*}[itemjoin={{, }},
+\begin{inlineenum}[itemjoin={{, }},
                   itemjoin*={{, and }},
@@
     \item each step has an intrinsic hierarchical cognitive structure (see \cref{fig:lego:hierarchical}), affording multiple levels of cognitive control. 
-\end{enumerate*}
+\end{inlineenum}
 
@@
 The participants wore devices to acquire four physiological measures:
-\begin{enumerate*}[itemjoin={{, }},
+\begin{inlineenum}[itemjoin={{, }},
                   itemjoin*={{, and }},
@@
   \item heart rate.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 The signals were sampled at \SI{200}{\hertz} and postprocessed to 
-\begin{enumerate*}[itemjoin={{, }},
+\begin{inlineenum}[itemjoin={{, }},
                   itemjoin*={{, and }},
@@
   \item smooth out noise (using the technique proposed by Agarwal et al.~\cite{agarwal2017eeg}).
-\end{enumerate*}
+\end{inlineenum}
 
@@
 We performed analyses where participants transitioned from either a short- or long-delay block into a
-\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
   \item no delay block (36 subjects)
@@
   \item \SI{3.0}{\second} delay block (37 subjects).
-\end{enumerate*}
+\end{inlineenum}
 The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.

### 2023-02-03 publications/2021ImpactDelayedResponse/body.tex
fixing formatting

@@
 Accordingly, in the following we review background work in the areas of 
-\begin{enumerate*}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}[label={}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
   \item[(\cref{paper:olguinmunoz2021impact:ssec:timeperception})] time perception
   \item[(\cref{paper:olguinmunoz2021impact:ssec:potentialmechs})] mechanisms relating human performance to delay.
-\end{enumerate*}
+\end{inlineenum}
 
@@
     \item arousal can expand durations~\cite{droit_volet2011emotion}.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{chen2015early}.
-This application belongs to a category of \gls{WCA}s designed to guide users through the execution of a sequential task.
+This application belongs to a category of \gls{WCA}'s designed to guide users through the execution of a sequential task.
 % Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
@@
     \item each step has an intrinsic hierarchical cognitive structure (see \cref{paper:olguinmunoz2021impact:fig:lego:hierarchical}), affording multiple levels of cognitive control.
-\end{enumerate*}
+\end{inlineenum}
 
@@
   \item \SI{3.0}{\second} delay block (37 subjects).
-\end{enumerate*}
+\end{inlineenum}
 The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
@@
 
-\subsection{\glsfmtshort{EEG}}
+\subsection{\acs{EEG}}
 Analyses were conducted on the log \gls{EEG} power in the alpha band, beta band, and total of all bands measured.
@@
 
-\subsection{\glsfmtshort{GSR} and Heart Rate}
+\subsection{\acs{GSR} and Heart Rate}
 The measure of \gls{GSR} was the log of total amplitude in the signal.

### 2023-02-03 publications/2022Ainur/ainur.tex
fixing formatting

@@
-\section{The Ainur Framework}\label{paper:olguinmunoz2022airnur:ainur}
+\section{The Ainur Framework}\label{paper:olguinmunoz2022ainur:ainur}
 
@@
     \includegraphics[width=\textwidth]{publications/2022Ainur/figures/overview}
-    \caption{Layered structure of an Ainur experiment}\label{paper:olguinmunoz2022airnur:fig:overview}
+    \caption{Layered structure of an Ainur experiment}\label{paper:olguinmunoz2022ainur:fig:overview}
 \end{figure}
@@
     \includegraphics[width=\textwidth]{publications/2022Ainur/figures/flow2}
-    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{paper:olguinmunoz2022airnur:fig:flow}
+    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{paper:olguinmunoz2022ainur:fig:flow}
 \end{figure*}
@@
     \includegraphics[width=\textwidth]{publications/2022Ainur/figures/network}
-    \caption{Network structure assumed by Ainur.}\label{paper:olguinmunoz2022airnur:fig:network}
+    \caption{Network structure assumed by Ainur.}\label{paper:olguinmunoz2022ainur:fig:network}
 \end{figure*}
@@
 Conceptually, in Ainur, an experiment is decomposed into a layered structure, consisting of
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item the distributed application (workload)
@@
     \item communication networks connecting the hosts
-\end{enumerate*}.
-In \cref{paper:olguinmunoz2022airnur:fig:overview}, the workload is represented by a client-server process pair.
+\end{inlineenum}.
+In \cref{paper:olguinmunoz2022ainur:fig:overview}, the workload is represented by a client-server process pair.
 These could correspond, for instance, to the emulation of an inverted pendulum and a matching controller.
@@
 Ainur can provision wired networks over ethernet, and supports a number of software-defined wireless communication stacks, currently including Wi-Fi, 4G \gls{LTE}, and 5G.
-To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \gls{USRP} \gls{SDR}s and some computation hosts dedicated to signal processing and/or radio management.
+To realize these various wireless communication protocols, it is assumed that the testbed is equipped with \gls{USRP} \gls{SDR}'s and some computation hosts dedicated to signal processing and/or radio management.
 
@@
     We leverage Docker containers to great extent for the virtualization and orchestration of workloads, as well as the encapsulation and management of complex network configurations.
-    See \cref{paper:olguinmunoz2022airnur:containers} for more details.
+    See \cref{paper:olguinmunoz2022ainur:containers} for more details.
 
@@
 
-\subsection{Containerization}\label{paper:olguinmunoz2022airnur:containers}
+\subsection{Containerization}\label{paper:olguinmunoz2022ainur:containers}
 
@@
 Ainur leverages containerization extensively across the framework, in order to
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{inlineenum}
     \item deploy and orchestrate workloads
@@
     \item support automated collection of logs
-\end{enumerate*}.
+\end{inlineenum}.
 Workloads in the framework are deployed packaged in Docker containers, orchestrated through \texttt{docker-compose} and Docker Swarm.
 We have selected Docker Swarm as our container orchestration layer instead of more advanced toolkits such as Kubernetes due to
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item it being more lightweight
@@
     \item the fact that it is included in the default Docker runtime installation
-\end{enumerate*}.
+\end{inlineenum}.
 This allows the framework to remain mostly agnostic to the nature of the workloads, and thus support a wide ranged of different applications and system architectures.
@@
     \includegraphics[height=25em]{publications/2022Ainur/figures/network-minimal}
-    \caption{Minimal testbed setup on which Ainur can be deployed.}\label{paper:olguinmunoz2022airnur:fig:network:minimal}
+    \caption{Minimal testbed setup on which Ainur can be deployed.}\label{paper:olguinmunoz2022ainur:fig:network:minimal}
 \end{figure}
@@
 
-A full-featured, and rather complex deployment is depicted in \cref{paper:olguinmunoz2022airnur:fig:network}, including a number of optional testbed features such as offload to the cloud through \gls{VPN} gateways.
-However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{paper:olguinmunoz2022airnur:fig:network:minimal}.
+A full-featured, and rather complex deployment is depicted in \cref{paper:olguinmunoz2022ainur:fig:network}, including a number of optional testbed features such as offload to the cloud through \gls{VPN} gateways.
+However, a minimal a setup can be achieved with a control host, at least two workload hosts, and a couple of managed (smart) switches, as illustrated in \cref{paper:olguinmunoz2022ainur:fig:network:minimal}.
 The only requirements for the aforementioned hosts are that a general purpose Linux distribution such as Ubuntu can be deployed on them, and --- in the case of the workload hosts --- the presence of two network interfaces.

### 2023-02-03 publications/2022Ainur/demo.tex
fixing formatting

@@
         Not pictured, in Phy Option 2:
-        \begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+        \begin{inlineenum}
             \item \gls{SDR} processing hosts
             \item additional base-station host for the Core Network and \gls{eNodeB}
-        \end{enumerate*}.
-    }\label{paper:olguinmunoz2022airnur:fig:democonfigs}
+        \end{inlineenum}.
+    }\label{paper:olguinmunoz2022ainur:fig:democonfigs}
 \end{figure*}
@@
         \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_1}
-        \caption{Workload 1: \acs{MNIST} image classifier}\label{paper:olguinmunoz2022airnur:fig:wkld:mnist}
+        \caption{Workload 1: \acs{MNIST} image classifier}\label{paper:olguinmunoz2022ainur:fig:wkld:mnist}
     \end{subfigure}%
@@
         \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_2}
-        \caption{Workload 2: inverted pendulum \acs{NCS}}\label{paper:olguinmunoz2022airnur:fig:wkld:ncs}
+        \caption{Workload 2: inverted pendulum \acs{NCS}}\label{paper:olguinmunoz2022ainur:fig:wkld:ncs}
     \end{subfigure}
-    \caption{Demonstration workloads}\label{paper:olguinmunoz2022airnur:fig:wkld}
+    \caption{Demonstration workloads}\label{paper:olguinmunoz2022ainur:fig:wkld}
 \end{figure}
 
-\section{Demo}\label{paper:olguinmunoz2022airnur:demo}
+\section{Demo}\label{paper:olguinmunoz2022ainur:demo}
 
@@
 
-\Cref{paper:olguinmunoz2022airnur:fig:democonfigs} illustrates the possible configurations for the testbed.
+\Cref{paper:olguinmunoz2022ainur:fig:democonfigs} illustrates the possible configurations for the testbed.
 Workloads are deployed on client hosts, and computation is offloaded either to an edge server (cloudlet), or to \gls{AWS} \gls{EC2} instances on the cloud.
@@
 \begin{description}
-    \item[WiFi:] \pgls{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board Wi-Fi.
+    \item[Wi-Fi:] \pgls{SDR} is configured as an IEEE 802.11n access point, and client hosts connect to it using on-board Wi-Fi.
     \item[4G \gls{LTE}:] \pgls{SDR} is configured as an 4G \gls{LTE} base station, and another is configured as an 4G \gls{LTE} \gls{UE}.
@@
 The number of workload instances (and therefore client hosts) deployed in each execution of this demonstration will range from \numrange[]{1}{10}.
-Workloads deployed to the cloud will be able to target any of \gls{AWS}s datacenters.
+Workloads deployed to the cloud will be able to target any of \gls{AWS}'s datacenters.
 
-We will employ two different workloads for this demonstration, illustrated in \cref{paper:olguinmunoz2022airnur:fig:wkld}.
-The first of these (\cref{paper:olguinmunoz2022airnur:fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{deng2012mnist}.
+We will employ two different workloads for this demonstration, illustrated in \cref{paper:olguinmunoz2022ainur:fig:wkld}.
+The first of these (\cref{paper:olguinmunoz2022ainur:fig:wkld:mnist}) consists of a proof-of-concept web application which implements a simple classifier to identify hand-drawn digits from the widely-used \gls{MNIST} dataset~\cite{deng2012mnist}.
 The application consists of a client with a web interface, and \pgls{HTTP} server which responds to requests from the client and performs the actual image recognition.
@@
 
-The second workload (\cref{paper:olguinmunoz2022airnur:fig:wkld:ncs}) corresponds to \pgls{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \glspl{NCS} using cloud-native technologies~\cite{olguinmunoz2022cleave}.
+The second workload (\cref{paper:olguinmunoz2022ainur:fig:wkld:ncs}) corresponds to \pgls{NCS} balancing an inverted pendulum, implemented on a software framework for the emulation of \glspl{NCS} using cloud-native technologies~\cite{olguinmunoz2022cleave}.
 It consists of an emulation of the physical inverted pendulum system plant and a software-implemented proportional-differential controller.
@@
 This demonstration will be performed on a testbed consisting of
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and finally }}]
+\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; and finally }}]
     \item \num{10} Raspberry Pi 4 Model B boards, acting as client-side workload hosts
@@
     \item a separate \texttt{i386} workstation hosting the Fluent server and on which Ainur is deployed as well.
-\end{enumerate*}
-These nodes are interconnected using a combination of managed switches, \glspl{SDR}, and \gls{VPN} gateways; please refer to \cref{paper:olguinmunoz2022airnur:fig:network} for an architectural overview of this setup.
+\end{inlineenum}
+These nodes are interconnected using a combination of managed switches, \glspl{SDR}, and \gls{VPN} gateways; please refer to \cref{paper:olguinmunoz2022ainur:fig:network} for an architectural overview of this setup.
 
@@
     \caption{Screenshot of the user interface of demo workload 1.
-After deciding on the placement of the compute backend and the type of physical layer connecting client and backend, participants will use this interface to interact with the system.}\label{paper:olguinmunoz2022airnur:fig:mockup}
+After deciding on the placement of the compute backend and the type of physical layer connecting client and backend, participants will use this interface to interact with the system.}\label{paper:olguinmunoz2022ainur:fig:mockup}
 \end{figure}
@@
 Participants will be asked to choose
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item a location to offload computation to (local (i.e.\ no offloading), edge, or any \gls{AWS} datacenter)
     \item a physical layer for the first hop of the network (Wi-Fi, 4G \gls{LTE}, or ethernet).
-\end{enumerate*}
+\end{inlineenum}
 Through the Ainur command line, we will deploy a single client-server pair according to the specifications.
 Next, participants will be able to access the client web interface and interact with the server by requesting classification of images from the \gls{MNIST} database.
-The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{paper:olguinmunoz2022airnur:fig:mockup}.
+The server will respond to these, and the assigned labels will be visible on the web interface together with timing statistics for each request, see \cref{paper:olguinmunoz2022ainur:fig:mockup}.
 

### 2023-02-03 publications/2022Ainur/introduction.tex
fixing formatting

@@
 Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the 
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{inlineenum}
     \item \gls{COSMOS}
@@
     \item Drexel Grid \gls{SDR}
-\end{enumerate*} testbeds.
+\end{inlineenum} testbeds.
 \gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
@@
 The rest of this paper is structured as follows.
-\cref{paper:olguinmunoz2022airnur:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
+\cref{paper:olguinmunoz2022ainur:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
 This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
-Next, \cref{paper:olguinmunoz2022airnur:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
-Finally, in \cref{paper:olguinmunoz2022airnur:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
+Next, \cref{paper:olguinmunoz2022ainur:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
+Finally, in \cref{paper:olguinmunoz2022ainur:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.
 

### 2023-02-03 publications/2022Ainur/software_layers/ip_layer.tex
glossary fixes
fixing formatting

@@
     This method takes a \mintinline{python}{PhysicalLayer} object instance and configures \gls{IP} layer connectivity on the hosts within.
-    This includes assigning \gls{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of WiFi.
+    This includes assigning \gls{IP} addresses to interfaces, configuring default and static routes, as well as authenticating and connecting to the desired access point in the case of Wi-Fi.
     The underlying implementation uses Ansible to write Netplan YAML configuration files on the target hosts and subsequently apply the network configuration.
@@
     \item The VPNCloud connections configured on the the gateways share the same pre-shared key.
-    \item The VPNCloud links belong to mutually \emph{disjoint} \acs{IP}v4 address ranges.
+    \item The VPNCloud links belong to mutually \emph{disjoint} \gls{IP}v4 address ranges.
     % \item The gateways are assigned the first host address in the available address space of their respective \gls{VPN} links.

### 2023-02-03 publications/2022Ainur/software_layers/physical.tex
fixing formatting

@@
 This layer creates and deploy the underlying physical connections of the data network and is implemented in the \mintinline{python}{PhysicalLayer} class. 
-Types of connections handled by this layer could be ethernet and WiFi links, or even an entire software-defined 4G \gls{LTE} network.
+Types of connections handled by this layer could be ethernet and Wi-Fi links, or even an entire software-defined 4G \gls{LTE} network.
 
@@
 
-Currently, WiFi and \gls{LTE} implementations are available.
+Currently, Wi-Fi and \gls{LTE} implementations are available.
 User-defined implementation for other wireless connections are possible and encouraged.
@@
 \begin{description}[style=nextline]
-    \item[SDR-based WiFi Network]
+    \item[SDR-based Wi-Fi Network]
     The \mintinline{python}{SDRWifi} class handles the configuration and deployment of a software-defined IEEE 802.11 network using the Mangocomm software package and its supported software-defined radios \footnote{See https://mangocomm.com/802.11-mac-phy/}.
     It lunches a Docker container on the same machine to communicate with the \glspl{SDR} over the control network.
-    This communication is for the configuration and monitoring the WiFi nodes.
-    In a WiFi network, a radio is either \emph{station} or \emph{access point}.
+    This communication is for the configuration and monitoring the Wi-Fi nodes.
+    In a Wi-Fi network, a radio is either \emph{station} or \emph{access point}.
     To initialize a \mintinline{python}{SDRWifi} class, the user is required to provide two lists containing the access points and station nodes instances as described below.
 
-    \mintinline{python}{APSoftwareDefinedRadio} class represents a WiFi access point.
+    \mintinline{python}{APSoftwareDefinedRadio} class represents a Wi-Fi access point.
     It advertises a network with a specific SSID on a radio channel and waits for the incoming connections.
 
-    \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station WiFi nodes.
-    It searches for an WiFi \gls{SSID} among the networks that are being advertised and attempts for establishing the connection.
+    \mintinline{python}{StationSoftwareDefinedRadio} class is used for the station Wi-Fi nodes.
+    It searches for an Wi-Fi \gls{SSID} among the networks that are being advertised and attempts for establishing the connection.
 
-    When a WiFi network is up and an station is connected, the data network interfaces of the radios behave as if an ethernet cable is connecting them.
+    When a Wi-Fi network is up and an station is connected, the data network interfaces of the radios behave as if an ethernet cable is connecting them.
     Thus, the client hosts and the server hosts can be on the same ip layer subnets.
@@
         \includegraphics[width=0.8\linewidth]{figures/wifi.png}
-        \caption{An example of WiFi network topology in data network and its corresponding \glspl{VLAN}}
+        \caption{An example of Wi-Fi network topology in data network and its corresponding \glspl{VLAN}}
         \label{fig:wifi}
@@
     The resulting topology of an \gls{LTE} network in Ainur is depicted in Figure \ref{fig:lte}.
-    Unlike WiFi wireless connection, due to the fact that \gls{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.
+    Unlike Wi-Fi wireless connection, due to the fact that \gls{LTE} creates an internal ip network and that introduces routing complexities, the server host and the client host cannot be on the same subnet in the ip layer.
 

### 2023-02-03 publications/2022CLEAVE/conclusion.tex
fixing formatting

@@
 We argue such a use case represents a realistic scenario and appropriate benchmark for the tool, since
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item the inverted pendulum plant is ubiquitous in \gls{NCS} research
@@
     \item video analytics has long been proposed as a ``killer app'' for edge computing.
-\end{enumerate*}
+\end{inlineenum}
 Our results showcase the ability of the framework to extract relevant metrics relating to the stability of the control system, as well as on the performance of the underlying network link.

### 2023-02-03 publications/2022CLEAVE/experiments.tex
fixing formatting

@@
 A simple \gls{NCS} consisting of an \emph{inverted pendulum} plant controlled by a \emph{proportional-differential} controller is to be deployed on an Edge server.
-The connection between plant and server is over a WiFi (IEEE 802.11n) link.
+The connection between plant and server is over a Wi-Fi (IEEE 802.11n) link.
 Moreover, there are a number of video-streaming applications running concurrently on the Edge setup.
@@
 Secondly, similar, albeit non-Edge-enabled, setups have been a reality in industry for almost two decades, as evidenced by\ \cite{gupta2010networked}.
-Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \glspl{RTT} offered by these technologies are higher, or at best equal, those offered by WiFi.
-4G, for instance, has average \glspl{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \gls{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for WiFi.
-Thus, we argue that using WiFi allows for more experimental freedom, as the system can be tweaked to obtain worse \glspl{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
+Thirdly, although real deployments would most assuredly employ mobile technologies due to the added benefits of mobility and range, the \glspl{RTT} offered by these technologies are higher, or at best equal, those offered by Wi-Fi.
+4G, for instance, has average \glspl{RTT} of around a few hundred milliseconds, which is orders of magnitude greater than the values measured in our scenario, and current 5G deployments offer \gls{RTT} values barely matching the sub-\SI{10}{\milli\second} values we obtain for Wi-Fi.
+Thus, we argue that using Wi-Fi allows for more experimental freedom, as the system can be tweaked to obtain worse \glspl{RTT} more akin to those of mobile technologies while still having the option to study best-case scenarios with very low-latencies.
 Finally, video analytics is one of the main proposed use cases for edge computing~\cite{ananthanarayanan2017real,yi2017lavea,wang2018bandwidth}, and thus we foresee edge \gls{NCS} deployments being deployed in parallel with such applications in the future.
@@
 For both of these setups, we execute a series of scenarios varying:
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item the sampling rate of the Plant state, setting it to \SIlist[list-final-separator={, or }]{5;10;20;40;60}{\hertz}
     \item the responsiveness of the Controller, adding fixed delays of \SIlist[list-final-separator={, or }]{0;25;50}{\milli\second} after the processing of each sample.
-\end{enumerate*}
+\end{inlineenum}
 
@@
 
-This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \gls{CLEAVE}s design.
+This repeated experimentation and data collection while ``zooming in'' on particular setups is facilitated by \gls{CLEAVE}'s design.
 Scenarios are executed automatically in batches using a simple Python script which interacts with Docker through the widely adopted \emph{docker-py}\footnote{Docker SDK for Python: \url{https://docker-py.readthedocs.io/en/stable/}} library.
-This is \gls{CLEAVE}s first advantage over existing frameworks;
+This is \gls{CLEAVE}'s first advantage over existing frameworks;
 it is designed with cloud and edge technology and paradigms in mind, making integration with existing systems convenient.

### 2023-02-03 publications/2022Ainur/demo.tex
fixing figs

@@
         \centering
-        \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_1}
+        \includegraphics[width=\textwidth]{publications/2022Ainur/figures/demo_workload_1}
         \caption{Workload 1: \acs{MNIST} image classifier}\label{paper:olguinmunoz2022ainur:fig:wkld:mnist}
@@
         \centering
-        \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/demo_workload_2}
+        \includegraphics[width=\textwidth]{publications/2022Ainur/figures/demo_workload_2}
         \caption{Workload 2: inverted pendulum \acs{NCS}}\label{paper:olguinmunoz2022ainur:fig:wkld:ncs}

### 2023-02-03 publications/2022CLEAVE/experiments.tex
fixing figs

@@
     \centering
-    \begin{subfigure}[h]{.5\textwidth}
+    \begin{subfigure}[h]{.45\textwidth}
         \centering
-        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
         \caption{Toppled plants.}\label{paper:olguinmunoz2022cleave:fig:video:toppled}
     \end{subfigure}%
-%    \hfill%
-    \begin{subfigure}[h]{.5\textwidth}
+    \hfill%
+    \begin{subfigure}[h]{.45\textwidth}
         \centering
-        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
         \caption{Angle \gls{RMS}.}\label{paper:olguinmunoz2022cleave:fig:video:rms}
     \end{subfigure}\\
-    \begin{subfigure}[h]{.5\textwidth}
+    \begin{subfigure}[h]{.45\textwidth}
         \centering
-        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
         \caption{Packet losses.}\label{paper:olguinmunoz2022cleave:fig:video:drop}
@@
 %    \hfill%
-    \begin{subfigure}[h]{.5\textwidth}
+    \begin{subfigure}[h]{.45\textwidth}
         \centering
-        \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
         \caption{\glspl{RTT}.}\label{paper:olguinmunoz2022cleave:fig:video:rtt}

### 2023-02-03 model.tex
thesis tweaks

@@
 For each video frame submitted to the \gls{WCA} during the tasks, we recorded
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item the raw video frame captured
@@
     \item a tag representing the result of the \gls{WCA} processing
-\end{enumerate*}.
+\end{inlineenum}.
 The tags assigned corresponded to:
@@
 For each step, given
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     \item a collection of random non-\texttt{SUCCESS}, non-\texttt{REPEAT} frames (at least one frame for each of the \texttt{BLANK} and \texttt{LOW\_CONFIDENCE} categories)
@@
     \item an appropriate \texttt{REPEAT} video frame containing the correct state for the \emph{previous} step
-\end{enumerate*},
+\end{inlineenum},
 we can then procedurally generate a trace by randomly selecting appropriate frames according to the distributions presented in \cref{fig:frameprobs}.

### 2023-02-03 sampling.tex
thesis tweaks

@@
 For this, we make use of the facts that
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{inlineenum}
     \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
     \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
-\end{enumerate*}
+\end{inlineenum}
 \begin{alignat*}{1}
@@
 Since any positive real value is valid for this ratio, and since the sampling instants are aperiodic and can take any positive real values, we have
-\begin{enumerate*}
+\begin{inlineenum}
     \item the bound will be tight at the optimum due to the continuous nature of possible sampling instants and sampling intervals
     \item one can achieve any arbitrary point $(\mathbb{E}[\mathcal{S}],\mathbb{E}[\mathcal{W}])$ via the unconstrained optimization discussed above simply by varying the ratio $\frac{\alpha}{\beta}$.
-\end{enumerate*}
+\end{inlineenum}
 Thus, can argue that any optimisation problem -- constrained or unconstrained -- can be boiled down to modifying the problem to the general format and finding the appropriate $\frac{\alpha}{\beta}$ ratio.

### 2023-02-04 body/part1/background.tex
writing

@@
 \section{Background \& related work}
+\glsresetall%
 \todo[inline]{%
@@
     Next, what applications benefit from the Edge?
-    \gls{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
+    \glsfmtshort{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
     Main points: prototypes and characterization of system latencies and load.
@@
 
-However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive applications.
+However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive and/or bandwidth-intensive applications.
 In order to achieve the necessary economies of scale, cloud datacenters are designed to serve users distributed across vast geographical areas.
@@
 for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
-This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{satyanarayanan2009case,varghese2016challenges,shi2016promise}.
+This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{tolia2006quantifying,lagar2007interactive,satyanarayanan2009case,varghese2016challenges,shi2016promise}.
 The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
 Such latencies are unfeasible to consistently achieve with cloud computing~\cite{dang2021cloudy}.
-
-Current cloud infrastructures are also ill-suited to deal with the bandwidth requirements of the \gls{IOT} and of mobile computing.
-As smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face significant challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
+On the other hand, as smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face increasing challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
 
@@
 
-Edge computing, emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
-This term refers to a novel distributed computing paradigm, complementary to cloud computing, which tackles the challenges of the cloud by moving computation closer to where it is needed.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter --- a \emph{cloudlet}~\cite{satyanarayanan2009case} --- at the \emph{edge} of the network.
-These coudlets feature most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, one or two hops away from the user (see \cref{fig:edgecomputing})
+\medskip
+Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
+The conceptual foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
+As an answer to the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
+Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
+\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
+This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
+Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure.
+This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.
 
-The architecture of edge computing offers several advantages.
-Thanks to their proximity to users, cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Mobile \gls{XR}, such as \gls{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application}, and 5G-enabled \glspl{CPS} and \glspl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient} are uniquely poised to take advantage of these characteristic of edge computing.
-By processing and aggregating data closer to where its needed, cloudlets can also drastically reduce the amount of traffic going to cloud datacenters.
+The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
+Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
+Other advantages include bandwidth reduction and improved privacy.
+By processing and aggregating data closer to where its needed, edge computing can drastically reduce the amount of traffic going to cloud datacenters.
 Cloudlets also present opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
 
-\todo[inline]{Transition to talk about research and tools for Edge computing research.}
+In later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
+This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
+Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \acl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
+These applications had been hitherto inviable to implement due to the limitations of cloud offloading, but have now become a practical reality thanks to \gls{MEC}.
+We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
 
-\subsection{\acl{WCA}}
+Finally, we must note a closely related concept; \emph{fog} computing.
+Introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, fog computing concerns itself with the distribution of computing power between the cloud and the edge.
+This stands in contrast to the stated primary goal of edge computing, which is the reduction of latency and jitter by offloading computation to infrastructure as close to the user as possible.
+The motivating factor for the fog paradigm is the aggregation of data and reduction of bandwidth requirements for the \gls{IOT}.
+By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive sensor networks~\cite{yi2015survey}.
+
+\subsection{\acl{WCA}}\label{background:wca}
 \glsreset{WCA}
@@
 
-\subsection{\acsp{NCS} and other \acsp{CPS}}
+\subsection{\acsp{NCS}}\label{background:ncs}
 

### 2023-02-04 body/part1/background.tex
opt

@@
 
+\todo[inline]{Something feels missing here}
+
 Finally, we must note a closely related concept; \emph{fog} computing.
@@
 This kind of compute- and energy-intensive processing is unfeasible to be performed on wearable devices, and thus these applications necessarily make use of compute offloading.
-As explained above, the cloud is unsuitable as an offloading point for these applications, given the high latencies incurred, and there is thus a broad understanding today that with the advent of edge computing, \glspl{WCA} will become a practical reality~\cite{bittmann2017edge,flinn2012cyber,chen2017empirical,ha2013just,wang2020scaling,chen2018application,ha2014towards}.
+As explained above, the cloud is unsuitable as an offloading point for these applications, given the high latencies incurred, and there is thus a broad understanding today that with the advent of edge computing, \glspl{WCA} will become a practical reality~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.
 
-There's more to end-to-end latency than merely the question of where the compute backend is placed, however.
+\medskip
+In literature, research on \gls{WCA} has focused mostly on the implementation of prototype applications and platforms.
+\citeauthor{ha2014towards}~\cite{ha2014towards} coined the term \acl{WCA} in\ \citeyear{ha2014towards}.
+Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
+\cite{chen2015early} and~\cite{chen2018application} further develop this platform, and produce a number of different assistive applications such as the aforementioned LEGO and IKEA assistants, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis.
+Yet another platform for \gls{WCA} is introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
+Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.
+
+In later years, an body of work studying the optimization of these applications has emerged.
+A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
+\citeauthor{wang2019towards}~\cite{wang2019towards} explore strategies for scalable \gls{WCA} on the edge through different adaptive sampling and resource allocation schemes;
+this is further developed in the authors' PhD thesis later on~\cite{wang2020scaling}.
+\cite{moothedath2021energy},\ \cite{moothedath2022energy1}, and~\cite{moothedath2022energy2} studied the potential for energy optimization in \gls{WCA} through convex optimization of the sampling strategies employed.
+In~\cite{george2020openrtist}, the authors introduce \emph{OpenRTiST}, a benchmarking tool for immersive, latency-sensitive, and bandwidth-hungry applications on edge computing leveraging \glspl{DNN} to perform real-time style transfer on a live video feed.
+
+The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
+There's more to end-to-end latency, \gls{WCA} performance, and scalability than merely the question of where the compute backend is placed, however.
 The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
 These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
-Existing studies of this class of applications have only recently started to delve more deeply upon these issues~\cite{chen2017empirical,wang2019towards,wang2020scaling,ha2014towards,chen2015early,satyanarayanan2009case,chatzopoulos2017hyperion}.
-Recently published models for end-to-end latency of edge computing architectures, on the other hand, are quite complex, while not accounting for the specifics of human-in-the-loop applications and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite,moothedath2022energy2,moothedath2022energy1,moothedath2021energy}.
-\todo[inline]{%
-    Maybe also present ``other'' research on WCA here? E.g.\ other research in WCA has focused on... So-and-so have... blah blah.
-}
-The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
+As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.add
+Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
+Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.
 

### 2023-02-04 body/part1/background.tex
great

@@
 Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
-The conceptual foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
-As an answer to the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
+The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
+To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
 Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
@@
 This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
-Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure.
+Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
 This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.
@@
 Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Other advantages include bandwidth reduction and improved privacy.
-By processing and aggregating data closer to where its needed, edge computing can drastically reduce the amount of traffic going to cloud datacenters.
-Cloudlets also present opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
+Another advantage is bandwidth reduction.
+%By processing and aggregating data closer to where its needed, edge computing can significantly reduce the amount of traffic going to cloud datacenters.
+A variant of edge computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for this purpose.
+By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
+Edge computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
 
-In later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
+Finally, in later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
 This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
-Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \acl{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
-These applications had been hitherto inviable to implement due to the limitations of cloud offloading, but have now become a practical reality thanks to \gls{MEC}.
+Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
+These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
 We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
-
-\todo[inline]{Something feels missing here}
-
-Finally, we must note a closely related concept; \emph{fog} computing.
-Introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, fog computing concerns itself with the distribution of computing power between the cloud and the edge.
-This stands in contrast to the stated primary goal of edge computing, which is the reduction of latency and jitter by offloading computation to infrastructure as close to the user as possible.
-The motivating factor for the fog paradigm is the aggregation of data and reduction of bandwidth requirements for the \gls{IOT}.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive sensor networks~\cite{yi2015survey}.
 
@@
 They represent a particular class of \gls{AR} applications, aiming to amplify human cognition in both day-to-day tasks and professional settings.
-\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
-These systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
-More recently, their scope has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards,ha2014towards}.
-Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}, and it is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
-\todo[inline]{Develop a bit further. Talk about more reactive setups like ping pong too.}
 
@@
 
-We identify a number of defining characteristics of \gls{WCA}.
-One, and as the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
+\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
+These applications operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
+Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
+The input is parsed to an internal symbolic representation of the state of the task;
+whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
+This feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
+If no change in state is detected, the application remains silent and out-of-the-way.
+In other words, the application silently discards samples which capture an intermediate or unfinished state, or simply are too noisy to be processed.
+\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
+
+As the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
 they are pervasive and mobile~\cite{ha2014towards}.
+This leads to interesting consequences for \gls{WCA} system design when combined with their requirement of seamless integration with the context of the user.
+The wearable nature of \gls{WCA} directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
+The requirements of immersive and seamless interaction, on the other hand, suggest a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
+This kind of compute- and energy-intensive processing, often involving \gls{ML} algorithms and \glspl{DNN}, is unfeasible to be performed on wearable devices, and thus these applications \emph{must} necessarily make use of compute offloading~\cite{ha2014towards,wang2020scaling}.
 
-Next, interaction with \glspl{WCA} is immersive and seamless.
-As mentioned above, \glspl{WCA} operate in a manner similar to \gls{GPS} guidance systems.
-Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
-The input is then parsed to an internal symbolic representation of the state of the task, and whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
-Feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
-If no change in state is detected, the application remains silent and out-of-the-way of the user;
-in other words, samples which capture an intermediate or unfinished state, or simply noise are silently discarded.
-\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
-Assistants should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
-\gls{WCA} is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
-
-Note that this mode interaction implies seamless integration of the application with the context of the user and the real world.
-The user does not trigger an update explicitly at any point, leading to one of the key characteristics of \gls{WCA}.
-The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
-An important corollary of this is that these are therefore the \emph{only} points in time at which users can notice changes in system responsiveness.
-
-Finally, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand) to the user.
-This goes hand in hand with the previous characteristic of seamless interaction, as users will have expectations of constant and immediate feedback as they interact with the system.
+However, it is understood today the cloud represents an unsuitable candidate for offloading computation in these applications, given the high latencies involved.
+Given their immersive nature, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand), as users will have expectations of constant and immediate feedback as they interact with the system.
 As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
 Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).
-
-The above characteristics together lead to interesting consequences for \gls{WCA} system design.
-Their wearable nature directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
-The requirement of immersive and seamless interaction suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-This kind of compute- and energy-intensive processing is unfeasible to be performed on wearable devices, and thus these applications necessarily make use of compute offloading.
-As explained above, the cloud is unsuitable as an offloading point for these applications, given the high latencies incurred, and there is thus a broad understanding today that with the advent of edge computing, \glspl{WCA} will become a practical reality~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.
+There is therefore a broad understanding today that edge computing is a key enabling technology for \gls{WCA} applications~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.
 
@@
 Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
-\cite{chen2015early} and~\cite{chen2018application} further develop this platform, and produce a number of different assistive applications such as the aforementioned LEGO and IKEA assistants, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis.
-Yet another platform for \gls{WCA} is introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
+The authors were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
+More recently, the scope of \glspl{WCA} has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards}.
+\cite{chen2015early} and~\cite{chen2018application} further developed the platform described in~\cite{ha2014towards}, and produced a number of different assistive applications such as the aforementioned LEGO assistant, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis, and more advanced assembly assistant for IKEA furniture.
+Yet another platform for \gls{WCA} was introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
 Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.
+Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
+It is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
 
-In later years, an body of work studying the optimization of these applications has emerged.
+In later years, a body of work studying the optimization of these applications has emerged.
 A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
@@
 These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
-As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.add
+As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.
+On the other hand, recently published models for end-to-end latency of edge computing architectures, are quite complex, while not accounting for the specifics of \gls{WCA} and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite}.
 Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
@@
 From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
+However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
+The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
+A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
 If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then

### 2023-02-06 body/part1/background.tex
writing

@@
 
-\subsection{\acl{WCA}}\label{background:wca}
-\glsreset{WCA}
-
-There is increasing interest from academia and industry in a novel category of wearable and mobile immersive applications called \gls{WCA}~\cite{ha2014towards,chen2018application,chen2015early,belletier2021wearable,chen2017empirical,wang2020scaling,chatzopoulos2017hyperion}
-They represent a particular class of \gls{AR} applications, aiming to amplify human cognition in both day-to-day tasks and professional settings.
+\subsection{\glsfmtlong{WCA}}\label{background:wca}
 
@@
     \caption{%
-        Mode of operation of \gls{WCA} applications.
+        Mode of operation of \acs{WCA} applications.
         The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
@@
 
-\glspl{WCA} operate in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
+\glsreset{WCA}
+\gls{WCA} refers to a novel category of wearable and mobile immersive applications which aim to amplify human cognition in both day-to-day tasks and professional settings.
+\glspl{WCA} work in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
 These applications operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
@@
 
-\subsection{\acsp{NCS}}\label{background:ncs}
+\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
+\todo[inline]{Needs citations.}
 
-The number and applications of \glspl{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
-However, this rapid increase in adoption has mostly been limited to industrial contexts.
-Although \glspl{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
-This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \glspl{CPS} will be made possible.
-These technologies meet two key requirements of \glspl{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
+\begin{figure}
+    \centering
+    \begin{subfigure}[t]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{Figs/control_system}
+        \caption{%
+            Traditional control system.
+        }
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{0.50\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{Figs/networked_control_system}
+        \caption{%
+            \acl{NCS}.
+        }
+    \end{subfigure}
+    \caption{%
+        Comparison between traditional control systems and \aclp{NCS}.
+        In an \aclp{NCS}, controller and plant are physically separated and both actuation commands and sensor inputs are shared through a general purpose \acs{TCP}/\acs{IP} network.
+    }\label{fig:csvsncs}
+\end{figure}
 
-\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
-This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
+\glsreset{NCS}
+\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
+A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
+Control systems are composed of
+\begin{inlineenum}
+    \item a \emph{plant}
+    \item \emph{sensors}
+    \item \emph{actuators}
+    \item a \emph{controller}
+\end{inlineenum}.
+Plant refers to the physical system being controlled.
+Sensors sample and encode the state of the plant and transmit it to the controller.
+The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
+The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
+
+\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
+see \cref{fig:csvsncs}.
+This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
+\glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
+This leads to enhanced scalability and flexibility, and allows improved coordination and collaboration among systems.
+Furthermore, \glspl{NCS} provide greater access to information, as they enable the exchange of data and information between control components and systems, leading to more informed decision-making and improved performance.
+
+\glspl{NCS} have become increasingly popular in recent years due to advances in communication and control technologies, as well as the increasing demand for real-time, remote, and distributed control.
+They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the cloud for centralized, distributed control in what has been called \emph{cloud control systems}~\cite{xia2015cloud}
+However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms simply cannot meet~\cite{wan2020efficient}.
+This has led their adoption to be limited mostly to industrial environments.
+This is about to change, however, as with the advent of edge computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
+These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.
+
+\medskip
+Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
+Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
+They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
+\todo[inline]{TODO FINISH}
+
+%\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
+%Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
+%This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
 
@@
 
-One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
-%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
-On the other, there exist several approaches when considering experimental methodologies.
-%A number of works concerning \glspl{NCS} deal with experimental studies.
-%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
-One such approach uses setups in which the complete system is built on top of real hardware.
-This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
-The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
-Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
-    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-\end{inlineenum}
+%One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
+%%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
+%On the other, there exist several approaches when considering experimental methodologies.
+%%A number of works concerning \glspl{NCS} deal with experimental studies.
+%%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
+%One such approach uses setups in which the complete system is built on top of real hardware.
+%This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
+%Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
+%The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
+%These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
+%Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
+%\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
+%    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
+%    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
+%\end{inlineenum}
+%
+%As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
+%This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
+%Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
+%This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
+%
+%Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
+%In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
+%Their methodology utilizes joint knowledge of control, computation, and communication.
+%In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
+%Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
+%This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
+%Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
 
-As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
-This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
-This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
+\subsubsection{Related methodologies and tools}
 
-Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
-In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
-Their methodology utilizes joint knowledge of control, computation, and communication.
-In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
-Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
-This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
-Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
+The work we present in this thesis lies at the intersection of research on edge computing, \gls{WCA} applications, and \acl{CPS} such as \glspl{NCS}, each of these topics representing a vast body of research.
+Of principal interest to us are works dealing with the optimization of \gls{WCA} and \glspl{NCS} deployed on distributed systems such as edge computing.
+Below we will summarize the key milestones in literature relating to this.
 
-\subsubsection{Benchmarking tools for edge computing}
+\paragraph{Experimental platforms for Edge Computing}
 

### 2023-02-07 body/glossary.tex
fix impact paper
relwork
great
fixing glossary problems agian
writing
exp edge research

@@
-%! suppress = MissingImport
+\usepackage{glossaries}%! suppress = MissingImport
 \newacronym[longplural={Wearable Cognitive Assistants}]{WCA}{WCA}{Wearable Cognitive Assistance}
@@
 \newacronym{MaaS}{MaaS}{Metal-as-a-Service}
-\newacronym{OEDL}{OEDL}{\glsfmtshort{OMF} Experiment Description Language}
+\newacronym{OEDL}{OEDL}{\acs{OMF} Experiment Description Language}
 \newacronym{RSpec}{RSpec}{Resource Specification}
@@
 \newacronym[longplural={Times-to-Feedback}]{TTF}{TTF}{Time-to-Feedback}
-\newacronym{MAR}{MAR}{Mobile \glsfmtshort{AR}}
+\newacronym{MAR}{MAR}{Mobile \acs{AR}}
 \newacronym{QoE}{QoE}{Quality of Experience}
@@
 \newacronym{CDF}{CDF}{Cumulative Distribution Function}
-\newacronym{CCDF}{CCDF}{Complementary \glsfmtshort{CDF}}
+\newacronym{CCDF}{CCDF}{Complementary \acs{CDF}}
 \newacronym{PDF}{PDF}{Probability Density Function}
-\newacronym{ECDF}{ECDF}{Empirical \glsfmtshort{CDF}}
+\newacronym{ECDF}{ECDF}{Empirical \acs{CDF}}
 \newacronym{NSF}{NSF}{United States National Science Foundation}
@@
 \newacronym{SEM}{SEM}{Standard Error of the Mean}
+\newacronym{SD}{SD}{Standard Deviation}
+\newacronym{BPM}{BPM}{Beats Per Minute}
+\newacronym{HR}{HR}{Heart Rate}
+\newacronym{IOT}{IoT}{Internet of Things}
+\newacronym{ML}{ML}{Machine Learning}
+\newacronym{MPC}{MPC}{Model-Predictive Control}
+\newacronym{USA}{USA}{United States of America}
+\newacronym{CLOUDRAFT}{CloudRAFT}{Cloud-based Remote Access For wireless Testbeds.}
+\newacronym{FIWI}{FiWi}{Fiber-Wireless}
+\newacronym{GENI}{GENI}{Global Environment for Networking Innovation}
 
-\glsaddall%
+%\glsaddall%

### 2023-02-07 body/part1/background.tex
exp edge research

@@
 The work we present in this thesis lies at the intersection of research on edge computing, \gls{WCA} applications, and \acl{CPS} such as \glspl{NCS}, each of these topics representing a vast body of research.
-Of principal interest to us are works dealing with the optimization of \gls{WCA} and \glspl{NCS} deployed on distributed systems such as edge computing.
+Of principal interest to us are works dealing with the experimental optimization of \gls{WCA} and \glspl{NCS} deployed on distributed systems such as edge computing.
 Below we will summarize the key milestones in literature relating to this.
@@
 
-Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications.
-On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
-The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.
+Research in edge computing is still in its early stages, although it has been steadily increasing in the last decade as researchers, businesses, and organizations recognize the benefits of bringing computing and storage capabilities closer to the source of data.
+Standardization is however still a major unsolved challenge in this field, and thus most research hitherto has relied on ad-hoc methodologies and/or platforms.\todo{Citation}
 
-Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
+Nonetheless, the last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
+Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
+Of particular interest to us are the
 \begin{inlineenum}
-    \item \gls{COSMOS}
-    \item \gls{POWDER}
-    \item \gls{ARA}
-    \item Drexel Grid \gls{SDR}
+    \item \acs{GENI}
+    \item Chameleon
+    \item \acs{COSMOS}
+    \item \acs{POWDER}
+    \item \acs{ARA}
+    \item EdgeNet
+%    \item Drexel Grid \gls{SDR}
 \end{inlineenum} testbeds.
-\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
-It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
-\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
 
-\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
-The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
-\gls{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \gls{VM} images containing the necessary software and configurations.
-These profiles are defined through using \gls{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
+\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
+Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
+The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
 
-The \gls{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
-It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-\gls{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \gls{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
-This affords the \gls{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
+The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
+Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
+The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
 
-Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
-Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
-However, the testbed extensively employs the \gls{LXC} runtime for the deployment of both experimental code and \gls{SDR} software, which affords users great freedom when it comes to the development of experiments.
+\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
+The testbeds main \emph{raison d'etre} is \gls{MEC} research, with a strong emphasis on wireless networks.
+The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
 
-Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
-Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
-Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
-This is, for instance, the case with \gls{COSMOS} and \gls{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
-These testbeds further leverage virtualization technology based on \glspl{VM} instead of more lightweight and edge-compatible solutions such as containers.
+\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
+It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
+These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
 
-To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
-CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \glspl{VM} containing the necessary software for experiments.
-Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
-Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \pgls{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.
+The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
+\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
+Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+
+EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
+It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
+The EdgeNet software is available for anyone to download;
+individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
+The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
+A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
+
+Finally, a number of other, small-scale approaches also exist.
+In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
+The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
+This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
+In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
+Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.

### 2023-02-07 body/part1/background.tex
fixing bookmarks;

@@
 We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
+
+\subsubsection{Experimental research in Edge Computing}
+
+The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
+Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
+Of particular interest to us are the
+\begin{inlineenum}
+    \item \acs{GENI}
+    \item Chameleon
+    \item \acs{COSMOS}
+    \item \acs{POWDER}
+    \item \acs{ARA}
+    \item EdgeNet
+%    \item Drexel Grid \gls{SDR}
+\end{inlineenum} testbeds.
+
+\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
+Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
+The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
+
+The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
+Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
+The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
+
+\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
+The testbeds main \emph{raison d'etre} is \gls{MEC} research, with a strong emphasis on wireless networks.
+The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
+
+\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
+It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
+These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
+
+The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
+\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
+Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+
+EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
+It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
+The EdgeNet software is available for anyone to download;
+individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
+The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
+A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
+
+Finally, a number of other, small-scale approaches also exist.
+In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
+The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
+This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
+In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
+Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
 
@@
 They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-\todo[inline]{TODO FINISH}
+
+In terms of research relating to the benchmarking and optimization of these systems, works in literature have for the most part leveraged either theoretical models~\cite{zhang2019networked} or simulations (for example ~\cite{ma2019optimal}).
+However, these approaches can only capture network behaviors at a coarse level, and thus a body of works has emerge in later years involving experimental approaches to \gls{NCS} benchmarking.
+\cite{baumann2018evaluating} and~\cite{cuenca2019periodic} employ purpose-built physical testbeds for the study of these systems.
+\cite{zoppi2020ncsbench} introduces \emph{NCSBench}, a platform for reproducible benchmarking of \gls{NCS} using a LEGO Mindstorms platform.
+
+Finally, hybrid approaches, mixing testbed or prototype implementations and simulations or emulations of \gls{NCS} components also exist in literature.
+These approaches tend to focus on a particular element of the system which is implemented using a prototype or testbed, and virtualize the rest of the system.
+For instance, in~\cite{wang2020inverter} the authors are concerned with realistic effects from the network, and thus they have an emulated \gls{NCS} interact with a real network.
+On the other hand, works such as~\cite{natale2004inverted} focus on the system being controlled and opt to instead emulate the network.
 
@@
 %Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
-
-\subsubsection{Related methodologies and tools}
-
-The work we present in this thesis lies at the intersection of research on edge computing, \gls{WCA} applications, and \acl{CPS} such as \glspl{NCS}, each of these topics representing a vast body of research.
-Of principal interest to us are works dealing with the experimental optimization of \gls{WCA} and \glspl{NCS} deployed on distributed systems such as edge computing.
-Below we will summarize the key milestones in literature relating to this.
-
-\paragraph{Experimental platforms for Edge Computing}
-
-Research in edge computing is still in its early stages, although it has been steadily increasing in the last decade as researchers, businesses, and organizations recognize the benefits of bringing computing and storage capabilities closer to the source of data.
-Standardization is however still a major unsolved challenge in this field, and thus most research hitherto has relied on ad-hoc methodologies and/or platforms.\todo{Citation}
-
-Nonetheless, the last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
-Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
-Of particular interest to us are the
-\begin{inlineenum}
-    \item \acs{GENI}
-    \item Chameleon
-    \item \acs{COSMOS}
-    \item \acs{POWDER}
-    \item \acs{ARA}
-    \item EdgeNet
-%    \item Drexel Grid \gls{SDR}
-\end{inlineenum} testbeds.
-
-\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
-Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
-
-The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
-Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
-
-\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
-The testbeds main \emph{raison d'etre} is \gls{MEC} research, with a strong emphasis on wireless networks.
-The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
-
-\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
-It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
-These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
-
-The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
-\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-
-EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
-It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
-The EdgeNet software is available for anyone to download;
-individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
-The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
-A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
-
-Finally, a number of other, small-scale approaches also exist.
-In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
-The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
-This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
-In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
-Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.

### 2023-02-10 body/part1/overview/methodology.tex
fixing formatting
fixing figs
writing

--- 2022-12-20 body/part1/overview/methodology.tex
@@
-\section{A methodology for the study of feedback-loop systems in Edge Computing}\label{summary:methodology}
+\section{A methodology for the study of latency-sensitive applications in Edge Computing}\label{summary:methodology}
+
+This thesis makes a contribution to the field of Edge Computing by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure.
+These applications, such as \gls{WCA} and \glspl{NCS}, are challenging to study and scale due to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
+
+The main focus of this thesis is to enhance the accuracy and realism of results related to Edge infrastructure, particularly in regard to the network.
+The proposed methodology aims to provide a more comprehensive and realistic assessment of the performance of these latency-sensitive applications, allowing for a better understanding of the strengths and limitations of Edge Computing in this context.
+Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing systems.
 
@@
     \centering
-    \includegraphics[width=.8\textwidth]{Figs/methodology.png}
+    \includegraphics[width=.8\textwidth]{Figs/methodology}
     \caption{%
-        Overview of the proposed methodology for the study of feedback-loop systems on edge computing infrastructure.
+        Overview of the proposed methodology for the study of latency-sensitive systems on edge computing infrastructure.
         The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
@@
 
-As discussed above, benchmarking infrastructures for edge-bound applications featuring feedback-loops is challenging, in particular when dealing with applications with high sensitivity to latency (such as \gls{NCS}) or with human involvement (such as \gls{WCA}).
-In the following, we introduce our methodological framework for tackling this challenge.
+Our methodology is based on the emulation of target workloads on actual Edge infrastructure;
+see \cref{fig:methodology}.
+We replace the client side of the system with a realistic emulation of the desired behaviors;
+this emulation is implemented in software deployed on \gls{COTS} general-purpose computing devices.
+In our initial implementation, these correspond to low-cost, easily replaceable and scalable Raspberry Pi 4 Model B~\cite{raspberrypi} \glspl{SBC}.
+We aim to maintain the backend software and hardware, as well as the network, unchanged with respect to the real-world deployment, in order to preserve as much realism as possible in the effects arising from the hardware and network.
 
-A general overview of our methodology is illustrated in \cref{fig:methodology}.
-It is based on the core idea of executing an emulation of the target workload on top of the real hardware on which it is to be deployed.
-We replace the client side of the system with a realistic emulation of the desired behaviors; this emulation is implemented in software deployed on \gls{COTS} general-purpose computing devices.
-The backend side and the network are not changed.
+As discussed in \cref{chap:relatedwork}, most research in this field falls into either simulated or fully-practical real-world approaches.
+Our emulated methodology presents several advantages over these alternatives.
+Firstly, it preserves the realism of effects stemming from the hardware and network.
+Employing emulations allows our methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
+Of particular interest to us are effects due to network factors such as contention, congestion control, and medium access.
+These effects are often of stochastic or chaotic natures, and are thus complex to capture in simulations.
 
-Two main arguments drive this approach.
-One, the difficulty of scaling the number of clients in these systems in a research context.
+Secondly, it allows for easier scaling of studies compared to real-world studies.
 The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale, either due to the involvement of humans (as in the case of \gls{WCA}) or due its cyber-physical nature (in \glspl{NCS}).
-Emulating this component reduces this complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, general-purpose hardware such as single-board computers (e.g.
-Raspberry Pis, Beaglebones, etc.).
+The client-side component of the applications of interest for this research is often complex to scale.
+In the case of interactive and immersive applications, scaling the number of human users can be a significant challenge, as recruiting participants is an expensive and time-consuming endeavor.
+On the other hand, scaling applications involving \glspl{CPS} such as \glspl{NCS} can be prohibitively complex and expensive depending on the specific hardware involved.
+In many cases, these systems are made-to-order, further increasing the cost as economies of scale cannot be employed.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as the aforementioned \glspl{SBC}.
 
-And two, it maintains the realism of the most complex parts of the system.
+Thirdly, in hand with the previous point, emulating the workload provides improved repeatability and replicability.
+Repeating a study becomes a matter of re-running the workload on the same testbed;
+in turn, studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
+These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
 
-\todo[inline]{Fix and finish}
+Finally, it improves extensibility significantly.
+Replicating the methodology for other workloads simply requires developing a new piece of software to emulate the new workload.
+This can further be simplified by employing softwareplatforms for the emulation of categories of workloads, as well will discuss later on in this thesis.
 
-In the following, we will discuss the initial implementation of this methodology to two different classes of applications.
-We first discuss the methodology in the context of a human-in-the-loop application, \gls{WCA}, in section \cref{ssec:methodology:wca}.
-Then, in \cref{ssec:methodology:ncs}, we present an exploration into it's utility for the study of \glspl{NCS}.
+In the following subsections we will discuss two illustrative use-cases for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
+\Cref{summary:methodology:usecase_wca} discusses the application of our methodology to \gls{WCA}.
+This work was originally published in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
+Next, in \cref{summary:methodology:usecase_ncs} we expand this discussion to \glspl{NCS}, a discussion originally approached in \cref{paper:olguinmunoz2022cleave}.
 
-\subsection{Applying the methodology to \gls{WCA}}\label{ssec:methodology:wca}
+\subsection{Use case: \acl{WCA}}\label{summary:methodology:usecase_wca}
 
@@
     \centering
-    \begin{subfigure}[b]{.47\textwidth}
+    \begin{subfigure}[b]{.45\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{Figs/trace_edgedroid.png}
+        \includegraphics[width=\textwidth]{Figs/trace_edgedroid}
         \caption{%
@@
             This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
-        }
+        }\label{fig:methodology:wca:conceptual}
     \end{subfigure}
     \hfill
-    \begin{subfigure}[b]{.47\textwidth}
+    \begin{subfigure}[b]{.45\textwidth}
         \centering
@@
         Overview of our first implementation of the methodology for \gls{WCA}.
-    }\label{fig:edgedroid1:trace}
+        Figures originally published in \cref{paper:olguinmunoz2019edgedroid}.
+    }
 \end{figure}
 
-We first introduce our methodology in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
-The former corresponds to an extended abstract paper which discusses a high level overview of our approach; the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
+We first introduce our methodology and its applicability to \gls{WCA} in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
+The former corresponds to an extended abstract paper which discusses a high level overview of our approach;
+the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
+As discussed in \cref{chap:introduction}\todo{Was it really here?}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
+Humans are unreliable, and, perhaps more importantly, hard to scale.
 
-The overall goal of these papers is to explore this methodological approach as applied to \gls{WCA}.
-As discussed in \cref{chap:introduction}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
-Humans are unreliable, and, perhaps more importantly, hard to scale.
-\todo[inline]{More reasons}
-
+\todo[inline]{Move part of this to the ``tools'' section?}
 The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
+\Cref{fig:methodology:wca:conceptual} presents a diagram of our conceptual approach to this implementation.
 We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \gls{WCA} backend.
-The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile devices on which real \gls{WCA} applications are intended to run in the future.
+The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile and/or wearable devices (e.g.\ Google Glass~\cite{googleglass}) on which real \gls{WCA} applications are intended to run in the future.
 Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
@@
 As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
+\todo[inline]{}
 
-\todo[]{}
+\medskip
+We proceed to demonstrate the practical utility of our methodology as applied to \gls{WCA} by setting up a series of scenarios.
+We deployed a \emph{Gabriel}-powered~\cite{chen2018application} \gls{WCA} application on a small testbed consisting of
+\begin{inlineenum}
+    \item a server computer acting as a cloudlet
+    \item a 802.11n Wi-Fi \gls{AP} connected to the cloudlet via ethernet
+    \item \num{10} Android smartphones connected wirelessly to the \gls{AP}, emulating users for the application
+\end{inlineenum}.
+Three scenarios were considered, corresponding to setups with \num{1} and \num{10} clients under ideal circumstances, plus an additional scenario with \num{10} clients located at a distance from the \gls{AP} such that the signal strength was significantly degraded.
+Each scenario was repeated a \num{100} times for statistical validity.
+
+\subsection{Use case: \acsp{NCS}}\label{summary:methodology:usecase_ncs}
+
+\todo[inline]{Discuss, at high level, the advantages ofthis methodology for NCSs.
+Refer back to related work. }
+
+\subsection{Tools for the application of the methodology}
+
+\todo[inline]{Here we actually present the tools}
+
+\subsubsection{\acs{WCA}: EdgeDroid}
+
+\subsubsection{\acsp{NCS}: CLEAVE}
+
+\subsubsection{Testbeds: Ainur}

### 2023-02-10 body/part1/overview/methodology.tex
reformulating;

@@
 Our emulated methodology presents several advantages over these alternatives.
-Firstly, it preserves the realism of effects stemming from the hardware and network.
+To begin with, and perhaps most importantly, it makes our approach portable and applicable to real edge computing deployments.
+By limiting the emulation to the workload, our approach can be directly employed on pre-existing edge computing infrastructure in order to benchmark applications before they are deployed.
+This makes our approach valuable to not only to the research community, but also to system designers and application developers.
+
+In hand with the previous point, it allows for easier scaling compared to real-world studies.
+The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
+The client-side component of the applications of interest for this research is often complex to scale.
+Consider a system owner interested in studying the performance of some edge computing setup for latency-sensitive applications.
+In the case of interactive and immersive applications, scaling the number of human users is a significant challenge, as recruiting participants is an expensive and time-consuming endeavor.
+Scaling applications involving \glspl{CPS} such as \glspl{NCS} can be prohibitively complex and expensive depending on the specific hardware involved.
+In many cases, these systems are made-to-order, further increasing the cost as economies of scale cannot be employed.
+On the other hand, a simulated approach requires modeling not only the application workload, but also the edge computing system and infrastructure.
+Due to the intricate nature of contemporary cloud and edge computing setups, which frequently involve multiple tiers of proxies, load balancers, virtualized computing units, and network functions, the modeling of such systems can quickly become an overly complex endeavor.
+
+Our approach circumvents both of these limitations.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as the aforementioned \glspl{SBC}.
+Furthermore, it preserves the realism of effects stemming from the hardware and network.
 Employing emulations allows our methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
@@
 
-Secondly, it allows for easier scaling of studies compared to real-world studies.
-The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
-The client-side component of the applications of interest for this research is often complex to scale.
-In the case of interactive and immersive applications, scaling the number of human users can be a significant challenge, as recruiting participants is an expensive and time-consuming endeavor.
-On the other hand, scaling applications involving \glspl{CPS} such as \glspl{NCS} can be prohibitively complex and expensive depending on the specific hardware involved.
-In many cases, these systems are made-to-order, further increasing the cost as economies of scale cannot be employed.
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as the aforementioned \glspl{SBC}.
-
-Thirdly, in hand with the previous point, emulating the workload provides improved repeatability and replicability.
+Emulating the workload also provides improved repeatability and replicability.
 Repeating a study becomes a matter of re-running the workload on the same testbed;
@@
 
-Finally, it improves extensibility significantly.
+Finally, our methodology improves extensibility significantly.
 Replicating the methodology for other workloads simply requires developing a new piece of software to emulate the new workload.
-This can further be simplified by employing softwareplatforms for the emulation of categories of workloads, as well will discuss later on in this thesis.
+This can further be simplified by employing software platforms for the emulation of categories of workloads, as well will discuss later on in this thesis.
 
+\medskip
 In the following subsections we will discuss two illustrative use-cases for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
@@
 \end{inlineenum}.
-Three scenarios were considered, corresponding to setups with \num{1} and \num{10} clients under ideal circumstances, plus an additional scenario with \num{10} clients located at a distance from the \gls{AP} such that the signal strength was significantly degraded.
+Three scenarios were considered, corresponding to setups under ideal network conditions with \num{1} and \num{10} clients, plus an additional scenario with \num{10} clients located at a distance from the \gls{AP} such that the signal strength was significantly degraded.
 Each scenario was repeated a \num{100} times for statistical validity.
+These setups already showcase the flexibility of our approach.
 

### 2023-02-11 body/part1/overview/methodology.tex
expeca

@@
 
-\subsubsection{Testbeds: Ainur}
+\subsubsection{ExPECA Testbed \& the Ainur software stack}
+
+After our initial implementation and experimentation with our methodology for \gls{WCA} applications, it became clear that our research would benefit greatly from the establishment of a testbed on which our methodology could be deployed.
+A testbed would
+\begin{inlineenum}
+    \item speed up subsequent research by removing the need for ad-hoc implementations for each new study
+    \item provide a standardized baseline, allowing for easier comparisons between studies.
+    \todo[inline]{More}
+\end{inlineenum}.
+Additionally, such a testbed would benefit other research happening in the group.
+
+These reasons led to the establishment of the \gls{EXPECA} testbed at \pgls{KTH}.
+\gls{EXPECA} is an \gls{SSF}-funded infrastructure project targeting the development and provisioning of an edge computing infrastructure for research into novel applications and network architectures.
+It consists of a cluster of hardware-reconfigurable \gls{COTS} computing nodes interconnected using managed switches and \glspl{SDR}.
+This allows us to quickly, on-the-fly, and in an automated fashion change the characteristics of the cluster and the network, in order to study different Edge- and Cloud-computing deployments and the applications that run on them.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/network}
+    \caption{
+        Prototype architecture of \acs{EXPECA}.
+        Figure originally published in \cref{paper:olguinmunoz2022ainur}.
+    }\label{fig:expeca}
+\end{figure}
+
+The initial prototype architecture of \gls{EXPECA} is illustrated in \cref{fig:expeca}.
+\emph{Workload hosts} act as client-side user devices.
+These devices connect to either a cloudlet or cloud computing instances over the \emph{workload data network}.
+Through intelligent configuration of \acsp{VLAN} and routing, traffic on this network can be forced to flow through the attached \acsp{SDR}, on of them acting as a wireless connection client and the other as an access point.
+This allows us to seamlessly switch out the physical layer for experiments.
+Alternatively, workload hosts can connect wirelessly via Wi-Fi to an \acs{SDR} acting as an \acs{AP}.
+Finally, the \emph{control network} is used for the out-of-band configuration of testbed resources, and the \acs{VPN} links allow direct tunneled connections to cloud compute nodes on \acs{AWS}.
+
+An implementation of this prototype was employed for the studies in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
+As an initial deployment of our testbed concept, it employed
+\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; and finally }}]
+    \item \num{10} Raspberry Pi 4 Model B boards, acting as workload hosts
+    \item two additional Raspberry Pi 4 Model boards acting as \gls{VPN} gateways as well as hosting necessary time synchronization and \gls{DNS} nameserver software
+    \item a cloudlet
+    \item a logging and management server
+    \item a configurable number of \gls{AWS} \gls{EC2} cloud instances acting as cloud compute nodes
+    \item two managed switches,
+    \item two \glspl{SDR} with associated radio stack processing hosts
+\end{inlineenum}.
+
+The above prototype was formally introduced in \cref{paper:olguinmunoz2022ainur}, together with the software framework employed for its automation and experiment orchestration.

### 2023-02-11 body/part1/overview/methodology.tex
more writing

@@
+This thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
+The primary contribution of this research is a methodology for analyzing and evaluating the performance of these applications.
+The methodology is validated through the examination of two initial case studies, \gls{WCA} and \glspl{NCS}.
+In addition to the methodology, this thesis includes several other key contributions.
+A prototype testbed and accompanying software framework have been developed to provide a platform for conducting experiments in Edge Computing.
+Furthermore, a further deeper exploration into the \gls{WCA} case studied yielded the first-ever model of human behavior in \gls{WCA}, which is a significant advancement in the field.
+The methodology, the case studies, and the prototype testbed provide valuable insights into the performance of latency-sensitive applications in Edge Computing and can be used as a basis for future research in this area.
+
 \section{A methodology for the study of latency-sensitive applications in Edge Computing}\label{summary:methodology}
@@
 \medskip
-In the following subsections we will discuss two illustrative use-cases for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
+In the following subsections we will discuss two illustrative case studies for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
 \Cref{summary:methodology:usecase_wca} discusses the application of our methodology to \gls{WCA}.
@@
 
-\subsection{Use case: \acl{WCA}}\label{summary:methodology:usecase_wca}
+\subsection{Case study: \acl{WCA}}\label{summary:methodology:usecase_wca}
 
@@
 
-\todo[inline]{Move part of this to the ``tools'' section?}
 The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
@@
 As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
-\todo[inline]{}
 
@@
 
-\subsection{Use case: \acsp{NCS}}\label{summary:methodology:usecase_ncs}
+\subsection{Case study: \acsp{NCS}}\label{summary:methodology:usecase_ncs}
 
-\todo[inline]{Discuss, at high level, the advantages ofthis methodology for NCSs.
+\todo[inline]{Discuss, at high level, the advantages of this methodology for NCSs.
 Refer back to related work. }
 
-\subsection{Tools for the application of the methodology}
-
-\todo[inline]{Here we actually present the tools}
-
-\subsubsection{\acs{WCA}: EdgeDroid}
-
-\subsubsection{\acsp{NCS}: CLEAVE}
-
-\subsubsection{ExPECA Testbed \& the Ainur software stack}
+\subsection{ExPECA Testbed \& the Ainur software stack}
 

### 2023-02-12 body/part1/overview/methodology.tex
blah

@@
 These reasons led to the establishment of the \gls{EXPECA} testbed at \pgls{KTH}.
-\gls{EXPECA} is an \gls{SSF}-funded infrastructure project targeting the development and provisioning of an edge computing infrastructure for research into novel applications and network architectures.
+\gls{EXPECA} is a \gls{SSF}-funded infrastructure project targeting the development and provisioning of an edge computing infrastructure for research into novel applications and network architectures.
 It consists of a cluster of hardware-reconfigurable \gls{COTS} computing nodes interconnected using managed switches and \glspl{SDR}.
@@
 The initial prototype architecture of \gls{EXPECA} is illustrated in \cref{fig:expeca}.
-\emph{Workload hosts} act as client-side user devices.
+\emph{Workload hosts} act as client-side user devices;
+in our prototype, these corresponded to \num{10} Raspberry Pi 4 Model B boards.
 These devices connect to either a cloudlet or cloud computing instances over the \emph{workload data network}.
-Through intelligent configuration of \acsp{VLAN} and routing, traffic on this network can be forced to flow through the attached \acsp{SDR}, on of them acting as a wireless connection client and the other as an access point.
+Through clever configuration of \acsp{VLAN} and routing, traffic on this network can be forced to flow through the attached \acsp{SDR} and their associated radio processing hosts.
 This allows us to seamlessly switch out the physical layer for experiments.
-Alternatively, workload hosts can connect wirelessly via Wi-Fi to an \acs{SDR} acting as an \acs{AP}.
+We implemented configurations wherein one of the \glspl{SDR} acted as a 4G or 5G base station and the other one as a cellular client.
+An alternative setup had the workload hosts connect wirelessly via Wi-Fi to one of the radios acting as an \gls{AP}
 Finally, the \emph{control network} is used for the out-of-band configuration of testbed resources, and the \acs{VPN} links allow direct tunneled connections to cloud compute nodes on \acs{AWS}.
 
-An implementation of this prototype was employed for the studies in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
-As an initial deployment of our testbed concept, it employed
-\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; and finally }}]
-    \item \num{10} Raspberry Pi 4 Model B boards, acting as workload hosts
-    \item two additional Raspberry Pi 4 Model boards acting as \gls{VPN} gateways as well as hosting necessary time synchronization and \gls{DNS} nameserver software
-    \item a cloudlet
-    \item a logging and management server
-    \item a configurable number of \gls{AWS} \gls{EC2} cloud instances acting as cloud compute nodes
-    \item two managed switches,
-    \item two \glspl{SDR} with associated radio stack processing hosts
-\end{inlineenum}.
+The above prototype was formally introduced in \cref{paper:olguinmunoz2022ainur}, together with the software framework we developed for its automation and experiment orchestration, \emph{Ainur}.
+Ainur is a highly flexible end-to-end wireless network testbed management framework.
+Its objective is to facilitate the study of the effects of communication and computation elements on the performance of distributed applications.
+To achieve this, Ainur provides a Python \gls{API} that allows researchers to describe experiments in a procedural manner, with plans to eventually offer toolkits for declarative configuration.
 
-The above prototype was formally introduced in \cref{paper:olguinmunoz2022ainur}, together with the software framework employed for its automation and experiment orchestration.
+The architecture of Ainur closely resembles the conceptual layers of the \acs{TCP}/\acs{IP} stack, and the framework utilizes containerization throughout.
+This includes the deployment and orchestration of workloads, support for different physical layer configurations, and automated collection of logs.
+To support this, the framework leverages Docker Swarm~\cite{docker,merkel2014docker,Swarm2021} as the container orchestration layer.
+Docker Swarm was chosen for its lightweight nature, simple configuration, and inclusion in the default Docker runtime installation.
+This allows Ainur to remain agnostic to the nature of the workloads and support a wide range of applications and system architectures, without having to consider library dependencies.
+
+Ainur also leverages containerization for the communication stack, which can nowadays be deployed on general-purpose processors using \gls{O-RAN} and \glspl{SDR}.
+This allows for the distribution of the communication stack across multiple hosts, providing researchers with the ability to study the effects of communication and computation elements on the performance of distributed applications.
+
+Both Ainur and \gls{EXPECA} were utilized in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
+\todo[inline]{finish}

### 2023-02-13 body/part1/overview/methodology.tex
cleave done?

@@
 
-\todo[inline]{Discuss, at high level, the advantages of this methodology for NCSs.
-Refer back to related work. }
+We discuss an initial case study of our methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
+In this work, we present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of edge-native \aclp{NCS}.
+The growing adoption of edge computing has made it important to understand the strengths and weaknesses of different edge concepts for the deployment of such \glspl{CPS}.
+
+\gls{CLEAVE} is a fully virtualized benchmarking framework, inspired by our previous work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} on benchmarking human-in-the-loop applications on the edge.
+In the case of \gls{CLEAVE} however, we forgo the trace-based approach of EdgeDroid in favor of direct modeling of the target systems.
+\gls{CLEAVE} includes a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
+These virtual \glspl{CPS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking.
+
+\gls{CLEAVE} is built using Python 3.8, making it highly extensible and able to use existing libraries.
+It is compatible with container technologies such as Docker~\cite{docker}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{publications/2022CLEAVE/images/CLEAVE_experiment_setup}
+    \caption{
+        The setup used for experimentation in the \gls{NCS} case study.
+        Figure originally published in \cref{paper:olguinmunoz2022cleave}.
+    }\label{fig:cleavesetup}
+\end{figure}
+
+We study the validity of our methodology and the implementation of the software framework through a simple scenario, illustrated in \cref{fig:cleavesetup}.
+An inverted pendulum plant controlled by a proportional-differential controller is deployed on an Edge server over a Wi-Fi link, while video-streaming applications run concurrently on the setup.
+The inverted pendulum is a popular benchmark in control-system literature, and the setup allows for straightforward reparametrization and a broad range of experiments.
+Video analytics, on the other hand, is a main proposed use case for edge computing, and it is likely that edge \gls{NCS} deployments will be deployed in parallel with such applications in the future.
+Using Wi-Fi allows for more experimental freedom, as the system can be adapted to study both best-case and worst-case scenarios with low and high latencies, respectively.
+
+We first establish a baseline by deploying a single control loop.
+This setup is repeated multiple times with different sampling rates to obtain statistically significant results.
+The results of the single-loop scenarios show, as expected, that higher sampling rates tend to correlate with better quality of control and that network delays can be compensated for by increasing the sampling rate.
+
+\begin{figure}[t]
+    \centering
+    \begin{subfigure}[t]{.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
+        \caption{Fraction of experimental runs with at least one toppled plant.}%\label{paper:olguinmunoz2022cleave:fig:video:toppled}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
+        \caption{\glspl{RTT}.}%\label{paper:olguinmunoz2022cleave:fig:video:rtt}
+    \end{subfigure}%
+    \caption{
+        Results from experimentation with multiple \gls{CLEAVE} control loops co-deployed with video streaming applications.
+        Error bars indicate \SI{95}{\percent} \glspl{CI} in all plots.
+        Figures originally published in \cref{paper:olguinmunoz2022cleave}.
+    }\label{fig:cleaveresults}
+\end{figure}
+
+Next, we study the interaction between the control systems and the video-streaming applications.
+Six control loops are deployed, as uplink \gls{UDP} traffic from four clients is generated to emulate the load generated by Full-HD video streaming.
+Based on the baseline results, we execute scenarios with varying plant sampling rates and repeat each scenario \num{30} times.
+Representative results from this experimentation are showcased in \cref{fig:cleaveresults}.
+While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, the results from these experiments show that network and resource contention are important factors when designing \glspl{NCS} for deployment on the edge.
+As sampling rates increase, a critical point is eventually reached at which the congestion on the network link is such that it severely impacts the quality of control.
+We conclude that adaptive sampling might be thus a viable method for optimizing resource utilization in these systems.
+
+The results of the experiments will be useful for optimizing resource utilization in edge-bound networked control systems.
+The results of the experiments with video-streaming applications will add to the understanding of the interaction between control systems and networked applications.
+Furthermore, the success of our experimentation shows that our methodology, implemented through \gls{CLEAVE}, makes the repeated experimentation and data collection convenient compared to fully physical setups, while at the same time providing enhanced realism with respect to simulated approaches.
 

### 2023-02-13 body/part1/futurework.tex
glossaries
restructuring for easier following of the red thread
future work

@@
-\chapter{Open questions \& future work}
+\section{Conclusions}
 
-Extending methodology to other classes of \ac{WCA}.
+\section{Open questions \& future work}
+
+The present thesis presents a comprehensive exploration into the application of our methodology for \glspl{NCS} and specially for \gls{WCA}.
+Nonetheless, a number of key questions remain open for future research in this field.
+
+The first of these relates to the applicability of the methodology to other systems and applications in an Edge Computing context.
+In this thesis, we have delved deep into its applicability for step-based \gls{WCA} and performed a shallower case study into its advantages for \gls{NCS} research.
+The question remains whether our methodology is suitable for
+\begin{inlineenum}
+%    \item other categories of \gls{WCA} which do not follow a pre-defined, linear task model (e.g.\ \citeauthor{chen2018application}s Ping-Pong assistant application)
+%    \item more complex \glspl{NCS}, for instance involving more complex controllers and system architectures
+    \item other, non-\gls{WCA} \gls{XR} applications, such as more general mobile \gls{AR} and \gls{VR}
+    \item other more general categories of \glspl{CPS}
+\end{inlineenum}.
+The results we have so far obtained would indicate that our methodology is in fact applicable to a broad spectrum of interactive and \gls{CPS} applications on Edge Computing, but these results should be validated by domain-specific studies.
+
+In line with the more general questions posed above, we identify a number of more specific questions related to the individual contributions to literature made by this thesis.
+Beginning with the tool, EdgeDroid, introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2023realistic}, we believe there are opportunities for its extension toward other classes of \gls{WCA} applications.
+Our current model only targets a particular class of step-based \gls{WCA} with a linear task model, but our tool --- and by extension, our methodology --- could be applied to similar applications without much difficulty.
+
+In that same vein, our data~\cite{olguinmunoz:impact2021} only considers young undergraduate students at a highly competitive university in the United States.
+An extension of this dataset and the model of human behavior developed in \cref{paper:olguinmunoz2023realistic} towards a more representative sample of the general population would surely be a valuable endeavor.
+Such research could also determine new individual difference factors that would substantially impact response to \gls{WCA} in novel populations, something which would represent a valuable contribution to literature in \gls{HCI} in general.
+
+The optimization framework employed in \cref{paper:olguinmunoz2023realistic} to study the implications of our methodology for the optimization of \gls{WCA} could be improved by expanding it towards distributions which better describe the behavior of human execution times.
+Currently, the framework assumes executions times are Rayleigh-distributed, however in literature these values are generally understood to follow an \glx{exGaussian} distribution.
+Moreover, the utility of the framework could be significantly increased by considering extensions towards scenarios with multiple clients and a distributed solution.
+
+\medskip
+Opportunities for future research exist also in regard to our work on applying the methodology to \gls{NCS}.
+Currently, \gls{CLEAVE} includes a small repertoire of emulations and controllers.
+We are planning to extend this with the goal of creating an open library of \glspl{NCS} to share with the community.
+
+\todo[inline]{Finish talking about CLEAVE.}
+
+\todo[inline]{Then talk about tremendous opportunities for EXPECA and Ainur. Talk about future of EXPECA.}
+
+At the moment, the interactions of \gls{CLEAVE} and tools such as Docker are still quite superficial.
+Our goal is to achieve a much tighter integration, e.g.\ by providing the toolkit as pre-packaged container images.
+Finally, the validity of the results obtained by the framework will have to be verified through more thorough, realistic scenarios than what we have been able to show in this work.
+In particular, we intend to perform large-scale experimentation targetting 5G cellular deployments, as this technology is set to become the backbone of edge networks in the near future.
+
+Extending methodology to other classes of \gls{WCA}.
 Ping-pong is the typical example --- look for other classes too.
@@
 Extending methodology to other systems.
-\ac{AR}, \ac{XR} in general?
+\gls{AR}, \gls{XR} in general?
 Closed-loop control? (Although we already started with CLEAVE).
 
-Further optimization of \ac{WCA} systems.
+Further optimization of \gls{WCA} systems.
 In this thesis we touch upon sampling, but other dimensions can also be optimized.

### 2023-02-13 body/part1/introduction.tex
restructuring for easier following of the red thread
writing
future work

@@
-\chapter{Introduction}\label{chap:introduction}
+\section{Introduction}
 
-\todo[inline]{
-    Introduction here. Why do we need this work?
-}
+The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
+One of the most promising of these is Edge Computing, which seeks to bring computing closer to the end user, reducing latency and increasing reliability.
+Another key player in this space is 5G, the latest generation of mobile communication networks that offers higher data rates, lower latency, and improved reliability.
+The combination of these two technologies holds great promise for the deployment of latency-sensitive applications, including \glspl{CPS}, \glspl{NCS}, and \gls{WCA} applications, among others.
 
-\todo[inline]{
-    Reorganize EdgeDroid, Cleave, Ainur, together.
+Despite these advancements, there are still significant challenges in understanding and scaling these systems, particularly in regard to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
+The aim of this thesis is to make a contribution to the field by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
+The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
+Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of \gls{MEC}.
 
-    Pull in multi-loop results for EdgeDroid?
+Our methodology is based on the emulation of target workloads on actual Edge infrastructure.
+We replace the client side of the system with a realistic emulation of the desired behaviors, implemented in software deployed on \gls{COTS} general-purpose computing devices.
+In our initial implementation, these correspond to low-cost, easily replaceable, and scalable Raspberry Pi 4 Model B \glspl{SBC}.
 
-    Think about experiments.
-}
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, COTS general-purpose hardware such as \glspl{SBC}.
+It also preserves the realism of effects stemming from the hardware and network.
+In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
 
-\section{Summary of the key contributions of this thesis}
+The methodology also provides improved repeatability and replicability.
+Repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
+These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
+
+Our methodology provides a comprehensive and realistic assessment of the performance of latency-sensitive applications deployed on Edge infrastructure.
+The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
+
+\section{Summary of the key contributions of this thesis}\label{sec:summary_contributions}
 
 This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
-Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \gls{WCA} and \gls{NCS}.
+Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound latency-sensitive applications such as \gls{WCA} and \gls{NCS}.
 This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
+We validate this methodology by example, presenting case studies of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
 

### 2023-02-13 body/part1/overview/methodology.tex
future work

@@
-This thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
+As summarized in \cref{sec:summary_contributions}, this thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
 The primary contribution of this research is a methodology for analyzing and evaluating the performance of these applications.
@@
 
-The main focus of this thesis is to enhance the accuracy and realism of results related to Edge infrastructure, particularly in regard to the network.
+The main focus of this work is to enhance the accuracy and realism of results related to Edge infrastructure, particularly in regard to the network.
 The proposed methodology aims to provide a more comprehensive and realistic assessment of the performance of these latency-sensitive applications, allowing for a better understanding of the strengths and limitations of Edge Computing in this context.
@@
 the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
-As discussed in \cref{chap:introduction}\todo{Was it really here?}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
+As discussed in \cref{chap:introduction,chap:relatedwork}\todo{Was it really here?}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
 Humans are unreliable, and, perhaps more importantly, hard to scale.
@@
 The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
+The trace-based approach is chosen for three reasons:
+\begin{inlineenum}
+    \item realism in terms of payloads sent over the network
+    \item simplicity without complex modeling of human behavior
+    \item extensibility through the recording of new traces for different tasks in the same category of WCA applications
+\end{inlineenum}.
+This approach provides a realistic execution path while being easily obtainable through instrumentation of existing WCA client applications and easily extensible to different tasks.
+
 \Cref{fig:methodology:wca:conceptual} presents a diagram of our conceptual approach to this implementation.
@@
 The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
-Once recorded, the trace is manually segmented into the logical component steps of the task, and these are replayed in sequence back to the backend by the EdgeDroid tool.
-
-We opt for a trace-based approach for our first implementation of the methodology for \gls{WCA} for three key reasons.
-The first of these is the level of realism if affords in terms of the payloads sent over the network and, in particular, processed by the backend.
-Using a trace ensures that the same computation is performed on the edge as if a human was involved, while also ensuring a reproducible application execution path.
-
-The second is simplicity, as such an approach does not require complex modeling of human behavioral patterns, merely the observation and recording of them.
-Compatible traces can be easily obtained simply by instrumenting existing \gls{WCA} client applications to record all captured inputs.
-
-The final advantage corresponds to extensibility.
-As long as the tasks belong to the same category of \gls{WCA} applications (in this case, step-based cognitive assistance), using a trace makes extending the tool to different tasks merely a matter of recording a new trace.
 
@@
 These setups already showcase the flexibility of our approach.
+
+\todo[inline]{Finish?}
 
@@
 
+\medskip
 We study the validity of our methodology and the implementation of the software framework through a simple scenario, illustrated in \cref{fig:cleavesetup}.

### 2023-02-13 body/part1/futurework.tex
futwork

@@
 The optimization framework employed in \cref{paper:olguinmunoz2023realistic} to study the implications of our methodology for the optimization of \gls{WCA} could be improved by expanding it towards distributions which better describe the behavior of human execution times.
-Currently, the framework assumes executions times are Rayleigh-distributed, however in literature these values are generally understood to follow an \glx{exGaussian} distribution.
+Currently, the framework assumes executions times are Rayleigh-distributed, however in literature these values are generally understood to follow an \gls{exGaussian} distribution.
 Moreover, the utility of the framework could be significantly increased by considering extensions towards scenarios with multiple clients and a distributed solution.

### 2023-02-14 body/part1/futurework.tex
futwork

--- 2022-11-18 body/part1/futurework.tex
@@
-\chapter{Open questions \& future work}
+\section{Conclusions}
 
-Extending methodology to other classes of \ac{WCA}.
-Ping-pong is the typical example --- look for other classes too.
+\section{Open questions \& future work}
 
-Extending methodology to other systems.
-\ac{AR}, \ac{XR} in general?
-Closed-loop control? (Although we already started with CLEAVE).
+The present thesis presents a comprehensive exploration into the application of our methodology for \glspl{NCS} and specially for \gls{WCA}.
+Nonetheless, a number of key questions remain open for future research in this field.
 
-Further optimization of \ac{WCA} systems.
-In this thesis we touch upon sampling, but other dimensions can also be optimized.
-Resource allocation, scheduling, user identification and classification, QoS/QoE guarantees?
+The first concern is the applicability of the methodology to other systems and applications in the context of Edge Computing.
+This thesis explores the methodology's use for step-based \gls{WCA} in depth, and also performs a less thorough case study on its benefits for \gls{NCS} research.
+The question remains if the methodology is suitable for
+\begin{inlineenum}
+    \item other, non-\gls{WCA} \gls{XR} applications, such as more general mobile \gls{AR} and \gls{VR}
+    \item other more general categories of \glspl{CPS}
+\end{inlineenum}.
+The results obtained thus far indicate that the methodology is applicable to a wide range of interactive and \gls{CPS} applications in Edge Computing, but these results should be confirmed by domain-specific studies.
+
+In accordance with the broader questions mentioned above, this thesis also raises specific questions related to its individual contributions to the literature.
+The EdgeDroid tool, introduced in papers \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2023realistic}, has the potential for further development to support other forms of \gls{WCA} applications.
+Currently, the tool is only designed for a specific class of step-based \gls{WCA} with a linear task model.
+However, the tool, and by extension the methodology, could be easily adapted to similar applications.
+
+We identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
+The current dataset is limited to young undergraduate students from a highly selective university in the \gls{USA}.
+Adding participants from different demographic groups would enhance the representativeness of the data and increase the generalizability of the findings.
+This would also provide an opportunity to extend the human timing model presented in \cref{paper:olguinmunoz2023realistic} to better reflect the behavior of a broader range of users.
+
+The optimization framework described in \cref{paper:olguinmunoz2023realistic} can be enhanced in two ways.
+Firstly, the framework assumes that human execution times follow a Rayleigh distribution, but in reality, they are often modeled with an \gls{exGaussian} distribution.
+Secondly, the framework can be made more practical by incorporating the optimization of scenarios with multiple clients and a distributed solution.
+
+There are numerous avenues for further investigation in the optimization and development of \gls{WCA}.
+While this thesis has explored the optimization of sampling and energy consumption, there are several other dimensions of these systems that can be examined.
+These include resource allocation, task scheduling, user identification and classification, as well as guaranteeing both \gls{QoE} and \gls{QoS}.
+Each of these areas presents opportunities for future work and could lead to significant advancements in the field of \gls{WCA}.
+
+\medskip
+Our research on applying the methodology to \glspl{NCS} presents promising potential for future work.
+While \cref{paper:olguinmunoz2022cleave} provides an initial exploration, further investigation through more realistic and thorough experimentation is necessary to validate the results obtained by the framework.
+To this end, we aim to conduct larger-scale experiments that utilize cutting-edge networking technologies, including 5G and future advancements such as 6G.
+By exploring the full range of networking technologies and scenarios, we can ensure that our methodology is robust and applicable to a wide range of \gls{NCS}.
+
+We also intend to extend the functionality of the \gls{CLEAVE} tool significantly, primarily to target a wider variety of \glspl{NCS} deployments.
+Currently, \gls{CLEAVE} includes a small repertoire of emulations and controllers.
+We are planning to extend this with the goal of creating an open library of \glspl{NCS} to share with the community.
+
+\medskip
+Finally, we see tremendous opportunities for testbed-driven research in Edge Computing.
+Our current efforts in this context aim towards the extension of our Ainur framework and its integration with \gls{CHI}~\cite{keahey2020lessons}.
+We believe Ainur's unique focus on fully-automated end-to-end wireless edge computing experimentation makes it valuable addition to the \gls{CHI} stack.
+At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainurs potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.
+
+The \gls{EXPECA} testbed is undergoing a major expansion and improvement.
+%The aim is to enhance its capabilities and hardware to meet the ever-increasing demands of the industry.
+The expansion consists of three main components;
+\begin{inlineenum}
+    \item a hardware upgrade
+    \item a small scale, private 5G deployment
+    \item a campus-wide 5G deployment
+\end{inlineenum}.
+Firstly, the \gls{EXPECA}s hardware is receiving a complete overhaul.
+We are replacing our initial Raspberry Pi-based prototype with a number of industry grade servers.
+Secondly, a small-scale private 5G deployment using industrial-grade equipment is currently being deployed.
+This will provide a controlled environment for testing and evaluating 5G technologies and applications.
+The use of industrial-grade equipment ensures that the results obtained from the tests are reliable and can be directly applied to real-world scenarios.
+Thirdly, the testbed will be extended to a campus-wide deployment at \gls{KTH} in Stockholm.
+This will provide an opportunity for a larger-scale deployment of 5G and \gls{MEC} technologies and applications, allowing for testing in real-world settings.
+The campus-wide deployment will also serve as a platform for researchers and industry professionals to collaborate and exchange ideas.
+The combined efforts of the small-scale private deployment and the campus-wide deployment at \gls{KTH} will significantly enhance the capabilities and knowledge of the \gls{EXPECA} testbed, positioning it as a leading platform for 5G and \gls{MEC} research and innovation.
+
+In conclusion, the present dissertation provides a comprehensive examination of the methodology for \glspl{NCS} and \gls{WCA}, but there are still many open questions that need to be addressed in future research.
+The EdgeDroid tool and the dataset have the potential to be extended, and the optimization framework employed in the thesis could be improved.
+We also plan to extend the functionality of the \gls{CLEAVE} tool and perform large-scale experimentation on novel networking technologies.
+There are also significant opportunities for testbed-driven research in Edge Computing through the \gls{EXPECA} testbed.

### 2023-02-14 body/part1/overview/extending-wca.tex
fixing formatting
fixing figs
futwork

--- 2023-01-24 body/part1/overview/extending-wca.tex
@@
 
-\todo[inline]{%
-    Rework this section.
-    First, discuss that we wish a more detailed study on the implications of the methodology for the optimization of WCA.
-    However, to achieve that we must first improve upon our model of human behavior.
-    This leads to us first characterizing human behavior, to then building an improved model.
-}
+In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
+To this end, we must first address a limitation of our current model of human behavior in the context of these applications.
+Our work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
+However, this is not an accurate representation of real human behavior, and a more realistic human model must take into consideration the perceived feedback latency and its effects on human reactions.
 
-\todo[inline]{%
-    Why do we need to extend the methodology? What was wrong with the approach in the previous section?
-}
+To build a more accurate model, we must first characterize human behavior in the context of \gls{WCA}.
+We achieve this through an experimental study measuring human performance in various scenarios, taking into account the impact of latency in the system to human behavior.
+The results of this study are then used to build a more representative model of human behavior, which we further use to provide a foundation for the optimization of these applications.
+
+Once we have established more accurate model of human behavior, we can then begin to explore the implications of our methodology for the optimization of \gls{WCA}.
+This includes evaluating the impact of various optimization strategies on human performance and identifying opportunities for improvement.
+By incorporating a realistic model of human behavior, we will be able to build a more comprehensive understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be applied to address these challenges.
 
@@
     \centering
-    \begin{subfigure}[t]{.49\textwidth}
+    \begin{subfigure}[t]{.45\textwidth}
         \centering
@@
     \hfill%
-    \begin{subfigure}[t]{.49\textwidth}
+    \begin{subfigure}[t]{.45\textwidth}
         \centering
@@
 The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
-\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
+\begin{inlineenum}
     \item agreeableness
@@
     \item openness
-\end{enumerate*}.
+\end{inlineenum}.
 The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
@@
 This array included devices to capture four physiological measures:
-\begin{enumerate*}
+\begin{inlineenum}
     [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
@@
     \item heart rate
-\end{enumerate*}.
+\end{inlineenum}.
 Unfortunately, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.

### 2023-02-14 body/part1/overview/extending-wca.tex
working

@@
 fewer samples per step translates into longer sampling periods, and lower energy consumption tends to be made possible by processing fewer bits of data per step.
-As such, we employ in these sections an optimization framework for the optimization of said trade-offs based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
-We note that objective metrics relating to sampling and the responsiveness of a \gls{WCA} system will present itself as a linear combination of the number of expected samples and expected wait time per step plus some constant term:
-\begin{align}
-    \mathcal{E} = \alpha\mathbb{E}[\mathcal{S}] + \beta\mathbb{E}[\mathcal{W}] + C\label{eq:genericopt}
-\end{align}
-The above objective function is generic, and can be adapted to any metric which relates to the sampling and responsiveness of \gls{WCA} by modifying the factors \ensuremath{\alpha}, \ensuremath{\beta}, and \ensuremath{C}.
 
-For the analyses in this subsection and \cref{ssec:implications:energy}, we use the solution
-\begin{align}
-    t_n = \left( 3\sigma\sqrt {\frac{\alpha}{2\beta}} \right)^{\frac{2}{3}} n^{\frac{2}{3}}
-\end{align}
-where \ensuremath{t_n} once again corresponds to the \ensuremath{n}-th sampling instant in a step.
-Note that any positive real value is valid for \ensuremath{\frac{\alpha}{2\beta}}.
-This ratio controls the optimization criteria --- as it goes to infinity, the criteria changes from minimizing wait time (i.e.\ maximizing responsiveness) to minimizing the number of samples per step (i.e.\ minimizing energy consumption).
-
-The full derivation and explanation of this framework is out of the scope of this summary.
-Please refer to \cref{paper:olguinmunoz2023realistic} and its appendix for a detailed discussion on this optimization approach.
-\todo{Include ref to section?}
+We employ in these sections an optimization framework based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
+The goal of this framework is to balance between increased responsiveness due to higher sampling rates and reduced resource consumption and network congestion caused by excessive sampling.
+In general, any objective metric in these applications which relates to sampling and the responsiveness of the system will present itself
+as a linear combination between \ensuremath{\mathbb{E}[\mathcal{S}]} and \ensuremath{\mathbb{E}[\mathcal{W}]} plus terms independent of the number of samples or wait time:
+\begin{alignat}{1}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C\;
+\end{alignat}
+The constants \ensuremath{\alpha}, \ensuremath{\beta}, and \ensuremath{C} modify the objective function from one metric to another.
 
 \medskip
+For the problem of minimizing the number of samples, an unconstrained optimization of the above equation leads to a single sample at \ensuremath{t \rightarrow \infty} and an infinite wait time.
+Thus, a constrained optimization is performed with an upper bound \ensuremath{w_0} for the expected wait time, i.e.\ \ensuremath{\mathbb{E}[\mathcal{W}] \leq w_0}.
+The below equation is derived in \cref{paper:olguinmunoz2023realistic} for finding appropriate \ensuremath{\alpha} and \ensuremath{\beta} for this problem:
+\begin{alignat}{1}\label{eq:sampling}
+\frac{\alpha}{\beta}=\frac{2\sqrt{2}\,w_0^2}{(\mathlarger{\Gamma}(\tfrac{3}{4}))^2\,\sigma}\approx1.9\frac{w_0^2}{\sigma}
+\end{alignat}
+This solution assumes execution times distributed according to a Rayleigh distribution with parameter \ensuremath{\sigma}.
+\ensuremath{\mathlarger{\Gamma}(x)} is the Gamma function.
+Note that the full derivation and explanation of this solution is out of the scope of this summary;
+please refer to the paper and its appendix for a detailed discussion on this optimization approach.
+
+A reference scheme from~\cite{wang2019towards} is introduced which adjusts the sampling rate at each sampling instant \ensuremath{t} according to the estimated likelihood of the user finishing the step.
+The scheme uses a static Gaussian distribution for all steps.
+
+We compare the effects of the optimization approach with the timing models to the reference scheme.
+We implement a sample-count-optimized aperiodic sampling scheme using \cref{eq:sampling} to determine the optimum sampling instants and the values for \ensuremath{\alpha} and \ensuremath{\beta}.
+The scheme includes an embedded human timing model that measures the perceived system responsiveness and updates the \ensuremath{\sigma} parameter at every step.
+The reference scheme is implemented using the Gaussian distribution fitted to the execution times in \cref{paper:olguinmunoz2021impact}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/sampling_optimization}
+    \caption{%
+        Round-trip time versus mean number of captured samples per step, averaged over all \num{100} repetitions of the experiment.
+        Note the logarithmic scale on the vertical axis.
+        Error bars indicate \SI{95}{\percent} \glspl{CI}.
+        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{wang2019towards}'s \gls{CDF}-based approach.
+        Figure originally published in \cref{paper:olguinmunoz2023realistic}.
+    }
+\end{figure}
+
+Simulations are performed using these sampling schemes implemented in Python.
+The sample-count-optimized aperiodic scheme results in a lower mean number of samples per step compared to the reference scheme, while still satisfying the expected wait time bound.
+In addition, the results show that the embedded timing model significantly improves the performance of the sample-count-optimized aperiodic scheme compared to using fixed values for \ensuremath{\alpha} and \ensuremath{\beta}.
+These results demonstrate the efficacy of the optimization approach, combined with our novel model of human behavior, in minimizing sampling rates in \gls{WCA} applications while still meeting responsiveness requirements.
 

### 2023-02-14 body/part1/overview/extending-wca.tex
implications

@@
 as a linear combination between \ensuremath{\mathbb{E}[\mathcal{S}]} and \ensuremath{\mathbb{E}[\mathcal{W}]} plus terms independent of the number of samples or wait time:
-\begin{alignat}{1}
+\begin{alignat}{1}\label{eq:tradeoff}
     \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C\;
@@
     \caption{%
-        Round-trip time versus mean number of captured samples per step, averaged over all \num{100} repetitions of the experiment.
-        Note the logarithmic scale on the vertical axis.
+        Round-trip time versus mean number of captured samples per step (logarithmic Y-axis).
         Error bars indicate \SI{95}{\percent} \glspl{CI}.
-        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{wang2019towards}'s \gls{CDF}-based approach.
         Figure originally published in \cref{paper:olguinmunoz2023realistic}.
-    }
+    }\label{fig:samplingresults}
 \end{figure}
 
-Simulations are performed using these sampling schemes implemented in Python.
-The sample-count-optimized aperiodic scheme results in a lower mean number of samples per step compared to the reference scheme, while still satisfying the expected wait time bound.
-In addition, the results show that the embedded timing model significantly improves the performance of the sample-count-optimized aperiodic scheme compared to using fixed values for \ensuremath{\alpha} and \ensuremath{\beta}.
-These results demonstrate the efficacy of the optimization approach, combined with our novel model of human behavior, in minimizing sampling rates in \gls{WCA} applications while still meeting responsiveness requirements.
+Simulations were performed in Python to test different sampling schemes.
+The results from these simulations are presented in \Cref{fig:samplingresults}.
+The figure compares the performance of four sampling schemes:
+\begin{inlineenum}
+    \item the sample-count-optimized approach described above
+    \item \citeauthor{wang2019towards}s adaptive approach based on a Gaussian distribution
+    \item a \emph{greedy}, completely unoptimized approach
+    \item an \emph{offline optimum} scheme which perfectly predicts execution times in order to only sample once per step
+\end{inlineenum}.
+
+The results indicate that the sample-count-optimized aperiodic scheme results in a lower mean number of samples per step compared to the reference scheme proposed by \citeauthor{wang2019towards}.
+This was achieved while still satisfying the expected wait time bound.
+
+The performance of the sample-count-optimized aperiodic scheme was further enhanced by incorporating the timing model that reflects human behavior.
+The timing model allows for a more precise optimization of the sampling rate in \gls{WCA} applications, which minimizes the number of samples needed while still meeting the responsiveness requirements.
+These results highlight the effectiveness of the optimization approach and the significance of the timing model in optimizing sampling rates, as well as provide further validation of our methodology.
 
 \subsection{Implications with respect to raw energy consumption}\label{ssec:implications:energy}
+
+We also optimize for energy consumption in \gls{WCA}.
+Energy consumption in these applications depends on various factors such as the number of samples captured, communication power and delays, among others.
+Hence, minimizing the number of samples captured is not a guarantee of reducing energy consumption.
+We therefore take the general solution for \cref{eq:tradeoff} and find the appropriate values of \ensuremath{\alpha} and \ensuremath{\beta} to minimize the energy consumed per step, represented by \ensuremath{\mathcal{E}=E}.
+With feedback given even to the discarded samples in our model and communication delay defined as the total delay in either direction, we have,
+\begin{alignat}{2}
+    \mathrm{E}=&\;\mathcal{S}\tau_cP_c+(\mathcal{T}+\mathcal{W}+\tau_\mathrm{p}+\tau_\mathrm{c}-\mathcal{S}\tau_c)P_0\nonumber\\
+    % &=s\tau_c(P_c-P_0^{(t)})+wP_0^{(t)}+(\tau+\tau_c+\tau_p) P_0^{(t)}+\tau_cP_c\\
+    =&\;\tau_{\text{c}}(P_{\text{c}} -P_0)\mathcal{S}+\mathcal{W}P_0+(\mathcal{T}+\tau_{\text{p}} +\tau_{\text{c}}) P_0\nonumber\\
+    &\Rightarrow \alpha=\tau_{\text{c}}(P_{\text{c}} -P_0),\text{ and }\beta=P_0
+\end{alignat}
+\( \tau_\text{p} \) and \( \tau_\text{c} \) correspond to the processing and two-way communication delay for each sample.
+\( P_\text{c} \) and \( P_0 \) correspond to the communication and idle power, respectively, of the \gls{WCA} client device.
+
+We perform an experiment to implement the energy-optimized aperiodic sampling scheme, replacing the sample-count-optimized sampling scheme.
+The energy-optimized scheme uses the calculated values of \ensuremath{\alpha} and \ensuremath{\beta} to minimize energy consumption.
+The values of communication power, \ensuremath{P_{\text{c}}}, and idle power, \ensuremath{P_0}, are estimated from previous research.
+The processing delay, \ensuremath{\tau_{\text{p}}}, is set to a constant value across all experiments, and the communication delay, \ensuremath{\tau_{\text{c}}}, is calculated as the \gls{RTT} minus the processing delay.
+As before, the optimized sampling scheme includes an embedded timing model to provide updated predictions for \ensuremath{\sigma} at every step.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/energy_optimization_diff}
+    \caption{%
+        Percentage difference in mean per step energy consumption by the energy-optimized sampling scheme with respect to the three reference schemes.
+        Error bars indicate \SI{95}{\percent} \glspl{CI}, calculated using a two-sided t-test.
+        Figure originally published in \cref{paper:olguinmunoz2023realistic}.
+    }\label{fig:energyresults}
+\end{figure}
+
+The results of our experiments, illustrated partially in \cref{fig:energyresults}, show that this energy-optimized scheme outperforms the reference schemes, with a significant reductions in energy consumption.
+Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{wang2019towards}'s approach, and is able to achieve up energy savings of up to \SI{40}{\percent} with respect to the greedy scheme.
+Furthermore, our approach is more consistent and reliable, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.
+
+Our findings have important implications for the development and deployment of \gls{WCA} systems.
+We have demonstrated that it is possible to significantly reduce the energy consumption of these systems while still maintaining a reasonable \gls{RTT}.
+This can help extend the battery life of wearable devices, making  systems more practical and usable for a wider range of users.
+The results of this study can serve as a valuable reference for future work in energy optimization for \acl{WCA} systems.

### 2023-02-16 body/part1/futurework.tex
conclusion

@@
 
-\section{Open questions \& future work}
+In this thesis, we have made three significant contributions to the field of edge computing, addressing the challenge of studying latency-sensitive applications in a scalable, repeatable manner.
 
-The present thesis presents a comprehensive exploration into the application of our methodology for \glspl{NCS} and specially for \gls{WCA}.
-Nonetheless, a number of key questions remain open for future research in this field.
+The overarching contribution of this dissertation is a new methodological approach for studying the trade-offs between system responsiveness and resource consumption in latency-sensitive applications at the edge.
+Our approach is based on the emulation of the client-side workload while maintaining the actual server-side process and network stack.
+This allows us to maintain a high degree of realism in studies, while at the same time affording significant potential for scalability and repeatability.
+We validate this approach and present case studies of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of edge computing testbeds necessary for its implementation.
+
+A follow-up contribution concerns a deeper exploration of this methodology for \gls{WCA} and introducing the first-ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
+We gather data from a comprehensive human-subject study and use it to build a model that provides a more accurate representation of human behavior in \gls{WCA}.
+This allows us to more accurately evaluate the advantages of our methodology when applied to these systems.
+
+Finally, as an extension and application of our first two contributions, we investigate the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
+In particular, we compare the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
+Through our methodology, we obtain reliable results indicating that the use of our human user model leads to significant improvements in system performance and energy consumption.
+
+Our work demonstrates the flexibility and advantages of an emulated approach for the study of latency-sensitve applications on edge computing.
+Further, it highlights the interplay between system responsiveness and resource consumption in the design and optimization of these applications.
+We believe that the methodology and human user model presented in this thesis provide a valuable tool for researchers and practitioners in this field, and we hope that our contributions will inspire further research in this area.
+
+\subsection{Open questions \& future work}
+
+Notwithstanding the important contributions of this dissertation, a number of key questions remain open for future research in this field.
 

### 2023-02-16 body/part1/introduction.tex
conclusion

@@
 
-\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \glspl{NCS}.
-As discussed above, these applications are similarly difficult as \gls{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
-\todo[inline]{More argumentation or refer to previous section.}
-We apply thus in this work our methodology to \gls{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
-The software acts as a middleware which abstracts away the network from the development of these workloads, allowing for quick prototyping and deployment.
-This work also showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.
+In \cref{paper:olguinmunoz2022cleave}, we extend our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
+These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
+To address these challenges, we introduce a tool, \acs{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
+The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
+We then present use-case scenarios demonstrating the scalability and flexibility of their approach by deploying scenarios with a large number of loops without the need for domain-specific hardware.
+This work showcases the applicability of the methodology beyond wearable cognitive assistance and highlights the potential of our approach for other edge computing applications.
 
@@
 
-\section{Structure of this thesis}
+\section{Structure of this dissertation}
 
-\todo[inline]{This thesis is structured as follows.}
+This dissertation is structured into two parts.
+\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
+Next, in \cref{part:publications} we present the publications that form the core of this thesis.
+
+\cref{part:summary} is further structured into four main chapters.
+\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
+It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance.
+
+In \cref{chap:relatedwork}, we present theoretical background and related work.
+The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
+We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
+
+\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
+We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
+We aim to demonstrate the significance of our work and how it contributes to the field.
+
+Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
+We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
+This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-02-16 body/part1/overview/extending-wca.tex
model

@@
 
-\todo[inline]{Probably needs to repeat a discussion on why the new model is necessary, in a shorter summarized form.
-This discussion should first be presented at the beginning of this section however.}
+Our current model of human behavior has a limitation that we need to address.
+The model we have discussed so far emulates a rigid and unyielding human who is not affected by low responsiveness in the system.
+This is not an accurate representation of real human behavior, and in the following we present a more realistic human model that takes into consideration the perceived feedback latency and its effects on human reactions.
 
+Our new model is based on the comprehensive study of human behavior in the presence of feedback latency outlined in \cref{paper:olguinmunoz2021impact}, and has significant implications for the optimization of edge computing systems.
+By incorporating a more accurate representation of human behavior into our methodology, we can better assess the impact of system design choices on the end-user experience.
+This, in turn, enables future research into optimization strategies that are more effective at improving the overall quality of service for edge computing applications.
+
+\medskip
 As discussed in \cref{impact:results}, the execution time for a step depends not only on the most recent perceived \gls{TTF}, but also on how long the system has been in the current level of responsiveness.
@@
 
-Our timing model produces realistic execution times for each step, according to the above effects.
+Our new timing model produces realistic execution times for each step, according to the above effects.
 We achieve this through the following design:

### 2023-02-16 body/part1/overview/methodology.tex
wca model

@@
 
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{.45\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/trace_edgedroid}
-        \caption{%
-            High level conceptual design of our methodology for \gls{WCA}.
-            We replace the human by a trace-driven emulation.
-            This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
-        }\label{fig:methodology:wca:conceptual}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{.45\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2018DemoScalingOnTheEdge/img/TraceReplay_GenArch}
-        \caption{%
-            Architectural overview of the EdgeDroid \num{1.0} tool.
-            The ``client emulator app'' implements the aforementioned user model together with the required networking functionality to connect to a \gls{WCA} backend running on a cloudlet.%
-        }
-    \end{subfigure}\\
-    \medskip
-    \begin{subfigure}[t]{\textwidth}
-        \centering
-        \adjustbox{scale=0.7}{
-            \begin{tikzpicture}[align=center,
-                node distance=.5cm and 1.5cm,
-                every initial by arrow/.style={-{Latex[length=2mm]}}]
-                % Place nodes
-                \node [initial, state, minimum size=6em, initial text=] (play) {Play};
-                \node [state, above right=of play, minimum size=6em] (change) {Change\\step};
-                \node [state, below right=of play, minimum size=6em] (rewind) {Rewind};
-                \node [state, accepting, above right=of rewind, minimum size=6em] (shutdown) {Shutdown};
-
-                % Draw edges
-                \path[draw, -{Latex[length=2mm]}]
-                (play) edge [bend right=20] node[left] {Step done} (rewind)
-                edge [bend left=20] node[left] {Got feedback\\(\emph{positive})} (change)
-                edge [out=140,in=220,looseness=6] node[left] {Step\\not done} (play)
-
-                (change) edge [bend left=20] node[right] {Step changed} (play)
-                edge [bend left=20] node[right] {All steps done} (shutdown)
-
-                (rewind) edge [bend right=20] node[right] {Rewound} (play)
-                edge [bend right=20] node[right] {Too many rewinds} (shutdown);
-
-            \end{tikzpicture}
-        }
-        \caption{State diagram of the user model governing the replay of the pre-recorded trace at runtime.
-This model approximates the behavior of an ``ideal'' human, one that is patient and makes no mistakes.}\label{fig:usermodel}
-    \end{subfigure}
-    \caption{%
-        Overview of our first implementation of the methodology for \gls{WCA}.
-        Figures originally published in \cref{paper:olguinmunoz2019edgedroid}.
-    }
-\end{figure}
-
 We first introduce our methodology and its applicability to \gls{WCA} in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
@@
 
-\Cref{fig:methodology:wca:conceptual} presents a diagram of our conceptual approach to this implementation.
-We introduce a tool, \emph{EdgeDroid \num{1.0}}, which replays a pre-recorded trace of sensory inputs to the original \gls{WCA} backend.
-The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile and/or wearable devices (e.g.\ Google Glass~\cite{googleglass}) on which real \gls{WCA} applications are intended to run in the future.
-Furthermore, this tool is instrumented, allowing us to collected a multitude of system-level metrics at runtime which can then be analyzed.
-On the other hand, the trace in question corresponds to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer.
-The trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
+While a trace-based emulation approach has advantages in providing realistic inputs without complex modeling of human behavior, it can have a significant drawback:
+if system responsiveness at trace replay time differs from system responsiveness at trace recording time, the trace can very easily end up out-of-sync.
+To address this, the concept of a user model is introduced, which adapts the replay of the trace to the responsiveness of the system by making slight modifications at runtime.
+The user model attempts to navigate the task model and reach the desired system state in light of changing system conditions, while still using the original trace as the source of input data.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{Figs/trace_edgedroid}
+    \caption{%
+        High level conceptual design of our methodology for \gls{WCA}.
+        We replace the human by a trace-driven emulation.
+        This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
+%        Figure originally published in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
+    }\label{fig:methodology:wca:conceptual}
+\end{figure}
+
+\Cref{fig:methodology:wca:conceptual} presents a conceptual diagram of this approach.
+The trace is fed to a model of human behavior, which then replays it to the real \gls{WCA} backend.
+Feedback from the backend is then interpreted by the model to adapt the runtime replay of the trace.
+
+We introduce a tool, \emph{EdgeDroid \num{1.0}}, which implements the concepts outlined above.
+The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile and/or wearable devices (e.g.\ Google Glass~\cite{googleglass}) on which real \gls{WCA} applications are intended to run in the future, and is instrumented for easy collection of a multitude of system-level metrics at runtime which can then be analyzed.
+We use a trace corresponding to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer;
+the trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
+Finally, we employ in this work a preliminary model of human behavior which is totally impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+This leads to a model of an ``ideal'' user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
 

### 2023-02-16 body/part1/overview/methodology.tex
writing

@@
 Each scenario was repeated a \num{100} times for statistical validity.
+
 These setups already showcase the flexibility of our approach.
+Our trace-based implementation of the methodology allows for detailed measurements of system performance at different levels, from individual components to specific task steps.
+This provides valuable insights into the bottlenecks and areas for optimization in the system.
+We find, for example, that \glspl{RTT} for feedback-rich inputs are up to an order of magnitude greater than feedback-less inputs\footnote{%
+    In the context of \gls{WCA}, \emph{feedback-rich} inputs refer to inputs which trigger a transition in step in the internal task model of the application and thus cause the generation of feedback which is then provided to the user.
+    Conversely, \emph{feedback-less} inputs are inputs which are silently discarded at the backend and do not generated noticeable feedback.
+}.\todo{Figure}
 
-\todo[inline]{Finish?}
+For example, the results suggest that improving the quality of the wireless link should be a priority for scaling the application, and that the transitions from task state 3 to 4 and from task state 4 to 5 are ripe for optimization.
+The flexibility of the methodology also allows researchers to zoom into specific components or steps of the system to gain deeper insights, which can inform the design and development of future systems.
+
+Our use case scenarios measure the round-trip times (RTTs) for inputs that caused a step transition versus inputs that did not, as well as the distribution of latencies across system components and individual task steps.
+The results show that RTTs for feedback-rich inputs are up to an order of magnitude greater than feedback-less inputs, and that the main component contributing to latency is backend processing.
+The study also reveals that the system scales linearly with respect to the number of clients, but the impaired Wi-Fi greatly affects the performance of feedback-rich inputs, with delays sometimes exceeding the threshold for acceptable user experience.
+
+Overall, the study demonstrates the usefulness of the trace-based methodology in evaluating system performance, and highlights the importance of considering feedback-rich inputs when testing the responsiveness of task-guidance applications.
+The results also underscore the need for optimization of system components and the importance of network quality in delivering a satisfactory user experience.
 

### 2023-02-16 body/part1/overview/methodology.tex
wca case

--- 2023-02-16 body/part1/overview/methodology.tex
@@
 Each scenario was repeated a \num{100} times for statistical validity.
+
+\begin{figure}
+    \begin{subfigure}[t]{.45\textwidth}
+        \centering%
+        \includegraphics[width=\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
+        \caption{
+            Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.
+        }\label{fig:wca:rtt}%
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{.45\textwidth}
+        \centering%
+        \includegraphics[width=\textwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
+        \caption{%
+            Distribution of latency across system components for inputs that triggered a state transition in the task model.
+        }\label{fig:wca:latency}
+    \end{subfigure}%
+    \caption{
+        Example results from the \acs{WCA} case study.
+        Figures originally published in \cref{paper:olguinmunoz2019edgedroid}.
+    }
+\end{figure}
+
 These setups already showcase the flexibility of our approach.
+Our trace-based implementation of the methodology allows for detailed measurements of system performance at different levels, from individual components to specific task steps.
+This provides valuable insights into the bottlenecks and areas for optimization in the system.
+We find, for example, that \glspl{RTT} for feedback-rich inputs are up to an order of magnitude greater than feedback-less inputs\footnote{%
+    In this context, \emph{feedback-rich} inputs refer to inputs which trigger a transition in step in the internal task model of the application and thus cause the generation of feedback which is then provided to the user.
+    Conversely, \emph{feedback-less} inputs are inputs which are silently discarded at the backend and do not generated noticeable feedback.
+}; see \cref{fig:wca:rtt}.
+The flexibility of the methodology also allows researchers to zoom into specific components or steps of the system to gain deeper insights, which can inform the design and development of future systems.
+For example, the results in \cref{fig:wca:latency} suggest that improving the quality of the wireless link should be a priority for scaling the application.
+The system scales linearly with respect to the number of clients, but the impaired Wi-Fi greatly affects the performance of feedback-rich inputs, with delays sometimes exceeding the threshold for acceptable user experience.
 
-\todo[inline]{Finish?}
+Overall, the study demonstrates the usefulness of the trace-based methodology in evaluating system performance.
+The results also underscore the need for optimization of system components and the importance of network quality in delivering a satisfactory user experience.
 

### 2023-02-16 body/part1/overview/methodology.tex
ainur

@@
 
-After our initial implementation and experimentation with our methodology for \gls{WCA} applications, it became clear that our research would benefit greatly from the establishment of a testbed on which our methodology could be deployed.
-A testbed would
-\begin{inlineenum}
-    \item speed up subsequent research by removing the need for ad-hoc implementations for each new study
-    \item provide a standardized baseline, allowing for easier comparisons between studies.
-    \todo[inline]{More}
-\end{inlineenum}.
-Additionally, such a testbed would benefit other research happening in the group.
+After our initial implementation and experimentation with our methodology for \gls{WCA} applications, we realized that creating a testbed for experimental Edge Computing research for would significantly enhance our work.
+Establishing such testbed would provide a dedicated and controlled environment that can accelerate future research by eliminating the need for ad-hoc implementations for each new study.
+Moreover, having a standardized baseline would make it easier to compare results across different studies.
+THe testbed could serve as a platform for researchers to test new ideas and compare them to the results of previous experiments.
+This would enable more efficient and accurate assessments of the methodology's performance and facilitate the development of new methods and techniques.
+Additionally, by establishing a testbed, we could provide a platform for other researchers within our group to conduct their research and experiments.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2022Ainur/figures/network}
+    \includegraphics[width=.9\textwidth]{publications/2022Ainur/figures/network}
     \caption{
@@
 
-The above prototype was formally introduced in \cref{paper:olguinmunoz2022ainur}, together with the software framework we developed for its automation and experiment orchestration, \emph{Ainur}.
+We also introduce a software framework we developed for its automation and experiment orchestration, \emph{Ainur}.
 Ainur is a highly flexible end-to-end wireless network testbed management framework.
@@
 
-Both Ainur and \gls{EXPECA} were utilized in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
-\todo[inline]{finish}
+Finally, we note that both Ainur and \gls{EXPECA} were utilized for experimentation in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
+The use of these tools was, however, during their development and prototyping, and they have thus not been referred to by name in these works.
+Their first formal appearance was in \cref{paper:olguinmunoz2022ainur}.

### 2023-02-16 body/part1/overview/extending-wca.tex
almost first draft!

@@
 
-\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \gls{WCA}.
+\Cref{paper:olguinmunoz2021impact} aims to investigate the effects of system responsiveness on human behavior in step-based \gls{WCA}.
 This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
+The study involved \num{40} undergraduate students who interacted with a \gls{WCA} application while the system responsiveness was systematically altered in real-time, and their behavioral and physiological reactions were recorded.
+The study attempted to answer a number of core research questions relating to human behavior in the context of these applications.
+These included
+\begin{inlineenum}
+    \item whether users change their actions in relation to system latency
+    \item whether they show signals of arousal in physiological responses
+    \item whether their responses to delay effects are mediated by cognition and/or emotion
+    \item whether these effects are mediated by personality indicators
+\end{inlineenum}.
 
-We present in this paper the design and execution of a study in which \num{40} undergraduate students of diverse fields of
-study\footnote{%
-    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \gls{CMU}.
-}, were asked to interact with and follow the instructions given to them by a \gls{WCA} application.
-Unbeknownst to the participants, the system responsiveness of the \gls{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.
-
-With this study, we intended to answer four core research questions relating to human responses to decreased application responsiveness.
-One, do subjects change the temporal profile of their actions in relation to system latency?
-We expected this to be the case, although the extent or form of these changes were unknown.
-Two, do users show signals of arousal in physiological responses to changes in system latency?
-As latency increased and due to the added annoyance of dealing with a less responsive system, we expected users to begin showing signs of stress and frustration.
-Three, are responses to delay effects in subjects mediated by cognition and/or emotion?
-We expected delay effects in subjects to be mediated primarily by emotion, and in particular expected these effects to be correlated with the level of the reduction in system responsiveness.
-And four, are these effects mediated by personality indicators in any way?
-It was our hypothesis that, among others, the individual trait of \emph{neuroticism}~\cite{john1999big} would play a particularly important role in mediating effects of reduced system responsiveness on subjects.
-
-Our results partially agree with our initial hypotheses.
-We found that reduced system responsiveness induces \emph{additional} behavioral slow-down in users of \gls{WCA}.
-This additional slow-down moreover scales with the strength of the reduction in system responsiveness.
-Furthermore, we note that humans tend to speed up as the task progresses.
-However, as system responsiveness decreases, this effect is strongly dampened.
-We also found that the effects of reduced system responsiveness on users linger for at least \num{4} steps after system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
-
-Finally, we also found that the above effects are modulated by measures of individual levels of personality.
-In particular, the trait of \emph{neuroticism} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
+We found that reduced system responsiveness induces additional behavioral slow-down in \gls{WCA} users, and this effect scales with the strength of the reduction in system responsiveness.
+Furthermore, we note that humans tend to speed up as the task progresses;
+however, as system responsiveness decreases, this effect is strongly dampened.
+The effects of reduced system responsiveness on users linger for at least four steps after the system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
+Finally, the trait of \emph{neuroticism}~\cite{john1999big} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
 
@@
         In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target interval of time denoted \emph{delay}.
+        Figures originally published in \cref{paper:olguinmunoz2021impact}.
     }\label{fig:regularwca-vs-delaywca}
@@
 \end{align}
-Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
+%Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
 In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note that the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
@@
     \includegraphics[width=.6\textwidth]{publications/2021ImpactDelayedResponse/Fig4c}
-    \caption{Structure of a \emph{block} in the experimental LEGO task.}\label{fig:stepblock}
+    \caption{%
+        Structure of a \emph{block} in the experimental LEGO task.
+        Figure originally published in \cref{paper:olguinmunoz2021impact}.
+    }\label{fig:stepblock}
     \todo[inline]{This figure may need reworking for clarity.}
@@
 
-The actual \gls{WCA} task employed for this study corresponded to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
+\medskip
+The \gls{WCA} task employed for this study corresponded to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
 Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consisted of a pseudo-random sequence of steps instructing subjects to add or remove pieces.
@@
 We introduced an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
-To study the effects of a delay applied across multiple steps, we manipulated the length of blocks during the task, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
-Furthermore, we defined a \emph{slice} as a continuous segment of \num{4} steps within a block.
-In other words, \num{4}-step blocks were composed of a single slice, whereas \num{8}- and \num{12}-step blocks were composed respectively of \num{2} and \num{3} consecutive slices.
-
-Using the above experimental design components, we then generated a task for each participant.
-We generated a pseudo-random permutation of the combinations of block length and delay values, and assigned a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
-Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reordered the initial permutation to generate \num{40} tasks.
+To study the effects of a delay applied across multiple steps, we manipulated the length of blocks during the task. %, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
+Furthermore, we defined a \emph{slice} as a continuous segment of \num{4} steps within a block;
+\num{4}-step blocks were composed of a single slice, whereas \num{8}- and \num{12}-step blocks were composed respectively of \num{2} and \num{3} consecutive slices.
+Using these experimental design components, we generated a task for each participant by arranging blocks according to a \emph{Latin-square}~\cite{keedwell2015latin} design.
+%We generated a pseudo-random permutation of the combinations of block length and delay values, and assigned a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
+%Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reordered the initial permutation to generate \num{40} tasks.
 
@@
 These correspond to well-established tools for the measurement of individual personality traits in humans.
-The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
-\begin{inlineenum}
-    \item agreeableness
-    \item conscientiousness
-    \item extroversion
-    \item neuroticism
-    \item openness
-\end{inlineenum}.
-The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
+%The \gls{BFI}~\cite{john1999big} consists of \num{44} questions assessing the traits of
+%\begin{inlineenum}
+%    \item agreeableness
+%    \item conscientiousness
+%    \item extroversion
+%    \item neuroticism
+%    \item openness
+%\end{inlineenum}.
+%The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
 The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
@@
 In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
-This array included devices to capture four physiological measures:
-\begin{inlineenum}
-    [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
-    \item \gls{GSR} (also known as electrodermal activity)
-    \item accelerometer data from the dominant wrist
-    \item brain activity in the form of \gls{EEG}
-    \item heart rate
-\end{inlineenum}.
+This array included devices to capture a number of physiological measures, including \gls{EEG}, \gls{HR}, and \gls{GSR}.
+%\begin{inlineenum}
+%    [itemjoin={{, }}, itemjoin*={{, and }}, label={(\arabic*)}]
+%    \item \gls{GSR} (also known as electrodermal activity)
+%    \item accelerometer data from the dominant wrist
+%    \item brain activity in the form of \gls{EEG}
+%    \item heart rate
+%\end{inlineenum}.
 Unfortunately, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
@@
             \gls{TTF} vs average execution time, grouped by step slice in a block.
-            Slice \num{1} corresponds to steps \numrange{1}{4}, \num{2} to steps \numrange{5}{8}, and \num{3} to steps \numrange{9}{12} in a block.
-            Blocks with \num{4} steps only contribute to slice \num{1}, and blocks with \num{8} steps to slices \num{1} and \num{2}.
+%            Slice \num{1} corresponds to steps \numrange{1}{4}, \num{2} to steps \numrange{5}{8}, and \num{3} to steps \numrange{9}{12} in a block.
+%            Blocks with \num{4} steps only contribute to slice \num{1}, and blocks with \num{8} steps to slices \num{1} and \num{2}.
             Transparent bands indicate \SI{95}{\percent} \glspl{CI}.
@@
 
-The results summarized above lead to interesting implications for the design and optimization of \gls{WCA} application systems.
-The behavioral slow-down in users as a response to reductions in system responsiveness translates into noticeably extended application lifetimes, even for short durations of impairment.
-Combined with the permanence of these effects for a period even after the system returns to an improved level of responsiveness, this could lead to unconventional consequences for resource allocation, in particular in multi-tenant deployment and in cases where a user may be able to finish the task before the effects of slow-down subside.
-In such cases, it might be more efficient to not divert resources to the impaired application, as fair degradation may not ultimately be beneficial to the system as a whole.
-It might even be more beneficial to divert resources \emph{away} from the impaired application, to ensure optimal system responsiveness for other tenants in the system.
+The findings presented above reveal interesting implications for the design and optimization of \gls{WCA} applications.
+The noticeable slow-down in user behavior as a response to a decrease in system responsiveness leads to extended application lifetimes even for brief periods of impairment.
+This effect is compounded by the persistence of the slow-down effect for a certain period even after the system returns to an improved level of responsiveness.
+As a result, this could have unconventional consequences for resource allocation, particularly in multi-tenant deployment and cases where users may be able to finish their task before the effects of slow-down subside.
 
-These implications, however, are merely speculative.
-Moreover and as previously discussed, due to the complexity of these systems, accurate study of the actual consequences of the results discussed in the present section can only be achieved through an experimental approach.
-\todo[inline]{Does this need a better transition?}
+In such instances, it might be more efficient not to redirect resources to the impaired application, as equitable degradation may not necessarily benefit the system as a whole.
+It could even be more beneficial to allocate resources \emph{away} from the impaired application to ensure optimal system responsiveness for other tenants in the system.
+This unconventional approach to resource allocation can potentially improve the overall system performance and provide better average \gls{QoE} and \gls{QoS}.
+
+However, the implications mentioned above are purely speculative, and an experimental approach is required to accurately study the consequences of these results.
+The complexity of \gls{WCA} application systems makes it challenging to predict the outcomes of resource allocation strategies, and experimentation is necessary to confirm or reject these speculations.
 
@@
 
-Our new timing model produces realistic execution times for each step, according to the above effects.
-We achieve this through the following design:
+Our new timing model generates realistic execution times for each step by taking two inputs: the most recent measured \gls{TTF} and a level of normalized neuroticism.
+The model is stateful and calculates a weighted average of the 12 most recent \glspl{TTF}, with the most recent step accounting for approximately \SI{50}{\percent} of the weighted result.
+The resulting weighted rolling \gls{TTF} is then used to look up potential values for the execution time in the original data and generate a realistic execution time for the step based on the selected level of neuroticism.
 
-\begin{enumerate}
-    \item Our model takes two inputs, the measured \gls{TTF} for the most recent step and a level of normalized neuroticism.
-    We use two levels of neuroticism: \emph{low}, for normalized values in the range \ensuremath{[0, 0.5)}; and \emph{high}, for the range \ensuremath{[0.5, 1.0]}.
-    \item The model is stateful, keeping the values of the \num{12} most recent \glspl{TTF} in memory.
-    \item Whenever a new \gls{TTF} is received, the thirteenth most recent \gls{TTF} is dropped, and a weighted average is calculated using the values in memory.
-    This value is obtained using the formula
-    \begin{equation}
-        \label{eq:modelweights}
-        w_{n - i} =
-        \left\{ \begin{array}{ll}
-                    \frac{e^{-0.7 i}}{\sum\limits^{12}_{j=1} e^{-0.7 j}} & 1 \leq i \leq 12 \\
-                    &                  \\
-                    0                                                    & i > 12
-        \end{array} \right.
-    \end{equation}
-    This results in the \gls{TTF} for the most recent step accounting for approximately \SI{50}{\percent} of the weighted result, with this percentage decaying by a factor of about \ensuremath{\frac{1}{2}} for each step further back in the sequence.
-
-    \item The resulting weighted rolling \gls{TTF} is used to look up potential values for the execution time in the original data from \cref{paper:olguinmunoz2021impact}.
-    We do this by pre-processing the original data from the \num{40} independent repetitions of the LEGO task with this same rolling weighted average formula, and then binning samples according to the resulting \num{7}-quantiles for weighted \glspl{TTF}.
-    The weighted \gls{TTF} calculated during runtime is then matched to the corresponding bin of samples.
-
-    \item The selected samples are further filtered according to the selected level of neuroticism for the model.
-    \item Finally, a realistic execution time for the step is generated either by directly sampling the execution time values from the selected samples, or first fitting a theoretical \gls{exGaussian} distribution to these values and then sampling the distribution instead.
-\end{enumerate}
+To achieve this, the original data from the \num{40} independent repetitions of the LEGO task from \cref{paper:olguinmunoz2021impact} is pre-processed using the same rolling weighted average formula and binned into \num{14} bins corresponding to the seven quantiles for weighted TTF and two levels of neuroticism.
+During runtime, the weighted \gls{TTF} calculated is matched to the corresponding bin of samples, which are then filtered based on the selected level of neuroticism.
+The execution time is generated either by directly sampling the execution time values from the selected samples or by fitting a theoretical \gls{exGaussian} distribution to these values and then sampling the distribution instead.
 
@@
 \Cref{fig:model:exectimesoversteps,fig:model:exectimestransitions} illustrate the adherence of the model to the conclusions obtained from our work in \cref{paper:olguinmunoz2021impact}.
-
 \Cref{fig:model:exectimesoversteps} shows the output of the execution time model described above over a series of steps at three levels of \gls{TTF};
@@
 
+\todo[inline]{Do we need more experiments?}
+
 In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
 The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
-
-\todo[inline]{Do we need more experiments?}
-\todo[inline]{Include discussions on the metrics here (grab inspiration from EdgeDroid2 paper.)}
 
@@
 Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
-In \cref{fig:taskcompletiontimesdiff} we present the results of an experimental exploration of this difference.
 We compare the total task durations of emulated \num{45}-step tasks using the EdgeDroid benchmarking tool with the improved model of human behavior to a baseline model.
@@
 
-These results show clear and significant differences in task durations between the realistic model and the baseline, with a reduction of up to \SI{11}{\percent} at low levels of resource contention for both levels of neuroticism.
-At higher degrees of resource contention, the high neuroticism parameterization also results in task durations more than \SI{8}{\percent} longer than the baseline.
+The results of the experiment, as presented in \cref{fig:taskcompletiontimesdiff}, show that there are clear and significant differences in task durations between the models.
+At low levels of resource contention, there was a reduction of up to \SI{12}{\percent} in task duration for both levels of neuroticism when using the realistic model.
+However, at higher levels of resource contention, the high neuroticism parameterization resulted in task durations that were more than \SI{8}{\percent} longer than the baseline.
+These findings demonstrate the importance of using a realistic model of human timing behavior to accurately predict and optimize task completion times.
 
-\todo[inline]{Briefly discuss implications of the above?}
+Overall, this case study highlights the potential benefits of using our methodology coupled with a realistic model of human timing behavior in edge computing systems.
+By considering higher order effects on execution times, such as the impact of neuroticism, we can improve the accuracy of our predictions and optimize system performance.
+However, it is important to note that the effects of using a realistic model may vary depending on the specific context and task being performed.
+Further research is needed to explore the generalizability of our findings and to investigate other factors that may impact task completion times in edge computing systems.
 

### 2023-02-17 background.tex
thesis tweaks
draft

@@
     \centering
-    \begin{subfigure}{\columnwidth}
+    \begin{subfigure}{\textwidth}
         \centering
-        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/task.png}
+        \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/task}
         \caption{%
@@
     \end{subfigure}\\
-    \begin{subfigure}{\columnwidth}
+    \begin{subfigure}{\textwidth}
         \centering
-        \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/step_time.png}
+        \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
         \caption{%
@@
 In this work we focus on a particular category --- ``step-based'' \gls{WCA} --- that have as their goal the guiding of a user through a sequential task.
-Examples of such applications are the LEGO and IKEA assistants~\cite{Chen2015LEGO,Chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
+Examples of such applications are the LEGO and IKEA assistants~\cite{chen2015early,chen2018application}, in which users are guided step-by-step through the process of assembling a LEGO model and an IKEA lamp, respectively.
 
@@
 Whenever the assistant detects that the user has correctly or incorrectly performed an instruction, it provides a new instruction to either advance the task or correct the detected mistake.
-The application otherwise remains silent and out-of-the-way of the user; that is, samples which do not generate a new instruction (e.g.~because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
+The application otherwise remains silent and out-of-the-way of the user;
+that is, samples which do not generate a new instruction (e.g.~because they captured an intermediate or unfinished state, or simply noise) are silently discarded.
 Herein lies one of the key characteristics of these applications: the user only consciously interacts with the application whenever they finish an instruction, and thus these are the \emph{only} points in time at which they can notice changes in system responsiveness.
@@
 
-In~\cite{olguinmunoz:impact2021} we studied the effects of system responsiveness on human behavior in step-based \gls{WCA} in a controlled experiment.
-We employed a modified and instrumented version of the above-mentioned, step-based LEGO \gls{WCA}~\cite{Chen2015LEGO}.
+In~\cite{olguinmunoz2021impact} we studied the effects of system responsiveness on human behavior in step-based \gls{WCA} in a controlled experiment.
+We employed a modified and instrumented version of the above-mentioned, step-based LEGO \gls{WCA}~\cite{chen2015early}.
 \num{40} subjects interacted with the assistant, performing a \num{169}-step task while we altered the responsiveness in realtime and captured key application and task performance metrics.
@@
 The correlation between this variable and execution time was then studied.
-To compensate for the distribution of \( t_c \)\footnote{\( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality.}~\cite{olguinmunoz:impact2021}, in the present work we adjust the nominal delay \( D \) by a factor \num{1.5}.
+To compensate for the distribution of \( t_c \)\footnote{\( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality.}~\cite{olguinmunoz2021impact}, in the present work we adjust the nominal delay \( D \) by a factor \num{1.5}.
 We will refer to this adjusted delay as \emph{\glspl{TTF}}.
@@
 % Delay, however, was merely an experimental tool, and if we consider
-% \begin{enumerate*}[itemjoin={{; }}, itemjoin={{; and }}]
+% \begin{inlineenum}
 %     \item that \( t_c \), and conversely the wait time \( \mathcal{W} \) of a step, can be assumed to be uniformly distributed in the interval \( [t_{n - 1}, t_n] \) without loss of generality~\cite{olguinmunoz:impact2021}
 %     \item that for a step subject to delay \( D \), \( t_k - t_{k - 1} = D \) for all \( k \in \{1, 2, \ldots, n\} \)
-% \end{enumerate*},
+% \end{inlineenum},
 % then it follows that the expected \gls{TTF} for steps subject to a delay \( D \) can be expressed as
@@
     For sequences of \num{12} steps in an unimpaired application state, humans executed the final four steps on average \SI{36}{\percent} faster than they did the first four.
-    However, this effect is dampened by reduced system responsiveness, and actually inverts at the highest levels of system impairment; humans actually become progressively slower the longer they spend in a degraded system state.
+    However, this effect is dampened by reduced system responsiveness, and actually inverts at the highest levels of system impairment;
+    humans actually become progressively slower the longer they spend in a degraded system state.
 
@@
 
-We recorded variables related to well-known individual differences, encompassing both the \glspl{BFI}~\cite{oliver:bfi1999}, and the \glspl{ITQ}~\cite{witmer1998measuring}.
+We recorded variables related to well-known individual differences, encompassing both the \glspl{BFI}~\cite{john1999big}, and the \glspl{ITQ}~\cite{witmer1998measuring}.
 Out of the individual-difference variables, the most salient effect on performance corresponded to \emph{neuroticism}, a \gls{BFI} trait linked to low tolerance for stress and high emotional reactivity, and which has previously been linked to higher \emph{delay discounting} rates~\cite{hirsh2008delay}.
@@
 %     \centering
-%     \includegraphics[width=\columnwidth]{figs/new_model/correlation_neuro_exectime.png}
+%     \includegraphics[width=\textwidth]{figs/new_model/correlation_neuro_exectime.png}
 %     \caption{%
@@
 
-In this work, we will use a normalized scale to describe neuroticism, derived from the minimum and maximum obtainable values for this variable in the \gls{BFI}~\cite{oliver:bfi1999}.
+In this work, we will use a normalized scale to describe neuroticism, derived from the minimum and maximum obtainable values for this variable in the \gls{BFI}~\cite{john1999big}.
 \emph{Low} and \emph{high} neuroticism will refer to the \( [0.0, 0.5) \) and \( [0.5, 1.0] \) ranges respectively.
@@
     \centering
-    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0.png}
+    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0}
     \caption{%
-        \( \mu \) parameter of \gls{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \gls{TTF} in~\cite{olguinmunoz:impact2021}.
+        \( \mu \) parameter of \gls{exGaussian} distributions fitted to execution times of the first four steps of segments of steps subject to the same \gls{TTF} in~\cite{olguinmunoz2021impact}.
         Distributions were fitted using \gls{MLE}.
@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/dist_fits_neuro.png}
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/dist_fits_neuro}
     \caption{%
@@
 
-Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \gls{TTF}, and continuous segments of steps subject to the same \gls{TTF}, were well-fit by an \gls{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey1951kolmogorov}.
+Furthermore, we found that execution times, when grouped by experimental variables such as neuroticism, \gls{TTF}, and continuous segments of steps subject to the same \gls{TTF}, were well-fit by an \gls{exGaussian} distribution, as verified using Kolmogorov-Smirnov goodness-of-fit tests~\cite{massey_jr1951kolmogorov}.
 When grouping by level of neuroticism, \gls{TTF}, and \emph{slice}\footnote{%
-In~\cite{olguinmunoz:impact2021} the \emph{slice} to which a step belongs to refers to whether the step occurred in the first, second, or third four-step segment of a sequence of steps subject to the same \gls{TTF}.
+In~\cite{olguinmunoz2021impact} the \emph{slice} to which a step belongs to refers to whether the step occurred in the first, second, or third four-step segment of a sequence of steps subject to the same \gls{TTF}.
 }, the best fit statistic was \ensuremath{0.028} (\ensuremath{p = 0.999}).
-This distribution has an ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+This distribution has an ample body of research supporting its suitability for the modeling of the timing of human actions and reaction times~\cite{rohrer1994analysis,palmer2011what,marmolejo_ramos2022generalised}.
 We found that the effects of neuroticism on execution times were clearly identifiable in the fitted distributions, in particular in their means and tails.

### 2023-02-17 implications_footprint.tex
draft

@@
 In order to illustrate the consequences of using a less realistic model that does not take into account higher order effects, we introduce here a reference model to which we will compare our more realistic models.
-This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \gls{exGaussian} distribution fitted to all execution time samples collected for~\cite{olguinmunoz:impact2021}.
+This model represents a first-order approximation to empirical execution time modeling, and consist simply of an \gls{exGaussian} distribution fitted to all execution time samples collected for~\cite{olguinmunoz2021impact}.
 This distribution is then randomly sampled at runtime to obtain execution times for each step, without any adjustment to the current state of the system.
@@
     \centering
-    \begin{subfigure}[]{\columnwidth}
+    \begin{subfigure}[t]{.45\textwidth}
         \centering
@@
         }
-    \end{subfigure}
-    \begin{subfigure}[]{\columnwidth}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{.45\textwidth}
         \centering
@@
     \centering
-    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/EdgeDroid2ExperimentalSetup.png}
+    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/EdgeDroid2ExperimentalSetup}
     \caption{%
@@
 The models, reference and realistic, are deployed on the Raspberry Pi clients of the testbed depicted in \cref{fig:expsetup}.
-For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \gls{WCA} platform~\cite{Chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
+For this, the timing models and frame generator are integrated into a custom Python3 client for the Gabriel \gls{WCA} platform~\cite{chen2018application}, which are then paired with real instances of Gabriel deployed on the cloudlet.
 Clients and cloudlet communicate over an \gls{IEEE} \num{802.11}b/g wireless network.
@@
     \centering
-    \begin{subfigure}[]{\columnwidth}
+    \begin{subfigure}[t]{.45\textwidth}
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed.png}
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed}
         \caption{%
@@
         }
-    \end{subfigure}
-    \par\bigskip
-    \begin{subtable}{\columnwidth}
+    \end{subfigure}%
+    \hfill%
+    \begin{subfigure}[t]{.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed_diff}
+        \caption{%
+            Percentage difference in mean application lifetimes with respect to the reference model.
+            Confidence intervals have been omitted due to the low number of samples and the use of the geometric mean.
+        }
+    \end{subfigure}\\
+    \bigskip
+    \begin{subtable}{\textwidth}
         \centering
@@
     \end{subtable}
-    \par\bigskip
-    \begin{subfigure}[]{\columnwidth}
-        \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed_diff.png}
-        \caption{%
-            Percentage difference in mean application lifetimes with respect to the reference model.
-            Confidence intervals have been omitted due to the low number of samples and the use of the geometric mean.
-        }
-    \end{subfigure}
     \caption{Application lifetimes in the realistic scenarios.}\label{fig:testbed_lifetimes}
@@
 
-We deploy configurations running \num{45}-step versions of the LEGO task described in~\cite{olguinmunoz:impact2021}.
+We deploy configurations running \num{45}-step versions of the LEGO task described in~\cite{olguinmunoz2021impact}.
 The testbed configurations include setups with \numlist{1;4;7;10} clients.

### 2023-02-17 implications_optimization.tex
draft

@@
 Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
-In \textcite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
+In \textcite{moothedath2021energy,moothedath2022energy1}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
 The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
@@
 With a value much smaller than the execution time of the event, the \gls{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
-Next, in~\cite{Moothedath2022Aperiodic}, the author retains the model and removes the constraint of periodicity to find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+Next, in~\cite{moothedath2022energy2}, the author retains the model and removes the constraint of periodicity to find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
 
-In this work, we use the modeling from~\cite{Moothedath2022Aperiodic} to find the optimum \emph{aperiodic} sampling interval.
+In this work, we use the modeling from~\cite{moothedath2022energy2} to find the optimum \emph{aperiodic} sampling interval.
 However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
@@
 Here, $\alpha, \beta$ and $C$ are constants responsible for modifying the objective function from one metric to another.
-For instance, in the modeling used by~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient,Moothedath2022Aperiodic}, the chosen characterization results in a metric which is equal to the energy penalty.
+For instance, in the modeling used by~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}, the chosen characterization results in a metric which is equal to the energy penalty.
 Our solution which minimizes \cref{eq:epsilon_terminal} is provided in \cref{tn_approx_rayleigh}.
@@
 We introduce here a reference scheme to which we will compare our approach.
-In~\cite{Wang2019Towards}, \citeauthor{Wang2019Towards} introduce an adaptive sampling scheme for \gls{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
+In~\cite{wang2019towards}, \citeauthor{wang2019towards} introduce an adaptive sampling scheme for \gls{WCA} intended to reduce the number of samples processed per step while still meeting application responsiveness bounds.
 At every sampling instant \( t \), the scheme adapts the sampling rate \( R(t) \) of the system according to the estimated likelihood of the user having finished the step,
@@
 \( \varphi \) corresponds to a scaling factor and \( t \) to the time of the current sampling instant with respect to the start of the step.
-Finally, \( CDF \) corresponds to the \gls{CDF} of a distribution describing the execution times for the current step; \citeauthor{Wang2019Towards} used a single static Gaussian distribution for all steps in their work.
+Finally, \( CDF \) corresponds to the \gls{CDF} of a distribution describing the execution times for the current step; \citeauthor{wang2019towards} used a single static Gaussian distribution for all steps in their work.
 
@@
 
-In the following, we will show the effects of our optimization approach combined with our timing models compared to the state-of-the-art approach~\cite{Wang2019Towards}.
+In the following, we will show the effects of our optimization approach combined with our timing models compared to the state-of-the-art approach~\cite{wang2019towards}.
 For this we implement these sampling schemes in Python and run a number of simulations with them.
@@
 
-We implement \citeauthor{Wang2019Towards}'s original design using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz:impact2021} for \gls{CDF} calculation.
+We implement \citeauthor{wang2019towards}'s original design using a Gaussian distribution fitted to all the execution times collected for~\cite{olguinmunoz2021impact} for \gls{CDF} calculation.
 This scheme does not include an embedded timing model, and uses the same \gls{CDF} for every step.
@@
         \( R_\text{min} \) & \SI{0.5}{\hertz} & Derived as \( R_\text{min} = {(2 w_0)}^{-1} \) \\
-        \( \varphi \) & \num{1.5} & Scaling factor, \textcite{Wang2019Towards}\\
+        \( \varphi \) & \num{1.5} & Scaling factor, \textcite{wang2019towards}\\
         \( P_0 \) & \SI{15}{\milli\watt} & Idle power \\
@@
 We set the \( w_0 \) and \( \beta \) factors of our sample-count-optimized scheme to \SI{1.0}{\second} and \num{1.0}, respectively, for mathematical simplicity, and derive \( \alpha = 1.9 \sigma^{-1} \).
-Next, we derive \( R_\text{min} \) for \textcite{Wang2019Towards}'s scheme from \( w_0 \).
+Next, we derive \( R_\text{min} \) for \textcite{wang2019towards}'s scheme from \( w_0 \).
 As discussed in \cref{sec:background}, we assume that wait times are uniformly distributed between \SI{0}{\second} and sampling interval of each step.
@@
     \caption{%
-        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \gls{CDF}-based approach.
+        Summary of results for experiment comparing the sample-count-optimized aperiodic sampling scheme to the reference schemes and \textcite{wang2019towards}'s \gls{CDF}-based approach.
     }\label{fig:optimization:samples}
@@
 The results of this investigation are presented in \cref{fig:optimization:samples}, and clearly show the advantages of using the sample-count-optimized scheme over the current state-of-the-art.
-The performance of \textcite{Wang2019Towards}'s approach appears to degrade with lower \glspl{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
+The performance of \textcite{wang2019towards}'s approach appears to degrade with lower \glspl{RTT}, exponentially oversampling as latency tends to zero and the maximum sampling rate of the system tends to infinity.
 On the other hand our sampling scheme consistently matches or beats the state-of-the-art while maintaining a relatively constant behavior with respect to \glspl{RTT}.
@@
 In the following, we thus take the general solution, \cref{eq:epsilon_terminal}, and find the appropriate \( \alpha \) and \( \beta \) to minimize the energy consumed per step, i.e.\ \( \mathcal{E}=E \).
-We directly take the modeling from~\cite{Moothedath2022Aperiodic} with a necessary modification in the  assumption of one-way communication in all but the final sample.
+We directly take the modeling from~\cite{moothedath2022energy2} with a necessary modification in the  assumption of one-way communication in all but the final sample.
 With feedback given even to the discarded samples in our model and communication delay defined as the total delay in either direction, we have, 
@@
 We refer to this scheme as the \emph{energy-optimized aperiodic} sampling scheme.
-For the power constants, we reuse the values estimated by the authors in~\cite{Moothedath2022EnergyEfficient}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
+For the power constants, we reuse the values estimated by the authors in~\cite{moothedath2022energy1}, \( P_\text{c} = 45\,\si{\milli\watt} \) and \( P_\text{0} = 15\,\si{\milli\watt} \).
 On the other hand, for the timing variables we define a constant processing delay \( \tau_\text{p} = 250\,\si{\milli\second} \) across all configurations and repetitions of the experiments; given then a constant \gls{RTT} for the task, we set \( \tau_\text{c} = \text{\gls{RTT}} - \tau_\text{p} \).
@@
     \caption{%
-        Summary of results for experiment comparing the energy-optimized aperiodic sampling scheme to the reference schemes and \textcite{Wang2019Towards}'s \gls{CDF}-based approach.
+        Summary of results for experiment comparing the energy-optimized aperiodic sampling scheme to the reference schemes and \textcite{wang2019towards}'s \gls{CDF}-based approach.
     }\label{fig:optimization:energy}
@@
 
-Again, we include the reference greedy and offline optimum schemes, as well as \textcite{Wang2019Towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \gls{RTT}.
+Again, we include the reference greedy and offline optimum schemes, as well as \textcite{wang2019towards}'s approach, and repeat the experiment \num{100} times for each combination of sampling scheme, neuroticism, and \gls{RTT}.
 
 The results of these experiments are presented in \cref{fig:optimization:energy}, and they clearly illustrate the advantages of the integration of our timing models with an adaptive, energy-optimized sampling scheme when compared to unoptimized and state-of-the-art sampling schemes.
-Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{Wang2019Towards}'s state-of-the-art, and is up to \SI{40}{\percent} more energy efficient than the greedy scheme.
+Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{wang2019towards}'s state-of-the-art, and is up to \SI{40}{\percent} more energy efficient than the greedy scheme.
 Furthermore, once again the behavior of our approach is more consistent and reliable than the competition, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.

### 2023-02-17 intro.tex
draft

@@
 These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
-\gls{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
-In particular, following the success of non-wearable \gls{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \gls{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.
+\gls{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,ha2014towards,satyanarayanan2019augmenting}, and have since expanded to a greater range of use cases.
+In particular, following the success of non-wearable \gls{XR} and cognitive assistance in industrial settings~\cite{funk2015cognitive,wang2022comprehensive}, there is increasing interest in the research community in the application of \gls{WCA} for step-by-step assistance for complex assembly tasks~\cite{chen2017empirical,belletier2021wearable}.
 
@@
 
-The combination of these opposing characteristics has led to \gls{WCA} applications being identified in the literature as prime candidates for offloading to the edge~\cite{Ha2014towards,Chen2017Empirical,Chen2018application}.
+The combination of these opposing characteristics has led to \gls{WCA} applications being identified in the literature as prime candidates for offloading to the edge~\cite{ha2014towards,chen2017empirical,chen2018application}.
 However, many unknowns still remain before consumer-scale adoption of these applications can become a reality.
@@
     \item\label{item:contrib:model} We introduce the first, to our knowledge, stochastic model for human timings in \gls{WCA} applications.
-    Using the data collected for~\cite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
+    Using the data collected for~\cite{olguinmunoz2021impact} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
     We also introduce a novel way to generate dynamic traces of frames for \gls{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
-    We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
+    We name this new model \emph{\edgedroid}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid}.
     \item\label{item:contrib:footprint} Using this model, we study the implications of realistic human behavior for the \emph{application lifetime footprint} of \gls{WCA}, understanding this term as the duration of a specific complete execution of the application task.
-    In accordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
+    In accordance with previous work~\cite{olguinmunoz2021impact}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
     \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \gls{WCA} when considering human behavior using our model.
@@
 \cref{sec:relwork} discusses related work in the field of \gls{WCA} modeling.
-In \cref{sec:background} we define and discuss key concepts in \gls{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz:impact2021}.
+In \cref{sec:background} we define and discuss key concepts in \gls{WCA}, as well as summarize the key conclusions from our previous work relating to the relationship between system responsiveness and human behavior in these applications~\cite{olguinmunoz2021impact}.
 In \cref{sec:model} we detail our model for the generation of realistic human timing in \gls{WCA}.

### 2023-02-17 model.tex
draft

@@
 
-In the following, we employ the above insights together with the data collected for~\cite{olguinmunoz:impact2021} to design and build a probabilistic model of human behavior for \gls{WCA}.
+In the following, we employ the above insights together with the data collected for~\cite{olguinmunoz2021impact} to design and build a probabilistic model of human behavior for \gls{WCA}.
 We detail the construction of a model which uses the data collected to generate, at runtime, realistic execution times. 
@@
 We detail this in \cref{ssec:model:exectimes}.
-Neuroticism is incorporated in this aspect of the model, as it is the most salient individual difference in the data; however, other factors such as immersive tendency could be treated similarly.
+Neuroticism is incorporated in this aspect of the model, as it is the most salient individual difference in the data;
+however, other factors such as immersive tendency could be treated similarly.
 Second, the model must produce sequences of input samples for each step mimicking what a real human would generate; this is explained in \cref{ssec:model:frames}. 
@@
 % For this paper, we use this data to construct a probabilistic model for the generation of realistic execution times.
-The processing of the data from~\cite{olguinmunoz:impact2021} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \gls{TTF}.
+The processing of the data from~\cite{olguinmunoz2021impact} for the generation of execution times can be summarized as a grouping according to discretized levels of neuroticism and a weighted rolling average of \gls{TTF}.
 The resulting collections of execution times represent the distributions of these values for users with specific levels of neuroticism, interacting with systems at specific states of impairment and recent histories of impairment.
@@
 
-We verify the behavior of the above described model with respect to the four main conclusions of our work in~\cite{olguinmunoz:impact2021}, which were previously discussed in \cref{ssec:plos}.
+We verify the behavior of the above described model with respect to the four main conclusions of our work in~\cite{olguinmunoz2021impact}, which were previously discussed in \cref{ssec:plos}.
 These are reformulated as objectives below:
@@
     \centering
-    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/ttf_to_exectime.png}
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/ttf_to_exectime}
     \caption{%
@@
     \centering
-    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/figs/new_model/exectime_over_steps.png}
+    \includegraphics[height=25em]{publications/2023EdgeDroid2/figs/new_model/exectime_over_steps}
     \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/transitions.png}
+    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/transitions}
     \caption{%
@@
 Conversely, execution times are consistently lower than otherwise immediately after a change from a lower \gls{TTF} into a higher one.
-These results are once again in line with our findings in~\cite{olguinmunoz:impact2021}, in which we found lingering effects of transitions between levels of system impairment on human execution times.
+These results are once again in line with our findings in~\cite{olguinmunoz2021impact}, in which we found lingering effects of transitions between levels of system impairment on human execution times.
 
@@
 
-Apart from the aforementioned timing and performance data, for~\cite{olguinmunoz:impact2021} we also recorded all collected video frame samples together with matching metadata.
+Apart from the aforementioned timing and performance data, for~\cite{olguinmunoz2021impact} we also recorded all collected video frame samples together with matching metadata.
 For each video frame submitted to the \gls{WCA} during the tasks, we recorded
@@
     \centering
-    \includegraphics[width=\columnwidth]{publications/2023EdgeDroid2/model_data/frame_probabilities.png}
+    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/model_data/frame_probabilities}
     \caption{%
@@
 
-Using the above insights, together with the corresponding recorded video frames, we devise a scheme for the procedural generation of a synthetic trace for any step in \gls{WCA} task in the same category as those used in~\cite{olguinmunoz:impact2021}.
+Using the above insights, together with the corresponding recorded video frames, we devise a scheme for the procedural generation of a synthetic trace for any step in \gls{WCA} task in the same category as those used in~\cite{olguinmunoz2021impact}.
 We first prepare a discretized representation of the probability density map in \cref{fig:frameprobs}.

### 2023-03-02 background.tex
using import package

@@
         \centering
-        \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/task}
+        \includegraphics[width=.9\textwidth]{figs/task}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
+        \includegraphics[width=.9\textwidth]{figs/step_time}
         \caption{%
@@
     \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/new_model/mu_fits_exgaussian_slice0}
+    \includegraphics[width=.9\textwidth]{figs/new_model/mu_fits_exgaussian_slice0}
     \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/dist_fits_neuro}
+    \includegraphics[width=\textwidth]{figs/new_model/dist_fits_neuro}
     \caption{%

### 2023-03-02 implications_footprint.tex
using import package

@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_all_ttfs.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_all_ttfs.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_diff.png}
         \caption{%
@@
     \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/EdgeDroid2ExperimentalSetup}
+    \includegraphics[width=.9\textwidth]{figs/EdgeDroid2ExperimentalSetup}
     \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/lifetime_testbed_diff}
+        \includegraphics[width=\textwidth]{figs/new_model/lifetime_testbed_diff}
         \caption{%

### 2023-03-02 model.tex
using import package

@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/ttf_to_exectime}
+    \includegraphics[width=\textwidth]{figs/new_model/ttf_to_exectime}
     \caption{%
@@
     \centering
-    \includegraphics[height=25em]{publications/2023EdgeDroid2/figs/new_model/exectime_over_steps}
+    \includegraphics[height=25em]{figs/new_model/exectime_over_steps}
     \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/transitions}
+    \includegraphics[width=\textwidth]{figs/new_model/transitions}
     \caption{%

### 2023-03-03 intro.tex
working on paper

@@
 
-\glspl{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
+\todo[inline]{Needs rewrite}
+
+\gls{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
 These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.

### 2023-03-04 publications/2023EdgeDroid2/appendices.tex

--- 2022-12-19 appendices.tex
@@
-\appendix{}\label[appendixwithoutnumber]{appx1}
+%\appendix{}
+\section{Appendix}\label[appendixwithoutnumber]{appx1}
 % \subsection*{Sampling optimisation}
@@
 
-We first borrow the idea of checkpointing density from~\cite{Satoshi1992Optimal} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
+We first borrow the idea of checkpointing density from~\cite{satoshi1992optimal} and define an instantaneous sampling rate function $r(t)$ which is related to $\{t_n\}$ such that,
 \begin{alignat}{1}
@@
 
-As per the Euler-Lagrange equation from the calculus of variations~\cite{Bellman1954Dynamic,Arfken2015Calculus}, the extreme value of $\mathcal{E}$ is obtained at 
+As per the Euler-Lagrange equation from the calculus of variations~\cite{bellman1954dynamic,arfken2013calculus}, the extreme value of $\mathcal{E}$ is obtained at
 \begin{alignat}{1}
@@
 % Note that throughout the paper we use a exGaussian distribution but not Rayleigh. This is because, 
-Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \gls{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{Rohrer1994analysis,Palmer2011shapes,Marmolejo2022generalised}.
+Note that, due to the close similarity in their density functions, the task times can be approximated fairly equally to an \gls{exGaussian} distribution as well as a Rayleigh distribution, with the former attracting more attention from works like~\cite{rohrer1994analysis,palmer2011what,marmolejo_ramos2022generalised}.
 This is the reason why the above results are applicable in this work where we have predominantly considered \gls{exGaussian} distribution.

### 2023-03-04 publications/2023EdgeDroid2/conclusion.tex

--- 2022-12-19 conclusion.tex
@@
 It first introduces the \edgedroid{} model of human timing behavior for \gls{WCA}.
-This model represents a stochastic approach to execution time modeling which builds upon prior data~\cite{olguinmunoz:impact2021}.
+This model represents a stochastic approach to execution time modeling which builds upon prior data~\cite{olguinmunoz2021impact}.
 It further introduces a novel procedure for the generation of synthetic traces of frames in step-based \gls{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.
@@
 For instance, our current model only targets a particular class of step-based \gls{WCA}, and extension of our methodology to other classes of these applications, or even more generally to \gls{AR} and \gls{XR}, is on our roadmap.
-Our data~\cite{olguinmunoz:impact2021} also only considers young undergraduate students at a highly competitive university in the United States.
+Our data~\cite{olguinmunoz2021impact} also only considers young undergraduate students at a highly competitive university in the United States.
 An extension of this dataset and the model towards a more representative sample of the general population would surely be a valuable endeavor.

### 2023-03-04 publications/2023EdgeDroid2/implications_optimization.tex

--- 2023-02-17 implications_optimization.tex
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/sampling_optimization.png}
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/sampling_optimization_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/sampling_optimization_diff.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/energy_optimization.png}
+        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization.png}
         \caption{%
@@
         \centering
-        \includegraphics[width=\textwidth]{publications/2023EdgeDroid2/figs/new_model/energy_optimization_diff.png}
+        \includegraphics[width=\textwidth]{figs/new_model/energy_optimization_diff.png}
         \caption{%

### 2023-03-04 publications/2023EdgeDroid2/optimization_framework.tex

--- 2023-03-04 publications/2023EdgeDroid2/appendices.tex
@@
-%\appendix{}
-\section{Appendix}\label[appendixwithoutnumber]{appx1}
-% \subsection*{Sampling optimisation}
+\section{An optimization framework for \gls{WCA}}\label{sec:optframework}
 
-In this appendix, we give the mathematical derivation of the optimum aperiodic sampling interval discussed in \cref{ssec:optframework}.
+Our optimization approach focuses on the individual steps which comprise a complete \gls{WCA} task.
+Thus, we begin by assuming a given execution time distribution \( \mathcal{T} \), and take the modeling and partial solution approach from recents works on energy efficient sampling in edge-based feedback systems.
+In \textcite{moothedath2021energy,moothedath2022energy1}, the authors model the energy in terms of the expected number of samples $\mathbb{E}[\mathcal{S}]$ and the expected wait time $\mathbb{E}[\mathcal{W}]$ experienced by the user.
+The authors then find the optimum periodic sampling interval that minimizes this energy, or equivalently the non-constant parts of this energy termed as \textit{Energy Penalty}.
+
+% In this optimization, the constant parts are excluded as it does not affect the process.
+With a value much smaller than the execution time of the event, the \gls{RTT} of the final sample is included in this constant part, which is why $\mathcal{W}$ is computed without it.
+Next, in~\cite{moothedath2022energy2}, the author retains the model and removes the constraint of periodicity to find the optimum aperiodic sampling instants $\{t_n,\,n=1,2,\dots\}$ that minimize the same energy penalty.
+
+In this work, we use the modeling from~\cite{moothedath2022energy2} to find the optimum \emph{aperiodic} sampling interval.
+However, instead of using their two-step approach to find the solution which includes a recursive solution followed by a bisection algorithm, we develop a novel, approximate, but easier solution to finding the set of the optimum aperiodic sampling intervals.
+Furthermore, instead of directly optimizing for energy, we start by noting that, in general, any objective metric in these applications which relates to sampling and the responsiveness of the system will present itself
+% that an objective metric of a variety of optimization problems that are related to sampling, RTT, responsiveness or wait time present themselves
+% % in a format similar to that of the energy,
+as a linear combination between $\mathbb{E}[\mathcal{S}]$ and $\mathbb{E}[\mathcal{W}]$ plus terms independent of the number of samples or wait time.
+Let $\mathcal{E}$ correspond to such an objective metric.
+Thus,
+\begin{alignat}{1}
+    \Rightarrow\mathcal{E}&=\alpha\mathbb{E}[\mathcal{S}]+\beta\mathbb{E}[\mathcal{W}]+C\;\label{eq:epsilon_terminal}
+\end{alignat}
+
+% This corresponds to a Lagrangian function of this optimization, with \ensuremath{\frac{\alpha}{\beta}} representing the Lagrangian parameter.
+Here, $\alpha, \beta$ and $C$ are constants responsible for modifying the objective function from one metric to another.
+For instance, in the modeling used by~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}, the chosen characterization results in a metric which is equal to the energy penalty.
+Our solution which minimizes \cref{eq:epsilon_terminal} is provided in \cref{tn_approx_rayleigh}.
+It assumes a $\mathcal{T}$ distributed according to a Rayleigh distribution with parameter $\sigma$.
+The complete mathematical derivation that leads to this formula is detailed in \cref{appx1}.
+
+\begin{alignat}{1}
+% r^*(t)&=\sqrt{\dfrac{\beta t}{2\alpha\sigma^2}}\nonumber\\
+% \Rightarrow
+    t_n&=\Big(3\sigma\!\sqrt{\tfrac{\alpha}{2\beta}}\Big)^{\frac{2}{3}}n^{\frac{2}{3}}\label{tn_approx_rayleigh}
+\end{alignat}
+
 We start with the general solution, where the objective function to be minimized is given by \cref{eq:epsilon_terminal}:

### 2023-03-04 publications/2023EdgeDroid2/related_work.tex

--- 2022-12-19 related_work.tex
@@
 It acts as an end-to-end realistic \gls{AR} workload, allowing thus researchers to measure the real end-to-end latencies of such systems.
-\textcite{Chen2017Empirical} use a number of prototype \gls{WCA} implementation to study and characterize real-world latency bounds in these systems.
+\textcite{chen2017empirical} use a number of prototype \gls{WCA} implementation to study and characterize real-world latency bounds in these systems.
 In~\cite{lecci2021open}, the authors collect and analyze large amounts of \gls{VR} network traffic, which they then use to construct a synthetic model for the generation of traces of such traffic.
@@
 We achieve this by building upon our own previous contributions to this field.
-In~\cite{olguin2018scaling,olguin2019edgedroid} we introduced a coarse approximation to human behavior modeling in \gls{WCA} we called \emph{EdgeDroid}.
-We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \gls{WCA} step-based task is replayed to a Gabriel~\cite{Chen2018application} backend.
+In~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} we introduced a coarse approximation to human behavior modeling in \gls{WCA} we called \emph{EdgeDroid}.
+We used a trace-driven approach, where a pre-recorded and pre-processed ``ideal'' trace of steps for a specific \gls{WCA} step-based task is replayed to a Gabriel~\cite{chen2018application} backend.
 In order to adapt to potential mismatches between system responsiveness at trace-capture time and trace-replay time, we used a simple \gls{FSM} to adapt the trace at the latter by replaying or skipping certain segments.
-In~\cite{olguinmunoz:impact2021} we studied the effects of reduced system responsiveness on human behavior in \gls{WCA} applications through human-subject studies.
+In~\cite{olguinmunoz2021impact} we studied the effects of reduced system responsiveness on human behavior in \gls{WCA} applications through human-subject studies.
 We found that humans generally pace themselves according to the perceived system responsiveness.
@@
 They use this characterization to develop a series of code optimizations which allow them to achieve a threefold increase in performance.
-\textcite{Wang2019Towards} develop a taxonomy for \gls{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
+\textcite{wang2019towards} develop a taxonomy for \gls{WCA} as well as strategies for workload reduction in these applications, including a novel adaptive sampling scheme with the goal of reducing the number of processed samples while maintaining application responsiveness.
 \textcite{huang2021proactive} present a framework for the proactive allocation of edge cloud resources while taking into account the inherent mobility of wearable and mobile \gls{AR}.
-\textcite{al2017energy} propose a resource allocation strategy which leverages collaboration between \gls{AR} applications to minimize energy consumption.
+\textcite{al_shuwaili2017energy} propose a resource allocation strategy which leverages collaboration between \gls{AR} applications to minimize energy consumption.
 As above, we have also previously directly contributed to this field.
-In~\cite{Moothedath2021EnergyOptimal,Moothedath2022EnergyEfficient} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \gls{WCA}.
-This is extended into an aperiodic strategy in~\cite{Moothedath2022Aperiodic}.
+In~\cite{moothedath2021energy,moothedath2022energy1} we find the optimum periodic sampling interval which minimizes the energy tracking human progress of a specific subtask in \gls{WCA}.
+This is extended into an aperiodic strategy in~\cite{moothedath2022energy2}.
 We further develop this approach in this work by deriving a more practical approximate solution to the aperiodic strategy, which we then use to implement a novel framework for the generic optimization of responsiveness-resource consumption trade-offs.

### 2023-03-05 body/part1/chapter1-introduction/background.tex
restructuring

--- 2023-02-07 body/part1/background.tex
@@
-\section{Background \& related work}
-\glsresetall%
-\todo[inline]{%
-    \textcite{satyanarayanan2009case}
-
-    Other papers detailing the advantages of Edge computing.
-
-    Next, what applications benefit from the Edge?
-    \glsfmtshort{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
-    Main points: prototypes and characterization of system latencies and load.
-}
+\section{Background}
 
 \subsection{Edge computing}
+\glsresetall%
 
@@
 \subsubsection{Experimental research in Edge Computing}
+\todo[inline]{Should be moved to related work or removed.}
 
@@
 
-\subsection{\glsfmtlong{WCA}}\label{background:wca}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{Figs/wca_state}
-    \caption{%
-        Mode of operation of \acs{WCA} applications.
-        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
-        Samples that do not result in feedback are silently discarded.
-    }\label{fig:wca}
-\end{figure}
+\subsection{\glsfmtlong{XR}}\label{background:xr}
+\todo[inline]{Rework into background XR}
 
@@
 \Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{Figs/wca_state}
+    \caption{%
+        Mode of operation of \acs{WCA} applications.
+        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
+        Samples that do not result in feedback are silently discarded.
+    }\label{fig:wca}
+\end{figure}
 
@@
 
-\subsubsection{Characterization of \acs{WCA}}
+\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
+\glsreset{NCS}
+\todo[inline]{Needs citations. Remove?}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
-    \caption[]{%
-        Breakdown of a step in a \gls{WCA} into its timing components.
-        \ensuremath{t_k | k \in \{1, \ldots, n \}} correspond to sampling instants.
-        \ensuremath{t_0} indicates the instant at which the instruction for step \ensuremath{M} is provided to the user and the first sample for said step is captured.
-        The instruction for step \ensuremath{M + 1} is provided at \ensuremath{t_{n+1}}.
-        Finally, \ensuremath{t_c} marks the instant at which the user finishes the instruction, and \ensuremath{t_n} the instant of capture of the first sample after the instruction has been completed;
-        this sample therefore also corresponds to the final sample of step \ensuremath{M}.
-        Figure originally included in \cref{paper:olguinmunoz2023realistic}.
-    }\label{fig:wcastep}
-\end{figure}
+\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
+A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
+Control systems are composed of
+\begin{inlineenum}
+    \item a \emph{plant}
+    \item \emph{sensors}
+    \item \emph{actuators}
+    \item a \emph{controller}
+\end{inlineenum}.
+Plant refers to the physical system being controlled.
+Sensors sample and encode the state of the plant and transmit it to the controller.
+The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
+The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
 
-An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
-We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
-In this section, we will provide some definitions relating to the operation of these applications.
-
-As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
-In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
-We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
-A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
-the time interval between these two events will be referred to as the \emph{step duration}.
-
-In the following, please refer to \cref{fig:wcastep}.
-\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
-\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
-this sample is the first sample to capture the finished state of the step.
-At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
-We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
-
-\glsreset{TTF}%
-From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
-However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
-The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
-A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
-If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
-\begin{inlineenum}
-    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
-    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
-    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
-\end{inlineenum}
-These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
-We will refer to them extensively in \cref{chap:contributions}.
-
-%\subsection{Mechanisms relating human behavior to system responsiveness}
-%
-%The question of how people respond to delay in a computer system is grounded in how people perceive time.
-%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-%The estimation of an event's duration varies with context of various types
-%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-%\item arousal can expand durations~\cite{droit_volet2011emotion}.
-%\end{inlineenum}
-%
-%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
-%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-%Violations of expectancy have a strong impact on the acceptability of computer systems.
-%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
-%
-%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
-%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
-%
-%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-%They then proposed a latency tolerance zone according to the task demands.
-%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
-%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
-%
-%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
-%These mechanisms could be cognitive or emotional in origin.
-%
-%A first possible explanation comes from research on cognitive and motor planning.
-%Delay may move users from relatively automatic to more attention-demanding processing.
-%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-%
-%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-%
-%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-%
-%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-%
-%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-%
-%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
-%These and other measures of individual variation were considered here.
-
-\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
-\todo[inline]{Needs citations.}
+\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
+see \cref{fig:csvsncs}.
+This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
+\glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
+This leads to enhanced scalability and flexibility, and allows improved coordination and collaboration among systems.
+Furthermore, \glspl{NCS} provide greater access to information, as they enable the exchange of data and information between control components and systems, leading to more informed decision-making and improved performance.
 
@@
 \end{figure}
-
-\glsreset{NCS}
-\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
-A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
-Control systems are composed of
-\begin{inlineenum}
-    \item a \emph{plant}
-    \item \emph{sensors}
-    \item \emph{actuators}
-    \item a \emph{controller}
-\end{inlineenum}.
-Plant refers to the physical system being controlled.
-Sensors sample and encode the state of the plant and transmit it to the controller.
-The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
-The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
-
-\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
-see \cref{fig:csvsncs}.
-This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
-\glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
-This leads to enhanced scalability and flexibility, and allows improved coordination and collaboration among systems.
-Furthermore, \glspl{NCS} provide greater access to information, as they enable the exchange of data and information between control components and systems, leading to more informed decision-making and improved performance.
 

### 2023-03-05 body/part1/chapter1-introduction/introduction.tex
restructuring

--- 2023-02-13 body/part1/introduction.tex
@@
-\section{Introduction}
-
 The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
@@
 The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
-
-\section{Summary of the key contributions of this thesis}\label{sec:summary_contributions}
-
-This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
-Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound latency-sensitive applications such as \gls{WCA} and \gls{NCS}.
-This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting case studies of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
-
-Secondly, we present a deeper exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
-This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
-
-Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
-In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
-We conclude that significant improvements can be achieved through the use of our human user model.
-
-\subsection{Overview of each included paper}
-
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
-This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
-\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
-We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-
-\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \glspl{NCS}.
-As discussed above, these applications are similarly difficult as \gls{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
-\todo[inline]{More argumentation or refer to previous section.}
-We apply thus in this work our methodology to \gls{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
-The software acts as a middleware which abstracts away the network from the development of these workloads, allowing for quick prototyping and deployment.
-This work also showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.
-
-\cref{paper:olguinmunoz2022ainur} presents a tangential contribution.
-It introduces the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
-This framework represents an ancillary contribution, crucial for the research presented in these works.
-Without it, the experimental approach described in these works would not have been feasible.
-
-Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
-Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
-In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
-
-\section{Structure of this thesis}
-
-\todo[inline]{This thesis is structured as follows.}

### 2023-03-05 body/part1/chapter2-relatedwork/contributions.tex
restructuring

--- 2023-02-16 body/part1/introduction.tex
@@
-\section{Introduction}
-
-The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
-One of the most promising of these is Edge Computing, which seeks to bring computing closer to the end user, reducing latency and increasing reliability.
-Another key player in this space is 5G, the latest generation of mobile communication networks that offers higher data rates, lower latency, and improved reliability.
-The combination of these two technologies holds great promise for the deployment of latency-sensitive applications, including \glspl{CPS}, \glspl{NCS}, and \gls{WCA} applications, among others.
-
-Despite these advancements, there are still significant challenges in understanding and scaling these systems, particularly in regard to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
-The aim of this thesis is to make a contribution to the field by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
-The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
-Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of \gls{MEC}.
-
-Our methodology is based on the emulation of target workloads on actual Edge infrastructure.
-We replace the client side of the system with a realistic emulation of the desired behaviors, implemented in software deployed on \gls{COTS} general-purpose computing devices.
-In our initial implementation, these correspond to low-cost, easily replaceable, and scalable Raspberry Pi 4 Model B \glspl{SBC}.
-
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, COTS general-purpose hardware such as \glspl{SBC}.
-It also preserves the realism of effects stemming from the hardware and network.
-In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
-
-The methodology also provides improved repeatability and replicability.
-Repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
-These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
-
-Our methodology provides a comprehensive and realistic assessment of the performance of latency-sensitive applications deployed on Edge infrastructure.
-The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
-
 \section{Summary of the key contributions of this thesis}\label{sec:summary_contributions}
+\todo[inline]{Needs tweaking to match scope and related work.}
 

### 2023-03-05 body/part1/chapter2-relatedwork/relatedwork.tex
restructuring

\section{Related work}

\subsection{Emulation of \glsfmtshort{XR} systems}

\subsection{Emulation of \glsfmtshortpl{NCS}}

### 2023-03-05 body/part1/chapter3-contributions/wca-implications.tex
restructuring

--- 2023-02-16 body/part1/overview/extending-wca.tex
@@
 By incorporating a realistic model of human behavior, we will be able to build a more comprehensive understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be applied to address these challenges.
+
+\subsection{Characterization of \glsfmtshort{WCA}}
+\todo[inline]{Rework}
+
+An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
+We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
+In this section, we will provide some definitions relating to the operation of these applications.
+
+As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
+In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
+We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
+A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
+the time interval between these two events will be referred to as the \emph{step duration}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
+    \caption[]{%
+        Breakdown of a step in a \gls{WCA} into its timing components.
+        \ensuremath{t_k | k \in \{1, \ldots, n \}} correspond to sampling instants.
+        \ensuremath{t_0} indicates the instant at which the instruction for step \ensuremath{M} is provided to the user and the first sample for said step is captured.
+        The instruction for step \ensuremath{M + 1} is provided at \ensuremath{t_{n+1}}.
+        Finally, \ensuremath{t_c} marks the instant at which the user finishes the instruction, and \ensuremath{t_n} the instant of capture of the first sample after the instruction has been completed;
+        this sample therefore also corresponds to the final sample of step \ensuremath{M}.
+        Figure originally included in \cref{paper:olguinmunoz2023realistic}.
+    }\label{fig:wcastep}
+\end{figure}
+
+In the following, please refer to \cref{fig:wcastep}.
+\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
+\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
+this sample is the first sample to capture the finished state of the step.
+At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
+We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
+
+\glsreset{TTF}%
+From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
+However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
+The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
+A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
+If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
+\begin{inlineenum}
+    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
+    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
+    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
+\end{inlineenum}
+These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
+We will refer to them extensively in \cref{chap:contributions}.
+
+%\subsection{Mechanisms relating human behavior to system responsiveness}
+%
+%The question of how people respond to delay in a computer system is grounded in how people perceive time.
+%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
+%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
+%The estimation of an event's duration varies with context of various types
+%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
+%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
+%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
+%\item arousal can expand durations~\cite{droit_volet2011emotion}.
+%\end{inlineenum}
+%
+%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
+%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
+%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
+%Violations of expectancy have a strong impact on the acceptability of computer systems.
+%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
+%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
+%
+%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
+%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
+%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
+%
+%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
+%They then proposed a latency tolerance zone according to the task demands.
+%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
+%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
+%
+%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
+%These mechanisms could be cognitive or emotional in origin.
+%
+%A first possible explanation comes from research on cognitive and motor planning.
+%Delay may move users from relatively automatic to more attention-demanding processing.
+%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
+%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
+%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
+%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
+%
+%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
+%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
+%
+%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
+%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
+%
+%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
+%
+%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
+%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
+%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
+%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
+%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
+%
+%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
+%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
+%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
+%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
+%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
+Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
+%These and other measures of individual variation were considered here.
 
@@
             Fits were achieved through \gls{MLE}.
+            Figure originally published in \cref{paper:olguinmunoz2023realistic}
         }\label{fig:ttfvsexgaussianmu}

### 2023-03-05 body/part1/chapter4-conclusions/conclusions.tex
restructuring

--- 2023-02-16 body/part1/futurework.tex
@@
-\section{Conclusions}
-
 In this thesis, we have made three significant contributions to the field of edge computing, addressing the challenge of studying latency-sensitive applications in a scalable, repeatable manner.
@@
 We believe that the methodology and human user model presented in this thesis provide a valuable tool for researchers and practitioners in this field, and we hope that our contributions will inspire further research in this area.
-
-\subsection{Open questions \& future work}
-
-Notwithstanding the important contributions of this dissertation, a number of key questions remain open for future research in this field.
-
-The first concern is the applicability of the methodology to other systems and applications in the context of Edge Computing.
-This thesis explores the methodology's use for step-based \gls{WCA} in depth, and also performs a less thorough case study on its benefits for \gls{NCS} research.
-The question remains if the methodology is suitable for
-\begin{inlineenum}
-    \item other, non-\gls{WCA} \gls{XR} applications, such as more general mobile \gls{AR} and \gls{VR}
-    \item other more general categories of \glspl{CPS}
-\end{inlineenum}.
-The results obtained thus far indicate that the methodology is applicable to a wide range of interactive and \gls{CPS} applications in Edge Computing, but these results should be confirmed by domain-specific studies.
-
-In accordance with the broader questions mentioned above, this thesis also raises specific questions related to its individual contributions to the literature.
-The EdgeDroid tool, introduced in papers \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2023realistic}, has the potential for further development to support other forms of \gls{WCA} applications.
-Currently, the tool is only designed for a specific class of step-based \gls{WCA} with a linear task model.
-However, the tool, and by extension the methodology, could be easily adapted to similar applications.
-
-We identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
-The current dataset is limited to young undergraduate students from a highly selective university in the \gls{USA}.
-Adding participants from different demographic groups would enhance the representativeness of the data and increase the generalizability of the findings.
-This would also provide an opportunity to extend the human timing model presented in \cref{paper:olguinmunoz2023realistic} to better reflect the behavior of a broader range of users.
-
-The optimization framework described in \cref{paper:olguinmunoz2023realistic} can be enhanced in two ways.
-Firstly, the framework assumes that human execution times follow a Rayleigh distribution, but in reality, they are often modeled with an \gls{exGaussian} distribution.
-Secondly, the framework can be made more practical by incorporating the optimization of scenarios with multiple clients and a distributed solution.
-
-There are numerous avenues for further investigation in the optimization and development of \gls{WCA}.
-While this thesis has explored the optimization of sampling and energy consumption, there are several other dimensions of these systems that can be examined.
-These include resource allocation, task scheduling, user identification and classification, as well as guaranteeing both \gls{QoE} and \gls{QoS}.
-Each of these areas presents opportunities for future work and could lead to significant advancements in the field of \gls{WCA}.
-
-\medskip
-Our research on applying the methodology to \glspl{NCS} presents promising potential for future work.
-While \cref{paper:olguinmunoz2022cleave} provides an initial exploration, further investigation through more realistic and thorough experimentation is necessary to validate the results obtained by the framework.
-To this end, we aim to conduct larger-scale experiments that utilize cutting-edge networking technologies, including 5G and future advancements such as 6G.
-By exploring the full range of networking technologies and scenarios, we can ensure that our methodology is robust and applicable to a wide range of \gls{NCS}.
-
-We also intend to extend the functionality of the \gls{CLEAVE} tool significantly, primarily to target a wider variety of \glspl{NCS} deployments.
-Currently, \gls{CLEAVE} includes a small repertoire of emulations and controllers.
-We are planning to extend this with the goal of creating an open library of \glspl{NCS} to share with the community.
-
-\medskip
-Finally, we see tremendous opportunities for testbed-driven research in Edge Computing.
-Our current efforts in this context aim towards the extension of our Ainur framework and its integration with \gls{CHI}~\cite{keahey2020lessons}.
-We believe Ainur's unique focus on fully-automated end-to-end wireless edge computing experimentation makes it valuable addition to the \gls{CHI} stack.
-At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainurs potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.
-
-The \gls{EXPECA} testbed is undergoing a major expansion and improvement.
-%The aim is to enhance its capabilities and hardware to meet the ever-increasing demands of the industry.
-The expansion consists of three main components;
-\begin{inlineenum}
-    \item a hardware upgrade
-    \item a small scale, private 5G deployment
-    \item a campus-wide 5G deployment
-\end{inlineenum}.
-Firstly, the \gls{EXPECA}s hardware is receiving a complete overhaul.
-We are replacing our initial Raspberry Pi-based prototype with a number of industry grade servers.
-Secondly, a small-scale private 5G deployment using industrial-grade equipment is currently being deployed.
-This will provide a controlled environment for testing and evaluating 5G technologies and applications.
-The use of industrial-grade equipment ensures that the results obtained from the tests are reliable and can be directly applied to real-world scenarios.
-Thirdly, the testbed will be extended to a campus-wide deployment at \gls{KTH} in Stockholm.
-This will provide an opportunity for a larger-scale deployment of 5G and \gls{MEC} technologies and applications, allowing for testing in real-world settings.
-The campus-wide deployment will also serve as a platform for researchers and industry professionals to collaborate and exchange ideas.
-The combined efforts of the small-scale private deployment and the campus-wide deployment at \gls{KTH} will significantly enhance the capabilities and knowledge of the \gls{EXPECA} testbed, positioning it as a leading platform for 5G and \gls{MEC} research and innovation.
-
-In conclusion, the present dissertation provides a comprehensive examination of the methodology for \glspl{NCS} and \gls{WCA}, but there are still many open questions that need to be addressed in future research.
-The EdgeDroid tool and the dataset have the potential to be extended, and the optimization framework employed in the thesis could be improved.
-We also plan to extend the functionality of the \gls{CLEAVE} tool and perform large-scale experimentation on novel networking technologies.
-There are also significant opportunities for testbed-driven research in Edge Computing through the \gls{EXPECA} testbed.

### 2023-03-06 body/part1/chapter1-introduction/background.tex
todonotes
reorg
xr

@@
     \centering
-    \includegraphics[height=30em]{Figs/edgecomputing}
+    \includegraphics[height=30em]{figures/edgecomputing}
     \caption{%
@@
 
-\subsubsection{Experimental research in Edge Computing}
-\todo[inline]{Should be moved to related work or removed.}
+\subsection{\glsfmtlong{XR}}\label{background:xr}
+\glsreset{XR}
+\glsreset{AR}
+\glsreset{VR}
+\glsreset{MR}
+\todo[inline]{Add some references}
 
-The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
-Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
-Of particular interest to us are the
-\begin{inlineenum}
-    \item \acs{GENI}
-    \item Chameleon
-    \item \acs{COSMOS}
-    \item \acs{POWDER}
-    \item \acs{ARA}
-    \item EdgeNet
-%    \item Drexel Grid \gls{SDR}
-\end{inlineenum} testbeds.
+\gls{XR} is an umbrella term used to refer to all immersive technologies that combine real and virtual environments, including technologies such as \gls{AR}, \gls{VR}, and \gls{MR}.
+\gls{AR} merges digital content with the physical world, enhancing human perception and cognition through the overlay of virtual objects and information on top of the real environment.
+This technology can superimpose tooltips, virtual \gls{2D} and \gls{3D} objects, or even full-fledged videos on top of what the user sees.
+These overlays are most often presented to the user through wearable or mobile devices, such as smartphones or smart glasses (e.g.
+Google Glass), fitted with a camera to capture a real-time video feed of the physical environment and a display on which the augmented feed is presented.
 
-\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
-Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
+\gls{VR}, on the other hand, takes it a step further and involves fully forgoing the real world, immersing the user in a fully digital, \emph{simulated} environment.
+Due to its deeply immersive nature, \gls{VR} requires hardware components specifically designed for this task, the most crucial of which is the \gls{VR} headset.
+The simplest type of such headset uses a pair of lenses to present an illusion of depth when looking through them at a small display.
+More advanced types combine microdisplays, accelerometers, and headphones to provide a comprehensively immersive experience, where the user truly feels as if they are inside the virtual world.
+Additionally, \gls{VR} systems often allow users to interact with the simulated environment through the use of instrumented gloves or special controllers, further increasing the illusion of immersion.
 
-The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
-Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
-
-\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
-The testbeds main \emph{raison d'etre} is \gls{MEC} research, with a strong emphasis on wireless networks.
-The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
-
-\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
-It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
-These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
-
-The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
-\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-
-EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
-It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
-The EdgeNet software is available for anyone to download;
-individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
-The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
-A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
-
-Finally, a number of other, small-scale approaches also exist.
-In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
-The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
-This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
-In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
-Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
-
-\subsection{\glsfmtlong{XR}}\label{background:xr}
-\todo[inline]{Rework into background XR}
-
-\glsreset{WCA}
-\gls{WCA} refers to a novel category of wearable and mobile immersive applications which aim to amplify human cognition in both day-to-day tasks and professional settings.
-\glspl{WCA} work in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
-These applications operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
-Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
-The input is parsed to an internal symbolic representation of the state of the task;
-whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
-This feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
-If no change in state is detected, the application remains silent and out-of-the-way.
-In other words, the application silently discards samples which capture an intermediate or unfinished state, or simply are too noisy to be processed.
-\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{Figs/wca_state}
-    \caption{%
-        Mode of operation of \acs{WCA} applications.
-        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
-        Samples that do not result in feedback are silently discarded.
-    }\label{fig:wca}
-\end{figure}
-
-As the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
-they are pervasive and mobile~\cite{ha2014towards}.
-This leads to interesting consequences for \gls{WCA} system design when combined with their requirement of seamless integration with the context of the user.
-The wearable nature of \gls{WCA} directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
-The requirements of immersive and seamless interaction, on the other hand, suggest a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-This kind of compute- and energy-intensive processing, often involving \gls{ML} algorithms and \glspl{DNN}, is unfeasible to be performed on wearable devices, and thus these applications \emph{must} necessarily make use of compute offloading~\cite{ha2014towards,wang2020scaling}.
-
-However, it is understood today the cloud represents an unsuitable candidate for offloading computation in these applications, given the high latencies involved.
-Given their immersive nature, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand), as users will have expectations of constant and immediate feedback as they interact with the system.
-As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
-Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).
-There is therefore a broad understanding today that edge computing is a key enabling technology for \gls{WCA} applications~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.
+Finally, \gls{MR} combines elements from both aforementioned technologies.
+As in \gls{AR}, \gls{MR} overlays virtual objects on the physical, real-world environment, while also involving the tools and techniques from \gls{VR} which allow the user to manipulate these.
+In this way, \gls{MR} creates a hybrid environment where users can interact with both virtual and real-world elements simultaneously, resulting in an illusion of natural interaction between users and digital elements.
 
 \medskip
-In literature, research on \gls{WCA} has focused mostly on the implementation of prototype applications and platforms.
-\citeauthor{ha2014towards}~\cite{ha2014towards} coined the term \acl{WCA} in\ \citeyear{ha2014towards}.
-Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
-The authors were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
-More recently, the scope of \glspl{WCA} has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards}.
-\cite{chen2015early} and~\cite{chen2018application} further developed the platform described in~\cite{ha2014towards}, and produced a number of different assistive applications such as the aforementioned LEGO assistant, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis, and more advanced assembly assistant for IKEA furniture.
-Yet another platform for \gls{WCA} was introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
-Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.
-Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
-It is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
 
-In later years, a body of work studying the optimization of these applications has emerged.
-A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
-\citeauthor{wang2019towards}~\cite{wang2019towards} explore strategies for scalable \gls{WCA} on the edge through different adaptive sampling and resource allocation schemes;
-this is further developed in the authors' PhD thesis later on~\cite{wang2020scaling}.
-\cite{moothedath2021energy},\ \cite{moothedath2022energy1}, and~\cite{moothedath2022energy2} studied the potential for energy optimization in \gls{WCA} through convex optimization of the sampling strategies employed.
-In~\cite{george2020openrtist}, the authors introduce \emph{OpenRTiST}, a benchmarking tool for immersive, latency-sensitive, and bandwidth-hungry applications on edge computing leveraging \glspl{DNN} to perform real-time style transfer on a live video feed.
+Due to their ability to create immersive experiences for users and closely mimic and simulate real-life scenarios, \gls{XR} systems have found applications across a wide spectrum of industries and fields.
+One of the most common uses of these technologies can be found in the domain of entertainment.
+\gls{VR} allows users to become fully immersed in virtual experiences, including games, concerts, movies, making them feel almost physically present without having to leave the comfort of their homes.
+Further, \gls{AR} and \gls{MR} allow for the merging of the virtual and real worlds, bringing entertainment to life in users day-to-day surroundings.
+The popularity of this approach can be clearly seen in the explosive success of modern mobile \gls{AR} games such \citetitle{pokemongo}~\cite{pokemongo}.
 
-The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
-There's more to end-to-end latency, \gls{WCA} performance, and scalability than merely the question of where the compute backend is placed, however.
-The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
-These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
-As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.
-On the other hand, recently published models for end-to-end latency of edge computing architectures, are quite complex, while not accounting for the specifics of \gls{WCA} and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite}.
-Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
-Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.
+\gls{XR} has also been leveraged for education and training.
+\gls{VR}
 
@@
 
-%\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-%Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
-%This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
-
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
-
-%One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
-%%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
-%On the other, there exist several approaches when considering experimental methodologies.
-%%A number of works concerning \glspl{NCS} deal with experimental studies.
-%%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
-%One such approach uses setups in which the complete system is built on top of real hardware.
-%This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-%Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
-%The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-%These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
-%Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-%\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
-%    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-%    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-%\end{inlineenum}
-%
-%As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
-%This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-%Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
-%This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
-%
-%Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
-%In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
-%Their methodology utilizes joint knowledge of control, computation, and communication.
-%In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
-%Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
-%This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
-%Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
+\subsection{Emulation as a research methodology}
+\todo[inline]{Rename?}

### 2023-03-06 recyclebin/experimental_research_edge_computing.tex
xr

--- 2023-02-07 body/part1/background.tex
@@
-\section{Background \& related work}
-\glsresetall%
-\todo[inline]{%
-    \textcite{satyanarayanan2009case}
-
-    Other papers detailing the advantages of Edge computing.
-
-    Next, what applications benefit from the Edge?
-    \glsfmtshort{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
-    Main points: prototypes and characterization of system latencies and load.
-}
-
-\subsection{Edge computing}
-
-Edge computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
-Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
-These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
-Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters~\citationeeded.
-These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
-
-Through this design, cloud computing affords significant advantages to users.
-As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
-Thanks to economies of scale and virtualization technologies, the cloud is highly scalable;
-services can be scaled simply by spawning more \glspl{VM}.
-The specialized nature of cloud providers, the scale of modern datacenters, and the use of virtualization also make the cloud highly reliable.
-When hardware fails, recovering is simply a matter of migrating the service container or \gls{VM} to an available compute node~\cite{endo2016high}.
-
-However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive and/or bandwidth-intensive applications.
-In order to achieve the necessary economies of scale, cloud datacenters are designed to serve users distributed across vast geographical areas.
-These installations are thus often located ``far'' from potential users;
-for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
-This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{tolia2006quantifying,lagar2007interactive,satyanarayanan2009case,varghese2016challenges,shi2016promise}.
-The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
-Such latencies are unfeasible to consistently achieve with cloud computing~\cite{dang2021cloudy}.
-On the other hand, as smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face increasing challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
-
-\begin{figure}
-    \centering
-    \includegraphics[height=30em]{Figs/edgecomputing}
-    \caption{%
-        Conceptual design of edge computing.
-        Micro-clouds (known as ``cloudlets''~\cite{satyanarayanan2009case}) are placed a at the edge of the network, a few hops away from end users.
-        In other words, these cloudlets are located between users and the internet and the cloud.
-    }\label{fig:edgecomputing}
-\end{figure}
-
-\medskip
-Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
-The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
-To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
-\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
-This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
-Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
-This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.
-
-The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
-Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Another advantage is bandwidth reduction.
-%By processing and aggregating data closer to where its needed, edge computing can significantly reduce the amount of traffic going to cloud datacenters.
-A variant of edge computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for this purpose.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
-Edge computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
-
-Finally, in later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
-This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
-Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
-These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
-We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
-
 \subsubsection{Experimental research in Edge Computing}
+\todo[inline]{Should be moved to related work or removed.}
 
@@
 The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
-
-\subsection{\glsfmtlong{WCA}}\label{background:wca}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{Figs/wca_state}
-    \caption{%
-        Mode of operation of \acs{WCA} applications.
-        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
-        Samples that do not result in feedback are silently discarded.
-    }\label{fig:wca}
-\end{figure}
-
-\glsreset{WCA}
-\gls{WCA} refers to a novel category of wearable and mobile immersive applications which aim to amplify human cognition in both day-to-day tasks and professional settings.
-\glspl{WCA} work in a mode analogous to how \gls{GPS} navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand and staying out of the way of the user when guidance is not required.
-These applications operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.
-Through the use of sensors (most commonly video inputs), the application follows the progress of the task in realtime by continuously sampling the state of the physical system.
-The input is parsed to an internal symbolic representation of the state of the task;
-whenever a change of state is detected the application provides appropriate feedback or instructions to the user.
-This feedback can be in the form of text, images, video, and/or audio, and guides the user towards the next desired system state.
-If no change in state is detected, the application remains silent and out-of-the-way.
-In other words, the application silently discards samples which capture an intermediate or unfinished state, or simply are too noisy to be processed.
-\Cref{fig:wca} illustrates this mode of operation in a \gls{WCA} guiding a user to manipulate a structure composed of LEGO bricks.
-
-As the name implies, \glspl{WCA} are available whenever the user requires them, without being tethered to a particular physical location;
-they are pervasive and mobile~\cite{ha2014towards}.
-This leads to interesting consequences for \gls{WCA} system design when combined with their requirement of seamless integration with the context of the user.
-The wearable nature of \gls{WCA} directly implies use of lightweight and battery-powered devices, greatly constrained in terms of energy and computing resources.
-The requirements of immersive and seamless interaction, on the other hand, suggest a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
-This kind of compute- and energy-intensive processing, often involving \gls{ML} algorithms and \glspl{DNN}, is unfeasible to be performed on wearable devices, and thus these applications \emph{must} necessarily make use of compute offloading~\cite{ha2014towards,wang2020scaling}.
-
-However, it is understood today the cloud represents an unsuitable candidate for offloading computation in these applications, given the high latencies involved.
-Given their immersive nature, feedback in \gls{WCA} systems should be provided ``quickly'' (relative to the task at hand), as users will have expectations of constant and immediate feedback as they interact with the system.
-As is the case for most \gls{XR} applications, delayed feedback in \gls{WCA} can have severe negative consequences for the quality of the user experience.
-Depending on the task, these consequences can range from simply distracting or annoying the user (in the case of less interactive tasks such as step-by-step assembly), to actively handicapping user performance (in the case of highly interactive tasks).
-There is therefore a broad understanding today that edge computing is a key enabling technology for \gls{WCA} applications~\cite{ha2014towards,wang2020scaling,chen2018application,olguinmunoz2021impact}.
-
-\medskip
-In literature, research on \gls{WCA} has focused mostly on the implementation of prototype applications and platforms.
-\citeauthor{ha2014towards}~\cite{ha2014towards} coined the term \acl{WCA} in\ \citeyear{ha2014towards}.
-Their work described a first prototype system employing a Google Glass wearable device, and identified the challenges of offloading the computation for such an application to the cloud.
-The authors were originally inspired by assistive use cases for people suffering from some form of cognitive decline due to aging, illness, or traumatic brain injury~\cite{ha2014towards,satyanarayanan2019augmenting}.
-More recently, the scope of \glspl{WCA} has been expanded to include a broader range of use cases, including complex assembly tasks~\cite{chen2017empirical,chen2018application,wang2020scaling,wang2019towards}.
-\cite{chen2015early} and~\cite{chen2018application} further developed the platform described in~\cite{ha2014towards}, and produced a number of different assistive applications such as the aforementioned LEGO assistant, and others such as a Ping-Pong assistant capable of providing real-time hints to players of table tennis, and more advanced assembly assistant for IKEA furniture.
-Yet another platform for \gls{WCA} was introduced in~\cite{chatzopoulos2017hyperion}, with the goal of providing real-time contextual information in text-form to users.
-Furthermore, important work has been done in the implementation of such systems for industrial and factory applications~\cite{belletier2021wearable}.
-Non-wearable cognitive assistance systems for assembly tasks have already been proven to be valuable tools in the industrial workplace~\cite{funk2015cognitive,gorecky2011cognito}.
-It is understood that the detethering of these system for their use in wearable scenarios opens up a multitude of possibilities.
-
-In later years, a body of work studying the optimization of these applications has emerged.
-A seminal work in this field is~\cite{chen2017empirical}, in which the authors establish hard and soft latency bounds for acceptable performance in \gls{WCA}.
-\citeauthor{wang2019towards}~\cite{wang2019towards} explore strategies for scalable \gls{WCA} on the edge through different adaptive sampling and resource allocation schemes;
-this is further developed in the authors' PhD thesis later on~\cite{wang2020scaling}.
-\cite{moothedath2021energy},\ \cite{moothedath2022energy1}, and~\cite{moothedath2022energy2} studied the potential for energy optimization in \gls{WCA} through convex optimization of the sampling strategies employed.
-In~\cite{george2020openrtist}, the authors introduce \emph{OpenRTiST}, a benchmarking tool for immersive, latency-sensitive, and bandwidth-hungry applications on edge computing leveraging \glspl{DNN} to perform real-time style transfer on a live video feed.
-
-The present thesis aims to contribute to this body of work by presenting and discussing a viable methodology for the design and evaluation of these systems.
-There's more to end-to-end latency, \gls{WCA} performance, and scalability than merely the question of where the compute backend is placed, however.
-The design of any complex application such as \gls{WCA} involves a multitude of decisions with the potential to influence the system responsiveness as experienced by the user.
-These decisions include those on the implementation side (compression standards, algorithms, protocols, etc.), as well as on the infrastructure side (physical network layer, traffic prioritization, etc.).
-As expressed above, existing studies of this class of applications have only recently started to delve more deeply upon these issues.
-On the other hand, recently published models for end-to-end latency of edge computing architectures, are quite complex, while not accounting for the specifics of \gls{WCA} and in many cases limited to very constrained scenarios and analyses~\cite{al_zubaidy2015performance,schiessl2017finite}.
-Our work in~\cite{olguinmunoz2018demoscaling,olguinmunoz2019edgedroid} introduces the first ever model for realistic emulation of these applications on real hardware.
-Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.
-
-\subsubsection{Characterization of \acs{WCA}}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
-    \caption[]{%
-        Breakdown of a step in a \gls{WCA} into its timing components.
-        \ensuremath{t_k | k \in \{1, \ldots, n \}} correspond to sampling instants.
-        \ensuremath{t_0} indicates the instant at which the instruction for step \ensuremath{M} is provided to the user and the first sample for said step is captured.
-        The instruction for step \ensuremath{M + 1} is provided at \ensuremath{t_{n+1}}.
-        Finally, \ensuremath{t_c} marks the instant at which the user finishes the instruction, and \ensuremath{t_n} the instant of capture of the first sample after the instruction has been completed;
-        this sample therefore also corresponds to the final sample of step \ensuremath{M}.
-        Figure originally included in \cref{paper:olguinmunoz2023realistic}.
-    }\label{fig:wcastep}
-\end{figure}
-
-An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
-We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
-In this section, we will provide some definitions relating to the operation of these applications.
-
-As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
-In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
-We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
-A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
-the time interval between these two events will be referred to as the \emph{step duration}.
-
-In the following, please refer to \cref{fig:wcastep}.
-\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
-\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
-this sample is the first sample to capture the finished state of the step.
-At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
-We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
-
-\glsreset{TTF}%
-From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
-However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
-The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
-A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
-If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
-\begin{inlineenum}
-    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
-    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
-    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
-\end{inlineenum}
-These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
-We will refer to them extensively in \cref{chap:contributions}.
-
-%\subsection{Mechanisms relating human behavior to system responsiveness}
-%
-%The question of how people respond to delay in a computer system is grounded in how people perceive time.
-%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-%The estimation of an event's duration varies with context of various types
-%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-%\item arousal can expand durations~\cite{droit_volet2011emotion}.
-%\end{inlineenum}
-%
-%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
-%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-%Violations of expectancy have a strong impact on the acceptability of computer systems.
-%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
-%
-%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
-%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
-%
-%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-%They then proposed a latency tolerance zone according to the task demands.
-%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
-%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
-%
-%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
-%These mechanisms could be cognitive or emotional in origin.
-%
-%A first possible explanation comes from research on cognitive and motor planning.
-%Delay may move users from relatively automatic to more attention-demanding processing.
-%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-%
-%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-%
-%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-%
-%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-%
-%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-%
-%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
-%These and other measures of individual variation were considered here.
-
-\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
-\todo[inline]{Needs citations.}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/control_system}
-        \caption{%
-            Traditional control system.
-        }
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[t]{0.50\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/networked_control_system}
-        \caption{%
-            \acl{NCS}.
-        }
-    \end{subfigure}
-    \caption{%
-        Comparison between traditional control systems and \aclp{NCS}.
-        In an \aclp{NCS}, controller and plant are physically separated and both actuation commands and sensor inputs are shared through a general purpose \acs{TCP}/\acs{IP} network.
-    }\label{fig:csvsncs}
-\end{figure}
-
-\glsreset{NCS}
-\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
-A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
-Control systems are composed of
-\begin{inlineenum}
-    \item a \emph{plant}
-    \item \emph{sensors}
-    \item \emph{actuators}
-    \item a \emph{controller}
-\end{inlineenum}.
-Plant refers to the physical system being controlled.
-Sensors sample and encode the state of the plant and transmit it to the controller.
-The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
-The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
-
-\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
-see \cref{fig:csvsncs}.
-This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
-\glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
-This leads to enhanced scalability and flexibility, and allows improved coordination and collaboration among systems.
-Furthermore, \glspl{NCS} provide greater access to information, as they enable the exchange of data and information between control components and systems, leading to more informed decision-making and improved performance.
-
-\glspl{NCS} have become increasingly popular in recent years due to advances in communication and control technologies, as well as the increasing demand for real-time, remote, and distributed control.
-They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the cloud for centralized, distributed control in what has been called \emph{cloud control systems}~\cite{xia2015cloud}
-However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms simply cannot meet~\cite{wan2020efficient}.
-This has led their adoption to be limited mostly to industrial environments.
-This is about to change, however, as with the advent of edge computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
-These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.
-
-\medskip
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-
-In terms of research relating to the benchmarking and optimization of these systems, works in literature have for the most part leveraged either theoretical models~\cite{zhang2019networked} or simulations (for example ~\cite{ma2019optimal}).
-However, these approaches can only capture network behaviors at a coarse level, and thus a body of works has emerge in later years involving experimental approaches to \gls{NCS} benchmarking.
-\cite{baumann2018evaluating} and~\cite{cuenca2019periodic} employ purpose-built physical testbeds for the study of these systems.
-\cite{zoppi2020ncsbench} introduces \emph{NCSBench}, a platform for reproducible benchmarking of \gls{NCS} using a LEGO Mindstorms platform.
-
-Finally, hybrid approaches, mixing testbed or prototype implementations and simulations or emulations of \gls{NCS} components also exist in literature.
-These approaches tend to focus on a particular element of the system which is implemented using a prototype or testbed, and virtualize the rest of the system.
-For instance, in~\cite{wang2020inverter} the authors are concerned with realistic effects from the network, and thus they have an emulated \gls{NCS} interact with a real network.
-On the other hand, works such as~\cite{natale2004inverted} focus on the system being controlled and opt to instead emulate the network.
-
-%\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-%Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
-%This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
-
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
-
-%One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
-%%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
-%On the other, there exist several approaches when considering experimental methodologies.
-%%A number of works concerning \glspl{NCS} deal with experimental studies.
-%%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
-%One such approach uses setups in which the complete system is built on top of real hardware.
-%This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-%Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
-%The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-%These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
-%Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-%\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
-%    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-%    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-%\end{inlineenum}
-%
-%As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
-%This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-%Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
-%This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
-%
-%Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
-%In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
-%Their methodology utilizes joint knowledge of control, computation, and communication.
-%In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
-%Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
-%This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
-%Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.

### 2023-03-06 recyclebin/ncs_background.tex
xr

--- 2022-12-19 publications/2022CLEAVE/intro.tex
@@
-\section{Introduction}\label{sec:intro}
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=.8\textwidth]{publications/2022CLEAVE/images/CLEAVE_NCS_structure}
-    \caption{
-        Structure of an emulated \glspl{NCS} in \gls{CLEAVE}.
-    }\label{fig:cleave:ncs:struct}
-\end{figure*}
-
-The number and applications of \glspl{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
-However, this rapid increase in adoption has mostly been limited to industrial contexts.
-Although \glspl{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
-This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \glspl{CPS} will be made possible.
-These technologies meet two key requirements of \glspl{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \gls{CPS} in the future.
-
 \glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
@@
 
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
+Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
 
@@
 Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
-
-We overcome this issue by proposing a completely virtual plant allowing for unparalleled flexibility in changing the plant model and characteristic features of the experiments.
-In this paper, we present the first fully-software-based framework for scalable and repeatable benchmarking of edge-native \gls{NCS}.
-As edge computing begins being adopted by industry, more and more variations have begun to appear in literature.
-``Near'', ``far'', ``core'', and ``telco'' edge describe variations of the original concept and are becoming ubiquitous in new research.
-While the core idea of edge computing is widely accepted as fundamental for pervasive \glspl{NCS} in general, understanding the strengths and weaknesses of such different edge concepts is of paramount importance.
-
-Our framework, \gls{CLEAVE}, aims to simplify the repeatable and scalable benchmarking of such systems.
-It is fully virtualized, inspired by our previous work on benchmarking human-in-the-loop applications on the edge~\cite{Olguin2019EdgeDroid}.
-The tool consists of a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
-These virtual \glspl{NCS} can then be deployed on real networks for reparametrizable, repeatable, and reproducible benchmarking of the complete system.
-
-\gls{CLEAVE} is built using \emph{Python 3.8}, making it highly extensible and able to harness the multitude of already existing user libraries.
-It is furthermore compatible with container technologies such as \emph{Docker}\footnote{Docker Engine: \url{https://www.docker.com/}}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
-
-The rest of this paper is structured as follows.
-\cref{sec:approach} presents the design principles and implementation of the framework.
-In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \gls{CLEAVE}.
-Finally, in \cref{sec:conclusion} we conclude this work.

### 2023-03-06 recyclebin/wca_background.tex
xr

--- 2023-02-06 body/part1/background.tex
@@
-\section{Background \& related work}
-\glsresetall%
-\todo[inline]{%
-    \textcite{satyanarayanan2009case}
-
-    Other papers detailing the advantages of Edge computing.
-
-    Next, what applications benefit from the Edge?
-    \glsfmtshort{WCA}! Mention both the main body of work (by Satya et.\ al.) but also other authors out there.
-    Main points: prototypes and characterization of system latencies and load.
-}
-
-\subsection{Edge computing}
-
-Edge computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
-Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
-These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
-Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters~\citationeeded.
-These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
-
-Through this design, cloud computing affords significant advantages to users.
-As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
-Thanks to economies of scale and virtualization technologies, the cloud is highly scalable;
-services can be scaled simply by spawning more \glspl{VM}.
-The specialized nature of cloud providers, the scale of modern datacenters, and the use of virtualization also make the cloud highly reliable.
-When hardware fails, recovering is simply a matter of migrating the service container or \gls{VM} to an available compute node~\cite{endo2016high}.
-
-However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive and/or bandwidth-intensive applications.
-In order to achieve the necessary economies of scale, cloud datacenters are designed to serve users distributed across vast geographical areas.
-These installations are thus often located ``far'' from potential users;
-for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
-This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{tolia2006quantifying,lagar2007interactive,satyanarayanan2009case,varghese2016challenges,shi2016promise}.
-The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
-Such latencies are unfeasible to consistently achieve with cloud computing~\cite{dang2021cloudy}.
-On the other hand, as smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face increasing challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
-
-\begin{figure}
-    \centering
-    \includegraphics[height=30em]{Figs/edgecomputing}
-    \caption{%
-        Conceptual design of edge computing.
-        Micro-clouds (known as ``cloudlets''~\cite{satyanarayanan2009case}) are placed a at the edge of the network, a few hops away from end users.
-        In other words, these cloudlets are located between users and the internet and the cloud.
-    }\label{fig:edgecomputing}
-\end{figure}
-
-\medskip
-Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
-The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
-To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
-\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
-This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
-Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
-This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.
-
-The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
-Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Another advantage is bandwidth reduction.
-%By processing and aggregating data closer to where its needed, edge computing can significantly reduce the amount of traffic going to cloud datacenters.
-A variant of edge computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for this purpose.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
-Edge computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
-
-Finally, in later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
-This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
-Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
-These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
-We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
-
-\subsection{\glsfmtlong{WCA}}\label{background:wca}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{Figs/wca_state}
-    \caption{%
-        Mode of operation of \acs{WCA} applications.
-        The assistant continuously samples the state of the monitored task, and provides appropriate feedback once a state change has been detected.
-        Samples that do not result in feedback are silently discarded.
-    }\label{fig:wca}
-\end{figure}
-
 \glsreset{WCA}
@@
 Our characterization of human timings in these systems in~\cite{olguinmunoz2021impact}, and our subsequent design of a realistic model for human behavior in \gls{WCA} constitute novel insights and tools for the optimization of these systems.
-
-\subsubsection{Characterization of \acs{WCA}}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/figs/step_time}
-    \caption[]{%
-        Breakdown of a step in a \gls{WCA} into its timing components.
-        \ensuremath{t_k | k \in \{1, \ldots, n \}} correspond to sampling instants.
-        \ensuremath{t_0} indicates the instant at which the instruction for step \ensuremath{M} is provided to the user and the first sample for said step is captured.
-        The instruction for step \ensuremath{M + 1} is provided at \ensuremath{t_{n+1}}.
-        Finally, \ensuremath{t_c} marks the instant at which the user finishes the instruction, and \ensuremath{t_n} the instant of capture of the first sample after the instruction has been completed;
-        this sample therefore also corresponds to the final sample of step \ensuremath{M}.
-        Figure originally included in \cref{paper:olguinmunoz2023realistic}.
-    }\label{fig:wcastep}
-\end{figure}
-
-An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
-We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
-In this section, we will provide some definitions relating to the operation of these applications.
-
-As discussed above, \glspl{WCA} track the state of the real world through sampling of inputs, most commonly of video feeds.
-In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
-We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
-A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
-the time interval between these two events will be referred to as the \emph{step duration}.
-
-In the following, please refer to \cref{fig:wcastep}.
-\ensuremath{\{ t_0, t_1, \ldots, t_{n + 1} \}} corresponds to a series of discrete and sequential sampling instants, such that \ensuremath{t_0} corresponds to the instant at which the instruction for step \ensuremath{M} is provided and the first sample of the step is taken.
-\ensuremath{t_n} corresponds to the instant at which the final sample of step \ensuremath{M} is taken;
-this sample is the first sample to capture the finished state of the step.
-At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
-We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
-
-\glsreset{TTF}%
-From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
-However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
-The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
-A corollary of this is that these are therefore also the \emph{only} points in time at which users can notice changes in system responsiveness.
-If we understand \ensuremath{t_c} as the point in time at which the user finishes executing the instruction for step \ensuremath{M}, then
-\begin{inlineenum}
-    \item \ensuremath{t_c - t_0} corresponds to the \emph{step execution time}, that is, the time it has taken the user to complete the instruction
-    \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
-    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
-\end{inlineenum}
-These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
-We will refer to them extensively in \cref{chap:contributions}.
-
-%\subsection{Mechanisms relating human behavior to system responsiveness}
-%
-%The question of how people respond to delay in a computer system is grounded in how people perceive time.
-%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-%The estimation of an event's duration varies with context of various types
-%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-%\item arousal can expand durations~\cite{droit_volet2011emotion}.
-%\end{inlineenum}
-%
-%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
-%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-%Violations of expectancy have a strong impact on the acceptability of computer systems.
-%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
-%
-%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
-%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
-%
-%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-%They then proposed a latency tolerance zone according to the task demands.
-%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
-%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
-%
-%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
-%These mechanisms could be cognitive or emotional in origin.
-%
-%A first possible explanation comes from research on cognitive and motor planning.
-%Delay may move users from relatively automatic to more attention-demanding processing.
-%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-%
-%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-%
-%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-%
-%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-%
-%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-%
-%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
-%These and other measures of individual variation were considered here.
-
-\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
-\todo[inline]{Needs citations.}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[t]{0.45\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/control_system}
-        \caption{%
-            Traditional control system.
-        }
-    \end{subfigure}%
-    \hfill%
-    \begin{subfigure}[t]{0.50\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{Figs/networked_control_system}
-        \caption{%
-            \acl{NCS}.
-        }
-    \end{subfigure}
-    \caption{%
-        Comparison between traditional control systems and \aclp{NCS}.
-        In an \aclp{NCS}, controller and plant are physically separated and both actuation commands and sensor inputs are shared through a general purpose \acs{TCP}/\acs{IP} network.
-    }\label{fig:csvsncs}
-\end{figure}
-
-\glsreset{NCS}
-\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
-A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
-Control systems are composed of
-\begin{inlineenum}
-    \item a \emph{plant}
-    \item \emph{sensors}
-    \item \emph{actuators}
-    \item a \emph{controller}
-\end{inlineenum}.
-Plant refers to the physical system being controlled.
-Sensors sample and encode the state of the plant and transmit it to the controller.
-The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
-The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
-
-\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
-see \cref{fig:csvsncs}.
-This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
-\glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
-This leads to enhanced scalability and flexibility, and allows improved coordination and collaboration among systems.
-Furthermore, \glspl{NCS} provide greater access to information, as they enable the exchange of data and information between control components and systems, leading to more informed decision-making and improved performance.
-
-\glspl{NCS} have become increasingly popular in recent years due to advances in communication and control technologies, as well as the increasing demand for real-time, remote, and distributed control.
-They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the cloud for centralized, distributed control in what has been called \emph{cloud control systems}~\cite{xia2015cloud}
-However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms simply cannot meet~\cite{wan2020efficient}.
-This has led their adoption to be limited mostly to industrial environments.
-This is about to change, however, as with the advent of edge computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
-These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.
-
-\medskip
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-\todo[inline]{TODO FINISH}
-
-%\glspl{NCS}~\cite{Gupta2010NCSOverview}, a type of \gls{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
-%Depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
-%This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.
-
-%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \glspl{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.
-
-%One the one hand, related literature in \glspl{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level.
-%%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
-%On the other, there exist several approaches when considering experimental methodologies.
-%%A number of works concerning \glspl{NCS} deal with experimental studies.
-%%\glspl{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
-%One such approach uses setups in which the complete system is built on top of real hardware.
-%This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
-%Conversely, other studies choose to instead use completely \emph{simulated} \gls{NCS} setups.
-%The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
-%These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \glspl{NCS}.
-%Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
-%\begin{inlineenum}[itemjoin={{; }}, itemjoin*={{; or }}]
-%    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
-%    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
-%\end{inlineenum}
-%
-%As evidenced above, experimental research in \glspl{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
-%This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
-%Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \gls{NCS}.
-%This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.
-%
-%Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
-%In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\
-%Their methodology utilizes joint knowledge of control, computation, and communication.
-%In their work various architectural elements and the corresponding delays associated with the NCS are modelled.
-%Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation.
-%This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
-%Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.
-
-\subsubsection{Related methodologies and tools}
-
-The work we present in this thesis lies at the intersection of research on edge computing, \gls{WCA} applications, and \acl{CPS} such as \glspl{NCS}, each of these topics representing a vast body of research.
-Of principal interest to us are works dealing with the optimization of \gls{WCA} and \glspl{NCS} deployed on distributed systems such as edge computing.
-Below we will summarize the key milestones in literature relating to this.
-
-\paragraph{Experimental platforms for Edge Computing}
-
-Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications.
-On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
-The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.
-
-Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the
-\begin{inlineenum}
-    \item \gls{COSMOS}
-    \item \gls{POWDER}
-    \item \gls{ARA}
-    \item Drexel Grid \gls{SDR}
-\end{inlineenum} testbeds.
-\gls{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
-It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
-\gls{COSMOS} relies on the \gls{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \gls{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.
-
-\gls{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
-The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes.
-\gls{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \gls{VM} images containing the necessary software and configurations.
-These profiles are defined through using \gls{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.
-
-The \gls{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
-It includes a broad range of wireless platforms ranging from low-\gls{UHF} massive \gls{MIMO} to \unit{\milli\meter}Wave, deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-\gls{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \gls{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
-This affords the \gls{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.
-
-Finally, the \emph{Drexel Grid \gls{SDR} Testbed} features \glspl{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
-Primarily intended for \gls{SDR}-centric research, it does not integrate any core, cloud or edge components.
-However, the testbed extensively employs the \gls{LXC} runtime for the deployment of both experimental code and \gls{SDR} software, which affords users great freedom when it comes to the development of experiments.
-
-Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
-Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
-Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
-This is, for instance, the case with \gls{COSMOS} and \gls{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
-These testbeds further leverage virtualization technology based on \glspl{VM} instead of more lightweight and edge-compatible solutions such as containers.
-
-To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
-CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \glspl{VM} containing the necessary software for experiments.
-Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
-Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \pgls{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.

### 2023-03-06 body/part1/chapter1-introduction/background.tex
more xr

@@
 \todo[inline]{Add some references}
+\todo[inline]{Discuss tethered versus untethered}
 
@@
 
-\gls{XR} has also been leveraged for education and training.
-\gls{VR}
+\gls{XR} also presents tremendous opportunities for education and training.
+Through the use of these technologies, students can get hands-on yet safe, simulated experience in a multitude of scenarios.
+\gls{VR} is leveraged, for instance, in the training of new medical professionals, allowing students to learn anatomy and practice surgical procedures in realistic virtual environments.
+On the other hand, \gls{AR} and \gls{MR} are employed in manufacturing, to train and assist technicians in highly specialized assembly lines.
+Workers can receive real-time guidance through \gls{AR}-enabled headsets, improving accuracy and safety while controlling machine operations on the factory floor.
+\gls{AR} can also provide augmented, step-by-step guidance for maintenance and repair of equipment, augmenting thus the capabilities of workers and the productivity of the workplace as a whole.
 

### 2023-03-07 body/part1/chapter2-relatedwork/contributions.tex
todonotes
writing

@@
 \todo[inline]{Needs tweaking to match scope and related work.}
+\todo[inline]{Missing discussion, presentation, of alternative approaches explicitly.}
+\todo[inline]{Challenges and how we tackled them.}
 
@@
 We conclude that significant improvements can be achieved through the use of our human user model.
-
-\subsection{Overview of each included paper}
-
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
-This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
-\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
-We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-
-In \cref{paper:olguinmunoz2022cleave}, we extend our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
-These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, we introduce a tool, \acs{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
-The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
-We then present use-case scenarios demonstrating the scalability and flexibility of their approach by deploying scenarios with a large number of loops without the need for domain-specific hardware.
-This work showcases the applicability of the methodology beyond wearable cognitive assistance and highlights the potential of our approach for other edge computing applications.
-
-\cref{paper:olguinmunoz2022ainur} presents a tangential contribution.
-It introduces the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
-This framework represents an ancillary contribution, crucial for the research presented in these works.
-Without it, the experimental approach described in these works would not have been feasible.
-
-Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
-Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
-In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
 

### 2023-03-07 body/part1/chapter2-relatedwork/scope.tex
writing

--- 2022-12-19 body/part1/introduction.tex
@@
-\chapter{Introduction}\label{chap:introduction}
+\section{Scope of this work}
 
-\todo[inline]{
-    Introduction here. Why do we need this work?
-}
+This dissertation presents a methodology for the study of feedback-loop \gls{CPS} systems deployed on Edge Computing and studies its potential applications and implications.
+This methodology is based on the emulation of the desired application workload on top of the actual Edge infrastructure, including backend computing hardware and network technology used for client-server communication.
+We do not in any way claim that this methodology is the best approach to studying these systems, but rather that it is a valid approach able to produce accurate, representative results.
 
-\todo[inline]{
-    Reorganize EdgeDroid, Cleave, Ainur, together.
+The scope of \gls{CPS} systems is vast and diverse, making it unfeasible to investigate all possible variations of these systems in this work. 
+Instead, we opt to study a representative subset of future applications that could potentially leverage Edge Computing resources to provide innovative solutions, step-based task-assistance \gls{WCA} and \glspl{NCS}.
+By considering these two application domains, we demonstrate the applicability of our methodology and showcase its efficacy.
 
-    Pull in multi-loop results for EdgeDroid?
+Specifically, \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
+\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} present a high level overview of this methodological approach as applied to step-based task-assistance \gls{WCA}.
+The \gls{WCA} system employed for these works corresponds to the LEGO assistant devised by \citeauthor{chen2015early}\cite{chen2015early}, using a task with \num{7} distinct steps.
 
-    Think about experiments.
-}
-
-\section{Summary of the key contributions of this thesis}
-
-This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
-Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \gls{WCA} and \gls{NCS}.
-This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
-
-Secondly, we present a deeper exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
-This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
-
-Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
-In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
-We conclude that significant improvements can be achieved through the use of our human user model.
-
-\subsection{Overview of each included paper}
-
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
+\cref{} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
 This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
@@
 This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
-
-\section{Structure of this thesis}
-
-\todo[inline]{This thesis is structured as follows.}

### 2023-03-08 body/part1/chapter1-introduction/introduction.tex
scope draft

@@
 The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
+
+\subsection{Structure of this dissertation}
+
+This dissertation is structured into two parts.
+\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
+Next, in \cref{part:publications} we present the publications that form the core of this thesis.
+
+\cref{part:summary} is further structured into four main chapters.
+\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
+It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance.
+
+In \cref{chap:relatedwork}, we present theoretical background and related work.
+The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
+We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
+
+\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
+We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
+We aim to demonstrate the significance of our work and how it contributes to the field.
+
+Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
+We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
+This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-03-08 body/part1/chapter2-relatedwork/contributions.tex
scope draft

--- 2023-03-05 body/part1/chapter2-relatedwork/contributions.tex
@@
 
-\subsection{Overview of each included paper}
+\subsection{Overview of the included publications}
 
@@
 This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
-
-\section{Structure of this dissertation}
-
-This dissertation is structured into two parts.
-\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
-Next, in \cref{part:publications} we present the publications that form the core of this thesis.
-
-\cref{part:summary} is further structured into four main chapters.
-\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
-It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance.
-
-In \cref{chap:relatedwork}, we present theoretical background and related work.
-The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
-We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
-
-\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
-We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
-We aim to demonstrate the significance of our work and how it contributes to the field.
-
-Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
-We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
-This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-03-08 body/part1/chapter2-relatedwork/scope.tex
scope draft

@@
 The scope of \gls{CPS} systems is vast and diverse, making it unfeasible to investigate all possible variations of these systems in this work. 
-Instead, we opt to study a representative subset of future applications that could potentially leverage Edge Computing resources to provide innovative solutions, step-based task-assistance \gls{WCA} and \glspl{NCS}.
+Instead, we opt to study a representative subset of future applications that could potentially leverage Edge Computing resources to provide innovative solutions: \gls{WCA} and \glspl{NCS}.
 By considering these two application domains, we demonstrate the applicability of our methodology and showcase its efficacy.
 
-Specifically, \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} present a high level overview of this methodological approach as applied to step-based task-assistance \gls{WCA}.
-The \gls{WCA} system employed for these works corresponds to the LEGO assistant devised by \citeauthor{chen2015early}\cite{chen2015early}, using a task with \num{7} distinct steps.
+Specifically, a subset of the research works included in this dissertation presents and studies the application of our methodology for \gls{WCA}.
+To limit the scope of our research in this area, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
+We focus on this category for a couple of reasons.
+Step-based task-assistance \gls{WCA} were among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represent a more mature application class compared to other \glspl{WCA}.
+These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
+The specific application used for these studies corresponds to the LEGO assistant devised by \citeauthor{chen2015early}\cite{chen2015early}.
 
-\cref{} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
-This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
-\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
-We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+A separate subset of the research presented in this work discusses a preliminary foray into the applications of our methodology for \glspl{NCS}.
+Given the vastness and complexity of this field, we limit our exploration to simple, small-scale, single-plant systems.
+In particular, the system used as a representative example for the effectiveness of our approach for \glspl{NCS} corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
 
-In \cref{paper:olguinmunoz2022cleave}, we extend our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
-These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, we introduce a tool, \acs{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
-The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
-We then present use-case scenarios demonstrating the scalability and flexibility of their approach by deploying scenarios with a large number of loops without the need for domain-specific hardware.
-This work showcases the applicability of the methodology beyond wearable cognitive assistance and highlights the potential of our approach for other edge computing applications.
-
-\cref{paper:olguinmunoz2022ainur} presents a tangential contribution.
-It introduces the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
-This framework represents an ancillary contribution, crucial for the research presented in these works.
-Without it, the experimental approach described in these works would not have been feasible.
-
-Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
-Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
-In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+Finally, this dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
+Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \gls{SDR} paired with general-purpose \gls{COTS} compute hardware.
+We do not target large-scale testbeds due to their complexity, both in terms of hardware and software.

### 2023-03-09 body/part1/chapter2-relatedwork/relatedwork.tex
scope draft
relatedwork

@@
 
+\textcite{munro2016aaremu}: Software framework for the emulation of AR on Android devices.
+A trace of pre-recorded inputs (video, gyro, GPS) are replayed to the real processing code by mocking Android SDK APIS.
+Recordings can be ``interactive'', in which case the trace corresponds to images indexed by directional bearings, i.e.\ a 3D sphere), or non-interactive, in which case the trace corresponds to a video replayed at a customizable framerate.
+
+\textcite{williams2013transform}: Toolkit for visualization of AR and evaluation of tracking algorithms.
+Recorded trace is replayed in a virtualized environment.
+
+\textcite{berning2013parnorama}: Inverted emulation.
+AR application is emulated, user is real.
+AR application emulation corresponds to 360 degree FOV video.
+Used for prototyping user interactions in AR.
+
 \subsection{Emulation of \glsfmtshortpl{NCS}}
+
+\subsection{Testbed research in Edge Computing}
+
+\todo[inline]{Needs harmonizing.}
+
+The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
+Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
+Of particular interest to us are the
+\begin{inlineenum}
+    \item \acs{GENI}
+    \item Chameleon
+    \item \acs{COSMOS}
+    \item \acs{POWDER}
+    \item \acs{ARA}
+    \item EdgeNet
+%    \item Drexel Grid \gls{SDR}
+\end{inlineenum} testbeds.
+
+\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
+Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
+The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
+
+The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
+Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
+The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
+
+\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
+The testbeds main \emph{raison d'être} is \gls{MEC} research, with a strong emphasis on wireless networks.
+The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
+
+\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
+It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
+These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
+
+The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
+\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
+Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+
+EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
+It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
+The EdgeNet software is available for anyone to download;
+individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
+The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
+A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
+
+Finally, a number of other, small-scale approaches also exist.
+In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
+The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
+This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
+In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
+Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.

### 2023-03-09 recyclebin/experimental_research_edge_computing.tex
relatedwork

@@
 \todo[inline]{Should be moved to related work or removed.}
-
-The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
-Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
-Of particular interest to us are the
-\begin{inlineenum}
-    \item \acs{GENI}
-    \item Chameleon
-    \item \acs{COSMOS}
-    \item \acs{POWDER}
-    \item \acs{ARA}
-    \item EdgeNet
-%    \item Drexel Grid \gls{SDR}
-\end{inlineenum} testbeds.
-
-\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
-Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
-
-The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
-Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
-
-\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
-The testbeds main \emph{raison d'etre} is \gls{MEC} research, with a strong emphasis on wireless networks.
-The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
-
-\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
-It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
-These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
-
-The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
-\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-
-EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
-It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
-The EdgeNet software is available for anyone to download;
-individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
-The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
-A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
-
-Finally, a number of other, small-scale approaches also exist.
-In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
-The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
-This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
-In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
-Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.

### 2023-03-09 body/part1/chapter2-relatedwork/relatedwork.tex
relwork

@@
 Used for prototyping user interactions in AR.
+
+\textcite{choi2022emulating}: XREmul, input emulation tool for XR.
+Similar to \cite{munro2016aaremu}, traces of inputs are pre-recorded and then replayed.
+Allows for abstraction of inputs in general classes for device-agnostic inputs.
+Paper highlights the challenges of deterministic trace (leaves solution to future work).
 

### 2023-03-13 body/part1/chapter2-relatedwork/relatedwork.tex
ncs research

@@
 \subsection{Emulation of \glsfmtshort{XR} systems}
+
+\todo[inline]{Maybe expand to evaluating performance of XR? Prototyping of XR?}
 
@@
 
-\subsection{Emulation of \glsfmtshortpl{NCS}}
+\textcite{chetoui2022arbench}: ARBench, a benchmarking suite for AR.
+Corresponds to a software package including a multitude of AR workloads, emulating the load placed on the the hardware.
+Only considers locally-running AR, not network.
+Does not emulate human behavior, uses traces of AR applications.
+
+\textcite{george2020openrtist}: OpenRTiST, benchmarking workload for Edge Computing, AR
+Based on stable diffusion.
+
+\subsection{Experimental research in \glsfmtshortpl{NCS}}
+
+A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
+Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies.
+\gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
+
+Within the body of works relating to experimental research on \glspl{NCS}, we identify three distinct approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
+
+In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
+This approach is straightforward to implement, and can be very useful for evaluating the real-world applicability of the system, but can be costly and time-consuming.
+Examples of the application of this approach can be found in studies such as~\cite{drew2005networked,baumann2018evaluating,li2014wireless,cuenca2019periodic}.
+These studies implemented wireless and microcontroller-based systems for control and validated them on a physical prototype.
+
+In~\cite{drew2005networked}, the authors present a \gls{NCS} model which considers both random packet delay and loss on both the sensor and actuator sides of the feedback loop.
+They validate their design on a physical testbed consisting of inverted pendula controlled from a central computation point over Wi-Fi.
+A similar testbed is employed for validation in~\cite{baumann2018evaluating}, in which the authors develop an evaluation methodology for wireless \gls{NCS}.\@
+\cite{li2014wireless} implements a wireless microcontroller-based system for vibration control and validates it on a physical prototype comprising a cantilever beam controlled using an \acs{IEEE} 802.15.4 \gls{WPAN} system.\@ \cite{cuenca2019periodic} discusses the design and implementation of a periodic event-triggered sampling and dual-rate control techniques for wireless control, and validate their contributions on a four-rotor autonomous drone platform controlled over WiFi.
+\emph{NCSbench}~\cite{zoppi2020ncsbench}, an open-source \gls{NCS} benchmarking platform designed with reproducibility in mind built using the \citetitle{LEGOMindstormsEV3}~\cite{LEGOMindstormsEV3} platform, is another such example.
+
+Completely simulated setups, on the other hand, rely on computer simulations to evaluate the \gls{NCS}'s performance.
+This approach is less expensive and faster than the practical approach but may not fully capture the real-world complexities of the system.
+Examples of tools and frameworks for the implementation of this approach can be found in works such as~\cite{andersson2005simulation,eyisi2012ncswt}.
+
+A simulated approach is employed in~\cite{du2009novel}, in which the authors develop a novel \emph{Smith} predictor to compensate for varying latencies in wireless \glspl{NCS} and validate it using the \emph{TrueTime}~\cite{henriksson2002truetime} simulator.
+In~\cite{chen2015synchronous} the authors validate and evaluate a number of synchronous control strategies for three-motor setups using completely simulated \glspl{NCS} environments.\@
+\cite{wu2012application} introduces the concept of \emph{network predictive control} in and uses it to balance a simulated inverted pendulum over a simulated wireless network.\@
+\cite{ma2019optimal} proposes an optimal dynamic scheduling strategy that optimizes performance of multi-loop control systems and validate it by simulating a four-loop control system over an \acs{IEEE} 802.15.4 \gls{WPAN}.
+
+Finally, network and hardware-in-the-loop approaches, which aim to strike a balance between realism and efficiency, have also been employed in literature.
+Network-in-the-loop refers to experimental setups in which the \emph{network} is instead simulated or emulated.
+Such approaches have, for instance, been used to study the effects of the \gls{CSMA/CD} medium access control mechanism on plant stability, as in~\cite{natale2004inverted}.
+Hardware-in-the-loop approaches, referring to \gls{NCS} setups in which a real network interacts with a simulated or emulated control system, are particularly prevalent in the field of \emph{smart grid} control.
+An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
 

### 2023-03-14 body/part1/chapter1-introduction/background.tex
background

@@
 
+\subsubsection{\glsfmtlong{WCA}}
+
+Of particular interest for this dissertation is a particular subset of \gls{AR} denoted \acf{WCA}, a technology designed to provide real-time support and guidance to individuals performing complex tasks in a plethora of domains.
+Designed to be deployed on wearable devices, \gls{WCA} overlays digital information onto the real world and provides context-aware feedback to users, in order to assist and guide them in to efficiently complete a task~\cite{ha2014towards,chen2015early,chen2018application,wang2020scaling}.
+
+These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users. 
+The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
+The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video feeds.
+Whenever the assistant detects a change in the context of the user, it provides a new instruction.
+The application otherwise remains silent and out-of-the-way.
+
+Originally motivated by assistive use-cases for individuals with reduced cognition due to illness, injury, or age, \gls{WCA} technology has been evolved to include the enhancement of the cognitive abilities of users in areas such as manufacturing, healthcare, and aviation.
+Among the primary benefits of this technology is that it can help reduce the cognitive load on individuals performing complex tasks.
+By providing real-time guidance and support, it can improve accuracy and speed of task completion, reduce errors and minimize the risk of accidents.
+For example, a \gls{WCA} could help a technician performing maintenance on an aircraft by displaying relevant technical information about the aircraft parts and step-by-step instructions on how to perform complex specific repairs.
+
+\gls{WCA} technology employs sensors, cameras, and other hardware in wearable devices to collect data from the environment and the user.
+This data is then processed to interpret the data and provide relevant feedback.
+As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
+The context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
+
+Overall, \gls{WCA} is a promising technology that has the potential to revolutionize the way we perform complex tasks.
+As the technology continues to evolve and become more sophisticated, it is expected to have a significant impact on many industries and improve the safety and efficiency of numerous processes.
+
 \subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
@@
 
-\medskip
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-
-In terms of research relating to the benchmarking and optimization of these systems, works in literature have for the most part leveraged either theoretical models~\cite{zhang2019networked} or simulations (for example ~\cite{ma2019optimal}).
-However, these approaches can only capture network behaviors at a coarse level, and thus a body of works has emerge in later years involving experimental approaches to \gls{NCS} benchmarking.
-\cite{baumann2018evaluating} and~\cite{cuenca2019periodic} employ purpose-built physical testbeds for the study of these systems.
-\cite{zoppi2020ncsbench} introduces \emph{NCSBench}, a platform for reproducible benchmarking of \gls{NCS} using a LEGO Mindstorms platform.
-
-Finally, hybrid approaches, mixing testbed or prototype implementations and simulations or emulations of \gls{NCS} components also exist in literature.
-These approaches tend to focus on a particular element of the system which is implemented using a prototype or testbed, and virtualize the rest of the system.
-For instance, in~\cite{wang2020inverter} the authors are concerned with realistic effects from the network, and thus they have an emulated \gls{NCS} interact with a real network.
-On the other hand, works such as~\cite{natale2004inverted} focus on the system being controlled and opt to instead emulate the network.
-
 \subsection{Emulation as a research methodology}

### 2023-03-14 body/part1/chapter2-relatedwork/relatedwork.tex
background

@@
 
-\subsection{Emulation of \glsfmtshort{XR} systems}
+\subsection{Wearable Cognitive Assistance}
 
-\todo[inline]{Maybe expand to evaluating performance of XR? Prototyping of XR?}
+\subsection{Benchmarking and emulation of \glsfmtshort{XR} systems}
 
@@
 A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
-Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies.
+Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
+Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
+They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
+
+Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies~\cite{zhang2019networked}.
 \gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
-
-Within the body of works relating to experimental research on \glspl{NCS}, we identify three distinct approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
+However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerge in later years involving experimental approaches to \gls{NCS} benchmarking.
+Within this body of work, we identify three distinct approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
 

### 2023-03-14 body/part1/chapter1-introduction/background.tex
background

@@
 We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
+
+\subsection{Emulation and simulation}
+
+The terms \emph{emulation} and \emph{simulation} are often used interchangeably.
+However, although both terms refer to methods of imitating and modeling systems, they are distinct in their goals, methods, and applications.
+
+Emulation refers to the process of reproducing or mimicking the observed behavior of a system through another system, 
+with the ultimate goal being the recreation of behavior that matches the original system as close as possible.
+Emulations run in \emph{real time}, and can thus interact with real environments;
+in turn, this means that emulations can be used to replace subcomponents of larger systems for research or benchmarking.
+This method does not aim to model the original system, nor aim to understand its underlying behavior.
+
+Simulation, on the other hand, is the process of modeling a system to study its behavior through the creation of a mathematical or logical model of the system.
+The model can then be adjusted to simulate and better represent the original system in different scenarios and conditions.
+Simulations run in entirely virtualized environments, and are not beholden to any real-time constraint.
+
+Finally, these approaches differ as well in the level of accuracy each of them affords.
+While simulation only attempts to approximate the behavior of the system in a given context, emulation aims to represent as close to a perfect copy of the observed behavior of the system.
+The former is therefore employed when a general understanding of the system is sufficient, while the latter is used when precision is required.
+However, thanks to its real time constraint and high degree of realism, emulation is typically more resource-intensive than simulation.
+This means that emulation requires more processing power, memory, and storage space than simulation.
 
@@
 \glsreset{NCS}
-\todo[inline]{Needs citations. Remove?}
 
-\glspl{NCS} are a type of control system that involves the integration of communication networks into traditional control systems.
+\glspl{NCS} are a class of control systems that involve the integration of communication networks into traditional control systems.
 A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
-Control systems are composed of
+These systems are composed of
 \begin{inlineenum}

### 2023-03-14 body/part1/chapter1-introduction/introduction.tex
background

@@
 
-\subsection{Structure of this dissertation}
+% \subsection{Structure of this dissertation}
+
+\medskip
 
@@
 
-\cref{part:summary} is further structured into four main chapters.
+\cref{part:summary} is itself structured into four main chapters.
 \cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
-It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance.
+It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
 
-In \cref{chap:relatedwork}, we present theoretical background and related work.
+In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
 The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.

### 2023-03-14 body/part1/chapter2-relatedwork/scope.tex
background

@@
 Finally, this dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
-Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \gls{SDR} paired with general-purpose \gls{COTS} compute hardware.
+Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \glspl{SDR} paired with general-purpose \gls{COTS} compute hardware.
 We do not target large-scale testbeds due to their complexity, both in terms of hardware and software.

### 2023-03-16 body/part1/chapter2-relatedwork/relatedwork.tex
background
relwork

@@
 
-\subsection{Wearable Cognitive Assistance}
+\subsection{Emulation in \glsfmtshort{XR} research}
 
-\subsection{Benchmarking and emulation of \glsfmtshort{XR} systems}
+Only a limited number of works addressing the partial or complete emulation of XR applications for benchmarking and research purposes exist in literature.
+One such work is~\cite{williams2013transform}.
+In this, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
+The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.
 
-\textcite{munro2016aaremu}: Software framework for the emulation of AR on Android devices.
-A trace of pre-recorded inputs (video, gyro, GPS) are replayed to the real processing code by mocking Android SDK APIS.
-Recordings can be ``interactive'', in which case the trace corresponds to images indexed by directional bearings, i.e.\ a 3D sphere), or non-interactive, in which case the trace corresponds to a video replayed at a customizable framerate.
+\cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{AR} application by mocking Android \gls{SDK} \glspl{API}.
+This approach allows for benchmarking and evaluation of algorithms and architectures on real data.
+A similar approach was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
 
-\textcite{williams2013transform}: Toolkit for visualization of AR and evaluation of tracking algorithms.
-Recorded trace is replayed in a virtualized environment.
+Finally,\ \cite{chetoui2022arbench} employed \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
+Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
+This approach is similar to the one presented in this dissertation, but it exclusively considers on-device \gls{AR} processing.
+Furthermore, the benchmarks are deterministic, and no emulation of realistic human behavior is attempted.
 
-\textcite{berning2013parnorama}: Inverted emulation.
-AR application is emulated, user is real.
-AR application emulation corresponds to 360 degree FOV video.
-Used for prototyping user interactions in AR.
+\subsection{Edge and Cloud Benchmarking}
 
-\textcite{choi2022emulating}: XREmul, input emulation tool for XR.
-Similar to \cite{munro2016aaremu}, traces of inputs are pre-recorded and then replayed.
-Allows for abstraction of inputs in general classes for device-agnostic inputs.
-Paper highlights the challenges of deterministic trace (leaves solution to future work).
-
-\textcite{chetoui2022arbench}: ARBench, a benchmarking suite for AR.
-Corresponds to a software package including a multitude of AR workloads, emulating the load placed on the the hardware.
-Only considers locally-running AR, not network.
-Does not emulate human behavior, uses traces of AR applications.
-
-\textcite{george2020openrtist}: OpenRTiST, benchmarking workload for Edge Computing, AR
-Based on stable diffusion.
+\emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
+Video is streamed from a source Android device to a compute node where neural style-transfer~\cite{gatys2016image} is applied to it.
+The transformed video is then fed back to the source to be presented, resulting in a compute-intensive, bandwidth-hungry, and latency-sensitive benchmark.
 
@@
 \subsection{Testbed research in Edge Computing}
-
-\todo[inline]{Needs harmonizing.}
 

### 2023-03-17 body/part1/chapter2-relatedwork/relatedwork.tex
more related work

@@
 
-Finally,\ \cite{chetoui2022arbench} employed \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
+Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
 Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
@@
 
-\emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
+This dissertation complements existing work on benchmarking distributed system architectures.
+A rich body of literature exists with examples of such approaches for Cloud computing.
+For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and cloud services.
+Workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
+Both these approaches heavily focus on data processing and downlink transmission in cloud computing scenarios.
+In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the cloud.
+The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and include generators which mimic remote clients accessing the system over the internet.\@
+\cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
+
+A number of works have tackled this challenge in the context of Fog and Edge Computing as well.
+\emph{EdgeBench}, introduced in~\cite{das2018edgebench}, corresponds to a benchmarking suite for the Edge.
+It emulates three different workloads, speech-to-text, image recognition, and sensor network application traffic, and targets exclusively server and specifically targets serverless deployments.\@
+\cite{baurle2022comb} presents \emph{ComB}, a flexible and extensible, application-oriented benchmarking suite for Edge computing.
+Its stated purpose is to assist system designers and application developers to evaluate the suitability of Edge deployments.
+The framework is designed to decouple orchestration and workload, allowing for easy extension or replacement of the built-in distributed video-analytics benchmark.
+
+\emph{IoTBench}~\cite{lee2019iotbench} is a benchmarking tool for edge-deployed \gls{IOT} applications.
+It includes seven representative workloads, and allows researchers to evaluate the performance of these applications on Edge infrastructure through the analysis of computational demands and resource consumption.\@
+\cite{mcchesney2019defog} presents a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud named \emph{DeFog}.
+\emph{DeFog} includes six Fog- and Edge-relevant workloads, including real-time video detection and a text-to-speech application, which are latency-critical and bandwidth-intensive.
+In~\cite{lockhart2020scission}, the authors introduce a tool, \emph{Scission}, for the benchmarking of \glspl{DNN} partitioned across the user device, edge, and cloud continuum.
+The tool can also be used to automatically determine the optimal partition of a \gls{DNN}.
+Finally, \emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
 Video is streamed from a source Android device to a compute node where neural style-transfer~\cite{gatys2016image} is applied to it.

### 2023-03-24 body/part1/chapter1-introduction/background.tex
background
work
more tweaks

@@
 These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
-Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters~\citationeeded.
+Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters.
 These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
@@
 These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
-We will discuss these in depth in their respective sections, \cref{background:wca} and \cref{background:ncs}.
 
@@
 This means that emulation requires more processing power, memory, and storage space than simulation.
+
+\subsubsection{Emulation as a research methodology}
+\todo[inline]{Remove?}
 
@@
 These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.
-
-\subsection{Emulation as a research methodology}
-\todo[inline]{Rename?}

### 2023-03-24 body/part1/chapter2-relatedwork/relatedwork.tex
more tweaks

@@
 Workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
-Both these approaches heavily focus on data processing and downlink transmission in cloud computing scenarios.
+Both these approaches focus heavily on data processing and downlink transmission in cloud computing scenarios.
+
 In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the cloud.
-The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and include generators which mimic remote clients accessing the system over the internet.\@
+The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
 \cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
@@
 \emph{EdgeBench}, introduced in~\cite{das2018edgebench}, corresponds to a benchmarking suite for the Edge.
-It emulates three different workloads, speech-to-text, image recognition, and sensor network application traffic, and targets exclusively server and specifically targets serverless deployments.\@
+It emulates three different workloads, speech-to-text, image recognition, and sensor network application traffic, and specifically targets serverless deployments.\@
 \cite{baurle2022comb} presents \emph{ComB}, a flexible and extensible, application-oriented benchmarking suite for Edge computing.
@@
 \cite{mcchesney2019defog} presents a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud named \emph{DeFog}.
-\emph{DeFog} includes six Fog- and Edge-relevant workloads, including real-time video detection and a text-to-speech application, which are latency-critical and bandwidth-intensive.
+\emph{DeFog} includes six latency-critical and bandwidth-intensive, Fog- and Edge-relevant workloads, including real-time video detection and a text-to-speech application.
 In~\cite{lockhart2020scission}, the authors introduce a tool, \emph{Scission}, for the benchmarking of \glspl{DNN} partitioned across the user device, edge, and cloud continuum.
@@
 Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing edge computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
+They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
 
@@
 \gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
-However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerge in later years involving experimental approaches to \gls{NCS} benchmarking.
+However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
 Within this body of work, we identify three distinct approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
@@
 
-The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for edge computing research.
+The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for Edge Computing research.
 Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
@@
 Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling edge computing platform.
+The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling Edge Computing platform.
 
-The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target edge computing research through the \acs{CHI}@Edge project.
+The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
 Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and edge computing scenarios.
+The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and Edge Computing scenarios.
 
@@
 \gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced edge computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+Its core goal is the study and deployment of advanced Edge Computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
 
-EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for edge computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
+EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for Edge Computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
 It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
@@
 Finally, a number of other, small-scale approaches also exist.
-In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for edge computing research.
-The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} edge computing.
+In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for Edge Computing research.
+The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} Edge Computing.
 This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
-In~\cite{yamanaka2021design}, the authors present an experimental edge computing testbed capable of determining the offload location of a workload based on desired latency parameters.
+In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
 \cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.

### 2023-03-24 body/part1/chapter2-relatedwork/scope.tex
work
more tweaks

@@
-\section{Scope of this work}
+\section{Scope of this dissertation}
 
@@
 We focus on this category for a couple of reasons.
-Step-based task-assistance \gls{WCA} were among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represent a more mature application class compared to other \glspl{WCA}.
+Step-based task-assistance \gls{WCA} was among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represents a more mature application class compared to other \glspl{WCA}.
 These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.

### 2023-03-27 body/part1/chapter1-introduction/introduction.tex
more tweaks
introduction

@@
 The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
-One of the most promising of these is Edge Computing, which seeks to bring computing closer to the end user, reducing latency and increasing reliability.
-Another key player in this space is 5G, the latest generation of mobile communication networks that offers higher data rates, lower latency, and improved reliability.
-The combination of these two technologies holds great promise for the deployment of latency-sensitive applications, including \glspl{CPS}, \glspl{NCS}, and \gls{WCA} applications, among others.
+One of the most promising of these is Edge Computing, a distributed computing paradigm which seeks to bring computing closer to the end user, reducing latency (i.e.\ the time between input and response) and increasing reliability.
 
-Despite these advancements, there are still significant challenges in understanding and scaling these systems, particularly in regard to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
-The aim of this thesis is to make a contribution to the field by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
+The current model for distributed computing, known as Cloud Computing, allows users to access shared pools of resources, services, and applications over the internet~\cite{gai2012towards}.
+Cloud computing offers many advantages, such as virtually unlimited storage and processing power, by virtue of employing economies of scale in large, geographically distributed datacenters.
+However, this design leads to trade-offs in bandwidth and, particularly, latency.
+The responsiveness of the Cloud is limited by the speed of light, and thus it can suffer from high latency due to the distance data must travel between the end user and the cloud.
+This can cause delays in applications that require quick response times, such as real-time video processing or online gaming.
+Additionally, the centralized nature of Cloud datacenters means that it can be vulnerable to outages or other disruptions, which can impact large numbers of users simultaneously.
+
+Edge Computing aims to improve upon the shortcomings of Cloud Computing by moving the processing closer to the end user, at the \emph{edge} of the network.
+In this context, the ``edge of the network'' refers to locations which are both geographically and topologically (in the networking sense) close to the user, such as cellular service provider towers/base stations or micro-datacenters located within the same urban area as the user.
+By placing computing resources in proximity to where data is generated and/or consumed, and thus reducing the distance that data must travel, Edge Computing can greatly decrease latency and improve the responsiveness of applications.
+It also allows for more efficient use of network resources, as data does not need to travel back and forth to a remote, central datacenter.
+
+The advent of 5G~\cite{5Gstandard}, the latest generation of mobile communication networks, offers even greater potential for Edge Computing.
+5G offers higher data rates, lower latency, and improved reliability compared to previous generations of mobile networks.
+This enables even more powerful and sophisticated applications to be run on the Edge, such as \acsp{CPS} and \acsp{NCS}, as well as \acs{MAR} applications.
+
+\acfp{CPS} and \acfp{NCS} refer to systems where physical processes, such as manufacturing or transportation, are controlled  and monitored by computers, usually over communication networks.
+Edge Computing, combined with 5G networking, allows these systems to operate in real-time, enabling near instantaneous response times, more efficient operation, and more precise control.
+These characteristics are critical for applications such as \glspl{ITS}, autonomous driving, or remote surgery.
+
+\gls{MAR}, on the other hand, is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
+This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
+\gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
+By using Edge Computing and 5G, \gls{MAR} can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.
+
+Overall, the combination of Edge Computing and 5G holds great promise for the massive deployment of latency-sensitive applications such as the ones discussed above.
+Nonetheless, significant challenges remain in the understanding and scaling of these systems, particularly in regard to their complexity and requirements in the context of low-latency, high-reliability computing.
+
+Latency-sensitive applications, such as \glspl{CPS} and \gls{MAR}, involve multiple processing steps that can influence the latency of the system.
+There are many trade-offs involved in optimizing the latency of these applications beyond the question of where the compute process is located. 
+These include the choice of wireless system for transmitting application data (e.g.\ 4G \gls{LTE} versus 5G), the protocols used to communicate over the network, and the hardware and operating system used in the backend.
+In the case of \gls{MAR}, the sensory inputs need to be pre-processed and compressed on-device, before being transmitted to the compute backend for processing.
+The backend algorithms need to be carefully designed and implemented to ensure efficient processing and minimal latency. 
+Similarly, in \glspl{CPS}, the communication network must be designed to minimize latency and ensure reliable data transmission between the components of the control system.
+Delayed or lost data can result in catastrophic failure of these systems.
+Other critical aspects with the potential to impact the latency of latency-sensitive applications on the Edge include congestion on the communication loop and fluctuations of the wireless channel.
+These issues can lead to delays in data transmission between devices on the network, leading to higher end-to-end latency.
+
+At the time of the writing of this dissertation, no widely adopted methodology exists for the study of these trade-offs.
+We aim to contribute to the field by investigating the applications of such a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
 The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
@@
 It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
-
 In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
@@
 We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
-
 \cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
@@
 We aim to demonstrate the significance of our work and how it contributes to the field.
-
 Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.

### 2023-03-27 body/part1/chapter1-introduction/introduction.tex
still working on intro

@@
 
+One of the main challenges to the study of latency-sensitive applications such as \gls{CPS} and \gls{MAR} on Edge Computing is that these applications interact with the real world, and thus must be studied in real-time.
+This implies that any study of these applications must be done in an environment that accurately reflects the real-world conditions in which they are deployed.
+Unfortunately, replicating these conditions in a controlled experimental environment is difficult and often leads to inconsistencies in results, and traditional methods that rely on simulations may not be sufficient to accurately capture and evaluate the performance of these applications in real-world scenarios.
+
+The scalability of studying these systems is also a challenge.
+It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, personnel and test subjects, as well as facilities.
+\glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{MAR} is time-consuming and expensive.
+
+The highly complex effects on the wireless network is another major challenge to studying latency-sensitive applications on Edge Computing.
+The presence of multiple devices accessing the network simultaneously can result in interference, network congestion, and other factors that can affect the performance of the system.
+These effects are difficult to simulate and/or emulate, particularly in multi-tenant deployments, where each tenant may have different requirements and resource usage patterns.
+
+Finally, the combination of the characteristics enumerated above makes the study of these applications hard to automate, making it challenging to conduct large-scale experiments or to collect data over an extended period of time.
+\gls{CPS} and \gls{MAR} applications are highly dynamic and adaptive, and their behavior cannot be easily predicted.
+Moreover, there is often a significant amount of variability in the behavior of different components of the application, such as sensors, actuators, and network connections, which makes it difficult to automate the testing process.
+
 Our methodology is based on the emulation of target workloads on actual Edge infrastructure.

### 2023-03-27 body/part1/chapter1-introduction/introduction.tex
introduction yay

@@
 At the time of the writing of this dissertation, no widely adopted methodology exists for the study of these trade-offs.
-We aim to contribute to the field by investigating the applications of such a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
+We aim to contribute to the field by introducing and subsequently investigating the applications of such a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
 The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
@@
 
-Our methodology is based on the emulation of target workloads on actual Edge infrastructure.
-We replace the client side of the system with a realistic emulation of the desired behaviors, implemented in software deployed on \gls{COTS} general-purpose computing devices.
-In our initial implementation, these correspond to low-cost, easily replaceable, and scalable Raspberry Pi 4 Model B \glspl{SBC}.
+Our methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{MAR} applications. 
+We emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
+    In our initial implementation, these correspond to low-cost and easily replaceable, scalable Raspberry Pi 4 Model B \acsp{SBC}.
+} running a realistic software imitation of the desired behaviors.
+At the same time, our methodology retains the real network as well as the real compute hardware and software at the backend.
+This design results in real-time benchmarking of applications, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.
 
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, COTS general-purpose hardware such as \glspl{SBC}.
-It also preserves the realism of effects stemming from the hardware and network.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
+This also reduces the barrier of entry to this research, as \glspl{SBC} are for the most part cheap and easily accessible.
+Additionally, it preserves the realism of effects stemming from the hardware and network.
 In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
 
-The methodology also provides improved repeatability and replicability.
+The methodology also provides automation as well as improved repeatability and replicability.
+The softwarized nature of the client-side emulation allows for straightforward automation through general-purpose programming and scripting languages.
 Repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
@@
 
-\medskip
+\bigskip
 

### 2023-03-28 body/part1/chapter1-introduction/background.tex
drafting

@@
 
-\subsection{Edge computing}
+\subsection{Edge Computing}
+\todo[inline]{Maybe add a table where you compare cloud and edge computing regrading different characteristics?}
 \glsresetall%
 
-Edge computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
+Edge Computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
 Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
@@
     \caption{%
-        Conceptual design of edge computing.
+        Conceptual design of Edge Computing.
         Micro-clouds (known as ``cloudlets''~\cite{satyanarayanan2009case}) are placed a at the edge of the network, a few hops away from end users.
@@
 \medskip
-Edge computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
+Edge Computing emerges as a potential answer to these challenges~\cite{satyanarayanan2009case,shi2016promise,shi2016edge,varghese2016challenges,satyanarayanan2017emergence,bittmann2017edge,wang2019towards}.
 The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
 To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in edge computing it is offloaded to a micro-datacenter at the \emph{edge} of the network close to the user.
-\citeauthor{satyanarayanan2009case} named these compute nodes \emph{cloudlets}, and envisioned them as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user (see \cref{fig:edgecomputing})
-This built on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
+Instead of offloading computation to a datacenter potentially thousands of kilometers away, in Edge Computing it is offloaded to a compute node at the \emph{edge} of the network close to the user, see \cref{fig:edgecomputing}.
+
+The architecture of Edge Computing offers several advantages, the primary being the significant reduction of latency.
+Edge Computing can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
+Another important advantage of this design is bandwidth reduction.
+Edge application traffic either does not need to traverse the public internet, or it can be aggregated and compressed at the Edge before uplink transmission.
+
+The original Edge Computing concept as introduced in~\citeauthor{satyanarayanan2009case} has come to be known as \emph{Cloudlet-based} Edge Computing.
+It is based on the idea that Edge compute capabilities will be realized through geographically distributed micro-datacenters known as \emph{cloudlets}.
+Cloudlets have been envisioned as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user.
+This design builds on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
 Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
-This reduces energy consumption and allows for the deployment of otherwise unfeasible workloads on mobile hardware.
 
-The architecture of edge computing offers several advantages, the primary being the significant reduction of latency.
-Cloudlets can serve highly latency-sensitive and resource-intensive applications with extremely low latencies while still providing orders of magnitude more computing and energy resources than those present on mobile and wearable devices.
-Another advantage is bandwidth reduction.
-%By processing and aggregating data closer to where its needed, edge computing can significantly reduce the amount of traffic going to cloud datacenters.
-A variant of edge computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for this purpose.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
-Edge computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
-
-Finally, in later years, edge computing has been paired with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
-This combination is often referred to as \gls{MEC}, and has the potential to enable unprecedented use-cases requiring stringent latency bounds and high bandwidths.
+In later years, an extension to Edge Computing denoted \gls{MEC} has emerged, which pairs Edge Computing with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
+A modern mobile network is composed of two main components: the \gls{RAN} and the core network.
+The former includes the base stations and other wireless communication equipment to which user's equipment (i.e.\ mobile phones, mobile internet dongles, etc.) connect. 
+The \gls{RAN} provides access to the core network, which in turn routes data between \glspl{RAN}, handles billing and authentication, and provides access to external services such as the internet and the Cloud.
+\gls{MEC} is defined as an Edge Computing deployment in which Cloud-like compute is deployed within the \gls{RAN} of a mobile network.
+Compute power is never more than a single wireless hop away from users application traffic thus never enters the core or the public internet.
+This architectural design has the potential to enable unprecedented use-cases requiring stringent, sub-\SI{10}{\milli\second} latency bounds and high bandwidths.
 Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
@@
 
+Finally, another variant of Edge Computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for the reduction of bandwidth requirements.
+By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
+Edge Computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
+
 \subsection{Emulation and simulation}
+\todo[inline]{
+    Section 1.1.2: This subsection is not very informative so far.
+Maybe first discuss extended reality, and then present a subsection on performance evaluation of systems, where simulation is a subsection, and experimentation is another subsection (and mathematical analysis is a third one).
+Focus on the usage of these concepts in the context of networked systems.
+Why do we need tools/approaches for performance evaluation, what are the advantages and disadvantages of different approaches, and also what is the historical context.
+You also need to have references.
+Finally, tables and figures are always welcome...
+}
 
@@
 This has led their adoption to be limited mostly to industrial environments.
-This is about to change, however, as with the advent of edge computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
+This is about to change, however, as with the advent of Edge Computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
 These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.

### 2023-03-28 resources/glossary.tex
xr
more related work
introduction
drafting

--- 2023-02-07 body/glossary.tex
@@
-\usepackage{glossaries}%! suppress = MissingImport
+%! suppress = MissingImport
 \newacronym[longplural={Wearable Cognitive Assistants}]{WCA}{WCA}{Wearable Cognitive Assistance}
@@
 \newacronym[longplural={eXtended Realities}]{XR}{XR}{eXtended Reality}
+\newacronym[longplural={Mixed Realities}]{MR}{MR}{Mixed Reality}
+\newacronym{MAR}{MAR}{Mobile \acs{AR}}
 \newacronym{GPU}{GPU}{Graphical Processing Unit}
@@
 \newacronym{AWS}{AWS}{Amazon Web Services}
+\newacronym{GCP}{GCP}{Google Cloud Platform}
 \newacronym{EC2}{EC2}{Elastic Compute 2}
@@
 \newacronym{YAML}{YAML}{YAML Ain't Markup Language}
-\newacronym{O-RAN}{O-RAN}{Open Radio Access Network}
+\newacronym{RAN}{RAN}{Radio Access Network}
+\newacronym{O-RAN}{O-RAN}{Open \acs{RAN}}
 \newacronym{FOSS}{FOSS}{Free and Open Source Software}
@@
 \newacronym[longplural={Times-to-Feedback}]{TTF}{TTF}{Time-to-Feedback}
-\newacronym{MAR}{MAR}{Mobile \acs{AR}}
 \newacronym{QoE}{QoE}{Quality of Experience}
@@
 \newacronym{GENI}{GENI}{Global Environment for Networking Innovation}
+\newacronym{SBC}{SBC}{Single-Board Computer}
+\newacronym{EXPECA}{ExPECA}{Experimental Platform for Edge Computing Applications}
+\newacronym[user1={Kungliga Tekniska Högskolan}, prefixfirst={the\ }]{KTH}{KTH}{Royal Institute of Technology of Sweden}
+\newacronym{HCI}{HCI}{Human-Computer Interaction}
+\newacronym{2D}{2-D}{two-dimensional}
+\newacronym{3D}{3-D}{three-dimensional}
+\newacronym{PD}{PD}{Proportional-Differential}
+\newacronym{PID}{PID}{Proportional-Integral-Differential}
+\newacronym{SDK}{SDK}{Software Development Kit}
+\newacronym{YCSB}{YCSB}{Yahoo! Cloud Serving Benchmark}
+\newacronym{ITS}{ITS}{Intelligent Transportation System}
 

### 2023-03-28 body/part1/chapter1-introduction/background.tex
cloud vs edge table

@@
 \subsection{Edge Computing}
-\todo[inline]{Maybe add a table where you compare cloud and edge computing regrading different characteristics?}
 \glsresetall%
-
-Edge Computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the cloud.
-Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
-These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
-Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters.
-These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
-
-Through this design, cloud computing affords significant advantages to users.
-As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
-Thanks to economies of scale and virtualization technologies, the cloud is highly scalable;
-services can be scaled simply by spawning more \glspl{VM}.
-The specialized nature of cloud providers, the scale of modern datacenters, and the use of virtualization also make the cloud highly reliable.
-When hardware fails, recovering is simply a matter of migrating the service container or \gls{VM} to an available compute node~\cite{endo2016high}.
-
-However, the cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive and/or bandwidth-intensive applications.
-In order to achieve the necessary economies of scale, cloud datacenters are designed to serve users distributed across vast geographical areas.
-These installations are thus often located ``far'' from potential users;
-for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
-This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{tolia2006quantifying,lagar2007interactive,satyanarayanan2009case,varghese2016challenges,shi2016promise}.
-The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
-Such latencies are unfeasible to consistently achieve with cloud computing~\cite{dang2021cloudy}.
-On the other hand, as smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face increasing challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
 
@@
         Conceptual design of Edge Computing.
-        Micro-clouds (known as ``cloudlets''~\cite{satyanarayanan2009case}) are placed a at the edge of the network, a few hops away from end users.
-        In other words, these cloudlets are located between users and the internet and the cloud.
+        Compute nodes are placed a at the edge of the network, a few hops away from end users.
+        In other words, compute is located between users and the internet and the Cloud.
     }\label{fig:edgecomputing}
 \end{figure}
+
+Edge Computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the Cloud.
+Cloud computing, the reigning distributed computing model, allows users to access shared pools of resources such as servers, databases, and applications, over the internet~\cite{gai2012towards}.
+These pools of resources are managed in a centralized manner by specialized providers, and users and businesses can access them on-demand, without having to invest in and manage infrastructure of their own.
+Providers in turn employ economies of scale, providing these services by deploying massive amounts of computing power and storage capacity in specialized locations known as datacenters.
+These hardware resources are then further compartmentalized through the use of virtualization technologies such as \glspl{VM} and containers~\cite{gai2012towards}.
+
+Through this design, Cloud computing affords significant advantages to users.
+As services are deployed in a centralized manner accessible over the internet, users can interact with their data and applications from anywhere in the world, from any device.
+Thanks to economies of scale and virtualization technologies, the Cloud is highly scalable;
+services can be scaled simply by spawning more \glspl{VM}.
+The specialized nature of Cloud providers, the scale of modern datacenters, and the use of virtualization also make the Cloud highly reliable.
+When hardware fails, recovering is simply a matter of migrating the service container or \gls{VM} to an available compute node~\cite{endo2016high}.
+
+However, the Cloud is not suitable for everything, and presents important drawbacks and challenges for latency-sensitive and/or bandwidth-intensive applications.
+In order to achieve the necessary economies of scale, Cloud datacenters are designed to serve users distributed across vast geographical areas.
+These installations are thus often located ``far'' from potential users;
+for instance, at the time of writing, \gls{AWS} processes traffic from all of Scandinavia and the Baltic countries through a single datacenter in Stockholm~\cite{awsregions}.
+This leads to prohibitively high latencies for both highly interactive immersive applications such as mobile \gls{XR} and for \glspl{CPS} and \glspl{NCS}~\cite{tolia2006quantifying,lagar2007interactive,satyanarayanan2009case,varghese2016challenges,shi2016promise}.
+The former category requires \emph{motion-to-photon} latencies (i.e.\ time between input capture and feedback) below \SI{60}{\milli\second} for interactions to be perceived as fluid and responsive by the user~\cite{chen2017empirical}; the latter can require sub-\SI{10}{\milli\second} latencies, for instance in the case of vehicular safety systems.
+Such latencies are unfeasible to consistently achieve with Cloud computing~\cite{dang2021cloudy}.
+On the other hand, as smart devices, appliances, and sensors become more and more ubiquitous, the network architectures of modern datacenters face increasing challenges to deal with the massively increasing volume of traffic~\cite{shi2016edge,wang2019towards}.
 
@@
 The foundation for this concept was laid by \citeauthor{satyanarayanan2009case}~\cite{satyanarayanan2009case} in\ \citeyear{satyanarayanan2009case}.
-To tackle the prohibitively high latencies of cloud computing, the authors proposed an extension of the existing cloud computing paradigm with compute nodes at the edge of the Internet.
+To tackle the prohibitively high latencies of Cloud computing, the authors proposed an extension of the existing Cloud computing paradigm with compute nodes at the edge of the Internet.
 Instead of offloading computation to a datacenter potentially thousands of kilometers away, in Edge Computing it is offloaded to a compute node at the \emph{edge} of the network close to the user, see \cref{fig:edgecomputing}.
@@
 
-The original Edge Computing concept as introduced in~\citeauthor{satyanarayanan2009case} has come to be known as \emph{Cloudlet-based} Edge Computing.
+\input{tables/cloud-vs-edge}
+
+\Cref{tab:cloud-vs-edge} shows an overview of the differences between Cloud Computing and three implementations of Edge Computing, \emph{Cloudlet}-based Computing, \gls{MEC}, and \emph{Fog} Computing.
+The first of these corresponds to the original Edge Computing concept as introduced in~\citeauthor{satyanarayanan2009case}.
 It is based on the idea that Edge compute capabilities will be realized through geographically distributed micro-datacenters known as \emph{cloudlets}.
-Cloudlets have been envisioned as featuring most of the key characteristics of cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user.
+Cloudlets have been envisioned as featuring most of the key characteristics of Cloud datacenters, such as multi-tenancy, virtualization of computing resources, virtually unrestricted access to energy, as well as limited scalability, all while being one or two hops away from the user.
 This design builds on previous work on \emph{cyber foraging}~\cite{noble1997agile,flinn1999energy,satyanarayanan2001pervasive}.
-Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the cloud).
+Cyber foraging refers to the extension and amplification of the capabilities of mobile and wearable devices by offloading computation and data manipulation to nearby infrastructure (as opposed to distant infrastructure such as the Cloud).
 
+\glsreset{MEC}%
 In later years, an extension to Edge Computing denoted \gls{MEC} has emerged, which pairs Edge Computing with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
@@
 Two examples of such applications, of relevance for the present work, are \acl{WCA}~\cite{ha2014towards,chen2018application,wang2020scaling,chen2017empirical,chen2018application} and \aclp{NCS}~\cite{sasaki2016vehicle,wang2018bandwidth,wan2020efficient}.
-These applications had been hitherto inviable to implement due to the limitations of cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
+These applications had been hitherto inviable to implement due to the limitations of Cloud offloading in mobile and wearable contexts, but have now become a practical reality thanks to \gls{MEC}.
 
-Finally, another variant of Edge Computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the cloud and the edge for the reduction of bandwidth requirements.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
-Edge Computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to cloud services~\cite{satyanarayanan2017emergence}.
+Finally, another variant of Edge Computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the Cloud and the edge for the reduction of bandwidth requirements.
+By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on Cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
+Edge Computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to Cloud services~\cite{satyanarayanan2017emergence}.
 
@@
 \glspl{NCS} have become increasingly popular in recent years due to advances in communication and control technologies, as well as the increasing demand for real-time, remote, and distributed control.
-They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the cloud for centralized, distributed control in what has been called \emph{cloud control systems}~\cite{xia2015cloud}
-However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms simply cannot meet~\cite{wan2020efficient}.
+They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the Cloud for centralized, distributed control in what has been called \emph{Cloud control systems}~\cite{xia2015cloud}
+However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional Cloud paradigms simply cannot meet~\cite{wan2020efficient}.
 This has led their adoption to be limited mostly to industrial environments.

### 2023-03-28 body/part1/chapter1-introduction/tables/cloud-vs-edge.tex
cloud vs edge table

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
% \usepackage{graphicx}
\begin{table}[]
    \centering
    \caption{Comparison between Cloud Computing and three implementations of Edge Computing.}\label{tab:cloud-vs-edge}
    \tiny
    \renewcommand{\arraystretch}{1.5}
    \begin{tabularx}{\textwidth}{@{}lYYYY@{}}
        \toprule
        &
        \textbf{Cloud} &
        \textbf{Cloudlet} &
        \textbf{\acs{MEC}} &
        \textbf{Fog} \\ \midrule
        \textbf{Location} &
        Data centers that are geographically distant from end-users. &
        Micro-datacenters close to end-users. &
        Integrated within the \acs{RAN}. &
        Distributed across the Edge-Cloud network continuum. \\
        \textbf{Latency} &
        \ensuremath{> \SI{50}{\milli\second}} &
        \ensuremath{\leq \SI{50}{\milli\second}} &
        \ensuremath{\leq \SI{10}{\milli\second}} &
        \ensuremath{\leq \SI{50}{\milli\second}} \\
        \textbf{Resources} &
        Limited only by network bandwidth and latency. &
        \multicolumn{3}{%
            >{\hsize=\dimexpr3\hsize+4\tabcolsep+2\arrayrulewidth\relax}Y%
        }{Limited, but orders of magnitude more than end-user devices.} \\
        \textbf{Security} &
        Relies on strong network and perimeter defenses. &
        \multicolumn{3}{%
            >{\hsize=\dimexpr3\hsize+4\tabcolsep+2\arrayrulewidth\relax}Y%
        }{Data can be kept close to the source and encrypted.} \\
        \textbf{Privacy} &
        Data is aggregated and stored centrally. &
        \multicolumn{3}{%
            >{\hsize=\dimexpr3\hsize+4\tabcolsep+2\arrayrulewidth\relax}Y%
        }{Data can be kept locally, reducing the risk of exposure.} \\
        \textbf{Scalability} &
        Easily scalable by adding more resources to datacenters. &
        \multicolumn{3}{%
            >{\hsize=\dimexpr3\hsize+4\tabcolsep+2\arrayrulewidth\relax}Y%
        }{Limited by virtue of Edge nodes being distributed and energy-constrained in comparison to datacenters.} \\
        \textbf{Use~Cases} &
        Large-scale systems that are not latency-sensitive or do not require real-time processing. &
        Latency-sensitive applications such as \acs{CPS} and \acs{AR}. &
        Latency-sensitive applications in mobile and wearable contexts. &
        \acs{IOT} \\ \bottomrule
    \end{tabularx}%
\end{table}

### 2023-03-28 body/part1/chapter1-introduction/background.tex
fixes

@@
 \glsreset{MEC}%
+\glsreset{RAN}%
 In later years, an extension to Edge Computing denoted \gls{MEC} has emerged, which pairs Edge Computing with novel mobile networking standards such as 5G~\cite{hassan2019edge,pham2020survey,wan2020efficient}.
@@
 
-Finally, another variant of Edge Computing called \emph{fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the Cloud and the edge for the reduction of bandwidth requirements.
-By aggregating, transforming, and filtering data at multiple levels in the network, fog computing aims to reduce the load placed on Cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
+Finally, another variant of Edge Computing called \emph{Fog} computing, introduced by \citeauthor{bonomi2012fog}~\cite{bonomi2012fog} in\ \citeyear{bonomi2012fog}, concerns itself specifically with the distribution of computing power between the Cloud and the edge for the reduction of bandwidth requirements.
+By aggregating, transforming, and filtering data at multiple levels in the network, Fog computing aims to reduce the load placed on Cloud datacenters by massive \gls{IOT} sensor networks~\cite{yi2015survey}.
 Edge Computing also presents opportunities for increased data security, privacy, and integrity, by keeping it geographically close to its origin and by allowing for anonymization through denaturing and local aggregation before offloading to Cloud services~\cite{satyanarayanan2017emergence}.

### 2023-03-29 body/part1/chapter1-introduction/background.tex
xr

@@
 
-\subsection{Emulation and simulation}
+\subsection{\glsfmtlong{XR}}\label{background:xr}
+\todo[inline]{Needs citations.}
+\glsreset{XR}
+\glsreset{AR}
+\glsreset{VR}
+\glsreset{MR}
+
+\gls{XR} is an umbrella term used to refer to all immersive technologies that combine real and virtual environments, including technologies such as \gls{AR}, \gls{VR}, and \gls{MR}.
+\emph{Immersiveness} refers to the extent to which a technology can fully engage a user's senses to create the illusion of being in a different environment or situation.
+A highly immersive experience can make the user feel like they are physically present in a different world or reality.
+
+\gls{AR} merges digital content with the physical world, enhancing human perception and cognition through the overlay of virtual objects and information on top of the real environment.
+This technology can superimpose tooltips, virtual \gls{2D} and \gls{3D} objects, or even full-fledged videos on top of what the user sees.
+These overlays are most often presented to the user through wearable or mobile devices, such as smartphones or smart glasses (e.g. Google Glass).
+To create an \gls{AR} experience, the device's camera captures the real-world environment, and the image feed is processed in real-time in software.
+The digital objects are then superimposed onto the real-world view and displayed on the device's screen, where they can optionally be interacted with by touching the screen or using voice commands.
+Other sensors can also be used to track the user's movements and position, and adjust the digital objects accordingly.
+
+\gls{VR}, on the other hand, takes it a step further and involves forgoing the real world, completely immersing the user in a fully virtual environment.
+Due to its deeply immersive nature, \gls{VR} requires users to wear specialized equipment such as a \gls{VR} headset.
+The headset contains two high-resolution displays (one for each eye) and a set of headphones, and is connected to a computer or a gaming console, which generates the virtual environment in real-time.
+The headset tracks the user's head movements and adjusts the virtual displays accordingly to create the illusion of being in a completely immersed in a different world.
+Objects in the virtual environment can be interacted with either through the use of special hardware such as instrumented gloves or controllers, or by using sensors which track the user's hands.
+Some advanced \gls{VR} platforms even include omnidirectional treadmills which allow the users navigate the virtual world by walking, running, and even jumping.
+
+Finally, \gls{MR} combines elements from both aforementioned technologies.
+Like \gls{VR}, \gls{MR} uses a headset with high-resolution displays and sensors to track the user's movements.
+However, \gls{MR} headsets also contain cameras that capture the real-world environment, in order to overlay digital objects onto the real world as in \gls{AR}.
+These objects are rendered in real-time and projected onto the real-world view in the user's headset in a seamless and realistic way.
+In this way, \gls{MR} creates a hybrid environment where users can interact with both virtual and real-world elements simultaneously, resulting in an illusion of natural interaction between users and digital elements.
+
+\input{tables/extendedreality.tex}
+
+A comparison of select characteristics of \gls{VR}, \gls{AR}, and \gls{MR} is presented in \cref{tab:xrcomparison}.
+A key factor in which these applications differ is in their respective \gls{QoS} requirements.
+Out of the three, \gls{AR} has the least stringent requirements.
+\gls{AR} in particular does not require a high bandwidth, as only individual digital objects are generated.
+However, low latencies are still a requirement, in order to ensure that the superimposition of the digital objects on top of the real-word view is accurate and in real-time.
+In addition, \gls{AR} requires a high level of accuracy in tracking the user's movements and the position of the device's camera, as inaccuracies can lead to misalignment between the real-world view and the digital objects, leading to degraded \gls{QoE}.
+
+\gls{VR}, on the other hand, requires a high level of \gls{QoS} in terms of both latency and bandwidth to deliver a smooth, highly immersive, and realistic experience.
+Moreover, delays or jitter in \gls{VR} experiences can lead to motion sickness in users.
+The user's movements and interactions must thus be tracked and rendered accurately in real-time, which requires low latencies between the headset and the computing device.
+A high bandwidth is also necessary to transmit high-resolution graphics and audio back to the headset.
+Given the similarities between the technologies, these same requirements also apply to \gls{MR}.
+
+\medskip
+
+Due to their ability to create immersive experiences for users and closely mimic and simulate real-life scenarios, \gls{XR} systems have found applications across a wide spectrum of industries and fields.
+One of the most common uses of these technologies can be found in the domain of entertainment.
+\gls{VR} allows users to become fully immersed in virtual experiences, including games, concerts, movies, making them feel almost physically present without having to leave the comfort of their homes.
+Further, \gls{AR} and \gls{MR} allow for the merging of the virtual and real worlds, bringing entertainment to life in users day-to-day surroundings.
+The popularity of this approach can be clearly seen in the explosive success of modern mobile \gls{AR} games such \citetitle{pokemongo}~\cite{pokemongo}.
+
+\gls{XR} also presents tremendous opportunities for education and training.
+Through the use of these technologies, students can get hands-on yet safe, simulated experience in a multitude of scenarios.
+\gls{VR} is leveraged, for instance, in the training of new medical professionals, allowing students to learn anatomy and practice surgical procedures in realistic virtual environments.
+On the other hand, \gls{AR} and \gls{MR} are employed in manufacturing, to train and assist technicians in highly specialized assembly lines.
+Workers can receive real-time guidance through \gls{AR}-enabled headsets, improving accuracy and safety while controlling machine operations on the factory floor.
+\gls{AR} can also provide augmented, step-by-step guidance for maintenance and repair of equipment, augmenting thus the capabilities of workers and the productivity of the workplace as a whole.
+
+\subsubsection{\glsfmtlong{WCA}}
+\todo[inline]{
+Then, when it comes to WCA, please also go more into the details regarding the operation, the footprints and the QoS requirements.
+Alternatively, you can also defer the discussion of WCA entirely to section 2 (you could claim that WCA is simply a type of AR, and thus in section you 'choose' WCA as example application to conduct your studies - in the same way as the inverted pendulum is a type of NCS, that we choose to study further).
+}
+
+Of particular interest for this dissertation is a particular subset of \gls{AR} denoted \acf{WCA}, a technology designed to provide real-time support and guidance to individuals performing complex tasks in a plethora of domains.
+Designed to be deployed on wearable devices, \gls{WCA} overlays digital information onto the real world and provides context-aware feedback to users, in order to assist and guide them in to efficiently complete a task~\cite{ha2014towards,chen2015early,chen2018application,wang2020scaling}.
+
+These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users. 
+The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
+The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video feeds.
+Whenever the assistant detects a change in the context of the user, it provides a new instruction.
+The application otherwise remains silent and out-of-the-way.
+
+Originally motivated by assistive use-cases for individuals with reduced cognition due to illness, injury, or age, \gls{WCA} technology has been evolved to include the enhancement of the cognitive abilities of users in areas such as manufacturing, healthcare, and aviation.
+Among the primary benefits of this technology is that it can help reduce the cognitive load on individuals performing complex tasks.
+By providing real-time guidance and support, it can improve accuracy and speed of task completion, reduce errors and minimize the risk of accidents.
+For example, a \gls{WCA} could help a technician performing maintenance on an aircraft by displaying relevant technical information about the aircraft parts and step-by-step instructions on how to perform complex specific repairs.
+
+\gls{WCA} technology employs sensors, cameras, and other hardware in wearable devices to collect data from the environment and the user.
+This data is then processed to interpret the data and provide relevant feedback.
+As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
+The context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
+
+Overall, \gls{WCA} is a promising technology that has the potential to revolutionize the way we perform complex tasks.
+As the technology continues to evolve and become more sophisticated, it is expected to have a significant impact on many industries and improve the safety and efficiency of numerous processes.
+
+\subsubsection{Emulation and simulation for \acs{XR}}
 \todo[inline]{
@@
 
-\subsubsection{Emulation as a research methodology}
-\todo[inline]{Remove?}
-
-\subsection{\glsfmtlong{XR}}\label{background:xr}
-\glsreset{XR}
-\glsreset{AR}
-\glsreset{VR}
-\glsreset{MR}
-\todo[inline]{Add some references}
-\todo[inline]{Discuss tethered versus untethered}
-
-\gls{XR} is an umbrella term used to refer to all immersive technologies that combine real and virtual environments, including technologies such as \gls{AR}, \gls{VR}, and \gls{MR}.
-\gls{AR} merges digital content with the physical world, enhancing human perception and cognition through the overlay of virtual objects and information on top of the real environment.
-This technology can superimpose tooltips, virtual \gls{2D} and \gls{3D} objects, or even full-fledged videos on top of what the user sees.
-These overlays are most often presented to the user through wearable or mobile devices, such as smartphones or smart glasses (e.g.
-Google Glass), fitted with a camera to capture a real-time video feed of the physical environment and a display on which the augmented feed is presented.
-
-\gls{VR}, on the other hand, takes it a step further and involves fully forgoing the real world, immersing the user in a fully digital, \emph{simulated} environment.
-Due to its deeply immersive nature, \gls{VR} requires hardware components specifically designed for this task, the most crucial of which is the \gls{VR} headset.
-The simplest type of such headset uses a pair of lenses to present an illusion of depth when looking through them at a small display.
-More advanced types combine microdisplays, accelerometers, and headphones to provide a comprehensively immersive experience, where the user truly feels as if they are inside the virtual world.
-Additionally, \gls{VR} systems often allow users to interact with the simulated environment through the use of instrumented gloves or special controllers, further increasing the illusion of immersion.
-
-Finally, \gls{MR} combines elements from both aforementioned technologies.
-As in \gls{AR}, \gls{MR} overlays virtual objects on the physical, real-world environment, while also involving the tools and techniques from \gls{VR} which allow the user to manipulate these.
-In this way, \gls{MR} creates a hybrid environment where users can interact with both virtual and real-world elements simultaneously, resulting in an illusion of natural interaction between users and digital elements.
-
-\medskip
-
-Due to their ability to create immersive experiences for users and closely mimic and simulate real-life scenarios, \gls{XR} systems have found applications across a wide spectrum of industries and fields.
-One of the most common uses of these technologies can be found in the domain of entertainment.
-\gls{VR} allows users to become fully immersed in virtual experiences, including games, concerts, movies, making them feel almost physically present without having to leave the comfort of their homes.
-Further, \gls{AR} and \gls{MR} allow for the merging of the virtual and real worlds, bringing entertainment to life in users day-to-day surroundings.
-The popularity of this approach can be clearly seen in the explosive success of modern mobile \gls{AR} games such \citetitle{pokemongo}~\cite{pokemongo}.
-
-\gls{XR} also presents tremendous opportunities for education and training.
-Through the use of these technologies, students can get hands-on yet safe, simulated experience in a multitude of scenarios.
-\gls{VR} is leveraged, for instance, in the training of new medical professionals, allowing students to learn anatomy and practice surgical procedures in realistic virtual environments.
-On the other hand, \gls{AR} and \gls{MR} are employed in manufacturing, to train and assist technicians in highly specialized assembly lines.
-Workers can receive real-time guidance through \gls{AR}-enabled headsets, improving accuracy and safety while controlling machine operations on the factory floor.
-\gls{AR} can also provide augmented, step-by-step guidance for maintenance and repair of equipment, augmenting thus the capabilities of workers and the productivity of the workplace as a whole.
-
-\subsubsection{\glsfmtlong{WCA}}
-
-Of particular interest for this dissertation is a particular subset of \gls{AR} denoted \acf{WCA}, a technology designed to provide real-time support and guidance to individuals performing complex tasks in a plethora of domains.
-Designed to be deployed on wearable devices, \gls{WCA} overlays digital information onto the real world and provides context-aware feedback to users, in order to assist and guide them in to efficiently complete a task~\cite{ha2014towards,chen2015early,chen2018application,wang2020scaling}.
-
-These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users. 
-The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
-The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video feeds.
-Whenever the assistant detects a change in the context of the user, it provides a new instruction.
-The application otherwise remains silent and out-of-the-way.
-
-Originally motivated by assistive use-cases for individuals with reduced cognition due to illness, injury, or age, \gls{WCA} technology has been evolved to include the enhancement of the cognitive abilities of users in areas such as manufacturing, healthcare, and aviation.
-Among the primary benefits of this technology is that it can help reduce the cognitive load on individuals performing complex tasks.
-By providing real-time guidance and support, it can improve accuracy and speed of task completion, reduce errors and minimize the risk of accidents.
-For example, a \gls{WCA} could help a technician performing maintenance on an aircraft by displaying relevant technical information about the aircraft parts and step-by-step instructions on how to perform complex specific repairs.
-
-\gls{WCA} technology employs sensors, cameras, and other hardware in wearable devices to collect data from the environment and the user.
-This data is then processed to interpret the data and provide relevant feedback.
-As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
-The context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
-
-Overall, \gls{WCA} is a promising technology that has the potential to revolutionize the way we perform complex tasks.
-As the technology continues to evolve and become more sophisticated, it is expected to have a significant impact on many industries and improve the safety and efficiency of numerous processes.
-
 \subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
 \glsreset{NCS}
+
+\todo[inline]{
+    - Section 1.1.4: Goes again in the right direction, but details are missing/off.
+First of all, NCS is really a research direction, not necessarily a system type (every control system is more or less networked, but networking effects have not been always captured in the theoretical analysis/design of the controller).
+Second, maybe you stress more the working principle in terms workloads and requirements.
+In a sense, you can work out a pattern how you structure the discussion in Section 1.1.3, and reuse that pattern to a large extent in 1.1.4, that would be neat (both are closed-loop applications with a lot of similar, but also some different properties and requirements).
+}
 

### 2023-03-29 body/part1/chapter1-introduction/tables/extendedreality.tex
xr

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
    \centering
    \caption{Comparison between different \acs{XR} technologies.}
    \label{tab:xrcomparison}
    \tiny
    \renewcommand{\arraystretch}{1.5}
    \begin{tabularx}{\textwidth}{@{}lYYY@{}}
        \toprule
        &
        \textbf{\acl{VR}} &
        \textbf{\acl{AR}} &
        \textbf{\acl{MR}} \\ \midrule
        % \textbf{Definition} &
        % A technology that creates a completely digital environment that simulates a real-world experience &
        % A technology that superimposes digital objects on the real world &
        % A technology that combines digital objects with the real world in a seamless way \\
        \textbf{Immersiveness} &
        High &
        Medium &
        Medium to high \\
        \textbf{User Interaction} &
        Only with the digital world &
        Both with the real and digital worlds &
        Both with the real and digital worlds in a seamless way \\
        \textbf{Equipment Required} &
        Headset and controllers &
        Smartphones, tablets, or \acs{AR} glasses &
        Headset and controllers \\
        % \textbf{Use Cases} &
        % Gaming, education, and training &
        % Retail, entertainment, education, and gaming &
        % Industrial design, engineering, and military training \\
        \textbf{Advantages} &
        Provides realistic immersive experience in a completely digital environment that can be easily adapted and changed. &
        Can be implemented cheaply on massively-available technology such as smartphones. &
        Provides a combination of AR and VR benefits, allowing seamless interaction between real and virtual worlds. \\
        \textbf{Drawbacks} &
        Isolates users from the real world, may cause motion sickness, and requires expensive equipment. &
        Limited field of view, reliance on external environment, and not fully immersive. &
        Expensive, complex, and may cause visual discomfort \\
        \textbf{\acs{QoS} Requirements} &
        High bandwidth, low latencies.
        Delay and jitter can lead to motion sickness. &
        Low latencies to ensure accurate, real-time overlay of digital assets. &
        High bandwidth, low latencies.
        Delay and jitter can lead to motion sickness. \\
        \bottomrule
    \end{tabularx}
\end{table}

### 2023-03-29 body/part1/chapter1-introduction/background.tex
wca

@@
 \subsubsection{\glsfmtlong{WCA}}
-\todo[inline]{
-Then, when it comes to WCA, please also go more into the details regarding the operation, the footprints and the QoS requirements.
-Alternatively, you can also defer the discussion of WCA entirely to section 2 (you could claim that WCA is simply a type of AR, and thus in section you 'choose' WCA as example application to conduct your studies - in the same way as the inverted pendulum is a type of NCS, that we choose to study further).
-}
+\glsreset{WCA}
 
@@
 
-These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users. 
-The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
-The application follows the progress of the task in ``realtime'' by repeatedly sampling the state of the physical system, most commonly through video feeds.
-Whenever the assistant detects a change in the context of the user, it provides a new instruction.
-The application otherwise remains silent and out-of-the-way.
-
-Originally motivated by assistive use-cases for individuals with reduced cognition due to illness, injury, or age, \gls{WCA} technology has been evolved to include the enhancement of the cognitive abilities of users in areas such as manufacturing, healthcare, and aviation.
+Originally motivated by assistive use-cases for individuals with reduced cognition due to illness, injury, or age, \gls{WCA} technology has evolved to include the enhancement of the cognitive abilities of users in areas such as manufacturing, healthcare, and aviation.
 Among the primary benefits of this technology is that it can help reduce the cognitive load on individuals performing complex tasks.
@@
 For example, a \gls{WCA} could help a technician performing maintenance on an aircraft by displaying relevant technical information about the aircraft parts and step-by-step instructions on how to perform complex specific repairs.
+\gls{WCA} is a promising technology that has the potential to revolutionize the way we perform complex tasks.
+As the technology continues to evolve and become more sophisticated, it is expected to have a significant impact on many industries and improve the safety and efficiency of numerous processes.
 
-\gls{WCA} technology employs sensors, cameras, and other hardware in wearable devices to collect data from the environment and the user.
-This data is then processed to interpret the data and provide relevant feedback.
-As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
-The context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
+\medskip
 
-Overall, \gls{WCA} is a promising technology that has the potential to revolutionize the way we perform complex tasks.
-As the technology continues to evolve and become more sophisticated, it is expected to have a significant impact on many industries and improve the safety and efficiency of numerous processes.
+\begin{figure}
+    \centering
+    \includegraphics[width=.9\textwidth]{figures/wca_state}
+    \caption{Mode of operation of step-based \acs{WCA} applications.}\label{fig:wca:operation}
+\end{figure}
+
+In this dissertation, we focus specifically on \emph{step-based} \gls{WCA}.
+These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users, illustrated in \cref{fig:wca:operation}.
+The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
+The application follows the progress of the task in real-time by repeatedly sampling the state of the physical system, most commonly through video feeds.
+Whenever the assistant detects a change in the context of the user, it provides a new instruction.
+The application otherwise remains silent and out-of-the-way; samples with unfinished states or noise are silently discarded.
+
+This mode of operation has novel consequences for the \gls{QoS} requirements of \gls{WCA} applications.
+As with general \gls{AR}, \gls{WCA} has relatively lax bandwidth requirements; however, latency requirements in \gls{WCA} can vary along the execution time of the application.
+Users of these applications can only notice system responsiveness when they are \emph{expecting} feedback, that is, immediately after the completion of a logical step or instruction.
+Latency bounds can thus be very lax immediately after an instruction is generated (as no feedback is expected yet), and on the other hand much more stringent right after the user has completed an instruction.
+In turn, this leads to interesting opportunities for resource consumption optimization in \gls{WCA} applications, such as the adaptive sampling scheme proposed in~\cite{wang2019towards}.
+The fact that users \emph{consciously} expect near-instantaneous feedback immediately after finishing an instruction also leads to strong and cumulative reactions to delays in the system, as evidenced by our work in \cref{paper:olguinmunoz2021impact}, implying that careful consideration to latencies must be taken when designing these systems.
+
+% As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
+% This context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
 

### 2023-03-30 body/part1/chapter1-introduction/introduction.tex
intro abstract

@@
 
-% \subsection{Structure of this dissertation}
+We validate our approach by example, exploring an initial applications of it to \gls{NCS}, and a deeper, more thorough application of it to \gls{WCA}.
+For the latter, we develop the first-ever model for emulating human timing behavior \gls{WCA}, created through a comprehensive human-subject study that characterizes these behaviors.
+We use our methodology and this model to explore the optimization potential of \gls{WCA} deployments.
+Specifically, they investigate the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior, and conclude that significant improvements can be achieved by incorporating our model.
+Overall, our results showcase the utility of our methodology, and the importance of considering human behavior in the design and optimization of \gls{WCA} systems to achieve maximum benefits.
 

### 2023-03-30 body/part1/chapter1-introduction/introduction.tex
reorg

--- 2023-03-27 body/part1/chapter1-introduction/introduction.tex
@@
 
+We validate our proposed methodology by presenting case studies of its application in two different contexts: \gls{WCA} and \glspl{NCS}.
+The case studies demonstrate the efficacy of the methodology in evaluating system performance and overall \gls{QoS} and \gls{QoE}.
 Our methodology provides a comprehensive and realistic assessment of the performance of latency-sensitive applications deployed on Edge infrastructure.
@@
 
-% \subsection{Structure of this dissertation}
+In addition to the methodology, this dissertation introduces a number of further contributions to the field. 
+We introduce a software framework that facilitates the orchestration of Edge Computing testbeds needed for the implementation of our methodology.
+This framework makes it easier for researchers and practitioners to deploy and test Edge-bound low-latency applications.
+We also present a deep exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
+This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
 
-\bigskip
-
-This dissertation is structured into two parts.
-\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
-Next, in \cref{part:publications} we present the publications that form the core of this thesis.
-
-\cref{part:summary} is itself structured into four main chapters.
-\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
-It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
-In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
-The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
-We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
-\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
-We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
-We aim to demonstrate the significance of our work and how it contributes to the field.
-Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
-We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
-This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.
+Finally, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
+In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
+We conclude that significant improvements can be achieved through the use of our human user model.
+Overall, our results showcase the utility of our methodology, and the importance of considering human behavior in the design and optimization of \gls{WCA} systems to achieve maximum benefits.

### 2023-03-30 body/part1/chapter1-introduction/structure.tex
reorg

--- 2023-02-16 body/part1/introduction.tex
@@
-\section{Introduction}
+\section{Structure of this dissertation}
 
-The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
-One of the most promising of these is Edge Computing, which seeks to bring computing closer to the end user, reducing latency and increasing reliability.
-Another key player in this space is 5G, the latest generation of mobile communication networks that offers higher data rates, lower latency, and improved reliability.
-The combination of these two technologies holds great promise for the deployment of latency-sensitive applications, including \glspl{CPS}, \glspl{NCS}, and \gls{WCA} applications, among others.
+This dissertation is structured into two parts.
+\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
+Next, in \cref{part:publications} we present the publications that form the core of this thesis.
 
-Despite these advancements, there are still significant challenges in understanding and scaling these systems, particularly in regard to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
-The aim of this thesis is to make a contribution to the field by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
-The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
-Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of \gls{MEC}.
+\cref{part:summary} is itself structured into four main chapters.
+\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
+It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
+In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
+The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
+We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
+\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
+We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
+We aim to demonstrate the significance of our work and how it contributes to the field.
+Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
+We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
+This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.
 
-Our methodology is based on the emulation of target workloads on actual Edge infrastructure.
-We replace the client side of the system with a realistic emulation of the desired behaviors, implemented in software deployed on \gls{COTS} general-purpose computing devices.
-In our initial implementation, these correspond to low-cost, easily replaceable, and scalable Raspberry Pi 4 Model B \glspl{SBC}.
-
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, COTS general-purpose hardware such as \glspl{SBC}.
-It also preserves the realism of effects stemming from the hardware and network.
-In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
-
-The methodology also provides improved repeatability and replicability.
-Repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
-These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
-
-Our methodology provides a comprehensive and realistic assessment of the performance of latency-sensitive applications deployed on Edge infrastructure.
-The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
-
-\section{Summary of the key contributions of this thesis}\label{sec:summary_contributions}
-
-This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
-Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound latency-sensitive applications such as \gls{WCA} and \gls{NCS}.
-This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
-We validate this methodology by example, presenting case studies of its application in the context of \gls{WCA} and \glspl{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.
-
-Secondly, we present a deeper exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
-This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
-
-Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
-In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
-We conclude that significant improvements can be achieved through the use of our human user model.
-
-\subsection{Overview of each included paper}
+\subsection{Overview of the included publications}
 

### 2023-03-30 recyclebin/contributions.tex
reorg

--- 2023-03-07 body/part1/chapter2-relatedwork/contributions.tex
@@
 \todo[inline]{Challenges and how we tackled them.}
+
+\subsection{Primary contributions}
+
+\subsection{Secondary contributions}
 
@@
 We conclude that significant improvements can be achieved through the use of our human user model.
-
-\section{Structure of this dissertation}
-
-This dissertation is structured into two parts.
-\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
-Next, in \cref{part:publications} we present the publications that form the core of this thesis.
-
-\cref{part:summary} is further structured into four main chapters.
-\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
-It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance.
-
-In \cref{chap:relatedwork}, we present theoretical background and related work.
-The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
-We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
-
-\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
-We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
-We aim to demonstrate the significance of our work and how it contributes to the field.
-
-Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
-We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
-This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-04-01 body/part1/chapter2-relatedwork/relatedwork.tex
wip
relwork

@@
+\todo[inline]{Still needs some work, don't review yet.}
+
 \section{Related work}
@@
 A rich body of literature exists with examples of such approaches for Cloud computing.
-For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and cloud services.
-Workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
-Both these approaches focus heavily on data processing and downlink transmission in cloud computing scenarios.
+However, existing approaches tend to focus heavily on large-scale data processing and downlink transmission in Cloud scenarios, and are not applicable to the latency-sensitive, uplink-heavy characteristics of Edge Computing.
+For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
+In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
+The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
+\cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major Cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
 
-In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the cloud.
-The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
-\cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
-
-A number of works have tackled this challenge in the context of Fog and Edge Computing as well.
+The limitations of previous approaches has in later years led to a number of works that have attempted to tackle this challenge in the context of Fog and Edge Computing as well.
+Even so, these works have generally leveraged static workloads that fail to account for the intensely dynamic nature of \gls{CPS} deployments.
+Furthermore, none of these approaches have attempted to understand the relationship between latency and human performance in \gls{XR} applications deployed on the Edge.
 \emph{EdgeBench}, introduced in~\cite{das2018edgebench}, corresponds to a benchmarking suite for the Edge.
@@
 \cite{baurle2022comb} presents \emph{ComB}, a flexible and extensible, application-oriented benchmarking suite for Edge computing.
-Its stated purpose is to assist system designers and application developers to evaluate the suitability of Edge deployments.
-The framework is designed to decouple orchestration and workload, allowing for easy extension or replacement of the built-in distributed video-analytics benchmark.
-
-\emph{IoTBench}~\cite{lee2019iotbench} is a benchmarking tool for edge-deployed \gls{IOT} applications.
+It includes a built-in distributed video-analytics benchmark, but is designed to decouple orchestration and workload, allowing for easy extension of the latter.\@
+\emph{IoTBench}~\cite{lee2019iotbench} is a benchmarking tool for Edge-deployed \gls{IOT} applications.
 It includes seven representative workloads, and allows researchers to evaluate the performance of these applications on Edge infrastructure through the analysis of computational demands and resource consumption.\@
@@
 \emph{DeFog} includes six latency-critical and bandwidth-intensive, Fog- and Edge-relevant workloads, including real-time video detection and a text-to-speech application.
-In~\cite{lockhart2020scission}, the authors introduce a tool, \emph{Scission}, for the benchmarking of \glspl{DNN} partitioned across the user device, edge, and cloud continuum.
+In~\cite{lockhart2020scission}, the authors introduce a tool, \emph{Scission}, for the benchmarking of \glspl{DNN} partitioned across the user device, Edge, and Cloud continuum.
 The tool can also be used to automatically determine the optimal partition of a \gls{DNN}.
@@
 \gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
-Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
+Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local Edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
 The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling Edge Computing platform.
 
-The Chameleon testbed is a research platform originally designed for the study and evaluation of cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
-Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various cloud and Edge Computing scenarios.
+The Chameleon testbed is a research platform originally designed for the study and evaluation of Cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
+Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various Cloud computing technologies, algorithms, and configurations.
+The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various Cloud and Edge Computing scenarios.
 
@@
 The testbeds main \emph{raison d'être} is \gls{MEC} research, with a strong emphasis on wireless networks.
-The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
+The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, Cloud integration, and Edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
 
@@
 It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
-These radio nodes are combined with edge-based compute nodes as well as cloud resources for full \gls{MEC} experimentation.
+These radio nodes are combined with Edge-based compute nodes as well as Cloud resources for full \gls{MEC} experimentation.
 
@@
 The EdgeNet software is available for anyone to download;
-individuals and organizations can quickly deploy self-hosted edge nodes which are then made available to anyone using the platform.
-The end result is a flexible and scalable global edge cloud simplifying research and prototype development in edge and cloud computing.
-A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the cloud to the edge.
+individuals and organizations can quickly deploy self-hosted Edge nodes which are then made available to anyone using the platform.
+The end result is a flexible and scalable global Edge Cloud simplifying research and prototype development in Edge and Cloud computing.
+A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the Cloud to the Edge.
 
 Finally, a number of other, small-scale approaches also exist.
-In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for Edge Computing research.
-The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered edge architectures in the context of \gls{FIWI} Edge Computing.
+In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for Edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for Edge Computing research.
+The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered Edge architectures in the context of \gls{FIWI} Edge Computing.
 This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
 In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
-Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
+Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
 The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.

### 2023-04-01 body/part1/chapter2-relatedwork/relatedwork.tex
relwork!!!!

@@
 
-\section{Related work}
+% \section{Related work}
+\phantomsection%
 
@@
 In this, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
-The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.
-
+The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.\@
 \cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{AR} application by mocking Android \gls{SDK} \glspl{API}.
@@
 A similar approach was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
-
 Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
 Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
-This approach is similar to the one presented in this dissertation, but it exclusively considers on-device \gls{AR} processing.
-Furthermore, the benchmarks are deterministic, and no emulation of realistic human behavior is attempted.
+
+While the above approaches to benchmarking and research in \gls{XR} are steps in the right direction, they fail to consider the dynamic nature of \gls{XR}.
+The use of rigid traces in trace-based approaches can lead to timing mismatches when replaying at runtime, as the system may respond differently to the same input depending on factors such as system load and network congestion.
+Additionally, these approaches do not take into consideration the effects of system responsiveness on human behavior, which can be critical for evaluating both \gls{QoS} and \gls{QoE}.
+To address these issues, future approaches will need to adapt trace replay at runtime to match expected behaviors and consider the impact of system responsiveness on user behavior.
+By doing so, we can develop more accurate and reliable benchmarks for \gls{XR} applications and improve our understanding of how users interact with these technologies.
 
@@
 A rich body of literature exists with examples of such approaches for Cloud computing.
-However, existing approaches tend to focus heavily on large-scale data processing and downlink transmission in Cloud scenarios, and are not applicable to the latency-sensitive, uplink-heavy characteristics of Edge Computing.
 For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
@@
 
-The limitations of previous approaches has in later years led to a number of works that have attempted to tackle this challenge in the context of Fog and Edge Computing as well.
+As evidenced by the given examples, existing approaches to Cloud benchmarking tend to focus heavily on large-scale data processing and downlink transmission in Cloud scenarios.
+This makes them unsuitable for Edge Computing, as they fail to account for important characteristics of this paradigm, such as latency-sensitivity of workloads and uplink-heavy network traffic.
+
+The limitations of previous approaches has in later years led to a number of works that have attempted to tackle the challenge of specifically benchmarking Fog and Edge deployments as well.
 Even so, these works have generally leveraged static workloads that fail to account for the intensely dynamic nature of \gls{CPS} deployments.
-Furthermore, none of these approaches have attempted to understand the relationship between latency and human performance in \gls{XR} applications deployed on the Edge.
-\emph{EdgeBench}, introduced in~\cite{das2018edgebench}, corresponds to a benchmarking suite for the Edge.
-It emulates three different workloads, speech-to-text, image recognition, and sensor network application traffic, and specifically targets serverless deployments.\@
-\cite{baurle2022comb} presents \emph{ComB}, a flexible and extensible, application-oriented benchmarking suite for Edge computing.
-It includes a built-in distributed video-analytics benchmark, but is designed to decouple orchestration and workload, allowing for easy extension of the latter.\@
+Examples of approaches which fall into this category include
+\begin{inlineenum}
+    \item \emph{EdgeBench}~\cite{das2018edgebench}
+    \item \emph{IoTBench}~\cite{lee2019iotbench}
+    \item \emph{DeFog}~\cite{mcchesney2019defog}
+    \item \emph{ComB}~\cite{baurle2022comb}
+    \item \emph{OpenRTiST}~\cite{george2020openrtist}
+\end{inlineenum}.
+\emph{EdgeBench}~\cite{das2018edgebench} corresponds to a benchmarking suite for the Edge which emulates three different workloads (speech-to-text, image recognition, and sensor network application traffic), specifically targeting serverless deployments.\@
 \emph{IoTBench}~\cite{lee2019iotbench} is a benchmarking tool for Edge-deployed \gls{IOT} applications.
 It includes seven representative workloads, and allows researchers to evaluate the performance of these applications on Edge infrastructure through the analysis of computational demands and resource consumption.\@
-\cite{mcchesney2019defog} presents a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud named \emph{DeFog}.
-\emph{DeFog} includes six latency-critical and bandwidth-intensive, Fog- and Edge-relevant workloads, including real-time video detection and a text-to-speech application.
-In~\cite{lockhart2020scission}, the authors introduce a tool, \emph{Scission}, for the benchmarking of \glspl{DNN} partitioned across the user device, Edge, and Cloud continuum.
-The tool can also be used to automatically determine the optimal partition of a \gls{DNN}.
+\emph{DeFog}~\cite{mcchesney2019defog} corresponds to a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud which includes six latency-critical and bandwidth-intensive workloads, including real-time video detection and a text-to-speech application.
+\emph{ComB}~\cite{baurle2022comb} is a flexible and extensible application-oriented benchmarking suite for Edge computing, featuring a built-in distributed video-analytics benchmark.
 Finally, \emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
@@
 The transformed video is then fed back to the source to be presented, resulting in a compute-intensive, bandwidth-hungry, and latency-sensitive benchmark.
+
+Furthermore, although some of these approaches address some of the challenges to benchmarking \gls{XR} on the Edge, none of them have tackled the topic in a comprehensive way.\@
+\cite{das2018edgebench,mcchesney2019defog,baurle2022comb,george2020openrtist} all target image or video recognition or manipulation in some manner.
+However, none of these approaches have considered the interactive component of \gls{MAR}, particularly when it comes to human behavior and its relationship to system responsiveness and performance.
 
@@
 However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
-Within this body of work, we identify three distinct approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
+
+Experimental research in \glspl{NCS} involves a diverse range of hardware and software platforms, methodologies, and key performance indicators.
+Specifically, we identify three distinct categories of approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
 
@@
 An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
+
+\medskip
+
+The diversity of experimental \glspl{NCS} research discussed above has led to fragmentation of hardware, software, and methodology, as studies tend to favor approaches popular in their respective communities.
+Additionally, studies tend to focus on individual aspects and components of the system, resulting in incomplete knowledge of the overall \glspl{NCS}.
+There is a gap in knowledge concerning the reproducibility and comparison of experimental studies on these systems.
+
+To our knowledge, the only published attempt at addressing methodological fragmentation and poor reproducibility of experimental \gls{NCS} research is the LEGO-based approach introduced in~\cite{zoppi2020ncsbench}.
+However, this approach has limitations.
+It is dependent on specific hardware --- the LEGO Mindstorms set --- which makes its adoption limited due to availability issues, reduces its flexibility, scalability, and limits the scope of experiments able to be conducted via \emph{NCSbench} only to what is possible to implement in LEGO.\@
+Scaling up experiments becomes cumbersome and costly, further limiting the practicality of the tool.
+Thus, a more general approach that tackles the scalability and reproducibility issues of NCS research is urgently needed.
+The development of such an approach would enable researchers to conduct experiments that generate reproducible results consistently and can be compared across different studies while still retaining the computational sophistication required when scaling up experiments.
 
@@
 In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
-Finally,\ \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.\@
+\cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
 The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
+
+\medskip
+
+The implementation of next-generation wireless systems, Cloud, and Edge computing paradigms requires experimentation to fully understand their implications.
+As evidenced above, testbeds have emerged for this purpose.
+However, there has been little focus on general-purpose, hardware-agnostic software frameworks for the management and automation of these testbeds.
+Most existing works use their own platform-specific solutions, which are not compatible with other testbeds and often do not comply with Cloud and/or Edge-native standards.
+\gls{COSMOS} and \gls{POWDER}, which use domain-specific languages and virtualization technology based on \glspl{VM} instead of modern Edge-compatible solutions such as containers or serverless computing frameworks.
+
+From the examples above, \gls{CLOUDRAFT} is the only work known to tackle the challenge of general-purpose, hardware-agnostic software frameworks for the management and automation of testbeds, albeit only to a certain extent.
+\gls{CLOUDRAFT} integrates, coordinates, shares, and improves upon existing testbeds, using pre-built VMs with necessary software for experiments.
+Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely on the sharing and partitioning of testbed systems.
+Testbeds that are currently working with CloudRAFT include an SDR-based testbed and a ground vehicular robot for mobility-related experimentation.
+
+The lack of a unified approach to testbed management and automation hinders the development and implementation of next-generation wireless systems, Cloud, and Edge computing paradigms.
+General-purpose, hardware-agnostic software frameworks such as \gls{CLOUDRAFT} can simplify the management and automation of testbeds, promote compatibility with Cloud- and Edge-native standards, and facilitate collaboration among research groups.
+There is still remains much work to be done in this area, particularly with respect to the adoption of Edge-compatible solutions that can enable faster and more reliable experimentation.

### 2023-04-03 glossary.tex
scope draft
relwork
scope

--- 2023-02-07 body/glossary.tex
@@
 \newacronym[prefixfirst={a\ }, prefix={an\ }]{NCS}{NCS}{Networked Control System}
-%\acroindefinite{NCS}{an}{a}
-
 \newacronym{LAN}{LAN}{Local-Area Network}
@@
 \newacronym{GENI}{GENI}{Global Environment for Networking Innovation}
-
+\newacronym{SBC}{SBC}{Single-Board Computer}
+\newacronym{EXPECA}{ExPECA}{Experimental Platform for Edge Computing Applications}
+\newacronym[user1={Kungliga Tekniska Högskolan}, prefixfirst={the\ }]{KTH}{KTH}{Royal Institute of Technology of Sweden}
+\newacronym{HCI}{HCI}{Human-Computer Interaction}
+\newacronym{2D}{2-D}{two-dimensional}
+\newacronym{3D}{3-D}{three-dimensional}
+\newacronym{PD}{PD}{Proportional-Differential}
+\newacronym{PID}{PID}{Proportional-Integral-Differential}
+\newacronym{SDK}{SDK}{Software Development Kit}
+\newacronym{YCSB}{YCSB}{Yahoo! Cloud Serving Benchmark}
+\newacronym{ITS}{ITS}{Intelligent Transportation System}
+\newacronym{OOS}{OoS}{Out-of-Scope}
 %\glsaddall%

### 2023-04-03 body/part1/chapter1-introduction/introduction.tex
scope

--- 2023-03-27 body/part1/chapter1-introduction/introduction.tex
@@
+\section{Introduction}
+
 The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
@@
 
-% \subsection{Structure of this dissertation}
+In addition to the methodology, this dissertation introduces a number of further contributions to the field. 
+We introduce a software framework that facilitates the orchestration of Edge Computing testbeds needed for the implementation of our methodology.
+This framework makes it easier for researchers and practitioners to deploy and test Edge-bound low-latency applications.
+We also present a deep exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
+This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
 
-\bigskip
+Finally, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
+In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
+We conclude that significant improvements can be achieved through the use of our human user model.
+Overall, our results showcase the utility of our methodology, and the importance of considering human behavior in the design and optimization of \gls{WCA} systems to achieve maximum benefits.
+
+\subsection{Structure of this dissertation}
 

### 2023-04-03 body/part1/chapter2-relatedwork/relatedwork.tex
scope

@@
-\todo[inline]{Still needs some work, don't review yet.}
-
-% \section{Related work}
-\phantomsection%
+\section{Related work}
 
@@
 Most existing works use their own platform-specific solutions, which are not compatible with other testbeds and often do not comply with Cloud and/or Edge-native standards.
-\gls{COSMOS} and \gls{POWDER}, which use domain-specific languages and virtualization technology based on \glspl{VM} instead of modern Edge-compatible solutions such as containers or serverless computing frameworks.
+Examples of this are \gls{COSMOS} and \gls{POWDER}, which use domain-specific languages and virtualization technology based on \glspl{VM} instead of modern Edge-compatible solutions such as containers or serverless computing frameworks.
 
@@
 Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely on the sharing and partitioning of testbed systems.
-Testbeds that are currently working with CloudRAFT include an SDR-based testbed and a ground vehicular robot for mobility-related experimentation.
+Testbeds that are currently working with \gls{CLOUDRAFT} include an \gls{SDR}-based testbed and a ground vehicular robot for mobility-related experimentation.
 

### 2023-04-03 body/part1/chapter2-relatedwork/scope.tex
more fixes
scope

@@
-\section{Scope of this dissertation}
+\section{Scope \& overview of the included publications}
 
@@
 This methodology is based on the emulation of the desired application workload on top of the actual Edge infrastructure, including backend computing hardware and network technology used for client-server communication.
-We do not in any way claim that this methodology is the best approach to studying these systems, but rather that it is a valid approach able to produce accurate, representative results.
+We do not in any way claim that this methodology is the best approach to studying these systems, but rather that it is a \emph{valid} approach able to produce accurate and representative results.
+
+\input{tables/scope}
 
@@
 By considering these two application domains, we demonstrate the applicability of our methodology and showcase its efficacy.
+\Cref{tab:scope} presents an overview of the scope of this dissertation, and in the following we will summarize the contributions and limitations of each included publication.
 
-Specifically, a subset of the research works included in this dissertation presents and studies the application of our methodology for \gls{WCA}.
-To limit the scope of our research in this area, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
+\medskip
+
+\Cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} in this dissertation present and study the application of our methodology for \gls{WCA}.
+\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} discuss initial explorative studies into the application of our methodology for \gls{WCA}.
+These works present the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research, and further discuss the necessary tooling and software frameworks required for the implementation of our methodology.
+In particular, we introduce in these works a trace-based emulation tool which we call EdgeDroid.
+It uses a trace of inputs recorded in optimal conditions, the replay of which is controlled at run-time by a simple model of human behavior based on a \gls{FSM}.
+In terms of hardware, the setup is composed of Android client devices connected to a \gls{COTS} {x86--64} server over a Wi-Fi network.
+
+In \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deeper into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
+\cref{paper:olguinmunoz2021impact} presents a characterization of human behavior in \gls{WCA}.
+Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
+We develop a characterization through a human subject study in which participants interact with an instrumented \gls{WCA}.
+System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
+Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
+
+\cref{paper:olguinmunoz2023realistic} finally presents two contributions.
+Firstly, it introduces the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
+The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+Secondly, it explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+
+\medskip
+
+This line of work has some limitations.
+First of all, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
 We focus on this category for a couple of reasons.
@@
 These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
-The specific application used for these studies corresponds to the LEGO assistant devised by \citeauthor{chen2015early}\cite{chen2015early}.
 
-A separate subset of the research presented in this work discusses a preliminary foray into the applications of our methodology for \glspl{NCS}.
-Given the vastness and complexity of this field, we limit our exploration to simple, small-scale, single-plant systems.
-In particular, the system used as a representative example for the effectiveness of our approach for \glspl{NCS} corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
+However, restricting our research to step-based \gls{WCA} means that our results may not be directly applicable to other types of \gls{WCA} and \gls{AR} applications.
+In particular, our characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
+Additionally, the timing characteristics of step-based \gls{WCA} are such that an instruction remains valid for as long as it takes the user to perform it.
+The conclusions from these works therefore may not apply to applications with different execution flows, or in which instructions have a deadline after which they may become stale or invalid.
+Examples of such applications include, for instance, the \emph{Ping-Pong} assistant introduced in~\cite{chen2015early}.
 
-Finally, this dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
+The user models are further limited in their realism.
+In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we assume a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+These limitations emerge from a lack of data of human behavior in these applications, as at the moment of the writing of these works no comprehensive model of human behavior existed in the relevant literature.
+As a consequence, these works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
+We attempt to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
+the resulting model is able to adapt its timing behavior in response to delay in the system.
+However, we were unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
+Furthermore, participants for the study were recruited from a pool of undergraduate students at \gls{CMU} in the \gls{USA}, ranging in age from \numrange{18}{25} years old.
+This means that findings may not be generalizable to other demographic groups.
+
+Finally, to emphasize the possible effects of network congestion and latency on \gls{WCA}, in these works we exclusively employ \acs{IEEE} 802.11n \SI{2.4}{\giga\hertz} Wi-Fi networks in an Cloudlet Computing-like setup.
+Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
+Likewise, we do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
+Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate our results and conclusions in the context of \gls{MEC} and Fog Computing.
+
+\bigskip
+
+A separate subset of the research presented in this dissertation discusses a preliminary foray into the applications of our methodology for \glspl{NCS}.
+\Cref{paper:olguinmunoz2022cleave} extends our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
+These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
+To address these challenges, we introduce a tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
+The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
+We then present use-case scenarios demonstrating the scalability and flexibility of their approach by deploying scenarios with a large number of loops without the need for domain-specific hardware.
+
+Given the vastness and complexity of the field of \glspl{NCS}, we limit our exploration in this dissertation to simple, small-scale, single-plant systems.
+The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
+Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that our conclusions may not directly translate to larger scale or more realistic systems.
+Additionally, our tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
+This does not match well with more complex controllers such as \gls{MPC} controllers, in which each sample produces multiple outputs from the controller.
+\gls{CLEAVE} will thus have to be modified and extended to work with such setups.
+Our study into the application of our methodology for \glspl{NCS} also relies, once again, exclusively on Wi-Fi-enabled Cloudlet Computing.
+As such, the same limitations with respect to \gls{MEC} and Fog Computing apply as with our line of work in \gls{WCA}.
+
+\bigskip
+
+This dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
+\Cref{paper:olguinmunoz2022ainur} formally introduces the Ainur framework for Edge Computing testbed automation and workload orchestration.
+This framework is employed in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the Edge Computing testbeds on which the developed tools were deployed.
+Ainur represents an ancillary contribution, crucial for the research presented in these works.
+Without it, the experimental approach described in these works would not have been feasible.
+
 Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \glspl{SDR} paired with general-purpose \gls{COTS} compute hardware.
-We do not target large-scale testbeds due to their complexity, both in terms of hardware and software.
+Medium and large testbeds such as \gls{COSMOS} and \gls{POWDER} are considered \gls{OOS} due to their cost and complexity, both in terms of hardware and software.
+In terms of networking technologies, Ainur is designed to be as agnostic as possible with respect to the network stack employed.
+This dissertation includes examples using Ethernet, Wi-Fi and 4G \gls{LTE}.
+Finally, Ainur does not currently fully integrate with modern Cloud and Edge infrastructure management and deployment technologies such as OpenStack~\cite{openstack}, although it does integrate with some core Cloud-native technologies such as Docker and Ansible.
+Work is currently ongoing to extend the framework to modern mobile network technologies such as 5G and beyond-5G, as well as to general-purpose, Cloud- and Edge-native infrastructure stacks.

### 2023-04-03 body/part1/chapter2-relatedwork/tables/scope.tex
scope

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
    \centering
    \caption{Overview of the scope and main contribution of each included publication.}\label{tab:scope}
    \tiny
    \renewcommand{\arraystretch}{1.5}
    \begin{tabularx}{\textwidth}{@{}ccXYX@{}}
        \toprule
        \textbf{Paper} & \textbf{Application} & \textbf{Emulation Approach} & \textbf{Network} & \textbf{Main Contribution}                                \\ \midrule
        \ref{paper:olguinmunoz2018demoscaling} & \acs{WCA} & Trace, \acs{FSM} user model & Wi-Fi & Initial demo of methodology. \\
        %
        \ref{paper:olguinmunoz2019edgedroid} & \acs{WCA} & Trace, \acs{FSM} user model & Wi-Fi & EdgeDroid tool for application of the methodology to \acs{WCA}. \\
        %
        \ref{paper:olguinmunoz2021impact} & \acs{WCA} & \acs{N/A} & No network & Characterization of human behavior in \acs{WCA}. \\
        %
        \ref{paper:olguinmunoz2022cleave} & \acs{NCS} & Software implementation of Control System & Wi-Fi & \acs{CLEAVE} tool for application of the methodology to \acs{NCS}. \\
        %
        \ref{paper:olguinmunoz2022ainur} & \acs{WCA}, \acs{NCS} & \acs{N/A} & Ethernet, Wi-Fi, 4G~\acs{LTE} & Ainur framework for Edge Computing testbed automation. \\
        \ref{paper:olguinmunoz2023realistic} &
        \acs{WCA} &
        Data-driven stochastic synthetic trace and user model &
        Wi-Fi &
        Model for end-to-end emulation of humans in \acs{WCA}, exploration of implications of methodology and model for \acs{WCA} optimization. \\
        \bottomrule
    \end{tabularx}
\end{table}

### 2023-04-03 body/part1/chapter1-introduction/background.tex
wooork

@@
 
-\subsubsection{\glsfmtlong{WCA}}
+\subsubsection{\glsfmtlong{WCA}}\label{sec:background:wca}
 \glsreset{WCA}
@@
 
-\todo[inline]{
-    - Section 1.1.4: Goes again in the right direction, but details are missing/off.
+% \todo[inline]{
+%     - Section 1.1.4: Goes again in the right direction, but details are missing/off.
 First of all, NCS is really a research direction, not necessarily a system type (every control system is more or less networked, but networking effects have not been always captured in the theoretical analysis/design of the controller).
@@
 In a sense, you can work out a pattern how you structure the discussion in Section 1.1.3, and reuse that pattern to a large extent in 1.1.4, that would be neat (both are closed-loop applications with a lot of similar, but also some different properties and requirements).
-}
+% }
 
-\glspl{NCS} are a class of control systems that involve the integration of communication networks into traditional control systems.
 A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.
@@
 
-\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators and interconnecting them with general-purpose \acs{TCP}/\acs{IP} networks;
-see \cref{fig:csvsncs}.
-This allows for remote monitoring and control, which enhances the reliability and safety of the control system, which is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and transportation systems.
+\gls{NCS} research is a research direction in the field of automatic control that focuses on the study and development of control systems that formally account for the effects of incorporating communication networks for data exchange between different components of the system.
+Every modern control system is, to some extent, networked in the sense that a communication medium is used to transmit information between components.
+The field of \gls{NCS} research formalizes and extends this characteristic by integrating traditional control systems with general-purpose \acs{TCP}/\acs{IP} networks.
+\glspl{NCS}~\cite{gupta2010networked} aim to enhance and extend the capabilities of traditional control systems by decoupling the controller from the plant, sensors, and actuators (see \cref{fig:csvsncs}).
+This allows for remote monitoring and control, which enhances the reliability and safety of the control system
+This is particularly useful in applications where real-time monitoring and control are critical, such as in industrial automation and in transportation systems.
 \glspl{NCS} also enable the integration of multiple control systems, as well as centralized control of multiple plants by a single controller.
@@
 
+\input{tables/ncs-examples}
+
 \glspl{NCS} have become increasingly popular in recent years due to advances in communication and control technologies, as well as the increasing demand for real-time, remote, and distributed control.
 They have found use in varied applications, such as industrial automation, transportation systems, and smart buildings, and some have attempted to leverage the Cloud for centralized, distributed control in what has been called \emph{Cloud control systems}~\cite{xia2015cloud}
-However, depending on the physical system being controlled, \glspl{NCS} can have stringent timing and reliability requirements for communication that conventional Cloud paradigms simply cannot meet~\cite{wan2020efficient}.
-This has led their adoption to be limited mostly to industrial environments.
-This is about to change, however, as with the advent of Edge Computing, paired with novel wireless communication technologies such as cellular 5G, consumer-grade \glspl{NCS} will be made possible.
-These technologies will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.
+However, \glspl{NCS} can have timing requirements for communication that conventional Cloud paradigms simply cannot meet~\cite{wan2020efficient}.
+Depending on the system being controlled, control systems can have sampling rates ranging from sub-\SI{1}{\hertz} to several \unit{\kilo\hertz}.
+Reliability is another concern, as many of the core use cases for \gls{NCS} involve application domains in which failure has the potential to cause bodily harm to humans or significant financial losses.
+\cref{tab:controlsystemqos} lists some example use cases for \gls{NCS} and their \gls{QoS} requirements.
+These stringent requirements have led their adoption to be limited mostly to industrial environments.
+This is about to change, however, as with the advent of \gls{MEC}, consumer-grade \glspl{NCS} will be made possible.
+This technology will likely become the backbone of consumer-grade, widely-deployed \glspl{NCS}, enabling real-time capabilities through extremely low end-to-end latencies together with context- and locality-awareness.

### 2023-04-03 body/part1/chapter1-introduction/tables/ncs-examples.tex
wooork

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[t]
    \centering
    \caption{Examples of use cases for NCS and their respective latency and reliability requirements.}\label{tab:controlsystemqos}
    \tiny
    \renewcommand{\arraystretch}{1.5}
    \begin{tabularx}{\textwidth}{@{}lYcX@{}}
        \toprule
        \textbf{} & \bfseries\makecell{Sampling~Rate} & \bfseries\makecell{Latency\\Bounds} & \bfseries\makecell{Reliability\\Requirements} \\ \midrule
        \bfseries\makecell[l]{Power\\Electronics} & Up to several \unit{\kilo\hertz} & \textless~\SI{1}{\milli\second} & High; failures can lead to damage to expensive equipment, large-scale blackouts. \\
        \bfseries\makecell[l]{Autonomous\\Vehicles} & Up to hundreds of \unit{\hertz} & \textless~\SI{10}{\milli\second} & High; failures can lead to injury or death. \\
        \bfseries\makecell[l]{Industrial\\Automation} & From sub-\SI{1}{\hertz} to several \unit{\kilo\hertz}. & \SI{1}{\second} to sub-\SI{1}{\milli\second} & High; failures can lead to injuries and/or significant impairment of production and loss of revenue. \\
        \bfseries\makecell[l]{Robotics} & Up to several \unit{\kilo\hertz} & \textless~\SI{1}{\milli\second} & Low to high, depending on the application. \\ \bottomrule
    \end{tabularx}
\end{table}

### 2023-04-03 body/part1/chapter3-contributions/methodology.tex
wooork

--- 2023-02-16 body/part1/overview/methodology.tex
@@
-As summarized in \cref{sec:summary_contributions}, this thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
+As summarized in \cref{chap:introduction}, this thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
 The primary contribution of this research is a methodology for analyzing and evaluating the performance of these applications.
@@
     \caption{%
-        Overview of the proposed methodology for the study of latency-sensitive systems on edge computing infrastructure.
+        Overview of the proposed methodology for the study of latency-sensitive systems on Edge computing infrastructure.
         The  client side of the system is replaced with an emulation executed on a general-purpose computing device.
@@
 
-As discussed in \cref{chap:relatedwork}, most research in this field falls into either simulated or fully-practical real-world approaches.
-Our emulated methodology presents several advantages over these alternatives.
-To begin with, and perhaps most importantly, it makes our approach portable and applicable to real edge computing deployments.
-By limiting the emulation to the workload, our approach can be directly employed on pre-existing edge computing infrastructure in order to benchmark applications before they are deployed.
+% As discussed in \cref{chap:relatedwork}, most research in this field falls into either simulated or fully-practical real-world approaches.
+Our emulated methodology presents several advantages over alternatives, simulated or fully-practical approaches.
+To begin with, and perhaps most importantly, it makes our approach portable and applicable to real Edge computing deployments.
+By limiting the emulation to the workload, our approach can be directly employed on pre-existing Edge computing infrastructure in order to benchmark applications before they are deployed.
 This makes our approach valuable to not only to the research community, but also to system designers and application developers.
@@
 The client-side component of the applications of interest for this research is often complex to scale.
-Consider a system owner interested in studying the performance of some edge computing setup for latency-sensitive applications.
+Consider a system owner interested in studying the performance of some Edge computing setup for latency-sensitive applications.
 In the case of interactive and immersive applications, scaling the number of human users is a significant challenge, as recruiting participants is an expensive and time-consuming endeavor.
@@
 In many cases, these systems are made-to-order, further increasing the cost as economies of scale cannot be employed.
-On the other hand, a simulated approach requires modeling not only the application workload, but also the edge computing system and infrastructure.
-Due to the intricate nature of contemporary cloud and edge computing setups, which frequently involve multiple tiers of proxies, load balancers, virtualized computing units, and network functions, the modeling of such systems can quickly become an overly complex endeavor.
+On the other hand, a simulated approach requires modeling not only the application workload, but also the Edge computing system and infrastructure.
+Due to the intricate nature of contemporary cloud and Edge computing setups, which frequently involve multiple tiers of proxies, load balancers, virtualized computing units, and network functions, the modeling of such systems can quickly become an overly complex endeavor.
 
@@
 the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
-As discussed in \cref{chap:introduction,chap:relatedwork}\todo{Was it really here?}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the edge --- concerns the involvement of human beings in their operation.
+As discussed in \cref{chap:introduction,sec:relwork:xremulation}, the main challenge to benchmarking \gls{WCA} --- and other ``human-in-the-loop'' applications on the Edge --- concerns the involvement of human beings in their operation.
 Humans are unreliable, and, perhaps more importantly, hard to scale.
@@
 We discuss an initial case study of our methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
-In this work, we present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of edge-native \aclp{NCS}.
-The growing adoption of edge computing has made it important to understand the strengths and weaknesses of different edge concepts for the deployment of such \glspl{CPS}.
+In this work, we present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of Edge-native \aclp{NCS}.
+The growing adoption of Edge computing has made it important to understand the strengths and weaknesses of different Edge concepts for the deployment of such \glspl{CPS}.
 
-\gls{CLEAVE} is a fully virtualized benchmarking framework, inspired by our previous work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} on benchmarking human-in-the-loop applications on the edge.
+\gls{CLEAVE} is a fully virtualized benchmarking framework, inspired by our previous work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} on benchmarking human-in-the-loop applications on the Edge.
 In the case of \gls{CLEAVE} however, we forgo the trace-based approach of EdgeDroid in favor of direct modeling of the target systems.
@@
 \gls{CLEAVE} is built using Python 3.8, making it highly extensible and able to use existing libraries.
-It is compatible with container technologies such as Docker~\cite{docker}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard edge setups using container orchestration solutions.
+It is compatible with container technologies such as Docker~\cite{docker}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard Edge setups using container orchestration solutions.
 
@@
 The inverted pendulum is a popular benchmark in control-system literature, and the setup allows for straightforward reparametrization and a broad range of experiments.
-Video analytics, on the other hand, is a main proposed use case for edge computing, and it is likely that edge \gls{NCS} deployments will be deployed in parallel with such applications in the future.
+Video analytics, on the other hand, is a main proposed use case for Edge computing, and it is likely that Edge \gls{NCS} deployments will be deployed in parallel with such applications in the future.
 Using Wi-Fi allows for more experimental freedom, as the system can be adapted to study both best-case and worst-case scenarios with low and high latencies, respectively.
@@
 Representative results from this experimentation are showcased in \cref{fig:cleaveresults}.
-While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, the results from these experiments show that network and resource contention are important factors when designing \glspl{NCS} for deployment on the edge.
+While the baselines might lead us to think that higher sampling rates are always better for the stability of control systems, the results from these experiments show that network and resource contention are important factors when designing \glspl{NCS} for deployment on the Edge.
 As sampling rates increase, a critical point is eventually reached at which the congestion on the network link is such that it severely impacts the quality of control.
@@
 
-The results of the experiments will be useful for optimizing resource utilization in edge-bound networked control systems.
+The results of the experiments will be useful for optimizing resource utilization in Edge-bound networked control systems.
 The results of the experiments with video-streaming applications will add to the understanding of the interaction between control systems and networked applications.
@@
 
-\subsection{ExPECA Testbed \& the Ainur software stack}
+\subsection{ExPECA Testbed \& the Ainur software stack}\label{sec:testbed}
 
@@
 Moreover, having a standardized baseline would make it easier to compare results across different studies.
-THe testbed could serve as a platform for researchers to test new ideas and compare them to the results of previous experiments.
+The testbed could serve as a platform for researchers to test new ideas and compare them to the results of previous experiments.
 This would enable more efficient and accurate assessments of the methodology's performance and facilitate the development of new methods and techniques.
@@
 These reasons led to the establishment of the \gls{EXPECA} testbed at \pgls{KTH}.
-\gls{EXPECA} is a \gls{SSF}-funded infrastructure project targeting the development and provisioning of an edge computing infrastructure for research into novel applications and network architectures.
+\gls{EXPECA} is a \gls{SSF}-funded infrastructure project targeting the development and provisioning of an Edge computing infrastructure for research into novel applications and network architectures.
 It consists of a cluster of hardware-reconfigurable \gls{COTS} computing nodes interconnected using managed switches and \glspl{SDR}.

### 2023-04-03 body/part1/chapter3-contributions/wca-implications.tex
wooork

@@
 
-An important portion of the present thesis deals with the modeling of human behavior for the benchmarking of \gls{WCA}.
+In order to more easily and accurately discuss the implications of our methodology for \gls{WCA}, we first characterize these applications and provide some definitions relating to their operation.
 We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
-In this section, we will provide some definitions relating to the operation of these applications.
 
@@
 From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
-However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in the previous section and in \cref{fig:wca}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
+However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in \cref{sec:background:wca} and in \cref{fig:wca:operation}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
 The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
@@
 
-\todo[inline]{Do we need more experiments?}
-
 In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.

### 2023-04-06 body/part1/chapter2-relatedwork/key_contributions.tex
notes

--- 2023-04-03 body/part1/chapter2-relatedwork/scope.tex
@@
-\section{Scope \& overview of the included publications}
+\section{Overview of the contributions of this dissertation}
+
+\todo[inline]{
+    first main contribution: introduce the idea of emulating the physical part of a CPS when it comes to experimental evaluation running over an integrated communication and computation infrastructure.
+    when we do that, we avoid many of the complications of experimental evaluation of these systems
+    in combination with Ainur, is a very elegant proposition
+    what is not currently captured precisely is that the approach is for the whole system (app + infra)
+    Need to mention the emulation
+    elegant and contemporary approach to efficient and repeatable evaluation of CPS
+
+    second contribution:
+    question of to which degree the emulation needs to capture certain details of the physical world?
+    to which degree is emulation ``good enough''?
+    for WCA specifically we propose a relatively detail of human user behavior based on empirical data
+    evaluate to which extent the emulation is ``good enough'' by optimization of these systems
+
+    Mention software, then we are on the safe.
+}
+
+\todo[inline]{Rework and rewrite.}
 
@@
 Work is currently ongoing to extend the framework to modern mobile network technologies such as 5G and beyond-5G, as well as to general-purpose, Cloud- and Edge-native infrastructure stacks.
+
+% \section{Key contributions}
+
+% \begin{enumerate}
+%     \item A methodology for the benchmarking of latency-sensitive, feedback-loop applications enabled by Edge Computing.
+%     This methodology is validated by example through representative use-cases, \gls{WCA} and \glspl{NCS}.
+%     \begin{description}
+%         \item[Novelty:] This is the first formalized method for this challenge in literature.
+%     \end{description}
+
+%     \item An exploration and characterization of the optimization potential of \gls{WCA} applications using the methodology, in particular with respect to sampling strategies and resulting resource and energy consumption profiles.
+%     \begin{description}
+%         \item[Novelty:] the study of \gls{WCA} in literature has, so far, focused primarily on prototyping applications and platforms.
+%         To the best of our knowledge, only a limited subset of works~\cite{chen2017empirical,wang2019towards,wang2020scaling} deal with the performance characterization of these applications.
+%         Additionally, only~\cite{wang2019towards,wang2020scaling} present heuristics for resource consumption optimization in \gls{WCA}, but no formal mathematical derivation of optimums is performed.
+%     \end{description}
+
+%     \item Tools for the implementation of the methodology, including end-to-end emulation frameworks for \gls{WCA} and \glspl{NCS} as well as a tool for Edge Computing testbed automation.
+%     \begin{description}
+%         \item[Novelty:] to the best of our knowledge, EdgeDroid is the first work in literature dealing with the dynamic emulation of \gls{XR} applications in networked contexts.
+%         Existing works are either completely forgo the networking aspect our use rigid, deterministic workloads which fail to capture the feedback-loop interactions between system responsiveness and human behavior.
+
+%         Likewise, \gls{CLEAVE} is the first tool for fully softwarized end-to-end emulation of generic \glspl{NCS} employing Edge and Cloud-native technologies.
+%         Most other experimental approaches to \gls{NCS} research either fully simulate the system or employ real physical control systems.
+%         Although a number of works exist employing a similar network-in-the-loop approach, they limit their functionality to specific control system use cases such as smart grid control, and do not represent general approaches.
+%         They also do not employ Cloud or Edge-native technologies.
+
+%         Finally, Ainur represents the first framework for Edge Computing testbed research with end-to-end automation and orchestration.
+%         Existing approaches have focused primarily on achieving feature-parity with public clouds and targeting large-scale deployments, and have forgone the end-to-end automation aspect of Edge Computing experimentation.
+%         Ainur is the first toolkit able to configure a testbed from scratch, run experiments, collect the data, and clean up afterwards in an automated fashion.
+%     \end{description}
+% \end{enumerate}

### 2023-04-06 body/part1/chapter2-relatedwork/relatedwork.tex
wooork
notes

@@
-\section{Related work}
+\section{State of the art}
 
-\subsection{Emulation in \glsfmtshort{XR} research}
+\subsection{Performance evaluation of Edge Computing systems}
 
-Only a limited number of works addressing the partial or complete emulation of XR applications for benchmarking and research purposes exist in literature.
-One such work is~\cite{williams2013transform}.
-In this, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
-The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.\@
-\cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{AR} application by mocking Android \gls{SDK} \glspl{API}.
-This approach allows for benchmarking and evaluation of algorithms and architectures on real data.
-A similar approach was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
-Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
-Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
-
-While the above approaches to benchmarking and research in \gls{XR} are steps in the right direction, they fail to consider the dynamic nature of \gls{XR}.
-The use of rigid traces in trace-based approaches can lead to timing mismatches when replaying at runtime, as the system may respond differently to the same input depending on factors such as system load and network congestion.
-Additionally, these approaches do not take into consideration the effects of system responsiveness on human behavior, which can be critical for evaluating both \gls{QoS} and \gls{QoE}.
-To address these issues, future approaches will need to adapt trace replay at runtime to match expected behaviors and consider the impact of system responsiveness on user behavior.
-By doing so, we can develop more accurate and reliable benchmarks for \gls{XR} applications and improve our understanding of how users interact with these technologies.
-
-\subsection{Edge and Cloud Benchmarking}
+\todo[inline]{
+    Needs some massaging.
+}
+\todo[inline]{
+    Right now includes only experimental approaches.
+    Need to add mathematical models (ask James for citations) and simulations.
+}
 
@@
 
-\subsection{Experimental research in \glsfmtshortpl{NCS}}
-
-A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-
-Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies~\cite{zhang2019networked}.
-\gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
-However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
-
-Experimental research in \glspl{NCS} involves a diverse range of hardware and software platforms, methodologies, and key performance indicators.
-Specifically, we identify three distinct categories of approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
-
-In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
-This approach is straightforward to implement, and can be very useful for evaluating the real-world applicability of the system, but can be costly and time-consuming.
-Examples of the application of this approach can be found in studies such as~\cite{drew2005networked,baumann2018evaluating,li2014wireless,cuenca2019periodic}.
-These studies implemented wireless and microcontroller-based systems for control and validated them on a physical prototype.
-
-In~\cite{drew2005networked}, the authors present a \gls{NCS} model which considers both random packet delay and loss on both the sensor and actuator sides of the feedback loop.
-They validate their design on a physical testbed consisting of inverted pendula controlled from a central computation point over Wi-Fi.
-A similar testbed is employed for validation in~\cite{baumann2018evaluating}, in which the authors develop an evaluation methodology for wireless \gls{NCS}.\@
-\cite{li2014wireless} implements a wireless microcontroller-based system for vibration control and validates it on a physical prototype comprising a cantilever beam controlled using an \acs{IEEE} 802.15.4 \gls{WPAN} system.\@ \cite{cuenca2019periodic} discusses the design and implementation of a periodic event-triggered sampling and dual-rate control techniques for wireless control, and validate their contributions on a four-rotor autonomous drone platform controlled over WiFi.
-\emph{NCSbench}~\cite{zoppi2020ncsbench}, an open-source \gls{NCS} benchmarking platform designed with reproducibility in mind built using the \citetitle{LEGOMindstormsEV3}~\cite{LEGOMindstormsEV3} platform, is another such example.
-
-Completely simulated setups, on the other hand, rely on computer simulations to evaluate the \gls{NCS}'s performance.
-This approach is less expensive and faster than the practical approach but may not fully capture the real-world complexities of the system.
-Examples of tools and frameworks for the implementation of this approach can be found in works such as~\cite{andersson2005simulation,eyisi2012ncswt}.
-
-A simulated approach is employed in~\cite{du2009novel}, in which the authors develop a novel \emph{Smith} predictor to compensate for varying latencies in wireless \glspl{NCS} and validate it using the \emph{TrueTime}~\cite{henriksson2002truetime} simulator.
-In~\cite{chen2015synchronous} the authors validate and evaluate a number of synchronous control strategies for three-motor setups using completely simulated \glspl{NCS} environments.\@
-\cite{wu2012application} introduces the concept of \emph{network predictive control} in and uses it to balance a simulated inverted pendulum over a simulated wireless network.\@
-\cite{ma2019optimal} proposes an optimal dynamic scheduling strategy that optimizes performance of multi-loop control systems and validate it by simulating a four-loop control system over an \acs{IEEE} 802.15.4 \gls{WPAN}.
-
-Finally, network and hardware-in-the-loop approaches, which aim to strike a balance between realism and efficiency, have also been employed in literature.
-Network-in-the-loop refers to experimental setups in which the \emph{network} is instead simulated or emulated.
-Such approaches have, for instance, been used to study the effects of the \gls{CSMA/CD} medium access control mechanism on plant stability, as in~\cite{natale2004inverted}.
-Hardware-in-the-loop approaches, referring to \gls{NCS} setups in which a real network interacts with a simulated or emulated control system, are particularly prevalent in the field of \emph{smart grid} control.
-An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
-
-\medskip
-
-The diversity of experimental \glspl{NCS} research discussed above has led to fragmentation of hardware, software, and methodology, as studies tend to favor approaches popular in their respective communities.
-Additionally, studies tend to focus on individual aspects and components of the system, resulting in incomplete knowledge of the overall \glspl{NCS}.
-There is a gap in knowledge concerning the reproducibility and comparison of experimental studies on these systems.
-
-To our knowledge, the only published attempt at addressing methodological fragmentation and poor reproducibility of experimental \gls{NCS} research is the LEGO-based approach introduced in~\cite{zoppi2020ncsbench}.
-However, this approach has limitations.
-It is dependent on specific hardware --- the LEGO Mindstorms set --- which makes its adoption limited due to availability issues, reduces its flexibility, scalability, and limits the scope of experiments able to be conducted via \emph{NCSbench} only to what is possible to implement in LEGO.\@
-Scaling up experiments becomes cumbersome and costly, further limiting the practicality of the tool.
-Thus, a more general approach that tackles the scalability and reproducibility issues of NCS research is urgently needed.
-The development of such an approach would enable researchers to conduct experiments that generate reproducible results consistently and can be compared across different studies while still retaining the computational sophistication required when scaling up experiments.
-
-\subsection{Testbed research in Edge Computing}
+\subsubsection{Testbed research in Edge Computing}
 
@@
 There is still remains much work to be done in this area, particularly with respect to the adoption of Edge-compatible solutions that can enable faster and more reliable experimentation.
+
+\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshort{CPS}}
+
+\todo[inline]{
+    At this point we have settled for experimental approach.
+    No need to delve deep in math models and simulations, right?
+}
+
+\todo[inline]{
+    Also, title again needs reworking.
+}
+
+\subsubsection{Experimental approaches to \gls{MAR} research}\label{sec:relwork:xremulation}
+\todo[inline]{Rework to talk more generally about experimental research in MAR?}
+
+Only a limited number of works addressing the partial or complete emulation of XR applications for benchmarking and research purposes exist in literature.
+One such work is~\cite{williams2013transform}.
+In this, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
+The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.\@
+\cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{AR} application by mocking Android \gls{SDK} \glspl{API}.
+This approach allows for benchmarking and evaluation of algorithms and architectures on real data.
+A similar approach was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
+Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
+Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
+
+While the above approaches to benchmarking and research in \gls{XR} are steps in the right direction, they fail to consider the dynamic nature of \gls{XR}.
+The use of rigid traces in trace-based approaches can lead to timing mismatches when replaying at runtime, as the system may respond differently to the same input depending on factors such as system load and network congestion.
+Additionally, these approaches do not take into consideration the effects of system responsiveness on human behavior, which can be critical for evaluating both \gls{QoS} and \gls{QoE}.
+To address these issues, future approaches will need to adapt trace replay at runtime to match expected behaviors and consider the impact of system responsiveness on user behavior.
+By doing so, we can develop more accurate and reliable benchmarks for \gls{XR} applications and improve our understanding of how users interact with these technologies.
+
+\subsubsection{Experimental approaches to research in \glsfmtshortpl{NCS}}
+\todo[inline]{
+    Rework.
+    Don't focus on repeatability argument.
+    Argue that a generic front-end-emulated approach is needed for flexibility.
+}
+
+A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
+Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
+Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
+They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
+
+Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies~\cite{zhang2019networked}.
+\gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
+However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
+
+Experimental research in \glspl{NCS} involves a diverse range of hardware and software platforms, methodologies, and key performance indicators.
+Specifically, we identify three distinct categories of approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
+
+In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
+This approach is straightforward to implement, and can be very useful for evaluating the real-world applicability of the system, but can be costly and time-consuming.
+Examples of the application of this approach can be found in studies such as~\cite{drew2005networked,baumann2018evaluating,li2014wireless,cuenca2019periodic}.
+These studies implemented wireless and microcontroller-based systems for control and validated them on a physical prototype.
+
+In~\cite{drew2005networked}, the authors present a \gls{NCS} model which considers both random packet delay and loss on both the sensor and actuator sides of the feedback loop.
+They validate their design on a physical testbed consisting of inverted pendula controlled from a central computation point over Wi-Fi.
+A similar testbed is employed for validation in~\cite{baumann2018evaluating}, in which the authors develop an evaluation methodology for wireless \gls{NCS}.\@
+\cite{li2014wireless} implements a wireless microcontroller-based system for vibration control and validates it on a physical prototype comprising a cantilever beam controlled using an \acs{IEEE} 802.15.4 \gls{WPAN} system.\@ \cite{cuenca2019periodic} discusses the design and implementation of a periodic event-triggered sampling and dual-rate control techniques for wireless control, and validate their contributions on a four-rotor autonomous drone platform controlled over WiFi.
+\emph{NCSbench}~\cite{zoppi2020ncsbench}, an open-source \gls{NCS} benchmarking platform designed with reproducibility in mind built using the \citetitle{LEGOMindstormsEV3}~\cite{LEGOMindstormsEV3} platform, is another such example.
+
+Completely simulated setups, on the other hand, rely on computer simulations to evaluate the \gls{NCS}'s performance.
+This approach is less expensive and faster than the practical approach but may not fully capture the real-world complexities of the system.
+Examples of tools and frameworks for the implementation of this approach can be found in works such as~\cite{andersson2005simulation,eyisi2012ncswt}.
+
+A simulated approach is employed in~\cite{du2009novel}, in which the authors develop a novel \emph{Smith} predictor to compensate for varying latencies in wireless \glspl{NCS} and validate it using the \emph{TrueTime}~\cite{henriksson2002truetime} simulator.
+In~\cite{chen2015synchronous} the authors validate and evaluate a number of synchronous control strategies for three-motor setups using completely simulated \glspl{NCS} environments.\@
+\cite{wu2012application} introduces the concept of \emph{network predictive control} in and uses it to balance a simulated inverted pendulum over a simulated wireless network.\@
+\cite{ma2019optimal} proposes an optimal dynamic scheduling strategy that optimizes performance of multi-loop control systems and validate it by simulating a four-loop control system over an \acs{IEEE} 802.15.4 \gls{WPAN}.
+
+Finally, network and hardware-in-the-loop approaches, which aim to strike a balance between realism and efficiency, have also been employed in literature.
+Network-in-the-loop refers to experimental setups in which the \emph{network} is instead simulated or emulated.
+Such approaches have, for instance, been used to study the effects of the \gls{CSMA/CD} medium access control mechanism on plant stability, as in~\cite{natale2004inverted}.
+Hardware-in-the-loop approaches, referring to \gls{NCS} setups in which a real network interacts with a simulated or emulated control system, are particularly prevalent in the field of \emph{smart grid} control.
+An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
+
+\medskip
+
+The diversity of experimental \glspl{NCS} research discussed above has led to fragmentation of hardware, software, and methodology, as studies tend to favor approaches popular in their respective communities.
+Additionally, studies tend to focus on individual aspects and components of the system, resulting in incomplete knowledge of the overall \glspl{NCS}.
+There is a gap in knowledge concerning the reproducibility and comparison of experimental studies on these systems.
+
+To our knowledge, the only published attempt at addressing methodological fragmentation and poor reproducibility of experimental \gls{NCS} research is the LEGO-based approach introduced in~\cite{zoppi2020ncsbench}.
+However, this approach has limitations.
+It is dependent on specific hardware --- the LEGO Mindstorms set --- which makes its adoption limited due to availability issues, reduces its flexibility, scalability, and limits the scope of experiments able to be conducted via \emph{NCSbench} only to what is possible to implement in LEGO.\@
+Scaling up experiments becomes cumbersome and costly, further limiting the practicality of the tool.
+Thus, a more general approach that tackles the scalability and reproducibility issues of NCS research is urgently needed.
+The development of such an approach would enable researchers to conduct experiments that generate reproducible results consistently and can be compared across different studies while still retaining the computational sophistication required when scaling up experiments.

### 2023-04-06 body/part1/chapter2-relatedwork/scope.tex
notes

--- 2023-03-14 body/part1/chapter2-relatedwork/scope.tex
@@
-\section{Scope of this work}
+\section{Scope of the thesis}
 
-This dissertation presents a methodology for the study of feedback-loop \gls{CPS} systems deployed on Edge Computing and studies its potential applications and implications.
-This methodology is based on the emulation of the desired application workload on top of the actual Edge infrastructure, including backend computing hardware and network technology used for client-server communication.
-We do not in any way claim that this methodology is the best approach to studying these systems, but rather that it is a valid approach able to produce accurate, representative results.
+\todo[inline]{Write}
 
-The scope of \gls{CPS} systems is vast and diverse, making it unfeasible to investigate all possible variations of these systems in this work. 
-Instead, we opt to study a representative subset of future applications that could potentially leverage Edge Computing resources to provide innovative solutions: \gls{WCA} and \glspl{NCS}.
-By considering these two application domains, we demonstrate the applicability of our methodology and showcase its efficacy.
-
-Specifically, a subset of the research works included in this dissertation presents and studies the application of our methodology for \gls{WCA}.
-To limit the scope of our research in this area, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
-We focus on this category for a couple of reasons.
-Step-based task-assistance \gls{WCA} were among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represent a more mature application class compared to other \glspl{WCA}.
-These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
-The specific application used for these studies corresponds to the LEGO assistant devised by \citeauthor{chen2015early}\cite{chen2015early}.
-
-A separate subset of the research presented in this work discusses a preliminary foray into the applications of our methodology for \glspl{NCS}.
-Given the vastness and complexity of this field, we limit our exploration to simple, small-scale, single-plant systems.
-In particular, the system used as a representative example for the effectiveness of our approach for \glspl{NCS} corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
-
-Finally, this dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
-Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \glspl{SDR} paired with general-purpose \gls{COTS} compute hardware.
-We do not target large-scale testbeds due to their complexity, both in terms of hardware and software.
+\begin{enumerate}
+    \item Performance evaluation.
+    \item \gls{CPS} and \gls{MAR}
+    \item Edge Computing 
+\end{enumerate}

### 2023-04-09 body/part1/chapter1-introduction/background.tex
notes
cps

@@
 
-\subsection{\glsfmtlong{XR}}\label{background:xr}
+\subsection{Performance evaluation of networked systems}
+\todo[inline]{
+    Too vague.
+    Performance of networked systems.
+
+    Look into book and use it as background.
+    Something we can cite.
+    2-3 pages.
+}
+\todo[inline]{
+    Write.
+}
+
+\subsection{Closed-loop feedback systems enabled by Edge Computing}
+
+\todo[inline]{
+    Needs a better title.
+    NCS and WCA?\@ Put everything under CPS?\@
+    Closed-loop feedback systems.
+}
+
+\todo[inline]{
+    Rework structure a bit.
+}
+
+\subsubsection{\glsfmtlong{XR}}\label{background:xr}
 \todo[inline]{Needs citations.}
@@
 
-\subsubsection{\glsfmtlong{WCA}}\label{sec:background:wca}
+\paragraph{\glsfmtlong{WCA}}\label{sec:background:wca}
 \glsreset{WCA}
@@
 
-% As with many other categories of \gls{AR} applications, the computational complexity of the processing of the real-time input data makes it unfeasible to do on wearable devices.
-% This context-rich and latency-sensitive nature of \gls{WCA} inputs further makes this category of applications a prime candidate for offloading to the Edge.
+\subsubsection{\glsfmtlong{CPS}}
+\glsreset{CPS}
 
-\subsubsection{Emulation and simulation for \acs{XR}}
-\todo[inline]{
-    Section 1.1.2: This subsection is not very informative so far.
-Maybe first discuss extended reality, and then present a subsection on performance evaluation of systems, where simulation is a subsection, and experimentation is another subsection (and mathematical analysis is a third one).
-Focus on the usage of these concepts in the context of networked systems.
-Why do we need tools/approaches for performance evaluation, what are the advantages and disadvantages of different approaches, and also what is the historical context.
-You also need to have references.
-Finally, tables and figures are always welcome...
-}
+The term \gls{CPS} applies to a range of systems in which the computational world is tightly integrated with the physical world.
+\glspl{CPS} are designed to monitor, control, and optimize physical processes.
+By continuously monitoring and analyzing data from the physical components, \gls{CPS} can make decisions and take actions to optimize the system's performance.
+For example, a \gls{CPS} controlling a building's heating, ventilation, and air conditioning system may adjust the temperature based on occupancy patterns, external weather conditions, and energy consumption goals.
 
-The terms \emph{emulation} and \emph{simulation} are often used interchangeably.
-However, although both terms refer to methods of imitating and modeling systems, they are distinct in their goals, methods, and applications.
+\glspl{CPS} have acquired increasing importance in a wide range of industries due to their potential to enable greater efficiency, safety, and reliability in complex processes.
+Examples of such applications of \glspl{CPS} include smart grids, \acs{ITS} and autonomous vehicles, and smart health monitoring.
+A \emph{smart grid} is a \gls{CPS} that enables two-way communication between power generation, distribution, and consumption systems.
+Smart grids can thus dynamically and in real-time adjust the distribution of power based on demand and supply conditions.
+\glsreset{ITS} An \gls{ITS} is a system that integrates compute, monitoring, and communication in the context of a transportation system.
+\glspl{ITS} enable real-time traffic monitoring and management, route optimization, and integration of autonomous vehicles.
+These systems aim to optimize the management of transportation infrastructure and improve the safety, efficiency, and reliability of transportation systems.
+Finally, \emph{smart health monitoring} is an application of \glspl{CPS} in which wearable devices equipped with sensors and communication capabilities collect real-time data on patients' vital signs, activity levels, and medication adherence.
+This data can then be analyzed to provide alerts to healthcare providers, improve diagnoses, and facilitate early intervention.
 
-Emulation refers to the process of reproducing or mimicking the observed behavior of a system through another system, 
-with the ultimate goal being the recreation of behavior that matches the original system as close as possible.
-Emulations run in \emph{real time}, and can thus interact with real environments;
-in turn, this means that emulations can be used to replace subcomponents of larger systems for research or benchmarking.
-This method does not aim to model the original system, nor aim to understand its underlying behavior.
+Although these systems can enable greater efficiency, safety, and reliability in complex processes, the adoption of \glspl{CPS} faces several technical, social, and regulatory issues.
+One of the key technical challenges relates to the reliability and safety of \glspl{CPS}.
+As these systems tightly integrate physical and computational components, a failure in one component can have significant consequences.
+As exemplified above, key use-cases for \glspl{CPS} include critical infrastructure such as power grids, transportation systems, and healthcare monitoring systems.
+Any failure in these systems can have serious consequences for public safety and well-being.
+Furthermore, \glspl{CPS} require strict security and confidentiality, given the sensitive data in these systems.
+Developing CPS that are resilient to attacks and failures, as well as developing methods for testing and verifying their safety, is therefore a major research challenge.
 
-Simulation, on the other hand, is the process of modeling a system to study its behavior through the creation of a mathematical or logical model of the system.
-The model can then be adjusted to simulate and better represent the original system in different scenarios and conditions.
-Simulations run in entirely virtualized environments, and are not beholden to any real-time constraint.
-
-Finally, these approaches differ as well in the level of accuracy each of them affords.
-While simulation only attempts to approximate the behavior of the system in a given context, emulation aims to represent as close to a perfect copy of the observed behavior of the system.
-The former is therefore employed when a general understanding of the system is sufficient, while the latter is used when precision is required.
-However, thanks to its real time constraint and high degree of realism, emulation is typically more resource-intensive than simulation.
-This means that emulation requires more processing power, memory, and storage space than simulation.
-
-\subsection{\glsfmtlongpl{NCS}}\label{background:ncs}
+\paragraph{\glsfmtlongpl{NCS}}\label{background:ncs}
 \glsreset{NCS}
-
-% \todo[inline]{
-%     - Section 1.1.4: Goes again in the right direction, but details are missing/off.
-First of all, NCS is really a research direction, not necessarily a system type (every control system is more or less networked, but networking effects have not been always captured in the theoretical analysis/design of the controller).
-Second, maybe you stress more the working principle in terms workloads and requirements.
-In a sense, you can work out a pattern how you structure the discussion in Section 1.1.3, and reuse that pattern to a large extent in 1.1.4, that would be neat (both are closed-loop applications with a lot of similar, but also some different properties and requirements).
-% }
 

### 2023-04-09 body/part1/chapter2-relatedwork/key_contributions.tex
cps

--- 2023-04-03 body/part1/chapter2-relatedwork/scope.tex
@@
-\section{Scope \& overview of the included publications}
+\section{Overview of the included publications}
 
-This dissertation presents a methodology for the study of feedback-loop \gls{CPS} systems deployed on Edge Computing and studies its potential applications and implications.
-This methodology is based on the emulation of the desired application workload on top of the actual Edge infrastructure, including backend computing hardware and network technology used for client-server communication.
-We do not in any way claim that this methodology is the best approach to studying these systems, but rather that it is a \emph{valid} approach able to produce accurate and representative results.
+\todo[inline]{
+    first main contribution: introduce the idea of emulating the physical part of a CPS when it comes to experimental evaluation running over an integrated communication and computation infrastructure.
+    when we do that, we avoid many of the complications of experimental evaluation of these systems
+    in combination with Ainur, is a very elegant proposition
+    what is not currently captured precisely is that the approach is for the whole system (app + infra)
+    Need to mention the emulation
+    elegant and contemporary approach to efficient and repeatable evaluation of CPS
 
-\input{tables/scope}
+    second contribution:
+    question of to which degree the emulation needs to capture certain details of the physical world?
+    to which degree is emulation ``good enough''?
+    for WCA specifically we propose a relatively detail of human user behavior based on empirical data
+    evaluate to which extent the emulation is ``good enough'' by optimization of these systems
 
-The scope of \gls{CPS} systems is vast and diverse, making it unfeasible to investigate all possible variations of these systems in this work. 
-Instead, we opt to study a representative subset of future applications that could potentially leverage Edge Computing resources to provide innovative solutions: \gls{WCA} and \glspl{NCS}.
-By considering these two application domains, we demonstrate the applicability of our methodology and showcase its efficacy.
-\Cref{tab:scope} presents an overview of the scope of this dissertation, and in the following we will summarize the contributions and limitations of each included publication.
-
-\medskip
+    Mention software, then we are on the safe.
+}
 

### 2023-04-09 body/part1/chapter2-relatedwork/scope.tex
cps

@@
 \todo[inline]{Write}
+
+\todo[inline]{
+    Emphasize that apps are standing for a new class of application class that the community is interested in developing.
+    Why?
+    Utility.
+    Weave into intro and scope.
+
+    Not just a random mashup.
+    Relevance of the combination!
+
+    Prime reason for experimental evaluation.
+    Capture effects of compute, network, and design of the application.
+    Hard to capture otherwise.
+
+    Build argument in state of the art that there is no current satisfying way to address this.
+    Potentially orders of magnitude off.
+
+    In scope: interested in meaningful performance evaluation (i.e.\ capture all the complexity!) Make sure we note this.
+    Show how apps and system are intimately related to each other and complex to evaluate.
+    Where do complexities arise.
+    Understanding system performance is key and complex.
+
+    Scope: define the challenges.
+}
 

### 2023-04-10 body/part1/chapter1-introduction/background.tex
perf eval

@@
 \subsection{Performance evaluation of networked systems}
-\todo[inline]{
-    Too vague.
-    Performance of networked systems.
 
-    Look into book and use it as background.
-    Something we can cite.
-    2-3 pages.
-}
-\todo[inline]{
-    Write.
-}
+Performance evaluation is an essential step in designing and optimizing networked systems, in order to ensure efficient and effective operation.
+Such evaluation involves measuring and analyzing system performance under a multitude of scenarios to identify potential bottlenecks.
+In practice, three main approaches exist to the performance evaluation of any complex system; analytical modeling, simulations, and experimental evaluation~\cite{fernandes2017performance}.
+These approaches each have their advantages and disadvantages.
+
+Analytical or mathematical modeling involves creating a mathematical model that represents the system's behavior and performance.
+This involves defining the system's components, their interactions, and the performance metrics used to measure the system's performance.
+The mathematical model can be solved analytically, providing insights into the behavior of the system under different conditions.
+
+Analytical modeling is often used as a first step when attempting to identify possible solutions to a problem and provide an initial reduction of the problem space.
+One of the key advantages of mathematical modeling is that their realism and resulting complexity can be arbitrarily constrained as required.
+This allows researchers to quickly evaluate a large number of scenarios by employing coarse, approximate models to identify and highlight latent issues.
+These models can then be refined to further narrow the solution space.
+This approach is particularly useful for systems with complex interactions between multiple components, where testing every possible scenario in the real world or even in simulations would be impractical or impossible due to time and cost constraints.
+
+However, analytical modeling also has its limitations, in particular with respect to highly complex systems.
+In systems involving a large number of interacting components and processes, the mathematical equations required to model the system become more complex and difficult to solve.
+Assumptions about system behavior also become more inaccurate and incomplete as this behavior grows in complexity.
+As mathematical models are based on assumptions about the system's behavior and performance, incorrect, incomplete, or inaccurate assumptions can in turn lead to erroneous predictions.
+Additionally, complex systems may involve non-linear interactions between components, further increasing the difficulty of accurately modeling these systems.
+
+\medskip
+
+The second possible approach to performance evaluation of networked systems is simulation.
+This involves creating a computational model of the system, and then using it to simulate a series of scenarios to measure the model's performance.
+Simulations are often used to validate mathematical models and further narrow down the problem and solution spaces.
+
+Simulations offer numerous advantages.
+They represent an inexpensive and reliable way to study systems, by providing safe, controlled, and repeatable environments in which to test the system's behavior~\cite{fernandes2017performance}.
+Simulations also allow researchers to evaluate the system's performance with a high degree of accuracy and granularity, as they can approximate the system's behavior at a very fine-grained level without necessarily having to resort to complex mathematical modeling.
+Additionally, simulations can be used to test the impact of changes to the system's design or configuration, allowing researchers and system designers to identify the most effective solutions to improve performance.
+
+Simulation is a widely used and well-established method in networked system research, and there exists a rich literature surrounding tools for its implementation as a methodology~\cite{wehrle2010modeling}.
+The availability of simulation frameworks has made it easier for researchers to evaluate the performance of networked systems controlled environments.
+These frameworks provide a set of tools and libraries that enable researchers to create models of networked systems with very little overhead to simulate their behavior.
+
+Examples of frameworks for network simulation include the widely adopted \acs{OMNET}~\cite{omnet}, \acs{ns2}, and \acs{ns3}~\cite{nsnam} discrete-event simulators.
+\acs{OMNET} is an open-source, modular, and extensible discrete-event simulator that provides an object-oriented framework for modeling and simulating networks, and includes a variety of modules for different network protocols and technologies.
+\acs{ns2}, and \acs{ns3} refer to two generations of the same project.
+These widely popular simulation frameworks provide comprehensive sets of tools for modeling and simulating networks, and include support for a range of network technologies.
+
+Nevertheless, simulations also have their limitations.
+The realism of simulations is inherently limited by the accuracy and completeness of the underlying models used.
+Simulations rely on mathematical models that are an abstraction of the real-world system, and these models are only as accurate as the assumptions made and the data used to construct them.
+Simulations also cannot account for the impact of external factors on the system's performance.
+Real-world systems are often subject to factors that cannot be modeled accurately, such as unpredictable weather conditions, human behavior, or hardware failures.
+Moreover, for complex systems with many interacting components, simulations can become prohibitively computationally expensive and time consuming.
+It is not unusual for simulations of highly complex systems to execute much slower than real-time due to the computational power required to calculate the interactions between elements in the system.
+
+\medskip
+
+The third and final approach to performance evaluation of networked systems is practical experimentation on the real system.
+Such an approach involves measuring the performance of the \emph{real} system using practical, real-world experiments in different scenarios.
+Experimental research is often the final stage in systematic performance benchmarking, employed once the problem space has been reduced to a handful of potential solutions through mathematical modeling and subsequent simulations.
+
+Experimental research provides the most accurate assessment of system performance.
+It allows for an extremely high level of realism and can capture factors that cannot be accurately modeled through mathematical modeling or simulations, such as non-linear behaviors, human behavior, and unexpected events.
+Experimental methods provide direct measures of the system's performance under real-world conditions, enabling researchers to validate the accuracy of their mathematical models and simulations.
+
+However, experimental approaches also come with challenges, such as cost, practicality, repeatability, and replicability.
+Experimental research is often time-consuming and expensive to perform, as its implementation requires access to the real system under study.
+Additionally, experimental research is also limited in terms of replicability and repeatability, particularly for systems subject to external factors or stochastic and chaotic effects.
+This can make it challenging to draw definitive conclusions from experimental research.
+
+When it comes to experimental approaches in research involving humans, there are additional challenges that must be considered beyond those mentioned above.
+Ethical concerns related to participant safety, informed consent, and privacy must be addressed, and studies involving human participants must be reviewed and approved by appropriate ethical review panels\footnote{%
+    In Sweden, ethical review is handled at the state level by the Ethical Review Authority (\emph{Etikprövningsmyndigheten}); in the \acs{USA}, it is handled by each institution's \gls{IRB}.
+}.
+Recruitment and retention of human participants can be challenging, as participants may drop out or not show up for study sessions, leading to data collection difficulties.
+Despite these challenges, research involving human participants provides invaluable insights into the behavior and performance of networked systems and is critical for fully understanding and optimizing the performance of these systems in real-world settings.
 
@@
 
-\todo[inline]{
-    Needs a better title.
-    NCS and WCA?\@ Put everything under CPS?\@
-    Closed-loop feedback systems.
-}
-
-\todo[inline]{
-    Rework structure a bit.
-}
+\todo[inline]{Write an intro?}
 
@@
 
-\subsubsection{\glsfmtlong{CPS}}
+\subsubsection{\glsfmtlongpl{CPS}}
 \glsreset{CPS}
@@
 
+A research topic within \glspl{CPS} that is of particular interest to this dissertation is \glspl{NCS}. 
 A control system is a collection of devices or subsystems that work together to achieve a desired behavior or outcome of a physical system.

### 2023-04-10 body/part1/chapter1-introduction/tables/perf-eval.tex
background!

\begin{table}[]
    \centering
    \caption{Comparison between the three main approaches to performance evaluation of networked systems.}\label{tab:performance-evaluation}
    \tiny
    \renewcommand{\arraystretch}{1.5}
    \begin{tabularx}{\textwidth}{@{}lXXX@{}}
        \toprule
        & \textbf{Analytical modeling} & \textbf{Simulations} & \textbf{Experimental research} \\
        \midrule

        \textbf{Characteristics} &
        Based on mathematical models. &
        Based on mathematical models implemented as software to simulate the behavior of the system over time. &
        Involves direct measurement of system performance under real-world conditions. \\

        \textbf{Advantages} &
        Efficient. Provides theoretical insights and understanding of the system. &
        More realistic representation of the system's behavior and performance.
        Easy to extend and combine with other models for additional realism.
        Allows for controlled and repeatable experimentation. &
        Provides the most accurate assessment of system performance.
        Able to capture factors that cannot be modeled or simulated, such as non-linearities and stochastic or chaotic factors. \\

        \textbf{Disadvantages} &
        Limited in complexity, accuracy, and realism.
        May be unable to account for all relevant real-world factors. &
        Limited by the accuracy of the models used and the assumptions made.
        May not capture all relevant factors.
        Often computationally intensive and time-consuming. &
        Can be expensive, time-consuming, and difficult to replicate.
        When involving humans, ethical concerns must be addressed.
        May not be feasible for all research questions. \\

        \bottomrule
    \end{tabularx}
\end{table}

### 2023-04-10 body/part1/chapter1-introduction/background.tex
background!
background

@@
 \section{Background}
+\glsresetall%
+
+\subsection{Performance evaluation of networked systems}
+
+Performance evaluation is an essential step in designing and optimizing networked systems, in order to ensure efficient and effective operation.
+Such evaluation involves measuring and analyzing system performance under a multitude of scenarios to identify potential bottlenecks.
+In practice, three main approaches exist to the performance evaluation of any complex system; analytical modeling, simulations, and experimental evaluation~\cite{fernandes2017performance}.
+These approaches each have their advantages and disadvantages, summarized in \cref{tab:performance-evaluation} and discussed in detail below.
+
+\input{tables/perf-eval}
+
+Analytical or mathematical modeling involves creating a mathematical model that represents the system's behavior and performance.
+This involves defining the system's components, their interactions, and the performance metrics used to measure the system's performance.
+The mathematical model can be solved analytically, providing insights into the behavior of the system under different conditions.
+
+Analytical modeling is often used as a first step when attempting to identify possible solutions to a problem and provide an initial reduction of the problem space.
+One of the key advantages of mathematical modeling is that their realism and resulting complexity can be arbitrarily constrained as required.
+This allows researchers to quickly evaluate a large number of scenarios by employing coarse, approximate models to identify and highlight latent issues.
+These models can then be refined to further narrow the solution space.
+This approach is particularly useful for systems with complex interactions between multiple components, where testing every possible scenario in the real world or even in simulations would be impractical or impossible due to time and cost constraints.
+
+However, analytical modeling also has its limitations, in particular with respect to highly complex systems.
+In systems involving a large number of interacting components and processes, the mathematical equations required to model the system become more complex and difficult to solve.
+Assumptions about system behavior also become more inaccurate and incomplete as this behavior grows in complexity.
+As mathematical models are based on assumptions about the system's behavior and performance, incorrect, incomplete, or inaccurate assumptions can in turn lead to erroneous predictions.
+Additionally, complex systems may involve non-linear interactions between components, further increasing the difficulty of accurately modeling these systems.
+
+\medskip
+
+The second possible approach to performance evaluation of networked systems is simulation.
+This involves creating a computational model of the system, and then using it to simulate a series of scenarios to measure the model's performance.
+Simulations are often used to validate mathematical models and further narrow down the problem and solution spaces.
+
+Simulations offer numerous advantages.
+They represent an inexpensive and reliable way to study systems, by providing safe, controlled, and repeatable environments in which to test the system's behavior~\cite{fernandes2017performance}.
+Simulations also allow researchers to evaluate the system's performance with a high degree of accuracy and granularity, as they can approximate the system's behavior at a very fine-grained level without necessarily having to resort to complex mathematical modeling.
+Additionally, simulations can be used to test the impact of changes to the system's design or configuration, allowing researchers and system designers to identify the most effective solutions to improve performance.
+
+Simulation is a widely used and well-established method in networked system research, and there exists a rich literature surrounding tools for its implementation as a methodology~\cite{wehrle2010modeling}.
+The availability of simulation frameworks has made it easier for researchers to evaluate the performance of networked systems controlled environments.
+These frameworks provide a set of tools and libraries that enable researchers to create models of networked systems with very little overhead to simulate their behavior.
+
+Examples of frameworks for network simulation include the widely adopted \acs{OMNET}~\cite{omnet}, \acs{ns2}, and \acs{ns3}~\cite{nsnam} discrete-event simulators.
+\acs{OMNET} is an open-source, modular, and extensible discrete-event simulator that provides an object-oriented framework for modeling and simulating networks, and includes a variety of modules for different network protocols and technologies.
+\acs{ns2}, and \acs{ns3} refer to two generations of the same project.
+These widely popular simulation frameworks provide comprehensive sets of tools for modeling and simulating networks, and include support for a range of network technologies.
+
+Nevertheless, simulations also have their limitations.
+The realism of simulations is inherently limited by the accuracy and completeness of the underlying models used.
+Simulations rely on mathematical models that are an abstraction of the real-world system, and these models are only as accurate as the assumptions made and the data used to construct them.
+Simulations also cannot account for the impact of external factors on the system's performance.
+Real-world systems are often subject to factors that cannot be modeled accurately, such as unpredictable weather conditions, human behavior, or hardware failures.
+Moreover, for complex systems with many interacting components, simulations can become prohibitively computationally expensive and time consuming.
+It is not unusual for simulations of highly complex systems to execute much slower than real-time due to the computational power required to calculate the interactions between elements in the system.
+
+\medskip
+
+The third and final approach to performance evaluation of networked systems is practical experimentation on the real system.
+Such an approach involves measuring the performance of the \emph{real} system using practical, real-world experiments in different scenarios.
+Experimental research is often the final stage in systematic performance benchmarking, employed once the problem space has been reduced to a handful of potential solutions through mathematical modeling and subsequent simulations.
+
+Experimental research provides the most accurate assessment of system performance.
+It allows for an extremely high level of realism and can capture factors that cannot be accurately modeled through mathematical modeling or simulations, such as non-linear behaviors, human behavior, and unexpected events.
+Experimental methods provide direct measures of the system's performance under real-world conditions, enabling researchers to validate the accuracy of their mathematical models and simulations.
+
+However, experimental approaches also come with challenges, such as cost, practicality, repeatability, and replicability.
+Experimental research is often time-consuming and expensive to perform, as its implementation requires access to the real system under study.
+Additionally, experimental research is also limited in terms of replicability and repeatability, particularly for systems subject to external factors or stochastic and chaotic effects.
+This can make it challenging to draw definitive conclusions from experimental research.
+
+When it comes to experimental approaches in research involving humans, there are additional challenges that must be considered beyond those mentioned above.
+Ethical concerns related to participant safety, informed consent, and privacy must be addressed, and studies involving human participants must be reviewed and approved by appropriate ethical review panels\footnote{%
+    In Sweden, ethical review is handled at the state level by the Ethical Review Authority (\emph{Etikprövningsmyndigheten}); in the \acs{USA}, it is handled by each institution's \gls{IRB}.
+}.
+Recruitment and retention of human participants can be challenging, as participants may drop out or not show up for study sessions, leading to data collection difficulties.
+Despite these challenges, research involving human participants provides invaluable insights into the behavior and performance of networked systems and is critical for fully understanding and optimizing the performance of these systems in real-world settings.
 
 \subsection{Edge Computing}
-\glsresetall%
 
@@
 
-\subsection{Performance evaluation of networked systems}
+\subsection{Closed-loop feedback systems enabled by Edge Computing}
+\glsreset{CPS}
+\glsreset{XR}
 
-Performance evaluation is an essential step in designing and optimizing networked systems, in order to ensure efficient and effective operation.
-Such evaluation involves measuring and analyzing system performance under a multitude of scenarios to identify potential bottlenecks.
-In practice, three main approaches exist to the performance evaluation of any complex system; analytical modeling, simulations, and experimental evaluation~\cite{fernandes2017performance}.
-These approaches each have their advantages and disadvantages.
+Closed-loop feedback systems, i.e.\ which involve some sort of feedback loop controlling system behavior, are found in a wide range of applications.
+These systems typically involve real-time monitoring and control of complex physical or human processes, with feedback loops enabling rapid response times and adjustments to changing conditions.
+Edge Computing has emerged as a key enabler of these systems, providing the low-latency, context-aware computing capabilities required for real-time decision-making and control.
 
-Analytical or mathematical modeling involves creating a mathematical model that represents the system's behavior and performance.
-This involves defining the system's components, their interactions, and the performance metrics used to measure the system's performance.
-The mathematical model can be solved analytically, providing insights into the behavior of the system under different conditions.
-
-Analytical modeling is often used as a first step when attempting to identify possible solutions to a problem and provide an initial reduction of the problem space.
-One of the key advantages of mathematical modeling is that their realism and resulting complexity can be arbitrarily constrained as required.
-This allows researchers to quickly evaluate a large number of scenarios by employing coarse, approximate models to identify and highlight latent issues.
-These models can then be refined to further narrow the solution space.
-This approach is particularly useful for systems with complex interactions between multiple components, where testing every possible scenario in the real world or even in simulations would be impractical or impossible due to time and cost constraints.
-
-However, analytical modeling also has its limitations, in particular with respect to highly complex systems.
-In systems involving a large number of interacting components and processes, the mathematical equations required to model the system become more complex and difficult to solve.
-Assumptions about system behavior also become more inaccurate and incomplete as this behavior grows in complexity.
-As mathematical models are based on assumptions about the system's behavior and performance, incorrect, incomplete, or inaccurate assumptions can in turn lead to erroneous predictions.
-Additionally, complex systems may involve non-linear interactions between components, further increasing the difficulty of accurately modeling these systems.
-
-\medskip
-
-The second possible approach to performance evaluation of networked systems is simulation.
-This involves creating a computational model of the system, and then using it to simulate a series of scenarios to measure the model's performance.
-Simulations are often used to validate mathematical models and further narrow down the problem and solution spaces.
-
-Simulations offer numerous advantages.
-They represent an inexpensive and reliable way to study systems, by providing safe, controlled, and repeatable environments in which to test the system's behavior~\cite{fernandes2017performance}.
-Simulations also allow researchers to evaluate the system's performance with a high degree of accuracy and granularity, as they can approximate the system's behavior at a very fine-grained level without necessarily having to resort to complex mathematical modeling.
-Additionally, simulations can be used to test the impact of changes to the system's design or configuration, allowing researchers and system designers to identify the most effective solutions to improve performance.
-
-Simulation is a widely used and well-established method in networked system research, and there exists a rich literature surrounding tools for its implementation as a methodology~\cite{wehrle2010modeling}.
-The availability of simulation frameworks has made it easier for researchers to evaluate the performance of networked systems controlled environments.
-These frameworks provide a set of tools and libraries that enable researchers to create models of networked systems with very little overhead to simulate their behavior.
-
-Examples of frameworks for network simulation include the widely adopted \acs{OMNET}~\cite{omnet}, \acs{ns2}, and \acs{ns3}~\cite{nsnam} discrete-event simulators.
-\acs{OMNET} is an open-source, modular, and extensible discrete-event simulator that provides an object-oriented framework for modeling and simulating networks, and includes a variety of modules for different network protocols and technologies.
-\acs{ns2}, and \acs{ns3} refer to two generations of the same project.
-These widely popular simulation frameworks provide comprehensive sets of tools for modeling and simulating networks, and include support for a range of network technologies.
-
-Nevertheless, simulations also have their limitations.
-The realism of simulations is inherently limited by the accuracy and completeness of the underlying models used.
-Simulations rely on mathematical models that are an abstraction of the real-world system, and these models are only as accurate as the assumptions made and the data used to construct them.
-Simulations also cannot account for the impact of external factors on the system's performance.
-Real-world systems are often subject to factors that cannot be modeled accurately, such as unpredictable weather conditions, human behavior, or hardware failures.
-Moreover, for complex systems with many interacting components, simulations can become prohibitively computationally expensive and time consuming.
-It is not unusual for simulations of highly complex systems to execute much slower than real-time due to the computational power required to calculate the interactions between elements in the system.
-
-\medskip
-
-The third and final approach to performance evaluation of networked systems is practical experimentation on the real system.
-Such an approach involves measuring the performance of the \emph{real} system using practical, real-world experiments in different scenarios.
-Experimental research is often the final stage in systematic performance benchmarking, employed once the problem space has been reduced to a handful of potential solutions through mathematical modeling and subsequent simulations.
-
-Experimental research provides the most accurate assessment of system performance.
-It allows for an extremely high level of realism and can capture factors that cannot be accurately modeled through mathematical modeling or simulations, such as non-linear behaviors, human behavior, and unexpected events.
-Experimental methods provide direct measures of the system's performance under real-world conditions, enabling researchers to validate the accuracy of their mathematical models and simulations.
-
-However, experimental approaches also come with challenges, such as cost, practicality, repeatability, and replicability.
-Experimental research is often time-consuming and expensive to perform, as its implementation requires access to the real system under study.
-Additionally, experimental research is also limited in terms of replicability and repeatability, particularly for systems subject to external factors or stochastic and chaotic effects.
-This can make it challenging to draw definitive conclusions from experimental research.
-
-When it comes to experimental approaches in research involving humans, there are additional challenges that must be considered beyond those mentioned above.
-Ethical concerns related to participant safety, informed consent, and privacy must be addressed, and studies involving human participants must be reviewed and approved by appropriate ethical review panels\footnote{%
-    In Sweden, ethical review is handled at the state level by the Ethical Review Authority (\emph{Etikprövningsmyndigheten}); in the \acs{USA}, it is handled by each institution's \gls{IRB}.
-}.
-Recruitment and retention of human participants can be challenging, as participants may drop out or not show up for study sessions, leading to data collection difficulties.
-Despite these challenges, research involving human participants provides invaluable insights into the behavior and performance of networked systems and is critical for fully understanding and optimizing the performance of these systems in real-world settings.
-
-\subsection{Closed-loop feedback systems enabled by Edge Computing}
-
-\todo[inline]{Write an intro?}
+In this section, we will discuss two key examples of such systems, \glspl{CPS} and \gls{XR} applications.
+\glspl{CPS} are systems that integrate physical and computational components, while \gls{XR} applications involve the creation of immersive virtual environment that combine the physical and virtual worlds.
+Both types of systems require real-time monitoring and control of complex processes through feedback loops of either sensor inputs and control outputs in the case of \glspl{CPS}, and video and image streams in the case of \gls{XR}.
+This section will introduce the key characteristics of these systems, providing a thorough background to understand the opportunities and challenges they present.
 

### 2023-04-10 body/part1/chapter2-relatedwork/scope.tex
background

@@
 \section{Scope of the thesis}
+\glsresetall%
 
@@
 
+This dissertation focuses on the performance evaluation of networked systems in the context of \glspl{CPS} and \gls{MAR} applications enabled by low-latency distributed computing paradigms.
+Particular emphasis will be put on the role of Edge Computing as an enabling technology for these categories of systems and applications.
+
+As discussed in \cref{sec:intro}, \glspl{CPS} and \gls{MAR} applications have both garnered the fascination of the research community due to their transformative potential and unique technical requirements.
+\glspl{CPS} have the potential to transform entire industries through real-time monitoring and control and hitherto inviable optimization of physical processes.
+\gls{MAR} applications are poised to transform the way we interact with the world around us, enabling novel forms of entertainment, education, training, and communication.
+Nevertheless, both these classes of systems have stringent latency and context-awareness requirements which until recently had not been met by existing distributing computing paradigm.
+
+Edge Computing plays therefore a crucial role in enabling these systems and applications.
+Edge Computing enables low to ultra-low latencies while providing the necessary compute capabilities and context-sensitivity necessitated by \glspl{CPS} and \gls{MAR}.
+However, Edge Computing represents a challenging area to benchmark, as the tight integration between compute and communication in these systems results in highly complex behaviors.
+Unlike traditional
+
+\todo[inline]{WIP}
+
+% Unlike traditional computing systems, edge computing involves a distributed network of devices that work together to perform complex tasks.
+This creates significant challenges for benchmarking, as it is difficult to isolate individual components and measure their performance in isolation.
+
+% Additionally, the performance of edge computing systems is highly dependent on the communication between devices, which can be impacted by a wide range of factors, including network topology, traffic patterns, and interference.
+As a result, benchmarking edge computing systems requires careful consideration of these factors, as well as the design and implementation of benchmarking tools that can accurately capture the complex interactions between compute and communication.
+Ultimately, benchmarking edge computing systems will be critical for ensuring that they meet the performance requirements of real-world applications and can deliver the benefits of low-latency, context-aware computing at scale.
+
+% By processing data locally on edge devices, rather than sending it to centralized cloud servers, edge computing can significantly reduce the latency of CPS and MAR applications.
+This low-latency is particularly important in real-time applications that require rapid decision-making and response times, such as autonomous vehicles and remote surgery.
+In addition, edge computing provides context-awareness by enabling data to be processed in close proximity to the source, allowing for more efficient and effective use of data and reducing the need for bandwidth-intensive data transfers.
+Overall, the combination of low-latency and context-awareness provided by edge computing is essential for the successful deployment and operation of CPS and MAR applications in a wide range of industries and use cases.
+
+% CPS have the ability to transform entire industries by optimizing physical processes, providing real-time monitoring and control, and making decisions based on data collected from various sources.
+In addition, the low-latency requirements of CPS enable rapid decision-making and response times, making them ideal for safety-critical applications.
+On the other hand, AR applications have the potential to transform how we interact with the world around us, enabling new forms of entertainment, education, and communication.
+The low-latency requirements of AR applications are also critical, as even a small delay in response time can lead to a subpar user experience.
+Overall, the study of CPS and AR applications presents exciting opportunities to explore cutting-edge technologies and their potential to transform various industries and aspects of our daily lives.
+
+% The dissertation aims to investigate the challenges associated with the performance evaluation of networked systems and to propose novel methodologies and tools to assess the performance of such systems in real-world scenarios.
+The research will focus on the evaluation of cyber-physical systems and augmented reality applications that rely on low-latency distributed computing paradigms, such as fog computing and edge computing.
+
+% The research will also explore the role of edge computing as an enabling technology for such applications, and will investigate the benefits and limitations of edge computing in terms of performance, scalability, and reliability.
+The dissertation will propose new approaches for optimizing the performance of edge computing-based systems, and will evaluate their effectiveness in real-world scenarios.
+
+% Overall, the scope of this dissertation is to contribute to the body of knowledge on the performance evaluation of networked systems, with a focus on cyber-physical systems and augmented reality applications enabled by low-latency distributed computing paradigms, and the role of edge computing as an enabling technology.
+
 \begin{enumerate}

### 2023-04-12 body/part1/chapter2-relatedwork/relatedwork.tex
writing

--- 2023-04-03 body/part1/chapter2-relatedwork/relatedwork.tex
@@
-\section{Related work}
+\section{State of the art}
 
-\subsection{Emulation in \glsfmtshort{XR} research}
+\subsection{Performance evaluation of Edge Computing systems}
 
-Only a limited number of works addressing the partial or complete emulation of XR applications for benchmarking and research purposes exist in literature.
-One such work is~\cite{williams2013transform}.
-In this, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
-The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.\@
-\cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{AR} application by mocking Android \gls{SDK} \glspl{API}.
-This approach allows for benchmarking and evaluation of algorithms and architectures on real data.
-A similar approach was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
-Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
-Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
+\todo[inline]{
+    Needs some massaging.
+}
+\todo[inline]{
+    Right now includes only experimental approaches.
+    Need to add mathematical models (ask James for citations) and simulations.
+}
 
-While the above approaches to benchmarking and research in \gls{XR} are steps in the right direction, they fail to consider the dynamic nature of \gls{XR}.
-The use of rigid traces in trace-based approaches can lead to timing mismatches when replaying at runtime, as the system may respond differently to the same input depending on factors such as system load and network congestion.
-Additionally, these approaches do not take into consideration the effects of system responsiveness on human behavior, which can be critical for evaluating both \gls{QoS} and \gls{QoE}.
-To address these issues, future approaches will need to adapt trace replay at runtime to match expected behaviors and consider the impact of system responsiveness on user behavior.
-By doing so, we can develop more accurate and reliable benchmarks for \gls{XR} applications and improve our understanding of how users interact with these technologies.
+\subsubsection{Analytical modeling and simulation approaches}
 
-\subsection{Edge and Cloud Benchmarking}
+The performance evaluation of Edge Computing deployments is a challenging task, with only a limited number of works attempting to tackle it through analytical models or simulations.
+One reason for this is the inherent complexity of modeling such intricate systems, which often involve multiple layers of proxies, load balancers, virtualized computing instances, and networking functions.
+The dynamic nature of Edge Computing deployments, with their distributed architectures and diverse set of devices, makes it difficult to develop accurate models that capture the complexities of real-world deployments.
 
-This dissertation complements existing work on benchmarking distributed system architectures.
-A rich body of literature exists with examples of such approaches for Cloud computing.
-For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
-In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
-The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
-\cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major Cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
+Most of the existing approaches for performance evaluation in Edge Computing have originated from the field of network performance evaluation.
+These approaches often focus on specific use-cases and simplify the processing and compute steps by modeling them as black boxes with fixed or simple distributions of latency.
 
-As evidenced by the given examples, existing approaches to Cloud benchmarking tend to focus heavily on large-scale data processing and downlink transmission in Cloud scenarios.
-This makes them unsuitable for Edge Computing, as they fail to account for important characteristics of this paradigm, such as latency-sensitivity of workloads and uplink-heavy network traffic.
+\todo[inline]{WIP}
 
-The limitations of previous approaches has in later years led to a number of works that have attempted to tackle the challenge of specifically benchmarking Fog and Edge deployments as well.
-Even so, these works have generally leveraged static workloads that fail to account for the intensely dynamic nature of \gls{CPS} deployments.
-Examples of approaches which fall into this category include
+% This simplistic approach may not accurately reflect the true performance characteristics of Edge Computing deployments, which involve dynamic interactions between software processes and hardware and network infrastructure.
+
+% Despite these challenges, recent research has made progress in developing more sophisticated models and simulations for performance evaluation in Edge Computing.
+These approaches aim to capture the complex dynamics of Edge Computing deployments, including the variability in processing and compute times, network delays, and resource availability.
+By considering the intricacies of Edge Computing deployments, these advanced models can provide more accurate and comprehensive evaluations of performance, enabling researchers and practitioners to gain deeper insights into the behavior of latency-sensitive applications in Edge Computing environments.
+
+\subsubsection{Experimental research}
+
+% This dissertation complements existing work on benchmarking distributed system architectures.
+% A rich body of literature exists with examples of such approaches for Cloud computing.
+% For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
+% In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
+% The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
+% \cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major Cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.
+
+% As evidenced by the given examples, existing approaches to Cloud benchmarking tend to focus heavily on large-scale data processing and downlink transmission in Cloud scenarios.
+% This makes them unsuitable for Edge Computing, as they fail to account for important characteristics of this paradigm, such as latency-sensitivity of workloads and uplink-heavy network traffic.
+
+A number of works have attempted to tackle the challenge of experimentally evaluation performance of Fog and Edge deployments, such as 
 \begin{inlineenum}
@@
 
-Furthermore, although some of these approaches address some of the challenges to benchmarking \gls{XR} on the Edge, none of them have tackled the topic in a comprehensive way.\@
+Although valuable contributions, these works have generally leveraged static workloads that fail to account for the intensely dynamic nature of \gls{CPS} and \gls{XR} deployments.
+Furthermore, although some of these approaches address some of the additional challenges to benchmarking \gls{XR} on the Edge, none of them have tackled the topic in a comprehensive way.\@
 \cite{das2018edgebench,mcchesney2019defog,baurle2022comb,george2020openrtist} all target image or video recognition or manipulation in some manner.
@@
 
-\subsection{Experimental research in \glsfmtshortpl{NCS}}
-
-A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
-They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
-
-Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies~\cite{zhang2019networked}.
-\gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
-However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
-
-Experimental research in \glspl{NCS} involves a diverse range of hardware and software platforms, methodologies, and key performance indicators.
-Specifically, we identify three distinct categories of approaches: fully practical, fully simulated, and emulated approaches that combine components of the first two.
-
-In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
-This approach is straightforward to implement, and can be very useful for evaluating the real-world applicability of the system, but can be costly and time-consuming.
-Examples of the application of this approach can be found in studies such as~\cite{drew2005networked,baumann2018evaluating,li2014wireless,cuenca2019periodic}.
-These studies implemented wireless and microcontroller-based systems for control and validated them on a physical prototype.
-
-In~\cite{drew2005networked}, the authors present a \gls{NCS} model which considers both random packet delay and loss on both the sensor and actuator sides of the feedback loop.
-They validate their design on a physical testbed consisting of inverted pendula controlled from a central computation point over Wi-Fi.
-A similar testbed is employed for validation in~\cite{baumann2018evaluating}, in which the authors develop an evaluation methodology for wireless \gls{NCS}.\@
-\cite{li2014wireless} implements a wireless microcontroller-based system for vibration control and validates it on a physical prototype comprising a cantilever beam controlled using an \acs{IEEE} 802.15.4 \gls{WPAN} system.\@ \cite{cuenca2019periodic} discusses the design and implementation of a periodic event-triggered sampling and dual-rate control techniques for wireless control, and validate their contributions on a four-rotor autonomous drone platform controlled over WiFi.
-\emph{NCSbench}~\cite{zoppi2020ncsbench}, an open-source \gls{NCS} benchmarking platform designed with reproducibility in mind built using the \citetitle{LEGOMindstormsEV3}~\cite{LEGOMindstormsEV3} platform, is another such example.
-
-Completely simulated setups, on the other hand, rely on computer simulations to evaluate the \gls{NCS}'s performance.
-This approach is less expensive and faster than the practical approach but may not fully capture the real-world complexities of the system.
-Examples of tools and frameworks for the implementation of this approach can be found in works such as~\cite{andersson2005simulation,eyisi2012ncswt}.
-
-A simulated approach is employed in~\cite{du2009novel}, in which the authors develop a novel \emph{Smith} predictor to compensate for varying latencies in wireless \glspl{NCS} and validate it using the \emph{TrueTime}~\cite{henriksson2002truetime} simulator.
-In~\cite{chen2015synchronous} the authors validate and evaluate a number of synchronous control strategies for three-motor setups using completely simulated \glspl{NCS} environments.\@
-\cite{wu2012application} introduces the concept of \emph{network predictive control} in and uses it to balance a simulated inverted pendulum over a simulated wireless network.\@
-\cite{ma2019optimal} proposes an optimal dynamic scheduling strategy that optimizes performance of multi-loop control systems and validate it by simulating a four-loop control system over an \acs{IEEE} 802.15.4 \gls{WPAN}.
-
-Finally, network and hardware-in-the-loop approaches, which aim to strike a balance between realism and efficiency, have also been employed in literature.
-Network-in-the-loop refers to experimental setups in which the \emph{network} is instead simulated or emulated.
-Such approaches have, for instance, been used to study the effects of the \gls{CSMA/CD} medium access control mechanism on plant stability, as in~\cite{natale2004inverted}.
-Hardware-in-the-loop approaches, referring to \gls{NCS} setups in which a real network interacts with a simulated or emulated control system, are particularly prevalent in the field of \emph{smart grid} control.
-An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
-
-\medskip
-
-The diversity of experimental \glspl{NCS} research discussed above has led to fragmentation of hardware, software, and methodology, as studies tend to favor approaches popular in their respective communities.
-Additionally, studies tend to focus on individual aspects and components of the system, resulting in incomplete knowledge of the overall \glspl{NCS}.
-There is a gap in knowledge concerning the reproducibility and comparison of experimental studies on these systems.
-
-To our knowledge, the only published attempt at addressing methodological fragmentation and poor reproducibility of experimental \gls{NCS} research is the LEGO-based approach introduced in~\cite{zoppi2020ncsbench}.
-However, this approach has limitations.
-It is dependent on specific hardware --- the LEGO Mindstorms set --- which makes its adoption limited due to availability issues, reduces its flexibility, scalability, and limits the scope of experiments able to be conducted via \emph{NCSbench} only to what is possible to implement in LEGO.\@
-Scaling up experiments becomes cumbersome and costly, further limiting the practicality of the tool.
-Thus, a more general approach that tackles the scalability and reproducibility issues of NCS research is urgently needed.
-The development of such an approach would enable researchers to conduct experiments that generate reproducible results consistently and can be compared across different studies while still retaining the computational sophistication required when scaling up experiments.
-
-\subsection{Testbed research in Edge Computing}
-
-The last few years have seen the emergence of a number of small- to mid-scale platforms and testbeds for Edge Computing research.
+\paragraph{Testbed research in Edge Computing} The last few years have also seen the emergence of a number of small- to mid-scale platforms and testbeds for Edge Computing research.
 Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
@@
 
-\gls{GENI} is a testbed deployed across more than fifty sites across the \gls{USA}.
-Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local Edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling Edge Computing platform.
-
+\gls{GENI}~\cite{berman2014geni,gosain2017geni} is a testbed deployed across more than fifty sites across the \gls{USA}.
+It combines local Edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation, with wireless-capable sites incorporate single-hop access to compute, storage, and network resources.
 The Chameleon testbed is a research platform originally designed for the study and evaluation of Cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
-Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various Cloud computing technologies, algorithms, and configurations.
-The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various Cloud and Edge Computing scenarios.
-
-\gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
-The testbeds main \emph{raison d'être} is \gls{MEC} research, with a strong emphasis on wireless networks.
+The testbed provides a flexible and reconfigurable infrastructure for researchers to experiment with hardware and software resources, including servers, storage systems, and networking equipment.
+\gls{COSMOS} is a \gls{MEC} testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
 The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, Cloud integration, and Edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
-
-\gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
-It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
+\gls{POWDER}~\cite{breen2020powder} features roughly \num{15} fixed programmable radio nodes distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
 These radio nodes are combined with Edge-based compute nodes as well as Cloud resources for full \gls{MEC} experimentation.
-
-The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
-\gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-Its core goal is the study and deployment of advanced Edge Computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-
-EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for Edge Computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
-It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
-The EdgeNet software is available for anyone to download;
-individuals and organizations can quickly deploy self-hosted Edge nodes which are then made available to anyone using the platform.
-The end result is a flexible and scalable global Edge Cloud simplifying research and prototype development in Edge and Cloud computing.
+\gls{ARA} includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors and and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara} in Iowa, \gls{USA}.
+Its core goal is the study and deployment of advanced Edge Computing and wireless platforms and technologies in real-world agricultural and rural settings.
+EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for Edge Computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2} based on the widely adopted Kubernetes~\cite{kubernetes} framework.
 A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the Cloud to the Edge.
@@
 The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
+
+% Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local Edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
+% The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling Edge Computing platform.
+
+% The Chameleon testbed is a research platform originally designed for the study and evaluation of Cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
+% Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various Cloud computing technologies, algorithms, and configurations.
+% The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various Cloud and Edge Computing scenarios.
+
+% \gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
+% The testbeds main \emph{raison d'être} is \gls{MEC} research, with a strong emphasis on wireless networks.
+% The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, Cloud integration, and Edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
+
+% \gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
+% It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
+% These radio nodes are combined with Edge-based compute nodes as well as Cloud resources for full \gls{MEC} experimentation.
+
+% The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
+% \gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
+% Its core goal is the study and deployment of advanced Edge Computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
+
+% EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for Edge Computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
+% It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
+% The EdgeNet software is available for anyone to download;
+% individuals and organizations can quickly deploy self-hosted Edge nodes which are then made available to anyone using the platform.
+% The end result is a flexible and scalable global Edge Cloud simplifying research and prototype development in Edge and Cloud computing.
+% A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the Cloud to the Edge.
+
+% Finally, a number of other, small-scale approaches also exist.
+% In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for Edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for Edge Computing research.
+% The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered Edge architectures in the context of \gls{FIWI} Edge Computing.
+% This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
+% In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
+% \cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.\@
+% \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
+% The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
 
@@
 There is still remains much work to be done in this area, particularly with respect to the adoption of Edge-compatible solutions that can enable faster and more reliable experimentation.
+
+\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshort{CPS}}
+
+\subsubsection{Experimental research in \glsfmtlong{MAR}}\label{sec:relwork:xremulation}
+
+\gls{MAR} is difficult to model or simulate due to its dynamic and interactive nature involving ``humans-in-the-loop''.
+\gls{MAR} experiences are highly dependent on real-time interactions between the user, the virtual objects, the communication network, the computation node, and the real-world.
+These interactions are exceedingly challenging to accurately model and reproduce in controlled, simulated, environments.
+As a result, the vast majority of research in this field employs experimental approaches to performance evaluation.
+Although simulation and model-based approaches such as~\cite{liubogoshchev2021adaptive,sundararajan2021performance} exist, they are far and few in between, and generally target specific use-cases.
+Performance evaluation in this field relies instead broadly on real-world experiments and benchmarks are conducted to evaluate the actual performance and user experience of \gls{MAR} applications in real-world conditions, providing more accurate and reliable results.
+
+Examples of such experimental research include~\cite{williams2013transform,munro2016aaremu,george2020openrtist,choi2022emulating,chetoui2022arbench}.
+In~\cite{williams2013transform}, the authors introduced a trace-based toolkit for the evaluation of tracking algorithms in \gls{AR} research and development.
+The trace consists of geographical coordinates and accelerometer data as well as video frames which can be replayed on a desktop visualization tool or a smartphone application for the evaluation of tracking algorithms in \gls{AR} research and development.\@
+\cite{munro2016aaremu} presents a trace-based framework that allows a previously recorded trace of inputs to be replayed to the processing components of a real \gls{MAR} application by mocking Android \gls{SDK} \glspl{API}.
+This approach allows for benchmarking and evaluation of algorithms and architectures on real data.
+The previously mentioned~\cite{george2020openrtist} approximates the workload generated by a \gls{MAR} on an Edge Computing deployment by performing real-time style-transfer on a live video feed. 
+A similar approach to~\cite{munro2016aaremu} was taken by~\cite{choi2022emulating}, but the authors instead provided developers of \gls{AR} with a framework that abstracts input into general classes, allowing for device-agnostic processing.
+Finally, the authors of~\cite{chetoui2022arbench} employ \gls{AR} emulation to evaluate hardware rather than the software implementation itself.
+Their tool, \emph{ARBench}, features emulated \gls{AR} workloads that mimic the load placed on the system by real applications.
+
+While the above approaches to performance evaluation and research in \gls{MAR} are steps in the right direction, they fail to consider the dynamic nature of \gls{MAR}.
+The use of rigid traces in trace-based approaches can lead to timing mismatches when replaying at runtime, as the system may respond differently to the same input depending on factors such as system load and network congestion.
+Additionally, these approaches do not take into consideration the effects of system responsiveness on human behavior, which can be critical for evaluating both \gls{QoS} and \gls{QoE}.
+To address these issues, future approaches will need to adapt trace replay at runtime to match expected behaviors and consider the impact of system responsiveness on user behavior.
+By doing so, we can develop more accurate and reliable benchmarks for \gls{MAR} applications and improve our understanding of how users interact with these technologies.
+
+\subsubsection{Experimental approaches to research in \glsfmtlongpl{NCS}}
+
+A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
+Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
+Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
+They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
+
+Most of the large literature concerning \glspl{NCS}, however, follows a theoretical approach, with only small fraction of it dealing with experimental studies~\cite{zhang2019networked}.
+\gls{NCS} studies are complex due to the inter-domain nature of these systems, requiring expertise in the fields of communication networks, computing, and control theory.
+However, theoretical approaches can only capture network behaviors at a coarse level, and thus a body of works has emerged in later years involving experimental approaches to \gls{NCS} benchmarking.
+
+We identify two main approaches: fully practical and partially-emulated approaches.
+In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
+Examples of the application of this approach can be found in studies such as~\cite{drew2005networked,baumann2018evaluating,li2014wireless,cuenca2019periodic}.
+These studies implemented wireless and microcontroller-based systems for control and validated them on a physical prototype.
+In~\cite{drew2005networked}, the authors present a \gls{NCS} model which considers both random packet delay and loss on both the sensor and actuator sides of the feedback loop.
+They validate their design on a physical testbed consisting of inverted pendula controlled from a central computation point over Wi-Fi.
+A similar testbed is employed for validation in~\cite{baumann2018evaluating}, in which the authors develop an evaluation methodology for wireless \gls{NCS}.\@
+\cite{li2014wireless} implements a wireless microcontroller-based system for vibration control and validates it on a physical prototype comprising a cantilever beam controlled using an \acs{IEEE} 802.15.4 \gls{WPAN} system.\@ \cite{cuenca2019periodic} discusses the design and implementation of a periodic event-triggered sampling and dual-rate control techniques for wireless control, and validate their contributions on a four-rotor autonomous drone platform controlled over WiFi.
+\emph{NCSbench}~\cite{zoppi2020ncsbench}, an open-source \gls{NCS} benchmarking platform designed with reproducibility in mind built using the \citetitle{LEGOMindstormsEV3}~\cite{LEGOMindstormsEV3} platform, is another such example.
+
+Partially-emulated approaches, including network and hardware-in-the-loop approaches and which aim to strike a balance between realism and efficiency, have on the other hand also been employed in literature.
+Network-in-the-loop refers to experimental setups in which the \emph{network} is instead simulated or emulated.
+Such approaches have, for instance, been used to study the effects of the \gls{CSMA/CD} medium access control mechanism on plant stability, as in~\cite{natale2004inverted}.
+Hardware-in-the-loop approaches, referring to \gls{NCS} setups in which a real network interacts with a simulated or emulated control system, are particularly prevalent in the field of \emph{smart grid} control.
+An example of the application of such an approach can be found in~\cite{wang2020inverter}, in which the authors introduce and validate a novel three-level coordinated control method for photovoltaic inverters.
+
+To the best of our knowledge, there exist currently no generic, hardware-independent framework available for repeatable experimental research in Edge-bound \glspl{NCS}.
+While there has been some notable work in this area, such as the work by~\cite{zoppi2020ncsbench}, existing contributions still mostly rely on proprietary and hardware that is difficult to acquire.
+This limitation restricts the accessibility of the framework to researchers and practitioners, hindering the reproducibility and comparability of experimental results.
+Moreover, the existing research's reliance on hardware also poses scalability limitations.
+As the field of Edge Computing-enabled \glspl{NCS} continues to evolve, there is a growing demand for scalable solutions that can be easily adapted to different hardware configurations and setups.
+A software-based framework that provides a flexible and scalable solution for performance evaluation of \glspl{NCS} in Edge-native contexts is currently lacking in the field.

### 2023-04-12 body/part1/chapter2-relatedwork/relatedwork.tex
sim

@@
 
-The performance evaluation of Edge Computing deployments is a challenging task, with only a limited number of works attempting to tackle it through analytical models or simulations.
+The performance evaluation of Edge Computing deployments is a daunting task, and only a limited number of works have attempted to tackle it through analytical models or simulations.
 One reason for this is the inherent complexity of modeling such intricate systems, which often involve multiple layers of proxies, load balancers, virtualized computing instances, and networking functions.
@@
 
-Most of the existing approaches for performance evaluation in Edge Computing have originated from the field of network performance evaluation.
-These approaches often focus on specific use-cases and simplify the processing and compute steps by modeling them as black boxes with fixed or simple distributions of latency.
+\todo[inline]{Analytical approaches}
+
+A number of approaches to performance evaluation of distributed systems exist in literature, particularly geared towards the Cloud.
+Examples of such approaches include the \emph{CloudSim}, \emph{iCanCloud}, and \emph{GreenCloud} simulators.
+CloudSim~\cite{calheiros2011cloudsim} is a widely used event-based simulation toolkit designed for generic \gls{IaaS} Cloud Computing environments.
+iCanCloud~\cite{nunez2012icancloud} similarly focuses on simulating \gls{AWS} \gls{EC2} compute infrastructure, the project's primary objectives being performance and scalability of simulations.
+GreenCloud, on the other hand, is built upon the \acs{ns2}~\cite{nsnam} network simulator and specifically captures energy consumption details of datacenter components and communication patterns between these.
+However, these simulators are not suitable for Edge Computing research as they do not fully capture the unique characteristics of Edge environments.
+They do not account for the tight integration of compute and network in Edge Computing and only coarsely approximate network effects on workloads.
+In these Cloud Computing simulators, end-to-end transmission delays are often treated as static values;
+however, the dependency on wireless interfaces in Edge Computing requires a much more detailed modeling of the network.
+This includes considerations for mobility, a key characteristic of users in an Edge context which is not accounted for in any of the above tools.
+Moreover, existing models for Cloud workloads are unsuitable for Edge Computing research, as Edge workloads present compute characteristic which are wholly distinct from traditional Cloud workloads.
+Cloud workloads tend to be stateless and long-lived, whereas Edge workloads are context-sensitive and ephemeral.
+
+% \cite{sonmez2018edgecloudsim}: fully integrated simulation. Modular. Based on CloudSim. FOSS.
+
+% Most of the existing approaches for performance evaluation in Edge Computing have originated from the field of network performance evaluation.
+% These approaches often focus on specific use-cases and simplify the processing and compute steps by modeling them as black boxes with fixed or simple distributions of latency.
 

### 2023-04-14 body/part1/chapter2-relatedwork/scope.tex
working on scope

--- 2023-04-09 body/part1/chapter2-relatedwork/scope.tex
@@
 
+\todo[inline]{
+    Write about in terms of relevance and challenges.
+    Motivate a bit. And then tell what we're not interested in.
+
+    First 3 paragraphs in a conventional papers are usually scope. 
+}
+
 \begin{enumerate}
@@
 \end{enumerate}
+
+\glsresetall%
+\glspl{CPS} and \gls{XR} applications are emerging technologies with the potential to transform a plethora of industries.
+However, both these categories of applications have stringent latency requirements which make their deployment on traditional Cloud infrastructure unfeasible.
+Delays in communication and processing can significantly impact their performance and user experience, and potentially even cause physical harm.
+Cloud Computing, with its centralized, physically distant datacenters, may not be able to meet these strict latency requirements.
+This is where Edge Computing, a novel paradigm for distributed and mobile computing, offers a promising solution.
+Edge Computing enables deployment of computing resources at the edge of the network, closer to end-users.
+This shift in paradigm with respect to Cloud Computing provides drastically reduced latency, bandwidth reduction, as well as context-sensitivity and data locality.
+These are key factors for real-time, context-sensitive, low-latency applications such as \gls{CPS} and \gls{XR}.
+
+Before these systems can become widely deployed, however, their real-world performance must be thoroughly understood.
+\glspl{CPS} often operate in safety-critical environments, and poor \gls{QoS} in \gls{XR} can easily lead to users experiencing discomfort while using the application.
+Accurate evaluation, characterization, and optimization of the performance of these systems is therefore critical to ensure reliability, safety, and user satisfaction.
+Additionally, accurate performance evaluation enables efficient resource utilization, and enhances system scalability, in turn leading to lower operation costs and a reduced barrier of entry.
+
+Nevertheless, evaluating the performance of Edge environments involving such applications is challenging, due to complex behaviors emerging from the interaction between computation and network in these systems.
+Existing literature has attempted to tackle the performance evaluation of Edge systems through both analytical and simulation approaches.
+However, the nature of Edge Computing environments, which involves tight integration between network and compute, vast and often unfeasible interdisciplinary competence.
+Existing works have been constrained to narrow their scope by prioritizing the comprehensive modeling of either compute or network, resulting in a potential trade-off that may neglect the other aspect.
+This has resulted in an incomplete understanding of the overall performance of Edge systems and applications in literature.
+
+In this dissertation we introduce and study the applicability of an alternative methodology for Edge performance evaluation based on a practical, experimental approach.
+We aim to complement literature by providing a method which bridges the gap in knowledge left by existing approaches.
+Or work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
+
+However, the performance evaluation of Edge-enabled \gls{XR} encompasses a wide range of factors, making the scope of research in this area extensive and complex.
+The scope of performance evaluation research for \glspl{CPS} is at least equally extensive.
+Attempting to cover both these scopes comprehensively in a single dissertation would be prohibitively challenging.
+Therefore, in this work, we approach these challenges by focusing on representative use-cases for these application classes.
+These correspond to \gls{WCA} applications for \gls{XR}, and \glspl{NCS} for \glspl{CPS}.
+Through these examples, we aim to gain insights into the performance evaluation challenges and opportunities offered by Edge Computing these applications and systems, and contribute to the understanding of the broader fields of \gls{XR} and \gls{CPS} performance evaluation in distributed computing environments.

### 2023-04-16 body/part1/chapter2-relatedwork/relatedwork.tex
working on scope
rw

@@
-\section{State of the art}
+\section{State of the art}\label{sec:state-of-the-art}
 
@@
 
-The performance evaluation of Edge Computing deployments is a daunting task, and only a limited number of works have attempted to tackle it through analytical models or simulations.
+The performance evaluation of Edge Computing deployments is a daunting task.
 One reason for this is the inherent complexity of modeling such intricate systems, which often involve multiple layers of proxies, load balancers, virtualized computing instances, and networking functions.
@@
 
-\todo[inline]{Analytical approaches}
+Nevertheless, a small body of works has attempted to tackle this challenge through analytical approaches.
+Examples include~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}.
+In~\cite{chen2015efficient}, the authors formulate the problem of computation offloading in \gls{MEC} as a multi-user computation offloading game, and design a distributed computation offloading algorithm that achieves a Nash equilibrium to solve it.
+They demonstrate through numerical results that the proposed algorithm achieves good computation offloading performance and has good scaling capabilities.
+Another work that tackles computation offloading is~\cite{champati2015one}, in which the authors propose a novel optimization algorithm for the cost optimization of hybrid clouds.
+The algorithm selects tasks for remote computation on the public cloud based on their estimated processing times and the known cost of offloading them; tasks with large processing times that offset their offloading cost are offloaded.\@
+\cite{champati2016semi} proposes semi-online algorithms to to minimize the time required to complete all computational tasks in a sequence, with communication delay, in Cloud Computing environments.
+This is achieved by optimizing the scheduling of independent tasks with unknown processing times.
+Finally, the authors of~\cite{al2017reliable} propose a model-based rate adaptation algorithm for scalable video streaming in multihop wireless networks.
+The algorithm is built on a novel performance model based on stochastic network calculus and allows fast near-optimal rate adaptation for fixed transmission paths and cross-layer optimized routing.
 
-A number of approaches to performance evaluation of distributed systems exist in literature, particularly geared towards the Cloud.
+Although these works represent valuable contributions to literature, they present significant limitations to their realism and applicability to \glspl{CPS} and \gls{XR} enabled by Edge Computing.
+In~\cite{chen2015efficient,champati2016semi}, the authors assume that task computation times are unknown, and estimate them based on the size of their associated data.
+However, applications such as \glspl{NCS} and \gls{MAR} tend to have mostly constant data sizes, with computation times depending strictly upon the semantic content of the inputs.\@
+\cite{champati2016semi,al2017reliable} assume independent tasks, but in interactive, feedback-loop-based applications, tasks are intrinsically ordered and dependent on previous tasks.
+In~\cite{champati2015one}, the authors propose canceling and re-scheduling a task as a viable scheduling strategy; however, in \glspl{CPS} the canceling of a control task could lead to catastrophic harm to equipment and/or human beings.
+
+Approaches to performance evaluation of distributed systems using simulations also exist in literature, particularly geared towards the Cloud.
 Examples of such approaches include the \emph{CloudSim}, \emph{iCanCloud}, and \emph{GreenCloud} simulators.
@@
 
-% \cite{sonmez2018edgecloudsim}: fully integrated simulation. Modular. Based on CloudSim. FOSS.
+A number of simulation approaches which aim to overcome these limitations have therefore emerged in later years.
+iFogSim~\cite{gupta2017ifogsim} is an extension of CloudSim~\cite{calheiros2011cloudsim} geared towards the simulation of \gls{IOT} fog environments by modeling such components as sensors, actuators, Fog nodes, as well as the Cloud.
+Although iFogSim allows for location-aware servicing of user-devices by Fog nodes, this information is static during a simulation, and no mobility of users is considered.
+Network links in iFogSim also do not account for the effects on network load on network times, further limiting the simulator's realism.\@
+EdgeCloudSim~\cite{sonmez2018edgecloudsim} also extends CloudSim~\cite{calheiros2011cloudsim} and provides a much more comprehensive simulation environment for Edge deployments, which includes mobility, network, and edge compute models.
+However, the models chosen limit the realism of the tool, as they may not be applicable to all applications classes.
+The compute model, for instance, assumes offloading to \glspl{VM}, which does not match the current broad understanding that Edge Computing applications will rely on more flexible and lightweight forms of offloading, such as serverless functions and containers.
+Moreover, the tool is highly complex and requires extensive domain knowledge in both communication networks and compute modeling to use effectively.
 
-% Most of the existing approaches for performance evaluation in Edge Computing have originated from the field of network performance evaluation.
-% These approaches often focus on specific use-cases and simplify the processing and compute steps by modeling them as black boxes with fixed or simple distributions of latency.
-
-\todo[inline]{WIP}
-
-% This simplistic approach may not accurately reflect the true performance characteristics of Edge Computing deployments, which involve dynamic interactions between software processes and hardware and network infrastructure.
-
-% Despite these challenges, recent research has made progress in developing more sophisticated models and simulations for performance evaluation in Edge Computing.
-These approaches aim to capture the complex dynamics of Edge Computing deployments, including the variability in processing and compute times, network delays, and resource availability.
-By considering the intricacies of Edge Computing deployments, these advanced models can provide more accurate and comprehensive evaluations of performance, enabling researchers and practitioners to gain deeper insights into the behavior of latency-sensitive applications in Edge Computing environments.
+\todo[inline]{Wrap up.}
 

### 2023-04-16 body/part1/chapter2-relatedwork/scope.tex
rw

@@
 
-\todo[inline]{
-    Emphasize that apps are standing for a new class of application class that the community is interested in developing.
-    Why?
-    Utility.
-    Weave into intro and scope.
+% \todo[inline]{
+%     Emphasize that apps are standing for a new class of application class that the community is interested in developing.
+%     Why?
+%     Utility.
+%     Weave into intro and scope.
 
-    Not just a random mashup.
-    Relevance of the combination!
+%     Not just a random mashup.
+%     Relevance of the combination!
 
-    Prime reason for experimental evaluation.
-    Capture effects of compute, network, and design of the application.
-    Hard to capture otherwise.
+%     Prime reason for experimental evaluation.
+%     Capture effects of compute, network, and design of the application.
+%     Hard to capture otherwise.
 
-    Build argument in state of the art that there is no current satisfying way to address this.
-    Potentially orders of magnitude off.
+%     Build argument in state of the art that there is no current satisfying way to address this.
+%     Potentially orders of magnitude off.
 
-    In scope: interested in meaningful performance evaluation (i.e.\ capture all the complexity!) Make sure we note this.
-    Show how apps and system are intimately related to each other and complex to evaluate.
-    Where do complexities arise.
-    Understanding system performance is key and complex.
+%     In scope: interested in meaningful performance evaluation (i.e.\ capture all the complexity!) Make sure we note this.
+%     Show how apps and system are intimately related to each other and complex to evaluate.
+%     Where do complexities arise.
+%     Understanding system performance is key and complex.
 
-    Scope: define the challenges.
-}
+%     Scope: define the challenges.
+% }
 
-\todo[inline]{
-    Write about in terms of relevance and challenges.
-    Motivate a bit. And then tell what we're not interested in.
+% \todo[inline]{
+%     Write about in terms of relevance and challenges.
+%     Motivate a bit. And then tell what we're not interested in.
 
-    First 3 paragraphs in a conventional papers are usually scope. 
-}
+%     First 3 paragraphs in a conventional papers are usually scope. 
+% }
 
-\begin{enumerate}
-    \item Performance evaluation.
-    \item \gls{CPS} and \gls{MAR}
-    \item Edge Computing 
-\end{enumerate}
+% \begin{enumerate}
+%     \item Performance evaluation.
+%     \item \gls{CPS} and \gls{MAR}
+%     \item Edge Computing 
+% \end{enumerate}
 

### 2023-04-17 body/part1/chapter2-relatedwork/key_contributions.tex
related work

@@
-\section{Overview of the included publications}
+\section{Overview of our contributions}\label{sec:contributions}
 
-\todo[inline]{
-    first main contribution: introduce the idea of emulating the physical part of a CPS when it comes to experimental evaluation running over an integrated communication and computation infrastructure.
-    when we do that, we avoid many of the complications of experimental evaluation of these systems
-    in combination with Ainur, is a very elegant proposition
-    what is not currently captured precisely is that the approach is for the whole system (app + infra)
-    Need to mention the emulation
-    elegant and contemporary approach to efficient and repeatable evaluation of CPS
+This thesis presents two key contributions to literature.
+First, we introduce and study a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
+Our approach is based on the novel idea of emulating \emph{only} the end-user component of such applications while using the \emph{real} network and compute components of the system.
+In other words, our methodology involves replacing the client side of the system with commercial off-the-shelf (COTS) general-purpose computing devices running realistic software imitations of the desired behaviors, while retaining the real network and compute hardware at the backend.
+This entails modeling and emulating human inputs in \gls{MAR} applications and the physical components of \glspl{CPS}.
 
-    second contribution:
-    question of to which degree the emulation needs to capture certain details of the physical world?
-    to which degree is emulation ``good enough''?
-    for WCA specifically we propose a relatively detail of human user behavior based on empirical data
-    evaluate to which extent the emulation is ``good enough'' by optimization of these systems
-
-    Mention software, then we are on the safe.
-}
-
-\Cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} in this dissertation present and study the application of our methodology for \gls{WCA}.
+We study and validate our methodology by example, investigating its applicability in two representative use-cases for \gls{XR} applications and \glspl{CPS}, respectively: \acf{WCA} and \acfp{NCS}.
 \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} discuss initial explorative studies into the application of our methodology for \gls{WCA}.
@@
 It uses a trace of inputs recorded in optimal conditions, the replay of which is controlled at run-time by a simple model of human behavior based on a \gls{FSM}.
-In terms of hardware, the setup is composed of Android client devices connected to a \gls{COTS} {x86--64} server over a Wi-Fi network.
 
-In \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deeper into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
-\cref{paper:olguinmunoz2021impact} presents a characterization of human behavior in \gls{WCA}.
-Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
-We develop a characterization through a human subject study in which participants interact with an instrumented \gls{WCA}.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} finally presents two contributions.
-Firstly, it introduces the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-Secondly, it explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
-
-\medskip
-
-This line of work has some limitations.
-First of all, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
-We focus on this category for a couple of reasons.
-Step-based task-assistance \gls{WCA} was among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represents a more mature application class compared to other \glspl{WCA}.
-These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
-
-However, restricting our research to step-based \gls{WCA} means that our results may not be directly applicable to other types of \gls{WCA} and \gls{AR} applications.
-In particular, our characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
-Additionally, the timing characteristics of step-based \gls{WCA} are such that an instruction remains valid for as long as it takes the user to perform it.
-The conclusions from these works therefore may not apply to applications with different execution flows, or in which instructions have a deadline after which they may become stale or invalid.
-Examples of such applications include, for instance, the \emph{Ping-Pong} assistant introduced in~\cite{chen2015early}.
-
-The user models are further limited in their realism.
-In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we assume a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-These limitations emerge from a lack of data of human behavior in these applications, as at the moment of the writing of these works no comprehensive model of human behavior existed in the relevant literature.
-As a consequence, these works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
-We attempt to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
-the resulting model is able to adapt its timing behavior in response to delay in the system.
-However, we were unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
-Furthermore, participants for the study were recruited from a pool of undergraduate students at \gls{CMU} in the \gls{USA}, ranging in age from \numrange{18}{25} years old.
-This means that findings may not be generalizable to other demographic groups.
-
-Finally, to emphasize the possible effects of network congestion and latency on \gls{WCA}, in these works we exclusively employ \acs{IEEE} 802.11n \SI{2.4}{\giga\hertz} Wi-Fi networks in an Cloudlet Computing-like setup.
-Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
-Likewise, we do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
-Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate our results and conclusions in the context of \gls{MEC} and Fog Computing.
-
-\bigskip
-
-A separate subset of the research presented in this dissertation discusses a preliminary foray into the applications of our methodology for \glspl{NCS}.
 \Cref{paper:olguinmunoz2022cleave} extends our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
 These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, we introduce a tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
+To address these challenges, we introduce yet another tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+\gls{CLEAVE} allows us both to emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
 The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
@@
 
-Given the vastness and complexity of the field of \glspl{NCS}, we limit our exploration in this dissertation to simple, small-scale, single-plant systems.
-The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
-Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that our conclusions may not directly translate to larger scale or more realistic systems.
-Additionally, our tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
-This does not match well with more complex controllers such as \gls{MPC} controllers, in which each sample produces multiple outputs from the controller.
-\gls{CLEAVE} will thus have to be modified and extended to work with such setups.
-Our study into the application of our methodology for \glspl{NCS} also relies, once again, exclusively on Wi-Fi-enabled Cloudlet Computing.
-As such, the same limitations with respect to \gls{MEC} and Fog Computing apply as with our line of work in \gls{WCA}.
-
-\bigskip
-
-This dissertation also briefly discusses Edge Computing testbed research and automation as an ancillary contribution to the methodology.
-\Cref{paper:olguinmunoz2022ainur} formally introduces the Ainur framework for Edge Computing testbed automation and workload orchestration.
+In \cref{paper:olguinmunoz2022ainur} we also briefly discuss Edge Computing testbed research and automation as an ancillary contribution to the methodology.
+This paper formally introduces the Ainur framework for Edge Computing testbed automation and workload orchestration.
 This framework is employed in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the Edge Computing testbeds on which the developed tools were deployed.
@@
 
-Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \glspl{SDR} paired with general-purpose \gls{COTS} compute hardware.
-Medium and large testbeds such as \gls{COSMOS} and \gls{POWDER} are considered \gls{OOS} due to their cost and complexity, both in terms of hardware and software.
-In terms of networking technologies, Ainur is designed to be as agnostic as possible with respect to the network stack employed.
-This dissertation includes examples using Ethernet, Wi-Fi and 4G \gls{LTE}.
-Finally, Ainur does not currently fully integrate with modern Cloud and Edge infrastructure management and deployment technologies such as OpenStack~\cite{openstack}, although it does integrate with some core Cloud-native technologies such as Docker and Ansible.
-Work is currently ongoing to extend the framework to modern mobile network technologies such as 5G and beyond-5G, as well as to general-purpose, Cloud- and Edge-native infrastructure stacks.
+We argue that our approach presents several advantages over existing methods for the study of trade-offs in latency-sensitive applications.
+Our methodology allows for efficient and repeatable evaluation of Edge environments while avoiding many of the complications associated with experimental evaluation.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
+This also reduces the barrier of entry to  research, as \glspl{SBC} are for the most part cheap and easily accessible, and further preserves the realism of effects stemming from the hardware and network.
+In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
+The softwarized nature of the client-side emulation allows for straightforward automation through general-purpose programming and scripting languages.
+Repeatability and replicability are also enhanced, as repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
+These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
+
+Furthermore, the introduced approach encompasses the entire system, including both the application and the infrastructure. 
+This holistic approach to performance evaluation of Edge environments is a notable contribution to the literature, as it provides a comprehensive view of the system's performance in realistic conditions.
+This elegant methodology, in combination with the EdgeDroid, \gls{CLEAVE}, and Ainur tools, provides a contemporary approach to studying latency-sensitive applications on Edge Computing.
+
+\medskip
+
+The second key contribution of this dissertation relates to the exploration of the question of to which degree a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
+We explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which we then combine with mathematical framework to study the potential for optimization of these applications.
+
+\cref{paper:olguinmunoz2021impact} begins this work by presenting a characterization of human behavior in \gls{WCA}, developed through a human subject study in which participants interact with an instrumented application.
+System responsiveness was altered in real-time during each execution of the task, and we recorded participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
+Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
+
+Next, in \cref{paper:olguinmunoz2023realistic} we introduce the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
+The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+The paper goes on to explore potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.
+
+We conclude that realistic, dynamic human behavior in the client-side emulation enables superior optimization approaches that would otherwise be inviable to implement.
+The findings from this research highlight that simply focusing on system-related metrics or performance optimizations may not be sufficient, as human behavior and reactions play a critical role in the overall performance of such applications.
+By considering human behavior as an integral part of the emulation process, we demonstrate that it is possible to develop more realistic and effective optimization approaches that take into account the intricacies of human-computer interaction.
+These findings have significant implications for the design and optimization of \gls{WCA} applications, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
+
+% \medskip
+
+% This line of work has some limitations.
+% First of all, we constrain our work to \emph{step-based task-assistance} \gls{WCA}, and no other categories of \gls{WCA} are considered.
+% We focus on this category for a couple of reasons.
+% Step-based task-assistance \gls{WCA} was among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represents a more mature application class compared to other \glspl{WCA}.
+% These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
+
+% However, restricting our research to step-based \gls{WCA} means that our results may not be directly applicable to other types of \gls{WCA} and \gls{AR} applications.
+% In particular, our characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
+% Additionally, the timing characteristics of step-based \gls{WCA} are such that an instruction remains valid for as long as it takes the user to perform it.
+% The conclusions from these works therefore may not apply to applications with different execution flows, or in which instructions have a deadline after which they may become stale or invalid.
+% Examples of such applications include, for instance, the \emph{Ping-Pong} assistant introduced in~\cite{chen2015early}.
+
+% The user models are further limited in their realism.
+% In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we assume a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+% The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+% These limitations emerge from a lack of data of human behavior in these applications, as at the moment of the writing of these works no comprehensive model of human behavior existed in the relevant literature.
+% As a consequence, these works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
+% We attempt to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
+% the resulting model is able to adapt its timing behavior in response to delay in the system.
+% However, we were unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
+% Furthermore, participants for the study were recruited from a pool of undergraduate students at \gls{CMU} in the \gls{USA}, ranging in age from \numrange{18}{25} years old.
+% This means that findings may not be generalizable to other demographic groups.
+
+% Finally, to emphasize the possible effects of network congestion and latency on \gls{WCA}, in these works we exclusively employ \acs{IEEE} 802.11n \SI{2.4}{\giga\hertz} Wi-Fi networks in an Cloudlet Computing-like setup.
+% Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
+% Likewise, we do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
+% Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate our results and conclusions in the context of \gls{MEC} and Fog Computing.
+
+% \bigskip
+
+% Given the vastness and complexity of the field of \glspl{NCS}, we limit our exploration in this dissertation to simple, small-scale, single-plant systems.
+% The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
+% Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that our conclusions may not directly translate to larger scale or more realistic systems.
+% Additionally, our tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
+% This does not match well with more complex controllers such as \gls{MPC} controllers, in which each sample produces multiple outputs from the controller.
+% \gls{CLEAVE} will thus have to be modified and extended to work with such setups.
+% Our study into the application of our methodology for \glspl{NCS} also relies, once again, exclusively on Wi-Fi-enabled Cloudlet Computing.
+% As such, the same limitations with respect to \gls{MEC} and Fog Computing apply as with our line of work in \gls{WCA}.
+
+% \bigskip
+
+% Our work in this domain is limited to small-scale testbeds with up to a few dozen nodes, leveraging \glspl{SDR} paired with general-purpose \gls{COTS} compute hardware.
+% Medium and large testbeds such as \gls{COSMOS} and \gls{POWDER} are considered \gls{OOS} due to their cost and complexity, both in terms of hardware and software.
+% In terms of networking technologies, Ainur is designed to be as agnostic as possible with respect to the network stack employed.
+% This dissertation includes examples using Ethernet, Wi-Fi and 4G \gls{LTE}.
+% Finally, Ainur does not currently fully integrate with modern Cloud and Edge infrastructure management and deployment technologies such as OpenStack~\cite{openstack}, although it does integrate with some core Cloud-native technologies such as Docker and Ansible.
+% Work is currently ongoing to extend the framework to modern mobile network technologies such as 5G and beyond-5G, as well as to general-purpose, Cloud- and Edge-native infrastructure stacks.

### 2023-04-17 body/part1/chapter2-relatedwork/relatedwork.tex
related work

@@
 \subsection{Performance evaluation of Edge Computing systems}
-
-\todo[inline]{
-    Needs some massaging.
-}
-\todo[inline]{
-    Right now includes only experimental approaches.
-    Need to add mathematical models (ask James for citations) and simulations.
-}
-
 \subsubsection{Analytical modeling and simulation approaches}
@@
 
-Approaches to performance evaluation of distributed systems using simulations also exist in literature, particularly geared towards the Cloud.
+Approaches to performance evaluation of distributed systems using simulations also exist in literature, particularly geared towards the Cloud~\cite{mansouri2020cloud}.
 Examples of such approaches include the \emph{CloudSim}, \emph{iCanCloud}, and \emph{GreenCloud} simulators.
@@
 GreenCloud, on the other hand, is built upon the \acs{ns2}~\cite{nsnam} network simulator and specifically captures energy consumption details of datacenter components and communication patterns between these.
+
 However, these simulators are not suitable for Edge Computing research as they do not fully capture the unique characteristics of Edge environments.
@@
 
-A number of simulation approaches which aim to overcome these limitations have therefore emerged in later years.
-iFogSim~\cite{gupta2017ifogsim} is an extension of CloudSim~\cite{calheiros2011cloudsim} geared towards the simulation of \gls{IOT} fog environments by modeling such components as sensors, actuators, Fog nodes, as well as the Cloud.
+A number of simulation approaches specifically targeted toward Edge Computing and which aim to overcome these limitations have therefore emerged in later years~\cite{svorobej2019simulating}.
+\emph{iFogSim}~\cite{gupta2017ifogsim}, for instance, is an extension of CloudSim~\cite{calheiros2011cloudsim} geared towards the simulation of \gls{IOT} Fog environments by modeling components such as sensors, actuators, Fog nodes, as well as the Cloud.
 Although iFogSim allows for location-aware servicing of user-devices by Fog nodes, this information is static during a simulation, and no mobility of users is considered.
 Network links in iFogSim also do not account for the effects on network load on network times, further limiting the simulator's realism.\@
-EdgeCloudSim~\cite{sonmez2018edgecloudsim} also extends CloudSim~\cite{calheiros2011cloudsim} and provides a much more comprehensive simulation environment for Edge deployments, which includes mobility, network, and edge compute models.
+
+\emph{EdgeCloudSim}~\cite{sonmez2018edgecloudsim} also extends CloudSim~\cite{calheiros2011cloudsim} and provides a much more comprehensive simulation environment for Edge deployments, which includes mobility, network, and edge compute models.
 However, the models chosen limit the realism of the tool, as they may not be applicable to all applications classes.
@@
 
-\todo[inline]{Wrap up.}
+Finally, another example is \emph{FogNetSim++}~\cite{qayyum2018fognetsimpp}, tool which builds upon the \acs{OMNET} network simulator.
+FogNetSim++ allows users to simulate large Fog networks while incorporating mobility and scheduling models.
+However, the tool does not include support for compute (either \gls{VM}- or container-based) migration within the network, a limitation which impacts the ability to accurately and simulate more complex scenarios in Fog and Edge Computing.
+
+It is clear then that, while there have been efforts to develop simulation approaches for evaluating the performance of Edge environments, existing approaches still fall short in capturing sufficient realism for accurate performance evaluation.
+Works such as iFogSim, EdgeCloudSim, and FogNetSim++ have attempted to address the limitations of traditional simulation approaches by incorporating components such as mobility, network, and compute models.
+However, these approaches have their own limitations, such as unrealistic network load effects or assumptions about offloading technologies.
+These limitations impact the ability to accurately simulate the dynamic and complex nature of Edge environments, including the unique challenges posed by complex applications such as \gls{XR} and \glspl{CPS}.
+These applications are interactive, real-time, present non-linear behaviors, and, in the case of \gls{XR}, involve human interaction, all factors which are adequately captured by existing simulation approaches.
 
@@
     \item \emph{IoTBench}~\cite{lee2019iotbench}
+    \item \emph{OpenRTiST}~\cite{george2020openrtist}
     \item \emph{DeFog}~\cite{mcchesney2019defog}
     \item \emph{ComB}~\cite{baurle2022comb}
-    \item \emph{OpenRTiST}~\cite{george2020openrtist}
 \end{inlineenum}.
-\emph{EdgeBench}~\cite{das2018edgebench} corresponds to a benchmarking suite for the Edge which emulates three different workloads (speech-to-text, image recognition, and sensor network application traffic), specifically targeting serverless deployments.\@
-\emph{IoTBench}~\cite{lee2019iotbench} is a benchmarking tool for Edge-deployed \gls{IOT} applications.
-It includes seven representative workloads, and allows researchers to evaluate the performance of these applications on Edge infrastructure through the analysis of computational demands and resource consumption.\@
-\emph{DeFog}~\cite{mcchesney2019defog} corresponds to a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud which includes six latency-critical and bandwidth-intensive workloads, including real-time video detection and a text-to-speech application.
-\emph{ComB}~\cite{baurle2022comb} is a flexible and extensible application-oriented benchmarking suite for Edge computing, featuring a built-in distributed video-analytics benchmark.
-Finally, \emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
+\emph{EdgeBench}~\cite{das2018edgebench} corresponds to a benchmarking suite for the Edge which emulates three different workloads (speech-to-text, image recognition, and sensor network application traffic), specifically targeting serverless deployments.
+For each workload, a corresponding batch of data is processed on an Edge node, and results are pushed to the cloud;
+cloud-only variants of the workloads also exist.
+\emph{IoTBench}~\cite{lee2019iotbench} is a similar benchmarking tool for Edge-deployed \gls{IOT} applications.
+It includes seven representative workloads, and allows researchers to evaluate the performance of these applications on Edge infrastructure through the analysis of computational demands and resource consumption.
+Both EdgeBench and IoTBench focus exclusively on benchmarking of compute in Edge and Fog deployments, and thus do not represent comprehensive evaluations of performance including network effects in Edge Computing.
+
+\emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
 Video is streamed from a source Android device to a compute node where neural style-transfer~\cite{gatys2016image} is applied to it.
 The transformed video is then fed back to the source to be presented, resulting in a compute-intensive, bandwidth-hungry, and latency-sensitive benchmark.
+However, OpenRTiST is limited by its non-replaceable workload which only targets machine learning-enhanced video processing and analytics.
 
-Although valuable contributions, these works have generally leveraged static workloads that fail to account for the intensely dynamic nature of \gls{CPS} and \gls{XR} deployments.
-Furthermore, although some of these approaches address some of the additional challenges to benchmarking \gls{XR} on the Edge, none of them have tackled the topic in a comprehensive way.\@
-\cite{das2018edgebench,mcchesney2019defog,baurle2022comb,george2020openrtist} all target image or video recognition or manipulation in some manner.
-However, none of these approaches have considered the interactive component of \gls{MAR}, particularly when it comes to human behavior and its relationship to system responsiveness and performance.
+\emph{DeFog} and \emph{ComB} represent more flexible and comprehensive tools for Edge performance evaluation.
+DeFog~\cite{mcchesney2019defog} corresponds to a benchmarking suite for the comparison of deployments on the Edge, Fog, or Cloud.
+It includes six latency-critical and bandwidth-intensive workloads built on containers, including real-time video detection and a text-to-speech application.
+ComB~\cite{baurle2022comb} is a flexible and extensible application-oriented benchmarking suite for Edge computing.
+In opposition to existing approaches, ComB relies on real representative applications instead of synthetic workloads.
+It features a built-in distributed video-analytics application that can easily be replaced.
+However, although valuable contributions, these works leverage static workloads that fail to account for the intensely dynamic nature of \gls{CPS} and \gls{XR} deployments.
 
-\paragraph{Testbed research in Edge Computing} The last few years have also seen the emergence of a number of small- to mid-scale platforms and testbeds for Edge Computing research.
+Finally, although some of the above approaches address some of the additional challenges to benchmarking \gls{XR} on the Edge, none of them have tackled the topic in a comprehensive way.\@
+\cite{das2018edgebench,mcchesney2019defog,baurle2022comb,george2020openrtist} all target image or video recognition or manipulation in some manner, but none of these approaches have considered the interactive component of \gls{MAR} particularly when it comes to human behavior and its relationship to system responsiveness and performance.
+
+\medskip
+
+The last few years have also seen the emergence of a number of small- to mid-scale platforms and testbeds for Edge Computing research.
 Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
@@
 In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.\@
-\cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
-The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds through a common interface.
-
-% Originally designed for ``the development, deployment, and validation of transformative, at-scale concepts in network science, services, and security''~\cite{berman2014geni}, \gls{GENI} today combines local Edge compute resources with mobile network last-hop connectivity for \gls{MEC} experimentation~\cite{gosain2017geni}.
-% The wireless-capable sites incorporate single-hop access to compute, storage, and network resource, making \gls{GENI} a compelling Edge Computing platform.
-
-% The Chameleon testbed is a research platform originally designed for the study and evaluation of Cloud computing systems~\cite{keahey2020lessons}, but which has recently been expanded to target Edge Computing research through the \acs{CHI}@Edge~\cite{chiatedge} project.
-% Chameleon is a collaborative project operated by a consortium of research institutions and industry partners, and is designed to provide a flexible and reconfigurable infrastructure for researchers to experiment with and test various Cloud computing technologies, algorithms, and configurations.
-% The testbed provides a highly reconfigurable array of hardware and software resources, including servers, storage systems, and networking equipment, able to simulate various Cloud and Edge Computing scenarios.
-
-% \gls{COSMOS} is a testbed deployed in New York City (State of New York, \gls{USA}), containing over \num{200} rooftop, intermediate, and mobile nodes.
-% The testbeds main \emph{raison d'être} is \gls{MEC} research, with a strong emphasis on wireless networks.
-% The system includes \glspl{SDR}, \si{\milli\meter}-wave equipment, optical fibers, Cloud integration, and Edge compute for core network functionality and application data processing~\cite{yu2019cosmos,raychaudhuri2020challenge}.
-
-% \gls{POWDER} is described by the authors of~\cite{breen2020powder} as a ``city-scale, remotely accessible, end-to-end software defined platform supporting a broad range of wireless and mobile related research''.
-% It features roughly \num{15} fixed programmable radio nodes, based on off-the-shelves \glspl{SDR}, distributed across a \SI{15}{\kilo\meter\squared} area in Salt Lake City (State of Utah, \gls{USA}).
-% These radio nodes are combined with Edge-based compute nodes as well as Cloud resources for full \gls{MEC} experimentation.
-
-% The \gls{ARA} platform~\cite{zhang2022ara} is a highly flexible testbed based on the \gls{CHI} software suite originally developed for the aforementioned Chameleon testbed~\cite{keahey2020lessons}.
-% \gls{ARA} is developed by researchers at Iowa State University in Ames, Iowa (\gls{USA}), and spans a rural area with a diameter of over \SI{60}{\kilo\meter}~\cite{zhang2022ara}.
-% Its core goal is the study and deployment of advanced Edge Computing and wireless platforms and technologies in real-world agricultural and rural settings, and includes a broad range of wireless technologies deployed through both \glspl{SDR} and programmable \gls{COTS} radios, as well as automated ground vehicles, cameras and sensors.
-
-% EdgeNet deviates from the approaches described above and represents a fully software-based, massively distributed public platform for Edge Computing research~\cite{cappos2018edgenet,senel2021edgenet1,senel2021edgenet2}.
-% It consists of a public, modified and customized Kubernetes~\cite{kubernetes} cluster.
-% The EdgeNet software is available for anyone to download;
-% individuals and organizations can quickly deploy self-hosted Edge nodes which are then made available to anyone using the platform.
-% The end result is a flexible and scalable global Edge Cloud simplifying research and prototype development in Edge and Cloud computing.
-% A similar, closely-related project is KubeEdge~\cite{xiong2018extend}, a software framework for extending Kubernetes clusters from the Cloud to the Edge.
-
-% Finally, a number of other, small-scale approaches also exist.
-% In~\cite{gedawy2016cumulus}, the authors propose Cumulus, a prototype testbed for Edge compute offloading in \gls{IOT} with the explicit goal of providing a generic and heterogeneous platform for Edge Computing research.
-% The authors of~\cite{rimal2018experimental} propose an experimental testbed for two-tiered Edge architectures in the context of \gls{FIWI} Edge Computing.
-% This testbed is then employed to validate wireless access and compute resource scheduling strategies in these systems.
-% In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-% \cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.\@
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
 % \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
@@
 
-\medskip
-
-The implementation of next-generation wireless systems, Cloud, and Edge computing paradigms requires experimentation to fully understand their implications.
-As evidenced above, testbeds have emerged for this purpose.
-However, there has been little focus on general-purpose, hardware-agnostic software frameworks for the management and automation of these testbeds.
-Most existing works use their own platform-specific solutions, which are not compatible with other testbeds and often do not comply with Cloud and/or Edge-native standards.
-Examples of this are \gls{COSMOS} and \gls{POWDER}, which use domain-specific languages and virtualization technology based on \glspl{VM} instead of modern Edge-compatible solutions such as containers or serverless computing frameworks.
-
-From the examples above, \gls{CLOUDRAFT} is the only work known to tackle the challenge of general-purpose, hardware-agnostic software frameworks for the management and automation of testbeds, albeit only to a certain extent.
-\gls{CLOUDRAFT} integrates, coordinates, shares, and improves upon existing testbeds, using pre-built VMs with necessary software for experiments.
-Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely on the sharing and partitioning of testbed systems.
-Testbeds that are currently working with \gls{CLOUDRAFT} include an \gls{SDR}-based testbed and a ground vehicular robot for mobility-related experimentation.
-
-The lack of a unified approach to testbed management and automation hinders the development and implementation of next-generation wireless systems, Cloud, and Edge computing paradigms.
-General-purpose, hardware-agnostic software frameworks such as \gls{CLOUDRAFT} can simplify the management and automation of testbeds, promote compatibility with Cloud- and Edge-native standards, and facilitate collaboration among research groups.
+Although the aforementioned testbeds represent valuable and powerful tools, the lack of a unified approach to testbed management and automation hinders the development and implementation of next-generation wireless systems, Cloud, and Edge computing paradigms.
+General-purpose, hardware-agnostic software frameworks such as \gls{CLOUDRAFT}~\cite{moorthy2022cloudraft}  can simplify the management and automation of testbeds, promote compatibility with Cloud- and Edge-native standards, and facilitate collaboration among research groups.
+\gls{CLOUDRAFT} is a Cloud-based framework for mobile network experimentation with the goal integrate, coordinate, share, and improve upon existing testbeds through a common interface.
 There is still remains much work to be done in this area, particularly with respect to the adoption of Edge-compatible solutions that can enable faster and more reliable experimentation.

### 2023-04-17 body/part1/chapter2-relatedwork/scope.tex
related work

--- 2023-04-14 body/part1/chapter2-relatedwork/scope.tex
@@
-\section{Scope of the thesis}
+\section{Scope of this dissertation}
 \glsresetall%
 
-\todo[inline]{Write}
-
-\todo[inline]{
-    Emphasize that apps are standing for a new class of application class that the community is interested in developing.
-    Why?
-    Utility.
-    Weave into intro and scope.
-
-    Not just a random mashup.
-    Relevance of the combination!
-
-    Prime reason for experimental evaluation.
-    Capture effects of compute, network, and design of the application.
-    Hard to capture otherwise.
-
-    Build argument in state of the art that there is no current satisfying way to address this.
-    Potentially orders of magnitude off.
-
-    In scope: interested in meaningful performance evaluation (i.e.\ capture all the complexity!) Make sure we note this.
-    Show how apps and system are intimately related to each other and complex to evaluate.
-    Where do complexities arise.
-    Understanding system performance is key and complex.
-
-    Scope: define the challenges.
-}
-
-\todo[inline]{
-    Write about in terms of relevance and challenges.
-    Motivate a bit. And then tell what we're not interested in.
-
-    First 3 paragraphs in a conventional papers are usually scope. 
-}
-
-\begin{enumerate}
-    \item Performance evaluation.
-    \item \gls{CPS} and \gls{MAR}
-    \item Edge Computing 
-\end{enumerate}
-
-\glsresetall%
 \glspl{CPS} and \gls{XR} applications are emerging technologies with the potential to transform a plethora of industries.
@@
 
-Nevertheless, evaluating the performance of Edge environments involving such applications is challenging, due to complex behaviors emerging from the interaction between computation and network in these systems.
-Existing literature has attempted to tackle the performance evaluation of Edge systems through both analytical and simulation approaches.
-However, the nature of Edge Computing environments, which involves tight integration between network and compute, vast and often unfeasible interdisciplinary competence.
+Evaluating the performance of Edge environments involving such applications is challenging, due to complex behaviors emerging from the interaction between computation and network in these systems.
+Existing literature has attempted to tackle the performance evaluation of Edge systems through analytical, simulation, and experimental approaches.
+However, the nature of Edge Computing environments, which involves tight integration between network and compute, requires vast and often unfeasible interdisciplinary competence.
 Existing works have been constrained to narrow their scope by prioritizing the comprehensive modeling of either compute or network, resulting in a potential trade-off that may neglect the other aspect.
-This has resulted in an incomplete understanding of the overall performance of Edge systems and applications in literature.
+As we will see in \cref{sec:state-of-the-art}, this has led to an incomplete understanding of the overall performance of Edge systems and applications in literature.
 
@@
 Or work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
+We also focus exclusively on Cloudlet Computing-like setups.
+Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
+Likewise, we do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
+Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate our results and conclusions in the context of \gls{MEC} and Fog Computing.
 
-However, the performance evaluation of Edge-enabled \gls{XR} encompasses a wide range of factors, making the scope of research in this area extensive and complex.
+The performance evaluation of Edge-enabled \gls{XR} encompasses a wide range of factors, making the scope of research in this area extensive and complex.
 The scope of performance evaluation research for \glspl{CPS} is at least equally extensive.
@@
 Therefore, in this work, we approach these challenges by focusing on representative use-cases for these application classes.
-These correspond to \gls{WCA} applications for \gls{XR}, and \glspl{NCS} for \glspl{CPS}.
+
+Specifically, we employ \emph{step-based task-assistance} \gls{WCA} as a representative use case for \gls{XR}, and \glspl{NCS} for \glspl{CPS}.
+We focus on these categories for a couple of reasons.
+In the case of \gls{WCA}, step-based task-assistance was among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represents a more mature application class compared to other \glspl{WCA}.
+These applications also feature a linear mode of execution, simplifying emulation, and their step-based nature makes the tasks easier to shorten or extend for experimental purposes.
+
+For \glspl{NCS}, we limit our exploration in this dissertation to simple, small-scale, single-plant systems.
+The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
+We have chosen this system for its relative simplicity as well as its broad use in the field of automatic control as one of the fundamental examples of linear control.
 Through these examples, we aim to gain insights into the performance evaluation challenges and opportunities offered by Edge Computing these applications and systems, and contribute to the understanding of the broader fields of \gls{XR} and \gls{CPS} performance evaluation in distributed computing environments.
+
+\bigskip
+
+In the following  \cref{sec:state-of-the-art}, we will review the state of the art in the context of performance evaluation of Edge Computing environments, including analytical, simulation, and experimental approaches.
+We will evidence how existing approaches fail to fully capture the intricacies of performance evaluation of latency-sensitive, feedback-loop applications on Edge Computing, and make the case for the need for an alternative approach.
+We will further discuss existing experimental approaches to \gls{MAR} and \gls{NCS} research, and discuss their limitations.
+
+Finally, we conclude this chapter in \cref{sec:contributions}, where we will discuss in details the contributions of this dissertation.
+This includes a discussion of the key sub-contributions of each included publication.

### 2023-04-17 body/part1/chapter1-introduction/background.tex
writing
fixes

@@
 \subsubsection{\glsfmtlong{XR}}\label{background:xr}
-\todo[inline]{Needs citations.}
 \glsreset{XR}
@@
 
-\gls{XR} is an umbrella term used to refer to all immersive technologies that combine real and virtual environments, including technologies such as \gls{AR}, \gls{VR}, and \gls{MR}.
+\gls{XR} is an umbrella term used to refer to all immersive technologies that combine real and virtual environments, including technologies such as \gls{AR}, \gls{VR}, and \gls{MR}~\cite{doerner2022virtual}.
 \emph{Immersiveness} refers to the extent to which a technology can fully engage a user's senses to create the illusion of being in a different environment or situation.
@@
 
-\gls{AR} merges digital content with the physical world, enhancing human perception and cognition through the overlay of virtual objects and information on top of the real environment.
+\gls{AR} merges digital content with the physical world, enhancing human perception and cognition through the overlay of virtual objects and information on top of the real environment~\cite{doerner2022virtual}.
 This technology can superimpose tooltips, virtual \gls{2D} and \gls{3D} objects, or even full-fledged videos on top of what the user sees.
@@
 
-\gls{VR}, on the other hand, takes it a step further and involves forgoing the real world, completely immersing the user in a fully virtual environment.
+\gls{VR}, on the other hand, takes it a step further and involves forgoing the real world, completely immersing the user in a fully virtual environment~\cite{doerner2022virtual}.
 Due to its deeply immersive nature, \gls{VR} requires users to wear specialized equipment such as a \gls{VR} headset.
@@
 These objects are rendered in real-time and projected onto the real-world view in the user's headset in a seamless and realistic way.
-In this way, \gls{MR} creates a hybrid environment where users can interact with both virtual and real-world elements simultaneously, resulting in an illusion of natural interaction between users and digital elements.
+In this way, \gls{MR} creates a hybrid environment where users can interact with both virtual and real-world elements simultaneously, resulting in an illusion of natural interaction between users and digital elements~\cite{speicher2019mixed}.
 

### 2023-04-22 body/part1/chapter2-relatedwork/key_contributions.tex
contrib tree

@@
 
-This thesis presents two key contributions to literature.
-First, we introduce and study a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
-Our approach is based on the novel idea of emulating \emph{only} the end-user component of such applications while using the \emph{real} network and compute components of the system.
+\begin{figure}
+    \centering
+    \begin{forest}
+        for tree={% style of tree
+            align=center,
+            grow=0,reversed, % tree direction
+            parent anchor=east,
+            child anchor=west, % edge anchors
+            edge={line cap=round},
+            outer sep=+1pt, % edge/node connection
+            rounded corners,minimum width=15mm,minimum height=8mm, % node shape
+            l sep=7mm % level distance
+        }
+        [{\scshape Contributions}
+            [Methodology
+                [{Use-case\\studies} [\acs{WCA}] [\acsp{NCS}]]
+                [{Tooling} [{EdgeDroid\\(\acs{WCA})}] [{\acs{CLEAVE}\\(\acsp{NCS})}] [Ainur\\(testbed)]]
+            ]
+            [{Implications\\of Realism}
+                [{Characterization of\\human behavior in \acs{WCA}}]
+                [{Realistic human\\model for \acs{WCA}}]
+                [Exploration of avenues\\for optimization of \acs{WCA}]
+            ]
+        ]
+        \end{forest}
+    \caption{Hierarchical overview of the main and secondary contributions of this dissertation.}\label{fig:contribs}
+\end{figure}
+
+This thesis presents two key contributions and a number of sub-contributions to literature, as illustrated in \cref{fig:contribs}.
+Firstly, we introduce a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
+This methodology is validated by example through two representative use-cases, and we provide tooling for its implementation.
+Secondly, we study the implications improved client-side emulation realism has on the results produced with our methodology.
+These implications are again explored by example.
+We design and implement of a detailed model of human user behavior in \gls{WCA}, and then use it to study the potential for optimization of these systems.
+
+\medskip
+
+Our first contribution corresponds to a methodological approach to the study and performance evaluation of Edge-enabled latency-sensitive applications. 
+This approach is based on the novel idea of emulating \emph{only} the end-user component of such applications while using the \emph{real} network and compute components of the system.
 In other words, our methodology involves replacing the client side of the system with commercial off-the-shelf (COTS) general-purpose computing devices running realistic software imitations of the desired behaviors, while retaining the real network and compute hardware at the backend.

### 2023-04-22 body/part1/chapter3-contributions/methodology.tex
contrib tree

@@
-As summarized in \cref{chap:introduction}, this thesis presents a comprehensive study on latency-sensitive applications in Edge Computing.
-The primary contribution of this research is a methodology for analyzing and evaluating the performance of these applications.
-The methodology is validated through the examination of two initial case studies, \gls{WCA} and \glspl{NCS}.
-In addition to the methodology, this thesis includes several other key contributions.
-A prototype testbed and accompanying software framework have been developed to provide a platform for conducting experiments in Edge Computing.
-Furthermore, a further deeper exploration into the \gls{WCA} case studied yielded the first-ever model of human behavior in \gls{WCA}, which is a significant advancement in the field.
-The methodology, the case studies, and the prototype testbed provide valuable insights into the performance of latency-sensitive applications in Edge Computing and can be used as a basis for future research in this area.
+As previously discussed, this dissertation makes two key contributions to the literature surrounding performance evaluation of Edge Computing environments.
+These contributions, together with ancillary contributions and results, will be summarized in the two sections presented in this chapter.
+
+In \cref{summary:methodology}, we discuss our methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
+This methodology involves emulating only the end-user component of such applications while using the real network and compute components of the system. 
+We further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.
+
+Next, in \cref{summary:realism}, we explore the extent to which our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
+We broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which we then employ in our methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.
 
@@
 
-This thesis makes a contribution to the field of Edge Computing by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge infrastructure.
+This thesis makes a contribution to the field of performance evaluation of Edge Computing systems by investigating the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
 These applications, such as \gls{WCA} and \glspl{NCS}, are challenging to study and scale due to their cyber-physical nature, complexity, and requirements for low latency and high reliability.
 
-The main focus of this work is to enhance the accuracy and realism of results related to Edge infrastructure, particularly in regard to the network.
-The proposed methodology aims to provide a more comprehensive and realistic assessment of the performance of these latency-sensitive applications, allowing for a better understanding of the strengths and limitations of Edge Computing in this context.
+The main focus of this work is to enhance the accuracy and realism of results related to Edge infrastructure.
+In particular, the proposed methodology aims to provide a more comprehensive and realistic assessment of the performance of these latency-sensitive applications with respect to previous approaches, allowing for a better understanding of the strengths and limitations of Edge Computing.
 Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing systems.
@@
 
-% As discussed in \cref{chap:relatedwork}, most research in this field falls into either simulated or fully-practical real-world approaches.
 Our emulated methodology presents several advantages over alternatives, simulated or fully-practical approaches.
@@
 \medskip
-In the following subsections we will discuss two illustrative case studies for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
+In the following subsections we will discuss two illustrative use-cases for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
 \Cref{summary:methodology:usecase_wca} discusses the application of our methodology to \gls{WCA}.
@@
 
-\subsection{Case study: \acl{WCA}}\label{summary:methodology:usecase_wca}
+\subsection{Use-case: \acl{WCA}}\label{summary:methodology:usecase_wca}
 
@@
 
-\subsection{Case study: \acsp{NCS}}\label{summary:methodology:usecase_ncs}
+\subsection{Use-case: \acsp{NCS}}\label{summary:methodology:usecase_ncs}
 
-We discuss an initial case study of our methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
+We discuss an initial use-case study of our methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
 In this work, we present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of Edge-native \aclp{NCS}.
@@
 
-After our initial implementation and experimentation with our methodology for \gls{WCA} applications, we realized that creating a testbed for experimental Edge Computing research for would significantly enhance our work.
+After our initial implementation and experimentation with our methodology for \gls{WCA} applications, we arrived at the realization that the creation of a dedicated testbed for experimental Edge Computing research would greatly augment the scope and impact of our work.
 Establishing such testbed would provide a dedicated and controlled environment that can accelerate future research by eliminating the need for ad-hoc implementations for each new study.

### 2023-04-22 body/part1/chapter3-contributions/wca-implications.tex
contrib tree

@@
-\section{Implications of the methodology in the context of \acs{WCA}}\label{sec:extendingmethodology}
+\section{Exploring the need for realism in client-side emulation}\label{summary:realism}
 
 In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
-To this end, we must first address a limitation of our current model of human behavior in the context of these applications.
+In particular, we wish to explore the impact improving the realism of the client-side emulation has on the potential for optimization of the applications under evaluation.
+We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.
+
 Our work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
 However, this is not an accurate representation of real human behavior, and a more realistic human model must take into consideration the perceived feedback latency and its effects on human reactions.
+In order to develop a more accurate human model in the context of \gls{WCA}, we first conduct an experimental study that measures human performance in different scenarios, taking into account the impact of system latency on the behavior of humans.
+The results of this study serve as the foundation for building a more representative model of human behavior.
 
-To build a more accurate model, we must first characterize human behavior in the context of \gls{WCA}.
-We achieve this through an experimental study measuring human performance in various scenarios, taking into account the impact of latency in the system to human behavior.
-The results of this study are then used to build a more representative model of human behavior, which we further use to provide a foundation for the optimization of these applications.
+Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
+By improving the realism of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
+This research highlights the importance of the realism of emulations as  crucial factor in the application of our methodology.
 
-Once we have established more accurate model of human behavior, we can then begin to explore the implications of our methodology for the optimization of \gls{WCA}.
-This includes evaluating the impact of various optimization strategies on human performance and identifying opportunities for improvement.
-By incorporating a realistic model of human behavior, we will be able to build a more comprehensive understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be applied to address these challenges.
-
-\subsection{Characterization of \glsfmtshort{WCA}}
-\todo[inline]{Rework}
+\subsection{Definitions relating to the operation of \glsfmtshort{WCA} applications}
 
@@
 
-\subsubsection{Implications}
+\subsubsection{Implications for \glsfmtshort{WCA}}
 
@@
 Built upon the data and results from \cref{paper:olguinmunoz2021impact}, our model is capable of dynamically generating realistic predictions of execution times for steps in a \gls{WCA} based on historical system state data and a parameterized level of neuroticism.
-Furthermore, we implement a new version of the EdgeDroid tool capable of using this new model, together with a novel procedure for the generation of realistic synthetic traces of frames for \gls{WCA}, to fully emulate a human in the context of these systems.
+Furthermore, we implement a new version of the EdgeDroid tool capable of using this new model, together with a novel procedure for the generation of realistic synthetic traces of frames for \gls{WCA}, to provide a fully end-to-end emulation of a human in the context of these systems.
 

### 2023-04-23 body/part1/chapter2-relatedwork/scope.tex
scope

@@
 These are key factors for real-time, context-sensitive, low-latency applications such as \gls{CPS} and \gls{XR}.
+
+\input{tables/scope}
 
@@
 We aim to complement literature by providing a method which bridges the gap in knowledge left by existing approaches.
-Or work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
-We also focus exclusively on Cloudlet Computing-like setups.
+Our work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
+We also focus exclusively on Cloudlet Computing-like setups (see \cref{tab:scope}).
 Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.

### 2023-04-23 body/part1/chapter2-relatedwork/tables/scope.tex
scope

@@
 % \usepackage{booktabs}
+% \usepackage[table,xcdraw]{xcolor}
+% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
 \begin{table}[]
     \centering
-    \caption{Overview of the scope and main contribution of each included publication.}\label{tab:scope}
-    \tiny
-    \renewcommand{\arraystretch}{1.5}
-    \begin{tabularx}{\textwidth}{@{}ccXYX@{}}
+    \caption{Visualization of the scope of this dissertation with respect to performance evaluation.
+    We focus exclusively on Edge Computing based on Cloudlets~\cite{satyanarayanan2009case} and on experimental performance evaluation of \gls{XR} and \glspl{CPS}.}
+    \label{tab:scope}
+    \begin{tabularx}{\textwidth}{@{}YYYY@{}}
         \toprule
-        \textbf{Paper} & \textbf{Application} & \textbf{Emulation Approach} & \textbf{Network} & \textbf{Main Contribution}                                \\ \midrule
-        \ref{paper:olguinmunoz2018demoscaling} & \acs{WCA} & Trace, \acs{FSM} user model & Wi-Fi & Initial demo of methodology. \\
-        %
-        \ref{paper:olguinmunoz2019edgedroid} & \acs{WCA} & Trace, \acs{FSM} user model & Wi-Fi & EdgeDroid tool for application of the methodology to \acs{WCA}. \\
-        %
-        \ref{paper:olguinmunoz2021impact} & \acs{WCA} & \acs{N/A} & No network & Characterization of human behavior in \acs{WCA}. \\
-        %
-        \ref{paper:olguinmunoz2022cleave} & \acs{NCS} & Software implementation of Control System & Wi-Fi & \acs{CLEAVE} tool for application of the methodology to \acs{NCS}. \\
-        %
-        \ref{paper:olguinmunoz2022ainur} & \acs{WCA}, \acs{NCS} & \acs{N/A} & Ethernet, Wi-Fi, 4G~\acs{LTE} & Ainur framework for Edge Computing testbed automation. \\
-        \ref{paper:olguinmunoz2023realistic} &
-        \acs{WCA} &
-        Data-driven stochastic synthetic trace and user model &
-        Wi-Fi &
-        Model for end-to-end emulation of humans in \acs{WCA}, exploration of implications of methodology and model for \acs{WCA} optimization. \\
-        \bottomrule
+        & {Analytical Modeling} & {Simulations} & {Experimental}                   \\ \midrule
+        Cloudlets & \cellcolor[HTML]{9AFF99}Yes & \cellcolor[HTML]{9AFF99}Yes & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\
+        \gls{MEC} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{FFCCC9}No                         \\
+        Fog & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{FFCCC9}No                         \\
+        \gls{XR}  & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\
+        \glspl{CPS} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\ \bottomrule
     \end{tabularx}

### 2023-04-23 glossary.tex
working on intro

@@
 \newacronym[longplural={Mixed Realities}]{MR}{MR}{Mixed Reality}
-\newacronym{MAR}{MAR}{Mobile \acs{AR}}
+\newacronym{MAR}{MAR}{Mobile \nestedgls{AR}}
 \newacronym{GPU}{GPU}{Graphical Processing Unit}
@@
 \newacronym{WPAN}{WPAN}{Wireless Personal Area Network}
-\newacronym{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
+\newacronym{CSMA}{CSMA}{Carrier-Sense Multiple Access}
+\newacronym{CSMA/CD}{CSMA/CD}{\nestedgls{CSMA} with Collision Detection}
 \newacronym{OS}{OS}{Operating System}
@@
 %\acroindefinite{SDR}{an}{a}
-\newacronym{VLAN}{VLAN}{Virtual Local-Area Network}
+% \newacronym{LAN}{LAN}{Local Area Network}
+\newacronym{VLAN}{VLAN}{Virtual \nestedgls{LAN}}
 \newacronym{APT}{APT}{Advaced Packaging Tool}
@@
 \newacronym{RAN}{RAN}{Radio Access Network}
-\newacronym{O-RAN}{O-RAN}{Open \acs{RAN}}
+\newacronym{O-RAN}{O-RAN}{Open \nestedgls{RAN}}
 \newacronym{FOSS}{FOSS}{Free and Open Source Software}
@@
 \newacronym{MaaS}{MaaS}{Metal-as-a-Service}
-\newacronym{OEDL}{OEDL}{\acs{OMF} Experiment Description Language}
+\newacronym{OEDL}{OEDL}{\nestedgls{OMF} Experiment Description Language}
 \newacronym{RSpec}{RSpec}{Resource Specification}
@@
 \newacronym{CDF}{CDF}{Cumulative Distribution Function}
-\newacronym{CCDF}{CCDF}{Complementary \acs{CDF}}
+\newacronym{CCDF}{CCDF}{Complementary \nestedgls{CDF}}
 \newacronym{PDF}{PDF}{Probability Density Function}
-\newacronym{ECDF}{ECDF}{Empirical \acs{CDF}}
+\newacronym{ECDF}{ECDF}{Empirical \nestedgls{CDF}}
 \newacronym{NSF}{NSF}{United States National Science Foundation}
@@
 \newacronym{OOS}{OoS}{Out-of-Scope}
+\newacronym{OMNET}{OMNeT++}{Objective Modular Network Testbed in C++}
+\newacronym{ns2}{ns-2}{network simulator 2}
+\newacronym{ns3}{ns-3}{network simulator 3}
+\newacronym{IRB}{IRB}{Institutional Review Board}
+\newacronym{IaaS}{IaaS}{Infrastructure-as-a-Service}
 %\glsaddall%

### 2023-04-23 body/part1/chapter1-introduction/background.tex
working on intro

@@
 
-\begin{figure}
-    \centering
-    \includegraphics[height=30em]{figures/edgecomputing}
-    \caption{%
-        Conceptual design of Edge Computing.
-        Compute nodes are placed a at the edge of the network, a few hops away from end users.
-        In other words, compute is located between users and the internet and the Cloud.
-    }\label{fig:edgecomputing}
-\end{figure}
-
 Edge Computing is a novel distributed computing paradigm, emerging from a need to overcome the drawbacks of offloading computation and data to the Cloud.
@@
 To tackle the prohibitively high latencies of Cloud computing, the authors proposed an extension of the existing Cloud computing paradigm with compute nodes at the edge of the Internet.
-Instead of offloading computation to a datacenter potentially thousands of kilometers away, in Edge Computing it is offloaded to a compute node at the \emph{edge} of the network close to the user, see \cref{fig:edgecomputing}.
+Instead of offloading computation to a datacenter potentially thousands of kilometers away, in Edge Computing it is offloaded to a compute node at the \emph{edge} of the network close to the user.
 

### 2023-04-23 body/part1/chapter1-introduction/introduction.tex
notes
working on intro

@@
 
+\todo[inline]{
+    Needs to be reworked based on scope, new formulation of two contributions.
+    Enumerate contributions at the end.
+}
+
 The advancement of technology has brought about new computing paradigms and network technologies that have the potential to revolutionize the way we think about and approach computing and communication.
-One of the most promising of these is Edge Computing, a distributed computing paradigm which seeks to bring computing closer to the end user, reducing latency (i.e.\ the time between input and response) and increasing reliability.
+One of the most promising of these is Edge Computing, a distributed computing paradigm which seeks to bring compute nodes closer to the end user, reducing latency (i.e.\ the time between input and response) and increasing reliability.
 
 The current model for distributed computing, known as Cloud Computing, allows users to access shared pools of resources, services, and applications over the internet~\cite{gai2012towards}.
-Cloud computing offers many advantages, such as virtually unlimited storage and processing power, by virtue of employing economies of scale in large, geographically distributed datacenters.
+Cloud computing offers many advantages, including as virtually unlimited storage and processing power
+This achieved by virtue of employing economies of scale in massive installations containing up to thousands of compute servers and networking equipment.
+These installations are known as \emph{datacenters}, and often serve massive amounts of users.
+
 However, this design leads to trade-offs in bandwidth and, particularly, latency.
-The responsiveness of the Cloud is limited by the speed of light, and thus it can suffer from high latency due to the distance data must travel between the end user and the cloud.
-This can cause delays in applications that require quick response times, such as real-time video processing or online gaming.
+Cloud datacenter are commonly located in geographically centralized locations to better serve thousands or even millions of users.
+Conversely, this results in these datacenters having a large average \emph{topological} and \emph{physical} distance to individual users.
+Topologically, end-user connections to the Cloud tend to transit a significant number of hops, i.e.\ intermediate connections through networking equipment belonging to service providers at multiple tiers of the internet.
+Data needs to be received, routed, and re-transmitted at each hop, adding delay.
+Physically, Cloud datacenters are often located hundreds or even thousands of kilometers from end-users.
+Connections are limited in their responsiveness by the speed of light, and these long physical distances directly translate into higher latencies.
+
+These limitations of the Cloud make it unsuitable for a number of novel applications that require quick response times, such as real-time video processing, \gls{MAR}, or smart cities.
 Additionally, the centralized nature of Cloud datacenters means that it can be vulnerable to outages or other disruptions, which can impact large numbers of users simultaneously.
 
+\begin{figure}
+    \centering
+    \includegraphics[height=30em]{figures/edgecomputing}
+    \caption{%
+        Conceptual design of Edge Computing.
+        Compute nodes are placed a at the edge of the network, a few hops away from end users.
+        In other words, compute is located between users and the internet and the Cloud.
+    }\label{fig:edgecomputing}
+\end{figure}
+
 Edge Computing aims to improve upon the shortcomings of Cloud Computing by moving the processing closer to the end user, at the \emph{edge} of the network.
-In this context, the ``edge of the network'' refers to locations which are both geographically and topologically (in the networking sense) close to the user, such as cellular service provider towers/base stations or micro-datacenters located within the same urban area as the user.
-By placing computing resources in proximity to where data is generated and/or consumed, and thus reducing the distance that data must travel, Edge Computing can greatly decrease latency and improve the responsiveness of applications.
-It also allows for more efficient use of network resources, as data does not need to travel back and forth to a remote, central datacenter.
+In this context, the ``edge of the network'' refers to locations which are both geographically and topologically close to the user, such as micro-datacenters located within the same urban area as the user (see \cref{fig:edgecomputing}).
+\gls{MEC}, a variant of the paradigm, even proposes the integration of compute power within 5G~\cite{5Gstandard} cellular networks, co-locating computed nodes with cellular service provider base stations.
 
-The advent of 5G~\cite{5Gstandard}, the latest generation of mobile communication networks, offers even greater potential for Edge Computing.
-5G offers higher data rates, lower latency, and improved reliability compared to previous generations of mobile networks.
-This enables even more powerful and sophisticated applications to be run on the Edge, such as \acsp{CPS} and \acsp{NCS}, as well as \acs{MAR} applications.
+By placing computing resources in proximity to where data is generated and/or consumed, and thus reducing the topological and physical distances that data must travel, Edge Computing can greatly decrease latency, improve the responsiveness of applications, and allow for greatly improved efficiency in the use of network resources.
+This enables powerful and sophisticated applications to be run on the Edge, such as \glspl{CPS} as well as \gls{XR} applications.
+These applications have the potential to have a profound impact in our day-to-day lives, but have stringent latency requirements which make their deployment on traditional paradigms such as Cloud Computing inviable.
 
-\acfp{CPS} and \acfp{NCS} refer to systems where physical processes, such as manufacturing or transportation, are controlled  and monitored by computers, usually over communication networks.
-Edge Computing, combined with 5G networking, allows these systems to operate in real-time, enabling near instantaneous response times, more efficient operation, and more precise control.
+\glspl{CPS} refers to systems where physical processes, such as manufacturing or transportation, are controlled  and monitored by computers, usually over communication networks.
+Edge Computing allows these systems to operate in real-time, enabling near instantaneous response times, more efficient operation, and more precise control.
 These characteristics are critical for applications such as \glspl{ITS}, autonomous driving, or remote surgery.
 
-\gls{MAR}, on the other hand, is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
+\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive application systems.
+\gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
+For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
 This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
 \gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
-By using Edge Computing and 5G, \gls{MAR} can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.
+By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.
 
@@
 Nonetheless, significant challenges remain in the understanding and scaling of these systems, particularly in regard to their complexity and requirements in the context of low-latency, high-reliability computing.
+Before widespread deployment of these systems, it is crucial to thoroughly understand their real-world performance.
 
-Latency-sensitive applications, such as \glspl{CPS} and \gls{MAR}, involve multiple processing steps that can influence the latency of the system.
+\glspl{CPS} often correspond to safety-critical components of industrial systems, and subpar \gls{QoS} can therefore lead to significant danger to equipment and operators.
+On the other hand, the immersive nature of \gls{XR} applications implies that poor \gls{QoS} can easily translate into significant user discomfort.
+To ensure reliability, safety, and user satisfaction in these systems, careful and accurate characterization and optimization of these systems must be realized.
+
+However, evaluating the performance of such applications on Edge environments represents a significant challenge.
+Latency-sensitive applications such as \glspl{CPS} and \gls{XR}, involve multiple processing steps that can influence the latency of the system.
 There are many trade-offs involved in optimizing the latency of these applications beyond the question of where the compute process is located. 
 These include the choice of wireless system for transmitting application data (e.g.\ 4G \gls{LTE} versus 5G), the protocols used to communicate over the network, and the hardware and operating system used in the backend.
-In the case of \gls{MAR}, the sensory inputs need to be pre-processed and compressed on-device, before being transmitted to the compute backend for processing.
+In the case of \gls{XR}, the sensory inputs need to be pre-processed and compressed on-device, before being transmitted to the compute backend for processing.
 The backend algorithms need to be carefully designed and implemented to ensure efficient processing and minimal latency. 
 Similarly, in \glspl{CPS}, the communication network must be designed to minimize latency and ensure reliable data transmission between the components of the control system.
-Delayed or lost data can result in catastrophic failure of these systems.
 Other critical aspects with the potential to impact the latency of latency-sensitive applications on the Edge include congestion on the communication loop and fluctuations of the wireless channel.
 These issues can lead to delays in data transmission between devices on the network, leading to higher end-to-end latency.
+
+The evaluation of performance in Edge computing systems has been addressed in existing literature through various approaches such as analytical, simulation, and experimental methods.
+However, the unique nature of Edge computing environments demands a wide range of interdisciplinary expertise spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
+Such expertise may not always be feasible to obtain within a single research group.
+This has led to prior research to often be constrained in its scope, focusing on comprehensive characterization of either compute or network aspects, resulting in potential trade-offs that may neglect one half of the larger picture.
+We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
+There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.
 

### 2023-04-23 body/part1/chapter2-relatedwork/tables/scope.tex
working on intro

@@
     \caption{Visualization of the scope of this dissertation with respect to performance evaluation.
-    We focus exclusively on Edge Computing based on Cloudlets~\cite{satyanarayanan2009case} and on experimental performance evaluation of \gls{XR} and \glspl{CPS}.}
+    We focus exclusively on Edge Computing based on Cloudlets~\cite{satyanarayanan2009case} and on experimental performance evaluation of \acs{XR} and \acsp{CPS}.}
     \label{tab:scope}
-    \begin{tabularx}{\textwidth}{@{}YYYY@{}}
+    \begin{tabular}{@{}cccc@{}}
         \toprule
-        & {Analytical Modeling} & {Simulations} & {Experimental}                   \\ \midrule
+        & \makecell{Analytical\\Modeling} & {Simulations} & {Experimental}                   \\ \midrule
         Cloudlets & \cellcolor[HTML]{9AFF99}Yes & \cellcolor[HTML]{9AFF99}Yes & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\
-        \gls{MEC} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{FFCCC9}No                         \\
+        \acs{MEC} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{FFCCC9}No                         \\
         Fog & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{FFCCC9}No                         \\
-        \gls{XR}  & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\
-        \glspl{CPS} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\ \bottomrule
-    \end{tabularx}
+        \acs{XR}  & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\
+        \acsp{CPS} & \cellcolor[HTML]{FFCCC9}No              & \cellcolor[HTML]{FFCCC9}No      & \cellcolor[HTML]{9AFF99}{\color[HTML]{333333} Yes} \\ \bottomrule
+    \end{tabular}
 \end{table}

### 2023-04-24 body/part1/chapter1-introduction/introduction.tex
introduction

@@
 \section{Introduction}
-
-\todo[inline]{
-    Needs to be reworked based on scope, new formulation of two contributions.
-    Enumerate contributions at the end.
-}
 
@@
 
-These limitations of the Cloud make it unsuitable for a number of novel applications that require quick response times, such as real-time video processing, \gls{MAR}, or smart cities.
+These limitations of the Cloud make it unsuitable for a number of novel applications that require quick response times, such as real-time video processing, \gls{MAR}, or smart city systems.
 Additionally, the centralized nature of Cloud datacenters means that it can be vulnerable to outages or other disruptions, which can impact large numbers of users simultaneously.
@@
 This enables powerful and sophisticated applications to be run on the Edge, such as \glspl{CPS} as well as \gls{XR} applications.
-These applications have the potential to have a profound impact in our day-to-day lives, but have stringent latency requirements which make their deployment on traditional paradigms such as Cloud Computing inviable.
+These applications have the potential to cause a profound impact in our day-to-day lives, but have stringent latency requirements which make their deployment on traditional paradigms such as Cloud Computing inviable.
 
@@
 Nonetheless, significant challenges remain in the understanding and scaling of these systems, particularly in regard to their complexity and requirements in the context of low-latency, high-reliability computing.
+
 Before widespread deployment of these systems, it is crucial to thoroughly understand their real-world performance.
-
 \glspl{CPS} often correspond to safety-critical components of industrial systems, and subpar \gls{QoS} can therefore lead to significant danger to equipment and operators.
@@
 
-The evaluation of performance in Edge computing systems has been addressed in existing literature through various approaches such as analytical, simulation, and experimental methods.
-However, the unique nature of Edge computing environments demands a wide range of interdisciplinary expertise spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
+The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical, simulation, and experimental approaches.
+However, these systems exhibit several characteristic that make their study complex.
+One of these is unique nature of Edge computing environments, which tightly integrates compute and network.
+This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
 Such expertise may not always be feasible to obtain within a single research group.
-This has led to prior research to often be constrained in its scope, focusing on comprehensive characterization of either compute or network aspects, resulting in potential trade-offs that may neglect one half of the larger picture.
+
+The highly complex effects emerging from the interaction between network and compute is another major hurdle to studying latency-sensitive applications on Edge Computing.
+These effects are specially difficult to address through analytical models and simulations, particularly in multi-tenant deployments, where each tenant may have different requirements and resource usage patterns.
+Although powerful approaches to the modeling of network effects and compute process scheduling exist in literature, existing analytical and simulation approaches have so far not managed to combine the two.
+
+Nevertheless, experimental approaches also face important challenges.
+One of these stems from the fact that latency-sensitive applications such as \gls{CPS} and \gls{XR} on Edge Computing interact with the real world.
+This implies that these applications thus must be studied in real-time to achieve comprehensive characterization.
+Thus, any study of these applications must be done in an environment that accurately reflects the real-world conditions in which they are deployed.
+Unfortunately, replicating these conditions in a controlled experimental environment is difficult and often leads to inconsistencies in results, and traditional methods that rely on simulations may not be sufficient to accurately capture and evaluate the performance of these applications in real-world scenarios.
+
+The scalability of these systems is also a challenge.
+It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, personnel and test subjects, as well as facilities.
+\glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{XR} is time-consuming and expensive.
+This also makes the study of these applications hard to automate, making it challenging to conduct large-scale experiments or to collect data over an extended period of time.
+
+This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects, resulting in potential trade-offs that may neglect one half of the larger picture.
 We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
@@
 
-At the time of the writing of this dissertation, no widely adopted methodology exists for the study of these trade-offs.
-We aim to contribute to the field by introducing and subsequently investigating the applications of such a methodology for the study of latency-sensitive applications deployed on Edge infrastructure and network technologies such as 5G.
-The proposed methodology aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
-Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of \gls{MEC}.
+This dissertation aims to contribute to the field by presenting two key contributions.
+First, we introduce and subsequently investigate the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
+The proposed methodology is based on the emulation of workloads and aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
+Next, we present a deeper exploration of the implications of improved realism in the emulated components of our methodology by studying potential avenues for optimization of a particular sub-class of \gls{MAR}.
 
-One of the main challenges to the study of latency-sensitive applications such as \gls{CPS} and \gls{MAR} on Edge Computing is that these applications interact with the real world, and thus must be studied in real-time.
-This implies that any study of these applications must be done in an environment that accurately reflects the real-world conditions in which they are deployed.
-Unfortunately, replicating these conditions in a controlled experimental environment is difficult and often leads to inconsistencies in results, and traditional methods that rely on simulations may not be sufficient to accurately capture and evaluate the performance of these applications in real-world scenarios.
-
-The scalability of studying these systems is also a challenge.
-It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, personnel and test subjects, as well as facilities.
-\glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{MAR} is time-consuming and expensive.
-
-The highly complex effects on the wireless network is another major challenge to studying latency-sensitive applications on Edge Computing.
-The presence of multiple devices accessing the network simultaneously can result in interference, network congestion, and other factors that can affect the performance of the system.
-These effects are difficult to simulate and/or emulate, particularly in multi-tenant deployments, where each tenant may have different requirements and resource usage patterns.
-
-Finally, the combination of the characteristics enumerated above makes the study of these applications hard to automate, making it challenging to conduct large-scale experiments or to collect data over an extended period of time.
-\gls{CPS} and \gls{MAR} applications are highly dynamic and adaptive, and their behavior cannot be easily predicted.
-Moreover, there is often a significant amount of variability in the behavior of different components of the application, such as sensors, actuators, and network connections, which makes it difficult to automate the testing process.
-
-Our methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{MAR} applications. 
+Our methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{XR} applications. 
 We emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
@@
 At the same time, our methodology retains the real network as well as the real compute hardware and software at the backend.
-This design results in real-time benchmarking of applications, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.
+This design results in real-time benchmarking and performance evaluation of applications and Edge infrastructure, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.
 
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for improved repeatability and replicability, as well as easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
 This also reduces the barrier of entry to this research, as \glspl{SBC} are for the most part cheap and easily accessible.
 Additionally, it preserves the realism of effects stemming from the hardware and network.
-In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
+In particular, the methodology allows us to capture effects due to the interaction of network factors such as contention, congestion control, and medium access, with effects arising due to process scheduling and/or resource contention at the compute level.
+These effects are often of stochastic or chaotic natures and complex to capture in models or simulations.
 
-The methodology also provides automation as well as improved repeatability and replicability.
-The softwarized nature of the client-side emulation allows for straightforward automation through general-purpose programming and scripting languages.
-Repeating a study becomes a matter of re-running the workload on the same testbed, and studies can be replicated simply by obtaining the same or equivalent software workload and deploying it on a comparable testbed.
-These are complex tasks to accomplish in real-world approaches, particularly when dealing with humans.
+Next, we study how improved realism in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
+To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
+We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
+We find that improved realism in the emulation enables otherwise inviable optimization approaches.
 
-We validate our proposed methodology by presenting case studies of its application in two different contexts: \gls{WCA} and \glspl{NCS}.
-The case studies demonstrate the efficacy of the methodology in evaluating system performance and overall \gls{QoS} and \gls{QoE}.
-Our methodology provides a comprehensive and realistic assessment of the performance of latency-sensitive applications deployed on Edge infrastructure.
-The approach provides valuable insights for researchers, system designers, and application developers, and will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
-
-In addition to the methodology, this dissertation introduces a number of further contributions to the field. 
-We introduce a software framework that facilitates the orchestration of Edge Computing testbeds needed for the implementation of our methodology.
-This framework makes it easier for researchers and practitioners to deploy and test Edge-bound low-latency applications.
-We also present a deep exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
-This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.
-
-Finally, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
-In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
-We conclude that significant improvements can be achieved through the use of our human user model.
-Overall, our results showcase the utility of our methodology, and the importance of considering human behavior in the design and optimization of \gls{WCA} systems to achieve maximum benefits.
+We believe our contributions represent valuable tools and insights for researchers, system designers, and application developers.
+The methodology represents a novel, holistic approach to the performance evaluation of Edge environments.
+Furthermore, our findings have significant implications for the design and optimization of applications enabled by Edge Computing, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
+We hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
 

### 2023-04-24 body/part1/chapter1-introduction/background.tex
images

@@
 \gls{VR}, on the other hand, takes it a step further and involves forgoing the real world, completely immersing the user in a fully virtual environment~\cite{doerner2022virtual}.
-Due to its deeply immersive nature, \gls{VR} requires users to wear specialized equipment such as a \gls{VR} headset.
+Due to its deeply immersive nature, \gls{VR} requires users to wear specialized equipment such as a \gls{VR} headset, as showcased in \cref{fig:vrheadset}.
 The headset contains two high-resolution displays (one for each eye) and a set of headphones, and is connected to a computer or a gaming console, which generates the virtual environment in real-time.
@@
 Some advanced \gls{VR} platforms even include omnidirectional treadmills which allow the users navigate the virtual world by walking, running, and even jumping.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=26em]{figures/girl_VR}
+    \caption[{ }]{
+        A young user wearing a \gls{VR} headset.
+        The headset completely encases the user's eyes and ears, providing a fully immersive virtual experience.\\
+
+        {
+            \footnotesize%
+            Image by Miki Yoshihito on Flickr~\cite{picture:VRheadset}.
+            Licensed under a \gls{CCBY} license.
+        }
+    }\label{fig:vrheadset}
+\end{figure}
 

### 2023-04-26 body/part1/chapter4-conclusions/conclusions.tex
abstract

--- 2023-02-14 body/part1/futurework.tex
@@
-\section{Conclusions}
+In this dissertation, we have presented two key contributions to the literature surrounding performance evaluation of Edge Computing systems.
+Firstly, we have introduced a novel methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing, such as \gls{MAR} and \glspl{CPS}.
+Our methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
+We have validated this methodology through two representative use-cases,\gls{WCA} and \glspl{NCS}.
+Further, we have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
+We have demonstrated the scalability and flexibility of our approach by deploying and studying scenarios with significant number of clients without the need for domain-specific hardware.
 
-\section{Open questions \& future work}
+Our approach presents several advantages over existing methods for the study of trade-offs in latency-sensitive applications in Edge environments.
+It minimizes the complications associated with experimental evaluation, improving key aspects such as efficiency and repeatability.
+The softwarized nature of the client-side emulation allows for straightforward automation and enhances repeatability and replicability of studies.
+Our methodology reduces complexity by moving workload components fully into the software domain, while at the same time capturing effects due to the interaction of network factors, such as congestion control and medium access, and compute factors, such as process scheduling.
+These effects are are often exceedingly complex to capture in simulations.
+Our approach thus represents a a notable contribution to the literature in its holistic approach to performance evaluation of Edge environments, encompassing both the application and the infrastructure.
 
-The present thesis presents a comprehensive exploration into the application of our methodology for \glspl{NCS} and specially for \gls{WCA}.
-Nonetheless, a number of key questions remain open for future research in this field.
+Secondly, we have explored the implications of improved client-side emulation realism on the results produced with our methodology.
+To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
+We have then used this model to study the potential for optimization of these systems.
+Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
+Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
+These behaviors play a critical role in the overall performance of such applications.
+
+\medskip
+
+Notwithstanding the important contributions of this dissertation, a number of key questions remain open for future research in this field.
 
 The first concern is the applicability of the methodology to other systems and applications in the context of Edge Computing.
-This thesis explores the methodology's use for step-based \gls{WCA} in depth, and also performs a less thorough case study on its benefits for \gls{NCS} research.
-The question remains if the methodology is suitable for
+This thesis explores the methodology's use for step-based \gls{WCA} in depth, and also performs an initial use-case study on its benefits for \gls{NCS} research.
+The question remains if the methodology is as suitable for
 \begin{inlineenum}
@@
 In accordance with the broader questions mentioned above, this thesis also raises specific questions related to its individual contributions to the literature.
-The EdgeDroid tool, introduced in papers \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2023realistic}, has the potential for further development to support other forms of \gls{WCA} applications.
-Currently, the tool is only designed for a specific class of step-based \gls{WCA} with a linear task model.
-However, the tool, and by extension the methodology, could be easily adapted to similar applications.
+Our characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
+Additionally, the timing characteristics of step-based \gls{WCA} are such that an instruction remains valid for as long as it takes the user to perform it.
+The conclusions from these works therefore may not apply to applications with different execution flows, or in which instructions have a deadline after which they may become stale or invalid.
+Examples of such applications include, for instance, the \emph{Ping-Pong} assistant introduced in~\cite{chen2015early}.
+We believe there is potential for further development of the EdgeDroid tool to support these other forms of \gls{WCA} applications.
 
-We identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
+The user models are further limited in their realism.
+In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we assume a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
+These limitations emerge from a lack of data of human behavior in these applications, as at the moment of the writing of these works no comprehensive model of human behavior existed in the relevant literature.
+As a consequence, these works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
+We attempt to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
+the resulting model is able to adapt its timing behavior in response to delay in the system.
+However, we were unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
+
+Furthermore, we identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
 The current dataset is limited to young undergraduate students from a highly selective university in the \gls{USA}.
@@
 These include resource allocation, task scheduling, user identification and classification, as well as guaranteeing both \gls{QoE} and \gls{QoS}.
-Each of these areas presents opportunities for future work and could lead to significant advancements in the field of \gls{WCA}.
+Each of these areas presents opportunities for future work and could lead to significant advancements in the field of \gls{WCA}, and \gls{MAR} and \gls{XR} in general.
 
-\medskip
-Our research on applying the methodology to \glspl{NCS} presents promising potential for future work.
+Our research on applying the methodology to \glspl{NCS} also presents promising potential for future work.
 While \cref{paper:olguinmunoz2022cleave} provides an initial exploration, further investigation through more realistic and thorough experimentation is necessary to validate the results obtained by the framework.
@@
 We also intend to extend the functionality of the \gls{CLEAVE} tool significantly, primarily to target a wider variety of \glspl{NCS} deployments.
-Currently, \gls{CLEAVE} includes a small repertoire of emulations and controllers.
-We are planning to extend this with the goal of creating an open library of \glspl{NCS} to share with the community.
+Given the vastness and complexity of the field of \glspl{NCS}, we have limited our exploration in this dissertation to simple, small-scale, single-plant systems.
+The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
+Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that our conclusions may not directly translate to larger scale or more realistic systems.
+Additionally, our tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
+This does not match well with more complex controllers such as \gls{MPC} controllers, in which each sample produces multiple outputs from the controller.
+\gls{CLEAVE} will thus have to be modified and extended to work with such setups.
 
-\medskip
 Finally, we see tremendous opportunities for testbed-driven research in Edge Computing.
@@
 
-In conclusion, the present dissertation provides a comprehensive examination of the methodology for \glspl{NCS} and \gls{WCA}, but there are still many open questions that need to be addressed in future research.
-The EdgeDroid tool and the dataset have the potential to be extended, and the optimization framework employed in the thesis could be improved.
-We also plan to extend the functionality of the \gls{CLEAVE} tool and perform large-scale experimentation on novel networking technologies.
-There are also significant opportunities for testbed-driven research in Edge Computing through the \gls{EXPECA} testbed.
+\medskip
+
+In conclusion, the present dissertation has provided a comprehensive examination of our methodology for \glspl{NCS} and \gls{WCA}.
+However, there are still many open questions that require further investigation in future research.
+The EdgeDroid tool and the dataset utilized in this study have the potential for further extension, and the optimization framework employed in the thesis could be refined.
+Plans are in place to enhance the functionality of the CLEAVE tool and conduct large-scale experimentation on novel networking technologies.
+Overall, this dissertation serves as a foundation for future research endeavors in the field of \gls{XR} applications and \glspl{CPS} on the Edge.
+The findings, tools, and methodologies presented in this study provide valuable insights and opportunities for further improvements and the exploration of novel technologies enabled by Edge Computing.

### 2023-05-05 body/part1/chapter1-introduction/introduction.tex
images
fixes suggested by markus

@@
-\section{Introduction}
+\section{Motivation}
 
@@
 \gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
+An  example of such an application is illustrated in \cref{fig:augmentedreality}.
 By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=26em]{figures/AR}
+    \caption[{ }]{
+        Example of an \gls{AR} application that overlays anatomical information on top of a live video feed of a model of human anatomy.\\
+
+        {
+            \footnotesize%
+            Photo source: Pixabay~\cite{picture:AR}.
+            Licensed under a \gls{CC0} license.
+        }
+    }\label{fig:augmentedreality}
+\end{figure}
 
@@
 
-The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical, simulation, and experimental approaches.
+The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
 However, these systems exhibit several characteristic that make their study complex.
@@
 
-This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects, resulting in potential trade-offs that may neglect one half of the larger picture.
+This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects (e.g.~\cite{chen2015efficient} and~\cite{al2017reliable}), resulting in potential trade-offs that may neglect one half of the larger picture.
 We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
@@
 We hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
-
-\subsection{Structure of this dissertation}
-
-This dissertation is structured into two parts.
-\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
-Next, in \cref{part:publications} we present the publications that form the core of this thesis.
-
-\cref{part:summary} is itself structured into four main chapters.
-\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
-It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
-In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
-The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
-We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
-\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
-We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
-We aim to demonstrate the significance of our work and how it contributes to the field.
-Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
-We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
-This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-05-05 body/part1/chapter1-introduction/structure.tex
fixes suggested by markus

@@
 This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.
-
-\subsection{Overview of the included publications}
-
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
-\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
-\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
-This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
-\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
-We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
-The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
-
-In \cref{paper:olguinmunoz2022cleave}, we extend our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
-These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, we introduce a tool, \acs{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
-This tool allows us both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
-The middleware software abstracts away the network from the development of these workloads, enabling quick prototyping and deployment.
-We then present use-case scenarios demonstrating the scalability and flexibility of their approach by deploying scenarios with a large number of loops without the need for domain-specific hardware.
-This work showcases the applicability of the methodology beyond wearable cognitive assistance and highlights the potential of our approach for other edge computing applications.
-
-\cref{paper:olguinmunoz2022ainur} presents a tangential contribution.
-It introduces the software framework used in \cref{paper:olguinmunoz2023realistic,paper:olguinmunoz2022cleave} for the orchestration of the edge computing testbeds on which the developed tools were deployed.
-This framework represents an ancillary contribution, crucial for the research presented in these works.
-Without it, the experimental approach described in these works would not have been feasible.
-
-Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
-Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
-In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
-We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
-System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
-
-\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
-This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.

### 2023-05-05 body/part1/chapter2-relatedwork/relatedwork.tex
contrib tree
images
data centers -> datacenters
fixes suggested by markus

@@
 \section{State of the art}\label{sec:state-of-the-art}
-
 \subsection{Performance evaluation of Edge Computing systems}
@@
 
-Finally, another example is \emph{FogNetSim++}~\cite{qayyum2018fognetsimpp}, tool which builds upon the \acs{OMNET} network simulator.
+Finally, another example is \emph{FogNetSim++}~\cite{qayyum2018fognetsimpp}, a tool which builds upon the \acs{OMNET} network simulator.
 FogNetSim++ allows users to simulate large Fog networks while incorporating mobility and scheduling models.
@@
 % A rich body of literature exists with examples of such approaches for Cloud computing.
-% For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
+% For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of datacenters and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
 % In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
@@
 
-\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshort{CPS}}
+\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshortpl{NCS}}
 

### 2023-05-05 body/part1/chapter3-contributions/wca-implications.tex
fixes suggested by markus

@@
 This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
-At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \textasciitilde\SI{26}{\percent} lower than for steps \num{1} to \num{4}.
+At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
 However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
@@
 This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
-Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \textasciitilde\SI{73}{\percent} of the variance in the execution time results, neuroticism was present in the first two.
-Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism was present in the first two.
+Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
 
@@
 
-\subsubsection{Design of a realistic model for human timings in \gls{WCA}}
+\subsubsection{Design of a realistic model for human timings in \gls{WCA}}\label{sec:improved_model}
 
@@
 
-In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
+In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:improved_model}.
 The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.

### 2023-05-05 body/part1/chapter4-conclusions/conclusions.tex
clean up before sending to Markus
fixes suggested by markus

@@
 Our methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
-We have validated this methodology through two representative use-cases,\gls{WCA} and \glspl{NCS}.
+We have validated this methodology through two representative use-cases, \gls{WCA} and \glspl{NCS}.
 Further, we have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
@@
 
-Secondly, we have explored the implications of improved client-side emulation realism on the results produced with our methodology.
+Secondly, we have explored the implications of improved realism on the client-side emulation on the results produced with our methodology.
 To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.

### 2023-05-05 body/abstract.tex
swedish abstract

\phantomsection\addcontentsline{toc}{chapter}{Abstract}
\begin{abstract}
    Cloud Computing, with its globally-accessible nature and virtually unlimited scalability, has revolutionized our daily lives since its widespread adoption in the early 2000s.
    It allows us to access our documents anywhere, keep in contact with our friends, back up our photos, and even remote-control some of our appliances.
    Despite this, Cloud Computing has limitations when it comes to novel applications requiring real-time processing or low-latencies.
    Applications such as \glspl{CPS} or mobile \gls{XR}, which in turn also have great transformative potential, are unable to run on the Cloud.

    Edge Computing is emerging as a potential solution to these limitations by bringing computation and data processing closer to the edge of the network, thereby reducing latency and enabling real-time decision making.
    The combination of Edge Computing and modern mobile network technologies such as 5G offers potential for massive deployments of latency-sensitive applications.
    However, scaling and understanding these deployments poses important challenges such the optimization of latency through multiple processing steps and trade-offs in wireless system choice, protocols, hardware, and algorithms.
    Existing approaches have so far been unsuccessful in capturing the complex effects arising from the interplay between network and compute in these systems.

    This dissertation addresses the challenge of performance evaluation of Edge Computing and the applications enabled by this paradigm with two key contributions to literature.
    First, a methodological approach to experimentally studying the trade-offs between system responsiveness and resource consumption in Edge-bound latency-sensitive applications such as \glspl{CPS} and \gls{XR} is introduced.
    These applications and systems feature characteristics, such as tight interaction with the physical world and the involvement of humans, that make them challenging to study through simulated approaches or analytical modeling.
    The approach presented in this thesis involves therefore the emulation of the client-side workload while maintaining the real server-side process and network stack to retain the realism of network and compute effects.

    Next, an exploration of the requirements for realism in the emulation is presented.
    Further, this work discusses the extent to which realism in the emulation can open new avenues for optimization of these systems.
    To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
    The model is combined with a mathematical framework to study the potential for optimization in these applications.

    Results indicate that the methodology introduced in this work offers advantages over existing methods by improving efficiency, repeatability, and replicability.
    By fully integrating workload components into the emulated software domain, this methodology reduces complexity while still capturing complex effects of network and compute factors that are challenging to model.
    This approach represents thus an important contribution to literature, as it consists of a comprehensive method for the performance evaluation of Edge environments, encompassing both the application and the infrastructure.
    Furthermore, results from the exploration into the implications of realism in the emulation suggest that incorporating enhanced realism in client-side emulation can enable the implementation of optimization approaches that would otherwise be infeasible.
    In particular, this work highlights the significance of considering human behavior and reactions in addition to system-related metrics and performance optimizations in the context of \gls{MAR}.
\end{abstract}

\vspace{3em}

\setlength{\leftskip}{0.3 cm} \textbf {Keywords:} Lorem, Ipsum, Dolor, Sit, Amet
\todo[inline]{Keywords!}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% SWEDISH ABSTRACT %%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\selectlanguage{swedish}
\phantomsection\addcontentsline{toc}{section}{Abstract (Swedish)}
\glsresetall%
\begin{abstract}
    Cloud Computing, med sin globalt tillgängliga natur och praktiskt taget obegränsade skalbarhet, har revolutionerat våra dagliga liv sedan dess allmänna antagande i början av 2000-talet.
    Det gör det möjligt för oss att få tillgång till våra dokument var som helst, hålla kontakt med våra vänner, säkerhetskopiera foton, och till och med fjärrstyra våra hushållsapparater. 
    Trots detta har Cloud Computing begränsningar när det gäller nya applikationer som kräver realtidsbehandling eller låg latens.
    Applikationer som cyberfysiska system (engelska \emph{\gls{CPS}}) eller mobilt utvidgad verklighet (engelska \emph{\gls{XR}}), som i sin tur också har stor omvandlingspotential, kan inte köras på molnet.

     Edge Computing är en potentiell lösning på dessa begränsningar som, genom att föra beräkningar och databehandling närmare kanten av nätverket, minskar latensen och möjliggör beslutsfattande i realtid.
     Kombinationen av Edge Computing och moderna mobila nätverksteknologier som 5G erbjuder potential för massiva utrullningar av latenskänsliga applikationer. 
     Men att skala och förstå dessa utrullningar utgör viktiga utmaningar såsom optimering av latens genom flera bearbetningssteg och avvägningar i val av trådlösa system, protokoll, hårdvara och algoritmer.
     Befintliga metoder har hittills inte lyckats fånga de komplexa effekter som uppstår från samspelet mellan nätverk och datorer i dessa system.

     Den här avhandlingen tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
     Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
     Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillvägagångssätt eller analytisk modellering.
     Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att bibehålla realismen i nätverks- och datoreffekter.

     Därefter presenteras en utforskning av kraven på realism i emuleringen.
     Vidare diskuterar detta arbete i vilken utsträckning realism i emuleringen kan öppna nya vägar för optimering av dessa system.
     För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
     Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i dessa applikationer.

     Resultaten indikerar att den metod som introducerats i detta arbete erbjuder fördelar jämfört med befintliga metoder genom att förbättra effektiviteten, repeterbarheten och replikerbarheten.
     Genom att helt integrera arbetsbelastningskomponenter i den emulerade mjukvarudomänen, minskar denna metodik komplexiteten samtidigt som den fångar komplexa effekter av nätverks- och beräkningsfaktorer som är utmanande att modellera.
     Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
     Dessutom tyder resultat från utforskningen av implikationerna av realism i emuleringen att inkorporering av förbättrad realism i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
     Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

\end{abstract}
\selectlanguage{english}
\glsresetall%

### 2023-05-08 publications/2023EdgeDroid2/resources/glossary.tex

--- 2023-02-07 body/glossary.tex
@@
 \newacronym[longplural={eXtended Realities}]{XR}{XR}{eXtended Reality}
+\newacronym[longplural={Mixed Realities}]{MR}{MR}{Mixed Reality}
+\newacronym{MAR}{MAR}{Mobile \nestedgls{AR}}
 \newacronym{GPU}{GPU}{Graphical Processing Unit}
@@
 \newacronym{YAML}{YAML}{YAML Ain't Markup Language}
-\newacronym{O-RAN}{O-RAN}{Open Radio Access Network}
+\newacronym{RAN}{RAN}{Radio Access Network}
+\newacronym{O-RAN}{O-RAN}{Open \nestedgls{RAN}}
 \newacronym{FOSS}{FOSS}{Free and Open Source Software}
@@
 \newacronym{GENI}{GENI}{Global Environment for Networking Innovation}
-
+\newacronym{SBC}{SBC}{Single-Board Computer}
+\newacronym{EXPECA}{ExPECA}{Experimental Platform for Edge Computing Applications}
+\newacronym[user1={Kungliga Tekniska Högskolan}, prefixfirst={the\ }]{KTH}{KTH}{Royal Institute of Technology of Sweden}
+\newacronym{HCI}{HCI}{Human-Computer Interaction}
+\newacronym{2D}{2-D}{two-dimensional}
+\newacronym{3D}{3-D}{three-dimensional}
+\newacronym{PD}{PD}{Proportional-Differential}
+\newacronym{PID}{PID}{Proportional-Integral-Differential}
+\newacronym{SDK}{SDK}{Software Development Kit}
+\newacronym{YCSB}{YCSB}{Yahoo! Cloud Serving Benchmark}
+\newacronym{ITS}{ITS}{Intelligent Transportation System}
+\newacronym{OOS}{OoS}{Out-of-Scope}
+\newacronym{OMNET}{OMNeT++}{Objective Modular Network Testbed in C++}
+\newacronym{ns2}{ns-2}{network simulator 2}
+\newacronym{ns3}{ns-3}{network simulator 3}
+\newacronym{IRB}{IRB}{Institutional Review Board}
+\newacronym{IaaS}{IaaS}{Infrastructure-as-a-Service}
+\newacronym{CC}{CC}{Creative Commons}
+\newacronym{CCBY}{CCBY}{\nestedgls{CC} Attribution 2.0 Generic~\cite{ccbylicense}}
+\newacronym{CC0}{CC0}{\nestedgls{CC} Zero 1.0 Universal Public Domain Dedication~\cite{cczerolicense}}
 %\glsaddall%

### 2023-05-10 body/acks.tex
acks

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ACKNOWLEDGMENT %%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Acknowledgements}
\glsresetall%

I would like to begin by thanking my supervisor, Prof. James Gross, for his guidance and patience during this long process.\todo{todo}

I would like to express my deepest gratitude to Prof. Mahadev ``Satya'' Satyanarayanan, for hosting me repeatedly, over long periods of time, at \gls{CMU} as a visiting researcher.
His tremendous experience and generosity have been invaluable, and his guidance and advise have certainly shaped my work.
Likewise, I'm deeply indebted to Prof. Roberta ``Bobby'' Klatzky of \gls{CMU} for the many helpful and insightful discussions, for introducing me to the field of Human-Computer Interaction, and for always greeting me with a warm smile.

I would also like to thank everybody involved in the defense of this thesis.
I am very grateful to Prof. Yu Xiao of Aalto University, for agreeing to act as my dissertation opponent.
My dissertation defence will certainly benefit greatly from her expertise and insights.
I would like to extend my sincere thanks to my grading committe, consisting of Prof. Ana Aguiar (Universidade do Porto), Prof. Per Gunningberg (Uppsala Universitet), and Prof. Klaus Wehrle (\gls{RWTH}), for investing time and effort into reading and grading this dissertation.
Special thanks also go to Prof. Markus Flierl for advance-reviewing this dissertation, and to Prof. Mats Bengtsson for agreeing to act as chairperson during the defense.

I also extend my thanks to my colleagues and co-authors.
To Samie, Vishnu, and Neel at \pgls{KTH}, for always being willing to engage in interesting discussions, but also for generally being fun to be around.
And to Junjue and Babu Pillai from \gls{CMU} for enriching my research with their interesting insights.

I am very lucky and grateful that I got to stay at both \pgls{KTH} and \gls{CMU} over the years, and I would like to thank everyone who made these spaces welcoming and engaging.
A big thank you goes out to my colleagues (both current and former) at the division of \gls{ISE} for all those fun lunches at Restaurang Q, Cypern, or SysterOBror over the years, which always helped to break the monotony of the day.
This includes but is not limited to Samie, Vishnu, Neel, Sahar, Antonios, Germán, Boules, Baptiste, Håkan, Pol, Marie, and Sebastian.
I would additionally like to extend a special thank you to Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first moths.
Another big thank you goes out to everyone (current and former) at Satya's group at \gls{CMU}, for always welcoming me with open arms and treating me like one of them whenever I was visiting.
This includes but is not limited to Junjue, Roger, Shilpa, Tom, Jan, Edmond, Jim, Babu, and Zhuo.

I would like to extend my sincere thanks to Prof. Sandra Céspedes, of Universidad de Chile and Concordia University, for her guidance and urging to embark on this journey towards a Ph.D., and for always being willing to offer help and advice.
Likewise, I'd like to thank Prof. Javier Bustos, of Universidad de Chile and NIC Chile Research Labs, for giving me an opportunity many years ago to get into distributed systems research at NIC Labs.
Also, I'd like to extend my gratitude to Prof. José Miguel Piquer, of Universidad de Chile, for taking me as his ``long-distance'' teaching assistant and allowing me to give back to my \emph{alma mater} from across the ocean.

\medskip

This dissertation would not have been possible without the help and support of my family.
First and foremost, I would like to thank the love of my life and wife~\footnote{\emph{future-wife} at the time of writing, but \emph{wife} at the time of the defense of this thesis}, Andresa.
I could not have done this without her companionship, love, and unconditional support.\todo{TODO}

I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
They have been my rocks, my pillars, since the day of my birth, and I am proud to be their son.
A big thank you goes out to my sister, Paola, as well.
Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.

### 2023-05-10 resources/glossary.tex
fixes
perf eval
sim
acks

@@
 \newacronym{ITS}{ITS}{Intelligent Transportation System}
-
+\newacronym{OOS}{OoS}{Out-of-Scope}
+\newacronym{OMNET}{OMNeT++}{Objective Modular Network Testbed in C++}
+\newacronym{ns2}{ns-2}{network simulator 2}
+\newacronym{ns3}{ns-3}{network simulator 3}
+\newacronym{IRB}{IRB}{Institutional Review Board}
+\newacronym{IaaS}{IaaS}{Infrastructure-as-a-Service}
+\newacronym{CC}{CC}{Creative Commons}
+\newacronym{CCBY}{CCBY}{\nestedgls{CC} Attribution 2.0 Generic~\cite{ccbylicense}}
+\newacronym{CC0}{CC0}{\nestedgls{CC} Zero 1.0 Universal Public Domain Dedication~\cite{cczerolicense}}
+\newacronym{ISE}{ISE}{Information Science and Engineering}
+\newacronym{RWTH}{RWTH Aachen}{Rheinisch-Westfälische Technische Hochschule Aachen}
 %\glsaddall%

### 2023-05-10 body/acks.tex
acks

@@
 \chapter{Acknowledgements}
+\epigraph{La educación es, tal vez, la forma más alta de buscar a Dios.}{\emph{Gabriela Mistral}}
 \glsresetall%
 
-I would like to begin by thanking my supervisor, Prof. James Gross, for his guidance and patience during this long process.\todo{todo}
+There are many, many people I owe gratitude to for making this dissertation possible.
+I would like to begin by thanking my supervisor, Prof. James Gross.
+This dissertation would not have been possible without his guidance and motivation during this long process.
+His supervision opened doors that I had never imagined would open for me, and his dry German humor has certainly helped cheer me up whenever I was overwhelmed by some deadline.
+Above all, I would like to thank James for his patience and understanding whenever I encountered difficulties, and for the flexibility he has afforded me in completing this Ph.D.
 
-I would like to express my deepest gratitude to Prof. Mahadev ``Satya'' Satyanarayanan, for hosting me repeatedly, over long periods of time, at \gls{CMU} as a visiting researcher.
-His tremendous experience and generosity have been invaluable, and his guidance and advise have certainly shaped my work.
+I would like to express my deepest gratitude to Prof. Mahadev ``Satya'' Satyanarayanan, for welcoming me repeatedly, over long periods of time, as a visiting researcher at \gls{CMU}.
 Likewise, I'm deeply indebted to Prof.
 Roberta ``Bobby'' Klatzky of \gls{CMU} for the many helpful and insightful discussions, for introducing me to the field of Human-Computer Interaction, and for always greeting me with a warm smile.
+Both Satya's and Bobby's tremendous experience and generosity have been invaluable, and their guidance and advice have certainly shaped my work for the better.
 
-I would also like to thank everybody involved in the defense of this thesis.
+I would also like to thank everybody involved with the defense of this thesis.
 I am very grateful to Prof. Yu Xiao of Aalto University, for agreeing to act as my dissertation opponent.
-My dissertation defence will certainly benefit greatly from her expertise and insights.
+My dissertation defense will certainly benefit greatly from her expertise and insights.
 I would like to extend my sincere thanks to my grading committe, consisting of Prof.
@@
 Klaus Wehrle (\gls{RWTH}), for investing time and effort into reading and grading this dissertation.
-Special thanks also go to Prof. Markus Flierl for advance-reviewing this dissertation, and to Prof. Mats Bengtsson for agreeing to act as chairperson during the defense.
+Special thanks also go to Prof.
+Markus Flierl of \pgls{KTH} for advance-reviewing this dissertation, and to Prof.
+Mats Bengtsson, also of \pgls{KTH}, for agreeing to act as chairperson for the defense.
 
-I also extend my thanks to my colleagues and co-authors.
-To Samie, Vishnu, and Neel at \pgls{KTH}, for always being willing to engage in interesting discussions, but also for generally being fun to be around.
-And to Junjue and Babu Pillai from \gls{CMU} for enriching my research with their interesting insights.
+Moreover, I extend my thanks to my co-authors and colleagues.
+To Samie Mostafavi, Vishnu N. Moothedath, and Neelabhro Roy at \pgls{KTH}, for always being willing to engage in interesting discussions, but also for generally being fun to be around.
+And to Dr. Jaya Prakash (IMDEA Networks Institute), Dr. Junjue Wang (\gls{CMU}), and Dr. Padmanabhan ``Babu'' Pillai (\gls{CMU}) for enriching my research with their interesting insights.
 
-I am very lucky and grateful that I got to stay at both \pgls{KTH} and \gls{CMU} over the years, and I would like to thank everyone who made these spaces welcoming and engaging.
-A big thank you goes out to my colleagues (both current and former) at the division of \gls{ISE} for all those fun lunches at Restaurang Q, Cypern, or SysterOBror over the years, which always helped to break the monotony of the day.
-This includes but is not limited to Samie, Vishnu, Neel, Sahar, Antonios, Germán, Boules, Baptiste, Håkan, Pol, Marie, and Sebastian.
-I would additionally like to extend a special thank you to Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first moths.
+I am very lucky and grateful that I got to stay at both \pgls{KTH} and \gls{CMU}, and I would like to thank everyone who made these spaces welcoming and engaging.
+A big thank you goes out to my colleagues (both current and former) at the division of \gls{ISE} for all those fun lunches at Restaurang Q, Cypern, and SysterOBror over the years, which always helped to break the monotony of the days.
+This includes but is not limited to Samie, Vishnu, Neel, Sahar, Antonios, Germán, Boules, Baptiste, Håkan, Pol, Marie, Wendi, and Sebastian.
+I would additionally like to extend a special thank you to Dr.
+Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first moths.
+Likewise, I extend my gratitute to Gerd Fransson, for greeting me with a huge smile every morning the years our offices faced each other and cheering up my days with her wholesome energy.
+I also thank Prof. Joakim Jaldén for giving me the opportunity to act as his teaching assistant.
 Another big thank you goes out to everyone (current and former) at Satya's group at \gls{CMU}, for always welcoming me with open arms and treating me like one of them whenever I was visiting.
-This includes but is not limited to Junjue, Roger, Shilpa, Tom, Jan, Edmond, Jim, Babu, and Zhuo.
+This includes but is not limited to Junjue, Tom, Roger, Shilpa, Jan, Edmond, Jim, Babu, and Zhuo.
+I will always remember our Monday morning talks and Friday lunch outings dearly.
 
-I would like to extend my sincere thanks to Prof.
+Furthermore, I would like to extend my sincere thanks to Prof.
 Sandra Céspedes, of Universidad de Chile and Concordia University, for her guidance and urging to embark on this journey towards a Ph.D., and for always being willing to offer help and advice.
-Likewise, I'd like to thank Prof. Javier Bustos, of Universidad de Chile and NIC Chile Research Labs, for giving me an opportunity many years ago to get into distributed systems research at NIC Labs.
+Likewise, I'd like to thank Prof. Javier Bustos, of Universidad de Chile and NIC Chile Research Labs, for giving me the opportunity many years ago to first get into distributed systems research.
 Also, I'd like to extend my gratitude to Prof.
@@
 This dissertation would not have been possible without the help and support of my family.
-First and foremost, I would like to thank the love of my life and wife~\footnote{\emph{future-wife} at the time of writing, but \emph{wife} at the time of the defense of this thesis}, Andresa.
-I could not have done this without her companionship, love, and unconditional support.\todo{TODO}
+First and foremost, I would like to thank the love of my life and wife~\footnote{\emph{Future-wife} at the time of writing, but \emph{wife} at the time of the defense of this thesis!}, Andresa.
+I could not have done this without her companionship and love.
+I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
 
 I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
-They have been my rocks, my pillars, since the day of my birth, and I am proud to be their son.
+They have been my rocks, my pillars, since the day I was born, and I am proud to be their son.
 A big thank you goes out to my sister, Paola, as well.
 Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
+Likewise, I extend my deepest gratitude to everyone in my immediate and extended family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
+I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.
+I hope he would have been proud seeing how far I have come.
+Another special mention goes to my uncle Manuel, who graciously took me in and gave me a place to stay at the first few months of my Ph.D. studies.
+
+I would also like to thank all of my friends who have made this long, ardous journey more amenable.
+I would particularly like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jennifer, Alyson, Kuai-Kuai, Carmen, Moataz and Nora, Eric, and many others.
+They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
+I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to hang with.
+Further, I am deeply grateful for my friends in Chile who have remained a source of support, even though we only get to see each other once a year when I visit.
+This includes, but is not limited to, Juan Pablo, Fernanda, Diego, Camila, George, Patricio, Kevin, Pablo, and many others.
+Yet another special mention goes to Diego's family, who have always treated me as one of their own.
+
+Finally, I would be remiss in not mentioning my pets and animal friends who have provided me with love and warmth during these years.
+I thank my dogs, Aros and Skatt, who I sadly had to leave behind in Chile with my sister, but who still greet me with enthusiastically wagging tails whenever I visit.
+And I thank my cats, Mochi and Tiramisú, who, although have only been in my life for about a year, have sweetened and made much more bearable the long days and nights I have spent writing this dissertation.
+
+\bigskip
+
+\noindent{}Manuel Olguín Muñoz\\
+Pittsburgh, May 2023

### 2023-05-10 body/abstract.tex
mark as prelim

@@
 
-\setlength{\leftskip}{0.3 cm} \textbf {Keywords:} Lorem, Ipsum, Dolor, Sit, Amet
-\todo[inline]{Keywords!}
+\setlength{\leftskip}{0.3 cm} \textbf {Keywords:} Cloud Computing; Edge Computing; Performance Evaluation; Latency-sensitive Applications; Distributed Systems; Emulation; Extended Reality; Cyber-Physical Systems; Networked Control Systems; Wearable Cognitive Assistance.
 

### 2023-05-11 body/acks.tex
acks

@@
 José Miguel Piquer, of Universidad de Chile, for taking me as his ``long-distance'' teaching assistant and allowing me to give back to my \emph{alma mater} from across the ocean.
+A very special thanks goes out to Prof.
+Jérémy Barbay of Universidad de Chile and Universidad de Concepción, for being a mentor and a friend since the day I took his course on Android programming while an undergraduate student.
+I am deeply appreciative of the fact that although our interactions have been sporadic at best these last few years, I can always reach out to him for help and advice when I need it.
 
@@
 They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
-I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to hang with.
-Further, I am deeply grateful for my friends in Chile who have remained a source of support, even though we only get to see each other once a year when I visit.
+I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to be around.
+Further, I am deeply grateful to my friends in Chile who have remained a source of support, even though we only get to see each other once a year when I visit.
 This includes, but is not limited to, Juan Pablo, Fernanda, Diego, Camila, George, Patricio, Kevin, Pablo, and many others.

### 2023-05-11 body/acks.tex
acks

@@
 A very special thanks goes out to Prof.
-Jérémy Barbay of Universidad de Chile and Universidad de Concepción, for being a mentor and a friend since the day I took his course on Android programming while an undergraduate student.
-I am deeply appreciative of the fact that although our interactions have been sporadic at best these last few years, I can always reach out to him for help and advice when I need it.
+Jérémy Barbay of Universidad de Chile and Universidad de Concepción, for being both a mentor and a friend since the day I took his course on Android programming as an undergraduate student.
+I am deeply appreciative of the fact that although our interactions these last few years have been sporadic at best, I can always reach out to him for help and advice when I need them.
 

### 2023-05-12 body/acks.tex
acks

@@
 I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
-
 I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
@@
 
-I would also like to thank all of my friends who have made this long, ardous journey more amenable.
-I would particularly like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jennifer, Alyson, Kuai-Kuai, Carmen, Moataz and Nora, Eric, and many others.
+I would also like to thank all of my friends who have made this long, arduous journey more amenable.
+I would particularly like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jennifer, Alyson, Carmen, Kuai-Kuai and Eric, Moataz and Nora, and many others.
+\todo[inline]{Special mention Jenn and Alyson?}
 They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
@@
 
-\bigskip
-
-\noindent{}Manuel Olguín Muñoz\\
-Pittsburgh, May 2023
+\medskip
+{\begin{flushright}
+    Manuel Olguín Muñoz, May 2023
+\end{flushright}}

### 2023-05-18 body/abstract.tex
final version

@@
 
-    Next, an exploration of the requirements for realism in the emulation is presented.
-    Further, this work discusses the extent to which realism in the emulation can open new avenues for optimization of these systems.
+    Next, an exploration of the requirements for accuracy in the emulation is presented.
+    Further, this work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
     To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
@@
 
-     Den här avhandlingen tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
+    Denna avhandling tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
     Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
@@
 
-     Därefter presenteras en utforskning av kraven på realism i emuleringen.
-     Vidare diskuterar detta arbete i vilken utsträckning realism i emuleringen kan öppna nya vägar för optimering av dessa system.
+    Därefter presenteras en utforskning av kraven på noggrannhet i emuleringen.
+    Vidare diskuterar detta arbete i vilken utsträckning noggrannhet i emuleringen kan öppna nya vägar för optimering av dessa system.
     För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
@@
     Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
-     Dessutom tyder resultat från utforskningen av implikationerna av realism i emuleringen att inkorporering av förbättrad realism i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
+    Dessutom tyder resultat från utforskningen av implikationerna av noggrannhet i emuleringen att inkorporering av förbättrad noggrannhet i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
     Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

### 2023-05-18 body/acks.tex
final version

--- 2023-05-11 body/acks.tex
@@
 
-There are many, many people I owe gratitude to for making this dissertation possible.
+There are many people I owe gratitude to for making this dissertation possible.
 I would like to begin by thanking my supervisor, Prof. James Gross.
@@
 This dissertation would not have been possible without the help and support of my family.
-First and foremost, I would like to thank the love of my life and wife~\footnote{\emph{Future-wife} at the time of writing, but \emph{wife} at the time of the defense of this thesis!}, Andresa.
+First and foremost, I would like to thank the love of my life and wife, Andresa.
 I could not have done this without her companionship and love.
 I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
-
 I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
@@
 I hope he would have been proud seeing how far I have come.
-Another special mention goes to my uncle Manuel, who graciously took me in and gave me a place to stay at the first few months of my Ph.D. studies.
 
-I would also like to thank all of my friends who have made this long, ardous journey more amenable.
-I would particularly like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jennifer, Alyson, Kuai-Kuai, Carmen, Moataz and Nora, Eric, and many others.
+I would also like to thank all of my friends who have made this long, arduous journey more amenable.
+I would specially like to thank everyone I met through Tartan Salsa, including, but not limited to, Andresa, Jenn, Alyson, Carmen, Kuai-Kuai and Eric, Moataz and Nora, Russell, and many others.
 They made Pittsburgh feel like home, even before I finally moved there permanently in 2022.
+Andresa and I are particularly indebted to Jenn and Alyson for helping organize our elopement ceremony during a very stressful period of time, and for always stepping up whenever we need a cat-sitter.
 I also want to thank Ann-Charlotte in Stockholm, for always being so kind, so considerate, and so much fun to be around.
@@
 I thank my dogs, Aros and Skatt, who I sadly had to leave behind in Chile with my sister, but who still greet me with enthusiastically wagging tails whenever I visit.
-And I thank my cats, Mochi and Tiramisú, who, although have only been in my life for about a year, have sweetened and made much more bearable the long days and nights I have spent writing this dissertation.
+And I thank my cats, Mochi and Tiramisú, who, although they have only been in my life for about a year, have sweetened and made much more bearable the long days and nights I have spent writing this dissertation.
 
-\bigskip
-
-\noindent{}Manuel Olguín Muñoz\\
-Pittsburgh, May 2023
+\medskip
+{\begin{flushright}
+    Manuel Olguín Muñoz\\
+    May 2023
+\end{flushright}}

### 2023-05-18 body/part1/chapter1-introduction/introduction.tex
final version

@@
 The current model for distributed computing, known as Cloud Computing, allows users to access shared pools of resources, services, and applications over the internet~\cite{gai2012towards}.
-Cloud computing offers many advantages, including as virtually unlimited storage and processing power
-This achieved by virtue of employing economies of scale in massive installations containing up to thousands of compute servers and networking equipment.
+Cloud computing offers many advantages, including as virtually unlimited storage and processing power.
+This is achieved by virtue of employing economies of scale in massive installations containing up to thousands of compute servers and networking equipment.
 These installations are known as \emph{datacenters}, and often serve massive amounts of users.
@@
 
-Next, we study how improved realism in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
+Next, we study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
 To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
 We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
-We find that improved realism in the emulation enables otherwise inviable optimization approaches.
+We find that improved accuracy in the emulation enables otherwise inviable optimization approaches.
 

### 2023-05-18 body/part1/chapter2-relatedwork/key_contributions.tex
final version

@@
             ]
-            [{Implications\\of Realism}
+            [{Implications of\\Accuracy in\\the Emulation}
                 [{Characterization of\\human behavior in \acs{WCA}}]
@@
 This methodology is validated by example through two representative use-cases, and we provide tooling for its implementation.
-Secondly, we study the implications improved client-side emulation realism has on the results produced with our methodology.
+Secondly, we study the implications improved client-side emulation accuracy has on the results produced with our methodology.
 These implications are again explored by example.
@@
 
-The second key contribution of this dissertation relates to the exploration of the question of to which degree a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
+The second key contribution of this dissertation relates to the exploration of the question of what degree of detail about the physical world must be captured by a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
 We explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which we then combine with mathematical framework to study the potential for optimization of these applications.

### 2023-05-18 body/part1/chapter3-contributions/methodology.tex
final version

@@
 In \cref{summary:methodology}, we discuss our methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
-This methodology involves emulating only the end-user component of such applications while using the real network and compute components of the system. 
+This methodology involves emulating only the end-user, client-side component of such applications while using the real network and compute components of the system. 
 We further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.
 
-Next, in \cref{summary:realism}, we explore the extent to which our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing needs to capture certain details of the physical world.
+Next, in \cref{summary:realism}, we explore the extent of the need for accuracy in our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
 We broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which we then employ in our methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.

### 2023-05-18 body/part1/chapter3-contributions/wca-implications.tex
mark as prelim
final version

@@
-\section{Exploring the need for realism in client-side emulation}\label{summary:realism}
+\section{Exploring the need for accuracy in client-side emulation}\label{summary:realism}
 
 In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
-In particular, we wish to explore the impact improving the realism of the client-side emulation has on the potential for optimization of the applications under evaluation.
+In particular, we wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
 We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.
@@
 Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
-By improving the realism of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
-This research highlights the importance of the realism of emulations as  crucial factor in the application of our methodology.
+By improving the accuracy of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
+This research highlights the importance of the accuracy of emulations as  crucial factor in the application of our methodology.
 
@@
     }\label{fig:stepblock}
-    \todo[inline]{This figure may need reworking for clarity.}
+    % \todo[inline]{This figure may need reworking for clarity.}
 \end{figure}
@@
 
-\subsubsection{Design of a realistic model for human timings in \gls{WCA}}\label{sec:improved_model}
+\subsubsection{Design of a more accurate model for human timings in \gls{WCA}}\label{sec:improved_model}
 

### 2023-05-18 body/part1/chapter4-conclusions/conclusions.tex
final version

@@
 
-Secondly, we have explored the implications of improved realism on the client-side emulation on the results produced with our methodology.
+Secondly, we have explored the implications of improved accuracy on the client-side emulation on the results produced with our methodology.
 To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
 We have then used this model to study the potential for optimization of these systems.
-Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
+Our findings indicate that incorporating improved accuracy in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
 Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.

### 2023-05-19 body/abstract.tex
cleaning up for publication

@@
     Cloud Computing, with its globally-accessible nature and virtually unlimited scalability, has revolutionized our daily lives since its widespread adoption in the early 2000s.
-    It allows us to access our documents anywhere, keep in contact with our friends, back up our photos, and even remote-control some of our appliances.
+    It allows us to access our documents anywhere, keep in touch with friends, back up photos, and even remotely control our appliances.
     Despite this, Cloud Computing has limitations when it comes to novel applications requiring real-time processing or low-latencies.
@@
 
-    Edge Computing is emerging as a potential solution to these limitations by bringing computation and data processing closer to the edge of the network, thereby reducing latency and enabling real-time decision making.
+    Edge Computing is emerging as a potential solution to these limitations by bringing computation closer to the edge of the network, thereby reducing latency and enabling real-time decision making.
     The combination of Edge Computing and modern mobile network technologies such as 5G offers potential for massive deployments of latency-sensitive applications.
@@
     This dissertation addresses the challenge of performance evaluation of Edge Computing and the applications enabled by this paradigm with two key contributions to literature.
-    First, a methodological approach to experimentally studying the trade-offs between system responsiveness and resource consumption in Edge-bound latency-sensitive applications such as \glspl{CPS} and \gls{XR} is introduced.
+    First, a methodological approach to experimentally studying the trade-offs between system responsiveness and resource consumption in latency-sensitive applications such as \glspl{CPS} and \gls{XR} is introduced.
     These applications and systems feature characteristics, such as tight interaction with the physical world and the involvement of humans, that make them challenging to study through simulated approaches or analytical modeling.
@@
     Next, an exploration of the requirements for accuracy in the emulation is presented.
-    Further, this work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
+    This work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
     To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
-    The model is combined with a mathematical framework to study the potential for optimization in these applications.
+    The model is combined with a mathematical framework to study the potential for optimization in Edge Computing applications.
 
@@
     Cloud Computing, med sin globalt tillgängliga natur och praktiskt taget obegränsade skalbarhet, har revolutionerat våra dagliga liv sedan dess allmänna antagande i början av 2000-talet.
-    Det gör det möjligt för oss att få tillgång till våra dokument var som helst, hålla kontakt med våra vänner, säkerhetskopiera foton, och till och med fjärrstyra våra hushållsapparater. 
+    Det gör det möjligt för oss att få tillgång till våra dokument var som helst, hålla kontakt med vänner, säkerhets-kopiera foton, och fjärrstyra våra hushållsapparater. 
     Trots detta har Cloud Computing begränsningar när det gäller nya applikationer som kräver realtidsbehandling eller låg latens.
@@
 
-    Edge Computing är en potentiell lösning på dessa begränsningar som, genom att föra beräkningar och databehandling närmare kanten av nätverket, minskar latensen och möjliggör beslutsfattande i realtid.
+    Edge Computing framstår som en potentiell lösning på dessa begräns-ningar genom att föra beräkningar närmare kanten av nätverket, vilket minskar latensen och möjliggör beslutsfattande i realtid.
     Kombinationen av Edge Computing och moderna mobila nätverksteknologier som 5G erbjuder potential för massiva utrullningar av latenskänsliga applikationer. 
@@
     Denna avhandling tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
-    Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
-    Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillvägagångssätt eller analytisk modellering.
-    Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att bibehålla realismen i nätverks- och datoreffekter.
+    Först introduceras ett metodologiskt tillväga-gångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
+    Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillväga-gångssätt eller analytisk modellering.
+    Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att behålla realismen i nätverks- och datoreffekter.
 
-    Därefter presenteras en utforskning av kraven på noggrannhet i emuleringen.
-    Vidare diskuterar detta arbete i vilken utsträckning noggrannhet i emuleringen kan öppna nya vägar för optimering av dessa system.
-    För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
-    Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i dessa applikationer.
+    Därefter presenteras en utforskning av kraven på precision i emuleringen.
+    Detta arbete diskuterar i vilken utsträckning precision i emuleringen kan öppna nya vägar för optimering av dessa system.
+    För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobil förstärkt verklighet (engelska \emph{\gls{MAR}}).
+    Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i Edge Computing applikationer.
 
@@
     Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
-    Dessutom tyder resultat från utforskningen av implikationerna av noggrannhet i emuleringen att inkorporering av förbättrad noggrannhet i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
+    Dessutom tyder resultat från utforskningen av implikationerna av precision i emuleringen att inkorporering av förbättrad precision i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
     Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

### 2023-05-19 body/acks.tex
cleaning up for publication

@@
 I am very grateful to Prof. Yu Xiao of Aalto University, for agreeing to act as my dissertation opponent.
-My dissertation defense will certainly benefit greatly from her expertise and insights.
+My defense will certainly benefit greatly from her expertise and insights.
 I would like to extend my sincere thanks to my grading committe, consisting of Prof.
@@
 Klaus Wehrle (\gls{RWTH}), for investing time and effort into reading and grading this dissertation.
-Special thanks also go to Prof.
-Markus Flierl of \pgls{KTH} for advance-reviewing this dissertation, and to Prof.
-Mats Bengtsson, also of \pgls{KTH}, for agreeing to act as chairperson for the defense.
+Special thanks also go to Prof. Markus Flierl of \pgls{KTH} for advance-reviewing this work, and to Prof. Mats Bengtsson, also of \pgls{KTH}, for agreeing to act as chairperson for the defense.
 
@@
 I would additionally like to extend a special thank you to Dr.
-Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first moths.
+Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first months.
 Likewise, I extend my gratitute to Gerd Fransson, for greeting me with a huge smile every morning the years our offices faced each other and cheering up my days with her wholesome energy.
@@
 Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
-Likewise, I extend my deepest gratitude to everyone in my immediate and extended family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
+Likewise, I extend my deepest gratitude to everyone in my family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
 I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.

### 2023-05-19 body/part1/chapter1-introduction/background.tex
cleaning up for publication

@@
 
-In this section, we will discuss two key examples of such systems, \glspl{CPS} and \gls{XR} applications.
+In this section, I will discuss two key examples of such systems, \glspl{CPS} and \gls{XR} applications.
 \glspl{CPS} are systems that integrate physical and computational components, while \gls{XR} applications involve the creation of immersive virtual environment that combine the physical and virtual worlds.
@@
 
-In this dissertation, we focus specifically on \emph{step-based} \gls{WCA}.
+In this dissertation, I focus specifically on \emph{step-based} \gls{WCA}.
 These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users, illustrated in \cref{fig:wca:operation}.

### 2023-05-19 body/part1/chapter1-introduction/introduction.tex
cleaning up for publication

@@
 Physically, Cloud datacenters are often located hundreds or even thousands of kilometers from end-users.
-Connections are limited in their responsiveness by the speed of light, and these long physical distances directly translate into higher latencies.
+Connections are limited in their responsiveness by the speed of light, and long physical distances directly translate into higher latencies.
 
@@
 
-\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive application systems.
+\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive applications.
 \gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
 For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
-This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
+This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}).
 \gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
@@
 
-The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
-However, these systems exhibit several characteristic that make their study complex.
-One of these is unique nature of Edge computing environments, which tightly integrates compute and network.
-This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
+The evaluation of performance in Edge Computing has been addressed in existing works through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
+However, these systems exhibit characteristic that make their study complex.
+One of these is the unique nature of Edge computing environments, which tightly integrates compute and network.
+This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, and queuing analysis.
 Such expertise may not always be feasible to obtain within a single research group.
@@
 The scalability of these systems is also a challenge.
-It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, personnel and test subjects, as well as facilities.
+It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, test subjects, as well as facilities.
 \glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{XR} is time-consuming and expensive.
@@
 This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects (e.g.~\cite{chen2015efficient} and~\cite{al2017reliable}), resulting in potential trade-offs that may neglect one half of the larger picture.
-We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
+I argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
 There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.
@@
 This dissertation aims to contribute to the field by presenting two key contributions.
-First, we introduce and subsequently investigate the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
+First, I introduce and subsequently investigate the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
 The proposed methodology is based on the emulation of workloads and aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
-Next, we present a deeper exploration of the implications of improved realism in the emulated components of our methodology by studying potential avenues for optimization of a particular sub-class of \gls{MAR}.
+Next, I present a deeper exploration of the implications of improved accuracy in the emulated components of my methodology by studying potential avenues for optimization of a particular sub-class of \gls{MAR}.
 
-Our methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{XR} applications. 
-We emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
-    In our initial implementation, these correspond to low-cost and easily replaceable, scalable Raspberry Pi 4 Model B \acsp{SBC}.
+My methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{XR} applications. 
+I emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
+    In my initial implementation, these correspond to low-cost and easily replaceable, scalable Raspberry Pi 4 Model B \acsp{SBC}.
 } running a realistic software imitation of the desired behaviors.
-At the same time, our methodology retains the real network as well as the real compute hardware and software at the backend.
+At the same time, my methodology retains the real network as well as the real compute hardware and software at the backend.
 This design results in real-time benchmarking and performance evaluation of applications and Edge infrastructure, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.
@@
 Additionally, it preserves the realism of effects stemming from the hardware and network.
-In particular, the methodology allows us to capture effects due to the interaction of network factors such as contention, congestion control, and medium access, with effects arising due to process scheduling and/or resource contention at the compute level.
+In particular, the methodology allows for the capture of effects due to the interaction of network factors such as contention, congestion control, and medium access, with factors due to process scheduling and/or resource contention at the compute level.
 These effects are often of stochastic or chaotic natures and complex to capture in models or simulations.
 
-Next, we study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
-To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
-We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
-We find that improved accuracy in the emulation enables otherwise inviable optimization approaches.
+Next, I study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
+To achieve this, I develop a model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
+I combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications, and find that improved accuracy in the emulation enables otherwise inviable optimization approaches.
 
-We believe our contributions represent valuable tools and insights for researchers, system designers, and application developers.
+I believe my contributions represent valuable tools and insights for researchers, system designers, and application developers.
 The methodology represents a novel, holistic approach to the performance evaluation of Edge environments.
-Furthermore, our findings have significant implications for the design and optimization of applications enabled by Edge Computing, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
-We hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.
+Furthermore, my findings have significant implications for the design and optimization of applications enabled by Edge Computing, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
+I hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing.

### 2023-05-19 body/part1/chapter1-introduction/structure.tex
cleaning up for publication

@@
 This dissertation is structured into two parts.
-\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
-Next, in \cref{part:publications} we present the publications that form the core of this thesis.
+\cref{part:summary} presents a summary, introducing the topic, discussing related work, and highlighting key contributions.
+Next, in \cref{part:publications} I present the publications that form the core of this thesis.
 
@@
 It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
-In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
+In \cref{chap:relwork}, I present the scope of the thesis, as well as relevant related work.
 The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
-We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
+I discuss context for my research, and establish the foundation for the contributions presented in the dissertation.
 \cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
-We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
-We aim to demonstrate the significance of our work and how it contributes to the field.
+I discuss methods, tools, and findings that were developed and obtained throughout the course of my research.
+I aim to demonstrate the significance of my work and how it contributes to the field.
 Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
-We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
+I summarize the key findings and contributions of the thesis, and briefly discuss the implications of my findings and contributions for the field.
 This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

### 2023-05-19 body/part1/chapter2-relatedwork/key_contributions.tex
cleaning up for publication

@@
-\section{Overview of our contributions}\label{sec:contributions}
+\section{Overview of the contributions of this dissertation}\label{sec:contributions}
 
@@
             ]
-            [{Implications of\\Accuracy in\\the Emulation}
+            [{Implications of\\accuracy in\\the emulation}
                 [{Characterization of\\human behavior in \acs{WCA}}]
@@
 This thesis presents two key contributions and a number of sub-contributions to literature, as illustrated in \cref{fig:contribs}.
-Firstly, we introduce a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
-This methodology is validated by example through two representative use-cases, and we provide tooling for its implementation.
-Secondly, we study the implications improved client-side emulation accuracy has on the results produced with our methodology.
+Firstly, I introduce a methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing such as \gls{MAR} and \glspl{CPS}.
+This methodology is validated by example through two representative use-cases, and I provide tooling for its implementation.
+Secondly, I study the implications improved client-side emulation accuracy has on the results produced with our methodology.
 These implications are again explored by example.
-We design and implement of a detailed model of human user behavior in \gls{WCA}, and then use it to study the potential for optimization of these systems.
+I design and implement of a detailed model of human user behavior in \gls{WCA}, and then use it to study the potential for optimization of these systems.
 
@@
 
-Our first contribution corresponds to a methodological approach to the study and performance evaluation of Edge-enabled latency-sensitive applications. 
+My first contribution corresponds to a methodological approach to the study and performance evaluation of Edge-enabled latency-sensitive applications. 
 This approach is based on the novel idea of emulating \emph{only} the end-user component of such applications while using the \emph{real} network and compute components of the system.
-In other words, our methodology involves replacing the client side of the system with commercial off-the-shelf (COTS) general-purpose computing devices running realistic software imitations of the desired behaviors, while retaining the real network and compute hardware at the backend.
+The methodology involves replacing the client side of the system with \gls{COTS}, general-purpose computing devices running realistic software imitations of the desired behaviors, while retaining the real network and compute hardware at the backend.
 This entails modeling and emulating human inputs in \gls{MAR} applications and the physical components of \glspl{CPS}.
 
-We study and validate our methodology by example, investigating its applicability in two representative use-cases for \gls{XR} applications and \glspl{CPS}, respectively: \acf{WCA} and \acfp{NCS}.
-\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} discuss initial explorative studies into the application of our methodology for \gls{WCA}.
-These works present the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research, and further discuss the necessary tooling and software frameworks required for the implementation of our methodology.
-In particular, we introduce in these works a trace-based emulation tool which we call EdgeDroid.
+I study and validate this methodology by example, investigating its applicability in two representative use-cases for \gls{XR} applications and \glspl{CPS}, respectively: \acf{WCA} and \acfp{NCS}.
+\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} discuss initial explorative studies into the application of the methodology for \gls{WCA}.
+These works present the first empirical results obtained with my methodological approach, providing an initial validation of its utility for research.
+Further, I discuss the necessary tooling and software frameworks required for the implementation of my methodology.
+In particular, my co-authors and I introduce in these works a trace-based emulation tool which we call EdgeDroid.
 It uses a trace of inputs recorded in optimal conditions, the replay of which is controlled at run-time by a simple model of human behavior based on a \gls{FSM}.
 
-\Cref{paper:olguinmunoz2022cleave} extends our methodology to the study of control systems for \glspl{NCS} on Edge Computing.
+\Cref{paper:olguinmunoz2022cleave} extends my methodology to the study of control systems for \glspl{NCS} on Edge Computing.
 These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, we introduce yet another tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+To address these challenges, my co-authors and I introduce yet another tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
 \gls{CLEAVE} allows us both to emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
@@
 
-In \cref{paper:olguinmunoz2022ainur} we also briefly discuss Edge Computing testbed research and automation as an ancillary contribution to the methodology.
+In \cref{paper:olguinmunoz2022ainur}, we also briefly discuss Edge Computing testbed research and automation as an ancillary contribution to the methodology.
 This paper formally introduces the Ainur framework for Edge Computing testbed automation and workload orchestration.
@@
 
-We argue that our approach presents several advantages over existing methods for the study of trade-offs in latency-sensitive applications.
-Our methodology allows for efficient and repeatable evaluation of Edge environments while avoiding many of the complications associated with experimental evaluation.
+I argue that my approach presents several advantages over existing methods for the study of trade-offs in latency-sensitive applications.
+The methodology allows for efficient and repeatable evaluation of Edge environments while avoiding many of the complications associated with experimental evaluation.
 Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
 This also reduces the barrier of entry to  research, as \glspl{SBC} are for the most part cheap and easily accessible, and further preserves the realism of effects stemming from the hardware and network.
-In particular, the methodology allows us to capture effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
+In particular, the methodology allows for the capture of effects due to network factors such as contention, congestion control, and medium access, which are often of stochastic or chaotic natures and complex to capture in simulations.
 The softwarized nature of the client-side emulation allows for straightforward automation through general-purpose programming and scripting languages.
@@
 The second key contribution of this dissertation relates to the exploration of the question of what degree of detail about the physical world must be captured by a client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
-We explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which we then combine with mathematical framework to study the potential for optimization of these applications.
+I explore this interrogative through the elaboration of a detailed model of human user behavior in \gls{WCA}, which I then combine with mathematical framework to study the potential for optimization of these applications.
 
 \cref{paper:olguinmunoz2021impact} begins this work by presenting a characterization of human behavior in \gls{WCA}, developed through a human subject study in which participants interact with an instrumented application.
-System responsiveness was altered in real-time during each execution of the task, and we recorded participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
-Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.
+System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
+Participants are also asked to fill out two personality indicator questionnaires, allowing to later correlate individual personality traits and measured reaction to changes in system responsiveness.
 
-Next, in \cref{paper:olguinmunoz2023realistic} we introduce the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
+Next, in \cref{paper:olguinmunoz2023realistic} my co-authors and I introduce the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
 The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
@@
 
-We conclude that realistic, dynamic human behavior in the client-side emulation enables superior optimization approaches that would otherwise be inviable to implement.
+I conclude that realistic, dynamic human behavior in the client-side emulation enables superior optimization approaches that would otherwise be inviable to implement.
 The findings from this research highlight that simply focusing on system-related metrics or performance optimizations may not be sufficient, as human behavior and reactions play a critical role in the overall performance of such applications.
-By considering human behavior as an integral part of the emulation process, we demonstrate that it is possible to develop more realistic and effective optimization approaches that take into account the intricacies of human-computer interaction.
+By considering human behavior as an integral part of the emulation process, I demonstrate that it is possible to develop more realistic and effective optimization approaches that take into account the intricacies of human-computer interaction.
 These findings have significant implications for the design and optimization of \gls{WCA} applications, particularly in the areas of energy consumption, sampling strategies, and application lifetime.

### 2023-05-19 body/part1/chapter2-relatedwork/relatedwork.tex
cleaning up for publication

@@
 
-Nevertheless, a small body of works has attempted to tackle this challenge through analytical approaches.
-Examples include~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}.
+Nevertheless, a small body of works has attempted to tackle this challenge through analytical approaches~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}.
 In~\cite{chen2015efficient}, the authors formulate the problem of computation offloading in \gls{MEC} as a multi-user computation offloading game, and design a distributed computation offloading algorithm that achieves a Nash equilibrium to solve it.
@@
 Approaches to performance evaluation of distributed systems using simulations also exist in literature, particularly geared towards the Cloud~\cite{mansouri2020cloud}.
-Examples of such approaches include the \emph{CloudSim}, \emph{iCanCloud}, and \emph{GreenCloud} simulators.
+Examples include the \emph{CloudSim}, \emph{iCanCloud}, and \emph{GreenCloud} simulators.
 CloudSim~\cite{calheiros2011cloudsim} is a widely used event-based simulation toolkit designed for generic \gls{IaaS} Cloud Computing environments.
@@
 
-\emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the benchmarking and performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
-Video is streamed from a source Android device to a compute node where neural style-transfer~\cite{gatys2016image} is applied to it.
+\emph{OpenRTiST}~\cite{george2020openrtist} is a tool for the performance evaluation of Edge Computing deployments using real-time processing of a live video feed.
+Video is streamed from a mobile device to a compute node where neural style-transfer~\cite{gatys2016image} is applied to it.
 The transformed video is then fed back to the source to be presented, resulting in a compute-intensive, bandwidth-hungry, and latency-sensitive benchmark.
@@
 Several of these are additionally driven by research interests in novel mobile networking technologies and \gls{MEC}, as well as in the virtualization of network functions.
-Of particular interest to us are the
+Of particular interest to this dissertation are the
 \begin{inlineenum}
@@
 In~\cite{yamanaka2021design}, the authors present an experimental Edge Computing testbed capable of determining the offload location of a workload based on desired latency parameters.
-\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power single-board computers such as Raspberry Pis and Jetson Nanos.
+\cite{diao2019scalable} propose yet another Kubernetes-based testbed based for Edge task offloading based on heterogeneous hardware including low-power \glspl{SBC}.
 % \cite{moorthy2022cloudraft} proposes \gls{CLOUDRAFT}, a Cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
@@
 
-Although the aforementioned testbeds represent valuable and powerful tools, the lack of a unified approach to testbed management and automation hinders the development and implementation of next-generation wireless systems, Cloud, and Edge computing paradigms.
+Although the aforementioned testbeds represent valuable and powerful tools, the lack of a unified approach to testbed management and automation hinders the development of next-generation wireless systems, Cloud, and Edge computing paradigms.
 General-purpose, hardware-agnostic software frameworks such as \gls{CLOUDRAFT}~\cite{moorthy2022cloudraft}  can simplify the management and automation of testbeds, promote compatibility with Cloud- and Edge-native standards, and facilitate collaboration among research groups.
@@
 A significant amount of work has been dedicated to the modeling and performance characterization of \glspl{NCS} due to their potential benefits for industrial and commercial settings~\cite{lu2016real,hespanha2007survey,zhang2013network,zhang2016survey}.
-Research in \glspl{NCS} has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
-Network delays (latency), losses, and \emph{jitter} --- referring to the variation in random network delays --- are at the forefront of these challenges.
+Research has mostly focused on addressing the challenges imposed by the introduction of best-effort communication networks into traditional control systems.
+Network delays (latency), \emph{jitter} --- the variation in random network delays ---, and losses are at the forefront of these challenges.
 They have partially been addressed through the clever placement of the controller (e.g.\ by employing Edge Computing)~\cite{sasaki2017layered,sasaki2016vehicle}, as well as through the development of control algorithms capable of accounting for the effects of the network~\cite{zhang2013network}.
@@
 
-We identify two main approaches: fully practical and partially-emulated approaches.
+Two main approaches can be identified: fully practical and partially-emulated approaches.
 In a fully practical approach, the experimental setup uses real systems and hardware to test and validate the performance of the \gls{NCS}.
@@
 
-To the best of our knowledge, there exist currently no generic, hardware-independent framework available for repeatable experimental research in Edge-bound \glspl{NCS}.
+To the best of my knowledge, there exist currently no generic, hardware-independent framework available for repeatable experimental research in Edge-bound \glspl{NCS}.
 While there has been some notable work in this area, such as the work by~\cite{zoppi2020ncsbench}, existing contributions still mostly rely on proprietary and hardware that is difficult to acquire.
 This limitation restricts the accessibility of the framework to researchers and practitioners, hindering the reproducibility and comparability of experimental results.
-Moreover, the existing research's reliance on hardware also poses scalability limitations.
+Moreover, existing research's reliance on hardware poses scalability limitations.
 As the field of Edge Computing-enabled \glspl{NCS} continues to evolve, there is a growing demand for scalable solutions that can be easily adapted to different hardware configurations and setups.

### 2023-05-19 body/part1/chapter2-relatedwork/scope.tex
cleaning up for publication

--- 2023-04-17 body/part1/chapter2-relatedwork/scope.tex
@@
 
+\input{tables/scope}
+
 Before these systems can become widely deployed, however, their real-world performance must be thoroughly understood.
-\glspl{CPS} often operate in safety-critical environments, and poor \gls{QoS} in \gls{XR} can easily lead to users experiencing discomfort while using the application.
+\glspl{CPS} often operate in safety-critical environments, and poor \gls{QoS} in \gls{XR} can easily lead to users experiencing discomfort.
 Accurate evaluation, characterization, and optimization of the performance of these systems is therefore critical to ensure reliability, safety, and user satisfaction.
@@
 
-Evaluating the performance of Edge environments involving such applications is challenging, due to complex behaviors emerging from the interaction between computation and network in these systems.
+Evaluating the performance of Edge environments involving such applications is challenging, due to complex behaviors emerging from the interaction between computation and network.
 Existing literature has attempted to tackle the performance evaluation of Edge systems through analytical, simulation, and experimental approaches.
 However, the nature of Edge Computing environments, which involves tight integration between network and compute, requires vast and often unfeasible interdisciplinary competence.
-Existing works have been constrained to narrow their scope by prioritizing the comprehensive modeling of either compute or network, resulting in a potential trade-off that may neglect the other aspect.
-As we will see in \cref{sec:state-of-the-art}, this has led to an incomplete understanding of the overall performance of Edge systems and applications in literature.
+Existing works have been constrained to narrow their scope by prioritizing the comprehensive modeling of either compute or network, resulting in a trade-off that may neglect the other aspect.
+As I will discuss in \cref{sec:state-of-the-art}, this has led to an incomplete understanding of the overall performance of Edge systems and applications in literature.
 
-In this dissertation we introduce and study the applicability of an alternative methodology for Edge performance evaluation based on a practical, experimental approach.
-We aim to complement literature by providing a method which bridges the gap in knowledge left by existing approaches.
-Or work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
-We also focus exclusively on Cloudlet Computing-like setups.
-Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, we consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
-Likewise, we do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
-Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate our results and conclusions in the context of \gls{MEC} and Fog Computing.
+In this dissertation I introduce and study the applicability of an alternative methodology for Edge performance evaluation based on a practical, experimental approach.
+I aim to complement literature by providing a method which bridges the gap in knowledge left by existing approaches.
+My work centers of Edge Computing environments involving latency-sensitive applications such as \gls{XR} and \glspl{CPS}, as these are broadly understood to represent ``killer'' use-cases for the Edge.
+I also focus exclusively on Cloudlet Computing-like setups (see \cref{tab:scope}).
+Although \gls{MEC} is one of the main enablers for \gls{WCA} applications, I consider such network deployments to be \gls{OOS} due to the complexity and financial investment required to deploy a mobile network.
+Likewise, I do not target Fog Computing deployments due to this paradigm being a poor fit for latency-sensitive applications.
+Thus, although the methodology and tools introduced are architecture-agnostic and should be directly applicable to other kinds of Edge Computing setups than the one used in this dissertation, further work may be needed to validate my results and conclusions in the context of \gls{MEC} and Fog Computing.
 
@@
 Attempting to cover both these scopes comprehensively in a single dissertation would be prohibitively challenging.
-Therefore, in this work, we approach these challenges by focusing on representative use-cases for these application classes.
+Therefore, in this work, I approach these challenges by focusing on representative use-cases for these application classes.
 
-Specifically, we employ \emph{step-based task-assistance} \gls{WCA} as a representative use case for \gls{XR}, and \glspl{NCS} for \glspl{CPS}.
-We focus on these categories for a couple of reasons.
+Specifically, I employ \emph{step-based task-assistance} \gls{WCA} as a representative use case for \gls{XR}, and \glspl{NCS} for \glspl{CPS}.
+I focus on these categories for a couple of reasons.
 In the case of \gls{WCA}, step-based task-assistance was among the first categories of \gls{WCA} to be developed~\cite{chen2015early}, and thus represents a more mature application class compared to other \glspl{WCA}.
@@
 
-For \glspl{NCS}, we limit our exploration in this dissertation to simple, small-scale, single-plant systems.
+For \glspl{NCS}, I limit my exploration in this dissertation to simple, small-scale, single-plant systems.
 The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
-We have chosen this system for its relative simplicity as well as its broad use in the field of automatic control as one of the fundamental examples of linear control.
-Through these examples, we aim to gain insights into the performance evaluation challenges and opportunities offered by Edge Computing these applications and systems, and contribute to the understanding of the broader fields of \gls{XR} and \gls{CPS} performance evaluation in distributed computing environments.
+I have chosen this system for its relative simplicity as well as its broad use in the field of automatic control as one of the fundamental examples of linear control.
+Through these examples, I aim to gain insights into the performance evaluation challenges and opportunities offered by Edge Computing these applications and systems, and contribute to the understanding of the broader fields of \gls{XR} and \gls{CPS} performance evaluation in distributed computing environments.
 
@@
 
-In the following  \cref{sec:state-of-the-art}, we will review the state of the art in the context of performance evaluation of Edge Computing environments, including analytical, simulation, and experimental approaches.
-We will evidence how existing approaches fail to fully capture the intricacies of performance evaluation of latency-sensitive, feedback-loop applications on Edge Computing, and make the case for the need for an alternative approach.
-We will further discuss existing experimental approaches to \gls{MAR} and \gls{NCS} research, and discuss their limitations.
+In \cref{sec:state-of-the-art}, I will review the state of the art in the context of performance evaluation of Edge Computing environments, including analytical, simulation, and experimental approaches.
+I will evidence how existing works fail to fully capture the intricacies of performance evaluation of latency-sensitive, feedback-loop applications on Edge Computing, and make the case for the need for an alternative approach.
+I will further discuss existing experimental approaches to \gls{MAR} and \gls{NCS} research, and discuss their limitations.
 
-Finally, we conclude this chapter in \cref{sec:contributions}, where we will discuss in details the contributions of this dissertation.
+Finally, I conclude this chapter in \cref{sec:contributions}, where I will discuss in details the contributions of this dissertation.
 This includes a discussion of the key sub-contributions of each included publication.

### 2023-05-19 body/part1/chapter2-relatedwork/tables/scope.tex
cleaning up for publication

@@
     \caption{Visualization of the scope of this dissertation with respect to performance evaluation.
-    We focus exclusively on Edge Computing based on Cloudlets~\cite{satyanarayanan2009case} and on experimental performance evaluation of \acs{XR} and \acsp{CPS}.}
+    I focus exclusively on Edge Computing based on Cloudlets~\cite{satyanarayanan2009case} and on experimental performance evaluation of \acs{XR} and \acsp{CPS}.}
     \label{tab:scope}

### 2023-05-19 body/part1/chapter3-contributions/methodology.tex
my

@@
 
-In \cref{summary:methodology}, we discuss our methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
+In \cref{summary:methodology}, I discuss my methodology for evaluating the performance of latency-sensitive, feedback-loop-based applications enabled by Edge Computing.
 This methodology involves emulating only the end-user, client-side component of such applications while using the real network and compute components of the system. 
-We further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.
+I further discuss the validation of this approach through two representative use-cases for \gls{XR} and \glspl{CPS}.
 
-Next, in \cref{summary:realism}, we explore the extent of the need for accuracy in our client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
-We broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which we then employ in our methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.
+Next, in \cref{summary:realism}, I explore the extent of the need for accuracy in my client-side emulation of a latency-sensitive, feedback-loop application on Edge Computing.
+I broach this question through the elaboration of a detailed model of human behavior in \gls{WCA}, which I then employ in the methodology, in conjunction with a mathematical framework, to study the the potential for optimization of \gls{WCA} applications.
 
@@
 In particular, the proposed methodology aims to provide a more comprehensive and realistic assessment of the performance of these latency-sensitive applications with respect to previous approaches, allowing for a better understanding of the strengths and limitations of Edge Computing.
-Our results will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing systems.
+The results presented in this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing systems.
 
@@
 
-Our methodology is based on the emulation of target workloads on actual Edge infrastructure;
+This methodology is based on the emulation of target workloads on actual Edge infrastructure;
 see \cref{fig:methodology}.
-We replace the client side of the system with a realistic emulation of the desired behaviors;
+I replace the client side of the system with a realistic emulation of the desired behaviors;
 this emulation is implemented in software deployed on \gls{COTS} general-purpose computing devices.
-In our initial implementation, these correspond to low-cost, easily replaceable and scalable Raspberry Pi 4 Model B~\cite{raspberrypi} \glspl{SBC}.
-We aim to maintain the backend software and hardware, as well as the network, unchanged with respect to the real-world deployment, in order to preserve as much realism as possible in the effects arising from the hardware and network.
+In my initial implementation, these correspond to low-cost, easily replaceable and scalable Raspberry Pi 4 Model B~\cite{raspberrypi} \glspl{SBC}.
+I aim to maintain the backend software and hardware, as well as the network, unchanged with respect to the real-world deployment, in order to preserve as much realism as possible in the effects arising from the hardware and network.
 
-Our emulated methodology presents several advantages over alternatives, simulated or fully-practical approaches.
-To begin with, and perhaps most importantly, it makes our approach portable and applicable to real Edge computing deployments.
-By limiting the emulation to the workload, our approach can be directly employed on pre-existing Edge computing infrastructure in order to benchmark applications before they are deployed.
-This makes our approach valuable to not only to the research community, but also to system designers and application developers.
+My emulated methodology presents several advantages over alternatives, simulated or fully-practical approaches.
+To begin with, and perhaps most importantly, it makes my approach portable and applicable to real Edge computing deployments.
+By limiting the emulation to the workload, my approach can be directly employed on pre-existing Edge computing infrastructure in order to benchmark applications before they are deployed.
+This makes my approach valuable to not only to the research community, but also to system designers and application developers.
 
 In hand with the previous point, it allows for easier scaling compared to real-world studies.
-The systems we target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
+The systems I target exhibit a centralized nature in which a potentially large number of clients offload computation to a single central compute node.
 The client-side component of the applications of interest for this research is often complex to scale.
@@
 
-Our approach circumvents both of these limitations.
+My approach circumvents both of these limitations.
 Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as the aforementioned \glspl{SBC}.
 Furthermore, it preserves the realism of effects stemming from the hardware and network.
-Employing emulations allows our methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
-Of particular interest to us are effects due to network factors such as contention, congestion control, and medium access.
+Employing emulations allows my methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
+Of particular interest to this dissertation are effects due to network factors such as contention, congestion control, and medium access.
 These effects are often of stochastic or chaotic natures, and are thus complex to capture in simulations.
@@
 
-Finally, our methodology improves extensibility significantly.
+Finally, my methodology improves extensibility significantly.
 Replicating the methodology for other workloads simply requires developing a new piece of software to emulate the new workload.
@@
 \medskip
-In the following subsections we will discuss two illustrative use-cases for our methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
-\Cref{summary:methodology:usecase_wca} discusses the application of our methodology to \gls{WCA}.
+In the following subsections I will discuss two illustrative use-cases for my methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
+\Cref{summary:methodology:usecase_wca} discusses the application of my methodology to \gls{WCA}.
 This work was originally published in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
-Next, in \cref{summary:methodology:usecase_ncs} we expand this discussion to \glspl{NCS}, a discussion originally approached in \cref{paper:olguinmunoz2022cleave}.
+Next, in \cref{summary:methodology:usecase_ncs} I expand this discussion to \glspl{NCS}, a discussion originally approached in \cref{paper:olguinmunoz2022cleave}.
 
@@
 
-We first introduce our methodology and its applicability to \gls{WCA} in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
-The former corresponds to an extended abstract paper which discusses a high level overview of our approach;
+I first introduce my methodology and its applicability to \gls{WCA} in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
+The former corresponds to an extended abstract paper which discusses a high level overview of my approach;
 the latter presents a deeper, more complete discussion about the implementation together with some first experimental results.
@@
 
-The implementation of the client-side emulation necessary for our methodology follows in these works a trace-based design.
+The implementation of the client-side emulation necessary for my methodology follows in these works a trace-based design.
 The trace-based approach is chosen for three reasons:
@@
     \caption{%
-        High level conceptual design of our methodology for \gls{WCA}.
-        We replace the human by a trace-driven emulation.
-        This allows us to maintain realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
+        High level conceptual design of my methodology for \gls{WCA}.
+        I replace the human by a trace-driven emulation.
+        This allows for the retention of realism in inputs sent over the network to the compute process on the backend, while simplifying the design of the client software.
 %        Figure originally published in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}.
@@
 
-We introduce a tool, \emph{EdgeDroid \num{1.0}}, which implements the concepts outlined above.
+I introduce a tool, \emph{EdgeDroid \num{1.0}}, which implements the concepts outlined above.
 The tool is implemented in Android, allowing for its easy deployment on the same kind of \gls{COTS} mobile and/or wearable devices (e.g.\ Google Glass~\cite{googleglass}) on which real \gls{WCA} applications are intended to run in the future, and is instrumented for easy collection of a multitude of system-level metrics at runtime which can then be analyzed.
-We use a trace corresponding to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer;
+I use a trace corresponding to a pre-recorded sequence of sensory inputs obtained from a real execution of the \gls{WCA} by a human volunteer;
 the trace is recorded in a near-ideal setting, such that it does not include either human mistakes or segments with degraded system responsiveness.
-Finally, we employ in this work a preliminary model of human behavior which is totally impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+Finally, I employ in this work a preliminary model of human behavior which is totally impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
 This leads to a model of an ``ideal'' user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
@@
 \medskip
-We proceed to demonstrate the practical utility of our methodology as applied to \gls{WCA} by setting up a series of scenarios.
-We deployed a \emph{Gabriel}-powered~\cite{chen2018application} \gls{WCA} application on a small testbed consisting of
+I proceed to demonstrate the practical utility of my methodology as applied to \gls{WCA} by setting up a series of scenarios.
+I deployed a \emph{Gabriel}-powered~\cite{chen2018application} \gls{WCA} application on a small testbed consisting of
 \begin{inlineenum}
@@
 
-These setups already showcase the flexibility of our approach.
-Our trace-based implementation of the methodology allows for detailed measurements of system performance at different levels, from individual components to specific task steps.
+These setups already showcase the flexibility of my approach.
+My trace-based implementation of the methodology allows for detailed measurements of system performance at different levels, from individual components to specific task steps.
 This provides valuable insights into the bottlenecks and areas for optimization in the system.
-We find, for example, that \glspl{RTT} for feedback-rich inputs are up to an order of magnitude greater than feedback-less inputs\footnote{%
+I find, for example, that \glspl{RTT} for feedback-rich inputs are up to an order of magnitude greater than feedback-less inputs\footnote{%
     In this context, \emph{feedback-rich} inputs refer to inputs which trigger a transition in step in the internal task model of the application and thus cause the generation of feedback which is then provided to the user.
@@
 
-We discuss an initial use-case study of our methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
-In this work, we present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of Edge-native \aclp{NCS}.
+I discuss an initial use-case study of my methodology in the context of \glspl{NCS} and similar \glspl{CPS} in \cref{paper:olguinmunoz2022cleave}.
+In this work, I present \gls{CLEAVE}, a software-based framework for scalable and repeatable benchmarking of Edge-native \aclp{NCS}.
 The growing adoption of Edge computing has made it important to understand the strengths and weaknesses of different Edge concepts for the deployment of such \glspl{CPS}.
 
-\gls{CLEAVE} is a fully virtualized benchmarking framework, inspired by our previous work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} on benchmarking human-in-the-loop applications on the Edge.
-In the case of \gls{CLEAVE} however, we forgo the trace-based approach of EdgeDroid in favor of direct modeling of the target systems.
+\gls{CLEAVE} is a fully virtualized benchmarking framework, inspired by my previous work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} on benchmarking human-in-the-loop applications on the Edge.
+In the case of \gls{CLEAVE} however, I forgo the trace-based approach of EdgeDroid in favor of direct modeling of the target systems.
 \gls{CLEAVE} includes a benchmarking framework and software development kit for the development of emulated physical systems and softwarized controllers.
@@
 \medskip
-We study the validity of our methodology and the implementation of the software framework through a simple scenario, illustrated in \cref{fig:cleavesetup}.
+I study the validity of my methodology and the implementation of the software framework through a simple scenario, illustrated in \cref{fig:cleavesetup}.
 An inverted pendulum plant controlled by a proportional-differential controller is deployed on an Edge server over a Wi-Fi link, while video-streaming applications run concurrently on the setup.
@@
 
-We first establish a baseline by deploying a single control loop.
+I first establish a baseline by deploying a single control loop.
 This setup is repeated multiple times with different sampling rates to obtain statistically significant results.
@@
 
-Next, we study the interaction between the control systems and the video-streaming applications.
+Next, I study the interaction between the control systems and the video-streaming applications.
 Six control loops are deployed, as uplink \gls{UDP} traffic from four clients is generated to emulate the load generated by Full-HD video streaming.
-Based on the baseline results, we execute scenarios with varying plant sampling rates and repeat each scenario \num{30} times.
+Based on the baseline results, I execute scenarios with varying plant sampling rates and repeat each scenario \num{30} times.
 Representative results from this experimentation are showcased in \cref{fig:cleaveresults}.
@@
 As sampling rates increase, a critical point is eventually reached at which the congestion on the network link is such that it severely impacts the quality of control.
-We conclude that adaptive sampling might be thus a viable method for optimizing resource utilization in these systems.
+I conclude that adaptive sampling might be thus a viable method for optimizing resource utilization in these systems.
 
@@
 The results of the experiments with video-streaming applications will add to the understanding of the interaction between control systems and networked applications.
-Furthermore, the success of our experimentation shows that our methodology, implemented through \gls{CLEAVE}, makes the repeated experimentation and data collection convenient compared to fully physical setups, while at the same time providing enhanced realism with respect to simulated approaches.
+Furthermore, the success of my experimentation shows that my methodology, implemented through \gls{CLEAVE}, makes the repeated experimentation and data collection convenient compared to fully physical setups, while at the same time providing enhanced realism with respect to simulated approaches.
 
@@
 
-After our initial implementation and experimentation with our methodology for \gls{WCA} applications, we arrived at the realization that the creation of a dedicated testbed for experimental Edge Computing research would greatly augment the scope and impact of our work.
+After my initial implementation and experimentation with my methodology for \gls{WCA} applications, I arrived at the realization that the creation of a dedicated testbed for experimental Edge Computing research would greatly augment the scope and impact of my work.
 Establishing such testbed would provide a dedicated and controlled environment that can accelerate future research by eliminating the need for ad-hoc implementations for each new study.
@@
 This would enable more efficient and accurate assessments of the methodology's performance and facilitate the development of new methods and techniques.
-Additionally, by establishing a testbed, we could provide a platform for other researchers within our group to conduct their research and experiments.
+Additionally, by establishing a testbed, I could provide a platform for other researchers within my group to conduct their research and experiments.
 
@@
 It consists of a cluster of hardware-reconfigurable \gls{COTS} computing nodes interconnected using managed switches and \glspl{SDR}.
-This allows us to quickly, on-the-fly, and in an automated fashion change the characteristics of the cluster and the network, in order to study different Edge- and Cloud-computing deployments and the applications that run on them.
+This allows for quick, on-the-fly, and automated changes to the characteristics of the cluster and the network, in order to study different Edge- and Cloud-computing deployments and the applications that run on them.
 
@@
 \emph{Workload hosts} act as client-side user devices;
-in our prototype, these corresponded to \num{10} Raspberry Pi 4 Model B boards.
+in my prototype, these corresponded to \num{10} Raspberry Pi 4 Model B boards.
 These devices connect to either a cloudlet or cloud computing instances over the \emph{workload data network}.
 Through clever configuration of \acsp{VLAN} and routing, traffic on this network can be forced to flow through the attached \acsp{SDR} and their associated radio processing hosts.
-This allows us to seamlessly switch out the physical layer for experiments.
-We implemented configurations wherein one of the \glspl{SDR} acted as a 4G or 5G base station and the other one as a cellular client.
+This allows for seamless switching of the physical layer for experiments.
+I implemented configurations wherein one of the \glspl{SDR} acted as a 4G or 5G base station and the other one as a cellular client.
 An alternative setup had the workload hosts connect wirelessly via Wi-Fi to one of the radios acting as an \gls{AP}
@@
 
-We also introduce a software framework we developed for its automation and experiment orchestration, \emph{Ainur}.
+I also introduce a software framework I developed for its automation and experiment orchestration, \emph{Ainur}.
 Ainur is a highly flexible end-to-end wireless network testbed management framework.
@@
 
-Finally, we note that both Ainur and \gls{EXPECA} were utilized for experimentation in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
+Finally, I note that both Ainur and \gls{EXPECA} were utilized for experimentation in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
 The use of these tools was, however, during their development and prototyping, and they have thus not been referred to by name in these works.

### 2023-05-19 body/part1/chapter3-contributions/methodology.tex
reformulating

@@
 My approach circumvents both of these limitations.
-Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as the aforementioned \glspl{SBC}.
+Emulating the workload component reduces complexity by moving it into the software domain, allowing for easier scaling through the use of cheap, \gls{COTS} general-purpose hardware such as \glspl{SBC}.
 Furthermore, it preserves the realism of effects stemming from the hardware and network.
-Employing emulations allows my methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
+Leveraging emulation allows my methodology to realize a more realistic testing environment than a simulated approach and can provide results that are closer to what would be seen in the real-world.
 Of particular interest to this dissertation are effects due to network factors such as contention, congestion control, and medium access.
@@
 
-Emulating the workload also provides improved repeatability and replicability.
+Emulating the workload also improves repeatability and replicability.
 Repeating a study becomes a matter of re-running the workload on the same testbed;
@@
 \medskip
-In the following subsections I will discuss two illustrative use-cases for my methodology; \acl{WCA} and \acl{NCS} research and benchmarking.
+In the following subsections I will discuss two illustrative use-cases for my methodology; \glsfirst{WCA} and \glsfirst{NCS} research and benchmarking.
 \Cref{summary:methodology:usecase_wca} discusses the application of my methodology to \gls{WCA}.
@@
 
-\Cref{fig:methodology:wca:conceptual} presents a conceptual diagram of this approach.
-The trace is fed to a model of human behavior, which then replays it to the real \gls{WCA} backend.
+\Cref{fig:methodology:wca:conceptual} illustrates this approach.
+The trace is replayed to the real \gls{WCA} backend through a model of human behavior.
 Feedback from the backend is then interpreted by the model to adapt the runtime replay of the trace.
@@
 \medskip
-I proceed to demonstrate the practical utility of my methodology as applied to \gls{WCA} by setting up a series of scenarios.
-I deployed a \emph{Gabriel}-powered~\cite{chen2018application} \gls{WCA} application on a small testbed consisting of
+The practical utility of my methodology as applied to \gls{WCA} is demonstrated through a series of scenarios.
+I deploy a \emph{Gabriel}-powered~\cite{chen2018application} \gls{WCA} application on a small testbed consisting of
 \begin{inlineenum}
@@
 \end{inlineenum}.
-Three scenarios were considered, corresponding to setups under ideal network conditions with \num{1} and \num{10} clients, plus an additional scenario with \num{10} clients located at a distance from the \gls{AP} such that the signal strength was significantly degraded.
-Each scenario was repeated a \num{100} times for statistical validity.
+Three scenarios are considered, corresponding to setups under ideal network conditions with \num{1} and \num{10} clients, plus an additional scenario with \num{10} clients located at a distance from the \gls{AP} such that the signal strength was significantly degraded.
+Each scenario is repeated \num{100} times for statistical validity.
 
@@
         \caption{
-            Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.
+            Comparison of \acsp{RTT} for inputs that trigger a state transition in the task model versus inputs that do not.
         }\label{fig:wca:rtt}%
@@
         \caption{%
-            Distribution of latency across system components for inputs that triggered a state transition in the task model.
+            Distribution of latency across system components for inputs that trigger a state transition in the task model.
         }\label{fig:wca:latency}
@@
 
-These setups already showcase the flexibility of my approach.
+These setups showcase the flexibility of the approach.
 My trace-based implementation of the methodology allows for detailed measurements of system performance at different levels, from individual components to specific task steps.
@@
 }; see \cref{fig:wca:rtt}.
-The flexibility of the methodology also allows researchers to zoom into specific components or steps of the system to gain deeper insights, which can inform the design and development of future systems.
-For example, the results in \cref{fig:wca:latency} suggest that improving the quality of the wireless link should be a priority for scaling the application.
+The flexibility of the methodology also allows researchers to zoom into specific components of the system to gain deeper insights, which can inform the design and development of future systems.
+The results in \cref{fig:wca:latency} suggest that improving the quality of the wireless link should be a priority for scaling the application.
 The system scales linearly with respect to the number of clients, but the impaired Wi-Fi greatly affects the performance of feedback-rich inputs, with delays sometimes exceeding the threshold for acceptable user experience.
@@
 
-\gls{CLEAVE} is built using Python 3.8, making it highly extensible and able to use existing libraries.
+\gls{CLEAVE} is built using Python, making it highly extensible and able to use existing libraries.
 It is compatible with container technologies such as Docker~\cite{docker}, making it suitable for automated deployment, scaling, and benchmarking on industry-standard Edge setups using container orchestration solutions.
@@
 
-After my initial implementation and experimentation with my methodology for \gls{WCA} applications, I arrived at the realization that the creation of a dedicated testbed for experimental Edge Computing research would greatly augment the scope and impact of my work.
+After my initial implementation and experimentation with my methodology for \gls{WCA} applications, my research group and I arrived at the realization that the creation of a dedicated testbed for experimental Edge Computing research would greatly augment the scope and impact of my work.
 Establishing such testbed would provide a dedicated and controlled environment that can accelerate future research by eliminating the need for ad-hoc implementations for each new study.
@@
 This would enable more efficient and accurate assessments of the methodology's performance and facilitate the development of new methods and techniques.
-Additionally, by establishing a testbed, I could provide a platform for other researchers within my group to conduct their research and experiments.
+Additionally, by establishing a testbed, we could provide a platform for other researchers within my group to conduct their research and experiments.
 
-These reasons led to the establishment of the \gls{EXPECA} testbed at \pgls{KTH}.
+These reasons were part of the motivation for the establishment of the \gls{EXPECA} testbed at \pgls{KTH}.
 \gls{EXPECA} is a \gls{SSF}-funded infrastructure project targeting the development and provisioning of an Edge computing infrastructure for research into novel applications and network architectures.
@@
 
-The initial prototype architecture of \gls{EXPECA} is illustrated in \cref{fig:expeca}.
+The initial prototype of \gls{EXPECA} is illustrated in \cref{fig:expeca}.
 \emph{Workload hosts} act as client-side user devices;
-in my prototype, these corresponded to \num{10} Raspberry Pi 4 Model B boards.
+in our prototype, these correspond to \num{10} Raspberry Pi 4 Model B boards.
 These devices connect to either a cloudlet or cloud computing instances over the \emph{workload data network}.
@@
 This allows for seamless switching of the physical layer for experiments.
-I implemented configurations wherein one of the \glspl{SDR} acted as a 4G or 5G base station and the other one as a cellular client.
-An alternative setup had the workload hosts connect wirelessly via Wi-Fi to one of the radios acting as an \gls{AP}
+We implement configurations wherein one of the \glspl{SDR} act as a 4G or 5G base station and the other one as a cellular client.
+An alternative setup has the workload hosts connect wirelessly via Wi-Fi to one of the radios acting as an \gls{AP}
 Finally, the \emph{control network} is used for the out-of-band configuration of testbed resources, and the \acs{VPN} links allow direct tunneled connections to cloud compute nodes on \acs{AWS}.
 
-I also introduce a software framework I developed for its automation and experiment orchestration, \emph{Ainur}.
+In \cref{paper:olguinmunoz2022ainur}, my co-authors and I introduce \emph{Ainur}, a software framework developed for the automation and experiment orchestration of the \gls{EXPECA} testbed prototype.
 Ainur is a highly flexible end-to-end wireless network testbed management framework.
@@
 
-The architecture of Ainur closely resembles the conceptual layers of the \acs{TCP}/\acs{IP} stack, and the framework utilizes containerization throughout.
+The architecture of Ainur resembles the conceptual layers of the \acs{TCP}/\acs{IP} stack, and the framework employs containerization throughout.
 This includes the deployment and orchestration of workloads, support for different physical layer configurations, and automated collection of logs.
@@
 
-Finally, I note that both Ainur and \gls{EXPECA} were utilized for experimentation in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
+Note that both Ainur and \gls{EXPECA} were utilized for experimentation in \cref{paper:olguinmunoz2022cleave,paper:olguinmunoz2023realistic}.
 The use of these tools was, however, during their development and prototyping, and they have thus not been referred to by name in these works.

### 2023-05-19 body/part1/chapter3-contributions/wca-implications.tex
reformulating

@@
 
-In this section, we aim to delve deeper into the implications of our methodology for the optimization of \gls{WCA}.
-In particular, we wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
-We approach this question by developing an improved, more detailed version of our model of human behavior in \gls{WCA} used in the EdgeDroid tool, which we then incorporate into our methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.
+In this section, I aim to delve deeper into the implications of the methodology for the optimization of \gls{WCA}.
+In particular, I wish to explore the impact improving the accuracy of the client-side emulation has on the potential for optimization of the applications under evaluation.
+I approach this question by developing an improved, more detailed version of the model of human behavior in \gls{WCA} used in the EdgeDroid tool, which I then incorporate into my methodology along with a mathematical framework to study the potential for optimizing \gls{WCA} applications.
 
-Our work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
+The work in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} has used a model that emulates a rigid and unyielding human who makes no mistakes and is not affected by low responsiveness in the system.
 However, this is not an accurate representation of real human behavior, and a more realistic human model must take into consideration the perceived feedback latency and its effects on human reactions.
-In order to develop a more accurate human model in the context of \gls{WCA}, we first conduct an experimental study that measures human performance in different scenarios, taking into account the impact of system latency on the behavior of humans.
+In order to develop a more accurate human model in the context of \gls{WCA}, I first conduct an experimental study that measures human performance in different scenarios, taking into account the impact of system latency on the behavior of humans.
 The results of this study serve as the foundation for building a more representative model of human behavior.
 
-Once we have established a more accurate model of human behavior, we can explore its implications for the optimization of \gls{WCA} through our methodology.
-By improving the accuracy of our model of human behavior, we aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how our methodology can be effectively applied to address these.
-This research highlights the importance of the accuracy of emulations as  crucial factor in the application of our methodology.
+Once I have established a more accurate model of human behavior, I can explore its implications for the optimization of \gls{WCA} through my methodology.
+By improving the accuracy of the model of human behavior, I aim to gain novel understanding of the challenges involved in optimizing \gls{WCA} and how the methodology can be effectively applied to address these.
+This research highlights the importance of the accuracy of emulations as  crucial factor in the application of my methodology.
 
@@
 
-In order to more easily and accurately discuss the implications of our methodology for \gls{WCA}, we first characterize these applications and provide some definitions relating to their operation.
-We focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistance applications which have as their goal the guidance of a user through a task, using a series of instructions.
+In order to more easily and accurately discuss the implications of my methodology for \gls{WCA}, I first characterize these applications and provide some definitions relating to their operation.
+I focus in particular on \emph{step-based} \gls{WCA}, i.e.\ assistive applications which have as their goal the guidance of a user through a task, using a series of instructions.
 
@@
 In step-based \glspl{WCA}, the state in question refers to the progress of some arbitrary, sequential task, such as the assembly of a LEGO model or of a piece of IKEA furniture~\cite{chen2018application}.
-We will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
+I will formally define \emph{task} as a well-defined sequence of instructions to be performed in order to achieve a final goal state, and further define a \emph{step} within a task as a specific action to be performed by the user, described by a single instruction.
 A step begins when the corresponding instruction is provided to the user, and ends when the instruction for the next step is provided;
@@
 At this point, the application transitions to the next step \ensuremath{M + 1}, and thus \ensuremath{t_{n + 1}} is the instant at which the next instruction is provided and step \ensuremath{M + 1} begins.
-We will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
+I will also refer to the interval \ensuremath{t_{n + 1} - t_n} as the \emph{last sample \gls{RTT}}.
 
 \glsreset{TTF}%
-From this characterization, we can define key variables relating to human behavior with respect to the sampling behavior of these applications.
-However, we first note that the combination of the characteristic of seamless interaction with the context of the user described in \cref{sec:background:wca} and in \cref{fig:wca:operation}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
+From this characterization, key variables relating to human behavior with respect to the sampling behavior of these applications can be defined.
+However, it must first be noted that the combination of the characteristic of seamless interaction with the context of the user described in \cref{sec:background:wca} and in \cref{fig:wca:operation}, combined with the sampling behavior described above, lead to an interesting consequence for the perception of responsiveness in \gls{WCA}.
 The only points in time at which a user can consciously notice and interact with the application is whenever they are \emph{expecting} feedback.
@@
     \item \ensuremath{t_n - t_c} is the \emph{wait time}, an interval of time during which the task has been finished, but no sample capturing the new state has been taken yet
-    \item finally, the interval \ensuremath{t_{n + 1} - t_c} we call \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
+    \item the interval \ensuremath{t_{n + 1} - t_c} I will term \emph{\gls{TTF}}, and refers to the time the user spends waiting for a new instruction.
 \end{inlineenum}
 These values are key to understanding human perception of responsiveness in these systems, and therefore also to the optimization of resource consumption in \gls{WCA}.
-We will refer to them extensively in \cref{chap:contributions}.
+I will refer to them extensively in \cref{ssec:characterizing} onwards.
 
-%\subsection{Mechanisms relating human behavior to system responsiveness}
-%
-%The question of how people respond to delay in a computer system is grounded in how people perceive time.
-%Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
-%More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}.
-%The estimation of an event's duration varies with context of various types
-%\begin{inlineenum}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
-%\item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
-%\item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
-%\item arousal can expand durations~\cite{droit_volet2011emotion}.
-%\end{inlineenum}
-%
-%Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
-%It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{loewenstein1992anomalies}.
-%Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
-%Violations of expectancy have a strong impact on the acceptability of computer systems.
-%Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
-%In immersive systems like \gls{WCA}, which aim to provide seamless interaction, delays are particularly noticeable.
-%
-%It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowski201140}.
-%However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
-%Today's user context, and \gls{WCA} in particular, often demand response times orders of magnitude shorter.
-%
-%For \gls{WCA} the acceptable range for latencies was explored by Chen et al.~\cite{chen2017empirical}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
-%They then proposed a latency tolerance zone according to the task demands.
-%For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}.
-%Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowski201140}.
-%
-%While behavioral changes and negative interaction outcomes have been well documented in prior research on system delay, the specific mechanisms that mediate these outcomes are less well understood.
-%These mechanisms could be cognitive or emotional in origin.
-%
-%A first possible explanation comes from research on cognitive and motor planning.
-%Delay may move users from relatively automatic to more attention-demanding processing.
-%Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
-%As competency in a task increases, execution of the hierarchy becomes increasingly automated.
-%Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
-%Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.
-%
-%Although activities guided by a \gls{WCA} are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
-%Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.
-%
-%A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
-%As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.
-%
-%Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.
-%
-%The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
-%If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
-%Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
-%Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
-%Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.
-%
-%Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality.
-%Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
-%High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
-%Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
-%Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting.
-Extraversion could also  be  a moderating factor among the broader target audience of \gls{WCA}, which are intended for relatively inexperienced users of an application.
-%These and other measures of individual variation were considered here.
-
-\subsection{Characterizing human behavior}
+\subsection{Characterizing human behavior}\label{ssec:characterizing}
 
 \Cref{paper:olguinmunoz2021impact} aims to investigate the effects of system responsiveness on human behavior in step-based \gls{WCA}.
-This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
-The study involved \num{40} undergraduate students who interacted with a \gls{WCA} application while the system responsiveness was systematically altered in real-time, and their behavioral and physiological reactions were recorded.
-The study attempted to answer a number of core research questions relating to human behavior in the context of these applications.
-These included
+This corresponds to the first step in extending the EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.
+The study involves \num{40} undergraduate students who interact with a \gls{WCA} application while system responsiveness is systematically altered in real-time, and their behavioral and physiological reactions are recorded.
+The study attempts to answer a number of core research questions relating to human behavior in the context of these applications.
+These include whether
 \begin{inlineenum}
-    \item whether users change their actions in relation to system latency
-    \item whether they show signals of arousal in physiological responses
-    \item whether their responses to delay effects are mediated by cognition and/or emotion
-    \item whether these effects are mediated by personality indicators
+    \item users change their actions in relation to system latency
+    \item they show signals of arousal in physiological responses
+    \item their responses to delay effects are mediated by cognition and/or emotion
+    \item these effects are mediated by personality indicators
 \end{inlineenum}.
 
-We found that reduced system responsiveness induces additional behavioral slow-down in \gls{WCA} users, and this effect scales with the strength of the reduction in system responsiveness.
-Furthermore, we note that humans tend to speed up as the task progresses;
+I find that reduced system responsiveness induces additional behavioral slow-down in \gls{WCA} users, and that this effect scales with the strength of the reduction in system responsiveness.
+Furthermore, humans tend to speed up as the task progresses;
 however, as system responsiveness decreases, this effect is strongly dampened.
 The effects of reduced system responsiveness on users linger for at least four steps after the system responsiveness improves, leading to longer execution times for steps immediately after a period of reduced responsiveness.
-Finally, the trait of \emph{neuroticism}~\cite{john1999big} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
+Finally, the personality trait of \emph{neuroticism}~\cite{john1999big} seems to play a significant role in modulating users' response to reduced responsiveness in \gls{WCA}.
 
@@
 
-We employed a version of the step-based \emph{Gabriel} cognitive assistant framework first introduced by \citeauthor{chen2018application}~\cite{chen2018application}.
-The assistant was instrumented to capture key application and task performance metrics in real time, in particular step execution time.
-Furthermore, the software was modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
+I employ a version of the step-based \emph{Gabriel} cognitive assistant framework first introduced by \citeauthor{chen2018application}~\cite{chen2018application}.
+The assistant is instrumented to capture key application and task performance metrics in real time, in particular step execution time.
+Furthermore, the software is modified to allow for the real-time alteration of system responsiveness by extending the processing interval of each input to a temporarily fixed value denoted \emph{delay}.
 This is illustrated in \cref{fig:regularwca-vs-delaywca}.
-If, during a series of steps, delay was set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame was provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
+If, during a series of steps, delay is set to a value \ensuremath{\mathbb{D}}, the feedback for each input frame is provided to the user exactly a time interval \ensuremath{\mathbb{D}} after frame capture.
 That is,
@@
 %Seven values were selected for \ensuremath{\mathbb{D}}: no added delay (which was referred to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second}.
-In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note that the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
+In order to fully characterize the resulting \emph{\glspl{TTF}} perceived by participants, we note the instant of final frame capture for a step, \ensuremath{t_c} (and conversely the wait time \( \mathcal{W} \) of the step), can be assumed to be uniformly distributed in \( [t_{n - 1}, t_n] \), without loss of generality.
 Thus the average \gls{TTF} for a step can be expressed as
@@
 \end{align}
-In the following, we will forgo the delay values and present results with respect to the adjusted \glspl{TTF}.
+In the following, I forgo the delay values and present results with respect to the adjusted \glspl{TTF}.
 
@@
 \medskip
-The \gls{WCA} task employed for this study corresponded to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
-Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consisted of a pseudo-random sequence of steps instructing subjects to add or remove pieces.
+The \gls{WCA} task employed for this study corresponds to a modified and extended version of the LEGO assembly task introduced in~\cite{chen2015early}.
+Unlike the original design of this task, in which the user is led through the assembly of a specific model, the modified LEGO task consists of a pseudo-random sequence of steps instructing subjects to add or remove pieces.
 
-We introduced an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
-To study the effects of a delay applied across multiple steps, we manipulated the length of blocks during the task. %, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
-Furthermore, we defined a \emph{slice} as a continuous segment of \num{4} steps within a block;
-\num{4}-step blocks were composed of a single slice, whereas \num{8}- and \num{12}-step blocks were composed respectively of \num{2} and \num{3} consecutive slices.
-Using these experimental design components, we generated a task for each participant by arranging blocks according to a \emph{Latin-square}~\cite{keedwell2015latin} design.
-%We generated a pseudo-random permutation of the combinations of block length and delay values, and assigned a unique sequence of steps to each combination in order to create \num{21} unique blocks of steps.
-%Using a \emph{Latin-square}~\cite{keedwell2015latin} design, we reordered the initial permutation to generate \num{40} tasks.
+I introduce an experimental design component denoted a \emph{block}, illustrated in \cref{fig:stepblock}, corresponding to a segment of consecutive steps subject to the same delay value.
+To study the effects of a delay applied across multiple steps, I manipulate the length of blocks during the task. %, using values of \numlist[list-final-separator={, or }]{4;8;12} steps.
+Furthermore, I define a \emph{slice} as a continuous segment of \num{4} steps within a block;
+\num{4}-step blocks are composed of a single slice, whereas \num{8}- and \num{12}-step blocks are composed respectively of \num{2} and \num{3} consecutive slices.
+Using these experimental design components, I generate a task for each participant by arranging blocks according to a \emph{Latin-square}~\cite{keedwell2015latin} design.
 
-To measure the extent to which individual differences in each subject affected obtained task-performance results, participants were asked to complete two questionnaires previous to beginning the task;
+To measure the extent to which individual differences in each subject affect obtained task-performance results, participants are asked to complete two questionnaires previous to beginning the task;
 the \gls{BFI}~\cite{john1999big} and the \gls{ITQ}~\cite{witmer1998measuring}.
@@
 %The \gls{ITQ}~\cite{witmer1998measuring}, on the other hand, assesses involvement, focus, and the tendency to play games, using \num{29} questions.
-The scores for these questionnaires were normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
+The scores for these questionnaires are normalized in post-processing to fall in the \ensuremath{[0, 1]} range for ease of interpretation.
 
-In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
-This array included devices to capture a number of physiological measures, including \gls{EEG}, \gls{HR}, and \gls{GSR}.
+In order to try to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants are asked to wear an array of biometric sensors.
+This array includes devices to capture a number of physiological measures, such as \gls{EEG}, \gls{HR}, and \gls{GSR}.
 %\begin{inlineenum}
@@
 %\end{inlineenum}.
-Unfortunately, no statistically significant effects were measured in any of these variables, and as such neither will be discussed in this summary.
+Unfortunately, no statistically significant effects are measured in any of these variables, and as such neither is discussed in this summary.
 Please refer to \cref{paper:olguinmunoz2021impact} for details.
 
-Finally, we collected video recordings of participants' faces and as well as of the live feed of frames provided as input to the \gls{WCA}.
-The facial recordings will also not be discussed in this section, as they were not employed in any analyses.
-However, the recording of the live feed was employed later on in the implementation of a procedure for the generation of dynamic synthetic traces for the LEGO task in the context of the EdgeDroid load generator for \gls{WCA}.
-This will be discussed in \cref{model:traces}.
+Finally, I collect video recordings of participants' faces and as well as of the live feed of frames provided as input to the \gls{WCA}.
+The facial recordings are also not discussed in this section, as they are not employed in any analyses.
+However, the recording of the live feed is discussed in \cref{model:traces}, in the context of an implementation of a procedure for the generation of dynamic synthetic traces for the LEGO task in the context of the EdgeDroid load generator for \gls{WCA}.
 
@@
 
-\Cref{fig:ttfvsexectime} shows effects on the measured execution times of participants as perceived \gls{TTF} increased.
+\Cref{fig:ttfvsexectime} shows effects on the measured execution times of participants as perceived \gls{TTF} increases.
 These results have been grouped by block length (\cref{fig:ttfvsexectime:block}), showing the cumulative effects of delay over series of steps, and slice number (\cref{fig:ttfvsexectime:slice}), showing the evolution of these effects within a series of steps at the same level of impairment.
 
-\cref{fig:ttfvsexectime:block} shows that as system responsiveness decreased, the speed at which participants completed instructions tended to slow down.
-Compared to the unimpaired case, participants were on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
+\cref{fig:ttfvsexectime:block} shows that as system responsiveness decreases, the speed at which participants complete instructions tends to slow down.
+Compared to the unimpaired case, participants are on average \SI{12}{\percent} slower when subject to a mean \gls{TTF} of \SI{2.475}{\second}, and \num{26} percentage points slower at a mean \gls{TTF} of \SI{4.5}{\second}.
 \Cref{fig:ttfvsexectime:block} illustrates this effect clearly;
-for all block lengths, as \gls{TTF} grows we see an accompanying increase in execution times.
-Moreover, this effect seemed stronger for longer blocks at higher levels of impairment, indicating a slow-down which compounds with the number of steps.
+for all block lengths, as \gls{TTF} grows an accompanying increase in execution times can be observed.
+Moreover, this effect seems stronger for longer blocks at higher levels of impairment, indicating a slow-down which compounds with the number of steps.
 Note that by measuring \emph{step execution time} we are \emph{per se} compensating for the added delay in the measure, and thus this trend must result from the participants' behavioral slow-down as a response to the perceived increase in \gls{TTF}.
-This was confirmed through an \gls{ANOVA}~\cite{fujikoshi1993two} including factors of block length and \gls{TTF}; significant effects were found for both factors and their interaction (\ensuremath{p < 0.01}).
+This is confirmed through an \gls{ANOVA}~\cite{fujikoshi1993two} including factors of block length and \gls{TTF}; significant effects are found for both factors and their interaction (\ensuremath{p < 0.01}).
 
-On the other hand, in \cref{fig:ttfvsexectime:slice} we observe the evolution of execution times over \emph{slices} of \num{4} steps within blocks.
-This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
-At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
+On the other hand, in \cref{fig:ttfvsexectime:slice} the evolution of execution times over \emph{slices} of \num{4} steps within blocks is illustrated.
+This figure illustrates how, at the lowest levels of impairment, users tend to get faster as the task progresses.
+At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block is roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
 However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
-This was also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.
+This is also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.
 
@@
             Illustration of the effect of neuroticism on the \ensuremath{\mu} parameter of \acs{exGaussian} distributions fitted to the execution times of steps in slice \num{1}.
-            Fits were achieved through \gls{MLE}.
+            Fits are achieved through \gls{MLE}.
             Figure originally published in \cref{paper:olguinmunoz2023realistic}
@@
 
-As mentioned before, we found that the effects of system responsiveness on execution times lingered on after the system moved to a different level of impairment.
+As mentioned before, the effects of system responsiveness on execution times linger on after the system moves to a different level of impairment.
 \Cref{fig:prevblockvsexectime} illustrates this effect.
-On average, the mean execution time of the first four steps of blocks was roughly \SI{7}{\percent} lower when the previous block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the previous block was at a \ensuremath{TTF > \SI{2.475}{\second}} (single-tailed \ensuremath{p < 0.001}).
+On average, the mean execution time of the first four steps of blocks is roughly \SI{7}{\percent} lower when the preceding block was at a \ensuremath{TTF \leq \SI{2.475}{\second}} compared to when the preceding block was at a \ensuremath{TTF > \SI{2.475}{\second}} (single-tailed \ensuremath{p < 0.001}).
 
-Analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
+Analyses of the correlation between individual subject differences and the above effects highlight the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
 This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
-Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism was present in the first two.
-Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
+Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism is present in the first two.
+Individual neuroticism scores are found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.
 
-Finally, we found that execution times were well-fit by \gls{exGaussian} distributions.
+Finally, I find that execution times are well-fit by \gls{exGaussian} distributions.
 Previous research has established strong arguments for the use of this distribution for the modeling of human reaction times and the timings of human actions~\cite{rohrer1994analysis,palmer2011what,marmolejo_ramos2022generalised}.
@@
     \emph{Low} refers to values less than the middle value obtainable for neuroticism in the \gls{BFI}; \emph{high} to values greater than or equal to it.
-}, \gls{TTF}, and slice, the best fit statistic obtained using a Kolmogorov-Smirnov goodness-of-fit test~\cite{massey_jr1951kolmogorov} was \num{0.028}, with a \ensuremath{p}-value of \num{0.999}.
-The effects of neuroticism were most easily observed in the \ensuremath{\mu} parameter and the tails of the fitted distribution;
-an example of this can be observed in \cref{fig:ttfvsexgaussianmu}.
+}, \gls{TTF}, and slice, the best fit statistic obtained using a Kolmogorov-Smirnov goodness-of-fit test~\cite{massey_jr1951kolmogorov} is \num{0.028}, with a \ensuremath{p}-value of \num{0.999}.
+The effects of neuroticism are most easily observed in the \ensuremath{\mu} parameter and the tails of the fitted distribution;
+an example of this is presented in \cref{fig:ttfvsexgaussianmu}.
 There is a clear and direct correlation between higher neuroticism and higher \ensuremath{\mu} values and longer tails.
@@
 \cref{paper:olguinmunoz2023realistic} discusses the elaboration of a model of human behavior for \gls{WCA}.
-Built upon the data and results from \cref{paper:olguinmunoz2021impact}, our model is capable of dynamically generating realistic predictions of execution times for steps in a \gls{WCA} based on historical system state data and a parameterized level of neuroticism.
-Furthermore, we implement a new version of the EdgeDroid tool capable of using this new model, together with a novel procedure for the generation of realistic synthetic traces of frames for \gls{WCA}, to provide a fully end-to-end emulation of a human in the context of these systems.
+Built upon the data and results from \cref{paper:olguinmunoz2021impact}, this model is capable of dynamically generating realistic predictions of execution times for steps in a \gls{WCA} based on historical system state data and a parameterized level of neuroticism.
+Furthermore, I implement a new version of the EdgeDroid tool capable of using this new model, together with a novel procedure for the generation of realistic synthetic traces of frames for \gls{WCA}, to provide a fully end-to-end emulation of a human in the context of these systems.
 
@@
 
-Our current model of human behavior has a limitation that we need to address.
-The model we have discussed so far emulates a rigid and unyielding human who is not affected by low responsiveness in the system.
-This is not an accurate representation of real human behavior, and in the following we present a more realistic human model that takes into consideration the perceived feedback latency and its effects on human reactions.
+The model of human behavior that has a limitation that needs to be addressed.
+It emulates a rigid and unyielding human who is not affected by low responsiveness in the system.
+This is not an accurate representation of real human behavior, and in the following I present a more realistic human model that takes into consideration the perceived feedback latency and its effects on human reactions.
 
-Our new model is based on the comprehensive study of human behavior in the presence of feedback latency outlined in \cref{paper:olguinmunoz2021impact}, and has significant implications for the optimization of edge computing systems.
-By incorporating a more accurate representation of human behavior into our methodology, we can better assess the impact of system design choices on the end-user experience.
+This new model is based on the comprehensive study of human behavior in the presence of feedback latency outlined in \cref{paper:olguinmunoz2021impact}, and has significant implications for the optimization of edge computing systems.
+By incorporating a more accurate representation of human behavior into the methodology, the impact of system design choices on the end-user experience can better be assessed.
 This, in turn, enables future research into optimization strategies that are more effective at improving the overall quality of service for edge computing applications.
@@
 
-Our new timing model generates realistic execution times for each step by taking two inputs: the most recent measured \gls{TTF} and a level of normalized neuroticism.
+The new timing model generates realistic execution times for each step by taking two inputs: the most recent measured \gls{TTF} and a level of normalized neuroticism.
 The model is stateful and calculates a weighted average of the 12 most recent \glspl{TTF}, with the most recent step accounting for approximately \SI{50}{\percent} of the weighted result.
@@
 
-\Cref{fig:model:exectimesoversteps,fig:model:exectimestransitions} illustrate the adherence of the model to the conclusions obtained from our work in \cref{paper:olguinmunoz2021impact}.
+\Cref{fig:model:exectimesoversteps,fig:model:exectimestransitions} illustrate the adherence of the model to the conclusions obtained from the work in \cref{paper:olguinmunoz2021impact}.
 \Cref{fig:model:exectimesoversteps} shows the output of the execution time model described above over a series of steps at three levels of \gls{TTF};
@@
 Conversely, at high levels of impairment, execution times depend on strongly on the parameterized level of neuroticism;
-as concluded from our previous discussion, higher neuroticism leads to a stronger reaction to reduced responsiveness and subsequently to longer execution times.
+as concluded from the previous discussion, higher neuroticism leads to a stronger reaction to reduced responsiveness and subsequently to longer execution times.
 
-In \cref{fig:model:exectimestransitions}, moreover, we present the lingering effects of reduced system responsiveness on \glspl{TTF} generated after the system returns to a more responsive state.
+In \cref{fig:model:exectimestransitions}, moreover, I present the lingering effects of reduced system responsiveness on \glspl{TTF} generated after the system returns to a more responsive state.
 These results were generated by first feeding a sequence of \num{25} identical \glspl{TTF} to the model, with values of either \SI{0}{\second} (i.e.\ instantaneous feedback) or \SI{5}{\second}.
@@
 This procedure was repeated \num{600} times.
-As expected from our discussion in \cref{impact:results}, transitions from a higher (\SI{5}{\second}) \gls{TTF} resulted in noticeably inflated execution times (\ensuremath{+\SI{8.5}{\percent}} at a high level of neuroticism) for the first step at \SI{2.5}{\second} \gls{TTF} compared to when the transition was from a lower \gls{TTF}.
+As expected from the discussion in \cref{impact:results}, transitions from a higher (\SI{5}{\second}) \gls{TTF} resulted in noticeably inflated execution times (\ensuremath{+\SI{8.5}{\percent}} at a high level of neuroticism) for the first step at \SI{2.5}{\second} \gls{TTF} compared to when the transition was from a lower \gls{TTF}.
 
@@
 
-Our design for a procedure for the generation of synthetic yet realistic traces of video inputs for \gls{WCA} follows a stochastic approach.
+The design of procedure for the generation of synthetic yet realistic traces of video inputs for \gls{WCA} follows a stochastic approach.
 Analysis of the evolution of the distribution of processing results of frames over their normalized capture instant with respect to step execution time highlighted the pattern illustrated in \cref{fig:modelframes}.
-We define the normalized capture instant \ensuremath{t_\text{norm}} for each frame as the timestamp at which it was captured with respect to the beginning of the step, normalized by the total execution time of the same step.
+I define the normalized capture instant \ensuremath{t_\text{norm}} for each frame as the timestamp at which it was captured with respect to the beginning of the step, normalized by the total execution time of the same step.
 The distribution of the four possible categories of outputs of the \gls{CV} algorithm in the Gabriel LEGO Task~\cite{chen2018application} for each frame (\emph{blank}, i.e.\ no discernible board state, noise; \emph{low confidence}; \emph{success}, a frame which triggers a step transition;
@@
 
-In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:improved_model}.
+In this section, I discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:improved_model}.
 The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.
@@
 The term \emph{task completion time}, in the context of \gls{WCA}, will refer to the time it takes a user to work their way through the task presented by an assistant application.
-In \cref{paper:olguinmunoz2023realistic} we refer interchangeably to task completion times as application lifetimes.
+In \cref{paper:olguinmunoz2023realistic} I refer interchangeably to task completion times as application lifetimes.
 This metric directly relates to system resource and energy consumption --- longer task completion times mean that the system is occupied for longer and thus uses more resources.
@@
 
-The main question we aim to ask relates to the consequences of using our model of human timing behaviors versus using a less realistic approximation.
-Specifically, we aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
-We compare the total task durations of emulated \num{45}-step tasks using the EdgeDroid benchmarking tool with the improved model of human behavior to a baseline model.
-Our reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping) which is randomly sampled to obtain execution time values.
-We perform this experiment on the edge computing testbed discussed in \cref{sec:testbed};
-for simplicity and to magnify the effects of latency, in this setup we use an \acs{IEEE} \num{802.11}b/g physical layer.
+The main question I aim to ask relates to the consequences of using this model of human timing behaviors versus using a less realistic approximation.
+Specifically, I aim to quantify the difference in task completion times when using the realistic model compared to a baseline which does not take into consideration higher order effects on execution times.
+I compare the total task durations of emulated \num{45}-step tasks using the EdgeDroid benchmarking tool with the improved model of human behavior to a baseline model.
+The reference baseline corresponds here to a first order approximation to empirical execution time modeling, using an \gls{exGaussian} distribution fitted to \emph{all} the data points from \cref{paper:olguinmunoz2021impact} (without any grouping) which is randomly sampled to obtain execution time values.
+I perform this experiment on the Edge Computing testbed discussed in \cref{sec:testbed};
+for simplicity and to magnify the effects of latency, in this setup I use an \acs{IEEE} \num{802.11}b/g physical layer.
 
 The results of the experiment, as presented in \cref{fig:taskcompletiontimesdiff}, show that there are clear and significant differences in task durations between the models.
-At low levels of resource contention, there was a reduction of up to \SI{12}{\percent} in task duration for both levels of neuroticism when using the realistic model.
-However, at higher levels of resource contention, the high neuroticism parameterization resulted in task durations that were more than \SI{8}{\percent} longer than the baseline.
+At low levels of resource contention, there is a reduction of up to \SI{12}{\percent} in task duration for both levels of neuroticism when using the realistic model.
+However, at higher levels of resource contention, the high neuroticism parameterization results in task durations that are more than \SI{8}{\percent} longer than the baseline.
 These findings demonstrate the importance of using a realistic model of human timing behavior to accurately predict and optimize task completion times.
 
-Overall, this case study highlights the potential benefits of using our methodology coupled with a realistic model of human timing behavior in edge computing systems.
-By considering higher order effects on execution times, such as the impact of neuroticism, we can improve the accuracy of our predictions and optimize system performance.
+Overall, this case study highlights the potential benefits of using my methodology coupled with a realistic model of human timing behavior in edge computing systems.
+By considering higher order effects on execution times, such as the impact of neuroticism, the accuracy of predictions can be improved and system performance can be optimized.
 However, it is important to note that the effects of using a realistic model may vary depending on the specific context and task being performed.
-Further research is needed to explore the generalizability of our findings and to investigate other factors that may impact task completion times in edge computing systems.
+Further research is needed to explore the generalizability of these findings and to investigate other factors that may impact task completion times in edge computing systems.
 
@@
 
-In the present section and the following we briefly discuss the implications of our methodology on the optimization potential of number of samples and energy consumption per step, respectively.
-Both these analyses depend on an understanding of the tradeoffs between resource consumption and responsiveness in \gls{WCA};
+In the present section and the following I discuss the implications of my methodology on the optimization potential of number of samples and energy consumption per step, respectively.
+These analyses depend on an understanding of the tradeoffs between resource consumption and responsiveness in \gls{WCA};
 fewer samples per step translates into longer sampling periods, and lower energy consumption tends to be made possible by processing fewer bits of data per step.
 
-We employ in these sections an optimization framework based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
+I employ in these sections an optimization framework based primarily upon the work of Moothedath et al.~\cite{moothedath2021energy,moothedath2022energy1,moothedath2022energy2}.
 The goal of this framework is to balance between increased responsiveness due to higher sampling rates and reduced resource consumption and network congestion caused by excessive sampling.
 In general, any objective metric in these applications which relates to sampling and the responsiveness of the system will present itself
-as a linear combination between \ensuremath{\mathbb{E}[\mathcal{S}]} and \ensuremath{\mathbb{E}[\mathcal{W}]} plus terms independent of the number of samples or wait time:
+as a linear combination between \ensuremath{\mathbb{E}[\mathcal{S}]} and \ensuremath{\mathbb{E}[\mathcal{W}]} plus terms independent of the number of samples or wait~time:
 \begin{alignat}{1}\label{eq:tradeoff}
@@
 
-We compare the effects of the optimization approach with the timing models to the reference scheme.
-We implement a sample-count-optimized aperiodic sampling scheme using \cref{eq:sampling} to determine the optimum sampling instants and the values for \ensuremath{\alpha} and \ensuremath{\beta}.
+I compare the effects of the optimization approach with the timing models to the reference scheme.
+I implement a sample-count-optimized aperiodic sampling scheme using \cref{eq:sampling} to determine the optimum sampling instants and the values for \ensuremath{\alpha} and \ensuremath{\beta}.
 The scheme includes an embedded human timing model that measures the perceived system responsiveness and updates the \ensuremath{\sigma} parameter at every step.
@@
 
-Simulations were performed in Python to test different sampling schemes.
+Simulations are performed in Python to test different sampling schemes.
 The results from these simulations are presented in \Cref{fig:samplingresults}.
@@
 \begin{inlineenum}
-    \item the sample-count-optimized approach described above
+    \item the optimized approach described above
     \item \citeauthor{wang2019towards}s adaptive approach based on a Gaussian distribution
@@
 
-The results indicate that the sample-count-optimized aperiodic scheme results in a lower mean number of samples per step compared to the reference scheme proposed by \citeauthor{wang2019towards}.
-This was achieved while still satisfying the expected wait time bound.
+Results indicate that the optimized aperiodic scheme results in a lower mean number of samples per step compared to the reference scheme proposed by \citeauthor{wang2019towards}.
+This is achieved while still satisfying the expected wait time bound.
 
-The performance of the sample-count-optimized aperiodic scheme was further enhanced by incorporating the timing model that reflects human behavior.
+The performance of the sample-count-optimized aperiodic scheme is further enhanced by incorporating the timing model that reflects human behavior.
 The timing model allows for a more precise optimization of the sampling rate in \gls{WCA} applications, which minimizes the number of samples needed while still meeting the responsiveness requirements.
-These results highlight the effectiveness of the optimization approach and the significance of the timing model in optimizing sampling rates, as well as provide further validation of our methodology.
+These results highlight the effectiveness of the optimization approach and the significance of the timing model in optimizing sampling rates, as well as provide further validation of my methodology.
 
@@
 
-We also optimize for energy consumption in \gls{WCA}.
+I also optimize for energy consumption in \gls{WCA}.
 Energy consumption in these applications depends on various factors such as the number of samples captured, communication power and delays, among others.
 Hence, minimizing the number of samples captured is not a guarantee of reducing energy consumption.
-We therefore take the general solution for \cref{eq:tradeoff} and find the appropriate values of \ensuremath{\alpha} and \ensuremath{\beta} to minimize the energy consumed per step, represented by \ensuremath{\mathcal{E}=E}.
-With feedback given even to the discarded samples in our model and communication delay defined as the total delay in either direction, we have,
+I therefore take the general solution for \cref{eq:tradeoff} and find the appropriate values of \ensuremath{\alpha} and \ensuremath{\beta} to minimize the energy consumed per step, represented by \ensuremath{\mathcal{E}=E}.
+With feedback given even to the discarded samples in the model and communication delay defined as the total delay in either direction, we have,
 \begin{alignat}{2}
@@
 
-We perform an experiment to implement the energy-optimized aperiodic sampling scheme, replacing the sample-count-optimized sampling scheme.
+I repeat the previous experiment with the energy-optimized aperiodic sampling scheme.
 The energy-optimized scheme uses the calculated values of \ensuremath{\alpha} and \ensuremath{\beta} to minimize energy consumption.
@@
 
-The results of our experiments, illustrated partially in \cref{fig:energyresults}, show that this energy-optimized scheme outperforms the reference schemes, with a significant reductions in energy consumption.
-Our approach is consistently consumes \SI{20}{\percent} less energy than \textcite{wang2019towards}'s approach, and is able to achieve up energy savings of up to \SI{40}{\percent} with respect to the greedy scheme.
-Furthermore, our approach is more consistent and reliable, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.
+The results of my experiments, illustrated partially in \cref{fig:energyresults}, show that this energy-optimized scheme outperforms the reference schemes, with significant reductions in energy consumption.
+My approach consistently consumes \SI{20}{\percent} less energy than \textcite{wang2019towards}'s approach, and is able to achieve up energy savings of up to \SI{40}{\percent} with respect to the greedy scheme.
+Furthermore, my approach is more consistent and reliable, exhibiting a flat curve of energy consumption much akin to that of the offline optimum in behavior.
 
-Our findings have important implications for the development and deployment of \gls{WCA} systems.
-We have demonstrated that it is possible to significantly reduce the energy consumption of these systems while still maintaining a reasonable \gls{RTT}.
+These findings have important implications for the development and deployment of \gls{WCA} systems.
+I have demonstrated that it is possible to significantly reduce the energy consumption of these systems while still maintaining a reasonable \gls{RTT}.
 This can help extend the battery life of wearable devices, making  systems more practical and usable for a wider range of users.

### 2023-05-19 body/part1/chapter4-conclusions/conclusions.tex
part 1 tweaks

--- 2023-04-26 body/part1/chapter4-conclusions/conclusions.tex
@@
-In this dissertation, we have presented two key contributions to the literature surrounding performance evaluation of Edge Computing systems.
-Firstly, we have introduced a novel methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing, such as \gls{MAR} and \glspl{CPS}.
-Our methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
-We have validated this methodology through two representative use-cases,\gls{WCA} and \glspl{NCS}.
-Further, we have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
-We have demonstrated the scalability and flexibility of our approach by deploying and studying scenarios with significant number of clients without the need for domain-specific hardware.
+\section{Conclusions}
 
-Our approach presents several advantages over existing methods for the study of trade-offs in latency-sensitive applications in Edge environments.
+In this dissertation, I have presented two key contributions to the literature surrounding performance evaluation of Edge Computing systems.
+Firstly, I have introduced a novel methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing, such as \gls{MAR} and \glspl{CPS}.
+This methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
+I have validated this methodology through two representative use-cases, \gls{WCA} and \glspl{NCS}.
+Further, I have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
+I have demonstrated the scalability and flexibility of our approach by deploying and studying scenarios with significant number of clients without the need for domain-specific hardware.
+
+My approach presents several advantages over existing methods for the performance evaluation of Edge environments.
 It minimizes the complications associated with experimental evaluation, improving key aspects such as efficiency and repeatability.
 The softwarized nature of the client-side emulation allows for straightforward automation and enhances repeatability and replicability of studies.
-Our methodology reduces complexity by moving workload components fully into the software domain, while at the same time capturing effects due to the interaction of network factors, such as congestion control and medium access, and compute factors, such as process scheduling.
+My methodology reduces complexity by moving workload components into the software domain, while at the same time capturing effects due to the interaction of network factors, such as congestion control and medium access, and compute factors, such as process scheduling.
 These effects are are often exceedingly complex to capture in simulations.
-Our approach thus represents a a notable contribution to the literature in its holistic approach to performance evaluation of Edge environments, encompassing both the application and the infrastructure.
+This approach represents thus a notable contribution to literature in its holistic approach to performance evaluation of Edge environments, encompassing both application and infrastructure.
 
-Secondly, we have explored the implications of improved client-side emulation realism on the results produced with our methodology.
-To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
-We have then used this model to study the potential for optimization of these systems.
-Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
-Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
+Secondly, I have explored the implications of improved accuracy on the client-side emulation on the results produced with our methodology.
+To this end, I have provided a novel characterization of human behavior in \gls{WCA}, which I have used to design and implement an unprecedented model of human behavior for these applications.
+I have then used this model to study the potential for optimization of these systems.
+My findings indicate that incorporating improved accuracy in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
+Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- my work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
 These behaviors play a critical role in the overall performance of such applications.
 
-\medskip
+\section{Future work}
 
@@
 In accordance with the broader questions mentioned above, this thesis also raises specific questions related to its individual contributions to the literature.
-Our characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
+My characterization of human behavior in \cref{paper:olguinmunoz2021impact} and both our models of human behavior --- the \gls{FSM}-based model introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} and the data-driven model from \cref{paper:olguinmunoz2023realistic} --- assume linear modes of operation.
 Additionally, the timing characteristics of step-based \gls{WCA} are such that an instruction remains valid for as long as it takes the user to perform it.
@@
 Examples of such applications include, for instance, the \emph{Ping-Pong} assistant introduced in~\cite{chen2015early}.
-We believe there is potential for further development of the EdgeDroid tool to support these other forms of \gls{WCA} applications.
+I believe there is potential for further development of the EdgeDroid tool to support these other forms of \gls{WCA} applications.
 
 The user models are further limited in their realism.
-In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} we assume a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
+In \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid} I have assumed a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
 The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
 These limitations emerge from a lack of data of human behavior in these applications, as at the moment of the writing of these works no comprehensive model of human behavior existed in the relevant literature.
-As a consequence, these works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
-We attempt to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
+These works do not consider the effects of human error on \gls{WCA} application runtimes and footprints, and neither do they consider the feedback effects of system responsiveness on human task performance.
+I have attempted to address some of these limitations in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic};
 the resulting model is able to adapt its timing behavior in response to delay in the system.
-However, we were unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
+However, I was unable to capture any instances of subjects making a mistake during data collection in \cref{paper:olguinmunoz2021impact}, and thus neither the characterization or subsequent model consider human error during task execution.
 
-Furthermore, we identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
+Furthermore, I identify the potential for expanding the dataset presented in \cref{paper:olguinmunoz2021impact} to include a more diverse range of participants.
 The current dataset is limited to young undergraduate students from a highly selective university in the \gls{USA}.
@@
 
-Our research on applying the methodology to \glspl{NCS} also presents promising potential for future work.
+My research on applying the methodology to \glspl{NCS} also presents promising potential for future work.
 While \cref{paper:olguinmunoz2022cleave} provides an initial exploration, further investigation through more realistic and thorough experimentation is necessary to validate the results obtained by the framework.
-To this end, we aim to conduct larger-scale experiments that utilize cutting-edge networking technologies, including 5G and future advancements such as 6G.
-By exploring the full range of networking technologies and scenarios, we can ensure that our methodology is robust and applicable to a wide range of \gls{NCS}.
+To this end, larger-scale experiments that utilize cutting-edge networking technologies, including 5G and future advancements such as 6G, should be conducted.
+By exploring the full range of networking technologies and scenarios, it can be ensured that my methodology is robust and applicable to a wide range of \gls{NCS}.
 
-We also intend to extend the functionality of the \gls{CLEAVE} tool significantly, primarily to target a wider variety of \glspl{NCS} deployments.
-Given the vastness and complexity of the field of \glspl{NCS}, we have limited our exploration in this dissertation to simple, small-scale, single-plant systems.
+The functionality of the \gls{CLEAVE} tool could also be extended significantly, primarily to target a wider variety of \glspl{NCS} deployments.
+Given the vastness and complexity of the field of \glspl{NCS}, I have limited my exploration in this dissertation to simple, small-scale, single-plant systems.
 The system used corresponds to the \gls{2D} inverted pendulum, using simple \gls{PD} and \gls{PID} controllers.
-Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that our conclusions may not directly translate to larger scale or more realistic systems.
-Additionally, our tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
+Although a representative example for the effectiveness of our approach for \glspl{NCS}, this means that my conclusions may not directly translate to larger scale or more realistic systems.
+Additionally, my tool, \gls{CLEAVE}, assumes a simple request-response design in the sensor-controller-actuator loop.
 This does not match well with more complex controllers such as \gls{MPC} controllers, in which each sample produces multiple outputs from the controller.
@@
 
-Finally, we see tremendous opportunities for testbed-driven research in Edge Computing.
-Our current efforts in this context aim towards the extension of our Ainur framework and its integration with \gls{CHI}~\cite{keahey2020lessons}.
-We believe Ainur's unique focus on fully-automated end-to-end wireless edge computing experimentation makes it valuable addition to the \gls{CHI} stack.
+Finally, I see tremendous opportunities for testbed-driven research in Edge Computing.
+My current efforts in this context aim towards the extension of our Ainur framework and its integration with \gls{CHI}~\cite{keahey2020lessons}.
+I believe Ainur's unique focus on fully-automated end-to-end wireless edge computing experimentation makes it valuable addition to the \gls{CHI} stack.
 At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainurs potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.
@@
 
-\medskip
-
-In conclusion, the present dissertation has provided a comprehensive examination of our methodology for \glspl{NCS} and \gls{WCA}.
+In conclusion, the present dissertation has provided a comprehensive examination of my methodology for \glspl{NCS} and \gls{WCA}.
 However, there are still many open questions that require further investigation in future research.

### 2023-05-19 body/part1/chapter2-relatedwork/key_contributions.tex
edge computing

@@
 These applications are similarly difficult to benchmark as \gls{WCA}, particularly at scale on multi-tenant systems, due to their client-side complexity and extreme sensitivity to latency.
-To address these challenges, my co-authors and I introduce yet another tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on edge computing infrastructure.
+To address these challenges, my co-authors and I introduce yet another tool, \gls{CLEAVE}, for the emulation and subsequent deployment of these systems on Edge Computing infrastructure.
 \gls{CLEAVE} allows us both to emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
@@
 Next, in \cref{paper:olguinmunoz2023realistic} my co-authors and I introduce the first ever data-driven model of human timings for \gls{WCA}, built upon the data and insights collected for \cref{paper:olguinmunoz2021impact}.
-The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
+The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real Edge Computing testbed.
 The paper goes on to explore potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.

### 2023-05-19 body/part1/chapter3-contributions/wca-implications.tex
edge computing

@@
 
-This new model is based on the comprehensive study of human behavior in the presence of feedback latency outlined in \cref{paper:olguinmunoz2021impact}, and has significant implications for the optimization of edge computing systems.
+This new model is based on the comprehensive study of human behavior in the presence of feedback latency outlined in \cref{paper:olguinmunoz2021impact}, and has significant implications for the optimization of Edge Computing systems.
 By incorporating a more accurate representation of human behavior into the methodology, the impact of system design choices on the end-user experience can better be assessed.
-This, in turn, enables future research into optimization strategies that are more effective at improving the overall quality of service for edge computing applications.
+This, in turn, enables future research into optimization strategies that are more effective at improving the overall quality of service for Edge Computing applications.
 
@@
 
-Overall, this case study highlights the potential benefits of using my methodology coupled with a realistic model of human timing behavior in edge computing systems.
+Overall, this case study highlights the potential benefits of using my methodology coupled with a realistic model of human timing behavior in Edge Computing systems.
 By considering higher order effects on execution times, such as the impact of neuroticism, the accuracy of predictions can be improved and system performance can be optimized.
 However, it is important to note that the effects of using a realistic model may vary depending on the specific context and task being performed.
-Further research is needed to explore the generalizability of these findings and to investigate other factors that may impact task completion times in edge computing systems.
+Further research is needed to explore the generalizability of these findings and to investigate other factors that may impact task completion times in Edge Computing systems.
 

### 2023-05-19 body/part1/chapter4-conclusions/conclusions.tex
edge computing

@@
 My current efforts in this context aim towards the extension of our Ainur framework and its integration with \gls{CHI}~\cite{keahey2020lessons}.
-I believe Ainur's unique focus on fully-automated end-to-end wireless edge computing experimentation makes it valuable addition to the \gls{CHI} stack.
+I believe Ainur's unique focus on fully-automated end-to-end wireless Edge Computing experimentation makes it valuable addition to the \gls{CHI} stack.
 At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainurs potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.

### 2023-05-19 body/part1/chapter4-conclusions/conclusions.tex
fixes

@@
 I believe Ainur's unique focus on fully-automated end-to-end wireless Edge Computing experimentation makes it valuable addition to the \gls{CHI} stack.
-At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainurs potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.
+At the same time, integrating with \gls{CHI} will provide a number of features and functionalities which will greatly expand Ainur's potential, such as resource reservation, multi-tenant testbed access, and automated bare-metal and \gls{VM}-host instantiation.
 

### 2023-05-19 publications/2023EdgeDroid2/abstract.tex
fixes

--- 2022-12-19 abstract.tex
@@
     We build this model through statistical analysis of data collected from previous work in this field, and demonstrate its utility by studying application task durations.
-    Compared to first-order approximations, our model shows a \textasciitilde\SI{36}{\percent} larger gap between step execution times at high system impairment versus low.
+    Compared to first-order approximations, our model shows a \ensuremath{\sim}\SI{36}{\percent} larger gap between step execution times at high system impairment versus low.
     We further introduce a novel framework for stochastic optimization of resource consumption-responsiveness tradeoffs in \gls{WCA}, and show that by combining this framework with our realistic model of human behavior, significant reductions of up to \SI{50}{\percent} in number processed frame samples and \SI{20}{\percent} in energy consumption can be achieved with respect to the state-of-the-art.
 \end{abstract}
-\acresetall%
+\glsresetall%

### 2023-05-19 publications/2023EdgeDroid2/implications_footprint.tex
fixes

--- 2023-03-02 implications_footprint.tex
@@
 At low neuroticism, the behavior of the realistic model results in total task durations that are basically indistinguishable from the reference model.
-However, at high neuroticism, the model once again results in a considerable difference in total task duration with respect to the reference --- this time extending durations by \textasciitilde\SI{11}{\percent} on average.
+However, at high neuroticism, the model once again results in a considerable difference in total task duration with respect to the reference --- this time extending durations by \ensuremath{\sim}\SI{11}{\percent} on average.
 
@@
 With just a single client and mean \gls{RTT} of around \SI{400}{\milli\second}, all parameterizations of the realistic model achieved an average task duration \SIrange{6}{12}{\percent} shorter than the reference.
-At \num{10} clients, the results mimic those at higher \glspl{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \textasciitilde\SI{8}{\percent} longer application lifetimes.
+At \num{10} clients, the results mimic those at higher \glspl{TTF} in the ideal controlled setup, with high neuroticism parameterizations reaching \ensuremath{\sim}\SI{8}{\percent} longer application lifetimes.
 

### 2023-05-19 publications/2023EdgeDroid2/intro.tex
mark as prelim
fixes

--- 2023-03-03 intro.tex
@@
 \section{Introduction}\label{sec:intro}
-
-\todo[inline]{Needs rewrite}
 
@@
     The economic feasibility of WCA, and hence its likelihood of commercial adoption depends on it not being a resource hog, and this work furthers that effort.
-    Leveraging our more involved user model, combined with the introduced optimization approach, we achieve a \textasciitilde\SI{60}{\percent} reduction in the number of samples processed per step.
+    Leveraging our more involved user model, combined with the introduced optimization approach, we achieve a \ensuremath{\sim}\SI{60}{\percent} reduction in the number of samples processed per step.
     We also achieve an improvement of \SI{20}{\percent} in energy consumption, all while maintaining comparable levels of system responsiveness.

### 2023-05-19 publications/2023EdgeDroid2/model.tex
fixes

--- 2023-03-02 model.tex
@@
 We begin by calculating rolling weighted averages of the \glspl{TTF} of each of the \num{40} individual repetitions of the data.
-We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \gls{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \textasciitilde\SI{25}{\percent}, the third-most \textasciitilde\SI{12}{\percent}, and so on.
+We use exponentially decaying weights for the most recent \num{12} steps, defined in \cref{eq:weights}, such that the most recent \gls{TTF} accounts for roughly \SI{50}{\percent} of the rolling average, the second-most recent for \ensuremath{\sim}\SI{25}{\percent}, the third-most \ensuremath{\sim}\SI{12}{\percent}, and so on.
 We additionally pad the data for each run with \num{12} copies of the first \gls{TTF} in order to ensure sensible values for the first twelve steps; this padding is removed after the weighted values have been calculated.
@@
 This procedure is repeated \num{600} times for each configuration and target \gls{TTF}.
-The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \glspl{TTF} reaching \SI{14}{\percent} (\textasciitilde\SI{5.6}{\second} to \textasciitilde\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
+The resulting mean execution times match precisely the desired behavior mentioned in \cref{it:ttftoexectime}, with the difference in mean execution times at low versus high \glspl{TTF} reaching \SI{14}{\percent} (\ensuremath{\sim}\SI{5.6}{\second} to \ensuremath{\sim}\SI{6.4}{\second}) in the worst case (high neuroticism configuration).
 Additionally, we can observe the effects of neuroticism on generated execution times, as specified in \cref{it:neuro}.
@@
     \centering
-    \includegraphics[width=.9\textwidth]{publications/2023EdgeDroid2/model_data/frame_probabilities}
+    \includegraphics[width=.9\textwidth]{model_data/frame_probabilities}
     \caption{%

### 2023-05-19 publications/2023EdgeDroid2/old_model.tex
fixes

--- 2022-12-19 old_model.tex
@@
 Specifically, we showcase how
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and}}]
+\begin{inlineenum}
     \item higher neuroticism leads to higher mean execution times as impairment goes up (\cref{fig:timing:impneurvsetime})
@@
     \item transitions from higher to lower levels of impairment lead to higher mean execution times when compared to steps far away from a transition, or from a transition from lower to higher levels of impairment (\cref{fig:timing:imptransvsetime})
-\end{enumerate*}.
+\end{inlineenum}.
 
@@
 We also implement two simpler reference timing models:
-\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; and }}]
+\begin{inlineenum}
     % \item a model which returns a constant execution time, corresponding to the mean execution time of all samples associated with the lowest impairment and the shortest duration
@@
     \item a model which fits an \gls{exGaussian} distribution to the raw data, without any grouping, and then randomly samples from it
-\end{enumerate*}.
+\end{inlineenum}.
 We will refer to these models individually as the \emph{naive} and \emph{fitted-naive} model, respectively, and collectively as the \emph{firt-order models}.
@@
 One, all realistic models are faster than the reference models at the lowest \gls{TTF}.
-When compared to the fitted-naive model, the reduction in total duration was on average \textasciitilde\SI{12}{\percent} for realistic models with low neuroticism, and \textasciitilde\SI{6.8}{\percent} for the same models with high neuroticism.
+When compared to the fitted-naive model, the reduction in total duration was on average \ensuremath{\sim}\SI{12}{\percent} for realistic models with low neuroticism, and \ensuremath{\sim}\SI{6.8}{\percent} for the same models with high neuroticism.
 
 Two, at the highest \gls{TTF}, \emph{only} the high neuroticism models show any significant difference in duration with respect to the reference models.
-For \gls{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \textasciitilde\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
+For \gls{TTF} \(=\) \SI{5.0}{s}, the high neuroticism realistic models were on average \ensuremath{\sim}\SI{14.18}{\percent} slower than the fitted-naive model, whereas the low neuroticism models showed no significant relative difference in duration.
 

### 2023-05-19 body/part1/chapter1-introduction/background.tex
further trims

@@
 
-\gls{XR} also presents tremendous opportunities for education and training.
+\Glsfmtlong{XR} also presents tremendous opportunities for education and training.
 Through the use of these technologies, students can get hands-on yet safe, simulated experience in a multitude of scenarios.
@@
 Users of these applications can only notice system responsiveness when they are \emph{expecting} feedback, that is, immediately after the completion of a logical step or instruction.
-Latency bounds can thus be very lax immediately after an instruction is generated (as no feedback is expected yet), and on the other hand much more stringent right after the user has completed an instruction.
+Latency bounds can thus be very lax immediately after an instruction is provided, and on the other hand much more stringent right after the user has completed it.
 In turn, this leads to interesting opportunities for resource consumption optimization in \gls{WCA} applications, such as the adaptive sampling scheme proposed in~\cite{wang2019towards}.
@@
 Sensors sample and encode the state of the plant and transmit it to the controller.
-The controller processes these inputs, and generates control signals to be sent to the actuators, which modify the plant's behavior.
-The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature, tracking a set trajectory, or regulating the speed of a machine.
+The controller processes inputs and generates control signals to be sent to the actuators, which modify the plant's behavior.
+The overall goal of a control system is to regulate a process or a physical system to meet desired specifications or objectives, such as maintaining a constant temperature or tracking a set trajectory.
 

### 2023-05-22 split.sh
split

@@ -1,0 +1,9 @@
+#!/usr/bin/env bash +ex
+
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/kapa.pdf" -dFirstPage=1 -dLastPage=100 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperA.pdf" -dFirstPage=101 -dLastPage=108 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperB.pdf" -dFirstPage=109 -dLastPage=124 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperC.pdf" -dFirstPage=125 -dLastPage=158 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperD.pdf" -dFirstPage=159 -dLastPage=176 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperE.pdf" -dFirstPage=177 -dLastPage=194 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperF.pdf" -dFirstPage=195 -dLastPage=231 -sDEVICE=pdfwrite "MAIN.pdf"

### 2023-05-22 split.sh
split

@@ -1,6 +1,6 @@
 #!/usr/bin/env bash +ex
 
-gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/kapa.pdf" -dFirstPage=1 -dLastPage=100 -sDEVICE=pdfwrite "MAIN.pdf"
+gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/kappa.pdf" -dFirstPage=1 -dLastPage=100 -sDEVICE=pdfwrite "MAIN.pdf"
 gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperA.pdf" -dFirstPage=101 -dLastPage=108 -sDEVICE=pdfwrite "MAIN.pdf"
 gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperB.pdf" -dFirstPage=109 -dLastPage=124 -sDEVICE=pdfwrite "MAIN.pdf"
 gs -dNOPAUSE -dQUIET -dBATCH -sOutputFile="split/paperC.pdf" -dFirstPage=125 -dLastPage=158 -sDEVICE=pdfwrite "MAIN.pdf"
